### 2024-05-16 paper/main.tex
start with paper draft

\documentclass[12pt]{article}

\usepackage{stylesheet}

\title{Randomized trace estimation of parameter-dependent matrices applied to spectral density approximation}

\begin{document}

\maketitle

\section{Introduction}
\label{sec:introduction}

Need for spectral density approximation

Connection to trace estimation

Analysis added to Lin Lin, also works for general matrix functions

\subsection{Notation}

$\mathbb{E}^{k}$

\section{Analysis of parameter-dependent trace estimators}
\label{sec:analysis}

\subsection{Parameter-dependent Girard-Hutchinson estimator}
\label{subsec:hutchinson}

\begin{theorem}{test}{test}
    $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ symmetric, continuous in $t \in [a, b]$.
    \textcolor{red}{$\delta \in (0, e^{-1})$}. Then with probability $\geq 1 - \delta$
    \begin{equation}
        \int_{a}^{b} \left| \Tr(\mtx{A}(t)) - H_{n_{\Psi}}(\mtx{A}(t)) \right| ~\mathrm{d}t \leq \textcolor{red}{32e^{-1}} \frac{\log(\delta^{-1})}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert  ~\mathrm{d}t
    \end{equation}
\end{theorem}

Consider any time-dependent $\mtx{B}(t)$ and random variable $\vct{\psi}$ standard Gaussian
\begin{equation}
    r(t) = \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi} - \Tr(\mtx{B}(t))
\end{equation}

\begin{proof}
By \cite[Proof of Lemma 3]{cortinovis-2022-randomized-trace} we know that $r(t)$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B}(t) \rVert _F^2, 2 \lVert \mtx{B}(t) \rVert _2)$

By \cite[Theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every integer $k \geq 1$
\begin{equation}
    \mathbb{E}\left[ r(t)^{2 k} \right]
    \leq k! \left( 16 \lVert \mtx{B}(t) \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B}(t) \rVert _2 \right)^{2 k}
    = k! 2^{4 k} \lVert \mtx{B}(t) \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B}(t) \rVert _2^{2 k}
\end{equation}

Since $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \geq 1$ we upper bound 
\begin{equation}
    \mathbb{E}\left[ r(t)^{2 k} \right] < \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B}(t) \rVert _F^{2k}
\end{equation}

Stirling approximation \cite{robbins-1955-remark-stirling} 
\begin{equation}
    (2 k)! < 2 \sqrt{\pi k}  e^{\frac{1}{24 k}} \left(\frac{2 k}{e} \right)^{2 k}
\end{equation}

With this, we get
\begin{equation}
    \left( \mathbb{E}\left[ r(t)^{2 k} \right] \right)^{\frac{1}{2k}}
    < \left( \frac{9}{4} \sqrt{\pi k} e^{\frac{1}{24 k}} \right)^{\frac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B}(t) \rVert _F
\end{equation}

It can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\frac{1}{24 k}})^{\frac{1}{2k}} < e$ (since decreasing for $k \geq 1$).
\begin{equation}
    \left( \mathbb{E}\left[ r(t)^{2 k} \right] \right)^{\frac{1}{2k}}
    < 16 k \lVert \mtx{B}(t) \rVert _F
\end{equation}

Minkowski integral inequality \cite[Theorem 2.2]{hardy-1952-inequalities}
\begin{equation}
    \mathbb{E}\left[ \int_{a}^{b} |r(t)|^{2 k}~\mathrm{d}t  \right]^{\frac{1}{2 k}}
    \leq \int_{a}^{b} \mathbb{E}\left[ |r(t)|^{2 k}\right]^{\frac{1}{2 k}}~\mathrm{d}t
    \leq 16 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
\end{equation}

By Markov's inequality (find citation) we have with probability at least $1 - \gamma^{-2 k}$ for some $\gamma \geq 1$
\begin{equation}
    \int_{a}^{b} |r(t)|^{2 k}~\mathrm{d}t > 16 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
\end{equation}

In particular

Use the technique from \cite[Proof of Theorem 1]{cortinovis-2022-randomized-trace}
\begin{equation}
    \mtx{B}(t)
    = \begin{pmatrix}
        \mtx{A}(t) & & \\
        & \ddots & \\
        & & \mtx{A}(t)
    \end{pmatrix}
    \qquad \text{and} \qquad
    \vct{\psi} = \begin{pmatrix}
        \vct{\psi}_1 \\
        \vdots \\
        \vct{\psi}_{n_{\Psi}}
    \end{pmatrix}
\end{equation}
Where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\Psi}}$ independent standard Gaussian. Then $\lVert \mtx{B}(t) \rVert _F = \frac{1}{\sqrt{n_{\Psi}}} \lVert \mtx{A}(t) \rVert _F$ and
\begin{equation}
    r(t) = \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Tr(\mtx{A}(t))
\end{equation}

So we conclude that with probability at least $1 - \gamma^{-2 k}$

\begin{equation}
    \int_{a}^{b} \left| \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Tr(\mtx{A}(t)) \right| ~ \mathrm{d}t \leq 16 k \gamma \frac{1}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t
\end{equation}

Setting $\delta = \gamma^{-2 k}$ (i.e. $\gamma = \delta^{-\frac{1}{2k}}$) and choosing $k = \lceil \log(\delta^{-1}) \rceil$ we get that for all $\delta \in (0, e^{-\frac{1}{2}})$
\begin{equation}
    k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\frac{1}{2\lceil \log(\delta^{-1}) \rceil}}
    = \lceil \log(\delta^{-1}) \rceil e^{\frac{1}{2}\frac{\log(\delta^{-1})}{\lceil \log(\delta^{-1}) \rceil}}
    \leq 2 \log(\delta^{-1}) e^{\frac{1}{2}}
\end{equation}

Consequently, with probability $1 - \delta$ we have
\begin{equation}
    \int_{a}^{b} \left| \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Tr(\mtx{A}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\frac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t
\end{equation}

\end{proof}

\subsection{Parameter-dependent Nyström approximation}
\label{subsec:nystrom}

Pointwise eigenvalue decomposition
\begin{equation}
    \mtx{A}(t) 
    = \mtx{U}(t) \mtx{\Lambda}(t) \mtx{U}(t)^{\top} 
\end{equation}

Partition matrices
\begin{equation}
    \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+2ex+\baselineskip}
    \mtx{U}(t) = \begin{bmatrix}
        \smash{\underbrace{\mtx{U}_1(t)}_{n \times k}} & \smash{\underbrace{\mtx{U}_2(t)}_{n \times (n-k)}}
    \end{bmatrix}
    \qquad 
    \mtx{\Lambda}(t) =
    \begin{bmatrix}
        \smash{\overbrace{\mtx{\Lambda}_1(t)}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2(t)}_{(n-k) \times (n-k)}}
    \end{bmatrix}
\end{equation}
and define $\mtx{\Omega}_1(t) = \mtx{U}_1(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times 2k}$ and $\mtx{\Omega}_2(t) = \mtx{U}_2(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times 2k}$ with $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$

\begin{lemma}{OSE lemma}{ose}
    $\mtx{\Omega}_1 \in \mathbb{R}^{k \times 2k}, \mtx{\Omega}_2 \in \mathbb{R}^{(n - k) \times 2k}$ standard normal i.i.d, $\mtx{A}$, $k$ \emph{even}\textcolor{red}{(really necessary?)}

    \begin{equation}
        \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] \leq C \left( \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F \right)
    \end{equation}
\end{lemma}

\begin{proof}
Using submultiplicativity and independence
\begin{equation}
    \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
    \leq \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \rVert _2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
    = \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \rVert _2 \right] \mathbb{E}^{k}\left[  \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
\end{equation}

First term $\mtx{\Omega}_2' = \mtx{\Omega}_2 / \sqrt{2k}$. This matrix has the \textcolor{red}{$(c_1\sqrt{q/m}, 9^d c_2e^{-q}, d, q)$} OSE moment property
\begin{align}
    \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2' \rVert _2 \right]
    &= \left(\mathbb{E}^{\frac{k}{2}}\left[ \lVert (\mtx{A} \mtx{\Omega}_2')^{\top} (\mtx{A} \mtx{\Omega}_2') \rVert _2 \right]\right)^{\frac{1}{2}} \notag \\
    &\leq \left( \mathbb{E}^{\frac{k}{2}}\left[ \lVert (\mtx{A} \mtx{\Omega}_2')^{\top} (\mtx{A} \mtx{\Omega}_2') - \mtx{A}^{\top} \mtx{A} \rVert _2 + \lVert \mtx{A}^{\top} \mtx{A} \rVert _2 \right]\right)^{\frac{1}{2}} \notag \\
    &\leq \left( \mathbb{E}^{\frac{k}{2}}\left[ \lVert (\mtx{A} \mtx{\Omega}_2')^{\top} (\mtx{A} \mtx{\Omega}_2') - \mtx{A}^{\top} \mtx{A} \rVert _2\right]\right)^{\frac{1}{2}} + \lVert \mtx{A} \rVert _2
\end{align}

By \cite{} $\mtx{\Omega}_2'$ satisfies the \textcolor{red}{$(c_1 \sqrt{\frac{q}{m}}, c_2 e^{-q}, q)$}-JL moment property for \textcolor{red}{any $q$ (in what?)} \textcolor{red}{$\geq \dots$}, hence, by \cite[Lemma 4]{cohen-2016-optimal-approximate} it satisfies the \textcolor{red}{$(2 c_1 \sqrt{\frac{q}{m}}, 9^d c_2 e^{-q}, d, q)$}-OSE moment property for any $d \in \mathbb{N}$. Therefore, we can apply an intermediary result from \cite[Proof of Theorem 1]{cohen-2016-optimal-approximate} to bound \textcolor{red}{(order of $A\Omega$ vs $\Omega A$ ?!)}
\begin{equation}
    \mathbb{E}^{\frac{k}{2}}\left[ \lVert (\mtx{A} \mtx{\Omega}_2')^{\top} (\mtx{A} \mtx{\Omega}_2') - \mtx{A}^{\top} \mtx{A} \rVert _2\right] \leq \varepsilon \delta^{\frac{2}{k}}\left( \lVert \mtx{A} \rVert _2^2 + \frac{\lVert \mtx{A} \rVert _F^2}{k} \right)
\end{equation}

Inserting this bound into the above inequality, we get
\begin{align}
    \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \rVert _2 \right]
    &= \sqrt{2k} \cdot \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2' \rVert _2 \right] \notag \\
    &\leq \sqrt{2k} \cdot \todo{c} \left( \lVert \mtx{A} \rVert _2^2  + \frac{\lVert \mtx{A} \rVert _F^2}{k} \right)^{\frac{1}{2}} + \sqrt{2k} \cdot \lVert \mtx{A} \rVert _2 \notag \\
    &\leq \sqrt{2k} \cdot 2 (1 + \todo{c})\left( \lVert \mtx{A} \rVert _2  + \frac{\lVert \mtx{A} \rVert _F}{\sqrt{k}} \right)
\end{align}

The second term can be bounded with help of the proof of \cite[Lemma B.3]{tropp-2023-randomized-algorithms}. We bound \todo{(confusion with reformulating pseudo inverse as inverse)}
\begin{equation}
    \mathbb{E}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^{k} \right]
    = \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\frac{k}{2}} \right]
    \leq \left( k + 1 \right) \left( \frac{1}{(k + 1)!}\right)^{\frac{k}{k + 1}} \left( \frac{3 k}{2}\right)^{\frac{k}{2}}
\end{equation}
The Taylor series expansion of the exponential function shows that $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, and consequently $\left( \frac{1}{n!} \right)^{\frac{1}{n}} \leq \frac{e}{n}$. Hence
\begin{equation}
    \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
    \leq (k+1)^{\frac{1}{k}} \frac{e}{k + 1}\sqrt{ \frac{3k}{2} }
    \leq e^2 \sqrt{\frac{3}{2k}}
\end{equation}

Putting everything together, we get
\begin{equation}
    \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
    \leq e^2\sqrt{\frac{3}{2k}}
\end{equation}

\end{proof}

\begin{theorem}{nystrom}{nystrom}
    $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ symmetric positive semi-definite and continuous in $t \in [a, b]$. $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\Omega}}$ standard Gaussian with \textcolor{red}{$n_{\Omega} \geq 8 \log(\delta^{-1})$}. Then with probability $\geq 1 - \delta$
    \begin{equation}
        \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F ~\mathrm{d}t
        \leq C_{\Omega}\frac{1}{\sqrt{n_{\Omega}}} \int_{a}^{b} \Tr(\mtx{A}(t)) ~\mathrm{d}t 
    \end{equation}
\end{theorem}

\begin{proof}
    \cite[Theorem B.1]{persson-2023-randomized-lowrank} and standard matrix norm manipulations \textcolor{red}{(cite?)}
    \begin{align}
        \lVert \mtx{A} - \widehat{\mtx{A}} \rVert _F 
        &\leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert (\mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F \notag \\
        &\leq \frac{1}{\sqrt{k}} \Tr(\mtx{A}) + \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F \notag \\
        &\leq \frac{1}{\sqrt{k}} \Tr(\mtx{A}) + \sqrt{k} \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 
    \end{align}

    By Cauchy-Schwarz \cite{},
    OSE Lemma, triangle inequality
    \begin{align}
        \mathbb{E}^{\frac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
        &\leq \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]^2 \notag \\
        &\leq C \left( \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _F \right)^2 \notag \\
        &\leq C \left( \lVert \mtx{\Lambda}_2 \rVert _2  + \frac{1}{k}\lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _F^2 \right) \notag \\
        &= C \left( \lambda_{k+1} + \frac{1}{k}\Tr(\mtx{\Lambda}_2) \right) \notag \\
        &\leq \frac{2 C}{k} \Tr(\mtx{A})
    \end{align}

    Using triangle inequality and 
    Minkowski's integral inequality \cite[Theorem 2.2]{hardy-1952-inequalities}    
    \begin{align}
        \mathbb{E}^{\frac{k}{2}}\left[ \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F~\mathrm{d}t \right]
        &\leq \mathbb{E}^{\frac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Tr(\mtx{A}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\frac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
        &\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Tr(\mtx{A}(t))~\mathrm{d}t + \mathbb{E}^{\frac{k}{2}} \left[ \int_{a}^{b}  \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\frac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
        &\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Tr(\mtx{A}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\frac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\frac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
        &\leq \frac{1 + 2C}{\sqrt{k}} \int_{a}^{b} \Tr(\mtx{A}(t))~\mathrm{d}t
    \end{align}
    With Markov inequality \cite{} with probability at least $1 - \gamma^{-\frac{k}{2}}$
    \begin{equation}
        \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{1 + 2C}{\sqrt{k}} \int_{a}^{b} \Tr(\mtx{A}(t))~\mathrm{d}t
    \end{equation}
    Setting $\delta = \gamma^{-\frac{k}{2}}$
\end{proof}

\begin{theorem}{test}{ŧest}
    $f(\mtx{A}, t)$ continuous in $t \in [a, b]$ and $\mtx{\Omega} \in \mathbb{R}^{n \times (r + p)}$ Gaussian random matrix with $r \geq 2, p \geq 4$. Then for all $\gamma \geq 1$ with probability $\geq 1 - \gamma^{-p}$
    \begin{equation}
        \int_{a}^{b} \left| \Tr(f(\mtx{A}, t)) - \Tr(\widehat{f}(\mtx{A}, t)) \right| ~\mathrm{d}t
        \leq \gamma^2 (1 + r) \int_{a}^{b} \sum_{j = r+1}^{n} \sigma_j(f(\mtx{A}, t)) ~\mathrm{d}t
    \end{equation}
\end{theorem}

\begin{proof}
    Notice that \cite[Lemma 2.1, item 3]{frangella-2023-randomized-nystrom} implies that $f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t)$ is positive semi-definite \textcolor{red}{also for all $t$?}
    \begin{equation}
        \left| \Tr(f(\mtx{A}, t)) - \Tr(\widehat{f}(\mtx{A}, t)) \right|
        = \lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _1
    \end{equation}

    From \cite[Proof of Corollary 8.2]{tropp-2023-randomized-algorithms} it follows that
    \begin{equation}
        \lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _1 = \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{1/2} \mtx{\Omega}}) f(\mtx{A}, t)^{1/2} \rVert _2^2
    \end{equation}
    Hence, by \cite[Theorem 9]{kressner-2023-randomized-lowrank} the result follows
    with $\lVert \cdot \rVert _2 \leq \lVert \cdot \rVert _F$ and $\sigma_j(f^{1/2}(\mtx{A}, t))^{2} = \sigma_j(f(\mtx{A}, t))$.
\end{proof}

Discuss rank of kernel matrix function and application to theorem.

\subsection{Parameter-dependent Nyström++ estimator}
\label{subsec:nystrom-pp}

\section{Application to spectral density approximation}
\label{sec:application}

Spectral density definition
\begin{equation}
    \phi(t) = \frac{1}{n} \sum_{i=1}^{n} \delta(t - \lambda_i)
    \label{equ:spectral-density}
\end{equation}

\subsection{Introduction to spectral density estimation}
\label{subsec:spectral-density}

Smooth spectral density definition
\begin{equation}
    \phi_{\sigma}(t) = \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i) = \Tr(g_{\sigma}(t\mtx{I}_n - \mtx{A}))
    \label{equ:smooth-spectral-density}
\end{equation}
with Gaussian smoothing kernel
\begin{equation}
    g_{\sigma}(s) = \frac{1}{n \sigma \sqrt{2\pi}} e^{-\frac{s^2}{2\sigma^2}}
    \label{smoothing-kernel}
\end{equation}

Can apply trace estimators from \refsec{sec:analysis} to $g_{\sigma}(t\mtx{I}_n - \mtx{A})$

Discuss error measures and baseline smoothing error

\subsection{Chebyshev expansion of smoothing kernel}
\label{subsec:chebyshev-expansion}

Affine approximation for evaluating matrix function is favorable
\begin{equation}
    g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) = \sum_{l=0}^{m} \mu_l(t) T_l(\mtx{A})
    \label{equ:matrix-expansion}
\end{equation}

\subsubsection{Chebyshev expansion error}
\label{subsubsec:expansion-error}

\begin{lemma}{Chebyshev expansion error}{chebyshev-error}
    Let $\mtx{A} \in \mathbb{R}^{n \times n}$ be a symmetric matrix whose spectrum is contained in $[-1, 1]$. Then the expansion $g_{\sigma}^{(m)}$ of the Gaussian smoothing kernel satisfies
    \begin{equation}
        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \right| \leq \sqrt{\frac{2e}{\pi}} \frac{1}{n \sigma^2} (1 + \sigma)^{-m} \equiv E_{\sigma, m}
        \label{equ:2-chebyshev-interpolation-sup-error-kernel}
        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
        %\label{equ:2-chebyshev-interpolation-error}
    \end{equation}
    for all degrees $m > 0$ and smoothing parameters $\sigma>0$.
\end{lemma}

Closely follows \cite[Theorem 2]{lin-2017-randomized-estimation}

\begin{proof}
From Bernstein's theorem \cite[Theorem 4.3]{trefethen-2008-gauss-quadrature} if $f$ analytic within Bernstein ellipse $\mathcal{E}_{\chi}$, then for any $m > 0$
\begin{equation}
    \sup_{t \in [-1, 1]} \left| f - f^{(m)} \right| \leq \frac{2}{\chi^m (\chi - 1)} \max_{z \in \mathcal{E}_{\chi}} |f(z)|
    \label{equ:bernstein-bound}
\end{equation}

In particular, for $g_{\sigma}(t - \cdot)$ we observe for $z = x + iy \in \mathcal{E}_{\chi}$
\begin{equation}
| g_{\sigma}(t - z) | 
= \frac{1}{n \sigma \sqrt{2 \pi}} \left| e^{-\frac{(t - z)^2}{2\sigma^2}} \right|
= \frac{1}{n \sigma \sqrt{2 \pi}} e^{-\frac{(t - x)^2}{2\sigma^2}}e^{\frac{y^2}{2\sigma^2}}
%\leq \frac{1}{n \sigma \sqrt{2 \pi}} \max_{x + iy \in \mathcal{E}_{\chi}} e^{\frac{y^2}{2\sigma^2}} 
\end{equation}

Since $e^{-\frac{(t - x)^2}{2\sigma^2}} \leq 1$ and the maximum absolute value of $y$ is limited by the imaginary half-axis of the Bernstein ellipse, which has length $(\chi - \chi^{-1}) / 2$, we upper bound

\begin{equation}
    \max_{z \in \mathcal{E}_{\chi}} | g_{\sigma}(t - z) | 
    \leq \frac{1}{n \sigma \sqrt{2 \pi}} e^{\frac{(\chi - \chi^{-1})^2}{8 \sigma^2}} 
\end{equation}

What turns out to be most useful in practice is the choice $\chi = 1 + \sigma$. Since in this case $\chi - \chi^{-1} \leq 2\sigma$, this implies
\begin{equation}
    \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \right| \leq \sqrt{\frac{2e}{\pi}} \frac{1}{n \sigma^2} (1 + \sigma)^{-m}
\end{equation}
\end{proof}

\subsubsection{Discrete cosine transform}
\label{subsubsec:dct}

Fast computation of expansion with DCT (duality with evaluations)

Squaring of Chebyshev expansion for consistent and affine Nyström approximation
\begin{equation}
    \left( \sum_{l=0}^{m} \mu_l T_l(t) \right)^2 = \sum_{l=0}^{2m} \nu_l T_l(t)
    \label{equ:squared-chebyshev-expansion}
\end{equation}
where $\nu = \DCT^{-1} \left\{ \DCT\{\mu\}^2 \right\}$

Non-negative Chebyshev expansion for results in Nyström approximation of positive semi-definite matrices (discuss alternatives: Jackson damping, Putinar representation)

\begin{lemma}{Non-negative Chebyshev expansion error}{non-negative-chebyshev-error}
    Let $\mtx{A} \in \mathbb{R}^{n \times n}$ be a symmetric matrix whose spectrum
    is contained in $[-1, 1]$. Then the non-negative expansion $g_{\sigma}^{(\underline{m})}$ of the Gaussian smoothing kernel satisfies
    \begin{equation}
        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - g_{\sigma}^{(\underline{m})}(t - s) \right| \leq 2\sqrt{2} \left(1 + n \sigma \sqrt{\pi} E_{\sqrt{2}\sigma, \frac{m}{2}}\right) E_{\sqrt{2}\sigma, \frac{m}{2}} \equiv \underline{E}_{\sigma, m}
        \label{equ:2-chebyshev-interpolation-sup-error-kernel}
        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
        %\label{equ:2-chebyshev-interpolation-error}
    \end{equation}
    for all \emph{even} degrees $m > 0$ and smoothing parameters $\sigma>0$.
\end{lemma}

\begin{proof}

For any numbers $a, b \in \mathbb{R}$ it holds
\begin{equation}
| a^2 - b^2 | = | (a + b)(a - b) |\leq | a + b | | a - b | \leq (2 | a | + | a - b |)  | a - b |
\end{equation}

Therefore, with $g_{\sigma} = \sqrt{g_{\sigma}}^2$ and $g_{\sigma}^{(\underline{m})} = (\sqrt{g_{\sigma}}^{(m/2)})^2$
we have
\begin{equation}
    \left| g_{\sigma} - g_{\sigma}^{(\underline{m})} \right| \leq \left( 2 \left| \sqrt{g_{\sigma}} \right| + \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(m/2)} \right| \right) \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(m/2)} \right|
\end{equation}

Since 
\begin{equation}
\sqrt{g_{\sigma}} = \sqrt{2 n \sigma \sqrt{2 \pi}} g_{\sqrt{2}\sigma}
\end{equation}
we can apply the results from \textcolor{red}{above} to get
\begin{equation}
    \left| g_{\sigma}(t - s) - g_{\sigma}^{(\underline{m})}(t - s) \right| \leq \left( 2 \frac{1}{\sqrt{n \sigma \sqrt{2\pi}}} + \sqrt{2 n \sigma \sqrt{2 \pi}} E_{\sqrt{2}\sigma, \frac{m}{2}}\right)\sqrt{2 n \sigma \sqrt{2 \pi}} E_{\sqrt{2}\sigma, \frac{m}{2}}
\end{equation}
from which follows the result with some simplification.
\end{proof}

\subsection{The Chebyshev-Nyström++ method for approximating spectral densities}
\label{subsec:chebyshev-nystrom}

\subsubsection{Implementation}
\label{subsubsec:chebyshev-nystrom-implementation}

\begin{algo}{Chebyshev-Nystrom++}{nystrom-chebyshev-pp}
\begin{algorithmic}[1]
    \Statex \textbf{Input:} Symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$, evaluation points $\{t_i\}_{i=1}^{n_t}$
    \Statex \textbf{Parameters:} Degree $m$, allocation $(n_{\Omega}, n_{\Psi})$,  smoothing parameter $\sigma$
    \Statex \textbf{Output:} Approximate evaluations of the spectral density $\{\breve{\phi}_{\sigma}^{(m)}(t_i)\}_{i=1}^{n_t}$
    \State Compute $\{\mu_l(t_i)\}_{l=0}^{m}$ and $\{\nu_l(t_i)\}_{l=0}^{2m}$ for all $t_i$ %using %\refalg{alg:2-chebyshev-chebyshev-expansion}
    %\State Compute  for all $t_i$ using %\refalg{alg:3-nystrom-chebyshev-exponentiation}
    \State Generate standard Gaussian matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\Omega}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\Psi}}$%\glsfirst{sketching-matrix} $\in \mathbb{R}^{n \times n_{\Omega}}$
    %\State Generate standard Gaussian %\glsfirst{random-matrix} $\in \mathbb{R}^{n \times n_{\Psi}}$
    \State Initialize $[\mtx{V}_1, \mtx{V}_2, \mtx{V}_3] \gets [\mtx{0}_{n \times n_{\Omega}}, \mtx{\mtx{\Omega}}, \mtx{0}_{n \times n_{\Omega}}]$
    \State Initialize $[\mtx{W}_1, \mtx{W}_2, \mtx{W}_3] \gets [\mtx{0}_{n \times n_{\Psi}}, \mtx{\Psi}, \mtx{0}_{n \times n_{\Psi}}]$
    \State Initialize $[\mtx{K}_1(t_i), \mtx{K}_2(t_i)] \gets [\mtx{0}_{n_{\Omega} \times n_{\Omega}}, \mtx{0}_{n_{\Omega} \times n_{\Omega}}]$ for all $t_i$
    \State Initialize $[\mtx{L}_1(t_i), \ell(t_i)] \gets [\mtx{0}_{n_{\Omega} \times n_{\Psi}}, 0]$ for all $t_i$
    %\State Set $\breve{\phi}_{\sigma}^{(m)}(t_i) \gets 0$ for all $t_i$
    \For {$l = 0, \dots, 2m$}
    \State $[\mtx{X}, \mtx{Y}] \gets \mtx{\mtx{\Omega}}^{\top} [\mtx{V}_2, \mtx{W}_2]$  
    %\State $\mtx{X} \gets \mtx{\mtx{\Omega}}^{\top} \mtx{V}_2$
      %\State $\mtx{Y} \gets \mtx{\mtx{\Omega}}^{\top} \mtx{W}_2$
      \State $z \gets \Tr(\mtx{\Psi}^{\top} \mtx{W}_2)$
      \For {$i = 1, \dots, n_t$}
        \If {$l \leq m$}
            \State $\mtx{K}_1(t_i) \gets \mtx{K}_1(t_i) + \mu_l(t_i) \mtx{X}$ \Comment{assemble $\mtx{\Omega}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$}
            \State $\mtx{L}_1(t_i) \gets \mtx{L}_1(t_i) + \mu_l(t_i) \mtx{Y}$ \Comment{assemble $\mtx{\Psi}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$}
            \State $\ell(t_i) \gets \ell(t_i) + \mu_l(t_i) z$ \Comment{assemble $\Tr(\mtx{\Psi}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi})$}
        \EndIf
        \State $\mtx{K}_2(t_i) \gets \mtx{K}_2(t_i) + \nu_l(t_i) \mtx{X}$ \Comment{assemble $\mtx{\Omega}^{\top} (g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}))^2 \mtx{\Omega}$}
      \EndFor
      \State $[\mtx{V}_3, \mtx{W}_3] \gets (2 - \delta_{l0}) \mtx{A} [\mtx{V}_2, \mtx{W}_2] - [\mtx{V}_1, \mtx{W}_1]$ \Comment{Chebyshev recurrence}
      \State $[\mtx{V}_1, \mtx{W}_1] \gets [\mtx{V}_2, \mtx{W}_2]$
      \State $[\mtx{V}_2, \mtx{W}_2] \gets [\mtx{V}_3, \mtx{W}_3]$
      %\State $\mtx{V}_3 \gets (2 - \delta_{l0}) \mtx{A} \mtx{V}_2 - \mtx{V}_1$ %\Comment{Chebyshev recurrence \refequ{equ:2-chebyshev-chebyshev-recursion}}
      %\State $\mtx{V}_1 \gets \mtx{V}_2, \mtx{V}_2 \gets \mtx{V}_3$
      %\State $\mtx{W}_3 \gets (2 - \delta_{l0}) \mtx{A} \mtx{W}_2 - \mtx{W}_1$ %\Comment{Chebyshev recurrence \refequ{equ:2-chebyshev-chebyshev-recursion}}
      %\State $\mtx{W}_1 \gets \mtx{W}_2, \mtx{W}_2 \gets \mtx{W}_3$
    \EndFor
    \For {$i = 1, \dots, n_t$}
      \State $\breve{\phi}_{\sigma}^{(m)}(t_i) \gets \Tr\left( \mtx{K}_1(t_i)^{\dagger}\mtx{K}_2(t_i) \right) + \frac{1}{n_{\Psi}} \left( \ell(t_i) + \Tr\left( \mtx{L}_1(t_i)^{\top} \mtx{K}_1(t_i)^{\dagger} \mtx{L}_1(t_i) \right)  \right) $ \label{lin:4-nystromchebyshev-nystrom-pp}
    \EndFor
\end{algorithmic}
\end{algo}

\subsubsection{Analysis}
\label{subsubsec:chebyshev-nystrom-analysis}

Corollary for expanded spectral density

\begin{lemma}{$L^1$-error of Chebyshev expansion of spectral density}{chebyshev-error}
    Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$. Then the spectal density constructed with a Chebyshev expansion $\phi_{\sigma}^{m}$ satisfies
    \begin{equation}
        \int_{a}^{b} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq 2 n E_{\sigma, m}
        \label{equ:chebyshev-interpolation-sup-error-kernel}
        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
        %\label{equ:2-chebyshev-interpolation-error}
    \end{equation}
    for all degrees $m > 0$ and smoothing parameters $\sigma>0$. The same result holds for the expansion of a non-negative Chebyshev expansion $\phi_{\sigma}^{(\underline{m})}$ with $\underline{E}_{\sigma, m}$.
\end{lemma}

\begin{proof}
    \begin{align}
        &\left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right| \notag \\
        &= \left| \Tr(g_{\sigma}(t\mtx{I}_n - \mtx{A})) - \Tr(g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})) \right|
        && \text{(definitions \refequ{equ:1-introduction-spectral-density-as-trace} and \refequ{equ:2-chebyshev-chebyshev-expansion})} \notag \\
        &= \left| \sum_{i=1}^n \left(g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i)\right) \right|
        && \text{($\lambda_1, \dots, \lambda_n$ eigenvalues of $\mtx{A}$)} \notag \\
        &\leq n \max_{i = 1, \dots, n} \left| g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i) \right|
        && \text{(conservative upper bound)} \notag \\
        &\leq n E_{\sigma, m}
        && \text{(using \refequ{equ:2-chebyshev-uniform-bound})} \notag \\
    \end{align}\\

    Finally, H\"older's inequality %\cite{klenke2013probability} 
    allows us to also show the last assertion with what we have found above:
    \begin{equation}
        \int_{a}^{b} | \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) | ~\mathrm{d}t
            \leq 2 \sup_{t \in [-1, 1]} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|
            \leq 2 n E_{\sigma, m}.
    \end{equation}
\end{proof}

\section{Numerical results}
\label{sec:results}

For our first example, we consider the matrix which arises from the second order
finite difference discretization of the Laplace operator $\Delta$ in a potential
field $V$,
\begin{equation}
    \mathcal{A} u(\vct{x}) = - \Delta u(\vct{x}) + V(\vct{x}) u(\vct{x}),
    \label{equ:5-experiments-electronic-hamiltonian}
\end{equation}
for a uniform mesh of size $h=0.6$. The potential $V$ results from a
lattice whose primitive cell is of side-length $L=6$ and in whose center a
potential
\begin{equation}
    \alpha \exp(-\frac{\lVert \vct{x} \rVert _2^2}{ 2 \beta^2 })
    \label{equ:5-experiments-gaussian-cell}
\end{equation}
with $\alpha = -4$, $\beta = 2$ is located. The computational domain is chosen
to span $n_c \in \mathbb{N}$ primitive cells in every spatial dimension, hence, yielding
discretization matrices which are growing in size with $n_c$. In our experiments
we consider the three-dimensional case, but for visualization purposes, we
illustrate the potential in FIGURE
in two dimensions.

\todo{[Plot with $1/\epsilon$ and $1/\epsilon^2$ result]}

\todo{[Plot with dependence on $\sigma$]}

\todo{[Timing comparison]}

\section{Discussion}
\label{sec:discussion}

\end{document}

### 2024-05-19 paper/main.tex
rewrite most proofs

@@
 
+\todo{
+\begin{itemize}
+    \item Unify index notation
+    \item Double-check proofs
+    \item Add all references
+    \item Run numerical experiments 
+    \item Write introduction / abstract
+\end{itemize}
+}
+
+\begin{abstract}
+    Abstract
+\end{abstract}
+
 \section{Introduction}
@@
 
-Need for spectral density approximation
+Need for spectral density approximation (cite some literature)
 
-Connection to trace estimation
+Connection to trace estimation (maybe only in words, not in formulas)
 
-Analysis added to Lin Lin, also works for general matrix functions
+For \emph{many} evaluations of parameter, i.e. try to decrease complexity with each evaluation
+
+Our work: Analysis added to Lin Lin, also works for general matrix functions
 
 \subsection{Notation}
-
-$\mathbb{E}^{k}$
+\begin{itemize}
+    \item $e$ and $\log$
+    \item the $k$-th moment of a random variable is denoted with $\mathbb{E}^{k}[X] = \left(\mathbb{E}\left[ | X |^{k} \right] \right)^{\frac{1}{k}}$;
+    \item in a Gaussian random vector and a Gaussian random matrix, all entries follow independently the standard normal distribution;
+    \item Schatten norms
+    \item $\mathbb{N}$ all positive whole numbers
+\end{itemize}
 
@@
 
-\subsection{Parameter-dependent Girard-Hutchinson estimator}
+We consider parameter-dependent matrices of the form
+\begin{equation}
+    \mtx{A}(t) = \begin{bmatrix}
+        a_{11}(t) & a_{12}(t) & \dots & a_{1n}(t) \\
+        a_{21}(t) & a_{22}(t) & \dots & a_{2n}(t) \\
+        \vdots & \vdots & \ddots & \vdots \\
+        a_{n1}(t) & a_{n2}(t) & \dots & a_{nn}(t) \\
+    \end{bmatrix}.
+\end{equation}
+where $t$ is a parameter taking values in the interval $[a,b]$.
+Mostly analogous to constant matrix.
+
+Trace is defined as
+\begin{equation}
+    \Trace(\mtx{A}(t)) = \sum_{i=1}^{n} a_{ii}(t).
+\end{equation}
+If we can only access the parameter-dependent matrix through matrix-vector products, we can estimate the trace with the \emph{quadratic trace estimator} \cite{}: We take $n_{\Psi} \in \mathbb{N}$ Gaussian random vectors $\vct{\psi}_1,\vct{\psi}_2 \dots, \vct{\psi}_{n_{\psi}} \in \mathbb{R}^{n}$ to form
+\begin{equation}
+    \Hutch{n_{\Psi}}(\mtx{A}(t)) = \frac{1}{n_{\Psi}} \sum_{j=1}^{n_{\Psi}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j.
+\end{equation}
+If we gather the random vectors in a matrix $\mtx{\Psi} = [\vct{\psi}_1 ~ \vct{\psi}_2 ~ \cdots ~ \vct{\psi}_{n_{\psi}}] \in \mathbb{R}^{n \times n_{\Psi}}$ we may rewrite $\Hutch{n_{\Psi}}(\mtx{A}(t)) = n_{\Psi}^{-1} \Trace( \mtx{\Psi}^{\top} \mtx{A}(t) \mtx{\Psi})$.
+Other choices for the distribution of the entries of the random vectors are possible \cite{}.
+However, our theoretical developments only hold in the Gaussian case.
+
+Alternatively, the trace of a symmetric matrix whose rank is significantly lower than its size can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\Omega}}$ to form the \emph{Nystr\"om approximation}
+\begin{equation}
+    \widehat{\mtx{A}}(t) = (\mtx{A}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} (\mtx{A}(t) \mtx{\Omega})^{\top}.
+\end{equation}
+Then we define the trace estimator $\Nystr{n_{\Omega}}(\mtx{A}(t)) = \Trace( \widehat{\mtx{A}}(t) )$.
+Thanks to the invariance of the trace under cyclic permutation of its arguments and the symmetry of the matrix, we may also rewrite
+\begin{equation}
+    \Nystr{n_{\Omega}}(\mtx{A}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{A}(t)^2 \mtx{\Omega})).
+\end{equation}
+
+Finally, a combination of the two estimators yields an analogue for the \emph{Nyström++ estimator} \cite{persson-2022-improved-variants} for parameter-dependent settings \todo{this is super ugly.
+find better notation}
+\begin{equation}
+    \Nystrpp{n_{\Omega}}{n_{\Psi}}(\mtx{A}(t)) = \Nystr{n_{\Omega}}(\mtx{A}(t)) + \Hutch{n_{\Psi}}(\mtx{A}(t) - \widehat{\mtx{A}}(t)).
+\end{equation}
+It is understood that $\widehat{\mtx{A}}$ is computed with the same sketching matrix as the one used for the trace estimator $\Nystr{n_{\Omega}}$.
+We can interpret this estimator as an interpolation between the trace of the Nystr\"om approximation and the quadratic trace estimator.
+
+\subsection{Parameter-dependent quadratic trace estimator}
 \label{subsec:hutchinson}
 
-\begin{theorem}{test}{test}
-    $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ symmetric, continuous in $t \in [a, b]$.
-    \textcolor{red}{$\delta \in (0, e^{-1})$}. Then with probability $\geq 1 - \delta$
+\begin{theorem}{Parameter-dependent quadratic trace estimator}{quadratic}
+    Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ continuously depend on $t \in [a, b]$. Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ it holds with probability at least $1 - \gamma^{-2k}$
     \begin{equation}
-        \int_{a}^{b} \left| \Tr(\mtx{A}(t)) - H_{n_{\Psi}}(\mtx{A}(t)) \right| ~\mathrm{d}t \leq \textcolor{red}{32e^{-1}} \frac{\log(\delta^{-1})}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert  ~\mathrm{d}t
+        \int_{a}^{b} \left| \Trace(\mtx{A}(t)) - \Hutch{n_{\Psi}}(\mtx{A}(t)) \right| ~\mathrm{d}t < 16 \gamma k \frac{1}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F  ~\mathrm{d}t.
     \end{equation}
@@
 
-Consider any time-dependent $\mtx{B}(t)$ and random variable $\vct{\psi}$ standard Gaussian
+The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Markov's inequality to transition to the continuous parameter case.
+We first only consider the $1$-query estimator and subsequently use a trick to carry the result over to the general case.
+\begin{proof} 
+    We define the 1-query quadratic trace estimate 
     \begin{equation}
-    r(t) = \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi} - \Tr(\mtx{B}(t))
+        r(t) = \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi} - \Trace(\mtx{B}(t))
+    \end{equation}
+    for a parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$ in a fixed $t \in [a,b]$.
+    From the proof of \cite[Lemma 3]{cortinovis-2022-randomized-trace} we know that $r(t)$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B}(t) \rVert _F^2, 2 \lVert \mtx{B}(t) \rVert _2)$.
+Thus, by \cite[Theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
+    \begin{align}
+        \mathbb{E}\left[ r(t)^{2 k} \right]
+        &\stackrel{\text{\cite{boucheron-2013-basic-inequalities}}}{\leq} k! \left( 16 \lVert \mtx{B}(t) \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B}(t) \rVert _2 \right)^{2 k} \notag \\
+        &= k! 2^{4 k} \lVert \mtx{B}(t) \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B}(t) \rVert _2^{2 k}.
+    \end{align}
+    Since $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$ we can upper bound 
+    \begin{equation}
+        \mathbb{E}\left[ r(t)^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B}(t) \rVert _F^{2k}.
+    \end{equation}
+    Stirling's approximation \cite{robbins-1955-remark-stirling} bonuds $(2 k)!
+< 2 \sqrt{\pi k}  e^{\frac{1}{24 k}} ( 2 k / e )^{2 k}$.
+Consequently, the moments of $r(t)$ for a fixed $t$ are limited by
+    \begin{equation}
+        \mathbb{E}^{2k}\left[ r(t) \right]
+        < \left( \frac{9}{4} \sqrt{\pi k} e^{\frac{1}{24 k}} \right)^{\frac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B}(t) \rVert _F < 16 k \lVert \mtx{B}(t) \rVert _F,
+        \label{equ:quadratic-trace-onequery-fixed}
+    \end{equation}
+    since it can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\frac{1}{24 k}})^{\frac{1}{2k}} < e$, due to the monotonous decrease of this expression in $k \in \mathbb{N}$.
+
+    Now we make the transition to the continuous.
+Minkowski's integral inequality \cite[Theorem 2.2]{hardy-1952-inequalities} allows us to apply \refequ{equ:quadratic-trace-onequery-fixed} in the continuous setting:
+    \begin{equation}
+        \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(t)|~\mathrm{d}t  \right]
+        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}\left[ |r(t)|^{2 k}\right]^{\frac{1}{2 k}}~\mathrm{d}t
+        < 16 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
+    \end{equation}
+    By Markov's inequality \cite{}, for any $k \in \mathbb{N}$ and $\gamma \geq 1$, it holds with probability at least $1 - \gamma^{-2 k}$
+    \begin{equation}
+        \int_{a}^{b} |r(t)|~\mathrm{d}t < 16 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
+        \label{equ:quadratic-trace-onequery-uniform}
     \end{equation}
 
-\begin{proof}
-By \cite[Proof of Lemma 3]{cortinovis-2022-randomized-trace} we know that $r(t)$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B}(t) \rVert _F^2, 2 \lVert \mtx{B}(t) \rVert _2)$
-
-By \cite[Theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every integer $k \geq 1$
-\begin{equation}
-    \mathbb{E}\left[ r(t)^{2 k} \right]
-    \leq k! \left( 16 \lVert \mtx{B}(t) \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B}(t) \rVert _2 \right)^{2 k}
-    = k! 2^{4 k} \lVert \mtx{B}(t) \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B}(t) \rVert _2^{2 k}
-\end{equation}
-
-Since $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \geq 1$ we upper bound 
-\begin{equation}
-    \mathbb{E}\left[ r(t)^{2 k} \right] < \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B}(t) \rVert _F^{2k}
-\end{equation}
-
-Stirling approximation \cite{robbins-1955-remark-stirling} 
-\begin{equation}
-    (2 k)! < 2 \sqrt{\pi k}  e^{\frac{1}{24 k}} \left(\frac{2 k}{e} \right)^{2 k}
-\end{equation}
-
-With this, we get
-\begin{equation}
-    \left( \mathbb{E}\left[ r(t)^{2 k} \right] \right)^{\frac{1}{2k}}
-    < \left( \frac{9}{4} \sqrt{\pi k} e^{\frac{1}{24 k}} \right)^{\frac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B}(t) \rVert _F
-\end{equation}
-
-It can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\frac{1}{24 k}})^{\frac{1}{2k}} < e$ (since decreasing for $k \geq 1$).
-\begin{equation}
-    \left( \mathbb{E}\left[ r(t)^{2 k} \right] \right)^{\frac{1}{2k}}
-    < 16 k \lVert \mtx{B}(t) \rVert _F
-\end{equation}
-
-Minkowski integral inequality \cite[Theorem 2.2]{hardy-1952-inequalities}
-\begin{equation}
-    \mathbb{E}\left[ \int_{a}^{b} |r(t)|^{2 k}~\mathrm{d}t  \right]^{\frac{1}{2 k}}
-    \leq \int_{a}^{b} \mathbb{E}\left[ |r(t)|^{2 k}\right]^{\frac{1}{2 k}}~\mathrm{d}t
-    \leq 16 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
-\end{equation}
-
-By Markov's inequality (find citation) we have with probability at least $1 - \gamma^{-2 k}$ for some $\gamma \geq 1$
-\begin{equation}
-    \int_{a}^{b} |r(t)|^{2 k}~\mathrm{d}t > 16 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
-\end{equation}
-
-In particular
-
-Use the technique from \cite[Proof of Theorem 1]{cortinovis-2022-randomized-trace}
+    To extend the bound to the $n_{\Psi}$-query quadratic trace estimator, we use a technique from the proof of \cite[Proof of Theorem 1]{cortinovis-2022-randomized-trace}. Let
     \begin{equation}
         \mtx{B}(t)
-    = \begin{pmatrix}
+        = \frac{1}{n_{\Psi}} \begin{pmatrix}
             \mtx{A}(t) & & \\
@@
     \end{equation}
-Where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\Psi}}$ independent standard Gaussian. Then $\lVert \mtx{B}(t) \rVert _F = \frac{1}{\sqrt{n_{\Psi}}} \lVert \mtx{A}(t) \rVert _F$ and
+    where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\Psi}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors.
+Then $\lVert \mtx{B}(t) \rVert _F = \frac{1}{\sqrt{n_{\Psi}}} \lVert \mtx{A}(t) \rVert _F$ and
     \begin{equation}
-    r(t) = \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Tr(\mtx{A}(t))
+        r(t) = \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Trace(\mtx{A}(t)).
+    \end{equation}
+    Hence, we conclude from \refequ{equ:quadratic-trace-onequery-uniform} that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
+    \begin{equation}
+        \int_{a}^{b} \left| \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Trace(\mtx{A}(t)) \right| ~ \mathrm{d}t
+        < 16 k \gamma \frac{1}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t.
     \end{equation}
 
-So we conclude that with probability at least $1 - \gamma^{-2 k}$
-
-\begin{equation}
-    \int_{a}^{b} \left| \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Tr(\mtx{A}(t)) \right| ~ \mathrm{d}t \leq 16 k \gamma \frac{1}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t
-\end{equation}
-
-Setting $\delta = \gamma^{-2 k}$ (i.e. $\gamma = \delta^{-\frac{1}{2k}}$) and choosing $k = \lceil \log(\delta^{-1}) \rceil$ we get that for all $\delta \in (0, e^{-\frac{1}{2}})$
-\begin{equation}
-    k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\frac{1}{2\lceil \log(\delta^{-1}) \rceil}}
-    = \lceil \log(\delta^{-1}) \rceil e^{\frac{1}{2}\frac{\log(\delta^{-1})}{\lceil \log(\delta^{-1}) \rceil}}
-    \leq 2 \log(\delta^{-1}) e^{\frac{1}{2}}
-\end{equation}
-
-Consequently, with probability $1 - \delta$ we have
-\begin{equation}
-    \int_{a}^{b} \left| \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Tr(\mtx{A}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\frac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t
-\end{equation}
+    %Setting $\delta = \gamma^{-2 k}$ (i.e. $\gamma = \delta^{-\frac{1}{2k}}$) and choosing $k = \lceil \log(\delta^{-1}) \rceil$ we get that for all $\delta \in (0, e^{-\frac{1}{2}})$
+    %\begin{equation}
+    %    k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\frac{1}{2\lceil \log(\delta^{-1}) \rceil}}
+    %    = \lceil \log(\delta^{-1}) \rceil e^{\frac{1}{2}\frac{\log(\delta^{-1})}{\lceil \log(\delta^{-1}) \rceil}}
+    %    \leq 2 \log(\delta^{-1}) e^{\frac{1}{2}}
+    %\end{equation}
+    %
+    %Consequently, with probability $1 - \delta$ we have
+    %\begin{equation}
+    %    \int_{a}^{b} \left| \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Trace(\mtx{A}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\frac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t
+    %\end{equation}
 
@@
 
-\subsection{Parameter-dependent Nyström approximation}
+\subsection{Parameter-dependent Nystr\"om approximation}
 \label{subsec:nystrom}
@@
 \end{equation}
+where $\mtx{\Lambda}(t) = \operatorname{diag}(\lambda_1(t), \dots, \lambda_n(t))$.
 
-Partition matrices
+We define the matrix partitions
 \begin{equation}
@@
 \end{equation}
-and define $\mtx{\Omega}_1(t) = \mtx{U}_1(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times 2k}$ and $\mtx{\Omega}_2(t) = \mtx{U}_2(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times 2k}$ with $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$
+and let $\mtx{\Omega}_1(t) = \mtx{U}_1(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times 2k}$ and $\mtx{\Omega}_2(t) = \mtx{U}_2(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times 2k}$ with $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$.
 
-\begin{lemma}{OSE lemma}{ose}
-    $\mtx{\Omega}_1 \in \mathbb{R}^{k \times 2k}, \mtx{\Omega}_2 \in \mathbb{R}^{(n - k) \times 2k}$ standard normal i.i.d, $\mtx{A}$, $k$ \emph{even}\textcolor{red}{(really necessary?)}
-
+\begin{lemma}{Moment tail bound}{ose}
+    A symmetric matrix $\mtx{A} \in \mathbb{R}^{l \times l}$ and stochastically independent Gaussian random matrices $\mtx{\Omega}_1 \in \mathbb{R}^{k \times 2k}, \mtx{\Omega}_2 \in \mathbb{R}^{l \times 2k}$ satisfy for all $k \in \mathbb{N}$ and a constant $c > 0$
     \begin{equation}
-        \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] \leq C \left( \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F \right)
+        \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] \leq c \left( \lVert \mtx{A} \rVert _2 + \frac{\lVert \mtx{A} \rVert _F}{\sqrt{k}} \right).
     \end{equation}
@@
 \begin{proof}
-Using submultiplicativity and independence
+    We first use the submultiplicativity of the spectral norm and the stochastic independendence of the random matrices
     \begin{equation}
@@
         \leq \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \rVert _2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    = \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \rVert _2 \right] \mathbb{E}^{k}\left[  \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
+        = \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \rVert _2 \right] \mathbb{E}^{k}\left[  \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right].
+        \label{equ:OSE-moment-bound-initial}
     \end{equation}
 
-First term $\mtx{\Omega}_2' = \mtx{\Omega}_2 / \sqrt{2k}$. This matrix has the \textcolor{red}{$(c_1\sqrt{q/m}, 9^d c_2e^{-q}, d, q)$} OSE moment property
+    To bound the first term, we define $\widetilde{\mtx{\Omega}}_2 = \mtx{\Omega}_2^{\top} / \sqrt{2k}$. We may use the triangle inequality to rewrite
     \begin{align}
-    \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2' \rVert _2 \right]
-    &= \left(\mathbb{E}^{\frac{k}{2}}\left[ \lVert (\mtx{A} \mtx{\Omega}_2')^{\top} (\mtx{A} \mtx{\Omega}_2') \rVert _2 \right]\right)^{\frac{1}{2}} \notag \\
-    &\leq \left( \mathbb{E}^{\frac{k}{2}}\left[ \lVert (\mtx{A} \mtx{\Omega}_2')^{\top} (\mtx{A} \mtx{\Omega}_2') - \mtx{A}^{\top} \mtx{A} \rVert _2 + \lVert \mtx{A}^{\top} \mtx{A} \rVert _2 \right]\right)^{\frac{1}{2}} \notag \\
-    &\leq \left( \mathbb{E}^{\frac{k}{2}}\left[ \lVert (\mtx{A} \mtx{\Omega}_2')^{\top} (\mtx{A} \mtx{\Omega}_2') - \mtx{A}^{\top} \mtx{A} \rVert _2\right]\right)^{\frac{1}{2}} + \lVert \mtx{A} \rVert _2
+        &\mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \rVert _2 \right] \notag \\
+        &= \sqrt{2k} \cdot \mathbb{E}^{k}\left[ \lVert \mtx{A} \widetilde{\mtx{\Omega}}_2^{\top} \rVert _2 \right] && \text{(definition of $\widetilde{\mtx{\Omega}}_2$)} \notag \\
+        &= \sqrt{2k} \cdot \left(\mathbb{E}^{\frac{k}{2}}\left[ \lVert (\mtx{A} \widetilde{\mtx{\Omega}}_2^{\top}) (\mtx{A} \widetilde{\mtx{\Omega}}_2^{\top})^{\top} \rVert _2 \right]\right)^{\frac{1}{2}} && \text{($\lVert \mtx{B} \rVert _2 = \lVert \mtx{B} \mtx{B}^{\top} \rVert _2^{\frac{1}{2}} $ )} \notag \\
+        &\leq \sqrt{2k} \cdot \left( \mathbb{E}^{\frac{k}{2}}\left[ \lVert (\mtx{A} \widetilde{\mtx{\Omega}}_2^{\top}) (\mtx{A} \widetilde{\mtx{\Omega}}_2^{\top})^{\top} - \mtx{A}^{\top} \mtx{A} \rVert _2 + \lVert \mtx{A}^{\top} \mtx{A} \rVert _2 \right]\right)^{\frac{1}{2}} && \text{(triangle inequality)} \notag \\
+        &\leq \sqrt{2k} \cdot \left( \mathbb{E}^{\frac{k}{2}}\left[ \lVert (\widetilde{\mtx{\Omega}}_2 \mtx{A} )^{\top} (\widetilde{\mtx{\Omega}}_2 \mtx{A}) - \mtx{A}^{\top} \mtx{A} \rVert _2\right]\right)^{\frac{1}{2}} + \sqrt{2k} \cdot \lVert \mtx{A} \rVert _2.
+&& \text{($(a+b)^{\frac{1}{2}} \leq a^{\frac{1}{2}} + b^{\frac{1}{2}}$)}
+    \end{align}
+    By \cite{} $\widetilde{\mtx{\Omega}}_2 \in \mathbb{R}^{2k \times (n - k)}$ satisfies the \textcolor{red}{$(c_1 \sqrt{\frac{q}{2k}}, c_2 e^{-q}, q)$}-JL moment property for \todo{any $q$ (in what?) $\geq \dots$} and some constants $c_1, c_2 > 0$, hence, by \cite[Lemma 4]{cohen-2016-optimal-approximate} it satisfies the \textcolor{red}{$(2 c_1 \sqrt{\frac{q}{2k}}, 9^d c_2 e^{-q}, d, q)$}-OSE moment property for any $d \in \mathbb{N}$.
+For convenience, we choose $d=2k$ and $q=\frac{k}{2}$, such that $\widetilde{\mtx{\Omega}}_2$ has $(c_1, c_2 (3^4e^{-\frac{1}{2}})^k, 2k, \frac{k}{2})$-OSE moment property.
+We may then apply an intermediary result from the proof of \cite[Theorem 1]{cohen-2016-optimal-approximate} to bound
+    \begin{equation}
+        \mathbb{E}^{\frac{k}{2}}\left[ \lVert (\widetilde{\mtx{\Omega}}_2 \mtx{A} )^{\top} (\widetilde{\mtx{\Omega}}_2 \mtx{A}) - \mtx{A}^{\top} \mtx{A} \rVert _2\right] \leq c_1 c_2^2 3^8e^{-1} \left( \lVert \mtx{A} \rVert _2^2 + \frac{\lVert \mtx{A} \rVert _F^2}{k} \right).
+    \end{equation}
+    Inserting this bound into the above inequality and generously bounding all terms with a single constant we get
+    \begin{align}
+        \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \rVert _2 \right]
+        &\leq \sqrt{2k} \cdot \sqrt{c_1 e^{-1}} c_2 3^4 \left( \lVert \mtx{A} \rVert _2^2  + \frac{\lVert \mtx{A} \rVert _F^2}{k} \right)^{\frac{1}{2}} + \sqrt{2k} \cdot \lVert \mtx{A} \rVert _2 \notag \\
+        &\leq 2 \sqrt{2} (1 + \sqrt{c_1 e^{-1}} c_2 3^4)\left( \sqrt{k} \cdot \lVert \mtx{A} \rVert _2  + \lVert \mtx{A} \rVert _F \right).
+        \label{equ:OSE-moment-bound-first}
     \end{align}
 
-By \cite{} $\mtx{\Omega}_2'$ satisfies the \textcolor{red}{$(c_1 \sqrt{\frac{q}{m}}, c_2 e^{-q}, q)$}-JL moment property for \textcolor{red}{any $q$ (in what?)} \textcolor{red}{$\geq \dots$}, hence, by \cite[Lemma 4]{cohen-2016-optimal-approximate} it satisfies the \textcolor{red}{$(2 c_1 \sqrt{\frac{q}{m}}, 9^d c_2 e^{-q}, d, q)$}-OSE moment property for any $d \in \mathbb{N}$.
-Therefore, we can apply an intermediary result from \cite[Proof of Theorem 1]{cohen-2016-optimal-approximate} to bound \textcolor{red}{(order of $A\Omega$ vs $\Omega A$ ?!)}
-\begin{equation}
-    \mathbb{E}^{\frac{k}{2}}\left[ \lVert (\mtx{A} \mtx{\Omega}_2')^{\top} (\mtx{A} \mtx{\Omega}_2') - \mtx{A}^{\top} \mtx{A} \rVert _2\right] \leq \varepsilon \delta^{\frac{2}{k}}\left( \lVert \mtx{A} \rVert _2^2 + \frac{\lVert \mtx{A} \rVert _F^2}{k} \right)
-\end{equation}
-
-Inserting this bound into the above inequality, we get
-\begin{align}
-    \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \rVert _2 \right]
-    &= \sqrt{2k} \cdot \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2' \rVert _2 \right] \notag \\
-    &\leq \sqrt{2k} \cdot \todo{c} \left( \lVert \mtx{A} \rVert _2^2  + \frac{\lVert \mtx{A} \rVert _F^2}{k} \right)^{\frac{1}{2}} + \sqrt{2k} \cdot \lVert \mtx{A} \rVert _2 \notag \\
-    &\leq \sqrt{2k} \cdot 2 (1 + \todo{c})\left( \lVert \mtx{A} \rVert _2  + \frac{\lVert \mtx{A} \rVert _F}{\sqrt{k}} \right)
-\end{align}
-
-The second term can be bounded with help of the proof of \cite[Lemma B.3]{tropp-2023-randomized-algorithms}. We bound \todo{(confusion with reformulating pseudo inverse as inverse)}
+    The second term in \refequ{equ:OSE-moment-bound-initial} can be bounded with help of the proof of \cite[Lemma B.3]{tropp-2023-randomized-algorithms}
     \begin{equation}
@@
         = \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\frac{k}{2}} \right]
-    \leq \left( k + 1 \right) \left( \frac{1}{(k + 1)!}\right)^{\frac{k}{k + 1}} \left( \frac{3 k}{2}\right)^{\frac{k}{2}}
+        \leq \left( k + 1 \right) \left( \frac{1}{(k + 1)!}\right)^{\frac{k}{k + 1}} \left( \frac{3 k}{2}\right)^{\frac{k}{2}}.
     \end{equation}
-The Taylor series expansion of the exponential function shows that $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, and consequently $\left( \frac{1}{n!} \right)^{\frac{1}{n}} \leq \frac{e}{n}$.
-Hence
+    With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\frac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\frac{1}{n}} \leq \frac{e}{n}$ follow respectively.
+Hence,
     \begin{equation}
         \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    \leq (k+1)^{\frac{1}{k}} \frac{e}{k + 1}\sqrt{ \frac{3k}{2} }
-    \leq e^2 \sqrt{\frac{3}{2k}}
+        \leq \frac{e^2}{k + 1}\sqrt{ \frac{3k}{2} }
+        \leq e^2 \sqrt{\frac{3}{2k}}.
+        \label{equ:OSE-moment-bound-second}
     \end{equation}
 
-Putting everything together, we get
+    Putting \refequ{equ:OSE-moment-bound-first} and  \refequ{equ:OSE-moment-bound-second} together, we get
     \begin{equation}
         \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    \leq e^2\sqrt{\frac{3}{2k}}
+        \leq \underbrace{2 e^2\sqrt{3} (1 + \sqrt{c_1 e^{-1}} c_2 3^4)}_{\equiv c}\left( \lVert \mtx{A} \rVert _2  + \frac{\lVert \mtx{A} \rVert _F}{\sqrt{k}} \right).
     \end{equation}
@@
 
-\begin{theorem}{nystrom}{nystrom}
+\begin{lemma}{Parameter-dependent Nystrom approximation}{nystrom}
     $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ symmetric positive semi-definite and continuous in $t \in [a, b]$.
 $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\Omega}}$ standard Gaussian with \textcolor{red}{$n_{\Omega} \geq 8 \log(\delta^{-1})$}.
-Then with probability $\geq 1 - \delta$
+Then with probability at least $1 - \delta$
     \begin{equation}
         \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F ~\mathrm{d}t
-        \leq C_{\Omega}\frac{1}{\sqrt{n_{\Omega}}} \int_{a}^{b} \Tr(\mtx{A}(t)) ~\mathrm{d}t 
+        \leq C_{\Omega}\frac{1}{\sqrt{n_{\Omega}}} \int_{a}^{b} \Trace(\mtx{A}(t)) ~\mathrm{d}t 
     \end{equation}
-\end{theorem}
+\end{lemma}
+
+\todo{proof idea}
 
 \begin{proof}
-    \cite[Theorem B.1]{persson-2023-randomized-lowrank} and standard matrix norm manipulations \textcolor{red}{(cite?)}
-    \begin{align}
+    Omitting the parameter dependence, we can apply \cite[Theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound the integrand with
+    \begin{equation}
         \lVert \mtx{A} - \widehat{\mtx{A}} \rVert _F 
-        &\leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert (\mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F \notag \\
-        &\leq \frac{1}{\sqrt{k}} \Tr(\mtx{A}) + \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F \notag \\
-        &\leq \frac{1}{\sqrt{k}} \Tr(\mtx{A}) + \sqrt{k} \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 
-    \end{align}
+        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert (\mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F.
+    \end{equation}
 
-    By Cauchy-Schwarz \cite{},
-    OSE Lemma, triangle inequality
+    As in the proof of \cite[Lemma 3]{meyer-2021-hutch-optimal}, the first term can be bonud with
+    \begin{equation}
+        \lVert \mtx{\Lambda}_2 \rVert _F
+        = \sqrt{\sum_{i=k+1}^{n} \lambda_i^2}
+        \leq \sqrt{ \lambda_{k+1} \sum_{i=k+1}^{n} \lambda_i}
+        \leq \sqrt{ \frac{\Trace(\mtx{A})}{k} \Trace(\mtx{A})}
+        \leq \frac{1}{\sqrt{k}} \Trace(\mtx{A}).
+    \end{equation}
+
+    The second term is first processed with standard matrix-norm inequalities:
+    \begin{equation}
+        \lVert (\mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F 
+        \leq \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F
+        \leq \sqrt{k} \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2.
+    \end{equation}
+    Then we consider its moments
     \begin{align}
         \mathbb{E}^{\frac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
-        &\leq \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]^2 \notag \\
-        &\leq C \left( \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _F \right)^2 \notag \\
-        &\leq C \left( \lVert \mtx{\Lambda}_2 \rVert _2  + \frac{1}{k}\lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _F^2 \right) \notag \\
-        &= C \left( \lambda_{k+1} + \frac{1}{k}\Tr(\mtx{\Lambda}_2) \right) \notag \\
-        &\leq \frac{2 C}{k} \Tr(\mtx{A})
+        &= \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]^2 && \text{(by definition of $\mathbb{E}^{k}$)} \notag \\
+        &\leq C \left( \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _F \right)^2 && \text{(\reflem{lem:ose} with $\mtx{A}=\mtx{\Lambda}_2^{\frac{1}{2}}$)} \notag \\
+        &\leq C \left( \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _2^2  + \frac{1}{k}\lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _F^2 \right) && \text{($(a + b)^2 \leq a^2 + b^2$)} \notag \\
+        &= C \left( \lambda_{k+1} + \frac{1}{k}\Trace(\mtx{\Lambda}_2) \right) && \text{(definition of $\lVert \cdot \rVert _2$ and $\lVert \cdot \rVert _F$)} \notag \\
+        &\leq \frac{2 C}{k} \Trace(\mtx{A}) && \text{($k \lambda_{k+1} \leq \Trace(\mtx{A})$)}
     \end{align}
+    Thus, we get
+    \begin{equation}
+        \mathbb{E}^{\frac{k}{2}} \left[\lVert \mtx{A} - \widehat{\mtx{A}} \rVert _F \right]
+        \leq 
+    \end{equation}
 
@@
     \begin{align}
-        \mathbb{E}^{\frac{k}{2}}\left[ \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F~\mathrm{d}t \right]
-        &\leq \mathbb{E}^{\frac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Tr(\mtx{A}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\frac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
-        &\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Tr(\mtx{A}(t))~\mathrm{d}t + \mathbb{E}^{\frac{k}{2}} \left[ \int_{a}^{b}  \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\frac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
-        &\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Tr(\mtx{A}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\frac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\frac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
-        &\leq \frac{1 + 2C}{\sqrt{k}} \int_{a}^{b} \Tr(\mtx{A}(t))~\mathrm{d}t
+        &\mathbb{E}^{\frac{k}{2}}\left[ \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F~\mathrm{d}t \right] \notag \\
+        &\leq \mathbb{E}^{\frac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\frac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
+        &\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t + \mathbb{E}^{\frac{k}{2}} \left[ \int_{a}^{b}  \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\frac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
+        &\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\frac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\frac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
+        &\leq \frac{1 + 2C}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t
     \end{align}
@@
     \begin{equation}
-        \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{1 + 2C}{\sqrt{k}} \int_{a}^{b} \Tr(\mtx{A}(t))~\mathrm{d}t
+        \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{1 + 2C}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t
     \end{equation}
@@
 
-\begin{theorem}{test}{ŧest}
-    $f(\mtx{A}, t)$ continuous in $t \in [a, b]$ and $\mtx{\Omega} \in \mathbb{R}^{n \times (r + p)}$ Gaussian random matrix with $r \geq 2, p \geq 4$.
-Then for all $\gamma \geq 1$ with probability $\geq 1 - \gamma^{-p}$
+\begin{theorem}{Parameter-dependent Nyström trace estimator}{nystrom}
+    Suppose $f(\mtx{A}, t)$ is continuous and non-negative in $t \in [a, b]$ and $r \geq 2, p \geq 4$. Then for all $\gamma \geq 1$ with probability at least $1 - \gamma^{-p}$
     \begin{equation}
-        \int_{a}^{b} \left| \Tr(f(\mtx{A}, t)) - \Tr(\widehat{f}(\mtx{A}, t)) \right| ~\mathrm{d}t
+        \int_{a}^{b} \left| \Trace(f(\mtx{A}, t)) - \Nystr{r+p}(f(\mtx{A}, t)) \right| ~\mathrm{d}t
         \leq \gamma^2 (1 + r) \int_{a}^{b} \sum_{j = r+1}^{n} \sigma_j(f(\mtx{A}, t)) ~\mathrm{d}t
@@
 \begin{proof}
+    Since $f$ is non-negative, $f(\mtx{A}, t)$ is positive definite for all $\mtx{A}$ and $t$.
     Notice that \cite[Lemma 2.1, item 3]{frangella-2023-randomized-nystrom} implies that $f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t)$ is positive semi-definite \textcolor{red}{also for all $t$?}
     \begin{equation}
-        \left| \Tr(f(\mtx{A}, t)) - \Tr(\widehat{f}(\mtx{A}, t)) \right|
-        = \lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _1
+        \left| \Trace(f(\mtx{A}, t)) - \Trace(\widehat{f}(\mtx{A}, t)) \right|
+        = \lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast}
     \end{equation}
@@
     \begin{equation}
-        \lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _1 = \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{1/2} \mtx{\Omega}}) f(\mtx{A}, t)^{1/2} \rVert _2^2
+        \lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{1/2} \mtx{\Omega}}) f(\mtx{A}, t)^{1/2} \rVert _2^2
     \end{equation}
@@
 
-\subsection{Parameter-dependent Nyström++ estimator}
+\subsection{Parameter-dependent Nystr\"om++ estimator}
 \label{subsec:nystrom-pp}
+
+\begin{theorem}{Parameter-dependent Nystr\"om++ trace estimator}{nystrom-pp}
+    If $n_{\Psi} = n_{\Omega} = \mathcal{O}(\varepsilon^{-1} \log(\delta^{-1}))$, with probability at least $1 - \delta$
+    \begin{equation}
+        \int_{a}^{b} | \Trace(\mtx{A}(t)) - \Nystrpp{n_{\Omega}}{n_{\Psi}}(\mtx{A}(t)) |~\mathrm{d}t
+        \leq \varepsilon \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t
+    \end{equation}
+\end{theorem}
+
+\begin{proof}
+    Combining with union bound.
+    \begin{align}
+        \int_{a}^{b} | \Trace(\mtx{A}(t)) - \Nystrpp{n_{\Omega}}{n_{\Psi}}(\mtx{A}(t)) |~\mathrm{d}t
+        &= \int_{a}^{b} | \Hutch{n_{\Psi}}(\mtx{A}(t) - \widehat{\mtx{A}}(t)) - \Trace(\mtx{A}(t) - \widehat{\mtx{A}}(t)) |~\mathrm{d}t \notag \\
+        &= \todo{c} \frac{1}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F ~\mathrm{d}t \notag \\
+        &= \todo{c} \frac{1}{\sqrt{n_{\Psi} n_{\Omega}}} \int_{a}^{b} \Trace(\mtx{A}(t)) ~\mathrm{d}t
+    \end{align}
+\end{proof}
 
@@
 \begin{equation}
-    \phi_{\sigma}(t) = \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i) = \Tr(g_{\sigma}(t\mtx{I}_n - \mtx{A}))
+    \phi_{\sigma}(t) = \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i) = \Trace(g_{\sigma}(t\mtx{I}_n - \mtx{A}))
     \label{equ:smooth-spectral-density}
@@
     g_{\sigma}(s) = \frac{1}{n \sigma \sqrt{2\pi}} e^{-\frac{s^2}{2\sigma^2}}
-    \label{smoothing-kernel}
+    \label{equ:smoothing-kernel}
 \end{equation}
@@
 \end{equation}
+Then
+\begin{equation}
+    \phi_{\sigma}^{(m)}(t) = \Trace(g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}))
+    \label{equ:expanded-spectral-density}
+\end{equation}
+
+Spectral transform (mention estimation of $[a,b]$ see notes, and maybe some buffer region to avoid cut-off)
 
@@
 \begin{lemma}{Chebyshev expansion error}{chebyshev-error}
-    Let $\mtx{A} \in \mathbb{R}^{n \times n}$ be a symmetric matrix whose spectrum is contained in $[-1, 1]$. Then the expansion $g_{\sigma}^{(m)}$ of the Gaussian smoothing kernel satisfies
+    The expansion $g_{\sigma}^{(m)}$ of the Gaussian smoothing kernel $g_{{\sigma}}$ satisfies
     \begin{equation}
-        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \right| \leq \sqrt{\frac{2e}{\pi}} \frac{1}{n \sigma^2} (1 + \sigma)^{-m} \equiv E_{\sigma, m}
+        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \right| \leq \sqrt{\frac{2e}{\pi}} \frac{1}{n \sigma^2} (1 + \sigma)^{-m} \equiv \frac{E_{\sigma, m}}{n}
         \label{equ:2-chebyshev-interpolation-sup-error-kernel}
@@
     \end{equation}
-    for all degrees $m > 0$ and smoothing parameters $\sigma>0$.
+    for all degrees $m > 0$, smoothing parameters $\sigma > 0$, and $s \in [-1, 1]$.
 \end{lemma}
 
-Closely follows \cite[Theorem 2]{lin-2017-randomized-estimation}
+The proof specializes Bernstein's theorem for Chebyshev interpolation to the case of a Gaussian smoothing kernel \refequ{equ:smoothing-kernel}.
+\todo{It loosely follows the proof give in the preprint of \cite[Theorem 2]{lin-2017-randomized-estimation}, but reaches slightly different conclusion.}
 
 \begin{proof}
-From Bernstein's theorem \cite[Theorem 4.3]{trefethen-2008-gauss-quadrature} if $f$ analytic within Bernstein ellipse $\mathcal{E}_{\chi}$, then for any $m > 0$
+    From Bernstein's theorem \cite[Theorem 4.3]{trefethen-2008-gauss-quadrature} it follows that if $f$ is analytic within the Bernstein ellipse $\mathcal{E}_{\chi} \subset \mathbb{C}$ with foci $\pm 1$ and sum of semi-axes $\chi > 1$, then for any $m \in \mathbb{N}$
     \begin{equation}
-    \sup_{t \in [-1, 1]} \left| f - f^{(m)} \right| \leq \frac{2}{\chi^m (\chi - 1)} \max_{z \in \mathcal{E}_{\chi}} |f(z)|
+        \sup_{t \in [-1, 1]} \left| f - f^{(m)} \right| \leq \frac{2}{\chi^m (\chi - 1)} \max_{z \in \mathcal{E}_{\chi}} |f(z)|.
         \label{equ:bernstein-bound}
     \end{equation}
-
-In particular, for $g_{\sigma}(t - \cdot)$ we observe for $z = x + iy \in \mathcal{E}_{\chi}$
+    In particular, when $f = g_{\sigma}(t - \cdot)$ we observe for $z = x + iy \in \mathcal{E}_{\chi}$
     \begin{equation}
@@
     = \frac{1}{n \sigma \sqrt{2 \pi}} \left| e^{-\frac{(t - z)^2}{2\sigma^2}} \right|
-= \frac{1}{n \sigma \sqrt{2 \pi}} e^{-\frac{(t - x)^2}{2\sigma^2}}e^{\frac{y^2}{2\sigma^2}}
+    = \frac{1}{n \sigma \sqrt{2 \pi}} e^{-\frac{(t - x)^2}{2\sigma^2}}e^{\frac{y^2}{2\sigma^2}}.
     %\leq \frac{1}{n \sigma \sqrt{2 \pi}} \max_{x + iy \in \mathcal{E}_{\chi}} e^{\frac{y^2}{2\sigma^2}} 
     \end{equation}
-
-Since $e^{-\frac{(t - x)^2}{2\sigma^2}} \leq 1$ and the maximum absolute value of $y$ is limited by the imaginary half-axis of the Bernstein ellipse, which has length $(\chi - \chi^{-1}) / 2$, we upper bound
-
+    Because $e^{-\frac{(t - x)^2}{2\sigma^2}} \leq 1$ for all $x \in \mathbb{R}$ and $t \in [-1,1]$, and the maximum absolute value of $y$ is limited by the length of the semi-axis of the Bernstein ellipse in the direction of the imaginary axis, which is $(\chi - \chi^{-1}) / 2$, we upper bound
     \begin{equation}
@@
 
-What turns out to be most useful in practice is the choice $\chi = 1 + \sigma$. Since in this case $\chi - \chi^{-1} \leq 2\sigma$, this implies
+    Choosing $\chi = 1 + \sigma$ gives a simple and good bound, since in this case $\chi - \chi^{-1} \leq 2\sigma$, which implies
     \begin{equation}
@@
 
-Squaring of Chebyshev expansion for consistent and affine Nyström approximation
+Squaring of Chebyshev expansion for consistent and affine Nystr\"om approximation
 \begin{equation}
@@
 \end{equation}
-where $\nu = \DCT^{-1} \left\{ \DCT\{\mu\}^2 \right\}$
+where $\nu = \DCT^{-1} \left\{ \DCT\{\mu\}^2 \right\}$ \todo{which is consistent unlike \cite{lin-2017-randomized-estimation}}.
 
-Non-negative Chebyshev expansion for results in Nyström approximation of positive semi-definite matrices (discuss alternatives: Jackson damping, Putinar representation)
+Non-negative Chebyshev expansion for results in Nystr\"om approximation of positive semi-definite matrices (discuss alternatives: Jackson damping, Putinar representation)
 
 \begin{lemma}{Non-negative Chebyshev expansion error}{non-negative-chebyshev-error}
-    Let $\mtx{A} \in \mathbb{R}^{n \times n}$ be a symmetric matrix whose spectrum
-    is contained in $[-1, 1]$. Then the non-negative expansion $g_{\sigma}^{(\underline{m})}$ of the Gaussian smoothing kernel satisfies
+    The non-negative expansion $\underline{g}_{\sigma}^{(m)}$ of the Gaussian smoothing kernel satisfies
     \begin{equation}
-        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - g_{\sigma}^{(\underline{m})}(t - s) \right| \leq 2\sqrt{2} \left(1 + n \sigma \sqrt{\pi} E_{\sqrt{2}\sigma, \frac{m}{2}}\right) E_{\sqrt{2}\sigma, \frac{m}{2}} \equiv \underline{E}_{\sigma, m}
+        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \leq 2\sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \frac{m}{2}}\right) \frac{E_{\sqrt{2}\sigma, \frac{m}{2}}}{n} \equiv \frac{\underline{E}_{\sigma, m}}{n}
         \label{equ:2-chebyshev-interpolation-sup-error-kernel}
@@
     \end{equation}
-    for all \emph{even} degrees $m > 0$ and smoothing parameters $\sigma>0$.
+    for all \emph{even} degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and $s \in [-1, 1]$.
 \end{lemma}
@@
 \begin{proof}
-
     For any numbers $a, b \in \mathbb{R}$ it holds
     \begin{equation}
-| a^2 - b^2 | = | (a + b)(a - b) |\leq | a + b | | a - b | \leq (2 | a | + | a - b |)  | a - b |
+    | a^2 - b^2 | = | (a + b)(a - b) | = | a + b | | a - b | \leq (2 | a | + | a - b |)  | a - b |
     \end{equation}
-
-Therefore, with $g_{\sigma} = \sqrt{g_{\sigma}}^2$ and $g_{\sigma}^{(\underline{m})} = (\sqrt{g_{\sigma}}^{(m/2)})^2$
+    Therefore, omitting the arguments and noting $g_{\sigma} = (\sqrt{g_{\sigma}})^2$ and $\underline{g}_{\sigma}^{(m)} = (\sqrt{g_{\sigma}}^{(m/2)})^2$.
     we have
     \begin{equation}
-    \left| g_{\sigma} - g_{\sigma}^{(\underline{m})} \right| \leq \left( 2 \left| \sqrt{g_{\sigma}} \right| + \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(m/2)} \right| \right) \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(m/2)} \right|
+        \left| g_{\sigma} - \underline{g}_{\sigma}^{(m)} \right| \leq \left( 2 \left| \sqrt{g_{\sigma}} \right| + \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(m/2)} \right| \right) \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(m/2)} \right|.
     \end{equation}
-
-Since 
-\begin{equation}
-\sqrt{g_{\sigma}} = \sqrt{2 n \sigma \sqrt{2 \pi}} g_{\sqrt{2}\sigma}
-\end{equation}
-we can apply the results from \textcolor{red}{above} to get
-\begin{equation}
-    \left| g_{\sigma}(t - s) - g_{\sigma}^{(\underline{m})}(t - s) \right| \leq \left( 2 \frac{1}{\sqrt{n \sigma \sqrt{2\pi}}} + \sqrt{2 n \sigma \sqrt{2 \pi}} E_{\sqrt{2}\sigma, \frac{m}{2}}\right)\sqrt{2 n \sigma \sqrt{2 \pi}} E_{\sqrt{2}\sigma, \frac{m}{2}}
-\end{equation}
-from which follows the result with some simplification.
+    Since $\sqrt{g_{\sigma}} = \sqrt{2 n \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$ and $\sqrt{g_{\sigma}} \leq 1/\sqrt{n \sigma \sqrt{2 \pi}}$ we can apply the result from \reflem{lem:chebyshev-error} to get
+    \begin{align}
+        &\sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \notag \\
+        &\leq \left( 2 \frac{1}{\sqrt{n \sigma \sqrt{2\pi}}} + \sqrt{2 n \sigma \sqrt{2 \pi}} \cdot \frac{E_{\sqrt{2}\sigma, \frac{m}{2}}}{n}\right)\sqrt{2 n \sigma \sqrt{2 \pi}} \cdot \frac{E_{\sqrt{2}\sigma, \frac{m}{2}}}{n},
+    \end{align}
+    from which follows the result with some minor simplifications.
 \end{proof}
 
-\subsection{The Chebyshev-Nyström++ method for approximating spectral densities}
+\subsection{The Chebyshev-Nystr\"om++ method for approximating spectral densities}
 \label{subsec:chebyshev-nystrom}
@@
     \Statex \textbf{Parameters:} Degree $m$, allocation $(n_{\Omega}, n_{\Psi})$,  smoothing parameter $\sigma$
-    \Statex \textbf{Output:} Approximate evaluations of the spectral density $\{\breve{\phi}_{\sigma}^{(m)}(t_i)\}_{i=1}^{n_t}$
+    \Statex \textbf{Output:} Approximate evaluations of the spectral density $\{\phi_{\sigma}(t_i)\}_{i=1}^{n_t}$
     \State Compute $\{\mu_l(t_i)\}_{l=0}^{m}$ and $\{\nu_l(t_i)\}_{l=0}^{2m}$ for all $t_i$ %using %\refalg{alg:2-chebyshev-chebyshev-expansion}
@@
       %\State $\mtx{Y} \gets \mtx{\mtx{\Omega}}^{\top} \mtx{W}_2$
-      \State $z \gets \Tr(\mtx{\Psi}^{\top} \mtx{W}_2)$
+      \State $z \gets \Trace(\mtx{\Psi}^{\top} \mtx{W}_2)$
       \For {$i = 1, \dots, n_t$}
@@
             \State $\mtx{L}_1(t_i) \gets \mtx{L}_1(t_i) + \mu_l(t_i) \mtx{Y}$ \Comment{assemble $\mtx{\Psi}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$}
-            \State $\ell(t_i) \gets \ell(t_i) + \mu_l(t_i) z$ \Comment{assemble $\Tr(\mtx{\Psi}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi})$}
+            \State $\ell(t_i) \gets \ell(t_i) + \mu_l(t_i) z$ \Comment{assemble $\Trace(\mtx{\Psi}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi})$}
         \EndIf
@@
     \For {$i = 1, \dots, n_t$}
-      \State $\breve{\phi}_{\sigma}^{(m)}(t_i) \gets \Tr\left( \mtx{K}_1(t_i)^{\dagger}\mtx{K}_2(t_i) \right) + \frac{1}{n_{\Psi}} \left( \ell(t_i) + \Tr\left( \mtx{L}_1(t_i)^{\top} \mtx{K}_1(t_i)^{\dagger} \mtx{L}_1(t_i) \right)  \right) $ \label{lin:4-nystromchebyshev-nystrom-pp}
+      \State $\phi_{\sigma}(t_i) \gets \Trace\left( \mtx{K}_1(t_i)^{\dagger}\mtx{K}_2(t_i) \right) + \frac{1}{n_{\Psi}} \left( \ell(t_i) + \Trace\left( \mtx{L}_1(t_i)^{\top} \mtx{K}_1(t_i)^{\dagger} \mtx{L}_1(t_i) \right)  \right) $ \label{lin:4-nystromchebyshev-nystrom-pp}
     \EndFor
@@
     Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the spectal density constructed with a Chebyshev expansion $\phi_{\sigma}^{m}$ satisfies
+Then the spectal density with Chebyshev expansion $\phi_{\sigma}^{(m)}$ satisfies
     \begin{equation}
-        \int_{a}^{b} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq 2 n E_{\sigma, m}
+        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq 2 E_{\sigma, m}
         \label{equ:chebyshev-interpolation-sup-error-kernel}
@@
     \end{equation}
-    for all degrees $m > 0$ and smoothing parameters $\sigma>0$.
-The same result holds for the expansion of a non-negative Chebyshev expansion $\phi_{\sigma}^{(\underline{m})}$ with $\underline{E}_{\sigma, m}$.
+    for all degrees $m \in \mathbb{N}$ and smoothing parameters $\sigma > 0$.
+The same result holds for the expansion of a non-negative Chebyshev expansion of \emph{even} degree $m \in \mathbb{N}$ but with $\underline{E}_{\sigma, m}$.
 \end{lemma}
@@
 \begin{proof}
+    We observe that for all $t \in [-1, 1]$
     \begin{align}
         &\left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right| \notag \\
-        &= \left| \Tr(g_{\sigma}(t\mtx{I}_n - \mtx{A})) - \Tr(g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})) \right|
-        && \text{(definitions \refequ{equ:1-introduction-spectral-density-as-trace} and \refequ{equ:2-chebyshev-chebyshev-expansion})} \notag \\
+        &= \left| \Trace(g_{\sigma}(t\mtx{I}_n - \mtx{A})) - \Trace(g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})) \right|
+        && \text{(definitions \refequ{equ:smooth-spectral-density} and \refequ{equ:expanded-spectral-density})} \notag \\
         &= \left| \sum_{i=1}^n \left(g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i)\right) \right|
@@
         && \text{(conservative upper bound)} \notag \\
-        &\leq n E_{\sigma, m}
-        && \text{(using \refequ{equ:2-chebyshev-uniform-bound})} \notag \\
-    \end{align}\\
-
-    Finally, H\"older's inequality %\cite{klenke2013probability} 
-    allows us to also show the last assertion with what we have found above:
+        &\leq E_{\sigma, m}.
+        && \text{(using \reflem{lem:chebyshev-error})} \notag \\
+    \end{align}
+    Then, H\"older's inequality \cite{} allows us to conclude%\cite{klenke2013probability} 
     \begin{equation}
-        \int_{a}^{b} | \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) | ~\mathrm{d}t
+        \int_{-1}^{1} | \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) | ~\mathrm{d}t
             \leq 2 \sup_{t \in [-1, 1]} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|
-            \leq 2 n E_{\sigma, m}.
+            \leq 2 E_{\sigma, m}.
     \end{equation}
@@
 
+\clearpage
+\bibliography{bibliography.bib}
+
+\appendix
+
+\clearpage
+\section{Proofs}
+
 \end{document}

### 2024-05-21 paper/figures/numerical-rank.tex
finish first draft of paper

\begin{tikzpicture}
    \fill[lightblue] (-2, -0.15) rectangle (1.5, 0.15);
    \draw[thick, ->] (-5, 0) to (5, 0);
    \fill[darkblue] (-4, 0) circle (0.075) node[above] {$\lambda_{(7)}$};
    \fill[darkblue] (-3, 0) circle (0.075) node[above] {$\lambda_{(5)}$};
    \fill[darkblue] (-2.3, 0) circle (0.075) node[above] {$\lambda_{(4)}$};
    \fill[darkblue] (-1, 0) circle (0.075) node[above] {$\lambda_{(2)}$};
    \fill[darkblue] (0.3, 0) circle (0.075) node[above] {$\lambda_{(1)}$};
    \fill[darkblue] (1, 0) circle (0.075) node[above] {$\lambda_{(3)}$};
    \fill[darkblue] (2.8, 0) circle (0.075) node[above] {$\lambda_{(6)}$};
    \fill[darkblue] (3.6, 0) circle (0.075) node[above] {$\lambda_{(8)}$};
    \fill[darkblue] (4.4, 0) circle (0.075) node[above] {$\lambda_{(9)}$};
    \draw[darkblue, ultra thick] (-0.155, 0.15) to (-0.155, -0.15) node[below] {$t$};
    \draw[darkblue, thick] (-2, 0.15) to (-2, -0.15) node[below] {$t - d_{\varepsilon, \sigma}$};
    \draw[darkblue, thick] (1.5, 0.15) to (1.5, -0.15) node[below] {$t + d_{\varepsilon, \sigma}$};
\end{tikzpicture}

### 2024-05-21 paper/main.tex
finish first draft of paper

@@
 
-\todo{
-\begin{itemize}
-    \item Unify index notation
-    \item Double-check proofs
-    \item Add all references
-    \item Run numerical experiments 
-    \item Write introduction / abstract
-\end{itemize}
-}
+%\todo{
+%\begin{itemize}
+%    \item Unify index notation
+%    \item Double-check proofs
+%    \item Add all references
+%    \item Run numerical experiments 
+%    \item Write introduction, abstract, and discussion
+%\end{itemize}
+%}
 
 \begin{abstract}
-    Abstract
+    \todo{Write abstract.}
 \end{abstract}
@@
 
-Need for spectral density approximation (cite some literature)
+\begin{itemize}
+    \item Definition of parameter-dependent matrix ($\mtx{B}(t) \in \mathbb{R}^{n \times n}$, every entry depends continuously on $t \in [a,b]$ \cite{kressner-2023-randomized-lowrank})
+    \item Trace estimation through mat-vecs (if matrix is only accessed implicitly [Hessian, Graph] or for matrix functions [partition function, spectral density, exponential integrators])
+    \item Mention existing trace estimation methods for constant matrices (quadratic [Hutchinson, Girard], Hutch++ \cite{meyer-2021-hutch-optimal}, Nyström++ \cite{persson-2022-improved-variants}, XTrace [Tropp], Krylov aware [Chen]...)
+\end{itemize}
 
-Connection to trace estimation (maybe only in words, not in formulas)
+\paragraph{Spectral density approximation}
+\begin{itemize}
+    \item Definition of spectral density ($\phi(t) = n^{-1}\sum_{i=1}^{m} \delta(t - \lambda_i)$ for matrix $\mtx{A}$ with eigenvalues $\lambda_1, \dots, \lambda_n$)
+    \item Applications of spectral density approximation (electronic structure calculations \cite{lin-2017-randomized-estimation}, neural network optimization [Hessian, ...], graph processing [filtering, ...])
+    \item Link to trace estimation of a matrix function (smoothing with Gaussian \cite{lin-2017-randomized-estimation} or Haydock [Haydock, Lin] kernel)
+    \item Existing theory for spectral density approximation (Wasserstein-1 error [Braverman, Chen], missing $1/\varepsilon$ guarantees)
+\end{itemize}
 
-For \emph{many} evaluations of parameter, i.e. try to decrease complexity with each evaluation
+\paragraph{Contribution}
+\begin{itemize}
+    \item Development of a unified family of estimators for many evaluations of parameter, i.e. try to keep complexity with each evaluation as low as possible
+    \item Extends \cite{kressner-2023-randomized-lowrank} to trace estimation
+    \item Provides theoretical guarantees for methods in \cite{lin-2017-randomized-estimation}
+\end{itemize}
 
-Our work: Analysis added to Lin Lin, also works for general matrix functions
-
-\subsection{Notation}
+\paragraph{Notation}
+We follow the notation and conventions which are used in contemporary literature in this field. Further,
 \begin{itemize}
-    \item $e$ and $\log$
-    \item the $k$-th moment of a random variable is denoted with $\mathbb{E}^{k}[X] = \left(\mathbb{E}\left[ | X |^{k} \right] \right)^{\frac{1}{k}}$;
-    \item in a Gaussian random vector and a Gaussian random matrix, all entries follow independently the standard normal distribution;
-    \item Schatten norms
-    \item $\mathbb{N}$ all positive whole numbers
+    \item we denote with $\mathbb{N}$ all positive whole numbers and with $\mathbb{R}$ all real numbers;
+    \item $e$ denotes Euler's number and $\log$ is the corresponding logartihm;
+    \item the ceiling function $\lceil x \rceil$ of a real number $x$ returns the greatest integer larger than or equal to $x$;
+    \item scalars are represented by lower case Greek or Latin letters ($s$, $\varepsilon$, \dots),
+    vectors are additionally printed in bold ($\vct{v}$, $\vct{\psi}$, \dots),
+    and matrices are additionally capitalized ($\mtx{A}$, $\mtx{\Omega}$, \dots);
+    \item the expected value of a random variable $x$ is denoted with $\mathbb{E}[x]$.
+The $k$-th moment of a random variable $x$ is defined as $\mathbb{E}^{k}[x] = \left(\mathbb{E}\left[ | x |^{k} \right] \right)^{\frac{1}{k}}$;
+    \item the identity matrix $\mtx{I}_n = \diag(1, \dots, 1) \in \mathbb{R}^{n \times n}$
+    carries ones on its diagonal and zeros everywhere else. The zero matrix
+    $\mtx{0}_{n \times m} \in \mathbb{R}^{n \times m}$ consists of only zero entries;
+    \item in a Gaussian random vector and a Gaussian random matrix, all entries are independently sampled from the standard normal distribution;
+    \item matrix norms are denoted with $\lVert \cdot \rVert _{\cdot}$. For symmetric matrices, the $p$-Schatten norm is
+    $\lVert \mtx{A} \rVert _{(p)} = (\sum_{i=1}^n |\lambda_i|^p)^{1/p}$.
+We refer to the spectral norm as $\lVert \mtx{A} \rVert _2 = \lVert \mtx{A} \rVert _{(\infty)} = \max_{i=1,\dots,n}|\lambda_i|$, the nuclear norm as $\lVert \mtx{A} \rVert _{\ast} = \lVert \mtx{A} \rVert _{(1)} = \sum_{i=1}^{n} |\lambda_i|$, and the Frobenius norm as $\lVert \mtx{A} \rVert _F = \lVert \mtx{A} \rVert _{(2)} = (\sum_{i=1}^{n} |\lambda_i|^2)^{\frac{1}{2}}$.
 \end{itemize}
@@
         a_{n1}(t) & a_{n2}(t) & \dots & a_{nn}(t) \\
-    \end{bmatrix}.
+    \end{bmatrix} \in \mathbb{R}^{n \times n}
 \end{equation}
-where $t$ is a parameter taking values in the interval $[a,b]$.
-Mostly analogous to constant matrix.
-
-Trace is defined as
+where $a_{ij}(t)$ are functions depending continuously on the parameter $t$ which takes values in the interval $[a,b]$. The trace of such a matrix is defined as
 \begin{equation}
@@
 \end{equation}
-If we can only access the parameter-dependent matrix through matrix-vector products, we can estimate the trace with the \emph{quadratic trace estimator} \cite{}: We take $n_{\Psi} \in \mathbb{N}$ Gaussian random vectors $\vct{\psi}_1,\vct{\psi}_2 \dots, \vct{\psi}_{n_{\psi}} \in \mathbb{R}^{n}$ to form
+
+\paragraph{Quadratic trace estimator} If we can only access the parameter-dependent matrix through matrix-vector products, we can estimate the trace with the quadratic trace estimator \cite{}: We take $n_{\Psi} \in \mathbb{N}$ Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\psi}} \in \mathbb{R}^{n}$ to form
 \begin{equation}
     \Hutch{n_{\Psi}}(\mtx{A}(t)) = \frac{1}{n_{\Psi}} \sum_{j=1}^{n_{\Psi}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j.
+    \label{equ:quadratic-trace-estimator}
 \end{equation}
-If we gather the random vectors in a matrix $\mtx{\Psi} = [\vct{\psi}_1 ~ \vct{\psi}_2 ~ \cdots ~ \vct{\psi}_{n_{\psi}}] \in \mathbb{R}^{n \times n_{\Psi}}$ we may rewrite $\Hutch{n_{\Psi}}(\mtx{A}(t)) = n_{\Psi}^{-1} \Trace( \mtx{\Psi}^{\top} \mtx{A}(t) \mtx{\Psi})$.
+If we gather the random vectors in a matrix $\mtx{\Psi} = [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\psi}}] \in \mathbb{R}^{n \times n_{\Psi}}$ we may rewrite $\Hutch{n_{\Psi}}(\mtx{A}(t)) = n_{\Psi}^{-1} \Trace( \mtx{\Psi}^{\top} \mtx{A}(t) \mtx{\Psi})$.
 Other choices for the distribution of the entries of the random vectors are possible \cite{}.
@@
 
-Alternatively, the trace of a symmetric matrix whose rank is significantly lower than its size can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\Omega}}$ to form the \emph{Nystr\"om approximation}
+\paragraph{Nyström trace approximation} Alternatively, the trace of a symmetric matrix whose rank is significantly lower than its size can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\Omega}}$ to form the Nyström approximation
 \begin{equation}
@@
     \Nystr{n_{\Omega}}(\mtx{A}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{A}(t)^2 \mtx{\Omega})).
+    \label{equ:nystrom-trace-estimator}
 \end{equation}
 
-Finally, a combination of the two estimators yields an analogue for the \emph{Nyström++ estimator} \cite{persson-2022-improved-variants} for parameter-dependent settings \todo{this is super ugly.
-find better notation}
+\paragraph{Nyström++ estimator} Finally, a combination of the two estimators yields an analogue for the Nyström++ estimator \cite{persson-2022-improved-variants} in the parameter-dependent setting
 \begin{equation}
     \Nystrpp{n_{\Omega}}{n_{\Psi}}(\mtx{A}(t)) = \Nystr{n_{\Omega}}(\mtx{A}(t)) + \Hutch{n_{\Psi}}(\mtx{A}(t) - \widehat{\mtx{A}}(t)).
+    \label{equ:nystrompp-trace-estimator}
 \end{equation}
 It is understood that $\widehat{\mtx{A}}$ is computed with the same sketching matrix as the one used for the trace estimator $\Nystr{n_{\Omega}}$.
-We can interpret this estimator as an interpolation between the trace of the Nystr\"om approximation and the quadratic trace estimator.
+We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the quadratic trace estimator.
+
+\paragraph{$L^{1}$ norm}
+\todo{Give definition of $L^1$ norm and reason for choosing it}
 
 \subsection{Parameter-dependent quadratic trace estimator}
-\label{subsec:hutchinson}
+\label{subsec:quadratic}
+
+We first consider \refequ{equ:quadratic-trace-estimator}.
+\todo{Discuss theory for constant matrix
+\begin{equation}
+    \lVert f \rVert _1 = \int_{a}^{b} | f(t) | \mathrm{d}t
+\end{equation}
+}
 
@@
     \end{equation}
+    In particular, if we choose $n_{\Psi} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{\frac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{A}(t)) - \Hutch{n_{\Psi}}(\mtx{A}(t)) | ~\mathrm{d}t < \varepsilon \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t$.
 \end{theorem}
@@
         \end{pmatrix}
-        \qquad \text{and} \qquad
+        \quad \text{and} \quad
         \vct{\psi} = \begin{pmatrix}
@@
     \begin{equation}
-        r(t) = \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Trace(\mtx{A}(t)).
+        r(t) = \frac{1}{n_{\Psi}} \sum_{j=1}^{n_{\Psi}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j - \Trace(\mtx{A}(t)).
     \end{equation}
@@
     \begin{equation}
-        \int_{a}^{b} \left| \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Trace(\mtx{A}(t)) \right| ~ \mathrm{d}t
+        \int_{a}^{b} \left| \frac{1}{n_{\Psi}} \sum_{j=1}^{n_{\Psi}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j - \Trace(\mtx{A}(t)) \right| ~ \mathrm{d}t
         < 16 k \gamma \frac{1}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t.
@@
 
-    %Setting $\delta = \gamma^{-2 k}$ (i.e. $\gamma = \delta^{-\frac{1}{2k}}$) and choosing $k = \lceil \log(\delta^{-1}) \rceil$ we get that for all $\delta \in (0, e^{-\frac{1}{2}})$
-    %\begin{equation}
-    %    k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\frac{1}{2\lceil \log(\delta^{-1}) \rceil}}
-    %    = \lceil \log(\delta^{-1}) \rceil e^{\frac{1}{2}\frac{\log(\delta^{-1})}{\lceil \log(\delta^{-1}) \rceil}}
-    %    \leq 2 \log(\delta^{-1}) e^{\frac{1}{2}}
-    %\end{equation}
-    %
+    Setting $\delta = \gamma^{-2 k}$ and choosing $k = \lceil \log(\delta^{-1}) \rceil$ we get that for all $\delta \in (0, e^{-\frac{1}{2}})$
+    \begin{equation}
+        k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\frac{1}{2\lceil \log(\delta^{-1}) \rceil}}
+        = \lceil \log(\delta^{-1}) \rceil e^{\frac{1}{2}\frac{\log(\delta^{-1})}{\lceil \log(\delta^{-1}) \rceil}}
+        \leq 2 \log(\delta^{-1}) e^{\frac{1}{2}},
+    \end{equation}
+    from which follows the second part of the theorem.
     %Consequently, with probability $1 - \delta$ we have
@@
     %\end{equation}
-
 \end{proof}
 
-\subsection{Parameter-dependent Nystr\"om approximation}
+\todo{Discuss $1/\varepsilon^2$ drawback}
+
+\subsection{Parameter-dependent Nyström approximation}
 \label{subsec:nystrom}
+
+Next we consider \refequ{equ:nystrom-trace-estimator}. \todo{Motivate Nyström approximation if matrix is low-rank}
 
@@
     \end{bmatrix}
-    \qquad 
+    \quad \text{and} \quad
     \mtx{\Lambda}(t) =
@@
 
-\begin{lemma}{Moment tail bound}{ose}
+\begin{lemma}{Moments of tail bound}{ose}
     A symmetric matrix $\mtx{A} \in \mathbb{R}^{l \times l}$ and stochastically independent Gaussian random matrices $\mtx{\Omega}_1 \in \mathbb{R}^{k \times 2k}, \mtx{\Omega}_2 \in \mathbb{R}^{l \times 2k}$ satisfy for all $k \in \mathbb{N}$ and a constant $c > 0$
@@
 \begin{lemma}{Parameter-dependent Nystrom approximation}{nystrom}
-    $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ symmetric positive semi-definite and continuous in $t \in [a, b]$.
-$\mtx{\Omega} \in \mathbb{R}^{n \times n_{\Omega}}$ standard Gaussian with \textcolor{red}{$n_{\Omega} \geq 8 \log(\delta^{-1})$}.
-Then with probability at least $1 - \delta$
+    Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
+Then its Nyström approximation $\widehat{\mtx{A}}(t)$ with Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\Omega}}$ for even $n_{\Omega} \in \mathbb{N}$ and $\gamma \geq 1$ verifies with probability at least $1 - \gamma^{-\frac{n_{\Omega}}{4}}$
     \begin{equation}
-        \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F ~\mathrm{d}t
-        \leq C_{\Omega}\frac{1}{\sqrt{n_{\Omega}}} \int_{a}^{b} \Trace(\mtx{A}(t)) ~\mathrm{d}t 
+        \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F~\mathrm{d}t \leq \gamma \sqrt{2} \cdot \frac{1 + 2c}{\sqrt{n_{\Omega}}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t.
     \end{equation}
+    for a universal constant $c > 0$.
+In particular, if we choose $n_{\Omega} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})), n_{\Omega} \geq 2$, then with probability at least $1-\delta$ for $\delta \in (0, e^{-e^{4 \varepsilon^2}})$ and $\varepsilon > 0$ we have $\int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F~\mathrm{d}t  < \varepsilon \int_{a}^{b} \Trace(A(t))~\mathrm{d}t$.
 \end{lemma}
 
-\todo{proof idea}
+\todo{proof idea (structural bound, then higher order moment bound to apply Markov's inequality)}
 
 \begin{proof}
-    Omitting the parameter dependence, we can apply \cite[Theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound the integrand with
+    We let $n_{\Omega} = 2k$ for some $k \in \mathbb{N}$, i.e.
+$\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$.
+After temporarily omitting the parameter dependence, we can apply \cite[Theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound the integrand with \todo{maybe explain where $\mtx{\Omega}_i$ come from...}
     \begin{equation}
@@
         \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert (\mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F.
+        \label{equ:nystrom-proof-persson-bonud}
     \end{equation}
@@
         \leq \frac{1}{\sqrt{k}} \Trace(\mtx{A}).
+        \label{equ:nystrom-proof-frobenius-trace}
     \end{equation}
@@
         \leq \sqrt{k} \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2.
+        \label{equ:nystrom-proof-processed-tail}
     \end{equation}
-    Then we consider its moments
+    Its higher order moments are bounded by
     \begin{align}
@@
         &= \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]^2 && \text{(by definition of $\mathbb{E}^{k}$)} \notag \\
-        &\leq C \left( \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _F \right)^2 && \text{(\reflem{lem:ose} with $\mtx{A}=\mtx{\Lambda}_2^{\frac{1}{2}}$)} \notag \\
-        &\leq C \left( \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _2^2  + \frac{1}{k}\lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _F^2 \right) && \text{($(a + b)^2 \leq a^2 + b^2$)} \notag \\
-        &= C \left( \lambda_{k+1} + \frac{1}{k}\Trace(\mtx{\Lambda}_2) \right) && \text{(definition of $\lVert \cdot \rVert _2$ and $\lVert \cdot \rVert _F$)} \notag \\
-        &\leq \frac{2 C}{k} \Trace(\mtx{A}) && \text{($k \lambda_{k+1} \leq \Trace(\mtx{A})$)}
+        &\leq c \left( \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _F \right)^2 && \text{(\reflem{lem:ose} with $\mtx{A}=\mtx{\Lambda}_2^{\frac{1}{2}}$)} \notag \\
+        &\leq c \left( \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _2^2  + \frac{1}{k}\lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _F^2 \right) && \text{($(a + b)^2 \leq a^2 + b^2$)} \notag \\
+        &= c \left( \lambda_{k+1} + \frac{1}{k}\Trace(\mtx{\Lambda}_2) \right) && \text{(definition of $\lVert \cdot \rVert _2$ and $\lVert \cdot \rVert _F$)} \notag \\
+        &\leq \frac{2 c}{k} \Trace(\mtx{A}) && \text{($\lambda_{k+1} \leq \frac{\Trace(\mtx{A})}{k}$, $\Trace(\mtx{\Lambda}_2) \leq \Trace(\mtx{A})$)}
+        \label{equ:nystrom-proof-tail-bound}
     \end{align}
-    Thus, we get
+    Thus, inserting \refequ{equ:nystrom-proof-tail-bound} into \refequ{equ:nystrom-proof-processed-tail} and using it along with \refequ{equ:nystrom-proof-frobenius-trace} into \refequ{equ:nystrom-proof-persson-bonud} we obtain
     \begin{equation}
         \mathbb{E}^{\frac{k}{2}} \left[\lVert \mtx{A} - \widehat{\mtx{A}} \rVert _F \right]
-        \leq 
+        \leq \frac{1 + 2c}{\sqrt{k}} \Trace(\mtx{A})
     \end{equation}
 
-    Using triangle inequality and 
-    Minkowski's integral inequality \cite[Theorem 2.2]{hardy-1952-inequalities}    
+    We use Minkowski's integral inequality \cite[Theorem 2.2]{hardy-1952-inequalities} to get
+    \begin{equation}
+        \mathbb{E}^{\frac{k}{2}}\left[ \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F~\mathrm{d}t \right] 
+        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}^{\frac{k}{2}} \left[\lVert \mtx{A} - \widehat{\mtx{A}} \rVert _F \right] ~\mathrm{d}t% \notag \\
+        %&\leq \mathbb{E}^{\frac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\frac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
+        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t + \mathbb{E}^{\frac{k}{2}} \left[ \int_{a}^{b}  \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\frac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
+        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\frac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\frac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
+        \leq \frac{1 + 2c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t.
+    \end{equation}
+    From Markov's inequality \cite{} follows with probability at least $1 - \gamma^{-\frac{k}{2}}$
+    \begin{equation}
+        \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{1 + 2c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t,
+    \end{equation}
+    from which follows the first statement by replacing $k = \frac{n_{\Omega}}{2}$.
+
+    Setting $\delta = \gamma^{-\frac{k}{2}}$ and choosing $k = \lceil \varepsilon^{-2}\log(\delta^{-1}) \rceil$ we get
     \begin{align}
-        &\mathbb{E}^{\frac{k}{2}}\left[ \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F~\mathrm{d}t \right] \notag \\
-        &\leq \mathbb{E}^{\frac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\frac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
-        &\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t + \mathbb{E}^{\frac{k}{2}} \left[ \int_{a}^{b}  \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\frac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
-        &\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\frac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\frac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
-        &\leq \frac{1 + 2C}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t
+        \gamma \frac{1}{\sqrt{k}}
+        &= \delta^{-\frac{2}{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} \frac{1}{\sqrt{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} && \text{(definition of $\delta$ and choice of $k$)} \notag \\
+        &= e^{\frac{2}{\varepsilon^{-2}} \frac{\varepsilon^{-2}\log(\delta^{-1})}{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} \frac{1}{\sqrt{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} && \text{($\delta = e^{-\log(\delta^{-1})}$)} \notag \\
+        &\leq e^{2\varepsilon^{2}} \frac{\varepsilon}{\sqrt{ \log(\delta^{-1})}} && \text{($\lceil x \rceil \geq x$ if $x \geq 0$)}
     \end{align}
-    With Markov inequality \cite{} with probability at least $1 - \gamma^{-\frac{k}{2}}$
-    \begin{equation}
-        \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{1 + 2C}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t
-    \end{equation}
-    Setting $\delta = \gamma^{-\frac{k}{2}}$
+    which is smaller than $\varepsilon$ if $\delta \leq e^{-e^{4 \varepsilon^2}}$from which follows the second part of the theorem.
 \end{proof}
@@
 \begin{theorem}{Parameter-dependent Nyström trace estimator}{nystrom}
-    Suppose $f(\mtx{A}, t)$ is continuous and non-negative in $t \in [a, b]$ and $r \geq 2, p \geq 4$. Then for all $\gamma \geq 1$ with probability at least $1 - \gamma^{-p}$
+    Suppose $f(\mtx{A}, t)$ is continuous and non-negative in $t \in [a, b]$ for a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ and $n_{\Omega} \geq 6, n_{\Omega} - 4 \geq r \geq 2$ two integers.
+Then for all $\gamma \geq 1$ with probability at least $1 - \gamma^{-(n_{\Omega} - r)}$
     \begin{equation}
-        \int_{a}^{b} \left| \Trace(f(\mtx{A}, t)) - \Nystr{r+p}(f(\mtx{A}, t)) \right| ~\mathrm{d}t
-        \leq \gamma^2 (1 + r) \int_{a}^{b} \sum_{j = r+1}^{n} \sigma_j(f(\mtx{A}, t)) ~\mathrm{d}t
+        \int_{a}^{b} \left| \Trace(f(\mtx{A}, t)) - \Nystr{n_{\Omega}}(f(\mtx{A}, t)) \right| ~\mathrm{d}t
+        \leq \gamma^2 (1 + r) \int_{a}^{b} \sum_{i = r+1}^{n} \sigma_i(f(\mtx{A}, t)) ~\mathrm{d}t.
     \end{equation}
@@
     Since $f$ is non-negative, $f(\mtx{A}, t)$ is positive definite for all $\mtx{A}$ and $t$.
-    Notice that \cite[Lemma 2.1, item 3]{frangella-2023-randomized-nystrom} implies that $f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t)$ is positive semi-definite \textcolor{red}{also for all $t$?}
+    Notice that by \cite[Lemma 2.1]{frangella-2023-randomized-nystrom}, $f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t)$ is also positive semi-definite for all $t$. Therefore,
     \begin{equation}
         \left| \Trace(f(\mtx{A}, t)) - \Trace(\widehat{f}(\mtx{A}, t)) \right|
+        = \left| \Trace(f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t)) \right|
         = \lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast}
     \end{equation}
-
-    From \cite[Proof of Corollary 8.2]{tropp-2023-randomized-algorithms} it follows that
+    From the proof of \cite[Corollary 8.2]{tropp-2023-randomized-algorithms} it follows that
     \begin{equation}
@@
     \end{equation}
-    Hence, by \cite[Theorem 9]{kressner-2023-randomized-lowrank} the result follows
-    with $\lVert \cdot \rVert _2 \leq \lVert \cdot \rVert _F$ and $\sigma_j(f^{1/2}(\mtx{A}, t))^{2} = \sigma_j(f(\mtx{A}, t))$.
+    Hence, by \cite[Theorem 9]{kressner-2023-randomized-lowrank}, with $n_{\Omega} = p + r$, the result follows
+    with $\lVert \cdot \rVert _2 \leq \lVert \cdot \rVert _F$ and $\sigma_i(f^{1/2}(\mtx{A}, t))^{2} = \sigma_i(f(\mtx{A}, t))$.
 \end{proof}
 
-Discuss rank of kernel matrix function and application to theorem.
-
-\subsection{Parameter-dependent Nystr\"om++ estimator}
+\subsection{Parameter-dependent Nyström++ estimator}
 \label{subsec:nystrom-pp}
 
-\begin{theorem}{Parameter-dependent Nystr\"om++ trace estimator}{nystrom-pp}
-    If $n_{\Psi} = n_{\Omega} = \mathcal{O}(\varepsilon^{-1} \log(\delta^{-1}))$, with probability at least $1 - \delta$
+Finally, we combine the analysis of the two estimators from \refsec{subsec:quadratic} and \refsec{subsec:nystrom} into a more efficient, general-purpose trace estimator.
+
+\begin{theorem}{Parameter-dependent Nyström++ trace estimator}{nystrom-pp}
+    Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
+If $n_{\Psi} = n_{\Omega} = \mathcal{O}(\varepsilon^{-1} \log(2\delta^{-1})^2)$, with probability at least $1 - \delta$ for $\delta \in (0, 2^{-1} \min\{e^{\frac{1}{2}}, e^{-e^{4\varepsilon}}\})$ we have
     \begin{equation}
         \int_{a}^{b} | \Trace(\mtx{A}(t)) - \Nystrpp{n_{\Omega}}{n_{\Psi}}(\mtx{A}(t)) |~\mathrm{d}t
-        \leq \varepsilon \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t
+        \leq \varepsilon \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t.
     \end{equation}
@@
 \begin{proof}
-    Combining with union bound.
+    By choosing $n_{\Psi} = n_{\Omega} = \mathcal{O}(\tilde{\varepsilon}^{-2} \log(\tilde{\delta}^{-1})^2)$ we get
     \begin{align}
-        \int_{a}^{b} | \Trace(\mtx{A}(t)) - \Nystrpp{n_{\Omega}}{n_{\Psi}}(\mtx{A}(t)) |~\mathrm{d}t
-        &= \int_{a}^{b} | \Hutch{n_{\Psi}}(\mtx{A}(t) - \widehat{\mtx{A}}(t)) - \Trace(\mtx{A}(t) - \widehat{\mtx{A}}(t)) |~\mathrm{d}t \notag \\
-        &= \todo{c} \frac{1}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F ~\mathrm{d}t \notag \\
-        &= \todo{c} \frac{1}{\sqrt{n_{\Psi} n_{\Omega}}} \int_{a}^{b} \Trace(\mtx{A}(t)) ~\mathrm{d}t
+        &\int_{a}^{b} | \Trace(\mtx{A}(t)) - \Nystrpp{n_{\Omega}}{n_{\Psi}}(\mtx{A}(t)) |~\mathrm{d}t \notag \\
+        &= \int_{a}^{b} | \Hutch{n_{\Psi}}(\mtx{A}(t) - \widehat{\mtx{A}}(t)) - \Trace(\mtx{A}(t) - \widehat{\mtx{A}}(t)) |~\mathrm{d}t && \text{(by definition of estimators)} \notag \\
+        &= \tilde{\varepsilon} \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F ~\mathrm{d}t && \text{(\refthm{thm:quadratic} w.p. $\geq 1 - \tilde{\delta}$)} \notag \\
+        &= \tilde{\varepsilon}^2 \int_{a}^{b} \Trace(\mtx{A}(t)) ~\mathrm{d}t && \text{(\reflem{lem:nystrom} w.p. $\geq 1 - \widetilde{\delta}$)} 
     \end{align}
+    with probability at least $1 - 2\tilde{\delta}$. Taking $\varepsilon = \tilde{\varepsilon}^2$ and $\delta = 2 \tilde{\delta}$
 \end{proof}
+
+% \begin{proof}
+%     Combining with union bound.
+%     \begin{align}
+%         &\int_{a}^{b} | \Trace(\mtx{A}(t)) - \Nystrpp{n_{\Omega}}{n_{\Psi}}(\mtx{A}(t)) |~\mathrm{d}t \notag \\
+%         &= \int_{a}^{b} | \Hutch{n_{\Psi}}(\mtx{A}(t) - \widehat{\mtx{A}}(t)) - \Trace(\mtx{A}(t) - \widehat{\mtx{A}}(t)) |~\mathrm{d}t && \text{(definition of estimators)} \notag \\
+%         &= \todo{c} \frac{1}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F ~\mathrm{d}t && \text{(with probability $1 - \delta$ if $n_{\Psi} = \mathcal{O}(\varepsilon^{-1})$)} \notag \\
+%         &= \todo{c} \frac{1}{\sqrt{n_{\Psi} n_{\Omega}}} \int_{a}^{b} \Trace(\mtx{A}(t)) ~\mathrm{d}t
+%     \end{align}
+% \end{proof}
 
@@
 
-Spectral density definition
+We consider the task of computing the spectral density of a real symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ with eigenvalues $\lambda_1, \dots, \lambda_n$, which is defined as
 \begin{equation}
-    \phi(t) = \frac{1}{n} \sum_{i=1}^{n} \delta(t - \lambda_i)
+    \phi(t) = \frac{1}{n} \sum_{i=1}^{n} \delta(t - \lambda_i),
     \label{equ:spectral-density}
 \end{equation}
+where $\delta$ is the Dirac delta distribution.
 
@@
 
-Smooth spectral density definition
+\todo{Motivate reason for smoothing?} Therefore, we seek approximations to the smooth spectral density $\phi_{\sigma}$ defined as
 \begin{equation}
@@
 \end{equation}
-with Gaussian smoothing kernel
+with a Gaussian smoothing kernel of width $\sigma > 0$
 \begin{equation}
-    g_{\sigma}(s) = \frac{1}{n \sigma \sqrt{2\pi}} e^{-\frac{s^2}{2\sigma^2}}
+    g_{\sigma}(s) = \frac{1}{n \sigma \sqrt{2\pi}} e^{-\frac{s^2}{2\sigma^2}}.
     \label{equ:smoothing-kernel}
 \end{equation}
+This conversion of the task to a trace estimation problem for the parameter-dependent matrix $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ allows us to apply the results for the estimators derived in \refsec{sec:analysis}.
 
-Can apply trace estimators from \refsec{sec:analysis} to $g_{\sigma}(t\mtx{I}_n - \mtx{A})$
-
-Discuss error measures and baseline smoothing error
+\todo{Discuss error measures and baseline smoothing error, choice of $\sigma$}
 
@@
 
-Affine approximation for evaluating matrix function is favorable
+To estimate the trace in \refequ{equ:smooth-spectral-density}, we first need to evaluate the matrix function.
+However, doing this exactly usually requires us to first diagonalize $\mtx{A}$, which is often a prohibitively expensive operation.
+For algorithmic reasons, we seek for approximations of the matrix function in terms of affine expansions, of which a Chebyshev expansion is most suitable due to its approximation error guarantees and ease of manipulation \todo{more specific}.
+In every $t \in \mathbb{R}$ we expand
 \begin{equation}
@@
 \end{equation}
-Then
+in terms of a set of $m+1$ coefficients $\mu_0(t), \dots, \mu_m(t)$ which linearly combine the Chebyshev polynomials
 \begin{equation}
-    \phi_{\sigma}^{(m)}(t) = \Trace(g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}))
+    \begin{cases}
+        T_l : [-1, 1] \to \mathbb{R} \\
+        T_l(s) = \cos(l \cdot \arccos(s))
+    \end{cases}
+    \label{equ:chebyshev-polynomial}
+\end{equation}
+evaluated at the matrix $\mtx{A}$. It is easy to verify that these polynomials satisfy the three-term recurrence relation
+\begin{equation}
+    T_l(s) =
+    \begin{cases}
+        s^l, & \text{if $l \in \{0, 1\}$} \\
+        T_l(s) = 2 s T_{l-1}(s) - T_{l-2}(s), & \text{if $l \geq 2$.}
+    \end{cases}
+    \label{equ:chebyshev-recurrence}
+\end{equation}
+\todo{Spectral transform (mention estimation of $[a,b]$ see notes, and maybe some buffer region to avoid cut-off).}
+
+This allows us to define the expanded spectral density
+\begin{equation}
+    \phi_{\sigma}^{(m)}(t) = \Trace(g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})).
     \label{equ:expanded-spectral-density}
@@
 
-Spectral transform (mention estimation of $[a,b]$ see notes, and maybe some buffer region to avoid cut-off)
+\subsubsection{Fast and consistent squaring}
+\label{subsubsec:dct}
 
-\subsubsection{Chebyshev expansion error}
+Fast computation of expansion with DCT (duality with evaluations)
+
+Squaring of Chebyshev expansion for consistent and affine Nyström approximation
+\begin{equation}
+    \left( \sum_{l=0}^{m} \mu_l T_l(t) \right)^2 = \sum_{l=0}^{2m} \nu_l T_l(t)
+    \label{equ:squared-chebyshev-expansion}
+\end{equation}
+where $\nu = \DCT^{-1} \left\{ \DCT\{\mu\}^2 \right\}$ \todo{which is consistent unlike \cite{lin-2017-randomized-estimation}}.
+
+Non-negative Chebyshev expansion for results in Nyström approximation of positive semi-definite matrices (discuss alternatives: Jackson damping, Putinar representation)
+
+\subsubsection{Expansion error}
 \label{subsubsec:expansion-error}
+
+We analyze the error we commit when working with the expanded spectral density \refequ{equ:expanded-spectral-density}. To this extent, we first analyze the expansion error of the smoothing kernel.
 
@@
     \end{equation}
-    for all degrees $m > 0$, smoothing parameters $\sigma > 0$, and $s \in [-1, 1]$.
+    for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and $s \in [-1, 1]$.
 \end{lemma}
@@
     \end{equation}
-    Because $e^{-\frac{(t - x)^2}{2\sigma^2}} \leq 1$ for all $x \in \mathbb{R}$ and $t \in [-1,1]$, and the maximum absolute value of $y$ is limited by the length of the semi-axis of the Bernstein ellipse in the direction of the imaginary axis, which is $(\chi - \chi^{-1}) / 2$, we upper bound
+    Because $e^{-\frac{(t - x)^2}{2\sigma^2}} \leq 1$ for all $x, t \in \mathbb{R}$, and the maximum absolute value of $y$ is limited by the length of the semi-axis of the Bernstein ellipse in the direction of the imaginary axis, which is $(\chi - \chi^{-1}) / 2$, we upper bound
     \begin{equation}
@@
 
-\subsubsection{Discrete cosine transform}
-\label{subsubsec:dct}
-
-Fast computation of expansion with DCT (duality with evaluations)
-
-Squaring of Chebyshev expansion for consistent and affine Nystr\"om approximation
-\begin{equation}
-    \left( \sum_{l=0}^{m} \mu_l T_l(t) \right)^2 = \sum_{l=0}^{2m} \nu_l T_l(t)
-    \label{equ:squared-chebyshev-expansion}
-\end{equation}
-where $\nu = \DCT^{-1} \left\{ \DCT\{\mu\}^2 \right\}$ \todo{which is consistent unlike \cite{lin-2017-randomized-estimation}}.
-
-Non-negative Chebyshev expansion for results in Nystr\"om approximation of positive semi-definite matrices (discuss alternatives: Jackson damping, Putinar representation)
+\todo{Explain need for non-negativity for Nyström approximation, and mention alternatives (Putinar, Jackson-damping, ...)}
 
@@
     \end{equation}
-    Since $\sqrt{g_{\sigma}} = \sqrt{2 n \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$ and $\sqrt{g_{\sigma}} \leq 1/\sqrt{n \sigma \sqrt{2 \pi}}$ we can apply the result from \reflem{lem:chebyshev-error} to get
+    Since $\sqrt{g_{\sigma}} = \sqrt{2 n \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$ and $\sqrt{g_{\sigma}} \leq 1/\sqrt{n \sigma \sqrt{2 \pi}}$ we can apply the result from \refthm{thm:chebyshev-error} to get
     \begin{align}
@@
 
-\subsection{The Chebyshev-Nystr\"om++ method for approximating spectral densities}
+The
+
+\begin{theorem}{$L^1$-error of Chebyshev expansion of spectral density}{chebyshev-error}
+    Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
+Then the spectal density with Chebyshev expansion $\phi_{\sigma}^{(m)}$ satisfies
+    \begin{equation}
+        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq 2 E_{\sigma, m}
+        \label{equ:chebyshev-interpolation-sup-error-kernel}
+        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
+        %\label{equ:2-chebyshev-interpolation-error}
+    \end{equation}
+    for all degrees $m \in \mathbb{N}$ and smoothing parameters $\sigma > 0$.
+The same result holds for the expansion of a non-negative Chebyshev expansion of \emph{even} degree $m \in \mathbb{N}$ but with $\underline{E}_{\sigma, m}$.
+\end{theorem}
+
+\begin{proof}
+    We observe that for all $t \in [-1, 1]$
+    \begin{align}
+        &\left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right| \notag \\
+        &= \left| \Trace(g_{\sigma}(t\mtx{I}_n - \mtx{A})) - \Trace(g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})) \right|
+        && \text{(definitions \refequ{equ:smooth-spectral-density} and \refequ{equ:expanded-spectral-density})} \notag \\
+        &= \left| \sum_{i=1}^n \left(g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i)\right) \right|
+        && \text{($\lambda_1, \dots, \lambda_n$ eigenvalues of $\mtx{A}$)} \notag \\
+        &\leq n \max_{i = 1, \dots, n} \left| g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i) \right|
+        && \text{(conservative upper bound)} \notag \\
+        &\leq E_{\sigma, m} ~ \text{(or $\underline{E}_{\sigma, m}$)}.
+        && \text{(using \refthm{thm:chebyshev-error})} \notag \\
+    \end{align}
+    Then, Hölder's inequality \cite{} allows us to conclude%\cite{klenke2013probability} 
+    \begin{equation}
+        \int_{-1}^{1} | \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) | ~\mathrm{d}t
+            \leq 2 \sup_{t \in [-1, 1]} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|
+            \leq 2 E_{\sigma, m} ~ \text{(or $2 \underline{E}_{\sigma, m}$)}.
+    \end{equation}
+\end{proof}
+
+\subsection{Chebyshev-Nyström++ method for spectral density approximation}
 \label{subsec:chebyshev-nystrom}
@@
 \label{subsubsec:chebyshev-nystrom-implementation}
+
+An efficient implementation can be achieved thanks to the recurrence relation \refequ{equ:chebyshev-recurrence} which the Chebyshev polynomials satisfy.
+However, it is usually prohibitively expensive to interpolate a large matrix $\mtx{A}$ as a whole, since alone the matrix-matrix multiplication in each step of the recurrence can cost up to $\mathcal{O}(n^3)$, and the evaluation of the expansion at $n_t$ values of the parameter could cost a further $\mathcal{O}(n^2 m n_t)$ operations.
+Therefore, in case we are only interested in the result of a linear mapping applied to the interpolant, a significant speed-up can be achieved by directly interpolating the result of this linear mapping applied to the interpolant.
+This is done by using the invariance of the trace under cyclic permutation of its arguments.
+\todo{Illustrate this better.}
+
+\todo{Explain idea behind algorithm: Two-phase, first assemble matrices with recurrence then compute approximate in every time step}
 
@@
 
+With the cost of a matrix-vector product denoted by $c(n)$, and supposing $n_{\Omega} \approx n_{\Psi}$, we determine the computational complexity of the Chebyshev-Nyström++ method to be $\mathcal{O}(m \log(m) n_t + m n_{\Omega}^2 n + m n_t n_{\Omega}^2 +  m c(n) n_{\Omega} + n_t n_{\Omega}^3)$, with $\mathcal{O}(m n_t + n n_{\Omega} + n_{\Omega}^2 n_t)$ required additional storage.
+
+\todo{Explain modifications for other estimators, and improved complexity for quadratic trace estimator}
+
+\todo{Explain short-circuit mechanism -> may not be too crucial for directly calculating pseudo-inverse}
+
+\todo{Discuss Lin Lin generalized eigenvalue problem for pseudo-inverse, other ways are not available, since we only have access to the \enquote{compressed} matrix function}
+
 \subsubsection{Analysis}
@@
 
-\begin{lemma}{$L^1$-error of Chebyshev expansion of spectral density}{chebyshev-error}
+\begin{theorem}{Chebyshev-Nyström++ method}{chebyshev-nystrom}
     Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the spectal density with Chebyshev expansion $\phi_{\sigma}^{(m)}$ satisfies
+Then the Chebyshev-Nyström++ estimate $\underline{\widetilde{\phi}}_{\sigma}^{(m)}$ with \emph{even} degree $m \in \mathbb{N}$, $n_{\Omega} = n_{\Psi} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2)$, and smoothing parameter $\sigma > 0$ satisfies with probability at least $1 - \delta$
     \begin{equation}
-        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq 2 E_{\sigma, m}
-        \label{equ:chebyshev-interpolation-sup-error-kernel}
+        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\widetilde{\phi}}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t < 2 (1 + \varepsilon) \underline{E}_{\sigma, m} + \varepsilon
+        \label{equ:chebyshev-nystrom-error}
         %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
@@
     \end{equation}
-    for all degrees $m \in \mathbb{N}$ and smoothing parameters $\sigma > 0$.
-The same result holds for the expansion of a non-negative Chebyshev expansion of \emph{even} degree $m \in \mathbb{N}$ but with $\underline{E}_{\sigma, m}$.
-\end{lemma}
+    for all $\delta \in (0, 2^{-1} \min\{e^{\frac{1}{2}}, e^{-e^{8\varepsilon}}\})$.
+\end{theorem}
 
 \begin{proof}
-    We observe that for all $t \in [-1, 1]$
+    The previously derived results can be applied to
     \begin{align}
-        &\left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right| \notag \\
-        &= \left| \Trace(g_{\sigma}(t\mtx{I}_n - \mtx{A})) - \Trace(g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})) \right|
-        && \text{(definitions \refequ{equ:smooth-spectral-density} and \refequ{equ:expanded-spectral-density})} \notag \\
-        &= \left| \sum_{i=1}^n \left(g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i)\right) \right|
-        && \text{($\lambda_1, \dots, \lambda_n$ eigenvalues of $\mtx{A}$)} \notag \\
-        &\leq n \max_{i = 1, \dots, n} \left| g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i) \right|
-        && \text{(conservative upper bound)} \notag \\
-        &\leq E_{\sigma, m}.
-        && \text{(using \reflem{lem:chebyshev-error})} \notag \\
+        &\int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\widetilde{\phi}}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \notag \\
+        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \int_{-1}^{1} \left| \underline{\phi}^{(m)}_{\sigma}(t) - \underline{\widetilde{\phi}}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
+        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \varepsilon \int_{-1}^{1} \left| \underline{\phi}^{(m)}_{\sigma}(t) \right|~\mathrm{d}t && \text{(\refthm{thm:nystrom-pp} w.p.
+$\geq 1 - \delta$)} \notag \\
+        &\leq (1 + \varepsilon) \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \varepsilon \int_{-1}^{1} \left| \phi_{\sigma}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
+        &\leq (1 + \varepsilon) \underline{E}_{\sigma, m} + \varepsilon. && \text{(\reflem{lem:non-negative-chebyshev-error} and normalization)} \notag \\
     \end{align}
-    Then, H\"older's inequality \cite{} allows us to conclude%\cite{klenke2013probability} 
+\end{proof}
+
+We now discuss what value to choose for $n_{\Omega}$ in the Nyström trace estimator introduced in \refsec{subsec:nystrom}.
+In order to guarantee a good approximation error based on \refthm{thm:nystrom}, we analyze the numerical rank of the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
+
+We can express the numerical rank of a matrix with respect to the nuclear norm in terms of its singular values $\sigma_1 \geq \dots \geq \sigma_n \geq 0$
 \begin{equation}
-        \int_{-1}^{1} | \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) | ~\mathrm{d}t
-            \leq 2 \sup_{t \in [-1, 1]} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|
-            \leq 2 E_{\sigma, m}.
+    r_{\varepsilon} = \min \{1 \leq r \leq n: \sum_{j=r+1}^n \sigma_{j} \leq \varepsilon \},
+    \label{equ:numerical-rank}
 \end{equation}
-\end{proof}
+$\varepsilon$ is usually taken to be the double machine precision, i.e.
+$10^{-16}$.
+When using a Gaussian smoothing-kernel, we can explicitly write the signular values of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ as
+\begin{equation}
+    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) = \frac{1}{n \sqrt{2 \pi \sigma^2}} e^{-\frac{(t - \lambda_{(i)})^2}{2 \sigma^2}},
+    \label{equ:gaussian-kernel-eigenvalues}
+\end{equation}
+where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from spectral-parameter, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t)$.
+Consequently, by using the closed-form expression of the eigenvalues \refequ{equ:gaussian-kernel-eigenvalues}, we may upper bound the numerical rank of a Gaussian smoothing kernel as
+\begin{equation}
+    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A})) \leq \#\left\{1\leq i\leq n: |t - \lambda_i| < d_{\varepsilon, \sigma} \right\},
+    \label{equ:gaussian-kernel-numerical-rank}
+\end{equation}
+with $d_{\varepsilon, \sigma} = \sigma \sqrt{-2 \log(\sqrt{2 \pi n} \sigma \varepsilon)}$ and the eigenvalues $\lambda_1, \dots, \lambda_n$ of $\mtx{A}$ and the constants.
+The expression \refequ{equ:gaussian-kernel-numerical-rank} has a nice visual interpretation: The numerical rank of a matrix is at most equal to the number of eigenvalues which are closer to the parameter $t$ than $d_{\varepsilon, \sigma}$.
+This is illustrated in \reffig{fig:numerical-rank}.
+\begin{figure}[ht]
+    \centering
+    \input{figures/numerical-rank.tex}
+    \caption{The numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ can be
+        approximately computed by counting the number of eigenvalues
+        $\lambda_{(1)}, \dots, \lambda_{(n)}$ of the matrix $\mtx{A}$ which lie less than
+        a constant $d_{\varepsilon, \sigma}$ away from $t$.}
+    \label{fig:numerical-rank}
+\end{figure}
+
+If we additionally assume the eigenvalues of the matrix $\mtx{A}$ to be evenly distributed within $[a, b]$, that is, in any subinterval of fixed length in $[a, b]$ we can expect to find roughly the same number of eigenvalues, then we can estimate the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ to be
+\begin{equation}
+    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A})) \lessapprox \frac{2 n}{b - a} \cdot d_{\varepsilon, \sigma}.
+    \label{equ:gaussian-kernel-numerical-rank-uniform}
+\end{equation}
+
+\todo{Discuss how this informs choice of $n_{\Omega}$.}
 
@@
 \begin{equation}
-    \alpha \exp(-\frac{\lVert \vct{x} \rVert _2^2}{ 2 \beta^2 })
+    \alpha e^{-\frac{\lVert \vct{x} \rVert _2^2}{ 2 \beta^2 }}
     \label{equ:5-experiments-gaussian-cell}
@@
 we consider the three-dimensional case, but for visualization purposes, we
-illustrate the potential in FIGURE
+illustrate the potential in \reffig{fig:gaussian-well}
 in two dimensions.
 
-\todo{[Plot with $1/\epsilon$ and $1/\epsilon^2$ result]}
+\begin{figure}[ht]
+    \begin{subfigure}[b]{0.32\columnwidth}
+        \input{plots/gaussian-well-1.pgf}
+        \caption{$n_c=1$}
+        \label{fig:gaussian-well-1}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.32\columnwidth}
+        \input{plots/gaussian-well-2.pgf}
+        \caption{$n_c=2$}
+        \label{fig:gaussian-well-2}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.32\columnwidth}
+        \input{plots/gaussian-well-5.pgf}
+        \caption{$n_c=5$}
+        \label{fig:gaussian-well-5}
+    \end{subfigure}
+    \caption{Two dimensional periodic potential $V$ for different sizes $n_c$ of the computational domain.}
+    \label{fig:gaussian-well}
+\end{figure}
 
-\todo{[Plot with dependence on $\sigma$]}
+\begin{figure}[ht]
+    \begin{subfigure}[b]{0.49\columnwidth}
+        \input{plots/convergence-degree.pgf}
+        \caption{$n_{\Omega} + n_{\Psi} = 160$}
+        \label{convergence-randvec}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.49\columnwidth}
+        \input{plots/convergence-randvec.pgf}
+        \caption{$m = 2400$}
+        \label{fig:convergence-degree}
+    \end{subfigure}
+    \caption{For increasing values of s$n_{\Omega} + n_{\Psi}$ but fixed $m$ we plot the $L^1$ relative approximation error for the model problem with $\sigma=0.05$.}
+    \label{fig:convergence}
+\end{figure}
 
-\todo{[Timing comparison]}
+\begin{figure}[ht]
+    \centering
+    \input{plots/matvec-mix.pgf}
+    \caption{The NC++ method for different ways of allocations a 
+    total of $n_{\Omega} + n_{\Psi}=80$ random vectors
+    to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation
+    for the Gaussian smoothing kernel with multiple different values of
+    the smoothing parameter. We make the approximation error made in the
+    Chebyshev expansion negligible by rescaling $m=120 / \sigma$.}
+    \label{fig:matvec-mix}
+\end{figure}
+
+\todo{[Comparison plots with other estimators]}
 
@@
 
+\begin{itemize}
+    \item Test
+\end{itemize}
+
 \clearpage
@@
 
-\appendix
+%\appendix
 
-\clearpage
-\section{Proofs}
+%\clearpage
+%\section{Proofs}
+
+\tableofcontents
 

### 2024-05-23 paper/main.tex
start writing introduction

@@
 %    \item Write introduction, abstract, and discussion
+%    \item ArXiv HTML support (check if packages)
 %\end{itemize}
 %}
+
+\todo{
+    \tableofcontents
+}
 
@@
 
+In numerous problems across physics, chemistry, engineering, and data science, eigenvalues and especially their distribution are crucial for understanding the properties of a system: In electronic structure calculations they represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}.
+The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be captured with the spectral density
+\begin{equation}
+    \phi(t) = \frac{1}{n} \sum_{i=1}^{n} \delta(t - \lambda_i).
+\end{equation}
+Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation which is often prohibitively expensive.
+Therefore, many techniques have been proposed for approximating $\phi$.
+For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature from the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}.
+In many applications, it is also acceptable to trade in some of the finer details of the spectrum for a better approximability by smoothing $\phi$, typically through convolution with a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}.
+This converts the problem to estimating the trace of a smooth parameter-dependent matrix function.
+
+\todo{
 \begin{itemize}
     \item Definition of parameter-dependent matrix ($\mtx{B}(t) \in \mathbb{R}^{n \times n}$, every entry depends continuously on $t \in [a,b]$ \cite{kressner-2023-randomized-lowrank})
-    \item Trace estimation through mat-vecs (if matrix is only accessed implicitly [Hessian, Graph] or for matrix functions [partition function, spectral density, exponential integrators])
-    \item Mention existing trace estimation methods for constant matrices (quadratic [Hutchinson, Girard], Hutch++ \cite{meyer-2021-hutch-optimal}, Nyström++ \cite{persson-2022-improved-variants}, XTrace [Tropp], Krylov aware [Chen]...)
-\end{itemize}
-
-\paragraph{Spectral density approximation}
-\begin{itemize}
-    \item Definition of spectral density ($\phi(t) = n^{-1}\sum_{i=1}^{m} \delta(t - \lambda_i)$ for matrix $\mtx{A}$ with eigenvalues $\lambda_1, \dots, \lambda_n$)
-    \item Applications of spectral density approximation (electronic structure calculations \cite{lin-2017-randomized-estimation}, neural network optimization [Hessian, ...], graph processing [filtering, ...])
-    \item Link to trace estimation of a matrix function (smoothing with Gaussian \cite{lin-2017-randomized-estimation} or Haydock [Haydock, Lin] kernel)
-    \item Existing theory for spectral density approximation (Wasserstein-1 error [Braverman, Chen], missing $1/\varepsilon$ guarantees)
+    \item Trace estimation through mat-vecs (if matrix is only accessed implicitly \cite{pearlmutter-1994-fast-exact} or for complicated matrix functions [partition function, spectral density, exponential integrators])
+    \item Existing trace estimation methods for constant matrices: Girard-Hutchinson \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making} which have only recently been specialized to trace estimation of matrix functions \cite{chen-2023-krylovaware-stochastic}
+    \item Trace estimation for parameter-dependent matrices (Nyström++ but no analysis \cite{lin-2017-randomized-estimation}, low-rank approximation \cite{kressner-2023-randomized-lowrank}, dynamic trace estimation but only discrete setting [Woodruff, Dharangutte])
+    %\item Error analysis for spectral density approximation (Wasserstein-1 error [Braverman, Chen], missing $1/\varepsilon$ guarantees)
 \end{itemize}
@@
 \begin{itemize}
-    \item Development of a unified family of estimators for many evaluations of parameter, i.e. try to keep complexity with each evaluation as low as possible
-    \item Extends \cite{kressner-2023-randomized-lowrank} to trace estimation
-    \item Provides theoretical guarantees for methods in \cite{lin-2017-randomized-estimation}
+    \item Theoretical guarantees in the $L^1$-norm for three trace estimators on parameter-dependent matrices (Girard-Hutchinson, Nyström, Nyström++), in particular, $\mathcal{O}(\varepsilon^{-1})$ scaling with number of mat-vecs
+    \item A simple and efficient algorithm for estimating spectral densities using any of the three estimators using Chebyshev expansion, including theoretical guarantees
 \end{itemize}
+}
 
@@
     \item the expected value of a random variable $x$ is denoted with $\mathbb{E}[x]$.
-The $k$-th moment of a random variable $x$ is defined as $\mathbb{E}^{k}[x] = \left(\mathbb{E}\left[ | x |^{k} \right] \right)^{\frac{1}{k}}$;
+The $k$-th moment of a random variable $x$ is defined as $\mathbb{E}^{k}[x] = \left(\mathbb{E}\left[ | x |^{k} \right] \right)^{\sfrac{1}{k}}$;
     \item the identity matrix $\mtx{I}_n = \diag(1, \dots, 1) \in \mathbb{R}^{n \times n}$
@@
     \item in a Gaussian random vector and a Gaussian random matrix, all entries are independently sampled from the standard normal distribution;
-    \item matrix norms are denoted with $\lVert \cdot \rVert _{\cdot}$. For symmetric matrices, the $p$-Schatten norm is
+    \item matrix norms are denoted with $\lVert \cdot \rVert$. For symmetric matrices, the $p$-Schatten norm is
     $\lVert \mtx{A} \rVert _{(p)} = (\sum_{i=1}^n |\lambda_i|^p)^{1/p}$.
-We refer to the spectral norm as $\lVert \mtx{A} \rVert _2 = \lVert \mtx{A} \rVert _{(\infty)} = \max_{i=1,\dots,n}|\lambda_i|$, the nuclear norm as $\lVert \mtx{A} \rVert _{\ast} = \lVert \mtx{A} \rVert _{(1)} = \sum_{i=1}^{n} |\lambda_i|$, and the Frobenius norm as $\lVert \mtx{A} \rVert _F = \lVert \mtx{A} \rVert _{(2)} = (\sum_{i=1}^{n} |\lambda_i|^2)^{\frac{1}{2}}$.
+We refer to the spectral norm as $\lVert \mtx{A} \rVert _2 = \lVert \mtx{A} \rVert _{(\infty)} = \max_{i=1,\dots,n}|\lambda_i|$, the nuclear norm as $\lVert \mtx{A} \rVert _{\ast} = \lVert \mtx{A} \rVert _{(1)} = \sum_{i=1}^{n} |\lambda_i|$, and the Frobenius norm as $\lVert \mtx{A} \rVert _F = \lVert \mtx{A} \rVert _{(2)} = (\sum_{i=1}^{n} |\lambda_i|^2)^{\sfrac{1}{2}}$.
 \end{itemize}
 
-\section{Analysis of parameter-dependent trace estimators}
+\todo{
+\paragraph{Reproducibility} Fully reproducible paper hosted on GitHub.
+}
+\section{Analysis of trace estimators for parameter-dependent matrices}
 \label{sec:analysis}
@@
 
-\paragraph{Quadratic trace estimator} If we can only access the parameter-dependent matrix through matrix-vector products, we can estimate the trace with the quadratic trace estimator \cite{}: We take $n_{\Psi} \in \mathbb{N}$ Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\psi}} \in \mathbb{R}^{n}$ to form
+\paragraph{Girard-Hutchinson estimator} If we can only access the parameter-dependent matrix through matrix-vector products, we can estimate the trace with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ to form
 \begin{equation}
-    \Hutch{n_{\Psi}}(\mtx{A}(t)) = \frac{1}{n_{\Psi}} \sum_{j=1}^{n_{\Psi}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j.
-    \label{equ:quadratic-trace-estimator}
+    \Hutch{\mtx{\Psi}}(\mtx{A}(t))
+    = \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j
+    = \frac{1}{n_{\mtx{\Psi}}} \Trace( \mtx{\Psi}^{\top} \mtx{A}(t) \mtx{\Psi})
+    \label{equ:hutchinson-trace-estimator}
 \end{equation}
-If we gather the random vectors in a matrix $\mtx{\Psi} = [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\psi}}] \in \mathbb{R}^{n \times n_{\Psi}}$ we may rewrite $\Hutch{n_{\Psi}}(\mtx{A}(t)) = n_{\Psi}^{-1} \Trace( \mtx{\Psi}^{\top} \mtx{A}(t) \mtx{\Psi})$.
-Other choices for the distribution of the entries of the random vectors are possible \cite{}.
+where $\mtx{\Psi} = [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\mtx{\Psi}}}] \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$.
+Other choices for the distribution of the entries of the random vectors are possible, for example by uniformly sampling from $\{-1, +1\}$ or from the $(n-1)$-sphere.
 However, our theoretical developments only hold in the Gaussian case.
 
-\paragraph{Nyström trace approximation} Alternatively, the trace of a symmetric matrix whose rank is significantly lower than its size can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\Omega}}$ to form the Nyström approximation
+\paragraph{Nyström approximation} Alternatively, the trace of a symmetric matrix whose rank is significantly lower than its size can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström approximation \cite{gittens-2013-revisiting-nystrom}
 \begin{equation}
-    \widehat{\mtx{A}}(t) = (\mtx{A}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} (\mtx{A}(t) \mtx{\Omega})^{\top}.
+    \Nystr{\mtx{\Omega}}{\mtx{A}}(t) = (\mtx{A}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} (\mtx{A}(t) \mtx{\Omega})^{\top}.
 \end{equation}
-Then we define the trace estimator $\Nystr{n_{\Omega}}(\mtx{A}(t)) = \Trace( \widehat{\mtx{A}}(t) )$.
+Then we use the trace estimator $\Trace(\Nystr{\mtx{\Omega}}{\mtx{A}}(t))$.
 Thanks to the invariance of the trace under cyclic permutation of its arguments and the symmetry of the matrix, we may also rewrite
 \begin{equation}
-    \Nystr{n_{\Omega}}(\mtx{A}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{A}(t)^2 \mtx{\Omega})).
+    \Trace(\Nystr{\mtx{\Omega}}{\mtx{A}}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{A}(t)^2 \mtx{\Omega})).
     \label{equ:nystrom-trace-estimator}
@@
 \begin{equation}
-    \Nystrpp{n_{\Omega}}{n_{\Psi}}(\mtx{A}(t)) = \Nystr{n_{\Omega}}(\mtx{A}(t)) + \Hutch{n_{\Psi}}(\mtx{A}(t) - \widehat{\mtx{A}}(t)).
+    \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{A}(t)) = \Trace(\Nystr{\mtx{\Omega}}{\mtx{A}}(t)) + \Hutch{\mtx{\Psi}}(\mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)),
     \label{equ:nystrompp-trace-estimator}
 \end{equation}
-It is understood that $\widehat{\mtx{A}}$ is computed with the same sketching matrix as the one used for the trace estimator $\Nystr{n_{\Omega}}$.
-We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the quadratic trace estimator.
+which has originally been proposed in \cite{lin-2017-randomized-estimation}.
+%We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.
 
 \paragraph{$L^{1}$ norm}
-\todo{Give definition of $L^1$ norm and reason for choosing it}
+We define the $L^{1}$ norm of a square-integrable function $f:[a,b] \to \mathbb{R}$ as
+\begin{equation}
+    \int_{a}^{b} | f(t) | \mathrm{d}t.
+\end{equation}
+\todo{Motivate why we use this.}
 
-\subsection{Parameter-dependent quadratic trace estimator}
-\label{subsec:quadratic}
+\subsection{Girard-Hutchinson estimator for parameter-dependent matrices}
+\label{subsec:hutchinson}
 
-We first consider \refequ{equ:quadratic-trace-estimator}.
+We first consider \refequ{equ:hutchinson-trace-estimator}.
 \todo{Discuss theory for constant matrix
-\begin{equation}
-    \lVert f \rVert _1 = \int_{a}^{b} | f(t) | \mathrm{d}t
-\end{equation}
 }
 
-\begin{theorem}{Parameter-dependent quadratic trace estimator}{quadratic}
+\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
     Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ continuously depend on $t \in [a, b]$. Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ it holds with probability at least $1 - \gamma^{-2k}$
     \begin{equation}
-        \int_{a}^{b} \left| \Trace(\mtx{A}(t)) - \Hutch{n_{\Psi}}(\mtx{A}(t)) \right| ~\mathrm{d}t < 16 \gamma k \frac{1}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F  ~\mathrm{d}t.
+        \int_{a}^{b} \left| \Trace(\mtx{A}(t)) - \Hutch{\mtx{\Psi}}(\mtx{A}(t)) \right| ~\mathrm{d}t < 16 \gamma k \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F  ~\mathrm{d}t.
     \end{equation}
-    In particular, if we choose $n_{\Psi} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{\frac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{A}(t)) - \Hutch{n_{\Psi}}(\mtx{A}(t)) | ~\mathrm{d}t < \varepsilon \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t$.
+    In particular, if we choose $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{\sfrac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{A}(t)) - \Hutch{\mtx{\Psi}}(\mtx{A}(t)) | ~\mathrm{d}t < \varepsilon \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t$.
 \end{theorem}
@@
     \end{equation}
-    for a parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$ in a fixed $t \in [a,b]$.
-    From the proof of \cite[Lemma 3]{cortinovis-2022-randomized-trace} we know that $r(t)$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B}(t) \rVert _F^2, 2 \lVert \mtx{B}(t) \rVert _2)$.
+    for a parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$.
+
+    First, we consider $r(t)$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence.
+From the proof of \cite[Lemma 3]{cortinovis-2022-randomized-trace} we know that $r$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$.
 Thus, by \cite[Theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
     \begin{align}
-        \mathbb{E}\left[ r(t)^{2 k} \right]
-        &\stackrel{\text{\cite{boucheron-2013-basic-inequalities}}}{\leq} k! \left( 16 \lVert \mtx{B}(t) \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B}(t) \rVert _2 \right)^{2 k} \notag \\
-        &= k! 2^{4 k} \lVert \mtx{B}(t) \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B}(t) \rVert _2^{2 k}.
+        \mathbb{E}\left[ r^{2 k} \right]
+        &\stackrel{\text{\cite{boucheron-2013-basic-inequalities}}}{\leq} k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k} \notag \\
+        &= k! 2^{4 k} \lVert \mtx{B} \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _2^{2 k}.
     \end{align}
-    Since $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$ we can upper bound 
+    Since $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$ we can upper bound 
     \begin{equation}
-        \mathbb{E}\left[ r(t)^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B}(t) \rVert _F^{2k}.
+        \mathbb{E}\left[ r^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
     \end{equation}
     Stirling's approximation \cite{robbins-1955-remark-stirling} bonuds $(2 k)!
-< 2 \sqrt{\pi k}  e^{\frac{1}{24 k}} ( 2 k / e )^{2 k}$.
-Consequently, the moments of $r(t)$ for a fixed $t$ are limited by
+< 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$.
+Consequently, the moments of $r$ for a fixed $t$ are limited by
     \begin{equation}
-        \mathbb{E}^{2k}\left[ r(t) \right]
-        < \left( \frac{9}{4} \sqrt{\pi k} e^{\frac{1}{24 k}} \right)^{\frac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B}(t) \rVert _F < 16 k \lVert \mtx{B}(t) \rVert _F,
-        \label{equ:quadratic-trace-onequery-fixed}
+        \mathbb{E}^{2k}\left[ r \right]
+        < \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F < 16 k \lVert \mtx{B} \rVert _F,
+        \label{equ:hutchinson-trace-onequery-fixed}
     \end{equation}
-    since it can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\frac{1}{24 k}})^{\frac{1}{2k}} < e$, due to the monotonous decrease of this expression in $k \in \mathbb{N}$.
+    since it can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, due to the monotonous decrease of this expression in $k \in \mathbb{N}$.
 
     Now we make the transition to the continuous.
-Minkowski's integral inequality \cite[Theorem 2.2]{hardy-1952-inequalities} allows us to apply \refequ{equ:quadratic-trace-onequery-fixed} in the continuous setting:
+Minkowski's integral inequality \cite[Theorem 2.2]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting:
     \begin{equation}
         \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(t)|~\mathrm{d}t  \right]
-        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}\left[ |r(t)|^{2 k}\right]^{\frac{1}{2 k}}~\mathrm{d}t
+        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}\left[ |r(t)|^{2 k}\right]^{\sfrac{1}{2 k}}~\mathrm{d}t
         < 16 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
     \end{equation}
-    By Markov's inequality \cite{}, for any $k \in \mathbb{N}$ and $\gamma \geq 1$, it holds with probability at least $1 - \gamma^{-2 k}$
+    By Markov's inequality, for any $k \in \mathbb{N}$ and $\gamma \geq 1$, it holds with probability at least $1 - \gamma^{-2 k}$
     \begin{equation}
         \int_{a}^{b} |r(t)|~\mathrm{d}t < 16 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
-        \label{equ:quadratic-trace-onequery-uniform}
+        \label{equ:hutchinson-trace-onequery-uniform}
     \end{equation}
 
-    To extend the bound to the $n_{\Psi}$-query quadratic trace estimator, we use a technique from the proof of \cite[Proof of Theorem 1]{cortinovis-2022-randomized-trace}. Let
+    To extend the bound to the $n_{\mtx{\Psi}}$-query quadratic trace estimator, we use a technique from the proof of \cite[Theorem 1]{cortinovis-2022-randomized-trace}. Let
     \begin{equation}
         \mtx{B}(t)
-        = \frac{1}{n_{\Psi}} \begin{pmatrix}
+        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
             \mtx{A}(t) & & \\
@@
             \vdots \\
-            \vct{\psi}_{n_{\Psi}}
+            \vct{\psi}_{n_{\mtx{\Psi}}}
         \end{pmatrix}
     \end{equation}
-    where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\Psi}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors.
-Then $\lVert \mtx{B}(t) \rVert _F = \frac{1}{\sqrt{n_{\Psi}}} \lVert \mtx{A}(t) \rVert _F$ and
+    where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors.
+Then $\lVert \mtx{B}(t) \rVert _F = \lVert \mtx{A}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ and
     \begin{equation}
-        r(t) = \frac{1}{n_{\Psi}} \sum_{j=1}^{n_{\Psi}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j - \Trace(\mtx{A}(t)).
+        r(t) = \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j - \Trace(\mtx{A}(t)).
     \end{equation}
-    Hence, we conclude from \refequ{equ:quadratic-trace-onequery-uniform} that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
+    Hence, we conclude from \refequ{equ:hutchinson-trace-onequery-uniform} that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
     \begin{equation}
-        \int_{a}^{b} \left| \frac{1}{n_{\Psi}} \sum_{j=1}^{n_{\Psi}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j - \Trace(\mtx{A}(t)) \right| ~ \mathrm{d}t
-        < 16 k \gamma \frac{1}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t.
+        \int_{a}^{b} \left| \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j - \Trace(\mtx{A}(t)) \right| ~ \mathrm{d}t
+        < 16 k \gamma \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t.
     \end{equation}
 
-    Setting $\delta = \gamma^{-2 k}$ and choosing $k = \lceil \log(\delta^{-1}) \rceil$ we get that for all $\delta \in (0, e^{-\frac{1}{2}})$
+    Setting $\delta = \gamma^{-2 k}$ and choosing $k = \lceil \log(\delta^{-1}) \rceil$ we get that for all $\delta \in (0, e^{-\sfrac{1}{2}})$
     \begin{equation}
-        k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\frac{1}{2\lceil \log(\delta^{-1}) \rceil}}
-        = \lceil \log(\delta^{-1}) \rceil e^{\frac{1}{2}\frac{\log(\delta^{-1})}{\lceil \log(\delta^{-1}) \rceil}}
-        \leq 2 \log(\delta^{-1}) e^{\frac{1}{2}},
+        k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\sfrac{1}{2\lceil \log(\delta^{-1}) \rceil}}
+        = \lceil \log(\delta^{-1}) \rceil e^{\sfrac{1}{2}\frac{\log(\delta^{-1})}{\lceil \log(\delta^{-1}) \rceil}}
+        \leq 2 \log(\delta^{-1}) e^{\sfrac{1}{2}},
     \end{equation}
@@
     %\begin{equation}
-    %    \int_{a}^{b} \left| \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Trace(\mtx{A}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\frac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t
+    %    \int_{a}^{b} \left| \frac{1}{n_{\mtx{\Psi}}} \sum_{i=1}^{n_{\mtx{\Psi}}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Trace(\mtx{A}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\sfrac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t
     %\end{equation}
@@
 
-\subsection{Parameter-dependent Nyström approximation}
+\subsection{Nyström approximation for parameter-dependent matrices}
 \label{subsec:nystrom}
@@
 
-    To bound the first term, we define $\widetilde{\mtx{\Omega}}_2 = \mtx{\Omega}_2^{\top} / \sqrt{2k}$. We may use the triangle inequality to rewrite
+    To bound the first term, we define $\widetilde{\mtx{\Omega}}_2 = \mtx{\Omega}_2^{\top} / \sqrt{2k}$ and rewrite
     \begin{align}
@@
         &= \sqrt{2k} \cdot \mathbb{E}^{k}\left[ \lVert \mtx{A} \widetilde{\mtx{\Omega}}_2^{\top} \rVert _2 \right] && \text{(definition of $\widetilde{\mtx{\Omega}}_2$)} \notag \\
-        &= \sqrt{2k} \cdot \left(\mathbb{E}^{\frac{k}{2}}\left[ \lVert (\mtx{A} \widetilde{\mtx{\Omega}}_2^{\top}) (\mtx{A} \widetilde{\mtx{\Omega}}_2^{\top})^{\top} \rVert _2 \right]\right)^{\frac{1}{2}} && \text{($\lVert \mtx{B} \rVert _2 = \lVert \mtx{B} \mtx{B}^{\top} \rVert _2^{\frac{1}{2}} $ )} \notag \\
-        &\leq \sqrt{2k} \cdot \left( \mathbb{E}^{\frac{k}{2}}\left[ \lVert (\mtx{A} \widetilde{\mtx{\Omega}}_2^{\top}) (\mtx{A} \widetilde{\mtx{\Omega}}_2^{\top})^{\top} - \mtx{A}^{\top} \mtx{A} \rVert _2 + \lVert \mtx{A}^{\top} \mtx{A} \rVert _2 \right]\right)^{\frac{1}{2}} && \text{(triangle inequality)} \notag \\
-        &\leq \sqrt{2k} \cdot \left( \mathbb{E}^{\frac{k}{2}}\left[ \lVert (\widetilde{\mtx{\Omega}}_2 \mtx{A} )^{\top} (\widetilde{\mtx{\Omega}}_2 \mtx{A}) - \mtx{A}^{\top} \mtx{A} \rVert _2\right]\right)^{\frac{1}{2}} + \sqrt{2k} \cdot \lVert \mtx{A} \rVert _2.
-&& \text{($(a+b)^{\frac{1}{2}} \leq a^{\frac{1}{2}} + b^{\frac{1}{2}}$)}
+        &= \sqrt{2k} \cdot \sqrt{ \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\mtx{A} \widetilde{\mtx{\Omega}}_2^{\top}) (\mtx{A} \widetilde{\mtx{\Omega}}_2^{\top})^{\top} \rVert _2 \right] } && \text{($\lVert \mtx{B} \rVert _2 = \lVert \mtx{B} \mtx{B}^{\top} \rVert _2^{\sfrac{1}{2}} $ )} \notag \\
+        &\leq \sqrt{2k} \cdot \sqrt{ \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\mtx{A} \widetilde{\mtx{\Omega}}_2^{\top}) (\mtx{A} \widetilde{\mtx{\Omega}}_2^{\top})^{\top} - \mtx{A}^{\top} \mtx{A} \rVert _2 \right] + \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{A}^{\top} \mtx{A} \rVert _2 \right]} && \text{(triangle inequality)} \notag \\
+        &\leq \sqrt{2k} \cdot \sqrt{ \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\widetilde{\mtx{\Omega}}_2 \mtx{A} )^{\top} (\widetilde{\mtx{\Omega}}_2 \mtx{A}) - \mtx{A}^{\top} \mtx{A} \rVert _2\right]} + \sqrt{2k} \cdot \lVert \mtx{A} \rVert _2.
+&& \text{($\sqrt{a+b} \leq \sqrt{a} + \sqrt{b}$)}
     \end{align}
-    By \cite{} $\widetilde{\mtx{\Omega}}_2 \in \mathbb{R}^{2k \times (n - k)}$ satisfies the \textcolor{red}{$(c_1 \sqrt{\frac{q}{2k}}, c_2 e^{-q}, q)$}-JL moment property for \todo{any $q$ (in what?) $\geq \dots$} and some constants $c_1, c_2 > 0$, hence, by \cite[Lemma 4]{cohen-2016-optimal-approximate} it satisfies the \textcolor{red}{$(2 c_1 \sqrt{\frac{q}{2k}}, 9^d c_2 e^{-q}, d, q)$}-OSE moment property for any $d \in \mathbb{N}$.
-For convenience, we choose $d=2k$ and $q=\frac{k}{2}$, such that $\widetilde{\mtx{\Omega}}_2$ has $(c_1, c_2 (3^4e^{-\frac{1}{2}})^k, 2k, \frac{k}{2})$-OSE moment property.
+    By \cite{} $\widetilde{\mtx{\Omega}}_2 \in \mathbb{R}^{2k \times l}$ satisfies the \todo{$(c_1 \sqrt{q/2k}, c_2 e^{-q}, q)$}-JL moment property for \todo{any $q$ (in what?) $\geq \dots$} and some constants $c_1, c_2 > 0$, hence, by \cite[Lemma 4]{cohen-2016-optimal-approximate} it satisfies the $(2 c_1 \sqrt{q/2k}, 9^d c_2 e^{-q}, d, q)$-OSE moment property for any $d \in \mathbb{N}$.
+For convenience, we choose $d=2k$ and $q=k/2$, such that $\widetilde{\mtx{\Omega}}_2$ has $(c_1, c_2 (3^4e^{-\sfrac{1}{2}})^k, 2k, k/2)$-OSE moment property.
 We may then apply an intermediary result from the proof of \cite[Theorem 1]{cohen-2016-optimal-approximate} to bound
     \begin{equation}
-        \mathbb{E}^{\frac{k}{2}}\left[ \lVert (\widetilde{\mtx{\Omega}}_2 \mtx{A} )^{\top} (\widetilde{\mtx{\Omega}}_2 \mtx{A}) - \mtx{A}^{\top} \mtx{A} \rVert _2\right] \leq c_1 c_2^2 3^8e^{-1} \left( \lVert \mtx{A} \rVert _2^2 + \frac{\lVert \mtx{A} \rVert _F^2}{k} \right).
+        \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\widetilde{\mtx{\Omega}}_2 \mtx{A} )^{\top} (\widetilde{\mtx{\Omega}}_2 \mtx{A}) - \mtx{A}^{\top} \mtx{A} \rVert _2\right] \leq c_1 c_2^2 3^8e^{-1} \left( \lVert \mtx{A} \rVert _2^2 + \frac{\lVert \mtx{A} \rVert _F^2}{k} \right).
     \end{equation}
@@
         \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \rVert _2 \right]
-        &\leq \sqrt{2k} \cdot \sqrt{c_1 e^{-1}} c_2 3^4 \left( \lVert \mtx{A} \rVert _2^2  + \frac{\lVert \mtx{A} \rVert _F^2}{k} \right)^{\frac{1}{2}} + \sqrt{2k} \cdot \lVert \mtx{A} \rVert _2 \notag \\
+        &\leq \sqrt{2k} \cdot \sqrt{c_1 e^{-1}} c_2 3^4 \left( \lVert \mtx{A} \rVert _2^2  + \frac{\lVert \mtx{A} \rVert _F^2}{k} \right)^{\sfrac{1}{2}} + \sqrt{2k} \cdot \lVert \mtx{A} \rVert _2 \notag \\
         &\leq 2 \sqrt{2} (1 + \sqrt{c_1 e^{-1}} c_2 3^4)\left( \sqrt{k} \cdot \lVert \mtx{A} \rVert _2  + \lVert \mtx{A} \rVert _F \right).
@@
         \mathbb{E}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^{k} \right]
-        = \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\frac{k}{2}} \right]
-        \leq \left( k + 1 \right) \left( \frac{1}{(k + 1)!}\right)^{\frac{k}{k + 1}} \left( \frac{3 k}{2}\right)^{\frac{k}{2}}.
+        = \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{k}{2}} \right]
+        \leq \left( k + 1 \right) \left( \frac{1}{(k + 1)!}\right)^{\sfrac{k}{k + 1}} \left( \frac{3 k}{2}\right)^{\sfrac{k}{2}}.
     \end{equation}
-    With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\frac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\frac{1}{n}} \leq \frac{e}{n}$ follow respectively.
+    With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\sfrac{1}{n}} \leq \frac{e}{n}$ follow respectively.
 Hence,
@@
 
-\begin{lemma}{Parameter-dependent Nystrom approximation}{nystrom}
+\begin{lemma}{Nyström approximation for parameter-dependent matrices}{nystrom}
     Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-Then its Nyström approximation $\widehat{\mtx{A}}(t)$ with Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\Omega}}$ for even $n_{\Omega} \in \mathbb{N}$ and $\gamma \geq 1$ verifies with probability at least $1 - \gamma^{-\frac{n_{\Omega}}{4}}$
+Then its Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{A}}(t)$ with Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ for even $n_{\mtx{\Omega}} \in \mathbb{N}$ and $\gamma \geq 1$ verifies with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
     \begin{equation}
-        \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F~\mathrm{d}t \leq \gamma \sqrt{2} \cdot \frac{1 + 2c}{\sqrt{n_{\Omega}}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t.
+        \int_{a}^{b} \lVert \mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F~\mathrm{d}t \leq \gamma \sqrt{2} \cdot \frac{1 + 2c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t.
     \end{equation}
     for a universal constant $c > 0$.
-In particular, if we choose $n_{\Omega} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})), n_{\Omega} \geq 2$, then with probability at least $1-\delta$ for $\delta \in (0, e^{-e^{4 \varepsilon^2}})$ and $\varepsilon > 0$ we have $\int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F~\mathrm{d}t  < \varepsilon \int_{a}^{b} \Trace(A(t))~\mathrm{d}t$.
+In particular, if we choose $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})), n_{\mtx{\Omega}} \geq 2$, then with probability at least $1-\delta$ for $\delta \in (0, e^{-e^{4 \varepsilon^2}})$ and $\varepsilon > 0$ we have $\int_{a}^{b} \lVert \mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F~\mathrm{d}t  < \varepsilon \int_{a}^{b} \Trace(A(t))~\mathrm{d}t$.
 \end{lemma}
@@
 \begin{proof}
-    We let $n_{\Omega} = 2k$ for some $k \in \mathbb{N}$, i.e.
+    We let $n_{\mtx{\Omega}} = 2k$ for some $k \in \mathbb{N}$, i.e.
 $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$.
@@
     \begin{equation}
-        \lVert \mtx{A} - \widehat{\mtx{A}} \rVert _F 
-        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert (\mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F.
+        \lVert \mtx{A} - \Nystr{\mtx{\Omega}}{\mtx{A}} \rVert _F 
+        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F.
         \label{equ:nystrom-proof-persson-bonud}
@@
     \begin{equation}
-        \lVert (\mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F 
-        \leq \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F
-        \leq \sqrt{k} \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2.
+        \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F 
+        \leq \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F
+        \leq \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2.
         \label{equ:nystrom-proof-processed-tail}
@@
     \begin{align}
-        \mathbb{E}^{\frac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
-        &= \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]^2 && \text{(by definition of $\mathbb{E}^{k}$)} \notag \\
-        &\leq c \left( \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _F \right)^2 && \text{(\reflem{lem:ose} with $\mtx{A}=\mtx{\Lambda}_2^{\frac{1}{2}}$)} \notag \\
-        &\leq c \left( \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _2^2  + \frac{1}{k}\lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _F^2 \right) && \text{($(a + b)^2 \leq a^2 + b^2$)} \notag \\
+        \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
+        &= \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]^2 && \text{(by definition of $\mathbb{E}^{k}$)} \notag \\
+        &\leq c \left( \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F \right)^2 && \text{(\reflem{lem:ose} with $\mtx{A}=\mtx{\Lambda}_2^{\sfrac{1}{2}}$)} \notag \\
+        &\leq c \left( \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2  + \frac{1}{k}\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right) && \text{($(a + b)^2 \leq a^2 + b^2$)} \notag \\
         &= c \left( \lambda_{k+1} + \frac{1}{k}\Trace(\mtx{\Lambda}_2) \right) && \text{(definition of $\lVert \cdot \rVert _2$ and $\lVert \cdot \rVert _F$)} \notag \\
@@
     \begin{equation}
-        \mathbb{E}^{\frac{k}{2}} \left[\lVert \mtx{A} - \widehat{\mtx{A}} \rVert _F \right]
+        \mathbb{E}^{\sfrac{k}{2}} \left[\lVert \mtx{A} - \Nystr{\mtx{\Omega}}{\mtx{A}} \rVert _F \right]
         \leq \frac{1 + 2c}{\sqrt{k}} \Trace(\mtx{A})
@@
     \begin{equation}
-        \mathbb{E}^{\frac{k}{2}}\left[ \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F~\mathrm{d}t \right] 
-        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}^{\frac{k}{2}} \left[\lVert \mtx{A} - \widehat{\mtx{A}} \rVert _F \right] ~\mathrm{d}t% \notag \\
-        %&\leq \mathbb{E}^{\frac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\frac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
-        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t + \mathbb{E}^{\frac{k}{2}} \left[ \int_{a}^{b}  \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\frac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
-        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\frac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\frac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
+        \mathbb{E}^{\sfrac{k}{2}}\left[ \int_{a}^{b} \lVert \mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F~\mathrm{d}t \right] 
+        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[\lVert \mtx{A} - \Nystr{\mtx{\Omega}}{\mtx{A}} \rVert _F \right] ~\mathrm{d}t% \notag \\
+        %&\leq \mathbb{E}^{\sfrac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
+        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t + \mathbb{E}^{\sfrac{k}{2}} \left[ \int_{a}^{b}  \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
+        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
         \leq \frac{1 + 2c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t.
     \end{equation}
-    From Markov's inequality \cite{} follows with probability at least $1 - \gamma^{-\frac{k}{2}}$
+    From Markov's inequality follows with probability at least $1 - \gamma^{-\sfrac{k}{2}}$
     \begin{equation}
-        \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{1 + 2c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t,
+        \int_{a}^{b} \lVert \mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{1 + 2c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t,
     \end{equation}
-    from which follows the first statement by replacing $k = \frac{n_{\Omega}}{2}$.
+    from which follows the first statement by replacing $k = \frac{n_{\mtx{\Omega}}}{2}$.
 
-    Setting $\delta = \gamma^{-\frac{k}{2}}$ and choosing $k = \lceil \varepsilon^{-2}\log(\delta^{-1}) \rceil$ we get
+    Setting $\delta = \gamma^{-\sfrac{k}{2}}$ and choosing $k = \lceil \varepsilon^{-2}\log(\delta^{-1}) \rceil$ we get
     \begin{align}
         \gamma \frac{1}{\sqrt{k}}
-        &= \delta^{-\frac{2}{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} \frac{1}{\sqrt{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} && \text{(definition of $\delta$ and choice of $k$)} \notag \\
-        &= e^{\frac{2}{\varepsilon^{-2}} \frac{\varepsilon^{-2}\log(\delta^{-1})}{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} \frac{1}{\sqrt{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} && \text{($\delta = e^{-\log(\delta^{-1})}$)} \notag \\
+        &= \delta^{-\sfrac{2}{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} \frac{1}{\sqrt{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} && \text{(definition of $\delta$ and choice of $k$)} \notag \\
+        &= e^{\sfrac{2}{\varepsilon^{-2}} \frac{\varepsilon^{-2}\log(\delta^{-1})}{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} \frac{1}{\sqrt{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} && \text{($\delta = e^{-\log(\delta^{-1})}$)} \notag \\
         &\leq e^{2\varepsilon^{2}} \frac{\varepsilon}{\sqrt{ \log(\delta^{-1})}} && \text{($\lceil x \rceil \geq x$ if $x \geq 0$)}
@@
 
-\begin{theorem}{Parameter-dependent Nyström trace estimator}{nystrom}
-    Suppose $f(\mtx{A}, t)$ is continuous and non-negative in $t \in [a, b]$ for a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ and $n_{\Omega} \geq 6, n_{\Omega} - 4 \geq r \geq 2$ two integers.
-Then for all $\gamma \geq 1$ with probability at least $1 - \gamma^{-(n_{\Omega} - r)}$
+\begin{theorem}{Nyström estimator for parameter-dependent matrices}{nystrom}
+    Suppose $f(\mtx{A}, t)$ is continuous and non-negative in $t \in [a, b]$ for a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ and $n_{\mtx{\Omega}} \geq 6, n_{\mtx{\Omega}} - 4 \geq r \geq 2$ two integers.
+Then for all $\gamma \geq 1$ with probability at least $1 - \gamma^{-(n_{\mtx{\Omega}} - r)}$
     \begin{equation}
-        \int_{a}^{b} \left| \Trace(f(\mtx{A}, t)) - \Nystr{n_{\Omega}}(f(\mtx{A}, t)) \right| ~\mathrm{d}t
+        \int_{a}^{b} \left| \Trace(f(\mtx{A}, t)) - \Trace(\Nystr{\mtx{\Omega}}{f}(\mtx{A}, t)) \right| ~\mathrm{d}t
         \leq \gamma^2 (1 + r) \int_{a}^{b} \sum_{i = r+1}^{n} \sigma_i(f(\mtx{A}, t)) ~\mathrm{d}t.
@@
     \begin{equation}
-        \lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{1/2} \mtx{\Omega}}) f(\mtx{A}, t)^{1/2} \rVert _2^2
+        \lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _2^2
     \end{equation}
-    Hence, by \cite[Theorem 9]{kressner-2023-randomized-lowrank}, with $n_{\Omega} = p + r$, the result follows
-    with $\lVert \cdot \rVert _2 \leq \lVert \cdot \rVert _F$ and $\sigma_i(f^{1/2}(\mtx{A}, t))^{2} = \sigma_i(f(\mtx{A}, t))$.
+    Hence, by \cite[Theorem 9]{kressner-2023-randomized-lowrank}, with $n_{\mtx{\Omega}} = p + r$, the result follows
+    with $\lVert \cdot \rVert _2 \leq \lVert \cdot \rVert _F$ and $\sigma_i(f^{\sfrac{1}{2}}(\mtx{A}, t))^{2} = \sigma_i(f(\mtx{A}, t))$.
 \end{proof}
 
-\subsection{Parameter-dependent Nyström++ estimator}
+\subsection{Nyström++ estimator for parameter-dependent matrices}
 \label{subsec:nystrom-pp}
 
-Finally, we combine the analysis of the two estimators from \refsec{subsec:quadratic} and \refsec{subsec:nystrom} into a more efficient, general-purpose trace estimator.
+Finally, we combine the analysis of the two estimators from \refsec{subsec:hutchinson} and \refsec{subsec:nystrom} into a more efficient, general-purpose trace estimator.
 
-\begin{theorem}{Parameter-dependent Nyström++ trace estimator}{nystrom-pp}
+\begin{theorem}{Nyström++ trace estimator for parameter-dependent matrices}{nystrom-pp}
     Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-If $n_{\Psi} = n_{\Omega} = \mathcal{O}(\varepsilon^{-1} \log(2\delta^{-1})^2)$, with probability at least $1 - \delta$ for $\delta \in (0, 2^{-1} \min\{e^{\frac{1}{2}}, e^{-e^{4\varepsilon}}\})$ we have
+If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(2\delta^{-1})^2)$, with probability at least $1 - \delta$ for $\delta \in (0, 2^{-1} \min\{e^{\sfrac{1}{2}}, e^{-e^{4\varepsilon}}\})$ we have
     \begin{equation}
-        \int_{a}^{b} | \Trace(\mtx{A}(t)) - \Nystrpp{n_{\Omega}}{n_{\Psi}}(\mtx{A}(t)) |~\mathrm{d}t
+        \int_{a}^{b} | \Trace(\mtx{A}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{A}(t)) |~\mathrm{d}t
         \leq \varepsilon \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t.
@@
 \begin{proof}
-    By choosing $n_{\Psi} = n_{\Omega} = \mathcal{O}(\tilde{\varepsilon}^{-2} \log(\tilde{\delta}^{-1})^2)$ we get
+    By choosing $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\tilde{\varepsilon}^{-2} \log(\tilde{\delta}^{-1})^2)$ we get
     \begin{align}
-        &\int_{a}^{b} | \Trace(\mtx{A}(t)) - \Nystrpp{n_{\Omega}}{n_{\Psi}}(\mtx{A}(t)) |~\mathrm{d}t \notag \\
-        &= \int_{a}^{b} | \Hutch{n_{\Psi}}(\mtx{A}(t) - \widehat{\mtx{A}}(t)) - \Trace(\mtx{A}(t) - \widehat{\mtx{A}}(t)) |~\mathrm{d}t && \text{(by definition of estimators)} \notag \\
-        &= \tilde{\varepsilon} \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F ~\mathrm{d}t && \text{(\refthm{thm:quadratic} w.p. $\geq 1 - \tilde{\delta}$)} \notag \\
-        &= \tilde{\varepsilon}^2 \int_{a}^{b} \Trace(\mtx{A}(t)) ~\mathrm{d}t && \text{(\reflem{lem:nystrom} w.p. $\geq 1 - \widetilde{\delta}$)} 
+        &\int_{a}^{b} | \Trace(\mtx{A}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{A}(t)) |~\mathrm{d}t \notag \\
+        &= \int_{a}^{b} | \Hutch{\mtx{\Psi}}(\mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)) - \Trace(\mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)) |~\mathrm{d}t && \text{(by definition of estimators)} \notag \\
+        &\leq \tilde{\varepsilon} \int_{a}^{b} \lVert \mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F ~\mathrm{d}t && \text{(\refthm{thm:hutchinson} w.p. $\geq 1 - \tilde{\delta}$)} \notag \\
+        &\leq \tilde{\varepsilon}^2 \int_{a}^{b} \Trace(\mtx{A}(t)) ~\mathrm{d}t && \text{(\reflem{lem:nystrom} w.p. $\geq 1 - \widetilde{\delta}$)} 
     \end{align}
@@
 %     \begin{align}
-%         &\int_{a}^{b} | \Trace(\mtx{A}(t)) - \Nystrpp{n_{\Omega}}{n_{\Psi}}(\mtx{A}(t)) |~\mathrm{d}t \notag \\
-%         &= \int_{a}^{b} | \Hutch{n_{\Psi}}(\mtx{A}(t) - \widehat{\mtx{A}}(t)) - \Trace(\mtx{A}(t) - \widehat{\mtx{A}}(t)) |~\mathrm{d}t && \text{(definition of estimators)} \notag \\
-%         &= \todo{c} \frac{1}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F ~\mathrm{d}t && \text{(with probability $1 - \delta$ if $n_{\Psi} = \mathcal{O}(\varepsilon^{-1})$)} \notag \\
-%         &= \todo{c} \frac{1}{\sqrt{n_{\Psi} n_{\Omega}}} \int_{a}^{b} \Trace(\mtx{A}(t)) ~\mathrm{d}t
+%         &\int_{a}^{b} | \Trace(\mtx{A}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{A}(t)) |~\mathrm{d}t \notag \\
+%         &= \int_{a}^{b} | \Hutch{\mtx{\Psi}}(\mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)) - \Trace(\mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)) |~\mathrm{d}t && \text{(definition of estimators)} \notag \\
+%         &= \todo{c} \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F ~\mathrm{d}t && \text{(with probability $1 - \delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1})$)} \notag \\
+%         &= \todo{c} \frac{1}{\sqrt{n_{\mtx{\Psi}} n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{A}(t)) ~\mathrm{d}t
 %     \end{align}
@@
 \begin{equation}
-    g_{\sigma}(s) = \frac{1}{n \sigma \sqrt{2\pi}} e^{-\frac{s^2}{2\sigma^2}}.
+    g_{\sigma}(s) = \frac{1}{n \sigma \sqrt{2\pi}} e^{-\sfrac{s^2}{2\sigma^2}}.
     \label{equ:smoothing-kernel}
@@
     | g_{\sigma}(t - z) | 
-    = \frac{1}{n \sigma \sqrt{2 \pi}} \left| e^{-\frac{(t - z)^2}{2\sigma^2}} \right|
-    = \frac{1}{n \sigma \sqrt{2 \pi}} e^{-\frac{(t - x)^2}{2\sigma^2}}e^{\frac{y^2}{2\sigma^2}}.
-    %\leq \frac{1}{n \sigma \sqrt{2 \pi}} \max_{x + iy \in \mathcal{E}_{\chi}} e^{\frac{y^2}{2\sigma^2}} 
+    = \frac{1}{n \sigma \sqrt{2 \pi}} \left| e^{-\sfrac{(t - z)^2}{2\sigma^2}} \right|
+    = \frac{1}{n \sigma \sqrt{2 \pi}} e^{-\sfrac{(t - x)^2}{2\sigma^2}}e^{\sfrac{y^2}{2\sigma^2}}.
+    %\leq \frac{1}{n \sigma \sqrt{2 \pi}} \max_{x + iy \in \mathcal{E}_{\chi}} e^{\sfrac{y^2}{2\sigma^2}} 
     \end{equation}
-    Because $e^{-\frac{(t - x)^2}{2\sigma^2}} \leq 1$ for all $x, t \in \mathbb{R}$, and the maximum absolute value of $y$ is limited by the length of the semi-axis of the Bernstein ellipse in the direction of the imaginary axis, which is $(\chi - \chi^{-1}) / 2$, we upper bound
+    Because $e^{-\sfrac{(t - x)^2}{2\sigma^2}} \leq 1$ for all $x, t \in \mathbb{R}$, and the maximum absolute value of $y$ is limited by the length of the semi-axis of the Bernstein ellipse in the direction of the imaginary axis, which is $(\chi - \chi^{-1}) / 2$, we upper bound
     \begin{equation}
         \max_{z \in \mathcal{E}_{\chi}} | g_{\sigma}(t - z) | 
-        \leq \frac{1}{n \sigma \sqrt{2 \pi}} e^{\frac{(\chi - \chi^{-1})^2}{8 \sigma^2}} 
+        \leq \frac{1}{n \sigma \sqrt{2 \pi}} e^{\sfrac{(\chi - \chi^{-1})^2}{8 \sigma^2}} 
     \end{equation}
@@
     \Statex \textbf{Input:} Symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$, evaluation points $\{t_i\}_{i=1}^{n_t}$
-    \Statex \textbf{Parameters:} Degree $m$, allocation $(n_{\Omega}, n_{\Psi})$,  smoothing parameter $\sigma$
+    \Statex \textbf{Parameters:} Degree $m$, allocation $(n_{\mtx{\Omega}}, n_{\mtx{\Psi}})$,  smoothing parameter $\sigma$
     \Statex \textbf{Output:} Approximate evaluations of the spectral density $\{\phi_{\sigma}(t_i)\}_{i=1}^{n_t}$
@@
     %\State Compute  for all $t_i$ using %\refalg{alg:3-nystrom-chebyshev-exponentiation}
-    \State Generate standard Gaussian matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\Omega}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\Psi}}$%\glsfirst{sketching-matrix} $\in \mathbb{R}^{n \times n_{\Omega}}$
-    %\State Generate standard Gaussian %\glsfirst{random-matrix} $\in \mathbb{R}^{n \times n_{\Psi}}$
-    \State Initialize $[\mtx{V}_1, \mtx{V}_2, \mtx{V}_3] \gets [\mtx{0}_{n \times n_{\Omega}}, \mtx{\mtx{\Omega}}, \mtx{0}_{n \times n_{\Omega}}]$
-    \State Initialize $[\mtx{W}_1, \mtx{W}_2, \mtx{W}_3] \gets [\mtx{0}_{n \times n_{\Psi}}, \mtx{\Psi}, \mtx{0}_{n \times n_{\Psi}}]$
-    \State Initialize $[\mtx{K}_1(t_i), \mtx{K}_2(t_i)] \gets [\mtx{0}_{n_{\Omega} \times n_{\Omega}}, \mtx{0}_{n_{\Omega} \times n_{\Omega}}]$ for all $t_i$
-    \State Initialize $[\mtx{L}_1(t_i), \ell(t_i)] \gets [\mtx{0}_{n_{\Omega} \times n_{\Psi}}, 0]$ for all $t_i$
+    \State Generate standard Gaussian matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$%\glsfirst{sketching-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$
+    %\State Generate standard Gaussian %\glsfirst{random-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$
+    \State Initialize $[\mtx{V}_1, \mtx{V}_2, \mtx{V}_3] \gets [\mtx{0}_{n \times n_{\mtx{\Omega}}}, \mtx{\mtx{\Omega}}, \mtx{0}_{n \times n_{\mtx{\Omega}}}]$
+    \State Initialize $[\mtx{W}_1, \mtx{W}_2, \mtx{W}_3] \gets [\mtx{0}_{n \times n_{\mtx{\Psi}}}, \mtx{\Psi}, \mtx{0}_{n \times n_{\mtx{\Psi}}}]$
+    \State Initialize $[\mtx{K}_1(t_i), \mtx{K}_2(t_i)] \gets [\mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}}, \mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}}]$ for all $t_i$
+    \State Initialize $[\mtx{L}_1(t_i), \ell(t_i)] \gets [\mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Psi}}}, 0]$ for all $t_i$
     %\State Set $\breve{\phi}_{\sigma}^{(m)}(t_i) \gets 0$ for all $t_i$
@@
     \For {$i = 1, \dots, n_t$}
-      \State $\phi_{\sigma}(t_i) \gets \Trace\left( \mtx{K}_1(t_i)^{\dagger}\mtx{K}_2(t_i) \right) + \frac{1}{n_{\Psi}} \left( \ell(t_i) + \Trace\left( \mtx{L}_1(t_i)^{\top} \mtx{K}_1(t_i)^{\dagger} \mtx{L}_1(t_i) \right)  \right) $ \label{lin:4-nystromchebyshev-nystrom-pp}
+      \State $\phi_{\sigma}(t_i) \gets \Trace\left( \mtx{K}_1(t_i)^{\dagger}\mtx{K}_2(t_i) \right) + \frac{1}{n_{\mtx{\Psi}}} \left( \ell(t_i) + \Trace\left( \mtx{L}_1(t_i)^{\top} \mtx{K}_1(t_i)^{\dagger} \mtx{L}_1(t_i) \right)  \right) $ \label{lin:4-nystromchebyshev-nystrom-pp}
     \EndFor
@@
 
-With the cost of a matrix-vector product denoted by $c(n)$, and supposing $n_{\Omega} \approx n_{\Psi}$, we determine the computational complexity of the Chebyshev-Nyström++ method to be $\mathcal{O}(m \log(m) n_t + m n_{\Omega}^2 n + m n_t n_{\Omega}^2 +  m c(n) n_{\Omega} + n_t n_{\Omega}^3)$, with $\mathcal{O}(m n_t + n n_{\Omega} + n_{\Omega}^2 n_t)$ required additional storage.
+With the cost of a matrix-vector product denoted by $c(n)$, and supposing $n_{\mtx{\Omega}} \approx n_{\mtx{\Psi}}$, we determine the computational complexity of the Chebyshev-Nyström++ method to be $\mathcal{O}(m \log(m) n_t + m n_{\mtx{\Omega}}^2 n + m n_t n_{\mtx{\Omega}}^2 +  m c(n) n_{\mtx{\Omega}} + n_t n_{\mtx{\Omega}}^3)$, with $\mathcal{O}(m n_t + n n_{\mtx{\Omega}} + n_{\mtx{\Omega}}^2 n_t)$ required additional storage.
 
@@
     Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the Chebyshev-Nyström++ estimate $\underline{\widetilde{\phi}}_{\sigma}^{(m)}$ with \emph{even} degree $m \in \mathbb{N}$, $n_{\Omega} = n_{\Psi} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2)$, and smoothing parameter $\sigma > 0$ satisfies with probability at least $1 - \delta$
+Then the Chebyshev-Nyström++ estimate $\underline{\widetilde{\phi}}_{\sigma}^{(m)}$ with \emph{even} degree $m \in \mathbb{N}$, $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2)$, and smoothing parameter $\sigma > 0$ satisfies with probability at least $1 - \delta$
     \begin{equation}
@@
     \end{equation}
-    for all $\delta \in (0, 2^{-1} \min\{e^{\frac{1}{2}}, e^{-e^{8\varepsilon}}\})$.
+    for all $\delta \in (0, 2^{-1} \min\{e^{\sfrac{1}{2}}, e^{-e^{8\varepsilon}}\})$.
 \end{theorem}
@@
 
-We now discuss what value to choose for $n_{\Omega}$ in the Nyström trace estimator introduced in \refsec{subsec:nystrom}.
+We now discuss what value to choose for $n_{\mtx{\Omega}}$ in the Nyström trace estimator introduced in \refsec{subsec:nystrom}.
 In order to guarantee a good approximation error based on \refthm{thm:nystrom}, we analyze the numerical rank of the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
@@
 \begin{equation}
-    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) = \frac{1}{n \sqrt{2 \pi \sigma^2}} e^{-\frac{(t - \lambda_{(i)})^2}{2 \sigma^2}},
+    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) = \frac{1}{n \sqrt{2 \pi \sigma^2}} e^{-\sfrac{(t - \lambda_{(i)})^2}{2 \sigma^2}},
     \label{equ:gaussian-kernel-eigenvalues}
@@
 
-\todo{Discuss how this informs choice of $n_{\Omega}$.}
+\todo{Discuss how this informs choice of $n_{\mtx{\Omega}}$.}
 
@@
 \begin{equation}
-    \alpha e^{-\frac{\lVert \vct{x} \rVert _2^2}{ 2 \beta^2 }}
+    \alpha e^{-\sfrac{\lVert \vct{x} \rVert _2^2}{ 2 \beta^2 }}
     \label{equ:5-experiments-gaussian-cell}
@@
         \input{plots/convergence-degree.pgf}
-        \caption{$n_{\Omega} + n_{\Psi} = 160$}
+        \caption{$n_{\mtx{\Omega}} + n_{\mtx{\Psi}} = 160$}
         \label{convergence-randvec}
@@
     \end{subfigure}
-    \caption{For increasing values of s$n_{\Omega} + n_{\Psi}$ but fixed $m$ we plot the $L^1$ relative approximation error for the model problem with $\sigma=0.05$.}
+    \caption{For increasing values of s$n_{\mtx{\Omega}} + n_{\mtx{\Psi}}$ but fixed $m$ we plot the $L^1$ relative approximation error for the model problem with $\sigma=0.05$.}
     \label{fig:convergence}
@@
     \caption{The NC++ method for different ways of allocations a 
-    total of $n_{\Omega} + n_{\Psi}=80$ random vectors
+    total of $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}=80$ random vectors
     to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation
@@
 
-\tableofcontents
-
 \end{document}

### 2024-05-24 paper/main.tex
add stable rank discussion

@@
 In numerous problems across physics, chemistry, engineering, and data science, eigenvalues and especially their distribution are crucial for understanding the properties of a system: In electronic structure calculations they represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}.
-The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be captured with the spectral density
+The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be captured in the spectral density
 \begin{equation}
@@
 In many applications, it is also acceptable to trade in some of the finer details of the spectrum for a better approximability by smoothing $\phi$, typically through convolution with a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}.
-This converts the problem to estimating the trace of a smooth parameter-dependent matrix function.
+This converts the problem to estimating the trace of a smooth parameter-dependent matrix.
 
@@
 \begin{itemize}
-    \item Definition of parameter-dependent matrix ($\mtx{B}(t) \in \mathbb{R}^{n \times n}$, every entry depends continuously on $t \in [a,b]$ \cite{kressner-2023-randomized-lowrank})
+    \item Brief definition of parameter-dependent matrix ($\mtx{B}(t) \in \mathbb{R}^{n \times n}$, every entry depends continuously on $t \in [a,b]$ \cite{kressner-2023-randomized-lowrank})
     \item Trace estimation through mat-vecs (if matrix is only accessed implicitly \cite{pearlmutter-1994-fast-exact} or for complicated matrix functions [partition function, spectral density, exponential integrators])
-    \item Existing trace estimation methods for constant matrices: Girard-Hutchinson \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making} which have only recently been specialized to trace estimation of matrix functions \cite{chen-2023-krylovaware-stochastic}
-    \item Trace estimation for parameter-dependent matrices (Nyström++ but no analysis \cite{lin-2017-randomized-estimation}, low-rank approximation \cite{kressner-2023-randomized-lowrank}, dynamic trace estimation but only discrete setting [Woodruff, Dharangutte])
+    \item Existing trace estimation methods for constant matrices: Girard-Hutchinson \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} and variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making} which have only recently been specialized to trace estimation of matrix functions \cite{chen-2023-krylovaware-stochastic}
+    \item Trace estimation in the parameter-dependent matrices (Nyström++ but no analysis \cite{lin-2017-randomized-estimation}, only low-rank approximation \cite{kressner-2023-randomized-lowrank}, dynamic trace estimation for sequence of similar matrices \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace})
     %\item Error analysis for spectral density approximation (Wasserstein-1 error [Braverman, Chen], missing $1/\varepsilon$ guarantees)
@@
 \begin{itemize}
-    \item Theoretical guarantees in the $L^1$-norm for three trace estimators on parameter-dependent matrices (Girard-Hutchinson, Nyström, Nyström++), in particular, $\mathcal{O}(\varepsilon^{-1})$ scaling with number of mat-vecs
+    \item Theoretical guarantees in the $L^1$-norm for three established trace estimators for constant matrices when applied to parameter-dependent matrices (Girard-Hutchinson, Nyström, Nyström++), in particular, $\mathcal{O}(\varepsilon^{-1})$ scaling with number of mat-vecs
     \item A simple and efficient algorithm for estimating spectral densities using any of the three estimators using Chebyshev expansion, including theoretical guarantees
@@
 \begin{itemize}
-    \item we denote with $\mathbb{N}$ all positive whole numbers and with $\mathbb{R}$ all real numbers;
+    \item we denote with $\mathbb{N}$ all positive whole numbers;
     \item $e$ denotes Euler's number and $\log$ is the corresponding logartihm;
     \item the ceiling function $\lceil x \rceil$ of a real number $x$ returns the greatest integer larger than or equal to $x$;
-    \item scalars are represented by lower case Greek or Latin letters ($s$, $\varepsilon$, \dots),
-    vectors are additionally printed in bold ($\vct{v}$, $\vct{\psi}$, \dots),
-    and matrices are additionally capitalized ($\mtx{A}$, $\mtx{\Omega}$, \dots);
+    \item scalars are represented by lower case Greek or Latin letters ($s$, $\varepsilon$, \dots), vectors are additionally printed in bold ($\vct{v}$, $\vct{\psi}$, \dots), and matrices are additionally capitalized ($\mtx{A}$, $\mtx{\Omega}$, \dots);
     \item the expected value of a random variable $x$ is denoted with $\mathbb{E}[x]$.
 The $k$-th moment of a random variable $x$ is defined as $\mathbb{E}^{k}[x] = \left(\mathbb{E}\left[ | x |^{k} \right] \right)^{\sfrac{1}{k}}$;
-    \item the identity matrix $\mtx{I}_n = \diag(1, \dots, 1) \in \mathbb{R}^{n \times n}$
-    carries ones on its diagonal and zeros everywhere else. The zero matrix
-    $\mtx{0}_{n \times m} \in \mathbb{R}^{n \times m}$ consists of only zero entries;
+    \item the identity matrix $\mtx{I}_n = \diag(1, \dots, 1) \in \mathbb{R}^{n \times n}$ carries ones on its diagonal and zeros everywhere else.
+The zero matrix $\mtx{0}_{n \times m} \in \mathbb{R}^{n \times m}$ consists of only zero entries;
     \item in a Gaussian random vector and a Gaussian random matrix, all entries are independently sampled from the standard normal distribution;
-    \item matrix norms are denoted with $\lVert \cdot \rVert$. For symmetric matrices, the $p$-Schatten norm is
-    $\lVert \mtx{A} \rVert _{(p)} = (\sum_{i=1}^n |\lambda_i|^p)^{1/p}$.
+    \item matrix norms are denoted with $\lVert \cdot \rVert$.
+For symmetric matrices, the $p$-Schatten norm is $\lVert \mtx{A} \rVert _{(p)} = (\sum_{i=1}^n |\lambda_i|^p)^{1/p}$.
 We refer to the spectral norm as $\lVert \mtx{A} \rVert _2 = \lVert \mtx{A} \rVert _{(\infty)} = \max_{i=1,\dots,n}|\lambda_i|$, the nuclear norm as $\lVert \mtx{A} \rVert _{\ast} = \lVert \mtx{A} \rVert _{(1)} = \sum_{i=1}^{n} |\lambda_i|$, and the Frobenius norm as $\lVert \mtx{A} \rVert _F = \lVert \mtx{A} \rVert _{(2)} = (\sum_{i=1}^{n} |\lambda_i|^2)^{\sfrac{1}{2}}$.
@@
 \end{equation}
-\todo{Motivate why we use this.}
+We use it to measure the error of the trace estimators. \todo{Motivate why we use this.}
 
@@
 
-We first consider \refequ{equ:hutchinson-trace-estimator}.
-\todo{Discuss theory for constant matrix
-}
+We first consider the estimator \refequ{equ:hutchinson-trace-estimator}.
 
@@
     \begin{equation}
-        \int_{a}^{b} \left| \Trace(\mtx{A}(t)) - \Hutch{\mtx{\Psi}}(\mtx{A}(t)) \right| ~\mathrm{d}t < 16 \gamma k \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F  ~\mathrm{d}t.
+        \int_{a}^{b} \left| \Trace(\mtx{A}(t)) - \Hutch{\mtx{\Psi}}(\mtx{A}(t)) \right| ~\mathrm{d}t < 2^4 \gamma k \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F  ~\mathrm{d}t.
     \end{equation}
@@
     \begin{equation}
-        r(t) = \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi} - \Trace(\mtx{B}(t))
+        r(t) = \Trace(\mtx{B}(t)) - \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi}
     \end{equation}
@@
     First, we consider $r(t)$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence.
-From the proof of \cite[Lemma 3]{cortinovis-2022-randomized-trace} we know that $r$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$.
-Thus, by \cite[Theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
+From the proof of \cite[lemma 3]{cortinovis-2022-randomized-trace} we know that $r$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$.
+Thus, by \cite[theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
     \begin{align}
@@
         \mathbb{E}^{2k}\left[ r \right]
-        < \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F < 16 k \lVert \mtx{B} \rVert _F,
+        < \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F < 2^4 k \lVert \mtx{B} \rVert _F,
         \label{equ:hutchinson-trace-onequery-fixed}
@@
     Now we make the transition to the continuous.
-Minkowski's integral inequality \cite[Theorem 2.2]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting:
+Minkowski's integral inequality \cite[theorem 2.2]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting:
     \begin{equation}
         \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(t)|~\mathrm{d}t  \right]
-        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}\left[ |r(t)|^{2 k}\right]^{\sfrac{1}{2 k}}~\mathrm{d}t
-        < 16 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
+        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}^{2 k}\left[ r(t) \right]~\mathrm{d}t
+        < 2^4 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
     \end{equation}
@@
     \begin{equation}
-        \int_{a}^{b} |r(t)|~\mathrm{d}t < 16 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
+        \int_{a}^{b} |r(t)|~\mathrm{d}t < 2^4 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
         \label{equ:hutchinson-trace-onequery-uniform}
@@
 
-    To extend the bound to the $n_{\mtx{\Psi}}$-query quadratic trace estimator, we use a technique from the proof of \cite[Theorem 1]{cortinovis-2022-randomized-trace}. Let
+    To extend the bound to the $n_{\mtx{\Psi}}$-query quadratic trace estimator, we use a technique from the proof of \cite[theorem 1]{cortinovis-2022-randomized-trace}. Let
     \begin{equation}
@@
     \begin{equation}
-        r(t) = \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j - \Trace(\mtx{A}(t)).
+        r(t) = \Trace(\mtx{A}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j.
     \end{equation}
@@
     \begin{equation}
-        \int_{a}^{b} \left| \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j - \Trace(\mtx{A}(t)) \right| ~ \mathrm{d}t
-        < 16 k \gamma \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t.
+        \int_{a}^{b} \left| \Trace(\mtx{A}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j \right| ~ \mathrm{d}t
+        < 2^4 k \gamma \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t.
     \end{equation}
@@
 
-\todo{Discuss $1/\varepsilon^2$ drawback}
+\begin{remark}
+    If additionally $\mtx{A}(t)$ is nonzero for all $t \in [a, b]$, then it follows from the proof of \refthm{thm:hutchinson} that with probability at least $1 - \delta$
+    \begin{equation}
+        \int_{a}^{b} \frac{| \Trace(\mtx{A}(t)) - \Hutch{\mtx{\Psi}}(\mtx{A}(t)) |}{\Trace(\mtx{A}(t))} ~\mathrm{d}t < \varepsilon,
+    \end{equation}
+    if $n_{\mtx{\Psi}} \geq 2^{10} \varepsilon^{-2} \log(\delta^{-1})^2 (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ with $\mu(t) = \lVert \mtx{A}(t) \rVert _2 / \Trace(\mtx{A}(t))$.
+Compared to the constant matrix equivalent \cite[remark 2]{cortinovis-2022-randomized-trace} we observe an additional factor of $\log(\delta^{-1})$.
+\end{remark}
+
+\todo{The drawback of the Girard-Hutchinson estimator is the $\mathcal{O}(\varepsilon^{-2})$-requirement on the number of queries. }
 
@@
 
-Next we consider \refequ{equ:nystrom-trace-estimator}. \todo{Motivate Nyström approximation if matrix is low-rank}
+Next we consider \refequ{equ:nystrom-trace-estimator}. \todo{Motivate Nyström approximation if matrix is low-rank.}
 
-Pointwise eigenvalue decomposition
+\begin{theorem}{Nyström estimator for parameter-dependent matrices}{nystrom}
+    Suppose $f(\mtx{A}, t)$ is continuous and non-negative in $t \in [a, b]$ for a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ and $n_{\mtx{\Omega}} \geq 6, n_{\mtx{\Omega}} - 4 \geq r \geq 2$ two integers.
+Then for all $\gamma \geq 1$ with probability at least $1 - \gamma^{-(n_{\mtx{\Omega}} - r)}$
+    \begin{equation}
+        \int_{a}^{b} \left| \Trace(f(\mtx{A}, t)) - \Trace(\Nystr{\mtx{\Omega}}{f}(\mtx{A}, t)) \right| ~\mathrm{d}t
+        \leq \gamma^2 (1 + r) \int_{a}^{b} \sum_{i = r+1}^{n} \sigma_i(f(\mtx{A}, t)) ~\mathrm{d}t.
+    \end{equation}
+\end{theorem}
+
+\begin{proof}
+    Since $f$ is non-negative, $f(\mtx{A}, t)$ is positive definite for all $\mtx{A}$ and $t$.
+    Notice that by \cite[lemma 2.1]{frangella-2023-randomized-nystrom}, $f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t)$ is also positive semi-definite for all $t$. Therefore,
+    \begin{equation}
+        \left| \Trace(f(\mtx{A}, t)) - \Trace(\widehat{f}(\mtx{A}, t)) \right|
+        = \left| \Trace(f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t)) \right|
+        = \lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast}
+    \end{equation}
+    From the proof of \cite[corollary 8.2]{tropp-2023-randomized-algorithms} it follows that
+    \begin{equation}
+        \lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _2^2
+    \end{equation}
+    Hence, by \cite[theorem 9]{kressner-2023-randomized-lowrank}, with $n_{\mtx{\Omega}} = p + r$, the result follows
+    with $\lVert \cdot \rVert _2 \leq \lVert \cdot \rVert _F$ and $\sigma_i(f^{\sfrac{1}{2}}(\mtx{A}, t))^{2} = \sigma_i(f(\mtx{A}, t))$.
+\end{proof}
+
+\todo{Explain that \reflem{lem:nystrom} is a weaker result, and mainly there to help proof \refthm{thm:nystrom-pp}.}
+
+For the coming pages, we employ the following notation. Let
 \begin{equation}
@@
 \end{equation}
-where $\mtx{\Lambda}(t) = \operatorname{diag}(\lambda_1(t), \dots, \lambda_n(t))$.
-
+be the eigenvalue decomposition of $\mtx{A}(t)$ at a fixed $t$, i.e.
+$\mtx{\Lambda}(t) = \operatorname{diag}(\lambda_1(t), \dots, \lambda_n(t))$ and $\mtx{U}(t)^{\top} \mtx{U}(t) = \mtx{I}_n$.
 We define the matrix partitions
@@
     \end{align}
-    By \cite{} $\widetilde{\mtx{\Omega}}_2 \in \mathbb{R}^{2k \times l}$ satisfies the \todo{$(c_1 \sqrt{q/2k}, c_2 e^{-q}, q)$}-JL moment property for \todo{any $q$ (in what?) $\geq \dots$} and some constants $c_1, c_2 > 0$, hence, by \cite[Lemma 4]{cohen-2016-optimal-approximate} it satisfies the $(2 c_1 \sqrt{q/2k}, 9^d c_2 e^{-q}, d, q)$-OSE moment property for any $d \in \mathbb{N}$.
+    By \cite{} $\widetilde{\mtx{\Omega}}_2 \in \mathbb{R}^{2k \times l}$ satisfies the \todo{$(c_1 \sqrt{q/2k}, c_2 e^{-q}, q)$}-JL moment property for \todo{any $q$ (in what?) $\geq \dots$} and some constants $c_1, c_2 > 0$, hence, by \cite[lemma 4]{cohen-2016-optimal-approximate} it satisfies the $(2 c_1 \sqrt{q/2k}, 9^d c_2 e^{-q}, d, q)$-OSE moment property for any $d \in \mathbb{N}$.
 For convenience, we choose $d=2k$ and $q=k/2$, such that $\widetilde{\mtx{\Omega}}_2$ has $(c_1, c_2 (3^4e^{-\sfrac{1}{2}})^k, 2k, k/2)$-OSE moment property.
-We may then apply an intermediary result from the proof of \cite[Theorem 1]{cohen-2016-optimal-approximate} to bound
+We may then apply an intermediary result from the proof of \cite[theorem 1]{cohen-2016-optimal-approximate} to bound
     \begin{equation}
@@
 
-    The second term in \refequ{equ:OSE-moment-bound-initial} can be bounded with help of the proof of \cite[Lemma B.3]{tropp-2023-randomized-algorithms}
+    The second term in \refequ{equ:OSE-moment-bound-initial} can be bounded with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
     \begin{equation}
@@
     Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-Then its Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{A}}(t)$ with Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ for even $n_{\mtx{\Omega}} \in \mathbb{N}$ and $\gamma \geq 1$ verifies with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
+Then its Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{A}}(t)$ with Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ for \emph{even} $n_{\mtx{\Omega}} \in \mathbb{N}$ and $\gamma \geq 1$ verifies with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
     \begin{equation}
@@
 
-\todo{proof idea (structural bound, then higher order moment bound to apply Markov's inequality)}
+\todo{Proof idea: structural bound, then higher order moment bound to apply Markov's inequality.}
 
@@
 $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$.
-After temporarily omitting the parameter dependence, we can apply \cite[Theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound the integrand with \todo{maybe explain where $\mtx{\Omega}_i$ come from...}
+After temporarily omitting the parameter dependence, we can apply \cite[theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound the integrand with \todo{maybe explain where $\mtx{\Omega}_i$ come from...}
     \begin{equation}
@@
 
-    As in the proof of \cite[Lemma 3]{meyer-2021-hutch-optimal}, the first term can be bonud with
+    As in the proof of \cite[lemma 3]{meyer-2021-hutch-optimal}, the first term can be bonud with
     \begin{equation}
@@
 
-    We use Minkowski's integral inequality \cite[Theorem 2.2]{hardy-1952-inequalities} to get
+    We use Minkowski's integral inequality \cite[theorem 2.2]{hardy-1952-inequalities} to get
     \begin{equation}
@@
 
-\begin{theorem}{Nyström estimator for parameter-dependent matrices}{nystrom}
-    Suppose $f(\mtx{A}, t)$ is continuous and non-negative in $t \in [a, b]$ for a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ and $n_{\mtx{\Omega}} \geq 6, n_{\mtx{\Omega}} - 4 \geq r \geq 2$ two integers.
-Then for all $\gamma \geq 1$ with probability at least $1 - \gamma^{-(n_{\mtx{\Omega}} - r)}$
-    \begin{equation}
-        \int_{a}^{b} \left| \Trace(f(\mtx{A}, t)) - \Trace(\Nystr{\mtx{\Omega}}{f}(\mtx{A}, t)) \right| ~\mathrm{d}t
-        \leq \gamma^2 (1 + r) \int_{a}^{b} \sum_{i = r+1}^{n} \sigma_i(f(\mtx{A}, t)) ~\mathrm{d}t.
-    \end{equation}
-\end{theorem}
-
-\begin{proof}
-    Since $f$ is non-negative, $f(\mtx{A}, t)$ is positive definite for all $\mtx{A}$ and $t$.
-    Notice that by \cite[Lemma 2.1]{frangella-2023-randomized-nystrom}, $f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t)$ is also positive semi-definite for all $t$. Therefore,
-    \begin{equation}
-        \left| \Trace(f(\mtx{A}, t)) - \Trace(\widehat{f}(\mtx{A}, t)) \right|
-        = \left| \Trace(f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t)) \right|
-        = \lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast}
-    \end{equation}
-    From the proof of \cite[Corollary 8.2]{tropp-2023-randomized-algorithms} it follows that
-    \begin{equation}
-        \lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _2^2
-    \end{equation}
-    Hence, by \cite[Theorem 9]{kressner-2023-randomized-lowrank}, with $n_{\mtx{\Omega}} = p + r$, the result follows
-    with $\lVert \cdot \rVert _2 \leq \lVert \cdot \rVert _F$ and $\sigma_i(f^{\sfrac{1}{2}}(\mtx{A}, t))^{2} = \sigma_i(f(\mtx{A}, t))$.
-\end{proof}
-
 \subsection{Nyström++ estimator for parameter-dependent matrices}
@@
     Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(2\delta^{-1})^2)$, with probability at least $1 - \delta$ for $\delta \in (0, 2^{-1} \min\{e^{\sfrac{1}{2}}, e^{-e^{4\varepsilon}}\})$ we have
+If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(2\delta^{-1})^2)$ with \emph{even} $n_{\mtx{\Omega}} \geq 2$, then for $\delta \in (0, 2^{-1} e^{-e^{4\varepsilon}})$ with probability at least $1 - \delta$
     \begin{equation}
@@
         &\leq \tilde{\varepsilon} \int_{a}^{b} \lVert \mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F ~\mathrm{d}t && \text{(\refthm{thm:hutchinson} w.p. $\geq 1 - \tilde{\delta}$)} \notag \\
-        &\leq \tilde{\varepsilon}^2 \int_{a}^{b} \Trace(\mtx{A}(t)) ~\mathrm{d}t && \text{(\reflem{lem:nystrom} w.p. $\geq 1 - \widetilde{\delta}$)} 
+        &\leq \tilde{\varepsilon}^2 \int_{a}^{b} \Trace(\mtx{A}(t)) ~\mathrm{d}t && \text{(\reflem{lem:nystrom} w.p. $\geq 1 - \tilde{\delta}$)} 
     \end{align}
@@
 \end{proof}
+
+Comparing \refthm{thm:nystrom-pp} with its equivalent for constant matrices \cite[theorem 3.4]{persson-2022-improved-variants}, we notice that our choice of $\delta$ is more restricted and the dependence of $n_{\mtx{\Psi}}$ and $n_{\mtx{\Omega}}$ on $\delta$ is worse.
 
@@
 where $\delta$ is the Dirac delta distribution.
+Finding approximations to this distribution is challenging.
+Furthermore, in most applications the exact location of every eigenvalue is not important, but more so their approximate locations relative to each other, such as eigenvalue clusters, spectral gaps, or outliers.
+In the following we will introduce a smooth version of the spectral density, which is significantly easier to approximate at the cost of losing some of the finer characteristics of the spectrum.
 
-\subsection{Introduction to spectral density estimation}
+\subsection{Smooth spectral density}
 \label{subsec:spectral-density}
 
-\todo{Motivate reason for smoothing?} Therefore, we seek approximations to the smooth spectral density $\phi_{\sigma}$ defined as
+The smooth spectral density $\phi_{\sigma}$ is defined as
 \begin{equation}
@@
 \end{equation}
-with a Gaussian smoothing kernel of width $\sigma > 0$
+for a smoothing kernel $g_{\sigma}$ parametrized by a smoothing parameter $\sigma > 0$, which controls by how much $\phi$ is smoothed.
+We choose a Gaussian smoothing kernel of width $\sigma > 0$, given by
 \begin{equation}
@@
 \end{equation}
-This conversion of the task to a trace estimation problem for the parameter-dependent matrix $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ allows us to apply the results for the estimators derived in \refsec{sec:analysis}.
+The task in \refequ{equ:smooth-spectral-density} is to approximate the trace of the parameter-dependent matrix $g_{\sigma}(t\mtx{I}_n - \mtx{A})$, which allows us to apply the results derived in \refsec{sec:analysis}.
+
+\todo{
+Measuring error between spectral density $\phi$ and approximation $\widetilde{\phi}$
+\begin{equation}
+    \sup_{f \in \mathcal{S}} \int_{-1}^{1} f(t) (\phi(t) - \widetilde{\phi}(t)) \mathrm{d}t
+    \label{equ:error-metric}
+\end{equation}
+
+In \cite{lin-2016-approximating-spectral} propose $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - \lambda), \lambda \in [\lambda_{\text{min}}, \lambda_{\text{max}}]\}$
+
+In \cite{chen-2021-analysis-stochastic} use $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [-1, 1] \}$ where $\Theta$ the unit step which is equivalent to Kolmogorov-Smirnov.
+
+In \cite{braverman-2022-sublinear-time} use $\mathcal{S} = \text{Lip}_1 = \{f : |f(x) - f(y)| \leq |x - y| \}$ which is equivalent to the Wasserstein-1 norm.
+
+For small enough $\sigma$ and well separated eigenvalues, the  error \refequ{equ:error-metric} when testing against Lipschitz-1 functions we make by smoothing is already $\mathcal{\sigma}$.
+
+Having noted this, we will from now on measure errors in $L^1$ norm from smooth spectral density.
+}
 
@@
 
+\todo{
 Fast computation of expansion with DCT (duality with evaluations)
+}
 
@@
 The proof specializes Bernstein's theorem for Chebyshev interpolation to the case of a Gaussian smoothing kernel \refequ{equ:smoothing-kernel}.
-\todo{It loosely follows the proof give in the preprint of \cite[Theorem 2]{lin-2017-randomized-estimation}, but reaches slightly different conclusion.}
+\todo{It loosely follows the proof give in the preprint of \cite[theorem 2]{lin-2017-randomized-estimation}, but reaches slightly different conclusion.}
 
 \begin{proof}
-    From Bernstein's theorem \cite[Theorem 4.3]{trefethen-2008-gauss-quadrature} it follows that if $f$ is analytic within the Bernstein ellipse $\mathcal{E}_{\chi} \subset \mathbb{C}$ with foci $\pm 1$ and sum of semi-axes $\chi > 1$, then for any $m \in \mathbb{N}$
+    From Bernstein's theorem \cite[theorem 4.3]{trefethen-2008-gauss-quadrature} it follows that if $f$ is analytic within the Bernstein ellipse $\mathcal{E}_{\chi} \subset \mathbb{C}$ with foci $\pm 1$ and sum of semi-axes $\chi > 1$, then for any $m \in \mathbb{N}$
     \begin{equation}
@@
 
-\todo{Explain need for non-negativity for Nyström approximation, and mention alternatives (Putinar, Jackson-damping, ...)}
+\todo{
+Explain need for non-negativity for Nyström approximation (for positive semi-definiteness in theorem).
+
+Converges too slow: polynomial interpolation with Jackson damping \cite{jackson-1912-approximation-trigonometric,braverman-2022-sublinear-time}
+
+Non-negative Chebyshev expansion: expand $\sqrt{g_{\sigma}}$ to degree $m/2$ and then square the expansion. Error guarantee very comparable to normal expansion, and proof below.
+
+Side-note: it is possible to refine the coefficients of the Chebyshev expansion to get a better approximation in the supremum-norm over the set of all non-negative polynomials \cite{fejer-1916-uber-trigonometrische}.
+The non-convex nature of this problem and the relatively small improvement we get with this technique made us decide to not pursue this further.
+}
 
@@
     \begin{equation}
-        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \leq 2\sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \frac{m}{2}}\right) \frac{E_{\sqrt{2}\sigma, \frac{m}{2}}}{n} \equiv \frac{\underline{E}_{\sigma, m}}{n}
+        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \leq 2\sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right) \frac{E_{\sqrt{2}\sigma, \sfrac{m}{2}}}{n} \equiv \frac{\underline{E}_{\sigma, m}}{n}
         \label{equ:2-chebyshev-interpolation-sup-error-kernel}
@@
     \end{equation}
-    Therefore, omitting the arguments and noting $g_{\sigma} = (\sqrt{g_{\sigma}})^2$ and $\underline{g}_{\sigma}^{(m)} = (\sqrt{g_{\sigma}}^{(m/2)})^2$.
+    Therefore, omitting the arguments and noting $g_{\sigma} = (\sqrt{g_{\sigma}})^2$ and $\underline{g}_{\sigma}^{(m)} = (\sqrt{g_{\sigma}}^{(\sfrac{m}{2})})^2$.
     we have
     \begin{equation}
-        \left| g_{\sigma} - \underline{g}_{\sigma}^{(m)} \right| \leq \left( 2 \left| \sqrt{g_{\sigma}} \right| + \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(m/2)} \right| \right) \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(m/2)} \right|.
+        \left| g_{\sigma} - \underline{g}_{\sigma}^{(m)} \right| \leq \left( 2 \left| \sqrt{g_{\sigma}} \right| + \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right| \right) \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right|.
     \end{equation}
@@
         &\sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \notag \\
-        &\leq \left( 2 \frac{1}{\sqrt{n \sigma \sqrt{2\pi}}} + \sqrt{2 n \sigma \sqrt{2 \pi}} \cdot \frac{E_{\sqrt{2}\sigma, \frac{m}{2}}}{n}\right)\sqrt{2 n \sigma \sqrt{2 \pi}} \cdot \frac{E_{\sqrt{2}\sigma, \frac{m}{2}}}{n},
+        &\leq \left( 2 \frac{1}{\sqrt{n \sigma \sqrt{2\pi}}} + \sqrt{2 n \sigma \sqrt{2 \pi}} \cdot \frac{E_{\sqrt{2}\sigma, \sfrac{m}{2}}}{n}\right)\sqrt{2 n \sigma \sqrt{2 \pi}} \cdot \frac{E_{\sqrt{2}\sigma, \sfrac{m}{2}}}{n},
     \end{align}
@@
 
-Corollary for expanded spectral density
-
 \begin{theorem}{Chebyshev-Nyström++ method}{chebyshev-nystrom}
@@
         &\leq (1 + \varepsilon) \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \varepsilon \int_{-1}^{1} \left| \phi_{\sigma}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
-        &\leq (1 + \varepsilon) \underline{E}_{\sigma, m} + \varepsilon. && \text{(\reflem{lem:non-negative-chebyshev-error} and normalization)} \notag \\
+        &\leq 2 (1 + \varepsilon) \underline{E}_{\sigma, m} + \varepsilon. && \text{(\reflem{lem:non-negative-chebyshev-error} and normalization)} \notag \\
     \end{align}
 \end{proof}
+
+\todo{Maybe specialize \refthm{thm:nystrom} to spectral densities?}
 

### 2024-05-27 paper/main.tex
start filling in gaps in chapter 3

@@
 Therefore, many techniques have been proposed for approximating $\phi$.
-For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature from the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}.
+For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature rule from the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}.
 In many applications, it is also acceptable to trade in some of the finer details of the spectrum for a better approximability by smoothing $\phi$, typically through convolution with a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}.
@@
     Now we make the transition to the continuous.
-Minkowski's integral inequality \cite[theorem 2.2]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting:
+Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting:
     \begin{equation}
@@
 
-    We use Minkowski's integral inequality \cite[theorem 2.2]{hardy-1952-inequalities} to get
+    We use Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities} to get
     \begin{equation}
@@
 
-Comparing \refthm{thm:nystrom-pp} with its equivalent for constant matrices \cite[theorem 3.4]{persson-2022-improved-variants}, we notice that our choice of $\delta$ is more restricted and the dependence of $n_{\mtx{\Psi}}$ and $n_{\mtx{\Omega}}$ on $\delta$ is worse.
+Comparing \refthm{thm:nystrom-pp} with its equivalent for constant matrices \cite[theorem 3.4]{persson-2022-improved-variants}, we notice that our choice of $\delta$ is more restricted and the dependence of $n_{\mtx{\Psi}}$ and $n_{\mtx{\Omega}}$ on $\delta$ is slightly worse.
 
@@
 \begin{equation}
-    \phi_{\sigma}(t) = \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i) = \Trace(g_{\sigma}(t\mtx{I}_n - \mtx{A}))
+    \phi_{\sigma}(t) = \frac{1}{n} \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i) = \frac{1}{n} \Trace(g_{\sigma}(t\mtx{I}_n - \mtx{A}))
     \label{equ:smooth-spectral-density}
@@
 \begin{equation}
-    g_{\sigma}(s) = \frac{1}{n \sigma \sqrt{2\pi}} e^{-\sfrac{s^2}{2\sigma^2}}.
+    g_{\sigma}(s) = \frac{1}{\sigma \sqrt{2\pi}} e^{-\sfrac{s^2}{2\sigma^2}}.
     \label{equ:smoothing-kernel}
 \end{equation}
-The task in \refequ{equ:smooth-spectral-density} is to approximate the trace of the parameter-dependent matrix $g_{\sigma}(t\mtx{I}_n - \mtx{A})$, which allows us to apply the results derived in \refsec{sec:analysis}.
+The task in \refequ{equ:smooth-spectral-density} is to approximate the trace of the parameter-dependent matrix $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
+Hence, we can apply the results derived in \refsec{sec:analysis}.
 
-\todo{
-Measuring error between spectral density $\phi$ and approximation $\widetilde{\phi}$
+Choosing the smoothing parameter $\sigma$ results in the following trade off: Typically, larger $\sigma$ make it easier to approximate the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$, whereas smaller $\sigma$ allow us to stay closer to the original spectral density \refequ{equ:spectral-density}.
+To see the latter, we measure the error between spectral density $\phi$ and its smoothened version $\phi_{\sigma}$ with
 \begin{equation}
-    \sup_{f \in \mathcal{S}} \int_{-1}^{1} f(t) (\phi(t) - \widetilde{\phi}(t)) \mathrm{d}t
+    \sup_{f \in \mathcal{S}} \int_{-1}^{1} f(t) (\phi(t) - \phi_{\sigma}(t))~\mathrm{d}t,
     \label{equ:error-metric}
 \end{equation}
-
-In \cite{lin-2016-approximating-spectral} propose $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - \lambda), \lambda \in [\lambda_{\text{min}}, \lambda_{\text{max}}]\}$
-
-In \cite{chen-2021-analysis-stochastic} use $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [-1, 1] \}$ where $\Theta$ the unit step which is equivalent to Kolmogorov-Smirnov.
-
-In \cite{braverman-2022-sublinear-time} use $\mathcal{S} = \text{Lip}_1 = \{f : |f(x) - f(y)| \leq |x - y| \}$ which is equivalent to the Wasserstein-1 norm.
-
-For small enough $\sigma$ and well separated eigenvalues, the  error \refequ{equ:error-metric} when testing against Lipschitz-1 functions we make by smoothing is already $\mathcal{\sigma}$.
-
-Having noted this, we will from now on measure errors in $L^1$ norm from smooth spectral density.
-}
-
-\todo{Discuss error measures and baseline smoothing error, choice of $\sigma$}
+where $\mathcal{S}$ represents an appropriately chosen space of test functions.
+Among others, \cite{lin-2016-approximating-spectral} proposes $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - \lambda), \lambda \in [\lambda_{\text{min}}, \lambda_{\text{max}}]\}$, \cite{chen-2021-analysis-stochastic} uses $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [-1, 1] \}$ where $\Theta$ is the Heaviside step function, and \cite{braverman-2022-sublinear-time} uses $\mathcal{S} = \{f : |f(x) - f(y)| \leq |x - y| \}$ for which \refequ{equ:error-metric} is equivalent to the Wasserstein-1 distance between $\phi$ and $\phi_{\sigma}$.
+In this last metric, if we assume the spectrum to be somewhat uniform and $\sigma$ to be rather small, a simple calculation allows us to roughly estimate the smoothing error \refequ{equ:error-metric} to be $\sigma$.
+Consequently, if we can only allow an approximation to deviate by at most a factor of $\varepsilon > 0$ from the original spectral density, we should choose a smoothing parameter $\sigma \approx \varepsilon$.
 
@@
 
-To estimate the trace in \refequ{equ:smooth-spectral-density}, we first need to evaluate the matrix function.
-However, doing this exactly usually requires us to first diagonalize $\mtx{A}$, which is often a prohibitively expensive operation.
-For algorithmic reasons, we seek for approximations of the matrix function in terms of affine expansions, of which a Chebyshev expansion is most suitable due to its approximation error guarantees and ease of manipulation \todo{more specific}.
+To estimate the trace in \refequ{equ:smooth-spectral-density}, we first need to evaluate the involved matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
+However, doing this exactly would require us to first diagonalize $\mtx{A}$, which is often a prohibitively expensive operation.
+For algorithmic reasons, we seek for approximations of this matrix function in terms of affine expansions, of which the Chebyshev expansion is most suitable due to its approximation error guarantees and ease of manipulation.
 In every $t \in \mathbb{R}$ we expand
@@
     \begin{cases}
-        s^l, & \text{if $l \in \{0, 1\}$} \\
+        s^l, & \text{if $l \in \{0, 1\}$}; \\
         T_l(s) = 2 s T_{l-1}(s) - T_{l-2}(s), & \text{if $l \geq 2$.}
@@
 \end{equation}
-\todo{Spectral transform (mention estimation of $[a,b]$ see notes, and maybe some buffer region to avoid cut-off).}
-
-This allows us to define the expanded spectral density
+With the help of \refequ{equ:matrix-expansion} we can now define the expanded spectral density
 \begin{equation}
-    \phi_{\sigma}^{(m)}(t) = \Trace(g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})).
+    \phi_{\sigma}^{(m)}(t) =  \frac{1}{n} \Trace(g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})).
     \label{equ:expanded-spectral-density}
 \end{equation}
+
+However, the Chebyshev polynomials are only well-defined for matrices whose spectrum is contained in $[-1, 1]$.
+When working with general matrices of bounded spectrum, we will first have to estimate a lower bound $a$ and upper bound $b$, on the spectrum \todo{[cite Saad, estimate spectrum]}.
+A spectral transform
+\begin{equation}
+    \begin{cases}
+        \tau : [a, b] \to [-1, 1],\\
+        \tau(t) = \frac{2t - a - b}{b - a}.
+    \end{cases}
+\end{equation}
+then lets us define the matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ whose spectrum is identical to the one of $\mtx{A}$ but shrunk to the interval $[-1, 1]$, for which \refequ{equ:matrix-expansion} is a valid expansion.
+Note that through this transformation we will also have to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when reverting the spectral transform $\tau$.
 
@@
 
+It is well known that the coefficients of a Chebyshev expansion of a function can be computed which a discrete cosine transform (DCT) of the function evaluations at the Chebyshev nodes $s_i = \cos(\pi i / m)$ for $i=0,\dots,m$
+\begin{equation}
+    g_{\sigma}(t - \cos(s_i)) = \sum_{l=0}^{m} \mu_l(t) \cos\left(\frac{\pi i l}{m} \right).
+    \label{equ:discrete-cosine-transform}
+\end{equation}
+This can be accomplished in $\mathcal{O}(m \log(m))$ operations.
+
+A highly useful property of the Chebyshev expansion -- of which we will make use twice in the following pages -- is the fact that we can (algebraically) square such an expansion in $\mathcal{O}(m \log(m))$ operations, which will be shown in the below algorithm.
+
+\begin{algo}{Fast squaring of Chebyshev expansion}{chebyshev-squaring}
+    \begin{algorithmic}[1]
+        \Statex \textbf{Input:} Coefficients $\{ \mu_l \}_{l=0}^{m}$ of the expansion $\sum_{l=0}^{m} \mu_l T_l$
+        \Statex \textbf{Output:} Coefficients $\{ \nu_l \}_{l=0}^{2m}$ such that $\sum_{l=0}^{2m} \nu_l T_l = (\sum_{l=0}^{m} \mu_l T_l)^2$
+        \State Define $\vct{\mu} \in \mathbb{R}^{2m + 1}$ whose first $m + 1$ entries are the coefficients $\mu_l$ and the remaining $m$ are $0$
+    \State Compute $\vct{f} = \DCT(\vct{\mu})$
+    \State Compute $\vct{\nu} = \DCT^{-1}(\vct{f}^{2})$ whose entries are the coefficients $\nu_l$
+    \end{algorithmic}
+    \end{algo}
+
 \todo{
-Fast computation of expansion with DCT (duality with evaluations)
-}
+consistent and affine Nyström approximation
 
-Squaring of Chebyshev expansion for consistent and affine Nyström approximation
-\begin{equation}
-    \left( \sum_{l=0}^{m} \mu_l T_l(t) \right)^2 = \sum_{l=0}^{2m} \nu_l T_l(t)
-    \label{equ:squared-chebyshev-expansion}
-\end{equation}
 where $\nu = \DCT^{-1} \left\{ \DCT\{\mu\}^2 \right\}$ \todo{which is consistent unlike \cite{lin-2017-randomized-estimation}}.
 
-Non-negative Chebyshev expansion for results in Nyström approximation of positive semi-definite matrices (discuss alternatives: Jackson damping, Putinar representation)
+Non-negative Chebyshev expansion for results in Nyström approximation of positive semi-definite matrices (discuss alternatives: Jackson damping, Putinar representation), or other techniques (Nakatsukasa, shift, ...) [Maybe mention those later]
+}
 
@@
 \begin{lemma}{Chebyshev expansion error}{chebyshev-error}
-    The expansion $g_{\sigma}^{(m)}$ of the Gaussian smoothing kernel $g_{{\sigma}}$ satisfies
+    The expansion $g_{\sigma}^{(m)}$ of the Gaussian smoothing kernel $g_{{\sigma}}$ \refequ{equ:smoothing-kernel} satisfies
     \begin{equation}
-        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \right| \leq \sqrt{\frac{2e}{\pi}} \frac{1}{n \sigma^2} (1 + \sigma)^{-m} \equiv \frac{E_{\sigma, m}}{n}
+        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \right| \leq \sqrt{\frac{2e}{\pi}} \frac{1}{\sigma^2} (1 + \sigma)^{-m} \equiv E_{\sigma, m}
         \label{equ:2-chebyshev-interpolation-sup-error-kernel}
@@
     \end{equation}
-    for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and $s \in [-1, 1]$.
+    for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, $t \in \mathbb{R}$, and $s \in [-1, 1]$.
 \end{lemma}
@@
     | g_{\sigma}(t - z) | 
-    = \frac{1}{n \sigma \sqrt{2 \pi}} \left| e^{-\sfrac{(t - z)^2}{2\sigma^2}} \right|
-    = \frac{1}{n \sigma \sqrt{2 \pi}} e^{-\sfrac{(t - x)^2}{2\sigma^2}}e^{\sfrac{y^2}{2\sigma^2}}.
+    = \frac{1}{\sigma \sqrt{2 \pi}} \left| e^{-\sfrac{(t - z)^2}{2\sigma^2}} \right|
+    = \frac{1}{\sigma \sqrt{2 \pi}} e^{-\sfrac{(t - x)^2}{2\sigma^2}}e^{\sfrac{y^2}{2\sigma^2}}.
     %\leq \frac{1}{n \sigma \sqrt{2 \pi}} \max_{x + iy \in \mathcal{E}_{\chi}} e^{\sfrac{y^2}{2\sigma^2}} 
@@
         \max_{z \in \mathcal{E}_{\chi}} | g_{\sigma}(t - z) | 
-        \leq \frac{1}{n \sigma \sqrt{2 \pi}} e^{\sfrac{(\chi - \chi^{-1})^2}{8 \sigma^2}} 
+        \leq \frac{1}{\sigma \sqrt{2 \pi}} e^{\sfrac{(\chi - \chi^{-1})^2}{8 \sigma^2}} 
     \end{equation}
@@
     \begin{equation}
-        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \right| \leq \sqrt{\frac{2e}{\pi}} \frac{1}{n \sigma^2} (1 + \sigma)^{-m}
+        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \right| \leq \sqrt{\frac{2e}{\pi}} \frac{1}{\sigma^2} (1 + \sigma)^{-m}
     \end{equation}
@@
     \begin{equation}
-        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \leq 2\sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right) \frac{E_{\sqrt{2}\sigma, \sfrac{m}{2}}}{n} \equiv \frac{\underline{E}_{\sigma, m}}{n}
+        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \leq 2\sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right) E_{\sqrt{2}\sigma, \sfrac{m}{2}} \equiv \underline{E}_{\sigma, m}
         \label{equ:2-chebyshev-interpolation-sup-error-kernel}
@@
     \end{equation}
-    for all \emph{even} degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and $s \in [-1, 1]$.
+    for all \emph{even} degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, $t \in \mathbb{R}$, and $s \in [-1, 1]$.
 \end{lemma}
@@
     \end{equation}
-    Since $\sqrt{g_{\sigma}} = \sqrt{2 n \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$ and $\sqrt{g_{\sigma}} \leq 1/\sqrt{n \sigma \sqrt{2 \pi}}$ we can apply the result from \refthm{thm:chebyshev-error} to get
+    Since $\sqrt{g_{\sigma}} = \sqrt{2 \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$ and $\sqrt{g_{\sigma}} \leq 1/\sqrt{\sigma \sqrt{2 \pi}}$ we can apply the result from \refthm{thm:chebyshev-error} to get
     \begin{align}
         &\sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \notag \\
-        &\leq \left( 2 \frac{1}{\sqrt{n \sigma \sqrt{2\pi}}} + \sqrt{2 n \sigma \sqrt{2 \pi}} \cdot \frac{E_{\sqrt{2}\sigma, \sfrac{m}{2}}}{n}\right)\sqrt{2 n \sigma \sqrt{2 \pi}} \cdot \frac{E_{\sqrt{2}\sigma, \sfrac{m}{2}}}{n},
+        &\leq \left( 2 \frac{1}{\sqrt{ \sigma \sqrt{2\pi}}} + \sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right)\sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}},
     \end{align}
@@
     \end{equation}
-    for all degrees $m \in \mathbb{N}$ and smoothing parameters $\sigma > 0$.
+    for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and $t \in \mathbb{R}$.
 The same result holds for the expansion of a non-negative Chebyshev expansion of \emph{even} degree $m \in \mathbb{N}$ but with $\underline{E}_{\sigma, m}$.
@@
         &\left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right| \notag \\
-        &= \left| \Trace(g_{\sigma}(t\mtx{I}_n - \mtx{A})) - \Trace(g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})) \right|
+        &= \left| \frac{1}{n} \Trace(g_{\sigma}(t\mtx{I}_n - \mtx{A})) - \frac{1}{n} \Trace(g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})) \right|
         && \text{(definitions \refequ{equ:smooth-spectral-density} and \refequ{equ:expanded-spectral-density})} \notag \\
-        &= \left| \sum_{i=1}^n \left(g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i)\right) \right|
+        &= \left| \frac{1}{n} \sum_{i=1}^n \left(g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i)\right) \right|
         && \text{($\lambda_1, \dots, \lambda_n$ eigenvalues of $\mtx{A}$)} \notag \\
-        &\leq n \max_{i = 1, \dots, n} \left| g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i) \right|
+        &\leq \max_{i = 1, \dots, n} \left| g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i) \right|
         && \text{(conservative upper bound)} \notag \\
@@
     \end{align}
-    Then, Hölder's inequality \cite{} allows us to conclude%\cite{klenke2013probability} 
+    Then, Hölder's inequality allows us to conclude
     \begin{equation}

### 2024-05-28 paper/main.tex
continue writing

@@
 
-In numerous problems across physics, chemistry, engineering, and data science, eigenvalues and especially their distribution are crucial for understanding the properties of a system: In electronic structure calculations they represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}.
-The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be captured in the spectral density
+In numerous problems across physics, chemistry, engineering, and data science, eigenvalues, particularly their distribution, are crucial for understanding the properties of a system: In electronic structure calculations they represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}.
+The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be represented by the spectral density
 \begin{equation}
@@
 \begin{itemize}
-    \item Brief definition of parameter-dependent matrix ($\mtx{B}(t) \in \mathbb{R}^{n \times n}$, every entry depends continuously on $t \in [a,b]$ \cite{kressner-2023-randomized-lowrank})
+    \item Brief definition of parameter-dependent matrix ($\mtx{A}(t) \in \mathbb{R}^{n \times n}$, every entry depends continuously on $t \in [a,b]$ \cite{kressner-2023-randomized-lowrank})
     \item Trace estimation through mat-vecs (if matrix is only accessed implicitly \cite{pearlmutter-1994-fast-exact} or for complicated matrix functions [partition function, spectral density, exponential integrators])
     \item Existing trace estimation methods for constant matrices: Girard-Hutchinson \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} and variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making} which have only recently been specialized to trace estimation of matrix functions \cite{chen-2023-krylovaware-stochastic}
-    \item Trace estimation in the parameter-dependent matrices (Nyström++ but no analysis \cite{lin-2017-randomized-estimation}, only low-rank approximation \cite{kressner-2023-randomized-lowrank}, dynamic trace estimation for sequence of similar matrices \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace})
+    \item Trace estimation in the parameter-dependent matrices (Nyström++ but no analysis \cite{lin-2017-randomized-estimation}, only low-rank approximation \cite{kressner-2023-randomized-lowrank}, dynamic trace estimation for sequence $\mtx{A}_1, \dots, \mtx{A}_m$ of similar matrices \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace}) their algorithm provides trace approximations $a_1, \dots , a_m$ for $\Trace(\mtx{A}_1), \dots, \Trace(\mtx{A}_m)$ such that, for each $i \in 1, \ldots, m$,
+    \begin{equation}
+        \mathbb{P}(|a_i - \Trace(\mtx{A}_i)| \geq \epsilon)\leq \delta.
+    \end{equation}
+    drawback of this type of guarantee is that when we aim for $\epsilon$ accuracy simultaneously for all $|a_i-\Trace(\mtx{A}_i)|$, the probability will grow linearly with $m$ by the union bound.
+On the other hand, in our setting, instead of discretizing $[a, b]$ with $t_1, \ldots, t_m$ and looking at the error $|a(t_i)-\Trace(\mtx{A}(t_i))|$ for each $i$, a much more natural way to quantify the error would be to look at $\int_{a}^{b} | a(t) - \mtx{A}(t) |~\mathrm{d}t$, where $a(t)$ is the trace estimation in each $t \in [a, b]$.
     %\item Error analysis for spectral density approximation (Wasserstein-1 error [Braverman, Chen], missing $1/\varepsilon$ guarantees)
@@
     \item Theoretical guarantees in the $L^1$-norm for three established trace estimators for constant matrices when applied to parameter-dependent matrices (Girard-Hutchinson, Nyström, Nyström++), in particular, $\mathcal{O}(\varepsilon^{-1})$ scaling with number of mat-vecs
-    \item A simple and efficient algorithm for estimating spectral densities using any of the three estimators using Chebyshev expansion, including theoretical guarantees
+    \item A simple and efficient algorithm for estimating spectral densities using any of the three estimators with Chebyshev expansion, including theoretical guarantees
 \end{itemize}
@@
     \Nystr{\mtx{\Omega}}{\mtx{A}}(t) = (\mtx{A}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} (\mtx{A}(t) \mtx{\Omega})^{\top}.
+    \label{equ:nystrom-approximation}
 \end{equation}
@@
 \end{equation}
-which has originally been proposed in \cite{lin-2017-randomized-estimation}.
+which was originally proposed by \cite{lin-2017-randomized-estimation}.
 %We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.
@@
 \begin{remark}
-    If additionally $\mtx{A}(t)$ is nonzero for all $t \in [a, b]$, then it follows from the proof of \refthm{thm:hutchinson} that with probability at least $1 - \delta$
+    If additionally $\mtx{A}(t)$ is nonzero and symmetric positive semi-definite for all $t \in [a, b]$, then it follows from the proof of \refthm{thm:hutchinson} that with probability at least $1 - \delta$
     \begin{equation}
@@
         &= \delta^{-\sfrac{2}{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} \frac{1}{\sqrt{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} && \text{(definition of $\delta$ and choice of $k$)} \notag \\
-        &= e^{\sfrac{2}{\varepsilon^{-2}} \frac{\varepsilon^{-2}\log(\delta^{-1})}{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} \frac{1}{\sqrt{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} && \text{($\delta = e^{-\log(\delta^{-1})}$)} \notag \\
+        &= e^{2\varepsilon^{2} \frac{\varepsilon^{-2}\log(\delta^{-1})}{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} \frac{1}{\sqrt{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} && \text{($\delta = e^{-\log(\delta^{-1})}$)} \notag \\
         &\leq e^{2\varepsilon^{2}} \frac{\varepsilon}{\sqrt{ \log(\delta^{-1})}} && \text{($\lceil x \rceil \geq x$ if $x \geq 0$)}
@@
 
-\todo{
-consistent and affine Nyström approximation
-
-where $\nu = \DCT^{-1} \left\{ \DCT\{\mu\}^2 \right\}$ \todo{which is consistent unlike \cite{lin-2017-randomized-estimation}}.
-
-Non-negative Chebyshev expansion for results in Nyström approximation of positive semi-definite matrices (discuss alternatives: Jackson damping, Putinar representation), or other techniques (Nakatsukasa, shift, ...) [Maybe mention those later]
-}
-
 \subsubsection{Expansion error}
@@
 
-We analyze the error we commit when working with the expanded spectral density \refequ{equ:expanded-spectral-density}. To this extent, we first analyze the expansion error of the smoothing kernel.
+We now analyze the error of the Chebyshev expansion of a Gaussian smoothing kernel $g_{\sigma}$.
 
@@
 The proof specializes Bernstein's theorem for Chebyshev interpolation to the case of a Gaussian smoothing kernel \refequ{equ:smoothing-kernel}.
-\todo{It loosely follows the proof give in the preprint of \cite[theorem 2]{lin-2017-randomized-estimation}, but reaches slightly different conclusion.}
+It loosely follows the proof give in the preprint of \cite[theorem 2]{lin-2017-randomized-estimation}, but reaches slightly different conclusion.
 
@@
         \max_{z \in \mathcal{E}_{\chi}} | g_{\sigma}(t - z) | 
-        \leq \frac{1}{\sigma \sqrt{2 \pi}} e^{\sfrac{(\chi - \chi^{-1})^2}{8 \sigma^2}} 
+        \leq \frac{1}{\sigma \sqrt{2 \pi}} e^{\sfrac{(\chi - \chi^{-1})^2}{8 \sigma^2}}.
     \end{equation}
@@
     \begin{equation}
-        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \right| \leq \sqrt{\frac{2e}{\pi}} \frac{1}{\sigma^2} (1 + \sigma)^{-m}
+        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \right| \leq \sqrt{\frac{2e}{\pi}} \frac{1}{\sigma^2} (1 + \sigma)^{-m}.
     \end{equation}
@@
 
-\todo{
-Explain need for non-negativity for Nyström approximation (for positive semi-definiteness in theorem).
-
-Converges too slow: polynomial interpolation with Jackson damping \cite{jackson-1912-approximation-trigonometric,braverman-2022-sublinear-time}
-
-Non-negative Chebyshev expansion: expand $\sqrt{g_{\sigma}}$ to degree $m/2$ and then square the expansion. Error guarantee very comparable to normal expansion, and proof below.
-
-Side-note: it is possible to refine the coefficients of the Chebyshev expansion to get a better approximation in the supremum-norm over the set of all non-negative polynomials \cite{fejer-1916-uber-trigonometrische}.
-The non-convex nature of this problem and the relatively small improvement we get with this technique made us decide to not pursue this further.
-}
-
-\begin{lemma}{Non-negative Chebyshev expansion error}{non-negative-chebyshev-error}
-    The non-negative expansion $\underline{g}_{\sigma}^{(m)}$ of the Gaussian smoothing kernel satisfies
-    \begin{equation}
-        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \leq 2\sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right) E_{\sqrt{2}\sigma, \sfrac{m}{2}} \equiv \underline{E}_{\sigma, m}
-        \label{equ:2-chebyshev-interpolation-sup-error-kernel}
-        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
-        %\label{equ:2-chebyshev-interpolation-error}
-    \end{equation}
-    for all \emph{even} degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, $t \in \mathbb{R}$, and $s \in [-1, 1]$.
-\end{lemma}
-
-\begin{proof}
-    For any numbers $a, b \in \mathbb{R}$ it holds
-    \begin{equation}
-    | a^2 - b^2 | = | (a + b)(a - b) | = | a + b | | a - b | \leq (2 | a | + | a - b |)  | a - b |
-    \end{equation}
-    Therefore, omitting the arguments and noting $g_{\sigma} = (\sqrt{g_{\sigma}})^2$ and $\underline{g}_{\sigma}^{(m)} = (\sqrt{g_{\sigma}}^{(\sfrac{m}{2})})^2$.
-    we have
-    \begin{equation}
-        \left| g_{\sigma} - \underline{g}_{\sigma}^{(m)} \right| \leq \left( 2 \left| \sqrt{g_{\sigma}} \right| + \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right| \right) \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right|.
-    \end{equation}
-    Since $\sqrt{g_{\sigma}} = \sqrt{2 \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$ and $\sqrt{g_{\sigma}} \leq 1/\sqrt{\sigma \sqrt{2 \pi}}$ we can apply the result from \refthm{thm:chebyshev-error} to get
-    \begin{align}
-        &\sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \notag \\
-        &\leq \left( 2 \frac{1}{\sqrt{ \sigma \sqrt{2\pi}}} + \sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right)\sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}},
-    \end{align}
-    from which follows the result with some minor simplifications.
-\end{proof}
-
-The
+\Reflem{lem:chebyshev-error} now also allows us to bound the error we make with the expanded spectral density $\phi_{\sigma}^{(m)}$ \refequ{equ:expanded-spectral-density}.
 
@@
     for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and $t \in \mathbb{R}$.
-The same result holds for the expansion of a non-negative Chebyshev expansion of \emph{even} degree $m \in \mathbb{N}$ but with $\underline{E}_{\sigma, m}$.
+%The same result holds for the expansion of a non-negative Chebyshev expansion of \emph{even} degree $m \in \mathbb{N}$ but with $\underline{E}_{\sigma, m}$.
 \end{theorem}
@@
         && \text{(conservative upper bound)} \notag \\
-        &\leq E_{\sigma, m} ~ \text{(or $\underline{E}_{\sigma, m}$)}.
-        && \text{(using \refthm{thm:chebyshev-error})} \notag \\
+        &\leq E_{\sigma, m}.% ~ \text{(or $\underline{E}_{\sigma, m}$)}.
+        && \text{(using \reflem{lem:chebyshev-error})} \notag \\
     \end{align}
@@
         \leq 2 \sup_{t \in [-1, 1]} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|
-        \leq 2 E_{\sigma, m} ~ \text{(or $2 \underline{E}_{\sigma, m}$)}.
+        \leq 2 E_{\sigma, m}.% ~ \text{(or $2 \underline{E}_{\sigma, m}$)}.
     \end{equation}
@@
 
+Our goal is to apply the trace estimator analyzed in \refsec{subsec:nystrom-pp} to the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ to approximate the smooth spectral density $\phi_{\sigma}$.
+However, due to the oscillatory nature of the Chebyshev expansion, we can no longer guarantee that $g_{\sigma}^{(m)} \geq 0$, and consequently, the matrix function $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ may be indefinite.
+To still apply \refthm{thm:nystrom-pp} in this scenario, we will have to ensure that the Chebyshev expansion is non-negative.
+We can think of many ways in which this can be achieved: by using Jackson damping \cite{jackson-1912-approximation-trigonometric,braverman-2022-sublinear-time}, albeit at a significantly worse rate of convergence; by shifting up the expansion $g_{\sigma}^{(m)} + \rho$ with a large enough $\rho$, at the loss of most of the low-rank structure of $g_{\sigma}(t \mtx{I}_n - \mtx{A})$; or by directly minimizing the approximation error over all non-negative Chebyshev polynomials \cite{fejer-1916-uber-trigonometrische}, which realistically only works for small $m$.
+In the end, we propose the following algorithm to compute a non-negative Chebyshev expansion.
+First, expand the square root of the smoothing kernel in the Chebyshev polynomials up to degree $m/2$
+\begin{equation}
+    \sqrt{g_{\sigma}(t - s)}^{(\sfrac{m}{2})} = \sum_{l=0}^{m/2} \xi_l(t) T_l(s).
+\end{equation}
+Second, square this expansion with \refalg{alg:chebyshev-squaring} to obtain the non-negative Chebyshev expansion
+\begin{equation}
+    \underline{g}^{(m)}_{\sigma}(t - s) = \sum_{l=0}^{m} \mu_l(t) T_l(s).
+\end{equation}
+With a complexity of $\mathcal{O}(m \log(m))$ this procedure is computationally inexpensive and allows for very similar convergence bounds to \reflem{lem:chebyshev-error}, which we will derive in \refsec{subsubsec:chebyshev-nystrom-analysis}.
+
+\todo{
+Maybe mention indefinite Nyström [Nakatsukasa], which won't give us desirable guarantees, and refinement of Chebyshev expansion [Francis Bach], which only slightly improves, but is much harder to implement.
+Side-note: it is possible to refine the coefficients of the Chebyshev expansion to get a better approximation in the supremum-norm over the set of all non-negative polynomials \cite{fejer-1916-uber-trigonometrische}.
+The non-convex nature of this problem and the relatively small improvement we get with this technique made us decide to not pursue this further.
+}
+
+In the end, we propose the following approximation for the smooth spectral density
+\begin{equation}
+    \widetilde{\phi}_{\sigma}^{(m)}(t) = \frac{1}{n} \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})).
+    \label{equ:chebyshev-nystrom-formula}
+\end{equation}
+
 \subsubsection{Implementation}
@@
 
-An efficient implementation can be achieved thanks to the recurrence relation \refequ{equ:chebyshev-recurrence} which the Chebyshev polynomials satisfy.
-However, it is usually prohibitively expensive to interpolate a large matrix $\mtx{A}$ as a whole, since alone the matrix-matrix multiplication in each step of the recurrence can cost up to $\mathcal{O}(n^3)$, and the evaluation of the expansion at $n_t$ values of the parameter could cost a further $\mathcal{O}(n^2 m n_t)$ operations.
-Therefore, in case we are only interested in the result of a linear mapping applied to the interpolant, a significant speed-up can be achieved by directly interpolating the result of this linear mapping applied to the interpolant.
-This is done by using the invariance of the trace under cyclic permutation of its arguments.
-\todo{Illustrate this better.}
+An efficient implementation of \refequ{equ:chebyshev-nystrom-formula} can be achieved thanks to the invariance of the trace under cyclic permutation of its arguments and the recurrence relation \refequ{equ:chebyshev-recurrence} which the Chebyshev polynomials satisfy.
+To illustrate this, consider the estimator \refequ{equ:chebyshev-nystrom-formula} when $\mtx{\Psi}$ is zero.
+Using the definition of the Nyström approximation \refequ{equ:nystrom-approximation}, the cyclic invariance of the trace, and the symmetry of $\mtx{A}$, we can rewrite it as
+\begin{equation}
+    \frac{1}{n} \Trace((\mtx{\Omega}^{\top} \underline{g}^{(m)}_{\sigma}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega})^{\dagger} (\mtx{\Omega}^{\top} \underline{g}^{(m)}_{\sigma}(t \mtx{I}_n - \mtx{A})^2 \mtx{\Omega})).
+    \label{equ:cyclic-property}
+\end{equation}
+Thanks to the affine form of the Chebyshev expansion, we can rewrite the first matrix as
+\begin{equation}
+    \mtx{\Omega}^{\top} \underline{g}^{(m)}_{\sigma}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega} = \sum_{l=0}^{m} \mu_l(t) (\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega})
+\end{equation}
+for which we first precompute the $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ for $l=0, \dots, m$ using the recurrence \refequ{equ:chebyshev-recurrence} and then inexpensively sum them up with the corresponding coefficients $\mu_l(t)$ for all $t$ in which we want to evaluate the approximation.
+Thanks to \refalg{alg:chebyshev-squaring}, the second part of \refequ{equ:cyclic-property} can be approximated in an equivalent way \todo{(which is consistent, unlike \cite{lin-2017-randomized-estimation})}.
 
@@
     \Statex \textbf{Output:} Approximate evaluations of the spectral density $\{\phi_{\sigma}(t_i)\}_{i=1}^{n_t}$
-    \State Compute $\{\mu_l(t_i)\}_{l=0}^{m}$ and $\{\nu_l(t_i)\}_{l=0}^{2m}$ for all $t_i$ %using %\refalg{alg:2-chebyshev-chebyshev-expansion}
+    \State Compute $\{\mu_l(t_i)\}_{l=0}^{m}$ for all $t_i$ using the $\DCT$ \refequ{equ:discrete-cosine-transform}
+    \State Compute $\{\nu_l(t_i)\}_{l=0}^{2m}$ using \refalg{alg:chebyshev-squaring}%\refalg{alg:2-chebyshev-chebyshev-expansion}
     %\State Compute  for all $t_i$ using %\refalg{alg:3-nystrom-chebyshev-exponentiation}
@@
     \For {$i = 1, \dots, n_t$}
-      \State $\phi_{\sigma}(t_i) \gets \Trace\left( \mtx{K}_1(t_i)^{\dagger}\mtx{K}_2(t_i) \right) + \frac{1}{n_{\mtx{\Psi}}} \left( \ell(t_i) + \Trace\left( \mtx{L}_1(t_i)^{\top} \mtx{K}_1(t_i)^{\dagger} \mtx{L}_1(t_i) \right)  \right) $ \label{lin:4-nystromchebyshev-nystrom-pp}
+      \State $\phi_{\sigma}(t_i) \gets \frac{1}{n} \Trace\left( \mtx{K}_1(t_i)^{\dagger}\mtx{K}_2(t_i) \right) + \frac{1}{n n_{\mtx{\Psi}}} \left( \ell(t_i) + \Trace\left( \mtx{L}_1(t_i)^{\top} \mtx{K}_1(t_i)^{\dagger} \mtx{L}_1(t_i) \right)  \right) $ \label{lin:4-nystromchebyshev-nystrom-pp}
     \EndFor
@@
 
+\begin{lemma}{Non-negative Chebyshev expansion error}{non-negative-chebyshev-error}
+    The non-negative expansion $\underline{g}_{\sigma}^{(m)}$ of the Gaussian smoothing kernel satisfies
+    \begin{equation}
+        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \leq 2\sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right) E_{\sqrt{2}\sigma, \sfrac{m}{2}} \equiv \underline{E}_{\sigma, m}
+        \label{equ:2-chebyshev-interpolation-sup-error-kernel}
+        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
+        %\label{equ:2-chebyshev-interpolation-error}
+    \end{equation}
+    for all \emph{even} degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, $t \in \mathbb{R}$, and $s \in [-1, 1]$.
+\end{lemma}
+
+\begin{proof}
+    For any numbers $a, b \in \mathbb{R}$ it holds
+    \begin{equation}
+    | a^2 - b^2 | = | (a + b)(a - b) | = | a + b | | a - b | \leq (2 | a | + | a - b |)  | a - b |
+    \end{equation}
+    Therefore, omitting the arguments and using $g_{\sigma} = (\sqrt{g_{\sigma}})^2$ and $\underline{g}_{\sigma}^{(m)} = (\sqrt{g_{\sigma}}^{(\sfrac{m}{2})})^2$.
+    we have
+    \begin{equation}
+        \left| g_{\sigma} - \underline{g}_{\sigma}^{(m)} \right| \leq \left( 2 \left| \sqrt{g_{\sigma}} \right| + \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right| \right) \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right|.
+    \end{equation}
+    Since $\sqrt{g_{\sigma}} = \sqrt{2 \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$ and $\sqrt{g_{\sigma}} \leq 1/\sqrt{\sigma \sqrt{2 \pi}}$ we can apply the result from \refthm{thm:chebyshev-error} to get
+    \begin{align}
+        &\sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \notag \\
+        &\leq \left( 2 \frac{1}{\sqrt{ \sigma \sqrt{2\pi}}} + \sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right)\sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}},
+    \end{align}
+    from which follows the result with some minor simplifications.
+\end{proof}
+
 \begin{theorem}{Chebyshev-Nyström++ method}{chebyshev-nystrom}
     Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the Chebyshev-Nyström++ estimate $\underline{\widetilde{\phi}}_{\sigma}^{(m)}$ with \emph{even} degree $m \in \mathbb{N}$, $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2)$, and smoothing parameter $\sigma > 0$ satisfies with probability at least $1 - \delta$
+Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m \in \mathbb{N}$, $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2)$, and smoothing parameter $\sigma > 0$ satisfies with probability at least $1 - \delta$
     \begin{equation}
-        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\widetilde{\phi}}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t < 2 (1 + \varepsilon) \underline{E}_{\sigma, m} + \varepsilon
+        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t < 2 (1 + \varepsilon) \underline{E}_{\sigma, m} + \varepsilon
         \label{equ:chebyshev-nystrom-error}
@@
     \begin{align}
-        &\int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\widetilde{\phi}}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \notag \\
-        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \int_{-1}^{1} \left| \underline{\phi}^{(m)}_{\sigma}(t) - \underline{\widetilde{\phi}}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
-        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \varepsilon \int_{-1}^{1} \left| \underline{\phi}^{(m)}_{\sigma}(t) \right|~\mathrm{d}t && \text{(\refthm{thm:nystrom-pp} w.p.
+        &\int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \notag \\
+        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \int_{-1}^{1} \left| \phi^{(m)}_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
+        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \varepsilon \int_{-1}^{1} \left| \phi^{(m)}_{\sigma}(t) \right|~\mathrm{d}t && \text{(\refthm{thm:nystrom-pp} w.p.
 $\geq 1 - \delta$)} \notag \\
-        &\leq (1 + \varepsilon) \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \varepsilon \int_{-1}^{1} \left| \phi_{\sigma}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
+        &\leq (1 + \varepsilon) \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \varepsilon \int_{-1}^{1} \left| \phi_{\sigma}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
         &\leq 2 (1 + \varepsilon) \underline{E}_{\sigma, m} + \varepsilon. && \text{(\reflem{lem:non-negative-chebyshev-error} and normalization)} \notag \\

### 2024-05-28 paper/main.tex
rewrite introduction

@@
 
-In numerous problems across physics, chemistry, engineering, and data science, eigenvalues, particularly their distribution, are crucial for understanding the properties of a system: In electronic structure calculations they represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}.
-The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be represented by the spectral density
-\begin{equation}
-    \phi(t) = \frac{1}{n} \sum_{i=1}^{n} \delta(t - \lambda_i).
-\end{equation}
+In numerous problems across physics, chemistry, engineering, and data science, how the eigenvalues of certain matrices are distributed can be indicative of what properties a system has: In electronic structure calculations they represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}.
+The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$.
 Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation which is often prohibitively expensive.
@@
 For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature rule from the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}.
-In many applications, it is also acceptable to trade in some of the finer details of the spectrum for a better approximability by smoothing $\phi$, typically through convolution with a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}.
-This converts the problem to estimating the trace of a smooth parameter-dependent matrix.
+Further, $\delta$ is hard to approximate and in many applications, a rough estimate of the spectral density is already acceptable.
+One standard approach is to instead approximate the smooth spectral density
+\begin{equation}
+    \phi_{\sigma}(t) = \frac{1}{n} \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i)
+\end{equation}
+for a smoothing kernel $g_{\sigma}$, typically a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}.
+This allows the problem to be framed in the setting of trace estimation of parameter-dependent matrices.
 
@@
 \begin{itemize}
+    \item Same random matrix for all $t$, which allows for favorable scaling in the number of evaluation points
     \item Theoretical guarantees in the $L^1$-norm for three established trace estimators for constant matrices when applied to parameter-dependent matrices (Girard-Hutchinson, Nyström, Nyström++), in particular, $\mathcal{O}(\varepsilon^{-1})$ scaling with number of mat-vecs
@@
     \end{align}
-    By \cite{} $\widetilde{\mtx{\Omega}}_2 \in \mathbb{R}^{2k \times l}$ satisfies the \todo{$(c_1 \sqrt{q/2k}, c_2 e^{-q}, q)$}-JL moment property for \todo{any $q$ (in what?) $\geq \dots$} and some constants $c_1, c_2 > 0$, hence, by \cite[lemma 4]{cohen-2016-optimal-approximate} it satisfies the $(2 c_1 \sqrt{q/2k}, 9^d c_2 e^{-q}, d, q)$-OSE moment property for any $d \in \mathbb{N}$.
+    By \cite{} $\widetilde{\mtx{\Omega}}_2 \in \mathbb{R}^{2k \times l}$ satisfies the \todo{$(c_1 \sqrt{q/2k}, c_2 e^{-q}, q)$}-JL moment property for \todo{any $q$ (in what?) $\geq \dots$} \todo{\cite[remark below lemma 4]{cohen-2016-optimal-approximate} or define} and some constants $c_1, c_2 > 0$, hence, by \cite[lemma 4]{cohen-2016-optimal-approximate} it satisfies the $(2 c_1 \sqrt{q/2k}, 9^d c_2 e^{-q}, d, q)$-OSE moment property for any $d \in \mathbb{N}$.
 For convenience, we choose $d=2k$ and $q=k/2$, such that $\widetilde{\mtx{\Omega}}_2$ has $(c_1, c_2 (3^4e^{-\sfrac{1}{2}})^k, 2k, k/2)$-OSE moment property.
@@
 However, the Chebyshev polynomials are only well-defined for matrices whose spectrum is contained in $[-1, 1]$.
-When working with general matrices of bounded spectrum, we will first have to estimate a lower bound $a$ and upper bound $b$, on the spectrum \todo{[cite Saad, estimate spectrum]}.
+When working with general matrices of bounded spectrum, we will first have to estimate a lower bound $a$ and upper bound $b$ on the spectrum \cite{zhou-2011-bounding-spectrum}.
 A spectral transform
@@
 \begin{proof}
-    From Bernstein's theorem \cite[theorem 4.3]{trefethen-2008-gauss-quadrature} it follows that if $f$ is analytic within the Bernstein ellipse $\mathcal{E}_{\chi} \subset \mathbb{C}$ with foci $\pm 1$ and sum of semi-axes $\chi > 1$, then for any $m \in \mathbb{N}$
+    From Bernstein's theorem \cite[theorem 4.3]{trefethen-2008-gauss-quadrature} it follows that if $f$ is analytic within the Bernstein ellipse $\mathcal{E}_{\chi} \subset \mathbb{C}$ with foci $\{-1, +1\}$ and sum of semi-axes $\chi > 1$, then for any $m \in \mathbb{N}$
     \begin{equation}
-        \sup_{t \in [-1, 1]} \left| f - f^{(m)} \right| \leq \frac{2}{\chi^m (\chi - 1)} \max_{z \in \mathcal{E}_{\chi}} |f(z)|.
+        \sup_{s \in [-1, 1]} \left| f(s) - f^{(m)}(s) \right| \leq \frac{2}{\chi^m (\chi - 1)} \max_{z \in \mathcal{E}_{\chi}} |f(z)|.
         \label{equ:bernstein-bound}
@@
 
-\todo{Explain idea behind algorithm: Two-phase, first assemble matrices with recurrence then compute approximate in every time step}
+Similar observations for the case when $\mtx{\Psi}$ is nonzero lead to the following algorithm.
 
@@
     \Statex \textbf{Parameters:} Degree $m$, allocation $(n_{\mtx{\Omega}}, n_{\mtx{\Psi}})$,  smoothing parameter $\sigma$
-    \Statex \textbf{Output:} Approximate evaluations of the spectral density $\{\phi_{\sigma}(t_i)\}_{i=1}^{n_t}$
+    \Statex \textbf{Output:} Approximate evaluations of the spectral density $\{\widetilde{\phi}_{\sigma}(t_i)\}_{i=1}^{n_t}$
     \State Compute $\{\mu_l(t_i)\}_{l=0}^{m}$ for all $t_i$ using the $\DCT$ \refequ{equ:discrete-cosine-transform}
@@
     %\State Compute  for all $t_i$ using %\refalg{alg:3-nystrom-chebyshev-exponentiation}
-    \State Generate standard Gaussian matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$%\glsfirst{sketching-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$
+    \State Generate Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$%\glsfirst{sketching-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$
     %\State Generate standard Gaussian %\glsfirst{random-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$
@@
             \State $\mtx{K}_1(t_i) \gets \mtx{K}_1(t_i) + \mu_l(t_i) \mtx{X}$ \Comment{assemble $\mtx{\Omega}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$}
-            \State $\mtx{L}_1(t_i) \gets \mtx{L}_1(t_i) + \mu_l(t_i) \mtx{Y}$ \Comment{assemble $\mtx{\Psi}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$}
+            \State $\mtx{L}_1(t_i) \gets \mtx{L}_1(t_i) + \mu_l(t_i) \mtx{Y}$ \Comment{assemble $\mtx{\Omega}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi}$}
             \State $\ell(t_i) \gets \ell(t_i) + \mu_l(t_i) z$ \Comment{assemble $\Trace(\mtx{\Psi}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi})$}
         \EndIf
-        \State $\mtx{K}_2(t_i) \gets \mtx{K}_2(t_i) + \nu_l(t_i) \mtx{X}$ \Comment{assemble $\mtx{\Omega}^{\top} (g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}))^2 \mtx{\Omega}$}
+        \State $\mtx{K}_2(t_i) \gets \mtx{K}_2(t_i) + \nu_l(t_i) \mtx{X}$ \Comment{assemble $\mtx{\Omega}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})^2 \mtx{\Omega}$}
       \EndFor
@@
     \For {$i = 1, \dots, n_t$}
-      \State $\phi_{\sigma}(t_i) \gets \frac{1}{n} \Trace\left( \mtx{K}_1(t_i)^{\dagger}\mtx{K}_2(t_i) \right) + \frac{1}{n n_{\mtx{\Psi}}} \left( \ell(t_i) + \Trace\left( \mtx{L}_1(t_i)^{\top} \mtx{K}_1(t_i)^{\dagger} \mtx{L}_1(t_i) \right)  \right) $ \label{lin:4-nystromchebyshev-nystrom-pp}
+      \State $\widetilde{\phi}_{\sigma}(t_i) \gets \frac{1}{n} \Trace\left( \mtx{K}_1(t_i)^{\dagger}\mtx{K}_2(t_i) \right) + \frac{1}{n n_{\mtx{\Psi}}} \left( \ell(t_i) + \Trace\left( \mtx{L}_1(t_i)^{\top} \mtx{K}_1(t_i)^{\dagger} \mtx{L}_1(t_i) \right)  \right) $ \label{lin:4-nystromchebyshev-nystrom-pp}
     \EndFor
@@
 
-With the cost of a matrix-vector product denoted by $c(n)$, and supposing $n_{\mtx{\Omega}} \approx n_{\mtx{\Psi}}$, we determine the computational complexity of the Chebyshev-Nyström++ method to be $\mathcal{O}(m \log(m) n_t + m n_{\mtx{\Omega}}^2 n + m n_t n_{\mtx{\Omega}}^2 +  m c(n) n_{\mtx{\Omega}} + n_t n_{\mtx{\Omega}}^3)$, with $\mathcal{O}(m n_t + n n_{\mtx{\Omega}} + n_{\mtx{\Omega}}^2 n_t)$ required additional storage.
+The dominating cost stems from the $(2m + 1)(n_{\mtx{\Psi}} + n_{\mtx{\Omega}})$ multiplications of $\mtx{A}$ with a vector and the cost associated with computing the estimate in each evaluation point $t_1, \dots, t_{n_t}$ given by $\mathcal{O}(m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + (n_{\mtx{\Omega}}^3 + n_{\mtx{\Omega}} n_{\mtx{\Psi}}^2 + 1))$.
 
-\todo{Explain modifications for other estimators, and improved complexity for quadratic trace estimator}
+The algorithm also accomodates the two other estimators introduced in \refsec{sec:analysis}: For $n_{\mtx{\Omega}} = 0$ we recover \refequ{equ:hutchinson-trace-estimator}, and the complexity for each evaluation of the estimate decreases significantly.
+For $n_{\mtx{\Psi}} = 0$ we end up with \refequ{equ:nystrom-trace-estimator}.
+%Chebyshev $\mathcal{O}(n_t (m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + n_{\mtx{\Omega}}^3) + mn^2(n_{\mtx{\Psi}} + n_{\mtx{\Omega}}))$.
+
+%\todo{Complexity in terms of both $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$.}
+%With the cost of a matrix-vector product denoted by $c(n)$, and supposing $n_{\mtx{\Omega}} \approx n_{\mtx{\Psi}}$, we determine the computational complexity of the Chebyshev-Nyström++ method to be $\mathcal{O}(m \log(m) n_t + m n_{\mtx{\Omega}}^2 n + m n_t n_{\mtx{\Omega}}^2 +  m c(n) n_{\mtx{\Omega}} + n_t n_{\mtx{\Omega}}^3)$, with $\mathcal{O}(m n_t + n n_{\mtx{\Omega}} + n_{\mtx{\Omega}}^2 n_t)$ required additional storage.
+
+%\todo{Explain modifications for other estimators, and improved complexity for quadratic trace estimator}
+
+\todo{Discuss computation of pseudo inverse: Generalized eigenvalue problem with thresholding \cite{lin-2017-randomized-estimation, epperly-2022-theory-quantuma}, }
 
 \todo{Explain short-circuit mechanism -> may not be too crucial for directly calculating pseudo-inverse}
-
-\todo{Discuss Lin Lin generalized eigenvalue problem for pseudo-inverse, other ways are not available, since we only have access to the \enquote{compressed} matrix function}
 
@@
 $10^{-16}$.
-When using a Gaussian smoothing-kernel, we can explicitly write the signular values of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ as
+When using a Gaussian smoothing-kernel, we can explicitly write the singular values of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ as
 \begin{equation}
-    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) = \frac{1}{n \sqrt{2 \pi \sigma^2}} e^{-\sfrac{(t - \lambda_{(i)})^2}{2 \sigma^2}},
+    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) = \frac{1}{\sqrt{2 \pi \sigma^2}} e^{-\sfrac{(t - \lambda_{(i)})^2}{2 \sigma^2}},
     \label{equ:gaussian-kernel-eigenvalues}
@@
 \end{equation}
-with $d_{\varepsilon, \sigma} = \sigma \sqrt{-2 \log(\sqrt{2 \pi n} \sigma \varepsilon)}$ and the eigenvalues $\lambda_1, \dots, \lambda_n$ of $\mtx{A}$ and the constants.
+with $d_{\varepsilon, \sigma} = \sigma \sqrt{-2 \log(\sqrt{2 \pi} \sigma \varepsilon)}$ and the eigenvalues $\lambda_1, \dots, \lambda_n$ of $\mtx{A}$ and the constants.
 The expression \refequ{equ:gaussian-kernel-numerical-rank} has a nice visual interpretation: The numerical rank of a matrix is at most equal to the number of eigenvalues which are closer to the parameter $t$ than $d_{\varepsilon, \sigma}$.

### 2024-05-29 paper/main.tex
finish first draft of introduction

@@
 
-\title{Randomized trace estimation of parameter-dependent matrices applied to spectral density approximation}
+\title{Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation}
 
@@
 
+%\todo{
+%    \tableofcontents
+%}
+
 \todo{
-    \tableofcontents
+
+\paragraph{TODO}
+\begin{itemize}
+    \item Change all parameter-dependent matrices to $\mtx{B}(t)$ to avoid confusion with $\mtx{A}$
+    \item Change citation style to \textbf{[1]}, \textbf{[2]}, \dots
+    \item Re-run numerical experiments with simplified and improved implementation
+\end{itemize}
 }
@@
 
-In numerous problems across physics, chemistry, engineering, and data science, how the eigenvalues of certain matrices are distributed can be indicative of what properties a system has: In electronic structure calculations they represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}.
+In numerous problems across physics, chemistry, engineering, and data science, how the eigenvalues of certain matrices are distributed can be indicative of what properties a system has: In electronic structure calculations eigenvalues represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}.
 The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$.
@@
 Therefore, many techniques have been proposed for approximating $\phi$.
-For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature rule from the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}.
-Further, $\delta$ is hard to approximate and in many applications, a rough estimate of the spectral density is already acceptable.
-One standard approach is to instead approximate the smooth spectral density
+For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature rule with the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}.
+Further, because $\delta$ is hard to approximate and in many applications, a rough estimate of the spectral density is already acceptable, a standard approach is to instead approximate the smooth spectral density
 \begin{equation}
-    \phi_{\sigma}(t) = \frac{1}{n} \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i)
+    \phi_{\sigma}(t) = \frac{1}{n} \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i) = \frac{1}{n} \Trace(g_{\sigma}(t \mtx{I}_n - \mtx{A}))
 \end{equation}
 for a smoothing kernel $g_{\sigma}$, typically a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}.
-This allows the problem to be framed in the setting of trace estimation of parameter-dependent matrices.
-
-\todo{
-\begin{itemize}
-    \item Brief definition of parameter-dependent matrix ($\mtx{A}(t) \in \mathbb{R}^{n \times n}$, every entry depends continuously on $t \in [a,b]$ \cite{kressner-2023-randomized-lowrank})
-    \item Trace estimation through mat-vecs (if matrix is only accessed implicitly \cite{pearlmutter-1994-fast-exact} or for complicated matrix functions [partition function, spectral density, exponential integrators])
-    \item Existing trace estimation methods for constant matrices: Girard-Hutchinson \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} and variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making} which have only recently been specialized to trace estimation of matrix functions \cite{chen-2023-krylovaware-stochastic}
-    \item Trace estimation in the parameter-dependent matrices (Nyström++ but no analysis \cite{lin-2017-randomized-estimation}, only low-rank approximation \cite{kressner-2023-randomized-lowrank}, dynamic trace estimation for sequence $\mtx{A}_1, \dots, \mtx{A}_m$ of similar matrices \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace}) their algorithm provides trace approximations $a_1, \dots , a_m$ for $\Trace(\mtx{A}_1), \dots, \Trace(\mtx{A}_m)$ such that, for each $i \in 1, \ldots, m$,
-    \begin{equation}
-        \mathbb{P}(|a_i - \Trace(\mtx{A}_i)| \geq \epsilon)\leq \delta.
-    \end{equation}
-    drawback of this type of guarantee is that when we aim for $\epsilon$ accuracy simultaneously for all $|a_i-\Trace(\mtx{A}_i)|$, the probability will grow linearly with $m$ by the union bound.
-On the other hand, in our setting, instead of discretizing $[a, b]$ with $t_1, \ldots, t_m$ and looking at the error $|a(t_i)-\Trace(\mtx{A}(t_i))|$ for each $i$, a much more natural way to quantify the error would be to look at $\int_{a}^{b} | a(t) - \mtx{A}(t) |~\mathrm{d}t$, where $a(t)$ is the trace estimation in each $t \in [a, b]$.
-    %\item Error analysis for spectral density approximation (Wasserstein-1 error [Braverman, Chen], missing $1/\varepsilon$ guarantees)
-\end{itemize}
+This translates the problem to computing the trace of the parameter-dependent matrix $\mtx{B}(t) \equiv g_{\sigma}(t \mtx{I}_n - \mtx{A}) \in \mathbb{R}^{n \times n}$.
+Directly assembling the matrix function requires diagonalizing the matrix $\mtx{A}$, which again is prohibitively expensive.
+Therefore, we estimate its trace implicitly by approximating products of the matrix function with vectors.
+For constant matrices, the most widespread methods for implicit trace estimation are the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} and the variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making, chen-2023-krylovaware-stochastic}.
+Some of these estimators have already been successfully applied to parameter-dependent matrices \cite{lin-2017-randomized-estimation}, but they have not yet been analyzed.
+A related technique is dynamic trace estimation for sequences $\mtx{B}_1, \dots, \mtx{B}_m$ of similar matrices \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace}, which achieves guarantees for the estimate $b_i$ of $\Trace(\mtx{B}_i)$ of the form $\mathbb{P}(|\Trace(\mtx{B}_i) - b_i| \geq \varepsilon) \leq \delta$ for $i=1,\dots, m$.
+A drawback of this type of bound is that when we aim for $\varepsilon$ accuracy for all $i=1,\dots, m$ simultaneously, the probability will grow linearly with $m$ by the union bound.
+Therefore, instead of discretizing $[a, b]$ with $t_1, \ldots, t_m$ and looking at the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way to quantify the error is to use $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate for each $t \in [a, b]$.
 
 \paragraph{Contribution}
-\begin{itemize}
-    \item Same random matrix for all $t$, which allows for favorable scaling in the number of evaluation points
-    \item Theoretical guarantees in the $L^1$-norm for three established trace estimators for constant matrices when applied to parameter-dependent matrices (Girard-Hutchinson, Nyström, Nyström++), in particular, $\mathcal{O}(\varepsilon^{-1})$ scaling with number of mat-vecs
-    \item A simple and efficient algorithm for estimating spectral densities using any of the three estimators with Chebyshev expansion, including theoretical guarantees
-\end{itemize}
-}
+We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices.
+In particular, we reuse the same randomization for each value of the parameter $t$ in the estimators, which makes them scale favorably with the number of parameter evaluations.
+Subsequently, we propose a simple, efficient, and provably correct procedure in which these estimators are employed for approximating the smooth spectral density $\phi_{\sigma}$ of a matrix.
 
@@
     \Trace(\mtx{A}(t)) = \sum_{i=1}^{n} a_{ii}(t).
+    \label{equ:definition-trace}
 \end{equation}
+However, we assume that we only have access to products of this matrix with vectors for each $t \in [a, b]$, so this definition will not be directly useful for computing the trace.
 
-\paragraph{Girard-Hutchinson estimator} If we can only access the parameter-dependent matrix through matrix-vector products, we can estimate the trace with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ to form
+\paragraph{Girard-Hutchinson estimator} We can estimate the trace with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ to form
 \begin{equation}
@@
 where $\mtx{\Psi} = [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\mtx{\Psi}}}] \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$.
-Other choices for the distribution of the entries of the random vectors are possible, for example by uniformly sampling from $\{-1, +1\}$ or from the $(n-1)$-sphere.
+Other choices for the distribution of the random vectors are possible, for example by uniformly sampling from $\{-1, +1\}$ or from the $(n-1)$-sphere.
 However, our theoretical developments only hold in the Gaussian case.
@@
 \end{equation}
-Then we use the trace estimator $\Trace(\Nystr{\mtx{\Omega}}{\mtx{A}}(t))$.
-Thanks to the invariance of the trace under cyclic permutation of its arguments and the symmetry of the matrix, we may also rewrite
+Then we can estimate the trace as $\Trace(\Nystr{\mtx{\Omega}}{\mtx{A}}(t))$.
+Thanks to the invariance of the trace under cyclic permutation of its arguments and the symmetry of the matrix, we may rewrite this estimator as
+
 \begin{equation}
@@
 
-\paragraph{Nyström++ estimator} Finally, a combination of the two estimators yields an analogue for the Nyström++ estimator \cite{persson-2022-improved-variants} in the parameter-dependent setting
+\paragraph{Nyström++ estimator} Finally, \cite{lin-2017-randomized-estimation} proposes an estimator which corrects for inaccuracies in the Nyström approximation \refequ{equ:nystrom-trace-estimator} by estimating the trace of its residual using the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}
 \begin{equation}
-    \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{A}(t)) = \Trace(\Nystr{\mtx{\Omega}}{\mtx{A}}(t)) + \Hutch{\mtx{\Psi}}(\mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)),
+    \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{A}(t)) = \Trace(\Nystr{\mtx{\Omega}}{\mtx{A}}(t)) + \Hutch{\mtx{\Psi}}(\mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)).
     \label{equ:nystrompp-trace-estimator}
 \end{equation}
-which was originally proposed by \cite{lin-2017-randomized-estimation}.
+This is the analogue of the Nyström++ estimator \cite{persson-2022-improved-variants} in the parameter-dependent setting.
 %We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.
@@
 \paragraph{$L^{1}$ norm}
-We define the $L^{1}$ norm of a square-integrable function $f:[a,b] \to \mathbb{R}$ as
+We define the $L^{1}$ norm of a measurable function $f:[a,b] \to \mathbb{R}$ as
 \begin{equation}
@@
 \end{equation}
-We use it to measure the error of the trace estimators. \todo{Motivate why we use this.}
+We use it to measure the error of the trace estimators. The following sections provide bounds on this error for each of the above-mentioned estimators.
 
@@
     \end{equation}
-    In particular, if we choose $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{\sfrac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{A}(t)) - \Hutch{\mtx{\Psi}}(\mtx{A}(t)) | ~\mathrm{d}t < \varepsilon \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t$.
+    In particular, if we choose $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{- \sfrac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{A}(t)) - \Hutch{\mtx{\Psi}}(\mtx{A}(t)) | ~\mathrm{d}t < \varepsilon \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t$.
 \end{theorem}
@@
 The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Markov's inequality to transition to the continuous parameter case.
-We first only consider the $1$-query estimator and subsequently use a trick to carry the result over to the general case.
+We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
 \begin{proof} 
-    We define the 1-query quadratic trace estimate 
+    We define the 1-query Girard-Hutchinson estimate 
     \begin{equation}
@@
     \end{align}
-    Since $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$ we can upper bound 
+    Since $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
     \begin{equation}
@@
     \end{equation}
-    By Markov's inequality, for any $k \in \mathbb{N}$ and $\gamma \geq 1$, it holds with probability at least $1 - \gamma^{-2 k}$
+    Consequently, by Markov's inequality, for any $k \in \mathbb{N}$ and $\gamma \geq 1$, it holds with probability at least $1 - \gamma^{-2 k}$
     \begin{equation}
@@
     if $n_{\mtx{\Psi}} \geq 2^{10} \varepsilon^{-2} \log(\delta^{-1})^2 (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ with $\mu(t) = \lVert \mtx{A}(t) \rVert _2 / \Trace(\mtx{A}(t))$.
-Compared to the constant matrix equivalent \cite[remark 2]{cortinovis-2022-randomized-trace} we observe an additional factor of $\log(\delta^{-1})$.
+Compared to the constant matrix equivalent \cite[remark 2]{cortinovis-2022-randomized-trace} we mainly observe a larger constant and an additional factor of $\log(\delta^{-1})$.
 \end{remark}
@@
 \begin{theorem}{Nyström estimator for parameter-dependent matrices}{nystrom}
-    Suppose $f(\mtx{A}, t)$ is continuous and non-negative in $t \in [a, b]$ for a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ and $n_{\mtx{\Omega}} \geq 6, n_{\mtx{\Omega}} - 4 \geq r \geq 2$ two integers.
+    Suppose $f(\mtx{A}, t)$ is continuous and non-negative in $t \in [a, b]$ for a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$.
+Let $n_{\mtx{\Omega}} \geq 6, n_{\mtx{\Omega}} - 4 \geq r \geq 2$ be two integers.
 Then for all $\gamma \geq 1$ with probability at least $1 - \gamma^{-(n_{\mtx{\Omega}} - r)}$
@@
 \begin{proof}
-    Since $f$ is non-negative, $f(\mtx{A}, t)$ is positive definite for all $\mtx{A}$ and $t$.
-    Notice that by \cite[lemma 2.1]{frangella-2023-randomized-nystrom}, $f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t)$ is also positive semi-definite for all $t$. Therefore,
+    Since $f$ is non-negative, $f(\mtx{A}, t)$ is positive semi-definite for all $t$.
+By \cite[lemma 2.1]{frangella-2023-randomized-nystrom}, $f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t)$ is also positive semi-definite for all $t$.
+Therefore, for fixed $t \in [a, b]$
     \begin{equation}
@@
     \begin{equation}
-        \lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _2^2
+        %\lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _2^2 \leq \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _F^2
+        \lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _F^2
     \end{equation}
-    Hence, by \cite[theorem 9]{kressner-2023-randomized-lowrank}, with $n_{\mtx{\Omega}} = p + r$, the result follows
-    with $\lVert \cdot \rVert _2 \leq \lVert \cdot \rVert _F$ and $\sigma_i(f^{\sfrac{1}{2}}(\mtx{A}, t))^{2} = \sigma_i(f(\mtx{A}, t))$.
+    Hence, by \cite[theorem 9]{kressner-2023-randomized-lowrank}, with $n_{\mtx{\Omega}} = p + r$, the result follows by noting $\sigma_i(f^{\sfrac{1}{2}}(\mtx{A}, t))^{2} = \sigma_i(f(\mtx{A}, t))$.
 \end{proof}
@@
 \begin{equation}
-    \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+2ex+\baselineskip}
+    \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+4ex+\baselineskip}
     \mtx{U}(t) = \begin{bmatrix}
@@
 \begin{lemma}{Moments of tail bound}{ose}
-    A symmetric matrix $\mtx{A} \in \mathbb{R}^{l \times l}$ and stochastically independent Gaussian random matrices $\mtx{\Omega}_1 \in \mathbb{R}^{k \times 2k}, \mtx{\Omega}_2 \in \mathbb{R}^{l \times 2k}$ satisfy for all $k \in \mathbb{N}$ and a constant $c > 0$
+    A symmetric matrix $\mtx{A} \in \mathbb{R}^{l \times l}$ and independent standard Gaussian random matrices $\mtx{\Omega}_1 \in \mathbb{R}^{k \times 2k}, \mtx{\Omega}_2 \in \mathbb{R}^{l \times 2k}$ satisfy for all $k \in \mathbb{N}$ and a constant $c > 0$
     \begin{equation}
@@
     for a universal constant $c > 0$.
-In particular, if we choose $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})), n_{\mtx{\Omega}} \geq 2$, then with probability at least $1-\delta$ for $\delta \in (0, e^{-e^{4 \varepsilon^2}})$ and $\varepsilon > 0$ we have $\int_{a}^{b} \lVert \mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F~\mathrm{d}t  < \varepsilon \int_{a}^{b} \Trace(A(t))~\mathrm{d}t$.
+In particular, if we choose $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})), n_{\mtx{\Omega}} \geq 2$, then with probability at least $1-\delta$ for $\delta \in (0, e^{-e^{4 \varepsilon^2}})$ and $\varepsilon > 0$ we have $\int_{a}^{b} \lVert \mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(A(t))~\mathrm{d}t$.
 \end{lemma}
@@
     \Statex \textbf{Input:} Symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$, evaluation points $\{t_i\}_{i=1}^{n_t}$
-    \Statex \textbf{Parameters:} Degree $m$, allocation $(n_{\mtx{\Omega}}, n_{\mtx{\Psi}})$,  smoothing parameter $\sigma$
+    \Statex \textbf{Parameters:} Degree $m$, allocation $(n_{\mtx{\Psi}}, n_{\mtx{\Omega}})$,  smoothing parameter $\sigma$
     \Statex \textbf{Output:} Approximate evaluations of the spectral density $\{\widetilde{\phi}_{\sigma}(t_i)\}_{i=1}^{n_t}$
@@
 
+% Acknowledge advice from David
+
 \clearpage

### 2024-05-29 src/simple.py
simplify algorithms

@@ -1,0 +1,185 @@
+import numpy as np
+import scipy as sp
+# TODO: Add Numba JIT!!!
+
+def gaussian_kernel(s, sigma):
+    """
+    Gaussian kernel.
+
+    Parameters
+    ----------
+    s : int, float, or np.ndarray of shape (n, m)
+        Point(s) where the Gaussian should be evaluated.
+    sigma : int or float
+        Standard deviation of the Gaussian.
+
+    Returns
+    -------
+    g(s) : np.ndarray of shape (n, dim)
+        The Gaussian kernel evaulated at all points s.
+    """
+    s = np.asarray(s, dtype=np.float64)
+
+    return np.exp(- s**2 / (2 * sigma**2)) / (np.sqrt(2 * np.pi) * sigma)
+
+def chebyshev_expansion(kernel, t, m, nonnegative=False):
+    """
+    Chebyshev expansion of a kernel.
+
+    Parameters
+    ----------
+    kernel : kernel
+        The kernel used to regularize the spectral density.
+    t : int, float, list, or np.ndarray of shape (n_t,)
+        Point(s) where the expansion should be evaluated.
+    m : int > 0
+        Degree of the Chebyshev polynomial.
+    nonnegative : bool
+        Force interpolant to be non-negative (for non-negative functions).
+
+    Returns
+    -------
+    mu : np.ndarray of shape (n_t, m + 1)
+        The coefficients of the Chebyshev polynomials. Format: mu[t, l].
+    """
+    t = np.asarray(t, dtype=np.float64)
+
+    if nonnegative:
+        m = m // 2
+        kernel = lambda x: np.sqrt(kernel(x))
+
+    # Compute the coefficients mu for all t and l simultaneously with DCT
+    chebyshev_nodes = np.cos(np.arange(m + 1) * np.pi / m)
+    mu = sp.fft.idct(kernel(np.subtract.outer(t, chebyshev_nodes)), type=1)
+
+    # Rescale coefficients due to type-1 DCT convention
+    mu[..., 1:-1] *= 2
+
+    # Square Chebyshev expansion
+    if nonnegative:
+        mu = square_chebyshev_expansion(mu)
+
+    return mu
+
+def square_chebyshev_expansion(mu):
+    """
+    Square a Chebyshev expansion with coefficients mu.
+
+    Parameters
+    ----------
[…]

### 2024-05-31 paper/main.tex
implement plots

@@
     \item Change citation style to \textbf{[1]}, \textbf{[2]}, \dots
-    \item Re-run numerical experiments with simplified and improved implementation
+    \item Re-run numerical experiments with simplified and improved implementation (test pinvh against lstsq)
 \end{itemize}
@@
 for a smoothing kernel $g_{\sigma}$, typically a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}.
-This translates the problem to computing the trace of the parameter-dependent matrix $\mtx{B}(t) \equiv g_{\sigma}(t \mtx{I}_n - \mtx{A}) \in \mathbb{R}^{n \times n}$.
+This translates the problem to computing the trace of a continuous parameter-dependent matrix $\mtx{B}(t) \equiv g_{\sigma}(t \mtx{I}_n - \mtx{A}) \in \mathbb{R}^{n \times n}$.
 Directly assembling the matrix function requires diagonalizing the matrix $\mtx{A}$, which again is prohibitively expensive.
@@
 For $n_{\mtx{\Psi}} = 0$ we end up with \refequ{equ:nystrom-trace-estimator}.
+
+\todo{Clearly need not divide by $n_{\mtx{\Psi}}=0$, hence just ignore this term. }
 %Chebyshev $\mathcal{O}(n_t (m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + n_{\mtx{\Omega}}^3) + mn^2(n_{\mtx{\Psi}} + n_{\mtx{\Omega}}))$.
@@
 
-\todo{Explain short-circuit mechanism -> may not be too crucial for directly calculating pseudo-inverse}
+\todo{Explain short-circuit mechanism -> very crucial if matrix can be zero}
 
@@
 
-% Acknowledge advice from David
+% Acknowledge advice from David and Data/Parameters from Lin Lin
 

### 2024-05-31 paper/plots/convergence.py

@@ -1,0 +1,59 @@
+import __context__
+
+import numpy as np
+import scipy as sp
+import matplotlib.cm as cm
+import matplotlib.pyplot as plt
+from src.simple import spectral_density, spectral_transformation, form_spectral_density, gaussian_kernel
+
+np.random.seed(0)
+
+# Load matrix
+A = sp.sparse.load_npz("matrices/ModES3D_1.npz")
+
+# Perform spectral transform with A and its eigenvalues
+eigvals = np.linalg.eigvalsh(A)#.toarray())
+min_ev, max_ev = eigvals[0], eigvals[-1]
+A_st = spectral_transformation(A, min_ev, max_ev)
+eigvals_st = spectral_transformation(eigvals, min_ev, max_ev)
+
+# Set parameter
+t = np.linspace(-1, 1, 100)
+sigma = 0.2
+m = 500
+n_Vec_list = np.logspace(0.95, 2.6, 2).astype(int) 
+
+plt.style.use("paper/plots/stylesheet.mplstyle")
+colors = cm.magma(np.arange(5) / 4.5)
+markers = ["o", "^", "s"]
+labels = [r"$n_{\mathbf{\Omega}} = 0$",
+          r"$n_{\mathbf{\Psi}} = n_{\mathbf{\Omega}}$",
+          r"$n_{\mathbf{\Psi}} = 0$"]
+
+# Determine the baseline spectral density
+kernel = lambda x: gaussian_kernel(x, sigma=sigma)
+baseline = form_spectral_density(eigvals_st, t, kernel)
+
+error = np.empty((3, len(n_Vec_list)))
+for j, n_Vec in enumerate(n_Vec_list):
+    estimate = spectral_density(A_st, t, m, n_Vec, 0, kernel)
+    error[0, j] = 2 * np.mean(np.abs(estimate - baseline))
+    estimate = spectral_density(A_st, t, m, n_Vec // 2, n_Vec // 2, kernel)
+    error[1, j] = 2 * np.mean(np.abs(estimate - baseline))
+    estimate = spectral_density(A_st, t, m, 0, n_Vec, kernel)
+    error[2, j] = 2 * np.mean(np.abs(estimate - baseline))
+
+for i in range(3):
[…]

### 2024-05-31 paper/plots/distribution.py

@@ -1,0 +1,53 @@
+import __context__
+
+import numpy as np
+import scipy as sp
+import matplotlib.cm as cm
+import matplotlib.pyplot as plt
+from src.simple import spectral_density, spectral_transformation, form_spectral_density, gaussian_kernel
+
+np.random.seed(0)
+
+# Load matrix
+A = sp.sparse.load_npz("matrices/ModES3D_1.npz")
+
+# Perform spectral transform with A and its eigenvalues
+eigvals = np.linalg.eigvalsh(A.toarray())
+min_ev, max_ev = eigvals[0], eigvals[-1]
+A_st = spectral_transformation(A, min_ev, max_ev)
+eigvals_st = spectral_transformation(eigvals, min_ev, max_ev)
+
+# Set parameter
+t = np.linspace(-1, 1, 100)
+sigma_list = np.logspace(-2.5, -0.5, 7)
+n_Vec = 80
+n_Psi_list = np.arange(n_Vec + 1, step=20).astype(np.int64)
+n_Omega_list = n_Vec - n_Psi_list
+
+plt.style.use("paper/plots/stylesheet.mplstyle")
+colors = cm.magma(np.arange(5) / 4.5)
+markers = ["o", "^", "s", "p", "d"]
+labels = [r"$n_{\mathbf{\Psi}} = " + "{}$, ".format(n_Psi) + r"$n_{\mathbf{\Omega}} = " + "{}$".format(n_Omega) for n_Psi, n_Omega in zip(n_Psi_list, n_Omega_list)]
+
+error = np.empty((len(n_Psi_list), len(sigma_list)))
+for j, sigma in enumerate(sigma_list):
+
+    # Determine the baseline spectral density
+    kernel = lambda x: gaussian_kernel(x, sigma=sigma)
+    baseline = form_spectral_density(eigvals_st, t, kernel)
+    m = int(120 / sigma)
+
+    for i, (n_Psi, n_Omega) in enumerate(zip(n_Psi_list, n_Omega_list)):
+        estimate = spectral_density(A_st, t, m, n_Psi, n_Omega, kernel)
+        error[i, j] = 2 * np.mean(np.abs(estimate - baseline))
+
+for i in range(len(n_Psi_list)):
+    plt.plot(sigma_list, error[i], color=colors[i], marker=markers[i], label=labels[i])
+
+plt.grid()
+plt.ylabel(r"$L^1$ error")
[…]

### 2024-05-31 src/simple.py

@@ -1,8 +1,34 @@
 import numpy as np
 import scipy as sp
-# TODO: Add Numba JIT!!!
 
-def gaussian_kernel(s, sigma):
+def spectral_transformation(A, min_ev=None, max_ev=None):
+    """
+    Perform a spectral transformation of a matrix, i.e. transform a spectrum
+    contained in (a, b) to (-1, 1).
+
+    Parameters
+    ----------
+    A : np.ndarray of shape (n, n) or (n,)
+        The matrix or vector to be spectrally transformed.
+    min_ev : int, float or None
+        The starting point of the spectrum. If None is specified, the smallest
+        eigenvalue of A is computed and used for min_ev.
+    max_ev : int, float or None
+        The ending point of the spectrum. If None is specified, the largest
+        eigenvalue of A is computed and used for max_ev.
+    """
+    I = 1
+    if len(A.shape) == 2:
+        if isinstance(A, sp.sparse.spmatrix):
+            I = sp.sparse.eye(*A.shape)
+        if isinstance(A, np.ndarray):
+            I = np.eye(*A.shape)
+
+    A_st = (2 * A - (min_ev + max_ev) * I) / (max_ev - min_ev)
+
+    return A_st
+
+def gaussian_kernel(s, sigma=0.1):
     """
     Gaussian kernel.
 
@@ -18,9 +44,28 @@ def gaussian_kernel(s, sigma):
     g(s) : np.ndarray of shape (n, dim)
         The Gaussian kernel evaulated at all points s.
     """
-    s = np.asarray(s, dtype=np.float64)
-
     return np.exp(- s**2 / (2 * sigma**2)) / (np.sqrt(2 * np.pi) * sigma)
+
+def form_spectral_density(eigvals, t, kernel=gaussian_kernel):
+    """
+    Compute the (regularized) spectral density of a (small) matrix A at n_t
+    evenly spaced grid-points within the interval [a, b].
+
+    Parameters
+    ----------
+    eigvals : np.ndarray of shape (n,)
+        The eigenvalues for which the spectral density should be formed.
+    t: np.ndarray (n_t,)
+        Parameter values at which the spectral density should be evaluated.
+    kernel : function
+        Smoothing kernel.
+
+    Returns
+    -------
[…]

### 2024-06-03 paper/main.tex
rewrite Nystrom proof

@@
 In particular, we reuse the same randomization for each value of the parameter $t$ in the estimators, which makes them scale favorably with the number of parameter evaluations.
-Subsequently, we propose a simple, efficient, and provably correct procedure in which these estimators are employed for approximating the smooth spectral density $\phi_{\sigma}$ of a matrix.
+Subsequently, we propose a simple, fast, robust, and provable procedure in which these estimators are employed for approximating the smooth spectral density $\phi_{\sigma}$ of a matrix.
 
 \paragraph{Notation}
-We follow the notation and conventions which are used in contemporary literature in this field. Further,
-\begin{itemize}
-    \item we denote with $\mathbb{N}$ all positive whole numbers;
-    \item $e$ denotes Euler's number and $\log$ is the corresponding logartihm;
-    \item the ceiling function $\lceil x \rceil$ of a real number $x$ returns the greatest integer larger than or equal to $x$;
-    \item scalars are represented by lower case Greek or Latin letters ($s$, $\varepsilon$, \dots), vectors are additionally printed in bold ($\vct{v}$, $\vct{\psi}$, \dots), and matrices are additionally capitalized ($\mtx{A}$, $\mtx{\Omega}$, \dots);
-    \item the expected value of a random variable $x$ is denoted with $\mathbb{E}[x]$.
-The $k$-th moment of a random variable $x$ is defined as $\mathbb{E}^{k}[x] = \left(\mathbb{E}\left[ | x |^{k} \right] \right)^{\sfrac{1}{k}}$;
-    \item the identity matrix $\mtx{I}_n = \diag(1, \dots, 1) \in \mathbb{R}^{n \times n}$ carries ones on its diagonal and zeros everywhere else.
-The zero matrix $\mtx{0}_{n \times m} \in \mathbb{R}^{n \times m}$ consists of only zero entries;
-    \item in a Gaussian random vector and a Gaussian random matrix, all entries are independently sampled from the standard normal distribution;
-    \item matrix norms are denoted with $\lVert \cdot \rVert$.
-For symmetric matrices, the $p$-Schatten norm is $\lVert \mtx{A} \rVert _{(p)} = (\sum_{i=1}^n |\lambda_i|^p)^{1/p}$.
-We refer to the spectral norm as $\lVert \mtx{A} \rVert _2 = \lVert \mtx{A} \rVert _{(\infty)} = \max_{i=1,\dots,n}|\lambda_i|$, the nuclear norm as $\lVert \mtx{A} \rVert _{\ast} = \lVert \mtx{A} \rVert _{(1)} = \sum_{i=1}^{n} |\lambda_i|$, and the Frobenius norm as $\lVert \mtx{A} \rVert _F = \lVert \mtx{A} \rVert _{(2)} = (\sum_{i=1}^{n} |\lambda_i|^2)^{\sfrac{1}{2}}$.
-\end{itemize}
+We follow the notation and conventions which are used in contemporary literature in this field.
+Further, we denote with $\mathbb{N}$ all positive whole numbers; $e$ denotes Euler's number and $\log$ is the corresponding logartihm; the ceiling function $\lceil x \rceil$ of a real number $x$ returns the greatest integer larger than or equal to $x$; scalars are represented by lower case Greek or Latin letters ($s$, $\varepsilon$, \dots), vectors are additionally printed in bold ($\vct{v}$, $\vct{\psi}$, \dots), and matrices are additionally capitalized ($\mtx{A}$, $\mtx{\Omega}$, \dots); the expected value of a random variable $x$ is denoted with $\mathbb{E}[x]$; the $k$-th moment of a random variable $x$ is defined as $\mathbb{E}^{k}[x] = \left(\mathbb{E}\left[ | x |^{k} \right] \right)^{\sfrac{1}{k}}$; the identity matrix $\mtx{I}_n = \diag(1, \dots, 1) \in \mathbb{R}^{n \times n}$ carries ones on its diagonal and zeros everywhere else; the zero matrix $\mtx{0}_{n \times m} \in \mathbb{R}^{n \times m}$ consists of only zero entries; in a Gaussian random vector and a Gaussian random matrix, all entries are independently sampled from the standard normal distribution; matrix norms are denoted with $\lVert \cdot \rVert$; for symmetric matrices, the $p$-Schatten norm is $\lVert \mtx{A} \rVert _{(p)} = (\sum_{i=1}^n |\lambda_i|^p)^{1/p}$; we refer to the spectral norm as $\lVert \mtx{A} \rVert _2 = \lVert \mtx{A} \rVert _{(\infty)} = \max_{i=1,\dots,n}|\lambda_i|$, the nuclear norm as $\lVert \mtx{A} \rVert _{\ast} = \lVert \mtx{A} \rVert _{(1)} = \sum_{i=1}^{n} |\lambda_i|$, and the Frobenius norm as $\lVert \mtx{A} \rVert _F = \lVert \mtx{A} \rVert _{(2)} = (\sum_{i=1}^{n} |\lambda_i|^2)^{\sfrac{1}{2}}$.
 
-\todo{
-\paragraph{Reproducibility} Fully reproducible paper hosted on GitHub.
-}
+\paragraph{Reproducibility} This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-SD/tree/ca391a8}{ca391a8} at \href{https://github.com/FMatti/Rand-TRACE}{github.com/FMatti/Rand-TRACE} on 2024-02-05 at 18:44:29 UTC.
+
 \section{Analysis of trace estimators for parameter-dependent matrices}
@@
 \begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
-    Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ continuously depend on $t \in [a, b]$. Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ it holds with probability at least $1 - \gamma^{-2k}$
+    Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ be symmetric and continuously depend on $t \in [a, b]$.
+Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ it holds with probability at least $1 - \gamma^{-2k}$
     \begin{equation}
@@
     \end{equation}
-    for a parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$.
+    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$.
 
@@
 
-\todo{The drawback of the Girard-Hutchinson estimator is the $\mathcal{O}(\varepsilon^{-2})$-requirement on the number of queries. }
+The Girard-Hutchinson estimator distinguishes itself for its simplicity and provability for all symmetric matrices.
+However, the $n_{\mtx{\Psi}} = \mathcal{\varepsilon^{-2}}$ requirement implies that if we aim to increase the accuracy of the estimate by one digit, we would need to increase the number of queries by a factor of $100$.
+The next two estimators improve on this by exploiting the structure of $\mtx{A}(t)$.
 
@@
 \begin{lemma}{Moments of tail bound}{ose}
-    A symmetric matrix $\mtx{A} \in \mathbb{R}^{l \times l}$ and independent standard Gaussian random matrices $\mtx{\Omega}_1 \in \mathbb{R}^{k \times 2k}, \mtx{\Omega}_2 \in \mathbb{R}^{l \times 2k}$ satisfy for all $k \in \mathbb{N}$ and a constant $c > 0$
+    A matrix $\mtx{A} \in \mathbb{R}^{l \times l}$ and independent standard Gaussian random matrices $\mtx{\Omega}_1 \in \mathbb{R}^{k \times 2k}, \mtx{\Omega}_2 \in \mathbb{R}^{l \times 2k}$ satisfy for all $k \in \mathbb{N}$ and a universal constant $c > 0$
     \begin{equation}
-        \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] \leq c \left( \lVert \mtx{A} \rVert _2 + \frac{\lVert \mtx{A} \rVert _F}{\sqrt{k}} \right).
+        \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] \leq c \lVert \mtx{A} \rVert _2 .
     \end{equation}
@@
         \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-        \leq \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \rVert _2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-        = \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \rVert _2 \right] \mathbb{E}^{k}\left[  \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right].
-        \label{equ:OSE-moment-bound-initial}
+        \leq \mathbb{E}^{k}\left[ \lVert \mtx{A} \rVert _2 \lVert \mtx{\Omega}_2 \rVert _2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
+        = \lVert \mtx{A} \rVert _2 \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_2 \rVert _2 \right] \mathbb{E}^{k}\left[  \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right].
+        \label{equ:OSE-moment-bound-initial-new}
     \end{equation}
 
-    To bound the first term, we define $\widetilde{\mtx{\Omega}}_2 = \mtx{\Omega}_2^{\top} / \sqrt{2k}$ and rewrite
+    To bound the moments of $\lVert \mtx{\Omega}_2 \rVert _2$ we define $\widetilde{\mtx{\Omega}}_2 = \mtx{\Omega}_2^{\top} / \sqrt{2k}$ and rewrite
     \begin{align}
-        &\mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \rVert _2 \right] \notag \\
-        &= \sqrt{2k} \cdot \mathbb{E}^{k}\left[ \lVert \mtx{A} \widetilde{\mtx{\Omega}}_2^{\top} \rVert _2 \right] && \text{(definition of $\widetilde{\mtx{\Omega}}_2$)} \notag \\
-        &= \sqrt{2k} \cdot \sqrt{ \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\mtx{A} \widetilde{\mtx{\Omega}}_2^{\top}) (\mtx{A} \widetilde{\mtx{\Omega}}_2^{\top})^{\top} \rVert _2 \right] } && \text{($\lVert \mtx{B} \rVert _2 = \lVert \mtx{B} \mtx{B}^{\top} \rVert _2^{\sfrac{1}{2}} $ )} \notag \\
-        &\leq \sqrt{2k} \cdot \sqrt{ \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\mtx{A} \widetilde{\mtx{\Omega}}_2^{\top}) (\mtx{A} \widetilde{\mtx{\Omega}}_2^{\top})^{\top} - \mtx{A}^{\top} \mtx{A} \rVert _2 \right] + \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{A}^{\top} \mtx{A} \rVert _2 \right]} && \text{(triangle inequality)} \notag \\
-        &\leq \sqrt{2k} \cdot \sqrt{ \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\widetilde{\mtx{\Omega}}_2 \mtx{A} )^{\top} (\widetilde{\mtx{\Omega}}_2 \mtx{A}) - \mtx{A}^{\top} \mtx{A} \rVert _2\right]} + \sqrt{2k} \cdot \lVert \mtx{A} \rVert _2.
-&& \text{($\sqrt{a+b} \leq \sqrt{a} + \sqrt{b}$)}
+        &\mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_2 \rVert _2 \right] \notag \\
+        &= \sqrt{2k} \cdot \mathbb{E}^{k}\left[ \lVert \widetilde{\mtx{\Omega}}_2^{\top} \rVert _2 \right] && \text{(definition of $\widetilde{\mtx{\Omega}}_2$)} \notag \\
+        &= \sqrt{2k} \cdot \sqrt{ \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \widetilde{\mtx{\Omega}}_2^{\top} \widetilde{\mtx{\Omega}}_2 \rVert _2 \right] } && \text{($\lVert \mtx{B} \rVert _2 = \lVert \mtx{B} \mtx{B}^{\top} \rVert _2^{\sfrac{1}{2}} $ )} \notag \\
+        &\leq \sqrt{2k} \cdot \sqrt{ \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \widetilde{\mtx{\Omega}}_2^{\top} \widetilde{\mtx{\Omega}}_2 - \mtx{I}_l \rVert _2 \right] + \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{I}_l \rVert _2 \right]} && \text{(triangle inequality)} \notag \\
+        &\leq \sqrt{2k} \cdot \sqrt{ \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \widetilde{\mtx{\Omega}}_2^{\top} \widetilde{\mtx{\Omega}}_2 - \mtx{I}_l \rVert _2 \right]} + \sqrt{2k} &&
+         \text{($\sqrt{a+b} \leq \sqrt{a} + \sqrt{b}$, $\lVert \mtx{I}_l \rVert _2 = 1$)}
     \end{align}
-    By \cite{} $\widetilde{\mtx{\Omega}}_2 \in \mathbb{R}^{2k \times l}$ satisfies the \todo{$(c_1 \sqrt{q/2k}, c_2 e^{-q}, q)$}-JL moment property for \todo{any $q$ (in what?) $\geq \dots$} \todo{\cite[remark below lemma 4]{cohen-2016-optimal-approximate} or define} and some constants $c_1, c_2 > 0$, hence, by \cite[lemma 4]{cohen-2016-optimal-approximate} it satisfies the $(2 c_1 \sqrt{q/2k}, 9^d c_2 e^{-q}, d, q)$-OSE moment property for any $d \in \mathbb{N}$.
-For convenience, we choose $d=2k$ and $q=k/2$, such that $\widetilde{\mtx{\Omega}}_2$ has $(c_1, c_2 (3^4e^{-\sfrac{1}{2}})^k, 2k, k/2)$-OSE moment property.
-We may then apply an intermediary result from the proof of \cite[theorem 1]{cohen-2016-optimal-approximate} to bound
+    By \cite[section 2.2]{nelson-2017-sketching-algorithms} $\widetilde{\mtx{\Omega}}_2 \in \mathbb{R}^{2k \times l}$ satisfies the $(c_1 \sqrt{q/2k}, e^{-q}, q)$-JL moment property for any $q > 0$ and a universal constant $c_1 > 0$.
+Hence, by \cite[lemma 4]{cohen-2016-optimal-approximate} it satisfies the $(2 c_1 \sqrt{q/2k}, 9^d e^{-q}, d, q)$-OSE moment property for any $d \in \mathbb{N}$.
+For convenience, we choose $d=2k$ and $q=k/2$, such that $\widetilde{\mtx{\Omega}}_2$ has $(c_1, (3^4 e^{-\sfrac{1}{2}})^k, 2k, k/2)$-OSE moment property, i.e.
+\cite[definition 3]{cohen-2016-optimal-approximate}
     \begin{equation}
-        \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\widetilde{\mtx{\Omega}}_2 \mtx{A} )^{\top} (\widetilde{\mtx{\Omega}}_2 \mtx{A}) - \mtx{A}^{\top} \mtx{A} \rVert _2\right] \leq c_1 c_2^2 3^8e^{-1} \left( \lVert \mtx{A} \rVert _2^2 + \frac{\lVert \mtx{A} \rVert _F^2}{k} \right).
+        \mathbb{E}^{\sfrac{k}{2}} \left[ \lVert \widetilde{\mtx{\Omega}}_2 \widetilde{\mtx{\Omega}}_2^{\top} - \mtx{I}_l \rVert _2 \right] \leq c_1 (3^4 e^{-\sfrac{1}{2}})^2 = c_1 3^8 e^{-1}.
+        %\mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\widetilde{\mtx{\Omega}}_2 \mtx{A} )^{\top} (\widetilde{\mtx{\Omega}}_2 \mtx{A}) - \mtx{A}^{\top} \mtx{A} \rVert _2\right] \leq c_1 c_2^2 3^8e^{-1} \left( \lVert \mtx{A} \rVert _2^2 + \frac{\lVert \mtx{A} \rVert _F^2}{k} \right).
     \end{equation}
-    Inserting this bound into the above inequality and generously bounding all terms with a single constant we get
-    \begin{align}
-        \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \rVert _2 \right]
-        &\leq \sqrt{2k} \cdot \sqrt{c_1 e^{-1}} c_2 3^4 \left( \lVert \mtx{A} \rVert _2^2  + \frac{\lVert \mtx{A} \rVert _F^2}{k} \right)^{\sfrac{1}{2}} + \sqrt{2k} \cdot \lVert \mtx{A} \rVert _2 \notag \\
-        &\leq 2 \sqrt{2} (1 + \sqrt{c_1 e^{-1}} c_2 3^4)\left( \sqrt{k} \cdot \lVert \mtx{A} \rVert _2  + \lVert \mtx{A} \rVert _F \right).
+    Inserting this bound into the above inequality we get
+    \begin{equation}
+        \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_2 \rVert _2 \right]
+        \leq \sqrt{2k} \cdot (3^4 \sqrt{c_1 e^{-1}} + 1).
         \label{equ:OSE-moment-bound-first}
-    \end{align}
+    \end{equation}
 
-    The second term in \refequ{equ:OSE-moment-bound-initial} can be bounded with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
+    The moments of $\lVert \mtx{\Omega}^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
     \begin{equation}
@@
         \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-        \leq \underbrace{2 e^2\sqrt{3} (1 + \sqrt{c_1 e^{-1}} c_2 3^4)}_{\equiv c}\left( \lVert \mtx{A} \rVert _2  + \frac{\lVert \mtx{A} \rVert _F}{\sqrt{k}} \right).
+        \leq \underbrace{e^2\sqrt{3} (3^4 \sqrt{c_1 e^{-1}} + 1)}_{\equiv c} \lVert \mtx{A} \rVert _2.
     \end{equation}
@@
     \begin{equation}
-        \int_{a}^{b} \lVert \mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F~\mathrm{d}t \leq \gamma \sqrt{2} \cdot \frac{1 + 2c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t.
+        \int_{a}^{b} \lVert \mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F~\mathrm{d}t \leq \gamma \sqrt{2} \cdot \frac{1 + c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t.
     \end{equation}
@@
 
-\todo{Proof idea: structural bound, then higher order moment bound to apply Markov's inequality.}
+%\todo{Proof idea: structural bound, then higher order moment bound to apply Markov's inequality.}
 
@@
 $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$.
-After temporarily omitting the parameter dependence, we can apply \cite[theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound the integrand with \todo{maybe explain where $\mtx{\Omega}_i$ come from...}
+After temporarily omitting the parameter dependence, we can apply \cite[theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound the integrand with% \todo{maybe explain where $\mtx{\Omega}_i$ come from...}
     \begin{equation}
@@
     \end{equation}
-    Its higher order moments are bounded by
+    Its higher order moments are therefore bounded by
     \begin{align}
-        \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
-        &= \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]^2 && \text{(by definition of $\mathbb{E}^{k}$)} \notag \\
-        &\leq c \left( \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F \right)^2 && \text{(\reflem{lem:ose} with $\mtx{A}=\mtx{\Lambda}_2^{\sfrac{1}{2}}$)} \notag \\
-        &\leq c \left( \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2  + \frac{1}{k}\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right) && \text{($(a + b)^2 \leq a^2 + b^2$)} \notag \\
-        &= c \left( \lambda_{k+1} + \frac{1}{k}\Trace(\mtx{\Lambda}_2) \right) && \text{(definition of $\lVert \cdot \rVert _2$ and $\lVert \cdot \rVert _F$)} \notag \\
-        &\leq \frac{2 c}{k} \Trace(\mtx{A}) && \text{($\lambda_{k+1} \leq \frac{\Trace(\mtx{A})}{k}$, $\Trace(\mtx{\Lambda}_2) \leq \Trace(\mtx{A})$)}
+        \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F \right]
+        &\leq \sqrt{k} \cdot \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right] && \text{(using \refequ{equ:nystrom-proof-processed-tail})} \notag \\
+        &= \sqrt{k} \cdot \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]^2 && \text{(by definition of $\mathbb{E}^{k}$)} \notag \\
+        &\leq \sqrt{k} c \cdot \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 && \text{(\reflem{lem:ose} with $\mtx{A}=\mtx{\Lambda}_2^{\sfrac{1}{2}}$)} \notag \\
+        &= \sqrt{k} c \cdot \lambda_{k+1} && \text{(from the definition of $\lVert \cdot \rVert _2$)} \notag \\
+        &\leq \frac{c}{\sqrt{k}} \Trace(\mtx{A}) && \text{($\lambda_{k+1} \leq \frac{\Trace(\mtx{A})}{k}$)}
         \label{equ:nystrom-proof-tail-bound}
     \end{align}
-    Thus, inserting \refequ{equ:nystrom-proof-tail-bound} into \refequ{equ:nystrom-proof-processed-tail} and using it along with \refequ{equ:nystrom-proof-frobenius-trace} into \refequ{equ:nystrom-proof-persson-bonud} we obtain
+    Thus, using the triangle inequality for $\mathbb{E}^{\sfrac{k}{2}}$ and inserting \refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud} we obtain
     \begin{equation}
         \mathbb{E}^{\sfrac{k}{2}} \left[\lVert \mtx{A} - \Nystr{\mtx{\Omega}}{\mtx{A}} \rVert _F \right]
-        \leq \frac{1 + 2c}{\sqrt{k}} \Trace(\mtx{A})
+        \leq \frac{1 + c}{\sqrt{k}} \Trace(\mtx{A})
     \end{equation}
@@
         %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
-        \leq \frac{1 + 2c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t.
+        \leq \frac{1 + c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t.
     \end{equation}
@@
     \begin{equation}
-        \int_{a}^{b} \lVert \mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{1 + 2c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t,
+        \int_{a}^{b} \lVert \mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{1 + c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t,
     \end{equation}
@@
 
-\begin{figure}[ht]
-    \begin{subfigure}[b]{0.32\columnwidth}
-        \input{plots/gaussian-well-1.pgf}
-        \caption{$n_c=1$}
-        \label{fig:gaussian-well-1}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.32\columnwidth}
-        \input{plots/gaussian-well-2.pgf}
-        \caption{$n_c=2$}
-        \label{fig:gaussian-well-2}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.32\columnwidth}
-        \input{plots/gaussian-well-5.pgf}
-        \caption{$n_c=5$}
-        \label{fig:gaussian-well-5}
-    \end{subfigure}
-    \caption{Two dimensional periodic potential $V$ for different sizes $n_c$ of the computational domain.}
-    \label{fig:gaussian-well}
-\end{figure}
+%\begin{figure}[ht]
+%    \begin{subfigure}[b]{0.32\columnwidth}
+%        \input{plots/gaussian-well-1.pgf}
+%        \caption{$n_c=1$}
+%        \label{fig:gaussian-well-1}
+%    \end{subfigure}
+%    \begin{subfigure}[b]{0.32\columnwidth}
+%        \input{plots/gaussian-well-2.pgf}
+%        \caption{$n_c=2$}
+%        \label{fig:gaussian-well-2}
+%    \end{subfigure}
+%    \begin{subfigure}[b]{0.32\columnwidth}
+%        \input{plots/gaussian-well-5.pgf}
+%        \caption{$n_c=5$}
+%        \label{fig:gaussian-well-5}
+%    \end{subfigure}
+%    \caption{Two dimensional periodic potential $V$ for different sizes $n_c$ of the computational domain.}
+%    \label{fig:gaussian-well}
+%\end{figure}
 
 \begin{figure}[ht]
-    \begin{subfigure}[b]{0.49\columnwidth}
-        \input{plots/convergence-degree.pgf}
-        \caption{$n_{\mtx{\Omega}} + n_{\mtx{\Psi}} = 160$}
-        \label{convergence-randvec}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.49\columnwidth}
-        \input{plots/convergence-randvec.pgf}
-        \caption{$m = 2400$}
-        \label{fig:convergence-degree}
-    \end{subfigure}
-    \caption{For increasing values of s$n_{\mtx{\Omega}} + n_{\mtx{\Psi}}$ but fixed $m$ we plot the $L^1$ relative approximation error for the model problem with $\sigma=0.05$.}
+    \centering
+    \input{plots/convergence.pgf}
+    \caption{For increasing values of $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}$ but fixed $m$ we plot the $L^1$ relative approximation error for the model problem with $\sigma=0.005$.}
     \label{fig:convergence}
@@
     \centering
-    \input{plots/matvec-mix.pgf}
+    \input{plots/distribution.pgf}
     \caption{The NC++ method for different ways of allocations a 
-    total of $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}=80$ random vectors
-    to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation
-    for the Gaussian smoothing kernel with multiple different values of
-    the smoothing parameter. We make the approximation error made in the
-    Chebyshev expansion negligible by rescaling $m=120 / \sigma$.}
+    total of $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation for the Gaussian smoothing kernel with multiple different values of the smoothing parameter.
+We make the approximation error made in the Chebyshev expansion negligible by rescaling $m=16 / \sigma$ (based on \refthm{thm:chebyshev-error}).}
     \label{fig:matvec-mix}
@@
 \begin{itemize}
-    \item Test
+    \item Robustness (can choose $m$ and $n_{\mtx{\Omega}}$, $n_{\mtx{\Psi}}$ as large as one wants, and not break algorithm, unlike Krylov-aware Lanczos)
 \end{itemize}

### 2024-06-03 paper/plots/convergence.py

@@ -12,19 +12,19 @@ np.random.seed(0)
 A = sp.sparse.load_npz("matrices/ModES3D_1.npz")
 
 # Perform spectral transform with A and its eigenvalues
-eigvals = np.linalg.eigvalsh(A)#.toarray())
+eigvals = np.linalg.eigvalsh(A.toarray())
 min_ev, max_ev = eigvals[0], eigvals[-1]
 A_st = spectral_transformation(A, min_ev, max_ev)
 eigvals_st = spectral_transformation(eigvals, min_ev, max_ev)
 
 # Set parameter
 t = np.linspace(-1, 1, 100)
-sigma = 0.2
-m = 500
-n_Vec_list = np.logspace(0.95, 2.6, 2).astype(int) 
+sigma = 0.005
+m = 2000
+n_Vec_list = np.logspace(1.3, 2.8, 7).astype(int) 
 
 plt.style.use("paper/plots/stylesheet.mplstyle")
-colors = cm.magma(np.arange(5) / 4.5)
+colors = cm.magma(np.arange(3) / 2.5)
 markers = ["o", "^", "s"]
 labels = [r"$n_{\mathbf{\Omega}} = 0$",
           r"$n_{\mathbf{\Psi}} = n_{\mathbf{\Omega}}$",
@@ -51,9 +51,10 @@ plt.text(8e+1, 5e-3, r"$\mathcal{O}(\varepsilon^{-1})$", color="#7a7a7a")
 plt.plot(n_Vec_list, 0.45/n_Vec_list**(0.5), linestyle="dashed", color="#7a7a7a", alpha=0.5)
 plt.text(6e+1, 9.5e-2, r"$\mathcal{O}(\varepsilon^{-2})$", color="#7a7a7a")
 
-plt.grid()
+plt.grid(True, which="both")
 plt.ylabel(r"$L^1$ error")
 plt.xlabel(r"estimator size $n_{\mathbf{\Psi}} + n_{\mathbf{\Omega}}$")
 plt.legend()
 plt.xscale("log")
 plt.yscale("log")
+plt.savefig("paper/plots/convergence.pgf", bbox_inches="tight")

### 2024-06-03 paper/plots/distribution.py

@@ -19,7 +19,7 @@ eigvals_st = spectral_transformation(eigvals, min_ev, max_ev)
 
 # Set parameter
 t = np.linspace(-1, 1, 100)
-sigma_list = np.logspace(-2.5, -0.5, 7)
+sigma_list = np.logspace(-3.0, -1.0, 7)
 n_Vec = 80
 n_Psi_list = np.arange(n_Vec + 1, step=20).astype(np.int64)
 n_Omega_list = n_Vec - n_Psi_list
@@ -35,19 +35,19 @@ for j, sigma in enumerate(sigma_list):
     # Determine the baseline spectral density
     kernel = lambda x: gaussian_kernel(x, sigma=sigma)
     baseline = form_spectral_density(eigvals_st, t, kernel)
-    m = int(120 / sigma)
+    m = int(16 / sigma)  # Double-check this, i.e. use error estimate for this
 
     for i, (n_Psi, n_Omega) in enumerate(zip(n_Psi_list, n_Omega_list)):
         estimate = spectral_density(A_st, t, m, n_Psi, n_Omega, kernel)
         error[i, j] = 2 * np.mean(np.abs(estimate - baseline))
 
-for i in range(len(n_Psi_list)):
+for i in reversed(range(len(n_Psi_list))):
     plt.plot(sigma_list, error[i], color=colors[i], marker=markers[i], label=labels[i])
 
-plt.grid()
+plt.grid(True, which="both")
 plt.ylabel(r"$L^1$ error")
 plt.xlabel(r"smoothing parameter $\sigma$")
 plt.legend()
 plt.xscale("log")
 plt.yscale("log")
-
+plt.savefig("paper/plots/distribution.pgf", bbox_inches="tight")

### 2024-06-03 src/algorithms.py

@@ -280,6 +280,58 @@ def NCPP(A, t, m, sigma, n_v, n_v_tilde=None, k=1, zeta=1e-7, kappa=1e-5, eta=1e
     return phi_breve - rho * n
 
 
+
+def ChebyshevNystrom(A, t, m=100, n_Omega=10, n_Psi=10, sigma=0.1, nonnegative=False, seed=0):
+    """
+    Chebyshev-Nystrom++ method
+    """
+    # Seed the random number generator
+    np.random.seed(seed)
+
+    # Convert evaluation point(s) to numpy array
+    if not isinstance(t, np.ndarray):
+        t = np.array(t).reshape(-1)
+
+    # Determine size of matrix i.e. number of eigenvalues 
+    n = A.shape[0]
+    n_t = t.shape[0]
+
+    # Compute coefficients of Chebyshev expansion of the smoothing kernel
+    g = lambda x: gaussian_kernel(x, n=n, sigma=sigma)
+    mu = chebyshev_coefficients(t, m, function=g, nonnegative=nonnegative)
+    nu = exponentiate_chebyshev_coefficients_cosine_transform(mu, k=2)
+
+    # Generate Gaussian random matrices
+    Omega = np.random.randn(n, n_Omega)
+    Psi = np.random.randn(n, n_Psi)
+
+    # Initialize Chebyshev recurrence and helper matrices
+    V_1, V_2, V_3 = np.zeros((n, n_Omega)), Omega, np.zeros((n, n_Omega))
+    W_1, W_2, W_3 = np.zeros((n, n_Psi)), Psi, np.zeros((n, n_Psi))
+    K_1, K_2 = np.zeros((n_t, n_Omega, n_Omega)), np.zeros((n_t, n_Omega, n_Omega))
+    L_1, l_2 = np.zeros((n_t, n_Omega, n_Psi)), np.zeros(n_t)
+
+    for l in range(2 * m + 1):
+        X, Y = Omega.T @ V_2, Omega.T @ W_2
+        z = np.sum(Psi * W_2)
+        for i in range(n_t):
+            if l <= m:
+                K_1[i] += mu[i, l] * X
+                L_1[i] += mu[i, l] * Y
+                l_2[i] += mu[i, l] * z
+            K_2[i] += nu[i, l] * X
[…]

### 2024-06-03 src/metrics.py

@@ -59,6 +59,65 @@ def p_norm(phi, phi_tilde, p=1, relative=True):
     return error
 
 
+def wasserstein_distance(phi, phi_tilde, a=-1, b=1):
+    """
+    Compute the Wasserstein-1 distance between the spectral densities.
+
+    Parameters
+    ----------
+    phi : np.ndarray of shape (n,)
+        The true DOS evaluated at a series of uniformly distributed points.
+    phi_tilde : np.ndarray of shape (n,)
+        The approximated DOS evaluated at the same points as phi.
+    a : float or int
+        The start-point of the support of the spectral density.
+    b : float or int > a
+        The end-point of the support of the spectral density.
+
+    Returns
+    -------
+    error : float
+        The error of the approximated DOS from the actual DOS.
+    """
+    dx = (b - a) / (len(phi) - 1)
+    Phi = sp.integrate.cumulative_simpson(phi, dx=dx)
+    Phi_tilde = sp.integrate.cumulative_simpson(phi_tilde, dx=dx)
+    error = sp.integrate.simpson(np.abs(Phi - Phi_tilde), dx=dx)
+    return error
+
+
+def wasserstein_error(eigenvalues, phi_tilde, a=-1, b=1, normalize=False):
+    """
+    Compute the Wasserstein-1 of the spectral density from the ground truth.
+
+    Parameters
+    ----------
+    eigenvalues : np.ndarray of shape (n,)
+        The true eigenvalues of the matrix.
+    phi_tilde : np.ndarray of shape (n,)
+        The approximated DOS evaluated at the same points as phi.
+    a : float or int
+        The start-point of the support of the spectral density.
+    b : float or int > a
+        The end-point of the support of the spectral density.
+    normalize : bool
+        Whether to enforce normalization of the approximate spectral density.
+
+    Returns
+    -------
+    error : float
+        The error of the approximated DOS from the actual DOS.
+    """
+    dx = (b - a) / (len(phi_tilde) - 1)
+    Phi_x = lambda x: np.sum(np.subtract.outer(eigenvalues, x) < 0, axis=0) / len(eigenvalues)
+    Phi = Phi_x(np.linspace(a, b, len(phi_tilde)))
+    Phi_tilde = sp.integrate.cumulative_simpson(phi_tilde, dx=dx, initial=0)
+    if normalize:
+        Phi_tilde /= Phi_tilde[-1]
[…]

### 2024-06-03 src/plots.py

@@ -139,7 +139,7 @@ def plot_spectral_densities(spectral_densities, parameters, variable_parameter=N
     return ax
 
 
-def compute_spectral_density_errors(A, methods, labels, variable_parameter, variable_parameter_values, parameters, kernel=gaussian_kernel, n_t=1000, error_metric=p_norm, correlated_parameter=None, correlated_parameter_values=None, eigenvalues=None):
+def compute_spectral_density_errors(A, methods, labels, variable_parameter, variable_parameter_values, parameters, kernel=gaussian_kernel, n_t=1000, error_metric=p_norm, correlated_parameter=None, correlated_parameter_values=None, eigenvalues=None, smooth_baseline=True):
     """
     Compute the errors of the spectral densities of a matrix A using different
     methods or parameters.
@@ -168,6 +168,8 @@ def compute_spectral_density_errors(A, methods, labels, variable_parameter, vari
         The values which the correlated parameter depending on the variable parameter.
     eigenvalues : np.ndarray
         Eigenvalues which are used to form the baseline spectral density.
+    smooth_baseline : bool
+        Compute the metric from the smoothened or original spectral density.
     
     Returns
     -------
@@ -222,8 +224,11 @@ def compute_spectral_density_errors(A, methods, labels, variable_parameter, vari
             except:
                 pass
             parameter[variable_parameter] = param
-            spectral_density_baseline = form_spectral_density(eigenvalues_transformed, kernel=kernel, n=A_transformed.shape[0], n_t=n_t, sigma=parameter["sigma"])
             spectral_density = method(A=A_transformed, t=t, **parameter)
+            if smooth_baseline:
+                spectral_density_baseline = form_spectral_density(eigenvalues_transformed, kernel=kernel, n=A_transformed.shape[0], n_t=n_t, sigma=parameter["sigma"])
+            else:
+                spectral_density_baseline = eigenvalues_transformed
             spectral_density_errors[label].append(error_metric(spectral_density_baseline, spectral_density))
 
     return spectral_density_errors
@@ -293,6 +298,7 @@ def plot_spectral_density_errors(spectral_density_errors, parameters, variable_p
         ax.plot(variable_parameter_values, spectral_density_error, linewidth=1, color=colors[i], label=label, marker=markers[i], linestyle=linestyles[i])
     ax.legend()
     return ax
+
 
 
 # --- Unused implementations ---

### 2024-06-04 paper/main.tex
make corrections after meeting

@@
 \begin{itemize}
-    \item Change all parameter-dependent matrices to $\mtx{B}(t)$ to avoid confusion with $\mtx{A}$
     \item Change citation style to \textbf{[1]}, \textbf{[2]}, \dots
@@
 Some of these estimators have already been successfully applied to parameter-dependent matrices \cite{lin-2017-randomized-estimation}, but they have not yet been analyzed.
-A related technique is dynamic trace estimation for sequences $\mtx{B}_1, \dots, \mtx{B}_m$ of similar matrices \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace}, which achieves guarantees for the estimate $b_i$ of $\Trace(\mtx{B}_i)$ of the form $\mathbb{P}(|\Trace(\mtx{B}_i) - b_i| \geq \varepsilon) \leq \delta$ for $i=1,\dots, m$.
+\todo{A related technique is dynamic trace estimation for sequences $\mtx{B}_1, \dots, \mtx{B}_m$ of similar matrices \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace}, which achieves guarantees for the estimate $b_i$ of $\Trace(\mtx{B}_i)$ of the form $\mathbb{P}(|\Trace(\mtx{B}_i) - b_i| \geq \varepsilon) \leq \delta$ for $i=1,\dots, m$.
 A drawback of this type of bound is that when we aim for $\varepsilon$ accuracy for all $i=1,\dots, m$ simultaneously, the probability will grow linearly with $m$ by the union bound.
-Therefore, instead of discretizing $[a, b]$ with $t_1, \ldots, t_m$ and looking at the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way to quantify the error is to use $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate for each $t \in [a, b]$.
+Therefore, instead of discretizing $[a, b]$ with $t_1, \ldots, t_m$ and looking at the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way to quantify the error is to use $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate for each $t \in [a, b]$.}
 
@@
 Subsequently, we propose a simple, fast, robust, and provable procedure in which these estimators are employed for approximating the smooth spectral density $\phi_{\sigma}$ of a matrix.
-
-\paragraph{Notation}
-We follow the notation and conventions which are used in contemporary literature in this field.
-Further, we denote with $\mathbb{N}$ all positive whole numbers; $e$ denotes Euler's number and $\log$ is the corresponding logartihm; the ceiling function $\lceil x \rceil$ of a real number $x$ returns the greatest integer larger than or equal to $x$; scalars are represented by lower case Greek or Latin letters ($s$, $\varepsilon$, \dots), vectors are additionally printed in bold ($\vct{v}$, $\vct{\psi}$, \dots), and matrices are additionally capitalized ($\mtx{A}$, $\mtx{\Omega}$, \dots); the expected value of a random variable $x$ is denoted with $\mathbb{E}[x]$; the $k$-th moment of a random variable $x$ is defined as $\mathbb{E}^{k}[x] = \left(\mathbb{E}\left[ | x |^{k} \right] \right)^{\sfrac{1}{k}}$; the identity matrix $\mtx{I}_n = \diag(1, \dots, 1) \in \mathbb{R}^{n \times n}$ carries ones on its diagonal and zeros everywhere else; the zero matrix $\mtx{0}_{n \times m} \in \mathbb{R}^{n \times m}$ consists of only zero entries; in a Gaussian random vector and a Gaussian random matrix, all entries are independently sampled from the standard normal distribution; matrix norms are denoted with $\lVert \cdot \rVert$; for symmetric matrices, the $p$-Schatten norm is $\lVert \mtx{A} \rVert _{(p)} = (\sum_{i=1}^n |\lambda_i|^p)^{1/p}$; we refer to the spectral norm as $\lVert \mtx{A} \rVert _2 = \lVert \mtx{A} \rVert _{(\infty)} = \max_{i=1,\dots,n}|\lambda_i|$, the nuclear norm as $\lVert \mtx{A} \rVert _{\ast} = \lVert \mtx{A} \rVert _{(1)} = \sum_{i=1}^{n} |\lambda_i|$, and the Frobenius norm as $\lVert \mtx{A} \rVert _F = \lVert \mtx{A} \rVert _{(2)} = (\sum_{i=1}^{n} |\lambda_i|^2)^{\sfrac{1}{2}}$.
 
@@
 \begin{equation}
-    \mtx{A}(t) = \begin{bmatrix}
-        a_{11}(t) & a_{12}(t) & \dots & a_{1n}(t) \\
-        a_{21}(t) & a_{22}(t) & \dots & a_{2n}(t) \\
+    \mtx{B}(t) = \begin{bmatrix}
+        b_{11}(t) & b_{12}(t) & \dots & b_{1n}(t) \\
+        b_{21}(t) & b_{22}(t) & \dots & b_{2n}(t) \\
         \vdots & \vdots & \ddots & \vdots \\
-        a_{n1}(t) & a_{n2}(t) & \dots & a_{nn}(t) \\
+        b_{n1}(t) & b_{n2}(t) & \dots & b_{nn}(t) \\
     \end{bmatrix} \in \mathbb{R}^{n \times n}
 \end{equation}
-where $a_{ij}(t)$ are functions depending continuously on the parameter $t$ which takes values in the interval $[a,b]$. The trace of such a matrix is defined as
+where $b_{ij}(t)$ are functions depending continuously on the parameter $t$ which takes values in the interval $[a,b]$. The trace of such a matrix is defined as
 \begin{equation}
-    \Trace(\mtx{A}(t)) = \sum_{i=1}^{n} a_{ii}(t).
+    \Trace(\mtx{B}(t)) = \sum_{i=1}^{n} b_{ii}(t).
     \label{equ:definition-trace}
@@
 
-\paragraph{Girard-Hutchinson estimator} We can estimate the trace with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ to form
+\paragraph{Girard-Hutchinson estimator} We can estimate the trace with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ stochastically independent standard Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ to form
 \begin{equation}
-    \Hutch{\mtx{\Psi}}(\mtx{A}(t))
-    = \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j
-    = \frac{1}{n_{\mtx{\Psi}}} \Trace( \mtx{\Psi}^{\top} \mtx{A}(t) \mtx{\Psi})
+    \Hutch{\mtx{\Psi}}(\mtx{B}(t))
+    = \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j
+    = \frac{1}{n_{\mtx{\Psi}}} \Trace( \mtx{\Psi}^{\top} \mtx{B}(t) \mtx{\Psi})
     \label{equ:hutchinson-trace-estimator}
@@
 \begin{equation}
-    \Nystr{\mtx{\Omega}}{\mtx{A}}(t) = (\mtx{A}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} (\mtx{A}(t) \mtx{\Omega})^{\top}.
+    \Nystr{\mtx{\Omega}}{\mtx{B}}(t) = (\mtx{B}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} (\mtx{B}(t) \mtx{\Omega})^{\top}.
     \label{equ:nystrom-approximation}
 \end{equation}
-Then we can estimate the trace as $\Trace(\Nystr{\mtx{\Omega}}{\mtx{A}}(t))$.
+Then we can estimate the trace as $\Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t))$.
 Thanks to the invariance of the trace under cyclic permutation of its arguments and the symmetry of the matrix, we may rewrite this estimator as
@@
 \begin{equation}
-    \Trace(\Nystr{\mtx{\Omega}}{\mtx{A}}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{A}(t)^2 \mtx{\Omega})).
+    \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{B}(t)^2 \mtx{\Omega})).
     \label{equ:nystrom-trace-estimator}
@@
 \begin{equation}
-    \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{A}(t)) = \Trace(\Nystr{\mtx{\Omega}}{\mtx{A}}(t)) + \Hutch{\mtx{\Psi}}(\mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)).
+    \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) = \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) + \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)).
     \label{equ:nystrompp-trace-estimator}
 \end{equation}
-This is the analogue of the Nyström++ estimator \cite{persson-2022-improved-variants} in the parameter-dependent setting.
+This is the parameter-dependent analogue of the Nyström++ estimator \cite{persson-2022-improved-variants}, which is based on the Hutch++ estimator \cite{meyer-2021-hutch-optimal}.
 %We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.
-
-\paragraph{$L^{1}$ norm}
-We define the $L^{1}$ norm of a measurable function $f:[a,b] \to \mathbb{R}$ as
-\begin{equation}
-    \int_{a}^{b} | f(t) | \mathrm{d}t.
-\end{equation}
-We use it to measure the error of the trace estimators. The following sections provide bounds on this error for each of the above-mentioned estimators.
 
@@
 
-We first consider the estimator \refequ{equ:hutchinson-trace-estimator}.
+We first consider the estimator \refequ{equ:hutchinson-trace-estimator}. \todo{Introduce $L^1$ norm and Frobenius norm}
 
 \begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
-    Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ be symmetric and continuously depend on $t \in [a, b]$.
+    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric and continuously depend on $t \in [a, b]$.
 Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ it holds with probability at least $1 - \gamma^{-2k}$
     \begin{equation}
-        \int_{a}^{b} \left| \Trace(\mtx{A}(t)) - \Hutch{\mtx{\Psi}}(\mtx{A}(t)) \right| ~\mathrm{d}t < 2^4 \gamma k \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F  ~\mathrm{d}t.
+        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) \right| ~\mathrm{d}t < 2^4 \gamma k \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F  ~\mathrm{d}t.
     \end{equation}
-    In particular, if we choose $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{- \sfrac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{A}(t)) - \Hutch{\mtx{\Psi}}(\mtx{A}(t)) | ~\mathrm{d}t < \varepsilon \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t$.
+    In particular, if we choose $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{- \sfrac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t < \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$.
 \end{theorem}
@@
     \begin{equation}
-        r(t) = \Trace(\mtx{B}(t)) - \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi}
+        r(\mtx{B}(t), \vct{\psi}) = \Trace(\mtx{B}(t)) - \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi}
     \end{equation}
@@
 
-    First, we consider $r(t)$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence.
-From the proof of \cite[lemma 3]{cortinovis-2022-randomized-trace} we know that $r$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$.
+    First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence.
+From the proof of \cite[lemma 3]{cortinovis-2022-randomized-trace} we know that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ \todo{introduce spectral norm}.
 Thus, by \cite[theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
     \begin{align}
-        \mathbb{E}\left[ r^{2 k} \right]
+        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
         &\stackrel{\text{\cite{boucheron-2013-basic-inequalities}}}{\leq} k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k} \notag \\
@@
     \begin{equation}
-        \mathbb{E}\left[ r^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
+        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
     \end{equation}
@@
 < 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$.
-Consequently, the moments of $r$ for a fixed $t$ are limited by
+Consequently, the moments $\mathbb{E}^{k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{k} \right] \right)^{\sfrac{1}{k}}$ of $r(\mtx{B}, \vct{\psi})$ are limited by
     \begin{equation}
-        \mathbb{E}^{2k}\left[ r \right]
+        \mathbb{E}^{2k}\left[ r(\mtx{B}, \vct{\psi}) \right]
         < \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F < 2^4 k \lVert \mtx{B} \rVert _F,
@@
     \begin{equation}
-        \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(t)|~\mathrm{d}t  \right]
-        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}^{2 k}\left[ r(t) \right]~\mathrm{d}t
+        \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t  \right]
+        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}^{2 k}\left[ r(\mtx{B}(t), \vct{\psi}) \right]~\mathrm{d}t
         < 2^4 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
@@
     \begin{equation}
-        \int_{a}^{b} |r(t)|~\mathrm{d}t < 2^4 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
+        \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t < 2^4 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
         \label{equ:hutchinson-trace-onequery-uniform}
@@
     \begin{equation}
-        \mtx{B}(t)
+        \widetilde{\mtx{B}}(t)
         = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
-            \mtx{A}(t) & & \\
+            \mtx{B}(t) & & \\
             & \ddots & \\
-            & & \mtx{A}(t)
+            & & \mtx{B}(t)
         \end{pmatrix}
         \quad \text{and} \quad
-        \vct{\psi} = \begin{pmatrix}
+        \widetilde{\vct{\psi}} = \begin{pmatrix}
             \vct{\psi}_1 \\
@@
     where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors.
-Then $\lVert \mtx{B}(t) \rVert _F = \lVert \mtx{A}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ and
+Then $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$, and
     \begin{equation}
-        r(t) = \Trace(\mtx{A}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j.
+        r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}}) = \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j.
     \end{equation}
-    Hence, we conclude from \refequ{equ:hutchinson-trace-onequery-uniform} that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
+    Hence, by applying \refequ{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$ we conclude that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
     \begin{equation}
-        \int_{a}^{b} \left| \Trace(\mtx{A}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j \right| ~ \mathrm{d}t
-        < 2^4 k \gamma \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t.
+        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j \right| ~ \mathrm{d}t
+        < 2^4 k \gamma \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
     \end{equation}
 
-    Setting $\delta = \gamma^{-2 k}$ and choosing $k = \lceil \log(\delta^{-1}) \rceil$ we get that for all $\delta \in (0, e^{-\sfrac{1}{2}})$
+    Setting $\delta = \gamma^{-2 k}$ and choosing $k = \lceil \log(\delta^{-1}) \rceil$, where $\lceil x \rceil$ is the ceiling function which returns the greatest integer larger than or equal to $x$, we get that for all $\delta \in (0, e^{-\sfrac{1}{2}})$
     \begin{equation}
@@
     %\begin{equation}
-    %    \int_{a}^{b} \left| \frac{1}{n_{\mtx{\Psi}}} \sum_{i=1}^{n_{\mtx{\Psi}}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Trace(\mtx{A}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\sfrac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t
+    %    \int_{a}^{b} \left| \frac{1}{n_{\mtx{\Psi}}} \sum_{i=1}^{n_{\mtx{\Psi}}} \vct{\psi}_i^{\top} \mtx{B}(t) \vct{\psi}_i - \Trace(\mtx{B}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\sfrac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
     %\end{equation}
@@
 \begin{remark}
-    If additionally $\mtx{A}(t)$ is nonzero and symmetric positive semi-definite for all $t \in [a, b]$, then it follows from the proof of \refthm{thm:hutchinson} that with probability at least $1 - \delta$
+    If additionally $\mtx{B}(t)$ is nonzero and symmetric positive semi-definite for all $t \in [a, b]$, then it follows from the proof of \refthm{thm:hutchinson} that with probability at least $1 - \delta$
     \begin{equation}
-        \int_{a}^{b} \frac{| \Trace(\mtx{A}(t)) - \Hutch{\mtx{\Psi}}(\mtx{A}(t)) |}{\Trace(\mtx{A}(t))} ~\mathrm{d}t < \varepsilon,
+        \int_{a}^{b} \frac{| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) |}{\Trace(\mtx{B}(t))} ~\mathrm{d}t < \varepsilon,
     \end{equation}
-    if $n_{\mtx{\Psi}} \geq 2^{10} \varepsilon^{-2} \log(\delta^{-1})^2 (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ with $\mu(t) = \lVert \mtx{A}(t) \rVert _2 / \Trace(\mtx{A}(t))$.
+    if $n_{\mtx{\Psi}} \geq 2^{10} \varepsilon^{-2} \log(\delta^{-1})^2 (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ with $\mu(t) = \lVert \mtx{B}(t) \rVert _2 / \Trace(\mtx{B}(t))$.
 Compared to the constant matrix equivalent \cite[remark 2]{cortinovis-2022-randomized-trace} we mainly observe a larger constant and an additional factor of $\log(\delta^{-1})$.
@@
 The Girard-Hutchinson estimator distinguishes itself for its simplicity and provability for all symmetric matrices.
-However, the $n_{\mtx{\Psi}} = \mathcal{\varepsilon^{-2}}$ requirement implies that if we aim to increase the accuracy of the estimate by one digit, we would need to increase the number of queries by a factor of $100$.
-The next two estimators improve on this by exploiting the structure of $\mtx{A}(t)$.
+However, the $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ requirement implies that if we aim to increase the accuracy of the estimate by one digit, we would need to increase the number of queries by a factor of $100$.
+The next two estimators improve on this by exploiting the structure of $\mtx{B}(t)$.
 
@@
 \begin{theorem}{Nyström estimator for parameter-dependent matrices}{nystrom}
-    Suppose $f(\mtx{A}, t)$ is continuous and non-negative in $t \in [a, b]$ for a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$.
+    Suppose $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ is continuous and positive semi-definite in $t \in [a, b]$.
 Let $n_{\mtx{\Omega}} \geq 6, n_{\mtx{\Omega}} - 4 \geq r \geq 2$ be two integers.
@@
     \begin{equation}
-        \int_{a}^{b} \left| \Trace(f(\mtx{A}, t)) - \Trace(\Nystr{\mtx{\Omega}}{f}(\mtx{A}, t)) \right| ~\mathrm{d}t
-        \leq \gamma^2 (1 + r) \int_{a}^{b} \sum_{i = r+1}^{n} \sigma_i(f(\mtx{A}, t)) ~\mathrm{d}t.
+        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right| ~\mathrm{d}t
+        \leq \gamma^2 (1 + r) \int_{a}^{b} \sum_{i = r+1}^{n} \sigma_i(\mtx{B}(t)) ~\mathrm{d}t.
     \end{equation}
@@
 \begin{proof}
-    Since $f$ is non-negative, $f(\mtx{A}, t)$ is positive semi-definite for all $t$.
-By \cite[lemma 2.1]{frangella-2023-randomized-nystrom}, $f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t)$ is also positive semi-definite for all $t$.
+    Since $f$ is non-negative, $\mtx{B}(t)$ is positive semi-definite for all $t$.
+By \cite[lemma 2.1]{frangella-2023-randomized-nystrom}, $\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ is also positive semi-definite for all $t$.
 Therefore, for fixed $t \in [a, b]$
     \begin{equation}
-        \left| \Trace(f(\mtx{A}, t)) - \Trace(\widehat{f}(\mtx{A}, t)) \right|
-        = \left| \Trace(f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t)) \right|
-        = \lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast}
+        \left| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right|
+        = \left| \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right|
+        = \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _{\ast},
     \end{equation}
-    From the proof of \cite[corollary 8.2]{tropp-2023-randomized-algorithms} it follows that
+    \todo{introduce nuclear norm}.  From the proof of \cite[corollary 8.2]{tropp-2023-randomized-algorithms} it follows that
     \begin{equation}
         %\lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _2^2 \leq \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _F^2
-        \lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _F^2
+        \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{\mtx{B}(t)^{\sfrac{1}{2}} \mtx{\Omega}}) \mtx{B}(t)^{\sfrac{1}{2}} \rVert _F^2.
     \end{equation}
-    Hence, by \cite[theorem 9]{kressner-2023-randomized-lowrank}, with $n_{\mtx{\Omega}} = p + r$, the result follows by noting $\sigma_i(f^{\sfrac{1}{2}}(\mtx{A}, t))^{2} = \sigma_i(f(\mtx{A}, t))$.
+    Hence, by \cite[theorem 9]{kressner-2023-randomized-lowrank}, with $n_{\mtx{\Omega}} = p + r$, the result follows by noting $\sigma_i(\mtx{B}(t)^{\sfrac{1}{2}})^{2} = \sigma_i(\mtx{B}(t))$.
 \end{proof}
 
-\todo{Explain that \reflem{lem:nystrom} is a weaker result, and mainly there to help proof \refthm{thm:nystrom-pp}.}
+\subsection{Nyström++ estimator for parameter-dependent matrices}
+\label{subsec:nystrom-pp}
+
+\todo{Motivate and explain proof technique (to mimic Hutch++)}
 
@@
 \begin{equation}
-    \mtx{A}(t) 
+    \mtx{B}(t) 
     = \mtx{U}(t) \mtx{\Lambda}(t) \mtx{U}(t)^{\top} 
 \end{equation}
-be the eigenvalue decomposition of $\mtx{A}(t)$ at a fixed $t$, i.e.
+be the eigenvalue decomposition of $\mtx{B}(t)$ at a fixed $t$, i.e.
 $\mtx{\Lambda}(t) = \operatorname{diag}(\lambda_1(t), \dots, \lambda_n(t))$ and $\mtx{U}(t)^{\top} \mtx{U}(t) = \mtx{I}_n$.
@@
 \begin{lemma}{Moments of tail bound}{ose}
-    A matrix $\mtx{A} \in \mathbb{R}^{l \times l}$ and independent standard Gaussian random matrices $\mtx{\Omega}_1 \in \mathbb{R}^{k \times 2k}, \mtx{\Omega}_2 \in \mathbb{R}^{l \times 2k}$ satisfy for all $k \in \mathbb{N}$ and a universal constant $c > 0$
+    \todo{A matrix $\mtx{A} \in \mathbb{R}^{l \times l}$ and }independent standard Gaussian random matrices $\mtx{\Omega}_1 \in \mathbb{R}^{k \times 2k}, \mtx{\Omega}_2 \in \mathbb{R}^{l \times 2k}$ satisfy for all $k \in \mathbb{N}$ and a universal constant $c > 0$
     \begin{equation}
@@
         &= \sqrt{2k} \cdot \mathbb{E}^{k}\left[ \lVert \widetilde{\mtx{\Omega}}_2^{\top} \rVert _2 \right] && \text{(definition of $\widetilde{\mtx{\Omega}}_2$)} \notag \\
-        &= \sqrt{2k} \cdot \sqrt{ \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \widetilde{\mtx{\Omega}}_2^{\top} \widetilde{\mtx{\Omega}}_2 \rVert _2 \right] } && \text{($\lVert \mtx{B} \rVert _2 = \lVert \mtx{B} \mtx{B}^{\top} \rVert _2^{\sfrac{1}{2}} $ )} \notag \\
+        &= \sqrt{2k} \cdot \sqrt{ \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \widetilde{\mtx{\Omega}}_2^{\top} \widetilde{\mtx{\Omega}}_2 \rVert _2 \right] } && \text{($\lVert \mtx{B} \rVert _2 = \lVert \mtx{B}^{\top} \mtx{B} \rVert _2^{\sfrac{1}{2}} $ )} \notag \\
         &\leq \sqrt{2k} \cdot \sqrt{ \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \widetilde{\mtx{\Omega}}_2^{\top} \widetilde{\mtx{\Omega}}_2 - \mtx{I}_l \rVert _2 \right] + \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{I}_l \rVert _2 \right]} && \text{(triangle inequality)} \notag \\
@@
 
-    The moments of $\lVert \mtx{\Omega}^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
+    The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
     \begin{equation}
@@
 \begin{lemma}{Nyström approximation for parameter-dependent matrices}{nystrom}
-    Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-Then its Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{A}}(t)$ with Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ for \emph{even} $n_{\mtx{\Omega}} \in \mathbb{N}$ and $\gamma \geq 1$ verifies with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
+    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
+Then its Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ with Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ for \emph{even} $n_{\mtx{\Omega}} \in \mathbb{N}$ and $\gamma \geq 1$ verifies with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
     \begin{equation}
-        \int_{a}^{b} \lVert \mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F~\mathrm{d}t \leq \gamma \sqrt{2} \cdot \frac{1 + c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t.
+        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \sqrt{2} \cdot \frac{1 + c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
     \end{equation}
     for a universal constant $c > 0$.
-In particular, if we choose $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})), n_{\mtx{\Omega}} \geq 2$, then with probability at least $1-\delta$ for $\delta \in (0, e^{-e^{4 \varepsilon^2}})$ and $\varepsilon > 0$ we have $\int_{a}^{b} \lVert \mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(A(t))~\mathrm{d}t$.
+In particular, if we choose $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})), n_{\mtx{\Omega}} \geq 2$, then with probability at least $1-\delta$ for $\delta \in (0, e^{-e^{4 \varepsilon^2}})$ and $\varepsilon > 0$ we have $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$.
 \end{lemma}
@@
     \begin{equation}
-        \lVert \mtx{A} - \Nystr{\mtx{\Omega}}{\mtx{A}} \rVert _F 
+        \lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F 
         \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F.
@@
         \leq \sqrt{ \lambda_{k+1} \sum_{i=k+1}^{n} \lambda_i}
-        \leq \sqrt{ \frac{\Trace(\mtx{A})}{k} \Trace(\mtx{A})}
-        \leq \frac{1}{\sqrt{k}} \Trace(\mtx{A}).
+        \leq \sqrt{ \frac{\Trace(\mtx{B})}{k} \Trace(\mtx{B})}
+        \leq \frac{1}{\sqrt{k}} \Trace(\mtx{B}).
         \label{equ:nystrom-proof-frobenius-trace}
@@
         &= \sqrt{k} c \cdot \lambda_{k+1} && \text{(from the definition of $\lVert \cdot \rVert _2$)} \notag \\
-        &\leq \frac{c}{\sqrt{k}} \Trace(\mtx{A}) && \text{($\lambda_{k+1} \leq \frac{\Trace(\mtx{A})}{k}$)}
+        &\leq \frac{c}{\sqrt{k}} \Trace(\mtx{B}) && \text{($\lambda_{k+1} \leq \Trace(\mtx{B})/k$)}
         \label{equ:nystrom-proof-tail-bound}
@@
     \begin{equation}
-        \mathbb{E}^{\sfrac{k}{2}} \left[\lVert \mtx{A} - \Nystr{\mtx{\Omega}}{\mtx{A}} \rVert _F \right]
-        \leq \frac{1 + c}{\sqrt{k}} \Trace(\mtx{A})
+        \mathbb{E}^{\sfrac{k}{2}} \left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right]
+        \leq \frac{1 + c}{\sqrt{k}} \Trace(\mtx{B})
     \end{equation}
@@
     \begin{equation}
-        \mathbb{E}^{\sfrac{k}{2}}\left[ \int_{a}^{b} \lVert \mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F~\mathrm{d}t \right] 
-        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[\lVert \mtx{A} - \Nystr{\mtx{\Omega}}{\mtx{A}} \rVert _F \right] ~\mathrm{d}t% \notag \\
-        %&\leq \mathbb{E}^{\sfrac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
-        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t + \mathbb{E}^{\sfrac{k}{2}} \left[ \int_{a}^{b}  \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
-        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
-        \leq \frac{1 + c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t.
+        \mathbb{E}^{\sfrac{k}{2}}\left[ \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \right] 
+        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right] ~\mathrm{d}t% \notag \\
+        %&\leq \mathbb{E}^{\sfrac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
+        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \mathbb{E}^{\sfrac{k}{2}} \left[ \int_{a}^{b}  \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
+        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
+        \leq \frac{1 + c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
     \end{equation}
@@
     \begin{equation}
-        \int_{a}^{b} \lVert \mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{1 + c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t,
+        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{1 + c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t,
     \end{equation}
@@
 
-\subsection{Nyström++ estimator for parameter-dependent matrices}
-\label{subsec:nystrom-pp}
-
 Finally, we combine the analysis of the two estimators from \refsec{subsec:hutchinson} and \refsec{subsec:nystrom} into a more efficient, general-purpose trace estimator.
@@
 \begin{theorem}{Nyström++ trace estimator for parameter-dependent matrices}{nystrom-pp}
-    Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
+    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
 If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(2\delta^{-1})^2)$ with \emph{even} $n_{\mtx{\Omega}} \geq 2$, then for $\delta \in (0, 2^{-1} e^{-e^{4\varepsilon}})$ with probability at least $1 - \delta$
     \begin{equation}
-        \int_{a}^{b} | \Trace(\mtx{A}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{A}(t)) |~\mathrm{d}t
-        \leq \varepsilon \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t.
+        \int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t
+        \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
     \end{equation}
@@
     \begin{align}
-        &\int_{a}^{b} | \Trace(\mtx{A}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{A}(t)) |~\mathrm{d}t \notag \\
-        &= \int_{a}^{b} | \Hutch{\mtx{\Psi}}(\mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)) - \Trace(\mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)) |~\mathrm{d}t && \text{(by definition of estimators)} \notag \\
-        &\leq \tilde{\varepsilon} \int_{a}^{b} \lVert \mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F ~\mathrm{d}t && \text{(\refthm{thm:hutchinson} w.p. $\geq 1 - \tilde{\delta}$)} \notag \\
-        &\leq \tilde{\varepsilon}^2 \int_{a}^{b} \Trace(\mtx{A}(t)) ~\mathrm{d}t && \text{(\reflem{lem:nystrom} w.p. $\geq 1 - \tilde{\delta}$)} 
+        &\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t \notag \\
+        &= \int_{a}^{b} | \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) - \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) |~\mathrm{d}t && \text{(by definition of estimators)} \notag \\
+        &\leq \tilde{\varepsilon} \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F ~\mathrm{d}t && \text{(\refthm{thm:hutchinson} w.p. $\geq 1 - \tilde{\delta}$)} \notag \\
+        &\leq \tilde{\varepsilon}^2 \int_{a}^{b} \Trace(\mtx{B}(t)) ~\mathrm{d}t && \text{(\reflem{lem:nystrom} w.p. $\geq 1 - \tilde{\delta}$)} 
     \end{align}
@@
 %     \begin{align}
-%         &\int_{a}^{b} | \Trace(\mtx{A}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{A}(t)) |~\mathrm{d}t \notag \\
-%         &= \int_{a}^{b} | \Hutch{\mtx{\Psi}}(\mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)) - \Trace(\mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)) |~\mathrm{d}t && \text{(definition of estimators)} \notag \\
-%         &= \todo{c} \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F ~\mathrm{d}t && \text{(with probability $1 - \delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1})$)} \notag \\
-%         &= \todo{c} \frac{1}{\sqrt{n_{\mtx{\Psi}} n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{A}(t)) ~\mathrm{d}t
+%         &\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t \notag \\
+%         &= \int_{a}^{b} | \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)) - \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)) |~\mathrm{d}t && \text{(definition of estimators)} \notag \\
+%         &= \todo{c} \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F ~\mathrm{d}t && \text{(with probability $1 - \delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1})$)} \notag \\
+%         &= \todo{c} \frac{1}{\sqrt{n_{\mtx{\Psi}} n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t)) ~\mathrm{d}t
 %     \end{align}
@@
 discretization matrices which are growing in size with $n_c$. In our experiments
-we consider the three-dimensional case, but for visualization purposes, we
-illustrate the potential in \reffig{fig:gaussian-well}
-in two dimensions.
+we consider the three-dimensional case.%, but for visualization purposes, we illustrate the potential in \reffig{fig:gaussian-well} in two dimensions.
 
@@
     \input{plots/distribution.pgf}
-    \caption{The NC++ method for different ways of allocations a 
-    total of $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation for the Gaussian smoothing kernel with multiple different values of the smoothing parameter.
+    \caption{The NC++ method for different ways of allocations a total of $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation for the Gaussian smoothing kernel with multiple different values of the smoothing parameter.
 We make the approximation error made in the Chebyshev expansion negligible by rescaling $m=16 / \sigma$ (based on \refthm{thm:chebyshev-error}).}
-    \label{fig:matvec-mix}
+    \label{fig:distribution}
 \end{figure}
@@
 
+\todo{
 \begin{itemize}
@@
 \end{itemize}
+}
 

### 2024-06-04 paper/main.tex
Updates from Overleaf

--- 2024-06-03 paper/main.tex
@@
 The Girard-Hutchinson estimator distinguishes itself for its simplicity and provability for all symmetric matrices.
-However, the $n_{\mtx{\Psi}} = \mathcal{\varepsilon^{-2}}$ requirement implies that if we aim to increase the accuracy of the estimate by one digit, we would need to increase the number of queries by a factor of $100$.
+However, the $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ requirement implies that if we aim to increase the accuracy of the estimate by one digit, we would need to increase the number of queries by a factor of $100$.
 The next two estimators improve on this by exploiting the structure of $\mtx{A}(t)$.

### 2024-06-17 paper/main.tex
add comments from Hysan

--- 2024-06-04 paper/main.tex
@@
 
-The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Markov's inequality to transition to the continuous parameter case.
+The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Markov's inequality to extend the results to the continuous parameter case.
 We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
@@
         \mathbb{E}^{\sfrac{k}{2}} \left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right]
-        \leq \frac{1 + c}{\sqrt{k}} \Trace(\mtx{B})
+        \leq \frac{1 + c}{\sqrt{k}} \Trace(\mtx{B}).
     \end{equation}
 
-    We use Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities} to get
+    As in \cite{kressner-2023-randomized-lowrank}, we can show that the error $\left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right]$ is measurable and by the Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities}, we get
     \begin{equation}

### 2024-06-17 paper/main.tex
replace AMM proof

@@
 \begin{lemma}{Moments of tail bound}{ose}
-    \todo{A matrix $\mtx{A} \in \mathbb{R}^{l \times l}$ and }independent standard Gaussian random matrices $\mtx{\Omega}_1 \in \mathbb{R}^{k \times 2k}, \mtx{\Omega}_2 \in \mathbb{R}^{l \times 2k}$ satisfy for all $k \in \mathbb{N}$ and a universal constant $c > 0$
+    Let $\mtx{\Sigma} = \operatorname{diag}(\sigma_1, \dots, \sigma_m) \in \mathbb{R}^{m \times m}$ such that $|\sigma_1| \geq \dots \geq |\sigma_m|$ and independent standard Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all \todo{$k \geq 2$}
     \begin{equation}
-        \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] \leq c \lVert \mtx{A} \rVert _2 .
+        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]
+        \leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F.
     \end{equation}
+    %\begin{equation}
+    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
+    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
+    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
+    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
+    %\end{equation}
+    %If we choose 
 \end{lemma}
@@
 \begin{proof}
-    We first use the submultiplicativity of the spectral norm and the stochastic independendence of the random matrices
+    We can rewrite
     \begin{equation}
-        \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-        \leq \mathbb{E}^{k}\left[ \lVert \mtx{A} \rVert _2 \lVert \mtx{\Omega}_2 \rVert _2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-        = \lVert \mtx{A} \rVert _2 \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_2 \rVert _2 \right] \mathbb{E}^{k}\left[  \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right].
-        \label{equ:OSE-moment-bound-initial-new}
+        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2 \right]
+        = \mathbb{E}\left[  \left( \lVert \mtx{\Sigma} \vct{\omega} \rVert _2^2 \right)^{\sfrac{p}{2}} \right]^{\sfrac{1}{2p}}
+        = \mathbb{E}\left[  \left( \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right)^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}}
+        = \sqrt{ \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right] }.
     \end{equation}
 
-    To bound the moments of $\lVert \mtx{\Omega}_2 \rVert _2$ we define $\widetilde{\mtx{\Omega}}_2 = \mtx{\Omega}_2^{\top} / \sqrt{2k}$ and rewrite
+    Inspired by the proof of \cite[theorem 1]{cohen-2016-optimal-approximate}, we split the diagonal entries of $\mtx{\Sigma}$ into $\ell = \lceil m/k \rceil$ groups of size $k \geq 2$ as follows
+    \begin{equation}
+        \underbrace{\sigma_1, \dots, \sigma_k}_{\leq \sigma_1}, \underbrace{\sigma_{k+1}, \dots, \sigma_{2k}}_{\leq \sigma_{k+1}}, \dots, \underbrace{\sigma_{1+(\ell - 1)k}, \dots, \sigma_{\ell k}}_{\leq \sigma_{1+(\ell - 1)k}}
+    \end{equation}
+    In case $m$ is not a multiple of $k$, we set $\sigma_i = 0$ if $i > m$. Now we see that
+    \begin{equation}
+        \sigma_1^2 \leq \lVert \mtx{\Sigma} \rVert _2^2
+        \quad \text{and} \quad
+        \sigma_{1 + (i-1)k}^2 \leq \frac{1}{k} \sum_{j=1}^{k} \sigma_{j + (i-2)k}^2, i = 2, \dots, \ell
+    \end{equation}
+    and therefore
+    \begin{equation}
+        \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \sum_{j=1}^{(\ell - 1)k} \sigma_j^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2.
+    \end{equation}
+    By the triangle inequality for moments \todo{[cite Hysan]}
     \begin{align}
-        &\mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_2 \rVert _2 \right] \notag \\
-        &= \sqrt{2k} \cdot \mathbb{E}^{k}\left[ \lVert \widetilde{\mtx{\Omega}}_2^{\top} \rVert _2 \right] && \text{(definition of $\widetilde{\mtx{\Omega}}_2$)} \notag \\
-        &= \sqrt{2k} \cdot \sqrt{ \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \widetilde{\mtx{\Omega}}_2^{\top} \widetilde{\mtx{\Omega}}_2 \rVert _2 \right] } && \text{($\lVert \mtx{B} \rVert _2 = \lVert \mtx{B}^{\top} \mtx{B} \rVert _2^{\sfrac{1}{2}} $ )} \notag \\
-        &\leq \sqrt{2k} \cdot \sqrt{ \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \widetilde{\mtx{\Omega}}_2^{\top} \widetilde{\mtx{\Omega}}_2 - \mtx{I}_l \rVert _2 \right] + \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{I}_l \rVert _2 \right]} && \text{(triangle inequality)} \notag \\
-        &\leq \sqrt{2k} \cdot \sqrt{ \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \widetilde{\mtx{\Omega}}_2^{\top} \widetilde{\mtx{\Omega}}_2 - \mtx{I}_l \rVert _2 \right]} + \sqrt{2k} &&
-         \text{($\sqrt{a+b} \leq \sqrt{a} + \sqrt{b}$, $\lVert \mtx{I}_l \rVert _2 = 1$)}
+        \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right]
+        &= \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{j + (i-1)k}^2}_{\leq \sigma_{1 + (i-1)k}^2} \omega_{j + (i-1)k}^2 \right] \notag \\
+        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] \notag \\
+        &\leq \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \cdot \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] \notag \\
+        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \right)^{\sfrac{2}{p}} \notag \\
+        &\leq \frac{k + p}{2} \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right)
     \end{align}
-    By \cite[section 2.2]{nelson-2017-sketching-algorithms} $\widetilde{\mtx{\Omega}}_2 \in \mathbb{R}^{2k \times l}$ satisfies the $(c_1 \sqrt{q/2k}, e^{-q}, q)$-JL moment property for any $q > 0$ and a universal constant $c_1 > 0$.
-Hence, by \cite[lemma 4]{cohen-2016-optimal-approximate} it satisfies the $(2 c_1 \sqrt{q/2k}, 9^d e^{-q}, d, q)$-OSE moment property for any $d \in \mathbb{N}$.
-For convenience, we choose $d=2k$ and $q=k/2$, such that $\widetilde{\mtx{\Omega}}_2$ has $(c_1, (3^4 e^{-\sfrac{1}{2}})^k, 2k, k/2)$-OSE moment property, i.e.
-\cite[definition 3]{cohen-2016-optimal-approximate}
+    Thus, bounding the constant up front with $1$, we get
     \begin{equation}
-        \mathbb{E}^{\sfrac{k}{2}} \left[ \lVert \widetilde{\mtx{\Omega}}_2 \widetilde{\mtx{\Omega}}_2^{\top} - \mtx{I}_l \rVert _2 \right] \leq c_1 (3^4 e^{-\sfrac{1}{2}})^2 = c_1 3^8 e^{-1}.
-        %\mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\widetilde{\mtx{\Omega}}_2 \mtx{A} )^{\top} (\widetilde{\mtx{\Omega}}_2 \mtx{A}) - \mtx{A}^{\top} \mtx{A} \rVert _2\right] \leq c_1 c_2^2 3^8e^{-1} \left( \lVert \mtx{A} \rVert _2^2 + \frac{\lVert \mtx{A} \rVert _F^2}{k} \right).
+        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2 \right]
+        \leq \sqrt{\frac{k + p}{2}} \left( \lVert \mtx{\Sigma} \rVert _2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F \right)
     \end{equation}
-    Inserting this bound into the above inequality we get
+    From the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms} we know that 
     \begin{equation}
-        \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_2 \rVert _2 \right]
-        \leq \sqrt{2k} \cdot (3^4 \sqrt{c_1 e^{-1}} + 1).
-        \label{equ:OSE-moment-bound-first}
+        \mathbb{E}^p\left[ \lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]
+        \leq \lVert \mtx{\Sigma} \rVert _2 \mathbb{E}^p\left[ \lVert \vct{\omega}_1 \rVert _2 \right] + \mathbb{E}^p\left[ \lVert \mtx{\Sigma} \vct{\omega}_2 \rVert _2 \right]
+    \end{equation}
+    for independent standard Gaussian random vectors $\vct{\omega}_1 \in \mathbb{R}^{k}$ and $\vct{\omega}_2 \in \mathbb{R}^{m}$.
+The second moment is bound by \todo{equation} while for the first moment we apply the above $\mtx{\Sigma} = \mtx{I}_k$ to get
+    \begin{equation}
+        \mathbb{E}^p\left[ \lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]
+        \leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F.
     \end{equation}
 
-    The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
-    \begin{equation}
-        \mathbb{E}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^{k} \right]
-        = \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{k}{2}} \right]
-        \leq \left( k + 1 \right) \left( \frac{1}{(k + 1)!}\right)^{\sfrac{k}{k + 1}} \left( \frac{3 k}{2}\right)^{\sfrac{k}{2}}.
-    \end{equation}
-    With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\sfrac{1}{n}} \leq \frac{e}{n}$ follow respectively.
+    %The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
+    %\begin{align}
+    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
+    %    &= \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}} \notag \\
+    %    &\leq \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{1}{\Gamma(k - r + 2)} \right)^{\frac{1}{k - r + 1}} \left( \frac{k + r}{2} \right) \notag \\
+    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right)
+    %\end{align}
+    %With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\sfrac{1}{n}} \leq \frac{e}{n}$ follow respectively.
 Hence,
-    \begin{equation}
-        \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-        \leq \frac{e^2}{k + 1}\sqrt{ \frac{3k}{2} }
-        \leq e^2 \sqrt{\frac{3}{2k}}.
-        \label{equ:OSE-moment-bound-second}
-    \end{equation}
-
-    Putting \refequ{equ:OSE-moment-bound-first} and  \refequ{equ:OSE-moment-bound-second} together, we get
-    \begin{equation}
-        \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-        \leq \underbrace{e^2\sqrt{3} (3^4 \sqrt{c_1 e^{-1}} + 1)}_{\equiv c} \lVert \mtx{A} \rVert _2.
-    \end{equation}
+    %\begin{equation}
+    %    \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
+    %    \leq \frac{e^2}{k + 1}\sqrt{ \frac{3k}{2} }
+    %    \leq e^2 \sqrt{\frac{3}{2k}}.
+    %    \label{equ:OSE-moment-bound-second}
+    %\end{equation}
 
@@
     Its higher order moments are therefore bounded by
+    \begin{equation}
+        \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F \right]
+        \leq \sqrt{k} \cdot \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
+        = \sqrt{k} \cdot \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]^2
+        \label{equ:nystrom-proof-tail-bound}
+    \end{equation}
+
+    We first use the submultiplicativity of the spectral norm and the stochastic independendence of the random matrices
+    \begin{equation}
+        \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
+        \leq \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
+        \leq \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2 \right] \mathbb{E}^{k}\left[  \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
+        \label{equ:OSE-moment-bound-initial-new}
+    \end{equation}
+    The first term can be bound with \reflem{lem:ose} with $\mtx{\Sigma} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$:
+    \begin{equation}
+        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]
+        \leq \sqrt{\frac{3}{2}} \cdot \left( 2 \sqrt{k} + \frac{1}{\sqrt{2}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{3}{2}} \cdot \lVert \mtx{\Sigma} \rVert _F.
+    \end{equation}
+    The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
+    \begin{equation}
+        \mathbb{E}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^{k} \right]
+        = \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{k}{2}} \right]
+        \leq \left( k + 1 \right) \left( \frac{1}{(k + 1)!}\right)^{\sfrac{k}{k + 1}} \left( \frac{3 k}{2}\right)^{\sfrac{k}{2}}.
+    \end{equation}
+    With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\sfrac{1}{n}} \leq \frac{e}{n}$ follow respectively.
+Hence,
+    \begin{equation}
+        \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
+        \leq \frac{e^2}{k + 1}\sqrt{ \frac{3k}{2} }
+        \leq e^2 \sqrt{\frac{3}{2k}}.
+        \label{equ:OSE-moment-bound-second}
+    \end{equation}
+
+    Putting \refequ{equ:OSE-moment-bound-first} and  \refequ{equ:OSE-moment-bound-second} together, we get
+    \begin{equation}
+        \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
+        \leq ...
+    \end{equation}
+
+    Just as in the proof of \cite[lemma 3.1]{meyer-2021-hutch-optimal} we can bound
     \begin{align}
-        \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F \right]
-        &\leq \sqrt{k} \cdot \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right] && \text{(using \refequ{equ:nystrom-proof-processed-tail})} \notag \\
-        &= \sqrt{k} \cdot \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]^2 && \text{(by definition of $\mathbb{E}^{k}$)} \notag \\
         &\leq \sqrt{k} c \cdot \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 && \text{(\reflem{lem:ose} with $\mtx{A}=\mtx{\Lambda}_2^{\sfrac{1}{2}}$)} \notag \\
@@
         &\leq \frac{c}{\sqrt{k}} \Trace(\mtx{B}) && \text{($\lambda_{k+1} \leq \Trace(\mtx{B})/k$)}
-        \label{equ:nystrom-proof-tail-bound}
     \end{align}
+
     Thus, using the triangle inequality for $\mathbb{E}^{\sfrac{k}{2}}$ and inserting \refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud} we obtain

### 2024-06-19 paper/main.tex
clean up proofs

@@
 
-For the coming pages, we employ the following notation. Let
-\begin{equation}
-    \mtx{B}(t) 
-    = \mtx{U}(t) \mtx{\Lambda}(t) \mtx{U}(t)^{\top} 
-\end{equation}
-be the eigenvalue decomposition of $\mtx{B}(t)$ at a fixed $t$, i.e.
-$\mtx{\Lambda}(t) = \operatorname{diag}(\lambda_1(t), \dots, \lambda_n(t))$ and $\mtx{U}(t)^{\top} \mtx{U}(t) = \mtx{I}_n$.
-We define the matrix partitions
-\begin{equation}
-    \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+4ex+\baselineskip}
-    \mtx{U}(t) = \begin{bmatrix}
-        \smash{\underbrace{\mtx{U}_1(t)}_{n \times k}} & \smash{\underbrace{\mtx{U}_2(t)}_{n \times (n-k)}}
-    \end{bmatrix}
-    \quad \text{and} \quad
-    \mtx{\Lambda}(t) =
-    \begin{bmatrix}
-        \smash{\overbrace{\mtx{\Lambda}_1(t)}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2(t)}_{(n-k) \times (n-k)}}
-    \end{bmatrix}
-\end{equation}
-and let $\mtx{\Omega}_1(t) = \mtx{U}_1(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times 2k}$ and $\mtx{\Omega}_2(t) = \mtx{U}_2(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times 2k}$ with $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$.
-
-\begin{lemma}{Moments of tail bound}{ose}
-    Let $\mtx{\Sigma} = \operatorname{diag}(\sigma_1, \dots, \sigma_m) \in \mathbb{R}^{m \times m}$ such that $|\sigma_1| \geq \dots \geq |\sigma_m|$ and independent standard Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all \todo{$k \geq 2$}
+\begin{lemma}{Moment of spectral norm of non-standard Gaussian random matrix}{ose}
+    The matrix $\mtx{\Sigma} = \operatorname{diag}(\sigma_1, \dots, \sigma_m) \in \mathbb{R}^{m \times m}$ with $\sigma_1 \geq \dots \geq \sigma_m \geq 0$ and the standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all $p \in \mathbb{N}$
     \begin{equation}
         \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]
-        \leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F.
+        \leq  \sqrt{k + p} \left( 3 \lVert \mtx{\Sigma} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{\Sigma} \rVert _F \right).
     \end{equation}
@@
 \begin{proof}
-    We can rewrite
+    First, consider the above moment of a vector $\vct{\omega} \in \mathbb{R}^{m}$. We can rewrite
     \begin{equation}
         \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2 \right]
-        = \mathbb{E}\left[  \left( \lVert \mtx{\Sigma} \vct{\omega} \rVert _2^2 \right)^{\sfrac{p}{2}} \right]^{\sfrac{1}{2p}}
+        = \mathbb{E}\left[  \left( \lVert \mtx{\Sigma} \vct{\omega} \rVert _2^2 \right)^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}}
         = \mathbb{E}\left[  \left( \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right)^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}}
         = \sqrt{ \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right] }.
-    \end{equation}
-
-    Inspired by the proof of \cite[theorem 1]{cohen-2016-optimal-approximate}, we split the diagonal entries of $\mtx{\Sigma}$ into $\ell = \lceil m/k \rceil$ groups of size $k \geq 2$ as follows
-    \begin{equation}
-        \underbrace{\sigma_1, \dots, \sigma_k}_{\leq \sigma_1}, \underbrace{\sigma_{k+1}, \dots, \sigma_{2k}}_{\leq \sigma_{k+1}}, \dots, \underbrace{\sigma_{1+(\ell - 1)k}, \dots, \sigma_{\ell k}}_{\leq \sigma_{1+(\ell - 1)k}}
-    \end{equation}
-    In case $m$ is not a multiple of $k$, we set $\sigma_i = 0$ if $i > m$. Now we see that
-    \begin{equation}
-        \sigma_1^2 \leq \lVert \mtx{\Sigma} \rVert _2^2
+        \label{equ:matvec-spectral-norm-simplification}
+    \end{equation}
+    Inspired by the proof of \cite[theorem 1]{cohen-2016-optimal-approximate}, we split the diagonal entries of $\mtx{\Sigma}$ into $\ell = \lceil m/k \rceil$ groups of size $k \geq 1$
+    \begin{equation}
+        \underbrace{\sigma_1, \dots, \sigma_k}_{\leq \sigma_1}, \underbrace{\sigma_{k+1}, \dots, \sigma_{2k}}_{\leq \sigma_{k+1}}, \dots, \underbrace{\sigma_{1+(\ell - 1)k}, \dots, \sigma_{\ell k}}_{\leq \sigma_{1+(\ell - 1)k}}.
+    \end{equation}
+    If $m$ is not a multiple of $k$, we let $\sigma_i = 0$ for $i > m$. Since $\sigma_1 \geq \dots \geq \sigma_{\ell k} \geq 0$,
+    \begin{equation}
+        \sigma_1^2 = \lVert \mtx{\Sigma} \rVert _2^2
         \quad \text{and} \quad
-        \sigma_{1 + (i-1)k}^2 \leq \frac{1}{k} \sum_{j=1}^{k} \sigma_{j + (i-2)k}^2, i = 2, \dots, \ell
+        \sigma_{1 + (i-1)k}^2 \leq \frac{1}{k} \sum_{j=1}^{k} \sigma_{j + (i-2)k}^2, i = 2, \dots, \ell,
     \end{equation}
@@
         \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \sum_{j=1}^{(\ell - 1)k} \sigma_j^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2.
-    \end{equation}
-    By the triangle inequality for moments \todo{[cite Hysan]}
+        \label{equ:singular-value-group-bound}
+    \end{equation}
+    This allows us to bound
     \begin{align}
         \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right]
-        &= \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{j + (i-1)k}^2}_{\leq \sigma_{1 + (i-1)k}^2} \omega_{j + (i-1)k}^2 \right] \notag \\
-        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] \notag \\
-        &\leq \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \cdot \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] \notag \\
-        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \right)^{\sfrac{2}{p}} \notag \\
-        &\leq \frac{k + p}{2} \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right)
+        &= \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{j + (i-1)k}^2}_{\leq \sigma_{1 + (i-1)k}^2} \omega_{j + (i-1)k}^2 \right] && \text{(separate sum into groups)} \notag \\
+        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{($\sigma_{j + (i-1)k}^2 \leq \sigma_{1 + (i-1)k}^2$)} \notag \\
+        &\leq \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \cdot \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{(triangle inequality)} \notag \\
+        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \right)^{\sfrac{2}{p}} && \text{(\refequ{equ:singular-value-group-bound} and $\chi_k^2$ moments \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
+        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) (k + p). && \text{(\reflem{lem:gamma})}
     \end{align}
-    Thus, bounding the constant up front with $1$, we get
+    Thus, together with \refequ{equ:matvec-spectral-norm-simplification} and using $\sqrt{a+b} \leq \sqrt{a} + \sqrt{b}$ for $a, b \geq 0$, we obtain
     \begin{equation}
         \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2 \right]
-        \leq \sqrt{\frac{k + p}{2}} \left( \lVert \mtx{\Sigma} \rVert _2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F \right)
-    \end{equation}
-    From the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms} we know that 
+        \leq \sqrt{k + p} \left( \lVert \mtx{\Sigma} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{\Sigma} \rVert _F \right).
+        \label{equ:matvec-spectral-norm-moment-bound}
+    \end{equation}
+
+    Now, we consider the matrix case. From the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms} we know that 
     \begin{equation}
@@
     for independent standard Gaussian random vectors $\vct{\omega}_1 \in \mathbb{R}^{k}$ and $\vct{\omega}_2 \in \mathbb{R}^{m}$.
-The second moment is bound by \todo{equation} while for the first moment we apply the above $\mtx{\Sigma} = \mtx{I}_k$ to get
+Both moments can be bound using \refequ{equ:matvec-spectral-norm-moment-bound}, one of which with $\mtx{\Sigma} = \mtx{I}_k$, to get
     \begin{equation}
         \mathbb{E}^p\left[ \lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]
-        \leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F.
+        \leq \sqrt{k + p} \left( 3 \lVert \mtx{\Sigma} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{\Sigma} \rVert _F \right).
     \end{equation}
@@
     \begin{equation}
-        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \sqrt{2} \cdot \frac{1 + c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
+        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
     \end{equation}
     for a universal constant $c > 0$.
-In particular, if we choose $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})), n_{\mtx{\Omega}} \geq 2$, then with probability at least $1-\delta$ for $\delta \in (0, e^{-e^{4 \varepsilon^2}})$ and $\varepsilon > 0$ we have $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$.
+In particular, if we choose $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1}))$, then with probability at least $1-\delta$ for $\delta \in (0, e^{-e^{4 \varepsilon^2}})$ and $\varepsilon > 0$ we have $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$.
 \end{lemma}
@@
 \begin{proof}
-    We let $n_{\mtx{\Omega}} = 2k$ for some $k \in \mathbb{N}$, i.e.
-$\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$.
-After temporarily omitting the parameter dependence, we can apply \cite[theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound the integrand with% \todo{maybe explain where $\mtx{\Omega}_i$ come from...}
+    First, we temporarily omit the parameter dependence.
+Let $\mtx{B} = \mtx{U} \mtx{\Lambda} \mtx{U}^{\top}$ be the eigenvalue decomposition of a symmetric positive semi-definite matrix $\mtx{B}$, i.e.
+$\mtx{\Lambda} = \operatorname{diag}(\lambda_1, \dots, \lambda_n)$ and $\mtx{U}^{\top} \mtx{U}= \mtx{I}_n$.
+We define the matrix partitions
+    \begin{equation}
+        \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+4ex+\baselineskip}
+        \mtx{U} = \begin{bmatrix}
+            \smash{\underbrace{\mtx{U}_1}_{n \times k}} & \smash{\underbrace{\mtx{U}_2}_{n \times (n-k)}}
+        \end{bmatrix}
+        \quad \text{and} \quad
+        \mtx{\Lambda} =
+        \begin{bmatrix}
+            \smash{\overbrace{\mtx{\Lambda}_1}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2}_{(n-k) \times (n-k)}}
+        \end{bmatrix}
+    \end{equation}
+    for some \todo{$k \leq n/2$}.
+Further, for a standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$ we define $\mtx{\Omega}_1 = \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times 2k}$ and $\mtx{\Omega}_2 = \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times 2k}$, which are likewise standard Gaussian.
+
+    We will set $n_{\mtx{\Omega}} = 2k$, such that we can apply \cite[theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound
     \begin{equation}
         \lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F 
-        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F.
+        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2
         \label{equ:nystrom-proof-persson-bonud}
     \end{equation}
-
-    As in the proof of \cite[lemma 3]{meyer-2021-hutch-optimal}, the first term can be bonud with
+    where $\lVert \cdot \rVert _{(4)}$ denotes the Schatten-4 norm.
+
+    As in the proof of \cite[lemma 3]{meyer-2021-hutch-optimal}, the first term is limited by
     \begin{equation}
@@
     \end{equation}
-
-    The second term is first processed with standard matrix-norm inequalities:
-    \begin{equation}
-        \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F 
-        \leq \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F
-        \leq \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2.
+    The second term is first processed with standard matrix norm manipulations to
+    \begin{align}
+        \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
+        &= \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )( \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger})^{\top} \rVert _F \right] && \text{($\lVert \mtx{A} \rVert _{(4)}^2 = \lVert \mtx{A} \mtx{A}^{\top} \rVert _{F}$)} \notag \\
+        &\leq \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F \right] && \text{($\lVert \mtx{A} \mtx{B} \rVert _F \leq \lVert \mtx{A} \rVert _F \lVert \mtx{B} \rVert _2$)} \notag \\
+        &\leq \mathbb{E}^{\sfrac{k}{2}}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{($\lVert \mtx{A} \rVert _F \leq \sqrt{k} \lVert \mtx{A} \rVert _2$)} \notag \\
+        &\leq \mathbb{E}^{\sfrac{k}{2}}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{(submultiplicativity)} \notag \\
+        &\leq \sqrt{k} \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2\right] \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2  \right]  && \text{(independence $\mtx{\Omega}_1, \mtx{\Omega}_2$)} \notag \\
+        &= \sqrt{k} \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2 \right]^2 \mathbb{E}^{k}\left[  \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]^2 && \text{(definition of $\mathbb{E}^{k}[\cdot]$)}
         \label{equ:nystrom-proof-processed-tail}
-    \end{equation}
-    Its higher order moments are therefore bounded by
-    \begin{equation}
-        \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F \right]
-        \leq \sqrt{k} \cdot \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
-        = \sqrt{k} \cdot \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]^2
-        \label{equ:nystrom-proof-tail-bound}
-    \end{equation}
-
-    We first use the submultiplicativity of the spectral norm and the stochastic independendence of the random matrices
-    \begin{equation}
-        \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-        \leq \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-        \leq \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2 \right] \mathbb{E}^{k}\left[  \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-        \label{equ:OSE-moment-bound-initial-new}
-    \end{equation}
-    The first term can be bound with \reflem{lem:ose} with $\mtx{\Sigma} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$:
-    \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]
-        \leq \sqrt{\frac{3}{2}} \cdot \left( 2 \sqrt{k} + \frac{1}{\sqrt{2}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{3}{2}} \cdot \lVert \mtx{\Sigma} \rVert _F.
+    \end{align}
+    We can apply \reflem{lem:ose} with $\mtx{\Sigma} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$ to bound the first term with
+    \begin{equation}
+        \mathbb{E}^k\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2 \right]
+        \leq \sqrt{3 k} \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2 + \frac{1}{\sqrt{2k}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F \right).
+        \label{equ:spectral-norm-bound-applied}
     \end{equation}
@@
         \leq e^2 \sqrt{\frac{3}{2k}}.
-        \label{equ:OSE-moment-bound-second}
-    \end{equation}
-
-    Putting \refequ{equ:OSE-moment-bound-first} and  \refequ{equ:OSE-moment-bound-second} together, we get
-    \begin{equation}
-        \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-        \leq ...
-    \end{equation}
-
-    Just as in the proof of \cite[lemma 3.1]{meyer-2021-hutch-optimal} we can bound
+        \label{equ:pinv-spectral-norm-bound}
+    \end{equation}
+    Plugging \refequ{equ:spectral-norm-bound-applied} and  \refequ{equ:pinv-spectral-norm-bound} into \refequ{equ:nystrom-proof-processed-tail}, and using $(a + b)^2 \leq 2a^2 + 2b^2$ for $a, b > 0$ gives us
+    \begin{equation}
+        \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F \right]
+        \leq  9 e^4 \sqrt{k} \left( 9 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{2k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right)
+    \end{equation}
+    Here, we can identify $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 = \lambda_{k+1}$ and $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 = \Trace(\mtx{\Lambda}_2)$, and just as in the proof of \cite[lemma 3.1]{meyer-2021-hutch-optimal} we bound $\lambda_{k+1} \leq \Trace(\mtx{B})/k$ and $\Trace(\mtx{\Lambda}_2) \leq \Trace(\mtx{B})$ to get
+    \begin{equation}
+        \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F \right]
+        \leq 9 e^4 \sqrt{k} \left( \frac{9}{k} \Trace(\mtx{B}) + \frac{1}{2k} \Trace(\mtx{B}) \right) \leq  9 e^4 \left(9 + \frac{1}{2} \right) \frac{1}{\sqrt{k}} \Trace(\mtx{B})
+        \label{equ:nystrom-proof-tail-bound}
+    \end{equation}
+    Finally, using the triangle inequality for $\mathbb{E}^{\sfrac{k}{2}}$ and inserting \refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud} we obtain
+    \begin{equation}
+        \mathbb{E}^{\sfrac{k}{2}} \left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right]
+        \leq \frac{c}{\sqrt{k}} \Trace(\mtx{B}).
+    \end{equation}
+    where $c = 1 + 9 e^4 (9 + \frac{1}{2})$.
+
+    As in \cite{kressner-2023-randomized-lowrank}, we can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable and by the Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities}, we get
     \begin{align}
-        &\leq \sqrt{k} c \cdot \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 && \text{(\reflem{lem:ose} with $\mtx{A}=\mtx{\Lambda}_2^{\sfrac{1}{2}}$)} \notag \\
-        &= \sqrt{k} c \cdot \lambda_{k+1} && \text{(from the definition of $\lVert \cdot \rVert _2$)} \notag \\
-        &\leq \frac{c}{\sqrt{k}} \Trace(\mtx{B}) && \text{($\lambda_{k+1} \leq \Trace(\mtx{B})/k$)}
-    \end{align}
-
-    Thus, using the triangle inequality for $\mathbb{E}^{\sfrac{k}{2}}$ and inserting \refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud} we obtain
-    \begin{equation}
-        \mathbb{E}^{\sfrac{k}{2}} \left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right]
-        \leq \frac{1 + c}{\sqrt{k}} \Trace(\mtx{B}).
-    \end{equation}
-
-    As in \cite{kressner-2023-randomized-lowrank}, we can show that the error $\left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right]$ is measurable and by the Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities}, we get
-    \begin{equation}
         \mathbb{E}^{\sfrac{k}{2}}\left[ \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \right] 
-        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right] ~\mathrm{d}t% \notag \\
+        &\stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F \right] ~\mathrm{d}t \notag \\
         %&\leq \mathbb{E}^{\sfrac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
@@
         %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
-        \leq \frac{1 + c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
-    \end{equation}
+        &\leq \frac{c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
+    \end{align}
     From Markov's inequality follows with probability at least $1 - \gamma^{-\sfrac{k}{2}}$
     \begin{equation}
-        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{1 + c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t,
-    \end{equation}
-    from which follows the first statement by replacing $k = \frac{n_{\mtx{\Omega}}}{2}$.
-
-    Setting $\delta = \gamma^{-\sfrac{k}{2}}$ and choosing $k = \lceil \varepsilon^{-2}\log(\delta^{-1}) \rceil$ we get
+        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t,
+    \end{equation}
+    from which follows the first statement by remembering $k = \frac{n_{\mtx{\Omega}}}{2}$ and redefining $c \to \sqrt{2}c$.
+
+    Setting $\delta = \gamma^{-\sfrac{k}{4}}$ and choosing $k = \lceil \varepsilon^{-2}\log(\delta^{-1}) \rceil$ we get
     \begin{align}
@@
     \end{align}
-    which is smaller than $\varepsilon$ if $\delta \leq e^{-e^{4 \varepsilon^2}}$from which follows the second part of the theorem.
+    which is smaller than $\varepsilon$ if $\delta \leq e^{-e^{4 \varepsilon^2}}$from which follows the second part of the theorem when using $n_{\mtx{\Omega}} = 2k$.
 \end{proof}
@@
     Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(2\delta^{-1})^2)$ with \emph{even} $n_{\mtx{\Omega}} \geq 2$, then for $\delta \in (0, 2^{-1} e^{-e^{4\varepsilon}})$ with probability at least $1 - \delta$
+If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(2\delta^{-1})^2)$ with \emph{even} $n_{\mtx{\Omega}} \geq 2$, then for $\delta \in (0, e^{-e^{4\varepsilon}} / 2)$ with probability at least $1 - \delta$
     \begin{equation}
@@
 
-%\appendix
-
-%\clearpage
-%\section{Proofs}
+\appendix
+
+\clearpage
+\section{Proofs}
+
+\begin{lemma}{Quotient of Gamma functions}{gamma}
+    For all $p, k \in \mathbb{N}$ it holds
+    \begin{equation}
+        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \leq \left( \frac{k + p}{2} \right)^{\sfrac{p}{2}}.
+    \end{equation}
+\end{lemma}
+
+\begin{proof}
+    We first treat the case $k=p=1$, for which indeed
+    \begin{equation}
+        \frac{\Gamma(1)}{\Gamma(\frac{1}{2})} = \frac{1}{\sqrt{\pi}} \leq 1
+    \end{equation}
+    So we now assume $k + p \geq 3$. We can use a telescoping product and the property $\Gamma(z+1)/\Gamma(z) = z$ to bound
+    \begin{align}
+        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} 
+        &= \frac{\Gamma(\frac{k}{2} + 1)}{\Gamma(\frac{k}{2})} \frac{\Gamma(\frac{k}{2} + 2)}{\Gamma(\frac{k}{2} + 1)} \cdots \frac{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor - 1)} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)} \notag \\
+        &= \left(\frac{k}{2}\right) \left(\frac{k}{2} + 1\right) \cdots \left( \frac{k}{2} + \bigg\lfloor \frac{p}{2} \bigg\rfloor - 1 \right) \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)} \notag \\
+        &\leq \left(\frac{k}{2} + \bigg\lfloor \frac{p}{2} \bigg\rfloor - 1\right)^{\lfloor \sfrac{p}{2} \rfloor} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)}
+    \end{align}
+    If $p$ is even, then $\lfloor p/2 \rfloor = p/2$ and hence
+    \begin{equation}
+        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} 
+        \leq \left( \frac{k + p}{2} - 1 \right)^{\sfrac{p}{2}}
+        \leq \left( \frac{k + p}{2}\right)^{\sfrac{p}{2}}.
+    \end{equation}
+    If $p$ is odd, then $\lfloor p/2 \rfloor = (p - 1)/2$, from which follows with Gautschi's inequality \cite{kershaw-1983-extensions-gautschi}
+    \begin{equation}
+        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})}
+        \leq \left(\frac{k + p}{2} - \frac{3}{2} \right)^{ \frac{p - 1}{2}} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k + p}{2} - \frac{1}{2})}
+        \leq \left(\frac{k + p}{2}\right)^{ \frac{p - 1}{2}} \left(\frac{k + p}{2}\right)^{\sfrac{1}{2}}
+        = \left(\frac{k + p}{2}\right)^{ \sfrac{p}{2}}
+    \end{equation}
+\end{proof}
 

### 2024-06-19 paper/main.tex
generalize spectral norm moment bound

@@
 \begin{lemma}{Moment of spectral norm of non-standard Gaussian random matrix}{ose}
-    The matrix $\mtx{\Sigma} = \operatorname{diag}(\sigma_1, \dots, \sigma_m) \in \mathbb{R}^{m \times m}$ with $\sigma_1 \geq \dots \geq \sigma_m \geq 0$ and the standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all $p \in \mathbb{N}$
-    \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]
+    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all $p \in \mathbb{N}$ 
+    \begin{equation}
+        \mathbb{E}^{p}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2 \right]
         \leq  \sqrt{k + p} \left( 3 \lVert \mtx{\Sigma} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{\Sigma} \rVert _F \right).
@@
 \begin{proof}
+    With the singular value decomposition $\mtx{A} = \mtx{U} \mtx{\Sigma} \mtx{V}^{\top}$ and the unitary invariance of the spectral norm, it can be seen that $\lVert \mtx{A} \mtx{\Omega} \rVert _2 = \lVert \mtx{\Sigma} \widetilde{\mtx{\Omega}} \rVert _2$ where $\widetilde{\mtx{\Omega}}$ is again a standard Gaussian random matrix.
+Hence, it is enough to bound $\mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]$ for a diagonal matrix $\mtx{\Sigma} = \operatorname{diag}(\sigma_1, \dots, \sigma_m)$ with $\sigma_1 \geq \dots \geq \sigma_m \geq 0$.
+
     First, consider the above moment of a vector $\vct{\omega} \in \mathbb{R}^{m}$. We can rewrite
@@
     \end{align}
-    We can apply \reflem{lem:ose} with $\mtx{\Sigma} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$ to bound the first term with
+    We can apply \reflem{lem:ose} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$ to bound the first term with
     \begin{equation}

### 2024-06-19 paper/main.tex
split lemma into two

--- 2024-06-19 paper/main.tex
@@
 
-\begin{lemma}{Moment of spectral norm of non-standard Gaussian random matrix}{ose}
-    The matrix $\mtx{\Sigma} = \operatorname{diag}(\sigma_1, \dots, \sigma_m) \in \mathbb{R}^{m \times m}$ with $\sigma_1 \geq \dots \geq \sigma_m \geq 0$ and the standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all $p \in \mathbb{N}$
-    \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]
-        \leq  \sqrt{k + p} \left( 3 \lVert \mtx{\Sigma} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{\Sigma} \rVert _F \right).
+\begin{lemma}{Spectral norm moments of non-standard Gaussian random vector}{spectral-norm-moment-vector}
+    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random vector $\vct{\omega} \in \mathbb{R}^{m}$ satisfy for all $k,p \in \mathbb{N}$ 
+    \begin{equation}
+        \mathbb{E}^{p}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2 \right]
+        \leq  \sqrt{k + p} \left( 3 \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{A} \rVert _F \right).
     \end{equation}
@@
 \begin{proof}
-    First, consider the above moment of a vector $\vct{\omega} \in \mathbb{R}^{m}$. We can rewrite
+    With the singular value decomposition $\mtx{A} = \mtx{U} \mtx{\Sigma} \mtx{V}^{\top}$ and the unitary invariance of the spectral norm, it can be seen that $\lVert \mtx{A} \mtx{\omega} \rVert _2 = \lVert \mtx{\Sigma} \widetilde{\mtx{\omega}} \rVert _2$ where $\widetilde{\mtx{\omega}}$ is again a standard Gaussian random vector.
+Hence, it is enough to bound $\mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\omega} \rVert _2 \right]$ for a diagonal matrix $\mtx{\Sigma} = \operatorname{diag}(\sigma_1, \dots, \sigma_m)$ with $\sigma_1 \geq \dots \geq \sigma_m \geq 0$.
+
+    First, we can rewrite
     \begin{equation}
@@
         \label{equ:matvec-spectral-norm-moment-bound}
-    \end{equation}
-
-    Now, we consider the matrix case. From the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms} we know that 
-    \begin{equation}
-        \mathbb{E}^p\left[ \lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]
-        \leq \lVert \mtx{\Sigma} \rVert _2 \mathbb{E}^p\left[ \lVert \vct{\omega}_1 \rVert _2 \right] + \mathbb{E}^p\left[ \lVert \mtx{\Sigma} \vct{\omega}_2 \rVert _2 \right]
-    \end{equation}
-    for independent standard Gaussian random vectors $\vct{\omega}_1 \in \mathbb{R}^{k}$ and $\vct{\omega}_2 \in \mathbb{R}^{m}$.
-Both moments can be bound using \refequ{equ:matvec-spectral-norm-moment-bound}, one of which with $\mtx{\Sigma} = \mtx{I}_k$, to get
-    \begin{equation}
-        \mathbb{E}^p\left[ \lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]
-        \leq \sqrt{k + p} \left( 3 \lVert \mtx{\Sigma} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{\Sigma} \rVert _F \right).
     \end{equation}
@@
 
+\begin{lemma}{Spectral norm moments of non-standard Gaussian random matrix}{spectral-norm-moment}
+    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all $p \in \mathbb{N}$ 
+    \begin{equation}
+        \mathbb{E}^{p}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2 \right]
+        \leq  \sqrt{k + p} \left( 3 \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{A} \rVert _F \right).
+    \end{equation}
+    %\begin{equation}
+    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
+    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
+    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
+    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
+    %\end{equation}
+    %If we choose 
+\end{lemma}
+
+\begin{proof}
+    From the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms} we know that 
+    \begin{equation}
+        \mathbb{E}^p\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2 \right]
+        \leq \lVert \mtx{A} \rVert _2 \mathbb{E}^p\left[ \lVert \vct{\omega}_1 \rVert _2 \right] + \mathbb{E}^p\left[ \lVert \mtx{A} \vct{\omega}_2 \rVert _2 \right]
+    \end{equation}
+    for independent standard Gaussian random vectors $\vct{\omega}_1 \in \mathbb{R}^{k}$ and $\vct{\omega}_2 \in \mathbb{R}^{m}$.
+Both moments can be bound using \reflem{lem:spectral-norm-moment-vector}, one of which with $\mtx{A} = \mtx{I}_k$, to get
+    \begin{equation}
+        \mathbb{E}^p\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2 \right]
+        \leq \sqrt{k + p} \left( 3 \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{A} \rVert _F \right).
+    \end{equation}
+
+    %The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
+    %\begin{align}
+    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
+    %    &= \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}} \notag \\
+    %    &\leq \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{1}{\Gamma(k - r + 2)} \right)^{\frac{1}{k - r + 1}} \left( \frac{k + r}{2} \right) \notag \\
+    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right)
+    %\end{align}
+    %With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\sfrac{1}{n}} \leq \frac{e}{n}$ follow respectively.
+Hence,
+    %\begin{equation}
+    %    \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
+    %    \leq \frac{e^2}{k + 1}\sqrt{ \frac{3k}{2} }
+    %    \leq e^2 \sqrt{\frac{3}{2k}}.
+    %    \label{equ:OSE-moment-bound-second}
+    %\end{equation}
+
+\end{proof}
+
 \begin{lemma}{Nyström approximation for parameter-dependent matrices}{nystrom}
@@
     \end{align}
-    We can apply \reflem{lem:ose} with $\mtx{\Sigma} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$ to bound the first term with
+    We can apply \reflem{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$ to bound the first term with
     \begin{equation}

### 2024-06-19 paper/main.tex
fill in gaps between theorems

@@
 
+In the remainder of this section, we will derive upper bounds on the error of these estimators.
+%
+%\begin{table}[ht]
+%\centering
+%\renewcommand{\arraystretch}{1.2}
+%\begin{tabular}{@{}lcc@{}}
+%\toprule
+%Estimator & Matrix & $m=1600$\\
+%\midrule
+%Girard-Hutchinson & symmetric & $\mathcal{O}(\varepsilon^{-2})%$ \\
+%Nyström & symmetric positive semi-definite & $\mathcal{O}%(\varepsilon^{-2})$ \\
+%Nyström++ & symmetric positive semi-definite & $\mathcal{O}%(\varepsilon^{-1})$  \\
+%\bottomrule
+%\end{tabular}
+%\end{table}
+
 \subsection{Girard-Hutchinson estimator for parameter-dependent matrices}
@@
 
-We first consider the estimator \refequ{equ:hutchinson-trace-estimator}. \todo{Introduce $L^1$ norm and Frobenius norm}
+We first consider the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}.
+We bound its $L^1$ norm in terms of the integral over the Frobenius norm $\lVert \cdot \rVert _F$ of the matrix whose trace we aim to approximate.
 
@@
 
-The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Markov's inequality to extend the results to the continuous parameter case.
+Compared to the constant case \cite[lemma 2.1]{meyer-2021-hutch-optimal} we need to choose the number of queries $n_{\mtx{\Psi}}$ by a factor of $\log(\delta^{-1})$ larger to achieve the same guarantee.
+
+The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Minkowski's integral inequality to extend the results to the continuous parameter case.
 We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
@@
     First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence.
-From the proof of \cite[lemma 3]{cortinovis-2022-randomized-trace} we know that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ \todo{introduce spectral norm}.
+From the proof of \cite[lemma 3]{cortinovis-2022-randomized-trace} we know that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm.
 Thus, by \cite[theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
@@
 
+\todo{
 \begin{remark}
@@
 \end{remark}
-
-The Girard-Hutchinson estimator distinguishes itself for its simplicity and provability for all symmetric matrices.
-However, the $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ requirement implies that if we aim to increase the accuracy of the estimate by one digit, we would need to increase the number of queries by a factor of $100$.
-The next two estimators improve on this by exploiting the structure of $\mtx{B}(t)$.
+}
+
+The Girard-Hutchinson estimator distinguishes itself for its simplicity.
+However, its bound does not take into account the structure of $\mtx{B}(t)$.
+The $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ requirement implies that if we aim to increase the accuracy of the estimate by one digit, we would need to increase the number of queries by a factor of $100$.
 
@@
 
-Next we consider \refequ{equ:nystrom-trace-estimator}. \todo{Motivate Nyström approximation if matrix is low-rank.}
+For symmetric positive semi-definite matrices whose singular values decay rapidly, the Nyström estimate \refequ{equ:nystrom-trace-estimator} can be significantly better than the Girard-Hutchinson estimate.
+This is reflected in the following theorem.
 
@@
     \end{equation}
-    \todo{introduce nuclear norm}.  From the proof of \cite[corollary 8.2]{tropp-2023-randomized-algorithms} it follows that
+    where $\lVert \cdot \rVert _{\ast}$ represents the nuclear norm.  From the proof of \cite[corollary 8.2]{tropp-2023-randomized-algorithms} it follows that
     \begin{equation}
@@
 
+\todo{Compare to constant matrix case: \cite[theorem 8.1]{tropp-2023-randomized-algorithms} (similar result)}
+
+The downside of the Nyström estimator is that it will -- in general -- not work well enough on matrices whose singular values do not decay, for example $\mtx{B}(t) \equiv \mtx{I}_n$.
+
 \subsection{Nyström++ estimator for parameter-dependent matrices}
@@
 
-\todo{Motivate and explain proof technique (to mimic Hutch++)}
+Just as in the constant matrix case, our goal is now to prove that the Nyström++ estimator \refequ{equ:nystrompp-trace-estimator} achieves an $\varepsilon$ relative error with just $\mathcal{O}(\varepsilon)$ random vectors to construct the estimate.
+To do so, we mimic the proof of a similar bound for the Hutch++ estimator \cite[theorem 3.1]{meyer-2021-hutch-optimal}.
+Correspondingly, we will need to show that with $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2})$ the Nyström approximation verifies $\int \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)\rVert _F~\mathrm{d}t \leq \varepsilon \int \Trace(\mtx{B}(t))~\mathrm{d}t$, regardless of the structure of the matrix $\mtx{B}(t)$.
+Just as in the corresponding proof for the Girard-Hutchinson estimator \refthm{thm:hutchinson}, we again bound the higher order moments of the approximation error for fixed $t$ and then use Minkowsi's integral inequality to carry over the results to the continuous case.
+We will do this in a sequence of three lemmas.
+
+For reasons which will become clear in the proof of \reflem{lem:nystrom}, we will need to bound $\mathbb{E}^{p}\left[\lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]$ for a diagonal matrix $\mtx{\Sigma}$ and a standard Gaussian random matrix $\mtx{\Omega}$.
+Several ways of bounding the moments of $\mtx{\Omega}$ already exist, however, most of them are either not sufficient for our purposes \cite{chen-2005-condition-numbers, edelman-1988-eigenvalues-condition, james-1964-distributions-matrix}.
+The corresponding result for the Frobenius norm has already been shown in \cite{kressner-2023-randomized-lowrank}.
+In \cite[lemma B.1]{tropp-2023-randomized-algorithms} a bound is achieved for $p = 2$ and $p = 4$.
+We will now generalize this result for arbitrary $p \in \mathbb{N}$.
+We first show the bound for Gaussian random vectors and then extend it to Gaussian random matrices in \reflem{lem:spectral-norm-moment}.
 
@@
         \mathbb{E}^{p}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2 \right]
-        \leq  \sqrt{k + p} \left( 3 \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{A} \rVert _F \right).
+        \leq  \sqrt{k + p} \left( \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{A} \rVert _F \right).
     \end{equation}
@@
     \begin{align}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right]
+        &\mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right] \notag \\
         &= \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{j + (i-1)k}^2}_{\leq \sigma_{1 + (i-1)k}^2} \omega_{j + (i-1)k}^2 \right] && \text{(separate sum into groups)} \notag \\
-        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{($\sigma_{j + (i-1)k}^2 \leq \sigma_{1 + (i-1)k}^2$)} \notag \\
+        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{($\sigma_{j + (i-1)k}^2 \leq \sigma_{1 + (i-1)k}^2$, $j=1, \dots, k$)} \notag \\
         &\leq \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \cdot \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{(triangle inequality)} \notag \\
-        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \right)^{\sfrac{2}{p}} && \text{(\refequ{equ:singular-value-group-bound} and $\chi_k^2$ moments \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
+        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \right)^{\sfrac{2}{p}} && \text{(\refequ{equ:singular-value-group-bound} and \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
         &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) (k + p). && \text{(\reflem{lem:gamma})}
@@
             \smash{\overbrace{\mtx{\Lambda}_1}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2}_{(n-k) \times (n-k)}}
-        \end{bmatrix}
-    \end{equation}
-    for some \todo{$k \leq n/2$}.
+        \end{bmatrix}.
+    \end{equation}
     Further, for a standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$ we define $\mtx{\Omega}_1 = \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times 2k}$ and $\mtx{\Omega}_2 = \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times 2k}$, which are likewise standard Gaussian.
@@
 
-Finally, we combine the analysis of the two estimators from \refsec{subsec:hutchinson} and \refsec{subsec:nystrom} into a more efficient, general-purpose trace estimator.
+Finally, we can combine the bound on the Girard-Hutchinson estimator \refthm{thm:hutchinson} and the one on the Nyström approximation \reflem{lem:nystrom} to obtain our main result.
 
@@
 
-Comparing \refthm{thm:nystrom-pp} with its equivalent for constant matrices \cite[theorem 3.4]{persson-2022-improved-variants}, we notice that our choice of $\delta$ is more restricted and the dependence of $n_{\mtx{\Psi}}$ and $n_{\mtx{\Omega}}$ on $\delta$ is slightly worse.
+Comparing \refthm{thm:nystrom-pp} with its equivalent for constant matrices \cite[theorem 3.4]{persson-2022-improved-variants}, we notice that our choice of $\delta$ is more restricted and the dependence of $n_{\mtx{\Psi}}$ and $n_{\mtx{\Omega}}$ on $\delta$ is worse.
 
@@
 
-The algorithm also accomodates the two other estimators introduced in \refsec{sec:analysis}: For $n_{\mtx{\Omega}} = 0$ we recover \refequ{equ:hutchinson-trace-estimator}, and the complexity for each evaluation of the estimate decreases significantly.
-For $n_{\mtx{\Psi}} = 0$ we end up with \refequ{equ:nystrom-trace-estimator}.
-
-\todo{Clearly need not divide by $n_{\mtx{\Psi}}=0$, hence just ignore this term. }
+The algorithm also accomodates the two other estimators introduced in \refsec{sec:analysis}: For $n_{\mtx{\Omega}} = 0$ we recover the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}, and the complexity for each evaluation of the estimate decreases significantly.
+For $n_{\mtx{\Psi}} = 0$ we end up with the Nyström estimator \refequ{equ:nystrom-trace-estimator}, provided the implementation ignores the second term on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
+
 %Chebyshev $\mathcal{O}(n_t (m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + n_{\mtx{\Omega}}^3) + mn^2(n_{\mtx{\Psi}} + n_{\mtx{\Omega}}))$.
@@
 
+The pseudoinverses on \reflin{lin:4-nystromchebyshev-nystrom-pp} need to be computed with care.
 \todo{Discuss computation of pseudo inverse: Generalized eigenvalue problem with thresholding \cite{lin-2017-randomized-estimation, epperly-2022-theory-quantuma}, }

### 2024-06-20 paper/main.tex
expand applications section

@@
 
+\todo{
 \begin{abstract}
-    \todo{Write abstract.}
+    Write abstract.
 \end{abstract}
+}
 
@@
 For constant matrices, the most widespread methods for implicit trace estimation are the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} and the variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making, chen-2023-krylovaware-stochastic}.
-Some of these estimators have already been successfully applied to parameter-dependent matrices \cite{lin-2017-randomized-estimation}, but they have not yet been analyzed.
-\todo{A related technique is dynamic trace estimation for sequences $\mtx{B}_1, \dots, \mtx{B}_m$ of similar matrices \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace}, which achieves guarantees for the estimate $b_i$ of $\Trace(\mtx{B}_i)$ of the form $\mathbb{P}(|\Trace(\mtx{B}_i) - b_i| \geq \varepsilon) \leq \delta$ for $i=1,\dots, m$.
-A drawback of this type of bound is that when we aim for $\varepsilon$ accuracy for all $i=1,\dots, m$ simultaneously, the probability will grow linearly with $m$ by the union bound.
-Therefore, instead of discretizing $[a, b]$ with $t_1, \ldots, t_m$ and looking at the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way to quantify the error is to use $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate for each $t \in [a, b]$.}
-
-\paragraph{Contribution}
-We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices.
+Some of these estimators have already been successfully applied to parameter-dependent matrices \cite{lin-2017-randomized-estimation,chen-2023-krylovaware-stochastic}, but they have not yet been analyzed in this setting.
+A related technique is dynamic trace estimation for sequences $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ of matrices which satisfy $\lVert \mtx{B}(t_i) \rVert _F \leq 1$ and which change dynamically, i.e.
+$\lVert \mtx{B}(t_{i+1}) - \mtx{B}(t_i) \rVert _F \leq \alpha < 1$ for the Frobenius norm $\lVert \cdot \rVert _F$ \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace}.
+However, in our setting we often cannot guarantee either of the two conditions.
+Instead of discretizing $[a, b]$ into $t_1, \ldots, t_m$, approximating $\Trace(\mtx{B}(t_i))$ with $b(t_i)$, and analyzing the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way is to quantify the error with $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate in each $t \in [a, b]$.
+This approach has already been taken for analyzing low-rank approximations of parameter-dependent matrices \cite{kressner-2023-randomized-lowrank}.
+
+\paragraph{Contribution} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices.
 In particular, we reuse the same randomization for each value of the parameter $t$ in the estimators, which makes them scale favorably with the number of parameter evaluations.
 Subsequently, we propose a simple, fast, robust, and provable procedure in which these estimators are employed for approximating the smooth spectral density $\phi_{\sigma}$ of a matrix.
+
+\paragraph{Structure} \todo{Write structure.}
 
@@
 
-\paragraph{Nyström approximation} Alternatively, the trace of a symmetric matrix whose rank is significantly lower than its size can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström approximation \cite{gittens-2013-revisiting-nystrom}
+\paragraph{Nyström estimator} Alternatively, the trace of a symmetric matrix whose singular values decay quickly can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström approximation \cite{gittens-2013-revisiting-nystrom}
 \begin{equation}
@@
 
-In the remainder of this section, we will derive upper bounds on the error of these estimators.
+%In the remainder of this section, we will derive upper bounds on the error of these estimators.
 %
@@
 
-We first consider the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}.
+We first analyze the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}.
 We bound its $L^1$ norm in terms of the integral over the Frobenius norm $\lVert \cdot \rVert _F$ of the matrix whose trace we aim to approximate.
@@
     \begin{equation}
-        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) \right| ~\mathrm{d}t < 2^4 \gamma k \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F  ~\mathrm{d}t.
-    \end{equation}
-    In particular, if we choose $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{- \sfrac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t < \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$.
+        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) \right| ~\mathrm{d}t \leq 2^4 \gamma k \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F  ~\mathrm{d}t.
+    \end{equation}
+    In particular, if we choose $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{- \sfrac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$.
 \end{theorem}
@@
         \mathbb{E}^{2k}\left[ r(\mtx{B}, \vct{\psi}) \right]
-        < \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F < 2^4 k \lVert \mtx{B} \rVert _F,
+        \leq \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F \leq 2^4 k \lVert \mtx{B} \rVert _F,
         \label{equ:hutchinson-trace-onequery-fixed}
@@
         \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}^{2 k}\left[ r(\mtx{B}(t), \vct{\psi}) \right]~\mathrm{d}t
-        < 2^4 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
+        \leq 2^4 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
     \end{equation}
@@
     \begin{equation}
-        \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t < 2^4 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
+        \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t \leq 2^4 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
         \label{equ:hutchinson-trace-onequery-uniform}
@@
         \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j \right| ~ \mathrm{d}t
-        < 2^4 k \gamma \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
+        \leq 2^4 k \gamma \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
     \end{equation}
@@
     \begin{equation}
-        \underbrace{\sigma_1, \dots, \sigma_k}_{\leq \sigma_1}, \underbrace{\sigma_{k+1}, \dots, \sigma_{2k}}_{\leq \sigma_{k+1}}, \dots, \underbrace{\sigma_{1+(\ell - 1)k}, \dots, \sigma_{\ell k}}_{\leq \sigma_{1+(\ell - 1)k}}.
+        \overbrace{\underbrace{\sigma_1, \dots, \sigma_k}_{\leq \sigma_1}}^{\geq \sigma_{k+1}}, \overbrace{\underbrace{\sigma_{k+1}, \dots, \sigma_{2k}}_{\leq \sigma_{k+1}}}^{\geq \sigma_{2k+1}}, \dots, \overbrace{\underbrace{\sigma_{1+(\ell - 1)k}, \dots, \sigma_{\ell k}}_{\leq \sigma_{1+(\ell - 1)k}}}^{\geq 0}.
     \end{equation}
@@
 Choosing the smoothing parameter $\sigma$ results in the following trade off: Typically, larger $\sigma$ make it easier to approximate the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$, whereas smaller $\sigma$ allow us to stay closer to the original spectral density \refequ{equ:spectral-density}.
-To see the latter, we measure the error between spectral density $\phi$ and its smoothened version $\phi_{\sigma}$ with
+To see the latter, we measure the error between the spectral density $\phi$ and its smoothened version $\phi_{\sigma}$ with
 \begin{equation}
@@
 where $\mathcal{S}$ represents an appropriately chosen space of test functions.
-Among others, \cite{lin-2016-approximating-spectral} proposes $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - \lambda), \lambda \in [\lambda_{\text{min}}, \lambda_{\text{max}}]\}$, \cite{chen-2021-analysis-stochastic} uses $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [-1, 1] \}$ where $\Theta$ is the Heaviside step function, and \cite{braverman-2022-sublinear-time} uses $\mathcal{S} = \{f : |f(x) - f(y)| \leq |x - y| \}$ for which \refequ{equ:error-metric} is equivalent to the Wasserstein-1 distance between $\phi$ and $\phi_{\sigma}$.
+Among others, \cite{lin-2016-approximating-spectral} proposes $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - \lambda), \lambda \in [\lambda_{\text{min}}, \lambda_{\text{max}}]\}$, \cite{chen-2021-analysis-stochastic} uses $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [-1, 1] \}$ where $\Theta$ is the Heaviside step function, and \cite{braverman-2022-sublinear-time} uses $\mathcal{S} = \{f : |f(x) - f(y)| \leq |x - y| \}$ for which the error \refequ{equ:error-metric} is equivalent to the Wasserstein-1 distance between $\phi$ and $\phi_{\sigma}$.
 In this last metric, if we assume the spectrum to be somewhat uniform and $\sigma$ to be rather small, a simple calculation allows us to roughly estimate the smoothing error \refequ{equ:error-metric} to be $\sigma$.
@@
 \end{equation}
-With the help of \refequ{equ:matrix-expansion} we can now define the expanded spectral density
+With the help of the Chebyshev expansion \refequ{equ:matrix-expansion} we can now define the expanded spectral density
 \begin{equation}
@@
     \begin{cases}
-        \tau : [a, b] \to [-1, 1],\\
-        \tau(t) = \frac{2t - a - b}{b - a}.
+        \tau : [a, b] \to [-1, 1]\\
+        \tau(t) = \frac{2t - a - b}{b - a}
     \end{cases}
 \end{equation}
-then lets us define the matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ whose spectrum is identical to the one of $\mtx{A}$ but shrunk to the interval $[-1, 1]$, for which \refequ{equ:matrix-expansion} is a valid expansion.
+then lets us define the matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ whose spectrum is identical to the one of $\mtx{A}$ but shrunk to the interval $[-1, 1]$, for which the expansion \refequ{equ:matrix-expansion} is valid.
 Note that through this transformation we will also have to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when reverting the spectral transform $\tau$.
@@
 
-Our goal is to apply the trace estimator analyzed in \refsec{subsec:nystrom-pp} to the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ to approximate the smooth spectral density $\phi_{\sigma}$.
-However, due to the oscillatory nature of the Chebyshev expansion, we can no longer guarantee that $g_{\sigma}^{(m)} \geq 0$, and consequently, the matrix function $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ may be indefinite.
-To still apply \refthm{thm:nystrom-pp} in this scenario, we will have to ensure that the Chebyshev expansion is non-negative.
+Our goal is to apply the Nyström++ trace estimator which we analyzed in \refsec{subsec:nystrom-pp} to the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ to approximate the smooth spectral density $\phi_{\sigma}$.
+However, due to the oscillatory nature of the Chebyshev expansion, we can no longer guarantee that $g_{\sigma}^{(m)} \geq 0$, and consequently, the matrix function $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ may be (slightly) indefinite.
+While in our experiments this never seemed to be an issue, it disallows us to directly use \refthm{thm:nystrom-pp}.
+While variants of the Nyström approximation designed for indefinite matrices have been analyzed, no practical algorithms for computing them exist \cite{nakatsukasa-2023-randomized-lowrank}.
+
+To still apply the bound from \refthm{thm:nystrom-pp} in this scenario, we will have to ensure that the Chebyshev expansion is non-negative.
 We can think of many ways in which this can be achieved: by using Jackson damping \cite{jackson-1912-approximation-trigonometric,braverman-2022-sublinear-time}, albeit at a significantly worse rate of convergence; by shifting up the expansion $g_{\sigma}^{(m)} + \rho$ with a large enough $\rho$, at the loss of most of the low-rank structure of $g_{\sigma}(t \mtx{I}_n - \mtx{A})$; or by directly minimizing the approximation error over all non-negative Chebyshev polynomials \cite{fejer-1916-uber-trigonometrische}, which realistically only works for small $m$.
-In the end, we propose the following algorithm to compute a non-negative Chebyshev expansion.
-First, expand the square root of the smoothing kernel in the Chebyshev polynomials up to degree $m/2$
-\begin{equation}
-    \sqrt{g_{\sigma}(t - s)}^{(\sfrac{m}{2})} = \sum_{l=0}^{m/2} \xi_l(t) T_l(s).
+In the end, we propose the following method to produce a non-negative Chebyshev expansion: First, expand the square root of the smoothing kernel in the Chebyshev polynomials up to degree $m/2$
+\begin{equation}
+    \sqrt{g_{\sigma}}^{(\sfrac{m}{2})}(t - s) = \sum_{l=0}^{m/2} \xi_l(t) T_l(s).
 \end{equation}
@@
     \underline{g}^{(m)}_{\sigma}(t - s) = \sum_{l=0}^{m} \mu_l(t) T_l(s).
+    \label{equ:non-negative-chebyshev-expansion}
 \end{equation}
@@
 
-\todo{
-Maybe mention indefinite Nyström [Nakatsukasa], which won't give us desirable guarantees, and refinement of Chebyshev expansion [Francis Bach], which only slightly improves, but is much harder to implement.
-Side-note: it is possible to refine the coefficients of the Chebyshev expansion to get a better approximation in the supremum-norm over the set of all non-negative polynomials \cite{fejer-1916-uber-trigonometrische}.
+%\todo{
+%Maybe mention indefinite Nyström [Nakatsukasa], which won't give us desirable guarantees, and refinement of Chebyshev expansion [Francis Bach], which only slightly improves, but is much harder to implement.
+Side-note: it is possible to refine the coefficients of the Chebyshev expansion to get a better approximation in the supremum-norm over the set of all non-negative polynomials \cite%{fejer-1916-uber-trigonometrische}.
 The non-convex nature of this problem and the relatively small improvement we get with this technique made us decide to not pursue this further.
-}
+%}
 
@@
 
-An efficient implementation of \refequ{equ:chebyshev-nystrom-formula} can be achieved thanks to the invariance of the trace under cyclic permutation of its arguments and the recurrence relation \refequ{equ:chebyshev-recurrence} which the Chebyshev polynomials satisfy.
-To illustrate this, consider the estimator \refequ{equ:chebyshev-nystrom-formula} when $\mtx{\Psi}$ is zero.
+An efficient implementation of the approximation \refequ{equ:chebyshev-nystrom-formula} can be achieved thanks to the invariance of the trace under cyclic permutation of its arguments and the recurrence relation \refequ{equ:chebyshev-recurrence} which the Chebyshev polynomials satisfy.
+To illustrate this, consider the approximation \refequ{equ:chebyshev-nystrom-formula} when $\mtx{\Psi}$ is zero.
 Using the definition of the Nyström approximation \refequ{equ:nystrom-approximation}, the cyclic invariance of the trace, and the symmetry of $\mtx{A}$, we can rewrite it as
@@
 for which we first precompute the $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ for $l=0, \dots, m$ using the recurrence \refequ{equ:chebyshev-recurrence} and then inexpensively sum them up with the corresponding coefficients $\mu_l(t)$ for all $t$ in which we want to evaluate the approximation.
-Thanks to \refalg{alg:chebyshev-squaring}, the second part of \refequ{equ:cyclic-property} can be approximated in an equivalent way \todo{(which is consistent, unlike \cite{lin-2017-randomized-estimation})}.
-
+Thanks to the fast and exact squaring of Chebyshev expansions (cf.
+\refalg{alg:chebyshev-squaring}), the second part of \refequ{equ:cyclic-property} can be approximated in an equivalent way.
+Unlike expanding the squared smoothing kernel $g_{\sigma}^2$ separately \cite{lin-2017-randomized-estimation}, this approach is consistent, which we have observed to make a noticeable difference in terms of accuracy and does not hinder provability.
 Similar observations for the case when $\mtx{\Psi}$ is nonzero lead to the following algorithm.
@@
 \begin{algorithmic}[1]
-    \Statex \textbf{Input:} Symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$, evaluation points $\{t_i\}_{i=1}^{n_t}$
-    \Statex \textbf{Parameters:} Degree $m$, allocation $(n_{\mtx{\Psi}}, n_{\mtx{\Omega}})$,  smoothing parameter $\sigma$
+    \Statex \textbf{Input:} Symmetric $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum in $[-1, 1]$, points $\{t_i\}_{i=1}^{n_t} \subset \mathbb{R}$
+    \Statex \textbf{Parameters:} Degree of expansion $m \in 2\mathbb{N}$, number of Girard-Hutchinson queries $n_{\mtx{\Psi}} \in \mathbb{N}_0$, Nyström sketch size $n_{\mtx{\Omega}} \in \mathbb{N}_0$,  smoothing parameter $\sigma > 0$
     \Statex \textbf{Output:} Approximate evaluations of the spectral density $\{\widetilde{\phi}_{\sigma}(t_i)\}_{i=1}^{n_t}$
-    \State Compute $\{\mu_l(t_i)\}_{l=0}^{m}$ for all $t_i$ using the $\DCT$ \refequ{equ:discrete-cosine-transform}
+    \State Compute $\{\mu_l(t_i)\}_{l=0}^{m}$ for all $t_i$ with non-negative Chebyshev expansion \refequ{equ:non-negative-chebyshev-expansion}
     \State Compute $\{\nu_l(t_i)\}_{l=0}^{2m}$ using \refalg{alg:chebyshev-squaring}%\refalg{alg:2-chebyshev-chebyshev-expansion}
     %\State Compute  for all $t_i$ using %\refalg{alg:3-nystrom-chebyshev-exponentiation}
-    \State Generate Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$%\glsfirst{sketching-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$
+    \State Generate standard Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$%\glsfirst{sketching-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$
     %\State Generate standard Gaussian %\glsfirst{random-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$
@@
 The algorithm also accomodates the two other estimators introduced in \refsec{sec:analysis}: For $n_{\mtx{\Omega}} = 0$ we recover the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}, and the complexity for each evaluation of the estimate decreases significantly.
-For $n_{\mtx{\Psi}} = 0$ we end up with the Nyström estimator \refequ{equ:nystrom-trace-estimator}, provided the implementation ignores the second term on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
+For $n_{\mtx{\Psi}} = 0$ we end up with the Nyström estimator \refequ{equ:nystrom-trace-estimator}, provided the implementation evaluates $0/0$ to $0$ in the second term on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
 
@@
 The pseudoinverses on \reflin{lin:4-nystromchebyshev-nystrom-pp} need to be computed with care.
-\todo{Discuss computation of pseudo inverse: Generalized eigenvalue problem with thresholding \cite{lin-2017-randomized-estimation, epperly-2022-theory-quantuma}, }
-
-\todo{Explain short-circuit mechanism -> very crucial if matrix can be zero}
+It is possible to reformulate the problem as a generalized eigenvalue problem and apply thresholding to the smallest eigenvalues of a certain matrix to enforce stability \cite{lin-2017-randomized-estimation, epperly-2022-theory-quantuma}.
+Since this renders the algorithm significantly more complicated but only improved the accuracy slightly during our experiments, we will instead evaluate these expressions in \reflin{lin:4-nystromchebyshev-nystrom-pp} using a least-squares solver which likewise uses a truncation of the smallest eigenvalues to enforce better conditioning.
+
+What we have found to be crucial for a successful approximation of the full spectral density was the detection of regions wherein lie no eigenvalues, i.e.
+regions where the spectral density nearly vanishes.
+For small smoothing parameters $\sigma$, the rapid decay of the Gaussian $g_{\sigma}$ will cause $\mtx{K}(t_i) \approx g_{\sigma}(t_i\mtx{I}_n - \mtx{A})$ to be close to the zero matrix for parameter values $t_i$ in such regions.
+Hence, on line \reflin{lin:4-nystromchebyshev-nystrom-pp} we would compute the pseudoinverse of an almost zero matrix, which can be extremely unstable.
+Notice that $\Trace(\mtx{K}_1(t_i)) / n_{\mtx{\Omega}}$ is the $n_{\mtx{\Omega}}$-query Girard-Hutchinson estimator.
+We may use it as a rough indicator whether $t_i$ lies within a region where the spectral density nearly vanishes, and subsequently set $\widetilde{\phi}_{\sigma}(t_i) = 0$ instead of computing the full expression on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
 
@@
 
-% Acknowledge advice from David and Data/Parameters from Lin Lin
+% Acknowledge advice from David and Data/Parameters from Lin Lin, Michael Herbst?
 

### 2024-06-20 paper/main.tex
remove dependency of delta on epsilon

@@
     \begin{equation}
-        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
+        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t
+        \label{qu:nystrompp-theorem-bound}
     \end{equation}
     for a universal constant $c > 0$.
-In particular, if we choose $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1}))$, then with probability at least $1-\delta$ for $\delta \in (0, e^{-e^{4 \varepsilon^2}})$ and $\varepsilon > 0$ we have $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$.
+In particular, if we choose $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1}) + \log(\delta^{-1}))$, then with probability at least $1-\delta$ for $\delta \in (0, 1)$ and $\varepsilon > 0$ we have $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$.
 \end{lemma}
@@
 
-    Setting $\delta = \gamma^{-\sfrac{k}{4}}$ and choosing $k = \lceil \varepsilon^{-2}\log(\delta^{-1}) \rceil$ we get
-    \begin{align}
-        \gamma \frac{1}{\sqrt{k}}
-        &= \delta^{-\sfrac{2}{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} \frac{1}{\sqrt{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} && \text{(definition of $\delta$ and choice of $k$)} \notag \\
-        &= e^{2\varepsilon^{2} \frac{\varepsilon^{-2}\log(\delta^{-1})}{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} \frac{1}{\sqrt{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} && \text{($\delta = e^{-\log(\delta^{-1})}$)} \notag \\
-        &\leq e^{2\varepsilon^{2}} \frac{\varepsilon}{\sqrt{ \log(\delta^{-1})}} && \text{($\lceil x \rceil \geq x$ if $x \geq 0$)}
-    \end{align}
-    which is smaller than $\varepsilon$ if $\delta \leq e^{-e^{4 \varepsilon^2}}$from which follows the second part of the theorem when using $n_{\mtx{\Omega}} = 2k$.
+    Fixing $\gamma = e$ and inserting $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1}) + \log(\delta^{-1}))$ into \refequ{equ:nystrompp-theorem-bound} we get the second part of the theorem.
+
+    %Setting $\delta = \gamma^{-\sfrac{k}{4}}$ and choosing $k = \lceil \varepsilon^{-2}\log(\delta^{-1}) \rceil$ we get
+    %\begin{align}
+    %    \gamma \frac{1}{\sqrt{k}}
+    %    &= \delta^{-\sfrac{2}{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} \frac{1}{\sqrt{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} && \text{(definition of $\delta$ and choice of $k$)} \notag \\
+    %    &= e^{2\varepsilon^{2} \frac{\varepsilon^{-2}\log(\delta^{-1})}{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} \frac{1}{\sqrt{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} && \text{($\delta = e^{-\log(\delta^{-1})}$)} \notag \\
+    %    &\leq e^{2\varepsilon^{2}} \frac{\varepsilon}{\sqrt{ \log(\delta^{-1})}} && \text{($\lceil x \rceil \geq x$ if $x \geq 0$)}
+    %\end{align}
+    %which is smaller than $\varepsilon$ if $\delta \leq e^{-e^{4 \varepsilon^2}}$from which follows the second part of the theorem when using $n_{\mtx{\Omega}} = 2k$.
 \end{proof}
@@
     Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(2\delta^{-1})^2)$ with \emph{even} $n_{\mtx{\Omega}} \geq 2$, then for $\delta \in (0, e^{-e^{4\varepsilon}} / 2)$ with probability at least $1 - \delta$
+If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(2\delta^{-1})^2 + \log(2\delta^{-1}))$ with \emph{even} $n_{\mtx{\Omega}} \geq 2$, then for $\delta \in (0, 1/2)$ and $\varepsilon > 0$ with probability at least $1 - \delta$
     \begin{equation}
@@
 
-Comparing \refthm{thm:nystrom-pp} with its equivalent for constant matrices \cite[theorem 3.4]{persson-2022-improved-variants}, we notice that our choice of $\delta$ is more restricted and the dependence of $n_{\mtx{\Psi}}$ and $n_{\mtx{\Omega}}$ on $\delta$ is worse.
+Comparing \refthm{thm:nystrom-pp} with its equivalent for constant matrices \cite[theorem 3.4]{persson-2022-improved-variants}, we notice a less favorable scaling in terms of $\log(2\delta^{-1})$.
 
@@
     Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m \in \mathbb{N}$, $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2)$, and smoothing parameter $\sigma > 0$ satisfies with probability at least $1 - \delta$
+Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m \in \mathbb{N}$, $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2 + \log(2\delta^{-1}))$, and smoothing parameter $\sigma > 0$ satisfies with probability at least $1 - \delta$
     \begin{equation}
@@
     \end{equation}
-    for all $\delta \in (0, 2^{-1} \min\{e^{\sfrac{1}{2}}, e^{-e^{8\varepsilon}}\})$.
+    for all $\varepsilon > 0$ and $\delta \in (0, 1/2)$.
 \end{theorem}

### 2024-06-20 paper/main.tex
rewrite experiment description

@@
     Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(2\delta^{-1})^2 + \log(2\delta^{-1}))$ with \emph{even} $n_{\mtx{\Omega}} \geq 2$, then for $\delta \in (0, 1/2)$ and $\varepsilon > 0$ with probability at least $1 - \delta$
+If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(2\delta^{-1})^2 + \log(2\delta^{-1}))$ with \emph{even} $n_{\mtx{\Omega}}$, then for $\delta \in (0, 1/2)$ and $\varepsilon > 0$ with probability at least $1 - \delta$
     \begin{equation}
@@
     Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m \in \mathbb{N}$, $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2 + \log(2\delta^{-1}))$, and smoothing parameter $\sigma > 0$ satisfies with probability at least $1 - \delta$
+Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m \in \mathbb{N}$, $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2 + \log(2\delta^{-1}))$, $n_{\mtx{\Omega}}$ \emph{even}, and smoothing parameter $\sigma > 0$ satisfies with probability at least $1 - \delta$
     \begin{equation}
@@
 \end{equation}
-where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from spectral-parameter, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t)$.
+where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from spectral-parameter $t$, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t)$.
 Consequently, by using the closed-form expression of the eigenvalues \refequ{equ:gaussian-kernel-eigenvalues}, we may upper bound the numerical rank of a Gaussian smoothing kernel as
@@
 
-For our first example, we consider the matrix which arises from the second order
-finite difference discretization of the Laplace operator $\Delta$ in a potential
-field $V$,
-\begin{equation}
-    \mathcal{A} u(\vct{x}) = - \Delta u(\vct{x}) + V(\vct{x}) u(\vct{x}),
+For our first example, we consider the example from \cite{lin-2017-randomized-estimation}, i.e. matrices which arise from the second order finite difference discretization of the Hamiltonian
+\begin{equation}
+    \mathcal{H} u(\vct{x}) = - \Delta u(\vct{x}) + V(\vct{x}) u(\vct{x}),
     \label{equ:5-experiments-electronic-hamiltonian}
 \end{equation}
-for a uniform mesh of size $h=0.6$. The potential $V$ results from a
-lattice whose primitive cell is of side-length $L=6$ and in whose center a
-potential
+in a cubic domain for a uniform mesh of size $h=0.6$.
+The potential $V$ arises from a primitive cell of length $L=6$ which is repeated $n_c \in \mathbb{N}$ times in every spatial dimension, yielding discretization matrices which are growing in size with $n_c$.
+A Gaussian well
 \begin{equation}
@@
 \end{equation}
-with $\alpha = -4$, $\beta = 2$ is located. The computational domain is chosen
-to span $n_c \in \mathbb{N}$ primitive cells in every spatial dimension, hence, yielding
-discretization matrices which are growing in size with $n_c$. In our experiments
-we consider the three-dimensional case.%, but for visualization purposes, we illustrate the potential in \reffig{fig:gaussian-well} in two dimensions.
-
-%\begin{figure}[ht]
-%    \begin{subfigure}[b]{0.32\columnwidth}
-%        \input{plots/gaussian-well-1.pgf}
-%        \caption{$n_c=1$}
-%        \label{fig:gaussian-well-1}
-%    \end{subfigure}
-%    \begin{subfigure}[b]{0.32\columnwidth}
-%        \input{plots/gaussian-well-2.pgf}
-%        \caption{$n_c=2$}
-%        \label{fig:gaussian-well-2}
-%    \end{subfigure}
-%    \begin{subfigure}[b]{0.32\columnwidth}
-%        \input{plots/gaussian-well-5.pgf}
-%        \caption{$n_c=5$}
-%        \label{fig:gaussian-well-5}
-%    \end{subfigure}
-%    \caption{Two dimensional periodic potential $V$ for different sizes $n_c$ of the computational domain.}
-%    \label{fig:gaussian-well}
-%\end{figure}
+with $\alpha = -4$, $\beta = 2$ is located in the center of each of these cells.
+The Hamiltonian models the interaction of a single electron in a crystal.
+In our experiments we consider the three-dimensional case, but for visualization purposes, we illustrate the potential in \reffig{fig:gaussian-well} in two dimensions.
+
+\begin{figure}[ht]
+    \begin{subfigure}[b]{0.32\columnwidth}
+        \input{plots/gaussian-well-1.pgf}
+        \caption{$n_c=1$}
+        \label{fig:gaussian-well-1}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.32\columnwidth}
+        \input{plots/gaussian-well-2.pgf}
+        \caption{$n_c=2$}
+        \label{fig:gaussian-well-2}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.32\columnwidth}
+        \input{plots/gaussian-well-5.pgf}
+        \caption{$n_c=5$}
+        \label{fig:gaussian-well-5}
+    \end{subfigure}
+    \caption{Periodic Gaussian well potential $V$ for different sizes $n_c$ of the computational domain.}
+    \label{fig:gaussian-well}
+\end{figure}
 
@@
 
-\todo{[Comparison plots with other estimators]}
+\todo{[Comparison plots with other estimators?]}
 

### 2024-06-20 paper/main.tex
rewrite numerical section

@@
 
-\paragraph{Structure} \todo{Write structure.}
+\paragraph{Structure} Write structure.
+In \refsec{sec:analysis} we will analyze three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter dependent matrices.
+We propose and analyze a method for using the aforementioned estimators to estimate the spectral density of real symmetric matrices in \refsec{sec:application}.
+Finally, we will test our developments on an example from electronic structure calculation.
 
@@
 
-Just as in the constant matrix case, our goal is now to prove that the Nyström++ estimator \refequ{equ:nystrompp-trace-estimator} achieves an $\varepsilon$ relative error with just $\mathcal{O}(\varepsilon)$ random vectors to construct the estimate.
+Just as in the constant matrix case, our goal is now to prove that the Nyström++ estimator \refequ{equ:nystrompp-trace-estimator} achieves an $\varepsilon$ relative error with just $\mathcal{O}(\varepsilon^{-1})$ random vectors to construct the estimate.
 To do so, we mimic the proof of a similar bound for the Hutch++ estimator \cite[theorem 3.1]{meyer-2021-hutch-optimal}.
@@
 
+\todo{Summarize analysis}
+
 \begin{lemma}{Non-negative Chebyshev expansion error}{non-negative-chebyshev-error}
@@
 
-For our first example, we consider the example from \cite{lin-2017-randomized-estimation}, i.e. matrices which arise from the second order finite difference discretization of the Hamiltonian
-\begin{equation}
-    \mathcal{H} u(\vct{x}) = - \Delta u(\vct{x}) + V(\vct{x}) u(\vct{x}),
+Our first numerical example comes from electronic structure interaction \cite{lin-2017-randomized-estimation}, more precisely from the second order finite difference discretization of the Hamiltonian
+\begin{equation}
+    \mathcal{H} = - \Delta + V.
     \label{equ:5-experiments-electronic-hamiltonian}
 \end{equation}
-in a cubic domain for a uniform mesh of size $h=0.6$.
-The potential $V$ arises from a primitive cell of length $L=6$ which is repeated $n_c \in \mathbb{N}$ times in every spatial dimension, yielding discretization matrices which are growing in size with $n_c$.
-A Gaussian well
-\begin{equation}
-    \alpha e^{-\sfrac{\lVert \vct{x} \rVert _2^2}{ 2 \beta^2 }}
+The potential $V$ interacting with the electrons is generated by Gaussian wells
+\begin{equation}
+    v(r) = v_0 e^{-\lambda r^2}
     \label{equ:5-experiments-gaussian-cell}
 \end{equation}
-with $\alpha = -4$, $\beta = 2$ is located in the center of each of these cells.
-The Hamiltonian models the interaction of a single electron in a crystal.
-In our experiments we consider the three-dimensional case, but for visualization purposes, we illustrate the potential in \reffig{fig:gaussian-well} in two dimensions.
+with $V_0 = -4$ and $\lambda = 8$, centered in cells of side length $L=6$ which are stacked $n_c \in \mathbb{N}$ times in each spatial dimension (cf.
+\reffig{fig:gaussian-well}).
+The discretization step is fixed to be $h=0.6$, such that the size of the matrix grows cubically with $n_c$.
+This is an idealized model for the interaction of nuclei on a regular grid with electrons for a $k$-vector in the center of the first Brillouin zone.
+The distribution of the eigenvalues of the Hamiltonian -- its spectral density -- allows us to interpret the system's energy levels.
 
@@
     \end{subfigure}
-    \caption{Periodic Gaussian well potential $V$ for different sizes $n_c$ of the computational domain.}
+    \caption{Periodic Gaussian well potential $V$ for different sizes of the supercell.}
     \label{fig:gaussian-well}

### 2024-06-24 paper/main.tex
do some small adjustments and extensions

@@
 
-\paragraph{Structure} Write structure.
-In \refsec{sec:analysis} we will analyze three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter dependent matrices.
+\paragraph{Structure} In \refsec{sec:analysis} we will analyze three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter dependent matrices.
 We propose and analyze a method for using the aforementioned estimators to estimate the spectral density of real symmetric matrices in \refsec{sec:application}.
@@
     Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(2\delta^{-1})^2 + \log(2\delta^{-1}))$ with \emph{even} $n_{\mtx{\Omega}}$, then for $\delta \in (0, 1/2)$ and $\varepsilon > 0$ with probability at least $1 - \delta$
+If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(2\delta^{-1})^2 + \log(2\delta^{-1}))$ with \emph{even} $n_{\mtx{\Omega}}$, then for $\delta \in (0, e^{-\sfrac{1}{2}}/2)$ and $\varepsilon > 0$ with probability at least $1 - \delta$
     \begin{equation}
@@
 \end{proof}
+
+This theorem shows that for $\sigma \ll 1$, if we choose $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, the error we induce with the Chebyshev expansion is at most $\mathcal{O}(\varepsilon)$.
 
@@
 
-\todo{Summarize analysis}
+\todo{Summarize analysis (on one hand, half the degree $m$, but easier to interpolate because $\sigma \to \sqrt{2} \sigma$)}
 
@@
     Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m \in \mathbb{N}$, $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2 + \log(2\delta^{-1}))$, $n_{\mtx{\Omega}}$ \emph{even}, and smoothing parameter $\sigma > 0$ satisfies with probability at least $1 - \delta$
-    \begin{equation}
-        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t < 2 (1 + \varepsilon) \underline{E}_{\sigma, m} + \varepsilon
+Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, \emph{even} $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2 + \log(2\delta^{-1}))$, and smoothing parameter $\sigma > 0$ satisfies with probability at least $1 - \delta$
+    \begin{equation}
+        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq \varepsilon
         \label{equ:chebyshev-nystrom-error}
@@
     \end{equation}
-    for all $\varepsilon > 0$ and $\delta \in (0, 1/2)$.
+    for all $0 < \varepsilon \leq 1$ and $\delta \in (0, e^{-\sfrac{1}{2}}/2)$.
 \end{theorem}
@@
         &\leq (1 + \varepsilon) \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \varepsilon \int_{-1}^{1} \left| \phi_{\sigma}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
-        &\leq 2 (1 + \varepsilon) \underline{E}_{\sigma, m} + \varepsilon. && \text{(\reflem{lem:non-negative-chebyshev-error} and normalization)} \notag \\
+        &\leq (1 + \varepsilon) 2 \underline{E}_{\sigma, m} + \varepsilon. && \text{(\reflem{lem:non-negative-chebyshev-error} and normalization)} \notag \\
     \end{align}
+    From \reflem{lem:non-negative-chebyshev-error} we see that choosing $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$ ensures $\underline{E}_{\sigma, m} \leq \varepsilon$, and thus, since $\mathcal{\varepsilon} \leq 1$,
+    \begin{equation}
+        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq (1+\varepsilon) 2 \varepsilon + \varepsilon \leq 5 \varepsilon.
+    \end{equation}
 \end{proof}
@@
 \end{equation}
-$\varepsilon$ is usually taken to be the double machine precision, i.e.
-$10^{-16}$.
+$\varepsilon$ is usually taken to be the double machine precision, i.e. $10^{-16}$.
+
+%\begin{theorem}{Nyström estimator for parameter-dependent matrices}%{chebyshev-nystrom-raw}
+%    Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric %matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is %contained in $[-1, 1]$.
+Then the Chebyshev-Nyström++ estimate %$\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m \in %\mathbb{N}$, $n_{\mtx{\Omega}} \geq \max(6, \max_{t \in [-1, 1]}(r_%{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A}))) + \log(\delta^%{-1}))$, $n_{\mtx{\Psi}} = 0$, and smoothing parameter $\sigma > 0$ %satisfies with for all $\gamma \geq 1$ with probability at least $1 %- \delta$
+%    \begin{equation}
+%        \int_{a}^{b} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}%(t) \right| ~\mathrm{d}t
+%        \leq 2 e^2 (1 + r_{\varepsilon}) (\varepsilon + (n - r_%{\varepsilon}) \underline{E}_{\sigma, m}) + \underline{E}_%{\sigma, m}.
+%    \end{equation}
+%\end{theorem}
+%
+%\begin{proof}
+%    \todo{TODO}
+%\end{proof}
+
 When using a Gaussian smoothing-kernel, we can explicitly write the singular values of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ as
@@
 \begin{equation}
-    \mathcal{H} = - \Delta + V.
+    \mathcal{H} = - \Delta + V
     \label{equ:5-experiments-electronic-hamiltonian}
 \end{equation}
-The potential $V$ interacting with the electrons is generated by Gaussian wells
+in three dimensions. The potential $V$ interacting with the electrons is generated by Gaussian wells
 \begin{equation}
@@
 \end{equation}
-with $V_0 = -4$ and $\lambda = 8$, centered in cells of side length $L=6$ which are stacked $n_c \in \mathbb{N}$ times in each spatial dimension (cf.
+with $v_0 = -4$ and $\lambda = 8$, centered in cells of side length $L=6$ which are stacked $n_c \in \mathbb{N}$ times in each spatial dimension (cf.
 \reffig{fig:gaussian-well}).
@@
     \end{subfigure}
-    \caption{Periodic Gaussian well potential $V$ for different sizes of the supercell.}
+    \caption{Cross-sections of the periodic Gaussian well potential $V$ for different sizes of the supercell.}
     \label{fig:gaussian-well}

### 2024-06-24 paper/main.tex
make proof of nystrom lemma more explicit

@@
     \end{equation}
-    Further, for a standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$ we define $\mtx{\Omega}_1 = \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times 2k}$ and $\mtx{\Omega}_2 = \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times 2k}$, which are likewise standard Gaussian.
-
-    We will set $n_{\mtx{\Omega}} = 2k$, such that we can apply \cite[theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound
+    Further, for a standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ we define $\mtx{\Omega}_1 = \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times n_{\mtx{\Omega}}}$ and $\mtx{\Omega}_2 = \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times n_{\mtx{\Omega}}}$, which are likewise standard Gaussian.
+
+    We can apply \cite[theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound
     \begin{equation}
         \lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F 
-        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2
+        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2,
         \label{equ:nystrom-proof-persson-bonud}
@@
     \end{equation}
-    The second term is first processed with standard matrix norm manipulations to
+    \todo{In particular, its moments are also bounded with this.} The $p$-th moment of the second term can be processed with standard matrix norm manipulations to
     \begin{align}
-        \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        &= \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )( \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger})^{\top} \rVert _F \right] && \text{($\lVert \mtx{A} \rVert _{(4)}^2 = \lVert \mtx{A} \mtx{A}^{\top} \rVert _{F}$)} \notag \\
-        &\leq \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F \right] && \text{($\lVert \mtx{A} \mtx{B} \rVert _F \leq \lVert \mtx{A} \rVert _F \lVert \mtx{B} \rVert _2$)} \notag \\
-        &\leq \mathbb{E}^{\sfrac{k}{2}}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{($\lVert \mtx{A} \rVert _F \leq \sqrt{k} \lVert \mtx{A} \rVert _2$)} \notag \\
-        &\leq \mathbb{E}^{\sfrac{k}{2}}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{(submultiplicativity)} \notag \\
-        &\leq \sqrt{k} \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2\right] \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2  \right]  && \text{(independence $\mtx{\Omega}_1, \mtx{\Omega}_2$)} \notag \\
-        &= \sqrt{k} \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2 \right]^2 \mathbb{E}^{k}\left[  \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]^2 && \text{(definition of $\mathbb{E}^{k}[\cdot]$)}
+        \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
+        &= \mathbb{E}^{p}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )( \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger})^{\top} \rVert _F \right] && \text{($\lVert \mtx{A} \rVert _{(4)}^2 = \lVert \mtx{A} \mtx{A}^{\top} \rVert _{F}$)} \notag \\
+        &\leq \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F \right] && \text{($\lVert \mtx{A} \mtx{B} \rVert _F \leq \lVert \mtx{A} \rVert _F \lVert \mtx{B} \rVert _2$)} \notag \\
+        &\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{($\lVert \mtx{A} \rVert _F \leq \sqrt{k} \lVert \mtx{A} \rVert _2$)} \notag \\
+        &\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{(submultiplicativity)} \notag \\
+        &\leq \sqrt{k} \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2\right] \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2  \right]  && \text{(independence $\mtx{\Omega}_1, \mtx{\Omega}_2$)} \notag \\
+        &= \sqrt{k} \mathbb{E}^{2p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2 \right]^2 \mathbb{E}^{2p}\left[  \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]^2 && \text{(definition of $\mathbb{E}^{k}[\cdot]$)}
         \label{equ:nystrom-proof-processed-tail}
     \end{align}
-    We can apply \reflem{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$ to bound the first term with
-    \begin{equation}
-        \mathbb{E}^k\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2 \right]
-        \leq \sqrt{3 k} \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2 + \frac{1}{\sqrt{2k}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F \right).
+    We can apply \reflem{lem:spectral-norm-moment} to $\mtx{\Lambda}_2^{\sfrac{1}{2}}$ to bound the first term with
+    \begin{equation}
+        \mathbb{E}^{2p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2 \right]
+        \leq \sqrt{2p + n_{\mtx{\Omega}}} \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2 + \frac{1}{\sqrt{n_{\mtx{\Omega}}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F \right).
         \label{equ:spectral-norm-bound-applied}
     \end{equation}
-    The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
-    \begin{equation}
-        \mathbb{E}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^{k} \right]
-        = \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{k}{2}} \right]
-        \leq \left( k + 1 \right) \left( \frac{1}{(k + 1)!}\right)^{\sfrac{k}{k + 1}} \left( \frac{3 k}{2}\right)^{\sfrac{k}{2}}.
-    \end{equation}
-    With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\sfrac{1}{n}} \leq \frac{e}{n}$ follow respectively.
-Hence,
-    \begin{equation}
-        \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-        \leq \frac{e^2}{k + 1}\sqrt{ \frac{3k}{2} }
-        \leq e^2 \sqrt{\frac{3}{2k}}.
+    The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms} when $n_{\mtx{\Omega}} \geq k + 2p$
+    \begin{align}
+        \mathbb{E}^{2p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
+        &= \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{p} \right]^{\sfrac{1}{2p}}  \notag \\
+        &\leq \left( 1 + \frac{2p}{n_{\mtx{\Omega}} - k + 1 - 2p} \right)^{\sfrac{1}{2p}} \left( \frac{1}{(n_{\mtx{\Omega}} - m + 1)!}\right)^{\frac{1}{n_{\mtx{\Omega}} - m + 1}} \left( \frac{k + n_{\mtx{\Omega}}}{2}\right)^{\sfrac{1}{2}} \notag \\
+        &\leq \frac{e^2}{\sqrt{2}} \frac{\sqrt{k + n_{\mtx{\Omega}}}}{n_{\mtx{\Omega}} - k + 1}
         \label{equ:pinv-spectral-norm-bound}
-    \end{equation}
-    Plugging \refequ{equ:spectral-norm-bound-applied} and  \refequ{equ:pinv-spectral-norm-bound} into \refequ{equ:nystrom-proof-processed-tail}, and using $(a + b)^2 \leq 2a^2 + 2b^2$ for $a, b > 0$ gives us
-    \begin{equation}
-        \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F \right]
-        \leq  9 e^4 \sqrt{k} \left( 9 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{2k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right)
+    \end{align}
+    with the last steps following from the expansion of the exponential function $e^n = 1 + n + \dots + n^n/n!
++ \dots$ for $n \in \mathbb{N}$, i.e.
+$e^n \geq 1 + n$ and $e^n \geq n^n / n!$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( 1/n!
+\right)^{\sfrac{1}{n}} \leq e/n$ follow respectively.
+Hence, plugging \refequ{equ:spectral-norm-bound-applied} and  \refequ{equ:pinv-spectral-norm-bound} into \refequ{equ:nystrom-proof-processed-tail}, and using $(a + b)^2 \leq 2a^2 + 2b^2$ for $a, b > 0$ gives us
+    \begin{equation}
+        \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
+        \leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 9 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right)
     \end{equation}
     Here, we can identify $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 = \lambda_{k+1}$ and $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 = \Trace(\mtx{\Lambda}_2)$, and just as in the proof of \cite[lemma 3.1]{meyer-2021-hutch-optimal} we bound $\lambda_{k+1} \leq \Trace(\mtx{B})/k$ and $\Trace(\mtx{\Lambda}_2) \leq \Trace(\mtx{B})$ to get
-    \begin{equation}
-        \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F \right]
-        \leq 9 e^4 \sqrt{k} \left( \frac{9}{k} \Trace(\mtx{B}) + \frac{1}{2k} \Trace(\mtx{B}) \right) \leq  9 e^4 \left(9 + \frac{1}{2} \right) \frac{1}{\sqrt{k}} \Trace(\mtx{B})
+    \begin{align}
+        \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
+        &\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( \frac{9}{k} \Trace(\mtx{B}) + \frac{1}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) \right) \notag \\
+        &= \left( \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2} \frac{9n_{\mtx{\Omega}} + k}{n_{\mtx{\Omega}}}\right) \frac{\Trace(\mtx{B})}{\sqrt{k}}
         \label{equ:nystrom-proof-tail-bound}
-    \end{equation}
-    Finally, using the triangle inequality for $\mathbb{E}^{\sfrac{k}{2}}$ and inserting \refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud} we obtain
-    \begin{equation}
-        \mathbb{E}^{\sfrac{k}{2}} \left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right]
-        \leq \frac{c}{\sqrt{k}} \Trace(\mtx{B}).
-    \end{equation}
-    where $c = 1 + 9 e^4 (9 + \frac{1}{2})$.
+    \end{align}
+    To match the decay rate of the first term \refequ{equ:nystrom-proof-frobenius-trace}, we need to ensure that the term enclosed in parentheses is constant.
+We are free to choose $p > 1$ and $k \in \mathbb{N}$ such that $n_{\mtx{\Omega}} \geq k + 2p$.
+This is the case if we let $k = n_{\mtx{\Omega}}/2$ and $p = n_{\mtx{\Omega}}/4$.
+Then, using the triangle inequality for $\mathbb{E}^{p}[\cdot]$ and inserting \refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud} we obtain
+    \begin{equation}
+        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right]
+        \leq \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
+    \end{equation}
+    where $c = 1 + 9 \sqrt{2} e^4 (9 + \frac{1}{2})$.
 
@@
     \begin{align}
-        \mathbb{E}^{\sfrac{k}{2}}\left[ \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \right] 
-        &\stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F \right] ~\mathrm{d}t \notag \\
+        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \right] 
+        &\stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F \right] ~\mathrm{d}t \notag \\
         %&\leq \mathbb{E}^{\sfrac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
@@
     \end{align}
-    From Markov's inequality follows with probability at least $1 - \gamma^{-\sfrac{k}{2}}$
-    \begin{equation}
-        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t,
-    \end{equation}
-    from which follows the first statement by remembering $k = \frac{n_{\mtx{\Omega}}}{2}$ and redefining $c \to \sqrt{2}c$.
+    From Markov's inequality follows with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
+    \begin{equation}
+        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t,
+    \end{equation}
+    from which follows the first statement.
 

### 2024-06-24 paper/main.tex
rewrite spectral norm bound to squared version

@@
     \item Re-run numerical experiments with simplified and improved implementation (test pinvh against lstsq)
+    \item Check that everywhere correct/consistent orders $\mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1})$)
+    \item 
 \end{itemize}
@@
     \end{equation}
-    from which follows the second part of the theorem.
+    from which follows the second part of the theorem when inserting $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$.
     %Consequently, with probability $1 - \delta$ we have
@@
     \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2 \right]
-        \leq  \sqrt{k + p} \left( \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{A} \rVert _F \right).
+        \mathbb{E}^{p}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2^2 \right]
+        \leq  (k + 2p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
     \end{equation}
@@
 
+\begin{remark}
+    \todo{Add remark (need to maybe convert back appendix lemma to $p/2$)}
+    \begin{equation}
+        \mathbb{E}^{p}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2 \right] = \sqrt{\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2^2 \right]} \leq \sqrt{k + p} \cdot \left(\lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F\right)
+    \end{equation}
+\end{remark}
+
 \begin{proof}
-    With the singular value decomposition $\mtx{A} = \mtx{U} \mtx{\Sigma} \mtx{V}^{\top}$ and the unitary invariance of the spectral norm, it can be seen that $\lVert \mtx{A} \mtx{\omega} \rVert _2 = \lVert \mtx{\Sigma} \widetilde{\mtx{\omega}} \rVert _2$ where $\widetilde{\mtx{\omega}}$ is again a standard Gaussian random vector.
-Hence, it is enough to bound $\mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\omega} \rVert _2 \right]$ for a diagonal matrix $\mtx{\Sigma} = \operatorname{diag}(\sigma_1, \dots, \sigma_m)$ with $\sigma_1 \geq \dots \geq \sigma_m \geq 0$.
+    With the singular value decomposition $\mtx{A} = \mtx{U} \mtx{\Sigma} \mtx{V}^{\top}$ and the unitary invariance of the spectral norm, it can be seen that $\lVert \mtx{A} \mtx{\omega} \rVert _2^2 = \lVert \mtx{\Sigma} \widetilde{\mtx{\omega}} \rVert _2^2$ where $\widetilde{\mtx{\omega}}$ is again a standard Gaussian random vector.
+Hence, it is enough to bound $\mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\omega} \rVert _2^2 \right]$ for a diagonal matrix $\mtx{\Sigma} = \operatorname{diag}(\sigma_1, \dots, \sigma_m)$ with $\sigma_1 \geq \dots \geq \sigma_m \geq 0$.
 
@@
     \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2 \right]
-        = \mathbb{E}\left[  \left( \lVert \mtx{\Sigma} \vct{\omega} \rVert _2^2 \right)^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}}
-        = \mathbb{E}\left[  \left( \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right)^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}}
-        = \sqrt{ \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right] }.
+        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2^2 \right]
+        = \mathbb{E}^{p}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right].
         \label{equ:matvec-spectral-norm-simplification}
@@
     \begin{align}
-        &\mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right] \notag \\
-        &= \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{j + (i-1)k}^2}_{\leq \sigma_{1 + (i-1)k}^2} \omega_{j + (i-1)k}^2 \right] && \text{(separate sum into groups)} \notag \\
-        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{($\sigma_{j + (i-1)k}^2 \leq \sigma_{1 + (i-1)k}^2$, $j=1, \dots, k$)} \notag \\
-        &\leq \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \cdot \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{(triangle inequality)} \notag \\
-        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \right)^{\sfrac{2}{p}} && \text{(\refequ{equ:singular-value-group-bound} and \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
-        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) (k + p). && \text{(\reflem{lem:gamma})}
+        \mathbb{E}^{p}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right]
+        &= \mathbb{E}^{p}\Bigg[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{j + (i-1)k}^2}_{\leq \sigma_{1 + (i-1)k}^2} \omega_{j + (i-1)k}^2 \Bigg] && \text{(separate sum into groups)} \notag \\
+        &\leq \mathbb{E}^{p}\left[ \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{($\sigma_{j + (i-1)k}^2 \leq \sigma_{1 + (i-1)k}^2$, $\forall i, j$)} \notag \\
+        &\leq \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 ~ \mathbb{E}^{p}\left[ \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{(triangle inequality)} \notag \\
+        %&\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k}{2} + p)}{\Gamma(\frac{k}{2})} \right)^{\sfrac{1}{p}} && \text{(\refequ{equ:singular-value-group-bound} and \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
+        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) (k + 2 p). && \text{(\refequ{equ:singular-value-group-bound} and \reflem{lem:gamma})}
     \end{align}
-    Thus, together with \refequ{equ:matvec-spectral-norm-simplification} and using $\sqrt{a+b} \leq \sqrt{a} + \sqrt{b}$ for $a, b \geq 0$, we obtain
-    \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2 \right]
-        \leq \sqrt{k + p} \left( \lVert \mtx{\Sigma} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{\Sigma} \rVert _F \right).
-        \label{equ:matvec-spectral-norm-moment-bound}
-    \end{equation}
 
@@
     \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2 \right]
-        \leq  \sqrt{k + p} \left( 3 \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{A} \rVert _F \right).
+        \mathbb{E}^{p}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
+        \leq  (k + 2p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
     \end{equation}
@@
 
+\begin{remark}
+    \todo{Add remark (need to maybe convert back appendix lemma to $p/2$)}
+    \begin{equation}
+        \mathbb{E}^{p}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2 \right] = \sqrt{\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]} \leq \sqrt{k + p} \cdot \left(3 \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F \right)
+    \end{equation}
+\end{remark}
+
 \begin{proof}
@@
     \begin{equation}
-        \mathbb{E}^p\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2 \right]
-        \leq \lVert \mtx{A} \rVert _2 \mathbb{E}^p\left[ \lVert \vct{\omega}_1 \rVert _2 \right] + \mathbb{E}^p\left[ \lVert \mtx{A} \vct{\omega}_2 \rVert _2 \right]
+        \mathbb{E}^p\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
+        \leq \lVert \mtx{A} \rVert _2^2 \mathbb{E}^p\left[ \lVert \vct{\omega}_1 \rVert _2^2 \right] + \mathbb{E}^p\left[ \lVert \mtx{A} \vct{\omega}_2 \rVert _2^2 \right]
     \end{equation}
@@
         \mathbb{E}^p\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2 \right]
-        \leq \sqrt{k + p} \left( 3 \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{A} \rVert _F \right).
+        \leq (k + 2p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
     \end{equation}
@@
     Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-Then its Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ with Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ for \emph{even} $n_{\mtx{\Omega}} \in \mathbb{N}$ and $\gamma \geq 1$ verifies with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
+Then its Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ with Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ for \emph{even} $n_{\mtx{\Omega}} \geq 4$ verifies with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
     \begin{equation}
@@
     \end{equation}
-    for a universal constant $c > 0$.
-In particular, if we choose $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1}) + \log(\delta^{-1}))$, then with probability at least $1-\delta$ for $\delta \in (0, 1)$ and $\varepsilon > 0$ we have $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$.
+    for any $\gamma \geq 1$ and a universal constant $c > 0$.
+In particular, if we choose $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1}))$, then with probability at least $1-\delta$ for $\delta \in (0, 1)$ and $\varepsilon \in (0, 1)$ we have $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$.
 \end{lemma}
@@
 
-    As in the proof of \cite[lemma 3]{meyer-2021-hutch-optimal}, the first term is limited by
+    As in the proof of \cite[lemma 3]{meyer-2021-hutch-optimal}, the first term -- and its moments -- are limited by
     \begin{equation}
@@
     \end{equation}
-    \todo{In particular, its moments are also bounded with this.} The $p$-th moment of the second term can be processed with standard matrix norm manipulations to
+    The $p$-th moment of the second term can be processed with standard matrix norm manipulations to
     \begin{align}
@@
         &\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{(submultiplicativity)} \notag \\
-        &\leq \sqrt{k} \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2\right] \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2  \right]  && \text{(independence $\mtx{\Omega}_1, \mtx{\Omega}_2$)} \notag \\
-        &= \sqrt{k} \mathbb{E}^{2p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2 \right]^2 \mathbb{E}^{2p}\left[  \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]^2 && \text{(definition of $\mathbb{E}^{k}[\cdot]$)}
+        &\leq \sqrt{k} \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2\right] \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2  \right]  && \text{(independence)}
         \label{equ:nystrom-proof-processed-tail}
     \end{align}
-    We can apply \reflem{lem:spectral-norm-moment} to $\mtx{\Lambda}_2^{\sfrac{1}{2}}$ to bound the first term with
-    \begin{equation}
-        \mathbb{E}^{2p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2 \right]
-        \leq \sqrt{2p + n_{\mtx{\Omega}}} \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2 + \frac{1}{\sqrt{n_{\mtx{\Omega}}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F \right).
+    We can apply \reflem{lem:spectral-norm-moment} to $\mtx{\Lambda}_2^{\sfrac{1}{2}}$ to bound
+    \begin{equation}
+        \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \right]
+        \leq (n_{\mtx{\Omega}} + 2p) \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
         \label{equ:spectral-norm-bound-applied}
@@
     \begin{align}
-        \mathbb{E}^{2p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-        &= \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{p} \right]^{\sfrac{1}{2p}}  \notag \\
-        &\leq \left( 1 + \frac{2p}{n_{\mtx{\Omega}} - k + 1 - 2p} \right)^{\sfrac{1}{2p}} \left( \frac{1}{(n_{\mtx{\Omega}} - m + 1)!}\right)^{\frac{1}{n_{\mtx{\Omega}} - m + 1}} \left( \frac{k + n_{\mtx{\Omega}}}{2}\right)^{\sfrac{1}{2}} \notag \\
-        &\leq \frac{e^2}{\sqrt{2}} \frac{\sqrt{k + n_{\mtx{\Omega}}}}{n_{\mtx{\Omega}} - k + 1}
+        \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
+        &= \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{p} \right]^{\sfrac{1}{p}}  \notag \\
+        &\leq \left( 1 + \frac{2p}{n_{\mtx{\Omega}} - k + 1 - 2p} \right)^{\sfrac{1}{p}} \left( \frac{1}{(n_{\mtx{\Omega}} - k + 1)!}\right)^{\frac{2}{n_{\mtx{\Omega}} - k + 1}} \left( \frac{n_{\mtx{\Omega}} + k}{2}\right) \notag \\
+        &\leq \frac{e^4}{2} \frac{n_{\mtx{\Omega}} + k}{(n_{\mtx{\Omega}} - k + 1)^2}.
         \label{equ:pinv-spectral-norm-bound}
     \end{align}
-    with the last steps following from the expansion of the exponential function $e^n = 1 + n + \dots + n^n/n!
-+ \dots$ for $n \in \mathbb{N}$, i.e.
-$e^n \geq 1 + n$ and $e^n \geq n^n / n!$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( 1/n!
+    The last step follows by expanding exponential function $e^n = 1 + n + \dots + n^n/n!
++ \dots$ for $n \in \mathbb{N}$ to see $e^n \geq 1 + n$ and $e^n \geq n^n / n!$ for all $n \in \mathbb{N}$, from which $(1 + n)^{\sfrac{1}{n}} \leq e$ and $\left( 1/n!
 \right)^{\sfrac{1}{n}} \leq e/n$ follow respectively.
-Hence, plugging \refequ{equ:spectral-norm-bound-applied} and  \refequ{equ:pinv-spectral-norm-bound} into \refequ{equ:nystrom-proof-processed-tail}, and using $(a + b)^2 \leq 2a^2 + 2b^2$ for $a, b > 0$ gives us
+Hence, plugging \refequ{equ:spectral-norm-bound-applied} and  \refequ{equ:pinv-spectral-norm-bound} into \refequ{equ:nystrom-proof-processed-tail} gives us
     \begin{equation}
         \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 9 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right)
-    \end{equation}
-    Here, we can identify $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 = \lambda_{k+1}$ and $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 = \Trace(\mtx{\Lambda}_2)$, and just as in the proof of \cite[lemma 3.1]{meyer-2021-hutch-optimal} we bound $\lambda_{k+1} \leq \Trace(\mtx{B})/k$ and $\Trace(\mtx{\Lambda}_2) \leq \Trace(\mtx{B})$ to get
+        \leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right)
+    \end{equation}
+    We can identify $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 = \lambda_{k+1}$ and $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 = \Trace(\mtx{\Lambda}_2)$, and just as in the proof of \cite[lemma 3.1]{meyer-2021-hutch-optimal} we bound $\lambda_{k+1} \leq \Trace(\mtx{B})/k$ and $\Trace(\mtx{\Lambda}_2) \leq \Trace(\mtx{B})$ to get
     \begin{align}
         \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        &\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( \frac{9}{k} \Trace(\mtx{B}) + \frac{1}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) \right) \notag \\
-        &= \left( \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2} \frac{9n_{\mtx{\Omega}} + k}{n_{\mtx{\Omega}}}\right) \frac{\Trace(\mtx{B})}{\sqrt{k}}
+        &\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( \frac{3}{k} \Trace(\mtx{B}) + \frac{1}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) \right) \notag \\
+        &= \frac{e^4}{2} \left( \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2} \frac{3n_{\mtx{\Omega}} + k}{n_{\mtx{\Omega}}\sqrt{k}} \right) \Trace(\mtx{B})
         \label{equ:nystrom-proof-tail-bound}
     \end{align}
-    To match the decay rate of the first term \refequ{equ:nystrom-proof-frobenius-trace}, we need to ensure that the term enclosed in parentheses is constant.
+    To match the decay rate of the first term \refequ{equ:nystrom-proof-frobenius-trace}, we need to ensure that the term enclosed in parentheses is of order $\mathcal{O}(1 / \sqrt{k})$.
 We are free to choose $p > 1$ and $k \in \mathbb{N}$ such that $n_{\mtx{\Omega}} \geq k + 2p$.
-This is the case if we let $k = n_{\mtx{\Omega}}/2$ and $p = n_{\mtx{\Omega}}/4$.
+This is the case if we let $k = n_{\mtx{\Omega}}/2$ and $p = n_{\mtx{\Omega}}/4$, which will make the term in parentheses become $9\sqrt{2}(3 + 1/2) / \sqrt{n_{\mtx{\Omega}}}$.
 Then, using the triangle inequality for $\mathbb{E}^{p}[\cdot]$ and inserting \refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud} we obtain
@@
     \end{equation}
-    where $c = 1 + 9 \sqrt{2} e^4 (9 + \frac{1}{2})$.
-
-    As in \cite{kressner-2023-randomized-lowrank}, we can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable and by the Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities}, we get
+    where $c = 1 + 9 e^4 / \sqrt{2} (3 + 1/2)$.
+
+    As in \cite{kressner-2023-randomized-lowrank}, we can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable and by the Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities}, we get for $n_{\mtx{\Omega}} \geq 4$
     \begin{align}
@@
         %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
-        &\leq \frac{c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
+        &\leq \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
     \end{align}
@@
 
-    Fixing $\gamma = e$ and inserting $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1}) + \log(\delta^{-1}))$ into \refequ{equ:nystrompp-theorem-bound} we get the second part of the theorem.
+    Fixing $\gamma = e$ and letting $n_{\mtx{\Omega}} = (c e \varepsilon)^{-2}  + 4 \log(\delta^{-1})$ gives us the second part of the theorem.
 
@@
 
+\todo{Compare to Nyström++}
+
 Finally, we can combine the bound on the Girard-Hutchinson estimator \refthm{thm:hutchinson} and the one on the Nyström approximation \reflem{lem:nystrom} to obtain our main result.
@@
     Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(2\delta^{-1})^2 + \log(2\delta^{-1}))$ with \emph{even} $n_{\mtx{\Omega}}$, then for $\delta \in (0, e^{-\sfrac{1}{2}}/2)$ and $\varepsilon > 0$ with probability at least $1 - \delta$
+If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(2\delta^{-1})^2)$ with \emph{even} $n_{\mtx{\Omega}}$, then for $\delta \in (0, e^{-\sfrac{1}{2}}/2)$ and $\varepsilon \in (0, 1)$ with probability at least $1 - \delta$
     \begin{equation}
@@
     Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, \emph{even} $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2 + \log(2\delta^{-1}))$, and smoothing parameter $\sigma > 0$ satisfies with probability at least $1 - \delta$
+Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, \emph{even} $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2)$, and smoothing parameter $\sigma > 0$ satisfies with probability at least $1 - \delta$
     \begin{equation}
@@
 
-\begin{lemma}{Quotient of Gamma functions}{gamma}
-    For all $p, k \in \mathbb{N}$ it holds
-    \begin{equation}
-        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \leq \left( \frac{k + p}{2} \right)^{\sfrac{p}{2}}.
+\begin{lemma}{Moment bound of $\chi^2$-random variable}{gamma}
+    Let $X$ follow a $\chi^2$ distribution with $k$ degrees of freedom. Then for all $p \in \mathbb{N}$ it holds
+    \begin{equation}
+        \mathbb{E}^{p}[X] \leq k + 2p
+        %\frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \leq \left( \frac{k + p}{2} \right)^{\sfrac{p}{2}}.
     \end{equation}
@@
 \begin{proof}
-    We first treat the case $k=p=1$, for which indeed
-    \begin{equation}
-        \frac{\Gamma(1)}{\Gamma(\frac{1}{2})} = \frac{1}{\sqrt{\pi}} \leq 1
-    \end{equation}
-    So we now assume $k + p \geq 3$. We can use a telescoping product and the property $\Gamma(z+1)/\Gamma(z) = z$ to bound
+    From \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical} we know
+    \begin{equation}
+        \mathbb{E}^{p}[X] = 2 \left( \frac{\Gamma(\frac{k}{2} + p)}{\Gamma(\frac{k}{2})} \right)^{\sfrac{1}{p}}.
+    \end{equation}
+    We can use a telescoping product and the property $\Gamma(z+1)/\Gamma(z) = z$ to bound
     \begin{align}
-        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} 
-        &= \frac{\Gamma(\frac{k}{2} + 1)}{\Gamma(\frac{k}{2})} \frac{\Gamma(\frac{k}{2} + 2)}{\Gamma(\frac{k}{2} + 1)} \cdots \frac{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor - 1)} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)} \notag \\
-        &= \left(\frac{k}{2}\right) \left(\frac{k}{2} + 1\right) \cdots \left( \frac{k}{2} + \bigg\lfloor \frac{p}{2} \bigg\rfloor - 1 \right) \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)} \notag \\
-        &\leq \left(\frac{k}{2} + \bigg\lfloor \frac{p}{2} \bigg\rfloor - 1\right)^{\lfloor \sfrac{p}{2} \rfloor} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)}
+        \frac{\Gamma(\frac{k}{2} + p)}{\Gamma(\frac{k}{2})} 
+        &= \frac{\Gamma(\frac{k}{2} + 1)}{\Gamma(\frac{k}{2})} \frac{\Gamma(\frac{k}{2} + 2)}{\Gamma(\frac{k}{2} + 1)} \cdots \frac{\Gamma(\frac{k}{2} + p)}{\Gamma(\frac{k}{2} + p - 1)} \notag \\
+        &= \left(\frac{k}{2}\right) \left(\frac{k}{2} + 1\right) \cdots \left( \frac{k}{2} + p - 1 \right) \notag \\
+        &\leq \left(\frac{k}{2} + p - 1\right)^{p} 
     \end{align}
-    If $p$ is even, then $\lfloor p/2 \rfloor = p/2$ and hence
-    \begin{equation}
-        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} 
-        \leq \left( \frac{k + p}{2} - 1 \right)^{\sfrac{p}{2}}
-        \leq \left( \frac{k + p}{2}\right)^{\sfrac{p}{2}}.
-    \end{equation}
-    If $p$ is odd, then $\lfloor p/2 \rfloor = (p - 1)/2$, from which follows with Gautschi's inequality \cite{kershaw-1983-extensions-gautschi}
-    \begin{equation}
-        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})}
-        \leq \left(\frac{k + p}{2} - \frac{3}{2} \right)^{ \frac{p - 1}{2}} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k + p}{2} - \frac{1}{2})}
-        \leq \left(\frac{k + p}{2}\right)^{ \frac{p - 1}{2}} \left(\frac{k + p}{2}\right)^{\sfrac{1}{2}}
-        = \left(\frac{k + p}{2}\right)^{ \sfrac{p}{2}}
+    Hence,
+    \begin{equation}
+        \mathbb{E}^{p}[X] \leq 2 \left( \frac{k}{2} + p - 1 \right) \leq k + 2p.
     \end{equation}

### 2024-06-25 paper/main.tex
add remark below lemmas

@@
 
-Just as in the constant matrix case, our goal is now to prove that the Nyström++ estimator \refequ{equ:nystrompp-trace-estimator} achieves an $\varepsilon$ relative error with just $\mathcal{O}(\varepsilon^{-1})$ random vectors to construct the estimate.
+Just as in the constant matrix case, our goal is now to prove that the Nyström++ estimator \refequ{equ:nystrompp-trace-estimator} achieves an $\varepsilon$-relative error with just $\mathcal{O}(\varepsilon^{-1})$ random vectors used to construct the estimate.
 To do so, we mimic the proof of a similar bound for the Hutch++ estimator \cite[theorem 3.1]{meyer-2021-hutch-optimal}.
@@
 
-For reasons which will become clear in the proof of \reflem{lem:nystrom}, we will need to bound $\mathbb{E}^{p}\left[\lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]$ for a diagonal matrix $\mtx{\Sigma}$ and a standard Gaussian random matrix $\mtx{\Omega}$.
-Several ways of bounding the moments of $\mtx{\Omega}$ already exist, however, most of them are either not sufficient for our purposes \cite{chen-2005-condition-numbers, edelman-1988-eigenvalues-condition, james-1964-distributions-matrix}.
-The corresponding result for the Frobenius norm has already been shown in \cite{kressner-2023-randomized-lowrank}.
-In \cite[lemma B.1]{tropp-2023-randomized-algorithms} a bound is achieved for $p = 2$ and $p = 4$.
+For reasons which will become clear in the proof of \reflem{lem:nystrom}, we will need to bound $\mathbb{E}^{p}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]$ for a constant matrix $\mtx{A}$ and a standard Gaussian random matrix $\mtx{\Omega}$.
+The corresponding result for the Frobenius norm $\mathbb{E}^{p}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _F^2 \right]$ has already been shown in \cite{kressner-2023-randomized-lowrank}.
+Several ways of separately bounding the spectral norm moments of the Wishart matrix $\mtx{\Omega} \mtx{\Omega}^{\top}$ already exist, however, most of them are not sufficient for our purposes \cite{chen-2005-condition-numbers, edelman-1988-eigenvalues-condition, james-1964-distributions-matrix}.
+In \cite[lemma B.1]{tropp-2023-randomized-algorithms} a bound is achieved for $p = 1$ and $p = 2$.
 We will now generalize this result for arbitrary $p \in \mathbb{N}$.
-We first show the bound for Gaussian random vectors and then extend it to Gaussian random matrices in \reflem{lem:spectral-norm-moment}.
+To do so, we first show the bound for Gaussian random vectors and then extend it to Gaussian random matrices in \reflem{lem:spectral-norm-moment}.
 
@@
     \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2^2 \right]
-        \leq  (k + 2p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
+        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2^2 \right]
+        \leq  (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
     \end{equation}
@@
 \begin{remark}
-    \todo{Add remark (need to maybe convert back appendix lemma to $p/2$)}
+    In particular, this result can also be used to bound
     \begin{equation}
@@
     \end{equation}
+    for any $k,p \in \mathbb{N}$.
 \end{remark}
@@
     \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2^2 \right]
-        = \mathbb{E}^{p}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right].
+        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2^2 \right]
+        = \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right].
         \label{equ:matvec-spectral-norm-simplification}
@@
     \begin{align}
-        \mathbb{E}^{p}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right]
-        &= \mathbb{E}^{p}\Bigg[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{j + (i-1)k}^2}_{\leq \sigma_{1 + (i-1)k}^2} \omega_{j + (i-1)k}^2 \Bigg] && \text{(separate sum into groups)} \notag \\
-        &\leq \mathbb{E}^{p}\left[ \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{($\sigma_{j + (i-1)k}^2 \leq \sigma_{1 + (i-1)k}^2$, $\forall i, j$)} \notag \\
-        &\leq \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 ~ \mathbb{E}^{p}\left[ \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{(triangle inequality)} \notag \\
+        \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right]
+        &= \mathbb{E}^{\sfrac{p}{2}}\Bigg[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{j + (i-1)k}^2}_{\leq \sigma_{1 + (i-1)k}^2} \omega_{j + (i-1)k}^2 \Bigg] && \text{(separate sum into groups)} \notag \\
+        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{($\sigma_{j + (i-1)k}^2 \leq \sigma_{1 + (i-1)k}^2$, $\forall i, j$)} \notag \\
+        &\leq \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 ~ \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{(triangle inequality)} \notag \\
         %&\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k}{2} + p)}{\Gamma(\frac{k}{2})} \right)^{\sfrac{1}{p}} && \text{(\refequ{equ:singular-value-group-bound} and \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
-        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) (k + 2 p). && \text{(\refequ{equ:singular-value-group-bound} and \reflem{lem:gamma})}
+        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) (k + p). && \text{(\refequ{equ:singular-value-group-bound} and \reflem{lem:gamma})}
     \end{align}
@@
 \begin{lemma}{Spectral norm moments of non-standard Gaussian random matrix}{spectral-norm-moment}
-    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all $p \in \mathbb{N}$ 
-    \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
-        \leq  (k + 2p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
+    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all $p \in \mathbb{N}, p \geq 2$ 
+    \begin{equation}
+        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
+        \leq  (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
     \end{equation}
@@
 \begin{remark}
-    \todo{Add remark (need to maybe convert back appendix lemma to $p/2$)}
-    \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2 \right] = \sqrt{\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]} \leq \sqrt{k + p} \cdot \left(3 \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F \right)
-    \end{equation}
+    In particular, this result can also be used to bound
+    \begin{equation}
+        \mathbb{E}^{p}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2 \right] = \sqrt{\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]} \leq \sqrt{k + p} \cdot \left(\sqrt{3} \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F\right)
+    \end{equation}
+    for any $k,p \in \mathbb{N}, p \geq 2$.
 \end{remark}
@@
 \begin{proof}
-    From the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms} we know that
-    \begin{equation}
-        \mathbb{E}^p\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
-        \leq \lVert \mtx{A} \rVert _2^2 \mathbb{E}^p\left[ \lVert \vct{\omega}_1 \rVert _2^2 \right] + \mathbb{E}^p\left[ \lVert \mtx{A} \vct{\omega}_2 \rVert _2^2 \right]
+    From the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms} we know that for $p \geq 2$
+    \begin{equation}
+        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
+        \leq \lVert \mtx{A} \rVert _2^2 \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \vct{\omega}_1 \rVert _2^2 \right] + \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega}_2 \rVert _2^2 \right]
     \end{equation}
@@
 Both moments can be bound using \reflem{lem:spectral-norm-moment-vector}, one of which with $\mtx{A} = \mtx{I}_k$, to get
-    \begin{equation}
-        \mathbb{E}^p\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2 \right]
-        \leq (k + 2p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
-    \end{equation}
+    \begin{align}
+        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
+        &\leq \lVert \mtx{A} \rVert _2^2 (k + p) \left( \lVert \mtx{I}_k \rVert _2^2 + \frac{1}{k} \lVert \mtx{I}_k \rVert _F^2 \right) + (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right) \notag \\
+        &\leq (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
+    \end{align}
 
@@
         &= \mathbb{E}^{p}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )( \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger})^{\top} \rVert _F \right] && \text{($\lVert \mtx{A} \rVert _{(4)}^2 = \lVert \mtx{A} \mtx{A}^{\top} \rVert _{F}$)} \notag \\
-        &\leq \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F \right] && \text{($\lVert \mtx{A} \mtx{B} \rVert _F \leq \lVert \mtx{A} \rVert _F \lVert \mtx{B} \rVert _2$)} \notag \\
+        &\leq \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] && \text{($\lVert \mtx{A} \mtx{B} \rVert _F \leq \lVert \mtx{A} \rVert _F \lVert \mtx{B} \rVert _2$)} \notag \\
         &\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{($\lVert \mtx{A} \rVert _F \leq \sqrt{k} \lVert \mtx{A} \rVert _2$)} \notag \\
         &\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{(submultiplicativity)} \notag \\
-        &\leq \sqrt{k} \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2\right] \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2  \right]  && \text{(independence)}
+        &\leq \sqrt{k} \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2\right] \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2  \right].
+&& \text{(independence)}
         \label{equ:nystrom-proof-processed-tail}
     \end{align}
-    We can apply \reflem{lem:spectral-norm-moment} to $\mtx{\Lambda}_2^{\sfrac{1}{2}}$ to bound
+    We can apply \reflem{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$ to bound
     \begin{equation}
@@
     \end{align}
-    The last step follows by expanding exponential function $e^n = 1 + n + \dots + n^n/n!
-+ \dots$ for $n \in \mathbb{N}$ to see $e^n \geq 1 + n$ and $e^n \geq n^n / n!$ for all $n \in \mathbb{N}$, from which $(1 + n)^{\sfrac{1}{n}} \leq e$ and $\left( 1/n!
+    The last step follows from the expansion of the exponential function $e^n = 1 + n + \dots + n^n/n!
++ \dots$ for $n \in \mathbb{N}$ to see $e^n \geq 1 + n$ and $e^n \geq n^n / n!$, from which $(1 + n)^{\sfrac{1}{n}} \leq e$ and $\left( 1/n!
 \right)^{\sfrac{1}{n}} \leq e/n$ follow respectively.
-Hence, plugging \refequ{equ:spectral-norm-bound-applied} and  \refequ{equ:pinv-spectral-norm-bound} into \refequ{equ:nystrom-proof-processed-tail} gives us
+Hence, inserting \refequ{equ:spectral-norm-bound-applied} and  \refequ{equ:pinv-spectral-norm-bound} in \refequ{equ:nystrom-proof-processed-tail} gives
     \begin{equation}
         \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right)
+        \leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
     \end{equation}
@@
         &\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( \frac{3}{k} \Trace(\mtx{B}) + \frac{1}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) \right) \notag \\
-        &= \frac{e^4}{2} \left( \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2} \frac{3n_{\mtx{\Omega}} + k}{n_{\mtx{\Omega}}\sqrt{k}} \right) \Trace(\mtx{B})
+        &= \frac{e^4}{2} \left( \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2} \frac{3n_{\mtx{\Omega}} + k}{n_{\mtx{\Omega}}\sqrt{k}} \right) \Trace(\mtx{B}).
         \label{equ:nystrom-proof-tail-bound}
     \end{align}
-    To match the decay rate of the first term \refequ{equ:nystrom-proof-frobenius-trace}, we need to ensure that the term enclosed in parentheses is of order $\mathcal{O}(1 / \sqrt{k})$.
+    To match the decay rate of the first term \refequ{equ:nystrom-proof-frobenius-trace}, we need to ensure the term enclosed in parentheses is of order $\mathcal{O}(1 / \sqrt{k})$.
 We are free to choose $p > 1$ and $k \in \mathbb{N}$ such that $n_{\mtx{\Omega}} \geq k + 2p$.
@@
 
-\todo{Compare to Nyström++}
+When compared to the equivalent result for constant matrices \cite[Lemma 3.2]{persson-2022-improved-variants}, our constant factor up front is around four times larger, and our failure probability decays half a power slower.
 
@@
 \begin{lemma}{Moment bound of $\chi^2$-random variable}{gamma}
-    Let $X$ follow a $\chi^2$ distribution with $k$ degrees of freedom. Then for all $p \in \mathbb{N}$ it holds
-    \begin{equation}
-        \mathbb{E}^{p}[X] \leq k + 2p
+    Let $X$ follow a $\chi^2$ distribution with $k \in \mathbb{N}$ degrees of freedom. Then for all $p \in \mathbb{N}$ it holds
+    \begin{equation}
+        \mathbb{E}^{\sfrac{p}{2}}[X] \leq k + p
         %\frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \leq \left( \frac{k + p}{2} \right)^{\sfrac{p}{2}}.
@@
     \begin{equation}
-        \mathbb{E}^{p}[X] = 2 \left( \frac{\Gamma(\frac{k}{2} + p)}{\Gamma(\frac{k}{2})} \right)^{\sfrac{1}{p}}.
-    \end{equation}
-    We can use a telescoping product and the property $\Gamma(z+1)/\Gamma(z) = z$ to bound
+        \mathbb{E}^{\sfrac{p}{2}}[X] = 2 \left( \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \right)^{\sfrac{2}{p}}.
+    \end{equation}
+    We first treat the case $k=p=1$, for which indeed
+    \begin{equation}
+        2 \left( \frac{\Gamma(\frac{1+1}{2})}{\Gamma(\frac{1}{2})} \right)^2 = \frac{2}{\pi} \leq 2 = 1 + 1
+    \end{equation}
+    So we now assume $k + p \geq 3$. We can use a telescoping product and the property $\Gamma(z+1)/\Gamma(z) = z$ to bound
     \begin{align}
-        \frac{\Gamma(\frac{k}{2} + p)}{\Gamma(\frac{k}{2})} 
-        &= \frac{\Gamma(\frac{k}{2} + 1)}{\Gamma(\frac{k}{2})} \frac{\Gamma(\frac{k}{2} + 2)}{\Gamma(\frac{k}{2} + 1)} \cdots \frac{\Gamma(\frac{k}{2} + p)}{\Gamma(\frac{k}{2} + p - 1)} \notag \\
-        &= \left(\frac{k}{2}\right) \left(\frac{k}{2} + 1\right) \cdots \left( \frac{k}{2} + p - 1 \right) \notag \\
-        &\leq \left(\frac{k}{2} + p - 1\right)^{p} 
+        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} 
+        &= \frac{\Gamma(\frac{k}{2} + 1)}{\Gamma(\frac{k}{2})} \frac{\Gamma(\frac{k}{2} + 2)}{\Gamma(\frac{k}{2} + 1)} \cdots \frac{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor - 1)} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)} \notag \\
+        &= \left(\frac{k}{2}\right) \left(\frac{k}{2} + 1\right) \cdots \left( \frac{k}{2} + \bigg\lfloor \frac{p}{2} \bigg\rfloor - 1 \right) \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)} \notag \\
+        &\leq \left(\frac{k}{2} + \bigg\lfloor \frac{p}{2} \bigg\rfloor - 1\right)^{\lfloor \sfrac{p}{2} \rfloor} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)}
     \end{align}
+    If $p$ is even, then $\lfloor p/2 \rfloor = p/2$ and hence
+    \begin{equation}
+        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} 
+        \leq \left( \frac{k + p}{2} - 1 \right)^{\sfrac{p}{2}}
+        \leq \left( \frac{k + p}{2}\right)^{\sfrac{p}{2}}.
+    \end{equation}
+    If $p$ is odd, then $\lfloor p/2 \rfloor = (p - 1)/2$, from which follows with Gautschi's inequality \cite{kershaw-1983-extensions-gautschi}
+    \begin{equation}
+        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})}
+        \leq \left(\frac{k + p}{2} - \frac{3}{2} \right)^{ \frac{p - 1}{2}} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k + p}{2} - \frac{1}{2})}
+        \leq \left(\frac{k + p}{2}\right)^{ \frac{p - 1}{2}} \left(\frac{k + p}{2}\right)^{\sfrac{1}{2}}
+        = \left(\frac{k + p}{2}\right)^{ \sfrac{p}{2}}
+    \end{equation}
     Hence,
     \begin{equation}
-        \mathbb{E}^{p}[X] \leq 2 \left( \frac{k}{2} + p - 1 \right) \leq k + 2p.
+        \mathbb{E}^{\sfrac{p}{2}}[X] \leq 2 \left( \frac{k + p}{2} \right) \leq k + p.
     \end{equation}

### 2024-06-25 paper/main.tex
fill in some gaps

@@
 
-Compared to the constant case \cite[lemma 2.1]{meyer-2021-hutch-optimal} we need to choose the number of queries $n_{\mtx{\Psi}}$ by a factor of $\log(\delta^{-1})$ larger to achieve the same guarantee.
-
 The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Minkowski's integral inequality to extend the results to the continuous parameter case.
@@
     \end{equation}
-    from which follows the second part of the theorem when inserting $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$.
+    from which follows the second part of the theorem when $n_{\mtx{\Psi}} = (2^5 e^{\sfrac{1}{2}} \varepsilon \log(\delta^{-1}))^2$.
     %Consequently, with probability $1 - \delta$ we have
@@
 
-\todo{
-\begin{remark}
-    If additionally $\mtx{B}(t)$ is nonzero and symmetric positive semi-definite for all $t \in [a, b]$, then it follows from the proof of \refthm{thm:hutchinson} that with probability at least $1 - \delta$
-    \begin{equation}
-        \int_{a}^{b} \frac{| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) |}{\Trace(\mtx{B}(t))} ~\mathrm{d}t < \varepsilon,
-    \end{equation}
-    if $n_{\mtx{\Psi}} \geq 2^{10} \varepsilon^{-2} \log(\delta^{-1})^2 (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ with $\mu(t) = \lVert \mtx{B}(t) \rVert _2 / \Trace(\mtx{B}(t))$.
+%\todo{
+%\begin{remark}
+%    If additionally $\mtx{B}(t)$ is nonzero and symmetric positive semi-definite for all $t \in [a, b]$, then it follows from the proof of \refthm{thm:hutchinson} that with probability at least $1 - \delta$
+%    \begin{equation}
+%        \int_{a}^{b} \frac{| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) |}{\Trace(\mtx{B}(t))} ~\mathrm{d}t < \varepsilon,
+%    \end{equation}
+%    if $n_{\mtx{\Psi}} \geq 2^{10} \varepsilon^{-2} \log(\delta^{-1})^2 (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ with $\mu(t) = \lVert \mtx{B}(t) \rVert _2 / \Trace(\mtx{B}(t))$.
 Compared to the constant matrix equivalent \cite[remark 2]{cortinovis-2022-randomized-trace} we mainly observe a larger constant and an additional factor of $\log(\delta^{-1})$.
-\end{remark}
-}
+%\end{remark}
+%}
+
+Compared to the constant case \cite[lemma 2.1]{meyer-2021-hutch-optimal} we need to choose the number of queries $n_{\mtx{\Psi}}$ by a factor of $\log(\delta^{-1})$ larger to achieve the same guarantee.
 
@@
     Suppose $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ is continuous and positive semi-definite in $t \in [a, b]$.
-Let $n_{\mtx{\Omega}} \geq 6, n_{\mtx{\Omega}} - 4 \geq r \geq 2$ be two integers.
+Let $r \geq 2$ and $n_{\mtx{\Omega}} \geq r + 4$ be two integers.
 Then for all $\gamma \geq 1$ with probability at least $1 - \gamma^{-(n_{\mtx{\Omega}} - r)}$
@@
 \begin{proof}
-    Since $f$ is non-negative, $\mtx{B}(t)$ is positive semi-definite for all $t$.
-By \cite[lemma 2.1]{frangella-2023-randomized-nystrom}, $\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ is also positive semi-definite for all $t$.
+    Since $\mtx{B}(t)$ is positive semi-definite for all $t$, then by \cite[lemma 2.1]{frangella-2023-randomized-nystrom}, $\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ is also positive semi-definite for all $t$.
 Therefore, for fixed $t \in [a, b]$
@@
 
-\todo{Compare to constant matrix case: \cite[theorem 8.1]{tropp-2023-randomized-algorithms} (similar result)}
+Compared to the constant matrix case \cite[theorem 8.1]{tropp-2023-randomized-algorithms}, our bound also scales with the best rank-$r$ approximation error in the nuclear norm, albeit with a factor approximately proportional to $r$, which cannot be compensated for with large enough oversampling.
 
@@
 The proof specializes Bernstein's theorem for Chebyshev interpolation to the case of a Gaussian smoothing kernel \refequ{equ:smoothing-kernel}.
-It loosely follows the proof give in the preprint of \cite[theorem 2]{lin-2017-randomized-estimation}, but reaches slightly different conclusion.
+It loosely follows the proof give in the preprint of \cite[theorem 2]{lin-2017-randomized-estimation}, but attains a slightly different dependence on $\sigma$.
 
@@
 
-This theorem shows that for $\sigma \ll 1$, if we choose $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, the error we induce with the Chebyshev expansion is at most $\mathcal{O}(\varepsilon)$.
+This theorem shows that when $0 < \sigma \ll 1$, if we choose $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, the error we induce with the Chebyshev expansion is at most $\mathcal{O}(\varepsilon)$.
 
@@
 
-\begin{algo}{Chebyshev-Nystrom++}{nystrom-chebyshev-pp}
+\begin{algo}{Chebyshev-Nyström++}{nystrom-chebyshev-pp}
 \begin{algorithmic}[1]
@@
 
-The dominating cost stems from the $(2m + 1)(n_{\mtx{\Psi}} + n_{\mtx{\Omega}})$ multiplications of $\mtx{A}$ with a vector and the cost associated with computing the estimate in each evaluation point $t_1, \dots, t_{n_t}$ given by $\mathcal{O}(m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + (n_{\mtx{\Omega}}^3 + n_{\mtx{\Omega}} n_{\mtx{\Psi}}^2 + 1))$.
+The dominating cost stems from the $(2m + 1)(n_{\mtx{\Psi}} + n_{\mtx{\Omega}})$ matrix-vector multiplications with $\mtx{A}$ and the cost associated with computing the estimate in each evaluation point $t_1, \dots, t_{n_t}$ given by $\mathcal{O}(m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + (n_{\mtx{\Omega}}^3 + n_{\mtx{\Omega}} n_{\mtx{\Psi}}^2 + 1))$.
 
@@
 
-\todo{Summarize analysis (on one hand, half the degree $m$, but easier to interpolate because $\sigma \to \sqrt{2} \sigma$)}
+In this section, we will upper bound the output of the Chebyshev-Nyström++ method (see. \refalg{alg:nystrom-chebyshev-pp}). To this extent, we will first analyze the non-negative Chebyshev expansion.
 
@@
 
+On one hand, in the non-negative expansion, the underlying computes Chebyshev expansion -- which is shown in \reflem{lem:chebyshev-error} to exponentially decay with its degree -- is only computed to degree $m/2$, whereas the standard expansion is of degree $m$.
+On the other hand, we approximate $\sqrt{g_{\sigma}}$ instead of $g_{\sigma}$, which is less \enquote{pointy} and therefore better approximated by polynomials.
+Overall, we will need to choose the degree $m$ of the expansion approximately a factor of $\sqrt{2}$ larger to match the error of the standard expansion.
+
 \begin{theorem}{Chebyshev-Nyström++ method}{chebyshev-nystrom}
     Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, \emph{even} $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2)$, and smoothing parameter $\sigma > 0$ satisfies with probability at least $1 - \delta$
+Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, \emph{even} $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2)$, and smoothing parameter $0 < \sigma \ll 1$ satisfies with probability at least $1 - \delta$
     \begin{equation}
@@
     \end{equation}
-    for all $0 < \varepsilon \leq 1$ and $\delta \in (0, e^{-\sfrac{1}{2}}/2)$.
+    for all $\varepsilon \in (0, 1)$ and $\delta \in (0, e^{-\sfrac{1}{2}}/2)$.
 \end{theorem}
@@
         &\leq (1 + \varepsilon) \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \varepsilon \int_{-1}^{1} \left| \phi_{\sigma}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
-        &\leq (1 + \varepsilon) 2 \underline{E}_{\sigma, m} + \varepsilon. && \text{(\reflem{lem:non-negative-chebyshev-error} and normalization)} \notag \\
+        &\leq (1 + \varepsilon) 2 \underline{E}_{\sigma, m} + \varepsilon. && \text{(\refthm{thm:chebyshev-error} and normalization)} \notag \\
     \end{align}
-    From \reflem{lem:non-negative-chebyshev-error} we see that choosing $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$ ensures $\underline{E}_{\sigma, m} \leq \varepsilon$, and thus, since $\mathcal{\varepsilon} \leq 1$,
+    From \reflem{lem:non-negative-chebyshev-error} we see that if $\sigma \ll 1$, choosing $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$ ensures $\underline{E}_{\sigma, m} \leq \varepsilon$, and thus, since $\mathcal{\varepsilon} \leq 1$,
     \begin{equation}
@@
 
-\todo{Maybe specialize \refthm{thm:nystrom} to spectral densities?}
-
-We now discuss what value to choose for $n_{\mtx{\Omega}}$ in the Nyström trace estimator introduced in \refsec{subsec:nystrom}.
-In order to guarantee a good approximation error based on \refthm{thm:nystrom}, we analyze the numerical rank of the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
+%\todo{Maybe specialize \refthm{thm:nystrom} to spectral densities?}
+
+%We now discuss what value to choose for $n_{\mtx{\Omega}}$ in the Nyström trace estimator introduced in \refsec{subsec:nystrom}.
+In order to guarantee an even better approximation error based on \refthm{thm:nystrom}, we analyze the numerical rank of the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
+
+In practice, the approximation error is often significantly better than guaranteed in \refthm{thm:chebyshev-nystrom}, particularly when $n_{\mtx{\Omega}}$ is chosen rather large.
+This is because the singular values of the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ decay so quickly that already the Nyström approximation alone is highly accurate (cf.
+\refthm{thm:nystrom}).
+To establish a relation between the singular value decay of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ and what choice of $n_{\mtx{\Omega}}$ is needed to achieve a better approximation than what we show in \refthm{thm:chebyshev-nystrom}, we study the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
 
@@
 
-\todo{Discuss how this informs choice of $n_{\mtx{\Omega}}$.}
+Hence, choosing \todo{how discuss how to choose n Omega}
 

### 2024-06-26 paper/main.tex
polish details before meeting

@@
 A related technique is dynamic trace estimation for sequences $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ of matrices which satisfy $\lVert \mtx{B}(t_i) \rVert _F \leq 1$ and which change dynamically, i.e.
-$\lVert \mtx{B}(t_{i+1}) - \mtx{B}(t_i) \rVert _F \leq \alpha < 1$ for the Frobenius norm $\lVert \cdot \rVert _F$ \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace}.
+$\lVert \mtx{B}(t_{i+1}) - \mtx{B}(t_i) \rVert _F \leq \alpha < 1$ for the Frobenius norm $\lVert \cdot \rVert _F$ and a small $\alpha$ \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace}.
 However, in our setting we often cannot guarantee either of the two conditions.
@@
     Now we make the transition to the continuous.
-Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting:
+Following \cite{kressner-2023-randomized-lowrank}, we can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable.
+Therefore, Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting:
     \begin{equation}
@@
 
-The downside of the Nyström estimator is that it will -- in general -- not work well enough on matrices whose singular values do not decay, for example $\mtx{B}(t) \equiv \mtx{I}_n$.
+The downside of the Nyström estimator is that it will -- in general -- not work well enough on matrices whose singular values do not decay quickly, for example $\mtx{B}(t) \equiv \mtx{I}_n$.
 
@@
 To do so, we mimic the proof of a similar bound for the Hutch++ estimator \cite[theorem 3.1]{meyer-2021-hutch-optimal}.
-Correspondingly, we will need to show that with $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2})$ the Nyström approximation verifies $\int \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)\rVert _F~\mathrm{d}t \leq \varepsilon \int \Trace(\mtx{B}(t))~\mathrm{d}t$, regardless of the structure of the matrix $\mtx{B}(t)$.
+Correspondingly, we will need to show that with $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2})$ random vectors, the Nyström approximation verifies $\int \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)\rVert _F~\mathrm{d}t \leq \varepsilon \int \Trace(\mtx{B}(t))~\mathrm{d}t$, regardless of the structure of the matrix $\mtx{B}(t)$.
 Just as in the corresponding proof for the Girard-Hutchinson estimator \refthm{thm:hutchinson}, we again bound the higher order moments of the approximation error for fixed $t$ and then use Minkowsi's integral inequality to carry over the results to the continuous case.
@@
 
-For reasons which will become clear in the proof of \reflem{lem:nystrom}, we will need to bound $\mathbb{E}^{p}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]$ for a constant matrix $\mtx{A}$ and a standard Gaussian random matrix $\mtx{\Omega}$.
-The corresponding result for the Frobenius norm $\mathbb{E}^{p}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _F^2 \right]$ has already been shown in \cite{kressner-2023-randomized-lowrank}.
+For reasons which will become clear in the proof of \reflem{lem:nystrom}, we will need to bound $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]$ for a constant matrix $\mtx{A}$ and a standard Gaussian random matrix $\mtx{\Omega}$.
+The corresponding result for the Frobenius norm $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _F^2 \right]$ has already been shown in \cite{kressner-2023-randomized-lowrank}.
 Several ways of separately bounding the spectral norm moments of the Wishart matrix $\mtx{\Omega} \mtx{\Omega}^{\top}$ already exist, however, most of them are not sufficient for our purposes \cite{chen-2005-condition-numbers, edelman-1988-eigenvalues-condition, james-1964-distributions-matrix}.
-In \cite[lemma B.1]{tropp-2023-randomized-algorithms} a bound is achieved for $p = 1$ and $p = 2$.
+In \cite[lemma B.1]{tropp-2023-randomized-algorithms} a bound is achieved for $p = 2$ and $p = 4$.
 We will now generalize this result for arbitrary $p \in \mathbb{N}$.
@@
 \begin{lemma}{Spectral norm moments of non-standard Gaussian random vector}{spectral-norm-moment-vector}
-    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random vector $\vct{\omega} \in \mathbb{R}^{m}$ satisfy for all $k,p \in \mathbb{N}$ 
+    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random vector $\vct{\omega} \in \mathbb{R}^{m}$ satisfy for all $k,p \in \mathbb{N}, p \geq 2$ 
     \begin{equation}
@@
     \end{equation}
-    for any $k,p \in \mathbb{N}$.
+    for any $k,p \in \mathbb{N}, p \geq 2$.
 \end{remark}
@@
     With the singular value decomposition $\mtx{A} = \mtx{U} \mtx{\Sigma} \mtx{V}^{\top}$ and the unitary invariance of the spectral norm, it can be seen that $\lVert \mtx{A} \mtx{\omega} \rVert _2^2 = \lVert \mtx{\Sigma} \widetilde{\mtx{\omega}} \rVert _2^2$ where $\widetilde{\mtx{\omega}}$ is again a standard Gaussian random vector.
-Hence, it is enough to bound $\mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\omega} \rVert _2^2 \right]$ for a diagonal matrix $\mtx{\Sigma} = \operatorname{diag}(\sigma_1, \dots, \sigma_m)$ with $\sigma_1 \geq \dots \geq \sigma_m \geq 0$.
+Hence, it is enough to bound $\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{\Sigma} \mtx{\omega} \rVert _2^2 \right]$ for a diagonal matrix $\mtx{\Sigma} = \operatorname{diag}(\sigma_1, \dots, \sigma_m)$ with $\sigma_1 \geq \dots \geq \sigma_m \geq 0$.
 
@@
         %&\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k}{2} + p)}{\Gamma(\frac{k}{2})} \right)^{\sfrac{1}{p}} && \text{(\refequ{equ:singular-value-group-bound} and \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
-        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) (k + p). && \text{(\refequ{equ:singular-value-group-bound} and \reflem{lem:gamma})}
+        &\leq (k + p) \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 && \text{(\reflem{lem:gamma})} \notag \\
+        &\leq (k + p) \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right). && \text{(using \refequ{equ:singular-value-group-bound})}
     \end{align}
@@
     \end{equation}
-    for any $k,p \in \mathbb{N}, p \geq 2$.
+    for any $k,p \in \mathbb{N}, p \geq 2$. Indeed, for $p=2$ this bound is slightly less tight than its equivalent in \cite[lemma B.1]{tropp-2023-randomized-algorithms}.
 \end{remark}
@@
 
-Comparing \refthm{thm:nystrom-pp} with its equivalent for constant matrices \cite[theorem 3.4]{persson-2022-improved-variants}, we notice a less favorable scaling in terms of $\log(2\delta^{-1})$.
+Comparing \refthm{thm:nystrom-pp} with its equivalent for constant matrices \cite[theorem 3.4]{persson-2022-improved-variants}, we notice a less favorable scaling of the required number of random vectors $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$ in terms of $\log(2\delta^{-1})$.
 
@@
 Among others, \cite{lin-2016-approximating-spectral} proposes $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - \lambda), \lambda \in [\lambda_{\text{min}}, \lambda_{\text{max}}]\}$, \cite{chen-2021-analysis-stochastic} uses $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [-1, 1] \}$ where $\Theta$ is the Heaviside step function, and \cite{braverman-2022-sublinear-time} uses $\mathcal{S} = \{f : |f(x) - f(y)| \leq |x - y| \}$ for which the error \refequ{equ:error-metric} is equivalent to the Wasserstein-1 distance between $\phi$ and $\phi_{\sigma}$.
-In this last metric, if we assume the spectrum to be somewhat uniform and $\sigma$ to be rather small, a simple calculation allows us to roughly estimate the smoothing error \refequ{equ:error-metric} to be $\sigma$.
+In this last metric, provided we assume the spectrum to be somewhat uniform and $\sigma$ to be rather small, a simple calculation allows us to roughly estimate the smoothing error \refequ{equ:error-metric} to be $\sigma$.
 Consequently, if we can only allow an approximation to deviate by at most a factor of $\varepsilon > 0$ from the original spectral density, we should choose a smoothing parameter $\sigma \approx \varepsilon$.
@@
     \end{equation}
-    for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, $t \in \mathbb{R}$, and $s \in [-1, 1]$.
+    for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and $s \in [-1, 1]$.
 \end{lemma}
@@
 
-\Reflem{lem:chebyshev-error} now also allows us to bound the error we make with the expanded spectral density $\phi_{\sigma}^{(m)}$ \refequ{equ:expanded-spectral-density}.
+\Reflem{lem:chebyshev-error} shows that when $0 < \sigma \ll 1$ and if we choose $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, the error we induce with the Chebyshev expansion is at most $\mathcal{O}(\varepsilon)$.
+It now also allows us to bound the error we make when expanding the spectral density.
 
@@
     \end{equation}
-    for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and $t \in \mathbb{R}$.
+    for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and the error $E_{\sigma, m}$ from \reflem{lem:chebyshev-error}.%, and $t \in \mathbb{R}$.
 %The same result holds for the expansion of a non-negative Chebyshev expansion of \emph{even} degree $m \in \mathbb{N}$ but with $\underline{E}_{\sigma, m}$.
@@
 
-This theorem shows that when $0 < \sigma \ll 1$, if we choose $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, the error we induce with the Chebyshev expansion is at most $\mathcal{O}(\varepsilon)$.
-
 \subsection{Chebyshev-Nyström++ method for spectral density approximation}
@@
 To still apply the bound from \refthm{thm:nystrom-pp} in this scenario, we will have to ensure that the Chebyshev expansion is non-negative.
-We can think of many ways in which this can be achieved: by using Jackson damping \cite{jackson-1912-approximation-trigonometric,braverman-2022-sublinear-time}, albeit at a significantly worse rate of convergence; by shifting up the expansion $g_{\sigma}^{(m)} + \rho$ with a large enough $\rho$, at the loss of most of the low-rank structure of $g_{\sigma}(t \mtx{I}_n - \mtx{A})$; or by directly minimizing the approximation error over all non-negative Chebyshev polynomials \cite{fejer-1916-uber-trigonometrische}, which realistically only works for small $m$.
-In the end, we propose the following method to produce a non-negative Chebyshev expansion: First, expand the square root of the smoothing kernel in the Chebyshev polynomials up to degree $m/2$
+We can think of many ways in which this can be achieved: by using Jackson damping \cite{jackson-1912-approximation-trigonometric,braverman-2022-sublinear-time}, albeit at a significantly worse rate of convergence; by shifting up the expansion $g_{\sigma}^{(m)} + \rho$ with a large enough $\rho > 0$, at the loss of most of the low-rank structure of $g_{\sigma}(t \mtx{I}_n - \mtx{A})$; or by directly minimizing the approximation error over all non-negative Chebyshev polynomials \cite{fejer-1916-uber-trigonometrische}, which realistically only works for small $m$.
+In the end, we propose the following method to produce a non-negative Chebyshev expansion: First, expand the square root of the smoothing kernel in a basis of Chebyshev polynomials up to degree $m/2$
 \begin{equation}
@@
 \begin{equation}
-    \widetilde{\phi}_{\sigma}^{(m)}(t) = \frac{1}{n} \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})).
+    \widetilde{\phi}_{\sigma}^{(m)}(t) = \frac{1}{n} \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}))
     \label{equ:chebyshev-nystrom-formula}
 \end{equation}
+which is the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$ \refequ{equ:nystrompp-trace-estimator} applied to the non-negative Chebyshev expansion $\underline{g}_{\sigma}^{(m)}$ \refequ{equ:non-negative-chebyshev-expansion}.
 
@@
 \end{equation}
-Thanks to the affine form of the Chebyshev expansion, we can rewrite the first matrix as
+Thanks to the affine form of the Chebyshev expansion, the first parenthesized term in the trace is
 \begin{equation}
@@
 \end{equation}
-for which we first precompute the $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ for $l=0, \dots, m$ using the recurrence \refequ{equ:chebyshev-recurrence} and then inexpensively sum them up with the corresponding coefficients $\mu_l(t)$ for all $t$ in which we want to evaluate the approximation.
+for which we can first precompute the $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ for $l=0, \dots, m$ using the recurrence \refequ{equ:chebyshev-recurrence} and then inexpensively sum them up with the corresponding coefficients $\mu_l(t)$ for all $t$ in which we want to evaluate the approximation.
 Thanks to the fast and exact squaring of Chebyshev expansions (cf.
-\refalg{alg:chebyshev-squaring}), the second part of \refequ{equ:cyclic-property} can be approximated in an equivalent way.
-Unlike expanding the squared smoothing kernel $g_{\sigma}^2$ separately \cite{lin-2017-randomized-estimation}, this approach is consistent, which we have observed to make a noticeable difference in terms of accuracy and does not hinder provability.
+\refalg{alg:chebyshev-squaring}), the second parenthesized term in \refequ{equ:cyclic-property} can be approximated in an equivalent way.
+Unlike expanding the squared smoothing kernel $g_{\sigma}^2$ separately \cite{lin-2017-randomized-estimation}, this approach is consistent, which we have observed to make a noticeable difference in terms of accuracy, and which enables provability.
 Similar observations for the case when $\mtx{\Psi}$ is nonzero lead to the following algorithm.
 
-\begin{algo}{Chebyshev-Nyström++}{nystrom-chebyshev-pp}
+\begin{algo}{Chebyshev-Nyström++ method}{nystrom-chebyshev-pp}
 \begin{algorithmic}[1]
@@
 regions where the spectral density nearly vanishes.
-For small smoothing parameters $\sigma$, the rapid decay of the Gaussian $g_{\sigma}$ will cause $\mtx{K}(t_i) \approx g_{\sigma}(t_i\mtx{I}_n - \mtx{A})$ to be close to the zero matrix for parameter values $t_i$ in such regions.
+For small smoothing parameters $\sigma$, the rapid decay of the Gaussian $g_{\sigma}$ will cause $\mtx{K}(t_i) \approx g_{\sigma}(t_i\mtx{I}_n - \mtx{A})$ to be close to the zero matrix for parameter values $t_i$ located in such regions.
 Hence, on line \reflin{lin:4-nystromchebyshev-nystrom-pp} we would compute the pseudoinverse of an almost zero matrix, which can be extremely unstable.
@@
     \end{equation}
-    for all \emph{even} degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, $t \in \mathbb{R}$, and $s \in [-1, 1]$.
+    for all \emph{even} degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, $s \in [-1, 1]$, and with $E_{\sigma, m}$ as defined in \reflem{lem:chebyshev-error}.
 \end{lemma}
@@
         &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \int_{-1}^{1} \left| \phi^{(m)}_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
-        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \varepsilon \int_{-1}^{1} \left| \phi^{(m)}_{\sigma}(t) \right|~\mathrm{d}t && \text{(\refthm{thm:nystrom-pp} w.p.
+        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \widetilde{\varepsilon} \int_{-1}^{1} \left| \phi^{(m)}_{\sigma}(t) \right|~\mathrm{d}t && \text{(\refthm{thm:nystrom-pp} w.p.
 $\geq 1 - \delta$)} \notag \\
-        &\leq (1 + \varepsilon) \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \varepsilon \int_{-1}^{1} \left| \phi_{\sigma}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
-        &\leq (1 + \varepsilon) 2 \underline{E}_{\sigma, m} + \varepsilon. && \text{(\refthm{thm:chebyshev-error} and normalization)} \notag \\
+        &\leq (1 + \widetilde{\varepsilon}) \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \varepsilon \int_{-1}^{1} \left| \phi_{\sigma}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
+        &\leq (1 + \widetilde{\varepsilon}) 2 \underline{E}_{\sigma, m} + \widetilde{\varepsilon}. && \text{(\refthm{thm:chebyshev-error} and normalization)} \notag \\
     \end{align}
-    From \reflem{lem:non-negative-chebyshev-error} we see that if $\sigma \ll 1$, choosing $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$ ensures $\underline{E}_{\sigma, m} \leq \varepsilon$, and thus, since $\mathcal{\varepsilon} \leq 1$,
-    \begin{equation}
-        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq (1+\varepsilon) 2 \varepsilon + \varepsilon \leq 5 \varepsilon.
-    \end{equation}
+    From \reflem{lem:non-negative-chebyshev-error} we see that if $0 < \sigma \ll 1$, choosing $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$ ensures $\underline{E}_{\sigma, m} \leq \widetilde{\varepsilon}$, and thus, since $\widetilde{\varepsilon} \leq 1$,
+    \begin{equation}
+        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq (1+\widetilde{\varepsilon}) 2 \widetilde{\varepsilon} + \widetilde{\varepsilon} \leq 5 \widetilde{\varepsilon}
+    \end{equation}
+    from which the result follows by identifying $\varepsilon = 5 \widetilde{\varepsilon}$.
 \end{proof}
@@
 
-If we additionally assume the eigenvalues of the matrix $\mtx{A}$ to be evenly distributed within $[a, b]$, that is, in any subinterval of fixed length in $[a, b]$ we can expect to find roughly the same number of eigenvalues, then we can estimate the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ to be
-\begin{equation}
-    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A})) \lessapprox \frac{2 n}{b - a} \cdot d_{\varepsilon, \sigma}.
+If we additionally assume the eigenvalues of the matrix $\mtx{A}$ to be evenly distributed within $[-1, 1]$, that is, in any subinterval of fixed length in $[a, b]$, $-1 \leq a < b \leq 1$, we can expect to find roughly the same number of eigenvalues, then we can estimate the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ to be
+\begin{equation}
+    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A})) \lessapprox n d_{\varepsilon, \sigma} \approx n \sigma.
     \label{equ:gaussian-kernel-numerical-rank-uniform}
 \end{equation}
-
-Hence, choosing \todo{how discuss how to choose n Omega}
+Hence, choosing $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma)$ will approximations which are significantly better than \refthm{thm:chebyshev-nystrom}.
 
@@
 
+\todo{Explain experimental setup.}
+
 \begin{figure}[ht]
@@
     \input{plots/convergence.pgf}
-    \caption{For increasing values of $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}$ but fixed $m$ we plot the $L^1$ relative approximation error for the model problem with $\sigma=0.005$.}
+    \caption{For increasing values of $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}$ but fixed $m$ we plot the $L^1$-approximation error for $\sigma=0.005$ on the model problem with $n_c = 1$.}
     \label{fig:convergence}
@@
 
-\todo{[Comparison plots with other estimators?]}
-
+\todo{Interpret results.}
+
+\todo{Comparison plots with other estimators (Krylov-Aware)?}
+
+\clearpage
 \section{Discussion}

### 2024-06-27 paper/main.tex
reformat and add plots

@@
 \begin{itemize}
-    \item Change citation style to \textbf{[1]}, \textbf{[2]}, \dots
-    \item Re-run numerical experiments with simplified and improved implementation (test pinvh against lstsq)
     \item Check that everywhere correct/consistent orders $\mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1})$)
@@
 For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature rule with the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}.
-Further, because $\delta$ is hard to approximate and in many applications, a rough estimate of the spectral density is already acceptable, a standard approach is to instead approximate the smooth spectral density
+Further, because $\delta$ is hard to approximate and in many applications, a rough estimate of the spectral density is already acceptable, a standard approach is to instead approximate the smoothed spectral density
 \begin{equation}
     \phi_{\sigma}(t) = \frac{1}{n} \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i) = \frac{1}{n} \Trace(g_{\sigma}(t \mtx{I}_n - \mtx{A}))
+    \label{equ:smoothed-spectral-density}
 \end{equation}
@@
 
-\paragraph{Contribution} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices.
-In particular, we reuse the same randomization for each value of the parameter $t$ in the estimators, which makes them scale favorably with the number of parameter evaluations.
-Subsequently, we propose a simple, fast, robust, and provable procedure in which these estimators are employed for approximating the smooth spectral density $\phi_{\sigma}$ of a matrix.
+\paragraph{New contributions} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}.
+Interestingly, we can reuse the same randomization for computing the estimate at each value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations.
+We propose certain modifications to the algorithms from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}.
+In particular, use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials, and realize that an efficient implementation of all of their methods can be summarized in a single algorithm.
+Based on the analysis of the randomized trace estimators for parameter-dependent matrices we provide bounds on the approximation error for our method.
 
@@
 
-\section{Analysis of trace estimators for parameter-dependent matrices}
+\section{Analysis of randomized trace estimators applied to parameter-dependent matrices}
 \label{sec:analysis}
@@
 
-\paragraph{Girard-Hutchinson estimator} We can estimate the trace with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ stochastically independent standard Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ to form
+\paragraph{Girard-Hutchinson estimator} We can approximate the trace with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ stochastically independent standard Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ to form
 \begin{equation}
@@
         \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
-        &\stackrel{\text{\cite{boucheron-2013-basic-inequalities}}}{\leq} k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k} \notag \\
+        &\leq k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k} \notag \\
         &= k! 2^{4 k} \lVert \mtx{B} \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _2^{2 k}.
@@
         \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t  \right]
-        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}^{2 k}\left[ r(\mtx{B}(t), \vct{\psi}) \right]~\mathrm{d}t
+        \leq \int_{a}^{b} \mathbb{E}^{2 k}\left[ r(\mtx{B}(t), \vct{\psi}) \right]~\mathrm{d}t
         \leq 2^4 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
@@
     \end{equation}
-    from which follows the second part of the theorem when $n_{\mtx{\Psi}} = (2^5 e^{\sfrac{1}{2}} \varepsilon \log(\delta^{-1}))^2$.
+    from which follows the second part of the theorem when $n_{\mtx{\Psi}} = (2^5 e^{\sfrac{1}{2}} \varepsilon^{-1} \log(\delta^{-1}))^2$.
     %Consequently, with probability $1 - \delta$ we have
@@
 \begin{lemma}{Spectral norm moments of non-standard Gaussian random vector}{spectral-norm-moment-vector}
-    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random vector $\vct{\omega} \in \mathbb{R}^{m}$ satisfy for all $k,p \in \mathbb{N}, p \geq 2$ 
+    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random vector $\vct{\omega} \in \mathbb{R}^{m}$ satisfy for all $k,p \in \mathbb{N}$ 
     \begin{equation}
@@
     \end{equation}
-    for any $k,p \in \mathbb{N}, p \geq 2$.
+    for any $k,p \in \mathbb{N}$.
 \end{remark}
@@
 \begin{lemma}{Spectral norm moments of non-standard Gaussian random matrix}{spectral-norm-moment}
-    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all $p \in \mathbb{N}, p \geq 2$ 
+    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all $p \in \mathbb{N}$ 
     \begin{equation}
@@
     \end{equation}
-    for any $k,p \in \mathbb{N}, p \geq 2$. Indeed, for $p=2$ this bound is slightly less tight than its equivalent in \cite[lemma B.1]{tropp-2023-randomized-algorithms}.
+    for any $k,p \in \mathbb{N}$. Indeed, for $p=2$ this bound is slightly less tight than its equivalent in \cite[lemma B.1]{tropp-2023-randomized-algorithms}.
 \end{remark}
@@
 \begin{proof}
-    From the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms} we know that for $p \geq 2$
+    From the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms} we know that
     \begin{equation}
@@
         \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        &= \mathbb{E}^{p}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )( \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger})^{\top} \rVert _F \right] && \text{($\lVert \mtx{A} \rVert _{(4)}^2 = \lVert \mtx{A} \mtx{A}^{\top} \rVert _{F}$)} \notag \\
-        &\leq \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] && \text{($\lVert \mtx{A} \mtx{B} \rVert _F \leq \lVert \mtx{A} \rVert _F \lVert \mtx{B} \rVert _2$)} \notag \\
-        &\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{($\lVert \mtx{A} \rVert _F \leq \sqrt{k} \lVert \mtx{A} \rVert _2$)} \notag \\
+        %&= \mathbb{E}^{p}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )( \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger})^{\top} \rVert _F \right] && \text{($\lVert \mtx{A} \rVert _{(4)}^2 = \lVert \mtx{A} \mtx{A}^{\top} \rVert _{F}$)} \notag \\
+        %&\leq \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] && \text{($\lVert \mtx{A} \mtx{B} \rVert _F \leq \lVert \mtx{A} \rVert _F \lVert \mtx{B} \rVert _2$)} \notag \\
+        &\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{($\lVert \mtx{A} \rVert _{(4)}^2 \leq \sqrt{k} \lVert \mtx{A} \rVert _2^2$)} \notag \\
         &\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{(submultiplicativity)} \notag \\
@@
     \end{align}
+    To match the decay rate of the moments of the first term in \refequ{equ:nystrom-proof-persson-bonud}, we will need to ensure that this term also is of order $\mathcal{O}(\Trace{\mtx{B}} / \sqrt{k})$.
+To this purpose, and to simplify the notation, we choose $k = n_{\mtx{\Omega}}/2$ and $p = n_{\mtx{\Omega}}/4$ for even $n_{\mtx{\Omega}}$ to ensure $k \in \mathbb{N}$.
 We can apply \reflem{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$ to bound
     \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \right]
-        \leq (n_{\mtx{\Omega}} + 2p) \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
+        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \right]
+        \leq \frac{3 n_{\mtx{\Omega}}}{2} \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
         \label{equ:spectral-norm-bound-applied}
     \end{equation}
-    The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms} when $n_{\mtx{\Omega}} \geq k + 2p$
-    \begin{align}
-        \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
-        &= \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{p} \right]^{\sfrac{1}{p}}  \notag \\
-        &\leq \left( 1 + \frac{2p}{n_{\mtx{\Omega}} - k + 1 - 2p} \right)^{\sfrac{1}{p}} \left( \frac{1}{(n_{\mtx{\Omega}} - k + 1)!}\right)^{\frac{2}{n_{\mtx{\Omega}} - k + 1}} \left( \frac{n_{\mtx{\Omega}} + k}{2}\right) \notag \\
-        &\leq \frac{e^4}{2} \frac{n_{\mtx{\Omega}} + k}{(n_{\mtx{\Omega}} - k + 1)^2}.
+    The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
+    \begin{equation}
+        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
+        = \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{n_{\mtx{\Omega}}}{4}} \right]^{\sfrac{4}{n_{\mtx{\Omega}}}}%  \notag \\
+        \leq \left( 1 + \frac{n_{\mtx{\Omega}}}{2} \right)^{\sfrac{4}{n_{\mtx{\Omega}}}} \left( \frac{1}{(\frac{n_{\mtx{\Omega}}}{2} + 1)!}\right)^{\frac{2}{\frac{n_{\mtx{\Omega}}}{2} + 1}} \left( \frac{3 n_{\mtx{\Omega}}}{4}\right).
+    \end{equation}
+    From the expansion of the exponential function as a power series $e^n = 1 + n + \dots + n^n/n!
++ \dots$ for $n \in \mathbb{N}$ we see $e^n \geq 1 + n$ and $e^n \geq n^n / n!$, from which $(1 + n)^{\sfrac{1}{n}} \leq e$ and $\left( 1/n!
+\right)^{\sfrac{1}{n}} \leq e/n$ follow respectively.
+Hence,
+    \begin{equation}
+        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
+        \leq e^2 \frac{e^2}{(\frac{n_{\mtx{\Omega}}}{2}+ 1)^2} \left( \frac{3 n_{\mtx{\Omega}}}{4}\right)
+        \leq \frac{3 e^4}{n_{\mtx{\Omega}}}.
         \label{equ:pinv-spectral-norm-bound}
-    \end{align}
-    The last step follows from the expansion of the exponential function $e^n = 1 + n + \dots + n^n/n!
-+ \dots$ for $n \in \mathbb{N}$ to see $e^n \geq 1 + n$ and $e^n \geq n^n / n!$, from which $(1 + n)^{\sfrac{1}{n}} \leq e$ and $\left( 1/n!
-\right)^{\sfrac{1}{n}} \leq e/n$ follow respectively.
-Hence, inserting \refequ{equ:spectral-norm-bound-applied} and  \refequ{equ:pinv-spectral-norm-bound} in \refequ{equ:nystrom-proof-processed-tail} gives
+    \end{equation}
+    As opposed to \cite[lemma B.3]{tropp-2023-randomized-algorithms}, our bound holds for all even $n_{\mtx{\Omega}} \in \mathbb{N}$, at the cost of a factor of $e^2$.
+Inserting \refequ{equ:spectral-norm-bound-applied} and  \refequ{equ:pinv-spectral-norm-bound} in \refequ{equ:nystrom-proof-processed-tail} gives
     \begin{equation}
         \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
-    \end{equation}
-    We can identify $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 = \lambda_{k+1}$ and $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 = \Trace(\mtx{\Lambda}_2)$, and just as in the proof of \cite[lemma 3.1]{meyer-2021-hutch-optimal} we bound $\lambda_{k+1} \leq \Trace(\mtx{B})/k$ and $\Trace(\mtx{\Lambda}_2) \leq \Trace(\mtx{B})$ to get
-    \begin{align}
+        \leq \frac{9 e^4}{2 \sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
+        %\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
+    \end{equation}
+    We can identify $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 = \lambda_{k+1}$ and $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 = \Trace(\mtx{\Lambda}_2)$, and just as in the proof of \cite[lemma 3.1]{meyer-2021-hutch-optimal} we bound $\lambda_{k+1} \leq \Trace(\mtx{B})/k$ and $\Trace(\mtx{\Lambda}_2) \leq \Trace(\mtx{B})$ -- remembering $k = n_{\mtx{\Omega}}/2$ -- to get
+    \begin{equation}
         \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        &\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( \frac{3}{k} \Trace(\mtx{B}) + \frac{1}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) \right) \notag \\
-        &= \frac{e^4}{2} \left( \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2} \frac{3n_{\mtx{\Omega}} + k}{n_{\mtx{\Omega}}\sqrt{k}} \right) \Trace(\mtx{B}).
+        \leq \frac{9 e^4}{2 \sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \left( \frac{6}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) + \frac{1}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) \right)
+        \leq \frac{9e^4}{2 \sqrt{2}} \frac{7}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
         \label{equ:nystrom-proof-tail-bound}
-    \end{align}
-    To match the decay rate of the first term \refequ{equ:nystrom-proof-frobenius-trace}, we need to ensure the term enclosed in parentheses is of order $\mathcal{O}(1 / \sqrt{k})$.
-We are free to choose $p > 1$ and $k \in \mathbb{N}$ such that $n_{\mtx{\Omega}} \geq k + 2p$.
-This is the case if we let $k = n_{\mtx{\Omega}}/2$ and $p = n_{\mtx{\Omega}}/4$, which will make the term in parentheses become $9\sqrt{2}(3 + 1/2) / \sqrt{n_{\mtx{\Omega}}}$.
-Then, using the triangle inequality for $\mathbb{E}^{p}[\cdot]$ and inserting \refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud} we obtain
+        %\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
+    \end{equation}
+    Inserting \refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud} and using the triangle inequality for $\mathbb{E}^{p}[\cdot]$ we obtain
     \begin{equation}
         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right]
+        \leq \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[ \lVert \mtx{\Lambda}_2 \rVert _F \right] + \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
         \leq \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
     \end{equation}
-    where $c = 1 + 9 e^4 / \sqrt{2} (3 + 1/2)$.
+    where $c = 1 + 63 e^4 / 2 \sqrt{2}$.
 
@@
         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \right] 
-        &\stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F \right] ~\mathrm{d}t \notag \\
+        &\leq \int_{a}^{b} \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F \right] ~\mathrm{d}t \notag \\
         %&\leq \mathbb{E}^{\sfrac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
@@
     Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(2\delta^{-1})^2)$ with \emph{even} $n_{\mtx{\Omega}}$, then for $\delta \in (0, e^{-\sfrac{1}{2}}/2)$ and $\varepsilon \in (0, 1)$ with probability at least $1 - \delta$
+If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(\delta^{-1})^2)$ with \emph{even} $n_{\mtx{\Omega}}$, then for $\delta \in (0, e^{-\sfrac{1}{2}}/2)$ and $\varepsilon \in (0, 1)$ with probability at least $1 - \delta$
     \begin{equation}
@@
 
-Comparing \refthm{thm:nystrom-pp} with its equivalent for constant matrices \cite[theorem 3.4]{persson-2022-improved-variants}, we notice a less favorable scaling of the required number of random vectors $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$ in terms of $\log(2\delta^{-1})$.
+Comparing \refthm{thm:nystrom-pp} with its equivalent for constant matrices \cite[theorem 3.4]{persson-2022-improved-variants}, we notice a less favorable scaling of the required number of random vectors $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$ in terms of $\log(\delta^{-1})$.
 
@@
 Furthermore, in most applications the exact location of every eigenvalue is not important, but more so their approximate locations relative to each other, such as eigenvalue clusters, spectral gaps, or outliers.
-In the following we will introduce a smooth version of the spectral density, which is significantly easier to approximate at the cost of losing some of the finer characteristics of the spectrum.
-
-\subsection{Smooth spectral density}
+In the following we will introduce a smoothed version of the spectral density, which is significantly easier to approximate at the cost of losing some of the finer characteristics of the spectrum.
+
+\subsection{Smoothed spectral density}
 \label{subsec:spectral-density}
 
-The smooth spectral density $\phi_{\sigma}$ is defined as
+The smoothed spectral density $\phi_{\sigma}$ is defined as
 \begin{equation}
@@
 
-It is well known that the coefficients of a Chebyshev expansion of a function can be computed which a discrete cosine transform (DCT) of the function evaluations at the Chebyshev nodes $s_i = \cos(\pi i / m)$ for $i=0,\dots,m$
+It is well known that the coefficients of a Chebyshev expansion of a function can -- in exact arithmetic -- be computed with a discrete cosine transform (DCT) of the function evaluations at the Chebyshev nodes $s_i = \cos(\pi i / m)$ for $i=0,\dots,m$
 \begin{equation}
@@
 This can be accomplished in $\mathcal{O}(m \log(m))$ operations.
-
-A highly useful property of the Chebyshev expansion -- of which we will make use twice in the following pages -- is the fact that we can (algebraically) square such an expansion in $\mathcal{O}(m \log(m))$ operations, which will be shown in the below algorithm.
+\cite[algorithm 1]{lin-2017-randomized-estimation} uses a slightly different approach, which only approximates the coefficients and is noticeably slower (cf.
+\reftab{tab:chebyshev-timing-interpolation}).
+
+A highly useful property of the Chebyshev expansion -- of which we will make use twice in the following pages -- is the fact that we can square such an expansion in $\mathcal{O}(m \log(m))$ operations \cite{baszenski-1997-fast-polynomial}, which will be used in the below algorithm.
 
@@
         \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \right| \leq \sqrt{\frac{2e}{\pi}} \frac{1}{\sigma^2} (1 + \sigma)^{-m} \equiv E_{\sigma, m}
-        \label{equ:2-chebyshev-interpolation-sup-error-kernel}
+        \label{equ:chebyshev-interpolation-sup-error-kernel}
         %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
@@
 \begin{theorem}{$L^1$-error of Chebyshev expansion of spectral density}{chebyshev-error}
-    Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
+    Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
 Then the spectal density with Chebyshev expansion $\phi_{\sigma}^{(m)}$ satisfies
@@
         \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq 2 E_{\sigma, m}
-        \label{equ:chebyshev-interpolation-sup-error-kernel}
+        \label{equ:chebyshev-interpolation-spectral-density}
         %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
@@
 
-Our goal is to apply the Nyström++ trace estimator which we analyzed in \refsec{subsec:nystrom-pp} to the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ to approximate the smooth spectral density $\phi_{\sigma}$.
+Our goal is to apply the Nyström++ trace estimator which we analyzed in \refsec{subsec:nystrom-pp} to the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ to approximate the smoothed spectral density $\phi_{\sigma}$.
 However, due to the oscillatory nature of the Chebyshev expansion, we can no longer guarantee that $g_{\sigma}^{(m)} \geq 0$, and consequently, the matrix function $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ may be (slightly) indefinite.
@@
 With a complexity of $\mathcal{O}(m \log(m))$ this procedure is computationally inexpensive and allows for very similar convergence bounds to \reflem{lem:chebyshev-error}, which we will derive in \refsec{subsubsec:chebyshev-nystrom-analysis}.
+In \reftab{tab:chebyshev-timing-interpolation}, the non-negative expansion can be seen to only be slightly slower than the normal one.
+
+\begin{table}[ht]
+    \caption{Runtime in milliseconds for computing the coefficients of the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ \refequ{equ:smoothing-kernel} with $\sigma=0.05$, at $n_t=1000$ points, and for various degrees $m$.
+We average over 7 runs of the algorithms and repeat these runs 1000 times to form the mean and standard deviation which are given in the below table.}
+    \label{tab:chebyshev-timing-interpolation}
+   \input{tables/interpolation.tex}
+\end{table}
 
@@
 
-In the end, we propose the following approximation for the smooth spectral density
+In the end, we propose the following approximation for the smoothed spectral density
 \begin{equation}
@@
 \end{equation}
-for which we can first precompute the $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ for $l=0, \dots, m$ using the recurrence \refequ{equ:chebyshev-recurrence} and then inexpensively sum them up with the corresponding coefficients $\mu_l(t)$ for all $t$ in which we want to evaluate the approximation.
+for which we can first precompute the small $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ for $l=0, \dots, m$ using the recurrence \refequ{equ:chebyshev-recurrence} and then inexpensively sum them up with the corresponding coefficients $\mu_l(t)$ for all $t$ in which we want to evaluate the approximation.
 Thanks to the fast and exact squaring of Chebyshev expansions (cf.
 \refalg{alg:chebyshev-squaring}), the second parenthesized term in \refequ{equ:cyclic-property} can be approximated in an equivalent way.
-Unlike expanding the squared smoothing kernel $g_{\sigma}^2$ separately \cite{lin-2017-randomized-estimation}, this approach is consistent, which we have observed to make a noticeable difference in terms of accuracy, and which enables provability.
+Unlike expanding the squared smoothing kernel $g_{\sigma}^2$ separately \cite{lin-2017-randomized-estimation}, this approach is consistent, which we have observed to make a noticeable difference in terms of accuracy (cf.
+\reffig{fig:interpolation-issue}), and which enables provability.
+
+\begin{figure}[ht]
+    \centering
+    \input{plots/interpolation.pgf}
+    \caption{Difference in the accuracy when computing \refequ{equ:3-nystrom-spectral-density} using a separate expansion (interpolation) versus explicitly squaring the matrix function (squaring).
+With the NC, which we will introduce on the next few pages, the spectral density of a 2D model problem from \refsec{sec:5-experiments-density-function} is computed for a Gaussian smoothing-kernel with smoothing-parameter $=0.05$ and fixed sketch-size $=80$.
+An estimate of the relative error in the $L^1$-norm is computed based on num-evaluation-points $=100$ evaluation of the spectral density.}
+    \label{fig:interpolation-issue}
+\end{figure}
+
 Similar observations for the case when $\mtx{\Psi}$ is nonzero lead to the following algorithm.
@@
         \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \leq 2\sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right) E_{\sqrt{2}\sigma, \sfrac{m}{2}} \equiv \underline{E}_{\sigma, m}
-        \label{equ:2-chebyshev-interpolation-sup-error-kernel}
+        \label{equ:chebyshev-interpolation-sup-error-kernel-nonneg}
         %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
@@
 \begin{theorem}{Chebyshev-Nyström++ method}{chebyshev-nystrom}
-    Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
+    Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
 Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, \emph{even} $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2)$, and smoothing parameter $0 < \sigma \ll 1$ satisfies with probability at least $1 - \delta$
@@
         &\leq (1 + \widetilde{\varepsilon}) \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \varepsilon \int_{-1}^{1} \left| \phi_{\sigma}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
-        &\leq (1 + \widetilde{\varepsilon}) 2 \underline{E}_{\sigma, m} + \widetilde{\varepsilon}. && \text{(\refthm{thm:chebyshev-error} and normalization)} \notag \\
+        &\leq (1 + \widetilde{\varepsilon}) 2 \underline{E}_{\sigma, m} + \widetilde{\varepsilon}. && \text{(\refthm{thm:chebyshev-error}, normalization)} \notag \\
     \end{align}
@@
     \input{plots/distribution.pgf}
-    \caption{The NC++ method for different ways of allocations a total of $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation for the Gaussian smoothing kernel with multiple different values of the smoothing parameter.
+    \caption{The Chebyshev-Nyström++ method for different ways of allocations a total of $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation for the Gaussian smoothing kernel with multiple different values of the smoothing parameter.
 We make the approximation error made in the Chebyshev expansion negligible by rescaling $m=16 / \sigma$ (based on \refthm{thm:chebyshev-error}).}
@@
 \clearpage
-\section{Discussion}
-\label{sec:discussion}
+\section{Conclusion}
+\label{sec:conclusion}
 
@@
 \clearpage
-\bibliography{bibliography.bib}
+%\bibliography{bibliography.bib}
+\printbibliography
 

### 2024-06-27 paper/plots/convergence.py

@@ -24,7 +24,7 @@ m = 2000
 n_Vec_list = np.logspace(1.3, 2.8, 7).astype(int) 
 
 plt.style.use("paper/plots/stylesheet.mplstyle")
-colors = cm.magma(np.arange(3) / 2.5)
+colors = ["#FFB000", "#648FFF", "#DC267F"]
 markers = ["o", "^", "s"]
 labels = [r"$n_{\mathbf{\Omega}} = 0$",
           r"$n_{\mathbf{\Psi}} = n_{\mathbf{\Omega}}$",

### 2024-06-27 paper/plots/distribution.py

@@ -19,13 +19,13 @@ eigvals_st = spectral_transformation(eigvals, min_ev, max_ev)
 
 # Set parameter
 t = np.linspace(-1, 1, 100)
-sigma_list = np.logspace(-3.0, -1.0, 7)
+sigma_list = np.logspace(-3.5, -1.0, 7)
 n_Vec = 80
 n_Psi_list = np.arange(n_Vec + 1, step=20).astype(np.int64)
 n_Omega_list = n_Vec - n_Psi_list
 
 plt.style.use("paper/plots/stylesheet.mplstyle")
-colors = cm.magma(np.arange(5) / 4.5)
+colors = ["#FFB000", "#FE6100", "#DC267F", "#785EF0", "#648FFF"]
 markers = ["o", "^", "s", "p", "d"]
 labels = [r"$n_{\mathbf{\Psi}} = " + "{}$, ".format(n_Psi) + r"$n_{\mathbf{\Omega}} = " + "{}$".format(n_Omega) for n_Psi, n_Omega in zip(n_Psi_list, n_Omega_list)]
 

### 2024-06-27 paper/plots/interpolation.py

@@ -1,0 +1,30 @@
+import __context__
+
+import numpy as np
+import matplotlib.pyplot as plt
+
+from src.algorithms import NC
+from src.matrices import ModES3D
+from src.plots import compute_spectral_density_errors, plot_spectral_density_errors
+
+plt.style.use("paper/plots/stylesheet.mplstyle")
+
+np.random.seed(0)
+
+methods = NC
+labels = ["interpolation", "squaring", "non-negative"]
+fixed_parameters = [{"n_v": 80, "sigma": 0.05, "square_coefficients": "interpolation"},
+                    {"n_v": 80, "sigma": 0.05, "square_coefficients": "transformation"},
+                    {"n_v": 80, "sigma": 0.05, "square_coefficients": "transformation", "nonnegative": True}]
+variable_parameters = "m"
+variable_parameters_values = (np.logspace(2.3, 3.5, 7).astype(int) // 2) * 2
+
+A = ModES3D(dim=2)
+colors = ["#FFB000", "#648FFF", "#DC267F"]
+markers = ["o", "^", "s"]
+spectral_density_errors = compute_spectral_density_errors(A, methods, labels, variable_parameters, variable_parameters_values, fixed_parameters, n_t=100)
+
+fig, ax = plt.subplots(figsize=(4, 3))
+plot_spectral_density_errors(spectral_density_errors, fixed_parameters, variable_parameters, variable_parameters_values, ignored_parameters=["square_coefficients", "n_v", "sigma", "nonnegative"], colors=colors, error_metric_name="$L^1$ error", x_label="expansion degree $m$", markers=markers, ax=ax)
+
+plt.savefig("paper/plots/interpolation.pgf", bbox_inches="tight")

### 2024-06-27 paper/plots/potential.py

@@ -1,0 +1,24 @@
+import __context__
+
+import matplotlib.pyplot as plt
+
+from src.matrices import regular_grid, periodic_gaussian_well
+
+plt.style.use("paper/plots/stylesheet.mplstyle")
+
+# Define grid parameters
+N = 100
+L = 6
+alpha = -4.0
+beta = 2.0
+
+# Plot periodic Gaussian wells for different number of repetitions
+ns = [1, 2, 5]
+for n in ns:
+    fig, ax = plt.subplots(figsize=(2, 1.75))
+    #ax.set_xlim([0, 1])
+    #ax.set_ylim(-11, 80)
+    grid_points = regular_grid(a=0, b=n*L, N=N, dim=2)
+    grid_values = periodic_gaussian_well(grid_points, L=L, n=n, var=beta**2, scaling_factor=alpha)
+    plt.contourf(grid_points[:, 0].reshape(N, N), grid_points[:, 1].reshape(N, N), grid_values.reshape(N, N), levels=8, cmap="plasma_r")
+    plt.savefig("paper/plots/gaussian-well-{:d}.pgf".format(n), bbox_inches="tight")

### 2024-06-27 paper/tables/interpolation.py

@@ -1,0 +1,35 @@
+import __context__
+
+import numpy as np
+
+from src.interpolation import chebyshev_coefficients, _chebyshev_coefficients_quadrature
+from src.kernel import gaussian_kernel
+from src.utils import time_method, generate_tex_tabular
+
+methods = [_chebyshev_coefficients_quadrature, chebyshev_coefficients, chebyshev_coefficients]
+labels = ["\\cite[algorithm 1]{lin-2017-randomized-estimation}", "\\textbf{DCT}", "\\textbf{DCT (non-negative)}"]
+
+n_t = 1000
+t = np.arange(-1, 1, n_t)
+sigma = 0.05
+n = 1000
+g = lambda x: gaussian_kernel(x, n=n, sigma=sigma)
+parameters = [{"t": t, "m": 800, "function": g},
+              {"t": t, "m": 1600, "function": g},
+              {"t": t, "m": 2400, "function": g},
+              {"t": t, "m": 3200, "function": g}]
+
+means = np.empty((len(methods), len(parameters)))
+errors = np.empty((len(methods), len(parameters)))
+
+for i in range(len(methods)):
+    for j in range(len(parameters)):
+        if i == 2:
+            parameters[j]["nonnegative"] = True
+        mean, error = time_method(methods[i], parameters[j], num_times=1000, num_repeats=7)
+        means[i, j] = 1e3 * mean
+        errors[i, j] = 1e3 * error
+
+headline = ["", r"$m=800$", r"$m=1600$", r"$m=2400$", r"$m=3200$"]
+
+generate_tex_tabular(means, "paper/tables/interpolation.tex", headline, labels, errors, fmt=r"${:.1f}$")

### 2024-06-27 paper/main.tex
generate plots which illustrate improvements

@@
 %}
-
-\todo{
-
-\paragraph{TODO}
-\begin{itemize}
-    \item Check that everywhere correct/consistent orders $\mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1})$)
-    \item 
-\end{itemize}
-}
 
@@
 \begin{table}[ht]
-    \caption{Runtime in milliseconds for computing the coefficients of the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ \refequ{equ:smoothing-kernel} with $\sigma=0.05$, at $n_t=1000$ points, and for various degrees $m$.
-We average over 7 runs of the algorithms and repeat these runs 1000 times to form the mean and standard deviation which are given in the below table.}
+    \caption{We compare the runtime in milliseconds for computing the coefficients of the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ \refequ{equ:smoothing-kernel} in three different ways: as proposed in \cite[algorithm 1]{lin-2017-randomized-estimation} (\emph{oversampled FFT}), using the discrete cosine transform (\emph{DCT}), and computing a non-negative Chebyshev expansion with \refalg{alg:chebyshev-squaring} (\emph{non-negative DCT}).
+We use $\sigma=0.005$, $n_t=1000$ parameter values, and various degrees $m$.
+For stability, we average over 7 runs of the algorithms and repeat these runs 1000 times to form the mean and standard deviation which are given in the below table.}
     \label{tab:chebyshev-timing-interpolation}
@@
     \input{plots/interpolation.pgf}
-    \caption{Difference in the accuracy when computing \refequ{equ:3-nystrom-spectral-density} using a separate expansion (interpolation) versus explicitly squaring the matrix function (squaring).
-With the NC, which we will introduce on the next few pages, the spectral density of a 2D model problem from \refsec{sec:5-experiments-density-function} is computed for a Gaussian smoothing-kernel with smoothing-parameter $=0.05$ and fixed sketch-size $=80$.
-An estimate of the relative error in the $L^1$-norm is computed based on num-evaluation-points $=100$ evaluation of the spectral density.}
+    \caption{Difference in the accuracy when computing the estimate \refequ{equ:cyclic-property} once by separately expanding the squared smoothing kernel $g_{\sigma}^2$ (\emph{inconsistent}), once by squaring the Chebyshev expansion $g_{\sigma}^{(m)}$ using \refalg{alg:chebyshev-squaring} (\emph{consistent}), and once with a non-negative Chebyshev expansion $\underline{g}_{\sigma}^{(m)}$ \refequ{equ:non-negative-chebyshev-expansion} which is also squared with \refalg{alg:chebyshev-squaring} (\emph{non-negative}).
+We use a small example matrix from \refsec{sec:results}, fix the size of the Nyström approximation to $n_{\mtx{\Omega}} = 80$, let $\sigma = 0.005$, and compute the $L^1$ error from the approximation of the spectral density at $n_t = 100$ uniformly spaced values of $t$.}
     \label{fig:interpolation-issue}
@@
 For small smoothing parameters $\sigma$, the rapid decay of the Gaussian $g_{\sigma}$ will cause $\mtx{K}(t_i) \approx g_{\sigma}(t_i\mtx{I}_n - \mtx{A})$ to be close to the zero matrix for parameter values $t_i$ located in such regions.
-Hence, on line \reflin{lin:4-nystromchebyshev-nystrom-pp} we would compute the pseudoinverse of an almost zero matrix, which can be extremely unstable.
+Hence, on line \reflin{lin:4-nystromchebyshev-nystrom-pp} we would compute the pseudoinverse of an almost zero matrix, which can be extremely unstable (see \reffig{fig:zerocheck}).
 Notice that $\Trace(\mtx{K}_1(t_i)) / n_{\mtx{\Omega}}$ is the $n_{\mtx{\Omega}}$-query Girard-Hutchinson estimator.
 We may use it as a rough indicator whether $t_i$ lies within a region where the spectral density nearly vanishes, and subsequently set $\widetilde{\phi}_{\sigma}(t_i) = 0$ instead of computing the full expression on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
+
+\begin{figure}[ht]
+    \centering
+    \input{plots/zerocheck.pgf}
+    \caption{The difference the non-zero check can make when approximating a spectral density using \refalg{alg:nystrom-chebyshev-pp}.
+Here, we used a matrix described in \refsec{sec:results} and ran the Chebyshev-Nyström++ method with and without estimating if the matrix function is zero before taking its pseudo-inverse.
+We let $m=2000$, $n_{\mtx{\Omega}}=80$, $n_{\mtx{\Psi}}=0$, and $\sigma = 0.004$.}
+    \label{fig:zerocheck}
+\end{figure}
 
@@
 \begin{equation}
-    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A})) \lessapprox n d_{\varepsilon, \sigma} \approx n \sigma.
+    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A})) \lessapprox n d_{\varepsilon, \sigma} \approx 6 n \sigma,
     \label{equ:gaussian-kernel-numerical-rank-uniform}
 \end{equation}
+provided we set $\varepsilon = 10^{-16}$ and choose $\sigma$ roughly between $10^{-8}$ and $1$.
 Hence, choosing $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma)$ will approximations which are significantly better than \refthm{thm:chebyshev-nystrom}.

### 2024-06-27 paper/plots/convergence.py

@@ -2,7 +2,6 @@ import __context__
 
 import numpy as np
 import scipy as sp
-import matplotlib.cm as cm
 import matplotlib.pyplot as plt
 from src.simple import spectral_density, spectral_transformation, form_spectral_density, gaussian_kernel
 
@@ -24,11 +23,9 @@ m = 2000
 n_Vec_list = np.logspace(1.3, 2.8, 7).astype(int) 
 
 plt.style.use("paper/plots/stylesheet.mplstyle")
-colors = ["#FFB000", "#648FFF", "#DC267F"]
-markers = ["o", "^", "s"]
-labels = [r"$n_{\mathbf{\Omega}} = 0$",
-          r"$n_{\mathbf{\Psi}} = n_{\mathbf{\Omega}}$",
-          r"$n_{\mathbf{\Psi}} = 0$"]
+colors = ["#648FFF", "#DC267F", "#FFB000"]
+markers = ["o", "s", "d"]
+labels = [r"$n_{\mathbf{\Omega}} = 0$", r"$n_{\mathbf{\Psi}} = n_{\mathbf{\Omega}}$", r"$n_{\mathbf{\Psi}} = 0$"]
 
 # Determine the baseline spectral density
 kernel = lambda x: gaussian_kernel(x, sigma=sigma)

### 2024-06-27 paper/plots/distribution.py

@@ -2,7 +2,6 @@ import __context__
 
 import numpy as np
 import scipy as sp
-import matplotlib.cm as cm
 import matplotlib.pyplot as plt
 from src.simple import spectral_density, spectral_transformation, form_spectral_density, gaussian_kernel
 

### 2024-06-27 paper/plots/interpolation.py

@@ -2,29 +2,51 @@ import __context__
 
 import numpy as np
 import matplotlib.pyplot as plt
-
-from src.algorithms import NC
-from src.matrices import ModES3D
-from src.plots import compute_spectral_density_errors, plot_spectral_density_errors
-
-plt.style.use("paper/plots/stylesheet.mplstyle")
+from src.matrices import hamiltonian
+from src.simple import spectral_density, spectral_transformation, form_spectral_density, gaussian_kernel
 
 np.random.seed(0)
 
-methods = NC
-labels = ["interpolation", "squaring", "non-negative"]
-fixed_parameters = [{"n_v": 80, "sigma": 0.05, "square_coefficients": "interpolation"},
-                    {"n_v": 80, "sigma": 0.05, "square_coefficients": "transformation"},
-                    {"n_v": 80, "sigma": 0.05, "square_coefficients": "transformation", "nonnegative": True}]
-variable_parameters = "m"
-variable_parameters_values = (np.logspace(2.3, 3.5, 7).astype(int) // 2) * 2
+# Load matrix
+A = hamiltonian(dim=2)
 
-A = ModES3D(dim=2)
-colors = ["#FFB000", "#648FFF", "#DC267F"]
-markers = ["o", "^", "s"]
-spectral_density_errors = compute_spectral_density_errors(A, methods, labels, variable_parameters, variable_parameters_values, fixed_parameters, n_t=100)
+# Perform spectral transform with A and its eigenvalues
+eigvals = np.linalg.eigvalsh(A.toarray())
+min_ev, max_ev = eigvals[0], eigvals[-1]
+A_st = spectral_transformation(A, min_ev, max_ev)
+eigvals_st = spectral_transformation(eigvals, min_ev, max_ev)
 
-fig, ax = plt.subplots(figsize=(4, 3))
-plot_spectral_density_errors(spectral_density_errors, fixed_parameters, variable_parameters, variable_parameters_values, ignored_parameters=["square_coefficients", "n_v", "sigma", "nonnegative"], colors=colors, error_metric_name="$L^1$ error", x_label="expansion degree $m$", markers=markers, ax=ax)
+# Set parameter
+t = np.linspace(-1, 1, 100)
+sigma = 0.005
+n_Omega = 80
+m_list = (np.logspace(1.8, 3.3, 7).astype(int) // 2) * 2
 
[…]

### 2024-06-27 paper/plots/potential.py

@@ -9,16 +9,14 @@ plt.style.use("paper/plots/stylesheet.mplstyle")
 # Define grid parameters
 N = 100
 L = 6
-alpha = -4.0
+V_0 = -4.0
 beta = 2.0
 
 # Plot periodic Gaussian wells for different number of repetitions
 ns = [1, 2, 5]
 for n in ns:
     fig, ax = plt.subplots(figsize=(2, 1.75))
-    #ax.set_xlim([0, 1])
-    #ax.set_ylim(-11, 80)
     grid_points = regular_grid(a=0, b=n*L, N=N, dim=2)
-    grid_values = periodic_gaussian_well(grid_points, L=L, n=n, var=beta**2, scaling_factor=alpha)
+    grid_values = periodic_gaussian_well(grid_points, L=L, n=n, var=beta**2, scaling_factor=V_0)
     plt.contourf(grid_points[:, 0].reshape(N, N), grid_points[:, 1].reshape(N, N), grid_values.reshape(N, N), levels=8, cmap="plasma_r")
     plt.savefig("paper/plots/gaussian-well-{:d}.pgf".format(n), bbox_inches="tight")

### 2024-06-27 paper/plots/zerocheck.py

@@ -1,0 +1,79 @@
+#import __context__
+#
+#import numpy as np
+#import scipy as sp
+#import matplotlib.pyplot as plt
+#
+#from src.simple import spectral_density
+#from src.plots import compute_spectral_densities, plot_spectral_densities
+#
+#import matplotlib
+#matplotlib.rc("text", usetex=True)
+#matplotlib.rcParams["pgf.texsystem"] = "pdflatex"
+#matplotlib.rcParams["font.family"] = r"serif"
+#matplotlib.rcParams["font.serif"] = r"Palatino"
+#matplotlib.rcParams["font.size"] = 12
+#
+#np.random.seed(0)
+#
+#methods = spectral_density
+#labels = ["no non-zero check", "non-zero check"]
+#parameters = [{"m": 2000, "sigma": 0.05, "n_v": 80, "kappa": -1},
+#              {"m": 2000, "sigma": 0.05, "n_v": 80, "kappa": 1e-5}]
+#
+#A = sp.sparse.load_npz("matrices/ModES3D_1.npz")
+#colors = ["#648FFF", "#DC267F", "#FFB000"]
+#spectral_densities = compute_spectral_densities(A, methods, labels, parameters, add_baseline=False, n_t=500)
+#
+#fig, ax = plt.subplots(figsize=(6, 3))
+#plot_spectral_densities(spectral_densities, parameters, variable_parameter="kappa", ignored_parameters=["m", "sigma", "n_v", "kappa"], colors=colors, ax=ax)
+#
+#plt.savefig("thesis/plots/short_circuit_mechanism.pgf", bbox_inches="tight")
+#
+#
+import __context__
+
+import numpy as np
+import scipy as sp
+import matplotlib.pyplot as plt
+from src.simple import spectral_density, spectral_transformation, form_spectral_density, gaussian_kernel
+from src.matrices import hamiltonian
+
+np.random.seed(0)
+
+# Load matrix
+A = hamiltonian(dim=3)
+
+# Perform spectral transform with A and its eigenvalues
+eigvals = np.linalg.eigvalsh(A.toarray())
+min_ev, max_ev = eigvals[0], eigvals[-1]
+A_st = spectral_transformation(A, min_ev, max_ev)
+eigvals_st = spectral_transformation(eigvals, min_ev, max_ev)
+
+# Set parameter
+t = np.linspace(-1, 1, 500)
+sigma = 0.004
+m = 2000
[…]

### 2024-06-27 paper/tables/interpolation.py

@@ -2,22 +2,65 @@ import __context__
 
 import numpy as np
 
-from src.interpolation import chebyshev_coefficients, _chebyshev_coefficients_quadrature
-from src.kernel import gaussian_kernel
+from src.simple import gaussian_kernel, chebyshev_expansion
 from src.utils import time_method, generate_tex_tabular
 
-methods = [_chebyshev_coefficients_quadrature, chebyshev_coefficients, chebyshev_coefficients]
-labels = ["\\cite[algorithm 1]{lin-2017-randomized-estimation}", "\\textbf{DCT}", "\\textbf{DCT (non-negative)}"]
+def chebyshev_coefficients_quadrature(t, m, kernel, n_theta=None):
+    """
+    Delta-Gauss-Chebyshev polynomial expansion.
+
+    Parameters
+    ----------
+    t : int, float, list, or np.ndarray of shape (n,)
+        Point(s) where the expansion should be evaluated.
+    m : int > 0
+        Degree of the Chebyshev polynomial.
+    kernel : callable
+        The kernel used to regularize the spectral density.
+    n_theta : int > M
+        The (half) number of integration points.
+
+    Returns
+    -------
+    mu : np.ndarray of shape (N_t, M + 1)
+        The coefficients of the Chebyshev polynomials. Format: mu[t, l].
+
+    References
+    ----------
+    [2] Lin, L. Randomized estimation of spectral densities of large matrices
+        made accurate. Numer. Math. 136, 203-204 (2017). Algorithm 1.
+        DOI: https://doi.org/10.1007/s00211-016-0837-7
+    """
+    # If t is a scalar, we convert it to a 1d array to make computation work
+    if not isinstance(t, np.ndarray):
+        t = np.array(t).reshape(-1)
+
+    # If not specified, take minimum number of quadrature nodes
+    if n_theta is None:
+        n_theta = 2*(m + 1)
+
+    theta = np.arange(2 * n_theta) * np.pi / n_theta
+
+    # Can be computed via Fourier transform:
+    t_minus_theta = np.subtract.outer(t, np.cos(theta))
+    mu = np.real(np.fft.fft(kernel(t_minus_theta), axis=1)[:, :m+1])
+
+    # Rescale the coefficients (as required by the definition)
+    mu[:, 0] /= 2 * n_theta
+    mu[:, 1:] /= n_theta
+    return mu
+
[…]

### 2024-06-27 setup/generate_matrices.py

@@ -6,7 +6,7 @@ import scipy as sp
 import numpy as np
 
 from src.utils import download_matrix
-from src.matrices import ModES3D, uniform, gaussian_orthogonal_ensemble
+from src.matrices import hamiltonian, uniform, gaussian_orthogonal_ensemble
 
 matrix_dir = "matrices"
 
@@ -29,7 +29,7 @@ ns = [1, 2]#, 3, 4]
 
 print("\nGenerating ModES3D matrices")
 for i, n in enumerate(ns):
-    matrix = ModES3D(n=n, L=6, h=0.6, dim=3, bc="periodic", beta=2.0, alpha=-4.0)
+    matrix = hamiltonian(n=n, L=6, h=0.6, dim=3, bc="periodic", beta=2.0, alpha=-4.0)
     sp.sparse.save_npz(os.path.join(matrix_dir, "ModES3D_{}".format(n**3)), matrix)
     print("\u2713 Generated matrix ({}/{})".format(i + 1, len(ns)))
 

### 2024-06-27 src/matrices.py

@@ -221,7 +221,7 @@ def periodic_gaussian_well(X, n=1, L=6, var=1.0, scaling_factor=1.0):
     return potential
 
 
-def ModES3D(n=1, L=6, h=0.6, dim=3, bc="periodic", beta=2.0, alpha=-4.0):
+def hamiltonian(n=1, L=6, h=0.6, dim=3, bc="periodic", beta=2.0, alpha=-4.0):
     """
     Generate the example matrices 'ModES3D_X' from [2].
 

### 2024-06-27 src/simple.py

@@ -138,7 +138,7 @@ def square_chebyshev_expansion(mu):
 
     return nu
 
-def spectral_density(A, t=0, m=100, n_Psi=10, n_Omega=10, kernel=gaussian_kernel, nonnegative=False, kappa=1e-5, rcond=1e-5, seed=0):
+def spectral_density(A, t=0, m=100, n_Psi=10, n_Omega=10, kernel=gaussian_kernel, nonnegative=False, consistent=True, kappa=1e-5, rcond=1e-5, seed=0):
     """
     Chebyshev-Nyström++ method for estimating the spectral density.
 
@@ -184,7 +184,10 @@ def spectral_density(A, t=0, m=100, n_Psi=10, n_Omega=10, kernel=gaussian_kernel
 
     # Compute coefficients of Chebyshev expansion of the smoothing kernel
     mu = chebyshev_expansion(kernel, t, m, nonnegative=nonnegative)
-    nu = square_chebyshev_expansion(mu)
+    if consistent:
+        nu = square_chebyshev_expansion(mu)
+    else:
+        nu = chebyshev_expansion(lambda s: kernel(s) ** 2, t, 2 * m, nonnegative=nonnegative)
 
     # Generate Gaussian random matrices
     Omega = np.random.randn(n, n_Omega)

### 2024-06-27 paper/re-pro-badge.tex
add reproducibility pipeline

This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-SD/tree/ca391a8}{ca391a8} at \href{https://github.com/FMatti/Rand-TRACE}{github.com/FMatti/Rand-TRACE} on 2024-02-05 at 18:44:29 UTC.

### 2024-06-27 paper/plots/distribution.py

@@ -4,11 +4,12 @@ import numpy as np
 import scipy as sp
 import matplotlib.pyplot as plt
 from src.simple import spectral_density, spectral_transformation, form_spectral_density, gaussian_kernel
+from src.matrices import hamiltonian
 
 np.random.seed(0)
 
 # Load matrix
-A = sp.sparse.load_npz("matrices/ModES3D_1.npz")
+A = hamiltonian()
 
 # Perform spectral transform with A and its eigenvalues
 eigvals = np.linalg.eigvalsh(A.toarray())

### 2024-06-27 paper/plots/zerocheck.py

@@ -34,7 +34,6 @@
 import __context__
 
 import numpy as np
-import scipy as sp
 import matplotlib.pyplot as plt
 from src.simple import spectral_density, spectral_transformation, form_spectral_density, gaussian_kernel
 from src.matrices import hamiltonian

### 2024-06-27 setup/generate_paper.py

@@ -1,0 +1,28 @@
+import __context__
+
+import os
+import glob
+
+plot_files = glob.glob("paper/plots/*.py")
+table_files = glob.glob("paper/tables/*.py")
+
+for plot_file in plot_files:
+    os.system("python " + plot_file)
+
+for table_file in table_files:
+    os.system("python " + table_file)
+
+project_name = "paper"
+os.chdir(os.path.join(os.path.abspath(""), project_name))
+
+def send_command(command, file="", args=""):
+    cmd = " ".join([command, file, *args])
+    os.system(cmd)
+
+pdflatex_args = ["-synctex=1", "-interaction=nonstopmode", "-file-line-error"]
+
+send_command("pdflatex", project_name, pdflatex_args)
+send_command("bibtex", project_name)
+send_command("makeglossaries", project_name)
+send_command("pdflatex", project_name, pdflatex_args)
+send_command("pdflatex", project_name, pdflatex_args)

### 2024-06-27 paper/main.tex
add reproducibility pipeline
make proof-read adjustments

@@
 In numerous problems across physics, chemistry, engineering, and data science, how the eigenvalues of certain matrices are distributed can be indicative of what properties a system has: In electronic structure calculations eigenvalues represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}.
-The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$.
+The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$ with the Dirac delta distribution $\delta$.
 Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation which is often prohibitively expensive.
@@
 Interestingly, we can reuse the same randomization for computing the estimate at each value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations.
-We propose certain modifications to the algorithms from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}.
-In particular, use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials, and realize that an efficient implementation of all of their methods can be summarized in a single algorithm.
+We propose certain modifications to the methods from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}.
+In particular, we use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials and unify all of their methods in a single algorithm, with the desirable side effect of speeding up their \enquote{spectrum sweeping} method by orders of magnitude.
 Based on the analysis of the randomized trace estimators for parameter-dependent matrices we provide bounds on the approximation error for our method.
@@
 
-\paragraph{Reproducibility} This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-SD/tree/ca391a8}{ca391a8} at \href{https://github.com/FMatti/Rand-TRACE}{github.com/FMatti/Rand-TRACE} on 2024-02-05 at 18:44:29 UTC.
+\paragraph{Reproducibility} \input{re-pro-badge.tex}
 
@@
 
-The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Minkowski's integral inequality to extend the results to the continuous parameter case.
+The idea behind the proof is to bound the higher order moments of the residual $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ for fixed values of $t$ and then use Minkowski's integral inequality to extend the results to the continuous parameter case.
 We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
 \begin{proof} 
-    We define the 1-query Girard-Hutchinson estimate 
+    We define the residual of the 1-query Girard-Hutchinson estimate 
     \begin{equation}
@@
 \begin{proof}
-    From the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms} we know that
+    In the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms}, it is shown that Slepian's inequality \cite[theorem 7.2.1]{vershynin-2018-random-processes} applied to two appropriately constructed Gaussian fields yields
     \begin{equation}
@@
 For $n_{\mtx{\Psi}} = 0$ we end up with the Nyström estimator \refequ{equ:nystrom-trace-estimator}, provided the implementation evaluates $0/0$ to $0$ in the second term on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
+Notice that in this case, \refalg{alg:nystrom-chebyshev-pp} is equivalent to the \enquote{spectrum sweeping} method from \cite[algorithm 5]{lin-2017-randomized-estimation}, but has a significantly lower computational complexity.
 

### 2024-06-27 paper/plots/zerocheck.py

@@ -1,36 +1,3 @@
-#import __context__
-#
-#import numpy as np
-#import scipy as sp
-#import matplotlib.pyplot as plt
-#
-#from src.simple import spectral_density
-#from src.plots import compute_spectral_densities, plot_spectral_densities
-#
-#import matplotlib
-#matplotlib.rc("text", usetex=True)
-#matplotlib.rcParams["pgf.texsystem"] = "pdflatex"
-#matplotlib.rcParams["font.family"] = r"serif"
-#matplotlib.rcParams["font.serif"] = r"Palatino"
-#matplotlib.rcParams["font.size"] = 12
-#
-#np.random.seed(0)
-#
-#methods = spectral_density
-#labels = ["no non-zero check", "non-zero check"]
-#parameters = [{"m": 2000, "sigma": 0.05, "n_v": 80, "kappa": -1},
-#              {"m": 2000, "sigma": 0.05, "n_v": 80, "kappa": 1e-5}]
-#
-#A = sp.sparse.load_npz("matrices/ModES3D_1.npz")
-#colors = ["#648FFF", "#DC267F", "#FFB000"]
-#spectral_densities = compute_spectral_densities(A, methods, labels, parameters, add_baseline=False, n_t=500)
-#
-#fig, ax = plt.subplots(figsize=(6, 3))
-#plot_spectral_densities(spectral_densities, parameters, variable_parameter="kappa", ignored_parameters=["m", "sigma", "n_v", "kappa"], colors=colors, ax=ax)
-#
-#plt.savefig("thesis/plots/short_circuit_mechanism.pgf", bbox_inches="tight")
-#
-#
 import __context__
 
 import numpy as np

### 2024-06-27 setup/generate_paper.py

@@ -1,16 +1,6 @@
 import __context__
 
 import os
-import glob
-
-plot_files = glob.glob("paper/plots/*.py")
-table_files = glob.glob("paper/tables/*.py")
-
-for plot_file in plot_files:
-    os.system("python " + plot_file)
-
-for table_file in table_files:
-    os.system("python " + table_file)
 
 project_name = "paper"
 os.chdir(os.path.join(os.path.abspath(""), project_name))

### 2024-06-27 paper/re-pro-badge.tex
automatically reproduce paper

@@
-This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-SD/tree/ca391a8}{ca391a8} at \href{https://github.com/FMatti/Rand-TRACE}{github.com/FMatti/Rand-TRACE} on 2024-02-05 at 18:44:29 UTC.
+This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-SD/tree/ca391a8}{ca391a8} at \href{https://github.com/FMatti/Rand-TRACE}{github.com/FMatti/Rand-TRACE} on 2024-02-05 at 18:44:29 UTC.This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-TRACE/tree/4cf9b6b}{4cf9b6b}}} at \url{https://github.com/FMatti/Rand-TRACE}}} on 2024-06-27 at 15:01:59 UTC.

### 2024-06-28 paper/re-pro-badge.tex
automatically reproduce paper
fix bibtex issue in github action

--- 2024-06-27 paper/re-pro-badge.tex
@@
-This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-SD/tree/ca391a8}{ca391a8} at \href{https://github.com/FMatti/Rand-TRACE}{github.com/FMatti/Rand-TRACE} on 2024-02-05 at 18:44:29 UTC.
+This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-TRACE/tree/4d43b2e}{4d43b2e} at \url{https://github.com/FMatti/Rand-TRACE} on 2024-06-27 at 17:10:54 UTC.

### 2024-06-28 paper/paper.tex
write abstract and clean up bibliography

--- 2024-06-27 paper/main.tex
@@
 
-\todo{
 \begin{abstract}
-    Write abstract.
+    Stochastic trace estimation is a family of well established techniques for approximating the trace of constant matrices.
+However, oftentimes the matrix whose trace we aim to approximate continuously depends on a real parameter $t \in [a,b]$.
+This is the case when approximating the spectral density of a matrix or computing partition functions of quantum systems.
+Clearly, we could separately approximate the traces of the constant matrices $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ at the fixed values $t_1, \dots, t_m \in [a, b]$.
+However, this approach does not scale well if $m$ is large.
+We analyze three trace estimators for constant matrices -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter-dependent matrices with the same randomization for all values of the parameter $t$.
+We show that all of these estimators closely match their corresponding error guarantees for constant matrices.
+In particular, we show that for the Nyström++ method, we can guarantee an error of order $\varepsilon$ by just using $\mathcal{O}(\varepsilon^{-1})$ random vectors to build the estimate in all values of $t$, independent of what structure the symmetric positive semi-definite matrix $\mtx{B}(t)$ has.
+We employ these trace estimators to improve upon a technique for approximating the spectral density of a matrix and provide a detailed analysis of the resulting method.
+Finally, we confirm our theoretical developments in multiple numerical experiments.
 \end{abstract}
-}
 
@@
 for a smoothing kernel $g_{\sigma}$, typically a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}.
-This translates the problem to computing the trace of a continuous parameter-dependent matrix $\mtx{B}(t) \equiv g_{\sigma}(t \mtx{I}_n - \mtx{A}) \in \mathbb{R}^{n \times n}$.
+This translates the problem to computing the trace of a symmetric parameter-dependent matrix $\mtx{B}(t) \equiv g_{\sigma}(t \mtx{I}_n - \mtx{A}) \in \mathbb{R}^{n \times n}$.
 Directly assembling the matrix function requires diagonalizing the matrix $\mtx{A}$, which again is prohibitively expensive.

### 2024-06-28 paper/plots/distribution.py

@@ -26,7 +26,7 @@ n_Omega_list = n_Vec - n_Psi_list
 
 plt.style.use("paper/plots/stylesheet.mplstyle")
 colors = ["#FFB000", "#FE6100", "#DC267F", "#785EF0", "#648FFF"]
-markers = ["o", "^", "s", "p", "d"]
+markers = ["d", "p", "s", "^", "o"]
 labels = [r"$n_{\mathbf{\Psi}} = " + "{}$, ".format(n_Psi) + r"$n_{\mathbf{\Omega}} = " + "{}$".format(n_Omega) for n_Psi, n_Omega in zip(n_Psi_list, n_Omega_list)]
 
 error = np.empty((len(n_Psi_list), len(sigma_list)))

### 2024-06-28 paper/plots/zerocheck.py

@@ -18,7 +18,7 @@ eigvals_st = spectral_transformation(eigvals, min_ev, max_ev)
 
 # Set parameter
 t = np.linspace(-1, 1, 500)
-sigma = 0.004
+sigma = 0.005
 m = 2000
 n_Omega = 80
 

### 2024-06-28 setup.py

@@ -1,15 +1,0 @@
-import os
-import argparse
-
-parser = argparse.ArgumentParser(
-                    prog="setup.py",
-                    description="Usage of the automated setup",
-                    epilog="Text at the bottom of help")
-
-parser.add_argument("-a", "--generate-all", action="store_true", help="reproduce the whole project (matrices, ...)")
-parser.add_argument("-m", "--generate-matrices", action="store_true", help="download and create all matrices")
-
-args = parser.parse_args()
-
-if args.generate_matrices or args.generate_all:
-    os.system("python setup/generate_matrices.py")

### 2024-06-28 paper/re-pro-badge.tex
automatically reproduce paper

@@
-This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-TRACE/tree/4d43b2e}{4d43b2e} at \url{https://github.com/FMatti/Rand-TRACE} on 2024-06-27 at 17:10:54 UTC.
+This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-TRACE/tree/170f339}{170f339} at \url{https://github.com/FMatti/Rand-TRACE} on 2024-06-28 at 12:22:33 UTC.

### 2024-07-01 paper/paper.tex
implement proof-reading changes

@@
 \begin{abstract}
-    Stochastic trace estimation is a family of well established techniques for approximating the trace of constant matrices.
-However, oftentimes the matrix whose trace we aim to approximate continuously depends on a real parameter $t \in [a,b]$.
-This is the case when approximating the spectral density of a matrix or computing partition functions of quantum systems.
+    Stochastic trace estimators are a family of well established techniques for approximating the trace of constant matrices.
+Oftentimes the matrix whose trace we aim to approximate continuously depends on a real parameter $t \in [a,b]$.
+This is for example the case when approximating the spectral density of a matrix or computing partition functions of quantum systems.
 Clearly, we could separately approximate the traces of the constant matrices $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ at the fixed values $t_1, \dots, t_m \in [a, b]$.
 However, this approach does not scale well if $m$ is large.
-We analyze three trace estimators for constant matrices -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter-dependent matrices with the same randomization for all values of the parameter $t$.
+We analyze three trace estimators for constant matrices -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter-dependent matrices with the same randomization reused for all values of the parameter $t$.
 We show that all of these estimators closely match their corresponding error guarantees for constant matrices.
-In particular, we show that for the Nyström++ method, we can guarantee an error of order $\varepsilon$ by just using $\mathcal{O}(\varepsilon^{-1})$ random vectors to build the estimate in all values of $t$, independent of what structure the symmetric positive semi-definite matrix $\mtx{B}(t)$ has.
+In particular, for the Nyström++ method, we can guarantee an error of at most $\varepsilon$ by computing the estimate with just $\mathcal{O}(\varepsilon^{-1})$ random vectors, independent of what structure the symmetric positive semi-definite matrix $\mtx{B}(t)$ has.
 We employ these trace estimators to improve upon a technique for approximating the spectral density of a matrix and provide a detailed analysis of the resulting method.
@@
 In numerous problems across physics, chemistry, engineering, and data science, how the eigenvalues of certain matrices are distributed can be indicative of what properties a system has: In electronic structure calculations eigenvalues represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}.
-The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$ with the Dirac delta distribution $\delta$.
+The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$ which places a rescaled Dirac delta distribution $\delta$ at each eigenvalue.
 Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation which is often prohibitively expensive.
-Therefore, many techniques have been proposed for approximating $\phi$.
+Therefore, many techniques have been developed for approximating $\phi$.
 For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature rule with the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}.
-Further, because $\delta$ is hard to approximate and in many applications, a rough estimate of the spectral density is already acceptable, a standard approach is to instead approximate the smoothed spectral density
+Further, because $\delta$ is hard to approximate and because in many applications, a rough estimate of the spectral density is already acceptable, a standard approach is to instead approximate the smoothed spectral density
 \begin{equation}
@@
 Instead of discretizing $[a, b]$ into $t_1, \ldots, t_m$, approximating $\Trace(\mtx{B}(t_i))$ with $b(t_i)$, and analyzing the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way is to quantify the error with $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate in each $t \in [a, b]$.
-This approach has already been taken for analyzing low-rank approximations of parameter-dependent matrices \cite{kressner-2023-randomized-lowrank}.
+This approach has already been used in the analysis of low-rank approximations of parameter-dependent matrices \cite{kressner-2023-randomized-lowrank}.
 
@@
 In particular, we use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials and unify all of their methods in a single algorithm, with the desirable side effect of speeding up their \enquote{spectrum sweeping} method by orders of magnitude.
-Based on the analysis of the randomized trace estimators for parameter-dependent matrices we provide bounds on the approximation error for our method.
+Based on the analysis of the randomized trace estimators for parameter-dependent matrices, we provide bounds on the approximation error for our method.
 
 \paragraph{Structure} In \refsec{sec:analysis} we will analyze three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter dependent matrices.
-We propose and analyze a method for using the aforementioned estimators to estimate the spectral density of real symmetric matrices in \refsec{sec:application}.
-Finally, we will test our developments on an example from electronic structure calculation.
+We propose and analyze an algorithm for using the aforementioned estimators to approximate the spectral density of real symmetric matrices in \refsec{sec:application}.
+Finally, we will test our developments on an example from electronic structure calculation in \refsec{sec:results}.
 
@@
 \end{equation}
-However, we assume that we only have access to products of this matrix with vectors for each $t \in [a, b]$, so this definition will not be directly useful for computing the trace.
+However, as is often the case for matrix functions or implicit matrices, we assume that we can only efficiently access this matrix through products with vectors for each $t \in [a, b]$, so this definition will not directly be useful for computing the trace.
 
@@
 \end{equation}
-This is the parameter-dependent analogue of the Nyström++ estimator \cite{persson-2022-improved-variants}, which is based on the Hutch++ estimator \cite{meyer-2021-hutch-optimal}.
+This is the parameter-dependent analogue of the Nyström++ estimator \cite{persson-2022-improved-variants}, which itself is based on the Hutch++ estimator \cite{meyer-2021-hutch-optimal}.
 %We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.
@@
 We first analyze the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}.
-We bound its $L^1$ norm in terms of the integral over the Frobenius norm $\lVert \cdot \rVert _F$ of the matrix whose trace we aim to approximate.
+We bound its $L^1$-error in terms of the $L^1$-norm of the Frobenius norm $\lVert \cdot \rVert _F$ of the matrix whose trace we approximate.
 
@@
     \end{equation}
-    In particular, if we choose $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{- \sfrac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$.
+    In particular, if we choose $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, 1)$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$.
 \end{theorem}
@@
     \end{align}
-    Since $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
+    Because $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
     \begin{equation}
@@
     Now we make the transition to the continuous.
-Following \cite{kressner-2023-randomized-lowrank}, we can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable.
+Following \cite{kressner-2023-randomized-lowrank}, we can show that $r(\mtx{B}(t), \vct{\psi})$ is measurable.
 Therefore, Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting:
@@
 
-    Setting $\delta = \gamma^{-2 k}$ and choosing $k = \lceil \log(\delta^{-1}) \rceil$, where $\lceil x \rceil$ is the ceiling function which returns the greatest integer larger than or equal to $x$, we get that for all $\delta \in (0, e^{-\sfrac{1}{2}})$
-    \begin{equation}
-        k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\sfrac{1}{2\lceil \log(\delta^{-1}) \rceil}}
-        = \lceil \log(\delta^{-1}) \rceil e^{\sfrac{1}{2}\frac{\log(\delta^{-1})}{\lceil \log(\delta^{-1}) \rceil}}
-        \leq 2 \log(\delta^{-1}) e^{\sfrac{1}{2}},
-    \end{equation}
-    from which follows the second part of the theorem when $n_{\mtx{\Psi}} = (2^5 e^{\sfrac{1}{2}} \varepsilon^{-1} \log(\delta^{-1}))^2$.
+    %Setting $\delta = \gamma^{-2 k}$ and choosing $k = \lceil \log(\delta^{-1}) %\rceil$, where $\lceil x \rceil$ is the ceiling function which returns the greatest %integer larger than or equal to $x$, we get that for all $\delta \in (0, e^{-\sfrac%{1}{2}})$
+    %\begin{equation}
+    %    k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\sfrac{1}{2\lceil \log%(\delta^{-1}) \rceil}}
+    %    = \lceil \log(\delta^{-1}) \rceil e^{\sfrac{1}{2}\frac{\log(\delta^{-1})}%{\lceil \log(\delta^{-1}) \rceil}}
+    %    \leq 2 \log(\delta^{-1}) e^{\sfrac{1}{2}},
+    %\end{equation}
+    %from which follows the second part of the theorem when $n_{\mtx{\Psi}} = (2^5 e^%{\sfrac{1}{2}} \varepsilon^{-1} \log(\delta^{-1}))^2$.
+
+    Setting $\gamma = e$ and choosing $k = \lfloor \log(\delta^{-1}) / 2 \rfloor$, where $\lfloor x \rfloor$ is the floor function which returns the largest integer smaller than or equal to $x$, we get that for all $\delta \in (0, 1)$ the second part of the theorem follows when taking $n_{\mtx{\Psi}} = \lceil (2^3 e \varepsilon^{-1} \log(\delta^{-1}))^2 \rceil$, where $\lceil x \rceil$ is the ceiling function which returns the smallest integer larger than or equal to $x$.
+
     %Consequently, with probability $1 - \delta$ we have
@@
 
-%\todo{
 %\begin{remark}
@@
 %\end{remark}
-%}
 
@@
     Hence, by \cite[theorem 9]{kressner-2023-randomized-lowrank}, with $n_{\mtx{\Omega}} = p + r$, the result follows by noting $\sigma_i(\mtx{B}(t)^{\sfrac{1}{2}})^{2} = \sigma_i(\mtx{B}(t))$.
+
 \end{proof}
@@
 
-For reasons which will become clear in the proof of \reflem{lem:nystrom}, we will need to bound $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]$ for a constant matrix $\mtx{A}$ and a standard Gaussian random matrix $\mtx{\Omega}$.
-The corresponding result for the Frobenius norm $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _F^2 \right]$ has already been shown in \cite{kressner-2023-randomized-lowrank}.
-Several ways of separately bounding the spectral norm moments of the Wishart matrix $\mtx{\Omega} \mtx{\Omega}^{\top}$ already exist, however, most of them are not sufficient for our purposes \cite{chen-2005-condition-numbers, edelman-1988-eigenvalues-condition, james-1964-distributions-matrix}.
+For reasons which will become clear in the proof of the final and most interesting of these lemmas (\reflem{lem:nystrom}), we will need to bound $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]$ for a constant matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and a standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$.
+The corresponding result for the Frobenius norm $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _F^2 \right]$ has already been shown in \cite[lemma 7]{kressner-2023-randomized-lowrank}.
+Several ways of separately bounding the spectral norm moments of the Wishart matrix $\mtx{\Omega} \mtx{\Omega}^{\top}$, i.e.
+when $\mtx{A} = \mtx{I}_m$ already exist, however, all of them are not sufficient for our purposes because they scale with $m$ \cite{chen-2005-condition-numbers, edelman-1988-eigenvalues-condition, james-1964-distributions-matrix}.
 In \cite[lemma B.1]{tropp-2023-randomized-algorithms} a bound is achieved for $p = 2$ and $p = 4$.
@@
     \begin{equation}
-        \overbrace{\underbrace{\sigma_1, \dots, \sigma_k}_{\leq \sigma_1}}^{\geq \sigma_{k+1}}, \overbrace{\underbrace{\sigma_{k+1}, \dots, \sigma_{2k}}_{\leq \sigma_{k+1}}}^{\geq \sigma_{2k+1}}, \dots, \overbrace{\underbrace{\sigma_{1+(\ell - 1)k}, \dots, \sigma_{\ell k}}_{\leq \sigma_{1+(\ell - 1)k}}}^{\geq 0}.
+        \overbrace{\underbrace{\sigma_1, \dots, \sigma_k}_{\leq \sigma_1}}^{\geq \sigma_{k+1}}, \overbrace{\underbrace{\sigma_{k+1}, \dots, \sigma_{2k}}_{\leq \sigma_{k+1}}}^{\geq \sigma_{2k+1}}, \dots, \overbrace{\underbrace{\sigma_{(\ell - 1)k + 1}, \dots, \sigma_{\ell k}}_{\leq \sigma_{(\ell - 1)k + 1}}}^{\geq 0}.
     \end{equation}
@@
         \quad \text{and} \quad
-        \sigma_{1 + (i-1)k}^2 \leq \frac{1}{k} \sum_{j=1}^{k} \sigma_{j + (i-2)k}^2, i = 2, \dots, \ell,
+        \sigma_{(i-1)k + 1}^2 \leq \frac{1}{k} \sum_{j=1}^{k} \sigma_{(i-2)k + j}^2, i = 2, \dots, \ell,
     \end{equation}
@@
     \begin{equation}
-        \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \sum_{j=1}^{(\ell - 1)k} \sigma_j^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2.
+        \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 = \lVert \mtx{\Sigma} \rVert _2^2 + \sum_{i=2}^{\ell} \sigma_{(i-1)k + 1}^2  \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \sum_{n=1}^{(\ell - 1)k} \sigma_n^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2.
         \label{equ:singular-value-group-bound}
@@
         \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right]
-        &= \mathbb{E}^{\sfrac{p}{2}}\Bigg[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{j + (i-1)k}^2}_{\leq \sigma_{1 + (i-1)k}^2} \omega_{j + (i-1)k}^2 \Bigg] && \text{(separate sum into groups)} \notag \\
-        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{($\sigma_{j + (i-1)k}^2 \leq \sigma_{1 + (i-1)k}^2$, $\forall i, j$)} \notag \\
-        &\leq \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 ~ \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{(triangle inequality)} \notag \\
+        &= \mathbb{E}^{\sfrac{p}{2}}\Bigg[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{(i-1)k + j}^2}_{\leq \sigma_{(i-1)k + 1}^2} \omega_{(i-1)k + j}^2 \Bigg] && \text{(separate sum into groups)} \notag \\
+        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 \sum_{j=1}^{k} \omega_{(i-1)k + j}^2 \right] && \text{($\sigma_{(i-1)k + j}^2 \leq \sigma_{(i-1)k + 1}^2$, $\forall i, j$)} \notag \\
+        &\leq \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 ~ \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{(i-1)k + j}^2 \right] && \text{(triangle inequality)} \notag \\
         %&\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k}{2} + p)}{\Gamma(\frac{k}{2})} \right)^{\sfrac{1}{p}} && \text{(\refequ{equ:singular-value-group-bound} and \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
-        &\leq (k + p) \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 && \text{(\reflem{lem:gamma})} \notag \\
+        &\leq (k + p) \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 && \text{(\reflem{lem:gamma})} \notag \\
         &\leq (k + p) \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right). && \text{(using \refequ{equ:singular-value-group-bound})}
@@
     \end{equation}
-    for any $k,p \in \mathbb{N}$. Indeed, for $p=2$ this bound is slightly less tight than its equivalent in \cite[lemma B.1]{tropp-2023-randomized-algorithms}.
+    for any $k,p \in \mathbb{N}$. For $p=2$ this bound is slightly less tight than its equivalent in \cite[lemma B.1]{tropp-2023-randomized-algorithms}.
 \end{remark}
@@
 \begin{proof}
-    In the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms}, it is shown that Slepian's inequality \cite[theorem 7.2.1]{vershynin-2018-random-processes} applied to two appropriately constructed Gaussian fields yields
+    In the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms}, it is shown that Slepian's inequality \cite[theorem 7.2.1]{vershynin-2018-random-processes} applied to two appropriately constructed Gaussian random fields yields
     \begin{equation}
@@
     for any $\gamma \geq 1$ and a universal constant $c > 0$.
-In particular, if we choose $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1}))$, then with probability at least $1-\delta$ for $\delta \in (0, 1)$ and $\varepsilon \in (0, 1)$ we have $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$.
+In particular, if we choose $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1}))$, then with probability at least $1-\delta$ for $\delta \in (0, 1)$ and $\varepsilon > 0$ we have $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$.
 \end{lemma}
@@
     First, we temporarily omit the parameter dependence.
-Let $\mtx{B} = \mtx{U} \mtx{\Lambda} \mtx{U}^{\top}$ be the eigenvalue decomposition of a symmetric positive semi-definite matrix $\mtx{B}$, i.e.
+Let $\mtx{B} = \mtx{U} \mtx{\Lambda} \mtx{U}^{\top}$ be the eigenvalue decomposition of a symmetric positive semi-definite matrix $\mtx{B} \in \mathbb{R}^{n \times n}$, i.e.
 $\mtx{\Lambda} = \operatorname{diag}(\lambda_1, \dots, \lambda_n)$ and $\mtx{U}^{\top} \mtx{U}= \mtx{I}_n$.
-We define the matrix partitions
+We partition these matrices into
     \begin{equation}
@@
     \end{equation}
-    The $p$-th moment of the second term can be processed with standard matrix norm manipulations to
-    \begin{align}
-        \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
+    The $q$-th moment of the second term can be processed with standard matrix norm manipulations and the stochastic independence of $\mtx{\Omega}_1$ and $\mtx{\Omega}_2$ to
+    \begin{equation}
+        \mathbb{E}^{q}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
         %&= \mathbb{E}^{p}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )( \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger})^{\top} \rVert _F \right] && \text{($\lVert \mtx{A} \rVert _{(4)}^2 = \lVert \mtx{A} \mtx{A}^{\top} \rVert _{F}$)} \notag \\
         %&\leq \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] && \text{($\lVert \mtx{A} \mtx{B} \rVert _F \leq \lVert \mtx{A} \rVert _F \lVert \mtx{B} \rVert _2$)} \notag \\
-        &\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{($\lVert \mtx{A} \rVert _{(4)}^2 \leq \sqrt{k} \lVert \mtx{A} \rVert _2^2$)} \notag \\
-        &\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{(submultiplicativity)} \notag \\
-        &\leq \sqrt{k} \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2\right] \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2  \right].
-&& \text{(independence)}
+        \leq \mathbb{E}^{q}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]%  && \text{($\lVert \mtx{A} \rVert _{(4)}^2 \leq \sqrt{k} \lVert \mtx{A} \rVert _2^2$)} \notag \\
+        %\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]%  && \text{(submultiplicativity)} \notag \\
+        \leq \sqrt{k} \mathbb{E}^{q}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2\right] \mathbb{E}^{q}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2  \right].% && \text{(independence)}
         \label{equ:nystrom-proof-processed-tail}
-    \end{align}
-    To match the decay rate of the moments of the first term in \refequ{equ:nystrom-proof-persson-bonud}, we will need to ensure that this term also is of order $\mathcal{O}(\Trace{\mtx{B}} / \sqrt{k})$.
-To this purpose, and to simplify the notation, we choose $k = n_{\mtx{\Omega}}/2$ and $p = n_{\mtx{\Omega}}/4$ for even $n_{\mtx{\Omega}}$ to ensure $k \in \mathbb{N}$.
-We can apply \reflem{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$ to bound
+    \end{equation}
+    To match the decay rate of the moments of the first term in \refequ{equ:nystrom-proof-persson-bonud}, we will need to ensure that this term also is of order $\mathcal{O}(\Trace(\mtx{B}) / \sqrt{k})$.
+To this purpose, and to simplify the notation, we choose $k = n_{\mtx{\Omega}}/2$ and $q = n_{\mtx{\Omega}}/4$ for even $n_{\mtx{\Omega}}$ to ensure $k \in \mathbb{N}$.
+We can apply \reflem{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$ and $p = n_{\mtx{\Omega}}/2$ to bound
     \begin{equation}
         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \right]
-        \leq \frac{3 n_{\mtx{\Omega}}}{2} \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
+        \leq n_{\mtx{\Omega}} \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{2}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
         \label{equ:spectral-norm-bound-applied}
@@
     As opposed to \cite[lemma B.3]{tropp-2023-randomized-algorithms}, our bound holds for all even $n_{\mtx{\Omega}} \in \mathbb{N}$, at the cost of a factor of $e^2$.
-Inserting \refequ{equ:spectral-norm-bound-applied} and  \refequ{equ:pinv-spectral-norm-bound} in \refequ{equ:nystrom-proof-processed-tail} gives
-    \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \frac{9 e^4}{2 \sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
+Inserting \refequ{equ:spectral-norm-bound-applied} and  \refequ{equ:pinv-spectral-norm-bound} in \refequ{equ:nystrom-proof-processed-tail} with $k = n_{\mtx{\Omega}} / 2$ gives
+    \begin{equation}
+        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
+        \leq \frac{3 e^4}{\sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{2}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
         %\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
@@
     \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \frac{9 e^4}{2 \sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \left( \frac{6}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) + \frac{1}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) \right)
-        \leq \frac{9e^4}{2 \sqrt{2}} \frac{7}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
+        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
+        \leq \frac{3 e^4}{\sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \left( \frac{6}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) + \frac{2}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) \right)
+        \leq 12 \sqrt{2} e^4 \frac{1}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
         \label{equ:nystrom-proof-tail-bound}
@@
     \end{equation}
-    Inserting \refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud} and using the triangle inequality for $\mathbb{E}^{p}[\cdot]$ we obtain
+    Inserting \refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud} and using the triangle inequality for $\mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}[\cdot]$ we obtain
     \begin{equation}
@@
     \end{equation}
-    where $c = 1 + 63 e^4 / 2 \sqrt{2}$.
+    where $c = 1 + 12 \sqrt{2} e^4 $.
 
@@
 
-    Fixing $\gamma = e$ and letting $n_{\mtx{\Omega}} = (c e \varepsilon)^{-2}  + 4 \log(\delta^{-1})$ gives us the second part of the theorem.
+    Fixing $\gamma = e$ and letting $n_{\mtx{\Omega}} = \lceil(c e \varepsilon)^{-2}  + 4 \log(\delta^{-1}) \rceil$ gives us the second part of the theorem.
 
@@
 
-When compared to the equivalent result for constant matrices \cite[Lemma 3.2]{persson-2022-improved-variants}, our constant factor up front is around four times larger, and our failure probability decays half a power slower.
-
-Finally, we can combine the bound on the Girard-Hutchinson estimator \refthm{thm:hutchinson} and the one on the Nyström approximation \reflem{lem:nystrom} to obtain our main result.
+When compared to the equivalent result for constant matrices \cite[Lemma 3.2]{persson-2022-improved-variants}, our constant factor up front is around three times larger and our failure probability decays half a power slower.
+
+Finally, we can now combine the bound on the Girard-Hutchinson estimator (\refthm{thm:hutchinson}) and the one on the Nyström approximation (\reflem{lem:nystrom}) to obtain our main result.
 
@@
     Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(\delta^{-1})^2)$ with \emph{even} $n_{\mtx{\Omega}}$, then for $\delta \in (0, e^{-\sfrac{1}{2}}/2)$ and $\varepsilon \in (0, 1)$ with probability at least $1 - \delta$
+If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(\delta^{-1})^2)$ with \emph{even} $n_{\mtx{\Omega}}$, then for $\delta \in (0, 1/2)$ and $\varepsilon > 0$ with probability at least $1 - \delta$
     \begin{equation}
@@
         &\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t \notag \\
-        &= \int_{a}^{b} | \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) - \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) |~\mathrm{d}t && \text{(by definition of estimators)} \notag \\
+        &= \int_{a}^{b} | \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) |~\mathrm{d}t && \text{(by definition of estimators)} \notag \\
         &\leq \tilde{\varepsilon} \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F ~\mathrm{d}t && \text{(\refthm{thm:hutchinson} w.p. $\geq 1 - \tilde{\delta}$)} \notag \\
@@
     \end{align}
-    with probability at least $1 - 2\tilde{\delta}$. Taking $\varepsilon = \tilde{\varepsilon}^2$ and $\delta = 2 \tilde{\delta}$
+    with probability at least $1 - 2\tilde{\delta}$ by a union bound. Taking $\varepsilon = \tilde{\varepsilon}^2$ and $\delta = 2 \tilde{\delta}$ we conclude the result.
+
 \end{proof}
@@
 Finding approximations to this distribution is challenging.
-Furthermore, in most applications the exact location of every eigenvalue is not important, but more so their approximate locations relative to each other, such as eigenvalue clusters, spectral gaps, or outliers.
-In the following we will introduce a smoothed version of the spectral density, which is significantly easier to approximate at the cost of losing some of the finer characteristics of the spectrum.
+Furthermore, in most applications the exact location of each individual eigenvalue is not important, but more so their approximate locations relative to each other, such as eigenvalue clusters, spectral gaps, or outliers.
+In the following we will work with a smoothed version of the spectral density, which is significantly easier to approximate at the cost of losing some of the finer characteristics of the spectrum.
 
@@
 \end{equation}
-The task in \refequ{equ:smooth-spectral-density} is to approximate the trace of the parameter-dependent matrix $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
+The task in \refequ{equ:smooth-spectral-density} is to approximate the trace of the parameter-dependent symmetric positive semi-definite matrix $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
 Hence, we can apply the results derived in \refsec{sec:analysis}.
@@
 \begin{equation}
-    \sup_{f \in \mathcal{S}} \int_{-1}^{1} f(t) (\phi(t) - \phi_{\sigma}(t))~\mathrm{d}t,
+    \sup_{f \in \mathcal{S}} \int_{a}^{b} f(t) (\phi(t) - \phi_{\sigma}(t))~\mathrm{d}t,
     \label{equ:error-metric}
@@
 where $\mathcal{S}$ represents an appropriately chosen space of test functions.
-Among others, \cite{lin-2016-approximating-spectral} proposes $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - \lambda), \lambda \in [\lambda_{\text{min}}, \lambda_{\text{max}}]\}$, \cite{chen-2021-analysis-stochastic} uses $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [-1, 1] \}$ where $\Theta$ is the Heaviside step function, and \cite{braverman-2022-sublinear-time} uses $\mathcal{S} = \{f : |f(x) - f(y)| \leq |x - y| \}$ for which the error \refequ{equ:error-metric} is equivalent to the Wasserstein-1 distance between $\phi$ and $\phi_{\sigma}$.
-In this last metric, provided we assume the spectrum to be somewhat uniform and $\sigma$ to be rather small, a simple calculation allows us to roughly estimate the smoothing error \refequ{equ:error-metric} to be $\sigma$.
+Among others, \cite{lin-2016-approximating-spectral} proposes $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - s), s \in [a, b]\}$, \cite{chen-2021-analysis-stochastic} uses $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [a, b] \}$ where $\Theta$ is the Heaviside step function, and \cite{braverman-2022-sublinear-time} uses $\mathcal{S} = \{f : |f(t) - f(s)| \leq |t - s| \}$ for which the error \refequ{equ:error-metric} is equivalent to the Wasserstein-1 distance between $\phi$ and $\phi_{\sigma}$.
+In this last metric, provided we assume the eigenvalues to be somewhat uniformly distributed and $\sigma$ to be rather small, a simple calculation allows us to roughly estimate the smoothing error \refequ{equ:error-metric} to be $\sigma$.
 Consequently, if we can only allow an approximation to deviate by at most a factor of $\varepsilon > 0$ from the original spectral density, we should choose a smoothing parameter $\sigma \approx \varepsilon$.
@@
 However, doing this exactly would require us to first diagonalize $\mtx{A}$, which is often a prohibitively expensive operation.
-For algorithmic reasons, we seek for approximations of this matrix function in terms of affine expansions, of which the Chebyshev expansion is most suitable due to its approximation error guarantees and ease of manipulation.
+For algorithmic reasons (see \refsec{subsubsec:chebyshev-nystrom-implementation}), we seek for approximations of this matrix function in terms of affine expansions \todo{explain more}, of which the Chebyshev expansion was determined to be most suitable due to its approximation error guarantees and ease of manipulation.
 In every $t \in \mathbb{R}$ we expand
@@
     \begin{cases}
-        T_l : [-1, 1] \to \mathbb{R} \\
-        T_l(s) = \cos(l \cdot \arccos(s))
+        T_l : [-1, 1] \to \mathbb{R}; \\
+        T_l(s) = \cos(l \cdot \arccos(s)),
     \end{cases}
@@
 Note that through this transformation we will also have to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when reverting the spectral transform $\tau$.
+From now on, we will assume the eigenvalues of the matrix $\mtx{A}$ to be contained in $[-1, 1]$.
 
@@
 
-It is well known that the coefficients of a Chebyshev expansion of a function can -- in exact arithmetic -- be computed with a discrete cosine transform (DCT) of the function evaluations at the Chebyshev nodes $s_i = \cos(\pi i / m)$ for $i=0,\dots,m$
-\begin{equation}
-    g_{\sigma}(t - \cos(s_i)) = \sum_{l=0}^{m} \mu_l(t) \cos\left(\frac{\pi i l}{m} \right).
+It is well known that the coefficients of a Chebyshev expansion of a function can -- in exact arithmetic -- be computed by inverting the discrete cosine transform (DCT)
+\begin{equation}
+    g_{\sigma}(t - \cos(s_i)) = \sum_{l=0}^{m} \mu_l(t) \cos\left(\frac{\pi i l}{m} \right)
     \label{equ:discrete-cosine-transform}
 \end{equation}
+between the expansion coefficients $\mu_l(t),~i=0,\dots,m$ and the function evaluations at the Chebyshev nodes $s_i = \cos(\pi i / m),~i=0,\dots,m$.
 This can be accomplished in $\mathcal{O}(m \log(m))$ operations.
@@
     \State Compute $\vct{f} = \DCT(\vct{\mu})$
-    \State Compute $\vct{\nu} = \DCT^{-1}(\vct{f}^{2})$ whose entries are the coefficients $\nu_l$
+    \State Compute $\vct{\nu} = \DCT^{-1}(\vct{f} \odot \vct{f})$ whose entries are the coefficients $\nu_l$
+    \label{lin:inverse-DCT}
     \end{algorithmic}
 \end{algo}
+On \reflin{lin:inverse-DCT}, $\odot$ denotes the entrywise product.
 
@@
 
-We now analyze the error of the Chebyshev expansion of a Gaussian smoothing kernel $g_{\sigma}$.
+We now analyze what error we make when using the expanded spectral density $\phi_{\sigma}^{(m)}$ to approximate the spectral density $\phi_{\sigma}$.
+To this extent, we start with the error of the Chebyshev expansion for a Gaussian smoothing kernel $g_{\sigma}$.
 
@@
 The proof specializes Bernstein's theorem for Chebyshev interpolation to the case of a Gaussian smoothing kernel \refequ{equ:smoothing-kernel}.
-It loosely follows the proof give in the preprint of \cite[theorem 2]{lin-2017-randomized-estimation}, but attains a slightly different dependence on $\sigma$.
+It loosely follows the proof given in the arXiv preprint of \cite[theorem 2]{lin-2017-randomized-estimation}, but attains a slightly different dependence on $\sigma$.
 
@@
     \end{equation}
-    Because $e^{-\sfrac{(t - x)^2}{2\sigma^2}} \leq 1$ for all $x, t \in \mathbb{R}$, and the maximum absolute value of $y$ is limited by the length of the semi-axis of the Bernstein ellipse in the direction of the imaginary axis, which is $(\chi - \chi^{-1}) / 2$, we upper bound
+    Because $e^{-\sfrac{(t - x)^2}{2\sigma^2}} \leq 1$ for all $x, t \in \mathbb{R}$, and the maximum absolute value of $y$ is limited by the length of the semi-axis of the Bernstein ellipse in the direction of the imaginary axis, $(\chi - \chi^{-1}) / 2$, we upper bound
     \begin{equation}
@@
 
-Our goal is to apply the Nyström++ trace estimator which we analyzed in \refsec{subsec:nystrom-pp} to the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ to approximate the smoothed spectral density $\phi_{\sigma}$.
+Our goal is to apply the Nyström++ trace estimator which we have analyzed in \refsec{subsec:nystrom-pp} to the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ to approximate the smoothed spectral density $\phi_{\sigma}$.
 However, due to the oscillatory nature of the Chebyshev expansion, we can no longer guarantee that $g_{\sigma}^{(m)} \geq 0$, and consequently, the matrix function $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ may be (slightly) indefinite.
@@
 To still apply the bound from \refthm{thm:nystrom-pp} in this scenario, we will have to ensure that the Chebyshev expansion is non-negative.
-We can think of many ways in which this can be achieved: by using Jackson damping \cite{jackson-1912-approximation-trigonometric,braverman-2022-sublinear-time}, albeit at a significantly worse rate of convergence; by shifting up the expansion $g_{\sigma}^{(m)} + \rho$ with a large enough $\rho > 0$, at the loss of most of the low-rank structure of $g_{\sigma}(t \mtx{I}_n - \mtx{A})$; or by directly minimizing the approximation error over all non-negative Chebyshev polynomials \cite{fejer-1916-uber-trigonometrische}, which realistically only works for small $m$.
+We can think of many ways in which this can be achieved: by using Jackson damping \cite{jackson-1912-approximation-trigonometric,braverman-2022-sublinear-time}, albeit at a significantly worse rate of convergence; by shifting up the expansion $g_{\sigma}^{(m)} + \rho$ with a large enough $\rho > 0$, at the loss of some of the low-rank structure of $g_{\sigma}(t \mtx{I}_n - \mtx{A})$; or by directly minimizing the approximation error over all non-negative Chebyshev polynomials \cite{fejer-1916-uber-trigonometrische}, which realistically only works for small $m$.
 In the end, we propose the following method to produce a non-negative Chebyshev expansion: First, expand the square root of the smoothing kernel in a basis of Chebyshev polynomials up to degree $m/2$
@@
 \begin{table}[ht]
-    \caption{We compare the runtime in milliseconds for computing the coefficients of the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ \refequ{equ:smoothing-kernel} in three different ways: as proposed in \cite[algorithm 1]{lin-2017-randomized-estimation} (\emph{oversampled FFT}), using the discrete cosine transform (\emph{DCT}), and computing a non-negative Chebyshev expansion with \refalg{alg:chebyshev-squaring} (\emph{non-negative DCT}).
+    \caption{We compare the runtime in milliseconds for computing the coefficients of the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ \refequ{equ:smoothing-kernel} in three different ways: as proposed in \cite[algorithm 1]{lin-2017-randomized-estimation} (\emph{oversampled FFT}), using the discrete cosine transform (\emph{DCT}) \refequ{equ:discrete-cosine-transform}, and computing a non-negative Chebyshev expansion with \refalg{alg:chebyshev-squaring} (\emph{non-negative DCT}).
 We use $\sigma=0.005$, $n_t=1000$ parameter values, and various degrees $m$.
@@
 \begin{equation}
-    \mtx{\Omega}^{\top} \underline{g}^{(m)}_{\sigma}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega} = \sum_{l=0}^{m} \mu_l(t) (\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega})
-\end{equation}
-for which we can first precompute the small $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ for $l=0, \dots, m$ using the recurrence \refequ{equ:chebyshev-recurrence} and then inexpensively sum them up with the corresponding coefficients $\mu_l(t)$ for all $t$ in which we want to evaluate the approximation.
+    \mtx{\Omega}^{\top} \underline{g}^{(m)}_{\sigma}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega} = \sum_{l=0}^{m} \mu_l(t) (\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}),
+\end{equation}
+meaning we can first precompute the small $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ for $l=0, \dots, m$ using the Chebyshev recurrence \refequ{equ:chebyshev-recurrence} and then inexpensively sum them up with the corresponding coefficients $\mu_l(t)$ for all $t$ in which we want to evaluate the approximation.
 Thanks to the fast and exact squaring of Chebyshev expansions (cf.
@@
 
-The dominating cost stems from the $(2m + 1)(n_{\mtx{\Psi}} + n_{\mtx{\Omega}})$ matrix-vector multiplications with $\mtx{A}$ and the cost associated with computing the estimate in each evaluation point $t_1, \dots, t_{n_t}$ given by $\mathcal{O}(m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + (n_{\mtx{\Omega}}^3 + n_{\mtx{\Omega}} n_{\mtx{\Psi}}^2 + 1))$.
+The dominating cost in \refalg{alg:nystrom-chebyshev-pp} stems from the $(2m + 1)(n_{\mtx{\Psi}} + n_{\mtx{\Omega}})$ matrix-vector multiplications with $\mtx{A}$ and the cost associated with computing the estimate in each evaluation point $t_1, \dots, t_{n_t}$ given by $\mathcal{O}(m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + (n_{\mtx{\Omega}}^3 + n_{\mtx{\Omega}} n_{\mtx{\Psi}}^2 + 1))$.
 
@@
 regions where the spectral density nearly vanishes.
-For small smoothing parameters $\sigma$, the rapid decay of the Gaussian $g_{\sigma}$ will cause $\mtx{K}(t_i) \approx g_{\sigma}(t_i\mtx{I}_n - \mtx{A})$ to be close to the zero matrix for parameter values $t_i$ located in such regions.
-Hence, on line \reflin{lin:4-nystromchebyshev-nystrom-pp} we would compute the pseudoinverse of an almost zero matrix, which can be extremely unstable (see \reffig{fig:zerocheck}).
-Notice that $\Trace(\mtx{K}_1(t_i)) / n_{\mtx{\Omega}}$ is the $n_{\mtx{\Omega}}$-query Girard-Hutchinson estimator.
+For small smoothing parameters $\sigma$, the rapid decay of the Gaussian $g_{\sigma}$ will cause $\mtx{K}_1(t_i) \approx \mtx{\Omega}^{\top} g_{\sigma}(t_i\mtx{I}_n - \mtx{A}) \mtx{\Omega}$ to be close to the zero matrix for parameter values $t_i$ located in such regions.
+Hence, on \reflin{lin:4-nystromchebyshev-nystrom-pp} we would compute the pseudoinverse of an almost zero matrix, which can be extremely unstable (see \reffig{fig:zerocheck}).
+Notice that with $\mtx{K}_1(t_i)$ -- a matrix we need to assemble anyway -- we can cheaply compute $\Trace(\mtx{K}_1(t_i)) / n_{\mtx{\Omega}}$, which is the $n_{\mtx{\Omega}}$-query Girard-Hutchinson estimator.
 We may use it as a rough indicator whether $t_i$ lies within a region where the spectral density nearly vanishes, and subsequently set $\widetilde{\phi}_{\sigma}(t_i) = 0$ instead of computing the full expression on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
@@
     from which follows the result with some minor simplifications.
+
 \end{proof}
@@
 On the other hand, we approximate $\sqrt{g_{\sigma}}$ instead of $g_{\sigma}$, which is less \enquote{pointy} and therefore better approximated by polynomials.
-Overall, we will need to choose the degree $m$ of the expansion approximately a factor of $\sqrt{2}$ larger to match the error of the standard expansion.
+Overall, we will need to choose the degree $m$ of the expansion approximately a factor of $\sqrt{2}$ larger to match the error of the standard expansion (cf.
+\reffig{fig:interpolation-issue}).
 
@@
     Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, \emph{even} $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2)$, and smoothing parameter $0 < \sigma \ll 1$ satisfies with probability at least $1 - \delta$
+Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, \emph{even} $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(\delta^{-1})^2)$, and smoothing parameter $0 < \sigma \ll 1$ satisfies with probability at least $1 - \delta$
     \begin{equation}
@@
     \end{subfigure}
-    \caption{Cross-sections of the periodic Gaussian well potential $V$ for different sizes of the supercell.}
+    \caption{Cross-sections of the periodic Gaussian well potential $V$ for different sizes $n_c$ of the supercell.}
     \label{fig:gaussian-well}
@@
 
-\todo{Explain experimental setup.}
+While theoretically an important tool, we waive using the non-negative Chebyshev expansion, since the slight indefiniteness of $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ \refequ{equ:matrix-expansion} does not make a difference in finite precision arithmetic.
+We approximate the integrals for computing the $L^1$-errors in \refsec{sec:analysis} using a midpoint quadrature with $n_t \in \mathbb{N}$ nodes.
+For a fixed total number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$, we analyze the convergence behavior of the Chebyshev-Nyström++ method when run on the $1000 \times 1000$ matrix resulting from the finite difference discretization of the Hamiltonian for $n_c = 1$ described above (see \reffig{fig:convergence}).
 
@@
     \input{plots/convergence.pgf}
-    \caption{For increasing values of $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}$ but fixed $m$ we plot the $L^1$-approximation error for $\sigma=0.005$ on the model problem with $n_c = 1$.}
+    \caption{For increasing values of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ but fixed $m$ we plot the $L^1$-approximation error for $\sigma=0.005$ on the model problem with $n_c = 1$.
+The $L^1$-error is approximated by a quadrature with $n_t = 100$ evenly spaced nodes.}
     \label{fig:convergence}
 \end{figure}
+
+Indeed, we observe that for $n_{\mtx{\Omega}} = 0$, we require $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ to achieve an error of order $\varepsilon$, as shown in \refthm{thm:hutchinson}.
+Because the eigenvalues of the matrix are quite uniformly distributed, we observe the stronger convergence of the Nyström approximation discussed in the end of \refsec{sec:application} once $n_{\mtx{\Omega}}$ exceeds the numerical rank \refequ{equ:gaussian-kernel-numerical-rank-uniform}, which in this case would be somewhat larger than $30$.
 
@@
     \input{plots/distribution.pgf}
-    \caption{The Chebyshev-Nyström++ method for different ways of allocations a total of $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation for the Gaussian smoothing kernel with multiple different values of the smoothing parameter.
+    \caption{The Chebyshev-Nyström++ method for different ways of allocations a total of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation for the Gaussian smoothing kernel with multiple different values of the smoothing parameter.
 We make the approximation error made in the Chebyshev expansion negligible by rescaling $m=16 / \sigma$ (based on \refthm{thm:chebyshev-error}).}
@@
 \todo{Interpret results.}
+
+\todo{Hessian spectrum analysis?}
 

### 2024-07-01 paper/plots/convergence.py

@@ -50,7 +50,7 @@ plt.plot(n_Vec_list, 0.45/n_Vec_list**(0.5), linestyle="dashed", color="#7a7a7a"
 plt.text(6e+1, 9.5e-2, r"$\mathcal{O}(\varepsilon^{-2})$", color="#7a7a7a")
 
 plt.grid(True, which="both")
-plt.ylabel(r"$L^1$ error")
+plt.ylabel(r"$L^1$-error")
 plt.xlabel(r"estimator size $n_{\mathbf{\Psi}} + n_{\mathbf{\Omega}}$")
 plt.legend()
 plt.xscale("log")

### 2024-07-01 paper/plots/distribution.py

@@ -45,7 +45,7 @@ for i in reversed(range(len(n_Psi_list))):
     plt.plot(sigma_list, error[i], color=colors[i], marker=markers[i], label=labels[i])
 
 plt.grid(True, which="both")
-plt.ylabel(r"$L^1$ error")
+plt.ylabel(r"$L^1$-error")
 plt.xlabel(r"smoothing parameter $\sigma$")
 plt.legend()
 plt.xscale("log")

### 2024-07-01 paper/plots/interpolation.py

@@ -44,7 +44,7 @@ for i in range(3):
     plt.plot(m_list, error[i], color=colors[i], marker=markers[i], label=labels[i])
 
 plt.grid(True, which="both")
-plt.ylabel(r"$L^1$ error")
+plt.ylabel(r"$L^1$-error")
 plt.xlabel(r"expansion degree $m$")
 plt.legend()
 plt.xscale("log")

### 2024-07-01 paper/plots/zerocheck.py

@@ -18,7 +18,7 @@ eigvals_st = spectral_transformation(eigvals, min_ev, max_ev)
 
 # Set parameter
 t = np.linspace(-1, 1, 500)
-sigma = 0.005
+sigma = 0.004
 m = 2000
 n_Omega = 80
 

### 2024-07-01 paper/re-pro-badge.tex
automatically reproduce paper

@@
-This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-TRACE/tree/170f339}{170f339} at \url{https://github.com/FMatti/Rand-TRACE} on 2024-06-28 at 12:22:33 UTC.
+This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-TRACE/tree/b1d98c9}{b1d98c9} at \url{https://github.com/FMatti/Rand-TRACE} on 2024-07-01 at 10:20:08 UTC.

### 2024-07-05 algorithms/hessian_density.py
implement hessian training and spectral transform

@@ -6,6 +6,9 @@ Spectral density of Hessian matrix.
 """
 
 import torch
+import time
+import numpy as np
+from .krylov_aware import lanczos
 
 def compute_jacobian(model, loss_function, data):
     input, target = data
@@ -20,17 +23,28 @@ def hessian_vector_product(jacobian, params, X):
     return torch.autograd.grad(jacobian, params, X, retain_graph=True, is_grads_batched=True if X.ndim > 1 else False)
 
 class hessian():
-    def __init__(self, model, loss_function, data):
+    def __init__(self, model, loss_function, data, spectral_transform=True):
         self.parameters = list(model.parameters())
+        self.num_parameters = sum(p.numel() for p in model.parameters())
+        self.shape = (self.num_parameters, self.num_parameters)
         self.jacobian = compute_jacobian(model, loss_function, data)
+        self.scaling_parameter = 1.0
+        if spectral_transform:
+            k = 20
+            x = torch.randn(self.num_parameters)
+            Q, T = lanczos(self, x, k, dtype=np.float64)
+            self.scaling_parameter = np.max(np.linalg.eigvalsh(T[0, :-1, :])) + np.linalg.norm(T[0, -1, -1] * Q[:, :, -1])
 
     def __matmul__(self, X):
+        is_numpy = isinstance(X, np.ndarray)
+        if not torch.is_tensor(X):
+            X = torch.from_numpy(X)
         hvp = hessian_vector_product(self.jacobian, self.parameters, X.T)
         hvp = torch.cat([e.reshape(X.shape[-1], -1) for e in hvp], dim=1).T
-        return hvp
+        if is_numpy:
+            hvp = hvp.numpy()
+        return hvp / self.scaling_parameter
 
-import time
-import torch
 
 def train(model, data_loader, loss_function=torch.nn.MSELoss(), optimizer=torch.optim.SGD, optimizer_parameters={"lr": 0.01}, n_epochs=10, verbose=True, validator=None):
     """
@@ -61,13 +75,13 @@ def train(model, data_loader, loss_function=torch.nn.MSELoss(), optimizer=torch.
     for epoch in range(n_epochs):
         epoch_loss = 0
         start_time = time.time()
-        for i, (X, y) in enumerate(data_loader):
+        for batch in data_loader:
             # Reset gradients to zero
             optimizer.zero_grad()
 
             # Compute prediction error
-            outputs = model(X)
[…]

### 2024-07-05 algorithms/krylov_aware.py

@@ -60,8 +60,7 @@ def lanczos(A, X, k, reorth_tol : float = 0.7, return_matrix : bool = True, exte
         X = X[:, np.newaxis]
     n, m = X.shape
 
-    # Convert A to a function handle in case it is an array/matrix
-    matvec = (lambda x: A @ x) if isinstance(A, np.ndarray | sp.sparse.spmatrix) else A
+    # Determine the range of A in case it is not an array/matrix
     dtype = A.dtype if dtype is None else dtype
 
     # Initialize arrays for storing the block-tridiagonal elements
@@ -77,7 +76,7 @@ def lanczos(A, X, k, reorth_tol : float = 0.7, return_matrix : bool = True, exte
     for j in range(k):
 
         # Generate and orthogonalize next element(s) in Krylov-basis
-        w = matvec(U[j])
+        w = A @ U[j]
         a[j] = np.sum(U[j].conj() * w, axis=0)
         u_tilde = w - U[j] * a[j] - (U[j-1] * b[j] if j > 0 else 0)
 
@@ -160,8 +159,7 @@ def block_lanczos(A, X, k, reorth_steps : int = -1, return_matrix : bool = True,
         X = X[:, np.newaxis]
     n, m = X.shape
 
-    # Convert A to a function handle in case it is an array/matrix
-    matvec = (lambda x: A @ x) if isinstance(A, np.ndarray | sp.sparse.spmatrix) else A
+    # Determine the range of A in case it is not an array/matrix
     dtype = A.dtype if dtype is None else dtype
 
     # Initialize arrays for storing the block-tridiagonal elements
@@ -172,11 +170,10 @@ def block_lanczos(A, X, k, reorth_steps : int = -1, return_matrix : bool = True,
     # Specify initial iterate
     U[0], b[0] = np.linalg.qr(X)
 
-
     # Perform k iterations of the Lanczos algorithm
     for j in range(k):
 
-        w = matvec(U[j])
+        w = A @ U[j]
         a[j] = U[j].conj().T @ w
         u_tilde = w - U[j] @ a[j] - (U[j-1] @ b[j].conj().T if j > 0 else 0)
 

### 2024-07-05 paper/re-pro-badge.tex

@@
-This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-TRACE/tree/b1d98c9}{b1d98c9} at \url{https://github.com/FMatti/Rand-TRACE} on 2024-07-01 at 10:20:08 UTC.
+This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-TRACE/tree/a3299bf}{a3299bf} at \url{https://github.com/FMatti/Rand-TRACE} on 2024-07-04 at 14:51:50 UTC.

### 2024-07-05 paper/re-pro-badge.tex
automatically reproduce paper

@@
-This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-TRACE/tree/a3299bf}{a3299bf} at \url{https://github.com/FMatti/Rand-TRACE} on 2024-07-04 at 14:51:50 UTC.
+This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-TRACE/tree/6f952d3}{6f952d3} at \url{https://github.com/FMatti/Rand-TRACE} on 2024-07-05 at 12:56:53 UTC.

### 2024-07-07 paper/intro.tex
Updates from Overleaf

\section{Introduction}
\label{sec:introduction}

Spectral distributions can reveal important properties in numerous applications across physics, chemistry, engineering, and data science: For example, in electronic structure calculations eigenvalues represent the energy levels which electrons occupy\cite{drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure, lin-2017-randomized-estimation}, in machine learning they are indicative of the topology of the loss landscape \cite{ghorbani-2019-investigation-neural, yao-2020-pyhessian-neural}, and in graph processing they can uncover hidden graph motifs~\cite{huang-2021-density-states}. Given a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$, the distribution of its eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$, which places a rescaled Dirac delta distribution $\delta$ at each eigenvalue. Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation that is often prohibitively expensive. Several techniques have been developed for approximating $\phi$, including moment matching~\cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time} and polynomial approximation, either implicitly by a Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or explicitly by expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}.

Being composed of distributions, the function $\phi$ itself is hard to approximate. However, as many application only require a rough estimate of $\phi$, it usually suffices to instead approximate a smoothed spectral density of the form
\begin{equation}
    \phi_{\sigma}(t) = \frac{1}{n} \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i) = \frac{1}{n} \Trace(g_{\sigma}(t \mtx{I}_n - \mtx{A})),
    \label{equ:smoothed-spectral-density}
\end{equation}
for some fixed smoothing kernel $g_{\sigma}$, typically a Gaussian~\cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian~\cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}. We thus arrive at the problem of computing the trace of the parameter-dependent matrix function $\mtx{B}(t) \equiv g_{\sigma}(t \mtx{I}_n - \mtx{A}) \in \mathbb{R}^{n \times n}$. Evaluating this matrix function \emph{exactly} requires the diagonalization of $\mtx{A}$, which does not seem to yield any gains compared to the original problem of computing its spectral distribution. The crucial point is that we can now \emph{estimate} the trace by matrix-vector products with $\mtx{B}(t)$, which -- thanks to the smoothness of $g_\sigma$ -- can be well approximated by a few matrix-vector products with $A$.

For a \emph{constant} matrix $B$, one of the most popular trace estimators is the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} along with variance reduction techniques~\cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, chen-2023-krylovaware-stochastic, epperly-2024-xtrace-making}. Suitable extensions to parameter-dependent matrices have been considered, e.g., in~\cite{lin-2017-randomized-estimation,chen-2023-krylovaware-stochastic}, but we are not aware of an analysis providing insight and justification of these extensions. In passing, we note that dynamic trace estimation~\cite{dharangutte-2021-dynamic-trace,woodruff-2022-optimal-query} is an efficient technique for \textcolor{red}{subsequently ??? true? ???} estimating the traces of matrices $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ for wich the increments $\mtx{B}(t_{i+1}) - \mtx{B}(t_i)$ are relatively small in norm. The potential of dynamic trace estimation appears to be limited in our setting because $\mtx{B}(t)$ may change rapidly close to eigenvalues, with 
$g_\sigma$ approximating a Dirac delta function.

All methods considered in this work are based on the following simple idea: Apply an existing randomized trace estimator to $\Trace(\mtx{B}(t))$ with \emph{constant} random vectors, that is, the same randomization is used for each value of the parameter $t$. For example, the Girard-Hutchinson estimator ??? STOP HERE ???

\textcolor{red}{??? The following two paragraphs need to rewritten a bit at a later stage. It needs to be clearer what the approaches proposed in our paper are really about and the novelty wrt existing work has to come out more clearly and concretely. ???}
\color{black}
Instead of discretizing $[a, b]$ into $t_1, \ldots, t_m$, approximating $\Trace(\mtx{B}(t_i))$ with $b(t_i)$, and analyzing the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way is to quantify the error with $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate in each $t \in [a, b]$. This approach has already been used in the analysis of low-rank approximations of parameter-dependent matrices \cite{kressner-2023-randomized-lowrank}.

\paragraph{New contributions} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}. Interestingly, we can reuse the same randomization for computing the estimate at ea<ch value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations. We propose certain modifications to the methods from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}. In particular, we use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials and unify all of their methods in a single algorithm, with the desirable side effect of speeding up their \enquote{spectrum sweeping} method by orders of magnitude. Based on the analysis of the randomized trace estimators for parameter-dependent matrices, we provide bounds on the approximation error for our method.

\paragraph{Structure} In \refsec{sec:analysis} we will analyze three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter dependent matrices. We propose and analyze an algorithm for using the aforementioned estimators to approximate the spectral density of real symmetric matrices in \refsec{sec:application}. Finally, we will test our developments on an example from electronic structure calculation in \refsec{sec:results}.

\paragraph{Reproducibility} \textcolor{red}{This paragraph looks odd. Is this the wording required by SISC? If not, focus on the reproducibility of the numerical experiments and figures/tables in this statement.}  \input{re-pro-badge.tex}

### 2024-07-07 paper/paper.tex
Updates from Overleaf

--- 2024-06-27 paper/main.tex
@@
 
-\todo{
 \begin{abstract}
-    Write abstract.
+    Stochastic trace estimators are a family of well established techniques for approximating the trace of constant matrices.
+Oftentimes the matrix whose trace we aim to approximate continuously depends on a real parameter $t \in [a,b]$.
+This is for example the case when approximating the spectral density of a matrix or computing partition functions of quantum systems.
+Clearly, we could separately approximate the traces of the constant matrices $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ at the fixed values $t_1, \dots, t_m \in [a, b]$.
+However, this approach does not scale well if $m$ is large.
+We analyze three trace estimators for constant matrices -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter-dependent matrices with the same randomization reused for all values of the parameter $t$.
+We show that all of these estimators closely match their corresponding error guarantees for constant matrices.
+In particular, for the Nyström++ method, we can guarantee an error of at most $\varepsilon$ by computing the estimate with just $\mathcal{O}(\varepsilon^{-1})$ random vectors, independent of what structure the symmetric positive semi-definite matrix $\mtx{B}(t)$ has.
+We employ these trace estimators to improve upon a technique for approximating the spectral density of a matrix and provide a detailed analysis of the resulting method.
+Finally, we confirm our theoretical developments in multiple numerical experiments.
 \end{abstract}
+
+\textcolor{red}{TODO for Fabio:
+\begin{itemize}
+ \item Capitalize references to algorithms, lemma etc (that is, Lemma 3.2 instead of lemma 3.2 [Lin17, Algorithm 1] instead of [Lin17, algorithm 1]), etc.
+ \item Add full stops after paragraph titles.
+ \item It is not great that the references occupy so much space (more than 4 pages); consider using a more condensed style. The paper needs to get strictly below 30 pages prior to submission. 
+ \item Please include Acknowledgements as indicated in the comment.
+\end{itemize}
 }
+\color{blue}
+\input{intro.tex}
+\input{analysis.tex}
+\input{spectraldensity.tex}
+\input{numresults.tex}
 
-\section{Introduction}
-\label{sec:introduction}
-
-In numerous problems across physics, chemistry, engineering, and data science, how the eigenvalues of certain matrices are distributed can be indicative of what properties a system has: In electronic structure calculations eigenvalues represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}.
-The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$ with the Dirac delta distribution $\delta$.
-Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation which is often prohibitively expensive.
-Therefore, many techniques have been proposed for approximating $\phi$.
-For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature rule with the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}.
-Further, because $\delta$ is hard to approximate and in many applications, a rough estimate of the spectral density is already acceptable, a standard approach is to instead approximate the smoothed spectral density
-\begin{equation}
-    \phi_{\sigma}(t) = \frac{1}{n} \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i) = \frac{1}{n} \Trace(g_{\sigma}(t \mtx{I}_n - \mtx{A}))
-    \label{equ:smoothed-spectral-density}
-\end{equation}
-for a smoothing kernel $g_{\sigma}$, typically a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}.
-This translates the problem to computing the trace of a continuous parameter-dependent matrix $\mtx{B}(t) \equiv g_{\sigma}(t \mtx{I}_n - \mtx{A}) \in \mathbb{R}^{n \times n}$.
-Directly assembling the matrix function requires diagonalizing the matrix $\mtx{A}$, which again is prohibitively expensive.
-Therefore, we estimate its trace implicitly by approximating products of the matrix function with vectors.
-For constant matrices, the most widespread methods for implicit trace estimation are the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} and the variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making, chen-2023-krylovaware-stochastic}.
-Some of these estimators have already been successfully applied to parameter-dependent matrices \cite{lin-2017-randomized-estimation,chen-2023-krylovaware-stochastic}, but they have not yet been analyzed in this setting.
-A related technique is dynamic trace estimation for sequences $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ of matrices which satisfy $\lVert \mtx{B}(t_i) \rVert _F \leq 1$ and which change dynamically, i.e.
-$\lVert \mtx{B}(t_{i+1}) - \mtx{B}(t_i) \rVert _F \leq \alpha < 1$ for the Frobenius norm $\lVert \cdot \rVert _F$ and a small $\alpha$ \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace}.
-However, in our setting we often cannot guarantee either of the two conditions.
-Instead of discretizing $[a, b]$ into $t_1, \ldots, t_m$, approximating $\Trace(\mtx{B}(t_i))$ with $b(t_i)$, and analyzing the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way is to quantify the error with $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate in each $t \in [a, b]$.
-This approach has already been taken for analyzing low-rank approximations of parameter-dependent matrices \cite{kressner-2023-randomized-lowrank}.
-
-\paragraph{New contributions} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}.
-Interestingly, we can reuse the same randomization for computing the estimate at each value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations.
-We propose certain modifications to the methods from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}.
-In particular, we use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials and unify all of their methods in a single algorithm, with the desirable side effect of speeding up their \enquote{spectrum sweeping} method by orders of magnitude.
-Based on the analysis of the randomized trace estimators for parameter-dependent matrices we provide bounds on the approximation error for our method.
-
-\paragraph{Structure} In \refsec{sec:analysis} we will analyze three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter dependent matrices.
-We propose and analyze a method for using the aforementioned estimators to estimate the spectral density of real symmetric matrices in \refsec{sec:application}.
-Finally, we will test our developments on an example from electronic structure calculation.
-
-\paragraph{Reproducibility} \input{re-pro-badge.tex}
-
-\section{Analysis of randomized trace estimators applied to parameter-dependent matrices}
-\label{sec:analysis}
-
-We consider parameter-dependent matrices of the form
-\begin{equation}
-    \mtx{B}(t) = \begin{bmatrix}
-        b_{11}(t) & b_{12}(t) & \dots & b_{1n}(t) \\
-        b_{21}(t) & b_{22}(t) & \dots & b_{2n}(t) \\
-        \vdots & \vdots & \ddots & \vdots \\
-        b_{n1}(t) & b_{n2}(t) & \dots & b_{nn}(t) \\
-    \end{bmatrix} \in \mathbb{R}^{n \times n}
-\end{equation}
-where $b_{ij}(t)$ are functions depending continuously on the parameter $t$ which takes values in the interval $[a,b]$. The trace of such a matrix is defined as
-\begin{equation}
-    \Trace(\mtx{B}(t)) = \sum_{i=1}^{n} b_{ii}(t).
-    \label{equ:definition-trace}
-\end{equation}
-However, we assume that we only have access to products of this matrix with vectors for each $t \in [a, b]$, so this definition will not be directly useful for computing the trace.
-
-\paragraph{Girard-Hutchinson estimator} We can approximate the trace with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ stochastically independent standard Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ to form
-\begin{equation}
-    \Hutch{\mtx{\Psi}}(\mtx{B}(t))
-    = \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j
-    = \frac{1}{n_{\mtx{\Psi}}} \Trace( \mtx{\Psi}^{\top} \mtx{B}(t) \mtx{\Psi})
-    \label{equ:hutchinson-trace-estimator}
-\end{equation}
-where $\mtx{\Psi} = [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\mtx{\Psi}}}] \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$.
-Other choices for the distribution of the random vectors are possible, for example by uniformly sampling from $\{-1, +1\}$ or from the $(n-1)$-sphere.
-However, our theoretical developments only hold in the Gaussian case.
-
-\paragraph{Nyström estimator} Alternatively, the trace of a symmetric matrix whose singular values decay quickly can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström approximation \cite{gittens-2013-revisiting-nystrom}
-\begin{equation}
-    \Nystr{\mtx{\Omega}}{\mtx{B}}(t) = (\mtx{B}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} (\mtx{B}(t) \mtx{\Omega})^{\top}.
-    \label{equ:nystrom-approximation}
-\end{equation}
-Then we can estimate the trace as $\Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t))$.
-Thanks to the invariance of the trace under cyclic permutation of its arguments and the symmetry of the matrix, we may rewrite this estimator as
-
-\begin{equation}
-    \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{B}(t)^2 \mtx{\Omega})).
-    \label{equ:nystrom-trace-estimator}
-\end{equation}
-
-\paragraph{Nyström++ estimator} Finally, \cite{lin-2017-randomized-estimation} proposes an estimator which corrects for inaccuracies in the Nyström approximation \refequ{equ:nystrom-trace-estimator} by estimating the trace of its residual using the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}
-\begin{equation}
-    \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) = \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) + \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)).
-    \label{equ:nystrompp-trace-estimator}
-\end{equation}
-This is the parameter-dependent analogue of the Nyström++ estimator \cite{persson-2022-improved-variants}, which is based on the Hutch++ estimator \cite{meyer-2021-hutch-optimal}.
-%We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.
-
-%In the remainder of this section, we will derive upper bounds on the error of these estimators.
-%
-%\begin{table}[ht]
-%\centering
-%\renewcommand{\arraystretch}{1.2}
-%\begin{tabular}{@{}lcc@{}}
-%\toprule
-%Estimator & Matrix & $m=1600$\\
-%\midrule
-%Girard-Hutchinson & symmetric & $\mathcal{O}(\varepsilon^{-2})%$ \\
-%Nyström & symmetric positive semi-definite & $\mathcal{O}%(\varepsilon^{-2})$ \\
-%Nyström++ & symmetric positive semi-definite & $\mathcal{O}%(\varepsilon^{-1})$  \\
-%\bottomrule
-%\end{tabular}
-%\end{table}
-
-\subsection{Girard-Hutchinson estimator for parameter-dependent matrices}
-\label{subsec:hutchinson}
-
-We first analyze the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}.
-We bound its $L^1$ norm in terms of the integral over the Frobenius norm $\lVert \cdot \rVert _F$ of the matrix whose trace we aim to approximate.
-
-\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
-    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric and continuously depend on $t \in [a, b]$.
-Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ it holds with probability at least $1 - \gamma^{-2k}$
-    \begin{equation}
-        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) \right| ~\mathrm{d}t \leq 2^4 \gamma k \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F  ~\mathrm{d}t.
-    \end{equation}
-    In particular, if we choose $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{- \sfrac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$.
-\end{theorem}
-
-The idea behind the proof is to bound the higher order moments of the residual $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ for fixed values of $t$ and then use Minkowski's integral inequality to extend the results to the continuous parameter case.
-We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
-\begin{proof} 
-    We define the residual of the 1-query Girard-Hutchinson estimate 
-    \begin{equation}
-        r(\mtx{B}(t), \vct{\psi}) = \Trace(\mtx{B}(t)) - \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi}
-    \end{equation}
-    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$.
-
-    First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence.
-From the proof of \cite[lemma 3]{cortinovis-2022-randomized-trace} we know that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm.
-Thus, by \cite[theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
-    \begin{align}
-        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
-        &\leq k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k} \notag \\
-        &= k! 2^{4 k} \lVert \mtx{B} \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _2^{2 k}.
-    \end{align}
-    Since $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
-    \begin{equation}
-        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
-    \end{equation}
-    Stirling's approximation \cite{robbins-1955-remark-stirling} bonuds $(2 k)!
-< 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$.
-Consequently, the moments $\mathbb{E}^{k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{k} \right] \right)^{\sfrac{1}{k}}$ of $r(\mtx{B}, \vct{\psi})$ are limited by
-    \begin{equation}
-        \mathbb{E}^{2k}\left[ r(\mtx{B}, \vct{\psi}) \right]
-        \leq \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F \leq 2^4 k \lVert \mtx{B} \rVert _F,
-        \label{equ:hutchinson-trace-onequery-fixed}
-    \end{equation}
-    since it can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, due to the monotonous decrease of this expression in $k \in \mathbb{N}$.
-
-    Now we make the transition to the continuous.
-Following \cite{kressner-2023-randomized-lowrank}, we can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable.
-Therefore, Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting:
-    \begin{equation}
-        \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t  \right]
-        \leq \int_{a}^{b} \mathbb{E}^{2 k}\left[ r(\mtx{B}(t), \vct{\psi}) \right]~\mathrm{d}t
-        \leq 2^4 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
-    \end{equation}
-    Consequently, by Markov's inequality, for any $k \in \mathbb{N}$ and $\gamma \geq 1$, it holds with probability at least $1 - \gamma^{-2 k}$
-    \begin{equation}
-        \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t \leq 2^4 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
-        \label{equ:hutchinson-trace-onequery-uniform}
-    \end{equation}
-
-    To extend the bound to the $n_{\mtx{\Psi}}$-query quadratic trace estimator, we use a technique from the proof of \cite[theorem 1]{cortinovis-2022-randomized-trace}. Let
-    \begin{equation}
-        \widetilde{\mtx{B}}(t)
-        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
-            \mtx{B}(t) & & \\
-            & \ddots & \\
-            & & \mtx{B}(t)
-        \end{pmatrix}
-        \quad \text{and} \quad
-        \widetilde{\vct{\psi}} = \begin{pmatrix}
-            \vct{\psi}_1 \\
-            \vdots \\
-            \vct{\psi}_{n_{\mtx{\Psi}}}
-        \end{pmatrix}
-    \end{equation}
-    where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors.
-Then $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$, and
-    \begin{equation}
-        r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}}) = \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j.
-    \end{equation}
-    Hence, by applying \refequ{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$ we conclude that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
-    \begin{equation}
-        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j \right| ~ \mathrm{d}t
-        \leq 2^4 k \gamma \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
-    \end{equation}
-
-    Setting $\delta = \gamma^{-2 k}$ and choosing $k = \lceil \log(\delta^{-1}) \rceil$, where $\lceil x \rceil$ is the ceiling function which returns the greatest integer larger than or equal to $x$, we get that for all $\delta \in (0, e^{-\sfrac{1}{2}})$
-    \begin{equation}
-        k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\sfrac{1}{2\lceil \log(\delta^{-1}) \rceil}}
-        = \lceil \log(\delta^{-1}) \rceil e^{\sfrac{1}{2}\frac{\log(\delta^{-1})}{\lceil \log(\delta^{-1}) \rceil}}
-        \leq 2 \log(\delta^{-1}) e^{\sfrac{1}{2}},
-    \end{equation}
-    from which follows the second part of the theorem when $n_{\mtx{\Psi}} = (2^5 e^{\sfrac{1}{2}} \varepsilon^{-1} \log(\delta^{-1}))^2$.
-    %Consequently, with probability $1 - \delta$ we have
-    %\begin{equation}
-    %    \int_{a}^{b} \left| \frac{1}{n_{\mtx{\Psi}}} \sum_{i=1}^{n_{\mtx{\Psi}}} \vct{\psi}_i^{\top} \mtx{B}(t) \vct{\psi}_i - \Trace(\mtx{B}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\sfrac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
-    %\end{equation}
-\end{proof}
-
-%\todo{
-%\begin{remark}
-%    If additionally $\mtx{B}(t)$ is nonzero and symmetric positive semi-definite for all $t \in [a, b]$, then it follows from the proof of \refthm{thm:hutchinson} that with probability at least $1 - \delta$
-%    \begin{equation}
-%        \int_{a}^{b} \frac{| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) |}{\Trace(\mtx{B}(t))} ~\mathrm{d}t < \varepsilon,
-%    \end{equation}
-%    if $n_{\mtx{\Psi}} \geq 2^{10} \varepsilon^{-2} \log(\delta^{-1})^2 (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ with $\mu(t) = \lVert \mtx{B}(t) \rVert _2 / \Trace(\mtx{B}(t))$.
-Compared to the constant matrix equivalent \cite[remark 2]{cortinovis-2022-randomized-trace} we mainly observe a larger constant and an additional factor of $\log(\delta^{-1})$.
-%\end{remark}
-%}
-
-Compared to the constant case \cite[lemma 2.1]{meyer-2021-hutch-optimal} we need to choose the number of queries $n_{\mtx{\Psi}}$ by a factor of $\log(\delta^{-1})$ larger to achieve the same guarantee.
-
-The Girard-Hutchinson estimator distinguishes itself for its simplicity.
-However, its bound does not take into account the structure of $\mtx{B}(t)$.
-The $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ requirement implies that if we aim to increase the accuracy of the estimate by one digit, we would need to increase the number of queries by a factor of $100$.
-
-\subsection{Nyström approximation for parameter-dependent matrices}
-\label{subsec:nystrom}
-
-For symmetric positive semi-definite matrices whose singular values decay rapidly, the Nyström estimate \refequ{equ:nystrom-trace-estimator} can be significantly better than the Girard-Hutchinson estimate.
-This is reflected in the following theorem.
-
-\begin{theorem}{Nyström estimator for parameter-dependent matrices}{nystrom}
-    Suppose $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ is continuous and positive semi-definite in $t \in [a, b]$.
-Let $r \geq 2$ and $n_{\mtx{\Omega}} \geq r + 4$ be two integers.
-Then for all $\gamma \geq 1$ with probability at least $1 - \gamma^{-(n_{\mtx{\Omega}} - r)}$
-    \begin{equation}
-        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right| ~\mathrm{d}t
-        \leq \gamma^2 (1 + r) \int_{a}^{b} \sum_{i = r+1}^{n} \sigma_i(\mtx{B}(t)) ~\mathrm{d}t.
-    \end{equation}
-\end{theorem}
-
-\begin{proof}
-    Since $\mtx{B}(t)$ is positive semi-definite for all $t$, then by \cite[lemma 2.1]{frangella-2023-randomized-nystrom}, $\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ is also positive semi-definite for all $t$.
-Therefore, for fixed $t \in [a, b]$
-    \begin{equation}
-        \left| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right|
-        = \left| \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right|
-        = \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _{\ast},
-    \end{equation}
-    where $\lVert \cdot \rVert _{\ast}$ represents the nuclear norm.  From the proof of \cite[corollary 8.2]{tropp-2023-randomized-algorithms} it follows that
-    \begin{equation}
-        %\lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _2^2 \leq \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _F^2
-        \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{\mtx{B}(t)^{\sfrac{1}{2}} \mtx{\Omega}}) \mtx{B}(t)^{\sfrac{1}{2}} \rVert _F^2.
-    \end{equation}
-    Hence, by \cite[theorem 9]{kressner-2023-randomized-lowrank}, with $n_{\mtx{\Omega}} = p + r$, the result follows by noting $\sigma_i(\mtx{B}(t)^{\sfrac{1}{2}})^{2} = \sigma_i(\mtx{B}(t))$.
-\end{proof}
-
-Compared to the constant matrix case \cite[theorem 8.1]{tropp-2023-randomized-algorithms}, our bound also scales with the best rank-$r$ approximation error in the nuclear norm, albeit with a factor approximately proportional to $r$, which cannot be compensated for with large enough oversampling.
-
-The downside of the Nyström estimator is that it will -- in general -- not work well enough on matrices whose singular values do not decay quickly, for example $\mtx{B}(t) \equiv \mtx{I}_n$.
-
-\subsection{Nyström++ estimator for parameter-dependent matrices}
-\label{subsec:nystrom-pp}
-
-Just as in the constant matrix case, our goal is now to prove that the Nyström++ estimator \refequ{equ:nystrompp-trace-estimator} achieves an $\varepsilon$-relative error with just $\mathcal{O}(\varepsilon^{-1})$ random vectors used to construct the estimate.
-To do so, we mimic the proof of a similar bound for the Hutch++ estimator \cite[theorem 3.1]{meyer-2021-hutch-optimal}.
-Correspondingly, we will need to show that with $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2})$ random vectors, the Nyström approximation verifies $\int \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)\rVert _F~\mathrm{d}t \leq \varepsilon \int \Trace(\mtx{B}(t))~\mathrm{d}t$, regardless of the structure of the matrix $\mtx{B}(t)$.
-Just as in the corresponding proof for the Girard-Hutchinson estimator \refthm{thm:hutchinson}, we again bound the higher order moments of the approximation error for fixed $t$ and then use Minkowsi's integral inequality to carry over the results to the continuous case.
-We will do this in a sequence of three lemmas.
-
-For reasons which will become clear in the proof of \reflem{lem:nystrom}, we will need to bound $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]$ for a constant matrix $\mtx{A}$ and a standard Gaussian random matrix $\mtx{\Omega}$.
-The corresponding result for the Frobenius norm $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _F^2 \right]$ has already been shown in \cite{kressner-2023-randomized-lowrank}.
-Several ways of separately bounding the spectral norm moments of the Wishart matrix $\mtx{\Omega} \mtx{\Omega}^{\top}$ already exist, however, most of them are not sufficient for our purposes \cite{chen-2005-condition-numbers, edelman-1988-eigenvalues-condition, james-1964-distributions-matrix}.
-In \cite[lemma B.1]{tropp-2023-randomized-algorithms} a bound is achieved for $p = 2$ and $p = 4$.
-We will now generalize this result for arbitrary $p \in \mathbb{N}$.
-To do so, we first show the bound for Gaussian random vectors and then extend it to Gaussian random matrices in \reflem{lem:spectral-norm-moment}.
-
-\begin{lemma}{Spectral norm moments of non-standard Gaussian random vector}{spectral-norm-moment-vector}
-    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random vector $\vct{\omega} \in \mathbb{R}^{m}$ satisfy for all $k,p \in \mathbb{N}$ 
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2^2 \right]
-        \leq  (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
-    \end{equation}
-    %\begin{equation}
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
-    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
-    %\end{equation}
-    %If we choose 
-\end{lemma}
-
-\begin{remark}
-    In particular, this result can also be used to bound
-    \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2 \right] = \sqrt{\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2^2 \right]} \leq \sqrt{k + p} \cdot \left(\lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F\right)
-    \end{equation}
-    for any $k,p \in \mathbb{N}$.
-\end{remark}
-
-\begin{proof}
-    With the singular value decomposition $\mtx{A} = \mtx{U} \mtx{\Sigma} \mtx{V}^{\top}$ and the unitary invariance of the spectral norm, it can be seen that $\lVert \mtx{A} \mtx{\omega} \rVert _2^2 = \lVert \mtx{\Sigma} \widetilde{\mtx{\omega}} \rVert _2^2$ where $\widetilde{\mtx{\omega}}$ is again a standard Gaussian random vector.
-Hence, it is enough to bound $\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{\Sigma} \mtx{\omega} \rVert _2^2 \right]$ for a diagonal matrix $\mtx{\Sigma} = \operatorname{diag}(\sigma_1, \dots, \sigma_m)$ with $\sigma_1 \geq \dots \geq \sigma_m \geq 0$.
-
-    First, we can rewrite
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2^2 \right]
-        = \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right].
-        \label{equ:matvec-spectral-norm-simplification}
-    \end{equation}
-    Inspired by the proof of \cite[theorem 1]{cohen-2016-optimal-approximate}, we split the diagonal entries of $\mtx{\Sigma}$ into $\ell = \lceil m/k \rceil$ groups of size $k \geq 1$
-    \begin{equation}
-        \overbrace{\underbrace{\sigma_1, \dots, \sigma_k}_{\leq \sigma_1}}^{\geq \sigma_{k+1}}, \overbrace{\underbrace{\sigma_{k+1}, \dots, \sigma_{2k}}_{\leq \sigma_{k+1}}}^{\geq \sigma_{2k+1}}, \dots, \overbrace{\underbrace{\sigma_{1+(\ell - 1)k}, \dots, \sigma_{\ell k}}_{\leq \sigma_{1+(\ell - 1)k}}}^{\geq 0}.
-    \end{equation}
-    If $m$ is not a multiple of $k$, we let $\sigma_i = 0$ for $i > m$. Since $\sigma_1 \geq \dots \geq \sigma_{\ell k} \geq 0$,
-    \begin{equation}
-        \sigma_1^2 = \lVert \mtx{\Sigma} \rVert _2^2
-        \quad \text{and} \quad
-        \sigma_{1 + (i-1)k}^2 \leq \frac{1}{k} \sum_{j=1}^{k} \sigma_{j + (i-2)k}^2, i = 2, \dots, \ell,
-    \end{equation}
-    and therefore
-    \begin{equation}
-        \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \sum_{j=1}^{(\ell - 1)k} \sigma_j^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2.
-        \label{equ:singular-value-group-bound}
-    \end{equation}
-    This allows us to bound
-    \begin{align}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right]
-        &= \mathbb{E}^{\sfrac{p}{2}}\Bigg[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{j + (i-1)k}^2}_{\leq \sigma_{1 + (i-1)k}^2} \omega_{j + (i-1)k}^2 \Bigg] && \text{(separate sum into groups)} \notag \\
-        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{($\sigma_{j + (i-1)k}^2 \leq \sigma_{1 + (i-1)k}^2$, $\forall i, j$)} \notag \\
-        &\leq \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 ~ \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{(triangle inequality)} \notag \\
-        %&\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k}{2} + p)}{\Gamma(\frac{k}{2})} \right)^{\sfrac{1}{p}} && \text{(\refequ{equ:singular-value-group-bound} and \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
-        &\leq (k + p) \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 && \text{(\reflem{lem:gamma})} \notag \\
-        &\leq (k + p) \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right). && \text{(using \refequ{equ:singular-value-group-bound})}
-    \end{align}
-
-    %The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
-    %\begin{align}
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    &= \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}} \notag \\
-    %    &\leq \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{1}{\Gamma(k - r + 2)} \right)^{\frac{1}{k - r + 1}} \left( \frac{k + r}{2} \right) \notag \\
-    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right)
-    %\end{align}
-    %With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\sfrac{1}{n}} \leq \frac{e}{n}$ follow respectively.
-Hence,
-    %\begin{equation}
-    %    \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    \leq \frac{e^2}{k + 1}\sqrt{ \frac{3k}{2} }
-    %    \leq e^2 \sqrt{\frac{3}{2k}}.
-    %    \label{equ:OSE-moment-bound-second}
-    %\end{equation}
-
-\end{proof}
-
-\begin{lemma}{Spectral norm moments of non-standard Gaussian random matrix}{spectral-norm-moment}
-    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all $p \in \mathbb{N}$ 
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
-        \leq  (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
-    \end{equation}
-    %\begin{equation}
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
-    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
-    %\end{equation}
-    %If we choose 
-\end{lemma}
-
-\begin{remark}
-    In particular, this result can also be used to bound
-    \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2 \right] = \sqrt{\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]} \leq \sqrt{k + p} \cdot \left(\sqrt{3} \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F\right)
-    \end{equation}
-    for any $k,p \in \mathbb{N}$. Indeed, for $p=2$ this bound is slightly less tight than its equivalent in \cite[lemma B.1]{tropp-2023-randomized-algorithms}.
-\end{remark}
-
-\begin{proof}
-    In the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms}, it is shown that Slepian's inequality \cite[theorem 7.2.1]{vershynin-2018-random-processes} applied to two appropriately constructed Gaussian fields yields
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
-        \leq \lVert \mtx{A} \rVert _2^2 \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \vct{\omega}_1 \rVert _2^2 \right] + \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega}_2 \rVert _2^2 \right]
-    \end{equation}
-    for independent standard Gaussian random vectors $\vct{\omega}_1 \in \mathbb{R}^{k}$ and $\vct{\omega}_2 \in \mathbb{R}^{m}$.
-Both moments can be bound using \reflem{lem:spectral-norm-moment-vector}, one of which with $\mtx{A} = \mtx{I}_k$, to get
-    \begin{align}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
-        &\leq \lVert \mtx{A} \rVert _2^2 (k + p) \left( \lVert \mtx{I}_k \rVert _2^2 + \frac{1}{k} \lVert \mtx{I}_k \rVert _F^2 \right) + (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right) \notag \\
-        &\leq (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
-    \end{align}
-
-    %The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
-    %\begin{align}
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    &= \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}} \notag \\
-    %    &\leq \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{1}{\Gamma(k - r + 2)} \right)^{\frac{1}{k - r + 1}} \left( \frac{k + r}{2} \right) \notag \\
-    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right)
-    %\end{align}
-    %With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\sfrac{1}{n}} \leq \frac{e}{n}$ follow respectively.
-Hence,
-    %\begin{equation}
-    %    \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    \leq \frac{e^2}{k + 1}\sqrt{ \frac{3k}{2} }
-    %    \leq e^2 \sqrt{\frac{3}{2k}}.
-    %    \label{equ:OSE-moment-bound-second}
-    %\end{equation}
-
-\end{proof}
-
-\begin{lemma}{Nyström approximation for parameter-dependent matrices}{nystrom}
-    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-Then its Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ with Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ for \emph{even} $n_{\mtx{\Omega}} \geq 4$ verifies with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
-    \begin{equation}
-        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t
-        \label{qu:nystrompp-theorem-bound}
-    \end{equation}
-    for any $\gamma \geq 1$ and a universal constant $c > 0$.
-In particular, if we choose $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1}))$, then with probability at least $1-\delta$ for $\delta \in (0, 1)$ and $\varepsilon \in (0, 1)$ we have $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$.
-\end{lemma}
-
-%\todo{Proof idea: structural bound, then higher order moment bound to apply Markov's inequality.}
-
-\begin{proof}
-    First, we temporarily omit the parameter dependence.
-Let $\mtx{B} = \mtx{U} \mtx{\Lambda} \mtx{U}^{\top}$ be the eigenvalue decomposition of a symmetric positive semi-definite matrix $\mtx{B}$, i.e.
-$\mtx{\Lambda} = \operatorname{diag}(\lambda_1, \dots, \lambda_n)$ and $\mtx{U}^{\top} \mtx{U}= \mtx{I}_n$.
-We define the matrix partitions
-    \begin{equation}
-        \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+4ex+\baselineskip}
-        \mtx{U} = \begin{bmatrix}
-            \smash{\underbrace{\mtx{U}_1}_{n \times k}} & \smash{\underbrace{\mtx{U}_2}_{n \times (n-k)}}
-        \end{bmatrix}
-        \quad \text{and} \quad
-        \mtx{\Lambda} =
-        \begin{bmatrix}
-            \smash{\overbrace{\mtx{\Lambda}_1}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2}_{(n-k) \times (n-k)}}
-        \end{bmatrix}.
-    \end{equation}
-    Further, for a standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ we define $\mtx{\Omega}_1 = \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times n_{\mtx{\Omega}}}$ and $\mtx{\Omega}_2 = \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times n_{\mtx{\Omega}}}$, which are likewise standard Gaussian.
-
-    We can apply \cite[theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound
-    \begin{equation}
-        \lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F 
-        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2,
-        \label{equ:nystrom-proof-persson-bonud}
-    \end{equation}
-    where $\lVert \cdot \rVert _{(4)}$ denotes the Schatten-4 norm.
-
-    As in the proof of \cite[lemma 3]{meyer-2021-hutch-optimal}, the first term -- and its moments -- are limited by
-    \begin{equation}
-        \lVert \mtx{\Lambda}_2 \rVert _F
-        = \sqrt{\sum_{i=k+1}^{n} \lambda_i^2}
-        \leq \sqrt{ \lambda_{k+1} \sum_{i=k+1}^{n} \lambda_i}
-        \leq \sqrt{ \frac{\Trace(\mtx{B})}{k} \Trace(\mtx{B})}
-        \leq \frac{1}{\sqrt{k}} \Trace(\mtx{B}).
-        \label{equ:nystrom-proof-frobenius-trace}
-    \end{equation}
-    The $p$-th moment of the second term can be processed with standard matrix norm manipulations to
-    \begin{align}
-        \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        %&= \mathbb{E}^{p}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )( \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger})^{\top} \rVert _F \right] && \text{($\lVert \mtx{A} \rVert _{(4)}^2 = \lVert \mtx{A} \mtx{A}^{\top} \rVert _{F}$)} \notag \\
-        %&\leq \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] && \text{($\lVert \mtx{A} \mtx{B} \rVert _F \leq \lVert \mtx{A} \rVert _F \lVert \mtx{B} \rVert _2$)} \notag \\
-        &\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{($\lVert \mtx{A} \rVert _{(4)}^2 \leq \sqrt{k} \lVert \mtx{A} \rVert _2^2$)} \notag \\
-        &\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{(submultiplicativity)} \notag \\
-        &\leq \sqrt{k} \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2\right] \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2  \right].
-&& \text{(independence)}
-        \label{equ:nystrom-proof-processed-tail}
-    \end{align}
-    To match the decay rate of the moments of the first term in \refequ{equ:nystrom-proof-persson-bonud}, we will need to ensure that this term also is of order $\mathcal{O}(\Trace{\mtx{B}} / \sqrt{k})$.
-To this purpose, and to simplify the notation, we choose $k = n_{\mtx{\Omega}}/2$ and $p = n_{\mtx{\Omega}}/4$ for even $n_{\mtx{\Omega}}$ to ensure $k \in \mathbb{N}$.
-We can apply \reflem{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$ to bound
-    \begin{equation}
-        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \right]
-        \leq \frac{3 n_{\mtx{\Omega}}}{2} \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
-        \label{equ:spectral-norm-bound-applied}
-    \end{equation}
-    The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
-    \begin{equation}
-        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
-        = \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{n_{\mtx{\Omega}}}{4}} \right]^{\sfrac{4}{n_{\mtx{\Omega}}}}%  \notag \\
-        \leq \left( 1 + \frac{n_{\mtx{\Omega}}}{2} \right)^{\sfrac{4}{n_{\mtx{\Omega}}}} \left( \frac{1}{(\frac{n_{\mtx{\Omega}}}{2} + 1)!}\right)^{\frac{2}{\frac{n_{\mtx{\Omega}}}{2} + 1}} \left( \frac{3 n_{\mtx{\Omega}}}{4}\right).
-    \end{equation}
-    From the expansion of the exponential function as a power series $e^n = 1 + n + \dots + n^n/n!
-+ \dots$ for $n \in \mathbb{N}$ we see $e^n \geq 1 + n$ and $e^n \geq n^n / n!$, from which $(1 + n)^{\sfrac{1}{n}} \leq e$ and $\left( 1/n!
-\right)^{\sfrac{1}{n}} \leq e/n$ follow respectively.
-Hence,
-    \begin{equation}
-        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
-        \leq e^2 \frac{e^2}{(\frac{n_{\mtx{\Omega}}}{2}+ 1)^2} \left( \frac{3 n_{\mtx{\Omega}}}{4}\right)
-        \leq \frac{3 e^4}{n_{\mtx{\Omega}}}.
-        \label{equ:pinv-spectral-norm-bound}
-    \end{equation}
-    As opposed to \cite[lemma B.3]{tropp-2023-randomized-algorithms}, our bound holds for all even $n_{\mtx{\Omega}} \in \mathbb{N}$, at the cost of a factor of $e^2$.
-Inserting \refequ{equ:spectral-norm-bound-applied} and  \refequ{equ:pinv-spectral-norm-bound} in \refequ{equ:nystrom-proof-processed-tail} gives
-    \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \frac{9 e^4}{2 \sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
-        %\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
-    \end{equation}
-    We can identify $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 = \lambda_{k+1}$ and $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 = \Trace(\mtx{\Lambda}_2)$, and just as in the proof of \cite[lemma 3.1]{meyer-2021-hutch-optimal} we bound $\lambda_{k+1} \leq \Trace(\mtx{B})/k$ and $\Trace(\mtx{\Lambda}_2) \leq \Trace(\mtx{B})$ -- remembering $k = n_{\mtx{\Omega}}/2$ -- to get
-    \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \frac{9 e^4}{2 \sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \left( \frac{6}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) + \frac{1}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) \right)
-        \leq \frac{9e^4}{2 \sqrt{2}} \frac{7}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
-        \label{equ:nystrom-proof-tail-bound}
-        %\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
-    \end{equation}
-    Inserting \refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud} and using the triangle inequality for $\mathbb{E}^{p}[\cdot]$ we obtain
-    \begin{equation}
-        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right]
-        \leq \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[ \lVert \mtx{\Lambda}_2 \rVert _F \right] + \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
-    \end{equation}
-    where $c = 1 + 63 e^4 / 2 \sqrt{2}$.
-
-    As in \cite{kressner-2023-randomized-lowrank}, we can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable and by the Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities}, we get for $n_{\mtx{\Omega}} \geq 4$
-    \begin{align}
-        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \right] 
-        &\leq \int_{a}^{b} \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F \right] ~\mathrm{d}t \notag \\
-        %&\leq \mathbb{E}^{\sfrac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
-        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \mathbb{E}^{\sfrac{k}{2}} \left[ \int_{a}^{b}  \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
-        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
-        &\leq \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
-    \end{align}
-    From Markov's inequality follows with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
-    \begin{equation}
-        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t,
-    \end{equation}
-    from which follows the first statement.
-
-    Fixing $\gamma = e$ and letting $n_{\mtx{\Omega}} = (c e \varepsilon)^{-2}  + 4 \log(\delta^{-1})$ gives us the second part of the theorem.
-
-    %Setting $\delta = \gamma^{-\sfrac{k}{4}}$ and choosing $k = \lceil \varepsilon^{-2}\log(\delta^{-1}) \rceil$ we get
-    %\begin{align}
-    %    \gamma \frac{1}{\sqrt{k}}
-    %    &= \delta^{-\sfrac{2}{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} \frac{1}{\sqrt{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} && \text{(definition of $\delta$ and choice of $k$)} \notag \\
-    %    &= e^{2\varepsilon^{2} \frac{\varepsilon^{-2}\log(\delta^{-1})}{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} \frac{1}{\sqrt{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} && \text{($\delta = e^{-\log(\delta^{-1})}$)} \notag \\
-    %    &\leq e^{2\varepsilon^{2}} \frac{\varepsilon}{\sqrt{ \log(\delta^{-1})}} && \text{($\lceil x \rceil \geq x$ if $x \geq 0$)}
-    %\end{align}
-    %which is smaller than $\varepsilon$ if $\delta \leq e^{-e^{4 \varepsilon^2}}$from which follows the second part of the theorem when using $n_{\mtx{\Omega}} = 2k$.
-\end{proof}
-
-When compared to the equivalent result for constant matrices \cite[Lemma 3.2]{persson-2022-improved-variants}, our constant factor up front is around four times larger, and our failure probability decays half a power slower.
-
-Finally, we can combine the bound on the Girard-Hutchinson estimator \refthm{thm:hutchinson} and the one on the Nyström approximation \reflem{lem:nystrom} to obtain our main result.
-
-\begin{theorem}{Nyström++ trace estimator for parameter-dependent matrices}{nystrom-pp}
-    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(\delta^{-1})^2)$ with \emph{even} $n_{\mtx{\Omega}}$, then for $\delta \in (0, e^{-\sfrac{1}{2}}/2)$ and $\varepsilon \in (0, 1)$ with probability at least $1 - \delta$
-    \begin{equation}
-        \int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t
-        \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
-    \end{equation}
-\end{theorem}
-
-\begin{proof}
-    By choosing $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\tilde{\varepsilon}^{-2} \log(\tilde{\delta}^{-1})^2)$ we get
-    \begin{align}
-        &\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t \notag \\
-        &= \int_{a}^{b} | \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) - \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) |~\mathrm{d}t && \text{(by definition of estimators)} \notag \\
-        &\leq \tilde{\varepsilon} \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F ~\mathrm{d}t && \text{(\refthm{thm:hutchinson} w.p. $\geq 1 - \tilde{\delta}$)} \notag \\
-        &\leq \tilde{\varepsilon}^2 \int_{a}^{b} \Trace(\mtx{B}(t)) ~\mathrm{d}t && \text{(\reflem{lem:nystrom} w.p. $\geq 1 - \tilde{\delta}$)} 
-    \end{align}
-    with probability at least $1 - 2\tilde{\delta}$. Taking $\varepsilon = \tilde{\varepsilon}^2$ and $\delta = 2 \tilde{\delta}$
-\end{proof}
-
-Comparing \refthm{thm:nystrom-pp} with its equivalent for constant matrices \cite[theorem 3.4]{persson-2022-improved-variants}, we notice a less favorable scaling of the required number of random vectors $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$ in terms of $\log(\delta^{-1})$.
-
-% \begin{proof}
-%     Combining with union bound.
-%     \begin{align}
-%         &\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t \notag \\
-%         &= \int_{a}^{b} | \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)) - \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)) |~\mathrm{d}t && \text{(definition of estimators)} \notag \\
-%         &= \todo{c} \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F ~\mathrm{d}t && \text{(with probability $1 - \delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1})$)} \notag \\
-%         &= \todo{c} \frac{1}{\sqrt{n_{\mtx{\Psi}} n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t)) ~\mathrm{d}t
-%     \end{align}
-% \end{proof}
-
-\section{Application to spectral density approximation}
-\label{sec:application}
-
-We consider the task of computing the spectral density of a real symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ with eigenvalues $\lambda_1, \dots, \lambda_n$, which is defined as
-\begin{equation}
-    \phi(t) = \frac{1}{n} \sum_{i=1}^{n} \delta(t - \lambda_i),
-    \label{equ:spectral-density}
-\end{equation}
-where $\delta$ is the Dirac delta distribution.
-Finding approximations to this distribution is challenging.
-Furthermore, in most applications the exact location of every eigenvalue is not important, but more so their approximate locations relative to each other, such as eigenvalue clusters, spectral gaps, or outliers.
-In the following we will introduce a smoothed version of the spectral density, which is significantly easier to approximate at the cost of losing some of the finer characteristics of the spectrum.
-
-\subsection{Smoothed spectral density}
-\label{subsec:spectral-density}
-
-The smoothed spectral density $\phi_{\sigma}$ is defined as
-\begin{equation}
-    \phi_{\sigma}(t) = \frac{1}{n} \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i) = \frac{1}{n} \Trace(g_{\sigma}(t\mtx{I}_n - \mtx{A}))
-    \label{equ:smooth-spectral-density}
-\end{equation}
-for a smoothing kernel $g_{\sigma}$ parametrized by a smoothing parameter $\sigma > 0$, which controls by how much $\phi$ is smoothed.
-We choose a Gaussian smoothing kernel of width $\sigma > 0$, given by
-\begin{equation}
-    g_{\sigma}(s) = \frac{1}{\sigma \sqrt{2\pi}} e^{-\sfrac{s^2}{2\sigma^2}}.
-    \label{equ:smoothing-kernel}
-\end{equation}
-The task in \refequ{equ:smooth-spectral-density} is to approximate the trace of the parameter-dependent matrix $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
-Hence, we can apply the results derived in \refsec{sec:analysis}.
-
-Choosing the smoothing parameter $\sigma$ results in the following trade off: Typically, larger $\sigma$ make it easier to approximate the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$, whereas smaller $\sigma$ allow us to stay closer to the original spectral density \refequ{equ:spectral-density}.
-To see the latter, we measure the error between the spectral density $\phi$ and its smoothened version $\phi_{\sigma}$ with
-\begin{equation}
-    \sup_{f \in \mathcal{S}} \int_{-1}^{1} f(t) (\phi(t) - \phi_{\sigma}(t))~\mathrm{d}t,
-    \label{equ:error-metric}
-\end{equation}
-where $\mathcal{S}$ represents an appropriately chosen space of test functions.
-Among others, \cite{lin-2016-approximating-spectral} proposes $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - \lambda), \lambda \in [\lambda_{\text{min}}, \lambda_{\text{max}}]\}$, \cite{chen-2021-analysis-stochastic} uses $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [-1, 1] \}$ where $\Theta$ is the Heaviside step function, and \cite{braverman-2022-sublinear-time} uses $\mathcal{S} = \{f : |f(x) - f(y)| \leq |x - y| \}$ for which the error \refequ{equ:error-metric} is equivalent to the Wasserstein-1 distance between $\phi$ and $\phi_{\sigma}$.
-In this last metric, provided we assume the spectrum to be somewhat uniform and $\sigma$ to be rather small, a simple calculation allows us to roughly estimate the smoothing error \refequ{equ:error-metric} to be $\sigma$.
-Consequently, if we can only allow an approximation to deviate by at most a factor of $\varepsilon > 0$ from the original spectral density, we should choose a smoothing parameter $\sigma \approx \varepsilon$.
-
-\subsection{Chebyshev expansion of smoothing kernel}
-\label{subsec:chebyshev-expansion}
-
-To estimate the trace in \refequ{equ:smooth-spectral-density}, we first need to evaluate the involved matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
-However, doing this exactly would require us to first diagonalize $\mtx{A}$, which is often a prohibitively expensive operation.
-For algorithmic reasons, we seek for approximations of this matrix function in terms of affine expansions, of which the Chebyshev expansion is most suitable due to its approximation error guarantees and ease of manipulation.
-In every $t \in \mathbb{R}$ we expand
-\begin{equation}
-    g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) = \sum_{l=0}^{m} \mu_l(t) T_l(\mtx{A})
-    \label{equ:matrix-expansion}
-\end{equation}
-in terms of a set of $m+1$ coefficients $\mu_0(t), \dots, \mu_m(t)$ which linearly combine the Chebyshev polynomials
-\begin{equation}
-    \begin{cases}
-        T_l : [-1, 1] \to \mathbb{R} \\
-        T_l(s) = \cos(l \cdot \arccos(s))
-    \end{cases}
-    \label{equ:chebyshev-polynomial}
-\end{equation}
-evaluated at the matrix $\mtx{A}$. It is easy to verify that these polynomials satisfy the three-term recurrence relation
-\begin{equation}
-    T_l(s) =
-    \begin{cases}
-        s^l, & \text{if $l \in \{0, 1\}$}; \\
-        T_l(s) = 2 s T_{l-1}(s) - T_{l-2}(s), & \text{if $l \geq 2$.}
-    \end{cases}
-    \label{equ:chebyshev-recurrence}
-\end{equation}
-With the help of the Chebyshev expansion \refequ{equ:matrix-expansion} we can now define the expanded spectral density
-\begin{equation}
-    \phi_{\sigma}^{(m)}(t) =  \frac{1}{n} \Trace(g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})).
-    \label{equ:expanded-spectral-density}
-\end{equation}
-
-However, the Chebyshev polynomials are only well-defined for matrices whose spectrum is contained in $[-1, 1]$.
-When working with general matrices of bounded spectrum, we will first have to estimate a lower bound $a$ and upper bound $b$ on the spectrum \cite{zhou-2011-bounding-spectrum}.
-A spectral transform
-\begin{equation}
-    \begin{cases}
-        \tau : [a, b] \to [-1, 1]\\
-        \tau(t) = \frac{2t - a - b}{b - a}
-    \end{cases}
-\end{equation}
-then lets us define the matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ whose spectrum is identical to the one of $\mtx{A}$ but shrunk to the interval $[-1, 1]$, for which the expansion \refequ{equ:matrix-expansion} is valid.
-Note that through this transformation we will also have to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when reverting the spectral transform $\tau$.
-
-\subsubsection{Fast and consistent squaring}
-\label{subsubsec:dct}
-
-It is well known that the coefficients of a Chebyshev expansion of a function can -- in exact arithmetic -- be computed with a discrete cosine transform (DCT) of the function evaluations at the Chebyshev nodes $s_i = \cos(\pi i / m)$ for $i=0,\dots,m$
-\begin{equation}
-    g_{\sigma}(t - \cos(s_i)) = \sum_{l=0}^{m} \mu_l(t) \cos\left(\frac{\pi i l}{m} \right).
-    \label{equ:discrete-cosine-transform}
-\end{equation}
-This can be accomplished in $\mathcal{O}(m \log(m))$ operations.
-\cite[algorithm 1]{lin-2017-randomized-estimation} uses a slightly different approach, which only approximates the coefficients and is noticeably slower (cf.
-\reftab{tab:chebyshev-timing-interpolation}).
-
-A highly useful property of the Chebyshev expansion -- of which we will make use twice in the following pages -- is the fact that we can square such an expansion in $\mathcal{O}(m \log(m))$ operations \cite{baszenski-1997-fast-polynomial}, which will be used in the below algorithm.
-
-\begin{algo}{Fast squaring of Chebyshev expansion}{chebyshev-squaring}
-    \begin{algorithmic}[1]
-        \Statex \textbf{Input:} Coefficients $\{ \mu_l \}_{l=0}^{m}$ of the expansion $\sum_{l=0}^{m} \mu_l T_l$
-        \Statex \textbf{Output:} Coefficients $\{ \nu_l \}_{l=0}^{2m}$ such that $\sum_{l=0}^{2m} \nu_l T_l = (\sum_{l=0}^{m} \mu_l T_l)^2$
-        \State Define $\vct{\mu} \in \mathbb{R}^{2m + 1}$ whose first $m + 1$ entries are the coefficients $\mu_l$ and the remaining $m$ are $0$
-    \State Compute $\vct{f} = \DCT(\vct{\mu})$
-    \State Compute $\vct{\nu} = \DCT^{-1}(\vct{f}^{2})$ whose entries are the coefficients $\nu_l$
-    \end{algorithmic}
-    \end{algo}
-
-\subsubsection{Expansion error}
-\label{subsubsec:expansion-error}
-
-We now analyze the error of the Chebyshev expansion of a Gaussian smoothing kernel $g_{\sigma}$.
-
-\begin{lemma}{Chebyshev expansion error}{chebyshev-error}
-    The expansion $g_{\sigma}^{(m)}$ of the Gaussian smoothing kernel $g_{{\sigma}}$ \refequ{equ:smoothing-kernel} satisfies
-    \begin{equation}
-        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \right| \leq \sqrt{\frac{2e}{\pi}} \frac{1}{\sigma^2} (1 + \sigma)^{-m} \equiv E_{\sigma, m}
-        \label{equ:chebyshev-interpolation-sup-error-kernel}
-        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
-        %\label{equ:2-chebyshev-interpolation-error}
-    \end{equation}
-    for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and $s \in [-1, 1]$.
-\end{lemma}
-
-The proof specializes Bernstein's theorem for Chebyshev interpolation to the case of a Gaussian smoothing kernel \refequ{equ:smoothing-kernel}.
-It loosely follows the proof give in the preprint of \cite[theorem 2]{lin-2017-randomized-estimation}, but attains a slightly different dependence on $\sigma$.
-
-\begin{proof}
-    From Bernstein's theorem \cite[theorem 4.3]{trefethen-2008-gauss-quadrature} it follows that if $f$ is analytic within the Bernstein ellipse $\mathcal{E}_{\chi} \subset \mathbb{C}$ with foci $\{-1, +1\}$ and sum of semi-axes $\chi > 1$, then for any $m \in \mathbb{N}$
-    \begin{equation}
-        \sup_{s \in [-1, 1]} \left| f(s) - f^{(m)}(s) \right| \leq \frac{2}{\chi^m (\chi - 1)} \max_{z \in \mathcal{E}_{\chi}} |f(z)|.
-        \label{equ:bernstein-bound}
-    \end{equation}
-    In particular, when $f = g_{\sigma}(t - \cdot)$ we observe for $z = x + iy \in \mathcal{E}_{\chi}$
-    \begin{equation}
-    | g_{\sigma}(t - z) | 
-    = \frac{1}{\sigma \sqrt{2 \pi}} \left| e^{-\sfrac{(t - z)^2}{2\sigma^2}} \right|
-    = \frac{1}{\sigma \sqrt{2 \pi}} e^{-\sfrac{(t - x)^2}{2\sigma^2}}e^{\sfrac{y^2}{2\sigma^2}}.
-    %\leq \frac{1}{n \sigma \sqrt{2 \pi}} \max_{x + iy \in \mathcal{E}_{\chi}} e^{\sfrac{y^2}{2\sigma^2}} 
-    \end{equation}
-    Because $e^{-\sfrac{(t - x)^2}{2\sigma^2}} \leq 1$ for all $x, t \in \mathbb{R}$, and the maximum absolute value of $y$ is limited by the length of the semi-axis of the Bernstein ellipse in the direction of the imaginary axis, which is $(\chi - \chi^{-1}) / 2$, we upper bound
-    \begin{equation}
-        \max_{z \in \mathcal{E}_{\chi}} | g_{\sigma}(t - z) | 
-        \leq \frac{1}{\sigma \sqrt{2 \pi}} e^{\sfrac{(\chi - \chi^{-1})^2}{8 \sigma^2}}.
-    \end{equation}
-
-    Choosing $\chi = 1 + \sigma$ gives a simple and good bound, since in this case $\chi - \chi^{-1} \leq 2\sigma$, which implies
-    \begin{equation}
-        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \right| \leq \sqrt{\frac{2e}{\pi}} \frac{1}{\sigma^2} (1 + \sigma)^{-m}.
-    \end{equation}
-\end{proof}
-
-\Reflem{lem:chebyshev-error} shows that when $0 < \sigma \ll 1$ and if we choose $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, the error we induce with the Chebyshev expansion is at most $\mathcal{O}(\varepsilon)$.
-It now also allows us to bound the error we make when expanding the spectral density.
-
-\begin{theorem}{$L^1$-error of Chebyshev expansion of spectral density}{chebyshev-error}
-    Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the spectal density with Chebyshev expansion $\phi_{\sigma}^{(m)}$ satisfies
-    \begin{equation}
-        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq 2 E_{\sigma, m}
-        \label{equ:chebyshev-interpolation-spectral-density}
-        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
-        %\label{equ:2-chebyshev-interpolation-error}
-    \end{equation}
-    for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and the error $E_{\sigma, m}$ from \reflem{lem:chebyshev-error}.%, and $t \in \mathbb{R}$.
-%The same result holds for the expansion of a non-negative Chebyshev expansion of \emph{even} degree $m \in \mathbb{N}$ but with $\underline{E}_{\sigma, m}$.
-\end{theorem}
-
-\begin{proof}
-    We observe that for all $t \in [-1, 1]$
-    \begin{align}
-        &\left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right| \notag \\
-        &= \left| \frac{1}{n} \Trace(g_{\sigma}(t\mtx{I}_n - \mtx{A})) - \frac{1}{n} \Trace(g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})) \right|
-        && \text{(definitions \refequ{equ:smooth-spectral-density} and \refequ{equ:expanded-spectral-density})} \notag \\
-        &= \left| \frac{1}{n} \sum_{i=1}^n \left(g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i)\right) \right|
-        && \text{($\lambda_1, \dots, \lambda_n$ eigenvalues of $\mtx{A}$)} \notag \\
-        &\leq \max_{i = 1, \dots, n} \left| g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i) \right|
-        && \text{(conservative upper bound)} \notag \\
-        &\leq E_{\sigma, m}.% ~ \text{(or $\underline{E}_{\sigma, m}$)}.
-        && \text{(using \reflem{lem:chebyshev-error})} \notag \\
-    \end{align}
-    Then, Hölder's inequality allows us to conclude
-    \begin{equation}
-        \int_{-1}^{1} | \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) | ~\mathrm{d}t
-        \leq 2 \sup_{t \in [-1, 1]} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|
-        \leq 2 E_{\sigma, m}.% ~ \text{(or $2 \underline{E}_{\sigma, m}$)}.
-    \end{equation}
-\end{proof}
-
-\subsection{Chebyshev-Nyström++ method for spectral density approximation}
-\label{subsec:chebyshev-nystrom}
-
-Our goal is to apply the Nyström++ trace estimator which we analyzed in \refsec{subsec:nystrom-pp} to the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ to approximate the smoothed spectral density $\phi_{\sigma}$.
-However, due to the oscillatory nature of the Chebyshev expansion, we can no longer guarantee that $g_{\sigma}^{(m)} \geq 0$, and consequently, the matrix function $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ may be (slightly) indefinite.
-While in our experiments this never seemed to be an issue, it disallows us to directly use \refthm{thm:nystrom-pp}.
-While variants of the Nyström approximation designed for indefinite matrices have been analyzed, no practical algorithms for computing them exist \cite{nakatsukasa-2023-randomized-lowrank}.
-
-To still apply the bound from \refthm{thm:nystrom-pp} in this scenario, we will have to ensure that the Chebyshev expansion is non-negative.
-We can think of many ways in which this can be achieved: by using Jackson damping \cite{jackson-1912-approximation-trigonometric,braverman-2022-sublinear-time}, albeit at a significantly worse rate of convergence; by shifting up the expansion $g_{\sigma}^{(m)} + \rho$ with a large enough $\rho > 0$, at the loss of most of the low-rank structure of $g_{\sigma}(t \mtx{I}_n - \mtx{A})$; or by directly minimizing the approximation error over all non-negative Chebyshev polynomials \cite{fejer-1916-uber-trigonometrische}, which realistically only works for small $m$.
-In the end, we propose the following method to produce a non-negative Chebyshev expansion: First, expand the square root of the smoothing kernel in a basis of Chebyshev polynomials up to degree $m/2$
-\begin{equation}
-    \sqrt{g_{\sigma}}^{(\sfrac{m}{2})}(t - s) = \sum_{l=0}^{m/2} \xi_l(t) T_l(s).
-\end{equation}
-Second, square this expansion with \refalg{alg:chebyshev-squaring} to obtain the non-negative Chebyshev expansion
-\begin{equation}
-    \underline{g}^{(m)}_{\sigma}(t - s) = \sum_{l=0}^{m} \mu_l(t) T_l(s).
-    \label{equ:non-negative-chebyshev-expansion}
-\end{equation}
-With a complexity of $\mathcal{O}(m \log(m))$ this procedure is computationally inexpensive and allows for very similar convergence bounds to \reflem{lem:chebyshev-error}, which we will derive in \refsec{subsubsec:chebyshev-nystrom-analysis}.
-In \reftab{tab:chebyshev-timing-interpolation}, the non-negative expansion can be seen to only be slightly slower than the normal one.
-
-\begin{table}[ht]
-    \caption{We compare the runtime in milliseconds for computing the coefficients of the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ \refequ{equ:smoothing-kernel} in three different ways: as proposed in \cite[algorithm 1]{lin-2017-randomized-estimation} (\emph{oversampled FFT}), using the discrete cosine transform (\emph{DCT}), and computing a non-negative Chebyshev expansion with \refalg{alg:chebyshev-squaring} (\emph{non-negative DCT}).
-We use $\sigma=0.005$, $n_t=1000$ parameter values, and various degrees $m$.
-For stability, we average over 7 runs of the algorithms and repeat these runs 1000 times to form the mean and standard deviation which are given in the below table.}
-    \label{tab:chebyshev-timing-interpolation}
-   \input{tables/interpolation.tex}
-\end{table}
-
-%\todo{
-%Maybe mention indefinite Nyström [Nakatsukasa], which won't give us desirable guarantees, and refinement of Chebyshev expansion [Francis Bach], which only slightly improves, but is much harder to implement.
-Side-note: it is possible to refine the coefficients of the Chebyshev expansion to get a better approximation in the supremum-norm over the set of all non-negative polynomials \cite%{fejer-1916-uber-trigonometrische}.
-The non-convex nature of this problem and the relatively small improvement we get with this technique made us decide to not pursue this further.
-%}
-
-In the end, we propose the following approximation for the smoothed spectral density
-\begin{equation}
-    \widetilde{\phi}_{\sigma}^{(m)}(t) = \frac{1}{n} \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}))
-    \label{equ:chebyshev-nystrom-formula}
-\end{equation}
-which is the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$ \refequ{equ:nystrompp-trace-estimator} applied to the non-negative Chebyshev expansion $\underline{g}_{\sigma}^{(m)}$ \refequ{equ:non-negative-chebyshev-expansion}.
-
-\subsubsection{Implementation}
-\label{subsubsec:chebyshev-nystrom-implementation}
-
-An efficient implementation of the approximation \refequ{equ:chebyshev-nystrom-formula} can be achieved thanks to the invariance of the trace under cyclic permutation of its arguments and the recurrence relation \refequ{equ:chebyshev-recurrence} which the Chebyshev polynomials satisfy.
-To illustrate this, consider the approximation \refequ{equ:chebyshev-nystrom-formula} when $\mtx{\Psi}$ is zero.
-Using the definition of the Nyström approximation \refequ{equ:nystrom-approximation}, the cyclic invariance of the trace, and the symmetry of $\mtx{A}$, we can rewrite it as
-\begin{equation}
-    \frac{1}{n} \Trace((\mtx{\Omega}^{\top} \underline{g}^{(m)}_{\sigma}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega})^{\dagger} (\mtx{\Omega}^{\top} \underline{g}^{(m)}_{\sigma}(t \mtx{I}_n - \mtx{A})^2 \mtx{\Omega})).
-    \label{equ:cyclic-property}
-\end{equation}
-Thanks to the affine form of the Chebyshev expansion, the first parenthesized term in the trace is
-\begin{equation}
-    \mtx{\Omega}^{\top} \underline{g}^{(m)}_{\sigma}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega} = \sum_{l=0}^{m} \mu_l(t) (\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega})
-\end{equation}
-for which we can first precompute the small $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ for $l=0, \dots, m$ using the recurrence \refequ{equ:chebyshev-recurrence} and then inexpensively sum them up with the corresponding coefficients $\mu_l(t)$ for all $t$ in which we want to evaluate the approximation.
-Thanks to the fast and exact squaring of Chebyshev expansions (cf.
-\refalg{alg:chebyshev-squaring}), the second parenthesized term in \refequ{equ:cyclic-property} can be approximated in an equivalent way.
-Unlike expanding the squared smoothing kernel $g_{\sigma}^2$ separately \cite{lin-2017-randomized-estimation}, this approach is consistent, which we have observed to make a noticeable difference in terms of accuracy (cf.
-\reffig{fig:interpolation-issue}), and which enables provability.
-
-\begin{figure}[ht]
-    \centering
-    \input{plots/interpolation.pgf}
-    \caption{Difference in the accuracy when computing the estimate \refequ{equ:cyclic-property} once by separately expanding the squared smoothing kernel $g_{\sigma}^2$ (\emph{inconsistent}), once by squaring the Chebyshev expansion $g_{\sigma}^{(m)}$ using \refalg{alg:chebyshev-squaring} (\emph{consistent}), and once with a non-negative Chebyshev expansion $\underline{g}_{\sigma}^{(m)}$ \refequ{equ:non-negative-chebyshev-expansion} which is also squared with \refalg{alg:chebyshev-squaring} (\emph{non-negative}).
-We use a small example matrix from \refsec{sec:results}, fix the size of the Nyström approximation to $n_{\mtx{\Omega}} = 80$, let $\sigma = 0.005$, and compute the $L^1$ error from the approximation of the spectral density at $n_t = 100$ uniformly spaced values of $t$.}
-    \label{fig:interpolation-issue}
-\end{figure}
-
-Similar observations for the case when $\mtx{\Psi}$ is nonzero lead to the following algorithm.
-
-\begin{algo}{Chebyshev-Nyström++ method}{nystrom-chebyshev-pp}
-\begin{algorithmic}[1]
-    \Statex \textbf{Input:} Symmetric $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum in $[-1, 1]$, points $\{t_i\}_{i=1}^{n_t} \subset \mathbb{R}$
-    \Statex \textbf{Parameters:} Degree of expansion $m \in 2\mathbb{N}$, number of Girard-Hutchinson queries $n_{\mtx{\Psi}} \in \mathbb{N}_0$, Nyström sketch size $n_{\mtx{\Omega}} \in \mathbb{N}_0$,  smoothing parameter $\sigma > 0$
-    \Statex \textbf{Output:} Approximate evaluations of the spectral density $\{\widetilde{\phi}_{\sigma}(t_i)\}_{i=1}^{n_t}$
-    \State Compute $\{\mu_l(t_i)\}_{l=0}^{m}$ for all $t_i$ with non-negative Chebyshev expansion \refequ{equ:non-negative-chebyshev-expansion}
-    \State Compute $\{\nu_l(t_i)\}_{l=0}^{2m}$ using \refalg{alg:chebyshev-squaring}%\refalg{alg:2-chebyshev-chebyshev-expansion}
-    %\State Compute  for all $t_i$ using %\refalg{alg:3-nystrom-chebyshev-exponentiation}
-    \State Generate standard Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$%\glsfirst{sketching-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$
-    %\State Generate standard Gaussian %\glsfirst{random-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$
-    \State Initialize $[\mtx{V}_1, \mtx{V}_2, \mtx{V}_3] \gets [\mtx{0}_{n \times n_{\mtx{\Omega}}}, \mtx{\mtx{\Omega}}, \mtx{0}_{n \times n_{\mtx{\Omega}}}]$
-    \State Initialize $[\mtx{W}_1, \mtx{W}_2, \mtx{W}_3] \gets [\mtx{0}_{n \times n_{\mtx{\Psi}}}, \mtx{\Psi}, \mtx{0}_{n \times n_{\mtx{\Psi}}}]$
-    \State Initialize $[\mtx{K}_1(t_i), \mtx{K}_2(t_i)] \gets [\mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}}, \mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}}]$ for all $t_i$
-    \State Initialize $[\mtx{L}_1(t_i), \ell(t_i)] \gets [\mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Psi}}}, 0]$ for all $t_i$
-    %\State Set $\breve{\phi}_{\sigma}^{(m)}(t_i) \gets 0$ for all $t_i$
-    \For {$l = 0, \dots, 2m$}
-    \State $[\mtx{X}, \mtx{Y}] \gets \mtx{\mtx{\Omega}}^{\top} [\mtx{V}_2, \mtx{W}_2]$  
-    %\State $\mtx{X} \gets \mtx{\mtx{\Omega}}^{\top} \mtx{V}_2$
-      %\State $\mtx{Y} \gets \mtx{\mtx{\Omega}}^{\top} \mtx{W}_2$
-      \State $z \gets \Trace(\mtx{\Psi}^{\top} \mtx{W}_2)$
-      \For {$i = 1, \dots, n_t$}
-        \If {$l \leq m$}
-            \State $\mtx{K}_1(t_i) \gets \mtx{K}_1(t_i) + \mu_l(t_i) \mtx{X}$ \Comment{assemble $\mtx{\Omega}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$}
-            \State $\mtx{L}_1(t_i) \gets \mtx{L}_1(t_i) + \mu_l(t_i) \mtx{Y}$ \Comment{assemble $\mtx{\Omega}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi}$}
-            \State $\ell(t_i) \gets \ell(t_i) + \mu_l(t_i) z$ \Comment{assemble $\Trace(\mtx{\Psi}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi})$}
-        \EndIf
-        \State $\mtx{K}_2(t_i) \gets \mtx{K}_2(t_i) + \nu_l(t_i) \mtx{X}$ \Comment{assemble $\mtx{\Omega}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})^2 \mtx{\Omega}$}
-      \EndFor
-      \State $[\mtx{V}_3, \mtx{W}_3] \gets (2 - \delta_{l0}) \mtx{A} [\mtx{V}_2, \mtx{W}_2] - [\mtx{V}_1, \mtx{W}_1]$ \Comment{Chebyshev recurrence}
-      \State $[\mtx{V}_1, \mtx{W}_1] \gets [\mtx{V}_2, \mtx{W}_2]$
-      \State $[\mtx{V}_2, \mtx{W}_2] \gets [\mtx{V}_3, \mtx{W}_3]$
-      %\State $\mtx{V}_3 \gets (2 - \delta_{l0}) \mtx{A} \mtx{V}_2 - \mtx{V}_1$ %\Comment{Chebyshev recurrence \refequ{equ:2-chebyshev-chebyshev-recursion}}
-      %\State $\mtx{V}_1 \gets \mtx{V}_2, \mtx{V}_2 \gets \mtx{V}_3$
-      %\State $\mtx{W}_3 \gets (2 - \delta_{l0}) \mtx{A} \mtx{W}_2 - \mtx{W}_1$ %\Comment{Chebyshev recurrence \refequ{equ:2-chebyshev-chebyshev-recursion}}
-      %\State $\mtx{W}_1 \gets \mtx{W}_2, \mtx{W}_2 \gets \mtx{W}_3$
-    \EndFor
-    \For {$i = 1, \dots, n_t$}
-      \State $\widetilde{\phi}_{\sigma}(t_i) \gets \frac{1}{n} \Trace\left( \mtx{K}_1(t_i)^{\dagger}\mtx{K}_2(t_i) \right) + \frac{1}{n n_{\mtx{\Psi}}} \left( \ell(t_i) + \Trace\left( \mtx{L}_1(t_i)^{\top} \mtx{K}_1(t_i)^{\dagger} \mtx{L}_1(t_i) \right)  \right) $ \label{lin:4-nystromchebyshev-nystrom-pp}
-    \EndFor
-\end{algorithmic}
-\end{algo}
-
-The dominating cost stems from the $(2m + 1)(n_{\mtx{\Psi}} + n_{\mtx{\Omega}})$ matrix-vector multiplications with $\mtx{A}$ and the cost associated with computing the estimate in each evaluation point $t_1, \dots, t_{n_t}$ given by $\mathcal{O}(m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + (n_{\mtx{\Omega}}^3 + n_{\mtx{\Omega}} n_{\mtx{\Psi}}^2 + 1))$.
-
-The algorithm also accomodates the two other estimators introduced in \refsec{sec:analysis}: For $n_{\mtx{\Omega}} = 0$ we recover the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}, and the complexity for each evaluation of the estimate decreases significantly.
-For $n_{\mtx{\Psi}} = 0$ we end up with the Nyström estimator \refequ{equ:nystrom-trace-estimator}, provided the implementation evaluates $0/0$ to $0$ in the second term on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
-Notice that in this case, \refalg{alg:nystrom-chebyshev-pp} is equivalent to the \enquote{spectrum sweeping} method from \cite[algorithm 5]{lin-2017-randomized-estimation}, but has a significantly lower computational complexity.
-
-%Chebyshev $\mathcal{O}(n_t (m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + n_{\mtx{\Omega}}^3) + mn^2(n_{\mtx{\Psi}} + n_{\mtx{\Omega}}))$.
-
-%\todo{Complexity in terms of both $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$.}
-%With the cost of a matrix-vector product denoted by $c(n)$, and supposing $n_{\mtx{\Omega}} \approx n_{\mtx{\Psi}}$, we determine the computational complexity of the Chebyshev-Nyström++ method to be $\mathcal{O}(m \log(m) n_t + m n_{\mtx{\Omega}}^2 n + m n_t n_{\mtx{\Omega}}^2 +  m c(n) n_{\mtx{\Omega}} + n_t n_{\mtx{\Omega}}^3)$, with $\mathcal{O}(m n_t + n n_{\mtx{\Omega}} + n_{\mtx{\Omega}}^2 n_t)$ required additional storage.
-
-%\todo{Explain modifications for other estimators, and improved complexity for quadratic trace estimator}
-
-The pseudoinverses on \reflin{lin:4-nystromchebyshev-nystrom-pp} need to be computed with care.
-It is possible to reformulate the problem as a generalized eigenvalue problem and apply thresholding to the smallest eigenvalues of a certain matrix to enforce stability \cite{lin-2017-randomized-estimation, epperly-2022-theory-quantuma}.
-Since this renders the algorithm significantly more complicated but only improved the accuracy slightly during our experiments, we will instead evaluate these expressions in \reflin{lin:4-nystromchebyshev-nystrom-pp} using a least-squares solver which likewise uses a truncation of the smallest eigenvalues to enforce better conditioning.
-
-What we have found to be crucial for a successful approximation of the full spectral density was the detection of regions wherein lie no eigenvalues, i.e.
-regions where the spectral density nearly vanishes.
-For small smoothing parameters $\sigma$, the rapid decay of the Gaussian $g_{\sigma}$ will cause $\mtx{K}(t_i) \approx g_{\sigma}(t_i\mtx{I}_n - \mtx{A})$ to be close to the zero matrix for parameter values $t_i$ located in such regions.
-Hence, on line \reflin{lin:4-nystromchebyshev-nystrom-pp} we would compute the pseudoinverse of an almost zero matrix, which can be extremely unstable (see \reffig{fig:zerocheck}).
-Notice that $\Trace(\mtx{K}_1(t_i)) / n_{\mtx{\Omega}}$ is the $n_{\mtx{\Omega}}$-query Girard-Hutchinson estimator.
-We may use it as a rough indicator whether $t_i$ lies within a region where the spectral density nearly vanishes, and subsequently set $\widetilde{\phi}_{\sigma}(t_i) = 0$ instead of computing the full expression on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
-
-\begin{figure}[ht]
-    \centering
-    \input{plots/zerocheck.pgf}
-    \caption{The difference the non-zero check can make when approximating a spectral density using \refalg{alg:nystrom-chebyshev-pp}.
-Here, we used a matrix described in \refsec{sec:results} and ran the Chebyshev-Nyström++ method with and without estimating if the matrix function is zero before taking its pseudo-inverse.
-We let $m=2000$, $n_{\mtx{\Omega}}=80$, $n_{\mtx{\Psi}}=0$, and $\sigma = 0.004$.}
-    \label{fig:zerocheck}
-\end{figure}
-
-\subsubsection{Analysis}
-\label{subsubsec:chebyshev-nystrom-analysis}
-
-In this section, we will upper bound the output of the Chebyshev-Nyström++ method (see. \refalg{alg:nystrom-chebyshev-pp}). To this extent, we will first analyze the non-negative Chebyshev expansion.
-
-\begin{lemma}{Non-negative Chebyshev expansion error}{non-negative-chebyshev-error}
-    The non-negative expansion $\underline{g}_{\sigma}^{(m)}$ of the Gaussian smoothing kernel satisfies
-    \begin{equation}
-        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \leq 2\sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right) E_{\sqrt{2}\sigma, \sfrac{m}{2}} \equiv \underline{E}_{\sigma, m}
-        \label{equ:chebyshev-interpolation-sup-error-kernel-nonneg}
-        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
-        %\label{equ:2-chebyshev-interpolation-error}
-    \end{equation}
-    for all \emph{even} degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, $s \in [-1, 1]$, and with $E_{\sigma, m}$ as defined in \reflem{lem:chebyshev-error}.
-\end{lemma}
-
-\begin{proof}
-    For any numbers $a, b \in \mathbb{R}$ it holds
-    \begin{equation}
-    | a^2 - b^2 | = | (a + b)(a - b) | = | a + b | | a - b | \leq (2 | a | + | a - b |)  | a - b |
-    \end{equation}
-    Therefore, omitting the arguments and using $g_{\sigma} = (\sqrt{g_{\sigma}})^2$ and $\underline{g}_{\sigma}^{(m)} = (\sqrt{g_{\sigma}}^{(\sfrac{m}{2})})^2$.
-    we have
-    \begin{equation}
-        \left| g_{\sigma} - \underline{g}_{\sigma}^{(m)} \right| \leq \left( 2 \left| \sqrt{g_{\sigma}} \right| + \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right| \right) \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right|.
-    \end{equation}
-    Since $\sqrt{g_{\sigma}} = \sqrt{2 \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$ and $\sqrt{g_{\sigma}} \leq 1/\sqrt{\sigma \sqrt{2 \pi}}$ we can apply the result from \refthm{thm:chebyshev-error} to get
-    \begin{align}
-        &\sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \notag \\
-        &\leq \left( 2 \frac{1}{\sqrt{ \sigma \sqrt{2\pi}}} + \sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right)\sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}},
-    \end{align}
-    from which follows the result with some minor simplifications.
-\end{proof}
-
-On one hand, in the non-negative expansion, the underlying computes Chebyshev expansion -- which is shown in \reflem{lem:chebyshev-error} to exponentially decay with its degree -- is only computed to degree $m/2$, whereas the standard expansion is of degree $m$.
-On the other hand, we approximate $\sqrt{g_{\sigma}}$ instead of $g_{\sigma}$, which is less \enquote{pointy} and therefore better approximated by polynomials.
-Overall, we will need to choose the degree $m$ of the expansion approximately a factor of $\sqrt{2}$ larger to match the error of the standard expansion.
-
-\begin{theorem}{Chebyshev-Nyström++ method}{chebyshev-nystrom}
-    Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, \emph{even} $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2)$, and smoothing parameter $0 < \sigma \ll 1$ satisfies with probability at least $1 - \delta$
-    \begin{equation}
-        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq \varepsilon
-        \label{equ:chebyshev-nystrom-error}
-        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
-        %\label{equ:2-chebyshev-interpolation-error}
-    \end{equation}
-    for all $\varepsilon \in (0, 1)$ and $\delta \in (0, e^{-\sfrac{1}{2}}/2)$.
-\end{theorem}
-
-\begin{proof}
-    The previously derived results can be applied to
-    \begin{align}
-        &\int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \notag \\
-        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \int_{-1}^{1} \left| \phi^{(m)}_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
-        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \widetilde{\varepsilon} \int_{-1}^{1} \left| \phi^{(m)}_{\sigma}(t) \right|~\mathrm{d}t && \text{(\refthm{thm:nystrom-pp} w.p.
-$\geq 1 - \delta$)} \notag \\
-        &\leq (1 + \widetilde{\varepsilon}) \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \varepsilon \int_{-1}^{1} \left| \phi_{\sigma}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
-        &\leq (1 + \widetilde{\varepsilon}) 2 \underline{E}_{\sigma, m} + \widetilde{\varepsilon}. && \text{(\refthm{thm:chebyshev-error}, normalization)} \notag \\
-    \end{align}
-    From \reflem{lem:non-negative-chebyshev-error} we see that if $0 < \sigma \ll 1$, choosing $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$ ensures $\underline{E}_{\sigma, m} \leq \widetilde{\varepsilon}$, and thus, since $\widetilde{\varepsilon} \leq 1$,
-    \begin{equation}
-        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq (1+\widetilde{\varepsilon}) 2 \widetilde{\varepsilon} + \widetilde{\varepsilon} \leq 5 \widetilde{\varepsilon}
-    \end{equation}
-    from which the result follows by identifying $\varepsilon = 5 \widetilde{\varepsilon}$.
-\end{proof}
-
-%\todo{Maybe specialize \refthm{thm:nystrom} to spectral densities?}
-
-%We now discuss what value to choose for $n_{\mtx{\Omega}}$ in the Nyström trace estimator introduced in \refsec{subsec:nystrom}.
-In order to guarantee an even better approximation error based on \refthm{thm:nystrom}, we analyze the numerical rank of the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
-
-In practice, the approximation error is often significantly better than guaranteed in \refthm{thm:chebyshev-nystrom}, particularly when $n_{\mtx{\Omega}}$ is chosen rather large.
-This is because the singular values of the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ decay so quickly that already the Nyström approximation alone is highly accurate (cf.
-\refthm{thm:nystrom}).
-To establish a relation between the singular value decay of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ and what choice of $n_{\mtx{\Omega}}$ is needed to achieve a better approximation than what we show in \refthm{thm:chebyshev-nystrom}, we study the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
-
-We can express the numerical rank of a matrix with respect to the nuclear norm in terms of its singular values $\sigma_1 \geq \dots \geq \sigma_n \geq 0$
-\begin{equation}
-    r_{\varepsilon} = \min \{1 \leq r \leq n: \sum_{j=r+1}^n \sigma_{j} \leq \varepsilon \},
-    \label{equ:numerical-rank}
-\end{equation}
-$\varepsilon$ is usually taken to be the double machine precision, i.e. $10^{-16}$.
-
-%\begin{theorem}{Nyström estimator for parameter-dependent matrices}%{chebyshev-nystrom-raw}
-%    Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric %matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is %contained in $[-1, 1]$.
-Then the Chebyshev-Nyström++ estimate %$\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m \in %\mathbb{N}$, $n_{\mtx{\Omega}} \geq \max(6, \max_{t \in [-1, 1]}(r_%{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A}))) + \log(\delta^%{-1}))$, $n_{\mtx{\Psi}} = 0$, and smoothing parameter $\sigma > 0$ %satisfies with for all $\gamma \geq 1$ with probability at least $1 %- \delta$
-%    \begin{equation}
-%        \int_{a}^{b} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}%(t) \right| ~\mathrm{d}t
-%        \leq 2 e^2 (1 + r_{\varepsilon}) (\varepsilon + (n - r_%{\varepsilon}) \underline{E}_{\sigma, m}) + \underline{E}_%{\sigma, m}.
-%    \end{equation}
-%\end{theorem}
-%
-%\begin{proof}
-%    \todo{TODO}
-%\end{proof}
-
-When using a Gaussian smoothing-kernel, we can explicitly write the singular values of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ as
-\begin{equation}
-    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) = \frac{1}{\sqrt{2 \pi \sigma^2}} e^{-\sfrac{(t - \lambda_{(i)})^2}{2 \sigma^2}},
-    \label{equ:gaussian-kernel-eigenvalues}
-\end{equation}
-where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from spectral-parameter $t$, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t)$.
-Consequently, by using the closed-form expression of the eigenvalues \refequ{equ:gaussian-kernel-eigenvalues}, we may upper bound the numerical rank of a Gaussian smoothing kernel as
-\begin{equation}
-    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A})) \leq \#\left\{1\leq i\leq n: |t - \lambda_i| < d_{\varepsilon, \sigma} \right\},
-    \label{equ:gaussian-kernel-numerical-rank}
-\end{equation}
-with $d_{\varepsilon, \sigma} = \sigma \sqrt{-2 \log(\sqrt{2 \pi} \sigma \varepsilon)}$ and the eigenvalues $\lambda_1, \dots, \lambda_n$ of $\mtx{A}$ and the constants.
-The expression \refequ{equ:gaussian-kernel-numerical-rank} has a nice visual interpretation: The numerical rank of a matrix is at most equal to the number of eigenvalues which are closer to the parameter $t$ than $d_{\varepsilon, \sigma}$.
-This is illustrated in \reffig{fig:numerical-rank}.
-\begin{figure}[ht]
-    \centering
-    \input{figures/numerical-rank.tex}
-    \caption{The numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ can be
-        approximately computed by counting the number of eigenvalues
-        $\lambda_{(1)}, \dots, \lambda_{(n)}$ of the matrix $\mtx{A}$ which lie less than
-        a constant $d_{\varepsilon, \sigma}$ away from $t$.}
-    \label{fig:numerical-rank}
-\end{figure}
-
-If we additionally assume the eigenvalues of the matrix $\mtx{A}$ to be evenly distributed within $[-1, 1]$, that is, in any subinterval of fixed length in $[a, b]$, $-1 \leq a < b \leq 1$, we can expect to find roughly the same number of eigenvalues, then we can estimate the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ to be
-\begin{equation}
-    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A})) \lessapprox n d_{\varepsilon, \sigma} \approx 6 n \sigma,
-    \label{equ:gaussian-kernel-numerical-rank-uniform}
-\end{equation}
-provided we set $\varepsilon = 10^{-16}$ and choose $\sigma$ roughly between $10^{-8}$ and $1$.
-Hence, choosing $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma)$ will approximations which are significantly better than \refthm{thm:chebyshev-nystrom}.
-
-\section{Numerical results}
-\label{sec:results}
-
-Our first numerical example comes from electronic structure interaction \cite{lin-2017-randomized-estimation}, more precisely from the second order finite difference discretization of the Hamiltonian
-\begin{equation}
-    \mathcal{H} = - \Delta + V
-    \label{equ:5-experiments-electronic-hamiltonian}
-\end{equation}
-in three dimensions. The potential $V$ interacting with the electrons is generated by Gaussian wells
-\begin{equation}
-    v(r) = v_0 e^{-\lambda r^2}
-    \label{equ:5-experiments-gaussian-cell}
-\end{equation}
-with $v_0 = -4$ and $\lambda = 8$, centered in cells of side length $L=6$ which are stacked $n_c \in \mathbb{N}$ times in each spatial dimension (cf.
-\reffig{fig:gaussian-well}).
-The discretization step is fixed to be $h=0.6$, such that the size of the matrix grows cubically with $n_c$.
-This is an idealized model for the interaction of nuclei on a regular grid with electrons for a $k$-vector in the center of the first Brillouin zone.
-The distribution of the eigenvalues of the Hamiltonian -- its spectral density -- allows us to interpret the system's energy levels.
-
-\begin{figure}[ht]
-    \begin{subfigure}[b]{0.32\columnwidth}
-        \input{plots/gaussian-well-1.pgf}
-        \caption{$n_c=1$}
-        \label{fig:gaussian-well-1}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.32\columnwidth}
-        \input{plots/gaussian-well-2.pgf}
-        \caption{$n_c=2$}
-        \label{fig:gaussian-well-2}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.32\columnwidth}
-        \input{plots/gaussian-well-5.pgf}
-        \caption{$n_c=5$}
-        \label{fig:gaussian-well-5}
-    \end{subfigure}
-    \caption{Cross-sections of the periodic Gaussian well potential $V$ for different sizes of the supercell.}
-    \label{fig:gaussian-well}
-\end{figure}
-
-\todo{Explain experimental setup.}
-
-\begin{figure}[ht]
-    \centering
-    \input{plots/convergence.pgf}
-    \caption{For increasing values of $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}$ but fixed $m$ we plot the $L^1$-approximation error for $\sigma=0.005$ on the model problem with $n_c = 1$.}
-    \label{fig:convergence}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \input{plots/distribution.pgf}
-    \caption{The Chebyshev-Nyström++ method for different ways of allocations a total of $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation for the Gaussian smoothing kernel with multiple different values of the smoothing parameter.
-We make the approximation error made in the Chebyshev expansion negligible by rescaling $m=16 / \sigma$ (based on \refthm{thm:chebyshev-error}).}
-    \label{fig:distribution}
-\end{figure}
-
-\todo{Interpret results.}
-
-\todo{Comparison plots with other estimators (Krylov-Aware)?}
-
-\clearpage
 \section{Conclusion}
@@
 
-\clearpage
+%\clearpage
 %\bibliography{bibliography.bib}
@@
 
-\clearpage
+%\clearpage
 \section{Proofs}
+
+\textcolor{red}{??? Need to check carefully that this result is not known ???}
 

### 2024-07-07 paper/spectraldensity.tex
Updates from Overleaf

--- 2024-06-27 paper/main.tex
@@
-\documentclass[12pt]{article}
-
-\usepackage{stylesheet}
-
-\title{Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation}
-
-\begin{document}
-
-\maketitle
-
-%\todo{
-%\begin{itemize}
-%    \item Unify index notation
-%    \item Double-check proofs
-%    \item Add all references
-%    \item Run numerical experiments 
-%    \item Write introduction, abstract, and discussion
-%    \item ArXiv HTML support (check if packages)
-%\end{itemize}
-%}
-
-%\todo{
-%    \tableofcontents
-%}
-
-\todo{
-\begin{abstract}
-    Write abstract.
-\end{abstract}
-}
-
-\section{Introduction}
-\label{sec:introduction}
-
-In numerous problems across physics, chemistry, engineering, and data science, how the eigenvalues of certain matrices are distributed can be indicative of what properties a system has: In electronic structure calculations eigenvalues represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}.
-The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$ with the Dirac delta distribution $\delta$.
-Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation which is often prohibitively expensive.
-Therefore, many techniques have been proposed for approximating $\phi$.
-For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature rule with the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}.
-Further, because $\delta$ is hard to approximate and in many applications, a rough estimate of the spectral density is already acceptable, a standard approach is to instead approximate the smoothed spectral density
-\begin{equation}
-    \phi_{\sigma}(t) = \frac{1}{n} \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i) = \frac{1}{n} \Trace(g_{\sigma}(t \mtx{I}_n - \mtx{A}))
-    \label{equ:smoothed-spectral-density}
-\end{equation}
-for a smoothing kernel $g_{\sigma}$, typically a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}.
-This translates the problem to computing the trace of a continuous parameter-dependent matrix $\mtx{B}(t) \equiv g_{\sigma}(t \mtx{I}_n - \mtx{A}) \in \mathbb{R}^{n \times n}$.
-Directly assembling the matrix function requires diagonalizing the matrix $\mtx{A}$, which again is prohibitively expensive.
-Therefore, we estimate its trace implicitly by approximating products of the matrix function with vectors.
-For constant matrices, the most widespread methods for implicit trace estimation are the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} and the variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making, chen-2023-krylovaware-stochastic}.
-Some of these estimators have already been successfully applied to parameter-dependent matrices \cite{lin-2017-randomized-estimation,chen-2023-krylovaware-stochastic}, but they have not yet been analyzed in this setting.
-A related technique is dynamic trace estimation for sequences $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ of matrices which satisfy $\lVert \mtx{B}(t_i) \rVert _F \leq 1$ and which change dynamically, i.e.
-$\lVert \mtx{B}(t_{i+1}) - \mtx{B}(t_i) \rVert _F \leq \alpha < 1$ for the Frobenius norm $\lVert \cdot \rVert _F$ and a small $\alpha$ \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace}.
-However, in our setting we often cannot guarantee either of the two conditions.
-Instead of discretizing $[a, b]$ into $t_1, \ldots, t_m$, approximating $\Trace(\mtx{B}(t_i))$ with $b(t_i)$, and analyzing the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way is to quantify the error with $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate in each $t \in [a, b]$.
-This approach has already been taken for analyzing low-rank approximations of parameter-dependent matrices \cite{kressner-2023-randomized-lowrank}.
-
-\paragraph{New contributions} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}.
-Interestingly, we can reuse the same randomization for computing the estimate at each value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations.
-We propose certain modifications to the methods from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}.
-In particular, we use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials and unify all of their methods in a single algorithm, with the desirable side effect of speeding up their \enquote{spectrum sweeping} method by orders of magnitude.
-Based on the analysis of the randomized trace estimators for parameter-dependent matrices we provide bounds on the approximation error for our method.
-
-\paragraph{Structure} In \refsec{sec:analysis} we will analyze three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter dependent matrices.
-We propose and analyze a method for using the aforementioned estimators to estimate the spectral density of real symmetric matrices in \refsec{sec:application}.
-Finally, we will test our developments on an example from electronic structure calculation.
-
-\paragraph{Reproducibility} \input{re-pro-badge.tex}
-
-\section{Analysis of randomized trace estimators applied to parameter-dependent matrices}
-\label{sec:analysis}
-
-We consider parameter-dependent matrices of the form
-\begin{equation}
-    \mtx{B}(t) = \begin{bmatrix}
-        b_{11}(t) & b_{12}(t) & \dots & b_{1n}(t) \\
-        b_{21}(t) & b_{22}(t) & \dots & b_{2n}(t) \\
-        \vdots & \vdots & \ddots & \vdots \\
-        b_{n1}(t) & b_{n2}(t) & \dots & b_{nn}(t) \\
-    \end{bmatrix} \in \mathbb{R}^{n \times n}
-\end{equation}
-where $b_{ij}(t)$ are functions depending continuously on the parameter $t$ which takes values in the interval $[a,b]$. The trace of such a matrix is defined as
-\begin{equation}
-    \Trace(\mtx{B}(t)) = \sum_{i=1}^{n} b_{ii}(t).
-    \label{equ:definition-trace}
-\end{equation}
-However, we assume that we only have access to products of this matrix with vectors for each $t \in [a, b]$, so this definition will not be directly useful for computing the trace.
-
-\paragraph{Girard-Hutchinson estimator} We can approximate the trace with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ stochastically independent standard Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ to form
-\begin{equation}
-    \Hutch{\mtx{\Psi}}(\mtx{B}(t))
-    = \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j
-    = \frac{1}{n_{\mtx{\Psi}}} \Trace( \mtx{\Psi}^{\top} \mtx{B}(t) \mtx{\Psi})
-    \label{equ:hutchinson-trace-estimator}
-\end{equation}
-where $\mtx{\Psi} = [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\mtx{\Psi}}}] \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$.
-Other choices for the distribution of the random vectors are possible, for example by uniformly sampling from $\{-1, +1\}$ or from the $(n-1)$-sphere.
-However, our theoretical developments only hold in the Gaussian case.
-
-\paragraph{Nyström estimator} Alternatively, the trace of a symmetric matrix whose singular values decay quickly can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström approximation \cite{gittens-2013-revisiting-nystrom}
-\begin{equation}
-    \Nystr{\mtx{\Omega}}{\mtx{B}}(t) = (\mtx{B}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} (\mtx{B}(t) \mtx{\Omega})^{\top}.
-    \label{equ:nystrom-approximation}
-\end{equation}
-Then we can estimate the trace as $\Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t))$.
-Thanks to the invariance of the trace under cyclic permutation of its arguments and the symmetry of the matrix, we may rewrite this estimator as
-
-\begin{equation}
-    \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{B}(t)^2 \mtx{\Omega})).
-    \label{equ:nystrom-trace-estimator}
-\end{equation}
-
-\paragraph{Nyström++ estimator} Finally, \cite{lin-2017-randomized-estimation} proposes an estimator which corrects for inaccuracies in the Nyström approximation \refequ{equ:nystrom-trace-estimator} by estimating the trace of its residual using the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}
-\begin{equation}
-    \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) = \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) + \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)).
-    \label{equ:nystrompp-trace-estimator}
-\end{equation}
-This is the parameter-dependent analogue of the Nyström++ estimator \cite{persson-2022-improved-variants}, which is based on the Hutch++ estimator \cite{meyer-2021-hutch-optimal}.
-%We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.
-
-%In the remainder of this section, we will derive upper bounds on the error of these estimators.
-%
-%\begin{table}[ht]
-%\centering
-%\renewcommand{\arraystretch}{1.2}
-%\begin{tabular}{@{}lcc@{}}
-%\toprule
-%Estimator & Matrix & $m=1600$\\
-%\midrule
-%Girard-Hutchinson & symmetric & $\mathcal{O}(\varepsilon^{-2})%$ \\
-%Nyström & symmetric positive semi-definite & $\mathcal{O}%(\varepsilon^{-2})$ \\
-%Nyström++ & symmetric positive semi-definite & $\mathcal{O}%(\varepsilon^{-1})$  \\
-%\bottomrule
-%\end{tabular}
-%\end{table}
-
-\subsection{Girard-Hutchinson estimator for parameter-dependent matrices}
-\label{subsec:hutchinson}
-
-We first analyze the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}.
-We bound its $L^1$ norm in terms of the integral over the Frobenius norm $\lVert \cdot \rVert _F$ of the matrix whose trace we aim to approximate.
-
-\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
-    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric and continuously depend on $t \in [a, b]$.
-Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ it holds with probability at least $1 - \gamma^{-2k}$
-    \begin{equation}
-        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) \right| ~\mathrm{d}t \leq 2^4 \gamma k \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F  ~\mathrm{d}t.
-    \end{equation}
-    In particular, if we choose $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{- \sfrac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$.
-\end{theorem}
-
-The idea behind the proof is to bound the higher order moments of the residual $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ for fixed values of $t$ and then use Minkowski's integral inequality to extend the results to the continuous parameter case.
-We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
-\begin{proof} 
-    We define the residual of the 1-query Girard-Hutchinson estimate 
-    \begin{equation}
-        r(\mtx{B}(t), \vct{\psi}) = \Trace(\mtx{B}(t)) - \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi}
-    \end{equation}
-    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$.
-
-    First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence.
-From the proof of \cite[lemma 3]{cortinovis-2022-randomized-trace} we know that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm.
-Thus, by \cite[theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
-    \begin{align}
-        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
-        &\leq k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k} \notag \\
-        &= k! 2^{4 k} \lVert \mtx{B} \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _2^{2 k}.
-    \end{align}
-    Since $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
-    \begin{equation}
-        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
-    \end{equation}
-    Stirling's approximation \cite{robbins-1955-remark-stirling} bonuds $(2 k)!
-< 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$.
-Consequently, the moments $\mathbb{E}^{k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{k} \right] \right)^{\sfrac{1}{k}}$ of $r(\mtx{B}, \vct{\psi})$ are limited by
-    \begin{equation}
-        \mathbb{E}^{2k}\left[ r(\mtx{B}, \vct{\psi}) \right]
-        \leq \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F \leq 2^4 k \lVert \mtx{B} \rVert _F,
-        \label{equ:hutchinson-trace-onequery-fixed}
-    \end{equation}
-    since it can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, due to the monotonous decrease of this expression in $k \in \mathbb{N}$.
-
-    Now we make the transition to the continuous.
-Following \cite{kressner-2023-randomized-lowrank}, we can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable.
-Therefore, Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting:
-    \begin{equation}
-        \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t  \right]
-        \leq \int_{a}^{b} \mathbb{E}^{2 k}\left[ r(\mtx{B}(t), \vct{\psi}) \right]~\mathrm{d}t
-        \leq 2^4 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
-    \end{equation}
-    Consequently, by Markov's inequality, for any $k \in \mathbb{N}$ and $\gamma \geq 1$, it holds with probability at least $1 - \gamma^{-2 k}$
-    \begin{equation}
-        \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t \leq 2^4 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
-        \label{equ:hutchinson-trace-onequery-uniform}
-    \end{equation}
-
-    To extend the bound to the $n_{\mtx{\Psi}}$-query quadratic trace estimator, we use a technique from the proof of \cite[theorem 1]{cortinovis-2022-randomized-trace}. Let
-    \begin{equation}
-        \widetilde{\mtx{B}}(t)
-        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
-            \mtx{B}(t) & & \\
-            & \ddots & \\
-            & & \mtx{B}(t)
-        \end{pmatrix}
-        \quad \text{and} \quad
-        \widetilde{\vct{\psi}} = \begin{pmatrix}
-            \vct{\psi}_1 \\
-            \vdots \\
-            \vct{\psi}_{n_{\mtx{\Psi}}}
-        \end{pmatrix}
-    \end{equation}
-    where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors.
-Then $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$, and
-    \begin{equation}
-        r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}}) = \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j.
-    \end{equation}
-    Hence, by applying \refequ{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$ we conclude that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
-    \begin{equation}
-        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j \right| ~ \mathrm{d}t
-        \leq 2^4 k \gamma \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
-    \end{equation}
-
-    Setting $\delta = \gamma^{-2 k}$ and choosing $k = \lceil \log(\delta^{-1}) \rceil$, where $\lceil x \rceil$ is the ceiling function which returns the greatest integer larger than or equal to $x$, we get that for all $\delta \in (0, e^{-\sfrac{1}{2}})$
-    \begin{equation}
-        k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\sfrac{1}{2\lceil \log(\delta^{-1}) \rceil}}
-        = \lceil \log(\delta^{-1}) \rceil e^{\sfrac{1}{2}\frac{\log(\delta^{-1})}{\lceil \log(\delta^{-1}) \rceil}}
-        \leq 2 \log(\delta^{-1}) e^{\sfrac{1}{2}},
-    \end{equation}
-    from which follows the second part of the theorem when $n_{\mtx{\Psi}} = (2^5 e^{\sfrac{1}{2}} \varepsilon^{-1} \log(\delta^{-1}))^2$.
-    %Consequently, with probability $1 - \delta$ we have
-    %\begin{equation}
-    %    \int_{a}^{b} \left| \frac{1}{n_{\mtx{\Psi}}} \sum_{i=1}^{n_{\mtx{\Psi}}} \vct{\psi}_i^{\top} \mtx{B}(t) \vct{\psi}_i - \Trace(\mtx{B}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\sfrac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
-    %\end{equation}
-\end{proof}
-
-%\todo{
-%\begin{remark}
-%    If additionally $\mtx{B}(t)$ is nonzero and symmetric positive semi-definite for all $t \in [a, b]$, then it follows from the proof of \refthm{thm:hutchinson} that with probability at least $1 - \delta$
-%    \begin{equation}
-%        \int_{a}^{b} \frac{| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) |}{\Trace(\mtx{B}(t))} ~\mathrm{d}t < \varepsilon,
-%    \end{equation}
-%    if $n_{\mtx{\Psi}} \geq 2^{10} \varepsilon^{-2} \log(\delta^{-1})^2 (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ with $\mu(t) = \lVert \mtx{B}(t) \rVert _2 / \Trace(\mtx{B}(t))$.
-Compared to the constant matrix equivalent \cite[remark 2]{cortinovis-2022-randomized-trace} we mainly observe a larger constant and an additional factor of $\log(\delta^{-1})$.
-%\end{remark}
-%}
-
-Compared to the constant case \cite[lemma 2.1]{meyer-2021-hutch-optimal} we need to choose the number of queries $n_{\mtx{\Psi}}$ by a factor of $\log(\delta^{-1})$ larger to achieve the same guarantee.
-
-The Girard-Hutchinson estimator distinguishes itself for its simplicity.
-However, its bound does not take into account the structure of $\mtx{B}(t)$.
-The $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ requirement implies that if we aim to increase the accuracy of the estimate by one digit, we would need to increase the number of queries by a factor of $100$.
-
-\subsection{Nyström approximation for parameter-dependent matrices}
-\label{subsec:nystrom}
-
-For symmetric positive semi-definite matrices whose singular values decay rapidly, the Nyström estimate \refequ{equ:nystrom-trace-estimator} can be significantly better than the Girard-Hutchinson estimate.
-This is reflected in the following theorem.
-
-\begin{theorem}{Nyström estimator for parameter-dependent matrices}{nystrom}
-    Suppose $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ is continuous and positive semi-definite in $t \in [a, b]$.
-Let $r \geq 2$ and $n_{\mtx{\Omega}} \geq r + 4$ be two integers.
-Then for all $\gamma \geq 1$ with probability at least $1 - \gamma^{-(n_{\mtx{\Omega}} - r)}$
-    \begin{equation}
-        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right| ~\mathrm{d}t
-        \leq \gamma^2 (1 + r) \int_{a}^{b} \sum_{i = r+1}^{n} \sigma_i(\mtx{B}(t)) ~\mathrm{d}t.
-    \end{equation}
-\end{theorem}
-
-\begin{proof}
-    Since $\mtx{B}(t)$ is positive semi-definite for all $t$, then by \cite[lemma 2.1]{frangella-2023-randomized-nystrom}, $\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ is also positive semi-definite for all $t$.
-Therefore, for fixed $t \in [a, b]$
-    \begin{equation}
-        \left| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right|
-        = \left| \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right|
-        = \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _{\ast},
-    \end{equation}
-    where $\lVert \cdot \rVert _{\ast}$ represents the nuclear norm.  From the proof of \cite[corollary 8.2]{tropp-2023-randomized-algorithms} it follows that
-    \begin{equation}
-        %\lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _2^2 \leq \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _F^2
-        \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{\mtx{B}(t)^{\sfrac{1}{2}} \mtx{\Omega}}) \mtx{B}(t)^{\sfrac{1}{2}} \rVert _F^2.
-    \end{equation}
-    Hence, by \cite[theorem 9]{kressner-2023-randomized-lowrank}, with $n_{\mtx{\Omega}} = p + r$, the result follows by noting $\sigma_i(\mtx{B}(t)^{\sfrac{1}{2}})^{2} = \sigma_i(\mtx{B}(t))$.
-\end{proof}
-
-Compared to the constant matrix case \cite[theorem 8.1]{tropp-2023-randomized-algorithms}, our bound also scales with the best rank-$r$ approximation error in the nuclear norm, albeit with a factor approximately proportional to $r$, which cannot be compensated for with large enough oversampling.
-
-The downside of the Nyström estimator is that it will -- in general -- not work well enough on matrices whose singular values do not decay quickly, for example $\mtx{B}(t) \equiv \mtx{I}_n$.
-
-\subsection{Nyström++ estimator for parameter-dependent matrices}
-\label{subsec:nystrom-pp}
-
-Just as in the constant matrix case, our goal is now to prove that the Nyström++ estimator \refequ{equ:nystrompp-trace-estimator} achieves an $\varepsilon$-relative error with just $\mathcal{O}(\varepsilon^{-1})$ random vectors used to construct the estimate.
-To do so, we mimic the proof of a similar bound for the Hutch++ estimator \cite[theorem 3.1]{meyer-2021-hutch-optimal}.
-Correspondingly, we will need to show that with $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2})$ random vectors, the Nyström approximation verifies $\int \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)\rVert _F~\mathrm{d}t \leq \varepsilon \int \Trace(\mtx{B}(t))~\mathrm{d}t$, regardless of the structure of the matrix $\mtx{B}(t)$.
-Just as in the corresponding proof for the Girard-Hutchinson estimator \refthm{thm:hutchinson}, we again bound the higher order moments of the approximation error for fixed $t$ and then use Minkowsi's integral inequality to carry over the results to the continuous case.
-We will do this in a sequence of three lemmas.
-
-For reasons which will become clear in the proof of \reflem{lem:nystrom}, we will need to bound $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]$ for a constant matrix $\mtx{A}$ and a standard Gaussian random matrix $\mtx{\Omega}$.
-The corresponding result for the Frobenius norm $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _F^2 \right]$ has already been shown in \cite{kressner-2023-randomized-lowrank}.
-Several ways of separately bounding the spectral norm moments of the Wishart matrix $\mtx{\Omega} \mtx{\Omega}^{\top}$ already exist, however, most of them are not sufficient for our purposes \cite{chen-2005-condition-numbers, edelman-1988-eigenvalues-condition, james-1964-distributions-matrix}.
-In \cite[lemma B.1]{tropp-2023-randomized-algorithms} a bound is achieved for $p = 2$ and $p = 4$.
-We will now generalize this result for arbitrary $p \in \mathbb{N}$.
-To do so, we first show the bound for Gaussian random vectors and then extend it to Gaussian random matrices in \reflem{lem:spectral-norm-moment}.
-
-\begin{lemma}{Spectral norm moments of non-standard Gaussian random vector}{spectral-norm-moment-vector}
-    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random vector $\vct{\omega} \in \mathbb{R}^{m}$ satisfy for all $k,p \in \mathbb{N}$ 
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2^2 \right]
-        \leq  (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
-    \end{equation}
-    %\begin{equation}
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
-    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
-    %\end{equation}
-    %If we choose 
-\end{lemma}
-
-\begin{remark}
-    In particular, this result can also be used to bound
-    \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2 \right] = \sqrt{\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2^2 \right]} \leq \sqrt{k + p} \cdot \left(\lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F\right)
-    \end{equation}
-    for any $k,p \in \mathbb{N}$.
-\end{remark}
-
-\begin{proof}
-    With the singular value decomposition $\mtx{A} = \mtx{U} \mtx{\Sigma} \mtx{V}^{\top}$ and the unitary invariance of the spectral norm, it can be seen that $\lVert \mtx{A} \mtx{\omega} \rVert _2^2 = \lVert \mtx{\Sigma} \widetilde{\mtx{\omega}} \rVert _2^2$ where $\widetilde{\mtx{\omega}}$ is again a standard Gaussian random vector.
-Hence, it is enough to bound $\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{\Sigma} \mtx{\omega} \rVert _2^2 \right]$ for a diagonal matrix $\mtx{\Sigma} = \operatorname{diag}(\sigma_1, \dots, \sigma_m)$ with $\sigma_1 \geq \dots \geq \sigma_m \geq 0$.
-
-    First, we can rewrite
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2^2 \right]
-        = \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right].
-        \label{equ:matvec-spectral-norm-simplification}
-    \end{equation}
-    Inspired by the proof of \cite[theorem 1]{cohen-2016-optimal-approximate}, we split the diagonal entries of $\mtx{\Sigma}$ into $\ell = \lceil m/k \rceil$ groups of size $k \geq 1$
-    \begin{equation}
-        \overbrace{\underbrace{\sigma_1, \dots, \sigma_k}_{\leq \sigma_1}}^{\geq \sigma_{k+1}}, \overbrace{\underbrace{\sigma_{k+1}, \dots, \sigma_{2k}}_{\leq \sigma_{k+1}}}^{\geq \sigma_{2k+1}}, \dots, \overbrace{\underbrace{\sigma_{1+(\ell - 1)k}, \dots, \sigma_{\ell k}}_{\leq \sigma_{1+(\ell - 1)k}}}^{\geq 0}.
-    \end{equation}
-    If $m$ is not a multiple of $k$, we let $\sigma_i = 0$ for $i > m$. Since $\sigma_1 \geq \dots \geq \sigma_{\ell k} \geq 0$,
-    \begin{equation}
-        \sigma_1^2 = \lVert \mtx{\Sigma} \rVert _2^2
-        \quad \text{and} \quad
-        \sigma_{1 + (i-1)k}^2 \leq \frac{1}{k} \sum_{j=1}^{k} \sigma_{j + (i-2)k}^2, i = 2, \dots, \ell,
-    \end{equation}
-    and therefore
-    \begin{equation}
-        \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \sum_{j=1}^{(\ell - 1)k} \sigma_j^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2.
-        \label{equ:singular-value-group-bound}
-    \end{equation}
-    This allows us to bound
-    \begin{align}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right]
-        &= \mathbb{E}^{\sfrac{p}{2}}\Bigg[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{j + (i-1)k}^2}_{\leq \sigma_{1 + (i-1)k}^2} \omega_{j + (i-1)k}^2 \Bigg] && \text{(separate sum into groups)} \notag \\
-        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{($\sigma_{j + (i-1)k}^2 \leq \sigma_{1 + (i-1)k}^2$, $\forall i, j$)} \notag \\
-        &\leq \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 ~ \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{(triangle inequality)} \notag \\
-        %&\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k}{2} + p)}{\Gamma(\frac{k}{2})} \right)^{\sfrac{1}{p}} && \text{(\refequ{equ:singular-value-group-bound} and \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
-        &\leq (k + p) \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 && \text{(\reflem{lem:gamma})} \notag \\
-        &\leq (k + p) \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right). && \text{(using \refequ{equ:singular-value-group-bound})}
-    \end{align}
-
-    %The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
-    %\begin{align}
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    &= \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}} \notag \\
-    %    &\leq \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{1}{\Gamma(k - r + 2)} \right)^{\frac{1}{k - r + 1}} \left( \frac{k + r}{2} \right) \notag \\
-    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right)
-    %\end{align}
-    %With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\sfrac{1}{n}} \leq \frac{e}{n}$ follow respectively.
-Hence,
-    %\begin{equation}
-    %    \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    \leq \frac{e^2}{k + 1}\sqrt{ \frac{3k}{2} }
-    %    \leq e^2 \sqrt{\frac{3}{2k}}.
-    %    \label{equ:OSE-moment-bound-second}
-    %\end{equation}
-
-\end{proof}
-
-\begin{lemma}{Spectral norm moments of non-standard Gaussian random matrix}{spectral-norm-moment}
-    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all $p \in \mathbb{N}$ 
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
-        \leq  (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
-    \end{equation}
-    %\begin{equation}
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
-    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
-    %\end{equation}
-    %If we choose 
-\end{lemma}
-
-\begin{remark}
-    In particular, this result can also be used to bound
-    \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2 \right] = \sqrt{\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]} \leq \sqrt{k + p} \cdot \left(\sqrt{3} \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F\right)
-    \end{equation}
-    for any $k,p \in \mathbb{N}$. Indeed, for $p=2$ this bound is slightly less tight than its equivalent in \cite[lemma B.1]{tropp-2023-randomized-algorithms}.
-\end{remark}
-
-\begin{proof}
-    In the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms}, it is shown that Slepian's inequality \cite[theorem 7.2.1]{vershynin-2018-random-processes} applied to two appropriately constructed Gaussian fields yields
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
-        \leq \lVert \mtx{A} \rVert _2^2 \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \vct{\omega}_1 \rVert _2^2 \right] + \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega}_2 \rVert _2^2 \right]
-    \end{equation}
-    for independent standard Gaussian random vectors $\vct{\omega}_1 \in \mathbb{R}^{k}$ and $\vct{\omega}_2 \in \mathbb{R}^{m}$.
-Both moments can be bound using \reflem{lem:spectral-norm-moment-vector}, one of which with $\mtx{A} = \mtx{I}_k$, to get
-    \begin{align}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
-        &\leq \lVert \mtx{A} \rVert _2^2 (k + p) \left( \lVert \mtx{I}_k \rVert _2^2 + \frac{1}{k} \lVert \mtx{I}_k \rVert _F^2 \right) + (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right) \notag \\
-        &\leq (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
-    \end{align}
-
-    %The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
-    %\begin{align}
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    &= \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}} \notag \\
-    %    &\leq \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{1}{\Gamma(k - r + 2)} \right)^{\frac{1}{k - r + 1}} \left( \frac{k + r}{2} \right) \notag \\
-    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right)
-    %\end{align}
-    %With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\sfrac{1}{n}} \leq \frac{e}{n}$ follow respectively.
-Hence,
-    %\begin{equation}
-    %    \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    \leq \frac{e^2}{k + 1}\sqrt{ \frac{3k}{2} }
-    %    \leq e^2 \sqrt{\frac{3}{2k}}.
-    %    \label{equ:OSE-moment-bound-second}
-    %\end{equation}
-
-\end{proof}
-
-\begin{lemma}{Nyström approximation for parameter-dependent matrices}{nystrom}
-    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-Then its Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ with Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ for \emph{even} $n_{\mtx{\Omega}} \geq 4$ verifies with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
-    \begin{equation}
-        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t
-        \label{qu:nystrompp-theorem-bound}
-    \end{equation}
-    for any $\gamma \geq 1$ and a universal constant $c > 0$.
-In particular, if we choose $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1}))$, then with probability at least $1-\delta$ for $\delta \in (0, 1)$ and $\varepsilon \in (0, 1)$ we have $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$.
-\end{lemma}
-
-%\todo{Proof idea: structural bound, then higher order moment bound to apply Markov's inequality.}
-
-\begin{proof}
-    First, we temporarily omit the parameter dependence.
-Let $\mtx{B} = \mtx{U} \mtx{\Lambda} \mtx{U}^{\top}$ be the eigenvalue decomposition of a symmetric positive semi-definite matrix $\mtx{B}$, i.e.
-$\mtx{\Lambda} = \operatorname{diag}(\lambda_1, \dots, \lambda_n)$ and $\mtx{U}^{\top} \mtx{U}= \mtx{I}_n$.
-We define the matrix partitions
-    \begin{equation}
-        \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+4ex+\baselineskip}
-        \mtx{U} = \begin{bmatrix}
-            \smash{\underbrace{\mtx{U}_1}_{n \times k}} & \smash{\underbrace{\mtx{U}_2}_{n \times (n-k)}}
-        \end{bmatrix}
-        \quad \text{and} \quad
-        \mtx{\Lambda} =
-        \begin{bmatrix}
-            \smash{\overbrace{\mtx{\Lambda}_1}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2}_{(n-k) \times (n-k)}}
-        \end{bmatrix}.
-    \end{equation}
-    Further, for a standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ we define $\mtx{\Omega}_1 = \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times n_{\mtx{\Omega}}}$ and $\mtx{\Omega}_2 = \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times n_{\mtx{\Omega}}}$, which are likewise standard Gaussian.
-
-    We can apply \cite[theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound
-    \begin{equation}
-        \lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F 
-        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2,
-        \label{equ:nystrom-proof-persson-bonud}
-    \end{equation}
-    where $\lVert \cdot \rVert _{(4)}$ denotes the Schatten-4 norm.
-
-    As in the proof of \cite[lemma 3]{meyer-2021-hutch-optimal}, the first term -- and its moments -- are limited by
-    \begin{equation}
-        \lVert \mtx{\Lambda}_2 \rVert _F
-        = \sqrt{\sum_{i=k+1}^{n} \lambda_i^2}
-        \leq \sqrt{ \lambda_{k+1} \sum_{i=k+1}^{n} \lambda_i}
-        \leq \sqrt{ \frac{\Trace(\mtx{B})}{k} \Trace(\mtx{B})}
-        \leq \frac{1}{\sqrt{k}} \Trace(\mtx{B}).
-        \label{equ:nystrom-proof-frobenius-trace}
-    \end{equation}
-    The $p$-th moment of the second term can be processed with standard matrix norm manipulations to
-    \begin{align}
-        \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        %&= \mathbb{E}^{p}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )( \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger})^{\top} \rVert _F \right] && \text{($\lVert \mtx{A} \rVert _{(4)}^2 = \lVert \mtx{A} \mtx{A}^{\top} \rVert _{F}$)} \notag \\
-        %&\leq \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] && \text{($\lVert \mtx{A} \mtx{B} \rVert _F \leq \lVert \mtx{A} \rVert _F \lVert \mtx{B} \rVert _2$)} \notag \\
-        &\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{($\lVert \mtx{A} \rVert _{(4)}^2 \leq \sqrt{k} \lVert \mtx{A} \rVert _2^2$)} \notag \\
-        &\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{(submultiplicativity)} \notag \\
-        &\leq \sqrt{k} \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2\right] \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2  \right].
-&& \text{(independence)}
-        \label{equ:nystrom-proof-processed-tail}
-    \end{align}
-    To match the decay rate of the moments of the first term in \refequ{equ:nystrom-proof-persson-bonud}, we will need to ensure that this term also is of order $\mathcal{O}(\Trace{\mtx{B}} / \sqrt{k})$.
-To this purpose, and to simplify the notation, we choose $k = n_{\mtx{\Omega}}/2$ and $p = n_{\mtx{\Omega}}/4$ for even $n_{\mtx{\Omega}}$ to ensure $k \in \mathbb{N}$.
-We can apply \reflem{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$ to bound
-    \begin{equation}
-        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \right]
-        \leq \frac{3 n_{\mtx{\Omega}}}{2} \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
-        \label{equ:spectral-norm-bound-applied}
-    \end{equation}
-    The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
-    \begin{equation}
-        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
-        = \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{n_{\mtx{\Omega}}}{4}} \right]^{\sfrac{4}{n_{\mtx{\Omega}}}}%  \notag \\
-        \leq \left( 1 + \frac{n_{\mtx{\Omega}}}{2} \right)^{\sfrac{4}{n_{\mtx{\Omega}}}} \left( \frac{1}{(\frac{n_{\mtx{\Omega}}}{2} + 1)!}\right)^{\frac{2}{\frac{n_{\mtx{\Omega}}}{2} + 1}} \left( \frac{3 n_{\mtx{\Omega}}}{4}\right).
-    \end{equation}
-    From the expansion of the exponential function as a power series $e^n = 1 + n + \dots + n^n/n!
-+ \dots$ for $n \in \mathbb{N}$ we see $e^n \geq 1 + n$ and $e^n \geq n^n / n!$, from which $(1 + n)^{\sfrac{1}{n}} \leq e$ and $\left( 1/n!
-\right)^{\sfrac{1}{n}} \leq e/n$ follow respectively.
-Hence,
-    \begin{equation}
-        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
-        \leq e^2 \frac{e^2}{(\frac{n_{\mtx{\Omega}}}{2}+ 1)^2} \left( \frac{3 n_{\mtx{\Omega}}}{4}\right)
-        \leq \frac{3 e^4}{n_{\mtx{\Omega}}}.
-        \label{equ:pinv-spectral-norm-bound}
-    \end{equation}
-    As opposed to \cite[lemma B.3]{tropp-2023-randomized-algorithms}, our bound holds for all even $n_{\mtx{\Omega}} \in \mathbb{N}$, at the cost of a factor of $e^2$.
-Inserting \refequ{equ:spectral-norm-bound-applied} and  \refequ{equ:pinv-spectral-norm-bound} in \refequ{equ:nystrom-proof-processed-tail} gives
-    \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \frac{9 e^4}{2 \sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
-        %\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
-    \end{equation}
-    We can identify $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 = \lambda_{k+1}$ and $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 = \Trace(\mtx{\Lambda}_2)$, and just as in the proof of \cite[lemma 3.1]{meyer-2021-hutch-optimal} we bound $\lambda_{k+1} \leq \Trace(\mtx{B})/k$ and $\Trace(\mtx{\Lambda}_2) \leq \Trace(\mtx{B})$ -- remembering $k = n_{\mtx{\Omega}}/2$ -- to get
-    \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \frac{9 e^4}{2 \sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \left( \frac{6}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) + \frac{1}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) \right)
-        \leq \frac{9e^4}{2 \sqrt{2}} \frac{7}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
-        \label{equ:nystrom-proof-tail-bound}
-        %\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
-    \end{equation}
-    Inserting \refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud} and using the triangle inequality for $\mathbb{E}^{p}[\cdot]$ we obtain
-    \begin{equation}
-        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right]
-        \leq \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[ \lVert \mtx{\Lambda}_2 \rVert _F \right] + \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
-    \end{equation}
-    where $c = 1 + 63 e^4 / 2 \sqrt{2}$.
-
-    As in \cite{kressner-2023-randomized-lowrank}, we can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable and by the Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities}, we get for $n_{\mtx{\Omega}} \geq 4$
-    \begin{align}
-        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \right] 
-        &\leq \int_{a}^{b} \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F \right] ~\mathrm{d}t \notag \\
-        %&\leq \mathbb{E}^{\sfrac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
-        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \mathbb{E}^{\sfrac{k}{2}} \left[ \int_{a}^{b}  \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
-        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
-        &\leq \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
-    \end{align}
-    From Markov's inequality follows with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
-    \begin{equation}
-        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t,
-    \end{equation}
-    from which follows the first statement.
-
-    Fixing $\gamma = e$ and letting $n_{\mtx{\Omega}} = (c e \varepsilon)^{-2}  + 4 \log(\delta^{-1})$ gives us the second part of the theorem.
-
-    %Setting $\delta = \gamma^{-\sfrac{k}{4}}$ and choosing $k = \lceil \varepsilon^{-2}\log(\delta^{-1}) \rceil$ we get
-    %\begin{align}
-    %    \gamma \frac{1}{\sqrt{k}}
-    %    &= \delta^{-\sfrac{2}{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} \frac{1}{\sqrt{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} && \text{(definition of $\delta$ and choice of $k$)} \notag \\
-    %    &= e^{2\varepsilon^{2} \frac{\varepsilon^{-2}\log(\delta^{-1})}{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} \frac{1}{\sqrt{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} && \text{($\delta = e^{-\log(\delta^{-1})}$)} \notag \\
-    %    &\leq e^{2\varepsilon^{2}} \frac{\varepsilon}{\sqrt{ \log(\delta^{-1})}} && \text{($\lceil x \rceil \geq x$ if $x \geq 0$)}
-    %\end{align}
-    %which is smaller than $\varepsilon$ if $\delta \leq e^{-e^{4 \varepsilon^2}}$from which follows the second part of the theorem when using $n_{\mtx{\Omega}} = 2k$.
-\end{proof}
-
-When compared to the equivalent result for constant matrices \cite[Lemma 3.2]{persson-2022-improved-variants}, our constant factor up front is around four times larger, and our failure probability decays half a power slower.
-
-Finally, we can combine the bound on the Girard-Hutchinson estimator \refthm{thm:hutchinson} and the one on the Nyström approximation \reflem{lem:nystrom} to obtain our main result.
-
-\begin{theorem}{Nyström++ trace estimator for parameter-dependent matrices}{nystrom-pp}
-    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(\delta^{-1})^2)$ with \emph{even} $n_{\mtx{\Omega}}$, then for $\delta \in (0, e^{-\sfrac{1}{2}}/2)$ and $\varepsilon \in (0, 1)$ with probability at least $1 - \delta$
-    \begin{equation}
-        \int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t
-        \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
-    \end{equation}
-\end{theorem}
-
-\begin{proof}
-    By choosing $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\tilde{\varepsilon}^{-2} \log(\tilde{\delta}^{-1})^2)$ we get
-    \begin{align}
-        &\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t \notag \\
-        &= \int_{a}^{b} | \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) - \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) |~\mathrm{d}t && \text{(by definition of estimators)} \notag \\
-        &\leq \tilde{\varepsilon} \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F ~\mathrm{d}t && \text{(\refthm{thm:hutchinson} w.p. $\geq 1 - \tilde{\delta}$)} \notag \\
-        &\leq \tilde{\varepsilon}^2 \int_{a}^{b} \Trace(\mtx{B}(t)) ~\mathrm{d}t && \text{(\reflem{lem:nystrom} w.p. $\geq 1 - \tilde{\delta}$)} 
-    \end{align}
-    with probability at least $1 - 2\tilde{\delta}$. Taking $\varepsilon = \tilde{\varepsilon}^2$ and $\delta = 2 \tilde{\delta}$
-\end{proof}
-
-Comparing \refthm{thm:nystrom-pp} with its equivalent for constant matrices \cite[theorem 3.4]{persson-2022-improved-variants}, we notice a less favorable scaling of the required number of random vectors $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$ in terms of $\log(\delta^{-1})$.
-
-% \begin{proof}
-%     Combining with union bound.
-%     \begin{align}
-%         &\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t \notag \\
-%         &= \int_{a}^{b} | \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)) - \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)) |~\mathrm{d}t && \text{(definition of estimators)} \notag \\
-%         &= \todo{c} \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F ~\mathrm{d}t && \text{(with probability $1 - \delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1})$)} \notag \\
-%         &= \todo{c} \frac{1}{\sqrt{n_{\mtx{\Psi}} n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t)) ~\mathrm{d}t
-%     \end{align}
-% \end{proof}
-
 \section{Application to spectral density approximation}
@@
 
-It is well known that the coefficients of a Chebyshev expansion of a function can -- in exact arithmetic -- be computed with a discrete cosine transform (DCT) of the function evaluations at the Chebyshev nodes $s_i = \cos(\pi i / m)$ for $i=0,\dots,m$
+It is well known that the coefficients of a Chebyshev expansion of a function can -- in exact arithmetic -- be computed by inverting the discrete cosine transform (DCT)
 \begin{equation}
-    g_{\sigma}(t - \cos(s_i)) = \sum_{l=0}^{m} \mu_l(t) \cos\left(\frac{\pi i l}{m} \right).
+    g_{\sigma}(t - \cos(s_i)) = \sum_{l=0}^{m} \mu_l(t) \cos\left(\frac{\pi i l}{m} \right)
     \label{equ:discrete-cosine-transform}
 \end{equation}
+between the expansion coefficients $\mu_l(t),~i=0,\dots,m$ and the function evaluations at the Chebyshev nodes $s_i = \cos(\pi i / m),~i=0,\dots,m$.
 This can be accomplished in $\mathcal{O}(m \log(m))$ operations.
@@
 \begin{equation}
-    \mtx{\Omega}^{\top} \underline{g}^{(m)}_{\sigma}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega} = \sum_{l=0}^{m} \mu_l(t) (\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega})
+    \mtx{\Omega}^{\top} \underline{g}^{(m)}_{\sigma}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega} = \sum_{l=0}^{m} \mu_l(t) (\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}),
 \end{equation}
-for which we can first precompute the small $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ for $l=0, \dots, m$ using the recurrence \refequ{equ:chebyshev-recurrence} and then inexpensively sum them up with the corresponding coefficients $\mu_l(t)$ for all $t$ in which we want to evaluate the approximation.
+meaning we can first precompute the small $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ for $l=0, \dots, m$ using the Chebyshev recurrence \refequ{equ:chebyshev-recurrence} and then inexpensively sum them up with the corresponding coefficients $\mu_l(t)$ for all $t$ in which we want to evaluate the approximation.
 Thanks to the fast and exact squaring of Chebyshev expansions (cf.
@@
 Hence, choosing $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma)$ will approximations which are significantly better than \refthm{thm:chebyshev-nystrom}.
-
-\section{Numerical results}
-\label{sec:results}
-
-Our first numerical example comes from electronic structure interaction \cite{lin-2017-randomized-estimation}, more precisely from the second order finite difference discretization of the Hamiltonian
-\begin{equation}
-    \mathcal{H} = - \Delta + V
-    \label{equ:5-experiments-electronic-hamiltonian}
-\end{equation}
-in three dimensions. The potential $V$ interacting with the electrons is generated by Gaussian wells
-\begin{equation}
-    v(r) = v_0 e^{-\lambda r^2}
-    \label{equ:5-experiments-gaussian-cell}
-\end{equation}
-with $v_0 = -4$ and $\lambda = 8$, centered in cells of side length $L=6$ which are stacked $n_c \in \mathbb{N}$ times in each spatial dimension (cf.
-\reffig{fig:gaussian-well}).
-The discretization step is fixed to be $h=0.6$, such that the size of the matrix grows cubically with $n_c$.
-This is an idealized model for the interaction of nuclei on a regular grid with electrons for a $k$-vector in the center of the first Brillouin zone.
-The distribution of the eigenvalues of the Hamiltonian -- its spectral density -- allows us to interpret the system's energy levels.
-
-\begin{figure}[ht]
-    \begin{subfigure}[b]{0.32\columnwidth}
-        \input{plots/gaussian-well-1.pgf}
-        \caption{$n_c=1$}
-        \label{fig:gaussian-well-1}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.32\columnwidth}
-        \input{plots/gaussian-well-2.pgf}
-        \caption{$n_c=2$}
-        \label{fig:gaussian-well-2}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.32\columnwidth}
-        \input{plots/gaussian-well-5.pgf}
-        \caption{$n_c=5$}
-        \label{fig:gaussian-well-5}
-    \end{subfigure}
-    \caption{Cross-sections of the periodic Gaussian well potential $V$ for different sizes of the supercell.}
-    \label{fig:gaussian-well}
-\end{figure}
-
-\todo{Explain experimental setup.}
-
-\begin{figure}[ht]
-    \centering
-    \input{plots/convergence.pgf}
-    \caption{For increasing values of $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}$ but fixed $m$ we plot the $L^1$-approximation error for $\sigma=0.005$ on the model problem with $n_c = 1$.}
-    \label{fig:convergence}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \input{plots/distribution.pgf}
-    \caption{The Chebyshev-Nyström++ method for different ways of allocations a total of $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation for the Gaussian smoothing kernel with multiple different values of the smoothing parameter.
-We make the approximation error made in the Chebyshev expansion negligible by rescaling $m=16 / \sigma$ (based on \refthm{thm:chebyshev-error}).}
-    \label{fig:distribution}
-\end{figure}
-
-\todo{Interpret results.}
-
-\todo{Comparison plots with other estimators (Krylov-Aware)?}
-
-\clearpage
-\section{Conclusion}
-\label{sec:conclusion}
-
-\todo{
-\begin{itemize}
-    \item Robustness (can choose $m$ and $n_{\mtx{\Omega}}$, $n_{\mtx{\Psi}}$ as large as one wants, and not break algorithm, unlike Krylov-aware Lanczos)
-\end{itemize}
-}
-
-% Acknowledge advice from David and Data/Parameters from Lin Lin, Michael Herbst?
-
-\clearpage
-%\bibliography{bibliography.bib}
-\printbibliography
-
-\appendix
-
-\clearpage
-\section{Proofs}
-
-\begin{lemma}{Moment bound of $\chi^2$-random variable}{gamma}
-    Let $X$ follow a $\chi^2$ distribution with $k \in \mathbb{N}$ degrees of freedom. Then for all $p \in \mathbb{N}$ it holds
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}[X] \leq k + p
-        %\frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \leq \left( \frac{k + p}{2} \right)^{\sfrac{p}{2}}.
-    \end{equation}
-\end{lemma}
-
-\begin{proof}
-    From \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical} we know
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}[X] = 2 \left( \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \right)^{\sfrac{2}{p}}.
-    \end{equation}
-    We first treat the case $k=p=1$, for which indeed
-    \begin{equation}
-        2 \left( \frac{\Gamma(\frac{1+1}{2})}{\Gamma(\frac{1}{2})} \right)^2 = \frac{2}{\pi} \leq 2 = 1 + 1
-    \end{equation}
-    So we now assume $k + p \geq 3$. We can use a telescoping product and the property $\Gamma(z+1)/\Gamma(z) = z$ to bound
-    \begin{align}
-        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} 
-        &= \frac{\Gamma(\frac{k}{2} + 1)}{\Gamma(\frac{k}{2})} \frac{\Gamma(\frac{k}{2} + 2)}{\Gamma(\frac{k}{2} + 1)} \cdots \frac{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor - 1)} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)} \notag \\
-        &= \left(\frac{k}{2}\right) \left(\frac{k}{2} + 1\right) \cdots \left( \frac{k}{2} + \bigg\lfloor \frac{p}{2} \bigg\rfloor - 1 \right) \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)} \notag \\
-        &\leq \left(\frac{k}{2} + \bigg\lfloor \frac{p}{2} \bigg\rfloor - 1\right)^{\lfloor \sfrac{p}{2} \rfloor} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)}
-    \end{align}
-    If $p$ is even, then $\lfloor p/2 \rfloor = p/2$ and hence
-    \begin{equation}
-        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} 
-        \leq \left( \frac{k + p}{2} - 1 \right)^{\sfrac{p}{2}}
-        \leq \left( \frac{k + p}{2}\right)^{\sfrac{p}{2}}.
-    \end{equation}
-    If $p$ is odd, then $\lfloor p/2 \rfloor = (p - 1)/2$, from which follows with Gautschi's inequality \cite{kershaw-1983-extensions-gautschi}
-    \begin{equation}
-        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})}
-        \leq \left(\frac{k + p}{2} - \frac{3}{2} \right)^{ \frac{p - 1}{2}} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k + p}{2} - \frac{1}{2})}
-        \leq \left(\frac{k + p}{2}\right)^{ \frac{p - 1}{2}} \left(\frac{k + p}{2}\right)^{\sfrac{1}{2}}
-        = \left(\frac{k + p}{2}\right)^{ \sfrac{p}{2}}
-    \end{equation}
-    Hence,
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}[X] \leq 2 \left( \frac{k + p}{2} \right) \leq k + p.
-    \end{equation}
-\end{proof}
-
-\end{document}

### 2024-07-07 paper/analysis.tex

--- 2024-06-27 paper/main.tex
@@
-\documentclass[12pt]{article}
-
-\usepackage{stylesheet}
-
-\title{Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation}
-
-\begin{document}
-
-\maketitle
-
-%\todo{
-%\begin{itemize}
-%    \item Unify index notation
-%    \item Double-check proofs
-%    \item Add all references
-%    \item Run numerical experiments 
-%    \item Write introduction, abstract, and discussion
-%    \item ArXiv HTML support (check if packages)
-%\end{itemize}
-%}
-
-%\todo{
-%    \tableofcontents
-%}
-
-\todo{
-
-\paragraph{TODO}
-\begin{itemize}
-    \item Check that everywhere correct/consistent orders $\mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1})$)
-    \item 
-\end{itemize}
-}
-
-\todo{
-\begin{abstract}
-    Write abstract.
-\end{abstract}
-}
-
-\section{Introduction}
-\label{sec:introduction}
-
-In numerous problems across physics, chemistry, engineering, and data science, how the eigenvalues of certain matrices are distributed can be indicative of what properties a system has: In electronic structure calculations eigenvalues represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}.
-The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$.
-Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation which is often prohibitively expensive.
-Therefore, many techniques have been proposed for approximating $\phi$.
-For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature rule with the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}.
-Further, because $\delta$ is hard to approximate and in many applications, a rough estimate of the spectral density is already acceptable, a standard approach is to instead approximate the smoothed spectral density
-\begin{equation}
-    \phi_{\sigma}(t) = \frac{1}{n} \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i) = \frac{1}{n} \Trace(g_{\sigma}(t \mtx{I}_n - \mtx{A}))
-    \label{equ:smoothed-spectral-density}
-\end{equation}
-for a smoothing kernel $g_{\sigma}$, typically a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}.
-This translates the problem to computing the trace of a continuous parameter-dependent matrix $\mtx{B}(t) \equiv g_{\sigma}(t \mtx{I}_n - \mtx{A}) \in \mathbb{R}^{n \times n}$.
-Directly assembling the matrix function requires diagonalizing the matrix $\mtx{A}$, which again is prohibitively expensive.
-Therefore, we estimate its trace implicitly by approximating products of the matrix function with vectors.
-For constant matrices, the most widespread methods for implicit trace estimation are the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} and the variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making, chen-2023-krylovaware-stochastic}.
-Some of these estimators have already been successfully applied to parameter-dependent matrices \cite{lin-2017-randomized-estimation,chen-2023-krylovaware-stochastic}, but they have not yet been analyzed in this setting.
-A related technique is dynamic trace estimation for sequences $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ of matrices which satisfy $\lVert \mtx{B}(t_i) \rVert _F \leq 1$ and which change dynamically, i.e.
-$\lVert \mtx{B}(t_{i+1}) - \mtx{B}(t_i) \rVert _F \leq \alpha < 1$ for the Frobenius norm $\lVert \cdot \rVert _F$ and a small $\alpha$ \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace}.
-However, in our setting we often cannot guarantee either of the two conditions.
-Instead of discretizing $[a, b]$ into $t_1, \ldots, t_m$, approximating $\Trace(\mtx{B}(t_i))$ with $b(t_i)$, and analyzing the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way is to quantify the error with $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate in each $t \in [a, b]$.
-This approach has already been taken for analyzing low-rank approximations of parameter-dependent matrices \cite{kressner-2023-randomized-lowrank}.
-
-\paragraph{New contributions} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}.
-Interestingly, we can reuse the same randomization for computing the estimate at each value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations.
-We propose certain modifications to the algorithms from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}.
-In particular, use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials, and realize that an efficient implementation of all of their methods can be summarized in a single algorithm.
-Based on the analysis of the randomized trace estimators for parameter-dependent matrices we provide bounds on the approximation error for our method.
-
-\paragraph{Structure} In \refsec{sec:analysis} we will analyze three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter dependent matrices.
-We propose and analyze a method for using the aforementioned estimators to estimate the spectral density of real symmetric matrices in \refsec{sec:application}.
-Finally, we will test our developments on an example from electronic structure calculation.
-
-\paragraph{Reproducibility} This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-SD/tree/ca391a8}{ca391a8} at \href{https://github.com/FMatti/Rand-TRACE}{github.com/FMatti/Rand-TRACE} on 2024-02-05 at 18:44:29 UTC.
-
-\section{Analysis of randomized trace estimators applied to parameter-dependent matrices}
+\section{Randomized trace estimators for parameter-dependent matrices}
 \label{sec:analysis}
 
-We consider parameter-dependent matrices of the form
-\begin{equation}
+\color{blue}
+In this section, we consider a general parameter-dependent matrix
+\begin{equation*}
     \mtx{B}(t) = \begin{bmatrix}
@@
         b_{n1}(t) & b_{n2}(t) & \dots & b_{nn}(t) \\
-    \end{bmatrix} \in \mathbb{R}^{n \times n}
-\end{equation}
-where $b_{ij}(t)$ are functions depending continuously on the parameter $t$ which takes values in the interval $[a,b]$. The trace of such a matrix is defined as
-\begin{equation}
-    \Trace(\mtx{B}(t)) = \sum_{i=1}^{n} b_{ii}(t).
-    \label{equ:definition-trace}
-\end{equation}
-However, we assume that we only have access to products of this matrix with vectors for each $t \in [a, b]$, so this definition will not be directly useful for computing the trace.
+    \end{bmatrix} \in \mathbb{R}^{n \times n},
+\end{equation*}
+where each entry $b_{ij}(t)$ is a continuous function on a bounded interval $[a,b]$. \emph{If} these entries are given explicitly, we can simply compute the trace by its definition 
+$\Trace(\mtx{B}(t)) = b_{11}(t)+ \cdots + b_{nn}(t)$.
+This becomes more difficult when $\mtx{B}(t)$ is given implicitly (e.g., as a matrix function) and can only be accessed through matrix-vector products.
+The common theme of all methods analyzed in this section is to extract trace estimates from multiplying $\mtx{B}(t)$ with fixed random vectors
+
+as we will see in Section ???.
+\color{black}
 
@@
 \end{equation}
+??? Gaussian random matrix define! ???
 where $\mtx{\Psi} = [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\mtx{\Psi}}}] \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$.
@@
 However, our theoretical developments only hold in the Gaussian case.
+
+\color{black}
 
@@
 We first analyze the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}.
-We bound its $L^1$ norm in terms of the integral over the Frobenius norm $\lVert \cdot \rVert _F$ of the matrix whose trace we aim to approximate.
+We bound its $L^1$-error in terms of the $L^1$-norm of the Frobenius norm $\lVert \cdot \rVert _F$ of the matrix whose trace we approximate.
 
+??? START HERE ???`
 \begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
-    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric and continuously depend on $t \in [a, b]$.
-Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ it holds with probability at least $1 - \gamma^{-2k}$
+    For symmetric $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ depending continuously on $t \in [a, b]$, consider the .
+Girard-Hutchinson estimator~\eqref{equ:hutchinson-trace-estimator} with an $n\times n_{\mtx{\Psi}}$ Gaussian random matrix $\mtx{\Psi}$.
+Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ the error bound
     \begin{equation}
@@
     \end{equation}
-    In particular, if we choose $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{- \sfrac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$.
+    holds with probability at least $1 - \gamma^{-2k}$
+
+    In particular, for any $\varepsilon > 0$ we have that $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$ holds
+    with probability at least $1-\delta$ for $\delta \in (0, 1)$ when choosing $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then  for  and  we have .
 \end{theorem}
 
-The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Minkowski's integral inequality to extend the results to the continuous parameter case.
+The idea behind the proof is to bound the higher order moments of the residual $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ for fixed values of $t$ and then use Minkowski's integral inequality to extend the results to the continuous parameter case.
 We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
 \begin{proof} 
-    We define the 1-query Girard-Hutchinson estimate 
+    We define the residual of the 1-query Girard-Hutchinson estimate 
     \begin{equation}
@@
 \begin{proof}
-    From the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms} we know that
+    In the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms}, it is shown that Slepian's inequality \cite[theorem 7.2.1]{vershynin-2018-random-processes} applied to two appropriately constructed Gaussian random fields yields
     \begin{equation}
@@
     As opposed to \cite[lemma B.3]{tropp-2023-randomized-algorithms}, our bound holds for all even $n_{\mtx{\Omega}} \in \mathbb{N}$, at the cost of a factor of $e^2$.
-Inserting \refequ{equ:spectral-norm-bound-applied} and  \refequ{equ:pinv-spectral-norm-bound} in \refequ{equ:nystrom-proof-processed-tail} gives
+Inserting \refequ{equ:spectral-norm-bound-applied} and  \refequ{equ:pinv-spectral-norm-bound} in \refequ{equ:nystrom-proof-processed-tail} with $k = n_{\mtx{\Omega}} / 2$ gives
     \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \frac{9 e^4}{2 \sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
+        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
+        \leq \frac{3 e^4}{\sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{2}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
         %\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
@@
 
-When compared to the equivalent result for constant matrices \cite[Lemma 3.2]{persson-2022-improved-variants}, our constant factor up front is around four times larger, and our failure probability decays half a power slower.
+When compared to the equivalent result for constant matrices \cite[Lemma 3.2]{persson-2022-improved-variants}, our constant factor up front is around three times larger and our failure probability decays half a power slower.
 
-Finally, we can combine the bound on the Girard-Hutchinson estimator \refthm{thm:hutchinson} and the one on the Nyström approximation \reflem{lem:nystrom} to obtain our main result.
+Finally, we can now combine the bound on the Girard-Hutchinson estimator (\refthm{thm:hutchinson}) and the one on the Nyström approximation (\reflem{lem:nystrom}) to obtain our main result.
 
@@
 % \end{proof}
-
-\section{Application to spectral density approximation}
-\label{sec:application}
-
-We consider the task of computing the spectral density of a real symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ with eigenvalues $\lambda_1, \dots, \lambda_n$, which is defined as
-\begin{equation}
-    \phi(t) = \frac{1}{n} \sum_{i=1}^{n} \delta(t - \lambda_i),
-    \label{equ:spectral-density}
-\end{equation}
-where $\delta$ is the Dirac delta distribution.
-Finding approximations to this distribution is challenging.
-Furthermore, in most applications the exact location of every eigenvalue is not important, but more so their approximate locations relative to each other, such as eigenvalue clusters, spectral gaps, or outliers.
-In the following we will introduce a smoothed version of the spectral density, which is significantly easier to approximate at the cost of losing some of the finer characteristics of the spectrum.
-
-\subsection{Smoothed spectral density}
-\label{subsec:spectral-density}
-
-The smoothed spectral density $\phi_{\sigma}$ is defined as
-\begin{equation}
-    \phi_{\sigma}(t) = \frac{1}{n} \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i) = \frac{1}{n} \Trace(g_{\sigma}(t\mtx{I}_n - \mtx{A}))
-    \label{equ:smooth-spectral-density}
-\end{equation}
-for a smoothing kernel $g_{\sigma}$ parametrized by a smoothing parameter $\sigma > 0$, which controls by how much $\phi$ is smoothed.
-We choose a Gaussian smoothing kernel of width $\sigma > 0$, given by
-\begin{equation}
-    g_{\sigma}(s) = \frac{1}{\sigma \sqrt{2\pi}} e^{-\sfrac{s^2}{2\sigma^2}}.
-    \label{equ:smoothing-kernel}
-\end{equation}
-The task in \refequ{equ:smooth-spectral-density} is to approximate the trace of the parameter-dependent matrix $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
-Hence, we can apply the results derived in \refsec{sec:analysis}.
-
-Choosing the smoothing parameter $\sigma$ results in the following trade off: Typically, larger $\sigma$ make it easier to approximate the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$, whereas smaller $\sigma$ allow us to stay closer to the original spectral density \refequ{equ:spectral-density}.
-To see the latter, we measure the error between the spectral density $\phi$ and its smoothened version $\phi_{\sigma}$ with
-\begin{equation}
-    \sup_{f \in \mathcal{S}} \int_{-1}^{1} f(t) (\phi(t) - \phi_{\sigma}(t))~\mathrm{d}t,
-    \label{equ:error-metric}
-\end{equation}
-where $\mathcal{S}$ represents an appropriately chosen space of test functions.
-Among others, \cite{lin-2016-approximating-spectral} proposes $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - \lambda), \lambda \in [\lambda_{\text{min}}, \lambda_{\text{max}}]\}$, \cite{chen-2021-analysis-stochastic} uses $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [-1, 1] \}$ where $\Theta$ is the Heaviside step function, and \cite{braverman-2022-sublinear-time} uses $\mathcal{S} = \{f : |f(x) - f(y)| \leq |x - y| \}$ for which the error \refequ{equ:error-metric} is equivalent to the Wasserstein-1 distance between $\phi$ and $\phi_{\sigma}$.
-In this last metric, provided we assume the spectrum to be somewhat uniform and $\sigma$ to be rather small, a simple calculation allows us to roughly estimate the smoothing error \refequ{equ:error-metric} to be $\sigma$.
-Consequently, if we can only allow an approximation to deviate by at most a factor of $\varepsilon > 0$ from the original spectral density, we should choose a smoothing parameter $\sigma \approx \varepsilon$.
-
-\subsection{Chebyshev expansion of smoothing kernel}
-\label{subsec:chebyshev-expansion}
-
-To estimate the trace in \refequ{equ:smooth-spectral-density}, we first need to evaluate the involved matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
-However, doing this exactly would require us to first diagonalize $\mtx{A}$, which is often a prohibitively expensive operation.
-For algorithmic reasons, we seek for approximations of this matrix function in terms of affine expansions, of which the Chebyshev expansion is most suitable due to its approximation error guarantees and ease of manipulation.
-In every $t \in \mathbb{R}$ we expand
-\begin{equation}
-    g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) = \sum_{l=0}^{m} \mu_l(t) T_l(\mtx{A})
-    \label{equ:matrix-expansion}
-\end{equation}
-in terms of a set of $m+1$ coefficients $\mu_0(t), \dots, \mu_m(t)$ which linearly combine the Chebyshev polynomials
-\begin{equation}
-    \begin{cases}
-        T_l : [-1, 1] \to \mathbb{R} \\
-        T_l(s) = \cos(l \cdot \arccos(s))
-    \end{cases}
-    \label{equ:chebyshev-polynomial}
-\end{equation}
-evaluated at the matrix $\mtx{A}$. It is easy to verify that these polynomials satisfy the three-term recurrence relation
-\begin{equation}
-    T_l(s) =
-    \begin{cases}
-        s^l, & \text{if $l \in \{0, 1\}$}; \\
-        T_l(s) = 2 s T_{l-1}(s) - T_{l-2}(s), & \text{if $l \geq 2$.}
-    \end{cases}
-    \label{equ:chebyshev-recurrence}
-\end{equation}
-With the help of the Chebyshev expansion \refequ{equ:matrix-expansion} we can now define the expanded spectral density
-\begin{equation}
-    \phi_{\sigma}^{(m)}(t) =  \frac{1}{n} \Trace(g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})).
-    \label{equ:expanded-spectral-density}
-\end{equation}
-
-However, the Chebyshev polynomials are only well-defined for matrices whose spectrum is contained in $[-1, 1]$.
-When working with general matrices of bounded spectrum, we will first have to estimate a lower bound $a$ and upper bound $b$ on the spectrum \cite{zhou-2011-bounding-spectrum}.
-A spectral transform
-\begin{equation}
-    \begin{cases}
-        \tau : [a, b] \to [-1, 1]\\
-        \tau(t) = \frac{2t - a - b}{b - a}
-    \end{cases}
-\end{equation}
-then lets us define the matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ whose spectrum is identical to the one of $\mtx{A}$ but shrunk to the interval $[-1, 1]$, for which the expansion \refequ{equ:matrix-expansion} is valid.
-Note that through this transformation we will also have to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when reverting the spectral transform $\tau$.
-
-\subsubsection{Fast and consistent squaring}
-\label{subsubsec:dct}
-
-It is well known that the coefficients of a Chebyshev expansion of a function can -- in exact arithmetic -- be computed with a discrete cosine transform (DCT) of the function evaluations at the Chebyshev nodes $s_i = \cos(\pi i / m)$ for $i=0,\dots,m$
-\begin{equation}
-    g_{\sigma}(t - \cos(s_i)) = \sum_{l=0}^{m} \mu_l(t) \cos\left(\frac{\pi i l}{m} \right).
-    \label{equ:discrete-cosine-transform}
-\end{equation}
-This can be accomplished in $\mathcal{O}(m \log(m))$ operations.
-\cite[algorithm 1]{lin-2017-randomized-estimation} uses a slightly different approach, which only approximates the coefficients and is noticeably slower (cf.
-\reftab{tab:chebyshev-timing-interpolation}).
-
-A highly useful property of the Chebyshev expansion -- of which we will make use twice in the following pages -- is the fact that we can square such an expansion in $\mathcal{O}(m \log(m))$ operations \cite{baszenski-1997-fast-polynomial}, which will be used in the below algorithm.
-
-\begin{algo}{Fast squaring of Chebyshev expansion}{chebyshev-squaring}
-    \begin{algorithmic}[1]
-        \Statex \textbf{Input:} Coefficients $\{ \mu_l \}_{l=0}^{m}$ of the expansion $\sum_{l=0}^{m} \mu_l T_l$
-        \Statex \textbf{Output:} Coefficients $\{ \nu_l \}_{l=0}^{2m}$ such that $\sum_{l=0}^{2m} \nu_l T_l = (\sum_{l=0}^{m} \mu_l T_l)^2$
-        \State Define $\vct{\mu} \in \mathbb{R}^{2m + 1}$ whose first $m + 1$ entries are the coefficients $\mu_l$ and the remaining $m$ are $0$
-    \State Compute $\vct{f} = \DCT(\vct{\mu})$
-    \State Compute $\vct{\nu} = \DCT^{-1}(\vct{f}^{2})$ whose entries are the coefficients $\nu_l$
-    \end{algorithmic}
-    \end{algo}
-
-\subsubsection{Expansion error}
-\label{subsubsec:expansion-error}
-
-We now analyze the error of the Chebyshev expansion of a Gaussian smoothing kernel $g_{\sigma}$.
-
-\begin{lemma}{Chebyshev expansion error}{chebyshev-error}
-    The expansion $g_{\sigma}^{(m)}$ of the Gaussian smoothing kernel $g_{{\sigma}}$ \refequ{equ:smoothing-kernel} satisfies
-    \begin{equation}
-        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \right| \leq \sqrt{\frac{2e}{\pi}} \frac{1}{\sigma^2} (1 + \sigma)^{-m} \equiv E_{\sigma, m}
-        \label{equ:chebyshev-interpolation-sup-error-kernel}
-        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
-        %\label{equ:2-chebyshev-interpolation-error}
-    \end{equation}
-    for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and $s \in [-1, 1]$.
-\end{lemma}
-
-The proof specializes Bernstein's theorem for Chebyshev interpolation to the case of a Gaussian smoothing kernel \refequ{equ:smoothing-kernel}.
-It loosely follows the proof give in the preprint of \cite[theorem 2]{lin-2017-randomized-estimation}, but attains a slightly different dependence on $\sigma$.
-
-\begin{proof}
-    From Bernstein's theorem \cite[theorem 4.3]{trefethen-2008-gauss-quadrature} it follows that if $f$ is analytic within the Bernstein ellipse $\mathcal{E}_{\chi} \subset \mathbb{C}$ with foci $\{-1, +1\}$ and sum of semi-axes $\chi > 1$, then for any $m \in \mathbb{N}$
-    \begin{equation}
-        \sup_{s \in [-1, 1]} \left| f(s) - f^{(m)}(s) \right| \leq \frac{2}{\chi^m (\chi - 1)} \max_{z \in \mathcal{E}_{\chi}} |f(z)|.
-        \label{equ:bernstein-bound}
-    \end{equation}
-    In particular, when $f = g_{\sigma}(t - \cdot)$ we observe for $z = x + iy \in \mathcal{E}_{\chi}$
-    \begin{equation}
-    | g_{\sigma}(t - z) | 
-    = \frac{1}{\sigma \sqrt{2 \pi}} \left| e^{-\sfrac{(t - z)^2}{2\sigma^2}} \right|
-    = \frac{1}{\sigma \sqrt{2 \pi}} e^{-\sfrac{(t - x)^2}{2\sigma^2}}e^{\sfrac{y^2}{2\sigma^2}}.
-    %\leq \frac{1}{n \sigma \sqrt{2 \pi}} \max_{x + iy \in \mathcal{E}_{\chi}} e^{\sfrac{y^2}{2\sigma^2}} 
-    \end{equation}
-    Because $e^{-\sfrac{(t - x)^2}{2\sigma^2}} \leq 1$ for all $x, t \in \mathbb{R}$, and the maximum absolute value of $y$ is limited by the length of the semi-axis of the Bernstein ellipse in the direction of the imaginary axis, which is $(\chi - \chi^{-1}) / 2$, we upper bound
-    \begin{equation}
-        \max_{z \in \mathcal{E}_{\chi}} | g_{\sigma}(t - z) | 
-        \leq \frac{1}{\sigma \sqrt{2 \pi}} e^{\sfrac{(\chi - \chi^{-1})^2}{8 \sigma^2}}.
-    \end{equation}
-
-    Choosing $\chi = 1 + \sigma$ gives a simple and good bound, since in this case $\chi - \chi^{-1} \leq 2\sigma$, which implies
-    \begin{equation}
-        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \right| \leq \sqrt{\frac{2e}{\pi}} \frac{1}{\sigma^2} (1 + \sigma)^{-m}.
-    \end{equation}
-\end{proof}
-
-\Reflem{lem:chebyshev-error} shows that when $0 < \sigma \ll 1$ and if we choose $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, the error we induce with the Chebyshev expansion is at most $\mathcal{O}(\varepsilon)$.
-It now also allows us to bound the error we make when expanding the spectral density.
-
-\begin{theorem}{$L^1$-error of Chebyshev expansion of spectral density}{chebyshev-error}
-    Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the spectal density with Chebyshev expansion $\phi_{\sigma}^{(m)}$ satisfies
-    \begin{equation}
-        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq 2 E_{\sigma, m}
-        \label{equ:chebyshev-interpolation-spectral-density}
-        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
-        %\label{equ:2-chebyshev-interpolation-error}
-    \end{equation}
-    for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and the error $E_{\sigma, m}$ from \reflem{lem:chebyshev-error}.%, and $t \in \mathbb{R}$.
-%The same result holds for the expansion of a non-negative Chebyshev expansion of \emph{even} degree $m \in \mathbb{N}$ but with $\underline{E}_{\sigma, m}$.
-\end{theorem}
-
-\begin{proof}
-    We observe that for all $t \in [-1, 1]$
-    \begin{align}
-        &\left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right| \notag \\
-        &= \left| \frac{1}{n} \Trace(g_{\sigma}(t\mtx{I}_n - \mtx{A})) - \frac{1}{n} \Trace(g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})) \right|
-        && \text{(definitions \refequ{equ:smooth-spectral-density} and \refequ{equ:expanded-spectral-density})} \notag \\
-        &= \left| \frac{1}{n} \sum_{i=1}^n \left(g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i)\right) \right|
-        && \text{($\lambda_1, \dots, \lambda_n$ eigenvalues of $\mtx{A}$)} \notag \\
-        &\leq \max_{i = 1, \dots, n} \left| g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i) \right|
-        && \text{(conservative upper bound)} \notag \\
-        &\leq E_{\sigma, m}.% ~ \text{(or $\underline{E}_{\sigma, m}$)}.
-        && \text{(using \reflem{lem:chebyshev-error})} \notag \\
-    \end{align}
-    Then, Hölder's inequality allows us to conclude
-    \begin{equation}
-        \int_{-1}^{1} | \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) | ~\mathrm{d}t
-        \leq 2 \sup_{t \in [-1, 1]} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|
-        \leq 2 E_{\sigma, m}.% ~ \text{(or $2 \underline{E}_{\sigma, m}$)}.
-    \end{equation}
-\end{proof}
-
-\subsection{Chebyshev-Nyström++ method for spectral density approximation}
-\label{subsec:chebyshev-nystrom}
-
-Our goal is to apply the Nyström++ trace estimator which we analyzed in \refsec{subsec:nystrom-pp} to the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ to approximate the smoothed spectral density $\phi_{\sigma}$.
-However, due to the oscillatory nature of the Chebyshev expansion, we can no longer guarantee that $g_{\sigma}^{(m)} \geq 0$, and consequently, the matrix function $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ may be (slightly) indefinite.
-While in our experiments this never seemed to be an issue, it disallows us to directly use \refthm{thm:nystrom-pp}.
-While variants of the Nyström approximation designed for indefinite matrices have been analyzed, no practical algorithms for computing them exist \cite{nakatsukasa-2023-randomized-lowrank}.
-
-To still apply the bound from \refthm{thm:nystrom-pp} in this scenario, we will have to ensure that the Chebyshev expansion is non-negative.
-We can think of many ways in which this can be achieved: by using Jackson damping \cite{jackson-1912-approximation-trigonometric,braverman-2022-sublinear-time}, albeit at a significantly worse rate of convergence; by shifting up the expansion $g_{\sigma}^{(m)} + \rho$ with a large enough $\rho > 0$, at the loss of most of the low-rank structure of $g_{\sigma}(t \mtx{I}_n - \mtx{A})$; or by directly minimizing the approximation error over all non-negative Chebyshev polynomials \cite{fejer-1916-uber-trigonometrische}, which realistically only works for small $m$.
-In the end, we propose the following method to produce a non-negative Chebyshev expansion: First, expand the square root of the smoothing kernel in a basis of Chebyshev polynomials up to degree $m/2$
-\begin{equation}
-    \sqrt{g_{\sigma}}^{(\sfrac{m}{2})}(t - s) = \sum_{l=0}^{m/2} \xi_l(t) T_l(s).
-\end{equation}
-Second, square this expansion with \refalg{alg:chebyshev-squaring} to obtain the non-negative Chebyshev expansion
-\begin{equation}
-    \underline{g}^{(m)}_{\sigma}(t - s) = \sum_{l=0}^{m} \mu_l(t) T_l(s).
-    \label{equ:non-negative-chebyshev-expansion}
-\end{equation}
-With a complexity of $\mathcal{O}(m \log(m))$ this procedure is computationally inexpensive and allows for very similar convergence bounds to \reflem{lem:chebyshev-error}, which we will derive in \refsec{subsubsec:chebyshev-nystrom-analysis}.
-In \reftab{tab:chebyshev-timing-interpolation}, the non-negative expansion can be seen to only be slightly slower than the normal one.
-
-\begin{table}[ht]
-    \caption{Runtime in milliseconds for computing the coefficients of the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ \refequ{equ:smoothing-kernel} with $\sigma=0.05$, at $n_t=1000$ points, and for various degrees $m$.
-We average over 7 runs of the algorithms and repeat these runs 1000 times to form the mean and standard deviation which are given in the below table.}
-    \label{tab:chebyshev-timing-interpolation}
-   \input{tables/interpolation.tex}
-\end{table}
-
-%\todo{
-%Maybe mention indefinite Nyström [Nakatsukasa], which won't give us desirable guarantees, and refinement of Chebyshev expansion [Francis Bach], which only slightly improves, but is much harder to implement.
-Side-note: it is possible to refine the coefficients of the Chebyshev expansion to get a better approximation in the supremum-norm over the set of all non-negative polynomials \cite%{fejer-1916-uber-trigonometrische}.
-The non-convex nature of this problem and the relatively small improvement we get with this technique made us decide to not pursue this further.
-%}
-
-In the end, we propose the following approximation for the smoothed spectral density
-\begin{equation}
-    \widetilde{\phi}_{\sigma}^{(m)}(t) = \frac{1}{n} \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}))
-    \label{equ:chebyshev-nystrom-formula}
-\end{equation}
-which is the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$ \refequ{equ:nystrompp-trace-estimator} applied to the non-negative Chebyshev expansion $\underline{g}_{\sigma}^{(m)}$ \refequ{equ:non-negative-chebyshev-expansion}.
-
-\subsubsection{Implementation}
-\label{subsubsec:chebyshev-nystrom-implementation}
-
-An efficient implementation of the approximation \refequ{equ:chebyshev-nystrom-formula} can be achieved thanks to the invariance of the trace under cyclic permutation of its arguments and the recurrence relation \refequ{equ:chebyshev-recurrence} which the Chebyshev polynomials satisfy.
-To illustrate this, consider the approximation \refequ{equ:chebyshev-nystrom-formula} when $\mtx{\Psi}$ is zero.
-Using the definition of the Nyström approximation \refequ{equ:nystrom-approximation}, the cyclic invariance of the trace, and the symmetry of $\mtx{A}$, we can rewrite it as
-\begin{equation}
-    \frac{1}{n} \Trace((\mtx{\Omega}^{\top} \underline{g}^{(m)}_{\sigma}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega})^{\dagger} (\mtx{\Omega}^{\top} \underline{g}^{(m)}_{\sigma}(t \mtx{I}_n - \mtx{A})^2 \mtx{\Omega})).
-    \label{equ:cyclic-property}
-\end{equation}
-Thanks to the affine form of the Chebyshev expansion, the first parenthesized term in the trace is
-\begin{equation}
-    \mtx{\Omega}^{\top} \underline{g}^{(m)}_{\sigma}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega} = \sum_{l=0}^{m} \mu_l(t) (\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega})
-\end{equation}
-for which we can first precompute the small $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ for $l=0, \dots, m$ using the recurrence \refequ{equ:chebyshev-recurrence} and then inexpensively sum them up with the corresponding coefficients $\mu_l(t)$ for all $t$ in which we want to evaluate the approximation.
-Thanks to the fast and exact squaring of Chebyshev expansions (cf.
-\refalg{alg:chebyshev-squaring}), the second parenthesized term in \refequ{equ:cyclic-property} can be approximated in an equivalent way.
-Unlike expanding the squared smoothing kernel $g_{\sigma}^2$ separately \cite{lin-2017-randomized-estimation}, this approach is consistent, which we have observed to make a noticeable difference in terms of accuracy (cf.
-\reffig{fig:interpolation-issue}), and which enables provability.
-
-\begin{figure}[ht]
-    \centering
-    \input{plots/interpolation.pgf}
-    \caption{Difference in the accuracy when computing \refequ{equ:3-nystrom-spectral-density} using a separate expansion (interpolation) versus explicitly squaring the matrix function (squaring).
-With the NC, which we will introduce on the next few pages, the spectral density of a 2D model problem from \refsec{sec:5-experiments-density-function} is computed for a Gaussian smoothing-kernel with smoothing-parameter $=0.05$ and fixed sketch-size $=80$.
-An estimate of the relative error in the $L^1$-norm is computed based on num-evaluation-points $=100$ evaluation of the spectral density.}
-    \label{fig:interpolation-issue}
-\end{figure}
-
-Similar observations for the case when $\mtx{\Psi}$ is nonzero lead to the following algorithm.
-
-\begin{algo}{Chebyshev-Nyström++ method}{nystrom-chebyshev-pp}
-\begin{algorithmic}[1]
-    \Statex \textbf{Input:} Symmetric $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum in $[-1, 1]$, points $\{t_i\}_{i=1}^{n_t} \subset \mathbb{R}$
-    \Statex \textbf{Parameters:} Degree of expansion $m \in 2\mathbb{N}$, number of Girard-Hutchinson queries $n_{\mtx{\Psi}} \in \mathbb{N}_0$, Nyström sketch size $n_{\mtx{\Omega}} \in \mathbb{N}_0$,  smoothing parameter $\sigma > 0$
-    \Statex \textbf{Output:} Approximate evaluations of the spectral density $\{\widetilde{\phi}_{\sigma}(t_i)\}_{i=1}^{n_t}$
-    \State Compute $\{\mu_l(t_i)\}_{l=0}^{m}$ for all $t_i$ with non-negative Chebyshev expansion \refequ{equ:non-negative-chebyshev-expansion}
-    \State Compute $\{\nu_l(t_i)\}_{l=0}^{2m}$ using \refalg{alg:chebyshev-squaring}%\refalg{alg:2-chebyshev-chebyshev-expansion}
-    %\State Compute  for all $t_i$ using %\refalg{alg:3-nystrom-chebyshev-exponentiation}
-    \State Generate standard Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$%\glsfirst{sketching-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$
-    %\State Generate standard Gaussian %\glsfirst{random-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$
-    \State Initialize $[\mtx{V}_1, \mtx{V}_2, \mtx{V}_3] \gets [\mtx{0}_{n \times n_{\mtx{\Omega}}}, \mtx{\mtx{\Omega}}, \mtx{0}_{n \times n_{\mtx{\Omega}}}]$
-    \State Initialize $[\mtx{W}_1, \mtx{W}_2, \mtx{W}_3] \gets [\mtx{0}_{n \times n_{\mtx{\Psi}}}, \mtx{\Psi}, \mtx{0}_{n \times n_{\mtx{\Psi}}}]$
-    \State Initialize $[\mtx{K}_1(t_i), \mtx{K}_2(t_i)] \gets [\mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}}, \mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}}]$ for all $t_i$
-    \State Initialize $[\mtx{L}_1(t_i), \ell(t_i)] \gets [\mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Psi}}}, 0]$ for all $t_i$
-    %\State Set $\breve{\phi}_{\sigma}^{(m)}(t_i) \gets 0$ for all $t_i$
-    \For {$l = 0, \dots, 2m$}
-    \State $[\mtx{X}, \mtx{Y}] \gets \mtx{\mtx{\Omega}}^{\top} [\mtx{V}_2, \mtx{W}_2]$  
-    %\State $\mtx{X} \gets \mtx{\mtx{\Omega}}^{\top} \mtx{V}_2$
-      %\State $\mtx{Y} \gets \mtx{\mtx{\Omega}}^{\top} \mtx{W}_2$
-      \State $z \gets \Trace(\mtx{\Psi}^{\top} \mtx{W}_2)$
-      \For {$i = 1, \dots, n_t$}
-        \If {$l \leq m$}
-            \State $\mtx{K}_1(t_i) \gets \mtx{K}_1(t_i) + \mu_l(t_i) \mtx{X}$ \Comment{assemble $\mtx{\Omega}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$}
-            \State $\mtx{L}_1(t_i) \gets \mtx{L}_1(t_i) + \mu_l(t_i) \mtx{Y}$ \Comment{assemble $\mtx{\Omega}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi}$}
-            \State $\ell(t_i) \gets \ell(t_i) + \mu_l(t_i) z$ \Comment{assemble $\Trace(\mtx{\Psi}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi})$}
-        \EndIf
-        \State $\mtx{K}_2(t_i) \gets \mtx{K}_2(t_i) + \nu_l(t_i) \mtx{X}$ \Comment{assemble $\mtx{\Omega}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})^2 \mtx{\Omega}$}
-      \EndFor
-      \State $[\mtx{V}_3, \mtx{W}_3] \gets (2 - \delta_{l0}) \mtx{A} [\mtx{V}_2, \mtx{W}_2] - [\mtx{V}_1, \mtx{W}_1]$ \Comment{Chebyshev recurrence}
-      \State $[\mtx{V}_1, \mtx{W}_1] \gets [\mtx{V}_2, \mtx{W}_2]$
-      \State $[\mtx{V}_2, \mtx{W}_2] \gets [\mtx{V}_3, \mtx{W}_3]$
-      %\State $\mtx{V}_3 \gets (2 - \delta_{l0}) \mtx{A} \mtx{V}_2 - \mtx{V}_1$ %\Comment{Chebyshev recurrence \refequ{equ:2-chebyshev-chebyshev-recursion}}
-      %\State $\mtx{V}_1 \gets \mtx{V}_2, \mtx{V}_2 \gets \mtx{V}_3$
-      %\State $\mtx{W}_3 \gets (2 - \delta_{l0}) \mtx{A} \mtx{W}_2 - \mtx{W}_1$ %\Comment{Chebyshev recurrence \refequ{equ:2-chebyshev-chebyshev-recursion}}
-      %\State $\mtx{W}_1 \gets \mtx{W}_2, \mtx{W}_2 \gets \mtx{W}_3$
-    \EndFor
-    \For {$i = 1, \dots, n_t$}
-      \State $\widetilde{\phi}_{\sigma}(t_i) \gets \frac{1}{n} \Trace\left( \mtx{K}_1(t_i)^{\dagger}\mtx{K}_2(t_i) \right) + \frac{1}{n n_{\mtx{\Psi}}} \left( \ell(t_i) + \Trace\left( \mtx{L}_1(t_i)^{\top} \mtx{K}_1(t_i)^{\dagger} \mtx{L}_1(t_i) \right)  \right) $ \label{lin:4-nystromchebyshev-nystrom-pp}
-    \EndFor
-\end{algorithmic}
-\end{algo}
-
-The dominating cost stems from the $(2m + 1)(n_{\mtx{\Psi}} + n_{\mtx{\Omega}})$ matrix-vector multiplications with $\mtx{A}$ and the cost associated with computing the estimate in each evaluation point $t_1, \dots, t_{n_t}$ given by $\mathcal{O}(m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + (n_{\mtx{\Omega}}^3 + n_{\mtx{\Omega}} n_{\mtx{\Psi}}^2 + 1))$.
-
-The algorithm also accomodates the two other estimators introduced in \refsec{sec:analysis}: For $n_{\mtx{\Omega}} = 0$ we recover the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}, and the complexity for each evaluation of the estimate decreases significantly.
-For $n_{\mtx{\Psi}} = 0$ we end up with the Nyström estimator \refequ{equ:nystrom-trace-estimator}, provided the implementation evaluates $0/0$ to $0$ in the second term on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
-
-%Chebyshev $\mathcal{O}(n_t (m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + n_{\mtx{\Omega}}^3) + mn^2(n_{\mtx{\Psi}} + n_{\mtx{\Omega}}))$.
-
-%\todo{Complexity in terms of both $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$.}
-%With the cost of a matrix-vector product denoted by $c(n)$, and supposing $n_{\mtx{\Omega}} \approx n_{\mtx{\Psi}}$, we determine the computational complexity of the Chebyshev-Nyström++ method to be $\mathcal{O}(m \log(m) n_t + m n_{\mtx{\Omega}}^2 n + m n_t n_{\mtx{\Omega}}^2 +  m c(n) n_{\mtx{\Omega}} + n_t n_{\mtx{\Omega}}^3)$, with $\mathcal{O}(m n_t + n n_{\mtx{\Omega}} + n_{\mtx{\Omega}}^2 n_t)$ required additional storage.
-
-%\todo{Explain modifications for other estimators, and improved complexity for quadratic trace estimator}
-
-The pseudoinverses on \reflin{lin:4-nystromchebyshev-nystrom-pp} need to be computed with care.
-It is possible to reformulate the problem as a generalized eigenvalue problem and apply thresholding to the smallest eigenvalues of a certain matrix to enforce stability \cite{lin-2017-randomized-estimation, epperly-2022-theory-quantuma}.
-Since this renders the algorithm significantly more complicated but only improved the accuracy slightly during our experiments, we will instead evaluate these expressions in \reflin{lin:4-nystromchebyshev-nystrom-pp} using a least-squares solver which likewise uses a truncation of the smallest eigenvalues to enforce better conditioning.
-
-What we have found to be crucial for a successful approximation of the full spectral density was the detection of regions wherein lie no eigenvalues, i.e.
-regions where the spectral density nearly vanishes.
-For small smoothing parameters $\sigma$, the rapid decay of the Gaussian $g_{\sigma}$ will cause $\mtx{K}(t_i) \approx g_{\sigma}(t_i\mtx{I}_n - \mtx{A})$ to be close to the zero matrix for parameter values $t_i$ located in such regions.
-Hence, on line \reflin{lin:4-nystromchebyshev-nystrom-pp} we would compute the pseudoinverse of an almost zero matrix, which can be extremely unstable.
-Notice that $\Trace(\mtx{K}_1(t_i)) / n_{\mtx{\Omega}}$ is the $n_{\mtx{\Omega}}$-query Girard-Hutchinson estimator.
-We may use it as a rough indicator whether $t_i$ lies within a region where the spectral density nearly vanishes, and subsequently set $\widetilde{\phi}_{\sigma}(t_i) = 0$ instead of computing the full expression on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
-
-\subsubsection{Analysis}
-\label{subsubsec:chebyshev-nystrom-analysis}
-
-In this section, we will upper bound the output of the Chebyshev-Nyström++ method (see. \refalg{alg:nystrom-chebyshev-pp}). To this extent, we will first analyze the non-negative Chebyshev expansion.
-
-\begin{lemma}{Non-negative Chebyshev expansion error}{non-negative-chebyshev-error}
-    The non-negative expansion $\underline{g}_{\sigma}^{(m)}$ of the Gaussian smoothing kernel satisfies
-    \begin{equation}
-        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \leq 2\sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right) E_{\sqrt{2}\sigma, \sfrac{m}{2}} \equiv \underline{E}_{\sigma, m}
-        \label{equ:chebyshev-interpolation-sup-error-kernel-nonneg}
-        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
-        %\label{equ:2-chebyshev-interpolation-error}
-    \end{equation}
-    for all \emph{even} degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, $s \in [-1, 1]$, and with $E_{\sigma, m}$ as defined in \reflem{lem:chebyshev-error}.
-\end{lemma}
-
-\begin{proof}
-    For any numbers $a, b \in \mathbb{R}$ it holds
-    \begin{equation}
-    | a^2 - b^2 | = | (a + b)(a - b) | = | a + b | | a - b | \leq (2 | a | + | a - b |)  | a - b |
-    \end{equation}
-    Therefore, omitting the arguments and using $g_{\sigma} = (\sqrt{g_{\sigma}})^2$ and $\underline{g}_{\sigma}^{(m)} = (\sqrt{g_{\sigma}}^{(\sfrac{m}{2})})^2$.
-    we have
-    \begin{equation}
-        \left| g_{\sigma} - \underline{g}_{\sigma}^{(m)} \right| \leq \left( 2 \left| \sqrt{g_{\sigma}} \right| + \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right| \right) \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right|.
-    \end{equation}
-    Since $\sqrt{g_{\sigma}} = \sqrt{2 \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$ and $\sqrt{g_{\sigma}} \leq 1/\sqrt{\sigma \sqrt{2 \pi}}$ we can apply the result from \refthm{thm:chebyshev-error} to get
-    \begin{align}
-        &\sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \notag \\
-        &\leq \left( 2 \frac{1}{\sqrt{ \sigma \sqrt{2\pi}}} + \sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right)\sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}},
-    \end{align}
-    from which follows the result with some minor simplifications.
-\end{proof}
-
-On one hand, in the non-negative expansion, the underlying computes Chebyshev expansion -- which is shown in \reflem{lem:chebyshev-error} to exponentially decay with its degree -- is only computed to degree $m/2$, whereas the standard expansion is of degree $m$.
-On the other hand, we approximate $\sqrt{g_{\sigma}}$ instead of $g_{\sigma}$, which is less \enquote{pointy} and therefore better approximated by polynomials.
-Overall, we will need to choose the degree $m$ of the expansion approximately a factor of $\sqrt{2}$ larger to match the error of the standard expansion.
-
-\begin{theorem}{Chebyshev-Nyström++ method}{chebyshev-nystrom}
-    Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, \emph{even} $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2)$, and smoothing parameter $0 < \sigma \ll 1$ satisfies with probability at least $1 - \delta$
-    \begin{equation}
-        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq \varepsilon
-        \label{equ:chebyshev-nystrom-error}
-        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
-        %\label{equ:2-chebyshev-interpolation-error}
-    \end{equation}
-    for all $\varepsilon \in (0, 1)$ and $\delta \in (0, e^{-\sfrac{1}{2}}/2)$.
-\end{theorem}
-
-\begin{proof}
-    The previously derived results can be applied to
-    \begin{align}
-        &\int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \notag \\
-        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \int_{-1}^{1} \left| \phi^{(m)}_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
-        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \widetilde{\varepsilon} \int_{-1}^{1} \left| \phi^{(m)}_{\sigma}(t) \right|~\mathrm{d}t && \text{(\refthm{thm:nystrom-pp} w.p.
-$\geq 1 - \delta$)} \notag \\
-        &\leq (1 + \widetilde{\varepsilon}) \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \varepsilon \int_{-1}^{1} \left| \phi_{\sigma}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
-        &\leq (1 + \widetilde{\varepsilon}) 2 \underline{E}_{\sigma, m} + \widetilde{\varepsilon}. && \text{(\refthm{thm:chebyshev-error}, normalization)} \notag \\
-    \end{align}
-    From \reflem{lem:non-negative-chebyshev-error} we see that if $0 < \sigma \ll 1$, choosing $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$ ensures $\underline{E}_{\sigma, m} \leq \widetilde{\varepsilon}$, and thus, since $\widetilde{\varepsilon} \leq 1$,
-    \begin{equation}
-        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq (1+\widetilde{\varepsilon}) 2 \widetilde{\varepsilon} + \widetilde{\varepsilon} \leq 5 \widetilde{\varepsilon}
-    \end{equation}
-    from which the result follows by identifying $\varepsilon = 5 \widetilde{\varepsilon}$.
-\end{proof}
-
-%\todo{Maybe specialize \refthm{thm:nystrom} to spectral densities?}
-
-%We now discuss what value to choose for $n_{\mtx{\Omega}}$ in the Nyström trace estimator introduced in \refsec{subsec:nystrom}.
-In order to guarantee an even better approximation error based on \refthm{thm:nystrom}, we analyze the numerical rank of the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
-
-In practice, the approximation error is often significantly better than guaranteed in \refthm{thm:chebyshev-nystrom}, particularly when $n_{\mtx{\Omega}}$ is chosen rather large.
-This is because the singular values of the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ decay so quickly that already the Nyström approximation alone is highly accurate (cf.
-\refthm{thm:nystrom}).
-To establish a relation between the singular value decay of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ and what choice of $n_{\mtx{\Omega}}$ is needed to achieve a better approximation than what we show in \refthm{thm:chebyshev-nystrom}, we study the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
-
-We can express the numerical rank of a matrix with respect to the nuclear norm in terms of its singular values $\sigma_1 \geq \dots \geq \sigma_n \geq 0$
-\begin{equation}
-    r_{\varepsilon} = \min \{1 \leq r \leq n: \sum_{j=r+1}^n \sigma_{j} \leq \varepsilon \},
-    \label{equ:numerical-rank}
-\end{equation}
-$\varepsilon$ is usually taken to be the double machine precision, i.e. $10^{-16}$.
-
-%\begin{theorem}{Nyström estimator for parameter-dependent matrices}%{chebyshev-nystrom-raw}
-%    Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric %matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is %contained in $[-1, 1]$.
-Then the Chebyshev-Nyström++ estimate %$\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m \in %\mathbb{N}$, $n_{\mtx{\Omega}} \geq \max(6, \max_{t \in [-1, 1]}(r_%{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A}))) + \log(\delta^%{-1}))$, $n_{\mtx{\Psi}} = 0$, and smoothing parameter $\sigma > 0$ %satisfies with for all $\gamma \geq 1$ with probability at least $1 %- \delta$
-%    \begin{equation}
-%        \int_{a}^{b} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}%(t) \right| ~\mathrm{d}t
-%        \leq 2 e^2 (1 + r_{\varepsilon}) (\varepsilon + (n - r_%{\varepsilon}) \underline{E}_{\sigma, m}) + \underline{E}_%{\sigma, m}.
-%    \end{equation}
-%\end{theorem}
-%
-%\begin{proof}
-%    \todo{TODO}
-%\end{proof}
-
-When using a Gaussian smoothing-kernel, we can explicitly write the singular values of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ as
-\begin{equation}
-    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) = \frac{1}{\sqrt{2 \pi \sigma^2}} e^{-\sfrac{(t - \lambda_{(i)})^2}{2 \sigma^2}},
-    \label{equ:gaussian-kernel-eigenvalues}
-\end{equation}
-where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from spectral-parameter $t$, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t)$.
-Consequently, by using the closed-form expression of the eigenvalues \refequ{equ:gaussian-kernel-eigenvalues}, we may upper bound the numerical rank of a Gaussian smoothing kernel as
-\begin{equation}
-    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A})) \leq \#\left\{1\leq i\leq n: |t - \lambda_i| < d_{\varepsilon, \sigma} \right\},
-    \label{equ:gaussian-kernel-numerical-rank}
-\end{equation}
-with $d_{\varepsilon, \sigma} = \sigma \sqrt{-2 \log(\sqrt{2 \pi} \sigma \varepsilon)}$ and the eigenvalues $\lambda_1, \dots, \lambda_n$ of $\mtx{A}$ and the constants.
-The expression \refequ{equ:gaussian-kernel-numerical-rank} has a nice visual interpretation: The numerical rank of a matrix is at most equal to the number of eigenvalues which are closer to the parameter $t$ than $d_{\varepsilon, \sigma}$.
-This is illustrated in \reffig{fig:numerical-rank}.
-\begin{figure}[ht]
-    \centering
-    \input{figures/numerical-rank.tex}
-    \caption{The numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ can be
-        approximately computed by counting the number of eigenvalues
-        $\lambda_{(1)}, \dots, \lambda_{(n)}$ of the matrix $\mtx{A}$ which lie less than
-        a constant $d_{\varepsilon, \sigma}$ away from $t$.}
-    \label{fig:numerical-rank}
-\end{figure}
-
-If we additionally assume the eigenvalues of the matrix $\mtx{A}$ to be evenly distributed within $[-1, 1]$, that is, in any subinterval of fixed length in $[a, b]$, $-1 \leq a < b \leq 1$, we can expect to find roughly the same number of eigenvalues, then we can estimate the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ to be
-\begin{equation}
-    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A})) \lessapprox n d_{\varepsilon, \sigma} \approx n \sigma.
-    \label{equ:gaussian-kernel-numerical-rank-uniform}
-\end{equation}
-Hence, choosing $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma)$ will approximations which are significantly better than \refthm{thm:chebyshev-nystrom}.
-
-\section{Numerical results}
-\label{sec:results}
-
-Our first numerical example comes from electronic structure interaction \cite{lin-2017-randomized-estimation}, more precisely from the second order finite difference discretization of the Hamiltonian
-\begin{equation}
-    \mathcal{H} = - \Delta + V
-    \label{equ:5-experiments-electronic-hamiltonian}
-\end{equation}
-in three dimensions. The potential $V$ interacting with the electrons is generated by Gaussian wells
-\begin{equation}
-    v(r) = v_0 e^{-\lambda r^2}
-    \label{equ:5-experiments-gaussian-cell}
-\end{equation}
-with $v_0 = -4$ and $\lambda = 8$, centered in cells of side length $L=6$ which are stacked $n_c \in \mathbb{N}$ times in each spatial dimension (cf.
-\reffig{fig:gaussian-well}).
-The discretization step is fixed to be $h=0.6$, such that the size of the matrix grows cubically with $n_c$.
-This is an idealized model for the interaction of nuclei on a regular grid with electrons for a $k$-vector in the center of the first Brillouin zone.
-The distribution of the eigenvalues of the Hamiltonian -- its spectral density -- allows us to interpret the system's energy levels.
-
-\begin{figure}[ht]
-    \begin{subfigure}[b]{0.32\columnwidth}
-        \input{plots/gaussian-well-1.pgf}
-        \caption{$n_c=1$}
-        \label{fig:gaussian-well-1}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.32\columnwidth}
-        \input{plots/gaussian-well-2.pgf}
-        \caption{$n_c=2$}
-        \label{fig:gaussian-well-2}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.32\columnwidth}
-        \input{plots/gaussian-well-5.pgf}
-        \caption{$n_c=5$}
-        \label{fig:gaussian-well-5}
-    \end{subfigure}
-    \caption{Cross-sections of the periodic Gaussian well potential $V$ for different sizes of the supercell.}
-    \label{fig:gaussian-well}
-\end{figure}
-
-\todo{Explain experimental setup.}
-
-\begin{figure}[ht]
-    \centering
-    \input{plots/convergence.pgf}
-    \caption{For increasing values of $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}$ but fixed $m$ we plot the $L^1$-approximation error for $\sigma=0.005$ on the model problem with $n_c = 1$.}
-    \label{fig:convergence}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \input{plots/distribution.pgf}
-    \caption{The Chebyshev-Nyström++ method for different ways of allocations a total of $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation for the Gaussian smoothing kernel with multiple different values of the smoothing parameter.
-We make the approximation error made in the Chebyshev expansion negligible by rescaling $m=16 / \sigma$ (based on \refthm{thm:chebyshev-error}).}
-    \label{fig:distribution}
-\end{figure}
-
-\todo{Interpret results.}
-
-\todo{Comparison plots with other estimators (Krylov-Aware)?}
-
-\clearpage
-\section{Conclusion}
-\label{sec:conclusion}
-
-\todo{
-\begin{itemize}
-    \item Robustness (can choose $m$ and $n_{\mtx{\Omega}}$, $n_{\mtx{\Psi}}$ as large as one wants, and not break algorithm, unlike Krylov-aware Lanczos)
-\end{itemize}
-}
-
-% Acknowledge advice from David and Data/Parameters from Lin Lin, Michael Herbst?
-
-\clearpage
-%\bibliography{bibliography.bib}
-\printbibliography
-
-\appendix
-
-\clearpage
-\section{Proofs}
-
-\begin{lemma}{Moment bound of $\chi^2$-random variable}{gamma}
-    Let $X$ follow a $\chi^2$ distribution with $k \in \mathbb{N}$ degrees of freedom. Then for all $p \in \mathbb{N}$ it holds
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}[X] \leq k + p
-        %\frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \leq \left( \frac{k + p}{2} \right)^{\sfrac{p}{2}}.
-    \end{equation}
-\end{lemma}
-
-\begin{proof}
-    From \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical} we know
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}[X] = 2 \left( \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \right)^{\sfrac{2}{p}}.
-    \end{equation}
-    We first treat the case $k=p=1$, for which indeed
-    \begin{equation}
-        2 \left( \frac{\Gamma(\frac{1+1}{2})}{\Gamma(\frac{1}{2})} \right)^2 = \frac{2}{\pi} \leq 2 = 1 + 1
-    \end{equation}
-    So we now assume $k + p \geq 3$. We can use a telescoping product and the property $\Gamma(z+1)/\Gamma(z) = z$ to bound
-    \begin{align}
-        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} 
-        &= \frac{\Gamma(\frac{k}{2} + 1)}{\Gamma(\frac{k}{2})} \frac{\Gamma(\frac{k}{2} + 2)}{\Gamma(\frac{k}{2} + 1)} \cdots \frac{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor - 1)} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)} \notag \\
-        &= \left(\frac{k}{2}\right) \left(\frac{k}{2} + 1\right) \cdots \left( \frac{k}{2} + \bigg\lfloor \frac{p}{2} \bigg\rfloor - 1 \right) \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)} \notag \\
-        &\leq \left(\frac{k}{2} + \bigg\lfloor \frac{p}{2} \bigg\rfloor - 1\right)^{\lfloor \sfrac{p}{2} \rfloor} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)}
-    \end{align}
-    If $p$ is even, then $\lfloor p/2 \rfloor = p/2$ and hence
-    \begin{equation}
-        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} 
-        \leq \left( \frac{k + p}{2} - 1 \right)^{\sfrac{p}{2}}
-        \leq \left( \frac{k + p}{2}\right)^{\sfrac{p}{2}}.
-    \end{equation}
-    If $p$ is odd, then $\lfloor p/2 \rfloor = (p - 1)/2$, from which follows with Gautschi's inequality \cite{kershaw-1983-extensions-gautschi}
-    \begin{equation}
-        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})}
-        \leq \left(\frac{k + p}{2} - \frac{3}{2} \right)^{ \frac{p - 1}{2}} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k + p}{2} - \frac{1}{2})}
-        \leq \left(\frac{k + p}{2}\right)^{ \frac{p - 1}{2}} \left(\frac{k + p}{2}\right)^{\sfrac{1}{2}}
-        = \left(\frac{k + p}{2}\right)^{ \sfrac{p}{2}}
-    \end{equation}
-    Hence,
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}[X] \leq 2 \left( \frac{k + p}{2} \right) \leq k + p.
-    \end{equation}
-\end{proof}
-
-\end{document}

### 2024-07-07 paper/numresults.tex

--- 2024-06-27 paper/main.tex
@@
-\documentclass[12pt]{article}
-
-\usepackage{stylesheet}
-
-\title{Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation}
-
-\begin{document}
-
-\maketitle
-
-%\todo{
-%\begin{itemize}
-%    \item Unify index notation
-%    \item Double-check proofs
-%    \item Add all references
-%    \item Run numerical experiments 
-%    \item Write introduction, abstract, and discussion
-%    \item ArXiv HTML support (check if packages)
-%\end{itemize}
-%}
-
-%\todo{
-%    \tableofcontents
-%}
-
-\todo{
-
-\paragraph{TODO}
-\begin{itemize}
-    \item Check that everywhere correct/consistent orders $\mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1})$)
-    \item 
-\end{itemize}
-}
-
-\todo{
-\begin{abstract}
-    Write abstract.
-\end{abstract}
-}
-
-\section{Introduction}
-\label{sec:introduction}
-
-In numerous problems across physics, chemistry, engineering, and data science, how the eigenvalues of certain matrices are distributed can be indicative of what properties a system has: In electronic structure calculations eigenvalues represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}.
-The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$.
-Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation which is often prohibitively expensive.
-Therefore, many techniques have been proposed for approximating $\phi$.
-For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature rule with the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}.
-Further, because $\delta$ is hard to approximate and in many applications, a rough estimate of the spectral density is already acceptable, a standard approach is to instead approximate the smoothed spectral density
-\begin{equation}
-    \phi_{\sigma}(t) = \frac{1}{n} \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i) = \frac{1}{n} \Trace(g_{\sigma}(t \mtx{I}_n - \mtx{A}))
-    \label{equ:smoothed-spectral-density}
-\end{equation}
-for a smoothing kernel $g_{\sigma}$, typically a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}.
-This translates the problem to computing the trace of a continuous parameter-dependent matrix $\mtx{B}(t) \equiv g_{\sigma}(t \mtx{I}_n - \mtx{A}) \in \mathbb{R}^{n \times n}$.
-Directly assembling the matrix function requires diagonalizing the matrix $\mtx{A}$, which again is prohibitively expensive.
-Therefore, we estimate its trace implicitly by approximating products of the matrix function with vectors.
-For constant matrices, the most widespread methods for implicit trace estimation are the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} and the variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making, chen-2023-krylovaware-stochastic}.
-Some of these estimators have already been successfully applied to parameter-dependent matrices \cite{lin-2017-randomized-estimation,chen-2023-krylovaware-stochastic}, but they have not yet been analyzed in this setting.
-A related technique is dynamic trace estimation for sequences $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ of matrices which satisfy $\lVert \mtx{B}(t_i) \rVert _F \leq 1$ and which change dynamically, i.e.
-$\lVert \mtx{B}(t_{i+1}) - \mtx{B}(t_i) \rVert _F \leq \alpha < 1$ for the Frobenius norm $\lVert \cdot \rVert _F$ and a small $\alpha$ \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace}.
-However, in our setting we often cannot guarantee either of the two conditions.
-Instead of discretizing $[a, b]$ into $t_1, \ldots, t_m$, approximating $\Trace(\mtx{B}(t_i))$ with $b(t_i)$, and analyzing the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way is to quantify the error with $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate in each $t \in [a, b]$.
-This approach has already been taken for analyzing low-rank approximations of parameter-dependent matrices \cite{kressner-2023-randomized-lowrank}.
-
-\paragraph{New contributions} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}.
-Interestingly, we can reuse the same randomization for computing the estimate at each value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations.
-We propose certain modifications to the algorithms from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}.
-In particular, use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials, and realize that an efficient implementation of all of their methods can be summarized in a single algorithm.
-Based on the analysis of the randomized trace estimators for parameter-dependent matrices we provide bounds on the approximation error for our method.
-
-\paragraph{Structure} In \refsec{sec:analysis} we will analyze three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter dependent matrices.
-We propose and analyze a method for using the aforementioned estimators to estimate the spectral density of real symmetric matrices in \refsec{sec:application}.
-Finally, we will test our developments on an example from electronic structure calculation.
-
-\paragraph{Reproducibility} This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-SD/tree/ca391a8}{ca391a8} at \href{https://github.com/FMatti/Rand-TRACE}{github.com/FMatti/Rand-TRACE} on 2024-02-05 at 18:44:29 UTC.
-
-\section{Analysis of randomized trace estimators applied to parameter-dependent matrices}
-\label{sec:analysis}
-
-We consider parameter-dependent matrices of the form
-\begin{equation}
-    \mtx{B}(t) = \begin{bmatrix}
-        b_{11}(t) & b_{12}(t) & \dots & b_{1n}(t) \\
-        b_{21}(t) & b_{22}(t) & \dots & b_{2n}(t) \\
-        \vdots & \vdots & \ddots & \vdots \\
-        b_{n1}(t) & b_{n2}(t) & \dots & b_{nn}(t) \\
-    \end{bmatrix} \in \mathbb{R}^{n \times n}
-\end{equation}
-where $b_{ij}(t)$ are functions depending continuously on the parameter $t$ which takes values in the interval $[a,b]$. The trace of such a matrix is defined as
-\begin{equation}
-    \Trace(\mtx{B}(t)) = \sum_{i=1}^{n} b_{ii}(t).
-    \label{equ:definition-trace}
-\end{equation}
-However, we assume that we only have access to products of this matrix with vectors for each $t \in [a, b]$, so this definition will not be directly useful for computing the trace.
-
-\paragraph{Girard-Hutchinson estimator} We can approximate the trace with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ stochastically independent standard Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ to form
-\begin{equation}
-    \Hutch{\mtx{\Psi}}(\mtx{B}(t))
-    = \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j
-    = \frac{1}{n_{\mtx{\Psi}}} \Trace( \mtx{\Psi}^{\top} \mtx{B}(t) \mtx{\Psi})
-    \label{equ:hutchinson-trace-estimator}
-\end{equation}
-where $\mtx{\Psi} = [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\mtx{\Psi}}}] \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$.
-Other choices for the distribution of the random vectors are possible, for example by uniformly sampling from $\{-1, +1\}$ or from the $(n-1)$-sphere.
-However, our theoretical developments only hold in the Gaussian case.
-
-\paragraph{Nyström estimator} Alternatively, the trace of a symmetric matrix whose singular values decay quickly can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström approximation \cite{gittens-2013-revisiting-nystrom}
-\begin{equation}
-    \Nystr{\mtx{\Omega}}{\mtx{B}}(t) = (\mtx{B}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} (\mtx{B}(t) \mtx{\Omega})^{\top}.
-    \label{equ:nystrom-approximation}
-\end{equation}
-Then we can estimate the trace as $\Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t))$.
-Thanks to the invariance of the trace under cyclic permutation of its arguments and the symmetry of the matrix, we may rewrite this estimator as
-
-\begin{equation}
-    \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{B}(t)^2 \mtx{\Omega})).
-    \label{equ:nystrom-trace-estimator}
-\end{equation}
-
-\paragraph{Nyström++ estimator} Finally, \cite{lin-2017-randomized-estimation} proposes an estimator which corrects for inaccuracies in the Nyström approximation \refequ{equ:nystrom-trace-estimator} by estimating the trace of its residual using the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}
-\begin{equation}
-    \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) = \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) + \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)).
-    \label{equ:nystrompp-trace-estimator}
-\end{equation}
-This is the parameter-dependent analogue of the Nyström++ estimator \cite{persson-2022-improved-variants}, which is based on the Hutch++ estimator \cite{meyer-2021-hutch-optimal}.
-%We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.
-
-%In the remainder of this section, we will derive upper bounds on the error of these estimators.
-%
-%\begin{table}[ht]
-%\centering
-%\renewcommand{\arraystretch}{1.2}
-%\begin{tabular}{@{}lcc@{}}
-%\toprule
-%Estimator & Matrix & $m=1600$\\
-%\midrule
-%Girard-Hutchinson & symmetric & $\mathcal{O}(\varepsilon^{-2})%$ \\
-%Nyström & symmetric positive semi-definite & $\mathcal{O}%(\varepsilon^{-2})$ \\
-%Nyström++ & symmetric positive semi-definite & $\mathcal{O}%(\varepsilon^{-1})$  \\
-%\bottomrule
-%\end{tabular}
-%\end{table}
-
-\subsection{Girard-Hutchinson estimator for parameter-dependent matrices}
-\label{subsec:hutchinson}
-
-We first analyze the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}.
-We bound its $L^1$ norm in terms of the integral over the Frobenius norm $\lVert \cdot \rVert _F$ of the matrix whose trace we aim to approximate.
-
-\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
-    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric and continuously depend on $t \in [a, b]$.
-Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ it holds with probability at least $1 - \gamma^{-2k}$
-    \begin{equation}
-        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) \right| ~\mathrm{d}t \leq 2^4 \gamma k \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F  ~\mathrm{d}t.
-    \end{equation}
-    In particular, if we choose $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{- \sfrac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$.
-\end{theorem}
-
-The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Minkowski's integral inequality to extend the results to the continuous parameter case.
-We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
-\begin{proof} 
-    We define the 1-query Girard-Hutchinson estimate 
-    \begin{equation}
-        r(\mtx{B}(t), \vct{\psi}) = \Trace(\mtx{B}(t)) - \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi}
-    \end{equation}
-    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$.
-
-    First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence.
-From the proof of \cite[lemma 3]{cortinovis-2022-randomized-trace} we know that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm.
-Thus, by \cite[theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
-    \begin{align}
-        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
-        &\leq k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k} \notag \\
-        &= k! 2^{4 k} \lVert \mtx{B} \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _2^{2 k}.
-    \end{align}
-    Since $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
-    \begin{equation}
-        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
-    \end{equation}
-    Stirling's approximation \cite{robbins-1955-remark-stirling} bonuds $(2 k)!
-< 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$.
-Consequently, the moments $\mathbb{E}^{k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{k} \right] \right)^{\sfrac{1}{k}}$ of $r(\mtx{B}, \vct{\psi})$ are limited by
-    \begin{equation}
-        \mathbb{E}^{2k}\left[ r(\mtx{B}, \vct{\psi}) \right]
-        \leq \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F \leq 2^4 k \lVert \mtx{B} \rVert _F,
-        \label{equ:hutchinson-trace-onequery-fixed}
-    \end{equation}
-    since it can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, due to the monotonous decrease of this expression in $k \in \mathbb{N}$.
-
-    Now we make the transition to the continuous.
-Following \cite{kressner-2023-randomized-lowrank}, we can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable.
-Therefore, Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting:
-    \begin{equation}
-        \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t  \right]
-        \leq \int_{a}^{b} \mathbb{E}^{2 k}\left[ r(\mtx{B}(t), \vct{\psi}) \right]~\mathrm{d}t
-        \leq 2^4 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
-    \end{equation}
-    Consequently, by Markov's inequality, for any $k \in \mathbb{N}$ and $\gamma \geq 1$, it holds with probability at least $1 - \gamma^{-2 k}$
-    \begin{equation}
-        \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t \leq 2^4 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
-        \label{equ:hutchinson-trace-onequery-uniform}
-    \end{equation}
-
-    To extend the bound to the $n_{\mtx{\Psi}}$-query quadratic trace estimator, we use a technique from the proof of \cite[theorem 1]{cortinovis-2022-randomized-trace}. Let
-    \begin{equation}
-        \widetilde{\mtx{B}}(t)
-        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
-            \mtx{B}(t) & & \\
-            & \ddots & \\
-            & & \mtx{B}(t)
-        \end{pmatrix}
-        \quad \text{and} \quad
-        \widetilde{\vct{\psi}} = \begin{pmatrix}
-            \vct{\psi}_1 \\
-            \vdots \\
-            \vct{\psi}_{n_{\mtx{\Psi}}}
-        \end{pmatrix}
-    \end{equation}
-    where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors.
-Then $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$, and
-    \begin{equation}
-        r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}}) = \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j.
-    \end{equation}
-    Hence, by applying \refequ{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$ we conclude that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
-    \begin{equation}
-        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j \right| ~ \mathrm{d}t
-        \leq 2^4 k \gamma \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
-    \end{equation}
-
-    Setting $\delta = \gamma^{-2 k}$ and choosing $k = \lceil \log(\delta^{-1}) \rceil$, where $\lceil x \rceil$ is the ceiling function which returns the greatest integer larger than or equal to $x$, we get that for all $\delta \in (0, e^{-\sfrac{1}{2}})$
-    \begin{equation}
-        k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\sfrac{1}{2\lceil \log(\delta^{-1}) \rceil}}
-        = \lceil \log(\delta^{-1}) \rceil e^{\sfrac{1}{2}\frac{\log(\delta^{-1})}{\lceil \log(\delta^{-1}) \rceil}}
-        \leq 2 \log(\delta^{-1}) e^{\sfrac{1}{2}},
-    \end{equation}
-    from which follows the second part of the theorem when $n_{\mtx{\Psi}} = (2^5 e^{\sfrac{1}{2}} \varepsilon^{-1} \log(\delta^{-1}))^2$.
-    %Consequently, with probability $1 - \delta$ we have
-    %\begin{equation}
-    %    \int_{a}^{b} \left| \frac{1}{n_{\mtx{\Psi}}} \sum_{i=1}^{n_{\mtx{\Psi}}} \vct{\psi}_i^{\top} \mtx{B}(t) \vct{\psi}_i - \Trace(\mtx{B}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\sfrac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
-    %\end{equation}
-\end{proof}
-
-%\todo{
-%\begin{remark}
-%    If additionally $\mtx{B}(t)$ is nonzero and symmetric positive semi-definite for all $t \in [a, b]$, then it follows from the proof of \refthm{thm:hutchinson} that with probability at least $1 - \delta$
-%    \begin{equation}
-%        \int_{a}^{b} \frac{| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) |}{\Trace(\mtx{B}(t))} ~\mathrm{d}t < \varepsilon,
-%    \end{equation}
-%    if $n_{\mtx{\Psi}} \geq 2^{10} \varepsilon^{-2} \log(\delta^{-1})^2 (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ with $\mu(t) = \lVert \mtx{B}(t) \rVert _2 / \Trace(\mtx{B}(t))$.
-Compared to the constant matrix equivalent \cite[remark 2]{cortinovis-2022-randomized-trace} we mainly observe a larger constant and an additional factor of $\log(\delta^{-1})$.
-%\end{remark}
-%}
-
-Compared to the constant case \cite[lemma 2.1]{meyer-2021-hutch-optimal} we need to choose the number of queries $n_{\mtx{\Psi}}$ by a factor of $\log(\delta^{-1})$ larger to achieve the same guarantee.
-
-The Girard-Hutchinson estimator distinguishes itself for its simplicity.
-However, its bound does not take into account the structure of $\mtx{B}(t)$.
-The $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ requirement implies that if we aim to increase the accuracy of the estimate by one digit, we would need to increase the number of queries by a factor of $100$.
-
-\subsection{Nyström approximation for parameter-dependent matrices}
-\label{subsec:nystrom}
-
-For symmetric positive semi-definite matrices whose singular values decay rapidly, the Nyström estimate \refequ{equ:nystrom-trace-estimator} can be significantly better than the Girard-Hutchinson estimate.
-This is reflected in the following theorem.
-
-\begin{theorem}{Nyström estimator for parameter-dependent matrices}{nystrom}
-    Suppose $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ is continuous and positive semi-definite in $t \in [a, b]$.
-Let $r \geq 2$ and $n_{\mtx{\Omega}} \geq r + 4$ be two integers.
-Then for all $\gamma \geq 1$ with probability at least $1 - \gamma^{-(n_{\mtx{\Omega}} - r)}$
-    \begin{equation}
-        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right| ~\mathrm{d}t
-        \leq \gamma^2 (1 + r) \int_{a}^{b} \sum_{i = r+1}^{n} \sigma_i(\mtx{B}(t)) ~\mathrm{d}t.
-    \end{equation}
-\end{theorem}
-
-\begin{proof}
-    Since $\mtx{B}(t)$ is positive semi-definite for all $t$, then by \cite[lemma 2.1]{frangella-2023-randomized-nystrom}, $\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ is also positive semi-definite for all $t$.
-Therefore, for fixed $t \in [a, b]$
-    \begin{equation}
-        \left| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right|
-        = \left| \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right|
-        = \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _{\ast},
-    \end{equation}
-    where $\lVert \cdot \rVert _{\ast}$ represents the nuclear norm.  From the proof of \cite[corollary 8.2]{tropp-2023-randomized-algorithms} it follows that
-    \begin{equation}
-        %\lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _2^2 \leq \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _F^2
-        \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{\mtx{B}(t)^{\sfrac{1}{2}} \mtx{\Omega}}) \mtx{B}(t)^{\sfrac{1}{2}} \rVert _F^2.
-    \end{equation}
-    Hence, by \cite[theorem 9]{kressner-2023-randomized-lowrank}, with $n_{\mtx{\Omega}} = p + r$, the result follows by noting $\sigma_i(\mtx{B}(t)^{\sfrac{1}{2}})^{2} = \sigma_i(\mtx{B}(t))$.
-\end{proof}
-
-Compared to the constant matrix case \cite[theorem 8.1]{tropp-2023-randomized-algorithms}, our bound also scales with the best rank-$r$ approximation error in the nuclear norm, albeit with a factor approximately proportional to $r$, which cannot be compensated for with large enough oversampling.
-
-The downside of the Nyström estimator is that it will -- in general -- not work well enough on matrices whose singular values do not decay quickly, for example $\mtx{B}(t) \equiv \mtx{I}_n$.
-
-\subsection{Nyström++ estimator for parameter-dependent matrices}
-\label{subsec:nystrom-pp}
-
-Just as in the constant matrix case, our goal is now to prove that the Nyström++ estimator \refequ{equ:nystrompp-trace-estimator} achieves an $\varepsilon$-relative error with just $\mathcal{O}(\varepsilon^{-1})$ random vectors used to construct the estimate.
-To do so, we mimic the proof of a similar bound for the Hutch++ estimator \cite[theorem 3.1]{meyer-2021-hutch-optimal}.
-Correspondingly, we will need to show that with $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2})$ random vectors, the Nyström approximation verifies $\int \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)\rVert _F~\mathrm{d}t \leq \varepsilon \int \Trace(\mtx{B}(t))~\mathrm{d}t$, regardless of the structure of the matrix $\mtx{B}(t)$.
-Just as in the corresponding proof for the Girard-Hutchinson estimator \refthm{thm:hutchinson}, we again bound the higher order moments of the approximation error for fixed $t$ and then use Minkowsi's integral inequality to carry over the results to the continuous case.
-We will do this in a sequence of three lemmas.
-
-For reasons which will become clear in the proof of \reflem{lem:nystrom}, we will need to bound $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]$ for a constant matrix $\mtx{A}$ and a standard Gaussian random matrix $\mtx{\Omega}$.
-The corresponding result for the Frobenius norm $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _F^2 \right]$ has already been shown in \cite{kressner-2023-randomized-lowrank}.
-Several ways of separately bounding the spectral norm moments of the Wishart matrix $\mtx{\Omega} \mtx{\Omega}^{\top}$ already exist, however, most of them are not sufficient for our purposes \cite{chen-2005-condition-numbers, edelman-1988-eigenvalues-condition, james-1964-distributions-matrix}.
-In \cite[lemma B.1]{tropp-2023-randomized-algorithms} a bound is achieved for $p = 2$ and $p = 4$.
-We will now generalize this result for arbitrary $p \in \mathbb{N}$.
-To do so, we first show the bound for Gaussian random vectors and then extend it to Gaussian random matrices in \reflem{lem:spectral-norm-moment}.
-
-\begin{lemma}{Spectral norm moments of non-standard Gaussian random vector}{spectral-norm-moment-vector}
-    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random vector $\vct{\omega} \in \mathbb{R}^{m}$ satisfy for all $k,p \in \mathbb{N}$ 
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2^2 \right]
-        \leq  (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
-    \end{equation}
-    %\begin{equation}
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
-    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
-    %\end{equation}
-    %If we choose 
-\end{lemma}
-
-\begin{remark}
-    In particular, this result can also be used to bound
-    \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2 \right] = \sqrt{\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2^2 \right]} \leq \sqrt{k + p} \cdot \left(\lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F\right)
-    \end{equation}
-    for any $k,p \in \mathbb{N}$.
-\end{remark}
-
-\begin{proof}
-    With the singular value decomposition $\mtx{A} = \mtx{U} \mtx{\Sigma} \mtx{V}^{\top}$ and the unitary invariance of the spectral norm, it can be seen that $\lVert \mtx{A} \mtx{\omega} \rVert _2^2 = \lVert \mtx{\Sigma} \widetilde{\mtx{\omega}} \rVert _2^2$ where $\widetilde{\mtx{\omega}}$ is again a standard Gaussian random vector.
-Hence, it is enough to bound $\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{\Sigma} \mtx{\omega} \rVert _2^2 \right]$ for a diagonal matrix $\mtx{\Sigma} = \operatorname{diag}(\sigma_1, \dots, \sigma_m)$ with $\sigma_1 \geq \dots \geq \sigma_m \geq 0$.
-
-    First, we can rewrite
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2^2 \right]
-        = \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right].
-        \label{equ:matvec-spectral-norm-simplification}
-    \end{equation}
-    Inspired by the proof of \cite[theorem 1]{cohen-2016-optimal-approximate}, we split the diagonal entries of $\mtx{\Sigma}$ into $\ell = \lceil m/k \rceil$ groups of size $k \geq 1$
-    \begin{equation}
-        \overbrace{\underbrace{\sigma_1, \dots, \sigma_k}_{\leq \sigma_1}}^{\geq \sigma_{k+1}}, \overbrace{\underbrace{\sigma_{k+1}, \dots, \sigma_{2k}}_{\leq \sigma_{k+1}}}^{\geq \sigma_{2k+1}}, \dots, \overbrace{\underbrace{\sigma_{1+(\ell - 1)k}, \dots, \sigma_{\ell k}}_{\leq \sigma_{1+(\ell - 1)k}}}^{\geq 0}.
-    \end{equation}
-    If $m$ is not a multiple of $k$, we let $\sigma_i = 0$ for $i > m$. Since $\sigma_1 \geq \dots \geq \sigma_{\ell k} \geq 0$,
-    \begin{equation}
-        \sigma_1^2 = \lVert \mtx{\Sigma} \rVert _2^2
-        \quad \text{and} \quad
-        \sigma_{1 + (i-1)k}^2 \leq \frac{1}{k} \sum_{j=1}^{k} \sigma_{j + (i-2)k}^2, i = 2, \dots, \ell,
-    \end{equation}
-    and therefore
-    \begin{equation}
-        \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \sum_{j=1}^{(\ell - 1)k} \sigma_j^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2.
-        \label{equ:singular-value-group-bound}
-    \end{equation}
-    This allows us to bound
-    \begin{align}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right]
-        &= \mathbb{E}^{\sfrac{p}{2}}\Bigg[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{j + (i-1)k}^2}_{\leq \sigma_{1 + (i-1)k}^2} \omega_{j + (i-1)k}^2 \Bigg] && \text{(separate sum into groups)} \notag \\
-        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{($\sigma_{j + (i-1)k}^2 \leq \sigma_{1 + (i-1)k}^2$, $\forall i, j$)} \notag \\
-        &\leq \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 ~ \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{(triangle inequality)} \notag \\
-        %&\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k}{2} + p)}{\Gamma(\frac{k}{2})} \right)^{\sfrac{1}{p}} && \text{(\refequ{equ:singular-value-group-bound} and \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
-        &\leq (k + p) \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 && \text{(\reflem{lem:gamma})} \notag \\
-        &\leq (k + p) \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right). && \text{(using \refequ{equ:singular-value-group-bound})}
-    \end{align}
-
-    %The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
-    %\begin{align}
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    &= \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}} \notag \\
-    %    &\leq \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{1}{\Gamma(k - r + 2)} \right)^{\frac{1}{k - r + 1}} \left( \frac{k + r}{2} \right) \notag \\
-    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right)
-    %\end{align}
-    %With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\sfrac{1}{n}} \leq \frac{e}{n}$ follow respectively.
-Hence,
-    %\begin{equation}
-    %    \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    \leq \frac{e^2}{k + 1}\sqrt{ \frac{3k}{2} }
-    %    \leq e^2 \sqrt{\frac{3}{2k}}.
-    %    \label{equ:OSE-moment-bound-second}
-    %\end{equation}
-
-\end{proof}
-
-\begin{lemma}{Spectral norm moments of non-standard Gaussian random matrix}{spectral-norm-moment}
-    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all $p \in \mathbb{N}$ 
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
-        \leq  (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
-    \end{equation}
-    %\begin{equation}
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
-    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
-    %\end{equation}
-    %If we choose 
-\end{lemma}
-
-\begin{remark}
-    In particular, this result can also be used to bound
-    \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2 \right] = \sqrt{\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]} \leq \sqrt{k + p} \cdot \left(\sqrt{3} \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F\right)
-    \end{equation}
-    for any $k,p \in \mathbb{N}$. Indeed, for $p=2$ this bound is slightly less tight than its equivalent in \cite[lemma B.1]{tropp-2023-randomized-algorithms}.
-\end{remark}
-
-\begin{proof}
-    From the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms} we know that
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
-        \leq \lVert \mtx{A} \rVert _2^2 \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \vct{\omega}_1 \rVert _2^2 \right] + \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega}_2 \rVert _2^2 \right]
-    \end{equation}
-    for independent standard Gaussian random vectors $\vct{\omega}_1 \in \mathbb{R}^{k}$ and $\vct{\omega}_2 \in \mathbb{R}^{m}$.
-Both moments can be bound using \reflem{lem:spectral-norm-moment-vector}, one of which with $\mtx{A} = \mtx{I}_k$, to get
-    \begin{align}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
-        &\leq \lVert \mtx{A} \rVert _2^2 (k + p) \left( \lVert \mtx{I}_k \rVert _2^2 + \frac{1}{k} \lVert \mtx{I}_k \rVert _F^2 \right) + (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right) \notag \\
-        &\leq (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
-    \end{align}
-
-    %The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
-    %\begin{align}
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    &= \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}} \notag \\
-    %    &\leq \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{1}{\Gamma(k - r + 2)} \right)^{\frac{1}{k - r + 1}} \left( \frac{k + r}{2} \right) \notag \\
-    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right)
-    %\end{align}
-    %With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\sfrac{1}{n}} \leq \frac{e}{n}$ follow respectively.
-Hence,
-    %\begin{equation}
-    %    \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    \leq \frac{e^2}{k + 1}\sqrt{ \frac{3k}{2} }
-    %    \leq e^2 \sqrt{\frac{3}{2k}}.
-    %    \label{equ:OSE-moment-bound-second}
-    %\end{equation}
-
-\end{proof}
-
-\begin{lemma}{Nyström approximation for parameter-dependent matrices}{nystrom}
-    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-Then its Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ with Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ for \emph{even} $n_{\mtx{\Omega}} \geq 4$ verifies with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
-    \begin{equation}
-        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t
-        \label{qu:nystrompp-theorem-bound}
-    \end{equation}
-    for any $\gamma \geq 1$ and a universal constant $c > 0$.
-In particular, if we choose $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1}))$, then with probability at least $1-\delta$ for $\delta \in (0, 1)$ and $\varepsilon \in (0, 1)$ we have $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$.
-\end{lemma}
-
-%\todo{Proof idea: structural bound, then higher order moment bound to apply Markov's inequality.}
-
-\begin{proof}
-    First, we temporarily omit the parameter dependence.
-Let $\mtx{B} = \mtx{U} \mtx{\Lambda} \mtx{U}^{\top}$ be the eigenvalue decomposition of a symmetric positive semi-definite matrix $\mtx{B}$, i.e.
-$\mtx{\Lambda} = \operatorname{diag}(\lambda_1, \dots, \lambda_n)$ and $\mtx{U}^{\top} \mtx{U}= \mtx{I}_n$.
-We define the matrix partitions
-    \begin{equation}
-        \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+4ex+\baselineskip}
-        \mtx{U} = \begin{bmatrix}
-            \smash{\underbrace{\mtx{U}_1}_{n \times k}} & \smash{\underbrace{\mtx{U}_2}_{n \times (n-k)}}
-        \end{bmatrix}
-        \quad \text{and} \quad
-        \mtx{\Lambda} =
-        \begin{bmatrix}
-            \smash{\overbrace{\mtx{\Lambda}_1}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2}_{(n-k) \times (n-k)}}
-        \end{bmatrix}.
-    \end{equation}
-    Further, for a standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ we define $\mtx{\Omega}_1 = \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times n_{\mtx{\Omega}}}$ and $\mtx{\Omega}_2 = \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times n_{\mtx{\Omega}}}$, which are likewise standard Gaussian.
-
-    We can apply \cite[theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound
-    \begin{equation}
-        \lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F 
-        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2,
-        \label{equ:nystrom-proof-persson-bonud}
-    \end{equation}
-    where $\lVert \cdot \rVert _{(4)}$ denotes the Schatten-4 norm.
-
-    As in the proof of \cite[lemma 3]{meyer-2021-hutch-optimal}, the first term -- and its moments -- are limited by
-    \begin{equation}
-        \lVert \mtx{\Lambda}_2 \rVert _F
-        = \sqrt{\sum_{i=k+1}^{n} \lambda_i^2}
-        \leq \sqrt{ \lambda_{k+1} \sum_{i=k+1}^{n} \lambda_i}
-        \leq \sqrt{ \frac{\Trace(\mtx{B})}{k} \Trace(\mtx{B})}
-        \leq \frac{1}{\sqrt{k}} \Trace(\mtx{B}).
-        \label{equ:nystrom-proof-frobenius-trace}
-    \end{equation}
-    The $p$-th moment of the second term can be processed with standard matrix norm manipulations to
-    \begin{align}
-        \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        %&= \mathbb{E}^{p}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )( \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger})^{\top} \rVert _F \right] && \text{($\lVert \mtx{A} \rVert _{(4)}^2 = \lVert \mtx{A} \mtx{A}^{\top} \rVert _{F}$)} \notag \\
-        %&\leq \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] && \text{($\lVert \mtx{A} \mtx{B} \rVert _F \leq \lVert \mtx{A} \rVert _F \lVert \mtx{B} \rVert _2$)} \notag \\
-        &\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{($\lVert \mtx{A} \rVert _{(4)}^2 \leq \sqrt{k} \lVert \mtx{A} \rVert _2^2$)} \notag \\
-        &\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{(submultiplicativity)} \notag \\
-        &\leq \sqrt{k} \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2\right] \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2  \right].
-&& \text{(independence)}
-        \label{equ:nystrom-proof-processed-tail}
-    \end{align}
-    To match the decay rate of the moments of the first term in \refequ{equ:nystrom-proof-persson-bonud}, we will need to ensure that this term also is of order $\mathcal{O}(\Trace{\mtx{B}} / \sqrt{k})$.
-To this purpose, and to simplify the notation, we choose $k = n_{\mtx{\Omega}}/2$ and $p = n_{\mtx{\Omega}}/4$ for even $n_{\mtx{\Omega}}$ to ensure $k \in \mathbb{N}$.
-We can apply \reflem{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$ to bound
-    \begin{equation}
-        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \right]
-        \leq \frac{3 n_{\mtx{\Omega}}}{2} \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
-        \label{equ:spectral-norm-bound-applied}
-    \end{equation}
-    The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
-    \begin{equation}
-        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
-        = \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{n_{\mtx{\Omega}}}{4}} \right]^{\sfrac{4}{n_{\mtx{\Omega}}}}%  \notag \\
-        \leq \left( 1 + \frac{n_{\mtx{\Omega}}}{2} \right)^{\sfrac{4}{n_{\mtx{\Omega}}}} \left( \frac{1}{(\frac{n_{\mtx{\Omega}}}{2} + 1)!}\right)^{\frac{2}{\frac{n_{\mtx{\Omega}}}{2} + 1}} \left( \frac{3 n_{\mtx{\Omega}}}{4}\right).
-    \end{equation}
-    From the expansion of the exponential function as a power series $e^n = 1 + n + \dots + n^n/n!
-+ \dots$ for $n \in \mathbb{N}$ we see $e^n \geq 1 + n$ and $e^n \geq n^n / n!$, from which $(1 + n)^{\sfrac{1}{n}} \leq e$ and $\left( 1/n!
-\right)^{\sfrac{1}{n}} \leq e/n$ follow respectively.
-Hence,
-    \begin{equation}
-        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
-        \leq e^2 \frac{e^2}{(\frac{n_{\mtx{\Omega}}}{2}+ 1)^2} \left( \frac{3 n_{\mtx{\Omega}}}{4}\right)
-        \leq \frac{3 e^4}{n_{\mtx{\Omega}}}.
-        \label{equ:pinv-spectral-norm-bound}
-    \end{equation}
-    As opposed to \cite[lemma B.3]{tropp-2023-randomized-algorithms}, our bound holds for all even $n_{\mtx{\Omega}} \in \mathbb{N}$, at the cost of a factor of $e^2$.
-Inserting \refequ{equ:spectral-norm-bound-applied} and  \refequ{equ:pinv-spectral-norm-bound} in \refequ{equ:nystrom-proof-processed-tail} gives
-    \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \frac{9 e^4}{2 \sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
-        %\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
-    \end{equation}
-    We can identify $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 = \lambda_{k+1}$ and $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 = \Trace(\mtx{\Lambda}_2)$, and just as in the proof of \cite[lemma 3.1]{meyer-2021-hutch-optimal} we bound $\lambda_{k+1} \leq \Trace(\mtx{B})/k$ and $\Trace(\mtx{\Lambda}_2) \leq \Trace(\mtx{B})$ -- remembering $k = n_{\mtx{\Omega}}/2$ -- to get
-    \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \frac{9 e^4}{2 \sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \left( \frac{6}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) + \frac{1}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) \right)
-        \leq \frac{9e^4}{2 \sqrt{2}} \frac{7}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
-        \label{equ:nystrom-proof-tail-bound}
-        %\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
-    \end{equation}
-    Inserting \refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud} and using the triangle inequality for $\mathbb{E}^{p}[\cdot]$ we obtain
-    \begin{equation}
-        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right]
-        \leq \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[ \lVert \mtx{\Lambda}_2 \rVert _F \right] + \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
-    \end{equation}
-    where $c = 1 + 63 e^4 / 2 \sqrt{2}$.
-
-    As in \cite{kressner-2023-randomized-lowrank}, we can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable and by the Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities}, we get for $n_{\mtx{\Omega}} \geq 4$
-    \begin{align}
-        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \right] 
-        &\leq \int_{a}^{b} \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F \right] ~\mathrm{d}t \notag \\
-        %&\leq \mathbb{E}^{\sfrac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
-        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \mathbb{E}^{\sfrac{k}{2}} \left[ \int_{a}^{b}  \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
-        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
-        &\leq \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
-    \end{align}
-    From Markov's inequality follows with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
-    \begin{equation}
-        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t,
-    \end{equation}
-    from which follows the first statement.
-
-    Fixing $\gamma = e$ and letting $n_{\mtx{\Omega}} = (c e \varepsilon)^{-2}  + 4 \log(\delta^{-1})$ gives us the second part of the theorem.
-
-    %Setting $\delta = \gamma^{-\sfrac{k}{4}}$ and choosing $k = \lceil \varepsilon^{-2}\log(\delta^{-1}) \rceil$ we get
-    %\begin{align}
-    %    \gamma \frac{1}{\sqrt{k}}
-    %    &= \delta^{-\sfrac{2}{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} \frac{1}{\sqrt{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} && \text{(definition of $\delta$ and choice of $k$)} \notag \\
-    %    &= e^{2\varepsilon^{2} \frac{\varepsilon^{-2}\log(\delta^{-1})}{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} \frac{1}{\sqrt{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} && \text{($\delta = e^{-\log(\delta^{-1})}$)} \notag \\
-    %    &\leq e^{2\varepsilon^{2}} \frac{\varepsilon}{\sqrt{ \log(\delta^{-1})}} && \text{($\lceil x \rceil \geq x$ if $x \geq 0$)}
-    %\end{align}
-    %which is smaller than $\varepsilon$ if $\delta \leq e^{-e^{4 \varepsilon^2}}$from which follows the second part of the theorem when using $n_{\mtx{\Omega}} = 2k$.
-\end{proof}
-
-When compared to the equivalent result for constant matrices \cite[Lemma 3.2]{persson-2022-improved-variants}, our constant factor up front is around four times larger, and our failure probability decays half a power slower.
-
-Finally, we can combine the bound on the Girard-Hutchinson estimator \refthm{thm:hutchinson} and the one on the Nyström approximation \reflem{lem:nystrom} to obtain our main result.
-
-\begin{theorem}{Nyström++ trace estimator for parameter-dependent matrices}{nystrom-pp}
-    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(\delta^{-1})^2)$ with \emph{even} $n_{\mtx{\Omega}}$, then for $\delta \in (0, e^{-\sfrac{1}{2}}/2)$ and $\varepsilon \in (0, 1)$ with probability at least $1 - \delta$
-    \begin{equation}
-        \int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t
-        \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
-    \end{equation}
-\end{theorem}
-
-\begin{proof}
-    By choosing $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\tilde{\varepsilon}^{-2} \log(\tilde{\delta}^{-1})^2)$ we get
-    \begin{align}
-        &\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t \notag \\
-        &= \int_{a}^{b} | \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) - \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) |~\mathrm{d}t && \text{(by definition of estimators)} \notag \\
-        &\leq \tilde{\varepsilon} \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F ~\mathrm{d}t && \text{(\refthm{thm:hutchinson} w.p. $\geq 1 - \tilde{\delta}$)} \notag \\
-        &\leq \tilde{\varepsilon}^2 \int_{a}^{b} \Trace(\mtx{B}(t)) ~\mathrm{d}t && \text{(\reflem{lem:nystrom} w.p. $\geq 1 - \tilde{\delta}$)} 
-    \end{align}
-    with probability at least $1 - 2\tilde{\delta}$. Taking $\varepsilon = \tilde{\varepsilon}^2$ and $\delta = 2 \tilde{\delta}$
-\end{proof}
-
-Comparing \refthm{thm:nystrom-pp} with its equivalent for constant matrices \cite[theorem 3.4]{persson-2022-improved-variants}, we notice a less favorable scaling of the required number of random vectors $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$ in terms of $\log(\delta^{-1})$.
-
-% \begin{proof}
-%     Combining with union bound.
-%     \begin{align}
-%         &\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t \notag \\
-%         &= \int_{a}^{b} | \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)) - \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)) |~\mathrm{d}t && \text{(definition of estimators)} \notag \\
-%         &= \todo{c} \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F ~\mathrm{d}t && \text{(with probability $1 - \delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1})$)} \notag \\
-%         &= \todo{c} \frac{1}{\sqrt{n_{\mtx{\Psi}} n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t)) ~\mathrm{d}t
-%     \end{align}
-% \end{proof}
-
-\section{Application to spectral density approximation}
-\label{sec:application}
-
-We consider the task of computing the spectral density of a real symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ with eigenvalues $\lambda_1, \dots, \lambda_n$, which is defined as
-\begin{equation}
-    \phi(t) = \frac{1}{n} \sum_{i=1}^{n} \delta(t - \lambda_i),
-    \label{equ:spectral-density}
-\end{equation}
-where $\delta$ is the Dirac delta distribution.
-Finding approximations to this distribution is challenging.
-Furthermore, in most applications the exact location of every eigenvalue is not important, but more so their approximate locations relative to each other, such as eigenvalue clusters, spectral gaps, or outliers.
-In the following we will introduce a smoothed version of the spectral density, which is significantly easier to approximate at the cost of losing some of the finer characteristics of the spectrum.
-
-\subsection{Smoothed spectral density}
-\label{subsec:spectral-density}
-
-The smoothed spectral density $\phi_{\sigma}$ is defined as
-\begin{equation}
-    \phi_{\sigma}(t) = \frac{1}{n} \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i) = \frac{1}{n} \Trace(g_{\sigma}(t\mtx{I}_n - \mtx{A}))
-    \label{equ:smooth-spectral-density}
-\end{equation}
-for a smoothing kernel $g_{\sigma}$ parametrized by a smoothing parameter $\sigma > 0$, which controls by how much $\phi$ is smoothed.
-We choose a Gaussian smoothing kernel of width $\sigma > 0$, given by
-\begin{equation}
-    g_{\sigma}(s) = \frac{1}{\sigma \sqrt{2\pi}} e^{-\sfrac{s^2}{2\sigma^2}}.
-    \label{equ:smoothing-kernel}
-\end{equation}
-The task in \refequ{equ:smooth-spectral-density} is to approximate the trace of the parameter-dependent matrix $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
-Hence, we can apply the results derived in \refsec{sec:analysis}.
-
-Choosing the smoothing parameter $\sigma$ results in the following trade off: Typically, larger $\sigma$ make it easier to approximate the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$, whereas smaller $\sigma$ allow us to stay closer to the original spectral density \refequ{equ:spectral-density}.
-To see the latter, we measure the error between the spectral density $\phi$ and its smoothened version $\phi_{\sigma}$ with
-\begin{equation}
-    \sup_{f \in \mathcal{S}} \int_{-1}^{1} f(t) (\phi(t) - \phi_{\sigma}(t))~\mathrm{d}t,
-    \label{equ:error-metric}
-\end{equation}
-where $\mathcal{S}$ represents an appropriately chosen space of test functions.
-Among others, \cite{lin-2016-approximating-spectral} proposes $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - \lambda), \lambda \in [\lambda_{\text{min}}, \lambda_{\text{max}}]\}$, \cite{chen-2021-analysis-stochastic} uses $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [-1, 1] \}$ where $\Theta$ is the Heaviside step function, and \cite{braverman-2022-sublinear-time} uses $\mathcal{S} = \{f : |f(x) - f(y)| \leq |x - y| \}$ for which the error \refequ{equ:error-metric} is equivalent to the Wasserstein-1 distance between $\phi$ and $\phi_{\sigma}$.
-In this last metric, provided we assume the spectrum to be somewhat uniform and $\sigma$ to be rather small, a simple calculation allows us to roughly estimate the smoothing error \refequ{equ:error-metric} to be $\sigma$.
-Consequently, if we can only allow an approximation to deviate by at most a factor of $\varepsilon > 0$ from the original spectral density, we should choose a smoothing parameter $\sigma \approx \varepsilon$.
-
-\subsection{Chebyshev expansion of smoothing kernel}
-\label{subsec:chebyshev-expansion}
-
-To estimate the trace in \refequ{equ:smooth-spectral-density}, we first need to evaluate the involved matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
-However, doing this exactly would require us to first diagonalize $\mtx{A}$, which is often a prohibitively expensive operation.
-For algorithmic reasons, we seek for approximations of this matrix function in terms of affine expansions, of which the Chebyshev expansion is most suitable due to its approximation error guarantees and ease of manipulation.
-In every $t \in \mathbb{R}$ we expand
-\begin{equation}
-    g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) = \sum_{l=0}^{m} \mu_l(t) T_l(\mtx{A})
-    \label{equ:matrix-expansion}
-\end{equation}
-in terms of a set of $m+1$ coefficients $\mu_0(t), \dots, \mu_m(t)$ which linearly combine the Chebyshev polynomials
-\begin{equation}
-    \begin{cases}
-        T_l : [-1, 1] \to \mathbb{R} \\
-        T_l(s) = \cos(l \cdot \arccos(s))
-    \end{cases}
-    \label{equ:chebyshev-polynomial}
-\end{equation}
-evaluated at the matrix $\mtx{A}$. It is easy to verify that these polynomials satisfy the three-term recurrence relation
-\begin{equation}
-    T_l(s) =
-    \begin{cases}
-        s^l, & \text{if $l \in \{0, 1\}$}; \\
-        T_l(s) = 2 s T_{l-1}(s) - T_{l-2}(s), & \text{if $l \geq 2$.}
-    \end{cases}
-    \label{equ:chebyshev-recurrence}
-\end{equation}
-With the help of the Chebyshev expansion \refequ{equ:matrix-expansion} we can now define the expanded spectral density
-\begin{equation}
-    \phi_{\sigma}^{(m)}(t) =  \frac{1}{n} \Trace(g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})).
-    \label{equ:expanded-spectral-density}
-\end{equation}
-
-However, the Chebyshev polynomials are only well-defined for matrices whose spectrum is contained in $[-1, 1]$.
-When working with general matrices of bounded spectrum, we will first have to estimate a lower bound $a$ and upper bound $b$ on the spectrum \cite{zhou-2011-bounding-spectrum}.
-A spectral transform
-\begin{equation}
-    \begin{cases}
-        \tau : [a, b] \to [-1, 1]\\
-        \tau(t) = \frac{2t - a - b}{b - a}
-    \end{cases}
-\end{equation}
-then lets us define the matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ whose spectrum is identical to the one of $\mtx{A}$ but shrunk to the interval $[-1, 1]$, for which the expansion \refequ{equ:matrix-expansion} is valid.
-Note that through this transformation we will also have to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when reverting the spectral transform $\tau$.
-
-\subsubsection{Fast and consistent squaring}
-\label{subsubsec:dct}
-
-It is well known that the coefficients of a Chebyshev expansion of a function can -- in exact arithmetic -- be computed with a discrete cosine transform (DCT) of the function evaluations at the Chebyshev nodes $s_i = \cos(\pi i / m)$ for $i=0,\dots,m$
-\begin{equation}
-    g_{\sigma}(t - \cos(s_i)) = \sum_{l=0}^{m} \mu_l(t) \cos\left(\frac{\pi i l}{m} \right).
-    \label{equ:discrete-cosine-transform}
-\end{equation}
-This can be accomplished in $\mathcal{O}(m \log(m))$ operations.
-\cite[algorithm 1]{lin-2017-randomized-estimation} uses a slightly different approach, which only approximates the coefficients and is noticeably slower (cf.
-\reftab{tab:chebyshev-timing-interpolation}).
-
-A highly useful property of the Chebyshev expansion -- of which we will make use twice in the following pages -- is the fact that we can square such an expansion in $\mathcal{O}(m \log(m))$ operations \cite{baszenski-1997-fast-polynomial}, which will be used in the below algorithm.
-
-\begin{algo}{Fast squaring of Chebyshev expansion}{chebyshev-squaring}
-    \begin{algorithmic}[1]
-        \Statex \textbf{Input:} Coefficients $\{ \mu_l \}_{l=0}^{m}$ of the expansion $\sum_{l=0}^{m} \mu_l T_l$
-        \Statex \textbf{Output:} Coefficients $\{ \nu_l \}_{l=0}^{2m}$ such that $\sum_{l=0}^{2m} \nu_l T_l = (\sum_{l=0}^{m} \mu_l T_l)^2$
-        \State Define $\vct{\mu} \in \mathbb{R}^{2m + 1}$ whose first $m + 1$ entries are the coefficients $\mu_l$ and the remaining $m$ are $0$
-    \State Compute $\vct{f} = \DCT(\vct{\mu})$
-    \State Compute $\vct{\nu} = \DCT^{-1}(\vct{f}^{2})$ whose entries are the coefficients $\nu_l$
-    \end{algorithmic}
-    \end{algo}
-
-\subsubsection{Expansion error}
-\label{subsubsec:expansion-error}
-
-We now analyze the error of the Chebyshev expansion of a Gaussian smoothing kernel $g_{\sigma}$.
-
-\begin{lemma}{Chebyshev expansion error}{chebyshev-error}
-    The expansion $g_{\sigma}^{(m)}$ of the Gaussian smoothing kernel $g_{{\sigma}}$ \refequ{equ:smoothing-kernel} satisfies
-    \begin{equation}
-        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \right| \leq \sqrt{\frac{2e}{\pi}} \frac{1}{\sigma^2} (1 + \sigma)^{-m} \equiv E_{\sigma, m}
-        \label{equ:chebyshev-interpolation-sup-error-kernel}
-        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
-        %\label{equ:2-chebyshev-interpolation-error}
-    \end{equation}
-    for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and $s \in [-1, 1]$.
-\end{lemma}
-
-The proof specializes Bernstein's theorem for Chebyshev interpolation to the case of a Gaussian smoothing kernel \refequ{equ:smoothing-kernel}.
-It loosely follows the proof give in the preprint of \cite[theorem 2]{lin-2017-randomized-estimation}, but attains a slightly different dependence on $\sigma$.
-
-\begin{proof}
-    From Bernstein's theorem \cite[theorem 4.3]{trefethen-2008-gauss-quadrature} it follows that if $f$ is analytic within the Bernstein ellipse $\mathcal{E}_{\chi} \subset \mathbb{C}$ with foci $\{-1, +1\}$ and sum of semi-axes $\chi > 1$, then for any $m \in \mathbb{N}$
-    \begin{equation}
-        \sup_{s \in [-1, 1]} \left| f(s) - f^{(m)}(s) \right| \leq \frac{2}{\chi^m (\chi - 1)} \max_{z \in \mathcal{E}_{\chi}} |f(z)|.
-        \label{equ:bernstein-bound}
-    \end{equation}
-    In particular, when $f = g_{\sigma}(t - \cdot)$ we observe for $z = x + iy \in \mathcal{E}_{\chi}$
-    \begin{equation}
-    | g_{\sigma}(t - z) | 
-    = \frac{1}{\sigma \sqrt{2 \pi}} \left| e^{-\sfrac{(t - z)^2}{2\sigma^2}} \right|
-    = \frac{1}{\sigma \sqrt{2 \pi}} e^{-\sfrac{(t - x)^2}{2\sigma^2}}e^{\sfrac{y^2}{2\sigma^2}}.
-    %\leq \frac{1}{n \sigma \sqrt{2 \pi}} \max_{x + iy \in \mathcal{E}_{\chi}} e^{\sfrac{y^2}{2\sigma^2}} 
-    \end{equation}
-    Because $e^{-\sfrac{(t - x)^2}{2\sigma^2}} \leq 1$ for all $x, t \in \mathbb{R}$, and the maximum absolute value of $y$ is limited by the length of the semi-axis of the Bernstein ellipse in the direction of the imaginary axis, which is $(\chi - \chi^{-1}) / 2$, we upper bound
-    \begin{equation}
-        \max_{z \in \mathcal{E}_{\chi}} | g_{\sigma}(t - z) | 
-        \leq \frac{1}{\sigma \sqrt{2 \pi}} e^{\sfrac{(\chi - \chi^{-1})^2}{8 \sigma^2}}.
-    \end{equation}
-
-    Choosing $\chi = 1 + \sigma$ gives a simple and good bound, since in this case $\chi - \chi^{-1} \leq 2\sigma$, which implies
-    \begin{equation}
-        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \right| \leq \sqrt{\frac{2e}{\pi}} \frac{1}{\sigma^2} (1 + \sigma)^{-m}.
-    \end{equation}
-\end{proof}
-
-\Reflem{lem:chebyshev-error} shows that when $0 < \sigma \ll 1$ and if we choose $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, the error we induce with the Chebyshev expansion is at most $\mathcal{O}(\varepsilon)$.
-It now also allows us to bound the error we make when expanding the spectral density.
-
-\begin{theorem}{$L^1$-error of Chebyshev expansion of spectral density}{chebyshev-error}
-    Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the spectal density with Chebyshev expansion $\phi_{\sigma}^{(m)}$ satisfies
-    \begin{equation}
-        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq 2 E_{\sigma, m}
-        \label{equ:chebyshev-interpolation-spectral-density}
-        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
-        %\label{equ:2-chebyshev-interpolation-error}
-    \end{equation}
-    for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and the error $E_{\sigma, m}$ from \reflem{lem:chebyshev-error}.%, and $t \in \mathbb{R}$.
-%The same result holds for the expansion of a non-negative Chebyshev expansion of \emph{even} degree $m \in \mathbb{N}$ but with $\underline{E}_{\sigma, m}$.
-\end{theorem}
-
-\begin{proof}
-    We observe that for all $t \in [-1, 1]$
-    \begin{align}
-        &\left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right| \notag \\
-        &= \left| \frac{1}{n} \Trace(g_{\sigma}(t\mtx{I}_n - \mtx{A})) - \frac{1}{n} \Trace(g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})) \right|
-        && \text{(definitions \refequ{equ:smooth-spectral-density} and \refequ{equ:expanded-spectral-density})} \notag \\
-        &= \left| \frac{1}{n} \sum_{i=1}^n \left(g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i)\right) \right|
-        && \text{($\lambda_1, \dots, \lambda_n$ eigenvalues of $\mtx{A}$)} \notag \\
-        &\leq \max_{i = 1, \dots, n} \left| g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i) \right|
-        && \text{(conservative upper bound)} \notag \\
-        &\leq E_{\sigma, m}.% ~ \text{(or $\underline{E}_{\sigma, m}$)}.
-        && \text{(using \reflem{lem:chebyshev-error})} \notag \\
-    \end{align}
-    Then, Hölder's inequality allows us to conclude
-    \begin{equation}
-        \int_{-1}^{1} | \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) | ~\mathrm{d}t
-        \leq 2 \sup_{t \in [-1, 1]} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|
-        \leq 2 E_{\sigma, m}.% ~ \text{(or $2 \underline{E}_{\sigma, m}$)}.
-    \end{equation}
-\end{proof}
-
-\subsection{Chebyshev-Nyström++ method for spectral density approximation}
-\label{subsec:chebyshev-nystrom}
-
-Our goal is to apply the Nyström++ trace estimator which we analyzed in \refsec{subsec:nystrom-pp} to the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ to approximate the smoothed spectral density $\phi_{\sigma}$.
-However, due to the oscillatory nature of the Chebyshev expansion, we can no longer guarantee that $g_{\sigma}^{(m)} \geq 0$, and consequently, the matrix function $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ may be (slightly) indefinite.
-While in our experiments this never seemed to be an issue, it disallows us to directly use \refthm{thm:nystrom-pp}.
-While variants of the Nyström approximation designed for indefinite matrices have been analyzed, no practical algorithms for computing them exist \cite{nakatsukasa-2023-randomized-lowrank}.
-
-To still apply the bound from \refthm{thm:nystrom-pp} in this scenario, we will have to ensure that the Chebyshev expansion is non-negative.
-We can think of many ways in which this can be achieved: by using Jackson damping \cite{jackson-1912-approximation-trigonometric,braverman-2022-sublinear-time}, albeit at a significantly worse rate of convergence; by shifting up the expansion $g_{\sigma}^{(m)} + \rho$ with a large enough $\rho > 0$, at the loss of most of the low-rank structure of $g_{\sigma}(t \mtx{I}_n - \mtx{A})$; or by directly minimizing the approximation error over all non-negative Chebyshev polynomials \cite{fejer-1916-uber-trigonometrische}, which realistically only works for small $m$.
-In the end, we propose the following method to produce a non-negative Chebyshev expansion: First, expand the square root of the smoothing kernel in a basis of Chebyshev polynomials up to degree $m/2$
-\begin{equation}
-    \sqrt{g_{\sigma}}^{(\sfrac{m}{2})}(t - s) = \sum_{l=0}^{m/2} \xi_l(t) T_l(s).
-\end{equation}
-Second, square this expansion with \refalg{alg:chebyshev-squaring} to obtain the non-negative Chebyshev expansion
-\begin{equation}
-    \underline{g}^{(m)}_{\sigma}(t - s) = \sum_{l=0}^{m} \mu_l(t) T_l(s).
-    \label{equ:non-negative-chebyshev-expansion}
-\end{equation}
-With a complexity of $\mathcal{O}(m \log(m))$ this procedure is computationally inexpensive and allows for very similar convergence bounds to \reflem{lem:chebyshev-error}, which we will derive in \refsec{subsubsec:chebyshev-nystrom-analysis}.
-In \reftab{tab:chebyshev-timing-interpolation}, the non-negative expansion can be seen to only be slightly slower than the normal one.
-
-\begin{table}[ht]
-    \caption{Runtime in milliseconds for computing the coefficients of the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ \refequ{equ:smoothing-kernel} with $\sigma=0.05$, at $n_t=1000$ points, and for various degrees $m$.
-We average over 7 runs of the algorithms and repeat these runs 1000 times to form the mean and standard deviation which are given in the below table.}
-    \label{tab:chebyshev-timing-interpolation}
-   \input{tables/interpolation.tex}
-\end{table}
-
-%\todo{
-%Maybe mention indefinite Nyström [Nakatsukasa], which won't give us desirable guarantees, and refinement of Chebyshev expansion [Francis Bach], which only slightly improves, but is much harder to implement.
-Side-note: it is possible to refine the coefficients of the Chebyshev expansion to get a better approximation in the supremum-norm over the set of all non-negative polynomials \cite%{fejer-1916-uber-trigonometrische}.
-The non-convex nature of this problem and the relatively small improvement we get with this technique made us decide to not pursue this further.
-%}
-
-In the end, we propose the following approximation for the smoothed spectral density
-\begin{equation}
-    \widetilde{\phi}_{\sigma}^{(m)}(t) = \frac{1}{n} \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}))
-    \label{equ:chebyshev-nystrom-formula}
-\end{equation}
-which is the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$ \refequ{equ:nystrompp-trace-estimator} applied to the non-negative Chebyshev expansion $\underline{g}_{\sigma}^{(m)}$ \refequ{equ:non-negative-chebyshev-expansion}.
-
-\subsubsection{Implementation}
-\label{subsubsec:chebyshev-nystrom-implementation}
-
-An efficient implementation of the approximation \refequ{equ:chebyshev-nystrom-formula} can be achieved thanks to the invariance of the trace under cyclic permutation of its arguments and the recurrence relation \refequ{equ:chebyshev-recurrence} which the Chebyshev polynomials satisfy.
-To illustrate this, consider the approximation \refequ{equ:chebyshev-nystrom-formula} when $\mtx{\Psi}$ is zero.
-Using the definition of the Nyström approximation \refequ{equ:nystrom-approximation}, the cyclic invariance of the trace, and the symmetry of $\mtx{A}$, we can rewrite it as
-\begin{equation}
-    \frac{1}{n} \Trace((\mtx{\Omega}^{\top} \underline{g}^{(m)}_{\sigma}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega})^{\dagger} (\mtx{\Omega}^{\top} \underline{g}^{(m)}_{\sigma}(t \mtx{I}_n - \mtx{A})^2 \mtx{\Omega})).
-    \label{equ:cyclic-property}
-\end{equation}
-Thanks to the affine form of the Chebyshev expansion, the first parenthesized term in the trace is
-\begin{equation}
-    \mtx{\Omega}^{\top} \underline{g}^{(m)}_{\sigma}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega} = \sum_{l=0}^{m} \mu_l(t) (\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega})
-\end{equation}
-for which we can first precompute the small $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ for $l=0, \dots, m$ using the recurrence \refequ{equ:chebyshev-recurrence} and then inexpensively sum them up with the corresponding coefficients $\mu_l(t)$ for all $t$ in which we want to evaluate the approximation.
-Thanks to the fast and exact squaring of Chebyshev expansions (cf.
-\refalg{alg:chebyshev-squaring}), the second parenthesized term in \refequ{equ:cyclic-property} can be approximated in an equivalent way.
-Unlike expanding the squared smoothing kernel $g_{\sigma}^2$ separately \cite{lin-2017-randomized-estimation}, this approach is consistent, which we have observed to make a noticeable difference in terms of accuracy (cf.
-\reffig{fig:interpolation-issue}), and which enables provability.
-
-\begin{figure}[ht]
-    \centering
-    \input{plots/interpolation.pgf}
-    \caption{Difference in the accuracy when computing \refequ{equ:3-nystrom-spectral-density} using a separate expansion (interpolation) versus explicitly squaring the matrix function (squaring).
-With the NC, which we will introduce on the next few pages, the spectral density of a 2D model problem from \refsec{sec:5-experiments-density-function} is computed for a Gaussian smoothing-kernel with smoothing-parameter $=0.05$ and fixed sketch-size $=80$.
-An estimate of the relative error in the $L^1$-norm is computed based on num-evaluation-points $=100$ evaluation of the spectral density.}
-    \label{fig:interpolation-issue}
-\end{figure}
-
-Similar observations for the case when $\mtx{\Psi}$ is nonzero lead to the following algorithm.
-
-\begin{algo}{Chebyshev-Nyström++ method}{nystrom-chebyshev-pp}
-\begin{algorithmic}[1]
-    \Statex \textbf{Input:} Symmetric $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum in $[-1, 1]$, points $\{t_i\}_{i=1}^{n_t} \subset \mathbb{R}$
-    \Statex \textbf{Parameters:} Degree of expansion $m \in 2\mathbb{N}$, number of Girard-Hutchinson queries $n_{\mtx{\Psi}} \in \mathbb{N}_0$, Nyström sketch size $n_{\mtx{\Omega}} \in \mathbb{N}_0$,  smoothing parameter $\sigma > 0$
-    \Statex \textbf{Output:} Approximate evaluations of the spectral density $\{\widetilde{\phi}_{\sigma}(t_i)\}_{i=1}^{n_t}$
-    \State Compute $\{\mu_l(t_i)\}_{l=0}^{m}$ for all $t_i$ with non-negative Chebyshev expansion \refequ{equ:non-negative-chebyshev-expansion}
-    \State Compute $\{\nu_l(t_i)\}_{l=0}^{2m}$ using \refalg{alg:chebyshev-squaring}%\refalg{alg:2-chebyshev-chebyshev-expansion}
-    %\State Compute  for all $t_i$ using %\refalg{alg:3-nystrom-chebyshev-exponentiation}
-    \State Generate standard Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$%\glsfirst{sketching-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$
-    %\State Generate standard Gaussian %\glsfirst{random-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$
-    \State Initialize $[\mtx{V}_1, \mtx{V}_2, \mtx{V}_3] \gets [\mtx{0}_{n \times n_{\mtx{\Omega}}}, \mtx{\mtx{\Omega}}, \mtx{0}_{n \times n_{\mtx{\Omega}}}]$
-    \State Initialize $[\mtx{W}_1, \mtx{W}_2, \mtx{W}_3] \gets [\mtx{0}_{n \times n_{\mtx{\Psi}}}, \mtx{\Psi}, \mtx{0}_{n \times n_{\mtx{\Psi}}}]$
-    \State Initialize $[\mtx{K}_1(t_i), \mtx{K}_2(t_i)] \gets [\mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}}, \mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}}]$ for all $t_i$
-    \State Initialize $[\mtx{L}_1(t_i), \ell(t_i)] \gets [\mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Psi}}}, 0]$ for all $t_i$
-    %\State Set $\breve{\phi}_{\sigma}^{(m)}(t_i) \gets 0$ for all $t_i$
-    \For {$l = 0, \dots, 2m$}
-    \State $[\mtx{X}, \mtx{Y}] \gets \mtx{\mtx{\Omega}}^{\top} [\mtx{V}_2, \mtx{W}_2]$  
-    %\State $\mtx{X} \gets \mtx{\mtx{\Omega}}^{\top} \mtx{V}_2$
-      %\State $\mtx{Y} \gets \mtx{\mtx{\Omega}}^{\top} \mtx{W}_2$
-      \State $z \gets \Trace(\mtx{\Psi}^{\top} \mtx{W}_2)$
-      \For {$i = 1, \dots, n_t$}
-        \If {$l \leq m$}
-            \State $\mtx{K}_1(t_i) \gets \mtx{K}_1(t_i) + \mu_l(t_i) \mtx{X}$ \Comment{assemble $\mtx{\Omega}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$}
-            \State $\mtx{L}_1(t_i) \gets \mtx{L}_1(t_i) + \mu_l(t_i) \mtx{Y}$ \Comment{assemble $\mtx{\Omega}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi}$}
-            \State $\ell(t_i) \gets \ell(t_i) + \mu_l(t_i) z$ \Comment{assemble $\Trace(\mtx{\Psi}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi})$}
-        \EndIf
-        \State $\mtx{K}_2(t_i) \gets \mtx{K}_2(t_i) + \nu_l(t_i) \mtx{X}$ \Comment{assemble $\mtx{\Omega}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})^2 \mtx{\Omega}$}
-      \EndFor
-      \State $[\mtx{V}_3, \mtx{W}_3] \gets (2 - \delta_{l0}) \mtx{A} [\mtx{V}_2, \mtx{W}_2] - [\mtx{V}_1, \mtx{W}_1]$ \Comment{Chebyshev recurrence}
-      \State $[\mtx{V}_1, \mtx{W}_1] \gets [\mtx{V}_2, \mtx{W}_2]$
-      \State $[\mtx{V}_2, \mtx{W}_2] \gets [\mtx{V}_3, \mtx{W}_3]$
-      %\State $\mtx{V}_3 \gets (2 - \delta_{l0}) \mtx{A} \mtx{V}_2 - \mtx{V}_1$ %\Comment{Chebyshev recurrence \refequ{equ:2-chebyshev-chebyshev-recursion}}
-      %\State $\mtx{V}_1 \gets \mtx{V}_2, \mtx{V}_2 \gets \mtx{V}_3$
-      %\State $\mtx{W}_3 \gets (2 - \delta_{l0}) \mtx{A} \mtx{W}_2 - \mtx{W}_1$ %\Comment{Chebyshev recurrence \refequ{equ:2-chebyshev-chebyshev-recursion}}
-      %\State $\mtx{W}_1 \gets \mtx{W}_2, \mtx{W}_2 \gets \mtx{W}_3$
-    \EndFor
-    \For {$i = 1, \dots, n_t$}
-      \State $\widetilde{\phi}_{\sigma}(t_i) \gets \frac{1}{n} \Trace\left( \mtx{K}_1(t_i)^{\dagger}\mtx{K}_2(t_i) \right) + \frac{1}{n n_{\mtx{\Psi}}} \left( \ell(t_i) + \Trace\left( \mtx{L}_1(t_i)^{\top} \mtx{K}_1(t_i)^{\dagger} \mtx{L}_1(t_i) \right)  \right) $ \label{lin:4-nystromchebyshev-nystrom-pp}
-    \EndFor
-\end{algorithmic}
-\end{algo}
-
-The dominating cost stems from the $(2m + 1)(n_{\mtx{\Psi}} + n_{\mtx{\Omega}})$ matrix-vector multiplications with $\mtx{A}$ and the cost associated with computing the estimate in each evaluation point $t_1, \dots, t_{n_t}$ given by $\mathcal{O}(m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + (n_{\mtx{\Omega}}^3 + n_{\mtx{\Omega}} n_{\mtx{\Psi}}^2 + 1))$.
-
-The algorithm also accomodates the two other estimators introduced in \refsec{sec:analysis}: For $n_{\mtx{\Omega}} = 0$ we recover the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}, and the complexity for each evaluation of the estimate decreases significantly.
-For $n_{\mtx{\Psi}} = 0$ we end up with the Nyström estimator \refequ{equ:nystrom-trace-estimator}, provided the implementation evaluates $0/0$ to $0$ in the second term on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
-
-%Chebyshev $\mathcal{O}(n_t (m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + n_{\mtx{\Omega}}^3) + mn^2(n_{\mtx{\Psi}} + n_{\mtx{\Omega}}))$.
-
-%\todo{Complexity in terms of both $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$.}
-%With the cost of a matrix-vector product denoted by $c(n)$, and supposing $n_{\mtx{\Omega}} \approx n_{\mtx{\Psi}}$, we determine the computational complexity of the Chebyshev-Nyström++ method to be $\mathcal{O}(m \log(m) n_t + m n_{\mtx{\Omega}}^2 n + m n_t n_{\mtx{\Omega}}^2 +  m c(n) n_{\mtx{\Omega}} + n_t n_{\mtx{\Omega}}^3)$, with $\mathcal{O}(m n_t + n n_{\mtx{\Omega}} + n_{\mtx{\Omega}}^2 n_t)$ required additional storage.
-
-%\todo{Explain modifications for other estimators, and improved complexity for quadratic trace estimator}
-
-The pseudoinverses on \reflin{lin:4-nystromchebyshev-nystrom-pp} need to be computed with care.
-It is possible to reformulate the problem as a generalized eigenvalue problem and apply thresholding to the smallest eigenvalues of a certain matrix to enforce stability \cite{lin-2017-randomized-estimation, epperly-2022-theory-quantuma}.
-Since this renders the algorithm significantly more complicated but only improved the accuracy slightly during our experiments, we will instead evaluate these expressions in \reflin{lin:4-nystromchebyshev-nystrom-pp} using a least-squares solver which likewise uses a truncation of the smallest eigenvalues to enforce better conditioning.
-
-What we have found to be crucial for a successful approximation of the full spectral density was the detection of regions wherein lie no eigenvalues, i.e.
-regions where the spectral density nearly vanishes.
-For small smoothing parameters $\sigma$, the rapid decay of the Gaussian $g_{\sigma}$ will cause $\mtx{K}(t_i) \approx g_{\sigma}(t_i\mtx{I}_n - \mtx{A})$ to be close to the zero matrix for parameter values $t_i$ located in such regions.
-Hence, on line \reflin{lin:4-nystromchebyshev-nystrom-pp} we would compute the pseudoinverse of an almost zero matrix, which can be extremely unstable.
-Notice that $\Trace(\mtx{K}_1(t_i)) / n_{\mtx{\Omega}}$ is the $n_{\mtx{\Omega}}$-query Girard-Hutchinson estimator.
-We may use it as a rough indicator whether $t_i$ lies within a region where the spectral density nearly vanishes, and subsequently set $\widetilde{\phi}_{\sigma}(t_i) = 0$ instead of computing the full expression on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
-
-\subsubsection{Analysis}
-\label{subsubsec:chebyshev-nystrom-analysis}
-
-In this section, we will upper bound the output of the Chebyshev-Nyström++ method (see. \refalg{alg:nystrom-chebyshev-pp}). To this extent, we will first analyze the non-negative Chebyshev expansion.
-
-\begin{lemma}{Non-negative Chebyshev expansion error}{non-negative-chebyshev-error}
-    The non-negative expansion $\underline{g}_{\sigma}^{(m)}$ of the Gaussian smoothing kernel satisfies
-    \begin{equation}
-        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \leq 2\sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right) E_{\sqrt{2}\sigma, \sfrac{m}{2}} \equiv \underline{E}_{\sigma, m}
-        \label{equ:chebyshev-interpolation-sup-error-kernel-nonneg}
-        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
-        %\label{equ:2-chebyshev-interpolation-error}
-    \end{equation}
-    for all \emph{even} degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, $s \in [-1, 1]$, and with $E_{\sigma, m}$ as defined in \reflem{lem:chebyshev-error}.
-\end{lemma}
-
-\begin{proof}
-    For any numbers $a, b \in \mathbb{R}$ it holds
-    \begin{equation}
-    | a^2 - b^2 | = | (a + b)(a - b) | = | a + b | | a - b | \leq (2 | a | + | a - b |)  | a - b |
-    \end{equation}
-    Therefore, omitting the arguments and using $g_{\sigma} = (\sqrt{g_{\sigma}})^2$ and $\underline{g}_{\sigma}^{(m)} = (\sqrt{g_{\sigma}}^{(\sfrac{m}{2})})^2$.
-    we have
-    \begin{equation}
-        \left| g_{\sigma} - \underline{g}_{\sigma}^{(m)} \right| \leq \left( 2 \left| \sqrt{g_{\sigma}} \right| + \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right| \right) \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right|.
-    \end{equation}
-    Since $\sqrt{g_{\sigma}} = \sqrt{2 \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$ and $\sqrt{g_{\sigma}} \leq 1/\sqrt{\sigma \sqrt{2 \pi}}$ we can apply the result from \refthm{thm:chebyshev-error} to get
-    \begin{align}
-        &\sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \notag \\
-        &\leq \left( 2 \frac{1}{\sqrt{ \sigma \sqrt{2\pi}}} + \sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right)\sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}},
-    \end{align}
-    from which follows the result with some minor simplifications.
-\end{proof}
-
-On one hand, in the non-negative expansion, the underlying computes Chebyshev expansion -- which is shown in \reflem{lem:chebyshev-error} to exponentially decay with its degree -- is only computed to degree $m/2$, whereas the standard expansion is of degree $m$.
-On the other hand, we approximate $\sqrt{g_{\sigma}}$ instead of $g_{\sigma}$, which is less \enquote{pointy} and therefore better approximated by polynomials.
-Overall, we will need to choose the degree $m$ of the expansion approximately a factor of $\sqrt{2}$ larger to match the error of the standard expansion.
-
-\begin{theorem}{Chebyshev-Nyström++ method}{chebyshev-nystrom}
-    Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, \emph{even} $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2)$, and smoothing parameter $0 < \sigma \ll 1$ satisfies with probability at least $1 - \delta$
-    \begin{equation}
-        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq \varepsilon
-        \label{equ:chebyshev-nystrom-error}
-        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
-        %\label{equ:2-chebyshev-interpolation-error}
-    \end{equation}
-    for all $\varepsilon \in (0, 1)$ and $\delta \in (0, e^{-\sfrac{1}{2}}/2)$.
-\end{theorem}
-
-\begin{proof}
-    The previously derived results can be applied to
-    \begin{align}
-        &\int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \notag \\
-        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \int_{-1}^{1} \left| \phi^{(m)}_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
-        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \widetilde{\varepsilon} \int_{-1}^{1} \left| \phi^{(m)}_{\sigma}(t) \right|~\mathrm{d}t && \text{(\refthm{thm:nystrom-pp} w.p.
-$\geq 1 - \delta$)} \notag \\
-        &\leq (1 + \widetilde{\varepsilon}) \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \varepsilon \int_{-1}^{1} \left| \phi_{\sigma}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
-        &\leq (1 + \widetilde{\varepsilon}) 2 \underline{E}_{\sigma, m} + \widetilde{\varepsilon}. && \text{(\refthm{thm:chebyshev-error}, normalization)} \notag \\
-    \end{align}
-    From \reflem{lem:non-negative-chebyshev-error} we see that if $0 < \sigma \ll 1$, choosing $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$ ensures $\underline{E}_{\sigma, m} \leq \widetilde{\varepsilon}$, and thus, since $\widetilde{\varepsilon} \leq 1$,
-    \begin{equation}
-        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq (1+\widetilde{\varepsilon}) 2 \widetilde{\varepsilon} + \widetilde{\varepsilon} \leq 5 \widetilde{\varepsilon}
-    \end{equation}
-    from which the result follows by identifying $\varepsilon = 5 \widetilde{\varepsilon}$.
-\end{proof}
-
-%\todo{Maybe specialize \refthm{thm:nystrom} to spectral densities?}
-
-%We now discuss what value to choose for $n_{\mtx{\Omega}}$ in the Nyström trace estimator introduced in \refsec{subsec:nystrom}.
-In order to guarantee an even better approximation error based on \refthm{thm:nystrom}, we analyze the numerical rank of the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
-
-In practice, the approximation error is often significantly better than guaranteed in \refthm{thm:chebyshev-nystrom}, particularly when $n_{\mtx{\Omega}}$ is chosen rather large.
-This is because the singular values of the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ decay so quickly that already the Nyström approximation alone is highly accurate (cf.
-\refthm{thm:nystrom}).
-To establish a relation between the singular value decay of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ and what choice of $n_{\mtx{\Omega}}$ is needed to achieve a better approximation than what we show in \refthm{thm:chebyshev-nystrom}, we study the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
-
-We can express the numerical rank of a matrix with respect to the nuclear norm in terms of its singular values $\sigma_1 \geq \dots \geq \sigma_n \geq 0$
-\begin{equation}
-    r_{\varepsilon} = \min \{1 \leq r \leq n: \sum_{j=r+1}^n \sigma_{j} \leq \varepsilon \},
-    \label{equ:numerical-rank}
-\end{equation}
-$\varepsilon$ is usually taken to be the double machine precision, i.e. $10^{-16}$.
-
-%\begin{theorem}{Nyström estimator for parameter-dependent matrices}%{chebyshev-nystrom-raw}
-%    Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric %matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is %contained in $[-1, 1]$.
-Then the Chebyshev-Nyström++ estimate %$\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m \in %\mathbb{N}$, $n_{\mtx{\Omega}} \geq \max(6, \max_{t \in [-1, 1]}(r_%{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A}))) + \log(\delta^%{-1}))$, $n_{\mtx{\Psi}} = 0$, and smoothing parameter $\sigma > 0$ %satisfies with for all $\gamma \geq 1$ with probability at least $1 %- \delta$
-%    \begin{equation}
-%        \int_{a}^{b} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}%(t) \right| ~\mathrm{d}t
-%        \leq 2 e^2 (1 + r_{\varepsilon}) (\varepsilon + (n - r_%{\varepsilon}) \underline{E}_{\sigma, m}) + \underline{E}_%{\sigma, m}.
-%    \end{equation}
-%\end{theorem}
-%
-%\begin{proof}
-%    \todo{TODO}
-%\end{proof}
-
-When using a Gaussian smoothing-kernel, we can explicitly write the singular values of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ as
-\begin{equation}
-    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) = \frac{1}{\sqrt{2 \pi \sigma^2}} e^{-\sfrac{(t - \lambda_{(i)})^2}{2 \sigma^2}},
-    \label{equ:gaussian-kernel-eigenvalues}
-\end{equation}
-where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from spectral-parameter $t$, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t)$.
-Consequently, by using the closed-form expression of the eigenvalues \refequ{equ:gaussian-kernel-eigenvalues}, we may upper bound the numerical rank of a Gaussian smoothing kernel as
-\begin{equation}
-    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A})) \leq \#\left\{1\leq i\leq n: |t - \lambda_i| < d_{\varepsilon, \sigma} \right\},
-    \label{equ:gaussian-kernel-numerical-rank}
-\end{equation}
-with $d_{\varepsilon, \sigma} = \sigma \sqrt{-2 \log(\sqrt{2 \pi} \sigma \varepsilon)}$ and the eigenvalues $\lambda_1, \dots, \lambda_n$ of $\mtx{A}$ and the constants.
-The expression \refequ{equ:gaussian-kernel-numerical-rank} has a nice visual interpretation: The numerical rank of a matrix is at most equal to the number of eigenvalues which are closer to the parameter $t$ than $d_{\varepsilon, \sigma}$.
-This is illustrated in \reffig{fig:numerical-rank}.
-\begin{figure}[ht]
-    \centering
-    \input{figures/numerical-rank.tex}
-    \caption{The numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ can be
-        approximately computed by counting the number of eigenvalues
-        $\lambda_{(1)}, \dots, \lambda_{(n)}$ of the matrix $\mtx{A}$ which lie less than
-        a constant $d_{\varepsilon, \sigma}$ away from $t$.}
-    \label{fig:numerical-rank}
-\end{figure}
-
-If we additionally assume the eigenvalues of the matrix $\mtx{A}$ to be evenly distributed within $[-1, 1]$, that is, in any subinterval of fixed length in $[a, b]$, $-1 \leq a < b \leq 1$, we can expect to find roughly the same number of eigenvalues, then we can estimate the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ to be
-\begin{equation}
-    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A})) \lessapprox n d_{\varepsilon, \sigma} \approx n \sigma.
-    \label{equ:gaussian-kernel-numerical-rank-uniform}
-\end{equation}
-Hence, choosing $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma)$ will approximations which are significantly better than \refthm{thm:chebyshev-nystrom}.
-
 \section{Numerical results}
@@
 
+\todo{Hessian spectrum analysis?}
+
 \todo{Comparison plots with other estimators (Krylov-Aware)?}
-
-\clearpage
-\section{Conclusion}
-\label{sec:conclusion}
-
-\todo{
-\begin{itemize}
-    \item Robustness (can choose $m$ and $n_{\mtx{\Omega}}$, $n_{\mtx{\Psi}}$ as large as one wants, and not break algorithm, unlike Krylov-aware Lanczos)
-\end{itemize}
-}
-
-% Acknowledge advice from David and Data/Parameters from Lin Lin, Michael Herbst?
-
-\clearpage
-%\bibliography{bibliography.bib}
-\printbibliography
-
-\appendix
-
-\clearpage
-\section{Proofs}
-
-\begin{lemma}{Moment bound of $\chi^2$-random variable}{gamma}
-    Let $X$ follow a $\chi^2$ distribution with $k \in \mathbb{N}$ degrees of freedom. Then for all $p \in \mathbb{N}$ it holds
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}[X] \leq k + p
-        %\frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \leq \left( \frac{k + p}{2} \right)^{\sfrac{p}{2}}.
-    \end{equation}
-\end{lemma}
-
-\begin{proof}
-    From \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical} we know
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}[X] = 2 \left( \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \right)^{\sfrac{2}{p}}.
-    \end{equation}
-    We first treat the case $k=p=1$, for which indeed
-    \begin{equation}
-        2 \left( \frac{\Gamma(\frac{1+1}{2})}{\Gamma(\frac{1}{2})} \right)^2 = \frac{2}{\pi} \leq 2 = 1 + 1
-    \end{equation}
-    So we now assume $k + p \geq 3$. We can use a telescoping product and the property $\Gamma(z+1)/\Gamma(z) = z$ to bound
-    \begin{align}
-        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} 
-        &= \frac{\Gamma(\frac{k}{2} + 1)}{\Gamma(\frac{k}{2})} \frac{\Gamma(\frac{k}{2} + 2)}{\Gamma(\frac{k}{2} + 1)} \cdots \frac{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor - 1)} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)} \notag \\
-        &= \left(\frac{k}{2}\right) \left(\frac{k}{2} + 1\right) \cdots \left( \frac{k}{2} + \bigg\lfloor \frac{p}{2} \bigg\rfloor - 1 \right) \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)} \notag \\
-        &\leq \left(\frac{k}{2} + \bigg\lfloor \frac{p}{2} \bigg\rfloor - 1\right)^{\lfloor \sfrac{p}{2} \rfloor} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)}
-    \end{align}
-    If $p$ is even, then $\lfloor p/2 \rfloor = p/2$ and hence
-    \begin{equation}
-        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} 
-        \leq \left( \frac{k + p}{2} - 1 \right)^{\sfrac{p}{2}}
-        \leq \left( \frac{k + p}{2}\right)^{\sfrac{p}{2}}.
-    \end{equation}
-    If $p$ is odd, then $\lfloor p/2 \rfloor = (p - 1)/2$, from which follows with Gautschi's inequality \cite{kershaw-1983-extensions-gautschi}
-    \begin{equation}
-        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})}
-        \leq \left(\frac{k + p}{2} - \frac{3}{2} \right)^{ \frac{p - 1}{2}} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k + p}{2} - \frac{1}{2})}
-        \leq \left(\frac{k + p}{2}\right)^{ \frac{p - 1}{2}} \left(\frac{k + p}{2}\right)^{\sfrac{1}{2}}
-        = \left(\frac{k + p}{2}\right)^{ \sfrac{p}{2}}
-    \end{equation}
-    Hence,
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}[X] \leq 2 \left( \frac{k + p}{2} \right) \leq k + p.
-    \end{equation}
-\end{proof}
-
-\end{document}

### 2024-07-07 paper/analysis.tex
clean up bibliography and add acknowledgments

@@
 
-\paragraph{Girard-Hutchinson estimator} We can approximate the trace with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ stochastically independent standard Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ to form
+\paragraph{Girard-Hutchinson estimator.} We can approximate the trace with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ stochastically independent standard Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ to form
 \begin{equation}
@@
 
-\paragraph{Nyström estimator} Alternatively, the trace of a symmetric matrix whose singular values decay quickly can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström approximation \cite{gittens-2013-revisiting-nystrom}
+\paragraph{Nyström estimator.} Alternatively, the trace of a symmetric matrix whose singular values decay quickly can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström approximation \cite{gittens-2013-revisiting-nystrom}
 \begin{equation}
@@
 
-\paragraph{Nyström++ estimator} Finally, \cite{lin-2017-randomized-estimation} proposes an estimator which corrects for inaccuracies in the Nyström approximation \refequ{equ:nystrom-trace-estimator} by estimating the trace of its residual using the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}
+\paragraph{Nyström++ estimator.} Finally, \cite{lin-2017-randomized-estimation} proposes an estimator which corrects for inaccuracies in the Nyström approximation \refequ{equ:nystrom-trace-estimator} by estimating the trace of its residual using the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}
 \begin{equation}
@@
     First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence.
-From the proof of \cite[lemma 3]{cortinovis-2022-randomized-trace} we know that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm.
-Thus, by \cite[theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
+From the proof of \cite[Lemma 3]{cortinovis-2022-randomized-trace} we know that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm.
+Thus, by \cite[Theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
     \begin{align}
@@
 Following \cite{kressner-2023-randomized-lowrank}, we can show that $r(\mtx{B}(t), \vct{\psi})$ is measurable.
-Therefore, Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting:
+Therefore, Minkowski's integral inequality \cite[Theorem 202]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting:
     \begin{equation}
@@
 
-    To extend the bound to the $n_{\mtx{\Psi}}$-query quadratic trace estimator, we use a technique from the proof of \cite[theorem 1]{cortinovis-2022-randomized-trace}. Let
+    To extend the bound to the $n_{\mtx{\Psi}}$-query quadratic trace estimator, we use a technique from the proof of \cite[Theorem 1]{cortinovis-2022-randomized-trace}. Let
     \begin{equation}
@@
 %    if $n_{\mtx{\Psi}} \geq 2^{10} \varepsilon^{-2} \log(\delta^{-1})^2 (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ with $\mu(t) = \lVert \mtx{B}(t) \rVert _2 / \Trace(\mtx{B}(t))$.
-Compared to the constant matrix equivalent \cite[remark 2]{cortinovis-2022-randomized-trace} we mainly observe a larger constant and an additional factor of $\log(\delta^{-1})$.
+Compared to the constant matrix equivalent \cite[Remark 2]{cortinovis-2022-randomized-trace} we mainly observe a larger constant and an additional factor of $\log(\delta^{-1})$.
 %\end{remark}
 
-Compared to the constant case \cite[lemma 2.1]{meyer-2021-hutch-optimal} we need to choose the number of queries $n_{\mtx{\Psi}}$ by a factor of $\log(\delta^{-1})$ larger to achieve the same guarantee.
+Compared to the constant case \cite[Lemma 2.1]{meyer-2021-hutch-optimal} we need to choose the number of queries $n_{\mtx{\Psi}}$ by a factor of $\log(\delta^{-1})$ larger to achieve the same guarantee.
 
@@
 \begin{proof}
-    Since $\mtx{B}(t)$ is positive semi-definite for all $t$, then by \cite[lemma 2.1]{frangella-2023-randomized-nystrom}, $\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ is also positive semi-definite for all $t$.
+    Since $\mtx{B}(t)$ is positive semi-definite for all $t$, then by \cite[Lemma 2.1]{frangella-2023-randomized-nystrom}, $\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ is also positive semi-definite for all $t$.
 Therefore, for fixed $t \in [a, b]$
@@
     \end{equation}
-    where $\lVert \cdot \rVert _{\ast}$ represents the nuclear norm.  From the proof of \cite[corollary 8.2]{tropp-2023-randomized-algorithms} it follows that
+    where $\lVert \cdot \rVert _{\ast}$ represents the nuclear norm.  From the proof of \cite[Corollary 8.2]{tropp-2023-randomized-algorithms} it follows that
     \begin{equation}
@@
     \end{equation}
-    Hence, by \cite[theorem 9]{kressner-2023-randomized-lowrank}, with $n_{\mtx{\Omega}} = p + r$, the result follows by noting $\sigma_i(\mtx{B}(t)^{\sfrac{1}{2}})^{2} = \sigma_i(\mtx{B}(t))$.
+    Hence, by \cite[Theorem 9]{kressner-2023-randomized-lowrank}, with $n_{\mtx{\Omega}} = p + r$, the result follows by noting $\sigma_i(\mtx{B}(t)^{\sfrac{1}{2}})^{2} = \sigma_i(\mtx{B}(t))$.
 
@@
 
-Compared to the constant matrix case \cite[theorem 8.1]{tropp-2023-randomized-algorithms}, our bound also scales with the best rank-$r$ approximation error in the nuclear norm, albeit with a factor approximately proportional to $r$, which cannot be compensated for with large enough oversampling.
+Compared to the constant matrix case \cite[Theorem 8.1]{tropp-2023-randomized-algorithms}, our bound also scales with the best rank-$r$ approximation error in the nuclear norm, albeit with a factor approximately proportional to $r$, which cannot be compensated for with large enough oversampling.
 
@@
 Just as in the constant matrix case, our goal is now to prove that the Nyström++ estimator \refequ{equ:nystrompp-trace-estimator} achieves an $\varepsilon$-relative error with just $\mathcal{O}(\varepsilon^{-1})$ random vectors used to construct the estimate.
-To do so, we mimic the proof of a similar bound for the Hutch++ estimator \cite[theorem 3.1]{meyer-2021-hutch-optimal}.
+To do so, we mimic the proof of a similar bound for the Hutch++ estimator \cite[Theorem 3.1]{meyer-2021-hutch-optimal}.
 Correspondingly, we will need to show that with $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2})$ random vectors, the Nyström approximation verifies $\int \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)\rVert _F~\mathrm{d}t \leq \varepsilon \int \Trace(\mtx{B}(t))~\mathrm{d}t$, regardless of the structure of the matrix $\mtx{B}(t)$.
@@
 For reasons which will become clear in the proof of the final and most interesting of these lemmas (\reflem{lem:nystrom}), we will need to bound $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]$ for a constant matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and a standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$.
-The corresponding result for the Frobenius norm $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _F^2 \right]$ has already been shown in \cite[lemma 7]{kressner-2023-randomized-lowrank}.
+The corresponding result for the Frobenius norm $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _F^2 \right]$ has already been shown in \cite[Lemma 7]{kressner-2023-randomized-lowrank}.
 Several ways of separately bounding the spectral norm moments of the Wishart matrix $\mtx{\Omega} \mtx{\Omega}^{\top}$, i.e.
 when $\mtx{A} = \mtx{I}_m$ already exist, however, all of them are not sufficient for our purposes because they scale with $m$ \cite{chen-2005-condition-numbers, edelman-1988-eigenvalues-condition, james-1964-distributions-matrix}.
-In \cite[lemma B.1]{tropp-2023-randomized-algorithms} a bound is achieved for $p = 2$ and $p = 4$.
+In \cite[Lemma B.1]{tropp-2023-randomized-algorithms} a bound is achieved for $p = 2$ and $p = 4$.
 We will now generalize this result for arbitrary $p \in \mathbb{N}$.
@@
     \end{equation}
-    Inspired by the proof of \cite[theorem 1]{cohen-2016-optimal-approximate}, we split the diagonal entries of $\mtx{\Sigma}$ into $\ell = \lceil m/k \rceil$ groups of size $k \geq 1$
+    Inspired by the proof of \cite[Theorem 1]{cohen-2016-optimal-approximate}, we split the diagonal entries of $\mtx{\Sigma}$ into $\ell = \lceil m/k \rceil$ groups of size $k \geq 1$
     \begin{equation}
@@
         &\leq \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 ~ \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{(i-1)k + j}^2 \right] && \text{(triangle inequality)} \notag \\
-        %&\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k}{2} + p)}{\Gamma(\frac{k}{2})} \right)^{\sfrac{1}{p}} && \text{(\refequ{equ:singular-value-group-bound} and \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
+        %&\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k}{2} + p)}{\Gamma(\frac{k}{2})} \right)^{\sfrac{1}{p}} && \text{(\refequ{equ:singular-value-group-bound} and \cite[Theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
         &\leq (k + p) \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 && \text{(\reflem{lem:gamma})} \notag \\
@@
 
-    %The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
+    %The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[Lemma B.3]{tropp-2023-randomized-algorithms}
     %\begin{align}
@@
     \end{equation}
-    for any $k,p \in \mathbb{N}$. For $p=2$ this bound is slightly less tight than its equivalent in \cite[lemma B.1]{tropp-2023-randomized-algorithms}.
+    for any $k,p \in \mathbb{N}$. For $p=2$ this bound is slightly less tight than its equivalent in \cite[Lemma B.1]{tropp-2023-randomized-algorithms}.
 \end{remark}
@@
 \begin{proof}
-    In the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms}, it is shown that Slepian's inequality \cite[theorem 7.2.1]{vershynin-2018-random-processes} applied to two appropriately constructed Gaussian random fields yields
+    In the proof of \cite[Lemma B.1]{tropp-2023-randomized-algorithms}, it is shown that Slepian's inequality \cite[Theorem 7.2.1]{vershynin-2018-highdimensional-probability} applied to two appropriately constructed Gaussian random fields yields
     \begin{equation}
@@
 
-    We can apply \cite[theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound
+    We can apply \cite[Theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound
     \begin{equation}
@@
 
-    As in the proof of \cite[lemma 3]{meyer-2021-hutch-optimal}, the first term -- and its moments -- are limited by
+    As in the proof of \cite[Lemma 3]{meyer-2021-hutch-optimal}, the first term -- and its moments -- are limited by
     \begin{equation}
@@
     \end{equation}
-    The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
+    The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[Lemma B.3]{tropp-2023-randomized-algorithms}
     \begin{equation}
@@
     \end{equation}
-    As opposed to \cite[lemma B.3]{tropp-2023-randomized-algorithms}, our bound holds for all even $n_{\mtx{\Omega}} \in \mathbb{N}$, at the cost of a factor of $e^2$.
+    As opposed to \cite[Lemma B.3]{tropp-2023-randomized-algorithms}, our bound holds for all even $n_{\mtx{\Omega}} \in \mathbb{N}$, at the cost of a factor of $e^2$.
 Inserting \refequ{equ:spectral-norm-bound-applied} and  \refequ{equ:pinv-spectral-norm-bound} in \refequ{equ:nystrom-proof-processed-tail} with $k = n_{\mtx{\Omega}} / 2$ gives
@@
     \end{equation}
-    We can identify $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 = \lambda_{k+1}$ and $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 = \Trace(\mtx{\Lambda}_2)$, and just as in the proof of \cite[lemma 3.1]{meyer-2021-hutch-optimal} we bound $\lambda_{k+1} \leq \Trace(\mtx{B})/k$ and $\Trace(\mtx{\Lambda}_2) \leq \Trace(\mtx{B})$ -- remembering $k = n_{\mtx{\Omega}}/2$ -- to get
+    We can identify $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 = \lambda_{k+1}$ and $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 = \Trace(\mtx{\Lambda}_2)$, and just as in the proof of \cite[Lemma 3.1]{meyer-2021-hutch-optimal} we bound $\lambda_{k+1} \leq \Trace(\mtx{B})/k$ and $\Trace(\mtx{\Lambda}_2) \leq \Trace(\mtx{B})$ -- remembering $k = n_{\mtx{\Omega}}/2$ -- to get
     \begin{equation}
@@
 
-    As in \cite{kressner-2023-randomized-lowrank}, we can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable and by the Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities}, we get for $n_{\mtx{\Omega}} \geq 4$
+    As in \cite{kressner-2023-randomized-lowrank}, we can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable and by the Minkowski's integral inequality \cite[Theorem 202]{hardy-1952-inequalities}, we get for $n_{\mtx{\Omega}} \geq 4$
     \begin{align}
@@
 
-Comparing \refthm{thm:nystrom-pp} with its equivalent for constant matrices \cite[theorem 3.4]{persson-2022-improved-variants}, we notice a less favorable scaling of the required number of random vectors $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$ in terms of $\log(\delta^{-1})$.
+Comparing \refthm{thm:nystrom-pp} with its equivalent for constant matrices \cite[Theorem 3.4]{persson-2022-improved-variants}, we notice a less favorable scaling of the required number of random vectors $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$ in terms of $\log(\delta^{-1})$.
 

### 2024-07-07 paper/paper.tex
clean up bibliography and add acknowledgments

@@
 \begin{itemize}
- \item Capitalize references to algorithms, lemma etc (that is, Lemma 3.2 instead of lemma 3.2 [Lin17, Algorithm 1] instead of [Lin17, algorithm 1]), etc.
- \item Add full stops after paragraph titles.
- \item It is not great that the references occupy so much space (more than 4 pages); consider using a more condensed style. The paper needs to get strictly below 30 pages prior to submission. 
- \item Please include Acknowledgements as indicated in the comment.
+ \item Capitalize references to algorithms, lemma etc (that is, Lemma 3.2 instead of lemma 3.2 [Lin17, Algorithm 1] instead of [Lin17, algorithm 1]), etc. \textcolor{green}{Done.}
+ \item Add full stops after paragraph titles. \textcolor{green}{Done.}
+ \item It is not great that the references occupy so much space (more than 4 pages); consider using a more condensed style.
+The paper needs to get strictly below 30 pages prior to submission.
+\textcolor{green}{Journal abbreviations, max 2 authors, omit superfluous fields $\to$ 3 pages.}
+ \item Please include Acknowledgements as indicated in the comment. \textcolor{green}{Done.}
 \end{itemize}
@@
 
-% Acknowledge advice from David and Data/Parameters from Lin Lin, Michael Herbst?
+\paragraph{Acknowledgements.} The authors gratefully acknowledge the help of Lin Lin in reproducing the example in \refsec{subsec:hamiltonian}.
+They thank David Persson for many enlightening discussions relating the proofs in \refsec{subsec:nystrom-pp}.
 
-%\clearpage
-%\bibliography{bibliography.bib}
 \printbibliography
@@
 
-%\clearpage
 \section{Proofs}
@@
 \begin{proof}
-    From \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical} we know
+    From \cite[Theorem 3.3.2]{hogg-2013-introduction-mathematical} we know
     \begin{equation}

### 2024-07-07 paper/spectraldensity.tex
clean up bibliography and add acknowledgments

@@
 This can be accomplished in $\mathcal{O}(m \log(m))$ operations.
-\cite[algorithm 1]{lin-2017-randomized-estimation} uses a slightly different approach, which only approximates the coefficients and is noticeably slower (cf.
+\cite[Algorithm 1]{lin-2017-randomized-estimation} uses a slightly different approach, which only approximates the coefficients and is noticeably slower (cf.
 \reftab{tab:chebyshev-timing-interpolation}).
@@
 The proof specializes Bernstein's theorem for Chebyshev interpolation to the case of a Gaussian smoothing kernel \refequ{equ:smoothing-kernel}.
-It loosely follows the proof given in the arXiv preprint of \cite[theorem 2]{lin-2017-randomized-estimation}, but attains a slightly different dependence on $\sigma$.
+It loosely follows the proof given in the arXiv preprint of \cite[Theorem 2]{lin-2017-randomized-estimation}, but attains a slightly different dependence on $\sigma$.
 
 \begin{proof}
-    From Bernstein's theorem \cite[theorem 4.3]{trefethen-2008-gauss-quadrature} it follows that if $f$ is analytic within the Bernstein ellipse $\mathcal{E}_{\chi} \subset \mathbb{C}$ with foci $\{-1, +1\}$ and sum of semi-axes $\chi > 1$, then for any $m \in \mathbb{N}$
+    From Bernstein's theorem \cite[Theorem 4.3]{trefethen-2008-gauss-quadrature} it follows that if $f$ is analytic within the Bernstein ellipse $\mathcal{E}_{\chi} \subset \mathbb{C}$ with foci $\{-1, +1\}$ and sum of semi-axes $\chi > 1$, then for any $m \in \mathbb{N}$
     \begin{equation}
@@
 
-\Reflem{lem:chebyshev-error} shows that when $0 < \sigma \ll 1$ and if we choose $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, the error we induce with the Chebyshev expansion is at most $\mathcal{O}(\varepsilon)$.
+\reflem{lem:chebyshev-error} shows that when $0 < \sigma \ll 1$ and if we choose $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, the error we induce with the Chebyshev expansion is at most $\mathcal{O}(\varepsilon)$.
 It now also allows us to bound the error we make when expanding the spectral density.
@@
 \begin{table}[ht]
-    \caption{We compare the runtime in milliseconds for computing the coefficients of the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ \refequ{equ:smoothing-kernel} in three different ways: as proposed in \cite[algorithm 1]{lin-2017-randomized-estimation} (\emph{oversampled FFT}), using the discrete cosine transform (\emph{DCT}) \refequ{equ:discrete-cosine-transform}, and computing a non-negative Chebyshev expansion with \refalg{alg:chebyshev-squaring} (\emph{non-negative DCT}).
+    \caption{We compare the runtime in milliseconds for computing the coefficients of the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ \refequ{equ:smoothing-kernel} in three different ways: as proposed in \cite[Algorithm 1]{lin-2017-randomized-estimation} (\emph{oversampled FFT}), using the discrete cosine transform (\emph{DCT}) \refequ{equ:discrete-cosine-transform}, and computing a non-negative Chebyshev expansion with \refalg{alg:chebyshev-squaring} (\emph{non-negative DCT}).
 We use $\sigma=0.005$, $n_t=1000$ parameter values, and various degrees $m$.
@@
 For $n_{\mtx{\Psi}} = 0$ we end up with the Nyström estimator \refequ{equ:nystrom-trace-estimator}, provided the implementation evaluates $0/0$ to $0$ in the second term on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
-Notice that in this case, \refalg{alg:nystrom-chebyshev-pp} is equivalent to the \enquote{spectrum sweeping} method from \cite[algorithm 5]{lin-2017-randomized-estimation}, but has a significantly lower computational complexity.
+Notice that in this case, \refalg{alg:nystrom-chebyshev-pp} is equivalent to the \enquote{spectrum sweeping} method from \cite[Algorithm 5]{lin-2017-randomized-estimation}, but has a significantly lower computational complexity.
 
@@
     \end{equation}
-    for all $\varepsilon \in (0, 1)$ and $\delta \in (0, e^{-\sfrac{1}{2}}/2)$.
+    for all $\varepsilon \in (0, 1)$ and $\delta \in (0, 1/2)$.
 \end{theorem}

### 2024-07-07 paper/plots/convergence.py
migrate experiments to simplified code

@@ -1,10 +1,10 @@
 import __context__
 
 import numpy as np
-import scipy as sp
 import matplotlib.pyplot as plt
-from src.simple import spectral_density, spectral_transformation, form_spectral_density, gaussian_kernel
-from src.matrices import hamiltonian
+from algorithms.chebyshev_nystrom import chebyshev_nystrom
+from algorithms.helpers import spectral_transformation, form_spectral_density, gaussian_kernel
+from matrices.electronic_structure import hamiltonian
 
 np.random.seed(0)
 
@@ -34,11 +34,11 @@ baseline = form_spectral_density(eigvals_st, t, kernel)
 
 error = np.empty((3, len(n_Vec_list)))
 for j, n_Vec in enumerate(n_Vec_list):
-    estimate = spectral_density(A_st, t, m, n_Vec, 0, kernel)
+    estimate = chebyshev_nystrom(A_st, t, m, n_Vec, 0, kernel)
     error[0, j] = 2 * np.mean(np.abs(estimate - baseline))
-    estimate = spectral_density(A_st, t, m, n_Vec // 2, n_Vec // 2, kernel)
+    estimate = chebyshev_nystrom(A_st, t, m, n_Vec // 2, n_Vec // 2, kernel)
     error[1, j] = 2 * np.mean(np.abs(estimate - baseline))
-    estimate = spectral_density(A_st, t, m, 0, n_Vec, kernel)
+    estimate = chebyshev_nystrom(A_st, t, m, 0, n_Vec, kernel)
     error[2, j] = 2 * np.mean(np.abs(estimate - baseline))
 
 for i in range(3):

### 2024-07-07 paper/plots/distribution.py

@@ -1,10 +1,10 @@
 import __context__
 
 import numpy as np
-import scipy as sp
 import matplotlib.pyplot as plt
-from src.simple import spectral_density, spectral_transformation, form_spectral_density, gaussian_kernel
-from src.matrices import hamiltonian
+from algorithms.chebyshev_nystrom import chebyshev_nystrom
+from algorithms.helpers import spectral_transformation, form_spectral_density, gaussian_kernel
+from matrices.electronic_structure import hamiltonian
 
 np.random.seed(0)
 
@@ -38,7 +38,7 @@ for j, sigma in enumerate(sigma_list):
     m = int(16 / sigma)  # Double-check this, i.e. use error estimate for this
 
     for i, (n_Psi, n_Omega) in enumerate(zip(n_Psi_list, n_Omega_list)):
-        estimate = spectral_density(A_st, t, m, n_Psi, n_Omega, kernel)
+        estimate = chebyshev_nystrom(A_st, t, m, n_Psi, n_Omega, kernel)
         error[i, j] = 2 * np.mean(np.abs(estimate - baseline))
 
 for i in reversed(range(len(n_Psi_list))):

### 2024-07-07 paper/plots/interpolation.py

@@ -2,8 +2,9 @@ import __context__
 
 import numpy as np
 import matplotlib.pyplot as plt
-from src.matrices import hamiltonian
-from src.simple import spectral_density, spectral_transformation, form_spectral_density, gaussian_kernel
+from matrices.electronic_structure import hamiltonian
+from algorithms.chebyshev_nystrom import chebyshev_nystrom
+from algorithms.helpers import spectral_transformation, form_spectral_density, gaussian_kernel
 
 np.random.seed(0)
 
@@ -33,11 +34,11 @@ baseline = form_spectral_density(eigvals_st, t, kernel)
 
 error = np.empty((3, len(m_list)))
 for j, m in enumerate(m_list):
-    estimate = spectral_density(A_st, t, m, 0, n_Omega, kernel, consistent=False)
+    estimate = chebyshev_nystrom(A_st, t, m, 0, n_Omega, kernel, consistent=False)
     error[0, j] = 2 * np.mean(np.abs(estimate - baseline))
-    estimate = spectral_density(A_st, t, m, 0, n_Omega, kernel)
+    estimate = chebyshev_nystrom(A_st, t, m, 0, n_Omega, kernel)
     error[1, j] = 2 * np.mean(np.abs(estimate - baseline))
-    estimate = spectral_density(A_st, t, m, 0, n_Omega, kernel, nonnegative=True)
+    estimate = chebyshev_nystrom(A_st, t, m, 0, n_Omega, kernel, nonnegative=True)
     error[2, j] = 2 * np.mean(np.abs(estimate - baseline))
 
 for i in range(3):

### 2024-07-07 paper/plots/potential.py

@@ -2,7 +2,7 @@ import __context__
 
 import matplotlib.pyplot as plt
 
-from src.matrices import regular_grid, periodic_gaussian_well
+from matrices.electronic_structure import regular_grid, periodic_gaussian_well
 
 plt.style.use("paper/plots/stylesheet.mplstyle")
 

### 2024-07-07 paper/plots/zerocheck.py

@@ -2,8 +2,9 @@ import __context__
 
 import numpy as np
 import matplotlib.pyplot as plt
-from src.simple import spectral_density, spectral_transformation, form_spectral_density, gaussian_kernel
-from src.matrices import hamiltonian
+from algorithms.chebyshev_nystrom import chebyshev_nystrom
+from algorithms.helpers import spectral_transformation, form_spectral_density, gaussian_kernel
+from matrices.electronic_structure import hamiltonian
 
 np.random.seed(0)
 
@@ -33,7 +34,7 @@ baseline = form_spectral_density(eigvals_st, t, kernel)
 
 estimate = [baseline]
 for i, kappa in enumerate(kappa_list):
-    estimate.append(spectral_density(A_st, t, m, 0, n_Omega, kernel, kappa=kappa))
+    estimate.append(chebyshev_nystrom(A_st, t, m, 0, n_Omega, kernel, kappa=kappa))
 
 for i in range(3):
     plt.plot(t, estimate[i], color=colors[i], label=labels[i])

### 2024-07-07 paper/re-pro-badge.tex

@@
-This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-TRACE/tree/6f952d3}{6f952d3} at \url{https://github.com/FMatti/Rand-TRACE} on 2024-07-05 at 12:56:53 UTC.
+This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-TRACE/tree/ce5dd53}{ce5dd53} at \url{https://github.com/FMatti/Rand-TRACE} on 2024-07-07 at 11:46:28 UTC.

### 2024-07-08 paper/analysis.tex
Updates from Overleaf

--- 2024-07-07 paper/analysis.tex
@@
 \color{black}
+
+\textcolor{red}{Moved from below: The Girard-Hutchinson estimator distinguishes itself for its simplicity.
+However, its bound does not take into account the structure of $\mtx{B}(t)$.
+The $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ requirement implies that if we aim to increase the accuracy of the estimate by one digit, we would need to increase the number of queries by a factor of $100$.
+}
 
@@
 
-??? START HERE ???`
+??? START HERE ???
+\color{blue}
 \begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
-    For symmetric $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ depending continuously on $t \in [a, b]$, consider the .
-Girard-Hutchinson estimator~\eqref{equ:hutchinson-trace-estimator} with an $n\times n_{\mtx{\Psi}}$ Gaussian random matrix $\mtx{\Psi}$.
-Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ the error bound
-    \begin{equation}
+    For a symmetric matrix $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ that depends continuously on $t \in [a, b]$, consider the  Girard-Hutchinson estimator~\eqref{equ:hutchinson-trace-estimator} with an $n\times n_{\mtx{\Psi}}$ Gaussian random matrix $\mtx{\Psi}$.
+Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ the bound
+    \begin{equation} \label{eq:ghbound}
         \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) \right| ~\mathrm{d}t \leq 2^4 \gamma k \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F  ~\mathrm{d}t.
     \end{equation}
-    holds with probability at least $1 - \gamma^{-2k}$
-
-    In particular, for any $\varepsilon > 0$ we have that $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$ holds
-    with probability at least $1-\delta$ for $\delta \in (0, 1)$ when choosing $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then  for  and  we have .
+    holds with probability at least $1 - \gamma^{-2k}$.
+    In particular, given $\varepsilon > 0$ and $\delta \in (0, 1)$, the bound $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$ holds
+    with probability at least $1-\delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$.
 \end{theorem}
 
-The idea behind the proof is to bound the higher order moments of the residual $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ for fixed values of $t$ and then use Minkowski's integral inequality to extend the results to the continuous parameter case.
+%The idea behind the proof is to bound  to extend the results to the continuous parameter case.
 We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
 \begin{proof} 
-    We define the residual of the 1-query Girard-Hutchinson estimate 
+The proof proceeds by bounding moments for fixed values of $t$ and then use Minkowski's integral inequality.
+We start by considering the residual of the 1-query Girard-Hutchinson estimator for a fixed matrix $\mtx{B}$:
     \begin{equation}
-        r(\mtx{B}(t), \vct{\psi}) = \Trace(\mtx{B}(t)) - \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi}
+        r(\mtx{B}, \vct{\psi}) := \Trace(\mtx{B}) - \vct{\psi}^{\top} \mtx{B} \vct{\psi}
     \end{equation}
-    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$.
-
-    First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence.
-From the proof of \cite[lemma 3]{cortinovis-2022-randomized-trace} we know that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm.
-Thus, by \cite[theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
-    \begin{align}
+%    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers
+for a Gaussian random vector $\vct{\psi}$.
+%
+%    First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence.
+The proof of \cite[Lemma 3]{cortinovis-2022-randomized-trace} shows that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm.
+By \cite[Theorem 2.3]{Boucheron2013} this implies for every $k \in \mathbb{N}$ that
+    \begin{align*}
         \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
-        &\leq k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k} \notag \\
-        &= k! 2^{4 k} \lVert \mtx{B} \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _2^{2 k}.
-    \end{align}
-    Because $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
-    \begin{equation}
-        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
-    \end{equation}
-    Stirling's approximation \cite{robbins-1955-remark-stirling} bonuds $(2 k)!
+        &\leq k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k}  \\
+        &\le \big( k! 2^{4 k} + (2 k)! 2^{6 k}\big) \lVert \mtx{B} \rVert _F^{2 k} 
+        \le \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
+    \end{align*}
+%     Because $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
+%     \begin{equation}
+%         \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right] \leq 
+%     \end{equation}
+    Stirling's approximation \cite{robbins-1955-remark-stirling} gives $(2 k)!
 < 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$.
-Consequently, the moments $\mathbb{E}^{k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{k} \right] \right)^{\sfrac{1}{k}}$ of $r(\mtx{B}, \vct{\psi})$ are limited by
+Consequently, the $L^{2k}$ norm $\mathbb{E}^{2k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{2k} \right] \right)^{\sfrac{1}{2k}}$ of $r(\mtx{B}, \vct{\psi})$ is bounded by
     \begin{equation}
@@
     \end{equation}
-    since it can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, due to the monotonous decrease of this expression in $k \in \mathbb{N}$.
+    using the inequality $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, which can be verified using the monotonic decrease of the left-hand side with respect to $k \in \mathbb{N}$.
 
-    Now we make the transition to the continuous.
-Following \cite{kressner-2023-randomized-lowrank}, we can show that $r(\mtx{B}(t), \vct{\psi})$ is measurable.
-Therefore, Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting:
+    To address the parameter-dependent case, we first note that the continuity assumption implies that $r(\mtx{B}(t), \vct{\psi})$ is measurable.
+Therefore, we can apply Minkowski's integral inequality~\cite[Theorem 202]{hardy-1952-inequalities} to conclude from~\refequ{equ:hutchinson-trace-onequery-fixed} that
     \begin{equation}
@@
     \end{equation}
-    Consequently, by Markov's inequality, for any $k \in \mathbb{N}$ and $\gamma \geq 1$, it holds with probability at least $1 - \gamma^{-2 k}$
+    Consequently, by Markov's inequality,  the inequality 
     \begin{equation} 
@@
     \end{equation}
+    holds for any $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$.
 
-    To extend the bound to the $n_{\mtx{\Psi}}$-query quadratic trace estimator, we use a technique from the proof of \cite[theorem 1]{cortinovis-2022-randomized-trace}. Let
+    To extend the bound~\eqref{equ:hutchinson-trace-onequery-uniform} from $1$ to $n_{\mtx{\Psi}}$ queries, we use a standard embedding trick (see, e.g.,~\cite[Theorem 1]{cortinovis-2022-randomized-trace}).
+Letting
     \begin{equation}
@@
             \vct{\psi}_{n_{\mtx{\Psi}}}
-        \end{pmatrix}
+        \end{pmatrix},
     \end{equation}
-    where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors.
+    we note that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) = r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold.
+Thus, the claimed probabilistic bound~\eqref{eq:ghbound} follows from applying~\eqref{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$.
+%     Then 
+%     where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors.
 Then $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$, and
-    \begin{equation}
-        r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}}) = \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j.
-    \end{equation}
-    Hence, by applying \refequ{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$ we conclude that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
-    \begin{equation}
-        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j \right| ~ \mathrm{d}t
-        \leq 2^4 k \gamma \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
-    \end{equation}
+%     \begin{equation}
+%         r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}}) = \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j.
+%     \end{equation}
+%     Hence, by applying \refequ{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$ we conclude that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
+%     \begin{equation}
+%         \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j \right| ~ \mathrm{d}t
+%         \leq 2^4 k \gamma \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
+%     \end{equation}
 
@@
 
-    Setting $\gamma = e$ and choosing $k = \lfloor \log(\delta^{-1}) / 2 \rfloor$, where $\lfloor x \rfloor$ is the floor function which returns the largest integer smaller than or equal to $x$, we get that for all $\delta \in (0, 1)$ the second part of the theorem follows when taking $n_{\mtx{\Psi}} = \lceil (2^3 e \varepsilon^{-1} \log(\delta^{-1}))^2 \rceil$, where $\lceil x \rceil$ is the ceiling function which returns the smallest integer larger than or equal to $x$.
-
+    Setting $\gamma = e$ and choosing $k = \lfloor \log(\delta^{-1}) / 2 \rfloor$,
+    %where $\lfloor x \rfloor$ is the floor function which returns the largest integer smaller than or equal to %$x$,
+    %we get that for all $\delta \in (0, 1)$
+    the second part of the theorem follows when taking $n_{\mtx{\Psi}} = \lceil (2^3 e \varepsilon^{-1} \log(\delta^{-1}))^2 \rceil$.
+    %where $\lceil x \rceil$ is the ceiling function which returns the smallest integer larger than or equal to $x$.
     %Consequently, with probability $1 - \delta$ we have
@@
 \end{proof}
+\color{black}
 
@@
 
-Compared to the constant case \cite[lemma 2.1]{meyer-2021-hutch-optimal} we need to choose the number of queries $n_{\mtx{\Psi}}$ by a factor of $\log(\delta^{-1})$ larger to achieve the same guarantee.
-
-The Girard-Hutchinson estimator distinguishes itself for its simplicity.
-However, its bound does not take into account the structure of $\mtx{B}(t)$.
-The $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ requirement implies that if we aim to increase the accuracy of the estimate by one digit, we would need to increase the number of queries by a factor of $100$.
+Compared to the analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal},
+the result of~\Refthm{thm:hutchinson} requires one to choose the number of queries $n_{\mtx{\Psi}}$ by a factor $\log(\delta^{-1})$ larger to yield the same guarantee.
 

### 2024-07-08 paper/intro.tex
Updates from Overleaf

@@
 Evaluating this matrix function \emph{exactly} requires the diagonalization of $\mtx{A}$, which does not seem to yield any gains compared to the original problem of computing its spectral distribution.
-The crucial point is that we can now \emph{estimate} the trace by matrix-vector products with $\mtx{B}(t)$, which -- thanks to the smoothness of $g_\sigma$ -- can be well approximated by a few matrix-vector products with $A$.
+The crucial point is that we can now \emph{estimate} the trace by matrix-vector products with $\mtx{B}(t)$, which -- thanks to the smoothness of $g_\sigma$ -- can be well approximated by a few matrix-vector products with $\mtx{A}$.
 
-For a \emph{constant} matrix $B$, one of the most popular trace estimators is the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} along with variance reduction techniques~\cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, chen-2023-krylovaware-stochastic, epperly-2024-xtrace-making}.
-Suitable extensions to parameter-dependent matrices have been considered, e.g., in~\cite{lin-2017-randomized-estimation,chen-2023-krylovaware-stochastic}, but we are not aware of an analysis providing insight and justification of these extensions.
+For a \emph{constant} matrix $\mtx{B}$, one of the most popular trace estimators is the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} along with variance reduction techniques~\cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, chen-2023-krylovaware-stochastic, epperly-2024-xtrace-making}.
+Suitable extensions to parameter-dependent matrices have been considered, e.g., in~\cite{lin-2017-randomized-estimation,chen-2023-krylovaware-stochastic}, but we are not aware of an analysis providing insight and justification to these extensions.
 In passing, we note that dynamic trace estimation~\cite{dharangutte-2021-dynamic-trace,woodruff-2022-optimal-query} is an efficient technique for \textcolor{red}{subsequently ???

### 2024-07-08 paper/analysis.tex

@@
 \paragraph{Nyström estimator} Alternatively, the trace of a symmetric matrix whose singular values decay quickly can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström approximation \cite{gittens-2013-revisiting-nystrom}
+
 \begin{equation}
@@
     \begin{align*}
+
         \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
@@
 %    if $n_{\mtx{\Psi}} \geq 2^{10} \varepsilon^{-2} \log(\delta^{-1})^2 (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ with $\mu(t) = \lVert \mtx{B}(t) \rVert _2 / \Trace(\mtx{B}(t))$.
-Compared to the constant matrix equivalent \cite[remark 2]{cortinovis-2022-randomized-trace} we mainly observe a larger constant and an additional factor of $\log(\delta^{-1})$.
+Compared to the constant matrix equivalent \cite[Remark 2]{cortinovis-2022-randomized-trace} we mainly observe a larger constant and an additional factor of $\log(\delta^{-1})$.
 %\end{remark}
 
-Compared to the analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal},
-the result of~\Refthm{thm:hutchinson} requires one to choose the number of queries $n_{\mtx{\Psi}}$ by a factor $\log(\delta^{-1})$ larger to yield the same guarantee.
+Compared to the analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, the result of~\Refthm{thm:hutchinson} requires one to choose the number of queries $n_{\mtx{\Psi}}$ by a factor $\log(\delta^{-1})$ larger to yield the same guarantee.
 

### 2024-07-08 paper/intro.tex

@@
 For a \emph{constant} matrix $\mtx{B}$, one of the most popular trace estimators is the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} along with variance reduction techniques~\cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, chen-2023-krylovaware-stochastic, epperly-2024-xtrace-making}.
-Suitable extensions to parameter-dependent matrices have been considered, e.g., in~\cite{lin-2017-randomized-estimation,chen-2023-krylovaware-stochastic}, but we are not aware of an analysis providing insight and justification to these extensions.
-In passing, we note that dynamic trace estimation~\cite{dharangutte-2021-dynamic-trace,woodruff-2022-optimal-query} is an efficient technique for \textcolor{red}{subsequently ???
+Suitable extensions to parameter-dependent matrices have been considered, e.g., in~\cite{lin-2017-randomized-estimation,chen-2023-krylovaware-stochastic}, but we are not aware of an analysis providing insight and justification of these extensions.
+In passing, we note that dynamic trace estimation~\cite{dharangutte-2024-dynamic-trace,woodruff-2024-optimal-query} is an efficient technique for \textcolor{red}{subsequently ???
 true?
 ???} estimating the traces of matrices $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ for wich the increments $\mtx{B}(t_{i+1}) - \mtx{B}(t_i)$ are relatively small in norm.
-The potential of dynamic trace estimation appears to be limited in our setting because $\mtx{B}(t)$ may change rapidly close to eigenvalues, with
-$g_\sigma$ approximating a Dirac delta function.
+The potential of dynamic trace estimation appears to be limited in our setting because $\mtx{B}(t)$ may change rapidly close to eigenvalues, with $g_{\sigma}$ approximating a Dirac delta function.
 
@@
 
-\paragraph{New contributions} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}.
+\paragraph{New contributions.} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}.
 Interestingly, we can reuse the same randomization for computing the estimate at ea<ch value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations.
@@
 
-\paragraph{Structure} In \refsec{sec:analysis} we will analyze three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter dependent matrices.
+\paragraph{Structure.} In \refsec{sec:analysis} we will analyze three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter dependent matrices.
 We propose and analyze an algorithm for using the aforementioned estimators to approximate the spectral density of real symmetric matrices in \refsec{sec:application}.
@@
 
-\paragraph{Reproducibility} \textcolor{red}{This paragraph looks odd.
+\paragraph{Reproducibility.} \textcolor{red}{This paragraph looks odd.
 Is this the wording required by SISC?

### 2024-07-08 paper/re-pro-badge.tex

@@
-This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-TRACE/tree/ce5dd53}{ce5dd53} at \url{https://github.com/FMatti/Rand-TRACE} on 2024-07-07 at 11:46:28 UTC.
+This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-TRACE/tree/04376a9}{04376a9} at \url{https://github.com/FMatti/Rand-TRACE} on 2024-07-07 at 15:28:11 UTC.

### 2024-07-08 paper/re-pro-badge.tex
automatically reproduce paper

@@
-This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-TRACE/tree/04376a9}{04376a9} at \url{https://github.com/FMatti/Rand-TRACE} on 2024-07-07 at 15:28:11 UTC.
+This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-TRACE/tree/5832e99}{5832e99} at \url{https://github.com/FMatti/Rand-TRACE} on 2024-07-08 at 05:53:37 UTC.

### 2024-07-08 paper/re-pro-badge.tex
automatically reproduce paper

@@
-This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-TRACE/tree/5832e99}{5832e99} at \url{https://github.com/FMatti/Rand-TRACE} on 2024-07-08 at 05:53:37 UTC.
+This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-TRACE/tree/a000fde}{a000fde} at \url{https://github.com/FMatti/Rand-TRACE} on 2024-07-08 at 06:45:20 UTC.

### 2024-07-08 paper/analysis.tex

@@
     \begin{align*}
-
         \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
@@
 
-Compared to the analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, the result of~\Refthm{thm:hutchinson} requires one to choose the number of queries $n_{\mtx{\Psi}}$ by a factor $\log(\delta^{-1})$ larger to yield the same guarantee.
+Compared to the analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, the result of~\refthm{thm:hutchinson} requires one to choose the number of queries $n_{\mtx{\Psi}}$ by a factor $\log(\delta^{-1})$ larger to yield the same guarantee.
 

### 2024-07-08 paper/analysis.tex
pull feedback from Daniel

@@
 
-\paragraph{Nyström estimator} Alternatively, the trace of a symmetric matrix whose singular values decay quickly can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström approximation \cite{gittens-2013-revisiting-nystrom}
+\paragraph{Nyström estimator}
+
+Write:
+\begin{itemize}
+ \item fast singular value decay, intuitively sufficient to consider a low-rank approximation
+ \item motivated by our application we only consider symmetric positive semi-definite case here (smoothing kernels are positive) for which Nystroem approximation is the right choice
+\end{itemize}
+
+\textcolor{red}{Moved from below:
+For symmetric positive semi-definite matrices whose singular values decay rapidly, the Nyström estimate \refequ{equ:nystrom-trace-estimator} can be significantly better than the Girard-Hutchinson estimate.
+This is reflected in the following theorem.}
+
+Alternatively, the trace of a symmetric matrix whose singular values decay quickly can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström approximation \cite{gittens-2013-revisiting-nystrom}
 
@@
 \end{equation}
+
+\textcolor{red}{Moved from below: The downside of the Nyström estimator is that it will -- in general -- not work well enough on matrices whose singular values do not decay quickly, for example $\mtx{B}(t) \equiv \mtx{I}_n$.}
 
@@
 \end{proof}
-\color{black}
 
-%\begin{remark}
-%    If additionally $\mtx{B}(t)$ is nonzero and symmetric positive semi-definite for all $t \in [a, b]$, then it follows from the proof of \refthm{thm:hutchinson} that with probability at least $1 - \delta$
-%    \begin{equation}
-%        \int_{a}^{b} \frac{| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) |}{\Trace(\mtx{B}(t))} ~\mathrm{d}t < \varepsilon,
-%    \end{equation}
-%    if $n_{\mtx{\Psi}} \geq 2^{10} \varepsilon^{-2} \log(\delta^{-1})^2 (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ with $\mu(t) = \lVert \mtx{B}(t) \rVert _2 / \Trace(\mtx{B}(t))$.
-Compared to the constant matrix equivalent \cite[Remark 2]{cortinovis-2022-randomized-trace} we mainly observe a larger constant and an additional factor of $\log(\delta^{-1})$.
-%\end{remark}
-
-Compared to the analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, the result of~\refthm{thm:hutchinson} requires one to choose the number of queries $n_{\mtx{\Psi}}$ by a factor $\log(\delta^{-1})$ larger to yield the same guarantee.
+Compared to the analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, the number of queries $n_{\mtx{\Psi}}$ needed by~\refthm{thm:hutchinson} is roughly a factor $\log(\delta^{-1})$ larger.
 
@@
 
-For symmetric positive semi-definite matrices whose singular values decay rapidly, the Nyström estimate \refequ{equ:nystrom-trace-estimator} can be significantly better than the Girard-Hutchinson estimate.
-This is reflected in the following theorem.
+The following result confirms the inituition that the Nystr\"om approximation is very well suited in the presence of decaying singular values: If the nuclear norm error to the best rank-$r$ approximation to $B(t)$ is bounded by $\varepsilon$ for every $t$ then the trace of the Nystr\"om approximation (with a bit of oversampling) results, with high probability, in an error that is only modestly larger than $\varepsilon$.
 
 \begin{theorem}{Nyström estimator for parameter-dependent matrices}{nystrom}
-    Suppose $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ is continuous and positive semi-definite in $t \in [a, b]$.
-Let $r \geq 2$ and $n_{\mtx{\Omega}} \geq r + 4$ be two integers.
-Then for all $\gamma \geq 1$ with probability at least $1 - \gamma^{-(n_{\mtx{\Omega}} - r)}$
-    \begin{equation}
-        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right| ~\mathrm{d}t
+    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuous for $t \in [a, b]$.
+For integers $r \geq 2$ and $n_{\mtx{\Omega}} \geq r + 4$, consider the Nystr\"om approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ defined in~\eqref{equ:nystrom-approximation} with an $n\times n_{\mtx{\Omega}}$ Gaussian random matrix $\mtx{\Omega}$.
+Then for any $\gamma \geq 1$ the bound
+    \[
+        \int_{a}^{b} \big| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \big| ~\mathrm{d}t
         \leq \gamma^2 (1 + r) \int_{a}^{b} \sum_{i = r+1}^{n} \sigma_i(\mtx{B}(t)) ~\mathrm{d}t.
-    \end{equation}
+    \]
+    holds with probability at least $1 - \gamma^{-(n_{\mtx{\Omega}} - r)}$.
 \end{theorem}
-
 \begin{proof}
-    Since $\mtx{B}(t)$ is positive semi-definite for all $t$, then by \cite[Lemma 2.1]{frangella-2023-randomized-nystrom}, $\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ is also positive semi-definite for all $t$.
-Therefore, for fixed $t \in [a, b]$
-    \begin{equation}
-        \left| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right|
-        = \left| \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right|
-        = \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _{\ast},
-    \end{equation}
-    where $\lVert \cdot \rVert _{\ast}$ represents the nuclear norm.  From the proof of \cite[Corollary 8.2]{tropp-2023-randomized-algorithms} it follows that
-    \begin{equation}
-        %\lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _2^2 \leq \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _F^2
-        \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{\mtx{B}(t)^{\sfrac{1}{2}} \mtx{\Omega}}) \mtx{B}(t)^{\sfrac{1}{2}} \rVert _F^2.
-    \end{equation}
-    Hence, by \cite[Theorem 9]{kressner-2023-randomized-lowrank}, with $n_{\mtx{\Omega}} = p + r$, the result follows by noting $\sigma_i(\mtx{B}(t)^{\sfrac{1}{2}})^{2} = \sigma_i(\mtx{B}(t))$.
-
+Using that $\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ is positive semi-definite (see, e.g.,~\cite[Lemma 2.1]{frangella-2023-randomized-nystrom}) we obtain
+    \begin{align*}
+        \big| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \big|
+        & = \big| \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \big|
+        = \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _{\ast} \\
+        & = \lVert (\mtx{I}_n - \mtx{\Pi}_{\mtx{B}(t)^{\sfrac{1}{2}} \mtx{\Omega}}) \mtx{B}(t)^{\sfrac{1}{2}} \rVert _F^2,
+    \end{align*}
+    where $\lVert \cdot \rVert _{\ast}$ denotes the nuclear norm and we used a \textcolor{red}{result from the proof of \cite[Corollary 8.2]{tropp-2023-randomized-algorithms} ???
+Is this really the original source.
+Have a look whether David P mentions this somewhere?
+Also, this may be contained in the original Gittens Mahoney paper but I am not sure ???} in the last equality.
+This allows us to apply \textcolor{red}{Theorem 9 ???
+be careful that the theorem numbering may be different when the paper is published in NLAA ???} from \cite{kressner-2023-randomized-lowrank}, which immediately implies the statement claimed by the theorem.
 \end{proof}
 
-Compared to the constant matrix case \cite[Theorem 8.1]{tropp-2023-randomized-algorithms}, our bound also scales with the best rank-$r$ approximation error in the nuclear norm, albeit with a factor approximately proportional to $r$, which cannot be compensated for with large enough oversampling.
+\color{black}
 
-The downside of the Nyström estimator is that it will -- in general -- not work well enough on matrices whose singular values do not decay quickly, for example $\mtx{B}(t) \equiv \mtx{I}_n$.
+Compared to the constant case~\cite[Theorem 8.1]{tropp-2023-randomized-algorithms}, the bound of Theorem~\ref{thm:nystrom} features an additional factor $1+r$, which cannot be compensated with oversampling.
 

### 2024-07-08 paper/paper.tex
Updates from Overleaf
pull feedback from Daniel

@@
 \usepackage{stylesheet}
+
+% to speed up compilation speed a bit while drafting
+\pdfcompresslevel=0
+\pdfobjcompresslevel=0
 
@@
 \begin{itemize}
- \item Capitalize references to algorithms, lemma etc (that is, Lemma 3.2 instead of lemma 3.2 [Lin17, Algorithm 1] instead of [Lin17, algorithm 1]), etc. \textcolor{green}{Done.}
- \item Add full stops after paragraph titles. \textcolor{green}{Done.}
  \item It is not great that the references occupy so much space (more than 4 pages); consider using a more condensed style.
@@
 \textcolor{green}{Journal abbreviations, max 2 authors, omit superfluous fields $\to$ 3 pages.}
- \item Please include Acknowledgements as indicated in the comment. \textcolor{green}{Done.}
+ Journal abbreviations and removing superfluous fields is good. max 2 authors is not coming in math, please reverse this. The text style is a bit too generous but no need to change this now.
 \end{itemize}
@@
 \input{spectraldensity.tex}
-\input{numresults.tex}
+%\input{numresults.tex}
 

### 2024-07-08 paper/analysis.tex
address feedback

@@
 The proof of \cite[Lemma 3]{cortinovis-2022-randomized-trace} shows that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm.
-By \cite[Theorem 2.3]{Boucheron2013} this implies for every $k \in \mathbb{N}$ that
+By \cite[Theorem 2.3]{boucheron-2013-concentration-inequalities} this implies for every $k \in \mathbb{N}$ that
     \begin{align*}
@@
 Have a look whether David P mentions this somewhere?
-Also, this may be contained in the original Gittens Mahoney paper but I am not sure ???} in the last equality.
+Also, this may be contained in the original Gittens Mahoney paper but I am not sure ???} \textcolor{green}{The original is in Gittens' arXiv preprint \cite[Theorem 1]{gittens-2011-spectral-norm} but also in Gittens/Mahoney \cite[Equation 4]{gittens-2013-revisiting-nystrom}} in the last equality.
 This allows us to apply \textcolor{red}{Theorem 9 ???

### 2024-07-08 paper/paper.tex
address feedback

@@
 \textcolor{green}{Journal abbreviations, max 2 authors, omit superfluous fields $\to$ 3 pages.}
- Journal abbreviations and removing superfluous fields is good. max 2 authors is not coming in math, please reverse this. The text style is a bit too generous but no need to change this now.
+ Journal abbreviations and removing superfluous fields is good.
+max 2 authors is not coming in math, please reverse this.
+\textcolor{green}{Done.} The text style is a bit too generous but no need to change this now.
+\textcolor{green}{Ok!
+Let's talk about it at some point.}
 \end{itemize}
@@
 
-\textcolor{red}{??? Need to check carefully that this result is not known ???}
+\textcolor{red}{???
+Need to check carefully that this result is not known ???} \textcolor{green}{I haven't found anything so far.
+The fact that also Bujanović/Grubišić/Kressner/Lam in Lemma 20 use a very similar technique, but merely to bound $\Gamma(\frac{p+1}{2})$ makes me suspect there's not much, except for Stirling approximation at the cost of some simplicity.
+Will keep investigating.}
 

### 2024-07-08 paper/spectraldensity.tex
address feedback

@@
 However, doing this exactly would require us to first diagonalize $\mtx{A}$, which is often a prohibitively expensive operation.
-For algorithmic reasons (see \refsec{subsubsec:chebyshev-nystrom-implementation}), we seek for approximations of this matrix function in terms of affine expansions \todo{explain more}, of which the Chebyshev expansion was determined to be most suitable due to its approximation error guarantees and ease of manipulation.
+For algorithmic reasons (see \refsec{subsubsec:chebyshev-nystrom-implementation}), we seek for approximations of this matrix function in terms of affine linear expansions, of which the Chebyshev expansion was determined to be most suitable due to its approximation error guarantees and ease of manipulation.
 In every $t \in \mathbb{R}$ we expand

### 2024-07-08 paper/re-pro-badge.tex
automatically reproduce paper

@@
-This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-TRACE/tree/a000fde}{a000fde} at \url{https://github.com/FMatti/Rand-TRACE} on 2024-07-08 at 06:45:20 UTC.
+The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/68c3ef6}{68c3ef6} on 2024-07-08 at 18:58:43 UTC.

### 2024-07-09 paper/analysis.tex
pull changes from overleaf

@@
 
-\color{black}
-
 Compared to the constant case~\cite[Theorem 8.1]{tropp-2023-randomized-algorithms}, the bound of Theorem~\ref{thm:nystrom} features an additional factor $1+r$, which cannot be compensated with oversampling.
@@
 
-Just as in the constant matrix case, our goal is now to prove that the Nyström++ estimator \refequ{equ:nystrompp-trace-estimator} achieves an $\varepsilon$-relative error with just $\mathcal{O}(\varepsilon^{-1})$ random vectors used to construct the estimate.
-To do so, we mimic the proof of a similar bound for the Hutch++ estimator \cite[Theorem 3.1]{meyer-2021-hutch-optimal}.
+\refthm{thm:hutchinson} has established a relative $\varepsilon$-error for the Girard-Hutchinson estimator with $\mathcal{O}(\varepsilon^{-2})$ queries.
+The aim of this section is to show that the Nyström++ estimator~\refequ{equ:nystrompp-trace-estimator} improves this to $\mathcal{O}(\varepsilon^{-1})$ matrix-vector products with random vectors, \emph{without} requiring any assumption on the singular values of $\mtx{B}(t)$.
+This parallels existing results for the constant case
+for Hutch++\cite[Theorem 3.1]{meyer-2021-hutch-optimal} and Nyström++\textcolor{red}{??? Include specific result ???}.
+
+\color{black}
+
+To do so, we mimic the proof of a similar bound for the Hutch++ estimator .
 Correspondingly, we will need to show that with $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2})$ random vectors, the Nyström approximation verifies $\int \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)\rVert _F~\mathrm{d}t \leq \varepsilon \int \Trace(\mtx{B}(t))~\mathrm{d}t$, regardless of the structure of the matrix $\mtx{B}(t)$.
@@
     for independent standard Gaussian random vectors $\vct{\omega}_1 \in \mathbb{R}^{k}$ and $\vct{\omega}_2 \in \mathbb{R}^{m}$.
-Both moments can be bound using \reflem{lem:spectral-norm-moment-vector}, one of which with $\mtx{A} = \mtx{I}_k$, to get
+\textcolor{red}{Doesn't A.1 give a tighter bound for the first term?} Both moments can be bound using \reflem{lem:spectral-norm-moment-vector}, one of which with $\mtx{A} = \mtx{I}_k$, to get
     \begin{align}
@@
 
+\color{blue}
+
 \begin{lemma}{Nyström approximation for parameter-dependent matrices}{nystrom}
     Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-Then its Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ with Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ for \emph{even} $n_{\mtx{\Omega}} \geq 4$ verifies with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
+Consider the Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ with a Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and \emph{even} $n_{\mtx{\Omega}} \geq 4$.
+    Then for any $\gamma \geq 1$ the bound
     \begin{equation}
@@
     \end{equation}
-    for any $\gamma \geq 1$ and a universal constant $c > 0$.
-In particular, if we choose $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1}))$, then with probability at least $1-\delta$ for $\delta \in (0, 1)$ and $\varepsilon > 0$ we have $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$.
+        holds with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$ for 
+        a universal constant $c > 0$.
+In particular, given $\varepsilon > 0$ and $\delta \in (0, 1)$, the bound $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$ holds with probability at least $1-\delta$ if $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1}))$.
 \end{lemma}

### 2024-07-09 paper/re-pro-badge.tex
automatically reproduce paper

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/68c3ef6}{68c3ef6} on 2024-07-08 at 18:58:43 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/3f5eb52}{3f5eb52} on 2024-07-09 at 06:06:30 UTC.

### 2024-07-10 paper/analysis.tex
Updates from Overleaf

@@
 \color{black}
+
+\subsection{Methods}
 
@@
 
-\subsection{Girard-Hutchinson estimator for parameter-dependent matrices}
+\subsection{Error bounds}
+
+\subsubsection{Girard-Hutchinson estimator}
 \label{subsec:hutchinson}
 
-We first analyze the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}.
-We bound its $L^1$-error in terms of the $L^1$-norm of the Frobenius norm $\lVert \cdot \rVert _F$ of the matrix whose trace we approximate.
+\color{blue}
 
-??? START HERE ???
-\color{blue}
+We start our analysis of the methods presented above by deriving an $L^1$-error bound for the Girard-Hutchinson estimator~\refequ{equ:hutchinson-trace-estimator} that is relative to the $L^1$-norm of the Frobenius norm $\lVert \mtx{B}(t) \rVert _F$.
+
 \begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
@@
 
-\subsection{Nyström approximation for parameter-dependent matrices}
+\subsubsection{Nyström approximation}
 \label{subsec:nystrom}
@@
 
-Compared to the constant case~\cite[Theorem 8.1]{tropp-2023-randomized-algorithms}, the bound of Theorem~\ref{thm:nystrom} features an additional factor $1+r$, which cannot be compensated with oversampling.
+Compared to the constant case~\cite[Theorem 8.1]{tropp-2023-randomized-algorithms}, the bound of Theorem~\ref{thm:nystrom} features an additional factor $1+r$ that cannot be compensated with oversampling.
 
@@
 
-\refthm{thm:hutchinson} has established a relative $\varepsilon$-error for the Girard-Hutchinson estimator with $\mathcal{O}(\varepsilon^{-2})$ queries.
+\refthm{thm:hutchinson} has shown that the the Girard-Hutchinson estimator achieves a relative error $\varepsilon$ when using $\mathcal{O}(\varepsilon^{-2})$ queries.
 The aim of this section is to show that the Nyström++ estimator~\refequ{equ:nystrompp-trace-estimator} improves this to $\mathcal{O}(\varepsilon^{-1})$ matrix-vector products with random vectors, \emph{without} requiring any assumption on the singular values of $\mtx{B}(t)$.
@@
 We will do this in a sequence of three lemmas.
-
-For reasons which will become clear in the proof of the final and most interesting of these lemmas (\reflem{lem:nystrom}), we will need to bound $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]$ for a constant matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and a standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$.
-The corresponding result for the Frobenius norm $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _F^2 \right]$ has already been shown in \cite[Lemma 7]{kressner-2023-randomized-lowrank}.
-Several ways of separately bounding the spectral norm moments of the Wishart matrix $\mtx{\Omega} \mtx{\Omega}^{\top}$, i.e.
-when $\mtx{A} = \mtx{I}_m$ already exist, however, all of them are not sufficient for our purposes because they scale with $m$ \cite{chen-2005-condition-numbers, edelman-1988-eigenvalues-condition, james-1964-distributions-matrix}.
-In \cite[Lemma B.1]{tropp-2023-randomized-algorithms} a bound is achieved for $p = 2$ and $p = 4$.
-We will now generalize this result for arbitrary $p \in \mathbb{N}$.
-To do so, we first show the bound for Gaussian random vectors and then extend it to Gaussian random matrices in \reflem{lem:spectral-norm-moment}.
-
-\begin{lemma}{Spectral norm moments of non-standard Gaussian random vector}{spectral-norm-moment-vector}
-    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random vector $\vct{\omega} \in \mathbb{R}^{m}$ satisfy for all $k,p \in \mathbb{N}$ 
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2^2 \right]
-        \leq  (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
-    \end{equation}
-    %\begin{equation}
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
-    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
-    %\end{equation}
-    %If we choose 
-\end{lemma}
-
-\begin{remark}
-    In particular, this result can also be used to bound
-    \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2 \right] = \sqrt{\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2^2 \right]} \leq \sqrt{k + p} \cdot \left(\lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F\right)
-    \end{equation}
-    for any $k,p \in \mathbb{N}$.
-\end{remark}
-
-\begin{proof}
-    With the singular value decomposition $\mtx{A} = \mtx{U} \mtx{\Sigma} \mtx{V}^{\top}$ and the unitary invariance of the spectral norm, it can be seen that $\lVert \mtx{A} \mtx{\omega} \rVert _2^2 = \lVert \mtx{\Sigma} \widetilde{\mtx{\omega}} \rVert _2^2$ where $\widetilde{\mtx{\omega}}$ is again a standard Gaussian random vector.
-Hence, it is enough to bound $\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{\Sigma} \mtx{\omega} \rVert _2^2 \right]$ for a diagonal matrix $\mtx{\Sigma} = \operatorname{diag}(\sigma_1, \dots, \sigma_m)$ with $\sigma_1 \geq \dots \geq \sigma_m \geq 0$.
-
-    First, we can rewrite
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2^2 \right]
-        = \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right].
-        \label{equ:matvec-spectral-norm-simplification}
-    \end{equation}
-    Inspired by the proof of \cite[Theorem 1]{cohen-2016-optimal-approximate}, we split the diagonal entries of $\mtx{\Sigma}$ into $\ell = \lceil m/k \rceil$ groups of size $k \geq 1$
-    \begin{equation}
-        \overbrace{\underbrace{\sigma_1, \dots, \sigma_k}_{\leq \sigma_1}}^{\geq \sigma_{k+1}}, \overbrace{\underbrace{\sigma_{k+1}, \dots, \sigma_{2k}}_{\leq \sigma_{k+1}}}^{\geq \sigma_{2k+1}}, \dots, \overbrace{\underbrace{\sigma_{(\ell - 1)k + 1}, \dots, \sigma_{\ell k}}_{\leq \sigma_{(\ell - 1)k + 1}}}^{\geq 0}.
-    \end{equation}
-    If $m$ is not a multiple of $k$, we let $\sigma_i = 0$ for $i > m$. Since $\sigma_1 \geq \dots \geq \sigma_{\ell k} \geq 0$,
-    \begin{equation}
-        \sigma_1^2 = \lVert \mtx{\Sigma} \rVert _2^2
-        \quad \text{and} \quad
-        \sigma_{(i-1)k + 1}^2 \leq \frac{1}{k} \sum_{j=1}^{k} \sigma_{(i-2)k + j}^2, i = 2, \dots, \ell,
-    \end{equation}
-    and therefore
-    \begin{equation}
-        \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 = \lVert \mtx{\Sigma} \rVert _2^2 + \sum_{i=2}^{\ell} \sigma_{(i-1)k + 1}^2  \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \sum_{n=1}^{(\ell - 1)k} \sigma_n^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2.
-        \label{equ:singular-value-group-bound}
-    \end{equation}
-    This allows us to bound
-    \begin{align}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right]
-        &= \mathbb{E}^{\sfrac{p}{2}}\Bigg[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{(i-1)k + j}^2}_{\leq \sigma_{(i-1)k + 1}^2} \omega_{(i-1)k + j}^2 \Bigg] && \text{(separate sum into groups)} \notag \\
-        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 \sum_{j=1}^{k} \omega_{(i-1)k + j}^2 \right] && \text{($\sigma_{(i-1)k + j}^2 \leq \sigma_{(i-1)k + 1}^2$, $\forall i, j$)} \notag \\
-        &\leq \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 ~ \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{(i-1)k + j}^2 \right] && \text{(triangle inequality)} \notag \\
-        %&\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k}{2} + p)}{\Gamma(\frac{k}{2})} \right)^{\sfrac{1}{p}} && \text{(\refequ{equ:singular-value-group-bound} and \cite[Theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
-        &\leq (k + p) \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 && \text{(\reflem{lem:gamma})} \notag \\
-        &\leq (k + p) \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right). && \text{(using \refequ{equ:singular-value-group-bound})}
-    \end{align}
-
-    %The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[Lemma B.3]{tropp-2023-randomized-algorithms}
-    %\begin{align}
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    &= \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}} \notag \\
-    %    &\leq \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{1}{\Gamma(k - r + 2)} \right)^{\frac{1}{k - r + 1}} \left( \frac{k + r}{2} \right) \notag \\
-    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right)
-    %\end{align}
-    %With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\sfrac{1}{n}} \leq \frac{e}{n}$ follow respectively.
-Hence,
-    %\begin{equation}
-    %    \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    \leq \frac{e^2}{k + 1}\sqrt{ \frac{3k}{2} }
-    %    \leq e^2 \sqrt{\frac{3}{2k}}.
-    %    \label{equ:OSE-moment-bound-second}
-    %\end{equation}
-
-\end{proof}
-
-\begin{lemma}{Spectral norm moments of non-standard Gaussian random matrix}{spectral-norm-moment}
-    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all $p \in \mathbb{N}$ 
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
-        \leq  (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
-    \end{equation}
-    %\begin{equation}
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
-    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
-    %\end{equation}
-    %If we choose 
-\end{lemma}
-
-\begin{remark}
-    In particular, this result can also be used to bound
-    \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2 \right] = \sqrt{\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]} \leq \sqrt{k + p} \cdot \left(\sqrt{3} \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F\right)
-    \end{equation}
-    for any $k,p \in \mathbb{N}$. For $p=2$ this bound is slightly less tight than its equivalent in \cite[Lemma B.1]{tropp-2023-randomized-algorithms}.
-\end{remark}
-
-\begin{proof}
-    In the proof of \cite[Lemma B.1]{tropp-2023-randomized-algorithms}, it is shown that Slepian's inequality \cite[Theorem 7.2.1]{vershynin-2018-highdimensional-probability} applied to two appropriately constructed Gaussian random fields yields
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
-        \leq \lVert \mtx{A} \rVert _2^2 \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \vct{\omega}_1 \rVert _2^2 \right] + \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega}_2 \rVert _2^2 \right]
-    \end{equation}
-    for independent standard Gaussian random vectors $\vct{\omega}_1 \in \mathbb{R}^{k}$ and $\vct{\omega}_2 \in \mathbb{R}^{m}$.
-\textcolor{red}{Doesn't A.1 give a tighter bound for the first term?} Both moments can be bound using \reflem{lem:spectral-norm-moment-vector}, one of which with $\mtx{A} = \mtx{I}_k$, to get
-    \begin{align}
-        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
-        &\leq \lVert \mtx{A} \rVert _2^2 (k + p) \left( \lVert \mtx{I}_k \rVert _2^2 + \frac{1}{k} \lVert \mtx{I}_k \rVert _F^2 \right) + (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right) \notag \\
-        &\leq (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
-    \end{align}
-
-    %The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[Lemma B.3]{tropp-2023-randomized-algorithms}
-    %\begin{align}
-    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    &= \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}} \notag \\
-    %    &\leq \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{1}{\Gamma(k - r + 2)} \right)^{\frac{1}{k - r + 1}} \left( \frac{k + r}{2} \right) \notag \\
-    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right)
-    %\end{align}
-    %With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\sfrac{1}{n}} \leq \frac{e}{n}$ follow respectively.
-Hence,
-    %\begin{equation}
-    %    \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
-    %    \leq \frac{e^2}{k + 1}\sqrt{ \frac{3k}{2} }
-    %    \leq e^2 \sqrt{\frac{3}{2k}}.
-    %    \label{equ:OSE-moment-bound-second}
-    %\end{equation}
-
-\end{proof}
 
@@
 \begin{proof}
+\color{black}
+
     First, we temporarily omit the parameter dependence.

### 2024-07-10 paper/paper.tex
Updates from Overleaf

@@
 
-\section{Proofs}
+\section{Moment bounds for non-standard Gaussian random vectors and matrices}
+
+The goal of this section is to establish moment bounds for $\lVert \mtx{A} \mtx{\Omega} \rVert _2^2$ with a fixed matrix $\mtx{A}$ and a Gaussian random matrix $\mtx{\Omega}$.
+These bounds are needed in the proof of \reflem{lem:nystrom}, but may also be of independent interest.
+For reasons which will become clear in the proof of the final and most interesting of these lemmas (\reflem{lem:nystrom}), we will need to bound $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]$ for a constant matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and a standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$.
+The corresponding result for the Frobenius norm $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _F^2 \right]$ has already been shown in \cite[Lemma 7]{kressner-2023-randomized-lowrank}.
+Several ways of separately bounding the spectral norm moments of the Wishart matrix $\mtx{\Omega} \mtx{\Omega}^{\top}$, i.e.
+when $\mtx{A} = \mtx{I}_m$ already exist, however, all of them are not sufficient for our purposes because they scale with $m$ \cite{chen-2005-condition-numbers, edelman-1988-eigenvalues-condition, james-1964-distributions-matrix}.
+In \cite[Lemma B.1]{tropp-2023-randomized-algorithms} a bound is achieved for $p = 2$ and $p = 4$.
+We will now generalize this result for arbitrary $p \in \mathbb{N}$.
+To do so, we first show the bound for Gaussian random vectors and then extend it to Gaussian random matrices in \reflem{lem:spectral-norm-moment}.
 
@@
 
+\begin{lemma}{Spectral norm moments of non-standard Gaussian random vector}{spectral-norm-moment-vector}
+    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random vector $\vct{\omega} \in \mathbb{R}^{m}$ satisfy for all $k,p \in \mathbb{N}$ 
+    \begin{equation}
+        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2^2 \right]
+        \leq  (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
+    \end{equation}
+    %\begin{equation}
+    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
+    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
+    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
+    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
+    %\end{equation}
+    %If we choose 
+\end{lemma}
+
+\begin{remark}
+    In particular, this result can also be used to bound
+    \begin{equation}
+        \mathbb{E}^{p}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2 \right] = \sqrt{\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2^2 \right]} \leq \sqrt{k + p} \cdot \left(\lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F\right)
+    \end{equation}
+    for any $k,p \in \mathbb{N}$.
+\end{remark}
+
+\begin{proof}
+    With the singular value decomposition $\mtx{A} = \mtx{U} \mtx{\Sigma} \mtx{V}^{\top}$ and the unitary invariance of the spectral norm, it can be seen that $\lVert \mtx{A} \mtx{\omega} \rVert _2^2 = \lVert \mtx{\Sigma} \widetilde{\mtx{\omega}} \rVert _2^2$ where $\widetilde{\mtx{\omega}}$ is again a standard Gaussian random vector.
+Hence, it is enough to bound $\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{\Sigma} \mtx{\omega} \rVert _2^2 \right]$ for a diagonal matrix $\mtx{\Sigma} = \operatorname{diag}(\sigma_1, \dots, \sigma_m)$ with $\sigma_1 \geq \dots \geq \sigma_m \geq 0$.
+
+    First, we can rewrite
+    \begin{equation}
+        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2^2 \right]
+        = \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right].
+        \label{equ:matvec-spectral-norm-simplification}
+    \end{equation}
+    Inspired by the proof of \cite[Theorem 1]{cohen-2016-optimal-approximate}, we split the diagonal entries of $\mtx{\Sigma}$ into $\ell = \lceil m/k \rceil$ groups of size $k \geq 1$
+    \begin{equation}
+        \overbrace{\underbrace{\sigma_1, \dots, \sigma_k}_{\leq \sigma_1}}^{\geq \sigma_{k+1}}, \overbrace{\underbrace{\sigma_{k+1}, \dots, \sigma_{2k}}_{\leq \sigma_{k+1}}}^{\geq \sigma_{2k+1}}, \dots, \overbrace{\underbrace{\sigma_{(\ell - 1)k + 1}, \dots, \sigma_{\ell k}}_{\leq \sigma_{(\ell - 1)k + 1}}}^{\geq 0}.
+    \end{equation}
+    If $m$ is not a multiple of $k$, we let $\sigma_i = 0$ for $i > m$. Since $\sigma_1 \geq \dots \geq \sigma_{\ell k} \geq 0$,
+    \begin{equation}
+        \sigma_1^2 = \lVert \mtx{\Sigma} \rVert _2^2
+        \quad \text{and} \quad
+        \sigma_{(i-1)k + 1}^2 \leq \frac{1}{k} \sum_{j=1}^{k} \sigma_{(i-2)k + j}^2, i = 2, \dots, \ell,
+    \end{equation}
+    and therefore
+    \begin{equation}
+        \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 = \lVert \mtx{\Sigma} \rVert _2^2 + \sum_{i=2}^{\ell} \sigma_{(i-1)k + 1}^2  \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \sum_{n=1}^{(\ell - 1)k} \sigma_n^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2.
+        \label{equ:singular-value-group-bound}
+    \end{equation}
+    This allows us to bound
+    \begin{align}
+        \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right]
+        &= \mathbb{E}^{\sfrac{p}{2}}\Bigg[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{(i-1)k + j}^2}_{\leq \sigma_{(i-1)k + 1}^2} \omega_{(i-1)k + j}^2 \Bigg] && \text{(separate sum into groups)} \notag \\
+        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 \sum_{j=1}^{k} \omega_{(i-1)k + j}^2 \right] && \text{($\sigma_{(i-1)k + j}^2 \leq \sigma_{(i-1)k + 1}^2$, $\forall i, j$)} \notag \\
+        &\leq \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 ~ \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{(i-1)k + j}^2 \right] && \text{(triangle inequality)} \notag \\
+        %&\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k}{2} + p)}{\Gamma(\frac{k}{2})} \right)^{\sfrac{1}{p}} && \text{(\refequ{equ:singular-value-group-bound} and \cite[Theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
+        &\leq (k + p) \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 && \text{(\reflem{lem:gamma})} \notag \\
+        &\leq (k + p) \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right). && \text{(using \refequ{equ:singular-value-group-bound})}
+    \end{align}
+
+    %The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[Lemma B.3]{tropp-2023-randomized-algorithms}
+    %\begin{align}
+    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
+    %    &= \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}} \notag \\
+    %    &\leq \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{1}{\Gamma(k - r + 2)} \right)^{\frac{1}{k - r + 1}} \left( \frac{k + r}{2} \right) \notag \\
+    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right)
+    %\end{align}
+    %With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\sfrac{1}{n}} \leq \frac{e}{n}$ follow respectively.
+Hence,
+    %\begin{equation}
+    %    \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
+    %    \leq \frac{e^2}{k + 1}\sqrt{ \frac{3k}{2} }
+    %    \leq e^2 \sqrt{\frac{3}{2k}}.
+    %    \label{equ:OSE-moment-bound-second}
+    %\end{equation}
+
+\end{proof}
+
+\begin{lemma}{Spectral norm moments of non-standard Gaussian random matrix}{spectral-norm-moment}
+    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all $p \in \mathbb{N}$ 
+    \begin{equation}
+        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
+        \leq  (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
+    \end{equation}
+    %\begin{equation}
+    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
+    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
+    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
+    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
+    %\end{equation}
+    %If we choose 
+\end{lemma}
+
+\begin{remark}
+    In particular, this result can also be used to bound
+    \begin{equation}
+        \mathbb{E}^{p}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2 \right] = \sqrt{\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]} \leq \sqrt{k + p} \cdot \left(\sqrt{3} \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F\right)
+    \end{equation}
+    for any $k,p \in \mathbb{N}$. For $p=2$ this bound is slightly less tight than its equivalent in \cite[Lemma B.1]{tropp-2023-randomized-algorithms}.
+\end{remark}
+
+\begin{proof}
+    In the proof of \cite[Lemma B.1]{tropp-2023-randomized-algorithms}, it is shown that Slepian's inequality \cite[Theorem 7.2.1]{vershynin-2018-highdimensional-probability} applied to two appropriately constructed Gaussian random fields yields
+    \begin{equation}
+        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
+        \leq \lVert \mtx{A} \rVert _2^2 \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \vct{\omega}_1 \rVert _2^2 \right] + \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega}_2 \rVert _2^2 \right]
+    \end{equation}
+    for independent standard Gaussian random vectors $\vct{\omega}_1 \in \mathbb{R}^{k}$ and $\vct{\omega}_2 \in \mathbb{R}^{m}$.
+\textcolor{red}{Doesn't A.1 give a tighter bound for the first term?} Both moments can be bound using \reflem{lem:spectral-norm-moment-vector}, one of which with $\mtx{A} = \mtx{I}_k$, to get
+    \begin{align}
+        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
+        &\leq \lVert \mtx{A} \rVert _2^2 (k + p) \left( \lVert \mtx{I}_k \rVert _2^2 + \frac{1}{k} \lVert \mtx{I}_k \rVert _F^2 \right) + (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right) \notag \\
+        &\leq (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
+    \end{align}
+
+    %The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[Lemma B.3]{tropp-2023-randomized-algorithms}
+    %\begin{align}
+    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
+    %    &= \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}} \notag \\
+    %    &\leq \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{1}{\Gamma(k - r + 2)} \right)^{\frac{1}{k - r + 1}} \left( \frac{k + r}{2} \right) \notag \\
+    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right)
+    %\end{align}
+    %With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\sfrac{1}{n}} \leq \frac{e}{n}$ follow respectively.
+Hence,
+    %\begin{equation}
+    %    \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
+    %    \leq \frac{e^2}{k + 1}\sqrt{ \frac{3k}{2} }
+    %    \leq e^2 \sqrt{\frac{3}{2k}}.
+    %    \label{equ:OSE-moment-bound-second}
+    %\end{equation}
+
+\end{proof}
+
 \end{document}

### 2024-07-10 paper/analysis.tex
make some minor changes to text

@@
 This allows us to apply \textcolor{red}{Theorem 9 ???
-be careful that the theorem numbering may be different when the paper is published in NLAA ???} from \cite{kressner-2023-randomized-lowrank}, which immediately implies the statement claimed by the theorem.
+be careful that the theorem numbering may be different when the paper is published in NLAA ???} \textcolor{green}{changed it to Theorem 5} from \cite[Theorem 5]{kressner-2024-randomized-lowrank}, which immediately implies the statement claimed by the theorem.
 \end{proof}
@@
         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \right]
-        \leq n_{\mtx{\Omega}} \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{2}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
+        \leq n_{\mtx{\Omega}} \left( 2 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{2}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
         \label{equ:spectral-norm-bound-applied}
@@
         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \frac{3 e^4}{\sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{2}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
+        \leq \frac{3 e^4}{\sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \left( 2 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{2}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
         %\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
@@
         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \frac{3 e^4}{\sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \left( \frac{6}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) + \frac{2}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) \right)
-        \leq 12 \sqrt{2} e^4 \frac{1}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
+        \leq \frac{3 e^4}{\sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \left( \frac{4}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) + \frac{2}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) \right)
+        \leq 9 \sqrt{2} e^4 \frac{1}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
         \label{equ:nystrom-proof-tail-bound}
@@
     \end{equation}
-    where $c = 1 + 12 \sqrt{2} e^4 $.
+    where $c = 1 + 9 \sqrt{2} e^4$.
 
-    As in \cite{kressner-2023-randomized-lowrank}, we can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable and by the Minkowski's integral inequality \cite[Theorem 202]{hardy-1952-inequalities}, we get for $n_{\mtx{\Omega}} \geq 4$
+    As in \cite{kressner-2024-randomized-lowrank}, we can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable and by the Minkowski's integral inequality \cite[Theorem 202]{hardy-1952-inequalities}, we get for $n_{\mtx{\Omega}} \geq 4$
     \begin{align}
@@
 
-When compared to the equivalent result for constant matrices \cite[Lemma 3.2]{persson-2022-improved-variants}, our constant factor up front is around three times larger and our failure probability decays half a power slower.
+When compared to the equivalent result for constant matrices \cite[Lemma 3.2]{persson-2022-improved-variants}, our constant factor up front is a bit more than two times larger and our failure probability decays half a power slower.
 

### 2024-07-10 paper/intro.tex
address feedback
make some minor changes to text

@@
 Instead of discretizing $[a, b]$ into $t_1, \ldots, t_m$, approximating $\Trace(\mtx{B}(t_i))$ with $b(t_i)$, and analyzing the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way is to quantify the error with $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate in each $t \in [a, b]$.
-This approach has already been used in the analysis of low-rank approximations of parameter-dependent matrices \cite{kressner-2023-randomized-lowrank}.
+This approach has already been used in the analysis of low-rank approximations of parameter-dependent matrices \cite{kressner-2024-randomized-lowrank}.
 
@@
 Is this the wording required by SISC?
-If not, focus on the reproducibility of the numerical experiments and figures/tables in this statement.}  \input{re-pro-badge.tex}
+If not, focus on the reproducibility of the numerical experiments and figures/tables in this statement.} \textcolor{green}{Done.} \input{re-pro-badge.tex}

### 2024-07-10 paper/paper.tex
make some minor changes to text

@@
 For reasons which will become clear in the proof of the final and most interesting of these lemmas (\reflem{lem:nystrom}), we will need to bound $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]$ for a constant matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and a standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$.
-The corresponding result for the Frobenius norm $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _F^2 \right]$ has already been shown in \cite[Lemma 7]{kressner-2023-randomized-lowrank}.
+The corresponding result for the Frobenius norm $\mathbb{E}^{\sfrac{p}{2}}\left[\lVert \mtx{A} \mtx{\Omega} \rVert _F^2 \right]$ has already been shown in \cite[Lemma 3]{kressner-2024-randomized-lowrank}.
 Several ways of separately bounding the spectral norm moments of the Wishart matrix $\mtx{\Omega} \mtx{\Omega}^{\top}$, i.e.
@@
         \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
-        \leq  (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
+        \leq  (k + p) \left( 2 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
     \end{equation}
@@
     \begin{equation}
-        \mathbb{E}^{p}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2 \right] = \sqrt{\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]} \leq \sqrt{k + p} \cdot \left(\sqrt{3} \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F\right)
+        \mathbb{E}^{p}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2 \right] = \sqrt{\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]} \leq \sqrt{k + p} \cdot \left(\sqrt{2} \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F\right)
     \end{equation}
@@
     for independent standard Gaussian random vectors $\vct{\omega}_1 \in \mathbb{R}^{k}$ and $\vct{\omega}_2 \in \mathbb{R}^{m}$.
-\textcolor{red}{Doesn't A.1 give a tighter bound for the first term?} Both moments can be bound using \reflem{lem:spectral-norm-moment-vector}, one of which with $\mtx{A} = \mtx{I}_k$, to get
+\textcolor{red}{Doesn't A.1 give a tighter bound for the first term?} \textcolor{green}{Yes.
+Fixed it.} The first moment can be bound using \reflem{lem:gamma} and the second using \reflem{lem:spectral-norm-moment-vector} to get
     \begin{align}
         \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
-        &\leq \lVert \mtx{A} \rVert _2^2 (k + p) \left( \lVert \mtx{I}_k \rVert _2^2 + \frac{1}{k} \lVert \mtx{I}_k \rVert _F^2 \right) + (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right) \notag \\
-        &\leq (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
+        &\leq \lVert \mtx{A} \rVert _2^2 (k + p)  + (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right) \notag \\
+        &\leq (k + p) \left( 2 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
     \end{align}

### 2024-07-11 paper/analysis.tex
import changes from overleaf

@@
 \begin{proof} 
+\textcolor{red}{??? I feel that the block diagonal embedding trick should come first, like in the proof of Meyer et al. Hopefully, this gets rid of the additional $\log \delta$ factor ???}
 The proof proceeds by bounding moments for fixed values of $t$ and then use Minkowski's integral inequality.
@@
 < 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$.
-Consequently, the $L^{2k}$ norm $\mathbb{E}^{2k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{2k} \right] \right)^{\sfrac{1}{2k}}$ of $r(\mtx{B}, \vct{\psi})$ is bounded by
+\textcolor{red}{???
+this norm should be defined more prominently somewhere else ???} Consequently, the $L^{2k}$ norm $\mathbb{E}^{2k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{2k} \right] \right)^{\sfrac{1}{2k}}$ of $r(\mtx{B}, \vct{\psi})$ is bounded by
     \begin{equation}
@@
 
-The following result confirms the inituition that the Nystr\"om approximation is very well suited in the presence of decaying singular values: If the nuclear norm error to the best rank-$r$ approximation to $B(t)$ is bounded by $\varepsilon$ for every $t$ then the trace of the Nystr\"om approximation (with a bit of oversampling) results, with high probability, in an error that is only modestly larger than $\varepsilon$.
+A consequence of~\cite{kressner-2024-randomized-lowrank}, the following result confirms the inituition that the Nystr\"om approximation is very well suited in the presence of decaying singular values: If the nuclear norm error to the best rank-$r$ approximation to $B(t)$ is bounded by $\varepsilon$ for every $t$ then the trace of the Nystr\"om approximation (with a bit of oversampling) results, with high probability, in an error that is only modestly larger than $\varepsilon$.
 
@@
 
-\refthm{thm:hutchinson} has shown that the the Girard-Hutchinson estimator achieves a relative error $\varepsilon$ when using $\mathcal{O}(\varepsilon^{-2})$ queries.
-The aim of this section is to show that the Nyström++ estimator~\refequ{equ:nystrompp-trace-estimator} improves this to $\mathcal{O}(\varepsilon^{-1})$ matrix-vector products with random vectors, \emph{without} requiring any assumption on the singular values of $\mtx{B}(t)$.
+\refthm{thm:hutchinson} shows that the the Girard-Hutchinson estimator achieves a relative $\varepsilon$-error when using $\mathcal{O}(\varepsilon^{-2})$ queries.
+The aim of this section is to show that the Nyström++ estimator~\refequ{equ:nystrompp-trace-estimator} improves this to $\mathcal{O}(\varepsilon^{-1})$ queries (that is, matrix-vector products with random vectors), \emph{without} requiring any assumption on the singular values of $\mtx{B}(t)$.
 This parallels existing results for the constant case
-for Hutch++\cite[Theorem 3.1]{meyer-2021-hutch-optimal} and Nyström++\textcolor{red}{??? Include specific result ???}.
-
-\color{black}
-
-To do so, we mimic the proof of a similar bound for the Hutch++ estimator .
-Correspondingly, we will need to show that with $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2})$ random vectors, the Nyström approximation verifies $\int \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)\rVert _F~\mathrm{d}t \leq \varepsilon \int \Trace(\mtx{B}(t))~\mathrm{d}t$, regardless of the structure of the matrix $\mtx{B}(t)$.
-Just as in the corresponding proof for the Girard-Hutchinson estimator \refthm{thm:hutchinson}, we again bound the higher order moments of the approximation error for fixed $t$ and then use Minkowsi's integral inequality to carry over the results to the continuous case.
-We will do this in a sequence of three lemmas.
+for Hutch++\cite[Theorem 3.1]{meyer-2021-hutch-optimal} and Nyström++\textcolor{red}{???
+Include specific result ???}.
+Indeed, our proof follows the strategy from~\cite{meyer-2021-hutch-optimal}, by first showing that the parameter-dependent Nyström approximation achieves a relative $\varepsilon$-error when using $\mathcal{O}(\varepsilon^{-2})$ queries.
 
@@
 %\todo{Proof idea: structural bound, then higher order moment bound to apply Markov's inequality.}
-
 \begin{proof}
-\color{black}
-
-    First, we temporarily omit the parameter dependence.
-Let $\mtx{B} = \mtx{U} \mtx{\Lambda} \mtx{U}^{\top}$ be the eigenvalue decomposition of a symmetric positive semi-definite matrix $\mtx{B} \in \mathbb{R}^{n \times n}$, i.e.
-$\mtx{\Lambda} = \operatorname{diag}(\lambda_1, \dots, \lambda_n)$ and $\mtx{U}^{\top} \mtx{U}= \mtx{I}_n$.
-We partition these matrices into
-    \begin{equation}
+For fixed $t$, consider the spectral decomposition $\mtx{B}(t) = \mtx{U} \mtx{\Lambda} \mtx{U}^{\top}$ with  
+$\mtx{\Lambda} = \operatorname{diag}(\lambda_1, \dots, \lambda_n)$ and $\lambda_1 \ge \dots \ge \lambda_n \ge 0$. Partitioning
+    \begin{equation*}
         \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+4ex+\baselineskip}
@@
             \smash{\overbrace{\mtx{\Lambda}_1}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2}_{(n-k) \times (n-k)}}
-        \end{bmatrix}.
-    \end{equation}
-    Further, for a standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ we define $\mtx{\Omega}_1 = \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times n_{\mtx{\Omega}}}$ and $\mtx{\Omega}_2 = \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times n_{\mtx{\Omega}}}$, which are likewise standard Gaussian.
-
-    We can apply \cite[Theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound
+        \end{bmatrix},
+    \end{equation*}
+    we set
+    $\mtx{\Omega}_1 := \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times n_{\mtx{\Omega}}}$ and $\mtx{\Omega}_2 := \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times n_{\mtx{\Omega}}}$, which are independent Gaussian random matrices.
+Applying Theorem B.1 from~\cite{persson-2023-randomized-lowrank} for $f(x) = x$ yields the bound
     \begin{equation}
-        \lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F 
+        \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}(t)} \rVert _F 
         \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2,
@@
     \end{equation}
-    where $\lVert \cdot \rVert _{(4)}$ denotes the Schatten-4 norm.
-
-    As in the proof of \cite[Lemma 3]{meyer-2021-hutch-optimal}, the first term -- and its moments -- are limited by
+    where $\lVert \cdot \rVert _{(4)}$ denotes the Schatten-4 norm. Standard arguments (see, e.g., the proof of~\cite[Lemma 3]{meyer-2021-hutch-optimal}) provide a simple bound on the first term:
     \begin{equation}
         \lVert \mtx{\Lambda}_2 \rVert _F
-        = \sqrt{\sum_{i=k+1}^{n} \lambda_i^2}
-        \leq \sqrt{ \lambda_{k+1} \sum_{i=k+1}^{n} \lambda_i}
-        \leq \sqrt{ \frac{\Trace(\mtx{B})}{k} \Trace(\mtx{B})}
-        \leq \frac{1}{\sqrt{k}} \Trace(\mtx{B}).
+        \leq \sqrt{ \lambda_{k+1} (  \lambda_{k+1} + \cdots + \lambda_{n})}
+        \leq \sqrt{ \Trace(\mtx{B}) / k \cdot \Trace(\mtx{B})}
+        \leq \Trace(\mtx{B}) / \sqrt{k}.
         \label{equ:nystrom-proof-frobenius-trace}
     \end{equation}
+    \textcolor{red}{??? STOPPED HERE: Not clear to me why $n_\Omega/4$ should be the optimal exponent for the tail bound ???}
+
+    \color{black}
     The $q$-th moment of the second term can be processed with standard matrix norm manipulations and the stochastic independence of $\mtx{\Omega}_1$ and $\mtx{\Omega}_2$ to
@@
 
+\color{black}
+
 When compared to the equivalent result for constant matrices \cite[Lemma 3.2]{persson-2022-improved-variants}, our constant factor up front is a bit more than two times larger and our failure probability decays half a power slower.

### 2024-07-11 paper/paper.tex
import changes from overleaf

--- 2024-07-08 paper/paper.tex
@@
 \input{spectraldensity.tex}
-%\input{numresults.tex}
+\input{numresults.tex}
 
@@
 
-\appendix
-
-\section{Proofs}
-
-\textcolor{red}{???
-Need to check carefully that this result is not known ???} \textcolor{green}{I haven't found anything so far.
-The fact that also Bujanović/Grubišić/Kressner/Lam in Lemma 20 use a very similar technique, but merely to bound $\Gamma(\frac{p+1}{2})$ makes me suspect there's not much, except for Stirling approximation at the cost of some simplicity.
-Will keep investigating.}
-
-\begin{lemma}{Moment bound of $\chi^2$-random variable}{gamma}
-    Let $X$ follow a $\chi^2$ distribution with $k \in \mathbb{N}$ degrees of freedom. Then for all $p \in \mathbb{N}$ it holds
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}[X] \leq k + p
-        %\frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \leq \left( \frac{k + p}{2} \right)^{\sfrac{p}{2}}.
-    \end{equation}
-\end{lemma}
-
-\begin{proof}
-    From \cite[Theorem 3.3.2]{hogg-2013-introduction-mathematical} we know
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}[X] = 2 \left( \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \right)^{\sfrac{2}{p}}.
-    \end{equation}
-    We first treat the case $k=p=1$, for which indeed
-    \begin{equation}
-        2 \left( \frac{\Gamma(\frac{1+1}{2})}{\Gamma(\frac{1}{2})} \right)^2 = \frac{2}{\pi} \leq 2 = 1 + 1
-    \end{equation}
-    So we now assume $k + p \geq 3$. We can use a telescoping product and the property $\Gamma(z+1)/\Gamma(z) = z$ to bound
-    \begin{align}
-        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} 
-        &= \frac{\Gamma(\frac{k}{2} + 1)}{\Gamma(\frac{k}{2})} \frac{\Gamma(\frac{k}{2} + 2)}{\Gamma(\frac{k}{2} + 1)} \cdots \frac{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor - 1)} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)} \notag \\
-        &= \left(\frac{k}{2}\right) \left(\frac{k}{2} + 1\right) \cdots \left( \frac{k}{2} + \bigg\lfloor \frac{p}{2} \bigg\rfloor - 1 \right) \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)} \notag \\
-        &\leq \left(\frac{k}{2} + \bigg\lfloor \frac{p}{2} \bigg\rfloor - 1\right)^{\lfloor \sfrac{p}{2} \rfloor} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)}
-    \end{align}
-    If $p$ is even, then $\lfloor p/2 \rfloor = p/2$ and hence
-    \begin{equation}
-        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} 
-        \leq \left( \frac{k + p}{2} - 1 \right)^{\sfrac{p}{2}}
-        \leq \left( \frac{k + p}{2}\right)^{\sfrac{p}{2}}.
-    \end{equation}
-    If $p$ is odd, then $\lfloor p/2 \rfloor = (p - 1)/2$, from which follows with Gautschi's inequality \cite{kershaw-1983-extensions-gautschi}
-    \begin{equation}
-        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})}
-        \leq \left(\frac{k + p}{2} - \frac{3}{2} \right)^{ \frac{p - 1}{2}} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k + p}{2} - \frac{1}{2})}
-        \leq \left(\frac{k + p}{2}\right)^{ \frac{p - 1}{2}} \left(\frac{k + p}{2}\right)^{\sfrac{1}{2}}
-        = \left(\frac{k + p}{2}\right)^{ \sfrac{p}{2}}
-    \end{equation}
-    Hence,
-    \begin{equation}
-        \mathbb{E}^{\sfrac{p}{2}}[X] \leq 2 \left( \frac{k + p}{2} \right) \leq k + p.
-    \end{equation}
-\end{proof}
+\input{appendix.tex}
 

### 2024-07-11 paper/re-pro-badge.tex
automatically reproduce paper

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/3f5eb52}{3f5eb52} on 2024-07-09 at 06:06:30 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/75a23a4}{75a23a4} on 2024-07-11 at 06:15:48 UTC.

### 2024-07-11 paper/analysis.tex
rewrite Nystrom proof and address feedback

@@
 \subsection{Methods}
+\label{subsec:methods}
 
-\paragraph{Girard-Hutchinson estimator.} We can approximate the trace with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ stochastically independent standard Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ to form
+\paragraph{Girard-Hutchinson estimator.} We can approximate the trace with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ stochastically independent standard Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$, i.e.
+vectors whose entries are independent standard Gaussian random variables, to form
 \begin{equation}
@@
 \end{equation}
-??? Gaussian random matrix define! ???
-where $\mtx{\Psi} = [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\mtx{\Psi}}}] \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$.
+where $\mtx{\Psi} = [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\mtx{\Psi}}}] \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$ is a standard Gaussian random matrix.
 Other choices for the distribution of the random vectors are possible, for example by uniformly sampling from $\{-1, +1\}$ or from the $(n-1)$-sphere.
@@
 
+\textcolor{green}{In the following pages, we will upper bound the $L^1$-error of each of the trace estimators from \refsec{subsec:methods} with a suitable quantity which parallels the corresponding result in the constant case.
+As in \cite{kressner-2024-randomized-lowrank}, this often requires us to bound the high order moments $\mathbb{E}^p[X] = (\mathbb{E}[|X|^p])^{\sfrac{1}{p}}$ of certain random variables $X$.}
+
 \subsubsection{Girard-Hutchinson estimator}
@@
 \textcolor{red}{???
-this norm should be defined more prominently somewhere else ???} Consequently, the $L^{2k}$ norm $\mathbb{E}^{2k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{2k} \right] \right)^{\sfrac{1}{2k}}$ of $r(\mtx{B}, \vct{\psi})$ is bounded by
+this norm should be defined more prominently somewhere else Consequently, the $L^{2k}$ norm $\mathbb{E}^{2k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{2k} \right] \right)^{\sfrac{1}{2k}}$ ???
+} \textcolor{green}{Moved it to start of section.} Consequently, the high order moments of $r(\mtx{B}, \vct{\psi})$ are bounded by
     \begin{equation}
@@
 
-A consequence of~\cite{kressner-2024-randomized-lowrank}, the following result confirms the inituition that the Nystr\"om approximation is very well suited in the presence of decaying singular values: If the nuclear norm error to the best rank-$r$ approximation to $B(t)$ is bounded by $\varepsilon$ for every $t$ then the trace of the Nystr\"om approximation (with a bit of oversampling) results, with high probability, in an error that is only modestly larger than $\varepsilon$.
+As a consequence of~\cite{kressner-2024-randomized-lowrank}, the following result confirms the inituition that the Nystr\"om approximation is very well suited in the presence of decaying singular values: If the nuclear norm error to the best rank-$r$ approximation to $B(t)$ is bounded by $\varepsilon$ for every $t$ then the trace of the Nystr\"om approximation (with a bit of oversampling) results, with high probability, in an error that is only modestly larger than $\varepsilon$.
 
@@
     \end{align*}
-    where $\lVert \cdot \rVert _{\ast}$ denotes the nuclear norm and we used a \textcolor{red}{result from the proof of \cite[Corollary 8.2]{tropp-2023-randomized-algorithms} ???
-Is this really the original source.
-Have a look whether David P mentions this somewhere?
-Also, this may be contained in the original Gittens Mahoney paper but I am not sure ???} \textcolor{green}{The original is in Gittens' arXiv preprint \cite[Theorem 1]{gittens-2011-spectral-norm} but also in Gittens/Mahoney \cite[Equation 4]{gittens-2013-revisiting-nystrom}} in the last equality.
-This allows us to apply \textcolor{red}{Theorem 9 ???
-be careful that the theorem numbering may be different when the paper is published in NLAA ???} \textcolor{green}{changed it to Theorem 5} from \cite[Theorem 5]{kressner-2024-randomized-lowrank}, which immediately implies the statement claimed by the theorem.
+    where $\lVert \cdot \rVert _{\ast}$ denotes the nuclear norm and we used \cite[Theorem 1]{gittens-2011-spectral-norm} in the last equality.
+This allows us to apply \cite[Theorem 5]{kressner-2024-randomized-lowrank}, which immediately implies the statement claimed by the theorem.
 \end{proof}
 
-Compared to the constant case~\cite[Theorem 8.1]{tropp-2023-randomized-algorithms}, the bound of Theorem~\ref{thm:nystrom} features an additional factor $1+r$ that cannot be compensated with oversampling.
+Compared to the constant case~\cite[Theorem 8.1]{tropp-2023-randomized-algorithms}, the bound of \refthm{thm:nystrom} features an additional factor $1+r$ that cannot be compensated with oversampling.
 
@@
 The aim of this section is to show that the Nyström++ estimator~\refequ{equ:nystrompp-trace-estimator} improves this to $\mathcal{O}(\varepsilon^{-1})$ queries (that is, matrix-vector products with random vectors), \emph{without} requiring any assumption on the singular values of $\mtx{B}(t)$.
-This parallels existing results for the constant case
-for Hutch++\cite[Theorem 3.1]{meyer-2021-hutch-optimal} and Nyström++\textcolor{red}{???
-Include specific result ???}.
+This parallels existing results for the constant case for the Hutch++~\cite[Theorem 3.1]{meyer-2021-hutch-optimal} and Nyström++ estimators \cite[Theorem 3.4]{persson-2022-improved-variants}.
+\textcolor{red}{??
+Include specific result ???} \textcolor{green}{Specifically, a bound of the form $|\Trace(\mtx{B}) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})| \leq \varepsilon \Trace(\mtx{B})$ for any constant symmetric positive semi-definite matrix $\mtx{B}$ is shown to hold with high probability for the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})$ \refequ{equ:nystrompp-trace-estimator} when $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1})$ }.
 Indeed, our proof follows the strategy from~\cite{meyer-2021-hutch-optimal}, by first showing that the parameter-dependent Nyström approximation achieves a relative $\varepsilon$-error when using $\mathcal{O}(\varepsilon^{-2})$ queries.
@@
     \end{equation}
-    \textcolor{red}{??? STOPPED HERE: Not clear to me why $n_\Omega/4$ should be the optimal exponent for the tail bound ???}
+    \textcolor{red}{???
+STOPPED HERE: Not clear to me why $n_\Omega/4$ should be the optimal exponent for the tail bound ???} \textcolor{green}{I rearranged and reformulated things a bit to hopefully make it clearer:}
 
     \color{black}
-    The $q$-th moment of the second term can be processed with standard matrix norm manipulations and the stochastic independence of $\mtx{\Omega}_1$ and $\mtx{\Omega}_2$ to
+    The $q$-th moment of the second term in \refequ{equ:nystrom-proof-persson-bonud} can be processed with standard matrix norm manipulations and the stochastic independence of $\mtx{\Omega}_1$ and $\mtx{\Omega}_2$ to
     \begin{equation}
@@
     \end{equation}
+    \textcolor{green}{
     To match the decay rate of the moments of the first term in \refequ{equ:nystrom-proof-persson-bonud}, we will need to ensure that this term also is of order $\mathcal{O}(\Trace(\mtx{B}) / \sqrt{k})$.
-To this purpose, and to simplify the notation, we choose $k = n_{\mtx{\Omega}}/2$ and $q = n_{\mtx{\Omega}}/4$ for even $n_{\mtx{\Omega}}$ to ensure $k \in \mathbb{N}$.
-We can apply \reflem{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$ and $p = n_{\mtx{\Omega}}/2$ to bound
-    \begin{equation}
-        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \right]
-        \leq n_{\mtx{\Omega}} \left( 2 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{2}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
-        \label{equ:spectral-norm-bound-applied}
-    \end{equation}
-    The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[Lemma B.3]{tropp-2023-randomized-algorithms}
+We bound the moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ similarly to \cite[Lemma B.3]{tropp-2023-randomized-algorithms}, but without restricting the moments to $0 \leq q \leq 18$.
+The explicit integration of \cite[Equation B.7]{tropp-2023-randomized-algorithms} imposes the condition $n_{\mtx{\Omega}} - k - 2q \geq 0$.
+Both $n_{\mtx{\Omega}}$ and $k$ are integers and must be of the same order to ensure a decay of $\mathcal{O}(\Trace(\mtx{B}) / \sqrt{k})$.
+To avoid restricting the choice of $n_{\mtx{\Omega}}$ too much, we let it be even and set $k = n_{\mtx{\Omega}}/2$.
+The moment $q$ should be chosen as large as possible to ensure a fast decay of the failure probability, so we fix it to $q = n_{\mtx{\Omega}}/4$.
+Therefore, we get
     \begin{equation}
@@
     \end{equation}
-    As opposed to \cite[Lemma B.3]{tropp-2023-randomized-algorithms}, our bound holds for all even $n_{\mtx{\Omega}} \in \mathbb{N}$, at the cost of a factor of $e^2$.
+    As opposed to \cite[Lemma B.3]{tropp-2023-randomized-algorithms}, our bound holds for all even $n_{\mtx{\Omega}} \in \mathbb{N}$, at the cost of a factor of $e^2$.}
+
+    The remaining term can dealt with using \reflem{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$ and $p = n_{\mtx{\Omega}}/2$
+    \begin{equation}
+        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \right]
+        \leq n_{\mtx{\Omega}} \left( 2 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{2}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
+        \label{equ:spectral-norm-bound-applied}
+    \end{equation}
+
+    %The $q$-th moment of the second term can be processed with standard matrix norm manipulations and the stochastic independence of $\mtx{\Omega}_1$ and $\mtx{\Omega}_2$ to
+    %\begin{equation}
+    %    \mathbb{E}^{q}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
+        %&= \mathbb{E}^{p}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )( \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger})^{\top} \rVert _F \right] && \text{($\lVert \mtx{A} \rVert _{(4)}^2 = \lVert \mtx{A} \mtx{A}^{\top} \rVert _{F}$)} \notag \\
+        %&\leq \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] && \text{($\lVert \mtx{A} \mtx{B} \rVert _F \leq \lVert \mtx{A} \rVert _F \lVert \mtx{B} \rVert _2$)} \notag \\
+    %    \leq \mathbb{E}^{q}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]%  && \text{($\lVert \mtx{A} \rVert _{(4)}^2 \leq \sqrt{k} \lVert \mtx{A} \rVert _2^2$)} \notag \\
+        %\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]%  && \text{(submultiplicativity)} \notag \\
+    %    \leq \sqrt{k} \mathbb{E}^{q}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2\right] \mathbb{E}^{q}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2  \right].% && \text{(independence)}
+    %    \label{equ:nystrom-proof-processed-tail}
+    %\end{equation}
+    %To match the decay rate of the moments of the first term in \refequ{equ:nystrom-proof-persson-bonud}, we will need to ensure that this term also is of order $\mathcal{O}(\Trace(\mtx{B}) / \sqrt{k})$.
+To this purpose, and to simplify the notation, we choose $k = n_{\mtx{\Omega}}/2$ and $q = n_{\mtx{\Omega}}/4$ for even $n_{\mtx{\Omega}}$ to ensure $k \in \mathbb{N}$.
+We can apply \reflem{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$ and $p = n_{\mtx{\Omega}}/2$ to bound
+    %\begin{equation}
+    %    \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \right]
+    %    \leq n_{\mtx{\Omega}} \left( 2 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{2}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
+    %    \label{equ:spectral-norm-bound-applied}
+    %\end{equation}
+    %The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[Lemma B.3]{tropp-2023-randomized-algorithms}
+    %\begin{equation}
+    %    \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
+    %    = \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{n_{\mtx{\Omega}}}{4}} \right]^{\sfrac{4}{n_{\mtx{\Omega}}}}%  \notag \\
+    %    \leq \left( 1 + \frac{n_{\mtx{\Omega}}}{2} \right)^{\sfrac{4}{n_{\mtx{\Omega}}}} \left( \frac{1}{(\frac{n_{\mtx{\Omega}}}{2} + 1)!}\right)^{\frac{2}{\frac{n_{\mtx{\Omega}}}{2} + 1}} \left( \frac{3 n_{\mtx{\Omega}}}{4}\right).
+    %\end{equation}
+    %From the expansion of the exponential function as a power series $e^n = 1 + n + \dots + n^n/n!
++ \dots$ for $n \in \mathbb{N}$ we see $e^n \geq 1 + n$ and $e^n \geq n^n / n!$, from which $(1 + n)^{\sfrac{1}{n}} \leq e$ and $\left( 1/n!
+\right)^{\sfrac{1}{n}} \leq e/n$ follow respectively.
+Hence,
+    %\begin{equation}
+    %    \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
+    %    \leq e^2 \frac{e^2}{(\frac{n_{\mtx{\Omega}}}{2}+ 1)^2} \left( \frac{3 n_{\mtx{\Omega}}}{4}\right)
+    %    \leq \frac{3 e^4}{n_{\mtx{\Omega}}}.
+    %    \label{equ:pinv-spectral-norm-bound}
+    %\end{equation}
     Inserting \refequ{equ:spectral-norm-bound-applied} and  \refequ{equ:pinv-spectral-norm-bound} in \refequ{equ:nystrom-proof-processed-tail} with $k = n_{\mtx{\Omega}} / 2$ gives

### 2024-07-11 paper/intro.tex
rewrite Nystrom proof and address feedback

@@
 Suitable extensions to parameter-dependent matrices have been considered, e.g., in~\cite{lin-2017-randomized-estimation,chen-2023-krylovaware-stochastic}, but we are not aware of an analysis providing insight and justification of these extensions.
-In passing, we note that dynamic trace estimation~\cite{dharangutte-2024-dynamic-trace,woodruff-2024-optimal-query} is an efficient technique for \textcolor{red}{subsequently ???
-true?
-???}  estimating the traces of matrices $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ for wich the increments $\mtx{B}(t_{i+1}) - \mtx{B}(t_i)$ are relatively small in norm.
+In passing, we note that dynamic trace estimation~\cite{dharangutte-2024-dynamic-trace,woodruff-2024-optimal-query} is an efficient technique for subsequently estimating the traces of matrices $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ for wich the increments $\mtx{B}(t_{i+1}) - \mtx{B}(t_i)$ are relatively small in norm.
 The potential of dynamic trace estimation appears to be limited in our setting because $\mtx{B}(t)$ may change rapidly close to eigenvalues, with $g_{\sigma}$ approximating a Dirac delta function.
@@
 All methods considered in this work are based on the following simple idea: Apply an existing randomized trace estimator to $\Trace(\mtx{B}(t))$ with \emph{constant} random vectors, that is, the same randomization is used for each value of the parameter $t$.
-For example, the Girard-Hutchinson estimator ???
-STOP HERE ???
+\textcolor{green}{For example, the Girard-Hutchinson estimator becomes $n_{\mtx{\Psi}}^{-1} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j$ for $n_{\mtx{\Psi}}$ constant standard Gaussian random vectors $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}}$.}
+??? STOP HERE ???
 
@@
 
-\paragraph{Reproducibility.} \textcolor{red}{This paragraph looks odd.
-Is this the wording required by SISC?
-If not, focus on the reproducibility of the numerical experiments and figures/tables in this statement.} \textcolor{green}{Done.} \input{re-pro-badge.tex}
+\paragraph{Reproducibility.} \input{re-pro-badge.tex}

### 2024-07-11 paper/paper.tex
rewrite Nystrom proof and address feedback

@@
 \begin{itemize}
- \item It is not great that the references occupy so much space (more than 4 pages); consider using a more condensed style.
-The paper needs to get strictly below 30 pages prior to submission.
-\textcolor{green}{Journal abbreviations, max 2 authors, omit superfluous fields $\to$ 3 pages.}
- Journal abbreviations and removing superfluous fields is good.
-max 2 authors is not coming in math, please reverse this.
-\textcolor{green}{Done.} The text style is a bit too generous but no need to change this now.
-\textcolor{green}{Ok!
-Let's talk about it at some point.}
+ \item The text style is a bit too generous but no need to change this now. \textcolor{green}{Ok! Let's talk about it at some point.}
 \end{itemize}
@@
 \paragraph{Acknowledgements.} The authors gratefully acknowledge the help of Lin Lin in reproducing the example in \refsec{subsec:hamiltonian}.
-They thank David Persson for many enlightening discussions relating the proofs in \refsec{subsec:nystrom-pp}.
+They thank David Persson for many enlightening discussions relating the proof in \refsec{subsec:nystrom-pp}.
 

### 2024-07-11 paper/analysis.tex
rewrite proof of Girard-Hutchinson bound

@@
     For a symmetric matrix $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ that depends continuously on $t \in [a, b]$, consider the  Girard-Hutchinson estimator~\eqref{equ:hutchinson-trace-estimator} with an $n\times n_{\mtx{\Psi}}$ Gaussian random matrix $\mtx{\Psi}$.
-Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ the bound
+Then for any $p \geq 1, p \in \mathbb{R}$ and $\gamma \geq 1$ the bound
     \begin{equation} \label{eq:ghbound}
-        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) \right| ~\mathrm{d}t \leq 2^4 \gamma k \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F  ~\mathrm{d}t.
+        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq c \gamma \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}}  \right\} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
     \end{equation}
-    holds with probability at least $1 - \gamma^{-2k}$.
-    In particular, given $\varepsilon > 0$ and $\delta \in (0, 1)$, the bound $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$ holds
-    with probability at least $1-\delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$.
+    holds with probability at least $1 - \gamma^{-p}$.
+    In particular, given $\varepsilon \in (0, ce)$ and $\delta \in (0, e^{-1})$, the bound $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$ holds
+    with probability at least $1-\delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1}))$.
 \end{theorem}
 
-%The idea behind the proof is to bound  to extend the results to the continuous parameter case.
-We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
+\color{black}
 \begin{proof}
-\textcolor{red}{??? I feel that the block diagonal embedding trick should come first, like in the proof of Meyer et al. Hopefully, this gets rid of the additional $\log \delta$ factor ???}
-The proof proceeds by bounding moments for fixed values of $t$ and then use Minkowski's integral inequality.
-We start by considering the residual of the 1-query Girard-Hutchinson estimator for a fixed matrix $\mtx{B}$:
-    \begin{equation}
-        r(\mtx{B}, \vct{\psi}) := \Trace(\mtx{B}) - \vct{\psi}^{\top} \mtx{B} \vct{\psi}
-    \end{equation}
-%    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers
-for a Gaussian random vector $\vct{\psi}$.
-%
-%    First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence.
-The proof of \cite[Lemma 3]{cortinovis-2022-randomized-trace} shows that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm.
-By \cite[Theorem 2.3]{boucheron-2013-concentration-inequalities} this implies for every $k \in \mathbb{N}$ that
-    \begin{align*}
-        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
-        &\leq k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k}  \\
-        &\le \big( k! 2^{4 k} + (2 k)! 2^{6 k}\big) \lVert \mtx{B} \rVert _F^{2 k} 
-        \le \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
-    \end{align*}
-%     Because $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
-%     \begin{equation}
-%         \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right] \leq 
-%     \end{equation}
-    Stirling's approximation \cite{robbins-1955-remark-stirling} gives $(2 k)!
-< 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$.
-\textcolor{red}{???
-this norm should be defined more prominently somewhere else Consequently, the $L^{2k}$ norm $\mathbb{E}^{2k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{2k} \right] \right)^{\sfrac{1}{2k}}$ ???
-} \textcolor{green}{Moved it to start of section.} Consequently, the high order moments of $r(\mtx{B}, \vct{\psi})$ are bounded by
-    \begin{equation}
-        \mathbb{E}^{2k}\left[ r(\mtx{B}, \vct{\psi}) \right]
-        \leq \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F \leq 2^4 k \lVert \mtx{B} \rVert _F,
-        \label{equ:hutchinson-trace-onequery-fixed}
-    \end{equation}
-    using the inequality $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, which can be verified using the monotonic decrease of the left-hand side with respect to $k \in \mathbb{N}$.
-
-    To address the parameter-dependent case, we first note that the continuity assumption implies that $r(\mtx{B}(t), \vct{\psi})$ is measurable.
-Therefore, we can apply Minkowski's integral inequality~\cite[Theorem 202]{hardy-1952-inequalities} to conclude from~\refequ{equ:hutchinson-trace-onequery-fixed} that
-    \begin{equation}
-        \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t  \right]
-        \leq \int_{a}^{b} \mathbb{E}^{2 k}\left[ r(\mtx{B}(t), \vct{\psi}) \right]~\mathrm{d}t
-        \leq 2^4 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
-    \end{equation}
-    Consequently, by Markov's inequality,  the inequality 
-    \begin{equation} 
-        \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t \leq 2^4 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
-        \label{equ:hutchinson-trace-onequery-uniform}
-    \end{equation}
-    holds for any $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$.
-
-    To extend the bound~\eqref{equ:hutchinson-trace-onequery-uniform} from $1$ to $n_{\mtx{\Psi}}$ queries, we use a standard embedding trick (see, e.g.,~\cite[Theorem 1]{cortinovis-2022-randomized-trace}).
-Letting
+    The proof proceeds by bounding moments for fixed values of $t$ and then uses Minkowski's integral inequality.
+We start with a standard embedding trick (see, e.g.,~\cite[Theorem 1]{cortinovis-2022-randomized-trace}) for which we define
     \begin{equation}
@@
             \vct{\psi}_{n_{\mtx{\Psi}}}
-        \end{pmatrix},
+        \end{pmatrix}.
     \end{equation}
-    we note that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) = r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold.
-Thus, the claimed probabilistic bound~\eqref{eq:ghbound} follows from applying~\eqref{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$.
-%     Then 
-%     where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors.
-Then $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$, and
-%     \begin{equation}
-%         r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}}) = \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j.
-%     \end{equation}
-%     Hence, by applying \refequ{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$ we conclude that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
-%     \begin{equation}
-%         \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j \right| ~ \mathrm{d}t
-%         \leq 2^4 k \gamma \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
-%     \end{equation}
+    Note that $\Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}} = \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$, $\lVert \widetilde{\mtx{B}}(t) \rVert _2 = \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}}$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold.
+Therefore, by \reflem{lem:hanson-wright-moments} we have
+    \begin{align}
+        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))]
+        &= \mathbb{E}^{p}[\Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}}] \notag \\
+        &\leq c \max\left\{ \sqrt{p} \lVert \widetilde{\mtx{B}}(t) \rVert _F, p \lVert \widetilde{\mtx{B}}(t) \rVert _2 \right\} \notag \\
+        &= c \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F, \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 \right\} \notag \\
+        &\leq c \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}} \right\}  \lVert \mtx{B}(t) \rVert _F.
+        \label{equ:moment-bound-hanson-wright}
+    \end{align}
 
-    %Setting $\delta = \gamma^{-2 k}$ and choosing $k = \lceil \log(\delta^{-1}) %\rceil$, where $\lceil x \rceil$ is the ceiling function which returns the greatest %integer larger than or equal to $x$, we get that for all $\delta \in (0, e^{-\sfrac%{1}{2}})$
-    %\begin{equation}
-    %    k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\sfrac{1}{2\lceil \log%(\delta^{-1}) \rceil}}
-    %    = \lceil \log(\delta^{-1}) \rceil e^{\sfrac{1}{2}\frac{\log(\delta^{-1})}%{\lceil \log(\delta^{-1}) \rceil}}
-    %    \leq 2 \log(\delta^{-1}) e^{\sfrac{1}{2}},
-    %\end{equation}
-    %from which follows the second part of the theorem when $n_{\mtx{\Psi}} = (2^5 e^%{\sfrac{1}{2}} \varepsilon^{-1} \log(\delta^{-1}))^2$.
+    To address the parameter-dependent case, we first note that the continuity assumption implies that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is measurable.
+Therefore, we can apply Minkowski's integral inequality~\cite[Theorem 202]{hardy-1952-inequalities} to conclude from~\refequ{equ:moment-bound-hanson-wright} that
+    \begin{align}
+        \mathbb{E}^{p}\left[ \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t  \right]
+        &\leq \int_{a}^{b} \mathbb{E}^{p}\left[ \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) \right]~\mathrm{d}t \notag \\
+        &\leq c \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}}  \right\} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t .
+    \end{align}
+    Consequently, by Markov's inequality,  the inequality 
+    \begin{equation} 
+        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq c \gamma \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}}  \right\} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
+    \end{equation}
+    holds for any $\gamma \geq 1$ with probability at least $1 - \gamma^{-p}$.
 
-    Setting $\gamma = e$ and choosing $k = \lfloor \log(\delta^{-1}) / 2 \rfloor$,
+    Setting $\gamma = e$ and choosing $p = \log(\delta^{-1}) \geq 1$, the second part of the theorem follows when taking $n_{\mtx{\Psi}} = \lceil c^2 e^2 \log(\delta^{-1}) \varepsilon^{-2} \rceil$ and restricting $\varepsilon \leq c e$.
     %where $\lfloor x \rfloor$ is the floor function which returns the largest integer smaller than or equal to %$x$,
     %we get that for all $\delta \in (0, 1)$
-    the second part of the theorem follows when taking $n_{\mtx{\Psi}} = \lceil (2^3 e \varepsilon^{-1} \log(\delta^{-1}))^2 \rceil$.
-    %where $\lceil x \rceil$ is the ceiling function which returns the smallest integer larger than or equal to $x$.
-    %Consequently, with probability $1 - \delta$ we have
-    %\begin{equation}
-    %    \int_{a}^{b} \left| \frac{1}{n_{\mtx{\Psi}}} \sum_{i=1}^{n_{\mtx{\Psi}}} \vct{\psi}_i^{\top} \mtx{B}(t) \vct{\psi}_i - \Trace(\mtx{B}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\sfrac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
-    %\end{equation}
 \end{proof}
 
-Compared to the analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, the number of queries $n_{\mtx{\Psi}}$ needed by~\refthm{thm:hutchinson} is roughly a factor $\log(\delta^{-1})$ larger.
+\refthm{thm:hutchinson} coincides with the analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, except for a slightly more restricted choice of $\varepsilon$ and $\delta$.
 
+\color{blue}
+
+%The idea behind the proof is to bound  to extend the results to the continuous parameter case.
+We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
+%\begin{proof} 
+%\textcolor{red}{??? I feel that the block diagonal embedding trick should come first, like in the proof of Meyer et al. Hopefully, this gets rid of the additional $\log \delta$ factor ???}
+%The proof proceeds by bounding moments for fixed values of $t$ and then use Minkowski's integral inequality.
+We start by considering the residual of the 1-query Girard-Hutchinson estimator for a fixed matrix $\mtx{B}$:
+%    \begin{equation}
+%        r(\mtx{B}, \vct{\psi}) := \Trace(\mtx{B}) - \vct{\psi}^{\top} \mtx{B} \vct{\psi}
+%    \end{equation}
+%%    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers
+%for a Gaussian random vector $\vct{\psi}$.
+%%
+%%    First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence.
+%The proof of \cite[Lemma 3]{cortinovis-2022-randomized-trace} shows that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm.
+By \cite[Theorem 2.3]{boucheron-2013-concentration-inequalities} this implies for every $k \in \mathbb{N}$ that
+%    \begin{align*}
+%        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
+%        &\leq k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k}  \\
+%        &\le \big( k! 2^{4 k} + (2 k)! 2^{6 k}\big) \lVert \mtx{B} \rVert _F^{2 k} 
+%        \le \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
+%    \end{align*}
+%%     Because $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
+%%     \begin{equation}
+%%         \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right] \leq 
+%%     \end{equation}
+%    Stirling's approximation \cite{robbins-1955-remark-stirling} gives $(2 k)!
+< 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$.
+\textcolor{red}{???
+this norm should be defined more prominently somewhere else Consequently, the $L^{2k}$ norm $\mathbb{E}^{2k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{2k} \right] \right)^{\sfrac{1}{2k}}$ ???
+} \textcolor{green}{Moved it to start of section.} Consequently, the high order moments of $r(\mtx{B}, \vct{\psi})$ are bounded by
+%    \begin{equation}
+%        \mathbb{E}^{2k}\left[ r(\mtx{B}, \vct{\psi}) \right]
+%        \leq \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F \leq 2^4 k \lVert \mtx{B} \rVert _F,
+%        \label{equ:hutchinson-trace-onequery-fixed}
+%    \end{equation}
+%    using the inequality $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, which can be verified using the monotonic decrease of the left-hand side with respect to $k \in \mathbb{N}$.
+%
+%    To address the parameter-dependent case, we first note that the continuity assumption implies that $r(\mtx{B}(t), \vct{\psi})$ is measurable.
+Therefore, we can apply Minkowski's integral inequality~\cite[Theorem 202]{hardy-1952-inequalities} to conclude from~\refequ{equ:hutchinson-trace-onequery-fixed} that
+%    \begin{equation}
+%        \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t  \right]
+%        \leq \int_{a}^{b} \mathbb{E}^{2 k}\left[ r(\mtx{B}(t), \vct{\psi}) \right]~\mathrm{d}t
+%        \leq 2^4 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
+%    \end{equation}
+%    Consequently, by Markov's inequality,  the inequality 
+%    \begin{equation} 
+%        \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t \leq 2^4 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
+%        \label{equ:hutchinson-trace-onequery-uniform}
+%    \end{equation}
+%    holds for any $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$. 
+%
+%    To extend the bound~\eqref{equ:hutchinson-trace-onequery-uniform} from $1$ to $n_{\mtx{\Psi}}$ queries, we use a standard embedding trick (see, e.g.,~\cite[Theorem 1]{cortinovis-2022-randomized-trace}).
+Letting
+%    \begin{equation}
+%        \widetilde{\mtx{B}}(t)
+%        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
+%            \mtx{B}(t) & & \\
+%            & \ddots & \\
+%            & & \mtx{B}(t)
+%        \end{pmatrix}
+%        \quad \text{and} \quad
+%        \widetilde{\vct{\psi}} = \begin{pmatrix}
+%            \vct{\psi}_1 \\
+%            \vdots \\
+%            \vct{\psi}_{n_{\mtx{\Psi}}}
+%        \end{pmatrix},
+%    \end{equation}
+%    we note that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) = r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold.
+Thus, the claimed probabilistic bound~\eqref{eq:ghbound} follows from applying~\eqref{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$.
+%%     Then 
+%%     where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors.
+Then $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$, and
+%%     \begin{equation}
+%%         r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}}) = \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j.
+%%     \end{equation}
+%%     Hence, by applying \refequ{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$ we conclude that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
+%%     \begin{equation}
+%%         \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j \right| ~ \mathrm{d}t
+%%         \leq 2^4 k \gamma \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
+%%     \end{equation}
+%
+%    %Setting $\delta = \gamma^{-2 k}$ and choosing $k = \lceil \log(\delta^{-1}) %\rceil$, where $\lceil x \rceil$ is the ceiling function which returns the greatest %integer larger than or equal to $x$, we get that for all $\delta \in (0, e^{-\sfrac%{1}{2}})$
+%    %\begin{equation}
+%    %    k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\sfrac{1}{2\lceil \log%(\delta^{-1}) \rceil}}
+%    %    = \lceil \log(\delta^{-1}) \rceil e^{\sfrac{1}{2}\frac{\log(\delta^{-1})}%{\lceil \log(\delta^{-1}) \rceil}}
+%    %    \leq 2 \log(\delta^{-1}) e^{\sfrac{1}{2}},
+%    %\end{equation}
+%    %from which follows the second part of the theorem when $n_{\mtx{\Psi}} = (2^5 e^%{\sfrac{1}{2}} \varepsilon^{-1} \log(\delta^{-1}))^2$.
+%
+%    Setting $\gamma = e$ and choosing $k = \lfloor \log(\delta^{-1}) / 2 \rfloor$,
+%    %where $\lfloor x \rfloor$ is the floor function which returns the largest integer smaller than or equal to %$x$,
+%    %we get that for all $\delta \in (0, 1)$
+%    the second part of the theorem follows when taking $n_{\mtx{\Psi}} = \lceil (2^3 e \varepsilon^{-1} \log(\delta^{-1}))^2 \rceil$.
+%    %where $\lceil x \rceil$ is the ceiling function which returns the smallest integer larger than or equal to $x$.
+%    %Consequently, with probability $1 - \delta$ we have
+%    %\begin{equation}
+%    %    \int_{a}^{b} \left| \frac{1}{n_{\mtx{\Psi}}} \sum_{i=1}^{n_{\mtx{\Psi}}} \vct{\psi}_i^{\top} \mtx{B}(t) \vct{\psi}_i - \Trace(\mtx{B}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\sfrac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
+%    %\end{equation}
+%\end{proof}
+%
+%Compared to the analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, the number of queries $n_{\mtx{\Psi}}$ needed by~\refthm{thm:hutchinson} is roughly a factor $\log(\delta^{-1})$ larger.
+%
 \subsubsection{Nyström approximation}
@@
     \end{equation}
-    \textcolor{green}{
     To match the decay rate of the moments of the first term in \refequ{equ:nystrom-proof-persson-bonud}, we will need to ensure that this term also is of order $\mathcal{O}(\Trace(\mtx{B}) / \sqrt{k})$.
@@
     \end{equation}
-    As opposed to \cite[Lemma B.3]{tropp-2023-randomized-algorithms}, our bound holds for all even $n_{\mtx{\Omega}} \in \mathbb{N}$, at the cost of a factor of $e^2$.}
+    As opposed to \cite[Lemma B.3]{tropp-2023-randomized-algorithms}, our bound holds for all even $n_{\mtx{\Omega}} \in \mathbb{N}$, at the cost of a factor of $e^2$.
 
@@
     Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(\delta^{-1})^2)$ with \emph{even} $n_{\mtx{\Omega}}$, then for $\delta \in (0, 1/2)$ and $\varepsilon > 0$ with probability at least $1 - \delta$
+If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(\delta^{-1}))$ with \emph{even} $n_{\mtx{\Omega}} \geq 4$, then for $\delta \in (0, 2e^{-1})$ and $\varepsilon \in (0, \min\{1, c^2e^2\})$ with probability at least $1 - \delta$
     \begin{equation}
@@
 \begin{proof}
-    By choosing $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\tilde{\varepsilon}^{-2} \log(\tilde{\delta}^{-1})^2)$ we get
+    By choosing $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\tilde{\varepsilon}^{-2} \log(\tilde{\delta}^{-1}))$ we get
     \begin{align}
@@
 
-Comparing \refthm{thm:nystrom-pp} with its equivalent for constant matrices \cite[Theorem 3.4]{persson-2022-improved-variants}, we notice a less favorable scaling of the required number of random vectors $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$ in terms of $\log(\delta^{-1})$.
+Comparing \refthm{thm:nystrom-pp} with its equivalent for constant matrices \cite[Theorem 3.4]{persson-2022-improved-variants}, we notice an additional factor of $\sqrt{\log(\delta^{-1})}$ in the required number of random vectors $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$.
 

### 2024-07-12 paper/analysis.tex
rewrite introduction of methods

@@
 
-\paragraph{Girard-Hutchinson estimator.} We can approximate the trace with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ stochastically independent standard Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$, i.e.
+\paragraph{Girard-Hutchinson estimator.} The trace of a parameter-dependent matrix can be approximated with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ stochastically independent standard Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$, i.e.
 vectors whose entries are independent standard Gaussian random variables, to form
@@
 Other choices for the distribution of the random vectors are possible, for example by uniformly sampling from $\{-1, +1\}$ or from the $(n-1)$-sphere.
-However, our theoretical developments only hold in the Gaussian case.
+Our theoretical developments only hold in the Gaussian case.
 
-\color{black}
+The Girard-Hutchinson estimator distinguishes itself for its simplicity.
+It has already been used to approximate traces of parameter-dependent matrices for spectral density estimation in the Haydock method \cite{haydock-1972-electronic-structure}, as well as the Delta-Gauss-Legendre and Delta-Gauss-Chebyshev methods of \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation}.
+However, as a Monte Carlo estimate, its accuracy is often quite disappointing: if we aim to increase the accuracy of the estimate by one digit, we would need to increase the number of queries by a factor of one hundred.
 
-\textcolor{red}{Moved from below: The Girard-Hutchinson estimator distinguishes itself for its simplicity.
-However, its bound does not take into account the structure of $\mtx{B}(t)$.
-The $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ requirement implies that if we aim to increase the accuracy of the estimate by one digit, we would need to increase the number of queries by a factor of $100$.
-}
-
-\paragraph{Nyström estimator}
-
-Write:
-\begin{itemize}
- \item fast singular value decay, intuitively sufficient to consider a low-rank approximation
- \item motivated by our application we only consider symmetric positive semi-definite case here (smoothing kernels are positive) for which Nystroem approximation is the right choice
-\end{itemize}
-
-\textcolor{red}{Moved from below:
-For symmetric positive semi-definite matrices whose singular values decay rapidly, the Nyström estimate \refequ{equ:nystrom-trace-estimator} can be significantly better than the Girard-Hutchinson estimate.
-This is reflected in the following theorem.}
-
-Alternatively, the trace of a symmetric matrix whose singular values decay quickly can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström approximation \cite{gittens-2013-revisiting-nystrom}
-
+\paragraph{Nyström estimator.} Matrices with fast singular value decay can well be approximated with a low-rank approximation.
+Intuitively, it should be sufficient to approximate the trace of a quadratic matrix with the trace of a good low-rank approximation.
+Our application, the spectral density estimation (\refsec{sec:application}) of symmetric matrices, requires us to approximate the trace of a symmetric positive semi-definite matrix, originating from a non-negative smoothing kernel.
+In this scenario, the Nyström approximation is the most suitable low-rank approximation \cite{gittens-2013-revisiting-nystrom}.
+For a standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$, it is defined as
 \begin{equation}
@@
 \end{equation}
-Then we can estimate the trace as $\Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t))$.
+We can estimate the trace of $\mtx{B}(t)$ with $\Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t))$.
 Thanks to the invariance of the trace under cyclic permutation of its arguments and the symmetry of the matrix, we may rewrite this estimator as
-
 \begin{equation}
@@
 \end{equation}
+This estimator forms the basis of the so-called spectrum sweeping method \cite{lin-2017-randomized-estimation}.
 
-\textcolor{red}{Moved from below: The downside of the Nyström estimator is that it will -- in general -- not work well enough on matrices whose singular values do not decay quickly, for example $\mtx{B}(t) \equiv \mtx{I}_n$.}
+Two different low-rank approximation algorithms for general parameter-dependent matrices have already been analyzed in \cite{kressner-2024-randomized-lowrank}.
+However, the downside of the low-rank approximation approach is that it will -- in general -- not work well enough on matrices whose singular values do not decay quickly.
+%Write:
+%\begin{itemize}
+% \item fast singular value decay, intuitively sufficient to consider a low-rank %approximation
+% \item motivated by our application we only consider symmetric positive %semi-definite case here (smoothing kernels are positive) for which Nystroem %approximation is the right choice
+%\end{itemize}
 
-\paragraph{Nyström++ estimator.} Finally, \cite{lin-2017-randomized-estimation} proposes an estimator which corrects for inaccuracies in the Nyström approximation \refequ{equ:nystrom-trace-estimator} by estimating the trace of its residual using the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}
+%\textcolor{red}{Moved from below:
+%For symmetric positive semi-definite matrices whose singular values decay %rapidly, the Nyström estimate \refequ{equ:nystrom-trace-estimator} can be %significantly better than the Girard-Hutchinson estimate.
+This is reflected in %the following theorem.}
+%
+%
+%Alternatively, the trace of a symmetric matrix whose singular values decay %quickly can be approximated well by using a Gaussian sketching matrix $\mtx%{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström %approximation \cite{gittens-2013-revisiting-nystrom}
+
+\paragraph{Nyström++ estimator.} Finally, \cite{lin-2017-randomized-estimation} proposes an estimator which is more accurate than the Girard-Hutchinson estimator and, unlike the Nyström approximation, does not rely on a fast singular value decay of $\mtx{B}(t)$.
+The estimator first computes the trace of the Nyström approximation \refequ{equ:nystrom-trace-estimator} of the matrix.
+Because this result may not always be satisfactory, the estimator then corrects for eventual inaccuracies in the Nyström approximation by estimating the trace of the approximation residual with the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}.
+The full estimator is
 \begin{equation}
-    \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) = \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) + \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)).
+    \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) = \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) + \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)),
     \label{equ:nystrompp-trace-estimator}
 \end{equation}
-This is the parameter-dependent analogue of the Nyström++ estimator \cite{persson-2022-improved-variants}, which itself is based on the Hutch++ estimator \cite{meyer-2021-hutch-optimal}.
+where the matrix indices again emphasize the use of a constant standard Gaussian random matrix $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$ for the Girard-Hutchinson estimator and the dimension reduction matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ for the Nyström approximation.
+
+This estimator is the parameter-dependent analogue of the Nyström++ estimator \cite{persson-2022-improved-variants}, which itself is based on the Hutch++ estimator \cite{meyer-2021-hutch-optimal}.
+For constant matrices $\mtx{B}$, these estimators were both shown to achieve a relative $\varepsilon$-error with just using $\mathcal{O}(\varepsilon^{-1})$ matrix-vector products, independent of the singular value decay of $\mtx{B}$.
 %We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.
@@
     To match the decay rate of the moments of the first term in \refequ{equ:nystrom-proof-persson-bonud}, we will need to ensure that this term also is of order $\mathcal{O}(\Trace(\mtx{B}) / \sqrt{k})$.
-We bound the moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ similarly to \cite[Lemma B.3]{tropp-2023-randomized-algorithms}, but without restricting the moments to $0 \leq q \leq 18$.
+We bound the moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ similarly to \cite[Lemma B.3]{tropp-2023-randomized-algorithms}, but without restricting $q$ to be smaller than $18$.
 The explicit integration of \cite[Equation B.7]{tropp-2023-randomized-algorithms} imposes the condition $n_{\mtx{\Omega}} - k - 2q \geq 0$.

### 2024-07-12 paper/intro.tex
rewrite introduction of methods

@@
 
-\paragraph{Structure.} In \refsec{sec:analysis} we will analyze three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter dependent matrices.
+\paragraph{Structure.} In \refsec{sec:analysis} we will analyze three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter-dependent matrices.
 We propose and analyze an algorithm for using the aforementioned estimators to approximate the spectral density of real symmetric matrices in \refsec{sec:application}.

### 2024-07-16 paper/analysis.tex
derive hutchinson moment bound with explicit constants

@@
 
+%\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
+%    For a symmetric matrix $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ that depends continuously on $t \in [a, b]$, consider the  Girard-Hutchinson estimator~\eqref{equ:hutchinson-trace-estimator} with an $n\times n_{\mtx{\Psi}}$ Gaussian random matrix $\mtx{\Psi}$.
+Then for any $p \geq 1, p \in \mathbb{R}$ and $\gamma \geq 1$ the bound
+%    \begin{equation} \label{eq:ghbound}
+%        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq c \gamma \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}}  \right\} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
+%    \end{equation}
+%    holds with probability at least $1 - \gamma^{-p}$.
+%    In particular, given $\varepsilon \in (0, ce)$ and $\delta \in (0, e^{-1})$, the bound $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$ holds
+%    with probability at least $1-\delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1}))$.
+%\end{theorem}
+
 \begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
@@
     \begin{equation} \label{eq:ghbound}
-        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq c \gamma \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}}  \right\} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
+        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq C \gamma \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}}  \right) \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
     \end{equation}
-    holds with probability at least $1 - \gamma^{-p}$.
-    In particular, given $\varepsilon \in (0, ce)$ and $\delta \in (0, e^{-1})$, the bound $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$ holds
+    holds with probability at least $1 - \gamma^{-p}$ and with $C = 8 e^{1/e} (1 + \sqrt{\pi}/2)$.
+    In particular, given $\varepsilon \in (0, 1)$ and $\delta \in (0, e^{-1})$, the bound $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$ holds
     with probability at least $1-\delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1}))$.
@@
     \end{equation}
-    Note that $\Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}} = \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$, $\lVert \widetilde{\mtx{B}}(t) \rVert _2 = \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}}$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold.
-Therefore, by \reflem{lem:hanson-wright-moments} we have
-    \begin{align}
-        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))]
-        &= \mathbb{E}^{p}[\Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}}] \notag \\
-        &\leq c \max\left\{ \sqrt{p} \lVert \widetilde{\mtx{B}}(t) \rVert _F, p \lVert \widetilde{\mtx{B}}(t) \rVert _2 \right\} \notag \\
-        &= c \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F, \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 \right\} \notag \\
-        &\leq c \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}} \right\}  \lVert \mtx{B}(t) \rVert _F.
-        \label{equ:moment-bound-hanson-wright}
-    \end{align}
+    Note that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) =  \Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}}$, $\lVert \widetilde{\mtx{B}}(t) \rVert _2 = \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}}$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold.
+Therefore, by \reflem{lem:hutchinson-sub-exponential}, $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is a centered, $(2 \lVert \mtx{B}(t) \rVert _F^2 \frac{1}{1 - 1/c} / n_{\mtx{\Psi}}, 2 \lVert \mtx{A} \rVert _2 c / n_{\mtx{\Psi}})$-sub-exponential random variable for any $c > 1$.
+Thus, we can apply \reflem{lem:sub-exponential-moments} to bound
+    \begin{equation}
+        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq 2 e^{1/e} \left(\frac{\sqrt{\pi}}{1-1/c} \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F + 4 c \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 \right).
+    \end{equation}
+    Bounding $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$ and letting $c = 1 + \sqrt{\pi} / 2$ we get
+    \begin{equation}
+        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq \underbrace{8 e^{1/e} (1 + \sqrt{\pi} / 2)}_{=~C} \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}} \right) \lVert \mtx{B}(t) \rVert _F.
+        \label{equ:moment-bound-hutchinson}
+    \end{equation}
 
     To address the parameter-dependent case, we first note that the continuity assumption implies that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is measurable.
-Therefore, we can apply Minkowski's integral inequality~\cite[Theorem 202]{hardy-1952-inequalities} to conclude from~\refequ{equ:moment-bound-hanson-wright} that
+Therefore, we can apply Minkowski's integral inequality~\cite[Theorem 202]{hardy-1952-inequalities} to conclude from the moment bound~\refequ{equ:moment-bound-hutchinson} that
     \begin{align}
@@
         &\leq \int_{a}^{b} \mathbb{E}^{p}\left[ \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) \right]~\mathrm{d}t \notag \\
-        &\leq c \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}}  \right\} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t .
+        &\leq C \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}}  \right) \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t .
     \end{align}
@@
     \begin{equation} 
-        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq c \gamma \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}}  \right\} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
+        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq \gamma C \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}}  \right) \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
     \end{equation}
@@
 
-    Setting $\gamma = e$ and choosing $p = \log(\delta^{-1}) \geq 1$, the second part of the theorem follows when taking $n_{\mtx{\Psi}} = \lceil c^2 e^2 \log(\delta^{-1}) \varepsilon^{-2} \rceil$ and restricting $\varepsilon \leq c e$.
-    %where $\lfloor x \rfloor$ is the floor function which returns the largest integer smaller than or equal to %$x$,
-    %we get that for all $\delta \in (0, 1)$
+    Setting $\gamma = e$ and choosing $p = \log(\delta^{-1}) \geq 1$, the second part of the theorem follows when taking $n_{\mtx{\Psi}} = \lceil 2 C^2 e^2 \log(\delta^{-1}) \varepsilon^{-2} \rceil$.
 \end{proof}
+%\begin{proof}
+%    The proof proceeds by bounding moments for fixed values of $t$ and then uses Minkowski's integral inequality.
+We start with a standard embedding trick (see, e.g.,~\cite[Theorem 1]{cortinovis-2022-randomized-trace}) for which we define
+%    \begin{equation}
+%        \widetilde{\mtx{B}}(t)
+%        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
+%            \mtx{B}(t) & & \\
+%            & \ddots & \\
+%            & & \mtx{B}(t)
+%        \end{pmatrix}
+%        \quad \text{and} \quad
+%        \widetilde{\vct{\psi}} = \begin{pmatrix}
+%            \vct{\psi}_1 \\
+%            \vdots \\
+%            \vct{\psi}_{n_{\mtx{\Psi}}}
+%        \end{pmatrix}.
+%    \end{equation}
+%    Note that $\Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}} = \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$, $\lVert \widetilde{\mtx{B}}(t) \rVert _2 = \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}}$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold.
+Therefore, by \reflem{lem:hanson-wright-moments} we have
+%    \begin{align}
+%        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))]
+%        &= \mathbb{E}^{p}[\Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}}] \notag \\
+%        &\leq c \max\left\{ \sqrt{p} \lVert \widetilde{\mtx{B}}(t) \rVert _F, p \lVert \widetilde{\mtx{B}}(t) \rVert _2 \right\} \notag \\
+%        &= c \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F, \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 \right\} \notag \\
+%        &\leq c \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}} \right\}  \lVert \mtx{B}(t) \rVert _F.
+%        \label{equ:moment-bound-hanson-wright}
+%    \end{align}
+%
+%    To address the parameter-dependent case, we first note that the continuity assumption implies that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is measurable.
+Therefore, we can apply Minkowski's integral inequality~\cite[Theorem 202]{hardy-1952-inequalities} to conclude from~\refequ{equ:moment-bound-hanson-wright} that
+%    \begin{align}
+%        \mathbb{E}^{p}\left[ \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t  \right]
+%        &\leq \int_{a}^{b} \mathbb{E}^{p}\left[ \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) \right]~\mathrm{d}t \notag \\
+%        &\leq c \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}}  \right\} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t .
+%    \end{align}
+%    Consequently, by Markov's inequality,  the inequality 
+%    \begin{equation} 
+%        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq c \gamma \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}}  \right\} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
+%    \end{equation}
+%    holds for any $\gamma \geq 1$ with probability at least $1 - \gamma^{-p}$. 
+%
+%    Setting $\gamma = e$ and choosing $p = \log(\delta^{-1}) \geq 1$, the second part of the theorem follows when taking $n_{\mtx{\Psi}} = \lceil c^2 e^2 \log(\delta^{-1}) \varepsilon^{-2} \rceil$ and restricting $\varepsilon \leq c e$.
+%    %where $\lfloor x \rfloor$ is the floor function which returns the largest integer smaller than or equal to %$x$,
+%    %we get that for all $\delta \in (0, 1)$
+%\end{proof}
 
-\refthm{thm:hutchinson} coincides with the analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, except for a slightly more restricted choice of $\varepsilon$ and $\delta$.
+\refthm{thm:hutchinson} coincides with the analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, except for a slightly more restricted choice of $\delta$.
 
@@
     Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
-If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(\delta^{-1}))$ with \emph{even} $n_{\mtx{\Omega}} \geq 4$, then for $\delta \in (0, 2e^{-1})$ and $\varepsilon \in (0, \min\{1, c^2e^2\})$ with probability at least $1 - \delta$
+If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(\delta^{-1}))$ with \emph{even} $n_{\mtx{\Omega}} \geq 4$, then for $\delta \in (0, 2e^{-1})$ and $\varepsilon \in (0, 1)$ with probability at least $1 - \delta$
     \begin{equation}

### 2024-07-16 paper/analysis.tex
rewrite proof slightly and add missing paragraph in results

@@
     Note that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) =  \Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}}$, $\lVert \widetilde{\mtx{B}}(t) \rVert _2 = \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}}$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold.
-Therefore, by \reflem{lem:hutchinson-sub-exponential}, $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is a centered, $(2 \lVert \mtx{B}(t) \rVert _F^2 \frac{1}{1 - 1/c} / n_{\mtx{\Psi}}, 2 \lVert \mtx{A} \rVert _2 c / n_{\mtx{\Psi}})$-sub-exponential random variable for any $c > 1$.
+Therefore, by \reflem{lem:hutchinson-sub-exponential}, $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is a centered and $(2 \lVert \mtx{B}(t) \rVert _F^2 \frac{1}{1 - 1/c} / n_{\mtx{\Psi}}, 2 \lVert \mtx{A} \rVert _2 c / n_{\mtx{\Psi}})$-sub-exponential random variable for any $c > 1$.
 Thus, we can apply \reflem{lem:sub-exponential-moments} to bound

### 2024-07-16 paper/numresults.tex
clean up bibliography and add acknowledgments
rewrite proof slightly and add missing paragraph in results

@@
-\section{Numerical results}
+\section{Numerical experiments}
 \label{sec:results}
+
+\subsection{Spectral density for Hamiltonian of electronic structure}
+\label{subsec:hamiltonian}
 
@@
 
+For a fixed budget of random vectors $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}$, \reffig{fig:distribution} visualizes how for different values of the smoothing parameter $\sigma$ the random vectors should either be invested into the Nyström approximation $n_{\mtx{\Omega}}$ or the estimation of the residual trace $n_{\mtx{\Psi}}$.
+Again, since for small $\sigma$ the matrix function $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ has a small numerical rank, a Nyström approximation alone is already enough to achieve a good approximation.
+On the other hand, when $\sigma$ is large, the Nyström approximation alone is not effective, and the correction with the residual trace becomes indispensable.
+
 \begin{figure}[ht]
@@
 
-\todo{Interpret results.}
+\subsection{Spectral density of Hessian matrix of neural network}
+\label{subsec:hessian}
 
@@
 
+\subsection{Trace of parameter-dependent matrix functions}
+\label{subsec:function-trace}
+
 \todo{Comparison plots with other estimators (Krylov-Aware)?}

### 2024-07-17 paper/analysis.tex
make proof of sub-exponentiality more understandable

--- 2024-07-16 paper/analysis.tex
@@
     \begin{equation} \label{eq:ghbound}
-        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq C \gamma \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}}  \right) \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
+        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq \gamma c \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}}  \right) \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
     \end{equation}
-    holds with probability at least $1 - \gamma^{-p}$ and with $C = 8 e^{1/e} (1 + \sqrt{\pi}/2)$.
+    holds with probability at least $1 - \gamma^{-p}$ and for $c = 8 e^{\sfrac{1}{e}} (1 + \sqrt{\pi}/2)$.
     In particular, given $\varepsilon \in (0, 1)$ and $\delta \in (0, e^{-1})$, the bound $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$ holds
@@
     Note that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) =  \Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}}$, $\lVert \widetilde{\mtx{B}}(t) \rVert _2 = \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}}$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold.
-Therefore, by \reflem{lem:hutchinson-sub-exponential}, $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is a centered, $(2 \lVert \mtx{B}(t) \rVert _F^2 \frac{1}{1 - 1/c} / n_{\mtx{\Psi}}, 2 \lVert \mtx{A} \rVert _2 c / n_{\mtx{\Psi}})$-sub-exponential random variable for any $c > 1$.
+Therefore, by \reflem{lem:hutchinson-sub-exponential}, $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is a centered and $(2 \lVert \mtx{B}(t) \rVert _F^2 (1 - 1/\beta)^{-1} / n_{\mtx{\Psi}}, 2 \lVert \mtx{A} \rVert _2 \beta / n_{\mtx{\Psi}})$-sub-exponential random variable for any $\beta > 1$.
 Thus, we can apply \reflem{lem:sub-exponential-moments} to bound
     \begin{equation}
-        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq 2 e^{1/e} \left(\frac{\sqrt{\pi}}{1-1/c} \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F + 4 c \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 \right).
+        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq 2 e^{\sfrac{1}{e}} \left(\frac{\sqrt{\pi}}{1-1/\beta} \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F + 4 \beta \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 \right).
     \end{equation}
-    Bounding $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$ and letting $c = 1 + \sqrt{\pi} / 2$ we get
+    Bounding $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$ and letting $\beta = 1 + \sqrt{\pi} / 2$ we get
     \begin{equation}
-        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq \underbrace{8 e^{1/e} (1 + \sqrt{\pi} / 2)}_{=~C} \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}} \right) \lVert \mtx{B}(t) \rVert _F.
+        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq \underbrace{8 e^{\sfrac{1}{e}} (1 + \sqrt{\pi} / 2)}_{=~c} \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}} \right) \lVert \mtx{B}(t) \rVert _F.
         \label{equ:moment-bound-hutchinson}
@@
         &\leq \int_{a}^{b} \mathbb{E}^{p}\left[ \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) \right]~\mathrm{d}t \notag \\
-        &\leq C \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}}  \right) \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t .
+        &\leq c \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}}  \right) \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t .
     \end{align}
@@
     \begin{equation} 
-        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq \gamma C \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}}  \right) \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
+        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq \gamma c \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}}  \right) \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
     \end{equation}
@@
 
-    Setting $\gamma = e$ and choosing $p = \log(\delta^{-1}) \geq 1$, the second part of the theorem follows when taking $n_{\mtx{\Psi}} = \lceil 2 C^2 e^2 \log(\delta^{-1}) \varepsilon^{-2} \rceil$.
+    Setting $\gamma = e$ and choosing $p = \log(\delta^{-1}) \geq 1$, the second part of the theorem follows when taking $n_{\mtx{\Psi}} = \lceil 2 c^2 e^2 \log(\delta^{-1}) \varepsilon^{-2} \rceil$.
 \end{proof}

### 2024-07-17 paper/re-pro-badge.tex

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/75a23a4}{75a23a4} on 2024-07-11 at 06:15:48 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/a3fd6a9}{a3fd6a9} on 2024-07-16 at 16:59:46 UTC.

### 2024-07-18 paper/analysis.tex
improve and discuss sub-exponential bound

@@
 %\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
-%    For a symmetric matrix $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ that depends continuously on $t \in [a, b]$, consider the  Girard-Hutchinson estimator~\eqref{equ:hutchinson-trace-estimator} with an $n\times n_{\mtx{\Psi}}$ Gaussian random matrix $\mtx{\Psi}$.
+%    For a symmetric matrix $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ that depends continuously on $t \in [a, b]$, consider the  Girard-Hutchinson estimator~\refequ{equ:hutchinson-trace-estimator} with an $n\times n_{\mtx{\Psi}}$ Gaussian random matrix $\mtx{\Psi}$.
 Then for any $p \geq 1, p \in \mathbb{R}$ and $\gamma \geq 1$ the bound
@@
 \begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
-    For a symmetric matrix $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ that depends continuously on $t \in [a, b]$, consider the  Girard-Hutchinson estimator~\eqref{equ:hutchinson-trace-estimator} with an $n\times n_{\mtx{\Psi}}$ Gaussian random matrix $\mtx{\Psi}$.
+    For a symmetric matrix $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ that depends continuously on $t \in [a, b]$, consider the  Girard-Hutchinson estimator~\refequ{equ:hutchinson-trace-estimator} with an $n\times n_{\mtx{\Psi}}$ Gaussian random matrix $\mtx{\Psi}$.
 Then for any $p \geq 1, p \in \mathbb{R}$ and $\gamma \geq 1$ the bound
@@
     Note that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) =  \Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}}$, $\lVert \widetilde{\mtx{B}}(t) \rVert _2 = \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}}$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold.
-Therefore, by \reflem{lem:hutchinson-sub-exponential}, $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is a centered and $(2 \lVert \mtx{B}(t) \rVert _F^2 (1 - 1/\beta)^{-1} / n_{\mtx{\Psi}}, 2 \lVert \mtx{A} \rVert _2 \beta / n_{\mtx{\Psi}})$-sub-exponential random variable for any $\beta > 1$.
+Therefore, by \reflem{lem:hutchinson-sub-exponential}, $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is a centered and $(2 \lVert \mtx{B}(t) \rVert _F^2 (1 - 1/\beta)^{-1} / n_{\mtx{\Psi}}, 2 \lVert \mtx{B}(t) \rVert _2 \beta / n_{\mtx{\Psi}})$-sub-exponential random variable for any $\beta > 1$.
 Thus, we can apply \reflem{lem:sub-exponential-moments} to bound
     \begin{equation}
-        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq 2 e^{\sfrac{1}{e}} \left(\frac{\sqrt{\pi}}{1-1/\beta} \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F + 4 \beta \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 \right).
+        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq 2 \left(\sqrt{\frac{\pi}{1-1/\beta}} \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F + 4 \beta \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 \right).
     \end{equation}
-    Bounding $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$ and letting $\beta = 1 + \sqrt{\pi} / 2$ we get
+    Bounding $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$ and letting $\beta = (1 + \sqrt{1 + \pi/4}) / 2$ we get
     \begin{equation}
-        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq \underbrace{8 e^{\sfrac{1}{e}} (1 + \sqrt{\pi} / 2)}_{=~c} \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}} \right) \lVert \mtx{B}(t) \rVert _F.
+        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq \underbrace{4 (1 + \sqrt{1 + \pi/4})}_{=~c} \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}} \right) \lVert \mtx{B}(t) \rVert _F.
         \label{equ:moment-bound-hutchinson}
@@
 %
-%    To extend the bound~\eqref{equ:hutchinson-trace-onequery-uniform} from $1$ to $n_{\mtx{\Psi}}$ queries, we use a standard embedding trick (see, e.g.,~\cite[Theorem 1]{cortinovis-2022-randomized-trace}).
+%    To extend the bound~\refequ{equ:hutchinson-trace-onequery-uniform} from $1$ to $n_{\mtx{\Psi}}$ queries, we use a standard embedding trick (see, e.g.,~\cite[Theorem 1]{cortinovis-2022-randomized-trace}).
 Letting
@@
 %    we note that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) = r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold.
-Thus, the claimed probabilistic bound~\eqref{eq:ghbound} follows from applying~\eqref{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$.
+Thus, the claimed probabilistic bound~\refequ{eq:ghbound} follows from applying~\refequ{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$.
 %%     Then 
@@
     Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuous for $t \in [a, b]$.
-For integers $r \geq 2$ and $n_{\mtx{\Omega}} \geq r + 4$, consider the Nystr\"om approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ defined in~\eqref{equ:nystrom-approximation} with an $n\times n_{\mtx{\Omega}}$ Gaussian random matrix $\mtx{\Omega}$.
+For integers $r \geq 2$ and $n_{\mtx{\Omega}} \geq r + 4$, consider the Nystr\"om approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ defined in~\refequ{equ:nystrom-approximation} with an $n\times n_{\mtx{\Omega}}$ Gaussian random matrix $\mtx{\Omega}$.
 Then for any $\gamma \geq 1$ the bound

### 2024-07-18 algorithms/chebyshev_nystrom.py
make chebyshev-nystrom compatible with general parametric trace estimation

@@ -35,11 +35,11 @@ def chebyshev_expansion(kernel, t, m, nonnegative=False):
     if nonnegative:
         m = m // 2
         kernel_ = kernel # Avoid recursion
-        kernel = lambda x: np.sqrt(kernel_(x))
+        kernel = lambda t, x: np.sqrt(kernel_(t, x))
 
     # Compute the coefficients mu for all t and l simultaneously with DCT
     chebyshev_nodes = np.cos(np.arange(m + 1) * np.pi / m)
-    mu = sp.fft.idct(kernel(np.subtract.outer(t, chebyshev_nodes)), type=1)
+    mu = sp.fft.idct(kernel(t, chebyshev_nodes), type=1)
 
     # Rescale coefficients due to type-1 DCT convention
     mu[..., 1:-1] *= 2
@@ -97,7 +97,7 @@ def chebyshev_nystrom(A, t=0, m=100, n_Psi=10, n_Omega=10, kernel=gaussian_kerne
         Number of queries with Girard-Hutchinson estimator.
     n_Omega : int > 0
         Size of sketching matrix in Nyström approximation.
-    kernel : function
+    kernel : callable
         Smoothing kernel.
     nonnegative : bool
         Use non-negative Chebyshev expansion.
@@ -130,7 +130,7 @@ def chebyshev_nystrom(A, t=0, m=100, n_Psi=10, n_Omega=10, kernel=gaussian_kerne
     if consistent:
         nu = square_chebyshev_expansion(mu)
     else:
-        nu = chebyshev_expansion(lambda s: kernel(s) ** 2, t, 2 * m, nonnegative=nonnegative)
+        nu = chebyshev_expansion(lambda t, x: kernel(t, x) ** 2, t, 2 * m, nonnegative=nonnegative)
 
     # Generate Gaussian random matrices
     Omega = np.random.randn(n, n_Omega)

### 2024-07-18 algorithms/helpers.py

@@ -35,23 +35,27 @@ def spectral_transformation(A, min_ev=None, max_ev=None):
 
     return A_st
 
-def gaussian_kernel(s, sigma=0.1):
+def gaussian_kernel(t, x=None, sigma=0.1):
     """
     Gaussian kernel.
 
     Parameters
     ----------
-    s : int, float, or np.ndarray of shape (n, m)
-        Point(s) where the Gaussian should be evaluated.
+    t : int, float, or np.ndarray of shape (n, m)
+        Parameter value(s) where the Gaussian should be evaluated.
+    x : int, float, or np.ndarray of shape (n, m)
+        Value(s) outer-subtracted from paramter value(s) 
     sigma : int or float
         Standard deviation of the Gaussian.
 
     Returns
     -------
-    g(s) : np.ndarray of shape (n, dim)
-        The Gaussian kernel evaulated at all points s.
+    g(t, x) : np.ndarray of shape (n, dim)
+        The Gaussian kernel evaulated at all combinations of points t - x
     """
-    return np.exp(- s**2 / (2 * sigma**2)) / (np.sqrt(2 * np.pi) * sigma)
+    if x is not None:
+        t = np.subtract.outer(t, x)
+    return np.exp(- t**2 / (2 * sigma**2)) / (np.sqrt(2 * np.pi) * sigma)
 
 def form_spectral_density(eigvals, t, kernel=gaussian_kernel):
     """
@@ -72,4 +76,4 @@ def form_spectral_density(eigvals, t, kernel=gaussian_kernel):
     spectral_density : np.ndarray of shape (n_t,)
         The value of the spectral density evaluated at the grid points.
     """
-    return kernel(np.subtract.outer(t, eigvals)).sum(axis=1) / len(eigvals)
+    return kernel(t, eigvals).sum(axis=1) / len(eigvals)

### 2024-07-18 algorithms/krylov_aware.py

@@ -243,6 +243,6 @@ def krylov_aware(A, t=0, n_iter=10, n_reorth=5, n_Omega=10, n_Psi=10, kernel=gau
     if kernel is None:
         fun = lambda x: x
     else:
-        fun = lambda x: kernel(np.subtract.outer(t, x)) / A.shape[0]
+        fun = lambda x: kernel(x) / A.shape[0]
 
     return fun(nodes) @ weights

### 2024-07-18 matrices/neural_network.py

@@ -158,7 +158,8 @@ def hessian_transpose_product(gradient, params, x):
     ----------
     [1] TODO
     """
-    htp = torch.autograd.grad(gradient, params, x.T, retain_graph=True, is_grads_batched=X.ndim > 1)
+    htp = torch.autograd.grad(gradient, params, x.T, retain_graph=True, is_grads_batched=x.ndim > 1)
+    htp = torch.cat([e.reshape(x.shape[-1], -1) for e in htp], dim=1)
     return htp.T
 
 
@@ -223,8 +224,7 @@ class hessian(object):
         # Compute the product of the Hessian transpose with x
         htp = hessian_transpose_product(self.gradient, self.parameters, x)
 
-        # Flatten and rescale the result (for normalized Hessian)
-        htp = torch.cat([e.reshape(x.shape[-1], -1) for e in htp], dim=1)
+        # Rescale the result (for normalized Hessian)
         htp /= self.scaling_parameter
 
         # Convert back to numpy array

### 2024-07-18 paper/plots/convergence.py

@@ -29,7 +29,7 @@ markers = ["o", "s", "d"]
 labels = [r"$n_{\mathbf{\Omega}} = 0$", r"$n_{\mathbf{\Psi}} = n_{\mathbf{\Omega}}$", r"$n_{\mathbf{\Psi}} = 0$"]
 
 # Determine the baseline spectral density
-kernel = lambda x: gaussian_kernel(x, sigma=sigma)
+kernel = lambda t, x: gaussian_kernel(t, x, sigma=sigma)
 baseline = form_spectral_density(eigvals_st, t, kernel)
 
 error = np.empty((3, len(n_Vec_list)))

### 2024-07-18 paper/plots/interpolation.py

@@ -29,7 +29,7 @@ markers = ["o", "s", "d"]
 labels = ["inconsistent", "consistent", "non-negative"]
 
 # Determine the baseline spectral density
-kernel = lambda x: gaussian_kernel(x, sigma=sigma)
+kernel = lambda t, x: gaussian_kernel(t, x, sigma=sigma)
 baseline = form_spectral_density(eigvals_st, t, kernel)
 
 error = np.empty((3, len(m_list)))

### 2024-07-18 paper/plots/zerocheck.py

@@ -29,7 +29,7 @@ labels = ["baseline", "without zero-check", "with zero-check"]
 kappa_list = [-1, 1e-5]
 
 # Determine the baseline spectral density
-kernel = lambda x: gaussian_kernel(x, sigma=sigma)
+kernel = lambda t, x: gaussian_kernel(t, x, sigma=sigma)
 baseline = form_spectral_density(eigvals_st, t, kernel)
 
 estimate = [baseline]

### 2024-07-18 test/test_chebyshev_nystrom.py

@@ -15,7 +15,7 @@ def test_chebyshev_expansion():
     # For some linearly spaced t in [-1, 1], check if approximation is exact
     t_list = np.linspace(-1, 1, 100)
     s_list = np.linspace(-1, 1, 100)
-    kernel = lambda x: gaussian_kernel(x, sigma=sigma)
+    kernel = lambda t, x: gaussian_kernel(t, x, sigma=sigma)
 
     # Test standard Chebyshev expansion of kernel against theoretical bound
     bound = lambda sigma, m: np.sqrt(2 * np.e / np.pi) / sigma ** 2 * (1 + sigma) ** (-m)
@@ -24,7 +24,7 @@ def test_chebyshev_expansion():
         # Compute function values from Chebyshev approximation
         mu = chebyshev_expansion(kernel, t_list, m)
         for i, t in enumerate(t_list):
-            truth = gaussian_kernel(t - s_list, sigma=sigma)
+            truth = gaussian_kernel(t, s_list, sigma=sigma)
             approx = np.polynomial.chebyshev.Chebyshev(mu[i])(s_list)
 
             np.testing.assert_allclose(approx, truth, atol=bound(sigma, m))
@@ -36,7 +36,7 @@ def test_chebyshev_expansion():
         # Compute function values from Chebyshev approximation
         mu = chebyshev_expansion(kernel, t_list, m, nonnegative=True)
         for i, t in enumerate(t_list):
-            truth = gaussian_kernel(t - s_list, sigma=sigma)
+            truth = gaussian_kernel(t, s_list, sigma=sigma)
             approx = np.polynomial.chebyshev.Chebyshev(mu[i])(s_list)
 
             np.testing.assert_allclose(approx, truth, atol=nonnegative_bound(sigma, m))
@@ -67,13 +67,13 @@ def test_square_chebyshev_expansion():
     squared_bound = lambda sigma, m:  (2 / (sigma * np.sqrt(2 * np.pi)) + bound(sigma, m)) * bound(sigma, m)
 
     for m in m_list:
-        kernel = lambda x: gaussian_kernel(x, sigma=sigma)
+        kernel = lambda t, x: gaussian_kernel(t, x, sigma=sigma)
 
         # Compute function values from Chebyshev approximation
         mu = chebyshev_expansion(kernel, t_list, m)
         nu = square_chebyshev_expansion(mu)
         for i, t in enumerate(t_list):
[…]

### 2024-07-18 paper/re-pro-badge.tex
automatically reproduce paper

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/a3fd6a9}{a3fd6a9} on 2024-07-16 at 16:59:46 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/bca2ff7}{bca2ff7} on 2024-07-18 at 14:29:53 UTC.

### 2024-07-19 algorithms/krylov_aware.py
run numerical experiments

@@ -241,8 +241,8 @@ def krylov_aware(A, t=0, n_iter=10, n_reorth=5, n_Omega=10, n_Psi=10, kernel=gau
         weights = np.append(weights, weights_)
 
     if kernel is None:
-        fun = lambda x: x
+        fun = lambda t, x: x
     else:
-        fun = lambda x: kernel(x) / A.shape[0]
+        fun = lambda t, x: kernel(t, x) / A.shape[0]
 
-    return fun(nodes) @ weights
+    return fun(t, nodes) @ weights

### 2024-07-19 paper/plots/hessian_density.py

@@ -1,3 +1,9 @@
+"""
+hessian_density.py
+
+TODO: Only look at spectrum at range far in the negative or positive
+"""
+
 import __context__
 
 import numpy as np
@@ -7,6 +13,17 @@ import matplotlib.pyplot as plt
 from algorithms.chebyshev_nystrom import chebyshev_nystrom
 from algorithms.helpers import gaussian_kernel
 from matrices.neural_network import model, train, accuracy_validator, hessian
+
+model = torch.nn.Sequential(
+    torch.nn.Conv2d(1, 1, kernel_size=(3, 3)),
+    torch.nn.Softplus(),
+    torch.nn.BatchNorm2d(1),
+    torch.nn.Flatten(),
+    torch.nn.Linear(676, 10),
+    #torch.nn.Softplus(),
+    #torch.nn.Linear(256, 10),
+    torch.nn.Sigmoid(),
+)
 
 # For now, don't execute this
 if __name__ == "__main__":
@@ -33,17 +50,17 @@ test_loader = torch.utils.data.DataLoader(torchvision.datasets.MNIST("matrices/m
 # Train the model on the MNIST data set
 loss_function = lambda x, y: torch.nn.MSELoss()(x, torch.nn.functional.one_hot(y, num_classes=10).to(dtype=torch.float))
 validator = lambda model: accuracy_validator(model, test_loader)
-train(model, train_loader, loss_function, validator=validator, n_epochs=30)
+train(model, train_loader, loss_function, validator=validator, n_epochs=5)
 
 data = next(iter(train_loader))
 H = hessian(model, loss_function, data, spectral_transform=True)
 
 # Set parameter
-t = np.linspace(-0.1, 0.1, 100)
-sigma = 0.02
-m = 700
-n_Omega = 80
-n_Psi = 20
+t = np.linspace(-0.02, 0.02, 30)
+sigma = 0.003
+m = 1500
+n_Omega = 30
+n_Psi = 10
 
 plt.style.use("paper/plots/stylesheet.mplstyle")
 

### 2024-07-19 paper/plots/quantum_spin.py

@@ -1,0 +1,55 @@
+import __context__
+
+import scipy as sp
+import numpy as np
+
+from matrices.quantum_spin import hamiltonian, partition_function
+from algorithms.chebyshev_nystrom import chebyshev_nystrom
+from algorithms.krylov_aware import krylov_aware
+from algorithms.helpers import spectral_transformation
+import matplotlib.pyplot as plt
+
+N = 20
+s = 0.5
+h = 0.3
+J = 1
+A = hamiltonian(N, s, h, J)
+
+# For now, don't execute this
+if __name__ == "__main__":
+    exit()
+
+E_min, = sp.sparse.linalg.eigsh(A, k=1, which="SA", return_eigenvectors=False, tol=1e-5)
+E_max, = sp.sparse.linalg.eigsh(A, k=1, which="LA", return_eigenvectors=False, tol=1e-5)
+
+betas = 1 / np.logspace(-2.5, 3, 50)
+Z_true = partition_function(betas, N, h, J, E_min)
+
+A_transformed = spectral_transformation(A, E_min, E_max)
+m = 30
+n_Omega = 150
+n_Psi = 150
+
+function = lambda beta, x: np.exp(-np.multiply.outer(beta, x + 1))
+Z_nycheb = chebyshev_nystrom(A_transformed, betas * (E_max - E_min) / 2, m, n_Psi, n_Omega, function, kappa=-1, rcond=1e-10)
+Z_nycheb *= np.exp(- betas * (E_min + E_max) / 2)
+
+n_iter = 80
+n_reorth = 50
+n_Omega = 8
+n_Psi = 13
+
+function = lambda beta, x: np.exp(-np.multiply.outer(beta, x - E_min))
+Z_krylov = krylov_aware(A, betas, n_iter, n_reorth, n_Omega, n_Psi, function)
+
+plt.style.use("paper/plots/stylesheet.mplstyle")
+
+plt.plot(1 / betas, np.abs(1 - Z_nycheb * A.shape[0] / Z_true), color="#648FFF")  # TODO: CHANGE KERNEL TO NORMALIZED
+plt.plot(1 / betas, np.abs(1 - Z_krylov * A.shape[0] / Z_true), color="#FFB000")  # TODO: CHANGE KERNEL TO NORMALIZED
+
+plt.grid(True, which="both")
+plt.ylabel(r"relative approximation error")
+plt.xlabel(r"inverse temperature $\beta^{-1}$")
[…]

### 2024-07-19 paper/plots/quantum_spin.py
fix wrong directory name in workflow

@@ -1,3 +1,9 @@
+"""
+quantum_spin.py
+
+TODO: Maybe try smaller range of betas, or only Hutchinsons? (find example of A ill behaved and f(A) low-rank)
+"""
+
 import __context__
 
 import scipy as sp

### 2024-07-19 paper/re-pro-badge.tex
automatically reproduce paper

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/bca2ff7}{bca2ff7} on 2024-07-18 at 14:29:53 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/3aeb08c}{3aeb08c} on 2024-07-19 at 15:04:36 UTC.

### 2024-08-13 algorithms/chebyshev_nystrom.py
push away matrix size into kernel definition

@@ -152,8 +152,8 @@ def chebyshev_nystrom(A, t=0, m=100, n_Psi=10, n_Omega=10, kernel=gaussian_kerne
             # Accumulation
             if l <= m:
                 K_1[i] += mu[i, l] * X
-                L_1[i] += mu[i, l] * Y
                 l_2[i] += mu[i, l] * z
+                L_1[i] += mu[i, l] * Y
             K_2[i] += nu[i, l] * X
 
         # Chebyshev recurrence
@@ -169,8 +169,8 @@ def chebyshev_nystrom(A, t=0, m=100, n_Psi=10, n_Omega=10, kernel=gaussian_kerne
         if n_Omega and np.trace(K_1[i]) <= kappa * n_Omega:
             continue
 
-        phi[i] = np.trace(np.linalg.lstsq(K_1[i], K_2[i], rcond=rcond)[0]) / n
+        phi[i] = np.trace(np.linalg.lstsq(K_1[i], K_2[i], rcond=rcond)[0])
         if n_Psi > 0:
-            phi[i] += (l_2[i] - np.trace(L_1[i].T @ np.linalg.lstsq(K_1[i], L_1[i], rcond=rcond)[0])) / (n * n_Psi)
+            phi[i] += (l_2[i] - np.trace(L_1[i].T @ np.linalg.lstsq(K_1[i], L_1[i], rcond=rcond)[0])) / n_Psi
 
     return phi

### 2024-08-13 algorithms/helpers.py

@@ -35,7 +35,7 @@ def spectral_transformation(A, min_ev=None, max_ev=None):
 
     return A_st
 
-def gaussian_kernel(t, x=None, sigma=0.1):
+def gaussian_kernel(t, x=None, sigma=0.1, n=1):
     """
     Gaussian kernel.
 
@@ -47,6 +47,9 @@ def gaussian_kernel(t, x=None, sigma=0.1):
         Value(s) outer-subtracted from paramter value(s) 
     sigma : int or float
         Standard deviation of the Gaussian.
+    n : int
+        Normalization "g(t, x) / n" for simpler spectral density computations.
+        Here, n is the size of the matrix.
 
     Returns
     -------
@@ -55,7 +58,7 @@ def gaussian_kernel(t, x=None, sigma=0.1):
     """
     if x is not None:
         t = np.subtract.outer(t, x)
-    return np.exp(- t**2 / (2 * sigma**2)) / (np.sqrt(2 * np.pi) * sigma)
+    return np.exp(- t**2 / (2 * sigma**2)) / (np.sqrt(2 * np.pi) * sigma * n)
 
 def form_spectral_density(eigvals, t, kernel=gaussian_kernel):
     """
@@ -76,4 +79,4 @@ def form_spectral_density(eigvals, t, kernel=gaussian_kernel):
     spectral_density : np.ndarray of shape (n_t,)
         The value of the spectral density evaluated at the grid points.
     """
-    return kernel(t, eigvals).sum(axis=1) / len(eigvals)
+    return kernel(t, eigvals).sum(axis=1)

### 2024-08-13 algorithms/krylov_aware.py

@@ -243,6 +243,6 @@ def krylov_aware(A, t=0, n_iter=10, n_reorth=5, n_Omega=10, n_Psi=10, kernel=gau
     if kernel is None:
         fun = lambda t, x: x
     else:
-        fun = lambda t, x: kernel(t, x) / A.shape[0]
+        fun = lambda t, x: kernel(t, x)
 
     return fun(t, nodes) @ weights

### 2024-08-13 paper/plots/convergence.py

@@ -29,7 +29,7 @@ markers = ["o", "s", "d"]
 labels = [r"$n_{\mathbf{\Omega}} = 0$", r"$n_{\mathbf{\Psi}} = n_{\mathbf{\Omega}}$", r"$n_{\mathbf{\Psi}} = 0$"]
 
 # Determine the baseline spectral density
-kernel = lambda t, x: gaussian_kernel(t, x, sigma=sigma)
+kernel = lambda t, x: gaussian_kernel(t, x, sigma=sigma, n=A.shape[0])
 baseline = form_spectral_density(eigvals_st, t, kernel)
 
 error = np.empty((3, len(n_Vec_list)))

### 2024-08-13 paper/plots/distribution.py

@@ -33,7 +33,7 @@ error = np.empty((len(n_Psi_list), len(sigma_list)))
 for j, sigma in enumerate(sigma_list):
 
     # Determine the baseline spectral density
-    kernel = lambda x: gaussian_kernel(x, sigma=sigma)
+    kernel = lambda x: gaussian_kernel(x, sigma=sigma, n=A.shape[0])
     baseline = form_spectral_density(eigvals_st, t, kernel)
     m = int(16 / sigma)  # Double-check this, i.e. use error estimate for this
 

### 2024-08-13 paper/plots/hessian_density.py

@@ -65,7 +65,7 @@ n_Psi = 10
 plt.style.use("paper/plots/stylesheet.mplstyle")
 
 # Approximate the Hessian's spectral density
-kernel = lambda x: gaussian_kernel(x, sigma=sigma)
+kernel = lambda x: gaussian_kernel(x, sigma=sigma, n=H.shape[0])
 phi = chebyshev_nystrom(H, t, m, n_Psi, n_Omega, kernel)
 
 plt.plot(t, phi, color="#648FFF")

### 2024-08-13 paper/plots/zerocheck.py

@@ -29,7 +29,7 @@ labels = ["baseline", "without zero-check", "with zero-check"]
 kappa_list = [-1, 1e-5]
 
 # Determine the baseline spectral density
-kernel = lambda t, x: gaussian_kernel(t, x, sigma=sigma)
+kernel = lambda t, x: gaussian_kernel(t, x, sigma=sigma, n=A.shape[0])
 baseline = form_spectral_density(eigvals_st, t, kernel)
 
 estimate = [baseline]

### 2024-08-13 paper/plots/interpolation.py
adjust forgotten file

@@ -29,7 +29,7 @@ markers = ["o", "s", "d"]
 labels = ["inconsistent", "consistent", "non-negative"]
 
 # Determine the baseline spectral density
-kernel = lambda t, x: gaussian_kernel(t, x, sigma=sigma)
+kernel = lambda t, x: gaussian_kernel(t, x, sigma=sigma, n=A.shape[0])
 baseline = form_spectral_density(eigvals_st, t, kernel)
 
 error = np.empty((3, len(m_list)))

### 2024-08-13 paper/re-pro-badge.tex
automatically reproduce paper

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/3aeb08c}{3aeb08c} on 2024-07-19 at 15:04:36 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/841c60a}{841c60a} on 2024-08-13 at 15:16:51 UTC.

### 2024-08-19 paper/analysis.tex
make chebyshev-nystrom compatible with general parametric trace estimation
add remark for stable-rank bound in Hutchinson

@@
     \begin{equation} \label{eq:ghbound}
-        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq \gamma c \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}}  \right) \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
+        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq \gamma c \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}}  \right) \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
     \end{equation}
-    holds with probability at least $1 - \gamma^{-p}$ and for $c = 8 e^{\sfrac{1}{e}} (1 + \sqrt{\pi}/2)$.
+    holds with probability at least $1 - \gamma^{-p}$ and for $c = 4 (1 + \sqrt{1 + \pi/4})$.
     In particular, given $\varepsilon \in (0, 1)$ and $\delta \in (0, e^{-1})$, the bound $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$ holds
@@
 \refthm{thm:hutchinson} coincides with the analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, except for a slightly more restricted choice of $\delta$.
+
+\begin{remark}
+    If $\mtx{A}(t)$ is nonzero and symmetric positive semi-definite for all $t \in [a, b]$, then it follows that with probability at least $1 - \delta$
+    \begin{equation}
+        \int_{a}^{b} \frac{| \Trace(\mtx{A}(t)) - \Hutch{\mtx{\Psi}}(\mtx{A}(t)) |}{\Trace(\mtx{A}(t))} ~\mathrm{d}t < \varepsilon,
+    \end{equation}
+    if $n_{\mtx{\Psi}} \geq 2 c^2 e^2 \log(\delta^{-1}) \varepsilon^{-2} (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ where $\mu(t) = \lVert \mtx{A}(t) \rVert _2 / \Trace(\mtx{A}(t))$.
+Up to constant factors, this matches the equivalent result for constant matrices from \cite[Remark 2]{cortinovis-2022-randomized-trace}.
+\end{remark}
 

### 2024-08-19 paper/spectraldensity.tex
correct typo in bibliography key

@@
 The pseudoinverses on \reflin{lin:4-nystromchebyshev-nystrom-pp} need to be computed with care.
-It is possible to reformulate the problem as a generalized eigenvalue problem and apply thresholding to the smallest eigenvalues of a certain matrix to enforce stability \cite{lin-2017-randomized-estimation, epperly-2022-theory-quantuma}.
+It is possible to reformulate the problem as a generalized eigenvalue problem and apply thresholding to the smallest eigenvalues of a certain matrix to enforce stability \cite{lin-2017-randomized-estimation, epperly-2022-theory-quantum}.
 Since this renders the algorithm significantly more complicated but only improved the accuracy slightly during our experiments, we will instead evaluate these expressions in \reflin{lin:4-nystromchebyshev-nystrom-pp} using a least-squares solver which likewise uses a truncation of the smallest eigenvalues to enforce better conditioning.

### 2024-08-20 paper/re-pro-badge.tex
automatically reproduce paper

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/841c60a}{841c60a} on 2024-08-13 at 15:16:51 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/c1a5f4f}{c1a5f4f} on 2024-08-20 at 15:38:04 UTC.

### 2024-08-21 paper/tables/interpolation.py
fix wrong kernel definition in interpolation timing

@@ -133,13 +133,15 @@ labels = ["oversampled FFT", "DCT", "non-negative DCT"]
 n_t = 1000
 t = np.arange(-1, 1, n_t)
 sigma = 0.005
-g = lambda x: gaussian_kernel(x, sigma=sigma)
-parameters = [{"t": t, "m": 800, "kernel": g},
-              {"t": t, "m": 1600, "kernel": g},
-              {"t": t, "m": 2400, "kernel": g},
-              {"t": t, "m": 3200, "kernel": g}]
+kernel = lambda t, x: gaussian_kernel(t, x, sigma=sigma)
+    
+parameters = [{"t": t, "m": 800, "kernel": kernel},
+              {"t": t, "m": 1600, "kernel": kernel},
+              {"t": t, "m": 2400, "kernel": kernel},
+              {"t": t, "m": 3200, "kernel": kernel}]
 
-means = np.empty((len(methods), len(parameters)))
+means = np.em
+pty((len(methods), len(parameters)))
 errors = np.empty((len(methods), len(parameters)))
 
 for i in range(len(methods)):

### 2024-08-21 paper/re-pro-badge.tex
automatically reproduce paper

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/c1a5f4f}{c1a5f4f} on 2024-08-20 at 15:38:04 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/c2d5cd1}{c2d5cd1} on 2024-08-21 at 09:23:43 UTC.

### 2024-08-21 paper/tables/interpolation.py
fix small typo

@@ -140,8 +140,7 @@ parameters = [{"t": t, "m": 800, "kernel": kernel},
               {"t": t, "m": 2400, "kernel": kernel},
               {"t": t, "m": 3200, "kernel": kernel}]
 
-means = np.em
-pty((len(methods), len(parameters)))
+means = np.empty((len(methods), len(parameters)))
 errors = np.empty((len(methods), len(parameters)))
 
 for i in range(len(methods)):

### 2024-08-21 paper/re-pro-badge.tex
automatically reproduce paper

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/c2d5cd1}{c2d5cd1} on 2024-08-21 at 09:23:43 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/d86c14f}{d86c14f} on 2024-08-21 at 12:47:34 UTC.

### 2024-09-07 paper/tables/interpolation.py
fix kernel usage in timing table

@@ -118,9 +118,8 @@ def chebyshev_coefficients_quadrature(t, m, kernel, n_theta=None):
 
     theta = np.arange(2 * n_theta) * np.pi / n_theta
 
-    # Can be computed via Fourier transform:
-    t_minus_theta = np.subtract.outer(t, np.cos(theta))
-    mu = np.real(np.fft.fft(kernel(t_minus_theta), axis=1)[:, :m+1])
+    # Can be computed via Fourier transform
+    mu = np.real(np.fft.fft(kernel(t, np.cos(theta)), axis=1)[:, :m+1])
 
     # Rescale the coefficients (as required by the definition)
     mu[:, 0] /= 2 * n_theta

### 2024-09-07 paper/re-pro-badge.tex
automatically reproduce paper

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/d86c14f}{d86c14f} on 2024-08-21 at 12:47:34 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/9c36faf}{9c36faf} on 2024-09-07 at 07:01:11 UTC.

### 2024-09-10 paper/plots/convergence.py
change range of mat-vec sweep

@@ -21,7 +21,7 @@ eigvals_st = spectral_transformation(eigvals, min_ev, max_ev)
 t = np.linspace(-1, 1, 100)
 sigma = 0.005
 m = 2000
-n_Vec_list = np.logspace(1.3, 2.8, 7).astype(int) 
+n_Vec_list = np.logspace(0.8, 2.8, 7).astype(int) 
 
 plt.style.use("paper/plots/stylesheet.mplstyle")
 colors = ["#648FFF", "#DC267F", "#FFB000"]

### 2024-09-18 paper/analysis.tex
rework and correct stable rank remark

@@
     \end{equation}
-    if $n_{\mtx{\Psi}} \geq 2 c^2 e^2 \log(\delta^{-1}) \varepsilon^{-2} (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ where $\mu(t) = \lVert \mtx{A}(t) \rVert _2 / \Trace(\mtx{A}(t))$.
-Up to constant factors, this matches the equivalent result for constant matrices from \cite[Remark 2]{cortinovis-2022-randomized-trace}.
+    if $n_{\mtx{\Psi}} \geq 2 c^2 e^2 \log(\delta^{-1}) \varepsilon^{-2} (\int_{a}^{b} \eta(t)~\mathrm{d}t)^2$ where $\eta(t) = \lVert \mtx{A}(t) \rVert _F / \Trace(\mtx{A}(t))$.
+The quantity $\eta(t) \in [1/\sqrt{n}, 1]$ is small when the spectrum of $\mtx{A}(t)$ decays slowly and large when $\mtx{A}(t)$ when it decays quickly, i.e.
+$\mtx{A}(t)$ has low numerical rank.
+This parallels a similar result for constant matrices in~\cite[Remark 2]{cortinovis-2022-randomized-trace}.
 \end{remark}

### 2024-09-19 paper/re-pro-badge.tex

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/9c36faf}{9c36faf} on 2024-09-07 at 07:01:11 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/9a5f9ad}{9a5f9ad} on 2024-09-10 at 13:14:05 UTC.

### 2024-09-19 paper/re-pro-badge.tex
automatically reproduce paper

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/9a5f9ad}{9a5f9ad} on 2024-09-10 at 13:14:05 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/ab895e9}{ab895e9} on 2024-09-19 at 07:36:41 UTC.

### 2024-09-23 paper/analysis.tex
adjust paper according to feedback

--- 2024-07-18 paper/analysis.tex
@@
     \end{align}
-    Consequently, by Markov's inequality, the inequality 
+    Consequently, by Markov's inequality, the bound 
     \begin{equation} 
@@
 \refthm{thm:hutchinson} coincides with the analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, except for a slightly more restricted choice of $\delta$.
+
+\begin{remark}
+    A slightly different version of \refthm{thm:hutchinson} can be derived when $\mtx{B}(t)$ is nonzero and symmetric positive semi-definite for all $t \in [a, b]$: after dividing both sides of \refequ{equ:moment-bound-hutchinson} by $\Trace(\mtx{B}(t))$, one can apply the remaining steps in the proof to deduce with probability at least $1 - \delta$
+    \begin{equation}
+        \int_{a}^{b} \frac{| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) |}{\Trace(\mtx{B}(t))} ~\mathrm{d}t < \varepsilon,
+    \end{equation}
+    if $n_{\mtx{\Psi}} \geq 2 c^2 e^2 \log(\delta^{-1}) \varepsilon^{-2} \rho^2$, where we identify
+    \begin{equation}
+        \rho = \int_{a}^{b} \frac{\lVert \mtx{B}(t) \rVert _F}{\Trace(\mtx{B}(t))} ~\mathrm{d}t
+    \end{equation}
+    The quantity $\rho$ is small when on the interval $[a, b]$ the singular values of $\mtx{B}(t)$ decay slowly and large when they decay quickly.
+This parallels a similar result for constant matrices in~\cite[Remark 2]{cortinovis-2022-randomized-trace}.
+\end{remark}
 

### 2024-09-23 paper/spectraldensity.tex
adjust paper according to feedback

@@
 \begin{table}[ht]
-    \caption{We compare the runtime in milliseconds for computing the coefficients of the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ \refequ{equ:smoothing-kernel} in three different ways: as proposed in \cite[Algorithm 1]{lin-2017-randomized-estimation} (\emph{oversampled FFT}), using the discrete cosine transform (\emph{DCT}) \refequ{equ:discrete-cosine-transform}, and computing a non-negative Chebyshev expansion with \refalg{alg:chebyshev-squaring} (\emph{non-negative DCT}).
+    \caption{We compare the runtime in milliseconds for computing the coefficients of the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ \refequ{equ:smoothing-kernel} in three different ways: as proposed in \cite[Algorithm 1]{lin-2017-randomized-estimation} (\emph{quadrature FFT}), using the discrete cosine transform (\emph{DCT}) \refequ{equ:discrete-cosine-transform}, and computing a non-negative Chebyshev expansion with \refalg{alg:chebyshev-squaring} (\emph{non-negative DCT}).
 We use $\sigma=0.005$, $n_t=1000$ parameter values, and various degrees $m$.
@@
 An efficient implementation of the approximation \refequ{equ:chebyshev-nystrom-formula} can be achieved thanks to the invariance of the trace under cyclic permutation of its arguments and the recurrence relation \refequ{equ:chebyshev-recurrence} which the Chebyshev polynomials satisfy.
-To illustrate this, consider the approximation \refequ{equ:chebyshev-nystrom-formula} when $\mtx{\Psi}$ is zero.
+To illustrate this, consider the approximation \refequ{equ:chebyshev-nystrom-formula} when $n_{\mtx{\Psi}}$ is zero, i.e.
+only a low-rank approximation is computed.
 Using the definition of the Nyström approximation \refequ{equ:nystrom-approximation}, the cyclic invariance of the trace, and the symmetry of $\mtx{A}$, we can rewrite it as
@@
 $\geq 1 - \delta$)} \notag \\
-        &\leq (1 + \widetilde{\varepsilon}) \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \varepsilon \int_{-1}^{1} \left| \phi_{\sigma}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
+        &\leq (1 + \widetilde{\varepsilon}) \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \widetilde{\varepsilon} \int_{-1}^{1} \left| \phi_{\sigma}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
         &\leq (1 + \widetilde{\varepsilon}) 2 \underline{E}_{\sigma, m} + \widetilde{\varepsilon}. && \text{(\refthm{thm:chebyshev-error}, normalization)} \notag \\
@@
 In practice, the approximation error is often significantly better than guaranteed in \refthm{thm:chebyshev-nystrom}, particularly when $n_{\mtx{\Omega}}$ is chosen rather large.
-This is because the singular values of the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ decay so quickly that already the Nyström approximation alone is highly accurate (cf.
+The reason is that for small $\sigma$ the singular values of the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ decay so quickly that already the Nyström approximation alone is highly accurate (cf.
 \refthm{thm:nystrom}).
-To establish a relation between the singular value decay of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ and what choice of $n_{\mtx{\Omega}}$ is needed to achieve a better approximation than what we show in \refthm{thm:chebyshev-nystrom}, we study the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
+To establish a relation between the singular value decay of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ and what choice of $n_{\mtx{\Omega}}$ is needed for a significantly better approximation than ensured by \refthm{thm:chebyshev-nystrom}, we study the numerical rank of $n^{-1} g_{\sigma}(t\mtx{I}_n - \mtx{A})$\footnote{The factor $n^{-1}$ comes from the definition of the spectral density \refequ{equ:smooth-spectral-density} when it is written as the trace of a matrix function.}.
 
-We can express the numerical rank of a matrix with respect to the nuclear norm in terms of its singular values $\sigma_1 \geq \dots \geq \sigma_n \geq 0$
+We can express the numerical rank of a matrix with respect to the nuclear norm in terms of its singular values $\sigma_1 \geq \dots \geq \sigma_n \geq 0$ as
 \begin{equation}
@@
 \end{equation}
-$\varepsilon$ is usually taken to be the double machine precision, i.e. $10^{-16}$.
+where $\varepsilon$ is usually taken to be the double machine precision, i.e. $10^{-16}$.
 
@@
 \begin{equation}
-    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) = \frac{1}{\sqrt{2 \pi \sigma^2}} e^{-\sfrac{(t - \lambda_{(i)})^2}{2 \sigma^2}},
+    \sigma_i(t) = n^{-1} g_{\sigma}(t - \lambda_{(i)}) = \frac{1}{n \sqrt{2 \pi \sigma^2}} e^{-\sfrac{(t - \lambda_{(i)})^2}{2 \sigma^2}},
     \label{equ:gaussian-kernel-eigenvalues}
@@
 where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from spectral-parameter $t$, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t)$.
-Consequently, by using the closed-form expression of the eigenvalues \refequ{equ:gaussian-kernel-eigenvalues}, we may upper bound the numerical rank of a Gaussian smoothing kernel as
+Consequently, by using the closed-form expression of the eigenvalues \refequ{equ:gaussian-kernel-eigenvalues}, we may generously upper bound the numerical rank by
 \begin{equation}
-    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A})) \leq \#\left\{1\leq i\leq n: |t - \lambda_i| < d_{\varepsilon, \sigma} \right\},
+    r_{\varepsilon}(n^{-1} g_{\sigma}(t\mtx{I}_n - \mtx{A})) \leq \#\left\{1\leq i\leq n: |t - \lambda_i| < d_{\varepsilon, \sigma} \right\},
     \label{equ:gaussian-kernel-numerical-rank}
@@
     \input{figures/numerical-rank.tex}
-    \caption{The numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ can be
-        approximately computed by counting the number of eigenvalues
-        $\lambda_{(1)}, \dots, \lambda_{(n)}$ of the matrix $\mtx{A}$ which lie less than
-        a constant $d_{\varepsilon, \sigma}$ away from $t$.}
+    \caption{The numerical rank of $n^{-1} g_{\sigma}(t\mtx{I}_n - \mtx{A})$ can be approximately computed by counting the number of eigenvalues $\lambda_{(1)}, \dots, \lambda_{(n)}$ of the matrix $\mtx{A}$ which lie less than a constant $d_{\varepsilon, \sigma}$ away from $t$.}
     \label{fig:numerical-rank}
@@
 
-If we additionally assume the eigenvalues of the matrix $\mtx{A}$ to be evenly distributed within $[-1, 1]$, that is, in any subinterval of fixed length in $[a, b]$, $-1 \leq a < b \leq 1$, we can expect to find roughly the same number of eigenvalues, then we can estimate the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ to be
+If we additionally assume the eigenvalues of the matrix $\mtx{A}$ to be evenly distributed within $[-1, 1]$, that is, in any subinterval of fixed length in $[a, b]$, $-1 \leq a < b \leq 1$, we can expect to find roughly the same number of eigenvalues, then we can estimate the numerical rank of $n^{-1} g_{\sigma}(t\mtx{I}_n - \mtx{A})$ to be
 \begin{equation}
-    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A})) \lessapprox n d_{\varepsilon, \sigma} \approx 6 n \sigma,
+    r_{\varepsilon}(n^{-1} g_{\sigma}(t\mtx{I}_n - \mtx{A})) \lessapprox n d_{\varepsilon, \sigma} \approx 6 n \sigma,
     \label{equ:gaussian-kernel-numerical-rank-uniform}
@@
 provided we set $\varepsilon = 10^{-16}$ and choose $\sigma$ roughly between $10^{-8}$ and $1$.
-Hence, choosing $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma)$ will approximations which are significantly better than \refthm{thm:chebyshev-nystrom}.
+Hence, choosing $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma)$ will approximations which can be significantly better than \refthm{thm:chebyshev-nystrom}.

### 2024-09-23 paper/tables/interpolation.py

@@ -127,7 +127,7 @@ def chebyshev_coefficients_quadrature(t, m, kernel, n_theta=None):
     return mu
 
 methods = [chebyshev_coefficients_quadrature, chebyshev_expansion, chebyshev_expansion]
-labels = ["oversampled FFT", "DCT", "non-negative DCT"]
+labels = ["quadrature FFT", "DCT", "non-negative DCT"]
 
 n_t = 1000
 t = np.arange(-1, 1, n_t)

### 2024-09-23 paper/re-pro-badge.tex
automatically reproduce paper

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/ab895e9}{ab895e9} on 2024-09-19 at 07:36:41 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/7c053ed}{7c053ed} on 2024-09-23 at 17:06:23 UTC.

### 2024-09-24 paper/numresults.tex
add krylov-aware numerical results

@@
 
+\subsection{Comparison to Krylov-aware stochastic trace estimation}
+\label{subsec:krylov-aware}
+
+The Krylov-aware stochastic trace estimation \cite{chen-2023-krylovaware-stochastic} also works well for the approximation of the trace of a matrix function $f(\mtx{A}, t)$ which separately depends on a parameter $t$.
+It also samples two standard Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$, of which $\mtx{\Omega}$ is used to extract a low-rank approximation to $f(\mtx{A}, t)$ from a Krylov-subspace of $\mtx{A}$ and $\mtx{\Psi}$ to estimate the trace of the approximation residual.
+
+We run an optimized implementation of the Krylov-aware stochastic trace estimator \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} in multiple configurations on the example from \refsec{subsec:hamiltonian} and plot their approximation errors for various values of the smoothing parameter $\sigma$ in \reffig{fig:krylov-aware-density}.
+For reference, we add the error of the Chebyshev-Nyström++ method on parameters which lead to a comparable run-time.
+
+\begin{figure}[ht]
+    \begin{minipage}[c]{.45\linewidth}
+        \centering
+        \input{plots/krylov_aware_density.pgf}
+    \end{minipage}\hfill%
+    \begin{minipage}[c]{.45\linewidth}
+        %\label{tab:krylov-aware-density}
+        \renewcommand{\arraystretch}{1.2}
+        \begin{tabular}{@{}lccccc@{}}
+        \toprule
+        & $b$ & $q$ & $m$ & $n$ & $t$ (s)\\
+        \midrule
+        KA I & 10 & 30 & 60 & 20 & 9.13 \\
+        KA II & 20 & 30 & 30 & 20 & 9.13 \\
+        KA III & 10 & 60 & 30 & 40 & 9.13 \\
+        KA IV & 10 & 30 & 30 & 20 & 9.13 \\
+        CN++ & 10 & - & - & 20 & 9.13 \\
+        \bottomrule
+        \end{tabular}
+        %\caption{Comparison with Krylov-Aware stochastic trace estimation.}
+    \end{minipage}
+    \caption{Comparison with Krylov-Aware stochastic trace estimation.}
+    \label{fig:krylov-aware-density}
+\end{figure}
+
+To demonstrate that the Chebyshev-Nyström++ estimator is also effective on problems unrelated to spectral density estimation, we consider the problem of approximating $\Trace(\exp(-\beta \mtx{A}))$ for the Hamiltonian $\mtx{A}$ of a planar Heisenberg spin chain \cite{chen-2023-krylovaware-stochastic}.
+We compute the pointwise error from the analytic solution as a function of the temperature $\beta^{-1}$ and compare it to the error achieved by the Krylov-aware trace estimator in \reffig{fig:krylov-aware-spin}.
+
+\begin{figure}[ht]
+    \centering
+    \input{plots/krylov_aware_spin.pgf}
+    \caption{Comparison with Krylov-Aware stochastic trace estimation.}
+    \label{fig:krylov-aware-spin}
+\end{figure}
+
+Unlike in the spectral density example from \refsec{subsec:hamiltonian}, the numerical rank of the matrix we approximate wildly changes with the parameter $\beta$.
+Since the Chebyshev-Nyström++ method requires a fixed allocation of random vectors to the low-rank approximation for all values of the parameter $\beta$.
+\todo{As a consequence, ...}
+
 \subsection{Spectral density of Hessian matrix of neural network}
@@
 \todo{Hessian spectrum analysis?}
-
-\subsection{Trace of parameter-dependent matrix functions}
-\label{subsec:function-trace}
-
-\todo{Comparison plots with other estimators (Krylov-Aware)?}

### 2024-09-24 paper/plots/krylov_aware_density.py

@@ -1,0 +1,65 @@
+import __context__
+
+import time
+import numpy as np
+import matplotlib.pyplot as plt
+from algorithms.chebyshev_nystrom import chebyshev_nystrom
+from algorithms.krylov_aware import krylov_aware
+from algorithms.helpers import spectral_transformation, form_spectral_density, gaussian_kernel
+from matrices.electronic_structure import hamiltonian
+
+np.random.seed(0)
+
+# Load matrix
+A = hamiltonian()
+
+# Perform spectral transform with A and its eigenvalues
+eigvals = np.linalg.eigvalsh(A.toarray())
+min_ev, max_ev = eigvals[0], eigvals[-1]
+A_st = spectral_transformation(A, min_ev, max_ev)
+eigvals_st = spectral_transformation(eigvals, min_ev, max_ev)
+
+# Set parameter
+t = np.linspace(-1, 1, 100)
+sigma_list = np.logspace(-3.5, -1.0, 7)
+n_Psi_list = np.array([ 60, 30, 30, 30, 40])    # m
+n_Omega_list = np.array([ 10, 20, 10, 10, 40])    # b
+n_iter_list = np.array([ 30, 30, 60, 30, None])  # q
+n_reorth_list = np.array([20, 20, 20, 40, None])  # n
+m = 2000
+
+plt.style.use("paper/plots/stylesheet.mplstyle")
+plt.figure(figsize=(3, 3))
+colors = ["#FFB000", "#FE6100", "#785EF0", "#648FFF", "#DC267F"]
+markers = ["d", "p", "^", "o", "s"]
+labels = ["KA I", "KA II", "KA III", "KA IV", "CN++"]
+
+error = np.empty((len(n_Psi_list), len(sigma_list)))
+times = np.zeros(len(n_Psi_list))
+for j, sigma in enumerate(sigma_list):
+
+    # Determine the baseline spectral density
+    kernel = lambda t, x: gaussian_kernel(t, x, sigma=sigma, n=A.shape[0])
+    baseline = form_spectral_density(eigvals_st, t, kernel)
+
+    for i, (n_Psi, n_Omega, n_iter, n_reorth) in enumerate(zip(n_Psi_list, n_Omega_list, n_iter_list, n_reorth_list)):
+        t0 = time.time()
+        if i < len(n_Psi_list) - 1:
+            estimate = krylov_aware(A_st, t, n_iter, n_reorth, n_Psi, n_Omega, kernel)
[…]

### 2024-09-24 paper/plots/krylov_aware_spin.py

@@ -1,0 +1,60 @@
+import __context__
+from matrices.quantum_spin import hamiltonian, partition_function
+from algorithms.chebyshev_nystrom import chebyshev_nystrom
+from algorithms.krylov_aware import krylov_aware
+from algorithms.helpers import spectral_transformation
+import matplotlib.pyplot as plt
+import scipy as sp
+import numpy as np
+import time
+
+np.random.seed(0)
+
+N = 20
+s = 0.5
+h = 0.3
+J = 1
+A = hamiltonian(N, s, h, J)
+
+E_min, = sp.sparse.linalg.eigsh(A, k=1, which="SA", return_eigenvectors=False, tol=1e-5)
+E_max, = sp.sparse.linalg.eigsh(A, k=1, which="LA", return_eigenvectors=False, tol=1e-5)
+
+betas = 1 / np.logspace(-2.5, 3, 20)
+Z_true = partition_function(betas, N, h, J, E_min)
+
+A_transformed = spectral_transformation(A, E_min, E_max)
+m = 100
+n_Omega = 40
+n_Psi = 40
+
+function = lambda beta, x: np.exp(-np.multiply.outer(beta, x + 1))
+t0 = time.time()
+Z_nycheb = chebyshev_nystrom(A_transformed, betas * (E_max - E_min) / 2, m, n_Psi, n_Omega, function, kappa=-1, rcond=1e-10)
+print(time.time() - t0)
+Z_nycheb *= np.exp(- betas * (E_min + E_max) / 2)
+
+n_iter = 30 + 50
+n_reorth = 50
+n_Omega = 8
+n_Psi = 13
+
+function = lambda beta, x: np.exp(-np.multiply.outer(beta, x - E_min))
+t0 = time.time()
+Z_krylov = krylov_aware(A, betas, n_iter, n_reorth, n_Omega, n_Psi, function)
+print(time.time() - t0)
+
+plt.style.use("paper/plots/stylesheet.mplstyle")
+colors = ["#FFB000", "#FE6100", "#785EF0", "#648FFF", "#DC267F"]
+markers = ["d", "p", "^", "o", "s"]
+labels = ["KA I", "KA II", "KA III", "KA IV", "CN++"]
+
+plt.plot(1 / betas, np.abs(1 - Z_krylov / Z_true), label="KA", color="#648FFF", marker="^")
[…]

### 2024-09-25 paper/numresults.tex
finish plots for Krylov-aware comparison

@@
 \begin{figure}[ht]
-    \begin{minipage}[c]{.45\linewidth}
+    \begin{minipage}[c]{.475\linewidth}
         \centering
@@
     \end{minipage}\hfill%
-    \begin{minipage}[c]{.45\linewidth}
+    \begin{minipage}[c]{.475\linewidth}
         %\label{tab:krylov-aware-density}
-        \renewcommand{\arraystretch}{1.2}
-        \begin{tabular}{@{}lccccc@{}}
-        \toprule
-        & $b$ & $q$ & $m$ & $n$ & $t$ (s)\\
-        \midrule
-        KA I & 10 & 30 & 60 & 20 & 9.13 \\
-        KA II & 20 & 30 & 30 & 20 & 9.13 \\
-        KA III & 10 & 60 & 30 & 40 & 9.13 \\
-        KA IV & 10 & 30 & 30 & 20 & 9.13 \\
-        CN++ & 10 & - & - & 20 & 9.13 \\
-        \bottomrule
-        \end{tabular}
+        \vspace{-35pt}
+        \input{tables/krylov_aware_density_KA.tex}
+        \newline
+        \newline
+        \input{tables/krylov_aware_density_CN.tex}
+        %\renewcommand{\arraystretch}{1.2}
+        %\begin{tabular}{@{}lccccc@{}}
+        %\toprule
+        %& $n_{\mtx{\Omega}}$ & $n_{\mtx{\Psi}}$ & $q$ & $n$ & runtime (s)\\
+        %\midrule
+        %KA I & 10 & 60 & 30 & 20 & 9.13 \\
+        %KA II & 20 & 30 & 30 & 20 & 9.13 \\
+        %KA III & 10 & 30 & 60 & 40 & 9.13 \\
+        %KA IV & 10 & 30 & 30 & 20 & 9.13 \\
+        %\bottomrule
+        %\end{tabular}
+        %\begin{tabular}{@{}lcccc@{}}
+        %\\
+        %\toprule
+        %& $n_{\mtx{\Omega}}$ & $n_{\mtx{\Psi}}$ & $m$ & runtime (s)\\
+        %\midrule
+        %CN++ & 40  & 40 & 2000 & 9.13 \\
+        %\bottomrule
+        %\end{tabular}
         %\caption{Comparison with Krylov-Aware stochastic trace estimation.}
@@
 \begin{figure}[ht]
+    \begin{minipage}[c]{.475\linewidth}
         \centering
         \input{plots/krylov_aware_spin.pgf}
+    \end{minipage}\hfill%
+    \begin{minipage}[c]{.475\linewidth}
+        \vspace{-35pt}
+        \input{tables/krylov_aware_spin_KA.tex}
+        \newline
+        \vspace{13pt}
+        \newline
+        \input{tables/krylov_aware_spin_CN.tex}
+    \end{minipage}
     \caption{Comparison with Krylov-Aware stochastic trace estimation.}

### 2024-09-25 algorithms/helpers.py

@@ -80,3 +80,54 @@ def form_spectral_density(eigvals, t, kernel=gaussian_kernel):
         The value of the spectral density evaluated at the grid points.
     """
     return kernel(t, eigvals).sum(axis=1)
+
+def generate_tex_tabular(values, filepath, headline=None, row_labels=None, errors=None, fmt=r"${:.3f}$"):
+    """
+    Generate a LaTeX tabular from a numpy array.
+
+    Parameters
+    ----------
+    values : np.ndarray (n, m)
+        The values to be displayed in the tabular.
+    filepath : str 
+        The path to the file in which the table should be stored.
+    headline : list of str, int, or float (m,)
+        The values in the headline (column labels).
+    row_labels : list of str (n,)
+        The labels of the rows.
+    errors : np.ndarray (n, m)
+        The errors associated to the values.
+    fmt : str 
+        The format in which the values are displayed in the table.
+    """
+    f = open(filepath, "w")
+
+    num_rows = values.shape[0]
+    num_cols = values.shape[1]
+
+    if isinstance(fmt, str):
+        fmt = [fmt] * num_cols
+
+    f.write(r"\centering" + "\n")
+    f.write(r"\renewcommand{\arraystretch}{1.2}" + "\n")
+    f.write(r"\begin{tabular}{@{}" + ("l" if row_labels else "") + num_cols*"c" + r"@{}}" + "\n")
+    f.write(r"\toprule" + "\n")
+    if headline:
+        f.write(r" & ".join(headline) + r"\\" + "\n")
+        f.write(r"\midrule" + "\n")
+
+    for i in range(num_rows):
+        if row_labels:
+            f.write(row_labels[i] + r" & ")
+        for j in range(num_cols):
+            f.write(fmt[j].format(values[i, j]))
+            if errors is not None:
+                f.write(r" $\pm$ " + fmt[j].format(errors[i, j]))
+            if j < num_cols - 1:
+                f.write(r" & ")
+        f.write(r" \\" + "\n")
+
+    f.write(r"\bottomrule" + "\n")
+    f.write(r"\end{tabular}" + "\n")
+
+    f.close()

### 2024-09-25 paper/plots/krylov_aware_density.py

@@ -5,7 +5,7 @@ import numpy as np
 import matplotlib.pyplot as plt
 from algorithms.chebyshev_nystrom import chebyshev_nystrom
 from algorithms.krylov_aware import krylov_aware
-from algorithms.helpers import spectral_transformation, form_spectral_density, gaussian_kernel
+from algorithms.helpers import spectral_transformation, form_spectral_density, gaussian_kernel, generate_tex_tabular
 from matrices.electronic_structure import hamiltonian
 
 np.random.seed(0)
@@ -22,8 +22,8 @@ eigvals_st = spectral_transformation(eigvals, min_ev, max_ev)
 # Set parameter
 t = np.linspace(-1, 1, 100)
 sigma_list = np.logspace(-3.5, -1.0, 7)
-n_Psi_list = np.array([ 60, 30, 30, 30, 40])    # m
 n_Omega_list = np.array([ 10, 20, 10, 10, 40])    # b
+n_Psi_list = np.array([ 60, 30, 30, 30, 40])    # m
 n_iter_list = np.array([ 30, 30, 60, 30, None])  # q
 n_reorth_list = np.array([20, 20, 20, 40, None])  # n
 m = 2000
@@ -32,7 +32,7 @@ plt.style.use("paper/plots/stylesheet.mplstyle")
 plt.figure(figsize=(3, 3))
 colors = ["#FFB000", "#FE6100", "#785EF0", "#648FFF", "#DC267F"]
 markers = ["d", "p", "^", "o", "s"]
-labels = ["KA I", "KA II", "KA III", "KA IV", "CN++"]
+labels = ["KA (I)", "KA (II)", "KA (III)", "KA (IV)", "CN++"]
 
 error = np.empty((len(n_Psi_list), len(sigma_list)))
 times = np.zeros(len(n_Psi_list))
@@ -62,4 +62,14 @@ plt.xscale("log")
 plt.yscale("log")
 plt.savefig("paper/plots/krylov_aware_density.pgf", bbox_inches="tight")
 
-print(times)
+headline = ["", r"$n_{\mtx{\Omega}}$", r"$n_{\mtx{\Psi}}$", r"$q$", r"$n$", r"time (s)"]
+fmt = [r"${:0.0f}$", r"${:0.0f}$", r"${:0.0f}$", r"${:0.0f}$", r"${:.2f}$"]
[…]

### 2024-09-25 paper/plots/krylov_aware_spin.py

@@ -1,4 +1,84 @@
 import __context__
+
+import time
+import numpy as np
+import scipy as sp
+import matplotlib.pyplot as plt
+from algorithms.chebyshev_nystrom import chebyshev_nystrom
+from algorithms.krylov_aware import krylov_aware
+from algorithms.helpers import spectral_transformation, generate_tex_tabular
+from matrices.quantum_spin import hamiltonian, partition_function
+
+np.random.seed(0)
+
+N = 20
+s = 0.5
+h = 0.3
+J = 1
+A = hamiltonian(N, s, h, J)
+
+E_min, = sp.sparse.linalg.eigsh(A, k=1, which="SA", return_eigenvectors=False, tol=1e-5)
+E_max, = sp.sparse.linalg.eigsh(A, k=1, which="LA", return_eigenvectors=False, tol=1e-5)
+
+A_st = spectral_transformation(A, E_min, E_max)
+
+# Set parameter
+n_Omega_list = np.array([8, 8, 4, 4, 40])    # b
+n_Psi_list = np.array([0, 13, 6, 6, 40])    # m
+n_iter_list = np.array([30 + 50, 30 + 50, 30 + 50, 10 + 50, None])  # q
+n_reorth_list = np.array([50, 50, 50, 50, None])  # n
+m = 50
+
+plt.style.use("paper/plots/stylesheet.mplstyle")
+plt.figure(figsize=(3, 3))
+colors = ["#FFB000", "#FE6100", "#785EF0", "#648FFF", "#DC267F"]
+markers = ["d", "p", "^", "o", "s"]
+labels = ["KA (i)", "KA (iii)", "KA (iv)", "KA (v)", "CN++"]
+
+# Determine the baseline spectral density
+betas = 1 / np.logspace(-2.5, 3, 16)
+baseline = partition_function(betas, N, h, J, E_min)
+
+error = np.empty((len(n_Psi_list), len(betas)))
+times = np.zeros(len(n_Psi_list))
+
+for i, (n_Psi, n_Omega, n_iter, n_reorth) in enumerate(zip(n_Psi_list, n_Omega_list, n_iter_list, n_reorth_list)):
+    t0 = time.time()
+    if i < len(n_Psi_list) - 1:
+        function = lambda beta, x: np.exp(-np.multiply.outer(beta, x - E_min))
[…]

### 2024-09-25 paper/tables/interpolation.py

@@ -4,8 +4,7 @@ import timeit
 import numpy as np
 
 from algorithms.chebyshev_nystrom import chebyshev_expansion
-from algorithms.helpers import gaussian_kernel
-
+from algorithms.helpers import gaussian_kernel, generate_tex_tabular
 
 def time_method(method, parameters, num_times=1000, num_repeats=10):
     """
@@ -31,55 +30,6 @@ def time_method(method, parameters, num_times=1000, num_repeats=10):
     mean = np.mean(times)
     error = np.std(times)
     return mean, error
-
-
-def generate_tex_tabular(values, filepath, headline=None, row_labels=None, errors=None, fmt=r"${:.3f}$"):
-    """
-    Generate a LaTeX tabular from a numpy array.
-
-    Parameters
-    ----------
-    values : np.ndarray (n, m)
-        The values to be displayed in the tabular.
-    filepath : str 
-        The path to the file in which the table should be stored.
-    headline : list of str, int, or float (m,)
-        The values in the headline (column labels).
-    row_labels : list of str (n,)
-        The labels of the rows.
-    errors : np.ndarray (n, m)
-        The errors associated to the values.
-    fmt : str 
-        The format in which the values are displayed in the table.
-    """
-    f = open(filepath, "w")
-
-    num_rows = values.shape[0]
-    num_cols = values.shape[1]
-
-    f.write(r"\centering" + "\n")
-    f.write(r"\renewcommand{\arraystretch}{1.2}" + "\n")
-    f.write(r"\begin{tabular}{@{}" + ("l" if row_labels else "") + num_cols*"c" + r"@{}}" + "\n")
-    f.write(r"\toprule" + "\n")
-    if headline:
-        f.write(r" & ".join(headline) + r"\\" + "\n")
-        f.write(r"\midrule" + "\n")
-
-    for i in range(num_rows):
-        if row_labels:
-            f.write(row_labels[i] + r" & ")
-        for j in range(num_cols):
-            f.write(fmt.format(values[i, j]))
-            if errors is not None:
-                f.write(r" $\pm$ " + fmt.format(errors[i, j]))
-            if j < num_cols - 1:
-                f.write(r" & ")
-        f.write(r" \\" + "\n")
-
-    f.write(r"\bottomrule" + "\n")
-    f.write(r"\end{tabular}" + "\n")
-
-    f.close()
 
 
 def chebyshev_coefficients_quadrature(t, m, kernel, n_theta=None):

### 2024-09-25 algorithms/krylov_aware.py
add Hessian example and change Krylov-aware parameters

@@ -225,13 +225,18 @@ def krylov_aware(A, t=0, n_iter=10, n_reorth=5, n_Omega=10, n_Psi=10, kernel=gau
         Estimation". SIAM Journal on Matrix Analysis and Applications.
         Vol. 44 (3). doi:10.1137/22M1494257.
     """
-    Omega = np.random.randn(A.shape[0], n_Omega)
-    Q, T = block_lanczos(A, Omega, n_iter, extend_matrix=False, reorth_steps=n_reorth)
-    nodes, T_evec = np.linalg.eigh(T.toarray())
-    weights = np.linalg.norm(T_evec[:(n_reorth + 1) * n_Omega], axis=0)**2
+    if n_Omega > 0:
+        Omega = np.random.randn(A.shape[0], n_Omega)
+        Q, T = block_lanczos(A, Omega, n_iter, extend_matrix=False, reorth_steps=n_reorth)
+        nodes, T_evec = np.linalg.eigh(T.toarray())
+        weights = np.linalg.norm(T_evec[:(n_reorth + 1) * n_Omega], axis=0)**2
+    else:
+        nodes = []
+        weights = []
 
     Psi = np.random.randn(A.shape[0], n_Psi)
-    Psi -= Q[:, :(n_reorth + 1) * n_Omega] @ (Q[:, :(n_reorth + 1) * n_Omega].T @ Psi)
+    if n_Omega > 0:
+        Psi -= Q[:, :(n_reorth + 1) * n_Omega] @ (Q[:, :(n_reorth + 1) * n_Omega].T @ Psi)
 
     _, a_rem, b_rem = lanczos(A, Psi, n_iter, extend_matrix=False, return_matrix=False, reorth_tol=0)
     for a, b in zip(a_rem, b_rem):

### 2024-09-25 matrices/neural_network.py

@@ -215,6 +215,9 @@ class hessian(object):
         Q, T = lanczos(self, x, k, dtype=np.float64)
         return np.max(np.linalg.eigvalsh(T[0, :-1, :])) + np.linalg.norm(T[0, -1, -1] * Q[:, :, -1])
 
+    def update(self, model, loss_function, data):
+        self.gradient = compute_gradient(model, loss_function, data)
+
     def __matmul__(self, x):
         # Convert numpy array to torch tensor
         is_numpy = isinstance(x, np.ndarray)

### 2024-09-25 paper/plots/hessian_density.py

@@ -26,8 +26,8 @@ model = torch.nn.Sequential(
 )
 
 # For now, don't execute this
-if __name__ == "__main__":
-    exit()
+#if __name__ == "__main__":
+#    exit()
 
 np.random.seed(0)
 
@@ -47,30 +47,43 @@ test_loader = torch.utils.data.DataLoader(torchvision.datasets.MNIST("matrices/m
                                           shuffle=False
 )
 
-# Train the model on the MNIST data set
-loss_function = lambda x, y: torch.nn.MSELoss()(x, torch.nn.functional.one_hot(y, num_classes=10).to(dtype=torch.float))
-validator = lambda model: accuracy_validator(model, test_loader)
-train(model, train_loader, loss_function, validator=validator, n_epochs=5)
-
-data = next(iter(train_loader))
-H = hessian(model, loss_function, data, spectral_transform=True)
 
 # Set parameter
-t = np.linspace(-0.02, 0.02, 30)
-sigma = 0.003
-m = 1500
+t = np.linspace(0.2, 1.0, 100)
+sigma = 0.005
+m = 500
 n_Omega = 30
 n_Psi = 10
+data = next(iter(train_loader))
 
 plt.style.use("paper/plots/stylesheet.mplstyle")
 
+# Train the model on the MNIST data set
+loss_function = lambda x, y: torch.nn.MSELoss()(x, torch.nn.functional.one_hot(y, num_classes=10).to(dtype=torch.float))
+validator = lambda model: accuracy_validator(model, test_loader)
+
 # Approximate the Hessian's spectral density
-kernel = lambda x: gaussian_kernel(x, sigma=sigma, n=H.shape[0])
+kernel = lambda t, x: gaussian_kernel(t, x, sigma=sigma, n=H.shape[0])
+H = hessian(model, loss_function, data, spectral_transform=True)
 phi = chebyshev_nystrom(H, t, m, n_Psi, n_Omega, kernel)
 
-plt.plot(t, phi, color="#648FFF")
+plt.plot(t, phi, color="#648FFF", label=r"untrained")
 
+train(model, train_loader, loss_function, validator=validator, n_epochs=5)
+
+H.update(model, loss_function, data)
+phi = chebyshev_nystrom(H, t, m, n_Psi, n_Omega, kernel)
+
+plt.plot(t, phi, color="#DC267F", label=r"epoch $5$")
+
[…]

### 2024-09-25 paper/plots/krylov_aware_spin.py

@@ -23,9 +23,9 @@ E_max, = sp.sparse.linalg.eigsh(A, k=1, which="LA", return_eigenvectors=False, t
 A_st = spectral_transformation(A, E_min, E_max)
 
 # Set parameter
-n_Omega_list = np.array([8, 8, 4, 4, 40])    # b
-n_Psi_list = np.array([0, 13, 6, 6, 40])    # m
-n_iter_list = np.array([30 + 50, 30 + 50, 30 + 50, 10 + 50, None])  # q
+n_Omega_list = np.array([8, 0, 8, 4, 40])    # b
+n_Psi_list = np.array([0, 13, 13, 6, 40])    # m
+n_iter_list = np.array([30 + 50, 50, 30 + 50, 30 + 50, None])  # q
 n_reorth_list = np.array([50, 50, 50, 50, None])  # n
 m = 50
 
@@ -33,7 +33,7 @@ plt.style.use("paper/plots/stylesheet.mplstyle")
 plt.figure(figsize=(3, 3))
 colors = ["#FFB000", "#FE6100", "#785EF0", "#648FFF", "#DC267F"]
 markers = ["d", "p", "^", "o", "s"]
-labels = ["KA (i)", "KA (iii)", "KA (iv)", "KA (v)", "CN++"]
+labels = ["KA (i)", "KA (ii)", "KA (iii)", "KA (iv)", "CN++"]
 
 # Determine the baseline spectral density
 betas = 1 / np.logspace(-2.5, 3, 16)
@@ -76,66 +76,3 @@ fmt = [r"${:0.0f}$", r"${:0.0f}$", r"${:0.0f}$", r"${:.2f}$"]
 values = np.vstack((n_Omega_list[-1], n_Psi_list[-1], m, times[-1])).T
 
 generate_tex_tabular(values, "paper/tables/krylov_aware_spin_CN.tex", headline, labels[-1:], fmt=fmt)
-
-"""
-import __context__
-from matrices.quantum_spin import hamiltonian, partition_function
-from algorithms.chebyshev_nystrom import chebyshev_nystrom
-from algorithms.krylov_aware import krylov_aware
-from algorithms.helpers import spectral_transformation
-import matplotlib.pyplot as plt
-import scipy as sp
-import numpy as np
-import time
-
-np.random.seed(0)
-
-N = 20
-s = 0.5
-h = 0.3
-J = 1
-A = hamiltonian(N, s, h, J)
-
[…]

### 2024-09-26 paper/numresults.tex
add Hessian example and change Krylov-aware parameters
write down experimental setup

@@
     \begin{minipage}[c]{.475\linewidth}
-        %\label{tab:krylov-aware-density}
         \vspace{-35pt}
@@
         \newline
+        \vspace{15pt}
         \newline
         \input{tables/krylov_aware_density_CN.tex}
-        %\renewcommand{\arraystretch}{1.2}
-        %\begin{tabular}{@{}lccccc@{}}
-        %\toprule
-        %& $n_{\mtx{\Omega}}$ & $n_{\mtx{\Psi}}$ & $q$ & $n$ & runtime (s)\\
-        %\midrule
-        %KA I & 10 & 60 & 30 & 20 & 9.13 \\
-        %KA II & 20 & 30 & 30 & 20 & 9.13 \\
-        %KA III & 10 & 30 & 60 & 40 & 9.13 \\
-        %KA IV & 10 & 30 & 30 & 20 & 9.13 \\
-        %\bottomrule
-        %\end{tabular}
-        %\begin{tabular}{@{}lcccc@{}}
-        %\\
-        %\toprule
-        %& $n_{\mtx{\Omega}}$ & $n_{\mtx{\Psi}}$ & $m$ & runtime (s)\\
-        %\midrule
-        %CN++ & 40  & 40 & 2000 & 9.13 \\
-        %\bottomrule
-        %\end{tabular}
-        %\caption{Comparison with Krylov-Aware stochastic trace estimation.}
     \end{minipage}
-    \caption{Comparison with Krylov-Aware stochastic trace estimation.}
+    \caption{For the example from \refsec{subsec:hamiltonian} we compare the Chebyshev-Nyström++ method with the Krylov-Aware stochastic trace estimator in multiple configurations (right).
+To this extent, we compute the $L^1$-error of the methods for various levels of smoothing $\sigma$ (left).}
     \label{fig:krylov-aware-density}
@@
         \newline
-        \vspace{13pt}
+        \vspace{15pt}
         \newline
@@
     \end{minipage}
-    \caption{Comparison with Krylov-Aware stochastic trace estimation.}
+    \caption{For the Heisenberg spin chain example, we compare the Chebyshev-Nyström++ method with the Krylov-Aware stochastic trace estimator in the same configurations as in \cite[Table 5.1]{chen-2023-krylovaware-stochastic} (right).
+To this extent, we compute the $L^1$-error of the methods for various choices of the temperature parameter $\beta^{-1}$ (left).}
     \label{fig:krylov-aware-spin}
@@
 Since the Chebyshev-Nyström++ method requires a fixed allocation of random vectors to the low-rank approximation for all values of the parameter $\beta$.
-\todo{As a consequence, ...}
+Hence, both for large $\beta$, where the matrix $\exp(-\beta \mtx{A})$ is low-rank, and for small $\beta$, where the singular values only decay slowly, we cannot expect a good approximation.
 
@@
 
-\todo{Hessian spectrum analysis?}
+When optimizing a neural network over some data, it is of great interest to determine whether a minimum has been found and if this minimum is robust, i.e.
+a small perturbation of the parameters of the neural network will not lead to a significant decrease in the quality of the outputs.
+Both properties are reflected in the eigenvalues of the Hessian matrix $\mtx{A}$ with respect to an appropriately chosen loss function: if all eigenvalues are non-negative, then the loss attains an local minimum, and if additionally all of them are small, the minimum is robust.
+
+Since neural networks are usually parametrized by thousands if not millions of parameters, assembling the Hessian matrix and computing its eigenvalues is out of question.
+Luckily, there exists a method which can exactly compute matrix-vector products $\mtx{A} \vct{x}$ for any vector $\vct{x}$ which scales proportionally with the number of parameters \cite{pearlmutter-1994-fast-exact}.
+
+To demonstrate that our algorithm can effectively be applied in the setting of neural network optimization, we approximate the spectral density of a fully connected convolutional neural network which we train on the classification task given by the MNIST dataset\footnote{Handwritten digit classification; taken from \url{http://yann.lecun.com/exdb/mnist}.}.
+We plot the approximation of the spectral density of the Hessian matrix of the untrained neural network, and at different stages of training in \reffig{fig:hessian-density}.
+It can well be observed that the eigenvalues tend towards zero as the training proceeds.
+Furthermore, by the absence of large eigenvalues in the spectrum, we can tell that in the final epoch, the neural network attained a robust state.
+
+\begin{figure}
+    \input{plots/hessian_density.pgf}
+    \caption{Approximate spectral density of the Hessian matrix of a fully connected convolutional neural network in different epochs of training on the MNIST dataset.
+Computed with the Chebyshev-Nyström++ method (\refalg{alg:nystrom-chebyshev-pp}) on $n_{\mtx{\Omega}} = 30$, $n_{\mtx{\Psi}} = 10$, $m = 1500$, and $\sigma = 0.005$.}
+    \label{fig:hessian-density}
+\end{figure}

### 2024-09-26 paper/spectraldensity.tex
add Hessian example and change Krylov-aware parameters
write down experimental setup

@@
 
-Similar observations for the case when $\mtx{\Psi}$ is nonzero lead to the following algorithm.
+Similar observations for the case when $n_{\mtx{\Psi}}$ is nonzero lead to the following algorithm.
 
@@
 
-When using a Gaussian smoothing-kernel, we can explicitly write the singular values of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ as
+When using a Gaussian smoothing-kernel, we can explicitly write the singular values of $n^{-1} g_{\sigma}(t\mtx{I}_n - \mtx{A})$ as
 \begin{equation}

### 2024-09-26 paper/plots/hessian_density.py

@@ -49,9 +49,9 @@ test_loader = torch.utils.data.DataLoader(torchvision.datasets.MNIST("matrices/m
 
 
 # Set parameter
-t = np.linspace(0.2, 1.0, 100)
+t = np.linspace(0.1, 1.0, 150)
 sigma = 0.005
-m = 500
+m = 1000
 n_Omega = 30
 n_Psi = 10
 data = next(iter(train_loader))
@@ -67,23 +67,37 @@ kernel = lambda t, x: gaussian_kernel(t, x, sigma=sigma, n=H.shape[0])
 H = hessian(model, loss_function, data, spectral_transform=True)
 phi = chebyshev_nystrom(H, t, m, n_Psi, n_Omega, kernel)
 
-plt.plot(t, phi, color="#648FFF", label=r"untrained")
+plt.plot(t * H.scaling_parameter, phi, color="#648FFF", label=r"untrained")
 
-train(model, train_loader, loss_function, validator=validator, n_epochs=5)
-
-H.update(model, loss_function, data)
-phi = chebyshev_nystrom(H, t, m, n_Psi, n_Omega, kernel)
-
-plt.plot(t, phi, color="#DC267F", label=r"epoch $5$")
-
-train(model, train_loader, loss_function, validator=validator, n_epochs=5)
+train(model, train_loader, loss_function, validator=validator, n_epochs=2)
 
 H.update(model, loss_function, data)
 phi = chebyshev_nystrom(H, t, m, n_Psi, n_Omega, kernel)
 
-plt.plot(t, phi, color="#FFB000", label=r"epoch $10$")
+plt.plot(t * H.scaling_parameter, phi, color="#785EF0", label=r"epoch $2$")
+
+train(model, train_loader, loss_function, validator=validator, n_epochs=2)
+
+H.update(model, loss_function, data)
+phi = chebyshev_nystrom(H, t, m, n_Psi, n_Omega, kernel)
+
+plt.plot(t * H.scaling_parameter, phi, color="#DC267F", label=r"epoch $4$")
+
+train(model, train_loader, loss_function, validator=validator, n_epochs=2)
+
+H.update(model, loss_function, data)
+phi = chebyshev_nystrom(H, t, m, n_Psi, n_Omega, kernel)
+
+plt.plot(t * H.scaling_parameter, phi, color="#FE6100", label=r"epoch $6$")
+
[…]

### 2024-10-01 paper/intro.tex
try to fix pipeline

@@
 It needs to be clearer what the approaches proposed in our paper are really about and the novelty wrt existing work has to come out more clearly and concretely.
-???}
+???
+Mention \cite{park-2024-lowrank-approximation} ?.}
 \color{black}
 Instead of discretizing $[a, b]$ into $t_1, \ldots, t_m$, approximating $\Trace(\mtx{B}(t_i))$ with $b(t_i)$, and analyzing the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way is to quantify the error with $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate in each $t \in [a, b]$.
-This approach has already been used in the analysis of low-rank approximations of parameter-dependent matrices \cite{kressner-2024-randomized-lowrank}.
+This approach has already been used in the analysis of low-rank approximations to parameter-dependent matrices \cite{kressner-2024-randomized-lowrank}.
 
 \paragraph{New contributions.} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}.
-Interestingly, we can reuse the same randomization for computing the estimate at ea<ch value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations.
+Interestingly, we can reuse the same randomization for computing the estimate at each value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations.
 We propose certain modifications to the methods from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}.

### 2024-10-01 paper/plots/hessian_density.py

@@ -20,14 +20,8 @@ model = torch.nn.Sequential(
     torch.nn.BatchNorm2d(1),
     torch.nn.Flatten(),
     torch.nn.Linear(676, 10),
-    #torch.nn.Softplus(),
-    #torch.nn.Linear(256, 10),
     torch.nn.Sigmoid(),
 )
-
-# For now, don't execute this
-#if __name__ == "__main__":
-#    exit()
 
 np.random.seed(0)
 
@@ -49,7 +43,7 @@ test_loader = torch.utils.data.DataLoader(torchvision.datasets.MNIST("matrices/m
 
 
 # Set parameter
-t = np.linspace(0.1, 1.0, 150)
+t = np.linspace(0.05, 1.0, 150)
 sigma = 0.005
 m = 1000
 n_Omega = 30
@@ -100,4 +94,4 @@ plt.grid(True, which="both")
 plt.ylabel(r"smoothed spectral density $\phi_{\sigma}(t)$")
 plt.xlabel(r"spectral parameter $t$")
 plt.legend()
-plt.savefig("paper/plots/hessian_density_.pgf", bbox_inches="tight")
+plt.savefig("paper/plots/hessian_density.pgf", bbox_inches="tight")

### 2024-10-01 paper/plots/krylov_aware_spin.py

@@ -11,6 +11,10 @@ from matrices.quantum_spin import hamiltonian, partition_function
 
 np.random.seed(0)
 
+# For now, don't execute this
+if __name__ == "__main__":
+    exit()
+
 N = 20
 s = 0.5
 h = 0.3

### 2024-10-01 paper/re-pro-badge.tex
automatically reproduce paper

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/7c053ed}{7c053ed} on 2024-09-23 at 17:06:23 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/e763331}{e763331} on 2024-10-01 at 08:45:38 UTC.

### 2024-10-01 paper/plots/krylov_aware_density.py
make small adjustments to appendix

@@ -22,8 +22,8 @@ eigvals_st = spectral_transformation(eigvals, min_ev, max_ev)
 # Set parameter
 t = np.linspace(-1, 1, 100)
 sigma_list = np.logspace(-3.5, -1.0, 7)
-n_Omega_list = np.array([ 10, 20, 10, 10, 40])    # b
-n_Psi_list = np.array([ 60, 30, 30, 30, 40])    # m
+n_Omega_list = np.array([ 10, 20, 10, 10, 20])    # b
+n_Psi_list = np.array([ 60, 30, 30, 30, 20])    # m
 n_iter_list = np.array([ 30, 30, 60, 30, None])  # q
 n_reorth_list = np.array([20, 20, 20, 40, None])  # n
 m = 2000
@@ -60,16 +60,16 @@ plt.xlabel(r"smoothing parameter $\sigma$")
 plt.legend()
 plt.xscale("log")
 plt.yscale("log")
-plt.savefig("paper/plots/krylov_aware_density.pgf", bbox_inches="tight")
+plt.savefig("paper/plots/krylov_aware_density_.pgf", bbox_inches="tight")
 
 headline = ["", r"$n_{\mtx{\Omega}}$", r"$n_{\mtx{\Psi}}$", r"$q$", r"$n$", r"time (s)"]
 fmt = [r"${:0.0f}$", r"${:0.0f}$", r"${:0.0f}$", r"${:0.0f}$", r"${:.2f}$"]
 values = np.vstack((n_Omega_list[:-1], n_Psi_list[:-1], n_iter_list[:-1], n_reorth_list[:-1], times[:-1])).T
 
-generate_tex_tabular(values, "paper/tables/krylov_aware_density_KA.tex", headline, labels[:-1], fmt=fmt)
+generate_tex_tabular(values, "paper/tables/krylov_aware_density_KA_.tex", headline, labels[:-1], fmt=fmt)
 
 headline = ["", r"$n_{\mtx{\Omega}}$", r"$n_{\mtx{\Psi}}$", r"$m$", r"time (s)"]
 fmt = [r"${:0.0f}$", r"${:0.0f}$", r"${:0.0f}$", r"${:.2f}$"]
 values = np.vstack((n_Omega_list[-1], n_Psi_list[-1], m, times[-1])).T
 
-generate_tex_tabular(values, "paper/tables/krylov_aware_density_CN.tex", headline, labels[-1:], fmt=fmt)
[…]

### 2024-10-01 paper/re-pro-badge.tex
automatically reproduce paper

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/e763331}{e763331} on 2024-10-01 at 08:45:38 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/c37aaa8}{c37aaa8} on 2024-10-01 at 13:08:31 UTC.

### 2024-10-02 paper/numresults.tex
downsize and adjust parametres to match Krylov-aware

@@
 The Krylov-aware stochastic trace estimation \cite{chen-2023-krylovaware-stochastic} also works well for the approximation of the trace of a matrix function $f(\mtx{A}, t)$ which separately depends on a parameter $t$.
-It also samples two standard Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$, of which $\mtx{\Omega}$ is used to extract a low-rank approximation to $f(\mtx{A}, t)$ from a Krylov-subspace of $\mtx{A}$ and $\mtx{\Psi}$ to estimate the trace of the approximation residual.
+It also samples two standard Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$.
+$\mtx{\Omega}$ is used to extract a low-rank approximation to $f(\mtx{A}, t)$ from a block-Krylov subspace of $\mtx{A}$ generated from $r + q$ iterations of the block-Lanczos method of which the first $r$ are reorthogonalized.
+$\mtx{\Psi}$ is used to estimate the trace of the approximation residual with $r$ iterations of the Lanczos method.
 
@@
     \end{minipage}
-    \caption{For the example from \refsec{subsec:hamiltonian} we compare the Chebyshev-Nyström++ method with the Krylov-Aware stochastic trace estimator in multiple configurations (right).
+    \caption{For the example from \refsec{subsec:hamiltonian} we compare the Chebyshev-Nyström++ (CN++) method with the Krylov-Aware (KA) stochastic trace estimator in multiple configurations (right).
 To this extent, we compute the $L^1$-error of the methods for various levels of smoothing $\sigma$ (left).}
@@
     \end{minipage}
-    \caption{For the Heisenberg spin chain example, we compare the Chebyshev-Nyström++ method with the Krylov-Aware stochastic trace estimator in the same configurations as in \cite[Table 5.1]{chen-2023-krylovaware-stochastic} (right).
+    \caption{For the Heisenberg spin chain example, we compare the Chebyshev-Nyström++ (CN++) method with the Krylov-Aware stochastic trace estimator in the same configurations as in \cite[Table 5.1]{chen-2023-krylovaware-stochastic} (right).
 To this extent, we compute the $L^1$-error of the methods for various choices of the temperature parameter $\beta^{-1}$ (left).}

### 2024-10-02 paper/plots/krylov_aware_density.py

@@ -24,8 +24,8 @@ t = np.linspace(-1, 1, 100)
 sigma_list = np.logspace(-3.5, -1.0, 7)
 n_Omega_list = np.array([ 10, 20, 10, 10, 20])    # b
 n_Psi_list = np.array([ 60, 30, 30, 30, 20])    # m
-n_iter_list = np.array([ 30, 30, 60, 30, None])  # q
-n_reorth_list = np.array([20, 20, 20, 40, None])  # n
+q_list = np.array([20, 20, 40, 20, None])  # q
+n_list = np.array([10, 10, 10, 20, None])  # n
 m = 2000
 
 plt.style.use("paper/plots/stylesheet.mplstyle")
@@ -42,10 +42,10 @@ for j, sigma in enumerate(sigma_list):
     kernel = lambda t, x: gaussian_kernel(t, x, sigma=sigma, n=A.shape[0])
     baseline = form_spectral_density(eigvals_st, t, kernel)
 
-    for i, (n_Psi, n_Omega, n_iter, n_reorth) in enumerate(zip(n_Psi_list, n_Omega_list, n_iter_list, n_reorth_list)):
+    for i, (n_Psi, n_Omega, q, n) in enumerate(zip(n_Psi_list, n_Omega_list, q_list, n_list)):
         t0 = time.time()
         if i < len(n_Psi_list) - 1:
-            estimate = krylov_aware(A_st, t, n_iter, n_reorth, n_Psi, n_Omega, kernel)
+            estimate = krylov_aware(A_st, t, n + q, q, n_Psi, n_Omega, kernel)
         else:
             estimate = chebyshev_nystrom(A_st, t, m, n_Psi, n_Omega, kernel)
         error[i, j] = 2 * np.mean(np.abs(estimate - baseline))
@@ -62,9 +62,9 @@ plt.xscale("log")
 plt.yscale("log")
 plt.savefig("paper/plots/krylov_aware_density.pgf", bbox_inches="tight")
 
-headline = ["", r"$n_{\mtx{\Omega}}$", r"$n_{\mtx{\Psi}}$", r"$q$", r"$n$", r"time (s)"]
+headline = ["", r"$n_{\mtx{\Omega}}$", r"$n_{\mtx{\Psi}}$", r"$q$", r"$r$", r"time (s)"]
 fmt = [r"${:0.0f}$", r"${:0.0f}$", r"${:0.0f}$", r"${:0.0f}$", r"${:.2f}$"]
[…]

### 2024-10-02 paper/plots/krylov_aware_spin.py

@@ -11,10 +11,6 @@ from matrices.quantum_spin import hamiltonian, partition_function
 
 np.random.seed(0)
 
-# For now, don't execute this
-if __name__ == "__main__":
-    exit()
-
 N = 20
 s = 0.5
 h = 0.3
@@ -27,10 +23,10 @@ E_max, = sp.sparse.linalg.eigsh(A, k=1, which="LA", return_eigenvectors=False, t
 A_st = spectral_transformation(A, E_min, E_max)
 
 # Set parameter
-n_Omega_list = np.array([8, 0, 8, 4, 40])    # b
-n_Psi_list = np.array([0, 13, 13, 6, 40])    # m
-n_iter_list = np.array([30 + 50, 50, 30 + 50, 30 + 50, None])  # q
-n_reorth_list = np.array([50, 50, 50, 50, None])  # n
+n_Omega_list = np.array([8, 0, 8, 4, 10])  # b
+n_Psi_list = np.array([0, 13, 13, 6, 10])  # m
+q_list = np.array([30, 0, 30, 30, None])  # q
+n_list = np.array([50, 50, 50, 50, None])  # n
 m = 50
 
 plt.style.use("paper/plots/stylesheet.mplstyle")
@@ -46,11 +42,11 @@ baseline = partition_function(betas, N, h, J, E_min)
 error = np.empty((len(n_Psi_list), len(betas)))
 times = np.zeros(len(n_Psi_list))
 
-for i, (n_Psi, n_Omega, n_iter, n_reorth) in enumerate(zip(n_Psi_list, n_Omega_list, n_iter_list, n_reorth_list)):
+for i, (n_Psi, n_Omega, q, n) in enumerate(zip(n_Psi_list, n_Omega_list, q_list, n_list)):
     t0 = time.time()
     if i < len(n_Psi_list) - 1:
         function = lambda beta, x: np.exp(-np.multiply.outer(beta, x - E_min))
-        estimate = krylov_aware(A, betas, n_iter, n_reorth, n_Omega, n_Psi, function)
+        estimate = krylov_aware(A, betas, n + q, q, n_Omega, n_Psi, function)
     else:
         function = lambda beta, x: np.exp(-np.multiply.outer(beta, x + 1))
[…]

### 2024-10-04 paper/numresults.tex
improve plots and add loss to hessian spectral density example

--- 2024-09-25 paper/numresults.tex
@@
 \label{sec:results}
+
+In this last section, we test our algorithm in multiple scenarios.
+First, we numerically confirm our theoretical convergence bounds by approximating the spectral density of a Hamiltonian arising from electronic structure interaction.
+Subsequently, we compare our method to the Krylov-aware stochastic trace estimator \cite{chen-2023-krylovaware-stochastic} once on the electronic structure example and once for estimating the trace of a more general parameter-dependent matrix appearing in statistical thermodynamics.
+In the end, we demonstrate that our algorithm can also be used to monitor the optimization process of a neural network.
 
@@
 Again, since for small $\sigma$ the matrix function $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ has a small numerical rank, a Nyström approximation alone is already enough to achieve a good approximation.
-On the other hand, when $\sigma$ is large, the Nyström approximation alone is not effective, and the correction with the residual trace becomes indispensable.
+On the other hand, when $\sigma$ is large, the Nyström approximation by itself is not effective, and the correction with the residual trace becomes indispensable.
 
@@
 
-The Krylov-aware stochastic trace estimation \cite{chen-2023-krylovaware-stochastic} also works well for the approximation of the trace of a matrix function $f(\mtx{A}, t)$ which separately depends on a parameter $t$.
-It also samples two standard Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$, of which $\mtx{\Omega}$ is used to extract a low-rank approximation to $f(\mtx{A}, t)$ from a Krylov-subspace of $\mtx{A}$ and $\mtx{\Psi}$ to estimate the trace of the approximation residual.
+The Krylov-aware stochastic trace estimation \cite{chen-2023-krylovaware-stochastic} is also well suited for approximating the trace of a matrix function $f(\mtx{A}, t)$ which separately depends on a parameter $t$.
+It also samples two standard Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$.
+$\mtx{\Omega}$ is used to extract a low-rank approximation to $f(\mtx{A}, t)$ from a block-Krylov subspace of $\mtx{A}$ generated from $r$ iterations of the block-Lanczos method with reorthogonalization and subsequently $q$ without.
+$\mtx{\Psi}$ is then used to estimate the trace of the approximation residual with $r$ iterations of the Lanczos method.
 
-We run an optimized implementation of the Krylov-aware stochastic trace estimator \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} in multiple configurations on the example from \refsec{subsec:hamiltonian} and plot their approximation errors for various values of the smoothing parameter $\sigma$ in \reffig{fig:krylov-aware-density}.
+We run an optimized implementation of the Krylov-aware stochastic trace estimator \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} in multiple configurations on the example from \refsec{subsec:hamiltonian} and plot the approximation errors for logarithmically spaced values of the smoothing parameter $\sigma$ in \reffig{fig:krylov-aware-density}.
 For reference, we add the error of the Chebyshev-Nyström++ method on parameters which lead to a comparable run-time.
@@
         \newline
+        \vspace{15pt}
         \newline
         \input{tables/krylov_aware_density_CN.tex}
-        %\renewcommand{\arraystretch}{1.2}
-        %\begin{tabular}{@{}lccccc@{}}
-        %\toprule
-        %& $n_{\mtx{\Omega}}$ & $n_{\mtx{\Psi}}$ & $q$ & $n$ & runtime (s)\\
-        %\midrule
-        %KA I & 10 & 60 & 30 & 20 & 9.13 \\
-        %KA II & 20 & 30 & 30 & 20 & 9.13 \\
-        %KA III & 10 & 30 & 60 & 40 & 9.13 \\
-        %KA IV & 10 & 30 & 30 & 20 & 9.13 \\
-        %\bottomrule
-        %\end{tabular}
-        %\begin{tabular}{@{}lcccc@{}}
-        %\\
-        %\toprule
-        %& $n_{\mtx{\Omega}}$ & $n_{\mtx{\Psi}}$ & $m$ & runtime (s)\\
-        %\midrule
-        %CN++ & 40  & 40 & 2000 & 9.13 \\
-        %\bottomrule
-        %\end{tabular}
-        %\caption{Comparison with Krylov-Aware stochastic trace estimation.}
     \end{minipage}
-    \caption{Comparison with Krylov-Aware stochastic trace estimation.}
+    \caption{For the example from \refsec{subsec:hamiltonian} we compare the Chebyshev-Nyström++ (CN++) method with the Krylov-Aware (KA) stochastic trace estimator in multiple configurations.
+To this extent, we compute the $L^1$-error of the methods for various levels of smoothing $\sigma$ (left).
+The parameters of the algorithms and run-time averaged over all values of $\sigma$ can be found in the tables (right).}
     \label{fig:krylov-aware-density}
@@
     \end{minipage}
-    \caption{Comparison with Krylov-Aware stochastic trace estimation.}
+    \caption{For the Heisenberg spin chain example, we compare the Chebyshev-Nyström++ (CN++) method with the Krylov-Aware stochastic trace estimator in the same configurations as in \cite[Table 5.1]{chen-2023-krylovaware-stochastic} (right).
+To this extent, we compute the $L^1$-error of the methods for various choices of the temperature parameter $\beta^{-1}$ (left).}
     \label{fig:krylov-aware-spin}
@@
 
-\todo{Hessian spectrum analysis?}
+When optimizing a neural network over some data, it is of great interest to determine whether a minimum has been found and if this minimum is robust, i.e.
+a small perturbation of the parameters of the neural network will not lead to a significant decrease in the fit of the neural network.
+Both properties are reflected in the eigenvalues of the Hessian matrix $\mtx{A}$ with respect to some loss function: if all eigenvalues are non-negative, then the loss attains a local minimum, and if additionally all of them are small, the minimum is robust.
+
+Since neural networks are usually parametrized by thousands of parameters, assembling the Hessian matrix explicitly and then computing its eigenvalues is out of question.
+Luckily, there exists a method which can exactly compute matrix-vector products $\mtx{A} \vct{x}$ for any vector $\vct{x}$ which scales proportionally with the number of parameters \cite{pearlmutter-1994-fast-exact}.
+
+To demonstrate that our algorithm can effectively be applied in the setting of neural network optimization, we approximate the spectral density of a small fully connected convolutional neural network with 6782 parameters, which we train on the handwritten digit classification task given by the MNIST dataset\footnote{Handwritten digit classification; taken from \url{http://yann.lecun.com/exdb/mnist}.}.
+We plot the approximation of the spectral density of the Hessian matrix of the untrained neural network, and at different stages of training in \reffig{fig:hessian-density}, as well as the corresponding mean squared error loss.
+It can well be observed that the eigenvalues creep towards zero as the training proceeds.
+Furthermore, by the absence of large eigenvalues in the spectrum, we can tell that in the final epoch, the neural network attains a more robust state than in previous epochs.
+
+\begin{figure}
+    \begin{minipage}[c]{.49\linewidth}
+        \centering
+        \input{plots/hessian_density.pgf}
+    \end{minipage}\hfill%
+    \begin{minipage}[c]{.49\linewidth}
+        \centering
+        \input{plots/hessian_density_loss.pgf}
+    \end{minipage}
+    \caption{The mean squared error training loss (right) and the corresponding approximate spectral density of the Hessian matrix of a fully connected convolutional neural network in different epochs of training on the MNIST dataset (left).
+Computed with the Chebyshev-Nyström++ method (\refalg{alg:nystrom-chebyshev-pp}) on $n_{\mtx{\Omega}} = 10$, $n_{\mtx{\Psi}} = 10$, $m = 1000$, and $\sigma = 0.005$.}
+    \label{fig:hessian-density}
+\end{figure}

### 2024-10-04 paper/spectraldensity.tex
improve plots and add loss to hessian spectral density example

@@
 \end{equation}
-then lets us define the matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ whose spectrum is identical to the one of $\mtx{A}$ but shrunk to the interval $[-1, 1]$, for which the expansion \refequ{equ:matrix-expansion} is valid.
-Note that through this transformation we will also have to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when reverting the spectral transform $\tau$.
+then lets us define the matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ whose spectrum is identical to the one of $\mtx{A}$ but shrunk to the interval $[-1, 1]$, for which the expansion \refequ{equ:matrix-expansion} is valid\footnote{Note that we will have to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when inverting the transformation.}.
 From now on, we will assume the eigenvalues of the matrix $\mtx{A}$ to be contained in $[-1, 1]$.
@@
 provided we set $\varepsilon = 10^{-16}$ and choose $\sigma$ roughly between $10^{-8}$ and $1$.
-Hence, choosing $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma)$ will approximations which can be significantly better than \refthm{thm:chebyshev-nystrom}.
+Hence, choosing $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma)$ will yield approximations which can be significantly better than \refthm{thm:chebyshev-nystrom}.

### 2024-10-04 matrices/neural_network.py

@@ -53,6 +53,7 @@ def train(model, data_loader, loss_function=torch.nn.MSELoss(), optimizer=torch.
     """
 
     optimizer = optimizer(model.parameters(), **optimizer_parameters)
+    loss_list = []
 
     for epoch in range(n_epochs):
         epoch_loss = 0
@@ -72,8 +73,10 @@ def train(model, data_loader, loss_function=torch.nn.MSELoss(), optimizer=torch.
             # Accumulate the loss for this epoch
             epoch_loss += loss.item()
 
+        epoch_loss /= len(data_loader)
+        loss_list.append(epoch_loss)
+
         if verbose:
-            epoch_loss /= len(data_loader)
             msg = "[Epoch {:2d}/{}] \t".format(epoch + 1, n_epochs)
             msg += "Time: {:.2f} s - ".format(time.time() - start_time)
             msg += "Loss: {:.4f}".format(epoch_loss)
@@ -82,6 +85,8 @@ def train(model, data_loader, loss_function=torch.nn.MSELoss(), optimizer=torch.
                 msg += validator(model)
 
             print(msg)
+
+    return loss_list
 
 
 def accuracy_validator(model, data_loader):

### 2024-10-04 paper/plots/hessian_density.py

@@ -1,9 +1,3 @@
-"""
-hessian_density.py
-
-TODO: Only look at spectrum at range far in the negative or positive
-"""
-
 import __context__
 
 import numpy as np
@@ -30,7 +24,7 @@ train_loader = torch.utils.data.DataLoader(torchvision.datasets.MNIST("matrices/
                                            train=True,
                                            transform=torchvision.transforms.ToTensor()),
                                            batch_size=64,
-                                           shuffle=True
+                                           shuffle=False
 )
 
 test_loader = torch.utils.data.DataLoader(torchvision.datasets.MNIST("matrices/mnist_data", 
@@ -46,11 +40,12 @@ test_loader = torch.utils.data.DataLoader(torchvision.datasets.MNIST("matrices/m
 t = np.linspace(0.05, 1.0, 150)
 sigma = 0.005
 m = 1000
-n_Omega = 30
+n_Omega = 10
 n_Psi = 10
 data = next(iter(train_loader))
 
 plt.style.use("paper/plots/stylesheet.mplstyle")
+plt.figure(figsize=(3, 3))
 
 # Train the model on the MNIST data set
 loss_function = lambda x, y: torch.nn.MSELoss()(x, torch.nn.functional.one_hot(y, num_classes=10).to(dtype=torch.float))
@@ -63,28 +58,34 @@ phi = chebyshev_nystrom(H, t, m, n_Psi, n_Omega, kernel)
 
 plt.plot(t * H.scaling_parameter, phi, color="#648FFF", label=r"untrained")
 
-train(model, train_loader, loss_function, validator=validator, n_epochs=2)
+epoch_loss = 0
+for batch in train_loader:
+    epoch_loss += loss_function(model(*batch[:-1]), batch[-1]).item()
+
+loss_list = [epoch_loss / len(train_loader)]
+
+loss_list.extend(train(model, train_loader, loss_function, validator=validator, n_epochs=2))
 
 H.update(model, loss_function, data)
 phi = chebyshev_nystrom(H, t, m, n_Psi, n_Omega, kernel)
 
 plt.plot(t * H.scaling_parameter, phi, color="#785EF0", label=r"epoch $2$")
 
-train(model, train_loader, loss_function, validator=validator, n_epochs=2)
[…]

### 2024-10-04 paper/plots/krylov_aware_spin.py

@@ -11,6 +11,10 @@ from matrices.quantum_spin import hamiltonian, partition_function
 
 np.random.seed(0)
 
+# Don't execute this because memory overflow on GitHub due to private repository
+#if __name__ == "__main__":
+#    quit()
+
 N = 20
 s = 0.5
 h = 0.3
@@ -23,10 +27,10 @@ E_max, = sp.sparse.linalg.eigsh(A, k=1, which="LA", return_eigenvectors=False, t
 A_st = spectral_transformation(A, E_min, E_max)
 
 # Set parameter
-n_Omega_list = np.array([8, 0, 8, 4, 10])  # b
-n_Psi_list = np.array([0, 13, 13, 6, 10])  # m
+n_Omega_list = np.array([8, 0, 8, 4, 40])  # b
+n_Psi_list = np.array([0, 13, 13, 6, 40])  # m
 q_list = np.array([30, 0, 30, 30, None])  # q
-n_list = np.array([50, 50, 50, 50, None])  # n
+r_list = np.array([50, 50, 50, 50, None])  # n
 m = 50
 
 plt.style.use("paper/plots/stylesheet.mplstyle")
@@ -42,11 +46,11 @@ baseline = partition_function(betas, N, h, J, E_min)
 error = np.empty((len(n_Psi_list), len(betas)))
 times = np.zeros(len(n_Psi_list))
 
-for i, (n_Psi, n_Omega, q, n) in enumerate(zip(n_Psi_list, n_Omega_list, q_list, n_list)):
+for i, (n_Psi, n_Omega, q, r) in enumerate(zip(n_Psi_list, n_Omega_list, q_list, r_list)):
     t0 = time.time()
     if i < len(n_Psi_list) - 1:
         function = lambda beta, x: np.exp(-np.multiply.outer(beta, x - E_min))
-        estimate = krylov_aware(A, betas, n + q, q, n_Omega, n_Psi, function)
+        estimate = krylov_aware(A, betas, r + q, q, n_Omega, n_Psi, function)
     else:
         function = lambda beta, x: np.exp(-np.multiply.outer(beta, x + 1))
         estimate = chebyshev_nystrom(A_st, betas * (E_max - E_min) / 2, m, n_Psi, n_Omega, function, kappa=-1, rcond=1e-10)
[…]

### 2024-10-04 paper/re-pro-badge.tex
automatically reproduce paper

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/c37aaa8}{c37aaa8} on 2024-10-01 at 13:08:31 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/de1c340}{de1c340} on 2024-10-04 at 10:35:14 UTC.

### 2024-10-30 paper/analysis.tex
add citations of recently published preprint

@@
 This becomes more difficult when $\mtx{B}(t)$ is given implicitly (e.g., as a matrix function) and can only be accessed through matrix-vector products.
-The common theme of all methods analyzed in this section is to extract trace estimates from multiplying $\mtx{B}(t)$ with fixed random vectors
+The common theme of all methods analyzed in this section is to extract trace estimates from multiplying $\mtx{B}(t)$ with fixed random vectors.
 
-as we will see in Section ???.
+%as we will see in Section ???.
 \color{black}
@@
 %
-\subsubsection{Nyström approximation}
+\subsubsection{Nyström estimator}
 \label{subsec:nystrom}
@@
         holds with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$ for 
-        a universal constant $c > 0$.
+        $c = 1 + 9 \sqrt{2} e^4$.
 In particular, given $\varepsilon > 0$ and $\delta \in (0, 1)$, the bound $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$ holds with probability at least $1-\delta$ if $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1}))$.

### 2024-10-30 paper/intro.tex
add citations of recently published preprint

@@
 Several techniques have been developed for approximating $\phi$, including moment matching~\cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time} and polynomial approximation, either implicitly by a Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or explicitly by expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}.
+Some of these methods have also been combined with a deflation step to yield even more efficient algorithms \cite{lin-2017-randomized-estimation, bhattacharjee-2024-improved-spectral}.
 
@@
 \end{equation}
-for some fixed smoothing kernel $g_{\sigma}$, typically a Gaussian~\cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian~\cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}.
+for some fixed smoothing kernel $g_{\sigma}$, typically a Gaussian~\cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}.
 We thus arrive at the problem of computing the trace of the parameter-dependent matrix function $\mtx{B}(t) \equiv g_{\sigma}(t \mtx{I}_n - \mtx{A}) \in \mathbb{R}^{n \times n}$.

### 2024-10-30 paper/spectraldensity.tex
add citations of recently published preprint

@@
 where $\mathcal{S}$ represents an appropriately chosen space of test functions.
-Among others, \cite{lin-2016-approximating-spectral} proposes $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - s), s \in [a, b]\}$, \cite{chen-2021-analysis-stochastic} uses $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [a, b] \}$ where $\Theta$ is the Heaviside step function, and \cite{braverman-2022-sublinear-time} uses $\mathcal{S} = \{f : |f(t) - f(s)| \leq |t - s| \}$ for which the error \refequ{equ:error-metric} is equivalent to the Wasserstein-1 distance between $\phi$ and $\phi_{\sigma}$.
+Among others, \cite{lin-2016-approximating-spectral} proposes $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - s), s \in [a, b]\}$, \cite{chen-2021-analysis-stochastic} uses $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [a, b] \}$ where $\Theta$ is the Heaviside step function, and \cite{braverman-2022-sublinear-time,bhattacharjee-2024-improved-spectral} use $\mathcal{S} = \{f : |f(t) - f(s)| \leq |t - s| \}$ for which the error \refequ{equ:error-metric} is equivalent to the Wasserstein-1 distance between $\phi$ and $\phi_{\sigma}$.
 In this last metric, provided we assume the eigenvalues to be somewhat uniformly distributed and $\sigma$ to be rather small, a simple calculation allows us to roughly estimate the smoothing error \refequ{equ:error-metric} to be $\sigma$.
@@
         \State Define $\vct{\mu} \in \mathbb{R}^{2m + 1}$ whose first $m + 1$ entries are the coefficients $\mu_l$ and the remaining $m$ are $0$
-    \State Compute $\vct{f} = \DCT(\vct{\mu})$
-    \State Compute $\vct{\nu} = \DCT^{-1}(\vct{f} \odot \vct{f})$ whose entries are the coefficients $\nu_l$
+    \State Compute $\vct{f} = \operatorname{DCT}(\vct{\mu})$
+    \State Compute $\vct{\nu} = \operatorname{DCT}^{-1}(\vct{f} \odot \vct{f})$ whose entries are the coefficients $\nu_l$
     \label{lin:inverse-DCT}

### 2024-10-30 paper/re-pro-badge.tex
automatically reproduce paper

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/de1c340}{de1c340} on 2024-10-04 at 10:35:14 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/bd4e56b}{bd4e56b} on 2024-10-30 at 17:08:31 UTC.

### 2024-10-31 paper/re-pro-badge.tex
automatically reproduce paper

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/bd4e56b}{bd4e56b} on 2024-10-30 at 17:08:31 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/f30af3c}{f30af3c} on 2024-10-31 at 11:41:34 UTC.

### 2025-02-14 paper/analysis.tex
push changes from Overleaf

@@
 In this section, we consider a general parameter-dependent matrix
-\begin{equation*}
-    \mtx{B}(t) = \begin{bmatrix}
-        b_{11}(t) & b_{12}(t) & \dots & b_{1n}(t) \\
-        b_{21}(t) & b_{22}(t) & \dots & b_{2n}(t) \\
-        \vdots & \vdots & \ddots & \vdots \\
-        b_{n1}(t) & b_{n2}(t) & \dots & b_{nn}(t) \\
-    \end{bmatrix} \in \mathbb{R}^{n \times n},
-\end{equation*}
-where each entry $b_{ij}(t)$ is a continuous function on a bounded interval $[a,b]$. \emph{If} these entries are given explicitly, we can simply compute the trace by its definition 
+$\mtx{B}(t) \in \mathbb{R}^{n \times n}$, where each
+entry $b_{ij}(t)$ is a 
+%\begin{equation*}
+%     \mtx{B}(t) = \begin{bmatrix}
+%         b_{11}(t) & b_{12}(t) & \dots & b_{1n}(t) \\
+%         b_{21}(t) & b_{22}(t) & \dots & b_{2n}(t) \\
+%         \vdots & \vdots & \ddots & \vdots \\
+%         b_{n1}(t) & b_{n2}(t) & \dots & b_{nn}(t) \\
+%     \end{bmatrix} \in \mathbb{R}^{n \times n},
+% \end{equation*}
+continuous function on a bounded interval $[a,b]$. ??? UNTIL HERE ??? \emph{If} these entries are given explicitly, we can simply compute the trace by its definition 
 $\Trace(\mtx{B}(t)) = b_{11}(t)+ \cdots + b_{nn}(t)$.

### 2025-02-14 paper/analysis.tex
convert Hutchinson proof to sub-gamma

@@
     \begin{equation} \label{eq:ghbound}
-        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq \gamma c \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}}  \right) \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
+        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq 4 \gamma \int_{a}^{b} \sqrt{\frac{p}{n_{\mtx{\Psi}}}}  \lVert \mtx{B}(t) \rVert _F + \frac{2 p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 ~\mathrm{d}t
     \end{equation}
-    holds with probability at least $1 - \gamma^{-p}$ and for $c = 4 (1 + \sqrt{1 + \pi/4})$.
+    holds with probability at least $1 - \gamma^{-p}$.
     In particular, given $\varepsilon \in (0, 1)$ and $\delta \in (0, e^{-1})$, the bound $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$ holds
@@
 \begin{proof}
-    The proof proceeds by bounding moments for fixed values of $t$ and then uses Minkowski's integral inequality.
+    We first consider a fixed value for $t$.
+From the proof of \cite[Theorem 1]{cortinovis-2022-randomized-trace}, we know that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is a $(2 \lVert \mtx{B}(t) \rVert _F^2 / n_{\mtx{\Psi}}, 2 \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}})$-sub-gamma random variable.
+Therefore, by \reflem{lem:sub-gamma-moments}
+    \begin{equation}
+        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq 4  \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F + 8 \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2.
+        \label{equ:moment-bound-hutchinson}
+    \end{equation}
+%    The proof proceeds by bounding moments for fixed values of $t$ and then uses Minkowski's integral inequality.
 We start with a standard embedding trick (see, e.g.,~\cite[Theorem 1]{cortinovis-2022-randomized-trace}) for which we define
-    \begin{equation}
-        \widetilde{\mtx{B}}(t)
-        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
-            \mtx{B}(t) & & \\
-            & \ddots & \\
-            & & \mtx{B}(t)
-        \end{pmatrix}
-        \quad \text{and} \quad
-        \widetilde{\vct{\psi}} = \begin{pmatrix}
-            \vct{\psi}_1 \\
-            \vdots \\
-            \vct{\psi}_{n_{\mtx{\Psi}}}
-        \end{pmatrix}.
-    \end{equation}
-    Note that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) =  \Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}}$, $\lVert \widetilde{\mtx{B}}(t) \rVert _2 = \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}}$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold.
+%    \begin{equation}
+%        \widetilde{\mtx{B}}(t)
+%        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
+%            \mtx{B}(t) & & \\
+%            & \ddots & \\
+%            & & \mtx{B}(t)
+%        \end{pmatrix}
+%        \quad \text{and} \quad
+%        \widetilde{\vct{\psi}} = \begin{pmatrix}
+%            \vct{\psi}_1 \\
+%            \vdots \\
+%            \vct{\psi}_{n_{\mtx{\Psi}}}
+%        \end{pmatrix}.
+%    \end{equation}
+%    Note that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) =  \Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}}$, $\lVert \widetilde{\mtx{B}}(t) \rVert _2 = \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}}$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold.
 Therefore, by \reflem{lem:hutchinson-sub-exponential}, $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is a centered and $(2 \lVert \mtx{B}(t) \rVert _F^2 (1 - 1/\beta)^{-1} / n_{\mtx{\Psi}}, 2 \lVert \mtx{B}(t) \rVert _2 \beta / n_{\mtx{\Psi}})$-sub-exponential random variable for any $\beta > 1$.
 Thus, we can apply \reflem{lem:sub-exponential-moments} to bound
-    \begin{equation}
-        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq 2 \left(\sqrt{\frac{\pi}{1-1/\beta}} \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F + 4 \beta \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 \right).
-    \end{equation}
-    Bounding $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$ and letting $\beta = (1 + \sqrt{1 + \pi/4}) / 2$ we get
-    \begin{equation}
-        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq \underbrace{4 (1 + \sqrt{1 + \pi/4})}_{=~c} \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}} \right) \lVert \mtx{B}(t) \rVert _F.
-        \label{equ:moment-bound-hutchinson}
-    \end{equation}
+%    \begin{equation}
+%        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq 2 \left(\sqrt{\frac{\pi}{1-1/\beta}} \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F + 4 \beta \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 \right).
+%    \end{equation}
+%    Bounding $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$ and letting $\beta = (1 + \sqrt{1 + \pi/4}) / 2$ we get
+%    \begin{equation}
+%        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq \underbrace{4 (1 + \sqrt{1 + \pi/4})}_{=~c} \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}} \right) \lVert \mtx{B}(t) \rVert _F.
+%        \label{equ:moment-bound-hutchinson}
+%    \end{equation}
 
@@
         &\leq \int_{a}^{b} \mathbb{E}^{p}\left[ \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) \right]~\mathrm{d}t \notag \\
-        &\leq c \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}}  \right) \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t .
+        &\leq 4 \int_{a}^{b}  \sqrt{\frac{p}{n_{\mtx{\Psi}}}}  \lVert \mtx{B}(t) \rVert _F + \frac{2 p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 ~\mathrm{d}t.
     \end{align}
@@
     \begin{equation} 
-        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq \gamma c \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}}  \right) \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
+        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq 4 \gamma \int_{a}^{b} \sqrt{\frac{p}{n_{\mtx{\Psi}}}}  \lVert \mtx{B}(t) \rVert _F + \frac{2 p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 ~\mathrm{d}t.
     \end{equation}
@@
 
-    Setting $\gamma = e$ and choosing $p = \log(\delta^{-1}) \geq 1$, the second part of the theorem follows when taking $n_{\mtx{\Psi}} = \lceil 2 c^2 e^2 \log(\delta^{-1}) \varepsilon^{-2} \rceil$.
+    Bounding $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$, setting $\gamma = e$, and choosing $p = \log(\delta^{-1}) \geq 1$, the second part of the theorem follows for example when taking $n_{\mtx{\Psi}} = \lceil (8 e)^2 \log(\delta^{-1}) \varepsilon^{-2} \rceil$.
 \end{proof}
@@
 
-\refthm{thm:hutchinson} coincides with the analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, except for a slightly more restricted choice of $\delta$.
+\refthm{thm:hutchinson} coincides with the analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, except for a slightly more restricted choice of the failure probability $\delta$.
 
@@
     \end{equation}
-    if $n_{\mtx{\Psi}} \geq 2 c^2 e^2 \log(\delta^{-1}) \varepsilon^{-2} \rho^2$, where we identify
+    if $n_{\mtx{\Psi}} = \mathcal{O}(\log(\delta^{-1}) \varepsilon^{-2} \rho^2)$, where we identify
     \begin{equation}

### 2025-02-14 paper/paper.tex
add krylov-aware numerical results
write down experimental setup
improve plots and add loss to hessian spectral density example
add citations of recently published preprint
push changes from Overleaf
convert Hutchinson proof to sub-gamma

@@
 \title{Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation}
+
+\author{Fabio Matti\thanks{Institute of Mathematics, EPFL, 1015 Lausanne, Switzerland.
+{\textbf{(\href{mailto:fabio.matti@epfl.ch}{fabio.matti}, \href{mailto:haoze.he@epfl.ch}{haoze.he}, \href{mailto:daniel.kressner@epfl.ch}{daniel.kressner}, \href{mailto:hysan.lam@epfl.ch}{hysan.lam})@epfl.ch}}.}
+\and Haoze He\footnotemark[1]
+\and Daniel Kressner\footnotemark[1]
+\and Hei Yin Lam\footnotemark[1]}
 
@@
 \begin{itemize}
+%\item Add author list. Matti first, rest (He/Kressner/Lam) in alphabetical order.
  \item The text style is a bit too generous but no need to change this now. \textcolor{green}{Ok! Let's talk about it at some point.}
@@
 
-\section{Conclusion}
-\label{sec:conclusion}
-
-\todo{
-\begin{itemize}
-    \item Robustness (can choose $m$ and $n_{\mtx{\Omega}}$, $n_{\mtx{\Psi}}$ as large as one wants, and not break algorithm, unlike Krylov-aware Lanczos)
-\end{itemize}
-}
-
+%\section{Conclusion}
+%\label{sec:conclusion}
+%
+%\todo{
+%\begin{itemize}
+%    \item Robustness (can choose $m$ and $n_{\mtx{\Omega}}$, $n_{\mtx{\Psi}}$ as large as one wants, and not break algorithm, unlike Krylov-aware Lanczos)
+%\end{itemize}
+%}
 \paragraph{Acknowledgements.} The authors gratefully acknowledge the help of Lin Lin in reproducing the example in \refsec{subsec:hamiltonian}.

### 2025-02-17 paper/analysis.tex
review comments

@@
 \color{blue}
-In this section, we consider a general parameter-dependent matrix
-$\mtx{B}(t) \in \mathbb{R}^{n \times n}$, where each
+Consider a general parameter-dependent matrix
+$\mtx{B}(t) \in \mathbb{R}^{n \times n}$ for which each
 entry $b_{ij}(t)$ is a 
@@
 % \end{equation*}
-continuous function on a bounded interval $[a,b]$. ??? UNTIL HERE ??? \emph{If} these entries are given explicitly, we can simply compute the trace by its definition 
-$\Trace(\mtx{B}(t)) = b_{11}(t)+ \cdots + b_{nn}(t)$.
-This becomes more difficult when $\mtx{B}(t)$ is given implicitly (e.g., as a matrix function) and can only be accessed through matrix-vector products.
-The common theme of all methods analyzed in this section is to extract trace estimates from multiplying $\mtx{B}(t)$ with fixed random vectors.
-
-%as we will see in Section ???.
-\color{black}
+continuous function on a bounded interval $[a,b]$. In this section, we 
+describe and analyze methods that estimate, for several values of $t$, the 
+trace $\Trace(\mtx{B}(t)) = b_{11}(t)+ \cdots + b_{nn}(t)$
+from matrix-vector products of $B(t)$ with \emph{constant} random vectors.
+For one \emph{fixed} value of $t$, our methods reduce to known algorithms.
+Once one considers several, potentially many values of $t$, important differences arise from the fact that the random vectors do not depend on $t$.
+On the one hand, as we will see in~\refsec{sec:application}, this allows one to reuse computations.
+On the other hand, this significantly complicates the analysis, which will be the main focus of this section.
 
@@
 
-\paragraph{Girard-Hutchinson estimator.} The trace of a parameter-dependent matrix can be approximated with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ stochastically independent standard Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$, i.e.
-vectors whose entries are independent standard Gaussian random variables, to form
+\paragraph{Girard-Hutchinson estimator.} One of the simplest methods for trace estimation, the Girard-Hutchinson estimator~\cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator} proceeds by sampling $n_{\mtx{\Psi}}$ stochastically independent Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$
+and computes the approximation 
 \begin{equation}
-    \Hutch{\mtx{\Psi}}(\mtx{B}(t))
-    = \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j
-    = \frac{1}{n_{\mtx{\Psi}}} \Trace( \mtx{\Psi}^{\top} \mtx{B}(t) \mtx{\Psi})
+    \Trace(\mtx{B}(t)) \approx \Hutch{\mtx{\Psi}}(\mtx{B}(t))
+    := \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j
+    = \frac{1}{n_{\mtx{\Psi}}} \Trace( \mtx{\Psi}^{\top} \mtx{B}(t) \mtx{\Psi}).
     \label{equ:hutchinson-trace-estimator}
 \end{equation}
-where $\mtx{\Psi} = [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\mtx{\Psi}}}] \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$ is a standard Gaussian random matrix.
-Other choices for the distribution of the random vectors are possible, for example by uniformly sampling from $\{-1, +1\}$ or from the $(n-1)$-sphere.
-Our theoretical developments only hold in the Gaussian case.
+Here, $\mtx{\Psi} = [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\mtx{\Psi}}}] \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$ is a Gaussian random matrix, that is, 
+its entries are independent standard normal random variables.
+There are other popular choices for the probability distribution, such as the Rademacher distribution. Our theoretical developments, however, only apply to the Gaussian case.
 
-The Girard-Hutchinson estimator distinguishes itself for its simplicity.
-It has already been used to approximate traces of parameter-dependent matrices for spectral density estimation in the Haydock method \cite{haydock-1972-electronic-structure}, as well as the Delta-Gauss-Legendre and Delta-Gauss-Chebyshev methods of \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation}.
-However, as a Monte Carlo estimate, its accuracy is often quite disappointing: if we aim to increase the accuracy of the estimate by one digit, we would need to increase the number of queries by a factor of one hundred.
+The Girard-Hutchinson estimator is a popular choice for spectral density estimation; it has been used in the Haydock method \cite{haydock-1972-electronic-structure}, as well as the Delta-Gauss-Legendre and Delta-Gauss-Chebyshev methods from~\cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation}.
+However, it shares the disadvantage of Monte Carlo methods: Its error is proportional to $n^{-1/2}_{\mtx{\Psi}}$, making it expensive in applications that require decent accuracy.
 
-\paragraph{Nyström estimator.} Matrices with fast singular value decay can well be approximated with a low-rank approximation.
-Intuitively, it should be sufficient to approximate the trace of a quadratic matrix with the trace of a good low-rank approximation.
-Our application, the spectral density estimation (\refsec{sec:application}) of symmetric matrices, requires us to approximate the trace of a symmetric positive semi-definite matrix, originating from a non-negative smoothing kernel.
-In this scenario, the Nyström approximation is the most suitable low-rank approximation \cite{gittens-2013-revisiting-nystrom}.
-For a standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$, it is defined as
+\paragraph{Nyström estimator.} Given a good low-rank approximation of a matrix, its trace can be cheaply estimated by simply taking the trace of that low-rank approximation.
+In the context of  spectral density estimation, the use of a non-negative smoothing kernel implies that $\mtx{B}(t)$ is symmetric positive definite for every $t$.
+This makes the Nyström method
+from~\cite{gittens-2013-revisiting-nystrom} very well for constructing a (randomized) low-rank approximation.
+Given a Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$, Nyström computes the approximation
 \begin{equation}
-    \Nystr{\mtx{\Omega}}{\mtx{B}}(t) = (\mtx{B}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} (\mtx{B}(t) \mtx{\Omega})^{\top}.
+    \mtx{B}(t) \approx \Nystr{\mtx{\Omega}}{\mtx{B}}(t) := (\mtx{B}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} (\mtx{B}(t) \mtx{\Omega})^{\top}.
     \label{equ:nystrom-approximation}
 \end{equation}
-We can estimate the trace of $\mtx{B}(t)$ with $\Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t))$.
-Thanks to the invariance of the trace under cyclic permutation of its arguments and the symmetry of the matrix, we may rewrite this estimator as
+Related randomized low-rank approximations for general parameter-dependent matrices have been analyzed in~\cite{kressner-2024-randomized-lowrank}.
+Exploiting the invariance of the trace under cyclic permutations, taking the trace of~\eqref{equ:nystrom-approximation} gives rise to the Nyström estimator
 \begin{equation}
-    \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{B}(t)^2 \mtx{\Omega})).
+    \Trace(\mtx{B}(t)) \approx \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{B}(t)^2 \mtx{\Omega})).
     \label{equ:nystrom-trace-estimator}
@@
 
-Two different low-rank approximation algorithms for general parameter-dependent matrices have already been analyzed in \cite{kressner-2024-randomized-lowrank}.
-However, the downside of the low-rank approximation approach is that it will -- in general -- not work well enough on matrices whose singular values do not decay quickly.
 %Write:
@@
 
-\paragraph{Nyström++ estimator.} Finally, \cite{lin-2017-randomized-estimation} proposes an estimator which is more accurate than the Girard-Hutchinson estimator and, unlike the Nyström approximation, does not rely on a fast singular value decay of $\mtx{B}(t)$.
-The estimator first computes the trace of the Nyström approximation \refequ{equ:nystrom-trace-estimator} of the matrix.
-Because this result may not always be satisfactory, the estimator then corrects for eventual inaccuracies in the Nyström approximation by estimating the trace of the approximation residual with the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}.
-The full estimator is
+\paragraph{Nyström++ estimator.}
+The Nyström estimator~\eqref{equ:nystrom-trace-estimator} relies on the feasibility of low-rank approximation, that is, the singular values of $\mtx{B}(t)$ need to decay sufficiently quickly.
+If this property cannot be ensured, it is preferable to use an estimator that combines the Girard-Hutchinson and Nyström estimators.
+The combination suggested in~\cite{lin-2017-randomized-estimation} applies Girard-Hutchinson to the error of the Nyström approximation~\eqref{equ:nystrom-approximation} in order to
+reduce the error of the Nyström estimator:
 \begin{equation}
@@
 \end{equation}
-where the matrix indices again emphasize the use of a constant standard Gaussian random matrix $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$ for the Girard-Hutchinson estimator and the dimension reduction matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ for the Nyström approximation.
+where $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$, $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ are independent Gaussian random matrices. 
+\textcolor{red}{??? What about XTrace? Does it apply; should we mention it somewhere? ???}
 
-This estimator is the parameter-dependent analogue of the Nyström++ estimator \cite{persson-2022-improved-variants}, which itself is based on the Hutch++ estimator \cite{meyer-2021-hutch-optimal}.
-For constant matrices $\mtx{B}$, these estimators were both shown to achieve a relative $\varepsilon$-error with just using $\mathcal{O}(\varepsilon^{-1})$ matrix-vector products, independent of the singular value decay of $\mtx{B}$.
+For a fixed value of $t$, the estimator~\eqref{equ:nystrompp-trace-estimator} coincides with the Nyström++ estimator from~\cite{persson-2022-improved-variants}, which is based on the Hutch++ estimator~\cite{meyer-2021-hutch-optimal}.
+In this situation (constant $\mtx{B}$), these estimators were both shown to achieve a relative $\varepsilon$-error with just using $\mathcal{O}(\varepsilon^{-1})$ matrix-vector products, independent of the singular value decay of $\mtx{B}$.
 %We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.
@@
 
+%as we will see in Section ???.
+
 \subsection{Error bounds}
 
-\textcolor{green}{In the following pages, we will upper bound the $L^1$-error of each of the trace estimators from \refsec{subsec:methods} with a suitable quantity which parallels the corresponding result in the constant case.
-As in \cite{kressner-2024-randomized-lowrank}, this often requires us to bound the high order moments $\mathbb{E}^p[X] = (\mathbb{E}[|X|^p])^{\sfrac{1}{p}}$ of certain random variables $X$.}
+In this section, we derive bounds on the $L^1$-error (in $t$) for each of the trace estimators from \refsec{subsec:methods}.
+Our bounds parallel existing results in the constant case.
+As in~\cite{kressner-2024-randomized-lowrank}, our proofs incorporate parameter dependence by proceeding through moments $\mathbb{E}^p[X] = (\mathbb{E}[|X|^p])^{\sfrac{1}{p}}$ of certain random variables $X$.
 
@@
 
-\color{blue}
-
-We start our analysis of the methods presented above by deriving an $L^1$-error bound for the Girard-Hutchinson estimator~\refequ{equ:hutchinson-trace-estimator} that is relative to the $L^1$-norm of the Frobenius norm $\lVert \mtx{B}(t) \rVert _F$.
+We first derive an $L^1$-error bound for the Girard-Hutchinson estimator~\refequ{equ:hutchinson-trace-estimator} that is relative to the $L^1$-norms of the Frobenius and spectral norms of $\mtx{B}(t)$, defined by
+\[
+ \|\mtx{B}(t)\|_{F,1} := \int_{a}^{b} \|\mtx{B}(t)\|_{F}~\mathrm{d}t, \quad  \|\mtx{B}(t)\|_{2,1} := \int_{a}^{b} \|\mtx{B}(t)\|_{2}~\mathrm{d}t.
+\]
 
@@
     For a symmetric matrix $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ that depends continuously on $t \in [a, b]$, consider the  Girard-Hutchinson estimator~\refequ{equ:hutchinson-trace-estimator} with an $n\times n_{\mtx{\Psi}}$ Gaussian random matrix $\mtx{\Psi}$.
-Then for any $p \geq 1, p \in \mathbb{R}$ and $\gamma \geq 1$ the bound
+Then for any $p \geq 1$, $p \in \mathbb{R}$ and $\gamma \geq 1$, the following bound holds with probability at least $1 - \gamma^{-p}$:
     \begin{equation} \label{eq:ghbound}
-        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq 4 \gamma \int_{a}^{b} \sqrt{\frac{p}{n_{\mtx{\Psi}}}}  \lVert \mtx{B}(t) \rVert _F + \frac{2 p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 ~\mathrm{d}t
+        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq 4 \gamma \Big( \sqrt{\frac{p}{n_{\mtx{\Psi}}}}  \|\mtx{B}(t)\|_{F,1} + \frac{2 p}{n_{\mtx{\Psi}}} \|\mtx{B}(t)\|_{2,1} \Big).
     \end{equation}
-    holds with probability at least $1 - \gamma^{-p}$.
-    In particular, given $\varepsilon \in (0, 1)$ and $\delta \in (0, e^{-1})$, the bound $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$ holds
+In particular, given $\varepsilon \in (0, 1)$ and $\delta \in (0, e^{-1}]$, the bound $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$ holds
     with probability at least $1-\delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1}))$.
 \end{theorem}
-
-\color{black}
 \begin{proof}
-    We first consider a fixed value for $t$.
+    Let us first consider $t$ fixed.
 From the proof of \cite[Theorem 1]{cortinovis-2022-randomized-trace}, we know that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is a $(2 \lVert \mtx{B}(t) \rVert _F^2 / n_{\mtx{\Psi}}, 2 \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}})$-sub-gamma random variable.
-Therefore, by \reflem{lem:sub-gamma-moments}
+Therefore, \reflem{lem:sub-gamma-moments} yields
     \begin{equation}
@@
         &\leq \int_{a}^{b} \mathbb{E}^{p}\left[ \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) \right]~\mathrm{d}t \notag \\
-        &\leq 4 \int_{a}^{b}  \sqrt{\frac{p}{n_{\mtx{\Psi}}}}  \lVert \mtx{B}(t) \rVert _F + \frac{2 p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 ~\mathrm{d}t.
+        &\leq 4 \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert_{F,1} + 8 \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert_{2,1}.
     \end{align}
-    Consequently, by Markov's inequality, the bound 
-    \begin{equation} 
-        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq 4 \gamma \int_{a}^{b} \sqrt{\frac{p}{n_{\mtx{\Psi}}}}  \lVert \mtx{B}(t) \rVert _F + \frac{2 p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 ~\mathrm{d}t.
-    \end{equation}
-    holds for any $\gamma \geq 1$ with probability at least $1 - \gamma^{-p}$.
+    This implies~\eqref{eq:ghbound} by Markov's inequality.
 
-    Bounding $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$, setting $\gamma = e$, and choosing $p = \log(\delta^{-1}) \geq 1$, the second part of the theorem follows for example when taking $n_{\mtx{\Psi}} = \lceil (8 e)^2 \log(\delta^{-1}) \varepsilon^{-2} \rceil$.
+    Bounding $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$, setting $\gamma = e$, and choosing $p = \log(\delta^{-1}) \geq 1$, the second part of the theorem follows by taking,  e.g., $n_{\mtx{\Psi}} = \lceil (8 e)^2 \log(\delta^{-1}) \varepsilon^{-2} \rceil$.
 \end{proof}
@@
 
-\refthm{thm:hutchinson} coincides with the analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, except for a slightly more restricted choice of the failure probability $\delta$.
+\refthm{thm:hutchinson} parallels an analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, except for a slightly more restricted choice of the failure probability $\delta$.
 
 \begin{remark}
-    A slightly different version of \refthm{thm:hutchinson} can be derived when $\mtx{B}(t)$ is nonzero and symmetric positive semi-definite for all $t \in [a, b]$: after dividing both sides of \refequ{equ:moment-bound-hutchinson} by $\Trace(\mtx{B}(t))$, one can apply the remaining steps in the proof to deduce with probability at least $1 - \delta$
-    \begin{equation}
-        \int_{a}^{b} \frac{| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) |}{\Trace(\mtx{B}(t))} ~\mathrm{d}t < \varepsilon,
-    \end{equation}
-    if $n_{\mtx{\Psi}} = \mathcal{O}(\log(\delta^{-1}) \varepsilon^{-2} \rho^2)$, where we identify
-    \begin{equation}
-        \rho = \int_{a}^{b} \frac{\lVert \mtx{B}(t) \rVert _F}{\Trace(\mtx{B}(t))} ~\mathrm{d}t
-    \end{equation}
-    The quantity $\rho$ is small when on the interval $[a, b]$ the singular values of $\mtx{B}(t)$ decay slowly and large when they decay quickly.
-This parallels a similar result for constant matrices in~\cite[Remark 2]{cortinovis-2022-randomized-trace}.
+    A variant of \refthm{thm:hutchinson} can be derived when assuming that $\mtx{B}(t)$ is symmetric positive semi-definite and nonzero for every $t \in [a, b]$.
+After dividing both sides of \refequ{equ:moment-bound-hutchinson} by $\Trace(\mtx{B}(t))$, one can apply the remaining steps of the proof to deduce that
+    \begin{equation*}
+        \int_{a}^{b} \frac{| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) |}{\Trace(\mtx{B}(t))} ~\mathrm{d}t < \varepsilon
+    \end{equation*}
+    holds with probability at least $1 - \delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\log(\delta^{-1}) \varepsilon^{-2} \rho^2)$. Here, 
+    the quantity $
+        \rho := \frac{1}{b-a} \int_{a}^{b} \frac{\lVert \mtx{B}(t) \rVert _F}{\Trace(\mtx{B}(t))} ~\mathrm{d}t \in [1,n]
+    $
+is small when the singular values of $\mtx{B}(t)$ decay slowly for the majority of $t\in [a,b]$ and large otherwise.
+This parallels an analogous result for constant matrices; see~\cite[Remark 2]{cortinovis-2022-randomized-trace}.
 \end{remark}
-
-\color{blue}
 
@@
 
-As a consequence of~\cite{kressner-2024-randomized-lowrank}, the following result confirms the inituition that the Nystr\"om approximation is very well suited in the presence of decaying singular values: If the nuclear norm error to the best rank-$r$ approximation to $B(t)$ is bounded by $\varepsilon$ for every $t$ then the trace of the Nystr\"om approximation (with a bit of oversampling) results, with high probability, in an error that is only modestly larger than $\varepsilon$.
+A consequence of~\cite{kressner-2024-randomized-lowrank}, the following result confirms the inituition that the Nystr\"om estimator is very well suited when the singular values of $\mtx{B}(t)$ decay quickly.
+More specifically, it shows that the $L^1$-error of the estimator stays on the level of the $L^1$-norm of the best rank-$r$ approximation error of $B(t)$, measured in the nuclear norm.
 
@@
 For integers $r \geq 2$ and $n_{\mtx{\Omega}} \geq r + 4$, consider the Nystr\"om approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ defined in~\refequ{equ:nystrom-approximation} with an $n\times n_{\mtx{\Omega}}$ Gaussian random matrix $\mtx{\Omega}$.
-Then for any $\gamma \geq 1$ the bound
+Then, for any $\gamma \geq 1$, the bound
     \[
@@
 
-Compared to the constant case~\cite[Theorem 8.1]{tropp-2023-randomized-algorithms}, the bound of \refthm{thm:nystrom} features an additional factor $1+r$ that cannot be compensated with oversampling.
+Compared to the constant case~\cite[Theorem 8.1]{tropp-2023-randomized-algorithms}, the bound of \refthm{thm:nystrom} features an additional factor $1+r$, which cannot be compensated with oversampling.
 
@@
 
-\refthm{thm:hutchinson} shows that the the Girard-Hutchinson estimator achieves a relative $\varepsilon$-error when using $\mathcal{O}(\varepsilon^{-2})$ queries.
+\refthm{thm:hutchinson} shows that the Girard-Hutchinson estimator achieves a relative $\varepsilon$-error when using $\mathcal{O}(\varepsilon^{-2})$ queries.
 The aim of this section is to show that the Nyström++ estimator~\refequ{equ:nystrompp-trace-estimator} improves this to $\mathcal{O}(\varepsilon^{-1})$ queries (that is, matrix-vector products with random vectors), \emph{without} requiring any assumption on the singular values of $\mtx{B}(t)$.
-This parallels existing results for the constant case for the Hutch++~\cite[Theorem 3.1]{meyer-2021-hutch-optimal} and Nyström++ estimators \cite[Theorem 3.4]{persson-2022-improved-variants}.
-\textcolor{red}{??
-Include specific result ???} \textcolor{green}{Specifically, a bound of the form $|\Trace(\mtx{B}) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})| \leq \varepsilon \Trace(\mtx{B})$ for any constant symmetric positive semi-definite matrix $\mtx{B}$ is shown to hold with high probability for the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})$ \refequ{equ:nystrompp-trace-estimator} when $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1})$ }.
-Indeed, our proof follows the strategy from~\cite{meyer-2021-hutch-optimal}, by first showing that the parameter-dependent Nyström approximation achieves a relative $\varepsilon$-error when using $\mathcal{O}(\varepsilon^{-2})$ queries.
+This parallels existing results for the constant case~\cite{meyer-2021-hutch-optimal,persson-2022-improved-variants}.
+In particular, Theorem 3.4 from~\cite{persson-2022-improved-variants} shows that $|\Trace(\mtx{B}) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})| \leq \varepsilon \Trace(\mtx{B})$ holds with high probability for any (constant) SPSD matrix $\mtx{B}$ for the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})$ from~\refequ{equ:nystrompp-trace-estimator} when $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1})$.
 
-\color{blue}
+Our analysis mimicks the strategy from~\cite{meyer-2021-hutch-optimal}, by first establishing a relative Frobenius norm error proportional to $n_{\mtx{\Omega}}^{-1/2}$ for the parameter-dependent Nyström approximation.
 
 \begin{lemma}{Nyström approximation for parameter-dependent matrices}{nystrom}
-    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
+    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be SPSD and continuous for $t \in [a, b]$.
 Consider the Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ with a Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and \emph{even} $n_{\mtx{\Omega}} \geq 4$.
-    Then for any $\gamma \geq 1$ the bound
+    Then, for any $\gamma \geq 1$, the bound
     \begin{equation}
@@
     \end{equation}
-        holds with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$ for 
+        holds with probability at least $1 - \gamma^{-n_{\mtx{\Omega}} / 4}$ for 
         $c = 1 + 9 \sqrt{2} e^4$.
@@
     $\mtx{\Omega}_1 := \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times n_{\mtx{\Omega}}}$ and $\mtx{\Omega}_2 := \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times n_{\mtx{\Omega}}}$, which are independent Gaussian random matrices.
-Applying Theorem B.1 from~\cite{persson-2023-randomized-lowrank} for $f(x) = x$ yields the bound
+Applying Theorem B.1 from~\cite{persson-2023-randomized-lowrank} for $f(x) = x$, see also \textcolor{red}{???
+Also cite the same result from Tropp and Webber or Epperly and Tropp?
+???}, yields the bound
     \begin{equation}
@@
     \end{equation}
+    ??? UNTIL HERE ???
     \textcolor{red}{???

### 2025-02-17 paper/intro.tex
review comments

--- 2024-10-01 paper/intro.tex
@@
 
-Spectral distributions can reveal important properties in numerous applications across physics, chemistry, engineering, and data science: For example, in electronic structure calculations eigenvalues represent the energy levels which electrons occupy\cite{drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure, lin-2017-randomized-estimation}, in machine learning they are indicative of the topology of the loss landscape \cite{ghorbani-2019-investigation-neural, yao-2020-pyhessian-neural}, and in graph processing they can uncover hidden graph motifs~\cite{huang-2021-density-states}.
+Spectral distributions of matrices and linear operators reveal important properties in numerous applications across physics, chemistry, engineering, and data science: For example, in electronic structure calculations, eigenvalues represent the energy levels that electrons occupy~\cite{drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure, lin-2017-randomized-estimation}, in machine learning they are indicative of the topology of the loss landscape~\cite{ghorbani-2019-investigation-neural, yao-2020-pyhessian-neural}, and in graph processing they can uncover hidden graph motifs~\cite{huang-2021-density-states}.
 Given a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$, the distribution of its eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$, which places a rescaled Dirac delta distribution $\delta$ at each eigenvalue.
@@
 For a \emph{constant} matrix $\mtx{B}$, one of the most popular trace estimators is the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} along with variance reduction techniques~\cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, chen-2023-krylovaware-stochastic, epperly-2024-xtrace-making}.
-Suitable extensions to parameter-dependent matrices have been considered, e.g., in~\cite{lin-2017-randomized-estimation,chen-2023-krylovaware-stochastic}, but we are not aware of an analysis providing insight and justification of these extensions.
-In passing, we note that dynamic trace estimation~\cite{dharangutte-2024-dynamic-trace,woodruff-2024-optimal-query} is an efficient technique for subsequently estimating the traces of matrices $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ for wich the increments $\mtx{B}(t_{i+1}) - \mtx{B}(t_i)$ are relatively small in norm.
+Suitable extensions to parameter-dependent matrices have been considered, e.g., in~\cite{lin-2017-randomized-estimation,chen-2023-krylovaware-stochastic}, but we are not aware of an analysis providing rigorous justification and insight of these extensions.
+In passing, we note that dynamic trace estimation~\cite{dharangutte-2024-dynamic-trace,woodruff-2024-optimal-query} is an efficient technique for subsequently estimating the traces of matrices $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ when the increments $\mtx{B}(t_{i+1}) - \mtx{B}(t_i)$ are relatively small in norm.
 The potential of dynamic trace estimation appears to be limited in our setting because $\mtx{B}(t)$ may change rapidly close to eigenvalues, with $g_{\sigma}$ approximating a Dirac delta function.

### 2025-02-17 paper/analysis.tex
adapt nomenclature and fix README

@@
 \paragraph{Nyström estimator.} Given a good low-rank approximation of a matrix, its trace can be cheaply estimated by simply taking the trace of that low-rank approximation.
-In the context of  spectral density estimation, the use of a non-negative smoothing kernel implies that $\mtx{B}(t)$ is symmetric positive definite for every $t$.
+In the context of  spectral density estimation, the use of a non-negative smoothing kernel implies that $\mtx{B}(t)$ is symmetric positive semi-definite (SPSD) for every $t$.
 This makes the Nyström method
@@
 \begin{remark}
-    A variant of \refthm{thm:hutchinson} can be derived when assuming that $\mtx{B}(t)$ is symmetric positive semi-definite and nonzero for every $t \in [a, b]$.
+    A variant of \refthm{thm:hutchinson} can be derived when assuming that $\mtx{B}(t)$ isSPSD and nonzero for every $t \in [a, b]$.
 After dividing both sides of \refequ{equ:moment-bound-hutchinson} by $\Trace(\mtx{B}(t))$, one can apply the remaining steps of the proof to deduce that
@@
 \begin{theorem}{Nyström estimator for parameter-dependent matrices}{nystrom}
-    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuous for $t \in [a, b]$.
+    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be SPSD and continuous for $t \in [a, b]$.
 For integers $r \geq 2$ and $n_{\mtx{\Omega}} \geq r + 4$, consider the Nystr\"om approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ defined in~\refequ{equ:nystrom-approximation} with an $n\times n_{\mtx{\Omega}}$ Gaussian random matrix $\mtx{\Omega}$.
@@
 \begin{proof}
-Using that $\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ is positive semi-definite (see, e.g.,~\cite[Lemma 2.1]{frangella-2023-randomized-nystrom}) we obtain
+Using that $\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ is SPSD (see, e.g.,~\cite[Lemma 2.1]{frangella-2023-randomized-nystrom}) we obtain
     \begin{align*}
@@
 \begin{theorem}{Nyström++ trace estimator for parameter-dependent matrices}{nystrom-pp}
-    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$.
+    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be SPSD and continuously depend on $t \in [a, b]$.
 If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(\delta^{-1}))$ with \emph{even} $n_{\mtx{\Omega}} \geq 4$, then for $\delta \in (0, 2e^{-1})$ and $\varepsilon \in (0, 1)$ with probability at least $1 - \delta$

### 2025-02-17 paper/intro.tex
adapt nomenclature and fix README

@@
 All methods considered in this work are based on the following simple idea: Apply an existing randomized trace estimator to $\Trace(\mtx{B}(t))$ with \emph{constant} random vectors, that is, the same randomization is used for each value of the parameter $t$.
-\textcolor{green}{For example, the Girard-Hutchinson estimator becomes $n_{\mtx{\Psi}}^{-1} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j$ for $n_{\mtx{\Psi}}$ constant standard Gaussian random vectors $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}}$.}
+\textcolor{green}{For example, the Girard-Hutchinson estimator becomes $n_{\mtx{\Psi}}^{-1} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j$ for $n_{\mtx{\Psi}}$ constant Gaussian random vectors $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}}$.}
 ??? STOP HERE ???

### 2025-02-17 paper/numresults.tex
adapt nomenclature and fix README

@@
 The Krylov-aware stochastic trace estimation \cite{chen-2023-krylovaware-stochastic} is also well suited for approximating the trace of a matrix function $f(\mtx{A}, t)$ which separately depends on a parameter $t$.
-It also samples two standard Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$.
+It also samples two Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$.
 $\mtx{\Omega}$ is used to extract a low-rank approximation to $f(\mtx{A}, t)$ from a block-Krylov subspace of $\mtx{A}$ generated from $r$ iterations of the block-Lanczos method with reorthogonalization and subsequently $q$ without.

### 2025-02-17 paper/spectraldensity.tex
push changes from Overleaf
adapt nomenclature and fix README

@@
 \end{equation}
-The task in \refequ{equ:smooth-spectral-density} is to approximate the trace of the parameter-dependent symmetric positive semi-definite matrix $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
+The task in \refequ{equ:smooth-spectral-density} is to approximate the trace of the parameter-dependent SPSD matrix $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
 Hence, we can apply the results derived in \refsec{sec:analysis}.
@@
 It loosely follows the proof given in the arXiv preprint of \cite[Theorem 2]{lin-2017-randomized-estimation}, but attains a slightly different dependence on $\sigma$.
+\\ \vspace{\parskip}
 
@@
     %\State Compute  for all $t_i$ using %\refalg{alg:3-nystrom-chebyshev-exponentiation}
-    \State Generate standard Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$%\glsfirst{sketching-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$
+    \State Generate Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$%\glsfirst{sketching-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$
     %\State Generate standard Gaussian %\glsfirst{random-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$

### 2025-02-17 paper/analysis.tex
fix small typos

@@
     \end{equation}
-In particular, given $\varepsilon \in (0, 1)$ and $\delta \in (0, e^{-1}]$, the bound $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$ holds
+In particular, given $\varepsilon \in (0, 1)$ and $\delta \in (0, e^{-1}]$, the bound $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \|\mtx{B}(t)\|_{F,1}$ holds
     with probability at least $1-\delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1}))$.
@@
 \begin{remark}
-    A variant of \refthm{thm:hutchinson} can be derived when assuming that $\mtx{B}(t)$ isSPSD and nonzero for every $t \in [a, b]$.
+    A variant of \refthm{thm:hutchinson} can be derived when assuming that $\mtx{B}(t)$ is SPSD and nonzero for every $t \in [a, b]$.
 After dividing both sides of \refequ{equ:moment-bound-hutchinson} by $\Trace(\mtx{B}(t))$, one can apply the remaining steps of the proof to deduce that
@@
     \begin{equation}
-        \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}(t)} \rVert _F 
+        \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F 
         \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2,

### 2025-02-18 paper/analysis.tex
review of section 2 finished

@@
 trace $\Trace(\mtx{B}(t)) = b_{11}(t)+ \cdots + b_{nn}(t)$
-from matrix-vector products of $B(t)$ with \emph{constant} random vectors.
-For one \emph{fixed} value of $t$, our methods reduce to known algorithms.
-Once one considers several, potentially many values of $t$, important differences arise from the fact that the random vectors do not depend on $t$.
+from matrix-vector products of $\mtx{B}(t)$ with \emph{constant} random vectors.
+When $\mtx{B}$ is constant, our methods reduce to known algorithms.
+When $\mtx{B}$ is variable, important differences arise from the fact that the random vectors do not depend on $t$.
 On the one hand, as we will see in~\refsec{sec:application}, this allows one to reuse computations.
@@
 
-\paragraph{Girard-Hutchinson estimator.} One of the simplest methods for trace estimation, the Girard-Hutchinson estimator~\cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator} proceeds by sampling $n_{\mtx{\Psi}}$ stochastically independent Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$
+\paragraph{Girard-Hutchinson estimator.} One of the simplest methods for trace estimation, the Girard-Hutchinson estimator~\cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator} proceeds by sampling $n_{\mtx{\Psi}}$ independent Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$
 and computes the approximation 
@@
 \end{equation}
-Here, $\mtx{\Psi} = [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\mtx{\Psi}}}] \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$ is a Gaussian random matrix, that is, 
+Here, $\mtx{\Psi} := [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\mtx{\Psi}}}] \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$ is a Gaussian random matrix, that is, 
 its entries are independent standard normal random variables.
-There are other popular choices for the probability distribution, such as the Rademacher distribution. Our theoretical developments, however, only apply to the Gaussian case.
+There are other sensible choices for the probability distribution, such as the Rademacher distribution. Our theoretical developments, however, only apply to the Gaussian case.
 
@@
 
-\paragraph{Nyström estimator.} Given a good low-rank approximation of a matrix, its trace can be cheaply estimated by simply taking the trace of that low-rank approximation.
+\paragraph{Nyström estimator.} Given a good low-rank approximation of a matrix, the trace can be cheaply estimated by simply taking the trace of that low-rank approximation.
 In the context of  spectral density estimation, the use of a non-negative smoothing kernel implies that $\mtx{B}(t)$ is symmetric positive semi-definite (SPSD) for every $t$.
 This makes the Nyström method
-from~\cite{gittens-2013-revisiting-nystrom} very well for constructing a (randomized) low-rank approximation.
+from~\cite{gittens-2013-revisiting-nystrom} well suited for constructing a (randomized) low-rank approximation.
 Given a Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$, Nyström computes the approximation
@@
 If this property cannot be ensured, it is preferable to use an estimator that combines the Girard-Hutchinson and Nyström estimators.
-The combination suggested in~\cite{lin-2017-randomized-estimation} applies Girard-Hutchinson to the error of the Nyström approximation~\eqref{equ:nystrom-approximation} in order to
-reduce the error of the Nyström estimator:
+The combination suggested in~\cite{lin-2017-randomized-estimation} applies Girard-Hutchinson to correct the error of the Nyström approximation~\eqref{equ:nystrom-approximation}:
 \begin{equation}
@@
 
-For a fixed value of $t$, the estimator~\eqref{equ:nystrompp-trace-estimator} coincides with the Nyström++ estimator from~\cite{persson-2022-improved-variants}, which is based on the Hutch++ estimator~\cite{meyer-2021-hutch-optimal}.
-In this situation (constant $\mtx{B}$), these estimators were both shown to achieve a relative $\varepsilon$-error with just using $\mathcal{O}(\varepsilon^{-1})$ matrix-vector products, independent of the singular value decay of $\mtx{B}$.
+For constant $\mtx{\Omega}$, the estimator~\eqref{equ:nystrompp-trace-estimator} coincides with the Nyström++ estimator from~\cite{persson-2022-improved-variants}, which is based on the Hutch++ estimator~\cite{meyer-2021-hutch-optimal}.
+In this situation (constant $\mtx{B}$), these estimators were both shown to achieve a relative $\varepsilon$-error with $\mathcal{O}(\varepsilon^{-1})$ matrix-vector products only, independent of the singular value decay of $\mtx{B}$.
 %We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.
@@
 
-In this section, we derive bounds on the $L^1$-error (in $t$) for each of the trace estimators from \refsec{subsec:methods}.
+In this section, we derive bounds on the $L^1$-error (with respect to $t$) for each of the trace estimators from \refsec{subsec:methods}.
 Our bounds parallel existing results in the constant case.
@@
 A consequence of~\cite{kressner-2024-randomized-lowrank}, the following result confirms the inituition that the Nystr\"om estimator is very well suited when the singular values of $\mtx{B}(t)$ decay quickly.
-More specifically, it shows that the $L^1$-error of the estimator stays on the level of the $L^1$-norm of the best rank-$r$ approximation error of $B(t)$, measured in the nuclear norm.
+More specifically, it shows that the $L^1$-error of the estimator stays on the level of the best rank-$r$ approximation error of $B(t)$, measured in the nuclear norm.
 
@@
 This parallels existing results for the constant case~\cite{meyer-2021-hutch-optimal,persson-2022-improved-variants}.
-In particular, Theorem 3.4 from~\cite{persson-2022-improved-variants} shows that $|\Trace(\mtx{B}) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})| \leq \varepsilon \Trace(\mtx{B})$ holds with high probability for any (constant) SPSD matrix $\mtx{B}$ for the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})$ from~\refequ{equ:nystrompp-trace-estimator} when $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1})$.
+In particular, Theorem 3.4 from~\cite{persson-2022-improved-variants} shows that $|\Trace(\mtx{B}) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})| \leq \varepsilon \Trace(\mtx{B})$ holds with high probability for (constant) SPSD $\mtx{B}$ when $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1})$ in the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})$ from~\refequ{equ:nystrompp-trace-estimator}.
 
@@
         holds with probability at least $1 - \gamma^{-n_{\mtx{\Omega}} / 4}$ for 
-        $c = 1 + 9 \sqrt{2} e^4$.
+        $c = 108$.
 In particular, given $\varepsilon > 0$ and $\delta \in (0, 1)$, the bound $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$ holds with probability at least $1-\delta$ if $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1}))$.
@@
 For fixed $t$, consider the spectral decomposition $\mtx{B}(t) = \mtx{U} \mtx{\Lambda} \mtx{U}^{\top}$ with  
-$\mtx{\Lambda} = \operatorname{diag}(\lambda_1, \dots, \lambda_n)$ and $\lambda_1 \ge \dots \ge \lambda_n \ge 0$. Partitioning
+$\mtx{\Lambda} = \operatorname{diag}(\lambda_1, \dots, \lambda_n)$ and $\lambda_1 \ge \dots \ge \lambda_n \ge 0$. For $k < n_{\mtx{\Omega}}$ (chosen arbitrarily at the moment), partition
     \begin{equation*}
@@
     \end{equation}
-    where $\lVert \cdot \rVert _{(4)}$ denotes the Schatten-4 norm. Standard arguments (see, e.g., the proof of~\cite[Lemma 3]{meyer-2021-hutch-optimal}) provide a simple bound on the first term:
+    where $\lVert \cdot \rVert _{(4)}$ denotes the Schatten-4 norm. Following the proof of~\cite[Lemma 3]{meyer-2021-hutch-optimal}, we replace the the first term by a simpler bound:
     \begin{equation}
@@
     \end{equation}
-    ??? UNTIL HERE ???
-    \textcolor{red}{???
-STOPPED HERE: Not clear to me why $n_\Omega/4$ should be the optimal exponent for the tail bound ???} \textcolor{green}{I rearranged and reformulated things a bit to hopefully make it clearer:}
 
-    \color{black}
-    The $q$-th moment of the second term in \refequ{equ:nystrom-proof-persson-bonud} can be processed with standard matrix norm manipulations and the stochastic independence of $\mtx{\Omega}_1$ and $\mtx{\Omega}_2$ to
+To treat the second term in~\eqref{equ:nystrom-proof-persson-bonud}, we bound 
+its $q$th moment using the independence of $\mtx{\Omega}_1$ and $\mtx{\Omega}_2$:
     \begin{equation}
-        \mathbb{E}^{q}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
+        \mathbb{E}^{q}\left[ \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \big\|_{(4)}^2 \right]
         %&= \mathbb{E}^{p}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )( \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger})^{\top} \rVert _F \right] && \text{($\lVert \mtx{A} \rVert _{(4)}^2 = \lVert \mtx{A} \mtx{A}^{\top} \rVert _{F}$)} \notag \\
         %&\leq \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] && \text{($\lVert \mtx{A} \mtx{B} \rVert _F \leq \lVert \mtx{A} \rVert _F \lVert \mtx{B} \rVert _2$)} \notag \\
-        \leq \mathbb{E}^{q}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]%  && \text{($\lVert \mtx{A} \rVert _{(4)}^2 \leq \sqrt{k} \lVert \mtx{A} \rVert _2^2$)} \notag \\
+        \leq \mathbb{E}^{q}\left[ \sqrt{k} \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \big\| _2^2 \right]%  && \text{($\lVert \mtx{A} \rVert _{(4)}^2 \leq \sqrt{k} \lVert \mtx{A} \rVert _2^2$)} \notag \\
         %\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]%  && \text{(submultiplicativity)} \notag \\
-        \leq \sqrt{k} \mathbb{E}^{q}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2\right] \mathbb{E}^{q}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2  \right].% && \text{(independence)}
+        \leq \sqrt{k} \mathbb{E}^{q}\left[ \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2  \big\|_2^2\right] \mathbb{E}^{q}\left[ \big\| \mtx{\Omega}_1^{\dagger} \big\|_2^2  \right].% && \text{(independence)}
         \label{equ:nystrom-proof-processed-tail}
     \end{equation}
-    To match the decay rate of the moments of the first term in \refequ{equ:nystrom-proof-persson-bonud}, we will need to ensure that this term also is of order $\mathcal{O}(\Trace(\mtx{B}) / \sqrt{k})$.
+    To continue from here, we set $k = n_{\mtx{\Omega}}/2$ and $q = n_{\mtx{\Omega}}/4$. By the calculations after~\cite[Eqn~(B.7)]{tropp-2023-randomized-algorithms}, one has
+    \textcolor{red}{??? Fabio, the red terms are different from your calculations. Please double check that I am correct ???}
+    \[
+     \mathbb{E}\left[ \big\| ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \big\|_2^{\sfrac{n_{\mtx{\Omega}}}{4}} \right]%  \notag \\
+        \leq
+        \textcolor{red}{\frac{3}{2}}
+        \left( \frac{3}{4} n_{\mtx{\Omega}}\right)^{\sfrac{n_{\mtx{\Omega}}}{4}}
+        \big( ( n_{\mtx{\Omega}} / 2 + 1)!\big)^{-\frac{n_{\mtx{\Omega}}}{2+n_{\mtx{\Omega}}}}.
+    \]
+    Therefore,
+    \begin{equation} \label{equ:pinv-spectral-norm-bound}
+      \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \big\| \mtx{\Omega}_1^{\dagger} \big\|_2^2 \right]
+        = \mathbb{E}\left[ \big\| ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \big\|_2^{\sfrac{n_{\mtx{\Omega}}}{4}} \right]^{\sfrac{4}{n_{\mtx{\Omega}}}}
+        \le \frac{9}{8}  
+        \frac{e^2 n_{\mtx{\Omega}}}{(n_{\mtx{\Omega}} / 2+ 1)^2} \le 
+        \frac{9}{2}  \frac{e^2}{n_{\mtx{\Omega}}},
+    \end{equation}
+    where we used $(m!)^{-\sfrac{1}{m}} \leq e/m$. Note that, in contrast 
+    to the result of~\cite[Lemma B.3]{tropp-2023-randomized-algorithms}, this inequality is valid for arbitrarily large $n_{\mtx{\Omega}}$, at the expense of a slightly larger constant.
+%
+%     To match the decay rate of the moments of the first term in \refequ{equ:nystrom-proof-persson-bonud}, we will need to ensure that this term also is of order $\mathcal{O}(\Trace(\mtx{B}) / \sqrt{k})$.
 We bound the moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ similarly to \cite[Lemma B.3]{tropp-2023-randomized-algorithms}, but without restricting $q$ to be smaller than $18$.
@@
 Therefore, we get
-    \begin{equation}
-        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
-        = \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{n_{\mtx{\Omega}}}{4}} \right]^{\sfrac{4}{n_{\mtx{\Omega}}}}%  \notag \\
-        \leq \left( 1 + \frac{n_{\mtx{\Omega}}}{2} \right)^{\sfrac{4}{n_{\mtx{\Omega}}}} \left( \frac{1}{(\frac{n_{\mtx{\Omega}}}{2} + 1)!}\right)^{\frac{2}{\frac{n_{\mtx{\Omega}}}{2} + 1}} \left( \frac{3 n_{\mtx{\Omega}}}{4}\right).
-    \end{equation}
-    From the expansion of the exponential function as a power series $e^n = 1 + n + \dots + n^n/n!
+%     \begin{equation}
+%         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
+%         = \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{n_{\mtx{\Omega}}}{4}} \right]^{\sfrac{4}{n_{\mtx{\Omega}}}}%  \notag \\
+%         \leq \left( 1 + \frac{n_{\mtx{\Omega}}}{2} \right)^{\sfrac{4}{n_{\mtx{\Omega}}}} \left( \frac{1}{(\frac{n_{\mtx{\Omega}}}{2} + 1)!}\right)^{\frac{2}{\frac{n_{\mtx{\Omega}}}{2} + 1}} \left( \frac{3 n_{\mtx{\Omega}}}{4}\right).
+%     \end{equation}
+%     From the expansion of the exponential function as a power series $e^n = 1 + n + \dots + n^n/n!
 + \dots$ for $n \in \mathbb{N}$ we see $e^n \geq 1 + n$ and $e^n \geq n^n / n!$, from which $(1 + n)^{\sfrac{1}{n}} \leq e$ and $\left( 1/n!
@@
 Hence,
-    \begin{equation}
-        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
-        \leq e^2 \frac{e^2}{(\frac{n_{\mtx{\Omega}}}{2}+ 1)^2} \left( \frac{3 n_{\mtx{\Omega}}}{4}\right)
-        \leq \frac{3 e^4}{n_{\mtx{\Omega}}}.
-        \label{equ:pinv-spectral-norm-bound}
-    \end{equation}
-    As opposed to \cite[Lemma B.3]{tropp-2023-randomized-algorithms}, our bound holds for all even $n_{\mtx{\Omega}} \in \mathbb{N}$, at the cost of a factor of $e^2$.
-
-    The remaining term can dealt with using \reflem{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$ and $p = n_{\mtx{\Omega}}/2$
-    \begin{equation}
-        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \right]
-        \leq n_{\mtx{\Omega}} \left( 2 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{2}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
-        \label{equ:spectral-norm-bound-applied}
-    \end{equation}
-
+%     \begin{equation}
+%         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
+%         \leq e^2 \frac{e^2}{(\frac{n_{\mtx{\Omega}}}{2}+ 1)^2} \left( \frac{3 n_{\mtx{\Omega}}}{4}\right)
+%         \leq \frac{3 e^4}{n_{\mtx{\Omega}}}.
+%         \label{equ:pinv-spectral-norm-bound}
+%     \end{equation}
+%     As opposed to \cite[Lemma B.3]{tropp-2023-randomized-algorithms}, our bound holds for all even $n_{\mtx{\Omega}} \in \mathbb{N}$, at the cost of a factor of $e^2$.
+%
+    The second factor in~\eqref{equ:nystrom-proof-processed-tail} is bounded using \reflem{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$: % and $p = n_{\mtx{\Omega}}/2$:
+    \[
+        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \big\|_2^2 \right]
+        \leq \textcolor{red}{\frac{3}{2}} n_{\mtx{\Omega}} \Big( \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \big\|_2^2 + \frac{1}{n_{\mtx{\Omega}}} \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \big\|_F^2 \Big).
+      %  \label{equ:spectral-norm-bound-applied}
+    \]
     %The $q$-th moment of the second term can be processed with standard matrix norm manipulations and the stochastic independence of $\mtx{\Omega}_1$ and $\mtx{\Omega}_2$ to
@@
     %\end{equation}
-    Inserting \refequ{equ:spectral-norm-bound-applied} and  \refequ{equ:pinv-spectral-norm-bound} in \refequ{equ:nystrom-proof-processed-tail} with $k = n_{\mtx{\Omega}} / 2$ gives
+    Inserting this inequality and~\refequ{equ:pinv-spectral-norm-bound} into~\refequ{equ:nystrom-proof-processed-tail} gives
     \begin{equation}
         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \frac{3 e^4}{\sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \left( 2 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{2}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
+        \leq \frac{27 e^2}{4\sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \Big( \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \Big).
         %\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
     \end{equation}
-    We can identify $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 = \lambda_{k+1}$ and $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 = \Trace(\mtx{\Lambda}_2)$, and just as in the proof of \cite[Lemma 3.1]{meyer-2021-hutch-optimal} we bound $\lambda_{k+1} \leq \Trace(\mtx{B})/k$ and $\Trace(\mtx{\Lambda}_2) \leq \Trace(\mtx{B})$ -- remembering $k = n_{\mtx{\Omega}}/2$ -- to get
+    Bounding $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 = \lambda_{k+1}  \leq \Trace(\mtx{B})/k$ and $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 = \Trace(\mtx{\Lambda}_2) \le \Trace(\mtx{B})$ (recall that $k = n_{\mtx{\Omega}}/2$) yields
     \begin{equation}
         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \frac{3 e^4}{\sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \left( \frac{4}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) + \frac{2}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) \right)
-        \leq 9 \sqrt{2} e^4 \frac{1}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
+        \leq \frac{27 e^2}{4 \sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \Big( \frac{2}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) + \frac{1}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) \Big)
+        \leq  \frac{106}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
         \label{equ:nystrom-proof-tail-bound}
@@
     \end{equation}
-    Inserting \refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud} and using the triangle inequality for $\mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}[\cdot]$ we obtain
-    \begin{equation}
+
+    Inserting~\refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud} and using the triangle inequality for $\mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}[\cdot]$, we obtain
+    \begin{equation} \label{eq:blubber}
         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right]
@@
     \end{equation}
-    where $c = 1 + 9 \sqrt{2} e^4$.
+    As in \cite{kressner-2024-randomized-lowrank}, one can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable.
+    As in the proof of~\refthm{thm:hutchinson}, the claimed bound~\eqref{qu:nystrompp-theorem-bound}
+    follows from~\eqref{eq:blubber} using Minkowski's integral inequality and Markov's inequality.
+%     
+%     
+%         
+%     and by the Minkowski's integral inequality \cite[Theorem 202]{hardy-1952-inequalities}, we get for $n_{\mtx{\Omega}} \geq 4$
+%     \begin{align}
+%         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \right] 
+%         &\leq \int_{a}^{b} \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F \right] ~\mathrm{d}t \notag \\
+%         %&\leq \mathbb{E}^{\sfrac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
+%         %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \mathbb{E}^{\sfrac{k}{2}} \left[ \int_{a}^{b}  \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
+%         %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
+%         &\leq \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
+%     \end{align}
+%     From Markov's inequality follows with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
+%     \begin{equation}
+%         \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t,
+%     \end{equation}
+%     from which follows the first statement.
 
-    As in \cite{kressner-2024-randomized-lowrank}, we can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable and by the Minkowski's integral inequality \cite[Theorem 202]{hardy-1952-inequalities}, we get for $n_{\mtx{\Omega}} \geq 4$
-    \begin{align}
-        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \right] 
-        &\leq \int_{a}^{b} \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F \right] ~\mathrm{d}t \notag \\
-        %&\leq \mathbb{E}^{\sfrac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
-        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \mathbb{E}^{\sfrac{k}{2}} \left[ \int_{a}^{b}  \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
-        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
-        &\leq \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
-    \end{align}
-    From Markov's inequality follows with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
-    \begin{equation}
-        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t,
-    \end{equation}
-    from which follows the first statement.
-
-    Fixing $\gamma = e$ and letting $n_{\mtx{\Omega}} = \lceil(c e \varepsilon)^{-2}  + 4 \log(\delta^{-1}) \rceil$ gives us the second part of the theorem.
-
+    Fixing $\gamma = e$ and letting $n_{\mtx{\Omega}} = \lceil(c e \varepsilon)^{-2}  + 4 \log(\delta^{-1}) \rceil$ establishes the second part of the theorem.
     %Setting $\delta = \gamma^{-\sfrac{k}{4}}$ and choosing $k = \lceil \varepsilon^{-2}\log(\delta^{-1}) \rceil$ we get
@@
 
-\color{black}
+% When compared to the equivalent result for constant matrices \cite[Lemma 3.2]{persson-2022-improved-variants}, our constant factor up front is a bit more than two times larger and our failure probability decays half a power slower.
 
-When compared to the equivalent result for constant matrices \cite[Lemma 3.2]{persson-2022-improved-variants}, our constant factor up front is a bit more than two times larger and our failure probability decays half a power slower.
-
-Finally, we can now combine the bound on the Girard-Hutchinson estimator (\refthm{thm:hutchinson}) and the one on the Nyström approximation (\reflem{lem:nystrom}) to obtain our main result.
-
+Combining the bound on the Girard-Hutchinson estimator from~\refthm{thm:hutchinson} with the result of~\reflem{lem:nystrom}) finally gives the desired error bound for the Nyström++ trace estimator.
 \begin{theorem}{Nyström++ trace estimator for parameter-dependent matrices}{nystrom-pp}
-    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be SPSD and continuously depend on $t \in [a, b]$.
-If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(\delta^{-1}))$ with \emph{even} $n_{\mtx{\Omega}} \geq 4$, then for $\delta \in (0, 2e^{-1})$ and $\varepsilon \in (0, 1)$ with probability at least $1 - \delta$
+    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be SPSD and continuous for $t \in [a, b]$. Choose $\delta \in (0, 2e^{-1}]$ and $\varepsilon \in (0, 1)$.
+    Then, for $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(\delta^{-1}))$,
+    the inequality 
     \begin{equation}
         \int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t
-        \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
+        \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t
     \end{equation}
+    holds with probability at least $1 - \delta$.
 \end{theorem}
-
 \begin{proof}
-    By choosing $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\tilde{\varepsilon}^{-2} \log(\tilde{\delta}^{-1}))$ we get
-    \begin{align}
-        &\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t \notag \\
-        &= \int_{a}^{b} | \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) |~\mathrm{d}t && \text{(by definition of estimators)} \notag \\
-        &\leq \tilde{\varepsilon} \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F ~\mathrm{d}t && \text{(\refthm{thm:hutchinson} w.p. $\geq 1 - \tilde{\delta}$)} \notag \\
-        &\leq \tilde{\varepsilon}^2 \int_{a}^{b} \Trace(\mtx{B}(t)) ~\mathrm{d}t && \text{(\reflem{lem:nystrom} w.p. $\geq 1 - \tilde{\delta}$)} 
-    \end{align}
-    with probability at least $1 - 2\tilde{\delta}$ by a union bound. Taking $\varepsilon = \tilde{\varepsilon}^2$ and $\delta = 2 \tilde{\delta}$ we conclude the result.
-
+    Choosing $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\tilde{\varepsilon}^{-2} \log(\tilde{\delta}^{-1}))$, \refthm{thm:hutchinson} and~\reflem{lem:nystrom} imply that
+    \begin{align*}
+        &\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t 
+        = \int_{a}^{b} | \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) |~\mathrm{d}t  \\
+        \leq\ & \tilde{\varepsilon} \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F ~\mathrm{d}t       \leq \tilde{\varepsilon}^2 \int_{a}^{b} \Trace(\mtx{B}(t)) ~\mathrm{d}t
+    \end{align*}
+    holds with probability at least $1 - 2\tilde{\delta}$. Taking $\varepsilon = \tilde{\varepsilon}^2$ and $\delta = 2 \tilde{\delta}$ concludes the proof.
 \end{proof}
 
-Comparing \refthm{thm:nystrom-pp} with its equivalent for constant matrices \cite[Theorem 3.4]{persson-2022-improved-variants}, we notice an additional factor of $\sqrt{\log(\delta^{-1})}$ in the required number of random vectors $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$.
+Comparing \refthm{thm:nystrom-pp} with the analogous result for constant matrices \cite[Theorem 3.4]{persson-2022-improved-variants}, we notice an additional factor of $\sqrt{\log(\delta^{-1})}$ in the required number of random vectors $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$.
 

### 2025-02-18 paper/spectraldensity.tex
review of section 2 finished

@@
 
-We consider the task of computing the spectral density of a real symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ with eigenvalues $\lambda_1, \dots, \lambda_n$, which is defined as
-\begin{equation}
-    \phi(t) = \frac{1}{n} \sum_{i=1}^{n} \delta(t - \lambda_i),
-    \label{equ:spectral-density}
-\end{equation}
-where $\delta$ is the Dirac delta distribution.
-Finding approximations to this distribution is challenging.
+We now get back to the task of computing the spectral density $1/n \sum_{i=1}^{n} \delta(t - \lambda_i)$ of a real symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ with eigenvalues $\lambda_1, \dots, \lambda_n$, where $\delta(\cdot)$ denotes the Dirac delta distribution.
+As explained in the introduction, finding accurate approximations to this distribution is challenging and we will therefore work with
+a smoothed version. %This still yields enough information
+%.
 Furthermore, in most applications the exact location of each individual eigenvalue is not important, but more so their approximate locations relative to each other, such as eigenvalue clusters, spectral gaps, or outliers.
@@
 
-The smoothed spectral density $\phi_{\sigma}$ is defined as
+We recall that the smoothed spectral density $\phi_{\sigma}$ is defined as
 \begin{equation}
@@
 \end{equation}
-for a smoothing kernel $g_{\sigma}$ parametrized by a smoothing parameter $\sigma > 0$, which controls by how much $\phi$ is smoothed.
-We choose a Gaussian smoothing kernel of width $\sigma > 0$, given by
+for a smoothing kernel $g_{\sigma}$ parametrized by a smoothing parameter $\sigma > 0$. In this work, we choose a Gaussian smoothing kernel of width $\sigma > 0$:
 \begin{equation}
@@
 \end{equation}
-The task in \refequ{equ:smooth-spectral-density} is to approximate the trace of the parameter-dependent SPSD matrix $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
-Hence, we can apply the results derived in \refsec{sec:analysis}.
+The expression~\refequ{equ:smooth-spectral-density} gives rise to the task of estimating the trace of the parameter-dependent SPSD matrix $g_{\sigma}(t\mtx{I}_n - \mtx{A})$, which has been analyzed in \refsec{sec:analysis}.
 
-Choosing the smoothing parameter $\sigma$ results in the following trade off: Typically, larger $\sigma$ make it easier to approximate the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$, whereas smaller $\sigma$ allow us to stay closer to the original spectral density \refequ{equ:spectral-density}.
-To see the latter, we measure the error between the spectral density $\phi$ and its smoothened version $\phi_{\sigma}$ with
+Choosing the smoothing parameter $\sigma$ results in the following trade off: A larger $\sigma$ usually makes it  easier to approximate the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$, whereas a smaller $\sigma$ allow one to stay closer to the original spectral density $\phi$.
+To quantify the latter, one can measure the error between $\phi$ and its smoothened version $\phi_{\sigma}$ by
 \begin{equation}
@@
 where $\mathcal{S}$ represents an appropriately chosen space of test functions.
-Among others, \cite{lin-2016-approximating-spectral} proposes $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - s), s \in [a, b]\}$, \cite{chen-2021-analysis-stochastic} uses $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [a, b] \}$ where $\Theta$ is the Heaviside step function, and \cite{braverman-2022-sublinear-time,bhattacharjee-2024-improved-spectral} use $\mathcal{S} = \{f : |f(t) - f(s)| \leq |t - s| \}$ for which the error \refequ{equ:error-metric} is equivalent to the Wasserstein-1 distance between $\phi$ and $\phi_{\sigma}$.
-In this last metric, provided we assume the eigenvalues to be somewhat uniformly distributed and $\sigma$ to be rather small, a simple calculation allows us to roughly estimate the smoothing error \refequ{equ:error-metric} to be $\sigma$.
-Consequently, if we can only allow an approximation to deviate by at most a factor of $\varepsilon > 0$ from the original spectral density, we should choose a smoothing parameter $\sigma \approx \varepsilon$.
+Among others, the reference~\cite{lin-2016-approximating-spectral} uses $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - s), s \in [a, b]\}$, \cite{chen-2021-analysis-stochastic} uses $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [a, b] \}$ with the Heaviside step function $\Theta$, and \cite{braverman-2022-sublinear-time,bhattacharjee-2024-improved-spectral} use $\mathcal{S} = \{f : |f(t) - f(s)| \leq |t - s| \}$ for which the error \refequ{equ:error-metric} is equivalent to the Wasserstein-1 distance between $\phi$ and $\phi_{\sigma}$.
+When choosing 
+the last metric,
+assuming that the eigenvalues are somewhat uniformly distributed and $\sigma$ is rather small, a simple calculation allows one to conclude that the smoothing error \refequ{equ:error-metric} is of the order $\sigma$.
+Consequently, to ensure that the approximation deviates by at most a factor of $\varepsilon > 0$, one needs to choose $\sigma \sim \varepsilon$.
 
@@
 
-To estimate the trace in \refequ{equ:smooth-spectral-density}, we first need to evaluate the involved matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
+To estimate the trace in \refequ{equ:smooth-spectral-density} efficiently, we first approximate the function 
+$s\mapsto g_{\sigma}(t- s)$ by a truncated
+Chebyshev expansion:
+%need to evaluate the involved matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
 However, doing this exactly would require us to first diagonalize $\mtx{A}$, which is often a prohibitively expensive operation.
-For algorithmic reasons (see \refsec{subsubsec:chebyshev-nystrom-implementation}), we seek for approximations of this matrix function in terms of affine linear expansions, of which the Chebyshev expansion was determined to be most suitable due to its approximation error guarantees and ease of manipulation.
+For algorithmic reasons (see \refsec{subsubsec:chebyshev-nystrom-implementation}), we seek for approximations of this matrix function in terms of affine linear expansions, of which the was determined to be most suitable due to its approximation error guarantees and ease of manipulation.
 In every $t \in \mathbb{R}$ we expand
 \begin{equation}
-    g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) = \sum_{l=0}^{m} \mu_l(t) T_l(\mtx{A})
+    g_{\sigma}(t\mtx{I}_n - \mtx{A}) \approx g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) = \sum_{l=0}^{m} \mu_l(t) T_l(\mtx{A}),
     \label{equ:matrix-expansion}
 \end{equation}
-in terms of a set of $m+1$ coefficients $\mu_0(t), \dots, \mu_m(t)$ which linearly combine the Chebyshev polynomials
+with the expansion coefficients $\mu_0(t), \dots, \mu_m(t) \in \mathbb R$ and the Chebyshev polynomials
+$T_1,\ldots,T_m$. Assuming that the spectrum of $\mtx{A}$ is contained in $[-1,1]$, one has that
+$T_l(s) = \cos(l \cdot \arccos(s))$ satisfies the 
+three-term recurrence relation
 \begin{equation}
-    \begin{cases}
-        T_l : [-1, 1] \to \mathbb{R}; \\
-        T_l(s) = \cos(l \cdot \arccos(s)),
-    \end{cases}
-    \label{equ:chebyshev-polynomial}
-\end{equation}
-evaluated at the matrix $\mtx{A}$. It is easy to verify that these polynomials satisfy the three-term recurrence relation
-\begin{equation}
-    T_l(s) =
-    \begin{cases}
-        s^l, & \text{if $l \in \{0, 1\}$}; \\
-        T_l(s) = 2 s T_{l-1}(s) - T_{l-2}(s), & \text{if $l \geq 2$.}
-    \end{cases}
+    T_0(s) = 1,\quad  T_1(s) = s,\quad 
+    T_l(s) = 2 s T_{l-1}(s) - T_{l-2}(s) \text{ for $l \geq 2$.}
     \label{equ:chebyshev-recurrence}
 \end{equation}
-With the help of the Chebyshev expansion \refequ{equ:matrix-expansion} we can now define the expanded spectral density
+Inserting~\refequ{equ:matrix-expansion} into~\eqref{equ:smooth-spectral-density} yields the approximate smoothed spectral density
 \begin{equation}
-    \phi_{\sigma}^{(m)}(t) =  \frac{1}{n} \Trace(g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})).
+    \phi_{\sigma}(t) \approx \phi_{\sigma}^{(m)}(t) =  \frac{1}{n} \Trace(g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})).
     \label{equ:expanded-spectral-density}
@@
 
-However, the Chebyshev polynomials are only well-defined for matrices whose spectrum is contained in $[-1, 1]$.
-When working with general matrices of bounded spectrum, we will first have to estimate a lower bound $a$ and upper bound $b$ on the spectrum \cite{zhou-2011-bounding-spectrum}.
-A spectral transform
-\begin{equation}
-    \begin{cases}
-        \tau : [a, b] \to [-1, 1]\\
-        \tau(t) = \frac{2t - a - b}{b - a}
-    \end{cases}
-\end{equation}
-then lets us define the matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ whose spectrum is identical to the one of $\mtx{A}$ but shrunk to the interval $[-1, 1]$, for which the expansion \refequ{equ:matrix-expansion} is valid\footnote{Note that we will have to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when inverting the transformation.}.
-From now on, we will assume the eigenvalues of the matrix $\mtx{A}$ to be contained in $[-1, 1]$.
+For a general symmetric matrix $\mtx{A}$, one first estimates an interval $[a,b]$ containing the spectrum \cite{zhou-2011-bounding-spectrum}. The usual affine transform
+$\tau(t) = \frac{2t - a - b}{b - a}$
+then lets us define a matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ for which the spectrum is contained in $[-1, 1]$ and, hence, the expansion \refequ{equ:matrix-expansion} is valid.\footnote{Note that one has to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when inverting the transformation.} From now on, we will assume without loss of generality that the spectrum of $\mtx{A}$ is contained in $[-1, 1]$.
 
@@
 \label{subsubsec:dct}
+
+??? UNTIL HERE ???
+\color{black}
 

### 2025-02-18 paper/analysis.tex
change constant in Nyström proof and add references

@@
 Related randomized low-rank approximations for general parameter-dependent matrices have been analyzed in~\cite{kressner-2024-randomized-lowrank}.
-Exploiting the invariance of the trace under cyclic permutations, taking the trace of~\eqref{equ:nystrom-approximation} gives rise to the Nyström estimator
+Exploiting the invariance of the trace under cyclic permutations, taking the trace of~\refequ{equ:nystrom-approximation} gives rise to the Nyström estimator
 \begin{equation}
@@
 \paragraph{Nyström++ estimator.}
-The Nyström estimator~\eqref{equ:nystrom-trace-estimator} relies on the feasibility of low-rank approximation, that is, the singular values of $\mtx{B}(t)$ need to decay sufficiently quickly.
+The Nyström estimator~\refequ{equ:nystrom-trace-estimator} relies on the feasibility of low-rank approximation, that is, the singular values of $\mtx{B}(t)$ need to decay sufficiently quickly.
 If this property cannot be ensured, it is preferable to use an estimator that combines the Girard-Hutchinson and Nyström estimators.
-The combination suggested in~\cite{lin-2017-randomized-estimation} applies Girard-Hutchinson to correct the error of the Nyström approximation~\eqref{equ:nystrom-approximation}:
+The combination suggested in~\cite{lin-2017-randomized-estimation} applies Girard-Hutchinson to correct the error of the Nyström approximation~\refequ{equ:nystrom-approximation}:
 \begin{equation}
@@
 where $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$, $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ are independent Gaussian random matrices. 
-\textcolor{red}{??? What about XTrace? Does it apply; should we mention it somewhere? ???}
+%\textcolor{red}{??? What about XTrace? Does it apply; should we mention it somewhere? ???}
 
-For constant $\mtx{\Omega}$, the estimator~\eqref{equ:nystrompp-trace-estimator} coincides with the Nyström++ estimator from~\cite{persson-2022-improved-variants}, which is based on the Hutch++ estimator~\cite{meyer-2021-hutch-optimal}.
+For constant $\mtx{\Omega}$, the estimator~\refequ{equ:nystrompp-trace-estimator} coincides with the Nyström++ estimator from~\cite{persson-2022-improved-variants}, which is based on the Hutch++ estimator~\cite{meyer-2021-hutch-optimal}.
 In this situation (constant $\mtx{B}$), these estimators were both shown to achieve a relative $\varepsilon$-error with $\mathcal{O}(\varepsilon^{-1})$ matrix-vector products only, independent of the singular value decay of $\mtx{B}$.
+
+\textcolor{green}{At this point, we acknowledge the existence of the XNysTrace estimator from \cite{epperly-2024-xtrace-making} which oftentimes outperforms the Nyström++ estimator.
+Moreover, a straight-forward extension of the XNysTrace estimator to the parameter-dependent setting seems to be in reach.
+However, what distinguishes \refalg{alg:nystrom-chebyshev-pp} for estimating spectral densities are two key observations --- using the cyclic invariance of the trace and the affine linear form of Chebyshev expansions (see \refsec{subsubsec:chebyshev-nystrom-implementation} for details) --- which it exploits to significantly speed up the computation.
+Unfortunately, we do not see a way of marrying these observations with the efficient implementation of XNysTrace described in \cite[Section 2.2]{epperly-2024-xtrace-making}, which limits its suitability for spectral density estimation.} \textcolor{green}{???
+Maybe this would be more appropriate in the introduction?
+???}
 %We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.
@@
     \end{align}
-    This implies~\eqref{eq:ghbound} by Markov's inequality.
+    This implies~\refequ{eq:ghbound} by Markov's inequality.
 
@@
         holds with probability at least $1 - \gamma^{-n_{\mtx{\Omega}} / 4}$ for 
-        $c = 108$.
+        $c = \textcolor{green}{154}$.
 In particular, given $\varepsilon > 0$ and $\delta \in (0, 1)$, the bound $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$ holds with probability at least $1-\delta$ if $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1}))$.
@@
     $\mtx{\Omega}_1 := \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times n_{\mtx{\Omega}}}$ and $\mtx{\Omega}_2 := \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times n_{\mtx{\Omega}}}$, which are independent Gaussian random matrices.
-Applying Theorem B.1 from~\cite{persson-2023-randomized-lowrank} for $f(x) = x$, see also \textcolor{red}{???
-Also cite the same result from Tropp and Webber or Epperly and Tropp?
-???}, yields the bound
+Applying Theorem B.1 from~\cite{persson-2023-randomized-lowrank} for $f(x) = x$, see also \textcolor{green}{proof of \cite[Corollary 8.2]{tropp-2023-randomized-algorithms}}, yields the bound
     \begin{equation}
@@
 
-To treat the second term in~\eqref{equ:nystrom-proof-persson-bonud}, we bound 
+To treat the second term in~\refequ{equ:nystrom-proof-persson-bonud}, we bound 
 its $q$th moment using the independence of $\mtx{\Omega}_1$ and $\mtx{\Omega}_2$:
@@
     To continue from here, we set $k = n_{\mtx{\Omega}}/2$ and $q = n_{\mtx{\Omega}}/4$. By the calculations after~\cite[Eqn~(B.7)]{tropp-2023-randomized-algorithms}, one has
-    \textcolor{red}{??? Fabio, the red terms are different from your calculations. Please double check that I am correct ???}
     \[
@@
         \leq
-        \textcolor{red}{\frac{3}{2}}
+        \textcolor{green}{\left(1 + \frac{n_{\mtx{\Omega}}}{2}\right)}
         \left( \frac{3}{4} n_{\mtx{\Omega}}\right)^{\sfrac{n_{\mtx{\Omega}}}{4}}
@@
         = \mathbb{E}\left[ \big\| ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \big\|_2^{\sfrac{n_{\mtx{\Omega}}}{4}} \right]^{\sfrac{4}{n_{\mtx{\Omega}}}}
-        \le \frac{9}{8}  
-        \frac{e^2 n_{\mtx{\Omega}}}{(n_{\mtx{\Omega}} / 2+ 1)^2} \le 
-        \frac{9}{2}  \frac{e^2}{n_{\mtx{\Omega}}},
+        \le \frac{3}{4}  
+        \frac{e^4 n_{\mtx{\Omega}}}{(n_{\mtx{\Omega}} / 2+ 1)^2} \le 
+        \frac{3}{4}  \frac{e^4}{n_{\mtx{\Omega}}},
     \end{equation}
-    where we used $(m!)^{-\sfrac{1}{m}} \leq e/m$. Note that, in contrast 
+    where we used \textcolor{green}{$(1 + m)^{\sfrac{1}{m}} \leq e$ and} $(m!)^{-\sfrac{1}{m}} \leq e/m$. Note that, in contrast 
     to the result of~\cite[Lemma B.3]{tropp-2023-randomized-algorithms}, this inequality is valid for arbitrarily large $n_{\mtx{\Omega}}$, at the expense of a slightly larger constant.
@@
 %
-    The second factor in~\eqref{equ:nystrom-proof-processed-tail} is bounded using \reflem{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$: % and $p = n_{\mtx{\Omega}}/2$:
+    The second factor in~\refequ{equ:nystrom-proof-processed-tail} is bounded using \reflem{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$: % and $p = n_{\mtx{\Omega}}/2$:
     \[
         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \big\|_2^2 \right]
-        \leq \textcolor{red}{\frac{3}{2}} n_{\mtx{\Omega}} \Big( \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \big\|_2^2 + \frac{1}{n_{\mtx{\Omega}}} \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \big\|_F^2 \Big).
+        \leq \textcolor{green}{\frac{5}{4}} n_{\mtx{\Omega}} \Big( \textcolor{green}{2} \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \big\|_2^2 + \frac{1}{n_{\mtx{\Omega}}} \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \big\|_F^2 \Big).
       %  \label{equ:spectral-norm-bound-applied}
@@
         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \frac{27 e^2}{4\sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \Big( \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \Big).
+        \leq \textcolor{green}{\frac{15 e^4}{16}}  \sqrt{n_{\mtx{\Omega}}} \Big( \textcolor{green}{2} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \Big).
         %\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
@@
         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \frac{27 e^2}{4 \sqrt{2}}  \sqrt{n_{\mtx{\Omega}}} \Big( \frac{2}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) + \frac{1}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) \Big)
-        \leq  \frac{106}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
+        \leq \textcolor{green}{\frac{15 e^4}{16}}  \sqrt{n_{\mtx{\Omega}}} \Big( \frac{2}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) + \frac{1}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) \Big)
+        \leq  \frac{\textcolor{green}{154}}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
         \label{equ:nystrom-proof-tail-bound}
@@
 
-    Inserting~\refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud} and using the triangle inequality for $\mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}[\cdot]$, we obtain
+    Inserting~\refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud}\textcolor{green}{, letting $c=154$,} and using the triangle inequality for $\mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}[\cdot]$, we obtain
     \begin{equation} \label{eq:blubber}
@@
     As in \cite{kressner-2024-randomized-lowrank}, one can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable.
-    As in the proof of~\refthm{thm:hutchinson}, the claimed bound~\eqref{qu:nystrompp-theorem-bound}
-    follows from~\eqref{eq:blubber} using Minkowski's integral inequality and Markov's inequality.
+    As in the proof of~\refthm{thm:hutchinson}, the claimed bound~\refequ{qu:nystrompp-theorem-bound}
+    follows from~\refequ{eq:blubber} using Minkowski's integral inequality and Markov's inequality.
 %     

### 2025-02-18 paper/spectraldensity.tex
re-check chapter 3

@@
 \end{equation}
-with the expansion coefficients $\mu_0(t), \dots, \mu_m(t) \in \mathbb R$ and the Chebyshev polynomials
+with the expansion coefficients $\mu_0(t), \dots, \mu_m(t) \in \mathbb R$ and the Chebyshev polynomials of the first kind
 $T_1,\ldots,T_m$. Assuming that the spectrum of $\mtx{A}$ is contained in $[-1,1]$, one has that
@@
 \end{equation}
-Inserting~\refequ{equ:matrix-expansion} into~\eqref{equ:smooth-spectral-density} yields the approximate smoothed spectral density
+Inserting~\refequ{equ:matrix-expansion} into~\refequ{equ:smooth-spectral-density} yields the approximate smoothed spectral density
 \begin{equation}
@@
 
-It is well known that the coefficients of a Chebyshev expansion of a function can -- in exact arithmetic -- be computed by inverting the discrete cosine transform (DCT)
+It is well known that the coefficients of a Chebyshev expansion of a function can --- in exact arithmetic --- be computed by inverting the discrete cosine transform (DCT)
 \begin{equation}
@@
 \end{equation}
-between the expansion coefficients $\mu_l(t),~i=0,\dots,m$ and the function evaluations at the Chebyshev nodes $s_i = \cos(\pi i / m),~i=0,\dots,m$.
+between the expansion coefficients $\mu_l(t),~i=0,\dots,m$ and the function evaluations at the Chebyshev nodes of the second kind $s_i = \cos(\pi i / m),~i=0,\dots,m$.
 This can be accomplished in $\mathcal{O}(m \log(m))$ operations.
-\cite[Algorithm 1]{lin-2017-randomized-estimation} uses a slightly different approach, which only approximates the coefficients and is noticeably slower (cf.
+\cite[Algorithm 1]{lin-2017-randomized-estimation} uses a slightly different approach, which only approximates the expansion coefficients and is noticeably slower (cf.
 \reftab{tab:chebyshev-timing-interpolation}).
 
-A highly useful property of the Chebyshev expansion -- of which we will make use twice in the following pages -- is the fact that we can square such an expansion in $\mathcal{O}(m \log(m))$ operations \cite{baszenski-1997-fast-polynomial}, which will be used in the below algorithm.
+A highly useful property of the Chebyshev expansion --- of which we will make use twice in the following pages --- is the fact that we can square such an expansion in $\mathcal{O}(m \log(m))$ operations \cite{baszenski-1997-fast-polynomial}, which the below algorithm implements.
 
@@
         \Statex \textbf{Output:} Coefficients $\{ \nu_l \}_{l=0}^{2m}$ such that $\sum_{l=0}^{2m} \nu_l T_l = (\sum_{l=0}^{m} \mu_l T_l)^2$
-        \State Define $\vct{\mu} \in \mathbb{R}^{2m + 1}$ whose first $m + 1$ entries are the coefficients $\mu_l$ and the remaining $m$ are $0$
+        \State Define $\vct{\mu} \in \mathbb{R}^{2m + 1}$ whose first $m + 1$ entries are the coefficients $\mu_0, \dots, \mu_m$ and the remaining $m$ are $0$
     \State Compute $\vct{f} = \operatorname{DCT}(\vct{\mu})$
@@
 \reflem{lem:chebyshev-error} shows that when $0 < \sigma \ll 1$ and if we choose $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, the error we induce with the Chebyshev expansion is at most $\mathcal{O}(\varepsilon)$.
-It now also allows us to bound the error we make when expanding the spectral density.
+It now also allows us to bound the $L^1$-error we make when expanding the spectral density.
 
@@
 We can think of many ways in which this can be achieved: by using Jackson damping \cite{jackson-1912-approximation-trigonometric,braverman-2022-sublinear-time}, albeit at a significantly worse rate of convergence; by shifting up the expansion $g_{\sigma}^{(m)} + \rho$ with a large enough $\rho > 0$, at the loss of some of the low-rank structure of $g_{\sigma}(t \mtx{I}_n - \mtx{A})$; or by directly minimizing the approximation error over all non-negative Chebyshev polynomials \cite{fejer-1916-uber-trigonometrische}, which realistically only works for small $m$.
-In the end, we propose the following method to produce a non-negative Chebyshev expansion: First, expand the square root of the smoothing kernel in a basis of Chebyshev polynomials up to degree $m/2$
+In the end, we propose the following method to produce a non-negative Chebyshev expansion: Fix an even $m$ and expand the square root of the smoothing kernel in a basis of Chebyshev polynomials up to degree $m/2$
 \begin{equation}
@@
 \end{equation}
-Second, square this expansion with \refalg{alg:chebyshev-squaring} to obtain the non-negative Chebyshev expansion
+Then, square this expansion with \refalg{alg:chebyshev-squaring} to obtain the non-negative Chebyshev expansion
 \begin{equation}
@@
     \Statex \textbf{Input:} Symmetric $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum in $[-1, 1]$, points $\{t_i\}_{i=1}^{n_t} \subset \mathbb{R}$
-    \Statex \textbf{Parameters:} Degree of expansion $m \in 2\mathbb{N}$, number of Girard-Hutchinson queries $n_{\mtx{\Psi}} \in \mathbb{N}_0$, Nyström sketch size $n_{\mtx{\Omega}} \in \mathbb{N}_0$,  smoothing parameter $\sigma > 0$
+    \Statex \textbf{Parameters:} Even degree of expansion $m \in \mathbb{N}$, number of Girard-Hutchinson queries $n_{\mtx{\Psi}} \in \mathbb{N}_0$, Nyström sketch size $n_{\mtx{\Omega}} \in \mathbb{N}_0$,  smoothing parameter $\sigma > 0$
     \Statex \textbf{Output:} Approximate evaluations of the spectral density $\{\widetilde{\phi}_{\sigma}(t_i)\}_{i=1}^{n_t}$
@@
 \refthm{thm:nystrom}).
-To establish a relation between the singular value decay of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ and what choice of $n_{\mtx{\Omega}}$ is needed for a significantly better approximation than ensured by \refthm{thm:chebyshev-nystrom}, we study the numerical rank of $n^{-1} g_{\sigma}(t\mtx{I}_n - \mtx{A})$\footnote{The factor $n^{-1}$ comes from the definition of the spectral density \refequ{equ:smooth-spectral-density} when it is written as the trace of a matrix function.}.
+To establish a relation between the singular value decay of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ and what choice of $n_{\mtx{\Omega}}$ is needed for a significantly better approximation than ensured by \refthm{thm:chebyshev-nystrom}, we study the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$\footnote{The factor $1/n$ comes from the definition of the spectral density \refequ{equ:smooth-spectral-density} when it is written as the trace of a matrix function.}.
 
@@
 
-When using a Gaussian smoothing-kernel, we can explicitly write the singular values of $n^{-1} g_{\sigma}(t\mtx{I}_n - \mtx{A})$ as
+When using a Gaussian smoothing-kernel, we can explicitly write the singular values of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$ as
 \begin{equation}
-    \sigma_i(t) = n^{-1} g_{\sigma}(t - \lambda_{(i)}) = \frac{1}{n \sqrt{2 \pi \sigma^2}} e^{-\sfrac{(t - \lambda_{(i)})^2}{2 \sigma^2}},
+    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) / n = \frac{1}{n \sqrt{2 \pi \sigma^2}} e^{-\sfrac{(t - \lambda_{(i)})^2}{2 \sigma^2}},
     \label{equ:gaussian-kernel-eigenvalues}
@@
 \begin{equation}
-    r_{\varepsilon}(n^{-1} g_{\sigma}(t\mtx{I}_n - \mtx{A})) \leq \#\left\{1\leq i\leq n: |t - \lambda_i| < d_{\varepsilon, \sigma} \right\},
+    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n) \leq \#\left\{1\leq i\leq n: |t - \lambda_i| < d_{\varepsilon, \sigma} \right\},
     \label{equ:gaussian-kernel-numerical-rank}
@@
     \input{figures/numerical-rank.tex}
-    \caption{The numerical rank of $n^{-1} g_{\sigma}(t\mtx{I}_n - \mtx{A})$ can be approximately computed by counting the number of eigenvalues $\lambda_{(1)}, \dots, \lambda_{(n)}$ of the matrix $\mtx{A}$ which lie less than a constant $d_{\varepsilon, \sigma}$ away from $t$.}
+    \caption{The numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$ can be approximately computed by counting the number of eigenvalues $\lambda_{(1)}, \dots, \lambda_{(n)}$ of the matrix $\mtx{A}$ which lie less than a constant $d_{\varepsilon, \sigma}$ away from $t$.}
     \label{fig:numerical-rank}
@@
 
-If we additionally assume the eigenvalues of the matrix $\mtx{A}$ to be evenly distributed within $[-1, 1]$, that is, in any subinterval of fixed length in $[a, b]$, $-1 \leq a < b \leq 1$, we can expect to find roughly the same number of eigenvalues, then we can estimate the numerical rank of $n^{-1} g_{\sigma}(t\mtx{I}_n - \mtx{A})$ to be
+If we additionally assume the eigenvalues of the matrix $\mtx{A}$ to be evenly distributed within $[-1, 1]$, that is, in any subinterval of fixed length in $[a, b]$, $-1 \leq a < b \leq 1$, we can expect to find roughly the same number of eigenvalues, then we can estimate the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$ to be
 \begin{equation}
-    r_{\varepsilon}(n^{-1} g_{\sigma}(t\mtx{I}_n - \mtx{A})) \lessapprox n d_{\varepsilon, \sigma} \approx 6 n \sigma,
+    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n) \lessapprox n d_{\varepsilon, \sigma} \approx 6 n \sigma,
     \label{equ:gaussian-kernel-numerical-rank-uniform}

### 2025-02-18 paper/re-pro-badge.tex

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/f30af3c}{f30af3c} on 2024-10-31 at 11:41:34 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/c6a997e}{c6a997e} on 2025-02-18 at 14:03:19 UTC.

### 2025-02-18 paper/analysis.tex
remove green color

@@
 
-\textcolor{green}{At this point, we acknowledge the existence of the XNysTrace estimator from \cite{epperly-2024-xtrace-making} which oftentimes outperforms the Nyström++ estimator.
+At this point, we acknowledge the existence of the XNysTrace estimator from \cite{epperly-2024-xtrace-making} which oftentimes outperforms the Nyström++ estimator.
 Moreover, a straight-forward extension of the XNysTrace estimator to the parameter-dependent setting seems to be in reach.
 However, what distinguishes \refalg{alg:nystrom-chebyshev-pp} for estimating spectral densities are two key observations --- using the cyclic invariance of the trace and the affine linear form of Chebyshev expansions (see \refsec{subsubsec:chebyshev-nystrom-implementation} for details) --- which it exploits to significantly speed up the computation.
-Unfortunately, we do not see a way of marrying these observations with the efficient implementation of XNysTrace described in \cite[Section 2.2]{epperly-2024-xtrace-making}, which limits its suitability for spectral density estimation.} \textcolor{green}{???
-Maybe this would be more appropriate in the introduction?
-???}
+Unfortunately, we do not see a way of marrying these observations with the efficient implementation of XNysTrace described in \cite[Section 2.2]{epperly-2024-xtrace-making}, which limits its suitability for spectral density estimation.
 %We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.
@@
         holds with probability at least $1 - \gamma^{-n_{\mtx{\Omega}} / 4}$ for 
-        $c = \textcolor{green}{154}$.
+        $c = 154$.
 In particular, given $\varepsilon > 0$ and $\delta \in (0, 1)$, the bound $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$ holds with probability at least $1-\delta$ if $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1}))$.
@@
     $\mtx{\Omega}_1 := \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times n_{\mtx{\Omega}}}$ and $\mtx{\Omega}_2 := \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times n_{\mtx{\Omega}}}$, which are independent Gaussian random matrices.
-Applying Theorem B.1 from~\cite{persson-2023-randomized-lowrank} for $f(x) = x$, see also \textcolor{green}{proof of \cite[Corollary 8.2]{tropp-2023-randomized-algorithms}}, yields the bound
+Applying Theorem B.1 from~\cite{persson-2023-randomized-lowrank} for $f(x) = x$, see also proof of \cite[Corollary 8.2]{tropp-2023-randomized-algorithms}, yields the bound
     \begin{equation}
@@
         \leq
-        \textcolor{green}{\left(1 + \frac{n_{\mtx{\Omega}}}{2}\right)}
+        \left(1 + \frac{n_{\mtx{\Omega}}}{2}\right)
         \left( \frac{3}{4} n_{\mtx{\Omega}}\right)^{\sfrac{n_{\mtx{\Omega}}}{4}}
@@
     \end{equation}
-    where we used \textcolor{green}{$(1 + m)^{\sfrac{1}{m}} \leq e$ and} $(m!)^{-\sfrac{1}{m}} \leq e/m$. Note that, in contrast 
+    where we used $(1 + m)^{\sfrac{1}{m}} \leq e$ and $(m!)^{-\sfrac{1}{m}} \leq e/m$. Note that, in contrast 
     to the result of~\cite[Lemma B.3]{tropp-2023-randomized-algorithms}, this inequality is valid for arbitrarily large $n_{\mtx{\Omega}}$, at the expense of a slightly larger constant.
@@
         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \big\|_2^2 \right]
-        \leq \textcolor{green}{\frac{5}{4}} n_{\mtx{\Omega}} \Big( \textcolor{green}{2} \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \big\|_2^2 + \frac{1}{n_{\mtx{\Omega}}} \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \big\|_F^2 \Big).
+        \leq \frac{5}{4} n_{\mtx{\Omega}} \Big( 2 \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \big\|_2^2 + \frac{1}{n_{\mtx{\Omega}}} \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \big\|_F^2 \Big).
       %  \label{equ:spectral-norm-bound-applied}
@@
         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \textcolor{green}{\frac{15 e^4}{16}}  \sqrt{n_{\mtx{\Omega}}} \Big( \textcolor{green}{2} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \Big).
+        \leq \frac{15 e^4}{16}  \sqrt{n_{\mtx{\Omega}}} \Big( 2 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \Big).
         %\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
@@
         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        \leq \textcolor{green}{\frac{15 e^4}{16}}  \sqrt{n_{\mtx{\Omega}}} \Big( \frac{2}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) + \frac{1}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) \Big)
-        \leq  \frac{\textcolor{green}{154}}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
+        \leq \frac{15 e^4}{16}  \sqrt{n_{\mtx{\Omega}}} \Big( \frac{2}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) + \frac{1}{n_{\mtx{\Omega}}} \Trace(\mtx{B}) \Big)
+        \leq  \frac{154}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
         \label{equ:nystrom-proof-tail-bound}
@@
 
-    Inserting~\refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud}\textcolor{green}{, letting $c=154$,} and using the triangle inequality for $\mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}[\cdot]$, we obtain
+    Inserting~\refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud}, letting $c=154$, and using the triangle inequality for $\mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}[\cdot]$, we obtain
     \begin{equation} \label{eq:blubber}

### 2025-02-18 paper/intro.tex
remove green color

@@
 All methods considered in this work are based on the following simple idea: Apply an existing randomized trace estimator to $\Trace(\mtx{B}(t))$ with \emph{constant} random vectors, that is, the same randomization is used for each value of the parameter $t$.
-\textcolor{green}{For example, the Girard-Hutchinson estimator becomes $n_{\mtx{\Psi}}^{-1} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j$ for $n_{\mtx{\Psi}}$ constant Gaussian random vectors $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}}$.}
+For example, the Girard-Hutchinson estimator becomes $n_{\mtx{\Psi}}^{-1} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j$ for $n_{\mtx{\Psi}}$ constant Gaussian random vectors $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}}$.
 ??? STOP HERE ???

### 2025-02-19 paper/spectraldensity.tex
review pass through beginning of section 3

@@
 
-To estimate the trace in \refequ{equ:smooth-spectral-density} efficiently, we first approximate the function 
+To estimate the trace in \refequ{equ:smooth-spectral-density} efficiently, we approximate the function 
 $s\mapsto g_{\sigma}(t- s)$ by a truncated
-Chebyshev expansion:
+Chebyshev expansion $g_{\sigma}^{(m)}(t-s) = 
+\sum_{l=0}^{m} \mu_l(t) T_l(s)$
+with the expansion coefficients $\mu_l(t) \in \mathbb R$ and the Chebyshev polynomials
+$T_l(s) = \cos(l \cdot \arccos(s))$ for $l = 0,1,\ldots,m$.
+Assuming that the spectrum of $\mtx{A}$ is contained in $[-1,1]$, this gives rise to
 %need to evaluate the involved matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
@@
 \end{equation}
-with the expansion coefficients $\mu_0(t), \dots, \mu_m(t) \in \mathbb R$ and the Chebyshev polynomials of the first kind
-$T_1,\ldots,T_m$. Assuming that the spectrum of $\mtx{A}$ is contained in $[-1,1]$, one has that
-$T_l(s) = \cos(l \cdot \arccos(s))$ satisfies the 
+\textcolor{red}{??? MOVE BELOW! This is not at a good place
+For later purposes, we recall the well-known
 three-term recurrence relation
@@
     \label{equ:chebyshev-recurrence}
-\end{equation}
+\end{equation}}
 Inserting~\refequ{equ:matrix-expansion} into~\refequ{equ:smooth-spectral-density} yields the approximate smoothed spectral density
@@
 \end{equation}
-
-For a general symmetric matrix $\mtx{A}$, one first estimates an interval $[a,b]$ containing the spectrum \cite{zhou-2011-bounding-spectrum}. The usual affine transform
+\begin{remark}
+For a general symmetric matrix $\mtx{A}$, one first estimates an interval $[a,b]$ containing the spectrum of $A$~\cite{zhou-2011-bounding-spectrum}. The usual affine transform
 $\tau(t) = \frac{2t - a - b}{b - a}$
-then lets us define a matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ for which the spectrum is contained in $[-1, 1]$ and, hence, the expansion \refequ{equ:matrix-expansion} is valid.\footnote{Note that one has to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when inverting the transformation.} From now on, we will assume without loss of generality that the spectrum of $\mtx{A}$ is contained in $[-1, 1]$.
-
-\subsubsection{Fast and consistent squaring}
-\label{subsubsec:dct}
-
-??? UNTIL HERE ???
-\color{black}
-
-It is well known that the coefficients of a Chebyshev expansion of a function can --- in exact arithmetic --- be computed by inverting the discrete cosine transform (DCT)
-\begin{equation}
-    g_{\sigma}(t - \cos(s_i)) = \sum_{l=0}^{m} \mu_l(t) \cos\left(\frac{\pi i l}{m} \right)
-    \label{equ:discrete-cosine-transform}
-\end{equation}
-between the expansion coefficients $\mu_l(t),~i=0,\dots,m$ and the function evaluations at the Chebyshev nodes of the second kind $s_i = \cos(\pi i / m),~i=0,\dots,m$.
-This can be accomplished in $\mathcal{O}(m \log(m))$ operations.
-\cite[Algorithm 1]{lin-2017-randomized-estimation} uses a slightly different approach, which only approximates the expansion coefficients and is noticeably slower (cf.
-\reftab{tab:chebyshev-timing-interpolation}).
-
-A highly useful property of the Chebyshev expansion --- of which we will make use twice in the following pages --- is the fact that we can square such an expansion in $\mathcal{O}(m \log(m))$ operations \cite{baszenski-1997-fast-polynomial}, which the below algorithm implements.
-
-\begin{algo}{Fast squaring of Chebyshev expansion}{chebyshev-squaring}
-    \begin{algorithmic}[1]
-        \Statex \textbf{Input:} Coefficients $\{ \mu_l \}_{l=0}^{m}$ of the expansion $\sum_{l=0}^{m} \mu_l T_l$
-        \Statex \textbf{Output:} Coefficients $\{ \nu_l \}_{l=0}^{2m}$ such that $\sum_{l=0}^{2m} \nu_l T_l = (\sum_{l=0}^{m} \mu_l T_l)^2$
-        \State Define $\vct{\mu} \in \mathbb{R}^{2m + 1}$ whose first $m + 1$ entries are the coefficients $\mu_0, \dots, \mu_m$ and the remaining $m$ are $0$
-    \State Compute $\vct{f} = \operatorname{DCT}(\vct{\mu})$
-    \State Compute $\vct{\nu} = \operatorname{DCT}^{-1}(\vct{f} \odot \vct{f})$ whose entries are the coefficients $\nu_l$
-    \label{lin:inverse-DCT}
-    \end{algorithmic}
-\end{algo}
-On \reflin{lin:inverse-DCT}, $\odot$ denotes the entrywise product.
+then lets us define a matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ for which the spectrum is contained in $[-1, 1]$ and, hence, the expansion \refequ{equ:matrix-expansion} is valid.\footnote{Note that one has to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when inverting the transformation.} This allows to assume, from now on, without loss of generality that the spectrum of $\mtx{A}$ is contained in $[-1, 1]$.
+\end{remark}
 
@@
 
-\subsection{Chebyshev-Nyström++ method for spectral density approximation}
-\label{subsec:chebyshev-nystrom}
+\subsubsection{Preservation of non-negativity}
 
@@
 %}
+
+\subsection{Computation of expansion coefficients}
+
+\subsubsection{Fast and consistent squaring}
+\label{subsubsec:dct}
+
+??? UNTIL HERE ???
+\color{black}
+
+It is well known that the coefficients of a Chebyshev expansion of a function can --- in exact arithmetic --- be computed by inverting the discrete cosine transform (DCT)
+\begin{equation}
+    g_{\sigma}(t - \cos(s_i)) = \sum_{l=0}^{m} \mu_l(t) \cos\left(\frac{\pi i l}{m} \right)
+    \label{equ:discrete-cosine-transform}
+\end{equation}
+between the expansion coefficients $\mu_l(t),~i=0,\dots,m$ and the function evaluations at the Chebyshev nodes of the second kind $s_i = \cos(\pi i / m),~i=0,\dots,m$.
+This can be accomplished in $\mathcal{O}(m \log(m))$ operations.
+\cite[Algorithm 1]{lin-2017-randomized-estimation} uses a slightly different approach, which only approximates the expansion coefficients and is noticeably slower (cf.
+\reftab{tab:chebyshev-timing-interpolation}).
+
+A highly useful property of the Chebyshev expansion --- of which we will make use twice in the following pages --- is the fact that we can square such an expansion in $\mathcal{O}(m \log(m))$ operations \cite{baszenski-1997-fast-polynomial}, which the below algorithm implements.
+
+\begin{algo}{Fast squaring of Chebyshev expansion}{chebyshev-squaring}
+    \begin{algorithmic}[1]
+        \Statex \textbf{Input:} Coefficients $\{ \mu_l \}_{l=0}^{m}$ of the expansion $\sum_{l=0}^{m} \mu_l T_l$
+        \Statex \textbf{Output:} Coefficients $\{ \nu_l \}_{l=0}^{2m}$ such that $\sum_{l=0}^{2m} \nu_l T_l = (\sum_{l=0}^{m} \mu_l T_l)^2$
+        \State Define $\vct{\mu} \in \mathbb{R}^{2m + 1}$ whose first $m + 1$ entries are the coefficients $\mu_0, \dots, \mu_m$ and the remaining $m$ are $0$
+    \State Compute $\vct{f} = \operatorname{DCT}(\vct{\mu})$
+    \State Compute $\vct{\nu} = \operatorname{DCT}^{-1}(\vct{f} \odot \vct{f})$ whose entries are the coefficients $\nu_l$
+    \label{lin:inverse-DCT}
+    \end{algorithmic}
+\end{algo}
+On \reflin{lin:inverse-DCT}, $\odot$ denotes the entrywise product.
+
+\subsection{Chebyshev-Nyström++ method for spectral density approximation}
+\label{subsec:chebyshev-nystrom}
 
@@
 Similar observations for the case when $n_{\mtx{\Psi}}$ is nonzero lead to the following algorithm.
+\textcolor{red}{??? Lines 1 and 2 are quite confusing. Say more explicitly how $\mu$ is computed. Is Algorithm 3.3 applied twice? ???}
 
@@
     \end{equation}
-    Since $\sqrt{g_{\sigma}} = \sqrt{2 \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$ and $\sqrt{g_{\sigma}} \leq 1/\sqrt{\sigma \sqrt{2 \pi}}$ we can apply the result from \refthm{thm:chebyshev-error} to get
+    Since $\sqrt{g_{\sigma}} = \sqrt{2 \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$ and $\sqrt{g_{\sigma}} \leq 1/\sqrt{\sigma \sqrt{2 \pi}}$ \textcolor{red}{we can apply the result from \refthm{thm:chebyshev-error} ???
+the reference seems to be incorrect ???} to get
     \begin{align}

### 2025-02-19 paper/spectraldensity.tex
change symbols and write todo list for results

--- 2025-02-18 paper/spectraldensity.tex
@@
 \begin{equation}
-    \underline{g}^{(m)}_{\sigma}(t - s) = \sum_{l=0}^{m} \mu_l(t) T_l(s).
+    \underline{g}^{(m)}_{\sigma}(t - s) = \sum_{l=0}^{m} \underline{\mu}_l(t) T_l(s).
     \label{equ:non-negative-chebyshev-expansion}
@@
 \begin{equation}
-    \widetilde{\phi}_{\sigma}^{(m)}(t) = \frac{1}{n} \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}))
+    \underline{\phi}_{\sigma}^{(m)}(t) = \frac{1}{n} \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}))
     \label{equ:chebyshev-nystrom-formula}
@@
 \begin{equation}
-    \mtx{\Omega}^{\top} \underline{g}^{(m)}_{\sigma}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega} = \sum_{l=0}^{m} \mu_l(t) (\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}),
+    \mtx{\Omega}^{\top} \underline{g}^{(m)}_{\sigma}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega} = \sum_{l=0}^{m} \underline{\mu}_l(t) (\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}),
 \end{equation}
-meaning we can first precompute the small $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ for $l=0, \dots, m$ using the Chebyshev recurrence \refequ{equ:chebyshev-recurrence} and then inexpensively sum them up with the corresponding coefficients $\mu_l(t)$ for all $t$ in which we want to evaluate the approximation.
+meaning we can first precompute the small $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ for $l=0, \dots, m$ using the Chebyshev recurrence \refequ{equ:chebyshev-recurrence} and then inexpensively sum them up with the corresponding coefficients $\underline{\mu}_l(t)$ for all $t$ in which we want to evaluate the approximation.
 Thanks to the fast and exact squaring of Chebyshev expansions (cf.
@@
     \Statex \textbf{Parameters:} Even degree of expansion $m \in \mathbb{N}$, number of Girard-Hutchinson queries $n_{\mtx{\Psi}} \in \mathbb{N}_0$, Nyström sketch size $n_{\mtx{\Omega}} \in \mathbb{N}_0$,  smoothing parameter $\sigma > 0$
-    \Statex \textbf{Output:} Approximate evaluations of the spectral density $\{\widetilde{\phi}_{\sigma}(t_i)\}_{i=1}^{n_t}$
-    \State Compute $\{\mu_l(t_i)\}_{l=0}^{m}$ for all $t_i$ with non-negative Chebyshev expansion \refequ{equ:non-negative-chebyshev-expansion}
+    \Statex \textbf{Output:} Approximate evaluations of the spectral density $\{\underline{\phi}_{\sigma}^{(m)}(t_i)\}_{i=1}^{n_t}$
+    \State Compute $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ for all $t_i$ with non-negative Chebyshev expansion \refequ{equ:non-negative-chebyshev-expansion}
     \State Compute $\{\nu_l(t_i)\}_{l=0}^{2m}$ using \refalg{alg:chebyshev-squaring}%\refalg{alg:2-chebyshev-chebyshev-expansion}
@@
         \If {$l \leq m$}
-            \State $\mtx{K}_1(t_i) \gets \mtx{K}_1(t_i) + \mu_l(t_i) \mtx{X}$ \Comment{assemble $\mtx{\Omega}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$}
-            \State $\mtx{L}_1(t_i) \gets \mtx{L}_1(t_i) + \mu_l(t_i) \mtx{Y}$ \Comment{assemble $\mtx{\Omega}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi}$}
-            \State $\ell(t_i) \gets \ell(t_i) + \mu_l(t_i) z$ \Comment{assemble $\Trace(\mtx{\Psi}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi})$}
+            \State $\mtx{K}_1(t_i) \gets \mtx{K}_1(t_i) + \underline{\mu}_l(t_i) \mtx{X}$ \Comment{assemble $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$}
+            \State $\mtx{L}_1(t_i) \gets \mtx{L}_1(t_i) + \underline{\mu}_l(t_i) \mtx{Y}$ \Comment{assemble $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi}$}
+            \State $\ell(t_i) \gets \ell(t_i) + \underline{\mu}_l(t_i) z$ \Comment{assemble $\Trace(\mtx{\Psi}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi})$}
         \EndIf
-        \State $\mtx{K}_2(t_i) \gets \mtx{K}_2(t_i) + \nu_l(t_i) \mtx{X}$ \Comment{assemble $\mtx{\Omega}^{\top} g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})^2 \mtx{\Omega}$}
+        \State $\mtx{K}_2(t_i) \gets \mtx{K}_2(t_i) + \nu_l(t_i) \mtx{X}$ \Comment{assemble $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})^2 \mtx{\Omega}$}
       \EndFor
@@
     \For {$i = 1, \dots, n_t$}
-      \State $\widetilde{\phi}_{\sigma}(t_i) \gets \frac{1}{n} \Trace\left( \mtx{K}_1(t_i)^{\dagger}\mtx{K}_2(t_i) \right) + \frac{1}{n n_{\mtx{\Psi}}} \left( \ell(t_i) + \Trace\left( \mtx{L}_1(t_i)^{\top} \mtx{K}_1(t_i)^{\dagger} \mtx{L}_1(t_i) \right)  \right) $ \label{lin:4-nystromchebyshev-nystrom-pp}
+      \State $\underline{\phi}_{\sigma}(t_i) \gets \frac{1}{n} \Trace\left( \mtx{K}_1(t_i)^{\dagger}\mtx{K}_2(t_i) \right) + \frac{1}{n n_{\mtx{\Psi}}} \left( \ell(t_i) + \Trace\left( \mtx{L}_1(t_i)^{\top} \mtx{K}_1(t_i)^{\dagger} \mtx{L}_1(t_i) \right)  \right) $ \label{lin:4-nystromchebyshev-nystrom-pp}
     \EndFor
@@
 Notice that with $\mtx{K}_1(t_i)$ -- a matrix we need to assemble anyway -- we can cheaply compute $\Trace(\mtx{K}_1(t_i)) / n_{\mtx{\Omega}}$, which is the $n_{\mtx{\Omega}}$-query Girard-Hutchinson estimator.
-We may use it as a rough indicator whether $t_i$ lies within a region where the spectral density nearly vanishes, and subsequently set $\widetilde{\phi}_{\sigma}(t_i) = 0$ instead of computing the full expression on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
+We may use it as a rough indicator whether $t_i$ lies within a region where the spectral density nearly vanishes, and subsequently set $\underline{\phi}^{(m)}_{\sigma}(t_i) = 0$ instead of computing the full expression on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
 
@@
     Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, \emph{even} $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(\delta^{-1})^2)$, and smoothing parameter $0 < \sigma \ll 1$ satisfies with probability at least $1 - \delta$
+Then the Chebyshev-Nyström++ estimate $\underline{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, \emph{even} $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(\delta^{-1})^2)$, and smoothing parameter $0 < \sigma \ll 1$ satisfies with probability at least $1 - \delta$
     \begin{equation}
-        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq \varepsilon
+        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq \varepsilon
         \label{equ:chebyshev-nystrom-error}
@@
     \begin{align}
-        &\int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \notag \\
-        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \int_{-1}^{1} \left| \phi^{(m)}_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
+        &\int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \notag \\
+        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \int_{-1}^{1} \left| \phi^{(m)}_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
         &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \widetilde{\varepsilon} \int_{-1}^{1} \left| \phi^{(m)}_{\sigma}(t) \right|~\mathrm{d}t && \text{(\refthm{thm:nystrom-pp} w.p.
@@
     \begin{equation}
-        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq (1+\widetilde{\varepsilon}) 2 \widetilde{\varepsilon} + \widetilde{\varepsilon} \leq 5 \widetilde{\varepsilon}
+        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq (1+\widetilde{\varepsilon}) 2 \widetilde{\varepsilon} + \widetilde{\varepsilon} \leq 5 \widetilde{\varepsilon}
     \end{equation}

### 2025-02-19 paper/numresults.tex
review of section 2 finished
change symbols and write todo list for results
restructure Chebyshev expansion and adjust notation

@@
 \label{sec:results}
+
+\color{black}
+\textcolor{red}{
+To do in this section:
+\begin{itemize}
+    \item Specify implementation details (no non-negative Chebyshev expansion, speed-up tricks, thresholding, ...)
+    \item Clarify where code come from (faithful implementation, own code, ...)
+    \item Give experimental setup (software version, hardware, single-threaded, ...)
+\end{itemize}
+}
 

### 2025-02-19 paper/spectraldensity.tex
restructure Chebyshev expansion and adjust notation

@@
 \end{equation}
-\textcolor{red}{??? MOVE BELOW! This is not at a good place
-For later purposes, we recall the well-known
-three-term recurrence relation
-\begin{equation}
-    T_0(s) = 1,\quad  T_1(s) = s,\quad 
-    T_l(s) = 2 s T_{l-1}(s) - T_{l-2}(s) \text{ for $l \geq 2$.}
-    \label{equ:chebyshev-recurrence}
-\end{equation}}
 Inserting~\refequ{equ:matrix-expansion} into~\refequ{equ:smooth-spectral-density} yields the approximate smoothed spectral density
@@
 It now also allows us to bound the $L^1$-error we make when expanding the spectral density.
+\begin{align}
+    \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t 
+    &= \int_{-1}^{1} \left| \frac{1}{n} \sum_{i=1}^n \left(g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i)\right) \right|~\mathrm{d}t \notag \\
+    &\leq \int_{-1}^{1} \max_{i = 1, \dots, n} \left| g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i) \right|~\mathrm{d}t \notag \\
+    &\leq 2 E_{\sigma, m}
+    \label{equ:chebyshev-interpolation-spectral-density}
+\end{align}
 
-\begin{theorem}{$L^1$-error of Chebyshev expansion of spectral density}{chebyshev-error}
-    Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
+%\begin{theorem}{$L^1$-error of Chebyshev expansion of spectral density}{chebyshev-error}
+%    Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
 Then the spectal density with Chebyshev expansion $\phi_{\sigma}^{(m)}$ satisfies
-    \begin{equation}
-        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq 2 E_{\sigma, m}
-        \label{equ:chebyshev-interpolation-spectral-density}
-        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
-        %\label{equ:2-chebyshev-interpolation-error}
-    \end{equation}
-    for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and the error $E_{\sigma, m}$ from \reflem{lem:chebyshev-error}.%, and $t \in \mathbb{R}$.
+%    \begin{equation}
+%        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq 2 E_{\sigma, m}
+%        \label{equ:chebyshev-interpolation-spectral-density}
+%        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
+%        %\label{equ:2-chebyshev-interpolation-error}
+%    \end{equation}
+%    for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and the error $E_{\sigma, m}$ from \reflem{lem:chebyshev-error}.%, and $t \in \mathbb{R}$.
 %The same result holds for the expansion of a non-negative Chebyshev expansion of \emph{even} degree $m \in \mathbb{N}$ but with $\underline{E}_{\sigma, m}$.
-\end{theorem}
-
-\begin{proof}
-    We observe that for all $t \in [-1, 1]$
-    \begin{align}
-        &\left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right| \notag \\
-        &= \left| \frac{1}{n} \Trace(g_{\sigma}(t\mtx{I}_n - \mtx{A})) - \frac{1}{n} \Trace(g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})) \right|
-        && \text{(definitions \refequ{equ:smooth-spectral-density} and \refequ{equ:expanded-spectral-density})} \notag \\
-        &= \left| \frac{1}{n} \sum_{i=1}^n \left(g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i)\right) \right|
-        && \text{($\lambda_1, \dots, \lambda_n$ eigenvalues of $\mtx{A}$)} \notag \\
-        &\leq \max_{i = 1, \dots, n} \left| g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i) \right|
-        && \text{(conservative upper bound)} \notag \\
-        &\leq E_{\sigma, m}.% ~ \text{(or $\underline{E}_{\sigma, m}$)}.
-        && \text{(using \reflem{lem:chebyshev-error})} \notag \\
-    \end{align}
-    Then, Hölder's inequality allows us to conclude
-    \begin{equation}
-        \int_{-1}^{1} | \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) | ~\mathrm{d}t
-        \leq 2 \sup_{t \in [-1, 1]} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|
-        \leq 2 E_{\sigma, m}.% ~ \text{(or $2 \underline{E}_{\sigma, m}$)}.
-    \end{equation}
-\end{proof}
+%\end{theorem}
+%
+%\begin{proof}
+%    We observe that for all $t \in [-1, 1]$
+%    \begin{align}
+%        &\left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right| \notag \\
+%        &= \left| \frac{1}{n} \Trace(g_{\sigma}(t\mtx{I}_n - \mtx{A})) - \frac{1}{n} \Trace(g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})) \right|
+%        && \text{(definitions \refequ{equ:smooth-spectral-density} and \refequ{equ:expanded-spectral-density})} \notag \\
+%        &= \left| \frac{1}{n} \sum_{i=1}^n \left(g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i)\right) \right|
+%        && \text{($\lambda_1, \dots, \lambda_n$ eigenvalues of $\mtx{A}$)} \notag \\
+%        &\leq \max_{i = 1, \dots, n} \left| g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i) \right|
+%        && \text{(conservative upper bound)} \notag \\
+%        &\leq E_{\sigma, m}.% ~ \text{(or $\underline{E}_{\sigma, m}$)}.
+%        && \text{(using \reflem{lem:chebyshev-error})} \notag \\
+%    \end{align}
+%    Then, Hölder's inequality allows us to conclude
+%    \begin{equation}
+%        \int_{-1}^{1} | \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) | ~\mathrm{d}t
+%        \leq 2 \sup_{t \in [-1, 1]} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|
+%        \leq 2 E_{\sigma, m}.% ~ \text{(or $2 \underline{E}_{\sigma, m}$)}.
+%    \end{equation}
+%\end{proof}
 
@@
 To still apply the bound from \refthm{thm:nystrom-pp} in this scenario, we will have to ensure that the Chebyshev expansion is non-negative.
-We can think of many ways in which this can be achieved: by using Jackson damping \cite{jackson-1912-approximation-trigonometric,braverman-2022-sublinear-time}, albeit at a significantly worse rate of convergence; by shifting up the expansion $g_{\sigma}^{(m)} + \rho$ with a large enough $\rho > 0$, at the loss of some of the low-rank structure of $g_{\sigma}(t \mtx{I}_n - \mtx{A})$; or by directly minimizing the approximation error over all non-negative Chebyshev polynomials \cite{fejer-1916-uber-trigonometrische}, which realistically only works for small $m$.
-In the end, we propose the following method to produce a non-negative Chebyshev expansion: Fix an even $m$ and expand the square root of the smoothing kernel in a basis of Chebyshev polynomials up to degree $m/2$
+We propose the following method to produce a non-negative Chebyshev expansion \textcolor{red}{[REFERENCE NEEDED]}: Fix an even $m$ and expand the square root of the smoothing kernel in a basis of Chebyshev polynomials up to degree $m/2$
 \begin{equation}
     \sqrt{g_{\sigma}}^{(\sfrac{m}{2})}(t - s) = \sum_{l=0}^{m/2} \xi_l(t) T_l(s).
+    \label{equ:square-root-chebyshev-expansion}
 \end{equation}
@@
 \begin{equation}
-    \underline{g}^{(m)}_{\sigma}(t - s) = \sum_{l=0}^{m} \underline{\mu}_l(t) T_l(s).
+    \underline{g}_{\sigma}^{(m)}(t - s) = \sum_{l=0}^{m} \underline{\mu}_l(t) T_l(s).
     \label{equ:non-negative-chebyshev-expansion}
 \end{equation}
-With a complexity of $\mathcal{O}(m \log(m))$ this procedure is computationally inexpensive and allows for very similar convergence bounds to \reflem{lem:chebyshev-error}, which we will derive in \refsec{subsubsec:chebyshev-nystrom-analysis}.
-In \reftab{tab:chebyshev-timing-interpolation}, the non-negative expansion can be seen to only be slightly slower than the normal one.
 
-\begin{table}[ht]
-    \caption{We compare the runtime in milliseconds for computing the coefficients of the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ \refequ{equ:smoothing-kernel} in three different ways: as proposed in \cite[Algorithm 1]{lin-2017-randomized-estimation} (\emph{quadrature FFT}), using the discrete cosine transform (\emph{DCT}) \refequ{equ:discrete-cosine-transform}, and computing a non-negative Chebyshev expansion with \refalg{alg:chebyshev-squaring} (\emph{non-negative DCT}).
-We use $\sigma=0.005$, $n_t=1000$ parameter values, and various degrees $m$.
-For stability, we average over 7 runs of the algorithms and repeat these runs 1000 times to form the mean and standard deviation which are given in the below table.}
-    \label{tab:chebyshev-timing-interpolation}
-   \input{tables/interpolation.tex}
-\end{table}
+??? UNTIL HERE ???
 
+%With a complexity of $\mathcal{O}(m \log(m))$ this procedure is computationally inexpensive and allows for very similar convergence bounds to \reflem{lem:chebyshev-error}, which we will derive \textcolor{red}{in the following?} in \refsec{subsubsec:chebyshev-nystrom-analysis}.
+
+\color{black}
+
+%To this extent, we will first analyze the non-negative Chebyshev expansion.
+
+The resulting non-negative version of the approximate smoothed spectral density \refequ{equ:expanded-spectral-density}
+\begin{equation}
+    \phi_{\sigma}(t) \approx \underline{\phi}_{\sigma}^{(m)}(t) = \frac{1}{n} \Trace(\underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}))
+    \label{equ:non-negative-expanded-density}
+\end{equation}
+then satisfies the following bound:
+
+\begin{lemma}{$L^1$-error of non-negative approximate smooth spectral density}{non-negative-chebyshev-error}
+Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
+Then the non-negative approximate smoothed spectral density $\underline{\phi}_{\sigma}^{(m)}$ satisfies
+    \begin{equation}
+        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq 4 \sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right) E_{\sqrt{2}\sigma, \sfrac{m}{2}} \equiv \underline{E}_{\sigma, m}
+        \label{equ:chebyshev-interpolation-sup-error-kernel-nonneg}
+        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
+        %\label{equ:2-chebyshev-interpolation-error}
+    \end{equation}
+    for all \emph{even} degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and with $E_{\sigma, m}$ as defined in \reflem{lem:chebyshev-error}.
+\end{lemma}
+
+\begin{proof}
+    For any numbers $a, b \in \mathbb{R}$ it holds $| a^2 - b^2 | \leq (2 | a | + | a - b |)  | a - b |$. Therefore,
+    \begin{equation}
+        \left| g_{\sigma} - \underline{g}_{\sigma}^{(m)} \right| = \left| (\sqrt{g_{\sigma}})^2 - (\sqrt{g_{\sigma}}^{(\sfrac{m}{2})})^2 \right| \leq \left( 2 \left| \sqrt{g_{\sigma}} \right| + \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right| \right) \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right|.
+    \end{equation}
+    Since $\sqrt{g_{\sigma}} = \sqrt{2 \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$, we can apply \reflem{lem:chebyshev-error} with $\sigma \to \sqrt{2}\sigma$ and $m \to m/2$ along with the bound $|\sqrt{g_{\sigma}}| \leq 1/\sqrt{\sigma \sqrt{2 \pi}}$ to get
+    \begin{align}
+        &\sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \notag \\
+        &\leq \left( 2 \frac{1}{\sqrt{ \sigma \sqrt{2\pi}}} + \sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right)\sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}} \notag \\
+        &= 2\sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right) E_{\sqrt{2}\sigma, \sfrac{m}{2}}.
+    \end{align}
+    An identical treatment to \refequ{equ:chebyshev-interpolation-spectral-density} then leads to the statement of the lemma.
+\end{proof}
+
+On one hand, in the non-negative expansion of degree $m$ \refequ{equ:non-negative-chebyshev-expansion}, the underlying Chebyshev expansion -- whose error is shown in \reflem{lem:chebyshev-error} to exponentially decay with $m$ -- is only computed to half the degree (i.e.
+$m/2$) of the standard Chebyshev expansion, hence, usually not as accurate.
+On the other hand, in the non-negative expansion we approximate $\sqrt{g_{\sigma}} \propto g_{\sqrt{2}\sigma}$ which is less \enquote{pointy} than $g_{\sigma}$ and therefore easier to approximate with Chebyshev polynomials.
+Overall, we will need to choose the degree $m$ of the non-negative expansion roughly a factor of $\sqrt{2}$ larger to match the approximation error of the standard expansion (cf.
+\reffig{fig:interpolation-issue}).
+
+%In the following, we will show how the coefficients $\underline{\mu}_0(t), \dots, \underline{\mu}_m(t)$ of the non-negative Chebyshev expansion \refequ{equ:non-negative-chebyshev-expansion} can efficiently be computed and that this expansion exhibits similar convergence properties to the standard Chebyshev expansion.
 %\todo{
@@
 
-\subsection{Computation of expansion coefficients}
+%\subsection{Computation of expansion coefficients}
 
-\subsubsection{Fast and consistent squaring}
+\subsubsection{Computation of Chebyshev coefficients}
 \label{subsubsec:dct}
-
-??? UNTIL HERE ???
-\color{black}
 
@@
 \end{equation}
-between the expansion coefficients $\mu_l(t),~i=0,\dots,m$ and the function evaluations at the Chebyshev nodes of the second kind $s_i = \cos(\pi i / m),~i=0,\dots,m$.
+between the expansion coefficients $\mu_l(t),~i=0,\dots,m$ and the function evaluations at the Chebyshev nodes of the second kind $s_i = \cos(\pi i / m),~i=0,\dots,m$ \cite{vanloan-1992-computational-frameworks, baszenski-1997-fast-polynomial}.
 This can be accomplished in $\mathcal{O}(m \log(m))$ operations.
-\cite[Algorithm 1]{lin-2017-randomized-estimation} uses a slightly different approach, which only approximates the expansion coefficients and is noticeably slower (cf.
-\reftab{tab:chebyshev-timing-interpolation}).
 
-A highly useful property of the Chebyshev expansion --- of which we will make use twice in the following pages --- is the fact that we can square such an expansion in $\mathcal{O}(m \log(m))$ operations \cite{baszenski-1997-fast-polynomial}, which the below algorithm implements.
+A desirable property of the Chebyshev expansion is the fact that such an expansion can be squared in $\mathcal{O}(m \log(m))$ operations by using the DCT and its inverse \cite{baszenski-1997-fast-polynomial}.
+This allows us to efficiently compute the non-negative Chebyshev expansion \refequ{equ:non-negative-chebyshev-expansion}, and will also be useful in \refsec{subsec:chebyshev-nystrom}.
+The below algorithm implements this observation.
 
@@
     \State Compute $\vct{f} = \operatorname{DCT}(\vct{\mu})$
-    \State Compute $\vct{\nu} = \operatorname{DCT}^{-1}(\vct{f} \odot \vct{f})$ whose entries are the coefficients $\nu_l$
+    \State Compute $\vct{\nu} = \operatorname{DCT}^{-1}(\vct{f} \odot \vct{f})$ whose entries are the coefficients $\nu_0, \dots, \nu_{2m}$
     \label{lin:inverse-DCT}
@@
 
+In \cite{lin-2017-randomized-estimation}, a slightly different approach to computing the expansion coefficients in \refequ{equ:discrete-cosine-transform} is taken.
+They express the coefficients $\mu_l(t)$ in terms of an integral, which they subsequently approximate with a quadrature supported in at least $2m + 1$ nodes.
+They evaluate the resulting approximation with the fast Fourier transform (FFT).
+Besides not yielding the exact expansion coefficients $\mu_l(t)$, this procedure is also significantly slower due to the larger number of quadrature nodes and the implementational benefits of the DCT over the FFT.
+In \reftab{tab:chebyshev-timing-interpolation}, we compare a benevolent implementation of \cite[Algorithm 1]{lin-2017-randomized-estimation} with our approach, both for the standard and the non-negative Chebyshev expansion.
+
+\begin{table}[ht]
+    \caption{We compare the runtime in milliseconds for computing the coefficients of the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ \refequ{equ:smoothing-kernel} in three different ways: as proposed in \cite[Algorithm 1]{lin-2017-randomized-estimation} (\emph{FFT}), using the discrete cosine transform (\emph{DCT}) \refequ{equ:discrete-cosine-transform}, and using the discrete cosine transform to compute a non-negative Chebyshev expansion with \refalg{alg:chebyshev-squaring} (\emph{non-negative DCT}).
+We use $\sigma=0.005$, $n_t=1000$ parameter values, and various degrees $m$.
+For stability, we average over 7 runs of the algorithms and repeat these runs 1000 times to form the mean and standard deviation which are given in the below table.}
+    \label{tab:chebyshev-timing-interpolation}
+   \input{tables/interpolation.tex}
+\end{table}
+
 \subsection{Chebyshev-Nyström++ method for spectral density approximation}
@@
 
-In the end, we propose the following approximation for the smoothed spectral density
+Combining the results from \refsec{sec:analysis} for $\mtx{B}(t) = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ with the Chebyshev expansion framework from \refsec{subsec:chebyshev-expansion}, we now approximate the smoothed spectral density with
 \begin{equation}
-    \underline{\phi}_{\sigma}^{(m)}(t) = \frac{1}{n} \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}))
+    \phi_{\sigma}(t) \approx \widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) = \frac{1}{n} \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})),
     \label{equ:chebyshev-nystrom-formula}
@@
 
-An efficient implementation of the approximation \refequ{equ:chebyshev-nystrom-formula} can be achieved thanks to the invariance of the trace under cyclic permutation of its arguments and the recurrence relation \refequ{equ:chebyshev-recurrence} which the Chebyshev polynomials satisfy.
+An efficient implementation of the approximation \refequ{equ:chebyshev-nystrom-formula} can be achieved thanks to the invariance of the trace under cyclic permutation of its arguments and the three-term recurrence relation
+\begin{equation}
+    T_0(s) = 1,\quad  T_1(s) = s,\quad 
+    T_l(s) = 2 s T_{l-1}(s) - T_{l-2}(s) \text{ for $l \geq 2$,}
+    \label{equ:chebyshev-recurrence}
+\end{equation}
+which the Chebyshev polynomials satisfy.
 To illustrate this, consider the approximation \refequ{equ:chebyshev-nystrom-formula} when $n_{\mtx{\Psi}}$ is zero, i.e.
@@
 \begin{equation}
-    \frac{1}{n} \Trace((\mtx{\Omega}^{\top} \underline{g}^{(m)}_{\sigma}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega})^{\dagger} (\mtx{\Omega}^{\top} \underline{g}^{(m)}_{\sigma}(t \mtx{I}_n - \mtx{A})^2 \mtx{\Omega})).
+    \frac{1}{n} \Trace \big((\mtx{B}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} (\mtx{B}(t) \mtx{\Omega})^{\top} \big) =
+    \frac{1}{n} \Trace \big((\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega}}_{= \mtx{K}_1(t)})^{\dagger} (\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t)^2 \mtx{\Omega}}_{= \mtx{K}_2(t)}) \big).
+    %\frac{1}{n} \Trace((\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega})^{\dagger} (\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})^2 \mtx{\Omega})).
     \label{equ:cyclic-property}
 \end{equation}
-Thanks to the affine form of the Chebyshev expansion, the first parenthesized term in the trace is
+Thanks to the affine linear form of the Chebyshev expansion, we have
 \begin{equation}
-    \mtx{\Omega}^{\top} \underline{g}^{(m)}_{\sigma}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega} = \sum_{l=0}^{m} \underline{\mu}_l(t) (\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}),
+    \mtx{K}_1(t) = \mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega} = \sum_{l=0}^{m} \underline{\mu}_l(t) (\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}),
 \end{equation}
 meaning we can first precompute the small $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ for $l=0, \dots, m$ using the Chebyshev recurrence \refequ{equ:chebyshev-recurrence} and then inexpensively sum them up with the corresponding coefficients $\underline{\mu}_l(t)$ for all $t$ in which we want to evaluate the approximation.
-Thanks to the fast and exact squaring of Chebyshev expansions (cf.
-\refalg{alg:chebyshev-squaring}), the second parenthesized term in \refequ{equ:cyclic-property} can be approximated in an equivalent way.
-Unlike expanding the squared smoothing kernel $g_{\sigma}^2$ separately \cite{lin-2017-randomized-estimation}, this approach is consistent, which we have observed to make a noticeable difference in terms of accuracy (cf.
-\reffig{fig:interpolation-issue}), and which enables provability.
+The fast and exact squaring of Chebyshev expansions (cf.
+\refalg{alg:chebyshev-squaring}), allows us to also express $\underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})^2 = \sum_{l=0}^{2m} \underline{\nu}_l(t) T_l(\mtx{A})$ in terms of an affine linear sum of the small matrices $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ for $l=0, \dots, 2m$, such that the matrix $\mtx{K}_2(t)$ can be built in an analogous way to $\mtx{K}_1(t)$.
+Alternatively, \cite{lin-2017-randomized-estimation} proposes to expand the squared smoothing kernel $g_{\sigma}^2$ in a separate Chebyshev expansion of degree $2m$.
+While this approximation of $g_{\sigma}^2$ might be more accurate, it is \emph{inconsistent} in the sense that computing a Chebyshev expansion of $g_{\sigma}^2$ is in general not the same as squaring an existing Chebyshev expansion of $g_{\sigma}$.
+We have observed the consistency between the approximation of $g_{\sigma}$ and $g_{\sigma}^2$ to make a crucial difference in terms of accuracy (cf.
+\reffig{fig:interpolation-issue}).
+The consistency also allows bounds to be derived for the estimator \refequ{equ:chebyshev-nystrom-formula}.
 
@@
     \input{plots/interpolation.pgf}
-    \caption{Difference in the accuracy when computing the estimate \refequ{equ:cyclic-property} once by separately expanding the squared smoothing kernel $g_{\sigma}^2$ (\emph{inconsistent}), once by squaring the Chebyshev expansion $g_{\sigma}^{(m)}$ using \refalg{alg:chebyshev-squaring} (\emph{consistent}), and once with a non-negative Chebyshev expansion $\underline{g}_{\sigma}^{(m)}$ \refequ{equ:non-negative-chebyshev-expansion} which is also squared with \refalg{alg:chebyshev-squaring} (\emph{non-negative}).
+    \caption{Difference in the accuracy when computing the estimate \refequ{equ:cyclic-property} once by separately expanding the squared smoothing kernel $g_{\sigma}^2$ like \cite{lin-2017-randomized-estimation} (\emph{inconsistent}), once by squaring the Chebyshev expansion $g_{\sigma}^{(m)}$ using \refalg{alg:chebyshev-squaring} (\emph{consistent}), and once with a non-negative Chebyshev expansion $\underline{g}_{\sigma}^{(m)}$ \refequ{equ:non-negative-chebyshev-expansion} which is also squared with \refalg{alg:chebyshev-squaring} (\emph{non-negative}).
 We use a small example matrix from \refsec{sec:results}, fix the size of the Nyström approximation to $n_{\mtx{\Omega}} = 80$, let $\sigma = 0.005$, and compute the $L^1$ error from the approximation of the spectral density at $n_t = 100$ uniformly spaced values of $t$.}
@@
 
-Similar observations for the case when $n_{\mtx{\Psi}}$ is nonzero lead to the following algorithm.
-\textcolor{red}{??? Lines 1 and 2 are quite confusing. Say more explicitly how $\mu$ is computed. Is Algorithm 3.3 applied twice? ???}
+Analogous observations for the \enquote{Girard-Hutchinson} term which appears in $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$ when $n_{\mtx{\Psi}}$ is nonzero show that also then, the estimator can be expressed in terms of small quadratic forms involving the matrix $\mtx{B}(t) = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$:
+\begin{align}
+    &\frac{1}{n n_{\mtx{\Psi}}} \Trace \big(\mtx{\Psi}^{\top} (\mtx{B}(t) - (\mtx{B}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} (\mtx{B}(t) \mtx{\Omega})^{\top}) \mtx{\Psi}^{\top} \big) \notag \\
+    &= \frac{1}{n n_{\mtx{\Psi}}} \big( \underbrace{\Trace(\mtx{\Psi}^{\top} \mtx{B}(t)\mtx{\Psi})}_{= \ell(t)} - \Trace((\underbrace{\mtx{\Psi}^{\top} \mtx{B}(t) \mtx{\Omega}}_{= \mtx{L}_1(t)^{\top}}) (\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega}}_{= \mtx{K}_1(t)^{\top}})^{\dagger} (\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Psi}^{\top}}_{= \mtx{L}_1(t)})) \big).
+    \label{equ:cyclic-property-hutchinson}
+\end{align}
+
+An efficient implementation of the whole procedure described in this section can be found in \refalg{alg:nystrom-chebyshev-pp}.
 
@@
     \Statex \textbf{Parameters:} Even degree of expansion $m \in \mathbb{N}$, number of Girard-Hutchinson queries $n_{\mtx{\Psi}} \in \mathbb{N}_0$, Nyström sketch size $n_{\mtx{\Omega}} \in \mathbb{N}_0$,  smoothing parameter $\sigma > 0$
-    \Statex \textbf{Output:} Approximate evaluations of the spectral density $\{\underline{\phi}_{\sigma}^{(m)}(t_i)\}_{i=1}^{n_t}$
-    \State Compute $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ for all $t_i$ with non-negative Chebyshev expansion \refequ{equ:non-negative-chebyshev-expansion}
-    \State Compute $\{\nu_l(t_i)\}_{l=0}^{2m}$ using \refalg{alg:chebyshev-squaring}%\refalg{alg:2-chebyshev-chebyshev-expansion}
+    \Statex \textbf{Output:} Approximate evaluations of the spectral density $\{\widetilde{\underline{\phi}}_{\sigma}^{(m)}(t_i)\}_{i=1}^{n_t}$
+    \For {$i = 1, \dots, n_t$}
+        \State Compute the coefficients $\{\xi_l(t_i)\}_{l=0}^{m/2}$ of the Chebyshev expansion of $\sqrt{g_{\sigma}}$ \refequ{equ:square-root-chebyshev-expansion}
+        \State Compute the coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ of the non-negative Chebyshev expansion by applying \refalg{alg:chebyshev-squaring} to $\{\xi_l(t_i)\}_{l=0}^{m/2}$
+        \State Compute the coefficients $\{\underline{\nu}_l(t_i)\}_{l=0}^{2m}$ of the square of the non-negative Chebyshev expansion by applying \refalg{alg:chebyshev-squaring} to $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$
+    \EndFor
+    %\State Compute the coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ of the non-negative Chebyshev expansion \refequ{equ:non-negative-chebyshev-expansion} for all points $t_i$ \textcolor{red}{maybe more details needed (maybe compute $\xi$ for m/2, then twice Alg 3.3)}
+    %\State Compute the coefficients $\{\nu_l(t_i)\}_{l=0}^{2m}$ of the square of the non-negative Chebyshev expansion with coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ using \refalg{alg:chebyshev-squaring}%\refalg{alg:2-chebyshev-chebyshev-expansion}
     %\State Compute  for all $t_i$ using %\refalg{alg:3-nystrom-chebyshev-exponentiation}
@@
         \If {$l \leq m$}
-            \State $\mtx{K}_1(t_i) \gets \mtx{K}_1(t_i) + \underline{\mu}_l(t_i) \mtx{X}$ \Comment{assemble $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$}
-            \State $\mtx{L}_1(t_i) \gets \mtx{L}_1(t_i) + \underline{\mu}_l(t_i) \mtx{Y}$ \Comment{assemble $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi}$}
-            \State $\ell(t_i) \gets \ell(t_i) + \underline{\mu}_l(t_i) z$ \Comment{assemble $\Trace(\mtx{\Psi}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi})$}
+            \State $\mtx{K}_1(t_i) \gets \mtx{K}_1(t_i) + \underline{\mu}_l(t_i) \mtx{X}$ \Comment{assembles $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$}
+            \State $\mtx{L}_1(t_i) \gets \mtx{L}_1(t_i) + \underline{\mu}_l(t_i) \mtx{Y}$ \Comment{assembles $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi}$}
+            \State $\ell(t_i) \gets \ell(t_i) + \underline{\mu}_l(t_i) z$ \Comment{assembles $\Trace(\mtx{\Psi}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi})$}
         \EndIf
-        \State $\mtx{K}_2(t_i) \gets \mtx{K}_2(t_i) + \nu_l(t_i) \mtx{X}$ \Comment{assemble $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})^2 \mtx{\Omega}$}
+        \State $\mtx{K}_2(t_i) \gets \mtx{K}_2(t_i) + \underline{\nu}_l(t_i) \mtx{X}$ \Comment{assembles $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})^2 \mtx{\Omega}$}
       \EndFor
@@
     \For {$i = 1, \dots, n_t$}
-      \State $\underline{\phi}_{\sigma}(t_i) \gets \frac{1}{n} \Trace\left( \mtx{K}_1(t_i)^{\dagger}\mtx{K}_2(t_i) \right) + \frac{1}{n n_{\mtx{\Psi}}} \left( \ell(t_i) + \Trace\left( \mtx{L}_1(t_i)^{\top} \mtx{K}_1(t_i)^{\dagger} \mtx{L}_1(t_i) \right)  \right) $ \label{lin:4-nystromchebyshev-nystrom-pp}
+      \State $\widetilde{\underline{\phi}}_{\sigma}(t_i) \gets \frac{1}{n} \Trace\left( \mtx{K}_1(t_i)^{\dagger}\mtx{K}_2(t_i) \right) + \frac{1}{n n_{\mtx{\Psi}}} \left( \ell(t_i) + \Trace\left( \mtx{L}_1(t_i)^{\top} \mtx{K}_1(t_i)^{\dagger} \mtx{L}_1(t_i) \right)  \right) $ \label{lin:4-nystromchebyshev-nystrom-pp}
     \EndFor
@@
 regions where the spectral density nearly vanishes.
-For small smoothing parameters $\sigma$, the rapid decay of the Gaussian $g_{\sigma}$ will cause $\mtx{K}_1(t_i) \approx \mtx{\Omega}^{\top} g_{\sigma}(t_i\mtx{I}_n - \mtx{A}) \mtx{\Omega}$ to be close to the zero matrix for parameter values $t_i$ located in such regions.
+For small smoothing parameters $\sigma$, the rapid decay of the Gaussian $g_{\sigma}$ will cause $\mtx{K}_1(t) \approx \mtx{\Omega}^{\top} g_{\sigma}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$ to be close to the zero matrix for parameter values $t$ located in such regions.
 Hence, on \reflin{lin:4-nystromchebyshev-nystrom-pp} we would compute the pseudoinverse of an almost zero matrix, which can be extremely unstable (see \reffig{fig:zerocheck}).
-Notice that with $\mtx{K}_1(t_i)$ -- a matrix we need to assemble anyway -- we can cheaply compute $\Trace(\mtx{K}_1(t_i)) / n_{\mtx{\Omega}}$, which is the $n_{\mtx{\Omega}}$-query Girard-Hutchinson estimator.
-We may use it as a rough indicator whether $t_i$ lies within a region where the spectral density nearly vanishes, and subsequently set $\underline{\phi}^{(m)}_{\sigma}(t_i) = 0$ instead of computing the full expression on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
+Notice that with $\mtx{K}_1(t)$ -- a matrix we need to assemble anyway -- we can cheaply compute $\Trace(\mtx{K}_1(t)) / n_{\mtx{\Omega}}$, which is the $n_{\mtx{\Omega}}$-query Girard-Hutchinson estimator.
+We may use it as a rough indicator whether $t$ lies within a region where the spectral density nearly vanishes, and subsequently set $\widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) = 0$ instead of computing the full expression on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
 
@@
 
-In this section, we will upper bound the output of the Chebyshev-Nyström++ method (see. \refalg{alg:nystrom-chebyshev-pp}). To this extent, we will first analyze the non-negative Chebyshev expansion.
-
-\begin{lemma}{Non-negative Chebyshev expansion error}{non-negative-chebyshev-error}
-    The non-negative expansion $\underline{g}_{\sigma}^{(m)}$ of the Gaussian smoothing kernel satisfies
-    \begin{equation}
-        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \leq 2\sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right) E_{\sqrt{2}\sigma, \sfrac{m}{2}} \equiv \underline{E}_{\sigma, m}
-        \label{equ:chebyshev-interpolation-sup-error-kernel-nonneg}
-        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
-        %\label{equ:2-chebyshev-interpolation-error}
-    \end{equation}
-    for all \emph{even} degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, $s \in [-1, 1]$, and with $E_{\sigma, m}$ as defined in \reflem{lem:chebyshev-error}.
-\end{lemma}
-
-\begin{proof}
-    For any numbers $a, b \in \mathbb{R}$ it holds
-    \begin{equation}
-    | a^2 - b^2 | = | (a + b)(a - b) | = | a + b | | a - b | \leq (2 | a | + | a - b |)  | a - b |
-    \end{equation}
-    Therefore, omitting the arguments and using $g_{\sigma} = (\sqrt{g_{\sigma}})^2$ and $\underline{g}_{\sigma}^{(m)} = (\sqrt{g_{\sigma}}^{(\sfrac{m}{2})})^2$.
-    we have
-    \begin{equation}
-        \left| g_{\sigma} - \underline{g}_{\sigma}^{(m)} \right| \leq \left( 2 \left| \sqrt{g_{\sigma}} \right| + \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right| \right) \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right|.
-    \end{equation}
-    Since $\sqrt{g_{\sigma}} = \sqrt{2 \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$ and $\sqrt{g_{\sigma}} \leq 1/\sqrt{\sigma \sqrt{2 \pi}}$ \textcolor{red}{we can apply the result from \refthm{thm:chebyshev-error} ???
-the reference seems to be incorrect ???} to get
-    \begin{align}
-        &\sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \notag \\
-        &\leq \left( 2 \frac{1}{\sqrt{ \sigma \sqrt{2\pi}}} + \sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right)\sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}},
-    \end{align}
-    from which follows the result with some minor simplifications.
-
-\end{proof}
-
-On one hand, in the non-negative expansion, the underlying computes Chebyshev expansion -- which is shown in \reflem{lem:chebyshev-error} to exponentially decay with its degree -- is only computed to degree $m/2$, whereas the standard expansion is of degree $m$.
-On the other hand, we approximate $\sqrt{g_{\sigma}}$ instead of $g_{\sigma}$, which is less \enquote{pointy} and therefore better approximated by polynomials.
-Overall, we will need to choose the degree $m$ of the expansion approximately a factor of $\sqrt{2}$ larger to match the error of the standard expansion (cf.
-\reffig{fig:interpolation-issue}).
-
+In this section, we will upper bound the output of the Chebyshev-Nyström++ method (see. \refalg{alg:nystrom-chebyshev-pp}). 
 \begin{theorem}{Chebyshev-Nyström++ method}{chebyshev-nystrom}
     Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the Chebyshev-Nyström++ estimate $\underline{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, \emph{even} $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(\delta^{-1})^2)$, and smoothing parameter $0 < \sigma \ll 1$ satisfies with probability at least $1 - \delta$
+Then the Chebyshev-Nyström++ estimate $\widetilde{\underline{\phi}}_{\sigma}^{(m)}$ with \emph{even} degree $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, \emph{even} $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(\delta^{-1})^2)$, and smoothing parameter $0 < \sigma \ll 1$ satisfies with probability at least $1 - \delta$
     \begin{equation}
-        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq \varepsilon
+        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq \varepsilon
         \label{equ:chebyshev-nystrom-error}
@@
     \begin{align}
-        &\int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \notag \\
-        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \int_{-1}^{1} \left| \phi^{(m)}_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
-        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \widetilde{\varepsilon} \int_{-1}^{1} \left| \phi^{(m)}_{\sigma}(t) \right|~\mathrm{d}t && \text{(\refthm{thm:nystrom-pp} w.p.
+        &\int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \notag \\
+        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \int_{-1}^{1} \left| \underline{\phi}_{\sigma}^{(m)}(t) - \widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
+        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \widetilde{\varepsilon} \int_{-1}^{1} \left| \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t && \text{(\refthm{thm:nystrom-pp} w.p.
 $\geq 1 - \delta$)} \notag \\
-        &\leq (1 + \widetilde{\varepsilon}) \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \widetilde{\varepsilon} \int_{-1}^{1} \left| \phi_{\sigma}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
-        &\leq (1 + \widetilde{\varepsilon}) 2 \underline{E}_{\sigma, m} + \widetilde{\varepsilon}. && \text{(\refthm{thm:chebyshev-error}, normalization)} \notag \\
+        &\leq (1 + \widetilde{\varepsilon}) \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \widetilde{\varepsilon} \int_{-1}^{1} \left| \phi_{\sigma}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
+        &\leq (1 + \widetilde{\varepsilon}) \underline{E}_{\sigma, m} + \widetilde{\varepsilon}. && \text{(\reflem{lem:non-negative-chebyshev-error}, normalization)} \notag \\
     \end{align}
@@
     \begin{equation}
-        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq (1+\widetilde{\varepsilon}) 2 \widetilde{\varepsilon} + \widetilde{\varepsilon} \leq 5 \widetilde{\varepsilon}
+        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq (1+\widetilde{\varepsilon}) \widetilde{\varepsilon} + \widetilde{\varepsilon} \leq 3 \widetilde{\varepsilon}
     \end{equation}
-    from which the result follows by identifying $\varepsilon = 5 \widetilde{\varepsilon}$.
+    from which the result follows by identifying $\varepsilon = 3 \widetilde{\varepsilon}$.
 \end{proof}

### 2025-02-19 paper/plots/interpolation.py

@@ -26,7 +26,7 @@ m_list = (np.logspace(1.8, 3.3, 7).astype(int) // 2) * 2
 plt.style.use("paper/plots/stylesheet.mplstyle")
 colors = ["#648FFF", "#DC267F", "#FFB000"]
 markers = ["o", "s", "d"]
-labels = ["inconsistent", "consistent", "non-negative"]
+labels = ["inconsistent", "consistent", "consistent + non-negative"]
 
 # Determine the baseline spectral density
 kernel = lambda t, x: gaussian_kernel(t, x, sigma=sigma, n=A.shape[0])

### 2025-02-19 paper/tables/interpolation.py

@@ -77,7 +77,7 @@ def chebyshev_coefficients_quadrature(t, m, kernel, n_theta=None):
     return mu
 
 methods = [chebyshev_coefficients_quadrature, chebyshev_expansion, chebyshev_expansion]
-labels = ["quadrature FFT", "DCT", "non-negative DCT"]
+labels = ["FFT", "DCT", "non-negative DCT"]
 
 n_t = 1000
 t = np.arange(-1, 1, n_t)

### 2025-02-20 paper/spectraldensity.tex
review Chebyshev coefficient computation

@@
 
-We now analyze what error we make when using the expanded spectral density $\phi_{\sigma}^{(m)}$ to approximate the spectral density $\phi_{\sigma}$.
-To this extent, we start with the error of the Chebyshev expansion for a Gaussian smoothing kernel $g_{\sigma}$.
+We now analyze the impact on truncation on the accuracy of the spectral density $\phi_{\sigma}$. We start with a standard error estimate concerning the approximation of the Gaussian smoothing kernel.
 
 \begin{lemma}{Chebyshev expansion error}{chebyshev-error}
-    The expansion $g_{\sigma}^{(m)}$ of the Gaussian smoothing kernel $g_{{\sigma}}$ \refequ{equ:smoothing-kernel} satisfies
+    The truncated Chebyshev expansion $g_{\sigma}^{(m)}$ of the Gaussian smoothing kernel $g_{{\sigma}}$ from  \refequ{equ:smoothing-kernel} satisfies
     \begin{equation}
-        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \right| \leq \sqrt{\frac{2e}{\pi}} \frac{1}{\sigma^2} (1 + \sigma)^{-m} \equiv E_{\sigma, m}
+        \sup_{s,t \in [-1, 1]} \big| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \big| \leq \sqrt{\frac{2e}{\pi}} \frac{1}{\sigma^2} (1 + \sigma)^{-m} =: E_{\sigma, m}
         \label{equ:chebyshev-interpolation-sup-error-kernel}
@@
     \end{equation}
-    for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and $s \in [-1, 1]$.
+    for every degree $m \in \mathbb{N}$ and smoothing parameter $\sigma > 0$.
 \end{lemma}
+\begin{proof}
+The proof of this result is standard and follows the one given in the arXiv preprint of \cite[Theorem 2]{lin-2017-randomized-estimation}, but attains a slightly different dependence on $\sigma$.
 
-The proof specializes Bernstein's theorem for Chebyshev interpolation to the case of a Gaussian smoothing kernel \refequ{equ:smoothing-kernel}.
-It loosely follows the proof given in the arXiv preprint of \cite[Theorem 2]{lin-2017-randomized-estimation}, but attains a slightly different dependence on $\sigma$.
-\\ \vspace{\parskip}
-
-\begin{proof}
-    From Bernstein's theorem \cite[Theorem 4.3]{trefethen-2008-gauss-quadrature} it follows that if $f$ is analytic within the Bernstein ellipse $\mathcal{E}_{\chi} \subset \mathbb{C}$ with foci $\{-1, +1\}$ and sum of semi-axes $\chi > 1$, then for any $m \in \mathbb{N}$
-    \begin{equation}
-        \sup_{s \in [-1, 1]} \left| f(s) - f^{(m)}(s) \right| \leq \frac{2}{\chi^m (\chi - 1)} \max_{z \in \mathcal{E}_{\chi}} |f(z)|.
+Let $\mathcal{E}_{\chi} \subset \mathbb{C}$ denote the Bernstein ellipse  with foci $\{-1, +1\}$ and the sum of semi-axes equal to $\chi > 1$.
+By Bernstein's theorem \cite[Theorem 4.3]{trefethen-2008-gauss-quadrature},     \begin{equation}
+        \sup_{s \in [-1, 1]} \big| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \big| \leq \frac{2}{\chi^m (\chi - 1)} \max_{z \in \mathcal{E}_{\chi}} |g_{\sigma}(t - z)|,
         \label{equ:bernstein-bound}
     \end{equation}
-    In particular, when $f = g_{\sigma}(t - \cdot)$ we observe for $z = x + iy \in \mathcal{E}_{\chi}$
+    where used the analyticity of $g_{\sigma}(t- \cdot)$.
+    For $z = x + \mathrm{i} y \in \mathcal{E}_{\chi}$, we have
     \begin{equation}
     | g_{\sigma}(t - z) | 
-    = \frac{1}{\sigma \sqrt{2 \pi}} \left| e^{-\sfrac{(t - z)^2}{2\sigma^2}} \right|
-    = \frac{1}{\sigma \sqrt{2 \pi}} e^{-\sfrac{(t - x)^2}{2\sigma^2}}e^{\sfrac{y^2}{2\sigma^2}}.
+    = \frac{1}{\sigma \sqrt{2 \pi}} \big| e^{-\sfrac{(t - z)^2}{2\sigma^2}} \big|
+    = \frac{1}{\sigma \sqrt{2 \pi}} e^{-\sfrac{(t - x)^2}{2\sigma^2}}e^{\sfrac{y^2}{2\sigma^2}}
+    \le \frac{1}{\sigma \sqrt{2 \pi}} e^{\sfrac{(\chi - \chi^{-1})^2}{8 \sigma^2}},
     %\leq \frac{1}{n \sigma \sqrt{2 \pi}} \max_{x + iy \in \mathcal{E}_{\chi}} e^{\sfrac{y^2}{2\sigma^2}} 
     \end{equation}
-    Because $e^{-\sfrac{(t - x)^2}{2\sigma^2}} \leq 1$ for all $x, t \in \mathbb{R}$, and the maximum absolute value of $y$ is limited by the length of the semi-axis of the Bernstein ellipse in the direction of the imaginary axis, $(\chi - \chi^{-1}) / 2$, we upper bound
-    \begin{equation}
-        \max_{z \in \mathcal{E}_{\chi}} | g_{\sigma}(t - z) | 
-        \leq \frac{1}{\sigma \sqrt{2 \pi}} e^{\sfrac{(\chi - \chi^{-1})^2}{8 \sigma^2}}.
-    \end{equation}
-
-    Choosing $\chi = 1 + \sigma$ gives a simple and good bound, since in this case $\chi - \chi^{-1} \leq 2\sigma$, which implies
-    \begin{equation}
-        \sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \right| \leq \sqrt{\frac{2e}{\pi}} \frac{1}{\sigma^2} (1 + \sigma)^{-m}.
-    \end{equation}
+    where the last inequality uses 
+    $e^{-\sfrac{(t - x)^2}{2\sigma^2}} \leq 1$ and $|y| \le (\chi - \chi^{-1}) / 2$.
+    Choosing $\chi = 1 + \sigma$ and using $\chi - \chi^{-1} \leq 2\sigma$ completes the proof of~\eqref{equ:chebyshev-interpolation-sup-error-kernel}.
 \end{proof}
 
-\reflem{lem:chebyshev-error} shows that when $0 < \sigma \ll 1$ and if we choose $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, the error we induce with the Chebyshev expansion is at most $\mathcal{O}(\varepsilon)$.
-It now also allows us to bound the $L^1$-error we make when expanding the spectral density.
+For $0 < \sigma \ll 1$, \reflem{lem:chebyshev-error} implies that choosing a degree $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$ ensures an error $\mathcal{O}(\varepsilon)$.
+The lemma also allows us to bound the $L^1$-error of the spectral density approximation
+$\phi_{\sigma}^{(m)}$ defined in~\eqref{equ:expanded-spectral-density}:
 \begin{align}
     \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t 
-    &= \int_{-1}^{1} \left| \frac{1}{n} \sum_{i=1}^n \left(g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i)\right) \right|~\mathrm{d}t \notag \\
-    &\leq \int_{-1}^{1} \max_{i = 1, \dots, n} \left| g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i) \right|~\mathrm{d}t \notag \\
-    &\leq 2 E_{\sigma, m}
+    &= \int_{-1}^{1} \left| \frac{1}{n} \sum_{i=1}^n \big(g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i)\big) \right|~\mathrm{d}t \notag \\
+    &\leq \int_{-1}^{1} \max_{i = 1, \dots, n} \big| g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i) \big|~\mathrm{d}t    \leq 2 E_{\sigma, m}.
     \label{equ:chebyshev-interpolation-spectral-density}
@@
 
-\subsubsection{Preservation of non-negativity}
+\subsubsection{Preservation of non-negativity} \label{sec:preservnonneg}
 
-Our goal is to apply the Nyström++ trace estimator which we have analyzed in \refsec{subsec:nystrom-pp} to the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ to approximate the smoothed spectral density $\phi_{\sigma}$.
-However, due to the oscillatory nature of the Chebyshev expansion, we can no longer guarantee that $g_{\sigma}^{(m)} \geq 0$, and consequently, the matrix function $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ may be (slightly) indefinite.
-While in our experiments this never seemed to be an issue, it disallows us to directly use \refthm{thm:nystrom-pp}.
-While variants of the Nyström approximation designed for indefinite matrices have been analyzed, no practical algorithms for computing them exist \cite{nakatsukasa-2023-randomized-lowrank}.
+In general, the truncated Chebyshev expansion $g_{\sigma}^{(m)}$ does not inherit the  non-negativity of the smoothing kernel $g_{\sigma}$.
+Consequently, the matrix $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ may become
+(slightly) indefinite.
+This causes issues when attempting to apply the Nyström++ trace estimator (Algorithm~\ref{equ:nystrompp-trace-estimator}).
+While we have never observed this in our numerical experiments, the Nyström approximation may fail when the SPSD condition is not met; see also the discussion in~\cite{nakatsukasa-2023-randomized-lowrank}.
+On the theoretical side,~\refthm{thm:nystrom-pp}
+critically relies on the SPSD condition.
 
-To still apply the bound from \refthm{thm:nystrom-pp} in this scenario, we will have to ensure that the Chebyshev expansion is non-negative.
-We propose the following method to produce a non-negative Chebyshev expansion \textcolor{red}{[REFERENCE NEEDED]}: Fix an even $m$ and expand the square root of the smoothing kernel in a basis of Chebyshev polynomials up to degree $m/2$
+To ensure a non-negative Chebyshev expansion, we suggest the following simple method~\textcolor{red}{[REFERENCE NEEDED]}: Given an even degree $m$, expand the square root of the smoothing kernel up to degree $m/2$:
 \begin{equation}
-    \sqrt{g_{\sigma}}^{(\sfrac{m}{2})}(t - s) = \sum_{l=0}^{m/2} \xi_l(t) T_l(s).
+    \sqrt{g_{\sigma}}(t - s) \approx \sqrt{g_{\sigma}}^{(\sfrac{m}{2})}(t - s) = \sum_{l=0}^{m/2} \xi_l(t) T_l(s).
     \label{equ:square-root-chebyshev-expansion}
 \end{equation}
-Then, square this expansion with \refalg{alg:chebyshev-squaring} to obtain the non-negative Chebyshev expansion
+When squaring this expansion, once obtains a polynomial of degree $m$, which admits an exact expansion
+of the form 
+$(\sum_{l=0}^{m/2} \xi_l(t) T_l(s))^2 = \sum_{l=0}^{m} \underline{\mu}_l(t) T_l(s)$.
+Thus, we obtain the non-negative Chebyshev expansion
 \begin{equation}
-    \underline{g}_{\sigma}^{(m)}(t - s) = \sum_{l=0}^{m} \underline{\mu}_l(t) T_l(s).
+g_{\sigma}(t - s) \approx    \underline{g}_{\sigma}^{(m)}(t - s) = \sum_{l=0}^{m} \underline{\mu}_l(t) T_l(s).
     \label{equ:non-negative-chebyshev-expansion}
 \end{equation}
-
-??? UNTIL HERE ???
-
 %With a complexity of $\mathcal{O}(m \log(m))$ this procedure is computationally inexpensive and allows for very similar convergence bounds to \reflem{lem:chebyshev-error}, which we will derive \textcolor{red}{in the following?} in \refsec{subsubsec:chebyshev-nystrom-analysis}.
-
-\color{black}
-
 %To this extent, we will first analyze the non-negative Chebyshev expansion.
-
-The resulting non-negative version of the approximate smoothed spectral density \refequ{equ:expanded-spectral-density}
+The corresponding non-negative approximation of the smoothed spectral density,
 \begin{equation}
-    \phi_{\sigma}(t) \approx \underline{\phi}_{\sigma}^{(m)}(t) = \frac{1}{n} \Trace(\underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}))
+    \phi_{\sigma}(t) \approx \underline{\phi}_{\sigma}^{(m)}(t) := \frac{1}{n} \Trace(\underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})),
     \label{equ:non-negative-expanded-density}
 \end{equation}
-then satisfies the following bound:
+satisfies the following bound.
 
 \begin{lemma}{$L^1$-error of non-negative approximate smooth spectral density}{non-negative-chebyshev-error}
-Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the non-negative approximate smoothed spectral density $\underline{\phi}_{\sigma}^{(m)}$ satisfies
+Let $\phi_{\sigma}$ be the smoothed spectral density defined in~\eqref{equ:smooth-spectral-density}
+for a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum contained in $[-1, 1]$.
+Then the approximation $\underline{\phi}_{\sigma}^{(m)}$ defined in~\eqref{equ:non-negative-expanded-density} satisfies
     \begin{equation}
-        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq 4 \sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right) E_{\sqrt{2}\sigma, \sfrac{m}{2}} \equiv \underline{E}_{\sigma, m}
+        \int_{-1}^{1} \big| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \big|~\mathrm{d}t \leq 4 \sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right) E_{\sqrt{2}\sigma, \sfrac{m}{2}} =: \underline{E}_{\sigma, m}
         \label{equ:chebyshev-interpolation-sup-error-kernel-nonneg}
@@
     \end{equation}
-    for all \emph{even} degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and with $E_{\sigma, m}$ as defined in \reflem{lem:chebyshev-error}.
+    for every \emph{even} degree $m \in \mathbb{N}$, smoothing parameter $\sigma > 0$, and with $E_{\sigma, m}$ defined in \reflem{lem:chebyshev-error}.
 \end{lemma}
-
 \begin{proof}
-    For any numbers $a, b \in \mathbb{R}$ it holds $| a^2 - b^2 | \leq (2 | a | + | a - b |)  | a - b |$. Therefore,
+    Using that the inequality $| a - b | \leq (2 \sqrt{a} + | \sqrt{a} - \sqrt{b} |)  | \sqrt{a} - \sqrt{b} |$ holds for any non-negative numbers $a, b$, it follows that
     \begin{equation}
-        \left| g_{\sigma} - \underline{g}_{\sigma}^{(m)} \right| = \left| (\sqrt{g_{\sigma}})^2 - (\sqrt{g_{\sigma}}^{(\sfrac{m}{2})})^2 \right| \leq \left( 2 \left| \sqrt{g_{\sigma}} \right| + \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right| \right) \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right|.
+        \big| g_{\sigma} - \underline{g}_{\sigma}^{(m)} \big| \leq \big( 2 \big| \sqrt{g_{\sigma}} \big| + \big| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \big| \big) \big| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \big|.
     \end{equation}
-    Since $\sqrt{g_{\sigma}} = \sqrt{2 \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$, we can apply \reflem{lem:chebyshev-error} with $\sigma \to \sqrt{2}\sigma$ and $m \to m/2$ along with the bound $|\sqrt{g_{\sigma}}| \leq 1/\sqrt{\sigma \sqrt{2 \pi}}$ to get
+    Because of $\sqrt{g_{\sigma}} = \sqrt{2 \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$,
+we can apply \reflem{lem:chebyshev-error} with the substitutions
+    $\sigma \gets \sqrt{2}\sigma$ and $m \gets m/2$ to bound 
+    $\big| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \big|$.
+    Together with  $|\sqrt{g_{\sigma}}| \leq 1/\sqrt{\sigma \sqrt{2 \pi}}$, we get
     \begin{align}
-        &\sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \notag \\
-        &\leq \left( 2 \frac{1}{\sqrt{ \sigma \sqrt{2\pi}}} + \sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right)\sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}} \notag \\
-        &= 2\sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right) E_{\sqrt{2}\sigma, \sfrac{m}{2}}.
+        \big| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \big| \notag 
+        &\leq \Big( 2 \frac{1}{\sqrt{ \sigma \sqrt{2\pi}}} + \sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\Big)\sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}} \notag \\
+        &= 2\sqrt{2} \big(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\big) E_{\sqrt{2}\sigma, \sfrac{m}{2}}
     \end{align}
-    An identical treatment to \refequ{equ:chebyshev-interpolation-spectral-density} then leads to the statement of the lemma.
+    for every $s,t \in [-1, 1]$.
+    Bounding the integral as in~\refequ{equ:chebyshev-interpolation-spectral-density} then yields the desired result.
 \end{proof}
 
-On one hand, in the non-negative expansion of degree $m$ \refequ{equ:non-negative-chebyshev-expansion}, the underlying Chebyshev expansion -- whose error is shown in \reflem{lem:chebyshev-error} to exponentially decay with $m$ -- is only computed to half the degree (i.e.
-$m/2$) of the standard Chebyshev expansion, hence, usually not as accurate.
-On the other hand, in the non-negative expansion we approximate $\sqrt{g_{\sigma}} \propto g_{\sqrt{2}\sigma}$ which is less \enquote{pointy} than $g_{\sigma}$ and therefore easier to approximate with Chebyshev polynomials.
-Overall, we will need to choose the degree $m$ of the non-negative expansion roughly a factor of $\sqrt{2}$ larger to match the approximation error of the standard expansion (cf.
-\reffig{fig:interpolation-issue}).
+As the non-negative expansion~\refequ{equ:non-negative-chebyshev-expansion} ``wastes'' half of the degrees of freedom, it can be expected to be less accurate than the standard Chebyshev expansion.
+On the other hand, $\sqrt{g_{\sigma}} \propto g_{\sqrt{2}\sigma}$ is somewhat easier to approximate because it is less \enquote{pointy} compared to $g_{\sigma}$.
+Overall, it turns out that one needs to choose the degree of the non-negative expansion roughly a factor $\sqrt{2}$ larger to match the approximation error of the standard Chebyshev expansion (see also \reffig{fig:interpolation-issue}).
 
@@
 
+\textcolor{red}{???
+I don't understand this derivation.
+Don't you think that~\eqref{equ:discrete-cosine-transform} only holds approximately for the exact $g_\sigma$ and it should be something like $g_\sigma \approx g^{(m)} = ...$.
+I think there is a major confusion of two different approaches here:
+\begin{itemize}
+ \item The Chebyshev \emph{expansion} of $f$ is the $\ell^2$-orthogonal projection of $f$ on the first $m+1$ Chebyshev polynomials.
+The coefficients are computed from $\ell^2$ inner products between the Chebyshev polynomials and $f$.
+This is probably what Lin is doing.
+ \item The Chebyshev \emph{approximation} used by Chebfun and possibly others proceeds differently.
+It first replaces (or rather pretends to replace) $f$ by the interpolation of $f$ in (some) Chebyshev nodes.
+This interpolation admits an exact truncated expansion and the coefficients can be computed from a Fourier transform (or even DCT?) of the interpolation data.
+This might match what is written below, but I am not sure whether the interpolation nodes are the same.
+Please check with Trefethen's book.
+\end{itemize}
+There are two issues now
+\begin{enumerate}
+ \item I am now unsure how $g_\sigma^{(m)}$ is precisely defined. This should be spelled out clearer in the beginning of Section 3.2.
+ \item There could be a mismatch between the $g_\sigma^{(m)}$ of Lemma 3.1 and the one we are actually using.
+This mismatch can most likely be resolved, because a result of the form of Lemma 3.1 also holds for the Chebyshev interpolation.
+If it is easy to resolve, please do so.
+If not, at least point this out here.
+Have a look at Trefethen's book; it might have the right error bound.
+\end{enumerate}
+???
 It is well known that the coefficients of a Chebyshev expansion of a function can --- in exact arithmetic --- be computed by inverting the discrete cosine transform (DCT)
@@
 between the expansion coefficients $\mu_l(t),~i=0,\dots,m$ and the function evaluations at the Chebyshev nodes of the second kind $s_i = \cos(\pi i / m),~i=0,\dots,m$ \cite{vanloan-1992-computational-frameworks, baszenski-1997-fast-polynomial}.
-This can be accomplished in $\mathcal{O}(m \log(m))$ operations.
+This can be accomplished in $\mathcal{O}(m \log(m))$ operations.}
 
-A desirable property of the Chebyshev expansion is the fact that such an expansion can be squared in $\mathcal{O}(m \log(m))$ operations by using the DCT and its inverse \cite{baszenski-1997-fast-polynomial}.
-This allows us to efficiently compute the non-negative Chebyshev expansion \refequ{equ:non-negative-chebyshev-expansion}, and will also be useful in \refsec{subsec:chebyshev-nystrom}.
-The below algorithm implements this observation.
+A Chebyshev expansion can be squared in $\mathcal{O}(m \log(m))$ operations by using the DCT and its inverse \cite{baszenski-1997-fast-polynomial}; see Algorithm~\ref{lin:inverse-DCT}.
+This allows one to efficiently compute the non-negative Chebyshev expansion \refequ{equ:non-negative-chebyshev-expansion}, and it will find further use in \refsec{subsec:chebyshev-nystrom}.
 
-\begin{algo}{Fast squaring of Chebyshev expansion}{chebyshev-squaring}
+\begin{algo}{Fast squaring of Chebyshev expansions}{chebyshev-squaring}
     \begin{algorithmic}[1]
-        \Statex \textbf{Input:} Coefficients $\{ \mu_l \}_{l=0}^{m}$ of the expansion $\sum_{l=0}^{m} \mu_l T_l$
+        \Statex \textbf{Input:} Coefficients $\{ \mu_l \}_{l=0}^{m}$ of Chebyshev expansion $\sum_{l=0}^{m} \mu_l T_l$
         \Statex \textbf{Output:} Coefficients $\{ \nu_l \}_{l=0}^{2m}$ such that $\sum_{l=0}^{2m} \nu_l T_l = (\sum_{l=0}^{m} \mu_l T_l)^2$
-        \State Define $\vct{\mu} \in \mathbb{R}^{2m + 1}$ whose first $m + 1$ entries are the coefficients $\mu_0, \dots, \mu_m$ and the remaining $m$ are $0$
+        \State By appending $m$ zeros, define vector $\vct{\mu} = [\mu_0, \ldots, \mu_m,0,\ldots,0 ] \in \mathbb{R}^{2m + 1}$.
     \State Compute $\vct{f} = \operatorname{DCT}(\vct{\mu})$
-    \State Compute $\vct{\nu} = \operatorname{DCT}^{-1}(\vct{f} \odot \vct{f})$ whose entries are the coefficients $\nu_0, \dots, \nu_{2m}$
+    \State Compute $[\nu_0, \dots, \nu_{2m}] = \operatorname{DCT}^{-1}(\vct{f} \odot \vct{f})$
     \label{lin:inverse-DCT}
@@
 \end{algo}
-On \reflin{lin:inverse-DCT}, $\odot$ denotes the entrywise product.
+On \reflin{lin:inverse-DCT} of Algorithm~\ref{lin:inverse-DCT}, the symbol $\odot$ denotes the entrywise product.
 
-In \cite{lin-2017-randomized-estimation}, a slightly different approach to computing the expansion coefficients in \refequ{equ:discrete-cosine-transform} is taken.
+\textcolor{red}{??? To be modified after the notion of Chebyshev expansion is clarified ???
+In \cite{lin-2017-randomized-estimation}, a slightly different approach for computing the expansion coefficients of a standard Chebyshev  is taken.
 They express the coefficients $\mu_l(t)$ in terms of an integral, which they subsequently approximate with a quadrature supported in at least $2m + 1$ nodes.
@@
 Besides not yielding the exact expansion coefficients $\mu_l(t)$, this procedure is also significantly slower due to the larger number of quadrature nodes and the implementational benefits of the DCT over the FFT.
-In \reftab{tab:chebyshev-timing-interpolation}, we compare a benevolent implementation of \cite[Algorithm 1]{lin-2017-randomized-estimation} with our approach, both for the standard and the non-negative Chebyshev expansion.
+In \reftab{tab:chebyshev-timing-interpolation}, we compare a benevolent implementation of \cite[Algorithm 1]{lin-2017-randomized-estimation} with our approach, both for the standard and the non-negative Chebyshev expansion.}
 
 \begin{table}[ht]
-    \caption{We compare the runtime in milliseconds for computing the coefficients of the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ \refequ{equ:smoothing-kernel} in three different ways: as proposed in \cite[Algorithm 1]{lin-2017-randomized-estimation} (\emph{FFT}), using the discrete cosine transform (\emph{DCT}) \refequ{equ:discrete-cosine-transform}, and using the discrete cosine transform to compute a non-negative Chebyshev expansion with \refalg{alg:chebyshev-squaring} (\emph{non-negative DCT}).
-We use $\sigma=0.005$, $n_t=1000$ parameter values, and various degrees $m$.
-For stability, we average over 7 runs of the algorithms and repeat these runs 1000 times to form the mean and standard deviation which are given in the below table.}
+    \caption{Runtime in milliseconds for computing the coefficients of a Chebyshev expansion of the smoothing kernel $g_{\sigma}$ in three different ways: Using Algorithm 1 from~\cite{lin-2017-randomized-estimation} (\emph{FFT}), applying the discrete cosine transform direclty to $g_\sigma$ (\emph{DCT}), and applying the discrete cosine transform to $\sqrt{g_\sigma}$ together with \refalg{alg:chebyshev-squaring} (\emph{non-negative DCT}).
+\textcolor{red}{???
+It is not clear whether the degree from the non-negative DCT is $m$ or $2m$ ???} We use $\sigma=0.005$, $n_t=1000$ parameter values, and various degrees $m$.
+For stability, we average over 7 runs of the algorithms and repeat this $1000$ times to form the mean and standard deviation reported below.}
     \label{tab:chebyshev-timing-interpolation}
@@
 
-Combining the results from \refsec{sec:analysis} for $\mtx{B}(t) = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ with the Chebyshev expansion framework from \refsec{subsec:chebyshev-expansion}, we now approximate the smoothed spectral density with
+Combining the results of \refsec{sec:analysis} and \refsec{subsec:chebyshev-expansion}, we apply the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$ from~\refequ{equ:nystrompp-trace-estimator}
+to the non-negative 
+Chebyshev expansion  $\mtx{B}(t) = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$, which yields the approximation
 \begin{equation}
-    \phi_{\sigma}(t) \approx \widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) = \frac{1}{n} \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})),
+    \phi_{\sigma}(t) \approx \widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) := \frac{1}{n} \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})).
     \label{equ:chebyshev-nystrom-formula}
 \end{equation}
-which is the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$ \refequ{equ:nystrompp-trace-estimator} applied to the non-negative Chebyshev expansion $\underline{g}_{\sigma}^{(m)}$ \refequ{equ:non-negative-chebyshev-expansion}.
 
@@
 
-An efficient implementation of the approximation \refequ{equ:chebyshev-nystrom-formula} can be achieved thanks to the invariance of the trace under cyclic permutation of its arguments and the three-term recurrence relation
+One usually aims at evaluating~\eqref{equ:chebyshev-nystrom-formula} for many values of $t$. In the following, we explain how this can be done efficiently by exploiting the 
+the well-known three-term recurrence relation of Chebyshev polynomials:
 \begin{equation}
     T_0(s) = 1,\quad  T_1(s) = s,\quad 
-    T_l(s) = 2 s T_{l-1}(s) - T_{l-2}(s) \text{ for $l \geq 2$,}
+    T_l(s) = 2 s T_{l-1}(s) - T_{l-2}(s) \text{ for $l \geq 2$.}
     \label{equ:chebyshev-recurrence}
 \end{equation}
-which the Chebyshev polynomials satisfy.
-To illustrate this, consider the approximation \refequ{equ:chebyshev-nystrom-formula} when $n_{\mtx{\Psi}}$ is zero, i.e.
-only a low-rank approximation is computed.
-Using the definition of the Nyström approximation \refequ{equ:nystrom-approximation}, the cyclic invariance of the trace, and the symmetry of $\mtx{A}$, we can rewrite it as
+
+We start with the first term in the defintion~\refequ{equ:nystrompp-trace-estimator} of $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$, which requires evaluating the trace of the 
+Nyström approximation \refequ{equ:nystrom-approximation}:
 \begin{equation}
     \frac{1}{n} \Trace \big((\mtx{B}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} (\mtx{B}(t) \mtx{\Omega})^{\top} \big) =
-    \frac{1}{n} \Trace \big((\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega}}_{= \mtx{K}_1(t)})^{\dagger} (\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t)^2 \mtx{\Omega}}_{= \mtx{K}_2(t)}) \big).
-    %\frac{1}{n} \Trace((\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega})^{\dagger} (\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})^2 \mtx{\Omega})).
+    \frac{1}{n} \Trace \big((\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega}}_{=: \mtx{K}_1(t)})^{\dagger} (\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t)^2 \mtx{\Omega}}_{=: \mtx{K}_2(t)}) \big).
+    %\frac{1}{n} \Trace((\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega})^{\dagger} (\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})^2 \mtx{\Omega})),
     \label{equ:cyclic-property}
 \end{equation}
-Thanks to the affine linear form of the Chebyshev expansion, we have
+where we used the invariance of the trace under cyclic permutations.
+Inserting the Chebyshev expansion  $\mtx{B}(t) = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ gives
 \begin{equation}
-    \mtx{K}_1(t) = \mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega} = \sum_{l=0}^{m} \underline{\mu}_l(t) (\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}),
+    \mtx{K}_1(t) = \mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega} = \sum_{l=0}^{m} \underline{\mu}_l(t) (\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}).
 \end{equation}
-meaning we can first precompute the small $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ for $l=0, \dots, m$ using the Chebyshev recurrence \refequ{equ:chebyshev-recurrence} and then inexpensively sum them up with the corresponding coefficients $\underline{\mu}_l(t)$ for all $t$ in which we want to evaluate the approximation.
-The fast and exact squaring of Chebyshev expansions (cf.
-\refalg{alg:chebyshev-squaring}), allows us to also express $\underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})^2 = \sum_{l=0}^{2m} \underline{\nu}_l(t) T_l(\mtx{A})$ in terms of an affine linear sum of the small matrices $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ for $l=0, \dots, 2m$, such that the matrix $\mtx{K}_2(t)$ can be built in an analogous way to $\mtx{K}_1(t)$.
+This allows us to precompute the small $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ for $l=0, \dots, m$ using the Chebyshev recurrence \refequ{equ:chebyshev-recurrence}.
+For each value of $t$, we then only need to sum these matrices (multiplied by the corresponding coefficient $\underline{\mu}_l(t)$).
+
+The squaring of Chebyshev expansions discussed in Section~\ref{sec:preservnonneg} allows us to express $\mtx{B}(t)^2 = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})^2 = \sum_{l=0}^{2m} \underline{\nu}_l(t) T_l(\mtx{A})$, where the coefficients $\underline{\nu}_l(t)$ can be cheaply computed from $\underline{\mu}_l(t)$ using~\refalg{alg:chebyshev-squaring}).
+Hence, if we precompute
+$\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$
+also for $l=m+1, \dots, 2m$, this allows us to evaluate $\mtx{K}_2(t)$ cheaply, in an analogous way to $\mtx{K}_1(t)$.
 Alternatively, \cite{lin-2017-randomized-estimation} proposes to expand the squared smoothing kernel $g_{\sigma}^2$ in a separate Chebyshev expansion of degree $2m$.
 While this approximation of $g_{\sigma}^2$ might be more accurate, it is \emph{inconsistent} in the sense that computing a Chebyshev expansion of $g_{\sigma}^2$ is in general not the same as squaring an existing Chebyshev expansion of $g_{\sigma}$.
-We have observed the consistency between the approximation of $g_{\sigma}$ and $g_{\sigma}^2$ to make a crucial difference in terms of accuracy (cf.
+\textcolor{red}{???
+The discussion in this whole paragraph suffers a bit from that the notion of Chebyshev expansions is ambiguous.
+???} We have observed that ensuring consistency between the approximations of $g_{\sigma}$ and $g_{\sigma}^2$ makes a visible difference in terms of accuracy (cf.
 \reffig{fig:interpolation-issue}).
-The consistency also allows bounds to be derived for the estimator \refequ{equ:chebyshev-nystrom-formula}.
+This consistency will also allow us to derive bounds for the estimator \refequ{equ:chebyshev-nystrom-formula}.
 
@@
     \input{plots/interpolation.pgf}
-    \caption{Difference in the accuracy when computing the estimate \refequ{equ:cyclic-property} once by separately expanding the squared smoothing kernel $g_{\sigma}^2$ like \cite{lin-2017-randomized-estimation} (\emph{inconsistent}), once by squaring the Chebyshev expansion $g_{\sigma}^{(m)}$ using \refalg{alg:chebyshev-squaring} (\emph{consistent}), and once with a non-negative Chebyshev expansion $\underline{g}_{\sigma}^{(m)}$ \refequ{equ:non-negative-chebyshev-expansion} which is also squared with \refalg{alg:chebyshev-squaring} (\emph{non-negative}).
-We use a small example matrix from \refsec{sec:results}, fix the size of the Nyström approximation to $n_{\mtx{\Omega}} = 80$, let $\sigma = 0.005$, and compute the $L^1$ error from the approximation of the spectral density at $n_t = 100$ uniformly spaced values of $t$.}
+    \caption{Difference in the accuracy when computing the estimate \refequ{equ:cyclic-property} once by separately expanding the squared smoothing kernel $g_{\sigma}^2$ as proposed in~\cite{lin-2017-randomized-estimation} (\emph{inconsistent}), once by squaring the Chebyshev expansion $g_{\sigma}^{(m)}$ using \refalg{alg:chebyshev-squaring} (\emph{consistent}), and once with a non-negative Chebyshev expansion $\underline{g}_{\sigma}^{(m)}$ \refequ{equ:non-negative-chebyshev-expansion} which is also squared with \refalg{alg:chebyshev-squaring} (\emph{non-negative consistent}) \textcolor{red}{???
+Please fix legend in Figure.
+???}.
+We use \textcolor{red}{a small example matrix ???
+which one?
+???} from \refsec{sec:results}, fix $n_{\mtx{\Omega}} = 80$, $\sigma = 0.005$, and compute the $L^1$ error from the approximation of the spectral density at $n_t = 100$ uniformly spaced values of $t$ \textcolor{red}{???
+In which interval?
+???}.}
     \label{fig:interpolation-issue}
@@
 
+\color{black}
 Analogous observations for the \enquote{Girard-Hutchinson} term which appears in $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$ when $n_{\mtx{\Psi}}$ is nonzero show that also then, the estimator can be expressed in terms of small quadratic forms involving the matrix $\mtx{B}(t) = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$:
@@
 \end{figure}
+
+\color{black}
 

### 2025-02-20 paper/re-pro-badge.tex
automatically reproduce paper

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/c6a997e}{c6a997e} on 2025-02-18 at 14:03:19 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/a3b75b8}{a3b75b8} on 2025-02-20 at 08:34:56 UTC.

### 2025-02-20 paper/analysis.tex
re-explain Chebyshev approximation

@@
 Moreover, a straight-forward extension of the XNysTrace estimator to the parameter-dependent setting seems to be in reach.
-However, what distinguishes \refalg{alg:nystrom-chebyshev-pp} for estimating spectral densities are two key observations --- using the cyclic invariance of the trace and the affine linear form of Chebyshev expansions (see \refsec{subsubsec:chebyshev-nystrom-implementation} for details) --- which it exploits to significantly speed up the computation.
+However, what distinguishes \refalg{alg:nystrom-chebyshev-pp} for estimating spectral densities are two key observations --- using the cyclic invariance of the trace and the affine linear form of Chebyshev approximations (see \refsec{subsubsec:chebyshev-nystrom-implementation} for details) --- which it exploits to significantly speed up the computation.
 Unfortunately, we do not see a way of marrying these observations with the efficient implementation of XNysTrace described in \cite[Section 2.2]{epperly-2024-xtrace-making}, which limits its suitability for spectral density estimation.
@@
 This parallels existing results for the constant case~\cite{meyer-2021-hutch-optimal,persson-2022-improved-variants}.
-In particular, Theorem 3.4 from~\cite{persson-2022-improved-variants} shows that $|\Trace(\mtx{B}) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})| \leq \varepsilon \Trace(\mtx{B})$ holds with high probability for (constant) SPSD $\mtx{B}$ when $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1})$ in the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})$ from~\refequ{equ:nystrompp-trace-estimator}.
+In particular, \cite[Theorem 3.4]{persson-2022-improved-variants} shows that $|\Trace(\mtx{B}) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})| \leq \varepsilon \Trace(\mtx{B})$ holds with high probability for (constant) SPSD $\mtx{B}$ when $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1})$ in the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})$ from~\refequ{equ:nystrompp-trace-estimator}.
 
@@
     $\mtx{\Omega}_1 := \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times n_{\mtx{\Omega}}}$ and $\mtx{\Omega}_2 := \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times n_{\mtx{\Omega}}}$, which are independent Gaussian random matrices.
-Applying Theorem B.1 from~\cite{persson-2023-randomized-lowrank} for $f(x) = x$, see also proof of \cite[Corollary 8.2]{tropp-2023-randomized-algorithms}, yields the bound
+Applying \cite[Theorem B.1]{persson-2023-randomized-lowrank} for $f(x) = x$, see also proof of \cite[Corollary 8.2]{tropp-2023-randomized-algorithms}, yields the bound
     \begin{equation}

### 2025-02-20 paper/intro.tex
re-explain Chebyshev approximation

@@
 We propose certain modifications to the methods from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}.
-In particular, we use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials and unify all of their methods in a single algorithm, with the desirable side effect of speeding up their \enquote{spectrum sweeping} method by orders of magnitude.
+In particular, we use a more rigorous approach for approximating matrix functions in terms of Chebyshev polynomials and unify all of their methods in a single algorithm, with the desirable side effect of speeding up their \enquote{spectrum sweeping} method by orders of magnitude.
 Based on the analysis of the randomized trace estimators for parameter-dependent matrices, we provide bounds on the approximation error for our method.

### 2025-02-20 paper/numresults.tex
re-explain Chebyshev approximation

@@
 \begin{itemize}
-    \item Specify implementation details (no non-negative Chebyshev expansion, speed-up tricks, thresholding, ...)
+    \item Specify implementation details (no non-negative Chebyshev approximation, speed-up tricks, thresholding, ...)
     \item Clarify where code come from (faithful implementation, own code, ...)
@@
 
-While theoretically an important tool, we waive using the non-negative Chebyshev expansion, since the slight indefiniteness of $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ \refequ{equ:matrix-expansion} does not make a difference in finite precision arithmetic.
+While theoretically an important tool, we waive using the non-negative Chebyshev approximation, since the slight indefiniteness of $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ \refequ{equ:matrix-approximation} does not make a difference in finite precision arithmetic.
 We approximate the integrals for computing the $L^1$-errors in \refsec{sec:analysis} using a midpoint quadrature with $n_t \in \mathbb{N}$ nodes.
@@
     \caption{The Chebyshev-Nyström++ method for different ways of allocations a total of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation for the Gaussian smoothing kernel with multiple different values of the smoothing parameter.
-We make the approximation error made in the Chebyshev expansion negligible by rescaling $m=16 / \sigma$ (based on \refthm{thm:chebyshev-error}).}
+We make the approximation error made in the Chebyshev approximation negligible by rescaling $m=16 / \sigma$ (based on \reflem{lem:non-negative-chebyshev-error}).}
     \label{fig:distribution}

### 2025-02-20 paper/spectraldensity.tex
re-explain Chebyshev approximation

@@
 We now get back to the task of computing the spectral density $1/n \sum_{i=1}^{n} \delta(t - \lambda_i)$ of a real symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ with eigenvalues $\lambda_1, \dots, \lambda_n$, where $\delta(\cdot)$ denotes the Dirac delta distribution.
-As explained in the introduction, finding accurate approximations to this distribution is challenging and we will therefore work with
+As explained in the introduction, finding accurate approximations to this distribution is challenging, and we will therefore work with
 a smoothed version. %This still yields enough information
@@
 assuming that the eigenvalues are somewhat uniformly distributed and $\sigma$ is rather small, a simple calculation allows one to conclude that the smoothing error \refequ{equ:error-metric} is of the order $\sigma$.
-Consequently, to ensure that the approximation deviates by at most a factor of $\varepsilon > 0$, one needs to choose $\sigma \sim \varepsilon$.
+Consequently, to ensure that the approximation deviates by at most a factor of $\varepsilon > 0$, one needs to choose $\sigma \approx \varepsilon$.
 
-\subsection{Chebyshev expansion of smoothing kernel}
-\label{subsec:chebyshev-expansion}
+\subsection{Chebyshev approximation of smoothing kernel}
+\label{subsec:chebyshev-approximation}
 
 To estimate the trace in \refequ{equ:smooth-spectral-density} efficiently, we approximate the function 
-$s\mapsto g_{\sigma}(t- s)$ by a truncated
-Chebyshev expansion $g_{\sigma}^{(m)}(t-s) = 
+$s\mapsto g_{\sigma}(t- s)$ by a
+Chebyshev approximation $g_{\sigma}^{(m)}(t-s) = 
 \sum_{l=0}^{m} \mu_l(t) T_l(s)$
-with the expansion coefficients $\mu_l(t) \in \mathbb R$ and the Chebyshev polynomials
+with the approximation coefficients $\mu_l(t) \in \mathbb R$ and the Chebyshev polynomials
 $T_l(s) = \cos(l \cdot \arccos(s))$ for $l = 0,1,\ldots,m$.
+We choose the coefficients $\mu_l(t)$ in such a way that $g_{\sigma}^{(m)}(t-s)$ interpolates $g_{\sigma}(t-s)$ in the Chebyshev points $s_i = \cos(\pi i / m),~i=0,1,\dots,m$.
+In this case, $g_{\sigma}^{(m)}$ is often referred to the Chebyshev interpolant \cite[Chapter 2]{trefethen-2020-approximation-theory}.
 Assuming that the spectrum of $\mtx{A}$ is contained in $[-1,1]$, this gives rise to
@@
 However, doing this exactly would require us to first diagonalize $\mtx{A}$, which is often a prohibitively expensive operation.
-For algorithmic reasons (see \refsec{subsubsec:chebyshev-nystrom-implementation}), we seek for approximations of this matrix function in terms of affine linear expansions, of which the was determined to be most suitable due to its approximation error guarantees and ease of manipulation.
+For algorithmic reasons (see \refsec{subsubsec:chebyshev-nystrom-implementation}), we seek for approximations of this matrix function in terms of affine linear approximations, of which the was determined to be most suitable due to its approximation error guarantees and ease of manipulation.
 In every $t \in \mathbb{R}$ we expand
 \begin{equation}
-    g_{\sigma}(t\mtx{I}_n - \mtx{A}) \approx g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) = \sum_{l=0}^{m} \mu_l(t) T_l(\mtx{A}),
-    \label{equ:matrix-expansion}
+    g_{\sigma}(t\mtx{I}_n - \mtx{A}) \approx g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) = \sum_{l=0}^{m} \mu_l(t) T_l(\mtx{A}).
+    \label{equ:matrix-approximation}
 \end{equation}
-Inserting~\refequ{equ:matrix-expansion} into~\refequ{equ:smooth-spectral-density} yields the approximate smoothed spectral density
+Inserting~\refequ{equ:matrix-approximation} into~\refequ{equ:smooth-spectral-density} yields the approximate smoothed spectral density
 \begin{equation}
@@
 \begin{remark}
-For a general symmetric matrix $\mtx{A}$, one first estimates an interval $[a,b]$ containing the spectrum of $A$~\cite{zhou-2011-bounding-spectrum}. The usual affine transform
+For a general symmetric matrix $\mtx{A}$, one first estimates an interval $[a,b]$ containing the spectrum of $\mtx{A}$~\cite{zhou-2011-bounding-spectrum}. The usual affine transform
 $\tau(t) = \frac{2t - a - b}{b - a}$
-then lets us define a matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ for which the spectrum is contained in $[-1, 1]$ and, hence, the expansion \refequ{equ:matrix-expansion} is valid.\footnote{Note that one has to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when inverting the transformation.} This allows to assume, from now on, without loss of generality that the spectrum of $\mtx{A}$ is contained in $[-1, 1]$.
+then lets us define a matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ for which the spectrum is contained in $[-1, 1]$ and, hence, the approximation \refequ{equ:matrix-approximation} is valid.\footnote{Note that one has to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when inverting the transformation.} This allows to assume, from now on and without loss of generality, that the spectrum of $\mtx{A}$ is contained in $[-1, 1]$.
 \end{remark}
 
-\subsubsection{Expansion error}
-\label{subsubsec:expansion-error}
+\subsubsection{Approximation error}
+\label{subsubsec:approximation-error}
 
-We now analyze the impact on truncation on the accuracy of the spectral density $\phi_{\sigma}$. We start with a standard error estimate concerning the approximation of the Gaussian smoothing kernel.
+We now analyze the impact of the Chebyshev approximation of the Gaussian smoothing kernel $g_{\sigma}$ in the approximation of the spectral density $\phi_{\sigma}$.
+We start with a standard error estimate concerning the approximation of $g_{\sigma}$.
 
-\begin{lemma}{Chebyshev expansion error}{chebyshev-error}
-    The truncated Chebyshev expansion $g_{\sigma}^{(m)}$ of the Gaussian smoothing kernel $g_{{\sigma}}$ from  \refequ{equ:smoothing-kernel} satisfies
+\begin{lemma}{Chebyshev approximation error}{chebyshev-error}
+    The Chebyshev approximation $g_{\sigma}^{(m)}$ of the Gaussian smoothing kernel $g_{{\sigma}}$ from  \refequ{equ:smoothing-kernel} satisfies
     \begin{equation}
-        \sup_{s,t \in [-1, 1]} \big| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \big| \leq \sqrt{\frac{2e}{\pi}} \frac{1}{\sigma^2} (1 + \sigma)^{-m} =: E_{\sigma, m}
+        \sup_{s,t \in [-1, 1]} \big| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \big| \leq 2\sqrt{\frac{2e}{\pi}} \frac{1}{\sigma^2} (1 + \sigma)^{-m} =: E_{\sigma, m}
         \label{equ:chebyshev-interpolation-sup-error-kernel}
@@
     Let $\mathcal{E}_{\chi} \subset \mathbb{C}$ denote the Bernstein ellipse  with foci $\{-1, +1\}$ and the sum of semi-axes equal to $\chi > 1$.
-By Bernstein's theorem \cite[Theorem 4.3]{trefethen-2008-gauss-quadrature},     \begin{equation}
-        \sup_{s \in [-1, 1]} \big| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \big| \leq \frac{2}{\chi^m (\chi - 1)} \max_{z \in \mathcal{E}_{\chi}} |g_{\sigma}(t - z)|,
+By\cite[Theorem 8.2]{trefethen-2020-approximation-theory},
+    \begin{equation}
+        \sup_{s \in [-1, 1]} \big| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \big| \leq \frac{4}{\chi^m (\chi - 1)} \max_{z \in \mathcal{E}_{\chi}} |g_{\sigma}(t - z)|,
         \label{equ:bernstein-bound}
@@
     $e^{-\sfrac{(t - x)^2}{2\sigma^2}} \leq 1$ and $|y| \le (\chi - \chi^{-1}) / 2$.
-    Choosing $\chi = 1 + \sigma$ and using $\chi - \chi^{-1} \leq 2\sigma$ completes the proof of~\eqref{equ:chebyshev-interpolation-sup-error-kernel}.
+    Choosing $\chi = 1 + \sigma$ and using $\chi - \chi^{-1} \leq 2\sigma$ completes the proof of~\refequ{equ:chebyshev-interpolation-sup-error-kernel}.
 \end{proof}
@@
 The lemma also allows us to bound the $L^1$-error of the spectral density approximation
-$\phi_{\sigma}^{(m)}$ defined in~\eqref{equ:expanded-spectral-density}:
+$\phi_{\sigma}^{(m)}$ defined in~\refequ{equ:expanded-spectral-density}:
 \begin{align}
@@
 
-%\begin{theorem}{$L^1$-error of Chebyshev expansion of spectral density}{chebyshev-error}
+%\begin{theorem}{$L^1$-error of Chebyshev approximation of spectral density}{chebyshev-error}
 %    Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the spectal density with Chebyshev expansion $\phi_{\sigma}^{(m)}$ satisfies
+Then the spectal density with Chebyshev approximation $\phi_{\sigma}^{(m)}$ satisfies
 %    \begin{equation}
@@
 %    for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and the error $E_{\sigma, m}$ from \reflem{lem:chebyshev-error}.%, and $t \in \mathbb{R}$.
-%The same result holds for the expansion of a non-negative Chebyshev expansion of \emph{even} degree $m \in \mathbb{N}$ but with $\underline{E}_{\sigma, m}$.
+%The same result holds for the approximation of a non-negative Chebyshev approximation of \emph{even} degree $m \in \mathbb{N}$ but with $\underline{E}_{\sigma, m}$.
 %\end{theorem}
@@
 
-In general, the truncated Chebyshev expansion $g_{\sigma}^{(m)}$ does not inherit the  non-negativity of the smoothing kernel $g_{\sigma}$.
+In general, the truncated Chebyshev approximation $g_{\sigma}^{(m)}$ does not inherit the  non-negativity of the smoothing kernel $g_{\sigma}$.
 Consequently, the matrix $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ may become
 (slightly) indefinite.
-This causes issues when attempting to apply the Nyström++ trace estimator (Algorithm~\ref{equ:nystrompp-trace-estimator}).
+This causes issues when attempting to apply the Nyström++ trace estimator (\refalg{equ:nystrompp-trace-estimator}).
 While we have never observed this in our numerical experiments, the Nyström approximation may fail when the SPSD condition is not met; see also the discussion in~\cite{nakatsukasa-2023-randomized-lowrank}.
@@
 
-To ensure a non-negative Chebyshev expansion, we suggest the following simple method~\textcolor{red}{[REFERENCE NEEDED]}: Given an even degree $m$, expand the square root of the smoothing kernel up to degree $m/2$:
+To ensure a non-negative Chebyshev approximation, we suggest the following simple method: Given an even degree $m$, approximate the square root of the smoothing kernel up to degree $m/2$:
 \begin{equation}
     \sqrt{g_{\sigma}}(t - s) \approx \sqrt{g_{\sigma}}^{(\sfrac{m}{2})}(t - s) = \sum_{l=0}^{m/2} \xi_l(t) T_l(s).
-    \label{equ:square-root-chebyshev-expansion}
+    \label{equ:square-root-chebyshev-approximation}
 \end{equation}
-When squaring this expansion, once obtains a polynomial of degree $m$, which admits an exact expansion
+When squaring this approximation, one obtains a polynomial of degree $m$, which admits an exact approximation
 of the form 
 $(\sum_{l=0}^{m/2} \xi_l(t) T_l(s))^2 = \sum_{l=0}^{m} \underline{\mu}_l(t) T_l(s)$.
-Thus, we obtain the non-negative Chebyshev expansion
+Thus, we obtain the non-negative Chebyshev approximation
 \begin{equation}
 g_{\sigma}(t - s) \approx    \underline{g}_{\sigma}^{(m)}(t - s) = \sum_{l=0}^{m} \underline{\mu}_l(t) T_l(s).
-    \label{equ:non-negative-chebyshev-expansion}
+    \label{equ:non-negative-chebyshev-approximation}
 \end{equation}
 %With a complexity of $\mathcal{O}(m \log(m))$ this procedure is computationally inexpensive and allows for very similar convergence bounds to \reflem{lem:chebyshev-error}, which we will derive \textcolor{red}{in the following?} in \refsec{subsubsec:chebyshev-nystrom-analysis}.
-%To this extent, we will first analyze the non-negative Chebyshev expansion.
+%To this extent, we will first analyze the non-negative Chebyshev approximation.
 The corresponding non-negative approximation of the smoothed spectral density,
@@
 \begin{lemma}{$L^1$-error of non-negative approximate smooth spectral density}{non-negative-chebyshev-error}
-Let $\phi_{\sigma}$ be the smoothed spectral density defined in~\eqref{equ:smooth-spectral-density}
+Let $\phi_{\sigma}$ be the smoothed spectral density defined in~\refequ{equ:smooth-spectral-density}
 for a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum contained in $[-1, 1]$.
-Then the approximation $\underline{\phi}_{\sigma}^{(m)}$ defined in~\eqref{equ:non-negative-expanded-density} satisfies
+Then the approximation $\underline{\phi}_{\sigma}^{(m)}$ defined in~\refequ{equ:non-negative-expanded-density} satisfies
     \begin{equation}
@@
 
-As the non-negative expansion~\refequ{equ:non-negative-chebyshev-expansion} ``wastes'' half of the degrees of freedom, it can be expected to be less accurate than the standard Chebyshev expansion.
+As the non-negative approximation~\refequ{equ:non-negative-chebyshev-approximation} ``wastes'' half of the degrees of freedom, it can be expected to be less accurate than the standard Chebyshev approximation.
 On the other hand, $\sqrt{g_{\sigma}} \propto g_{\sqrt{2}\sigma}$ is somewhat easier to approximate because it is less \enquote{pointy} compared to $g_{\sigma}$.
-Overall, it turns out that one needs to choose the degree of the non-negative expansion roughly a factor $\sqrt{2}$ larger to match the approximation error of the standard Chebyshev expansion (see also \reffig{fig:interpolation-issue}).
+Overall, it turns out that one needs to choose the degree of the non-negative approximation roughly a factor $\sqrt{2}$ larger to match the approximation error of the standard Chebyshev approximation (see also \reffig{fig:interpolation-issue}).
 
-%In the following, we will show how the coefficients $\underline{\mu}_0(t), \dots, \underline{\mu}_m(t)$ of the non-negative Chebyshev expansion \refequ{equ:non-negative-chebyshev-expansion} can efficiently be computed and that this expansion exhibits similar convergence properties to the standard Chebyshev expansion.
+%In the following, we will show how the coefficients $\underline{\mu}_0(t), \dots, \underline{\mu}_m(t)$ of the non-negative Chebyshev approximation \refequ{equ:non-negative-chebyshev-approximation} can efficiently be computed and that this approximation exhibits similar convergence properties to the standard Chebyshev approximation.
 %\todo{
-%Maybe mention indefinite Nyström [Nakatsukasa], which won't give us desirable guarantees, and refinement of Chebyshev expansion [Francis Bach], which only slightly improves, but is much harder to implement.
-Side-note: it is possible to refine the coefficients of the Chebyshev expansion to get a better approximation in the supremum-norm over the set of all non-negative polynomials \cite%{fejer-1916-uber-trigonometrische}.
+%Maybe mention indefinite Nyström [Nakatsukasa], which won't give us desirable guarantees, and refinement of Chebyshev approximation [Francis Bach], which only slightly improves, but is much harder to implement.
+Side-note: it is possible to refine the coefficients of the Chebyshev approximation to get a better approximation in the supremum-norm over the set of all non-negative polynomials \cite%{fejer-1916-uber-trigonometrische}.
 The non-convex nature of this problem and the relatively small improvement we get with this technique made us decide to not pursue this further.
@@
 
-%\subsection{Computation of expansion coefficients}
+%\subsection{Computation of approximation coefficients}
 
@@
 
-\textcolor{red}{???
+%\textcolor{red}{???
 I don't understand this derivation.
-Don't you think that~\eqref{equ:discrete-cosine-transform} only holds approximately for the exact $g_\sigma$ and it should be something like $g_\sigma \approx g^{(m)} = ...$.
+Don't you think that~\refequ{equ:discrete-cosine-transform} only holds approximately for the exact $g_\sigma$ and it should be something like $g_\sigma \approx g^{(m)} = ...$.
 I think there is a major confusion of two different approaches here:
-\begin{itemize}
- \item The Chebyshev \emph{expansion} of $f$ is the $\ell^2$-orthogonal projection of $f$ on the first $m+1$ Chebyshev polynomials.
+%\begin{itemize}
+% \item The Chebyshev \emph{approximation} of $f$ is the $\ell^2$-orthogonal projection of $f$ on the first $m+1$ Chebyshev polynomials.
 The coefficients are computed from $\ell^2$ inner products between the Chebyshev polynomials and $f$.
 This is probably what Lin is doing.
- \item The Chebyshev \emph{approximation} used by Chebfun and possibly others proceeds differently.
+% \item The Chebyshev \emph{approximation} used by Chebfun and possibly others proceeds differently.
 It first replaces (or rather pretends to replace) $f$ by the interpolation of $f$ in (some) Chebyshev nodes.
-This interpolation admits an exact truncated expansion and the coefficients can be computed from a Fourier transform (or even DCT?) of the interpolation data.
+This interpolation admits an exact truncated approximation and the coefficients can be computed from a Fourier transform (or even DCT?) of the interpolation data.
 This might match what is written below, but I am not sure whether the interpolation nodes are the same.
 Please check with Trefethen's book.
-\end{itemize}
-There are two issues now
-\begin{enumerate}
- \item I am now unsure how $g_\sigma^{(m)}$ is precisely defined. This should be spelled out clearer in the beginning of Section 3.2.
- \item There could be a mismatch between the $g_\sigma^{(m)}$ of Lemma 3.1 and the one we are actually using.
+%\end{itemize}
+%There are two issues now
+%\begin{enumerate}
+% \item I am now unsure how $g_\sigma^{(m)}$ is precisely defined. This should be spelled out clearer in the beginning of Section 3.2.
+% \item There could be a mismatch between the $g_\sigma^{(m)}$ of Lemma 3.1 and the one we are actually using.
 This mismatch can most likely be resolved, because a result of the form of Lemma 3.1 also holds for the Chebyshev interpolation.
@@
 Have a look at Trefethen's book; it might have the right error bound.
-\end{enumerate}
-???
-It is well known that the coefficients of a Chebyshev expansion of a function can --- in exact arithmetic --- be computed by inverting the discrete cosine transform (DCT)
+%\end{enumerate}
+%???
+It is well known that the coefficients of the Chebyshev approximation \refequ{equ:matrix-approximation} can --- in exact arithmetic --- be computed by inverting the (type I) discrete cosine transform (DCT)
 \begin{equation}
@@
 \end{equation}
-between the expansion coefficients $\mu_l(t),~i=0,\dots,m$ and the function evaluations at the Chebyshev nodes of the second kind $s_i = \cos(\pi i / m),~i=0,\dots,m$ \cite{vanloan-1992-computational-frameworks, baszenski-1997-fast-polynomial}.
-This can be accomplished in $\mathcal{O}(m \log(m))$ operations.}
+between the approximation coefficients $\mu_l(t),~l=0,\dots,m$ and the function evaluations at the Chebyshev points $s_i = \cos(\pi i / m),~i=0,\dots,m$ \cite{baszenski-1997-fast-polynomial, plonka-2018-numerical-fourier, trefethen-2020-approximation-theory}.
+This can be accomplished in $\mathcal{O}(m \log(m))$ operations.%}
 
-A Chebyshev expansion can be squared in $\mathcal{O}(m \log(m))$ operations by using the DCT and its inverse \cite{baszenski-1997-fast-polynomial}; see Algorithm~\ref{lin:inverse-DCT}.
-This allows one to efficiently compute the non-negative Chebyshev expansion \refequ{equ:non-negative-chebyshev-expansion}, and it will find further use in \refsec{subsec:chebyshev-nystrom}.
+A Chebyshev approximation can be squared in $\mathcal{O}(m \log(m))$ operations by using the DCT and its inverse \cite{baszenski-1997-fast-polynomial}; see \refalg{alg:chebyshev-squaring}.
+This allows one to efficiently compute the non-negative Chebyshev approximation \refequ{equ:non-negative-chebyshev-approximation}, and it will find further use in \refsec{subsec:chebyshev-nystrom}.
 
-\begin{algo}{Fast squaring of Chebyshev expansions}{chebyshev-squaring}
+\begin{algo}{Fast squaring of Chebyshev approximations}{chebyshev-squaring}
     \begin{algorithmic}[1]
-        \Statex \textbf{Input:} Coefficients $\{ \mu_l \}_{l=0}^{m}$ of Chebyshev expansion $\sum_{l=0}^{m} \mu_l T_l$
+        \Statex \textbf{Input:} Coefficients $\{ \mu_l \}_{l=0}^{m}$ of Chebyshev approximation $\sum_{l=0}^{m} \mu_l T_l$
         \Statex \textbf{Output:} Coefficients $\{ \nu_l \}_{l=0}^{2m}$ such that $\sum_{l=0}^{2m} \nu_l T_l = (\sum_{l=0}^{m} \mu_l T_l)^2$
@@
 \end{algo}
-On \reflin{lin:inverse-DCT} of Algorithm~\ref{lin:inverse-DCT}, the symbol $\odot$ denotes the entrywise product.
+On \reflin{lin:inverse-DCT} of \refalg{alg:chebyshev-squaring}, the symbol $\odot$ denotes the entrywise product.
 
-\textcolor{red}{??? To be modified after the notion of Chebyshev expansion is clarified ???
-In \cite{lin-2017-randomized-estimation}, a slightly different approach for computing the expansion coefficients of a standard Chebyshev  is taken.
-They express the coefficients $\mu_l(t)$ in terms of an integral, which they subsequently approximate with a quadrature supported in at least $2m + 1$ nodes.
-They evaluate the resulting approximation with the fast Fourier transform (FFT).
-Besides not yielding the exact expansion coefficients $\mu_l(t)$, this procedure is also significantly slower due to the larger number of quadrature nodes and the implementational benefits of the DCT over the FFT.
-In \reftab{tab:chebyshev-timing-interpolation}, we compare a benevolent implementation of \cite[Algorithm 1]{lin-2017-randomized-estimation} with our approach, both for the standard and the non-negative Chebyshev expansion.}
+%\textcolor{red}{??? To be modified after the notion of Chebyshev approximation is clarified ???
+In \cite{lin-2017-randomized-estimation}, a slightly different approach to approximating the smoothing kernel $g_{\sigma}$ is taken: They expand $g_{\sigma}$ in an (infinite) Chebyshev series and truncate it after $m$ terms \cite[Chapter 4]{trefethen-2020-approximation-theory}.
+Subsequently, they approximate the integral representation of the coefficients of the series with a quadrature rule supported in at least $2m + 1$ nodes, which they can compute with the fast Fourier transform (FFT).
+Our approach and the one from \cite{lin-2017-randomized-estimation} satisfy almost identical convergence bounds \cite[Theorem 8.2]{trefethen-2020-approximation-theory}.
+However, the procedure from \cite{lin-2017-randomized-estimation} is often significantly slower due to the larger input to the FFT and the implementational disadvantages of the FFT when compared to the DCT.
+Furthermore, it introduces an additional error when approximating the integral representation of the coefficients with a quadrature.
+In \reftab{tab:chebyshev-timing-interpolation}, we compare a benevolent implementation of \cite[Algorithm 1]{lin-2017-randomized-estimation} with our approach, both for the standard and the non-negative Chebyshev approximation.%}
 
 \begin{table}[ht]
-    \caption{Runtime in milliseconds for computing the coefficients of a Chebyshev expansion of the smoothing kernel $g_{\sigma}$ in three different ways: Using Algorithm 1 from~\cite{lin-2017-randomized-estimation} (\emph{FFT}), applying the discrete cosine transform direclty to $g_\sigma$ (\emph{DCT}), and applying the discrete cosine transform to $\sqrt{g_\sigma}$ together with \refalg{alg:chebyshev-squaring} (\emph{non-negative DCT}).
-\textcolor{red}{???
-It is not clear whether the degree from the non-negative DCT is $m$ or $2m$ ???} We use $\sigma=0.005$, $n_t=1000$ parameter values, and various degrees $m$.
+    \caption{Runtime in milliseconds for computing the coefficients of a Chebyshev approximation with degree $m$ of the smoothing kernel $g_{\sigma}$ in three different ways: Using \cite[Algorithm 1]{lin-2017-randomized-estimation} (\emph{FFT}), applying the discrete cosine transform direclty to $g_\sigma$ (\emph{DCT}), and applying the discrete cosine transform to $\sqrt{g_\sigma}$ together with \refalg{alg:chebyshev-squaring} (\emph{non-negative DCT}).
+We use $\sigma=0.005$, $n_t=1000$ parameter values, and various degrees $m$.
 For stability, we average over 7 runs of the algorithms and repeat this $1000$ times to form the mean and standard deviation reported below.}
@@
 
-Combining the results of \refsec{sec:analysis} and \refsec{subsec:chebyshev-expansion}, we apply the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$ from~\refequ{equ:nystrompp-trace-estimator}
+Combining the results of \refsec{sec:analysis} and \refsec{subsec:chebyshev-approximation}, we apply the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$ from~\refequ{equ:nystrompp-trace-estimator}
 to the non-negative 
-Chebyshev expansion  $\mtx{B}(t) = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$, which yields the approximation
+Chebyshev approximation  $\mtx{B}(t) = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$, which yields the approximation
 \begin{equation}
@@
 
-One usually aims at evaluating~\eqref{equ:chebyshev-nystrom-formula} for many values of $t$. In the following, we explain how this can be done efficiently by exploiting the 
+One usually aims at evaluating~\refequ{equ:chebyshev-nystrom-formula} for many values of $t$. In the following, we explain how this can be done efficiently by exploiting the 
 the well-known three-term recurrence relation of Chebyshev polynomials:
@@
 where we used the invariance of the trace under cyclic permutations.
-Inserting the Chebyshev expansion  $\mtx{B}(t) = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ gives
+Inserting the Chebyshev approximation  $\mtx{B}(t) = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ gives
 \begin{equation}
@@
 
-The squaring of Chebyshev expansions discussed in Section~\ref{sec:preservnonneg} allows us to express $\mtx{B}(t)^2 = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})^2 = \sum_{l=0}^{2m} \underline{\nu}_l(t) T_l(\mtx{A})$, where the coefficients $\underline{\nu}_l(t)$ can be cheaply computed from $\underline{\mu}_l(t)$ using~\refalg{alg:chebyshev-squaring}).
+The squaring of Chebyshev approximations discussed in \refsec{sec:preservnonneg} allows us to express $\mtx{B}(t)^2 = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})^2 = \sum_{l=0}^{2m} \underline{\nu}_l(t) T_l(\mtx{A})$, where the coefficients $\underline{\nu}_l(t)$ can be cheaply computed from $\underline{\mu}_l(t)$ using~\refalg{alg:chebyshev-squaring}.
 Hence, if we precompute
@@
 also for $l=m+1, \dots, 2m$, this allows us to evaluate $\mtx{K}_2(t)$ cheaply, in an analogous way to $\mtx{K}_1(t)$.
-Alternatively, \cite{lin-2017-randomized-estimation} proposes to expand the squared smoothing kernel $g_{\sigma}^2$ in a separate Chebyshev expansion of degree $2m$.
-While this approximation of $g_{\sigma}^2$ might be more accurate, it is \emph{inconsistent} in the sense that computing a Chebyshev expansion of $g_{\sigma}^2$ is in general not the same as squaring an existing Chebyshev expansion of $g_{\sigma}$.
-\textcolor{red}{???
-The discussion in this whole paragraph suffers a bit from that the notion of Chebyshev expansions is ambiguous.
-???} We have observed that ensuring consistency between the approximations of $g_{\sigma}$ and $g_{\sigma}^2$ makes a visible difference in terms of accuracy (cf.
+Alternatively, \cite{lin-2017-randomized-estimation} proposes to instead approximate the squared smoothing kernel $g_{\sigma}^2$ in a separate "auxiliary" Chebyshev series truncated after degree $2m$.
+While this direct approximation of $g_{\sigma}^2$ might be more accurate than our squaring approach, it is \emph{inconsistent} in the sense that numerically computing the truncated Chebyshev series of $g_{\sigma}^2$ gives --- in general --- not the same result as squaring the truncated Chebyshev series of $g_{\sigma}$.
+We have observed that meticulously ensuring consistency between the approximations of $g_{\sigma}$ and $g_{\sigma}^2$ makes a visible difference in terms of accuracy (cf.
 \reffig{fig:interpolation-issue}).
@@
     \input{plots/interpolation.pgf}
-    \caption{Difference in the accuracy when computing the estimate \refequ{equ:cyclic-property} once by separately expanding the squared smoothing kernel $g_{\sigma}^2$ as proposed in~\cite{lin-2017-randomized-estimation} (\emph{inconsistent}), once by squaring the Chebyshev expansion $g_{\sigma}^{(m)}$ using \refalg{alg:chebyshev-squaring} (\emph{consistent}), and once with a non-negative Chebyshev expansion $\underline{g}_{\sigma}^{(m)}$ \refequ{equ:non-negative-chebyshev-expansion} which is also squared with \refalg{alg:chebyshev-squaring} (\emph{non-negative consistent}) \textcolor{red}{???
-Please fix legend in Figure.
-???}.
-We use \textcolor{red}{a small example matrix ???
-which one?
-???} from \refsec{sec:results}, fix $n_{\mtx{\Omega}} = 80$, $\sigma = 0.005$, and compute the $L^1$ error from the approximation of the spectral density at $n_t = 100$ uniformly spaced values of $t$ \textcolor{red}{???
-In which interval?
-???}.}
+    \caption{Difference in the accuracy when computing the estimate \refequ{equ:cyclic-property} once where $g_{\sigma}^2$ is approximated with an auxiliary truncated Chebyshev series as proposed in~\cite{lin-2017-randomized-estimation} (\emph{inconsistent}), once by squaring the Chebyshev approximation $g_{\sigma}^{(m)}$ using \refalg{alg:chebyshev-squaring} (\emph{consistent}), and once by using the non-negative Chebyshev approximation $\underline{g}_{\sigma}^{(m)}$ \refequ{equ:non-negative-chebyshev-approximation} which is also squared with \refalg{alg:chebyshev-squaring} (\emph{non-negative consistent}).
+We use a coarse discretization of the Hamiltonian \refequ{equ:5-experiments-electronic-hamiltonian} in two dimensions, fix $n_{\mtx{\Omega}} = 80$, $\sigma = 0.005$, and compute the $L^1$ error from the approximation of the spectral density at $n_t = 100$ uniformly spaced values of $t$ in the interval $[-1, 1]$.}
     \label{fig:interpolation-issue}
@@
     &\frac{1}{n n_{\mtx{\Psi}}} \Trace \big(\mtx{\Psi}^{\top} (\mtx{B}(t) - (\mtx{B}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} (\mtx{B}(t) \mtx{\Omega})^{\top}) \mtx{\Psi}^{\top} \big) \notag \\
-    &= \frac{1}{n n_{\mtx{\Psi}}} \big( \underbrace{\Trace(\mtx{\Psi}^{\top} \mtx{B}(t)\mtx{\Psi})}_{= \ell(t)} - \Trace((\underbrace{\mtx{\Psi}^{\top} \mtx{B}(t) \mtx{\Omega}}_{= \mtx{L}_1(t)^{\top}}) (\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega}}_{= \mtx{K}_1(t)^{\top}})^{\dagger} (\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Psi}^{\top}}_{= \mtx{L}_1(t)})) \big).
+    &= \frac{1}{n n_{\mtx{\Psi}}} \big( \underbrace{\Trace(\mtx{\Psi}^{\top} \mtx{B}(t)\mtx{\Psi})}_{=: \ell(t)} - \Trace((\underbrace{\mtx{\Psi}^{\top} \mtx{B}(t) \mtx{\Omega}}_{=: \mtx{L}_1(t)^{\top}}) (\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega}}_{=: \mtx{K}_1(t)^{\top}})^{\dagger} (\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Psi}^{\top}}_{=: \mtx{L}_1(t)})) \big).
     \label{equ:cyclic-property-hutchinson}
@@
     \Statex \textbf{Input:} Symmetric $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum in $[-1, 1]$, points $\{t_i\}_{i=1}^{n_t} \subset \mathbb{R}$
-    \Statex \textbf{Parameters:} Even degree of expansion $m \in \mathbb{N}$, number of Girard-Hutchinson queries $n_{\mtx{\Psi}} \in \mathbb{N}_0$, Nyström sketch size $n_{\mtx{\Omega}} \in \mathbb{N}_0$,  smoothing parameter $\sigma > 0$
+    \Statex \textbf{Parameters:} Even degree of approximationtion $m \in \mathbb{N}$, number of Girard-Hutchinson queries $n_{\mtx{\Psi}} \in \mathbb{N}_0$, Nyström sketch size $n_{\mtx{\Omega}} \in \mathbb{N}_0$,  smoothing parameter $\sigma > 0$
     \Statex \textbf{Output:} Approximate evaluations of the spectral density $\{\widetilde{\underline{\phi}}_{\sigma}^{(m)}(t_i)\}_{i=1}^{n_t}$
     \For {$i = 1, \dots, n_t$}
-        \State Compute the coefficients $\{\xi_l(t_i)\}_{l=0}^{m/2}$ of the Chebyshev expansion of $\sqrt{g_{\sigma}}$ \refequ{equ:square-root-chebyshev-expansion}
-        \State Compute the coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ of the non-negative Chebyshev expansion by applying \refalg{alg:chebyshev-squaring} to $\{\xi_l(t_i)\}_{l=0}^{m/2}$
-        \State Compute the coefficients $\{\underline{\nu}_l(t_i)\}_{l=0}^{2m}$ of the square of the non-negative Chebyshev expansion by applying \refalg{alg:chebyshev-squaring} to $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$
+        \State Compute the coefficients $\{\xi_l(t_i)\}_{l=0}^{m/2}$ of the Chebyshev approximation of $\sqrt{g_{\sigma}}$ \refequ{equ:square-root-chebyshev-approximation}
+        \State Compute the coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ of the non-negative Chebyshev approximation by applying \refalg{alg:chebyshev-squaring} to $\{\xi_l(t_i)\}_{l=0}^{m/2}$
+        \State Compute the coefficients $\{\underline{\nu}_l(t_i)\}_{l=0}^{2m}$ of the square of the non-negative Chebyshev approximation by applying \refalg{alg:chebyshev-squaring} to $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$
     \EndFor
-    %\State Compute the coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ of the non-negative Chebyshev expansion \refequ{equ:non-negative-chebyshev-expansion} for all points $t_i$ \textcolor{red}{maybe more details needed (maybe compute $\xi$ for m/2, then twice Alg 3.3)}
-    %\State Compute the coefficients $\{\nu_l(t_i)\}_{l=0}^{2m}$ of the square of the non-negative Chebyshev expansion with coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ using \refalg{alg:chebyshev-squaring}%\refalg{alg:2-chebyshev-chebyshev-expansion}
+    %\State Compute the coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ of the non-negative Chebyshev approximation \refequ{equ:non-negative-chebyshev-approximation} for all points $t_i$ \textcolor{red}{maybe more details needed (maybe compute $\xi$ for m/2, then twice Alg 3.3)}
+    %\State Compute the coefficients $\{\nu_l(t_i)\}_{l=0}^{2m}$ of the square of the non-negative Chebyshev approximation with coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ using \refalg{alg:chebyshev-squaring}%\refalg{alg:2-chebyshev-chebyshev-approximation}
     %\State Compute  for all $t_i$ using %\refalg{alg:3-nystrom-chebyshev-exponentiation}
@@
     \end{equation}
-    for all $\varepsilon \in (0, 1)$ and $\delta \in (0, 1/2)$.
+    for all $\varepsilon \in (0, 3)$ and $\delta \in (0, 1/2)$.
 \end{theorem}
@@
     \end{align}
-    From \reflem{lem:non-negative-chebyshev-error} we see that if $0 < \sigma \ll 1$, choosing $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$ ensures $\underline{E}_{\sigma, m} \leq \widetilde{\varepsilon}$, and thus, since $\widetilde{\varepsilon} \leq 1$,
+    From \reflem{lem:non-negative-chebyshev-error} we see that if $0 < \sigma \ll 1$, choosing $m = \mathcal{O}(\sigma^{-1}(\log(\widetilde{\varepsilon}^{-1}) + \log(\sigma^{-1})))$ ensures $\underline{E}_{\sigma, m} \leq \widetilde{\varepsilon}$, and thus, if $\widetilde{\varepsilon} \leq 1$,
     \begin{equation}
@@
 \refthm{thm:nystrom}).
-To establish a relation between the singular value decay of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ and what choice of $n_{\mtx{\Omega}}$ is needed for a significantly better approximation than ensured by \refthm{thm:chebyshev-nystrom}, we study the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$\footnote{The factor $1/n$ comes from the definition of the spectral density \refequ{equ:smooth-spectral-density} when it is written as the trace of a matrix function.}.
+To establish a relation between the singular value decay of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ and what choice of $n_{\mtx{\Omega}}$ is needed for a significantly better approximation than ensured by \refthm{thm:chebyshev-nystrom}, we study the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$.\footnote{The factor $1/n$
+does not change the conclusion, but is added to simplify the calculations and also reflects how we implement the method: We absorb the factor $1/n$ from the smoothed spectral density \refequ{equ:smooth-spectral-density} in the definition of $g_{\sigma}$, such that in reality, the Chebyshev approximation and the Nyström++ estimator are applied to $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$.}
 
@@
 \begin{equation}
-    r_{\varepsilon} = \min \{1 \leq r \leq n: \sum_{j=r+1}^n \sigma_{j} \leq \varepsilon \},
+    r_{\varepsilon_m} = \min \{1 \leq r \leq n: \sum_{j=r+1}^n \sigma_{j} \leq \varepsilon_m \},
     \label{equ:numerical-rank}
 \end{equation}
-where $\varepsilon$ is usually taken to be the double machine precision, i.e. $10^{-16}$.
+where $\varepsilon_m$ is usually taken to be the double machine precision, i.e. $\varepsilon_m \approx 10^{-16}$.
 
@@
 \begin{equation}
-    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) / n = \frac{1}{n \sqrt{2 \pi \sigma^2}} e^{-\sfrac{(t - \lambda_{(i)})^2}{2 \sigma^2}},
+    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) / n = \frac{1}{n \sqrt{2 \pi \sigma^2}} e^{-\sfrac{\left(t - \lambda_{(i)}\right)^2}{2 \sigma^2}},
     \label{equ:gaussian-kernel-eigenvalues}
@@
 \begin{equation}
-    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n) \leq \#\left\{1\leq i\leq n: |t - \lambda_i| < d_{\varepsilon, \sigma} \right\},
+    r_{\varepsilon_m}(g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n) \leq \#\left\{1\leq i\leq n: |t - \lambda_i| < d_{\varepsilon_m, \sigma} \right\},
     \label{equ:gaussian-kernel-numerical-rank}
 \end{equation}
-with $d_{\varepsilon, \sigma} = \sigma \sqrt{-2 \log(\sqrt{2 \pi} \sigma \varepsilon)}$ and the eigenvalues $\lambda_1, \dots, \lambda_n$ of $\mtx{A}$ and the constants.
-The expression \refequ{equ:gaussian-kernel-numerical-rank} has a nice visual interpretation: The numerical rank of a matrix is at most equal to the number of eigenvalues which are closer to the parameter $t$ than $d_{\varepsilon, \sigma}$.
+with $d_{\varepsilon_m, \sigma} = \sigma \sqrt{-2 \log(\sqrt{2 \pi} \sigma \varepsilon_m)}$ and the eigenvalues $\lambda_1, \dots, \lambda_n$ of $\mtx{A}$ and the constants.
+The expression \refequ{equ:gaussian-kernel-numerical-rank} has a nice visual interpretation: The numerical rank of a matrix is at most equal to the number of eigenvalues which are closer to the parameter $t$ than $d_{\varepsilon_m, \sigma}$.
 This is illustrated in \reffig{fig:numerical-rank}.
@@
     \input{figures/numerical-rank.tex}
-    \caption{The numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$ can be approximately computed by counting the number of eigenvalues $\lambda_{(1)}, \dots, \lambda_{(n)}$ of the matrix $\mtx{A}$ which lie less than a constant $d_{\varepsilon, \sigma}$ away from $t$.}
+    \caption{The numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$ can be approximately computed by counting the number of eigenvalues $\lambda_{(1)}, \dots, \lambda_{(n)}$ of the matrix $\mtx{A}$ which lie less than a constant $d_{\varepsilon_m, \sigma}$ away from $t$.}
     \label{fig:numerical-rank}
@@
 \begin{equation}
-    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n) \lessapprox n d_{\varepsilon, \sigma} \approx 6 n \sigma,
+    r_{\varepsilon_m}(g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n) \lessapprox n d_{\varepsilon_m, \sigma} \approx 6 n \sigma,
     \label{equ:gaussian-kernel-numerical-rank-uniform}
 \end{equation}
-provided we set $\varepsilon = 10^{-16}$ and choose $\sigma$ roughly between $10^{-8}$ and $1$.
+provided we set $\varepsilon_m = 10^{-16}$ and choose $\sigma$ roughly between $10^{-8}$ and $1$.
 Hence, choosing $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma)$ will yield approximations which can be significantly better than \refthm{thm:chebyshev-nystrom}.

### 2025-02-20 algorithms/chebyshev_nystrom.py

@@ -10,16 +10,16 @@ import scipy as sp
 
 from .helpers import gaussian_kernel
 
-def chebyshev_expansion(kernel, t, m, nonnegative=False):
+def chebyshev_approximation(kernel, t, m, nonnegative=False):
     """
-    Chebyshev expansion of a kernel.
+    Chebyshev approximation of a kernel.
 
     Parameters
     ----------
     kernel : kernel
         The kernel used to regularize the spectral density.
     t : int, float, list, or np.ndarray of shape (n_t,)
-        Point(s) where the expansion should be evaluated.
+        Point(s) where the approximation should be evaluated.
     m : int > 0
         Degree of the Chebyshev polynomial.
     nonnegative : bool
@@ -44,15 +44,15 @@ def chebyshev_expansion(kernel, t, m, nonnegative=False):
     # Rescale coefficients due to type-1 DCT convention
     mu[..., 1:-1] *= 2
 
-    # Square Chebyshev expansion
+    # Square Chebyshev approximation
     if nonnegative:
-        mu = square_chebyshev_expansion(mu)
+        mu = square_chebyshev_approximation(mu)
 
     return mu
 
-def square_chebyshev_expansion(mu):
+def square_chebyshev_approximation(mu):
     """
-    Square a Chebyshev expansion with coefficients mu.
+    Square a Chebyshev approximation with coefficients mu.
 
     Parameters
     ----------
@@ -100,7 +100,7 @@ def chebyshev_nystrom(A, t=0, m=100, n_Psi=10, n_Omega=10, kernel=gaussian_kerne
     kernel : callable
         Smoothing kernel.
     nonnegative : bool
-        Use non-negative Chebyshev expansion.
+        Use non-negative Chebyshev approximation.
     kappa : float > 0
         The threshold on the Hutchinson estimate of g_sigma. If it is below this
         value, instead of solving the possibly ill-conditioned generalized
@@ -125,12 +125,12 @@ def chebyshev_nystrom(A, t=0, m=100, n_Psi=10, n_Omega=10, kernel=gaussian_kerne
     n = A.shape[0]
     n_t = t.shape[0]
 
-    # Compute coefficients of Chebyshev expansion of the smoothing kernel
-    mu = chebyshev_expansion(kernel, t, m, nonnegative=nonnegative)
+    # Compute coefficients of Chebyshev approximation of the smoothing kernel
[…]

### 2025-02-20 paper/plots/interpolation.py

@@ -46,7 +46,7 @@ for i in range(3):
 
 plt.grid(True, which="both")
 plt.ylabel(r"$L^1$-error")
-plt.xlabel(r"expansion degree $m$")
+plt.xlabel(r"approximation degree $m$")
 plt.legend()
 plt.xscale("log")
 plt.yscale("log")

### 2025-02-20 paper/tables/interpolation.py

@@ -3,7 +3,7 @@ import __context__
 import timeit
 import numpy as np
 
-from algorithms.chebyshev_nystrom import chebyshev_expansion
+from algorithms.chebyshev_nystrom import chebyshev_approximation
 from algorithms.helpers import gaussian_kernel, generate_tex_tabular
 
 def time_method(method, parameters, num_times=1000, num_repeats=10):
@@ -34,12 +34,12 @@ def time_method(method, parameters, num_times=1000, num_repeats=10):
 
 def chebyshev_coefficients_quadrature(t, m, kernel, n_theta=None):
     """
-    Delta-Gauss-Chebyshev polynomial expansion.
+    Delta-Gauss-Chebyshev polynomial approximation.
 
     Parameters
     ----------
     t : int, float, list, or np.ndarray of shape (n,)
-        Point(s) where the expansion should be evaluated.
+        Point(s) where the approximation should be evaluated.
     m : int > 0
         Degree of the Chebyshev polynomial.
     kernel : callable
@@ -76,7 +76,7 @@ def chebyshev_coefficients_quadrature(t, m, kernel, n_theta=None):
     mu[:, 1:] /= n_theta
     return mu
 
-methods = [chebyshev_coefficients_quadrature, chebyshev_expansion, chebyshev_expansion]
+methods = [chebyshev_coefficients_quadrature, chebyshev_approximation, chebyshev_approximation]
 labels = ["FFT", "DCT", "non-negative DCT"]
 
 n_t = 1000

### 2025-02-20 test/test_chebyshev_nystrom.py

@@ -1,11 +1,11 @@
 import __context__
 
 import numpy as np
-from algorithms.chebyshev_nystrom import chebyshev_expansion, square_chebyshev_expansion
+from algorithms.chebyshev_nystrom import chebyshev_approximation, square_chebyshev_approximation
 from algorithms.helpers import gaussian_kernel
 
 
-def test_chebyshev_expansion():
+def test_chebyshev_approximation():
     np.random.seed(0)
 
     # Fixed parameters
@@ -17,38 +17,38 @@ def test_chebyshev_expansion():
     s_list = np.linspace(-1, 1, 100)
     kernel = lambda t, x: gaussian_kernel(t, x, sigma=sigma)
 
-    # Test standard Chebyshev expansion of kernel against theoretical bound
+    # Test standard Chebyshev approximation of kernel against theoretical bound
     bound = lambda sigma, m: np.sqrt(2 * np.e / np.pi) / sigma ** 2 * (1 + sigma) ** (-m)
 
     for m in m_list:
         # Compute function values from Chebyshev approximation
-        mu = chebyshev_expansion(kernel, t_list, m)
+        mu = chebyshev_approximation(kernel, t_list, m)
         for i, t in enumerate(t_list):
             truth = gaussian_kernel(t, s_list, sigma=sigma)
             approx = np.polynomial.chebyshev.Chebyshev(mu[i])(s_list)
 
             np.testing.assert_allclose(approx, truth, atol=bound(sigma, m))
 
-    # Test non-negative Chebyshev expansion of kernel against theoretical bound
+    # Test non-negative Chebyshev approximation of kernel against theoretical bound
     nonnegative_bound = lambda sigma, m:  2 * np.sqrt(2) * (1 + sigma * np.sqrt(np.pi) * bound(np.sqrt(2)*sigma, m/2)) * bound(np.sqrt(2)*sigma, m/2)
 
     for m in m_list:
         # Compute function values from Chebyshev approximation
-        mu = chebyshev_expansion(kernel, t_list, m, nonnegative=True)
+        mu = chebyshev_approximation(kernel, t_list, m, nonnegative=True)
         for i, t in enumerate(t_list):
             truth = gaussian_kernel(t, s_list, sigma=sigma)
             approx = np.polynomial.chebyshev.Chebyshev(mu[i])(s_list)
 
[…]

### 2025-02-20 paper/re-pro-badge.tex
automatically reproduce paper

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/a3b75b8}{a3b75b8} on 2025-02-20 at 08:34:56 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/7e1dc43}{7e1dc43} on 2025-02-20 at 14:53:12 UTC.

### 2025-02-20 paper/conclusion.tex
polish results and add conclusion

\section{Conclusion}
\label{sec:conclusion}

We have analyzed three stochastic trace estimators for constant symmetric matrices $\mtx{B} \in \mathbb{R}^{n \times n}$ when they are applied straightforwardly to a parameter-dependent symmetric matrix $\mtx{B}(t) \in \mathbb{R}^{n \times n}$. We have showed that in the $L^1$-norm, these estimators satisfy bounds which parallel existing results for constant matrices. The results for the estimators allow us to analyze the Nyström-Chebyshev++ method, an efficient technique for approximating the spectral density of a symmetric matrix. Further, we have proposed multiple improvements, both algorithmic and theoretical, to an existing implementation of this method.

### 2025-02-20 paper/numresults.tex
polish results and add conclusion

--- 2025-02-17 paper/numresults.tex
@@
 
+\color{black}
+
 In this last section, we test our algorithm in multiple scenarios.
 First, we numerically confirm our theoretical convergence bounds by approximating the spectral density of a Hamiltonian arising from electronic structure interaction.
-Subsequently, we compare our method to the Krylov-aware stochastic trace estimator \cite{chen-2023-krylovaware-stochastic} once on the electronic structure example and once for estimating the trace of a more general parameter-dependent matrix appearing in statistical thermodynamics.
+Subsequently, we compare our method to the Krylov-aware stochastic trace estimator, once on the electronic structure example, and once for estimating the trace of a more general parameter-dependent matrix appearing in statistical thermodynamics.
 In the end, we demonstrate that our algorithm can also be used to monitor the optimization process of a neural network.
+
+All our implementations are written from scratch in Python 3.12.3 using the packages NumPy 2.0.0 and SciPy 1.14.1.
+They are executed on a single thread of a GitHub-hosted Ubuntu runner with a 64-bit processor and 16 GB of RAM.
+\input{re-pro-badge.tex}
 
@@
 This is an idealized model for the interaction of nuclei on a regular grid with electrons for a $k$-vector in the center of the first Brillouin zone.
-The distribution of the eigenvalues of the Hamiltonian -- its spectral density -- allows us to interpret the system's energy levels.
+The distribution of the eigenvalues of the Hamiltonian --- its spectral density --- allows us to interpret the system's energy levels.
 
@@
 
-While theoretically an important tool, we waive using the non-negative Chebyshev expansion, since the slight indefiniteness of $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ \refequ{equ:matrix-expansion} does not make a difference in finite precision arithmetic.
+While theoretically an important tool, for the remainder of this section, we waive using the non-negative Chebyshev approximation, since the (slight) indefiniteness of $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ \refequ{equ:matrix-approximation} does not seem to matter when computing the Nyström++ estimator in finite precision arithmetic.
 We approximate the integrals for computing the $L^1$-errors in \refsec{sec:analysis} using a midpoint quadrature with $n_t \in \mathbb{N}$ nodes.
@@
     \input{plots/convergence.pgf}
-    \caption{For increasing values of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ but fixed $m$ we plot the $L^1$-approximation error for $\sigma=0.005$ on the model problem with $n_c = 1$.
-The $L^1$-error is approximated by a quadrature with $n_t = 100$ evenly spaced nodes.}
+    \caption{For increasing values of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ but fixed $m$ in the Chebyshev-Nyström++ method, we plot the $L^1$-approximation error for $\sigma=0.005$ on the model problem with $n_c = 1$.
+The $L^1$-error is approximated by a quadrature with $n_t = 100$ evenly spaced nodes in $[-1, 1]$.}
     \label{fig:convergence}
@@
     \input{plots/distribution.pgf}
-    \caption{The Chebyshev-Nyström++ method for different ways of allocations a total of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation for the Gaussian smoothing kernel with multiple different values of the smoothing parameter.
-We make the approximation error made in the Chebyshev expansion negligible by rescaling $m=16 / \sigma$ (based on \refthm{thm:chebyshev-error}).}
+    \caption{The Chebyshev-Nyström++ method is applied to the model problem with $n_c = 1$ for different ways of allocations a total of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation for the Gaussian smoothing kernel with multiple different values of the smoothing parameter.
+We make the approximation error made in the Chebyshev approximation negligible by rescaling $m=16 / \sigma$ (based on \reflem{lem:non-negative-chebyshev-error}).
+The $L^1$-error is approximated by a quadrature with $n_t = 100$ evenly spaced nodes in $[-1, 1]$.}
     \label{fig:distribution}
@@
 
-To demonstrate that our algorithm can effectively be applied in the setting of neural network optimization, we approximate the spectral density of a small fully connected convolutional neural network with 6782 parameters, which we train on the handwritten digit classification task given by the MNIST dataset\footnote{Handwritten digit classification; taken from \url{http://yann.lecun.com/exdb/mnist}.}.
+To demonstrate that our algorithm can effectively be applied in the setting of neural network optimization, we approximate the spectral density of a small fully connected convolutional neural network with 6782 parameters.
+We train this network on the handwritten digit classification task given by the MNIST dataset\footnote{Handwritten digit classification; taken from \url{http://yann.lecun.com/exdb/mnist}.} in PyTorch 2.4.1 in a standard way.
 We plot the approximation of the spectral density of the Hessian matrix of the untrained neural network, and at different stages of training in \reffig{fig:hessian-density}, as well as the corresponding mean squared error loss.
 It can well be observed that the eigenvalues creep towards zero as the training proceeds.
-Furthermore, by the absence of large eigenvalues in the spectrum, we can tell that in the final epoch, the neural network attains a more robust state than in previous epochs.
+Furthermore, despite the loss steadily decreasing, the presence of large eigenvalues in some epochs indicates a sharp minimum of the network, hence, unfavorable generalization properties.
 

### 2025-02-20 paper/spectraldensity.tex
polish results and add conclusion

@@
 It is possible to reformulate the problem as a generalized eigenvalue problem and apply thresholding to the smallest eigenvalues of a certain matrix to enforce stability \cite{lin-2017-randomized-estimation, epperly-2022-theory-quantum}.
-Since this renders the algorithm significantly more complicated but only improved the accuracy slightly during our experiments, we will instead evaluate these expressions in \reflin{lin:4-nystromchebyshev-nystrom-pp} using a least-squares solver which likewise uses a truncation of the smallest eigenvalues to enforce better conditioning.
+Since this renders the algorithm significantly more complicated but only improved the accuracy slightly during our numerical experiments, we will instead evaluate these expressions in \reflin{lin:4-nystromchebyshev-nystrom-pp} using a least-squares solver, which likewise uses a truncates the smallest eigenvalues by treating them as zero if they are smaller than $10^{-5}$ times the largest eigenvalue.
 
@@
 Hence, on \reflin{lin:4-nystromchebyshev-nystrom-pp} we would compute the pseudoinverse of an almost zero matrix, which can be extremely unstable (see \reffig{fig:zerocheck}).
-Notice that with $\mtx{K}_1(t)$ -- a matrix we need to assemble anyway -- we can cheaply compute $\Trace(\mtx{K}_1(t)) / n_{\mtx{\Omega}}$, which is the $n_{\mtx{\Omega}}$-query Girard-Hutchinson estimator.
-We may use it as a rough indicator whether $t$ lies within a region where the spectral density nearly vanishes, and subsequently set $\widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) = 0$ instead of computing the full expression on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
+Notice that with $\mtx{K}_1(t)$ --- a matrix we need to assemble anyway --- we can cheaply compute $\Trace(\mtx{K}_1(t)) / n_{\mtx{\Omega}}$, which is the $n_{\mtx{\Omega}}$-query Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}.
+If this estimate is significantly smaller than $1/n$, in our experiments usually $10^{-5}$, we may use it as an indicator that $t$ lies within a region where the spectral density nearly vanishes, and therefore set $\widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) = 0$ instead of computing the full expression on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
 
@@
     \caption{The difference the non-zero check can make when approximating a spectral density using \refalg{alg:nystrom-chebyshev-pp}.
-Here, we used a matrix described in \refsec{sec:results} and ran the Chebyshev-Nyström++ method with and without estimating if the matrix function is zero before taking its pseudo-inverse.
-We let $m=2000$, $n_{\mtx{\Omega}}=80$, $n_{\mtx{\Psi}}=0$, and $\sigma = 0.004$.}
+Here, we use a coarse discretization of the Hamiltonian \refequ{equ:5-experiments-electronic-hamiltonian} in three dimensions and ran the Chebyshev-Nyström++ method with and without estimating if the matrix function is zero before taking its pseudo-inverse.
+We let $m=2000$, $n_{\mtx{\Omega}}=80$, $n_{\mtx{\Psi}}=0$, and $\sigma = 0.004$.
+We compute the $L^1$ error from the approximation of the spectral density at $n_t = 100$ uniformly spaced values of $t$ in the interval $[-1, 1]$.}
     \label{fig:zerocheck}

### 2025-02-21 paper/spectraldensity.tex
review remainder of section 3

@@
 $T_l(s) = \cos(l \cdot \arccos(s))$ for $l = 0,1,\ldots,m$.
-We choose the coefficients $\mu_l(t)$ in such a way that $g_{\sigma}^{(m)}(t-s)$ interpolates $g_{\sigma}(t-s)$ in the Chebyshev points $s_i = \cos(\pi i / m),~i=0,1,\dots,m$.
-In this case, $g_{\sigma}^{(m)}$ is often referred to the Chebyshev interpolant \cite[Chapter 2]{trefethen-2020-approximation-theory}.
+We choose the coefficients $\mu_l(t)$ in such a way that $g_{\sigma}^{(m)}(t-s)$ interpolates $g_{\sigma}(t-s)$ at the Chebyshev nodes $s_i = \cos(\pi i / m),~i=0,1,\dots,m$.
+In this case, $g_{\sigma}^{(m)}$ is often referred to as the Chebyshev interpolant~\cite[Chapter 2]{trefethen-2020-approximation-theory}.
 Assuming that the spectrum of $\mtx{A}$ is contained in $[-1,1]$, this gives rise to
@@
 $\tau(t) = \frac{2t - a - b}{b - a}$
-then lets us define a matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ for which the spectrum is contained in $[-1, 1]$ and, hence, the approximation \refequ{equ:matrix-approximation} is valid.\footnote{Note that one has to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when inverting the transformation.} This allows to assume, from now on and without loss of generality, that the spectrum of $\mtx{A}$ is contained in $[-1, 1]$.
+then lets us define a matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ for which the spectrum is contained in $[-1, 1]$ and, hence, the approximation \refequ{equ:matrix-approximation} is sensible.\footnote{Note that one has to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when inverting the transformation.} This allows to assume, from now on and without loss of generality, that the spectrum of $\mtx{A}$ is contained in $[-1, 1]$.
 \end{remark}
@@
 \begin{proof}
-    The proof of this result is standard and follows the one given in the arXiv preprint of \cite[Theorem 2]{lin-2017-randomized-estimation}, but attains a slightly different dependence on $\sigma$.
-
+    The proof of this result is standard.
+% and follows the one given in the arXiv preprint of \cite[Theorem 2]{lin-2017-randomized-estimation}, but attains a slightly different dependence on $\sigma$.
     Let $\mathcal{E}_{\chi} \subset \mathbb{C}$ denote the Bernstein ellipse  with foci $\{-1, +1\}$ and the sum of semi-axes equal to $\chi > 1$.
@@
 
-In general, the truncated Chebyshev approximation $g_{\sigma}^{(m)}$ does not inherit the  non-negativity of the smoothing kernel $g_{\sigma}$.
+In general, the Chebyshev approximation $g_{\sigma}^{(m)}$ does not inherit the  non-negativity of the smoothing kernel $g_{\sigma}$.
 Consequently, the matrix $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ may become
@@
 %\textcolor{red}{??? To be modified after the notion of Chebyshev approximation is clarified ???
-In \cite{lin-2017-randomized-estimation}, a slightly different approach to approximating the smoothing kernel $g_{\sigma}$ is taken: They expand $g_{\sigma}$ in an (infinite) Chebyshev series and truncate it after $m$ terms \cite[Chapter 4]{trefethen-2020-approximation-theory}.
-Subsequently, they approximate the integral representation of the coefficients of the series with a quadrature rule supported in at least $2m + 1$ nodes, which they can compute with the fast Fourier transform (FFT).
-Our approach and the one from \cite{lin-2017-randomized-estimation} satisfy almost identical convergence bounds \cite[Theorem 8.2]{trefethen-2020-approximation-theory}.
-However, the procedure from \cite{lin-2017-randomized-estimation} is often significantly slower due to the larger input to the FFT and the implementational disadvantages of the FFT when compared to the DCT.
-Furthermore, it introduces an additional error when approximating the integral representation of the coefficients with a quadrature.
-In \reftab{tab:chebyshev-timing-interpolation}, we compare a benevolent implementation of \cite[Algorithm 1]{lin-2017-randomized-estimation} with our approach, both for the standard and the non-negative Chebyshev approximation.%}
+In \cite{lin-2017-randomized-estimation}, a slightly different approach to approximating $g_{\sigma}$ is taken: Instead of Chebyshev interpolation,
+the kernel $g_{\sigma}$ is approximated by a Chebyshev series~\cite[Chapter 4]{trefethen-2020-approximation-theory} truncated after term $m$.
+In constrast to interpolation, the expansion coefficents involve integrals, which need to be approximated by quadrature.
+By choosing $2m+1$ Chebyshev quadrature nodes, this approximation can be carried using the fast Fourier transform (FFT).
+Although interpolation and truncated expansion satisfy nearly identical convergence bounds~\cite[Theorem 8.2]{trefethen-2020-approximation-theory}, we observed the procedure from~\cite{lin-2017-randomized-estimation} to be significantly slower due to need of applying the FFT (instead of DCT) to longer vectors.
+See~\reftab{tab:chebyshev-timing-interpolation}, where we compare a benevolent implementation of \cite[Algorithm 1]{lin-2017-randomized-estimation} with our approach, both for the standard and the non-negative Chebyshev approximation.%}
 
 \begin{table}[ht]
-    \caption{Runtime in milliseconds for computing the coefficients of a Chebyshev approximation with degree $m$ of the smoothing kernel $g_{\sigma}$ in three different ways: Using \cite[Algorithm 1]{lin-2017-randomized-estimation} (\emph{FFT}), applying the discrete cosine transform direclty to $g_\sigma$ (\emph{DCT}), and applying the discrete cosine transform to $\sqrt{g_\sigma}$ together with \refalg{alg:chebyshev-squaring} (\emph{non-negative DCT}).
+    \caption{Runtime in milliseconds for computing the coefficients of a degree-$m$ Chebyshev approximation to the smoothing kernel $g_{\sigma}$ in three different ways.
+\emph{FFT}: Using~\cite[Algorithm 1]{lin-2017-randomized-estimation}.
+\emph{DCT}: Applying the discrete cosine transform directly to $g_\sigma$.
+\emph{Non-negative DCT}: Applying the discrete cosine transform to $\sqrt{g_\sigma}$ together with \refalg{alg:chebyshev-squaring}.
 We use $\sigma=0.005$, $n_t=1000$ parameter values, and various degrees $m$.
-For stability, we average over 7 runs of the algorithms and repeat this $1000$ times to form the mean and standard deviation reported below.}
+We average over 7 runs of the algorithms and repeat this $1000$ times to form the mean and standard deviation reported below.}
     \label{tab:chebyshev-timing-interpolation}
@@
 to the non-negative 
-Chebyshev approximation  $\mtx{B}(t) = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$, which yields the approximation
+Chebyshev approximation, which yields the approximation
 \begin{equation}
-    \phi_{\sigma}(t) \approx \widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) := \frac{1}{n} \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})).
+    \phi_{\sigma}(t) \approx \widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) := \frac{1}{n} \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)), \quad \mtx{B}(t) = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}).
     \label{equ:chebyshev-nystrom-formula}
@@
 
-One usually aims at evaluating~\refequ{equ:chebyshev-nystrom-formula} for many values of $t$. In the following, we explain how this can be done efficiently by exploiting the 
+One usually aims at evaluating~\refequ{equ:chebyshev-nystrom-formula} for many values of $t$. In the following, we explain how this can be done efficiently by exploiting 
 the well-known three-term recurrence relation of Chebyshev polynomials:
@@
 
-We start with the first term in the defintion~\refequ{equ:nystrompp-trace-estimator} of $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$, which requires evaluating the trace of the 
+We start with the first term in the definition~\refequ{equ:nystrompp-trace-estimator} of $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$, which requires evaluating the trace of the 
 Nyström approximation \refequ{equ:nystrom-approximation}:
@@
 \end{equation}
-This allows us to precompute the small $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ for $l=0, \dots, m$ using the Chebyshev recurrence \refequ{equ:chebyshev-recurrence}.
+Observing that $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ is independent of $t$ allows us to precompute these (small) $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices for $l=0, \dots, m$ using the Chebyshev recurrence \refequ{equ:chebyshev-recurrence}.
 For each value of $t$, we then only need to sum these matrices (multiplied by the corresponding coefficient $\underline{\mu}_l(t)$).
@@
 also for $l=m+1, \dots, 2m$, this allows us to evaluate $\mtx{K}_2(t)$ cheaply, in an analogous way to $\mtx{K}_1(t)$.
-Alternatively, \cite{lin-2017-randomized-estimation} proposes to instead approximate the squared smoothing kernel $g_{\sigma}^2$ in a separate "auxiliary" Chebyshev series truncated after degree $2m$.
+Alternatively, \cite{lin-2017-randomized-estimation} proposes to instead approximate the squared smoothing kernel $g_{\sigma}^2$ in a separate "auxiliary" Chebyshev series truncated after term $2m$.
 While this direct approximation of $g_{\sigma}^2$ might be more accurate than our squaring approach, it is \emph{inconsistent} in the sense that numerically computing the truncated Chebyshev series of $g_{\sigma}^2$ gives --- in general --- not the same result as squaring the truncated Chebyshev series of $g_{\sigma}$.
@@
     \input{plots/interpolation.pgf}
-    \caption{Difference in the accuracy when computing the estimate \refequ{equ:cyclic-property} once where $g_{\sigma}^2$ is approximated with an auxiliary truncated Chebyshev series as proposed in~\cite{lin-2017-randomized-estimation} (\emph{inconsistent}), once by squaring the Chebyshev approximation $g_{\sigma}^{(m)}$ using \refalg{alg:chebyshev-squaring} (\emph{consistent}), and once by using the non-negative Chebyshev approximation $\underline{g}_{\sigma}^{(m)}$ \refequ{equ:non-negative-chebyshev-approximation} which is also squared with \refalg{alg:chebyshev-squaring} (\emph{non-negative consistent}).
-We use a coarse discretization of the Hamiltonian \refequ{equ:5-experiments-electronic-hamiltonian} in two dimensions, fix $n_{\mtx{\Omega}} = 80$, $\sigma = 0.005$, and compute the $L^1$ error from the approximation of the spectral density at $n_t = 100$ uniformly spaced values of $t$ in the interval $[-1, 1]$.}
+    \caption{Accuracy when computing the estimate \refequ{equ:cyclic-property} in three different ways.
+\emph{Inconsistent}: Applying an auxiliary truncated Chebyshev series to $g_{\sigma}^2$, as proposed in~\cite{lin-2017-randomized-estimation}.
+\emph{Consistent}: Squaring the Chebyshev approximation $g_{\sigma}^{(m)}$ using \refalg{alg:chebyshev-squaring}.
+\emph{Non-negative consistent}: Squaring the non-negative Chebyshev approximation $\underline{g}_{\sigma}^{(m)}$ with~\refalg{alg:chebyshev-squaring}.
+We use a \textcolor{red}{???
+MORE DETAIL: At least, say what is $N$.
+Same in Figure 3.2 ???} coarse discretization of the Hamiltonian defined in~\refequ{equ:5-experiments-electronic-hamiltonian} in two dimensions, fix $n_{\mtx{\Omega}} = 80$, $\sigma = 0.005$, and compute the $L^1$ error from the approximation of the spectral density at $n_t = 100$ uniformly spaced values of $t$ in the interval $[-1, 1]$.}
     \label{fig:interpolation-issue}
@@
 
-\color{black}
-Analogous observations for the \enquote{Girard-Hutchinson} term which appears in $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$ when $n_{\mtx{\Psi}}$ is nonzero show that also then, the estimator can be expressed in terms of small quadratic forms involving the matrix $\mtx{B}(t) = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$:
+To treat the second term in the definition~\refequ{equ:nystrompp-trace-estimator} of $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$, we observe that 
 \begin{align}
     &\frac{1}{n n_{\mtx{\Psi}}} \Trace \big(\mtx{\Psi}^{\top} (\mtx{B}(t) - (\mtx{B}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} (\mtx{B}(t) \mtx{\Omega})^{\top}) \mtx{\Psi}^{\top} \big) \notag \\
-    &= \frac{1}{n n_{\mtx{\Psi}}} \big( \underbrace{\Trace(\mtx{\Psi}^{\top} \mtx{B}(t)\mtx{\Psi})}_{=: \ell(t)} - \Trace((\underbrace{\mtx{\Psi}^{\top} \mtx{B}(t) \mtx{\Omega}}_{=: \mtx{L}_1(t)^{\top}}) (\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega}}_{=: \mtx{K}_1(t)^{\top}})^{\dagger} (\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Psi}^{\top}}_{=: \mtx{L}_1(t)})) \big).
+    &= \frac{1}{n n_{\mtx{\Psi}}} \big( \underbrace{\Trace(\mtx{\Psi}^{\top} \mtx{B}(t)\mtx{\Psi})}_{=: \ell(t)} - \Trace((\underbrace{\mtx{\Psi}^{\top} \mtx{B}(t) \mtx{\Omega}}_{=: \mtx{L}_1(t)^{\top}}) (\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega}}_{= \mtx{K}_1(t)^{\top}})^{\dagger} (\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Psi}^{\top}}_{=: \mtx{L}_1(t)})) \big).
     \label{equ:cyclic-property-hutchinson}
 \end{align}
+Each of the annotated terms can be evaluated efficiently for many values of $t$, by precomputing quantities using the Chebyshev reccurrence 
+$\mtx{B}(t) = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$, as explained above.
 
-An efficient implementation of the whole procedure described in this section can be found in \refalg{alg:nystrom-chebyshev-pp}.
+\refalg{alg:nystrom-chebyshev-pp} summarizes the considerations of this section.
 
@@
     \Statex \textbf{Input:} Symmetric $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum in $[-1, 1]$, points $\{t_i\}_{i=1}^{n_t} \subset \mathbb{R}$
-    \Statex \textbf{Parameters:} Even degree of approximationtion $m \in \mathbb{N}$, number of Girard-Hutchinson queries $n_{\mtx{\Psi}} \in \mathbb{N}_0$, Nyström sketch size $n_{\mtx{\Omega}} \in \mathbb{N}_0$,  smoothing parameter $\sigma > 0$
-    \Statex \textbf{Output:} Approximate evaluations of the spectral density $\{\widetilde{\underline{\phi}}_{\sigma}^{(m)}(t_i)\}_{i=1}^{n_t}$
+    \Statex \textbf{Parameters:} Even approximation degree $m \in \mathbb{N}$, number of Girard-Hutchinson queries $n_{\mtx{\Psi}} \in \mathbb{N}_0$, Nyström sketch size $n_{\mtx{\Omega}} \in \mathbb{N}_0$,  smoothing parameter $\sigma > 0$
+    \Statex \textbf{Output:} Spectral density approximation $\{\widetilde{\underline{\phi}}_{\sigma}^{(m)}(t_i)\}_{i=1}^{n_t}$
     \For {$i = 1, \dots, n_t$}
-        \State Compute the coefficients $\{\xi_l(t_i)\}_{l=0}^{m/2}$ of the Chebyshev approximation of $\sqrt{g_{\sigma}}$ \refequ{equ:square-root-chebyshev-approximation}
-        \State Compute the coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ of the non-negative Chebyshev approximation by applying \refalg{alg:chebyshev-squaring} to $\{\xi_l(t_i)\}_{l=0}^{m/2}$
-        \State Compute the coefficients $\{\underline{\nu}_l(t_i)\}_{l=0}^{2m}$ of the square of the non-negative Chebyshev approximation by applying \refalg{alg:chebyshev-squaring} to $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$
+        \State Compute coefficients $\{\xi_l(t_i)\}_{l=0}^{m/2}$ of Chebyshev approx.~\refequ{equ:square-root-chebyshev-approximation} to $\sqrt{g_{\sigma}}$ with DCT
+        \State Compute coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ of non-negative Chebyshev approx. to $g_{\sigma}$ by applying \refalg{alg:chebyshev-squaring} to $\{\xi_l(t_i)\}_{l=0}^{m/2}$
+        \State Compute coefficients $\{\underline{\nu}_l(t_i)\}_{l=0}^{2m}$ of squared non-negative Chebyshev approxim.
+by applying \refalg{alg:chebyshev-squaring} to $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$
     \EndFor
@@
     %\State Compute  for all $t_i$ using %\refalg{alg:3-nystrom-chebyshev-exponentiation}
-    \State Generate Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$%\glsfirst{sketching-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$
+    \State Generate Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$, $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$%\glsfirst{sketching-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$
     %\State Generate standard Gaussian %\glsfirst{random-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$
@@
 
-The dominating cost in \refalg{alg:nystrom-chebyshev-pp} stems from the $(2m + 1)(n_{\mtx{\Psi}} + n_{\mtx{\Omega}})$ matrix-vector multiplications with $\mtx{A}$ and the cost associated with computing the estimate in each evaluation point $t_1, \dots, t_{n_t}$ given by $\mathcal{O}(m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + (n_{\mtx{\Omega}}^3 + n_{\mtx{\Omega}} n_{\mtx{\Psi}}^2 + 1))$.
+\refalg{alg:nystrom-chebyshev-pp} requires $(2m + 1)(n_{\mtx{\Psi}} + n_{\mtx{\Omega}})$ matrix-vector multiplications with $\mtx{A}$, which can be expected to dominate its cost.
+For each of the evaluation points $t_1, \dots, t_{n_t}$ another $\mathcal{O}(m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + (n_{\mtx{\Omega}}^3 + n_{\mtx{\Omega}} n_{\mtx{\Psi}}^2 + 1))$ operations are needed, which is independent of $n$ -- thanks to the precomputations discussed above.
 
-The algorithm also accomodates the two other estimators introduced in \refsec{sec:analysis}: For $n_{\mtx{\Omega}} = 0$ we recover the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}, and the complexity for each evaluation of the estimate decreases significantly.
-For $n_{\mtx{\Psi}} = 0$ we end up with the Nyström estimator \refequ{equ:nystrom-trace-estimator}, provided the implementation evaluates $0/0$ to $0$ in the second term on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
-Notice that in this case, \refalg{alg:nystrom-chebyshev-pp} is equivalent to the \enquote{spectrum sweeping} method from \cite[Algorithm 5]{lin-2017-randomized-estimation}, but has a significantly lower computational complexity.
+\refalg{alg:nystrom-chebyshev-pp} also accommodates the other two estimators discussed in \refsec{sec:analysis}: For $n_{\mtx{\Omega}} = 0$ we recover the Girard-Hutchinson estimator~\refequ{equ:hutchinson-trace-estimator}.
+For $n_{\mtx{\Psi}} = 0$ we recover the Nyström estimator~\refequ{equ:nystrom-trace-estimator}, provided that $0/0$ is evaluated to $0$ in the second term on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
+In this case, \refalg{alg:nystrom-chebyshev-pp} essentially amounts to the \enquote{spectrum sweeping} method from \cite[Algorithm 5]{lin-2017-randomized-estimation}, featuring a significantly lower computational complexity.
 
@@
 
-The pseudoinverses on \reflin{lin:4-nystromchebyshev-nystrom-pp} need to be computed with care.
-It is possible to reformulate the problem as a generalized eigenvalue problem and apply thresholding to the smallest eigenvalues of a certain matrix to enforce stability \cite{lin-2017-randomized-estimation, epperly-2022-theory-quantum}.
-Since this renders the algorithm significantly more complicated but only improved the accuracy slightly during our numerical experiments, we will instead evaluate these expressions in \reflin{lin:4-nystromchebyshev-nystrom-pp} using a least-squares solver, which likewise uses a truncates the smallest eigenvalues by treating them as zero if they are smaller than $10^{-5}$ times the largest eigenvalue.
+The two pseudoinverses on \reflin{lin:4-nystromchebyshev-nystrom-pp} need to be computed with care. One way to ensure numerical stability is to 
+consider an associated eigenvalue problem and threshold the smallest eigenvalues~\cite{lin-2017-randomized-estimation, epperly-2022-theory-quantum}.
+However, this renders the algorithm significantly more complicated and we found that it improves accuracy only slightly.
+Instead, we  use a least-squares solver when applying $\mtx{K}_1(t_i)^{\dagger}$ \textcolor{red}{???
+Please add more detail here.
+Which Python function do you use?
+Is $10^{-5}$ your choice?
+???}, which likewise truncates the smallest eigenvalues by treating them as zero if they are smaller than $10^{-5}$ times the largest eigenvalue.
+
+\color{black}
 
@@
 
-In this section, we will upper bound the output of the Chebyshev-Nyström++ method (see. \refalg{alg:nystrom-chebyshev-pp}). 
-\begin{theorem}{Chebyshev-Nyström++ method}{chebyshev-nystrom}
-    Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the Chebyshev-Nyström++ estimate $\widetilde{\underline{\phi}}_{\sigma}^{(m)}$ with \emph{even} degree $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, \emph{even} $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(\delta^{-1})^2)$, and smoothing parameter $0 < \sigma \ll 1$ satisfies with probability at least $1 - \delta$
+\color{blue}
+
+In this section, we analyze the accuracy of Chebyshev-Nyström++ estimate returned by~\refalg{alg:nystrom-chebyshev-pp}. 
+\begin{theorem}{Error of Chebyshev-Nyström++ estimate}{chebyshev-nystrom}
+    For a smoothing parameter $0 < \sigma \ll 1$ \textcolor{red}{???
+Why do we need $\ll$ here?
+This is quite confusing because this is not a mathematical statement.
+Can this be simply replaced by $<$ or not?
+???}, let $\phi_{\sigma}$ be the smoothed spectral density~\eqref{equ:smooth-spectral-density} of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum contained in $[-1, 1]$.
+Consider $\varepsilon \in (0, 3)$ and $\delta \in (0, 1/2)$.
+    Then the Chebyshev-Nyström++ estimate $\widetilde{\underline{\phi}}_{\sigma}^{(m)}$ 
+    satisfies the error bound
     \begin{equation}
@@
     \end{equation}
-    for all $\varepsilon \in (0, 3)$ and $\delta \in (0, 1/2)$.
+    with probability at least $1 - \delta$ for $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$  and $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(\delta^{-1})^2)$.
 \end{theorem}
@@
 \begin{proof}
-    The previously derived results can be applied to
+Combining results derived above, we obtain that \color{black}
     \begin{align}
         &\int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \notag \\
-        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \int_{-1}^{1} \left| \underline{\phi}_{\sigma}^{(m)}(t) - \widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
+        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \int_{-1}^{1} \left| \underline{\phi}_{\sigma}^{(m)}(t) - \widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t &&  \notag \\
         &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \widetilde{\varepsilon} \int_{-1}^{1} \left| \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t && \text{(\refthm{thm:nystrom-pp} w.p.
 $\geq 1 - \delta$)} \notag \\
-        &\leq (1 + \widetilde{\varepsilon}) \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \widetilde{\varepsilon} \int_{-1}^{1} \left| \phi_{\sigma}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
-        &\leq (1 + \widetilde{\varepsilon}) \underline{E}_{\sigma, m} + \widetilde{\varepsilon}. && \text{(\reflem{lem:non-negative-chebyshev-error}, normalization)} \notag \\
+        &\leq (1 + \widetilde{\varepsilon}) \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \widetilde{\varepsilon} \int_{-1}^{1} \left| \phi_{\sigma}(t) \right|~\mathrm{d}t &&  \notag \\
+        &\leq (1 + \widetilde{\varepsilon}) \underline{E}_{\sigma, m} + \widetilde{\varepsilon}. && \text{(\reflem{lem:non-negative-chebyshev-error})} \notag 
     \end{align}
-    From \reflem{lem:non-negative-chebyshev-error} we see that if $0 < \sigma \ll 1$, choosing $m = \mathcal{O}(\sigma^{-1}(\log(\widetilde{\varepsilon}^{-1}) + \log(\sigma^{-1})))$ ensures $\underline{E}_{\sigma, m} \leq \widetilde{\varepsilon}$, and thus, if $\widetilde{\varepsilon} \leq 1$,
+    From \reflem{lem:non-negative-chebyshev-error} we see that if \textcolor{red}{$0 < \sigma \ll 1$ ???
+Why $\ll$?
+???}, choosing $m = \mathcal{O}(\sigma^{-1}(\log(\widetilde{\varepsilon}^{-1}) + \log(\sigma^{-1})))$ ensures $\underline{E}_{\sigma, m} \leq \widetilde{\varepsilon}$, and thus, if $\widetilde{\varepsilon} \leq 1$,
     \begin{equation}

### 2025-02-21 paper/numresults.tex
review remainder of section 3
rewrite numerical results

@@
 They are executed on a single thread of a GitHub-hosted Ubuntu runner with a 64-bit processor and 16 GB of RAM.
-\input{re-pro-badge.tex}
 
@@
 \reffig{fig:gaussian-well}).
-The discretization step is fixed to be $h=0.6$, such that the size of the matrix grows cubically with $n_c$.
+The discretization step is fixed to be $h=0.6$.
 This is an idealized model for the interaction of nuclei on a regular grid with electrons for a $k$-vector in the center of the first Brillouin zone.
@@
 
-\begin{figure}[ht]
+\begin{figure}[!htpb]
     \centering
     \input{plots/convergence.pgf}
-    \caption{For increasing values of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ but fixed $m$ in the Chebyshev-Nyström++ method, we plot the $L^1$-approximation error for $\sigma=0.005$ on the model problem with $n_c = 1$.
+    \caption{For increasing values of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ but fixed $m$, we apply the Chebyshev-Nyström++ method to the Hamiltonian matrix described in \refsec{subsec:hamiltonian}.
+We plot the $L^1$-approximation error for $\sigma=0.005$ on the model problem with $n_c = 1$.
 The $L^1$-error is approximated by a quadrature with $n_t = 100$ evenly spaced nodes in $[-1, 1]$.}
@@
 
-\begin{figure}[ht]
+\begin{figure}[!htpb]
     \centering

### 2025-02-21 paper/spectraldensity.tex
rewrite numerical results

@@
 
-For $0 < \sigma \ll 1$, \reflem{lem:chebyshev-error} implies that choosing a degree $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$ ensures an error $\mathcal{O}(\varepsilon)$.
+A simple algebraic manipulation shows that \reflem{lem:chebyshev-error} implies that choosing a degree $m = \mathcal{O}((\log(\varepsilon^{-1}) + \log(\sigma^{-1})) / \log(1 + \sigma))$ ensures an error $\mathcal{O}(\varepsilon)$.
 The lemma also allows us to bound the $L^1$-error of the spectral density approximation
@@
 \end{proof}
+
+Again, \reflem{lem:non-negative-chebyshev-error} implies that choosing a degree $m = \mathcal{O}((\log(\varepsilon^{-1}) + \log(\sigma^{-1})) / \log(1 + \sigma))$ ensures an error $\mathcal{O}(\varepsilon)$.
 
@@
 \emph{Non-negative consistent}: Squaring the non-negative Chebyshev approximation $\underline{g}_{\sigma}^{(m)}$ with~\refalg{alg:chebyshev-squaring}.
-We use a \textcolor{red}{???
-MORE DETAIL: At least, say what is $N$.
-Same in Figure 3.2 ???} coarse discretization of the Hamiltonian defined in~\refequ{equ:5-experiments-electronic-hamiltonian} in two dimensions, fix $n_{\mtx{\Omega}} = 80$, $\sigma = 0.005$, and compute the $L^1$ error from the approximation of the spectral density at $n_t = 100$ uniformly spaced values of $t$ in the interval $[-1, 1]$.}
+As input matrix we use the $100 \times 100$ matrix $\mtx{A}$ coming from a two-dimensional finite difference discretization of the Hamiltonian defined in~\refequ{equ:5-experiments-electronic-hamiltonian} on a $10 \times 10$ grid.
+We fix $n_{\mtx{\Omega}} = 80$ to make sure the low-rank approximation is accurate, set $\sigma = 0.005$, and compute the $L^1$ error from the approximation of the spectral density at $n_t = 100$ uniformly spaced values of $t$ in the interval $[-1, 1]$.}
     \label{fig:interpolation-issue}
@@
         \State Compute coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ of non-negative Chebyshev approx. to $g_{\sigma}$ by applying \refalg{alg:chebyshev-squaring} to $\{\xi_l(t_i)\}_{l=0}^{m/2}$
-        \State Compute coefficients $\{\underline{\nu}_l(t_i)\}_{l=0}^{2m}$ of squared non-negative Chebyshev approxim.
+        \State Compute coefficients $\{\underline{\nu}_l(t_i)\}_{l=0}^{2m}$ of squared non-negative Chebyshev approx.
 by applying \refalg{alg:chebyshev-squaring} to $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$
@@
 consider an associated eigenvalue problem and threshold the smallest eigenvalues~\cite{lin-2017-randomized-estimation, epperly-2022-theory-quantum}.
-However, this renders the algorithm significantly more complicated and we found that it improves accuracy only slightly.
-Instead, we  use a least-squares solver when applying $\mtx{K}_1(t_i)^{\dagger}$ \textcolor{red}{???
-Please add more detail here.
-Which Python function do you use?
-Is $10^{-5}$ your choice?
-???}, which likewise truncates the smallest eigenvalues by treating them as zero if they are smaller than $10^{-5}$ times the largest eigenvalue.
+However, this renders the algorithm significantly more complicated, and we found that it improves accuracy only slightly.
+Instead, we use the standard NumPy least-squares solver when applying $\mtx{K}_1(t_i)^{\dagger}$, which is based on the DGELSD routine from LAPACK, and likewise truncates the smallest eigenvalues by treating them as zero if they are smaller than a certain threshold.
+We have found a threshold of $10^{-5}$ times the largest eigenvalue to give the best results in our numerical experiments.
 
@@
 Notice that with $\mtx{K}_1(t)$ --- a matrix we need to assemble anyway --- we can cheaply compute $\Trace(\mtx{K}_1(t)) / n_{\mtx{\Omega}}$, which is the $n_{\mtx{\Omega}}$-query Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}.
-If this estimate is significantly smaller than $1/n$, in our experiments usually $10^{-5}$, we may use it as an indicator that $t$ lies within a region where the spectral density nearly vanishes, and therefore set $\widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) = 0$ instead of computing the full expression on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
+If this estimate is significantly smaller than $1/n$, in our experiments usually around $10^{-5}$, we may use it as an indicator that $t$ lies within a region where the spectral density nearly vanishes, and therefore set $\widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) = 0$ instead of computing the full expression on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
 
@@
     \caption{The difference the non-zero check can make when approximating a spectral density using \refalg{alg:nystrom-chebyshev-pp}.
-Here, we use a coarse discretization of the Hamiltonian \refequ{equ:5-experiments-electronic-hamiltonian} in three dimensions and ran the Chebyshev-Nyström++ method with and without estimating if the matrix function is zero before taking its pseudo-inverse.
+As input matrix we use the $1000 \times 1000$ matrix $\mtx{A}$ coming from a three-dimensional finite difference discretization of the Hamiltonian defined in~\refequ{equ:5-experiments-electronic-hamiltonian} on a $10 \times 10$ grid.
+We ran the Chebyshev-Nyström++ method with and without estimating if the matrix function is zero before taking its pseudo-inverse.
 We let $m=2000$, $n_{\mtx{\Omega}}=80$, $n_{\mtx{\Psi}}=0$, and $\sigma = 0.004$.
@@
 \begin{theorem}{Error of Chebyshev-Nyström++ estimate}{chebyshev-nystrom}
-    For a smoothing parameter $0 < \sigma \ll 1$ \textcolor{red}{???
-Why do we need $\ll$ here?
-This is quite confusing because this is not a mathematical statement.
-Can this be simply replaced by $<$ or not?
-???}, let $\phi_{\sigma}$ be the smoothed spectral density~\eqref{equ:smooth-spectral-density} of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum contained in $[-1, 1]$.
-Consider $\varepsilon \in (0, 3)$ and $\delta \in (0, 1/2)$.
+    For a smoothing parameter $\sigma > 0$, let $\phi_{\sigma}$ be the smoothed spectral density~\refequ{equ:smooth-spectral-density} of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum contained in $[-1, 1]$.
+Consider $\varepsilon \in (0, 3)$ and $\delta \in (0, 2 e^{-1})$.
     Then the Chebyshev-Nyström++ estimate $\widetilde{\underline{\phi}}_{\sigma}^{(m)}$ 
@@
     \end{equation}
-    with probability at least $1 - \delta$ for $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$  and $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(\delta^{-1})^2)$.
+    with probability at least $1 - \delta$ for $m = \mathcal{O}((\log(\varepsilon^{-1}) + \log(\sigma^{-1})) / \log(\sigma + 1))$  and $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(\delta^{-1}))$.
 \end{theorem}
@@
     \end{align}
-    From \reflem{lem:non-negative-chebyshev-error} we see that if \textcolor{red}{$0 < \sigma \ll 1$ ???
-Why $\ll$?
-???}, choosing $m = \mathcal{O}(\sigma^{-1}(\log(\widetilde{\varepsilon}^{-1}) + \log(\sigma^{-1})))$ ensures $\underline{E}_{\sigma, m} \leq \widetilde{\varepsilon}$, and thus, if $\widetilde{\varepsilon} \leq 1$,
+    From \reflem{lem:non-negative-chebyshev-error} and the discussion after \reflem{lem:chebyshev-error} we see that the choice $m = \mathcal{O}((\log(\widetilde{\varepsilon}^{-1}) + \log(\sigma^{-1})) / \log(\sigma + 1))$ ensures $\underline{E}_{\sigma, m} \leq \widetilde{\varepsilon}$, and thus, if $\widetilde{\varepsilon} \leq 1$,
     \begin{equation}

### 2025-02-21 paper/numresults.tex
review numerical results

@@
 
-All our implementations are written from scratch in Python 3.12.3 using the packages NumPy 2.0.0 and SciPy 1.14.1.
+All our implementations are written in Python 3.12.3 using the packages NumPy 2.0.0 and SciPy 1.14.1.
 They are executed on a single thread of a GitHub-hosted Ubuntu runner with a 64-bit processor and 16 GB of RAM.
+\textcolor{red}{???
+Provide all parameter choices algorithmic choices common toa all experiments here.
+For example, the dismissal of non-negative approx ???}
+(evtl. unless otherwise mentioned 
+$n_t$ sampling points uniformly distributed.
+The error is calculcated with the composite midpoint rule using this sampling points.
+For transforming the interval to $[-1,1]$ exterior eigenvalues of $A$ are estimated using the Lanczos method (Python's ....).
 
@@
 
-Our first numerical example comes from electronic structure interaction \cite{lin-2017-randomized-estimation}, more precisely from the second order finite difference discretization of the Hamiltonian
+Our first numerical example comes from electronic structure interaction \textcolor{red}{???
+Provide more specific reference ???} \cite{lin-2017-randomized-estimation}, more precisely from the second order finite difference discretization of the Hamiltonian
 \begin{equation}
@@
 \reffig{fig:gaussian-well}).
-The discretization step is fixed to be $h=0.6$.
+The grid width is fixed to be $h=0.6$.
+For example, for $n_c = 1$, this leads to a matrix of size \textcolor{red}{???
+???}.
 This is an idealized model for the interaction of nuclei on a regular grid with electrons for a $k$-vector in the center of the first Brillouin zone.
@@
 While theoretically an important tool, for the remainder of this section, we waive using the non-negative Chebyshev approximation, since the (slight) indefiniteness of $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ \refequ{equ:matrix-approximation} does not seem to matter when computing the Nyström++ estimator in finite precision arithmetic.
-We approximate the integrals for computing the $L^1$-errors in \refsec{sec:analysis} using a midpoint quadrature with $n_t \in \mathbb{N}$ nodes.
+We approximate the integrals for computing the $L^1$-errors in \refsec{sec:analysis} using composite midpoint quadrature with $n_t \in \mathbb{N}$ uniformly distributed nodes.
 For a fixed total number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$, we analyze the convergence behavior of the Chebyshev-Nyström++ method when run on the $1000 \times 1000$ matrix resulting from the finite difference discretization of the Hamiltonian for $n_c = 1$ described above (see \reffig{fig:convergence}).
@@
 
-We run an optimized implementation of the Krylov-aware stochastic trace estimator \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} in multiple configurations on the example from \refsec{subsec:hamiltonian} and plot the approximation errors for logarithmically spaced values of the smoothing parameter $\sigma$ in \reffig{fig:krylov-aware-density}.
+We run our own, faithfully optimized implementation of the Krylov-aware stochastic trace estimator \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} in multiple configurations on the example from \refsec{subsec:hamiltonian} and plot the approximation errors for logarithmically spaced values of the smoothing parameter $\sigma$ in \reffig{fig:krylov-aware-density}.
 For reference, we add the error of the Chebyshev-Nyström++ method on parameters which lead to a comparable run-time.

### 2025-02-21 paper/spectraldensity.tex
review numerical results

@@
 
-\color{black}
-
-What we have found to be crucial for a successful approximation of the full spectral density was the detection of regions wherein lie no eigenvalues, i.e.
-regions where the spectral density nearly vanishes.
-For small smoothing parameters $\sigma$, the rapid decay of the Gaussian $g_{\sigma}$ will cause $\mtx{K}_1(t) \approx \mtx{\Omega}^{\top} g_{\sigma}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$ to be close to the zero matrix for parameter values $t$ located in such regions.
-Hence, on \reflin{lin:4-nystromchebyshev-nystrom-pp} we would compute the pseudoinverse of an almost zero matrix, which can be extremely unstable (see \reffig{fig:zerocheck}).
-Notice that with $\mtx{K}_1(t)$ --- a matrix we need to assemble anyway --- we can cheaply compute $\Trace(\mtx{K}_1(t)) / n_{\mtx{\Omega}}$, which is the $n_{\mtx{\Omega}}$-query Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}.
-If this estimate is significantly smaller than $1/n$, in our experiments usually around $10^{-5}$, we may use it as an indicator that $t$ lies within a region where the spectral density nearly vanishes, and therefore set $\widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) = 0$ instead of computing the full expression on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
+To ensure numerical accuracy, we found it crucial to separately treat regions
+where the spectral density nearly vanishes, because there are no eigenvalues in or close to the region.
+Especially for small smoothing parameters $\sigma$, the rapid decay of the Gaussian $g_{\sigma}$ will cause $\mtx{K}_1(t) \approx \mtx{\Omega}^{\top} g_{\sigma}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$ to be close to zero for parameter values $t$ located in such regions.
+Hence, on \reflin{lin:4-nystromchebyshev-nystrom-pp} we would compute the pseudoinverse of an almost zero matrix, which can be extremely unstable; see \reffig{fig:zerocheck}.
+To detect this situation, we can cheaply compute the Girard-Hutchinson
+estimate $\Trace(\mtx{K}_1(t)) / n_{\mtx{\Omega}}$ of the nuclear norm.
+If this estimate is small (in our experiments, smaller than $10^{-5}$), we use it as an indicator that $t$ is within a region with a nearly vanishing spectral density, and therefore set $\widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) = 0$ instead of computing the full expression on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
 
@@
 \begin{proof}
-Combining results derived above, we obtain that \color{black}
+Combining results derived above, we obtain that 
     \begin{align}
@@
     \end{align}
-    From \reflem{lem:non-negative-chebyshev-error} and the discussion after \reflem{lem:chebyshev-error} we see that the choice $m = \mathcal{O}((\log(\widetilde{\varepsilon}^{-1}) + \log(\sigma^{-1})) / \log(\sigma + 1))$ ensures $\underline{E}_{\sigma, m} \leq \widetilde{\varepsilon}$, and thus, if $\widetilde{\varepsilon} \leq 1$,
-    \begin{equation}
-        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq (1+\widetilde{\varepsilon}) \widetilde{\varepsilon} + \widetilde{\varepsilon} \leq 3 \widetilde{\varepsilon}
-    \end{equation}
-    from which the result follows by identifying $\varepsilon = 3 \widetilde{\varepsilon}$.
+    From \reflem{lem:non-negative-chebyshev-error} and the discussion after \reflem{lem:chebyshev-error} we see that the choice $m = \mathcal{O}((\log(\widetilde{\varepsilon}^{-1}) + \log(\sigma^{-1})) / \log(\sigma + 1))$ ensures $\underline{E}_{\sigma, m} \leq \widetilde{\varepsilon}$, and thus
+    the result follows by setting $\varepsilon = 3 \widetilde{\varepsilon}$.
+%    if $\widetilde{\varepsilon} \leq 1$,
+%    \begin{equation}
+%        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq (1+\widetilde{\varepsilon}) \widetilde{\varepsilon} + \widetilde{\varepsilon} \leq 3 \widetilde{\varepsilon}
+%    \end{equation}
+%    from which the result follows by identifying $\varepsilon = 3 \widetilde{\varepsilon}$.
 \end{proof}
@@
 
-In practice, the approximation error is often significantly better than guaranteed in \refthm{thm:chebyshev-nystrom}, particularly when $n_{\mtx{\Omega}}$ is chosen rather large.
-The reason is that for small $\sigma$ the singular values of the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ decay so quickly that already the Nyström approximation alone is highly accurate (cf.
+\color{black}
+In practice, the $\varepsilon^{-1}$ complexity of $n_{\mtx{\Omega}}$, $n_{\mtx{\Psi}}$ in \refthm{thm:chebyshev-nystrom} is often pessimistic.
+\textcolor{red}{Make briefer:
+\begin{itemize}
+ \item Quick decay of eigenvalues makes Nyström alone look good.
+ \item State singular values.
+ \item Explain as quickly as possible how you arrive at estimate of numerical rank for a roughly uniform distribution.
+w.l.o.g.
+eigenvalue $t$ around zero $\leadsto$ roughly distribution count number of eigenvalues $\approx \pm i/n$ needed for numerical rank $\epsilon$.
+ \item try to reduce notation
+\end{itemize}
+}
+In particular, for small $\sigma$ the singular values of the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ decay so quickly that the Nyström approximation alone is highly accurate (cf.
 \refthm{thm:nystrom}).

### 2025-02-21 paper/re-pro-badge.tex
automatically reproduce paper

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/7e1dc43}{7e1dc43} on 2025-02-20 at 14:53:12 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/37ec0b7}{37ec0b7} on 2025-02-21 at 17:02:20 UTC.

### 2025-02-21 paper/figures/numerical-rank.tex
generate plots which illustrate improvements
rework numerical results section and numerical rank explanation

@@
 \begin{tikzpicture}
-    \fill[lightblue] (-2, -0.15) rectangle (1.5, 0.15);
+    \fill[black!15!white] (-2.0, -0.15) rectangle (2.0, 0.15);
     \draw[thick, ->] (-5, 0) to (5, 0);
-    \fill[darkblue] (-4, 0) circle (0.075) node[above] {$\lambda_{(7)}$};
-    \fill[darkblue] (-3, 0) circle (0.075) node[above] {$\lambda_{(5)}$};
-    \fill[darkblue] (-2.3, 0) circle (0.075) node[above] {$\lambda_{(4)}$};
-    \fill[darkblue] (-1, 0) circle (0.075) node[above] {$\lambda_{(2)}$};
-    \fill[darkblue] (0.3, 0) circle (0.075) node[above] {$\lambda_{(1)}$};
-    \fill[darkblue] (1, 0) circle (0.075) node[above] {$\lambda_{(3)}$};
-    \fill[darkblue] (2.8, 0) circle (0.075) node[above] {$\lambda_{(6)}$};
-    \fill[darkblue] (3.6, 0) circle (0.075) node[above] {$\lambda_{(8)}$};
-    \fill[darkblue] (4.4, 0) circle (0.075) node[above] {$\lambda_{(9)}$};
-    \draw[darkblue, ultra thick] (-0.155, 0.15) to (-0.155, -0.15) node[below] {$t$};
-    \draw[darkblue, thick] (-2, 0.15) to (-2, -0.15) node[below] {$t - d_{\varepsilon, \sigma}$};
-    \draw[darkblue, thick] (1.5, 0.15) to (1.5, -0.15) node[below] {$t + d_{\varepsilon, \sigma}$};
+    \fill[black] (-4, 0) circle (0.075) node[below] {$\lambda_{(7)}$};
+    \fill[black] (-3, 0) circle (0.075) node[below] {$\lambda_{(5)}$};
+    \fill[black] (-2.35, 0) circle (0.075) node[below] {$\lambda_{(4)}$};
+    \fill[black] (-0.8, 0) circle (0.075) node[below] {$\lambda_{(2)}$};
+    \fill[black] (0.3, 0) circle (0.075) node[below] {$\lambda_{(1)}$};
+    \fill[black] (1, 0) circle (0.075) node[below] {$\lambda_{(3)}$};
+    \fill[black] (3, 0) circle (0.075) node[below] {$\lambda_{(6)}$};
+    \fill[black] (4.2, 0) circle (0.075) node[below] {$\lambda_{(8)}$};
+    \fill[black] (4.8, 0) circle (0.075) node[below] {$\lambda_{(9)}$};
+    \fill[black] (-4, 0) circle (0.075) node[above] {$\sigma_{7}$};
+    \fill[black] (-3, 0) circle (0.075) node[above] {$\sigma_{5}$};
+    \fill[black] (-2.35, 0) circle (0.075) node[above] {$\sigma_{4}$};
+    \fill[black] (-0.8, 1.3) circle (0.075) node[above left] {$\sigma_{2}$};
+    \fill[black] (0.3, 2.28) circle (0.075) node[above right] {$\sigma_{1}$};
+    \fill[black] (1, 0.91) circle (0.075) node[above right] {$\sigma_{3}$};
+    \fill[black] (3, 0) circle (0.075) node[above] {$\sigma_{6}$};
+    \fill[black] (4.2, 0) circle (0.075) node[above] {$\sigma_{8}$};
+    \fill[black] (4.8, 0) circle (0.075) node[above] {$\sigma_{9}$};
+    \draw[black, ultra thick] (-0.05, 0.15) to (-0.05, -0.55) node[below] {$t$};
+    \draw[black, thick] (-2.0, 0.15) to (-2.0, -0.55) node[below] {$t - d$};
+    \draw[black, thick] (2.0, 0.15) to (2.0, -0.55) node[below] {$t + d$};
+    \draw[black, thick, dashed] (-0.8, 0) to (-0.8, 1.3);
+    \draw[black, thick, dashed] (0.3, 0) to (0.3, 2.28);
+    \draw[black, thick, dashed] (1, 0) to (1, 0.91);
+    \draw[black!60!white, thick] (-5.000, 0.000) to (-4.899, 0.000) to (-4.798, 0.000) to (-4.697, 0.000) to (-4.596, 0.000) to (-4.495, 0.000) to (-4.394, 0.000) to (-4.293, 0.000) to (-4.192, 0.000) to (-4.091, 0.000) to (-3.990, 0.000) to (-3.889, 0.000) to (-3.788, 0.000) to (-3.687, 0.000) to (-3.586, 0.000) to (-3.485, 0.000) to (-3.384, 0.000) to (-3.283, 0.000) to (-3.182, 0.000) to (-3.081, 0.000) to (-2.980, 0.000) to (-2.879, 0.001) to (-2.778, 0.001) to (-2.677, 0.002) to (-2.576, 0.003) to (-2.475, 0.005) to (-2.374, 0.009) to (-2.273, 0.014) to (-2.172, 0.022) to (-2.071, 0.034) to (-1.970, 0.052) to (-1.869, 0.076) to (-1.768, 0.110) to (-1.667, 0.155) to (-1.566, 0.215) to (-1.465, 0.293) to (-1.364, 0.389) to (-1.263, 0.508) to (-1.162, 0.649) to (-1.061, 0.812) to (-0.960, 0.995) to (-0.859, 1.196) to (-0.758, 1.408) to (-0.657, 1.625) to (-0.556, 1.836) to (-0.455, 2.033) to (-0.354, 2.206) to (-0.253, 2.346) to (-0.152, 2.443) to (-0.051, 2.494) to (0.051, 2.494) to (0.152, 2.443) to (0.253, 2.346) to (0.354, 2.206) to (0.455, 2.033) to (0.556, 1.836) to (0.657, 1.625) to (0.758, 1.408) to (0.859, 1.196) to (0.960, 0.995) to (1.061, 0.812) to (1.162, 0.649) to (1.263, 0.508) to (1.364, 0.389) to (1.465, 0.293) to (1.566, 0.215) to (1.667, 0.155) to (1.768, 0.110) to (1.869, 0.076) to (1.970, 0.052) to (2.071, 0.034) to (2.172, 0.022) to (2.273, 0.014) to (2.374, 0.009) to (2.475, 0.005) to (2.576, 0.003) to (2.677, 0.002) to (2.778, 0.001) to (2.879, 0.001) to (2.980, 0.000) to (3.081, 0.000) to (3.182, 0.000) to (3.283, 0.000) to (3.384, 0.000) to (3.485, 0.000) to (3.586, 0.000) to (3.687, 0.000) to (3.788, 0.000) to (3.889, 0.000) to (3.990, 0.000) to (4.091, 0.000) to (4.192, 0.000) to (4.293, 0.000) to (4.394, 0.000) to (4.495, 0.000) to (4.596, 0.000) to (4.697, 0.000) to (4.798, 0.000) to (4.899, 0.000) to (5.000, 0.000);
+    \node[black!60!white] at (-1.3, 2.3) {$g_{\sigma}(t - \cdot)$};
 \end{tikzpicture}

### 2025-02-21 paper/intro.tex
rework numerical results section and numerical rank explanation

--- 2025-02-18 paper/intro.tex
@@
 We propose certain modifications to the methods from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}.
-In particular, we use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials and unify all of their methods in a single algorithm, with the desirable side effect of speeding up their \enquote{spectrum sweeping} method by orders of magnitude.
+In particular, we use a more rigorous approach for approximating matrix functions in terms of Chebyshev polynomials and unify all of the methods of \cite{lin-2017-randomized-estimation} in a single algorithm, with the desirable side effect of speeding up the \enquote{spectrum sweeping} method by orders of magnitude.
 Based on the analysis of the randomized trace estimators for parameter-dependent matrices, we provide bounds on the approximation error for our method.

### 2025-02-21 paper/numresults.tex
rework numerical results section and numerical rank explanation

@@
 
-In this last section, we test our algorithm in multiple scenarios.
-First, we numerically confirm our theoretical convergence bounds by approximating the spectral density of a Hamiltonian arising from electronic structure interaction.
-Subsequently, we compare our method to the Krylov-aware stochastic trace estimator, once on the electronic structure example, and once for estimating the trace of a more general parameter-dependent matrix appearing in statistical thermodynamics.
-In the end, we demonstrate that our algorithm can also be used to monitor the optimization process of a neural network.
+In this last section, we test the Chebyshev-Nyström++ algorithm in multiple scenarios leading from electronic structure interaction, statistical thermodynamics, and neural network optimization.
+
+While theoretically an important tool, for the remainder of this section, we waive using the non-negative Chebyshev approximation, since the (slight) indefiniteness of $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ \refequ{equ:matrix-approximation} does not seem to matter when computing the Nyström++ estimator in finite precision arithmetic.
+Unless otherwise mentioned, we transform the matrices such that all their eigenvalues are contained in $[-1, 1]$ by approximating the spectrum with NumPy's Hessian eigenvalue solver.
+We compute the spectral densities at $n_t = 100$ uniformly distributed values of the parameter $t$ in $[-1, 1]$.
+The integral involved in computing the $L^1$-errors is approximated with the composite midpoint quadrature rule using the $n_t$ values of $t$ as nodes.
+Both parameters introduced at the end \refsec{subsubsec:chebyshev-nystrom-implementation},the eigenvalue truncation threshold and the parameter for detecting a vanishing spectral density, are both set to $10^{-5}$.
 
@@
 They are executed on a single thread of a GitHub-hosted Ubuntu runner with a 64-bit processor and 16 GB of RAM.
-\textcolor{red}{???
-Provide all parameter choices algorithmic choices common toa all experiments here.
-For example, the dismissal of non-negative approx ???}
-(evtl. unless otherwise mentioned 
-$n_t$ sampling points uniformly distributed.
-The error is calculcated with the composite midpoint rule using this sampling points.
-For transforming the interval to $[-1,1]$ exterior eigenvalues of $A$ are estimated using the Lanczos method (Python's ....).
 
@@
 
-Our first numerical example comes from electronic structure interaction \textcolor{red}{???
-Provide more specific reference ???} \cite{lin-2017-randomized-estimation}, more precisely from the second order finite difference discretization of the Hamiltonian
+Our first numerical example comes from electronic structure interaction \cite[Section 6]{lin-2017-randomized-estimation}, more precisely from the second order finite difference discretization of the Hamiltonian
 \begin{equation}
@@
 The grid width is fixed to be $h=0.6$.
-For example, for $n_c = 1$, this leads to a matrix of size \textcolor{red}{???
-???}.
+For example, for $n_c = 1$, this leads to a matrix of size $1000 \times 1000$.
 This is an idealized model for the interaction of nuclei on a regular grid with electrons for a $k$-vector in the center of the first Brillouin zone.
@@
 
-While theoretically an important tool, for the remainder of this section, we waive using the non-negative Chebyshev approximation, since the (slight) indefiniteness of $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ \refequ{equ:matrix-approximation} does not seem to matter when computing the Nyström++ estimator in finite precision arithmetic.
-We approximate the integrals for computing the $L^1$-errors in \refsec{sec:analysis} using composite midpoint quadrature with $n_t \in \mathbb{N}$ uniformly distributed nodes.
 For a fixed total number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$, we analyze the convergence behavior of the Chebyshev-Nyström++ method when run on the $1000 \times 1000$ matrix resulting from the finite difference discretization of the Hamiltonian for $n_c = 1$ described above (see \reffig{fig:convergence}).
+As a reference, we use the eigenvalues computed by NumPy's Hermitian eigenvalue solver applied to the matrix $\mtx{A}$.
 
-\begin{figure}[!htpb]
+\begin{figure}[ht]
     \centering
@@
     \caption{For increasing values of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ but fixed $m$, we apply the Chebyshev-Nyström++ method to the Hamiltonian matrix described in \refsec{subsec:hamiltonian}.
-We plot the $L^1$-approximation error for $\sigma=0.005$ on the model problem with $n_c = 1$.
-The $L^1$-error is approximated by a quadrature with $n_t = 100$ evenly spaced nodes in $[-1, 1]$.}
+We plot the $L^1$-approximation error for $\sigma=0.005$ on the model problem with $n_c = 1$.}
     \label{fig:convergence}
@@
 Indeed, we observe that for $n_{\mtx{\Omega}} = 0$, we require $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ to achieve an error of order $\varepsilon$, as shown in \refthm{thm:hutchinson}.
-Because the eigenvalues of the matrix are quite uniformly distributed, we observe the stronger convergence of the Nyström approximation discussed in the end of \refsec{sec:application} once $n_{\mtx{\Omega}}$ exceeds the numerical rank \refequ{equ:gaussian-kernel-numerical-rank-uniform}, which in this case would be somewhat larger than $30$.
+Because the eigenvalues of the matrix are quite uniformly distributed, we observe the stronger convergence of the Nyström approximation discussed in the end of \refsec{sec:application} once $n_{\mtx{\Omega}}$ is large enough.
 
@@
 
-\begin{figure}[!htpb]
+\begin{figure}[ht]
     \centering
     \input{plots/distribution.pgf}
-    \caption{The Chebyshev-Nyström++ method is applied to the model problem with $n_c = 1$ for different ways of allocations a total of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation for the Gaussian smoothing kernel with multiple different values of the smoothing parameter.
-We make the approximation error made in the Chebyshev approximation negligible by rescaling $m=16 / \sigma$ (based on \reflem{lem:non-negative-chebyshev-error}).
-The $L^1$-error is approximated by a quadrature with $n_t = 100$ evenly spaced nodes in $[-1, 1]$.}
+    \caption{For a range of values of $\sigma$, the Chebyshev-Nyström++ method is applied to the Hamiltonian matrix described in \refsec{subsec:hamiltonian} with $n_c = 1$ for different ways of allocating a total of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation.
+We make the approximation error committed by the Chebyshev approximation negligible by rescaling $m=16 / \sigma$ (based on \reflem{lem:non-negative-chebyshev-error}).}
     \label{fig:distribution}
@@
 
-The Krylov-aware stochastic trace estimation \cite{chen-2023-krylovaware-stochastic} is also well suited for approximating the trace of a matrix function $f(\mtx{A}, t)$ which separately depends on a parameter $t$.
+The Krylov-aware stochastic trace estimator \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} is also suitable in the context of spectral density estimation.
 It also samples two Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$.
-$\mtx{\Omega}$ is used to extract a low-rank approximation to $f(\mtx{A}, t)$ from a block-Krylov subspace of $\mtx{A}$ generated from $r$ iterations of the block-Lanczos method with reorthogonalization and subsequently $q$ without.
-$\mtx{\Psi}$ is then used to estimate the trace of the approximation residual with $r$ iterations of the Lanczos method.
+$\mtx{\Omega}$ is used to extract a low-rank approximation to $g_{\sigma}(t \mtx{I}_n - \mtx{A})$ by running the block-Lanczos algorithm on $\mtx{A}$ with starting block $\mtx{\Omega}$ for $r$ iterations with reorthogonalization and subsequently $q$ without.
+The columns of $\mtx{\Psi}$ are then used to estimate the trace of the approximation residual with $r$ iterations of the Lanczos algorithm on $\mtx{A}$.
 
@@
 
-To demonstrate that the Chebyshev-Nyström++ estimator is also effective on problems unrelated to spectral density estimation, we consider the problem of approximating $\Trace(\exp(-\beta \mtx{A}))$ for the Hamiltonian $\mtx{A}$ of a planar Heisenberg spin chain \cite{chen-2023-krylovaware-stochastic}.
+\subsection{Spectral density of Hessian matrix of neural network}
+\label{subsec:hessian}
+
+When optimizing a neural network over some data, it is of great interest to determine whether a minimum has been found and if this minimum is robust, i.e.
+a small perturbation of the parameters of the neural network will not lead to a significant decrease in the fit of the neural network.
+Both properties are reflected in the eigenvalues of the Hessian matrix $\mtx{A}$ with respect to some loss function: if all eigenvalues are non-negative, then the loss attains a local minimum, and if additionally all of them are small, the minimum is robust.
+
+Since neural networks are usually parametrized by thousands of parameters, assembling the Hessian matrix explicitly and then computing its eigenvalues is out of question.
+Luckily, there exists a method which can exactly compute matrix-vector products $\mtx{A} \vct{x}$ for any vector $\vct{x}$ which scales proportionally with the number of parameters \cite{pearlmutter-1994-fast-exact}.
+
+To demonstrate that our algorithm can effectively be applied in the setting of neural network optimization, we approximate the spectral density of a small fully connected convolutional neural network with 6782 parameters.
+We train this network on the handwritten digit classification task given by the MNIST dataset\footnote{Handwritten digit classification; taken from \url{http://yann.lecun.com/exdb/mnist}.} in PyTorch 2.4.1 in a standard way.
+We use the bound from \cite[Theorem 1]{zhou-2011-bounding-spectrum} to estimate a bound on the spectrum.
+
+We plot the approximation of the spectral density of the Hessian matrix of the untrained neural network, and at different stages of training in \reffig{fig:hessian-density}, as well as the corresponding mean squared error loss.
+It can well be observed that the eigenvalues creep towards zero as the training proceeds.
+Furthermore, despite the loss steadily decreasing, the presence of relatively large eigenvalues in some epochs indicates a sharp minimum of the network, hence, unfavorable generalization properties.
+
+\begin{figure}
+    \begin{minipage}[c]{.49\linewidth}
+        \centering
+        \input{plots/hessian_density.pgf}
+    \end{minipage}\hfill%
+    \begin{minipage}[c]{.49\linewidth}
+        \centering
+        \input{plots/hessian_density_loss.pgf}
+    \end{minipage}
+    \caption{The mean squared error training loss (right) and the corresponding approximate spectral density of the Hessian matrix of a fully connected convolutional neural network in different epochs of training on the MNIST dataset (left).
+The spectral density is approximated in $n_t=150$ uniformly spaced points using the Chebyshev-Nyström++ method (\refalg{alg:nystrom-chebyshev-pp}) with parameters $n_{\mtx{\Omega}} = 10$, $n_{\mtx{\Psi}} = 10$, $m = 1000$, and $\sigma = 0.005$.}
+    \label{fig:hessian-density}
+\end{figure}
+
+\subsection{Approximation of the partition function of a quantum system}
+\label{subsec:spin-chain}
+
+To demonstrate that the Chebyshev-Nyström++ estimator is also effective on problems unrelated to spectral density estimation, we consider the problem of approximating the partition function $\Trace(\exp(-\beta \mtx{A}))$ for the Hamiltonian $\mtx{A}$ of a planar Heisenberg spin chain \cite[Section 5.1]{chen-2023-krylovaware-stochastic}.
 We compute the pointwise error from the analytic solution as a function of the temperature $\beta^{-1}$ and compare it to the error achieved by the Krylov-aware trace estimator in \reffig{fig:krylov-aware-spin}.
@@
 
-Unlike in the spectral density example from \refsec{subsec:hamiltonian}, the numerical rank of the matrix we approximate wildly changes with the parameter $\beta$.
-Since the Chebyshev-Nyström++ method requires a fixed allocation of random vectors to the low-rank approximation for all values of the parameter $\beta$.
+Unlike in the spectral density example from \refsec{subsec:hamiltonian}, where the numerical rank of the matrix stays low for all $t$, it changes wildly with the parameter $\beta$.
+The Chebyshev-Nyström++ method requires a fixed allocation of random vectors to the low-rank approximation for all values of the parameter $\beta$.
 Hence, both for large $\beta$, where the matrix $\exp(-\beta \mtx{A})$ is low-rank, and for small $\beta$, where the singular values only decay slowly, we cannot expect a good approximation.
-
-\subsection{Spectral density of Hessian matrix of neural network}
-\label{subsec:hessian}
-
-When optimizing a neural network over some data, it is of great interest to determine whether a minimum has been found and if this minimum is robust, i.e.
-a small perturbation of the parameters of the neural network will not lead to a significant decrease in the fit of the neural network.
-Both properties are reflected in the eigenvalues of the Hessian matrix $\mtx{A}$ with respect to some loss function: if all eigenvalues are non-negative, then the loss attains a local minimum, and if additionally all of them are small, the minimum is robust.
-
-Since neural networks are usually parametrized by thousands of parameters, assembling the Hessian matrix explicitly and then computing its eigenvalues is out of question.
-Luckily, there exists a method which can exactly compute matrix-vector products $\mtx{A} \vct{x}$ for any vector $\vct{x}$ which scales proportionally with the number of parameters \cite{pearlmutter-1994-fast-exact}.
-
-To demonstrate that our algorithm can effectively be applied in the setting of neural network optimization, we approximate the spectral density of a small fully connected convolutional neural network with 6782 parameters.
-We train this network on the handwritten digit classification task given by the MNIST dataset\footnote{Handwritten digit classification; taken from \url{http://yann.lecun.com/exdb/mnist}.} in PyTorch 2.4.1 in a standard way.
-We plot the approximation of the spectral density of the Hessian matrix of the untrained neural network, and at different stages of training in \reffig{fig:hessian-density}, as well as the corresponding mean squared error loss.
-It can well be observed that the eigenvalues creep towards zero as the training proceeds.
-Furthermore, despite the loss steadily decreasing, the presence of large eigenvalues in some epochs indicates a sharp minimum of the network, hence, unfavorable generalization properties.
-
-\begin{figure}
-    \begin{minipage}[c]{.49\linewidth}
-        \centering
-        \input{plots/hessian_density.pgf}
-    \end{minipage}\hfill%
-    \begin{minipage}[c]{.49\linewidth}
-        \centering
-        \input{plots/hessian_density_loss.pgf}
-    \end{minipage}
-    \caption{The mean squared error training loss (right) and the corresponding approximate spectral density of the Hessian matrix of a fully connected convolutional neural network in different epochs of training on the MNIST dataset (left).
-Computed with the Chebyshev-Nyström++ method (\refalg{alg:nystrom-chebyshev-pp}) on $n_{\mtx{\Omega}} = 10$, $n_{\mtx{\Psi}} = 10$, $m = 1000$, and $\sigma = 0.005$.}
-    \label{fig:hessian-density}
-\end{figure}

### 2025-02-21 paper/spectraldensity.tex
rework numerical results section and numerical rank explanation

@@
 \color{black}
-In practice, the $\varepsilon^{-1}$ complexity of $n_{\mtx{\Omega}}$, $n_{\mtx{\Psi}}$ in \refthm{thm:chebyshev-nystrom} is often pessimistic.
-\textcolor{red}{Make briefer:
-\begin{itemize}
- \item Quick decay of eigenvalues makes Nyström alone look good.
- \item State singular values.
- \item Explain as quickly as possible how you arrive at estimate of numerical rank for a roughly uniform distribution.
+In practice, the $\varepsilon^{-1}$ complexity of $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$ in \refthm{thm:chebyshev-nystrom} is often pessimistic.
+The reason is that \refthm{thm:nystrom-pp} does not take advantage of the low-rank structure of the matrix $\mtx{B}(t) = g_{\sigma}(t\mtx{I}_n - \mtx{A})$: The singular values of the matrix $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ are given by
+\begin{equation}
+    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) = \frac{1}{\sqrt{2 \pi \sigma^2}} e^{-\sfrac{\left(t - \lambda_{(i)}\right)^2}{2 \sigma^2}},
+    \label{equ:gaussian-kernel-eigenvalues}
+\end{equation}
+where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from $t$, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t)$.
+Clearly, the further away an eigenvalue $\lambda_{(i)}$ is from $t$, the (exponentially) smaller its associated singular value $\sigma_i(t)$ will be.
+Hence, all singular values $\sigma_i(t)$ corresponding to eigenvalues outside of a certain window $[t - d, t + d]$ around $t$, where $d$ is a number depending on $\sigma$ and $n$, will be negligibly small (see \reffig{fig:numerical-rank}).
+\begin{figure}[ht]
+    \centering
+    \input{figures/numerical-rank.tex}
+    \caption{Singular values $\sigma_i$ of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ corresponding to eigenvalues $\lambda_{(i)}$ of $\mtx{A}$ which lie further away from $t$ than a certain distance $d$ are negligibly small.}
+    \label{fig:numerical-rank}
+\end{figure}
+If we choose $n_{\mtx{\Omega}}$ large enough such that the sum $\sum_{i=r+1}^{n} \sigma_i(t)$ which bounds the approximation error in \refthm{thm:nystrom} is negligibly small for all $t$, we achieve a much better approximation.
+In particular, if we assume the eigenvalues of $\mtx{A}$ to be uniformly distributed, a small calculation shows that $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma)$ will ensure this.
+
+%\textcolor{red}{Make briefer:
+%\begin{itemize}
+% \item Quick decay of eigenvalues makes Nyström alone look good.
+% \item State singular values.
+% \item Explain as quickly as possible how you arrive at estimate of numerical rank for a roughly uniform distribution.
 w.l.o.g.
 eigenvalue $t$ around zero $\leadsto$ roughly distribution count number of eigenvalues $\approx \pm i/n$ needed for numerical rank $\epsilon$.
- \item try to reduce notation
-\end{itemize}
-}
-In particular, for small $\sigma$ the singular values of the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ decay so quickly that the Nyström approximation alone is highly accurate (cf.
+% \item try to reduce notation
+%\end{itemize}
+%}
+%In particular, for small $\sigma$ the singular values of the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ decay so quickly that the Nyström approximation alone is highly accurate (cf.
 \refthm{thm:nystrom}).
 To establish a relation between the singular value decay of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ and what choice of $n_{\mtx{\Omega}}$ is needed for a significantly better approximation than ensured by \refthm{thm:chebyshev-nystrom}, we study the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$.\footnote{The factor $1/n$
-does not change the conclusion, but is added to simplify the calculations and also reflects how we implement the method: We absorb the factor $1/n$ from the smoothed spectral density \refequ{equ:smooth-spectral-density} in the definition of $g_{\sigma}$, such that in reality, the Chebyshev approximation and the Nyström++ estimator are applied to $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$.}
+%does not change the conclusion, but is added to simplify the calculations and also reflects how we implement the method: We absorb the factor $1/n$ from the smoothed spectral density \refequ{equ:smooth-spectral-density} in the definition of $g_{\sigma}$, such that in reality, the Chebyshev approximation and the Nyström++ estimator are applied to $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$.}
 
-We can express the numerical rank of a matrix with respect to the nuclear norm in terms of its singular values $\sigma_1 \geq \dots \geq \sigma_n \geq 0$ as
-\begin{equation}
-    r_{\varepsilon_m} = \min \{1 \leq r \leq n: \sum_{j=r+1}^n \sigma_{j} \leq \varepsilon_m \},
-    \label{equ:numerical-rank}
-\end{equation}
-where $\varepsilon_m$ is usually taken to be the double machine precision, i.e. $\varepsilon_m \approx 10^{-16}$.
+%We can express the numerical rank of a matrix with respect to the nuclear norm in terms of its singular values $\sigma_1 \geq \dots \geq \sigma_n \geq 0$ as
+%\begin{equation}
+%    r_{\varepsilon_m} = \min \{1 \leq r \leq n: \sum_{j=r+1}^n \sigma_{j} \leq \varepsilon_m \},
+%    \label{equ:numerical-rank}
+%\end{equation}
+%where $\varepsilon_m$ is usually taken to be the double machine precision, i.e. $\varepsilon_m \approx 10^{-16}$.
 
@@
 
-When using a Gaussian smoothing-kernel, we can explicitly write the singular values of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$ as
-\begin{equation}
-    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) / n = \frac{1}{n \sqrt{2 \pi \sigma^2}} e^{-\sfrac{\left(t - \lambda_{(i)}\right)^2}{2 \sigma^2}},
-    \label{equ:gaussian-kernel-eigenvalues}
-\end{equation}
-where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from spectral-parameter $t$, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t)$.
+%When using a Gaussian smoothing-kernel, we can explicitly write the singular values of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$ as
+%\begin{equation}
+%    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) / n = \frac{1}{n \sqrt{2 \pi \sigma^2}} e^{-\sfrac{\left(t - \lambda_{(i)}\right)^2}{2 \sigma^2}},
+%    \label{equ:gaussian-kernel-eigenvalues}
+%\end{equation}
+%where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from spectral-parameter $t$, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t)$.
 Consequently, by using the closed-form expression of the eigenvalues \refequ{equ:gaussian-kernel-eigenvalues}, we may generously upper bound the numerical rank by
-\begin{equation}
-    r_{\varepsilon_m}(g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n) \leq \#\left\{1\leq i\leq n: |t - \lambda_i| < d_{\varepsilon_m, \sigma} \right\},
-    \label{equ:gaussian-kernel-numerical-rank}
-\end{equation}
-with $d_{\varepsilon_m, \sigma} = \sigma \sqrt{-2 \log(\sqrt{2 \pi} \sigma \varepsilon_m)}$ and the eigenvalues $\lambda_1, \dots, \lambda_n$ of $\mtx{A}$ and the constants.
+%\begin{equation}
+%    r_{\varepsilon_m}(g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n) \leq \#\left\{1\leq i\leq n: |t - \lambda_i| < d_{\varepsilon_m, \sigma} \right\},
+%    \label{equ:gaussian-kernel-numerical-rank}
+%\end{equation}
+%with $d_{\varepsilon_m, \sigma} = \sigma \sqrt{-2 \log(\sqrt{2 \pi} \sigma \varepsilon_m)}$ and the eigenvalues $\lambda_1, \dots, \lambda_n$ of $\mtx{A}$ and the constants.
 The expression \refequ{equ:gaussian-kernel-numerical-rank} has a nice visual interpretation: The numerical rank of a matrix is at most equal to the number of eigenvalues which are closer to the parameter $t$ than $d_{\varepsilon_m, \sigma}$.
 This is illustrated in \reffig{fig:numerical-rank}.
-\begin{figure}[ht]
-    \centering
-    \input{figures/numerical-rank.tex}
-    \caption{The numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$ can be approximately computed by counting the number of eigenvalues $\lambda_{(1)}, \dots, \lambda_{(n)}$ of the matrix $\mtx{A}$ which lie less than a constant $d_{\varepsilon_m, \sigma}$ away from $t$.}
-    \label{fig:numerical-rank}
-\end{figure}
+%\begin{figure}[ht]
+%    \centering
+%    \input{figures/numerical-rank.tex}
+%    \caption{The numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$ can be approximately computed by counting the number of eigenvalues $\lambda_{(1)}, \dots, \lambda_{(n)}$ of the matrix $\mtx{A}$ which lie less than a constant $d_{\varepsilon_m, \sigma}$ away from $t$.}
+%    \label{fig:numerical-rank}
+%\end{figure}
 
-If we additionally assume the eigenvalues of the matrix $\mtx{A}$ to be evenly distributed within $[-1, 1]$, that is, in any subinterval of fixed length in $[a, b]$, $-1 \leq a < b \leq 1$, we can expect to find roughly the same number of eigenvalues, then we can estimate the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$ to be
-\begin{equation}
-    r_{\varepsilon_m}(g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n) \lessapprox n d_{\varepsilon_m, \sigma} \approx 6 n \sigma,
-    \label{equ:gaussian-kernel-numerical-rank-uniform}
-\end{equation}
-provided we set $\varepsilon_m = 10^{-16}$ and choose $\sigma$ roughly between $10^{-8}$ and $1$.
+%If we additionally assume the eigenvalues of the matrix $\mtx{A}$ to be evenly distributed within $[-1, 1]$, that is, in any subinterval of fixed length in $[a, b]$, $-1 \leq a < b \leq 1$, we can expect to find roughly the same number of eigenvalues, then we can estimate the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$ to be
+%\begin{equation}
+%    r_{\varepsilon_m}(g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n) \lessapprox n d_{\varepsilon_m, \sigma} \approx 6 n \sigma,
+%    \label{equ:gaussian-kernel-numerical-rank-uniform}
+%\end{equation}
+%provided we set $\varepsilon_m = 10^{-16}$ and choose $\sigma$ roughly between $10^{-8}$ and $1$.
 Hence, choosing $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma)$ will yield approximations which can be significantly better than \refthm{thm:chebyshev-nystrom}.

### 2025-02-24 paper/intro.tex
comments from final review pass

--- 2025-02-18 paper/intro.tex
@@
 All methods considered in this work are based on the following simple idea: Apply an existing randomized trace estimator to $\Trace(\mtx{B}(t))$ with \emph{constant} random vectors, that is, the same randomization is used for each value of the parameter $t$.
-For example, the Girard-Hutchinson estimator becomes $n_{\mtx{\Psi}}^{-1} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j$ for $n_{\mtx{\Psi}}$ constant Gaussian random vectors $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}}$.
-??? STOP HERE ???
+For example, the Girard-Hutchinson estimator becomes $\mathsf{est}(t):=n_{\mtx{\Psi}}^{-1} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j$ for $n_{\mtx{\Psi}}$ constant Gaussian random vectors $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}}$.
+We will use the $L^1$-norm $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \mathsf{est}(t) |~\mathrm{d}t$ to measure the error of such an estimator for all $t$ in an interval by $[a,b]$; similar to the approach taken in~\cite{kressner-2024-randomized-lowrank} for low-rank approximation.
 
-\textcolor{red}{???
-The following two paragraphs need to rewritten a bit at a later stage.
-It needs to be clearer what the approaches proposed in our paper are really about and the novelty wrt existing work has to come out more clearly and concretely.
+\paragraph{New contributions.} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices with constant randomness (with respect to $t$): The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}.
+Combined with Chebyshev approximation, this allows one to reuse the majority of computations across different values of the parameter $t$, making the estimators scale favorably with respect to the number of parameter evaluations.
+We propose several additional modifications to the methods from \cite{lin-2017-randomized-estimation} for the special case of approximating the smoothed spectral density $\phi_{\sigma}$ from~\refequ{equ:smoothed-spectral-density}.
+In particular, we use a more rigorous approach that preserves non-negativity for approximating matrix functions in terms of Chebyshev polynomials.
+The resulting Chebyshev-Nyström++ estimator
+unifies all methods from~\cite{lin-2017-randomized-estimation} in a single algorithm, with the desirable side effect of \textcolor{red}{speeding up the \enquote{spectrum sweeping} method by orders of magnitude.
 ???
-Mention \cite{park-2024-lowrank-approximation} ?.}
-\color{black}
-Instead of discretizing $[a, b]$ into $t_1, \ldots, t_m$, approximating $\Trace(\mtx{B}(t_i))$ with $b(t_i)$, and analyzing the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way is to quantify the error with $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate in each $t \in [a, b]$.
-This approach has already been used in the analysis of low-rank approximations to parameter-dependent matrices \cite{kressner-2024-randomized-lowrank}.
+Where do we actually demonstrate this?
+Illustrate/spell this out more explicitly in the numerical experiments ???} Combining our analysis of the randomized trace estimators with standard error bounds for Chebyshev approximation leads to Theorem~\refthm{thm:chebyshev-nystrom}, which shows that $\mathcal O(\varepsilon^{-1})$ matrix-vector products suffice to guarantee an error $\varepsilon$.
 
-\paragraph{New contributions.} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}.
-Interestingly, we can reuse the same randomization for computing the estimate at each value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations.
-We propose certain modifications to the methods from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}.
-In particular, we use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials and unify all of their methods in a single algorithm, with the desirable side effect of speeding up their \enquote{spectrum sweeping} method by orders of magnitude.
-Based on the analysis of the randomized trace estimators for parameter-dependent matrices, we provide bounds on the approximation error for our method.
-
-\paragraph{Structure.} In \refsec{sec:analysis} we will analyze three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter-dependent matrices.
-We propose and analyze an algorithm for using the aforementioned estimators to approximate the spectral density of real symmetric matrices in \refsec{sec:application}.
-Finally, we will test our developments on an example from electronic structure calculation in \refsec{sec:results}.
+\paragraph{Structure.} In \refsec{sec:analysis} we will analyze the error of three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter-dependent matrices with constant random matrix.
+In \refsec{sec:application}, we propose and analyze the Chebyshev-Nyström++ estimator, which uses the these estimators to approximate the smoothed spectral density of real symmetric matrices.
+Finally, we will illustrate and validate our developments for examples from various applications in \refsec{sec:results}.
 

### 2025-02-24 paper/numresults.tex
comments from final review pass

@@
 
-\color{black}
+In this section, we verify the performance the Chebyshev-Nyström++ algorithm proposed in this paper in various application scenarios, from electronic structure interaction, statistical thermodynamics, and neural network optimization.
 
-In this last section, we test the Chebyshev-Nyström++ algorithm in multiple scenarios leading from electronic structure interaction, statistical thermodynamics, and neural network optimization.
+While theoretically an important tool, we have observed that the preservation of non-negativity discussed in Section~\ref{sec:preservnonneg} is not needed in practice.
+The (slight) indefiniteness of the standard Chebyshev approximation $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ from~\refequ{equ:matrix-approximation} does not seem to impede the reliability of the Nyström++ estimator.
+Unless otherwise stated, we transform the input matrices such that their eigenvalues are contained in $[-1, 1]$, by approximating their spectrum with NumPy's \textcolor{red}{Hessian ???
+replace by sparse?
+???} eigenvalue solver.
+We compute spectral densities at $n_t = 100$ uniformly distributed values of the parameter $t \in [-1, 1]$.
+The integral involved in computing the $L^1$-errors is approximated with the composite midpoint quadrature rule using the $n_t$ values of $t$ as nodes.
+Both parameters introduced at the end \refsec{subsubsec:chebyshev-nystrom-implementation},the eigenvalue truncation threshold and the parameter for detecting a vanishing spectral density, are fixed to $10^{-5}$.
 
-While theoretically an important tool, for the remainder of this section, we waive using the non-negative Chebyshev approximation, since the (slight) indefiniteness of $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ \refequ{equ:matrix-approximation} does not seem to matter when computing the Nyström++ estimator in finite precision arithmetic.
-Unless otherwise mentioned, we transform the matrices such that all their eigenvalues are contained in $[-1, 1]$ by approximating the spectrum with NumPy's Hessian eigenvalue solver.
-We compute the spectral densities at $n_t = 100$ uniformly distributed values of the parameter $t$ in $[-1, 1]$.
-The integral involved in computing the $L^1$-errors is approximated with the composite midpoint quadrature rule using the $n_t$ values of $t$ as nodes.
-Both parameters introduced at the end \refsec{subsubsec:chebyshev-nystrom-implementation},the eigenvalue truncation threshold and the parameter for detecting a vanishing spectral density, are both set to $10^{-5}$.
-
-All our implementations are written in Python 3.12.3 using the packages NumPy 2.0.0 and SciPy 1.14.1.
+Our implementations are written in Python 3.12.3 using the packages NumPy 2.0.0 and SciPy 1.14.1.
 They are executed on a single thread of a GitHub-hosted Ubuntu runner with a 64-bit processor and 16 GB of RAM.
@@
 
-Our first numerical example comes from electronic structure interaction \cite[Section 6]{lin-2017-randomized-estimation}, more precisely from the second order finite difference discretization of the Hamiltonian
+We use the electronic structure interaction example from~\cite[Section 6]{lin-2017-randomized-estimation}, which involves the second order finite difference discretization of the Hamiltonian
 \begin{equation}
@@
 in three dimensions. The potential $V$ interacting with the electrons is generated by Gaussian wells
-\begin{equation}
-    v(r) = v_0 e^{-\lambda r^2}
-    \label{equ:5-experiments-gaussian-cell}
-\end{equation}
-with $v_0 = -4$ and $\lambda = 8$, centered in cells of side length $L=6$ which are stacked $n_c \in \mathbb{N}$ times in each spatial dimension (cf.
-\reffig{fig:gaussian-well}).
-The grid width is fixed to be $h=0.6$.
+%\begin{equation}
+    $v(r) = v_0 e^{-\lambda r^2}$, 
+%    \label{equ:5-experiments-gaussian-cell}
+%\end{equation}
+with $v_0 = -4$ and $\lambda = 8$, centered in cells of side length $L=6$ which are stacked $n_c \in \mathbb{N}$ times in each spatial dimension; see \reffig{fig:gaussian-well}.
+The grid width of the finite difference discretization is fixed to $h=0.6$.
 For example, for $n_c = 1$, this leads to a matrix of size $1000 \times 1000$.
-This is an idealized model for the interaction of nuclei on a regular grid with electrons for a $k$-vector in the center of the first Brillouin zone.
-The distribution of the eigenvalues of the Hamiltonian --- its spectral density --- allows us to interpret the system's energy levels.
+This example represents an idealized model for the interaction of nuclei on a regular grid with electrons for a $k$-vector in the center of the first Brillouin zone.
+The distribution of the eigenvalues of the Hamiltonian --- its spectral density --- allows one to interpret the system's energy levels.
+
+\textcolor{red}{???
+This discussion of $n_c$ only makes sense if we have at least one experiment with $n_c = 5$ or so.
+Could you please include such an experiment?
+For example, do Figure 4.2 again for larger $n_c$?
+Just use an overapproximation (large $n_\Psi$, $n_\Omega$) for approximating the true spectral distribution.
+???}
 
@@
     \end{subfigure}
-    \caption{Cross-sections of the periodic Gaussian well potential $V$ for different sizes $n_c$ of the supercell.}
+    \caption{Cross-sections of the periodic Gaussian wells potential $V$ for different $n_c$.}
     \label{fig:gaussian-well}
@@
 
-For a fixed total number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$, we analyze the convergence behavior of the Chebyshev-Nyström++ method when run on the $1000 \times 1000$ matrix resulting from the finite difference discretization of the Hamiltonian for $n_c = 1$ described above (see \reffig{fig:convergence}).
-As a reference, we use the eigenvalues computed by NumPy's Hermitian eigenvalue solver applied to the matrix $\mtx{A}$.
-
+\reffig{fig:convergence} displays the convergence behavior of Chebyshev-Nyström++ applied to the $1000 \times 1000$ matrix resulting from the discretization of the Hamiltonian for $n_c = 1$.
+As a reference, we use the eigenvalues computed by NumPy's dense eigenvalue solver.
 \begin{figure}[ht]
@@
     \input{plots/convergence.pgf}
-    \caption{For increasing values of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ but fixed $m$, we apply the Chebyshev-Nyström++ method to the Hamiltonian matrix described in \refsec{subsec:hamiltonian}.
-We plot the $L^1$-approximation error for $\sigma=0.005$ on the model problem with $n_c = 1$.}
+    \caption{Error vs.
+number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ (with fixed $\sigma=0.005$ and \textcolor{red}{$m$ ???
+Provide value???}) when applying  Chebyshev-Nyström++ to the Hamiltonian matrix from~\refsec{subsec:hamiltonian} with $n_c = 1$.}
     \label{fig:convergence}
-\end{figure}
+\end{figure}%
+As expected from \refthm{thm:hutchinson}, Girard-Hutchinson alone ($n_{\mtx{\Omega}} = 0$) require $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ samples to achieve an error of order $\varepsilon$.
+Because the eigenvalues of the matrix are spread out, we observe the better convergence of the Nyström approximation discussed at the end of \refsec{sec:application} once $n_{\mtx{\Omega}}$ is sufficiently large.
+\textcolor{red}{???
+Comment explicitly that and why the errors levels off at around $10^{-6}$.
+Is it because of Chebyshev or some other truncation, etc.?
+???}
 
-Indeed, we observe that for $n_{\mtx{\Omega}} = 0$, we require $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ to achieve an error of order $\varepsilon$, as shown in \refthm{thm:hutchinson}.
-Because the eigenvalues of the matrix are quite uniformly distributed, we observe the stronger convergence of the Nyström approximation discussed in the end of \refsec{sec:application} once $n_{\mtx{\Omega}}$ is large enough.
-
-For a fixed budget of random vectors $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}$, \reffig{fig:distribution} visualizes how for different values of the smoothing parameter $\sigma$ the random vectors should either be invested into the Nyström approximation $n_{\mtx{\Omega}}$ or the estimation of the residual trace $n_{\mtx{\Psi}}$.
-Again, since for small $\sigma$ the matrix function $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ has a small numerical rank, a Nyström approximation alone is already enough to achieve a good approximation.
-On the other hand, when $\sigma$ is large, the Nyström approximation by itself is not effective, and the correction with the residual trace becomes indispensable.
+For a fixed budget of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}} = 80$ of random vectors, \reffig{fig:distribution} shows how the value of the smoothing parameter $\sigma$ the choice between Nyström ($n_{\mtx{\Omega}}$) and
+Girard-Hutchinson ($n_{\mtx{\Psi}}$). As expected, for small $\sigma$ the Nyström approximation alone suffices, but 
+this is not an effective choice for large $\sigma$. Choosing $n_{\mtx{\Psi}} = n_{\mtx{\Omega}}$ constitutes a good compromise.
 
@@
     \input{plots/distribution.pgf}
-    \caption{For a range of values of $\sigma$, the Chebyshev-Nyström++ method is applied to the Hamiltonian matrix described in \refsec{subsec:hamiltonian} with $n_c = 1$ for different ways of allocating a total of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation.
-We make the approximation error committed by the Chebyshev approximation negligible by rescaling $m=16 / \sigma$ (based on \reflem{lem:non-negative-chebyshev-error}).}
+    \caption{Error vs. $\sigma$ for a fixed budget of $80$ random vectors 
+    when applying  Chebyshev-Nyström++ to the Hamiltonian matrix from~\refsec{subsec:hamiltonian} with $n_c = 1$.
+The Chebyshev approximation error is made negligible by setting $m=16 / \sigma$ (see \reflem{lem:non-negative-chebyshev-error}).}
     \label{fig:distribution}
@@
 
-The Krylov-aware stochastic trace estimator \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} is also suitable in the context of spectral density estimation.
+The Krylov-aware stochastic trace estimator from \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} is a viable alternative method for spectral density estimation.
+\textcolor{red}{???
+Please rewrite this.
+It does not come out whether~\cite{chen-2023-krylovaware-stochastic} actually cover spectral density estimation.
+Are the Gaussian matrices chosen constants in their paper/in our experiments?
+If this is not discussed in~\cite{chen-2023-krylovaware-stochastic}, comment more explicitly to which extent this algorithm benefits from constant random matrices.
+Do they have an analysis that applies to our situation, etc.
+???
 It also samples two Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$.
 $\mtx{\Omega}$ is used to extract a low-rank approximation to $g_{\sigma}(t \mtx{I}_n - \mtx{A})$ by running the block-Lanczos algorithm on $\mtx{A}$ with starting block $\mtx{\Omega}$ for $r$ iterations with reorthogonalization and subsequently $q$ without.
-The columns of $\mtx{\Psi}$ are then used to estimate the trace of the approximation residual with $r$ iterations of the Lanczos algorithm on $\mtx{A}$.
+The columns of $\mtx{\Psi}$ are then used to estimate the trace of the approximation residual with $r$ iterations of the Lanczos algorithm on $\mtx{A}$ ???
+Why $\mtx{A}$?
+???.}
 
-We run our own, faithfully optimized implementation of the Krylov-aware stochastic trace estimator \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} in multiple configurations on the example from \refsec{subsec:hamiltonian} and plot the approximation errors for logarithmically spaced values of the smoothing parameter $\sigma$ in \reffig{fig:krylov-aware-density}.
-For reference, we add the error of the Chebyshev-Nyström++ method on parameters which lead to a comparable run-time.
+We run our own, faithfully optimized implementation of the Krylov-aware stochastic trace estimator \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} for different parameter settings on the example from \refsec{subsec:hamiltonian} and plot the approximation errors for logarithmically spaced values of the smoothing parameter $\sigma$ in \reffig{fig:krylov-aware-density}.
+For reference, we also report the error of Chebyshev-Nyström++ for parameters that lead to a comparable run-time.
+While the the Krylov-aware estimator is clearly an attractive alternative, it requires a careful choice of parameters adapted to  different regimes of $\sigma$ and still leads to a significantly larger error for moderately small values $\sigma$.
 
@@
     \end{minipage}
-    \caption{For the example from \refsec{subsec:hamiltonian} we compare the Chebyshev-Nyström++ (CN++) method with the Krylov-Aware (KA) stochastic trace estimator in multiple configurations.
-To this extent, we compute the $L^1$-error of the methods for various levels of smoothing $\sigma$ (left).
-The parameters of the algorithms and run-time averaged over all values of $\sigma$ can be found in the tables (right).}
+    \caption{For the example from \refsec{subsec:hamiltonian}, error vs.
+$\sigma$ the Krylov-Aware (KA) stochastic trace estimator with different parameter settings, compared to the Chebyshev-Nyström++ (CN++) method.}
     \label{fig:krylov-aware-density}
@@
 
-When optimizing a neural network over some data, it is of great interest to determine whether a minimum has been found and if this minimum is robust, i.e.
-a small perturbation of the parameters of the neural network will not lead to a significant decrease in the fit of the neural network.
-Both properties are reflected in the eigenvalues of the Hessian matrix $\mtx{A}$ with respect to some loss function: if all eigenvalues are non-negative, then the loss attains a local minimum, and if additionally all of them are small, the minimum is robust.
+When fitting the parameters of a neural network, it can be of interest to determine whether a (local) minimum has been found and whether this minimum is robust, i.e., whether a small perturbation of the parameters leads to a significant decrease in the fit of the neural network.
+Both properties are reflected in the eigenvalues of the Hessian matrix $\mtx{A}$ with respect to some loss function: if all eigenvalues are non-negative, then the loss attains a local minimum, and if, additionally, all of them are small, the minimum is robust.
+\textcolor{red}{???
+Isn't it the opposite?
+If the eigenvalues are not close to zero, we get a robust local minimum?
+Please check!
+???}
 
-Since neural networks are usually parametrized by thousands of parameters, assembling the Hessian matrix explicitly and then computing its eigenvalues is out of question.
-Luckily, there exists a method which can exactly compute matrix-vector products $\mtx{A} \vct{x}$ for any vector $\vct{x}$ which scales proportionally with the number of parameters \cite{pearlmutter-1994-fast-exact}.
+Since neural networks are usually parametrized by a large number of parameters, assembling the Hessian matrix explicitly and then computing its eigenvalues is clearly infeasible.
+On the other hand, one can cheaply compute matrix-vector products $\mtx{A} \vct{x}$ with the Hessian, at a cost that scales proportionally with the number of parameters \cite{pearlmutter-1994-fast-exact}.
 
-To demonstrate that our algorithm can effectively be applied in the setting of neural network optimization, we approximate the spectral density of a small fully connected convolutional neural network with 6782 parameters.
+To demonstrate that our algorithm can effectively be applied in the setting of neural network optimization, we approximate the spectral density of a small fully connected convolutional neural network with $6\,782$ parameters.
 We train this network on the handwritten digit classification task given by the MNIST dataset\footnote{Handwritten digit classification; taken from \url{http://yann.lecun.com/exdb/mnist}.} in PyTorch 2.4.1 in a standard way.
-We use the bound from \cite[Theorem 1]{zhou-2011-bounding-spectrum} to estimate a bound on the spectrum.
+We use the bound from \cite[Theorem 1]{zhou-2011-bounding-spectrum} to estimate bounds on the spectrum.
 
 We plot the approximation of the spectral density of the Hessian matrix of the untrained neural network, and at different stages of training in \reffig{fig:hessian-density}, as well as the corresponding mean squared error loss.
-It can well be observed that the eigenvalues creep towards zero as the training proceeds.
+It can be observed that the eigenvalues creep towards zero as the training proceeds.
 Furthermore, despite the loss steadily decreasing, the presence of relatively large eigenvalues in some epochs indicates a sharp minimum of the network, hence, unfavorable generalization properties.
@@
     \end{minipage}
-    \caption{The mean squared error training loss (right) and the corresponding approximate spectral density of the Hessian matrix of a fully connected convolutional neural network in different epochs of training on the MNIST dataset (left).
-The spectral density is approximated in $n_t=150$ uniformly spaced points using the Chebyshev-Nyström++ method (\refalg{alg:nystrom-chebyshev-pp}) with parameters $n_{\mtx{\Omega}} = 10$, $n_{\mtx{\Psi}} = 10$, $m = 1000$, and $\sigma = 0.005$.}
+    \caption{Mean squared error training loss (right) and corresponding approximate spectral density of the Hessian matrix of a fully connected convolutional neural network in different epochs of training on the MNIST dataset (left).
+The spectral density is approximated in $n_t=150$ uniformly spaced points using Chebyshev-Nyström++ with parameters $n_{\mtx{\Omega}} = 10$, $n_{\mtx{\Psi}} = 10$, $m = 1000$, and $\sigma = 0.005$.}
     \label{fig:hessian-density}
@@
 
-To demonstrate that the Chebyshev-Nyström++ estimator is also effective on problems unrelated to spectral density estimation, we consider the problem of approximating the partition function $\Trace(\exp(-\beta \mtx{A}))$ for the Hamiltonian $\mtx{A}$ of a planar Heisenberg spin chain \cite[Section 5.1]{chen-2023-krylovaware-stochastic}.
-We compute the pointwise error from the analytic solution as a function of the temperature $\beta^{-1}$ and compare it to the error achieved by the Krylov-aware trace estimator in \reffig{fig:krylov-aware-spin}.
+To demonstrate that the Chebyshev-Nyström++ estimator is also effective for problems unrelated to spectral density estimation, we consider the problem of approximating the partition function $\Trace(\exp(-\beta \mtx{A}))$ for the Hamiltonian $\mtx{A}$ of a planar Heisenberg spin chain from~\cite[Section 5.1]{chen-2023-krylovaware-stochastic}.
+We compute the pointwise error from the analytic solution as a function of the temperature $t\equiv \beta^{-1}$ and compare it to the error achieved by the Krylov-aware trace estimator in \reffig{fig:krylov-aware-spin}.
 
@@
 
-Unlike in the spectral density example from \refsec{subsec:hamiltonian}, where the numerical rank of the matrix stays low for all $t$, it changes wildly with the parameter $\beta$.
-The Chebyshev-Nyström++ method requires a fixed allocation of random vectors to the low-rank approximation for all values of the parameter $\beta$.
-Hence, both for large $\beta$, where the matrix $\exp(-\beta \mtx{A})$ is low-rank, and for small $\beta$, where the singular values only decay slowly, we cannot expect a good approximation.
+Unlike for the spectral density example from \refsec{subsec:hamiltonian}, where the numerical rank of the matrix is relatively uniform across different $t$, this now changes changes wildly with the parameter $\beta$.
+Recall that the Chebyshev-Nyström++ method requires a fixed allocation of random vectors to the low-rank approximation for all values of the parameter $\beta$.
+Hence, both for large $\beta$, where the matrix $\exp(-\beta \mtx{A})$ has low numerical rank, and for small $\beta$, where the singular values decay very slowly, the approximation error is relatively large.
+\textcolor{red}{??? Could you please recheck this example? Why is CN++ extremely bad for small $\beta^{-1}$? and KA much better? Please comment on this aspect explicitly. ???}

### 2025-02-24 paper/paper.tex
review comments
adapt nomenclature and fix README
remove green color
review Chebyshev coefficient computation
polish results and add conclusion
comments from final review pass

@@
-\documentclass[12pt]{article}
+\documentclass[11pt]{article}
 
@@
 
-\title{Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation}
+\title{Randomized \textcolor{red}{I would prefer Stochastic} trace estimation for parameter-dependent matrices applied to spectral density approximation}
 
@@
 \begin{abstract}
-    Stochastic trace estimators are a family of well established techniques for approximating the trace of constant matrices.
-Oftentimes the matrix whose trace we aim to approximate continuously depends on a real parameter $t \in [a,b]$.
-This is for example the case when approximating the spectral density of a matrix or computing partition functions of quantum systems.
-Clearly, we could separately approximate the traces of the constant matrices $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ at the fixed values $t_1, \dots, t_m \in [a, b]$.
-However, this approach does not scale well if $m$ is large.
-We analyze three trace estimators for constant matrices -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter-dependent matrices with the same randomization reused for all values of the parameter $t$.
-We show that all of these estimators closely match their corresponding error guarantees for constant matrices.
-In particular, for the Nyström++ method, we can guarantee an error of at most $\varepsilon$ by computing the estimate with just $\mathcal{O}(\varepsilon^{-1})$ random vectors, independent of what structure the symmetric positive semi-definite matrix $\mtx{B}(t)$ has.
-We employ these trace estimators to improve upon a technique for approximating the spectral density of a matrix and provide a detailed analysis of the resulting method.
-Finally, we confirm our theoretical developments in multiple numerical experiments.
+    Stochastic trace estimation is a well-established tool for approximating the trace of a large symmetric positive semi-definite matrix $\mtx{B}$.
+Several applications involve a matrix that depends continuously on parameter $t \in [a,b]$, and require trace estimates of $\mtx{B}(t)$ for many different values of $t$.
+This is, for example, the case when approximating the spectral density of a matrix or computing partition functions of quantum systems.
+Approximating the trace separately for each matrix
+    $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ clearly incurs redundancies and a cost that scales linearly with $m$.
+To address this issue, we propose and analyze modifications for three stochastic trace estimators, the Girard-Hutchinson, Nyström, and Nyström++ estimators.
+Our modification uses \emph{constant} randomization across different values of $t$, that is,
+    every matrix $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ is multiplied with the \emph{same} set of random vectors.
+    When combined with Chebyshev approximation in $t$, the use of such constant random matrices allows one to reuse matrix vector-products across different values of $t$, leading to significant cost reduction.
+    Our analysis shows that the loss of stochastic independence across different $t$ does not lead to detoriation; in fact, our error bounds for these estimators closely match existing error bounds for the constant case.
+In particular, we show that $\mathcal{O}(\varepsilon^{-1})$ random matrix-vector products suffice to ensure an error of $\varepsilon > 0$ for Nyström++, independent of low-rank properties of $\mtx{B}(t)$.
+We discuss in detail how the combination of Nyström++ with
+    Chebyshev approximation applies to large-scale spectral density estimation and provide an analysis of the resulting method.
+This  improves various aspects of an existing stochastic estimator for spectral density estimation.
+Several numerical experiments for examples from electronic structure interaction, statistical thermodynamics, and neural network optimization illustrate and validate our findings.
 \end{abstract}
@@
 \begin{itemize}
-%\item Add author list. Matti first, rest (He/Kressner/Lam) in alphabetical order.
- \item The text style is a bit too generous but no need to change this now. \textcolor{green}{Ok! Let's talk about it at some point.}
+ \item Has~\cite{bhattacharjee-2024-improved-spectral} been published?
+ \item Fix discussion end of Section 3.3.2
+ \item Please reread entire Section 4 and address various red points.
+ \item Please reread abstract.
+ \item Please reread last part of Section 2 (startin from new contributions) and fix red remark.
+ \item Please remove all unreferenced equation labels (using checkref) at the end.
+ \item Perform a spell check at the end.
 \end{itemize}
 }
-\color{blue}
 \input{intro.tex}
@@
 \input{numresults.tex}
+\input{conclusion.tex}
 
-%\section{Conclusion}
-%\label{sec:conclusion}
-%
-%\todo{
-%\begin{itemize}
-%    \item Robustness (can choose $m$ and $n_{\mtx{\Omega}}$, $n_{\mtx{\Psi}}$ as large as one wants, and not break algorithm, unlike Krylov-aware Lanczos)
-%\end{itemize}
-%}
 \paragraph{Acknowledgements.} The authors gratefully acknowledge the help of Lin Lin in reproducing the example in \refsec{subsec:hamiltonian}.

### 2025-02-24 paper/spectraldensity.tex
comments from final review pass

@@
 
-\color{blue}
-
 In this section, we analyze the accuracy of Chebyshev-Nyström++ estimate returned by~\refalg{alg:nystrom-chebyshev-pp}. 
-\begin{theorem}{Error of Chebyshev-Nyström++ estimate}{chebyshev-nystrom}
+\begin{theorem}{Error of Chebyshev-Nyström++ estimate for smoothed spectral density estimation}{chebyshev-nystrom}
     For a smoothing parameter $\sigma > 0$, let $\phi_{\sigma}$ be the smoothed spectral density~\refequ{equ:smooth-spectral-density} of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum contained in $[-1, 1]$.
@@
 
-\color{black}
-In practice, the $\varepsilon^{-1}$ complexity of $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$ in \refthm{thm:chebyshev-nystrom} is often pessimistic.
-The reason is that \refthm{thm:nystrom-pp} does not take advantage of the low-rank structure of the matrix $\mtx{B}(t) = g_{\sigma}(t\mtx{I}_n - \mtx{A})$: The singular values of the matrix $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ are given by
+In practice, the $\varepsilon^{-1}$ complexity of $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$ in \refthm{thm:chebyshev-nystrom} is often pessimistic, because  the underlying result of \refthm{thm:nystrom-pp} is agnostic to the low-rank structure of the matrix $\mtx{B}(t) = g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
+The singular values of this matrix are given by
 \begin{equation}
-    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) = \frac{1}{\sqrt{2 \pi \sigma^2}} e^{-\sfrac{\left(t - \lambda_{(i)}\right)^2}{2 \sigma^2}},
+    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) = \frac{1}{\sigma\sqrt{2 \pi}} \exp\Big( -\frac{(t - \lambda_{(i)})^2}{2 \sigma^2} \Big),  
     \label{equ:gaussian-kernel-eigenvalues}
 \end{equation}
-where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from $t$, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t)$.
-Clearly, the further away an eigenvalue $\lambda_{(i)}$ is from $t$, the (exponentially) smaller its associated singular value $\sigma_i(t)$ will be.
-Hence, all singular values $\sigma_i(t)$ corresponding to eigenvalues outside of a certain window $[t - d, t + d]$ around $t$, where $d$ is a number depending on $\sigma$ and $n$, will be negligibly small (see \reffig{fig:numerical-rank}).
+where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from $t$, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t) \ge 0$.
+A distant eigenvalue $\lambda_{(i)}$ results in an exponentially small singular value $\sigma_i(t)$.
+In particular, for a small value of $\sigma$, all singular values corresponding to eigenvalues outside a certain window $[t - d, t + d]$ are negligibly small; see \reffig{fig:numerical-rank}).
 \begin{figure}[ht]
@@
 \end{figure}
-If we choose $n_{\mtx{\Omega}}$ large enough such that the sum $\sum_{i=r+1}^{n} \sigma_i(t)$ which bounds the approximation error in \refthm{thm:nystrom} is negligibly small for all $t$, we achieve a much better approximation.
-In particular, if we assume the eigenvalues of $\mtx{A}$ to be uniformly distributed, a small calculation shows that $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma)$ will ensure this.
+By choosing $n_{\mtx{\Omega}}$ such that the sum $\sum_{i=n_{\mtx{\Omega}}+5}^{n} \sigma_i(t)$ becomes negligibly small,  \refthm{thm:nystrom} implies an excellent approximation with high probability.
+In particular, if we assume the eigenvalues of $\mtx{A}$ to be roughly uniformly distributed, a small calculation shows that \textcolor{red}{???
+The point of the start of this discussion was the dependence on $\epsilon$.
+Please verify this dependence; it is hopefully something like $\log\varepsilon^{-1}$ ???} $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma)$ will ensure this.
 

### 2025-02-25 paper/analysis.tex
comments from final review pass
proof reading comments

@@
 
-\color{blue}
 Consider a general parameter-dependent matrix
@@
 
-For constant $\mtx{\Omega}$, the estimator~\refequ{equ:nystrompp-trace-estimator} coincides with the Nyström++ estimator from~\cite{persson-2022-improved-variants}, which is based on the Hutch++ estimator~\cite{meyer-2021-hutch-optimal}.
+For constant $\mtx{\Omega}$ \textcolor{blue}{Shoud it be constant $B$?}, the estimator~\refequ{equ:nystrompp-trace-estimator} coincides with the Nyström++ estimator from~\cite{persson-2022-improved-variants}, which is based on the Hutch++ estimator~\cite{meyer-2021-hutch-optimal}.
 In this situation (constant $\mtx{B}$), these estimators were both shown to achieve a relative $\varepsilon$-error with $\mathcal{O}(\varepsilon^{-1})$ matrix-vector products only, independent of the singular value decay of $\mtx{B}$.
@@
     \end{align*}
-    where $\lVert \cdot \rVert _{\ast}$ denotes the nuclear norm and we used \cite[Theorem 1]{gittens-2011-spectral-norm} in the last equality.
+    where $\lVert \cdot \rVert _{\ast}$ denotes the nuclear norm and we used \cite[Theorem 1]{gittens-2011-spectral-norm} in the \textcolor{blue}{second} equality.
 This allows us to apply \cite[Theorem 5]{kressner-2024-randomized-lowrank}, which immediately implies the statement claimed by the theorem.
@@
 
-Our analysis mimicks the strategy from~\cite{meyer-2021-hutch-optimal}, by first establishing a relative Frobenius norm error proportional to $n_{\mtx{\Omega}}^{-1/2}$ for the parameter-dependent Nyström approximation.
+Our analysis mimics the strategy from~\cite{meyer-2021-hutch-optimal}, by first establishing a relative Frobenius norm error proportional to $n_{\mtx{\Omega}}^{-1/2}$ for the parameter-dependent Nyström approximation.
 
@@
 
-Combining the bound on the Girard-Hutchinson estimator from~\refthm{thm:hutchinson} with the result of~\reflem{lem:nystrom}) finally gives the desired error bound for the Nyström++ trace estimator.
+Combining the bound on the Girard-Hutchinson estimator from~\refthm{thm:hutchinson} with the result of~\reflem{lem:nystrom} finally gives the desired error bound for the Nyström++ trace estimator.
 \begin{theorem}{Nyström++ trace estimator for parameter-dependent matrices}{nystrom-pp}

### 2025-02-25 paper/intro.tex
proof reading comments

@@
 
-Being composed of distributions, the function $\phi$ itself is hard to approximate.
-However, as many application only require a rough estimate of $\phi$, it usually suffices to instead approximate a smoothed spectral density of the form
+Being composed of distributions \textcolor{blue}{Why composed of distributions lead to bad approximation, I am wondering if we mean Dirac delta here?}, the function $\phi$ itself is hard to approximate.
+However, as many application\textcolor{blue}{s} only require a rough estimate of $\phi$, it usually suffices to instead approximate a smoothed spectral density of the form
 \begin{equation}
@@
 \paragraph{Structure.} In \refsec{sec:analysis} we will analyze the error of three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter-dependent matrices with constant random matrix.
-In \refsec{sec:application}, we propose and analyze the Chebyshev-Nyström++ estimator, which uses the these estimators to approximate the smoothed spectral density of real symmetric matrices.
+In \refsec{sec:application}, we propose and analyze the Chebyshev-Nyström++ estimator, which uses these estimators to approximate the smoothed spectral density of real symmetric matrices.
 Finally, we will illustrate and validate our developments for examples from various applications in \refsec{sec:results}.

### 2025-02-25 paper/numresults.tex
proof reading comments

@@
 
-In this section, we verify the performance the Chebyshev-Nyström++ algorithm proposed in this paper in various application scenarios, from electronic structure interaction, statistical thermodynamics, and neural network optimization.
+In this section, we verify the performance \textcolor{blue}{of} the Chebyshev-Nyström++ algorithm proposed in this paper in various application scenarios, from electronic structure interaction, statistical thermodynamics, and neural network optimization.
 

### 2025-02-25 paper/spectraldensity.tex
proof reading comments

@@
 also for $l=m+1, \dots, 2m$, this allows us to evaluate $\mtx{K}_2(t)$ cheaply, in an analogous way to $\mtx{K}_1(t)$.
-Alternatively, \cite{lin-2017-randomized-estimation} proposes to instead approximate the squared smoothing kernel $g_{\sigma}^2$ in a separate "auxiliary" Chebyshev series truncated after term $2m$.
+Alternatively, \cite{lin-2017-randomized-estimation} proposes to instead approximate the squared smoothing kernel $g_{\sigma}^2$ in a separate ``auxiliary" Chebyshev series truncated after term $2m$.
 While this direct approximation of $g_{\sigma}^2$ might be more accurate than our squaring approach, it is \emph{inconsistent} in the sense that numerically computing the truncated Chebyshev series of $g_{\sigma}^2$ gives --- in general --- not the same result as squaring the truncated Chebyshev series of $g_{\sigma}$.
@@
 where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from $t$, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t) \ge 0$.
-A distant eigenvalue $\lambda_{(i)}$ results in an exponentially small singular value $\sigma_i(t)$.
+An eigenvalue $\lambda_{(i)}$ \textcolor{blue}{distant from $t$} results in an exponentially small singular value $\sigma_i(t)$.
 In particular, for a small value of $\sigma$, all singular values corresponding to eigenvalues outside a certain window $[t - d, t + d]$ are negligibly small; see \reffig{fig:numerical-rank}).

### 2025-02-25 paper/analysis.tex
address review comments, clean up equations, create n_c=2 plot

--- 2025-02-20 paper/analysis.tex
@@
 We first derive an $L^1$-error bound for the Girard-Hutchinson estimator~\refequ{equ:hutchinson-trace-estimator} that is relative to the $L^1$-norms of the Frobenius and spectral norms of $\mtx{B}(t)$, defined by
-\[
+\begin{equation*}
  \|\mtx{B}(t)\|_{F,1} := \int_{a}^{b} \|\mtx{B}(t)\|_{F}~\mathrm{d}t, \quad  \|\mtx{B}(t)\|_{2,1} := \int_{a}^{b} \|\mtx{B}(t)\|_{2}~\mathrm{d}t.
-\]
+\end{equation*}
 
@@
 Then for any $p \geq 1, p \in \mathbb{R}$ and $\gamma \geq 1$ the bound
-%    \begin{equation} \label{eq:ghbound}
+%    \begin{equation} \label{equ:girard-hutchinson-bound}
 %        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq c \gamma \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}}  \right\} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
@@
 Then for any $p \geq 1$, $p \in \mathbb{R}$ and $\gamma \geq 1$, the following bound holds with probability at least $1 - \gamma^{-p}$:
-    \begin{equation} \label{eq:ghbound}
+\begin{equation}
     \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq 4 \gamma \Big( \sqrt{\frac{p}{n_{\mtx{\Psi}}}}  \|\mtx{B}(t)\|_{F,1} + \frac{2 p}{n_{\mtx{\Psi}}} \|\mtx{B}(t)\|_{2,1} \Big).
+    \label{equ:girard-hutchinson-bound}
 \end{equation}
@@
 Therefore, we can apply Minkowski's integral inequality~\cite[Theorem 202]{hardy-1952-inequalities} to conclude from the moment bound~\refequ{equ:moment-bound-hutchinson} that
-    \begin{align}
+    \begin{align*}
         \mathbb{E}^{p}\left[ \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t  \right]
@@
         &\leq 4 \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert_{F,1} + 8 \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert_{2,1}.
-    \end{align}
-    This implies~\refequ{eq:ghbound} by Markov's inequality.
+    \end{align*}
+    This implies~\refequ{equ:girard-hutchinson-bound} by Markov's inequality.
 
@@
 %    we note that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) = r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold.
-Thus, the claimed probabilistic bound~\refequ{eq:ghbound} follows from applying~\refequ{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$.
+Thus, the claimed probabilistic bound~\refequ{equ:girard-hutchinson-bound} follows from applying~\refequ{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$.
 %%     Then 
@@
         \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t
-        \label{qu:nystrompp-theorem-bound}
+        \label{equ:nystrompp-theorem-bound}
     \end{equation}
@@
     To continue from here, we set $k = n_{\mtx{\Omega}}/2$ and $q = n_{\mtx{\Omega}}/4$. By the calculations after~\cite[Eqn~(B.7)]{tropp-2023-randomized-algorithms}, one has
-    \[
+    \begin{equation*}
      \mathbb{E}\left[ \big\| ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \big\|_2^{\sfrac{n_{\mtx{\Omega}}}{4}} \right]%  \notag \\
@@
         \big( ( n_{\mtx{\Omega}} / 2 + 1)!\big)^{-\frac{n_{\mtx{\Omega}}}{2+n_{\mtx{\Omega}}}}.
-    \]
+    \end{equation*}
     Therefore,
-    \begin{equation} \label{equ:pinv-spectral-norm-bound}
+    \begin{equation}
       \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \big\| \mtx{\Omega}_1^{\dagger} \big\|_2^2 \right]
@@
         \frac{3}{4}  \frac{e^4}{n_{\mtx{\Omega}}},
+        \label{equ:pinv-spectral-norm-bound}
     \end{equation}
@@
     The second factor in~\refequ{equ:nystrom-proof-processed-tail} is bounded using \reflem{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$: % and $p = n_{\mtx{\Omega}}/2$:
-    \[
+    \begin{equation*}
         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \big\|_2^2 \right]
@@
       %  \label{equ:spectral-norm-bound-applied}
-    \]
+    \end{equation*}
     %The $q$-th moment of the second term can be processed with standard matrix norm manipulations and the stochastic independence of $\mtx{\Omega}_1$ and $\mtx{\Omega}_2$ to
@@
     Inserting this inequality and~\refequ{equ:pinv-spectral-norm-bound} into~\refequ{equ:nystrom-proof-processed-tail} gives
-    \begin{equation}
+    \begin{equation*}
         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
@@
         %\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
-    \end{equation}
+    \end{equation*}
     Bounding $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 = \lambda_{k+1}  \leq \Trace(\mtx{B})/k$ and $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 = \Trace(\mtx{\Lambda}_2) \le \Trace(\mtx{B})$ (recall that $k = n_{\mtx{\Omega}}/2$) yields
@@
     Inserting~\refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud}, letting $c=154$, and using the triangle inequality for $\mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}[\cdot]$, we obtain
-    \begin{equation} \label{eq:blubber}
+    \begin{equation}
         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right]
@@
         \leq \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
+        \label{equ:frobenius-moment-bound}
     \end{equation}
     As in \cite{kressner-2024-randomized-lowrank}, one can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable.
-    As in the proof of~\refthm{thm:hutchinson}, the claimed bound~\refequ{qu:nystrompp-theorem-bound}
-    follows from~\refequ{eq:blubber} using Minkowski's integral inequality and Markov's inequality.
+    As in the proof of~\refthm{thm:hutchinson}, the claimed bound~\refequ{equ:nystrompp-theorem-bound}
+    follows from~\refequ{equ:frobenius-moment-bound} using Minkowski's integral inequality and Markov's inequality.
 %     
@@
     the inequality 
-    \begin{equation}
+    \begin{equation*}
         \int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t
         \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t
-    \end{equation}
+    \end{equation*}
     holds with probability at least $1 - \delta$.

### 2025-02-25 paper/intro.tex
address review comments, clean up equations, create n_c=2 plot

--- 2025-02-24 paper/intro.tex
@@
 Several techniques have been developed for approximating $\phi$, including moment matching~\cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time} and polynomial approximation, either implicitly by a Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or explicitly by expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}.
-Some of these methods have also been combined with a deflation step to yield even more efficient algorithms \cite{lin-2017-randomized-estimation, bhattacharjee-2024-improved-spectral}.
+Some of these methods have also been combined with a deflation step to yield even more efficient algorithms \cite{lin-2017-randomized-estimation, bhattacharjee-2025-improved-spectral}.
 
@@
 The resulting Chebyshev-Nyström++ estimator
-unifies all methods from~\cite{lin-2017-randomized-estimation} in a single algorithm, with the desirable side effect of \textcolor{red}{speeding up the \enquote{spectrum sweeping} method by orders of magnitude.
+unifies all methods from~\cite{lin-2017-randomized-estimation} in a single algorithm.
+%, and improves multiple algorithmic aspects to make them faster and more robust.
+%\textcolor{red}{speeding up the \enquote{spectrum sweeping} method by orders of magnitude.
 ???
 Where do we actually demonstrate this?
-Illustrate/spell this out more explicitly in the numerical experiments ???} Combining our analysis of the randomized trace estimators with standard error bounds for Chebyshev approximation leads to Theorem~\refthm{thm:chebyshev-nystrom}, which shows that $\mathcal O(\varepsilon^{-1})$ matrix-vector products suffice to guarantee an error $\varepsilon$.
+Illustrate/spell this out more explicitly in the numerical experiments ???}
+Combining our analysis of the randomized trace estimators with standard error bounds for Chebyshev approximation leads to~\refthm{thm:chebyshev-nystrom}, which shows that $\mathcal O(\varepsilon^{-1})$ matrix-vector products suffice to guarantee an error $\varepsilon$.
 

### 2025-02-25 paper/numresults.tex
address review comments, clean up equations, create n_c=2 plot

--- 2025-02-24 paper/numresults.tex
@@
 
-While theoretically an important tool, we have observed that the preservation of non-negativity discussed in Section~\ref{sec:preservnonneg} is not needed in practice.
+While theoretically an important tool, we have observed that the preservation of non-negativity discussed in~\refsec{sec:preservnonneg} is not needed in practice.
 The (slight) indefiniteness of the standard Chebyshev approximation $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ from~\refequ{equ:matrix-approximation} does not seem to impede the reliability of the Nyström++ estimator.
-Unless otherwise stated, we transform the input matrices such that their eigenvalues are contained in $[-1, 1]$, by approximating their spectrum with NumPy's \textcolor{red}{Hessian ???
-replace by sparse?
-???} eigenvalue solver.
+Unless otherwise stated, we transform the input matrices such that their eigenvalues are contained in $[-1, 1]$, by approximating their spectrum with NumPy's Hermitian eigenvalue solver before we apply the methods to them.
 We compute spectral densities at $n_t = 100$ uniformly distributed values of the parameter $t \in [-1, 1]$.
@@
     \mathcal{H} = - \Delta + V
-    \label{equ:5-experiments-electronic-hamiltonian}
+    \label{equ:electronic-hamiltonian}
 \end{equation}
@@
 The grid width of the finite difference discretization is fixed to $h=0.6$.
-For example, for $n_c = 1$, this leads to a matrix of size $1000 \times 1000$.
+For $n_c = 1$, this leads to a sparse matrix of size $1\,000 \times 1\,000$ and for $n_c = 2$ of size $8\,000 \times 8\,000$.
 This example represents an idealized model for the interaction of nuclei on a regular grid with electrons for a $k$-vector in the center of the first Brillouin zone.
@@
 
-\textcolor{red}{???
+%\textcolor{red}{???
 This discussion of $n_c$ only makes sense if we have at least one experiment with $n_c = 5$ or so.
@@
 
-\reffig{fig:convergence} displays the convergence behavior of Chebyshev-Nyström++ applied to the $1000 \times 1000$ matrix resulting from the discretization of the Hamiltonian for $n_c = 1$.
+\reffig{fig:convergence} displays the convergence behavior of Chebyshev-Nyström++ applied to the matrix resulting from the discretization of the Hamiltonian for $n_c = 1$ and $n_c = 2$.
 As a reference, we use the eigenvalues computed by NumPy's dense eigenvalue solver.
@@
     \centering
-    \input{plots/convergence.pgf}
+    \begin{subfigure}[b]{0.495\textwidth}
+        \input{plots/convergence-1.pgf}
+        \caption{$n_c = 1$}
+        \label{fig:convergence-1}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.495\textwidth}
+        \input{plots/convergence-2.pgf}
+        \caption{$n_c = 2$}
+        \label{fig:convergence-2}
+    \end{subfigure}
     \caption{Error vs.
-number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ (with fixed $\sigma=0.005$ and \textcolor{red}{$m$ ???
-Provide value???}) when applying  Chebyshev-Nyström++ to the Hamiltonian matrix from~\refsec{subsec:hamiltonian} with $n_c = 1$.}
+number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ (with fixed $\sigma=0.005$ and $m=2000$) when applying  Chebyshev-Nyström++ to the Hamiltonian matrix from~\refsec{subsec:hamiltonian}.}
     \label{fig:convergence}
 \end{figure}%
+%\begin{figure}[ht]
+%    \centering
+%    \input{plots/convergence.pgf}
+%    \caption{Error vs.
+number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ (with fixed $\sigma=0.005$ and $m=2000$) when applying  Chebyshev-Nyström++ to the Hamiltonian matrix from~\refsec{subsec:hamiltonian} with $n_c = 1$.}
+%    \label{fig:convergence}
+%\end{figure}%
 As expected from \refthm{thm:hutchinson}, Girard-Hutchinson alone ($n_{\mtx{\Omega}} = 0$) require $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ samples to achieve an error of order $\varepsilon$.
 Because the eigenvalues of the matrix are spread out, we observe the better convergence of the Nyström approximation discussed at the end of \refsec{sec:application} once $n_{\mtx{\Omega}}$ is sufficiently large.
-\textcolor{red}{???
-Comment explicitly that and why the errors levels off at around $10^{-6}$.
-Is it because of Chebyshev or some other truncation, etc.?
-???}
+We can observe the errors stagnate at around $10^{-6}$ as a consequence of the error made in the Chebyshev approximation.
+
+%\begin{figure}[ht]
+%    \centering
+%    \input{plots/convergence_8_sigma005.pgf}
+%    \caption{Error vs.
+number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ (with fixed $\sigma=0.005$ and $m=2000$) when applying  Chebyshev-Nyström++ to the Hamiltonian matrix from~\refsec{subsec:hamiltonian} with $n_c = 2$.}
+%    \label{fig:convergence}
+%\end{figure}%
 
@@
 
-\subsection{Comparison to Krylov-aware stochastic trace estimation}
+\subsection{Comparison to Krylov-aware trace estimator}
 \label{subsec:krylov-aware}
 
-The Krylov-aware stochastic trace estimator from \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} is a viable alternative method for spectral density estimation.
-\textcolor{red}{???
+We notice that the Krylov-aware stochastic trace estimator from \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} can also effectively be used in the setting of spectral density approximation.
+It also samples two Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$.
+$\mtx{\Omega}$.
+It first runs block Lanczos algorithm on $\mtx{A}$ with starting block $\mtx{\Omega}$ for $r$ iterations with reorthogonalization and subsequently $q$ without.
+The columns of $\mtx{\Psi}$ are then projected onto the complement of the block Krylov subspace generated by the Lanczos algorithm, and are afterwards used as starting vectors for $r$ iterations of the Lanczos algorithm on $\mtx{A}$.
+What is remarkable is that all these heavy computations are completely independent of the smoothing kernel $g_{\sigma}$ and the parameter $t$.
+Further, some algebraic manipulations show that similarly to the stochastic Lanczos quadrature \cite[Section 3]{ubaru-2017-fast-estimation}, we can compress the estimator into a quadrature.
+Since this quadrature is independent of $g_{\sigma}$ and $t$, it can be applied to different smoothing kernels $g_{\sigma}$ and to as many values of $t$ as needed --- at little additional cost.
+
+%\textcolor{red}{???
 Please rewrite this.
@@
 When fitting the parameters of a neural network, it can be of interest to determine whether a (local) minimum has been found and whether this minimum is robust, i.e., whether a small perturbation of the parameters leads to a significant decrease in the fit of the neural network.
-Both properties are reflected in the eigenvalues of the Hessian matrix $\mtx{A}$ with respect to some loss function: if all eigenvalues are non-negative, then the loss attains a local minimum, and if, additionally, all of them are small, the minimum is robust.
-\textcolor{red}{???
-Isn't it the opposite?
-If the eigenvalues are not close to zero, we get a robust local minimum?
-Please check!
-???}
+Both properties are reflected in the eigenvalues of the Hessian matrix $\mtx{A}$ with respect to some loss function: if all eigenvalues are non-negative, then the loss attains a local minimum, and if, additionally, all of them are small, we talk about a flat minimum; one where small perturbations of the parameters do not cause large increases of the loss and are therefore associated with better generalization \cite{yao-2020-pyhessian-neural}.
+On the other hand, large eigenvalues in the spectrum indicate high sensitivity of the loss to small changes in the parameters.
+Those we want to avoid, which is the reason why we monitor them.
 
@@
 
-We plot the approximation of the spectral density of the Hessian matrix of the untrained neural network, and at different stages of training in \reffig{fig:hessian-density}, as well as the corresponding mean squared error loss.
+We plot the spectral density of the Hessian matrix of the untrained neural network, and at different stages of training in \reffig{fig:hessian-density}, as well as the corresponding mean squared error loss.
 It can be observed that the eigenvalues creep towards zero as the training proceeds.
@@
 
-To demonstrate that the Chebyshev-Nyström++ estimator is also effective for problems unrelated to spectral density estimation, we consider the problem of approximating the partition function $\Trace(\exp(-\beta \mtx{A}))$ for the Hamiltonian $\mtx{A}$ of a planar Heisenberg spin chain from~\cite[Section 5.1]{chen-2023-krylovaware-stochastic}.
+To demonstrate that the Chebyshev-Nyström++ estimator is also effective for problems unrelated to spectral density estimation, we consider the problem of approximating the partition function $\Trace(\exp(-\beta \mtx{A}))$ for the $1\,048\,576 \times 1\,048\,576$ Hamiltonian $\mtx{A}$ of a planar Heisenberg spin chain from~\cite[Section 5.1]{chen-2023-krylovaware-stochastic}.
 We compute the pointwise error from the analytic solution as a function of the temperature $t\equiv \beta^{-1}$ and compare it to the error achieved by the Krylov-aware trace estimator in \reffig{fig:krylov-aware-spin}.
@@
 Hence, both for large $\beta$, where the matrix $\exp(-\beta \mtx{A})$ has low numerical rank, and for small $\beta$, where the singular values decay very slowly, the approximation error is relatively large.
-\textcolor{red}{??? Could you please recheck this example? Why is CN++ extremely bad for small $\beta^{-1}$? and KA much better? Please comment on this aspect explicitly. ???}
+Similarly, one degree $m$ for the Chebyshev approximation needs to be fixed for all $\beta$.
+Clearly, for large values $\beta$, where $s \mapsto \exp(-\beta s)$ becomes difficult to approximate with polynomials, the fixed degree $m$ is not sufficient for a good Chebyshev approximation, and consequently, the Chebyshev-Nyström++ method is very inaccurate.

### 2025-02-25 paper/paper.tex
address review comments, clean up equations, create n_c=2 plot

@@
 
-\title{Randomized \textcolor{red}{I would prefer Stochastic} trace estimation for parameter-dependent matrices applied to spectral density approximation}
+\title{Stochastic trace estimation for parameter-dependent matrices applied to spectral density approximation}
 
@@
     Stochastic trace estimation is a well-established tool for approximating the trace of a large symmetric positive semi-definite matrix $\mtx{B}$.
-Several applications involve a matrix that depends continuously on parameter $t \in [a,b]$, and require trace estimates of $\mtx{B}(t)$ for many different values of $t$.
+Several applications involve a matrix that depends continuously on a parameter $t \in [a,b]$, and require trace estimates of $\mtx{B}(t)$ for many different values of $t$.
 This is, for example, the case when approximating the spectral density of a matrix or computing partition functions of quantum systems.
@@
     every matrix $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ is multiplied with the \emph{same} set of random vectors.
-    When combined with Chebyshev approximation in $t$, the use of such constant random matrices allows one to reuse matrix vector-products across different values of $t$, leading to significant cost reduction.
-    Our analysis shows that the loss of stochastic independence across different $t$ does not lead to detoriation; in fact, our error bounds for these estimators closely match existing error bounds for the constant case.
+    When combined with Chebyshev approximation in $t$, the use of such constant random matrices allows one to reuse matrix-vector products across different values of $t$, leading to significant cost reduction.
+    Our analysis shows that the loss of stochastic independence across different $t$ does not lead to deterioriation; in fact, our error bounds for these estimators closely match existing error bounds for the constant case.
 In particular, we show that $\mathcal{O}(\varepsilon^{-1})$ random matrix-vector products suffice to ensure an error of $\varepsilon > 0$ for Nyström++, independent of low-rank properties of $\mtx{B}(t)$.
@@
 
-\textcolor{red}{TODO for Fabio:
-\begin{itemize}
- \item Has~\cite{bhattacharjee-2024-improved-spectral} been published?
- \item Fix discussion end of Section 3.3.2
- \item Please reread entire Section 4 and address various red points.
- \item Please reread abstract.
- \item Please reread last part of Section 2 (startin from new contributions) and fix red remark.
- \item Please remove all unreferenced equation labels (using checkref) at the end.
- \item Perform a spell check at the end.
-\end{itemize}
-}
+%\textcolor{red}{TODO for Fabio:
+%\begin{itemize}
+ %\item Has~\cite{bhattacharjee-2024-improved-spectral} been published?
+ %\item Fix discussion end of Section 3.3.2
+ %\item Please reread entire Section 4 and address various red points.
+ %\item Please reread abstract.
+ %\item Please reread last part of Section 2 (startin from new contributions) and fix red remark.
+ %\item Please remove all unreferenced equation labels (using checkref) at the end.
+ %\item Perform a spell check at the end.
+%\end{itemize}
+%}
 \input{intro.tex}

### 2025-02-25 paper/spectraldensity.tex
address review comments, clean up equations, create n_c=2 plot

--- 2025-02-24 paper/spectraldensity.tex
@@
 where $\mathcal{S}$ represents an appropriately chosen space of test functions.
-Among others, the reference~\cite{lin-2016-approximating-spectral} uses $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - s), s \in [a, b]\}$, \cite{chen-2021-analysis-stochastic} uses $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [a, b] \}$ with the Heaviside step function $\Theta$, and \cite{braverman-2022-sublinear-time,bhattacharjee-2024-improved-spectral} use $\mathcal{S} = \{f : |f(t) - f(s)| \leq |t - s| \}$ for which the error \refequ{equ:error-metric} is equivalent to the Wasserstein-1 distance between $\phi$ and $\phi_{\sigma}$.
+Among others, the reference~\cite{lin-2016-approximating-spectral} uses $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - s), s \in [a, b]\}$, \cite{chen-2021-analysis-stochastic} uses $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [a, b] \}$ with the Heaviside step function $\Theta$, and \cite{braverman-2022-sublinear-time,bhattacharjee-2025-improved-spectral} use $\mathcal{S} = \{f : |f(t) - f(s)| \leq |t - s| \}$ for which the error \refequ{equ:error-metric} is equivalent to the Wasserstein-1 distance between $\phi$ and $\phi_{\sigma}$.
 When choosing 
@@
 By\cite[Theorem 8.2]{trefethen-2020-approximation-theory},
-    \begin{equation}
+    \begin{equation*}
         \sup_{s \in [-1, 1]} \big| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \big| \leq \frac{4}{\chi^m (\chi - 1)} \max_{z \in \mathcal{E}_{\chi}} |g_{\sigma}(t - z)|,
         \label{equ:bernstein-bound}
-    \end{equation}
+    \end{equation*}
     where used the analyticity of $g_{\sigma}(t- \cdot)$.
     For $z = x + \mathrm{i} y \in \mathcal{E}_{\chi}$, we have
-    \begin{equation}
+    \begin{equation*}
     | g_{\sigma}(t - z) | 
@@
     %\leq \frac{1}{n \sigma \sqrt{2 \pi}} \max_{x + iy \in \mathcal{E}_{\chi}} e^{\sfrac{y^2}{2\sigma^2}} 
-    \end{equation}
+    \end{equation*}
     where the last inequality uses 
@@
 Then the approximation $\underline{\phi}_{\sigma}^{(m)}$ defined in~\refequ{equ:non-negative-expanded-density} satisfies
-    \begin{equation}
+    \begin{equation*}
         \int_{-1}^{1} \big| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \big|~\mathrm{d}t \leq 4 \sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right) E_{\sqrt{2}\sigma, \sfrac{m}{2}} =: \underline{E}_{\sigma, m}
@@
         %\label{equ:2-chebyshev-interpolation-error}
-    \end{equation}
+    \end{equation*}
     for every \emph{even} degree $m \in \mathbb{N}$, smoothing parameter $\sigma > 0$, and with $E_{\sigma, m}$ defined in \reflem{lem:chebyshev-error}.
@@
     Using that the inequality $| a - b | \leq (2 \sqrt{a} + | \sqrt{a} - \sqrt{b} |)  | \sqrt{a} - \sqrt{b} |$ holds for any non-negative numbers $a, b$, it follows that
-    \begin{equation}
+    \begin{equation*}
         \big| g_{\sigma} - \underline{g}_{\sigma}^{(m)} \big| \leq \big( 2 \big| \sqrt{g_{\sigma}} \big| + \big| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \big| \big) \big| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \big|.
-    \end{equation}
+    \end{equation*}
     Because of $\sqrt{g_{\sigma}} = \sqrt{2 \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$,
@@
     Together with  $|\sqrt{g_{\sigma}}| \leq 1/\sqrt{\sigma \sqrt{2 \pi}}$, we get
-    \begin{align}
+    \begin{align*}
         \big| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \big| \notag 
@@
         &= 2\sqrt{2} \big(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\big) E_{\sqrt{2}\sigma, \sfrac{m}{2}}
-    \end{align}
+    \end{align*}
     for every $s,t \in [-1, 1]$.
@@
 It is well known that the coefficients of the Chebyshev approximation \refequ{equ:matrix-approximation} can --- in exact arithmetic --- be computed by inverting the (type I) discrete cosine transform (DCT)
-\begin{equation}
+\begin{equation*}
     g_{\sigma}(t - \cos(s_i)) = \sum_{l=0}^{m} \mu_l(t) \cos\left(\frac{\pi i l}{m} \right)
     \label{equ:discrete-cosine-transform}
-\end{equation}
+\end{equation*}
 between the approximation coefficients $\mu_l(t),~l=0,\dots,m$ and the function evaluations at the Chebyshev points $s_i = \cos(\pi i / m),~i=0,\dots,m$ \cite{baszenski-1997-fast-polynomial, plonka-2018-numerical-fourier, trefethen-2020-approximation-theory}.
@@
 Inserting the Chebyshev approximation  $\mtx{B}(t) = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ gives
-\begin{equation}
+\begin{equation*}
     \mtx{K}_1(t) = \mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega} = \sum_{l=0}^{m} \underline{\mu}_l(t) (\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}).
-\end{equation}
+\end{equation*}
 Observing that $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ is independent of $t$ allows us to precompute these (small) $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices for $l=0, \dots, m$ using the Chebyshev recurrence \refequ{equ:chebyshev-recurrence}.
@@
 This consistency will also allow us to derive bounds for the estimator \refequ{equ:chebyshev-nystrom-formula}.
-
 \begin{figure}[ht]
@@
 \emph{Non-negative consistent}: Squaring the non-negative Chebyshev approximation $\underline{g}_{\sigma}^{(m)}$ with~\refalg{alg:chebyshev-squaring}.
-As input matrix we use the $100 \times 100$ matrix $\mtx{A}$ coming from a two-dimensional finite difference discretization of the Hamiltonian defined in~\refequ{equ:5-experiments-electronic-hamiltonian} on a $10 \times 10$ grid.
+As input matrix we use the $100 \times 100$ matrix $\mtx{A}$ coming from a two-dimensional finite difference discretization of the Hamiltonian defined in~\refequ{equ:electronic-hamiltonian} on a $10 \times 10$ grid.
 We fix $n_{\mtx{\Omega}} = 80$ to make sure the low-rank approximation is accurate, set $\sigma = 0.005$, and compute the $L^1$ error from the approximation of the spectral density at $n_t = 100$ uniformly spaced values of $t$ in the interval $[-1, 1]$.}
@@
     \caption{The difference the non-zero check can make when approximating a spectral density using \refalg{alg:nystrom-chebyshev-pp}.
-As input matrix we use the $1000 \times 1000$ matrix $\mtx{A}$ coming from a three-dimensional finite difference discretization of the Hamiltonian defined in~\refequ{equ:5-experiments-electronic-hamiltonian} on a $10 \times 10$ grid.
+As input matrix we use the $1000 \times 1000$ matrix $\mtx{A}$ coming from a three-dimensional finite difference discretization of the Hamiltonian defined in~\refequ{equ:electronic-hamiltonian} on a $10 \times 10$ grid.
 We ran the Chebyshev-Nyström++ method with and without estimating if the matrix function is zero before taking its pseudo-inverse.
@@
 In this section, we analyze the accuracy of Chebyshev-Nyström++ estimate returned by~\refalg{alg:nystrom-chebyshev-pp}. 
-\begin{theorem}{Error of Chebyshev-Nyström++ estimate for smoothed spectral density estimation}{chebyshev-nystrom}
+\begin{theorem}{Error of Chebyshev-Nyström++ estimate for smoothed spectral density}{chebyshev-nystrom}
     For a smoothing parameter $\sigma > 0$, let $\phi_{\sigma}$ be the smoothed spectral density~\refequ{equ:smooth-spectral-density} of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum contained in $[-1, 1]$.
@@
     satisfies the error bound
-    \begin{equation}
+    \begin{equation*}
         \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq \varepsilon
@@
         %\label{equ:2-chebyshev-interpolation-error}
-    \end{equation}
+    \end{equation*}
     with probability at least $1 - \delta$ for $m = \mathcal{O}((\log(\varepsilon^{-1}) + \log(\sigma^{-1})) / \log(\sigma + 1))$  and $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(\delta^{-1}))$.
@@
 Combining results derived above, we obtain that 
-    \begin{align}
+    \begin{align*}
         &\int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \notag \\
@@
         &\leq (1 + \widetilde{\varepsilon}) \underline{E}_{\sigma, m} + \widetilde{\varepsilon}. && \text{(\reflem{lem:non-negative-chebyshev-error})} \notag 
-    \end{align}
+    \end{align*}
     From \reflem{lem:non-negative-chebyshev-error} and the discussion after \reflem{lem:chebyshev-error} we see that the choice $m = \mathcal{O}((\log(\widetilde{\varepsilon}^{-1}) + \log(\sigma^{-1})) / \log(\sigma + 1))$ ensures $\underline{E}_{\sigma, m} \leq \widetilde{\varepsilon}$, and thus
@@
 
-In practice, the $\varepsilon^{-1}$ complexity of $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$ in \refthm{thm:chebyshev-nystrom} is often pessimistic, because  the underlying result of \refthm{thm:nystrom-pp} is agnostic to the low-rank structure of the matrix $\mtx{B}(t) = g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
+In practice, the $\mathcal{O}(\varepsilon^{-1})$ complexity of $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$ in \refthm{thm:chebyshev-nystrom} is often pessimistic, because  the underlying result of \refthm{thm:nystrom-pp} is agnostic to the low-rank structure of the matrix $\mtx{B}(t) = g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
 The singular values of this matrix are given by
-\begin{equation}
+\begin{equation*}
     \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) = \frac{1}{\sigma\sqrt{2 \pi}} \exp\Big( -\frac{(t - \lambda_{(i)})^2}{2 \sigma^2} \Big),  
     \label{equ:gaussian-kernel-eigenvalues}
-\end{equation}
+\end{equation*}
 where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from $t$, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t) \ge 0$.
 A distant eigenvalue $\lambda_{(i)}$ results in an exponentially small singular value $\sigma_i(t)$.
-In particular, for a small value of $\sigma$, all singular values corresponding to eigenvalues outside a certain window $[t - d, t + d]$ are negligibly small; see \reffig{fig:numerical-rank}).
+In particular, for a small value of $\sigma$, all singular values corresponding to eigenvalues outside a certain window $[t - d, t + d]$ are negligibly small; see \reffig{fig:numerical-rank}.
 \begin{figure}[ht]
@@
 By choosing $n_{\mtx{\Omega}}$ such that the sum $\sum_{i=n_{\mtx{\Omega}}+5}^{n} \sigma_i(t)$ becomes negligibly small,  \refthm{thm:nystrom} implies an excellent approximation with high probability.
-In particular, if we assume the eigenvalues of $\mtx{A}$ to be roughly uniformly distributed, a small calculation shows that \textcolor{red}{???
-The point of the start of this discussion was the dependence on $\epsilon$.
-Please verify this dependence; it is hopefully something like $\log\varepsilon^{-1}$ ???} $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma)$ will ensure this.
+In particular, if we assume the eigenvalues of $\mtx{A}$ to be roughly uniformly distributed, a small calculation shows that $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma \sqrt{\log(\varepsilon^{-1}) + \log(\sigma^{-1})})$ will ensure this.
 

### 2025-02-25 paper/plots/convergence.py

@@ -9,50 +9,53 @@ from matrices.electronic_structure import hamiltonian
 np.random.seed(0)
 
 # Load matrix
-A = hamiltonian()
+for n in [1, 2]:
+    A = hamiltonian(n=n)
 
-# Perform spectral transform with A and its eigenvalues
-eigvals = np.linalg.eigvalsh(A.toarray())
-min_ev, max_ev = eigvals[0], eigvals[-1]
-A_st = spectral_transformation(A, min_ev, max_ev)
-eigvals_st = spectral_transformation(eigvals, min_ev, max_ev)
+    # Perform spectral transform with A and its eigenvalues
+    eigvals = np.linalg.eigvalsh(A.toarray())
+    min_ev, max_ev = eigvals[0], eigvals[-1]
+    A_st = spectral_transformation(A, min_ev, max_ev)
+    eigvals_st = spectral_transformation(eigvals, min_ev, max_ev)
 
-# Set parameter
-t = np.linspace(-1, 1, 100)
-sigma = 0.005
-m = 2000
-n_Vec_list = np.logspace(0.8, 2.8, 7).astype(int) 
+    # Set parameter
+    t = np.linspace(-1, 1, 100)
+    sigma = 0.005
+    m = 2000
+    n_Vec_list = np.logspace(0.8, 2.8, 7).astype(int) 
 
-plt.style.use("paper/plots/stylesheet.mplstyle")
-colors = ["#648FFF", "#DC267F", "#FFB000"]
-markers = ["o", "s", "d"]
-labels = [r"$n_{\mathbf{\Omega}} = 0$", r"$n_{\mathbf{\Psi}} = n_{\mathbf{\Omega}}$", r"$n_{\mathbf{\Psi}} = 0$"]
+    plt.style.use("paper/plots/stylesheet.mplstyle")
+    plt.figure(figsize=(3, 3))
+    colors = ["#648FFF", "#DC267F", "#FFB000"]
+    markers = ["o", "s", "d"]
+    labels = [r"$n_{\mathbf{\Omega}} = 0$", r"$n_{\mathbf{\Psi}} = n_{\mathbf{\Omega}}$", r"$n_{\mathbf{\Psi}} = 0$"]
 
-# Determine the baseline spectral density
-kernel = lambda t, x: gaussian_kernel(t, x, sigma=sigma, n=A.shape[0])
-baseline = form_spectral_density(eigvals_st, t, kernel)
[…]

### 2025-02-25 paper/spectraldensity.tex

@@
 This consistency will also allow us to derive bounds for the estimator \refequ{equ:chebyshev-nystrom-formula}.
+
 \begin{figure}[ht]
@@
 where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from $t$, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t) \ge 0$.
-A distant eigenvalue $\lambda_{(i)}$ results in an exponentially small singular value $\sigma_i(t)$.
+An eigenvalue $\lambda_{(i)}$ \textcolor{blue}{distant from $t$} results in an exponentially small singular value $\sigma_i(t)$.
 In particular, for a small value of $\sigma$, all singular values corresponding to eigenvalues outside a certain window $[t - d, t + d]$ are negligibly small; see \reffig{fig:numerical-rank}.

### 2025-02-25 paper/analysis.tex
incorporate proofreading changes and complete function docstrings

@@
 
-For constant $\mtx{\Omega}$ \textcolor{blue}{Shoud it be constant $B$?}, the estimator~\refequ{equ:nystrompp-trace-estimator} coincides with the Nyström++ estimator from~\cite{persson-2022-improved-variants}, which is based on the Hutch++ estimator~\cite{meyer-2021-hutch-optimal}.
+For constant $\mtx{B}$, the estimator~\refequ{equ:nystrompp-trace-estimator} coincides with the Nyström++ estimator from~\cite{persson-2022-improved-variants}, which is based on the Hutch++ estimator~\cite{meyer-2021-hutch-optimal}.
 In this situation (constant $\mtx{B}$), these estimators were both shown to achieve a relative $\varepsilon$-error with $\mathcal{O}(\varepsilon^{-1})$ matrix-vector products only, independent of the singular value decay of $\mtx{B}$.
@@
     \end{align*}
-    where $\lVert \cdot \rVert _{\ast}$ denotes the nuclear norm and we used \cite[Theorem 1]{gittens-2011-spectral-norm} in the \textcolor{blue}{second} equality.
+    where $\lVert \cdot \rVert _{\ast}$ denotes the nuclear norm and we used \cite[Theorem 1]{gittens-2011-spectral-norm} in the last equality.
 This allows us to apply \cite[Theorem 5]{kressner-2024-randomized-lowrank}, which immediately implies the statement claimed by the theorem.

### 2025-02-25 paper/intro.tex
incorporate proofreading changes and complete function docstrings

@@
 
-Being composed of distributions \textcolor{blue}{Why composed of distributions lead to bad approximation, I am wondering if we mean Dirac delta here?}, the function $\phi$ itself is hard to approximate.
-However, as many application\textcolor{blue}{s} only require a rough estimate of $\phi$, it usually suffices to instead approximate a smoothed spectral density of the form
+Being composed of Dirac delta distributions, the function $\phi$ itself is hard to approximate.
+However, as many applications only require a rough estimate of $\phi$, it usually suffices to instead approximate a smoothed spectral density of the form
 \begin{equation}

### 2025-02-25 paper/numresults.tex
incorporate proofreading changes and complete function docstrings

--- 2025-02-25 paper/numresults.tex
@@
 
-In this section, we verify the performance the Chebyshev-Nyström++ algorithm proposed in this paper in various application scenarios, from electronic structure interaction, statistical thermodynamics, and neural network optimization.
+In this section, we verify the performance of the Chebyshev-Nyström++ algorithm proposed in this paper in various application scenarios, from electronic structure interaction, statistical thermodynamics, and neural network optimization.
 

### 2025-02-25 algorithms/chebyshev_nystrom.py

@@ -105,12 +105,15 @@ def chebyshev_nystrom(A, t=0, m=100, n_Psi=10, n_Omega=10, kernel=gaussian_kerne
         The threshold on the Hutchinson estimate of g_sigma. If it is below this
         value, instead of solving the possibly ill-conditioned generalized
         eigenvalue problem, we set the spectral density at that point to zero.
+    rcond : float > 0
+        The relative threshold (with respect to largest eigenvalue) for
+        setting small eigenvalues to zero for computing pseudoinverse by lstsq.
     seed : int >= 0
-        The seed for generating the random matrix W.
+        The seed for generating the random matrix Omega.
 
     Returns
     -------
-    phi : np.ndarray
+    phi : np.ndarray (n_t, )
         Approximations of the spectral density at the points t.
 
     """

### 2025-02-25 algorithms/krylov_aware.py

@@ -18,37 +18,37 @@ def lanczos(A, X, k, reorth_tol=0.7, return_matrix=True, extend_matrix=True, dty
 
         A @ U[:, :k] = U[:, :k+1] @ T
 
-    after k iterations of the Lanczos method.
+    after k iterations of the Lanczos method. Multiple starting vectors can be
+    passed as X = [x_1, x_2, ..., x_m].
 
     Parameters
     ----------
+    A : np.ndarray (n, n)
+        Symmetric matrix.
+    X : np.ndarray (n, ) or (n, m,^)
+        Collection of starting vectors.
+    k : int > 0
+        Number of iterations.
     reorth_tol : float < 1
         The tolerance for reorthogonalizing the Krylov basis between iterations.
     return_matrix : bool
         Whether to return the (full) tridiagonal matrix H or arrays of its
         diagonal and off-diagonal elements.
-    return_matrix : bool
+    extend_matrix : bool
         Whether to extend the orthogonal matrix U with one more column after
         the last iteration.
+    dtype : type or str
+        The datatype of the matrix, such that auxiliary arrays are the same.
 
     Returns
     -------
-    TODO:
+    U : np.ndarray (m, n, k + 1)
+        Orthogonal basis of the block Krylov subspace(s) K(A, X).
+    T or a, b : np.ndarray (m, k + 1, k + 1) or (m, k + 1), (m, k + 1)
+        The block tridiagonal matrix T or its tridiagonal blocks a, b.
 
-    Example
-    -------
-    TODO:
-    >>> import numpy as np
-    >>> from roughly.approximate.krylov import LanczosDecomposition
-    >>> decomposition = LanczosDecomposition()
-    >>> A = np.random.randn(100, 100)
-    >>> A = A + A.T
-    >>> X = np.random.randn(100)
-    >>> U, H = decomposition.compute(A, X, 10)
-    >>> np.testing.assert_allclose(A @ U[:, :10] - U @ H, 0, atol=1e-10)
-    >>> U, H = decomposition.refine(1)
-    >>> np.testing.assert_allclose(A @ U[:, :11] - U @ H, 0, atol=1e-10)
-
+    References
+    ----------
     [1] Lanczos, C. (1950). "An iteration method for the solution of the
         eigenvalue problem of linear differential and integral operators".
[…]

### 2025-02-25 matrices/neural_network.py

@@ -158,10 +158,6 @@ def hessian_transpose_product(gradient, params, x):
     -------
     htp : torch.Tensor
         The result of the product of the Hessian transposed with x.
-
-    References
-    ----------
-    [1] TODO
     """
     htp = torch.autograd.grad(gradient, params, x.T, retain_graph=True, is_grads_batched=x.ndim > 1)
     htp = torch.cat([e.reshape(x.shape[-1], -1) for e in htp], dim=1)
@@ -211,10 +207,6 @@ class hessian(object):
         -------
         spectral_norm : float
             Approximation of spectral norm.
-
-        References
-        ----------
-        [2] TODO
         """
         x = torch.randn(self.num_parameters)
         Q, T = lanczos(self, x, k, dtype=np.float64)

### 2025-02-25 matrices/quantum_spin.py

@@ -2,7 +2,15 @@ import numpy as np
 import scipy as sp
 
 def hamiltonian(N=20, s=0.5, h=0.3, J=1):
+    """
+    Implements Hamiltonian matrix of Heisenberg spin chain [1].
 
+    References
+    ----------
+    [1] Chen T. and Hallman, E. (2023). "Krylov-Aware Stochastic Trace
+        Estimation". SIAM Journal on Matrix Analysis and Applications.
+        Vol. 44 (3). doi:10.1137/22M1494257.
+    """
     M = int(2*s+1)
 
     Jz_T = h*np.ones(N)
@@ -54,7 +62,15 @@ def hamiltonian(N=20, s=0.5, h=0.3, J=1):
     return out.real
 
 def partition_function(beta, N, h, J, E_min):
-    
+    """
+    Analytic solution to partition function of Heisenber spin chain [1].
+
+    References
+    ----------
+    [1] Chen T. and Hallman, E. (2023). "Krylov-Aware Stochastic Trace
+        Estimation". SIAM Journal on Matrix Analysis and Applications.
+        Vol. 44 (3). doi:10.1137/22M1494257.
+    """
     beta = np.asarray(beta, dtype=np.float64)
 
     k = np.arange(1, N+1) * np.pi / (N + 1)

### 2025-02-25 paper/intro.tex
fix bibliography entries and hyphenation

@@
 Evaluating this matrix function \emph{exactly} requires the diagonalization of $\mtx{A}$, which does not seem to yield any gains compared to the original problem of computing its spectral distribution.
-The crucial point is that we can now \emph{estimate} the trace by matrix-vector products with $\mtx{B}(t)$, which -- thanks to the smoothness of $g_\sigma$ -- can be well approximated by a few matrix-vector products with $\mtx{A}$.
+The crucial point is that we can now \emph{estimate} the trace by matrix-vector products with $\mtx{B}(t)$, which --- thanks to the smoothness of $g_\sigma$ --- can be well approximated by a few matrix-vector products with $\mtx{A}$.
 
@@
 
-\paragraph{Structure.} In \refsec{sec:analysis} we will analyze the error of three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter-dependent matrices with constant random matrix.
+\paragraph{Structure.} In \refsec{sec:analysis} we will analyze the error of three standard trace estimators --- the Girard-Hutchinson, Nyström, and Nyström++ estimators --- when they are applied to parameter-dependent matrices with constant random matrix.
 In \refsec{sec:application}, we propose and analyze the Chebyshev-Nyström++ estimator, which uses these estimators to approximate the smoothed spectral density of real symmetric matrices.

### 2025-02-25 paper/spectraldensity.tex
incorporate proofreading changes and complete function docstrings
fix bibliography entries and hyphenation

@@
 \refalg{alg:nystrom-chebyshev-pp} requires $(2m + 1)(n_{\mtx{\Psi}} + n_{\mtx{\Omega}})$ matrix-vector multiplications with $\mtx{A}$, which can be expected to dominate its cost.
-For each of the evaluation points $t_1, \dots, t_{n_t}$ another $\mathcal{O}(m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + (n_{\mtx{\Omega}}^3 + n_{\mtx{\Omega}} n_{\mtx{\Psi}}^2 + 1))$ operations are needed, which is independent of $n$ -- thanks to the precomputations discussed above.
+For each of the evaluation points $t_1, \dots, t_{n_t}$ another $\mathcal{O}(m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + (n_{\mtx{\Omega}}^3 + n_{\mtx{\Omega}} n_{\mtx{\Psi}}^2 + 1))$ operations are needed, which is independent of $n$ --- thanks to the precomputations discussed above.
 
@@
 where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from $t$, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t) \ge 0$.
-An eigenvalue $\lambda_{(i)}$ \textcolor{blue}{distant from $t$} results in an exponentially small singular value $\sigma_i(t)$.
+An eigenvalue $\lambda_{(i)}$ distant from $t$ results in an exponentially small singular value $\sigma_i(t)$.
 In particular, for a small value of $\sigma$, all singular values corresponding to eigenvalues outside a certain window $[t - d, t + d]$ are negligibly small; see \reffig{fig:numerical-rank}.

### 2025-02-25 paper/numresults.tex
add justification for choices in numerical experiments

@@
 
-\reffig{fig:convergence} displays the convergence behavior of Chebyshev-Nyström++ applied to the matrix resulting from the discretization of the Hamiltonian for $n_c = 1$ and $n_c = 2$.
-As a reference, we use the eigenvalues computed by NumPy's dense eigenvalue solver.
+\reffig{fig:convergence} displays the convergence behavior of Chebyshev-Nyström++ applied to the matrix $\mtx{A}$ resulting from the discretization of the Hamiltonian for $n_c = 1$ and $n_c = 2$.
+Unlike \cite{lin-2017-randomized-estimation}, where just a small portion of the spectral density is considered, we approximate the whole spectrum of $\mtx{A}$ (transformed to $[-1, 1]$) in accordance with our theoretical results (\refsec{subsubsec:chebyshev-nystrom-analysis}) and to demonstrate the effectiveness of our implementational particularities (\refsec{subsubsec:chebyshev-nystrom-implementation}).
+As a reference, we use the eigenvalues computed by NumPy's Hermitian eigenvalue solver.
 \begin{figure}[ht]
@@
         \input{plots/convergence-1.pgf}
-        \caption{$n_c = 1$}
+        \caption{$n_c = 1.$}
         \label{fig:convergence-1}
@@
         \input{plots/convergence-2.pgf}
-        \caption{$n_c = 2$}
+        \caption{$n_c = 2.$}
         \label{fig:convergence-2}
@@
 %\end{figure}%
-As expected from \refthm{thm:hutchinson}, Girard-Hutchinson alone ($n_{\mtx{\Omega}} = 0$) require $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ samples to achieve an error of order $\varepsilon$.
+As expected from \refthm{thm:hutchinson}, Girard-Hutchinson alone ($n_{\mtx{\Omega}} = 0$) requires $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ samples to achieve an error of order $\varepsilon$.
 Because the eigenvalues of the matrix are spread out, we observe the better convergence of the Nyström approximation discussed at the end of \refsec{sec:application} once $n_{\mtx{\Omega}}$ is sufficiently large.
-We can observe the errors stagnate at around $10^{-6}$ as a consequence of the error made in the Chebyshev approximation.
+At this point, we can observe the errors stagnate at around $10^{-6}$ as a consequence of the error made in the Chebyshev approximation.
+Due to the higher eigenvalue density of $\mtx{A}$ for $n_c = 2$, the stronger convergence can be observed to kick in later in this case, because a larger $n_{\mtx{\Omega}}$ is needed for a good Nyström approximation.
+In fact, for $n_c = 5$ the eigenvalue density is so high that to profit from this effect, $n_{\mtx{\Omega}}$ would need to be chosen so large that the computations become unfeasible.
 

### 2025-02-25 paper/analysis.tex
make notation consistent and clean up tex files

@@
 entry $b_{ij}(t)$ is a 
-%\begin{equation*}
-%     \mtx{B}(t) = \begin{bmatrix}
-%         b_{11}(t) & b_{12}(t) & \dots & b_{1n}(t) \\
-%         b_{21}(t) & b_{22}(t) & \dots & b_{2n}(t) \\
-%         \vdots & \vdots & \ddots & \vdots \\
-%         b_{n1}(t) & b_{n2}(t) & \dots & b_{nn}(t) \\
-%     \end{bmatrix} \in \mathbb{R}^{n \times n},
-% \end{equation*}
 continuous function on a bounded interval $[a,b]$. In this section, we 
@@
 
-%Write:
-%\begin{itemize}
-% \item fast singular value decay, intuitively sufficient to consider a low-rank %approximation
-% \item motivated by our application we only consider symmetric positive %semi-definite case here (smoothing kernels are positive) for which Nystroem %approximation is the right choice
-%\end{itemize}
-
-%\textcolor{red}{Moved from below:
-%For symmetric positive semi-definite matrices whose singular values decay %rapidly, the Nyström estimate \refequ{equ:nystrom-trace-estimator} can be %significantly better than the Girard-Hutchinson estimate.
-This is reflected in %the following theorem.}
-%
-%
-%Alternatively, the trace of a symmetric matrix whose singular values decay %quickly can be approximated well by using a Gaussian sketching matrix $\mtx%{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström %approximation \cite{gittens-2013-revisiting-nystrom}
-
 \paragraph{Nyström++ estimator.}
@@
 where $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$, $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ are independent Gaussian random matrices.
-%\textcolor{red}{??? What about XTrace? Does it apply; should we mention it somewhere? ???}
 
@@
 Unfortunately, we do not see a way of marrying these observations with the efficient implementation of XNysTrace described in \cite[Section 2.2]{epperly-2024-xtrace-making}, which limits its suitability for spectral density estimation.
-%We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.
-
-%In the remainder of this section, we will derive upper bounds on the error of these estimators.
-%
-%\begin{table}[ht]
-%\centering
-%\renewcommand{\arraystretch}{1.2}
-%\begin{tabular}{@{}lcc@{}}
-%\toprule
-%Estimator & Matrix & $m=1600$\\
-%\midrule
-%Girard-Hutchinson & symmetric & $\mathcal{O}(\varepsilon^{-2})%$ \\
-%Nyström & symmetric positive semi-definite & $\mathcal{O}%(\varepsilon^{-2})$ \\
-%Nyström++ & symmetric positive semi-definite & $\mathcal{O}%(\varepsilon^{-1})$  \\
-%\bottomrule
-%\end{tabular}
-%\end{table}
-
-%as we will see in Section ???.
 
@@
 \end{equation*}
-
-%\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
-%    For a symmetric matrix $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ that depends continuously on $t \in [a, b]$, consider the  Girard-Hutchinson estimator~\refequ{equ:hutchinson-trace-estimator} with an $n\times n_{\mtx{\Psi}}$ Gaussian random matrix $\mtx{\Psi}$.
-Then for any $p \geq 1, p \in \mathbb{R}$ and $\gamma \geq 1$ the bound
-%    \begin{equation} \label{equ:girard-hutchinson-bound}
-%        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq c \gamma \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}}  \right\} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
-%    \end{equation}
-%    holds with probability at least $1 - \gamma^{-p}$.
-%    In particular, given $\varepsilon \in (0, ce)$ and $\delta \in (0, e^{-1})$, the bound $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$ holds
-%    with probability at least $1-\delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1}))$.
-%\end{theorem}
 
@@
     \end{equation}
-%    The proof proceeds by bounding moments for fixed values of $t$ and then uses Minkowski's integral inequality.
-We start with a standard embedding trick (see, e.g.,~\cite[Theorem 1]{cortinovis-2022-randomized-trace}) for which we define
-%    \begin{equation}
-%        \widetilde{\mtx{B}}(t)
-%        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
-%            \mtx{B}(t) & & \\
-%            & \ddots & \\
-%            & & \mtx{B}(t)
-%        \end{pmatrix}
-%        \quad \text{and} \quad
-%        \widetilde{\vct{\psi}} = \begin{pmatrix}
-%            \vct{\psi}_1 \\
-%            \vdots \\
-%            \vct{\psi}_{n_{\mtx{\Psi}}}
-%        \end{pmatrix}.
-%    \end{equation}
-%    Note that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) =  \Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}}$, $\lVert \widetilde{\mtx{B}}(t) \rVert _2 = \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}}$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold.
-Therefore, by \reflem{lem:hutchinson-sub-exponential}, $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is a centered and $(2 \lVert \mtx{B}(t) \rVert _F^2 (1 - 1/\beta)^{-1} / n_{\mtx{\Psi}}, 2 \lVert \mtx{B}(t) \rVert _2 \beta / n_{\mtx{\Psi}})$-sub-exponential random variable for any $\beta > 1$.
-Thus, we can apply \reflem{lem:sub-exponential-moments} to bound
-%    \begin{equation}
-%        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq 2 \left(\sqrt{\frac{\pi}{1-1/\beta}} \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F + 4 \beta \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 \right).
-%    \end{equation}
-%    Bounding $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$ and letting $\beta = (1 + \sqrt{1 + \pi/4}) / 2$ we get
-%    \begin{equation}
-%        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq \underbrace{4 (1 + \sqrt{1 + \pi/4})}_{=~c} \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}} \right) \lVert \mtx{B}(t) \rVert _F.
-%        \label{equ:moment-bound-hutchinson}
-%    \end{equation}
 
@@
 \end{proof}
-%\begin{proof}
-%    The proof proceeds by bounding moments for fixed values of $t$ and then uses Minkowski's integral inequality.
-We start with a standard embedding trick (see, e.g.,~\cite[Theorem 1]{cortinovis-2022-randomized-trace}) for which we define
-%    \begin{equation}
-%        \widetilde{\mtx{B}}(t)
-%        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
-%            \mtx{B}(t) & & \\
-%            & \ddots & \\
-%            & & \mtx{B}(t)
-%        \end{pmatrix}
-%        \quad \text{and} \quad
-%        \widetilde{\vct{\psi}} = \begin{pmatrix}
-%            \vct{\psi}_1 \\
-%            \vdots \\
-%            \vct{\psi}_{n_{\mtx{\Psi}}}
-%        \end{pmatrix}.
-%    \end{equation}
-%    Note that $\Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}} = \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$, $\lVert \widetilde{\mtx{B}}(t) \rVert _2 = \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}}$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold.
-Therefore, by \reflem{lem:hanson-wright-moments} we have
-%    \begin{align}
-%        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))]
-%        &= \mathbb{E}^{p}[\Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}}] \notag \\
-%        &\leq c \max\left\{ \sqrt{p} \lVert \widetilde{\mtx{B}}(t) \rVert _F, p \lVert \widetilde{\mtx{B}}(t) \rVert _2 \right\} \notag \\
-%        &= c \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F, \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 \right\} \notag \\
-%        &\leq c \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}} \right\}  \lVert \mtx{B}(t) \rVert _F.
-%        \label{equ:moment-bound-hanson-wright}
-%    \end{align}
-%
-%    To address the parameter-dependent case, we first note that the continuity assumption implies that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is measurable.
-Therefore, we can apply Minkowski's integral inequality~\cite[Theorem 202]{hardy-1952-inequalities} to conclude from~\refequ{equ:moment-bound-hanson-wright} that
-%    \begin{align}
-%        \mathbb{E}^{p}\left[ \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t  \right]
-%        &\leq \int_{a}^{b} \mathbb{E}^{p}\left[ \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) \right]~\mathrm{d}t \notag \\
-%        &\leq c \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}}  \right\} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t .
-%    \end{align}
-%    Consequently, by Markov's inequality,  the inequality 
-%    \begin{equation} 
-%        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq c \gamma \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}}  \right\} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
-%    \end{equation}
-%    holds for any $\gamma \geq 1$ with probability at least $1 - \gamma^{-p}$. 
-%
-%    Setting $\gamma = e$ and choosing $p = \log(\delta^{-1}) \geq 1$, the second part of the theorem follows when taking $n_{\mtx{\Psi}} = \lceil c^2 e^2 \log(\delta^{-1}) \varepsilon^{-2} \rceil$ and restricting $\varepsilon \leq c e$.
-%    %where $\lfloor x \rfloor$ is the floor function which returns the largest integer smaller than or equal to %$x$,
-%    %we get that for all $\delta \in (0, 1)$
-%\end{proof}
 
@@
 
-%The idea behind the proof is to bound  to extend the results to the continuous parameter case.
-We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
-%\begin{proof} 
-%\textcolor{red}{??? I feel that the block diagonal embedding trick should come first, like in the proof of Meyer et al. Hopefully, this gets rid of the additional $\log \delta$ factor ???}
-%The proof proceeds by bounding moments for fixed values of $t$ and then use Minkowski's integral inequality.
-We start by considering the residual of the 1-query Girard-Hutchinson estimator for a fixed matrix $\mtx{B}$:
-%    \begin{equation}
-%        r(\mtx{B}, \vct{\psi}) := \Trace(\mtx{B}) - \vct{\psi}^{\top} \mtx{B} \vct{\psi}
-%    \end{equation}
-%%    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers
-%for a Gaussian random vector $\vct{\psi}$.
-%%
-%%    First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence.
-%The proof of \cite[Lemma 3]{cortinovis-2022-randomized-trace} shows that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm.
-By \cite[Theorem 2.3]{boucheron-2013-concentration-inequalities} this implies for every $k \in \mathbb{N}$ that
-%    \begin{align*}
-%        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
-%        &\leq k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k}  \\
-%        &\le \big( k! 2^{4 k} + (2 k)! 2^{6 k}\big) \lVert \mtx{B} \rVert _F^{2 k} 
-%        \le \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
-%    \end{align*}
-%%     Because $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
-%%     \begin{equation}
-%%         \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right] \leq 
-%%     \end{equation}
-%    Stirling's approximation \cite{robbins-1955-remark-stirling} gives $(2 k)!
-< 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$.
-\textcolor{red}{???
-this norm should be defined more prominently somewhere else Consequently, the $L^{2k}$ norm $\mathbb{E}^{2k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{2k} \right] \right)^{\sfrac{1}{2k}}$ ???
-} \textcolor{green}{Moved it to start of section.} Consequently, the high order moments of $r(\mtx{B}, \vct{\psi})$ are bounded by
-%    \begin{equation}
-%        \mathbb{E}^{2k}\left[ r(\mtx{B}, \vct{\psi}) \right]
-%        \leq \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F \leq 2^4 k \lVert \mtx{B} \rVert _F,
-%        \label{equ:hutchinson-trace-onequery-fixed}
-%    \end{equation}
-%    using the inequality $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, which can be verified using the monotonic decrease of the left-hand side with respect to $k \in \mathbb{N}$.
-%
-%    To address the parameter-dependent case, we first note that the continuity assumption implies that $r(\mtx{B}(t), \vct{\psi})$ is measurable.
-Therefore, we can apply Minkowski's integral inequality~\cite[Theorem 202]{hardy-1952-inequalities} to conclude from~\refequ{equ:hutchinson-trace-onequery-fixed} that
-%    \begin{equation}
-%        \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t  \right]
-%        \leq \int_{a}^{b} \mathbb{E}^{2 k}\left[ r(\mtx{B}(t), \vct{\psi}) \right]~\mathrm{d}t
-%        \leq 2^4 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
-%    \end{equation}
-%    Consequently, by Markov's inequality,  the inequality 
-%    \begin{equation} 
-%        \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t \leq 2^4 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
-%        \label{equ:hutchinson-trace-onequery-uniform}
-%    \end{equation}
-%    holds for any $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$. 
-%
-%    To extend the bound~\refequ{equ:hutchinson-trace-onequery-uniform} from $1$ to $n_{\mtx{\Psi}}$ queries, we use a standard embedding trick (see, e.g.,~\cite[Theorem 1]{cortinovis-2022-randomized-trace}).
-Letting
-%    \begin{equation}
-%        \widetilde{\mtx{B}}(t)
-%        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
-%            \mtx{B}(t) & & \\
-%            & \ddots & \\
-%            & & \mtx{B}(t)
-%        \end{pmatrix}
-%        \quad \text{and} \quad
-%        \widetilde{\vct{\psi}} = \begin{pmatrix}
-%            \vct{\psi}_1 \\
-%            \vdots \\
-%            \vct{\psi}_{n_{\mtx{\Psi}}}
-%        \end{pmatrix},
-%    \end{equation}
-%    we note that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) = r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold.
-Thus, the claimed probabilistic bound~\refequ{equ:girard-hutchinson-bound} follows from applying~\refequ{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$.
-%%     Then 
-%%     where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors.
-Then $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$, and
-%%     \begin{equation}
-%%         r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}}) = \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j.
-%%     \end{equation}
-%%     Hence, by applying \refequ{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$ we conclude that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
-%%     \begin{equation}
-%%         \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j \right| ~ \mathrm{d}t
-%%         \leq 2^4 k \gamma \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
-%%     \end{equation}
-%
-%    %Setting $\delta = \gamma^{-2 k}$ and choosing $k = \lceil \log(\delta^{-1}) %\rceil$, where $\lceil x \rceil$ is the ceiling function which returns the greatest %integer larger than or equal to $x$, we get that for all $\delta \in (0, e^{-\sfrac%{1}{2}})$
-%    %\begin{equation}
-%    %    k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\sfrac{1}{2\lceil \log%(\delta^{-1}) \rceil}}
-%    %    = \lceil \log(\delta^{-1}) \rceil e^{\sfrac{1}{2}\frac{\log(\delta^{-1})}%{\lceil \log(\delta^{-1}) \rceil}}
-%    %    \leq 2 \log(\delta^{-1}) e^{\sfrac{1}{2}},
-%    %\end{equation}
-%    %from which follows the second part of the theorem when $n_{\mtx{\Psi}} = (2^5 e^%{\sfrac{1}{2}} \varepsilon^{-1} \log(\delta^{-1}))^2$.
-%
-%    Setting $\gamma = e$ and choosing $k = \lfloor \log(\delta^{-1}) / 2 \rfloor$,
-%    %where $\lfloor x \rfloor$ is the floor function which returns the largest integer smaller than or equal to %$x$,
-%    %we get that for all $\delta \in (0, 1)$
-%    the second part of the theorem follows when taking $n_{\mtx{\Psi}} = \lceil (2^3 e \varepsilon^{-1} \log(\delta^{-1}))^2 \rceil$.
-%    %where $\lceil x \rceil$ is the ceiling function which returns the smallest integer larger than or equal to $x$.
-%    %Consequently, with probability $1 - \delta$ we have
-%    %\begin{equation}
-%    %    \int_{a}^{b} \left| \frac{1}{n_{\mtx{\Psi}}} \sum_{i=1}^{n_{\mtx{\Psi}}} \vct{\psi}_i^{\top} \mtx{B}(t) \vct{\psi}_i - \Trace(\mtx{B}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\sfrac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
-%    %\end{equation}
-%\end{proof}
-%
-%Compared to the analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, the number of queries $n_{\mtx{\Psi}}$ needed by~\refthm{thm:hutchinson} is roughly a factor $\log(\delta^{-1})$ larger.
-%
 \subsubsection{Nyström estimator}
@@
 Then, for any $\gamma \geq 1$, the bound
-    \[
+    \begin{equation*}
         \int_{a}^{b} \big| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \big| ~\mathrm{d}t
         \leq \gamma^2 (1 + r) \int_{a}^{b} \sum_{i = r+1}^{n} \sigma_i(\mtx{B}(t)) ~\mathrm{d}t.
-    \]
+    \end{equation*}
     holds with probability at least $1 - \gamma^{-(n_{\mtx{\Omega}} - r)}$.
@@
 
-Our analysis mimics the strategy from~\cite{meyer-2021-hutch-optimal}, by first establishing a relative Frobenius norm error proportional to $n_{\mtx{\Omega}}^{-1/2}$ for the parameter-dependent Nyström approximation.
+Our analysis mimics the strategy from~\cite{meyer-2021-hutch-optimal}, by first establishing a relative Frobenius norm error proportional to $n_{\mtx{\Omega}}^{-\sfrac{1}{2}}$ for the parameter-dependent Nyström approximation.
 
@@
     \end{equation}
-        holds with probability at least $1 - \gamma^{-n_{\mtx{\Omega}} / 4}$ for 
+        holds with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$ for 
         $c = 154$.
@@
 
-%\todo{Proof idea: structural bound, then higher order moment bound to apply Markov's inequality.}
 \begin{proof}
@@
         \mathbb{E}^{q}\left[ \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \big\|_{(4)}^2 \right]
-        %&= \mathbb{E}^{p}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )( \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger})^{\top} \rVert _F \right] && \text{($\lVert \mtx{A} \rVert _{(4)}^2 = \lVert \mtx{A} \mtx{A}^{\top} \rVert _{F}$)} \notag \\
-        %&\leq \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] && \text{($\lVert \mtx{A} \mtx{B} \rVert _F \leq \lVert \mtx{A} \rVert _F \lVert \mtx{B} \rVert _2$)} \notag \\
-        \leq \mathbb{E}^{q}\left[ \sqrt{k} \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \big\| _2^2 \right]%  && \text{($\lVert \mtx{A} \rVert _{(4)}^2 \leq \sqrt{k} \lVert \mtx{A} \rVert _2^2$)} \notag \\
-        %\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]%  && \text{(submultiplicativity)} \notag \\
-        \leq \sqrt{k} \mathbb{E}^{q}\left[ \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2  \big\|_2^2\right] \mathbb{E}^{q}\left[ \big\| \mtx{\Omega}_1^{\dagger} \big\|_2^2  \right].% && \text{(independence)}
+        \leq \mathbb{E}^{q}\left[ \sqrt{k} \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \big\| _2^2 \right]
+        \leq \sqrt{k} \mathbb{E}^{q}\left[ \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2  \big\|_2^2\right] \mathbb{E}^{q}\left[ \big\| \mtx{\Omega}_1^{\dagger} \big\|_2^2  \right].
         \label{equ:nystrom-proof-processed-tail}
@@
     \begin{equation*}
-     \mathbb{E}\left[ \big\| ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \big\|_2^{\sfrac{n_{\mtx{\Omega}}}{4}} \right]%  \notag \\
+     \mathbb{E}\left[ \big\| ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \big\|_2^{\sfrac{n_{\mtx{\Omega}}}{4}} \right]
         \leq
@@
     to the result of~\cite[Lemma B.3]{tropp-2023-randomized-algorithms}, this inequality is valid for arbitrarily large $n_{\mtx{\Omega}}$, at the expense of a slightly larger constant.
-%
-%     To match the decay rate of the moments of the first term in \refequ{equ:nystrom-proof-persson-bonud}, we will need to ensure that this term also is of order $\mathcal{O}(\Trace(\mtx{B}) / \sqrt{k})$.
-We bound the moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ similarly to \cite[Lemma B.3]{tropp-2023-randomized-algorithms}, but without restricting $q$ to be smaller than $18$.
-The explicit integration of \cite[Equation B.7]{tropp-2023-randomized-algorithms} imposes the condition $n_{\mtx{\Omega}} - k - 2q \geq 0$.
-Both $n_{\mtx{\Omega}}$ and $k$ are integers and must be of the same order to ensure a decay of $\mathcal{O}(\Trace(\mtx{B}) / \sqrt{k})$.
-To avoid restricting the choice of $n_{\mtx{\Omega}}$ too much, we let it be even and set $k = n_{\mtx{\Omega}}/2$.
-The moment $q$ should be chosen as large as possible to ensure a fast decay of the failure probability, so we fix it to $q = n_{\mtx{\Omega}}/4$.
-Therefore, we get
-%     \begin{equation}
-%         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
-%         = \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{n_{\mtx{\Omega}}}{4}} \right]^{\sfrac{4}{n_{\mtx{\Omega}}}}%  \notag \\
-%         \leq \left( 1 + \frac{n_{\mtx{\Omega}}}{2} \right)^{\sfrac{4}{n_{\mtx{\Omega}}}} \left( \frac{1}{(\frac{n_{\mtx{\Omega}}}{2} + 1)!}\right)^{\frac{2}{\frac{n_{\mtx{\Omega}}}{2} + 1}} \left( \frac{3 n_{\mtx{\Omega}}}{4}\right).
-%     \end{equation}
-%     From the expansion of the exponential function as a power series $e^n = 1 + n + \dots + n^n/n!
-+ \dots$ for $n \in \mathbb{N}$ we see $e^n \geq 1 + n$ and $e^n \geq n^n / n!$, from which $(1 + n)^{\sfrac{1}{n}} \leq e$ and $\left( 1/n!
-\right)^{\sfrac{1}{n}} \leq e/n$ follow respectively.
-Hence,
-%     \begin{equation}
-%         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
-%         \leq e^2 \frac{e^2}{(\frac{n_{\mtx{\Omega}}}{2}+ 1)^2} \left( \frac{3 n_{\mtx{\Omega}}}{4}\right)
-%         \leq \frac{3 e^4}{n_{\mtx{\Omega}}}.
-%         \label{equ:pinv-spectral-norm-bound}
-%     \end{equation}
-%     As opposed to \cite[Lemma B.3]{tropp-2023-randomized-algorithms}, our bound holds for all even $n_{\mtx{\Omega}} \in \mathbb{N}$, at the cost of a factor of $e^2$.
-%
-    The second factor in~\refequ{equ:nystrom-proof-processed-tail} is bounded using \reflem{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$: % and $p = n_{\mtx{\Omega}}/2$:
+
+    The second factor in~\refequ{equ:nystrom-proof-processed-tail} is bounded using \reflem{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$:
     \begin{equation*}
@@
         \leq \frac{5}{4} n_{\mtx{\Omega}} \Big( 2 \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \big\|_2^2 + \frac{1}{n_{\mtx{\Omega}}} \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \big\|_F^2 \Big).
-      %  \label{equ:spectral-norm-bound-applied}
     \end{equation*}
-    %The $q$-th moment of the second term can be processed with standard matrix norm manipulations and the stochastic independence of $\mtx{\Omega}_1$ and $\mtx{\Omega}_2$ to
-    %\begin{equation}
-    %    \mathbb{E}^{q}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
-        %&= \mathbb{E}^{p}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )( \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger})^{\top} \rVert _F \right] && \text{($\lVert \mtx{A} \rVert _{(4)}^2 = \lVert \mtx{A} \mtx{A}^{\top} \rVert _{F}$)} \notag \\
-        %&\leq \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] && \text{($\lVert \mtx{A} \mtx{B} \rVert _F \leq \lVert \mtx{A} \rVert _F \lVert \mtx{B} \rVert _2$)} \notag \\
-    %    \leq \mathbb{E}^{q}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]%  && \text{($\lVert \mtx{A} \rVert _{(4)}^2 \leq \sqrt{k} \lVert \mtx{A} \rVert _2^2$)} \notag \\
-        %\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]%  && \text{(submultiplicativity)} \notag \\
-    %    \leq \sqrt{k} \mathbb{E}^{q}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2\right] \mathbb{E}^{q}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2  \right].% && \text{(independence)}
-    %    \label{equ:nystrom-proof-processed-tail}
-    %\end{equation}
-    %To match the decay rate of the moments of the first term in \refequ{equ:nystrom-proof-persson-bonud}, we will need to ensure that this term also is of order $\mathcal{O}(\Trace(\mtx{B}) / \sqrt{k})$.
-To this purpose, and to simplify the notation, we choose $k = n_{\mtx{\Omega}}/2$ and $q = n_{\mtx{\Omega}}/4$ for even $n_{\mtx{\Omega}}$ to ensure $k \in \mathbb{N}$.
-We can apply \reflem{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$ and $p = n_{\mtx{\Omega}}/2$ to bound
-    %\begin{equation}
-    %    \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \right]
-    %    \leq n_{\mtx{\Omega}} \left( 2 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{2}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
-    %    \label{equ:spectral-norm-bound-applied}
-    %\end{equation}
-    %The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[Lemma B.3]{tropp-2023-randomized-algorithms}
-    %\begin{equation}
-    %    \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
-    %    = \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{n_{\mtx{\Omega}}}{4}} \right]^{\sfrac{4}{n_{\mtx{\Omega}}}}%  \notag \\
-    %    \leq \left( 1 + \frac{n_{\mtx{\Omega}}}{2} \right)^{\sfrac{4}{n_{\mtx{\Omega}}}} \left( \frac{1}{(\frac{n_{\mtx{\Omega}}}{2} + 1)!}\right)^{\frac{2}{\frac{n_{\mtx{\Omega}}}{2} + 1}} \left( \frac{3 n_{\mtx{\Omega}}}{4}\right).
-    %\end{equation}
-    %From the expansion of the exponential function as a power series $e^n = 1 + n + \dots + n^n/n!
-+ \dots$ for $n \in \mathbb{N}$ we see $e^n \geq 1 + n$ and $e^n \geq n^n / n!$, from which $(1 + n)^{\sfrac{1}{n}} \leq e$ and $\left( 1/n!
-\right)^{\sfrac{1}{n}} \leq e/n$ follow respectively.
-Hence,
-    %\begin{equation}
-    %    \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
-    %    \leq e^2 \frac{e^2}{(\frac{n_{\mtx{\Omega}}}{2}+ 1)^2} \left( \frac{3 n_{\mtx{\Omega}}}{4}\right)
-    %    \leq \frac{3 e^4}{n_{\mtx{\Omega}}}.
-    %    \label{equ:pinv-spectral-norm-bound}
-    %\end{equation}
     Inserting this inequality and~\refequ{equ:pinv-spectral-norm-bound} into~\refequ{equ:nystrom-proof-processed-tail} gives
@@
         \leq \frac{15 e^4}{16}  \sqrt{n_{\mtx{\Omega}}} \Big( 2 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \Big).
-        %\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
     \end{equation*}
@@
         \label{equ:nystrom-proof-tail-bound}
-        %\leq \sqrt{k} \frac{e^4}{2} \frac{(k + n_{\mtx{\Omega}})(2p + n_{\mtx{\Omega}})}{(n_{\mtx{\Omega}} - k + 1)^2}  \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 + \frac{1}{n_{\mtx{\Omega}}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 \right).
     \end{equation}
@@
     follows from~\refequ{equ:frobenius-moment-bound} using Minkowski's integral inequality and Markov's inequality.
-%     
-%     
-%         
-%     and by the Minkowski's integral inequality \cite[Theorem 202]{hardy-1952-inequalities}, we get for $n_{\mtx{\Omega}} \geq 4$
-%     \begin{align}
-%         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \right] 
-%         &\leq \int_{a}^{b} \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F \right] ~\mathrm{d}t \notag \\
-%         %&\leq \mathbb{E}^{\sfrac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
-%         %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \mathbb{E}^{\sfrac{k}{2}} \left[ \int_{a}^{b}  \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
-%         %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
-%         &\leq \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
-%     \end{align}
-%     From Markov's inequality follows with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
-%     \begin{equation}
-%         \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t,
-%     \end{equation}
-%     from which follows the first statement.
 
     Fixing $\gamma = e$ and letting $n_{\mtx{\Omega}} = \lceil(c e \varepsilon)^{-2}  + 4 \log(\delta^{-1}) \rceil$ establishes the second part of the theorem.
-    %Setting $\delta = \gamma^{-\sfrac{k}{4}}$ and choosing $k = \lceil \varepsilon^{-2}\log(\delta^{-1}) \rceil$ we get
-    %\begin{align}
-    %    \gamma \frac{1}{\sqrt{k}}
-    %    &= \delta^{-\sfrac{2}{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} \frac{1}{\sqrt{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} && \text{(definition of $\delta$ and choice of $k$)} \notag \\
-    %    &= e^{2\varepsilon^{2} \frac{\varepsilon^{-2}\log(\delta^{-1})}{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} \frac{1}{\sqrt{\lceil \varepsilon^{-2} \log(\delta^{-1}) \rceil}} && \text{($\delta = e^{-\log(\delta^{-1})}$)} \notag \\
-    %    &\leq e^{2\varepsilon^{2}} \frac{\varepsilon}{\sqrt{ \log(\delta^{-1})}} && \text{($\lceil x \rceil \geq x$ if $x \geq 0$)}
-    %\end{align}
-    %which is smaller than $\varepsilon$ if $\delta \leq e^{-e^{4 \varepsilon^2}}$from which follows the second part of the theorem when using $n_{\mtx{\Omega}} = 2k$.
 \end{proof}
-
-% When compared to the equivalent result for constant matrices \cite[Lemma 3.2]{persson-2022-improved-variants}, our constant factor up front is a bit more than two times larger and our failure probability decays half a power slower.
 
@@
 Comparing \refthm{thm:nystrom-pp} with the analogous result for constant matrices \cite[Theorem 3.4]{persson-2022-improved-variants}, we notice an additional factor of $\sqrt{\log(\delta^{-1})}$ in the required number of random vectors $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$.
-
-% \begin{proof}
-%     Combining with union bound.
-%     \begin{align}
-%         &\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t \notag \\
-%         &= \int_{a}^{b} | \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)) - \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)) |~\mathrm{d}t && \text{(definition of estimators)} \notag \\
-%         &= \todo{c} \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t) \rVert _F ~\mathrm{d}t && \text{(with probability $1 - \delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1})$)} \notag \\
-%         &= \todo{c} \frac{1}{\sqrt{n_{\mtx{\Psi}} n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t)) ~\mathrm{d}t
-%     \end{align}
-% \end{proof}

### 2025-02-25 paper/intro.tex
make notation consistent and clean up tex files

@@
 unifies all methods from~\cite{lin-2017-randomized-estimation} in a single algorithm.
-%, and improves multiple algorithmic aspects to make them faster and more robust.
-%\textcolor{red}{speeding up the \enquote{spectrum sweeping} method by orders of magnitude.
-???
-Where do we actually demonstrate this?
-Illustrate/spell this out more explicitly in the numerical experiments ???}
 Combining our analysis of the randomized trace estimators with standard error bounds for Chebyshev approximation leads to~\refthm{thm:chebyshev-nystrom}, which shows that $\mathcal O(\varepsilon^{-1})$ matrix-vector products suffice to guarantee an error $\varepsilon$.

### 2025-02-25 paper/numresults.tex
make notation consistent and clean up tex files

@@
 in three dimensions. The potential $V$ interacting with the electrons is generated by Gaussian wells
-%\begin{equation}
     $v(r) = v_0 e^{-\lambda r^2}$, 
-%    \label{equ:5-experiments-gaussian-cell}
-%\end{equation}
 with $v_0 = -4$ and $\lambda = 8$, centered in cells of side length $L=6$ which are stacked $n_c \in \mathbb{N}$ times in each spatial dimension; see \reffig{fig:gaussian-well}.
@@
 The distribution of the eigenvalues of the Hamiltonian --- its spectral density --- allows one to interpret the system's energy levels.
-
-%\textcolor{red}{???
-This discussion of $n_c$ only makes sense if we have at least one experiment with $n_c = 5$ or so.
-Could you please include such an experiment?
-For example, do Figure 4.2 again for larger $n_c$?
-Just use an overapproximation (large $n_\Psi$, $n_\Omega$) for approximating the true spectral distribution.
-???}
 
@@
 \end{figure}%
-%\begin{figure}[ht]
-%    \centering
-%    \input{plots/convergence.pgf}
-%    \caption{Error vs.
-number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ (with fixed $\sigma=0.005$ and $m=2000$) when applying  Chebyshev-Nyström++ to the Hamiltonian matrix from~\refsec{subsec:hamiltonian} with $n_c = 1$.}
-%    \label{fig:convergence}
-%\end{figure}%
 As expected from \refthm{thm:hutchinson}, Girard-Hutchinson alone ($n_{\mtx{\Omega}} = 0$) requires $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ samples to achieve an error of order $\varepsilon$.
@@
 In fact, for $n_c = 5$ the eigenvalue density is so high that to profit from this effect, $n_{\mtx{\Omega}}$ would need to be chosen so large that the computations become unfeasible.
-
-%\begin{figure}[ht]
-%    \centering
-%    \input{plots/convergence_8_sigma005.pgf}
-%    \caption{Error vs.
-number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ (with fixed $\sigma=0.005$ and $m=2000$) when applying  Chebyshev-Nyström++ to the Hamiltonian matrix from~\refsec{subsec:hamiltonian} with $n_c = 2$.}
-%    \label{fig:convergence}
-%\end{figure}%
 
@@
 Since this quadrature is independent of $g_{\sigma}$ and $t$, it can be applied to different smoothing kernels $g_{\sigma}$ and to as many values of $t$ as needed --- at little additional cost.
-
-%\textcolor{red}{???
-Please rewrite this.
-It does not come out whether~\cite{chen-2023-krylovaware-stochastic} actually cover spectral density estimation.
-Are the Gaussian matrices chosen constants in their paper/in our experiments?
-If this is not discussed in~\cite{chen-2023-krylovaware-stochastic}, comment more explicitly to which extent this algorithm benefits from constant random matrices.
-Do they have an analysis that applies to our situation, etc.
-???
-It also samples two Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$.
-$\mtx{\Omega}$ is used to extract a low-rank approximation to $g_{\sigma}(t \mtx{I}_n - \mtx{A})$ by running the block-Lanczos algorithm on $\mtx{A}$ with starting block $\mtx{\Omega}$ for $r$ iterations with reorthogonalization and subsequently $q$ without.
-The columns of $\mtx{\Psi}$ are then used to estimate the trace of the approximation residual with $r$ iterations of the Lanczos algorithm on $\mtx{A}$ ???
-Why $\mtx{A}$?
-???.}
 

### 2025-02-25 paper/spectraldensity.tex
make notation consistent and clean up tex files

@@
 As explained in the introduction, finding accurate approximations to this distribution is challenging, and we will therefore work with
-a smoothed version. %This still yields enough information
-%.
-Furthermore, in most applications the exact location of each individual eigenvalue is not important, but more so their approximate locations relative to each other, such as eigenvalue clusters, spectral gaps, or outliers.
-In the following we will work with a smoothed version of the spectral density, which is significantly easier to approximate at the cost of losing some of the finer characteristics of the spectrum.
+a smoothed version.
 
@@
 Assuming that the spectrum of $\mtx{A}$ is contained in $[-1,1]$, this gives rise to
-%need to evaluate the involved matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
-However, doing this exactly would require us to first diagonalize $\mtx{A}$, which is often a prohibitively expensive operation.
-For algorithmic reasons (see \refsec{subsubsec:chebyshev-nystrom-implementation}), we seek for approximations of this matrix function in terms of affine linear approximations, of which the was determined to be most suitable due to its approximation error guarantees and ease of manipulation.
-In every $t \in \mathbb{R}$ we expand
 \begin{equation}
@@
         \label{equ:chebyshev-interpolation-sup-error-kernel}
-        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
-        %\label{equ:2-chebyshev-interpolation-error}
     \end{equation}
@@
     The proof of this result is standard.
-% and follows the one given in the arXiv preprint of \cite[Theorem 2]{lin-2017-randomized-estimation}, but attains a slightly different dependence on $\sigma$.
 Let $\mathcal{E}_{\chi} \subset \mathbb{C}$ denote the Bernstein ellipse  with foci $\{-1, +1\}$ and the sum of semi-axes equal to $\chi > 1$.
@@
     \le \frac{1}{\sigma \sqrt{2 \pi}} e^{\sfrac{(\chi - \chi^{-1})^2}{8 \sigma^2}},
-    %\leq \frac{1}{n \sigma \sqrt{2 \pi}} \max_{x + iy \in \mathcal{E}_{\chi}} e^{\sfrac{y^2}{2\sigma^2}} 
     \end{equation*}
@@
 \end{align}
-
-%\begin{theorem}{$L^1$-error of Chebyshev approximation of spectral density}{chebyshev-error}
-%    Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$.
-Then the spectal density with Chebyshev approximation $\phi_{\sigma}^{(m)}$ satisfies
-%    \begin{equation}
-%        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq 2 E_{\sigma, m}
-%        \label{equ:chebyshev-interpolation-spectral-density}
-%        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
-%        %\label{equ:2-chebyshev-interpolation-error}
-%    \end{equation}
-%    for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and the error $E_{\sigma, m}$ from \reflem{lem:chebyshev-error}.%, and $t \in \mathbb{R}$.
-%The same result holds for the approximation of a non-negative Chebyshev approximation of \emph{even} degree $m \in \mathbb{N}$ but with $\underline{E}_{\sigma, m}$.
-%\end{theorem}
-%
-%\begin{proof}
-%    We observe that for all $t \in [-1, 1]$
-%    \begin{align}
-%        &\left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right| \notag \\
-%        &= \left| \frac{1}{n} \Trace(g_{\sigma}(t\mtx{I}_n - \mtx{A})) - \frac{1}{n} \Trace(g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})) \right|
-%        && \text{(definitions \refequ{equ:smooth-spectral-density} and \refequ{equ:expanded-spectral-density})} \notag \\
-%        &= \left| \frac{1}{n} \sum_{i=1}^n \left(g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i)\right) \right|
-%        && \text{($\lambda_1, \dots, \lambda_n$ eigenvalues of $\mtx{A}$)} \notag \\
-%        &\leq \max_{i = 1, \dots, n} \left| g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i) \right|
-%        && \text{(conservative upper bound)} \notag \\
-%        &\leq E_{\sigma, m}.% ~ \text{(or $\underline{E}_{\sigma, m}$)}.
-%        && \text{(using \reflem{lem:chebyshev-error})} \notag \\
-%    \end{align}
-%    Then, Hölder's inequality allows us to conclude
-%    \begin{equation}
-%        \int_{-1}^{1} | \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) | ~\mathrm{d}t
-%        \leq 2 \sup_{t \in [-1, 1]} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|
-%        \leq 2 E_{\sigma, m}.% ~ \text{(or $2 \underline{E}_{\sigma, m}$)}.
-%    \end{equation}
-%\end{proof}
 
@@
 \end{equation}
-%With a complexity of $\mathcal{O}(m \log(m))$ this procedure is computationally inexpensive and allows for very similar convergence bounds to \reflem{lem:chebyshev-error}, which we will derive \textcolor{red}{in the following?} in \refsec{subsubsec:chebyshev-nystrom-analysis}.
-%To this extent, we will first analyze the non-negative Chebyshev approximation.
 The corresponding non-negative approximation of the smoothed spectral density,
@@
         \label{equ:chebyshev-interpolation-sup-error-kernel-nonneg}
-        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
-        %\label{equ:2-chebyshev-interpolation-error}
     \end{equation*}
@@
 
-%In the following, we will show how the coefficients $\underline{\mu}_0(t), \dots, \underline{\mu}_m(t)$ of the non-negative Chebyshev approximation \refequ{equ:non-negative-chebyshev-approximation} can efficiently be computed and that this approximation exhibits similar convergence properties to the standard Chebyshev approximation.
-%\todo{
-%Maybe mention indefinite Nyström [Nakatsukasa], which won't give us desirable guarantees, and refinement of Chebyshev approximation [Francis Bach], which only slightly improves, but is much harder to implement.
-Side-note: it is possible to refine the coefficients of the Chebyshev approximation to get a better approximation in the supremum-norm over the set of all non-negative polynomials \cite%{fejer-1916-uber-trigonometrische}.
-The non-convex nature of this problem and the relatively small improvement we get with this technique made us decide to not pursue this further.
-%}
-
-%\subsection{Computation of approximation coefficients}
-
 \subsubsection{Computation of Chebyshev coefficients}
@@
 
-%\textcolor{red}{???
-I don't understand this derivation.
-Don't you think that~\refequ{equ:discrete-cosine-transform} only holds approximately for the exact $g_\sigma$ and it should be something like $g_\sigma \approx g^{(m)} = ...$.
-I think there is a major confusion of two different approaches here:
-%\begin{itemize}
-% \item The Chebyshev \emph{approximation} of $f$ is the $\ell^2$-orthogonal projection of $f$ on the first $m+1$ Chebyshev polynomials.
-The coefficients are computed from $\ell^2$ inner products between the Chebyshev polynomials and $f$.
-This is probably what Lin is doing.
-% \item The Chebyshev \emph{approximation} used by Chebfun and possibly others proceeds differently.
-It first replaces (or rather pretends to replace) $f$ by the interpolation of $f$ in (some) Chebyshev nodes.
-This interpolation admits an exact truncated approximation and the coefficients can be computed from a Fourier transform (or even DCT?) of the interpolation data.
-This might match what is written below, but I am not sure whether the interpolation nodes are the same.
-Please check with Trefethen's book.
-%\end{itemize}
-%There are two issues now
-%\begin{enumerate}
-% \item I am now unsure how $g_\sigma^{(m)}$ is precisely defined. This should be spelled out clearer in the beginning of Section 3.2.
-% \item There could be a mismatch between the $g_\sigma^{(m)}$ of Lemma 3.1 and the one we are actually using.
-This mismatch can most likely be resolved, because a result of the form of Lemma 3.1 also holds for the Chebyshev interpolation.
-If it is easy to resolve, please do so.
-If not, at least point this out here.
-Have a look at Trefethen's book; it might have the right error bound.
-%\end{enumerate}
-%???
 It is well known that the coefficients of the Chebyshev approximation \refequ{equ:matrix-approximation} can --- in exact arithmetic --- be computed by inverting the (type I) discrete cosine transform (DCT)
@@
 between the approximation coefficients $\mu_l(t),~l=0,\dots,m$ and the function evaluations at the Chebyshev points $s_i = \cos(\pi i / m),~i=0,\dots,m$ \cite{baszenski-1997-fast-polynomial, plonka-2018-numerical-fourier, trefethen-2020-approximation-theory}.
-This can be accomplished in $\mathcal{O}(m \log(m))$ operations.%}
+This can be accomplished in $\mathcal{O}(m \log(m))$ operations.
 
@@
 
-%\textcolor{red}{??? To be modified after the notion of Chebyshev approximation is clarified ???
 In \cite{lin-2017-randomized-estimation}, a slightly different approach to approximating $g_{\sigma}$ is taken: Instead of Chebyshev interpolation,
@@
     \frac{1}{n} \Trace \big((\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega}}_{=: \mtx{K}_1(t)})^{\dagger} (\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t)^2 \mtx{\Omega}}_{=: \mtx{K}_2(t)}) \big).
-    %\frac{1}{n} \Trace((\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega})^{\dagger} (\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})^2 \mtx{\Omega})),
     \label{equ:cyclic-property}
@@
     \For {$i = 1, \dots, n_t$}
-        \State Compute coefficients $\{\xi_l(t_i)\}_{l=0}^{m/2}$ of Chebyshev approx.~\refequ{equ:square-root-chebyshev-approximation} to $\sqrt{g_{\sigma}}$ with DCT
-        \State Compute coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ of non-negative Chebyshev approx. to $g_{\sigma}$ by applying \refalg{alg:chebyshev-squaring} to $\{\xi_l(t_i)\}_{l=0}^{m/2}$
+        \State Compute coefficients $\{\xi_l(t_i)\}_{l=0}^{\sfrac{m}{2}}$ of Chebyshev approx.~\refequ{equ:square-root-chebyshev-approximation} to $\sqrt{g_{\sigma}}$ with DCT
+        \State Compute coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ of non-negative Chebyshev approx.
+to $g_{\sigma}$ by applying \refalg{alg:chebyshev-squaring} to $\{\xi_l(t_i)\}_{l=0}^{\sfrac{m}{2}}$
         \State Compute coefficients $\{\underline{\nu}_l(t_i)\}_{l=0}^{2m}$ of squared non-negative Chebyshev approx.
@@
     \EndFor
-    %\State Compute the coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ of the non-negative Chebyshev approximation \refequ{equ:non-negative-chebyshev-approximation} for all points $t_i$ \textcolor{red}{maybe more details needed (maybe compute $\xi$ for m/2, then twice Alg 3.3)}
-    %\State Compute the coefficients $\{\nu_l(t_i)\}_{l=0}^{2m}$ of the square of the non-negative Chebyshev approximation with coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ using \refalg{alg:chebyshev-squaring}%\refalg{alg:2-chebyshev-chebyshev-approximation}
-    %\State Compute  for all $t_i$ using %\refalg{alg:3-nystrom-chebyshev-exponentiation}
-    \State Generate Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$, $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$%\glsfirst{sketching-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$
-    %\State Generate standard Gaussian %\glsfirst{random-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$
+    \State Generate Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$, $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$
     \State Initialize $[\mtx{V}_1, \mtx{V}_2, \mtx{V}_3] \gets [\mtx{0}_{n \times n_{\mtx{\Omega}}}, \mtx{\mtx{\Omega}}, \mtx{0}_{n \times n_{\mtx{\Omega}}}]$
@@
     \State Initialize $[\mtx{L}_1(t_i), \ell(t_i)] \gets [\mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Psi}}}, 0]$ for all $t_i$
-    %\State Set $\breve{\phi}_{\sigma}^{(m)}(t_i) \gets 0$ for all $t_i$
     \For {$l = 0, \dots, 2m$}
     \State $[\mtx{X}, \mtx{Y}] \gets \mtx{\mtx{\Omega}}^{\top} [\mtx{V}_2, \mtx{W}_2]$  
-    %\State $\mtx{X} \gets \mtx{\mtx{\Omega}}^{\top} \mtx{V}_2$
-      %\State $\mtx{Y} \gets \mtx{\mtx{\Omega}}^{\top} \mtx{W}_2$
       \State $z \gets \Trace(\mtx{\Psi}^{\top} \mtx{W}_2)$
@@
       \State $[\mtx{V}_2, \mtx{W}_2] \gets [\mtx{V}_3, \mtx{W}_3]$
-      %\State $\mtx{V}_3 \gets (2 - \delta_{l0}) \mtx{A} \mtx{V}_2 - \mtx{V}_1$ %\Comment{Chebyshev recurrence \refequ{equ:2-chebyshev-chebyshev-recursion}}
-      %\State $\mtx{V}_1 \gets \mtx{V}_2, \mtx{V}_2 \gets \mtx{V}_3$
-      %\State $\mtx{W}_3 \gets (2 - \delta_{l0}) \mtx{A} \mtx{W}_2 - \mtx{W}_1$ %\Comment{Chebyshev recurrence \refequ{equ:2-chebyshev-chebyshev-recursion}}
-      %\State $\mtx{W}_1 \gets \mtx{W}_2, \mtx{W}_2 \gets \mtx{W}_3$
     \EndFor
@@
 In this case, \refalg{alg:nystrom-chebyshev-pp} essentially amounts to the \enquote{spectrum sweeping} method from \cite[Algorithm 5]{lin-2017-randomized-estimation}, featuring a significantly lower computational complexity.
-
-%Chebyshev $\mathcal{O}(n_t (m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + n_{\mtx{\Omega}}^3) + mn^2(n_{\mtx{\Psi}} + n_{\mtx{\Omega}}))$.
-
-%\todo{Complexity in terms of both $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$.}
-%With the cost of a matrix-vector product denoted by $c(n)$, and supposing $n_{\mtx{\Omega}} \approx n_{\mtx{\Psi}}$, we determine the computational complexity of the Chebyshev-Nyström++ method to be $\mathcal{O}(m \log(m) n_t + m n_{\mtx{\Omega}}^2 n + m n_t n_{\mtx{\Omega}}^2 +  m c(n) n_{\mtx{\Omega}} + n_t n_{\mtx{\Omega}}^3)$, with $\mathcal{O}(m n_t + n n_{\mtx{\Omega}} + n_{\mtx{\Omega}}^2 n_t)$ required additional storage.
-
-%\todo{Explain modifications for other estimators, and improved complexity for quadratic trace estimator}
 
@@
         \label{equ:chebyshev-nystrom-error}
-        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
-        %\label{equ:2-chebyshev-interpolation-error}
     \end{equation*}
@@
     the result follows by setting $\varepsilon = 3 \widetilde{\varepsilon}$.
-%    if $\widetilde{\varepsilon} \leq 1$,
-%    \begin{equation}
-%        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq (1+\widetilde{\varepsilon}) \widetilde{\varepsilon} + \widetilde{\varepsilon} \leq 3 \widetilde{\varepsilon}
-%    \end{equation}
-%    from which the result follows by identifying $\varepsilon = 3 \widetilde{\varepsilon}$.
 \end{proof}
-
-%\todo{Maybe specialize \refthm{thm:nystrom} to spectral densities?}
-
-%We now discuss what value to choose for $n_{\mtx{\Omega}}$ in the Nyström trace estimator introduced in \refsec{subsec:nystrom}.
-In order to guarantee an even better approximation error based on \refthm{thm:nystrom}, we analyze the numerical rank of the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
 
@@
 In particular, if we assume the eigenvalues of $\mtx{A}$ to be roughly uniformly distributed, a small calculation shows that $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma \sqrt{\log(\varepsilon^{-1}) + \log(\sigma^{-1})})$ will ensure this.
-
-%\textcolor{red}{Make briefer:
-%\begin{itemize}
-% \item Quick decay of eigenvalues makes Nyström alone look good.
-% \item State singular values.
-% \item Explain as quickly as possible how you arrive at estimate of numerical rank for a roughly uniform distribution.
-w.l.o.g.
-eigenvalue $t$ around zero $\leadsto$ roughly distribution count number of eigenvalues $\approx \pm i/n$ needed for numerical rank $\epsilon$.
-% \item try to reduce notation
-%\end{itemize}
-%}
-%In particular, for small $\sigma$ the singular values of the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ decay so quickly that the Nyström approximation alone is highly accurate (cf.
-\refthm{thm:nystrom}).
-To establish a relation between the singular value decay of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ and what choice of $n_{\mtx{\Omega}}$ is needed for a significantly better approximation than ensured by \refthm{thm:chebyshev-nystrom}, we study the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$.\footnote{The factor $1/n$
-%does not change the conclusion, but is added to simplify the calculations and also reflects how we implement the method: We absorb the factor $1/n$ from the smoothed spectral density \refequ{equ:smooth-spectral-density} in the definition of $g_{\sigma}$, such that in reality, the Chebyshev approximation and the Nyström++ estimator are applied to $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$.}
-
-%We can express the numerical rank of a matrix with respect to the nuclear norm in terms of its singular values $\sigma_1 \geq \dots \geq \sigma_n \geq 0$ as
-%\begin{equation}
-%    r_{\varepsilon_m} = \min \{1 \leq r \leq n: \sum_{j=r+1}^n \sigma_{j} \leq \varepsilon_m \},
-%    \label{equ:numerical-rank}
-%\end{equation}
-%where $\varepsilon_m$ is usually taken to be the double machine precision, i.e. $\varepsilon_m \approx 10^{-16}$.
-
-%\begin{theorem}{Nyström estimator for parameter-dependent matrices}%{chebyshev-nystrom-raw}
-%    Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric %matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is %contained in $[-1, 1]$.
-Then the Chebyshev-Nyström++ estimate %$\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m \in %\mathbb{N}$, $n_{\mtx{\Omega}} \geq \max(6, \max_{t \in [-1, 1]}(r_%{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A}))) + \log(\delta^%{-1}))$, $n_{\mtx{\Psi}} = 0$, and smoothing parameter $\sigma > 0$ %satisfies with for all $\gamma \geq 1$ with probability at least $1 %- \delta$
-%    \begin{equation}
-%        \int_{a}^{b} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}%(t) \right| ~\mathrm{d}t
-%        \leq 2 e^2 (1 + r_{\varepsilon}) (\varepsilon + (n - r_%{\varepsilon}) \underline{E}_{\sigma, m}) + \underline{E}_%{\sigma, m}.
-%    \end{equation}
-%\end{theorem}
-%
-%\begin{proof}
-%    \todo{TODO}
-%\end{proof}
-
-%When using a Gaussian smoothing-kernel, we can explicitly write the singular values of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$ as
-%\begin{equation}
-%    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) / n = \frac{1}{n \sqrt{2 \pi \sigma^2}} e^{-\sfrac{\left(t - \lambda_{(i)}\right)^2}{2 \sigma^2}},
-%    \label{equ:gaussian-kernel-eigenvalues}
-%\end{equation}
-%where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from spectral-parameter $t$, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t)$.
-Consequently, by using the closed-form expression of the eigenvalues \refequ{equ:gaussian-kernel-eigenvalues}, we may generously upper bound the numerical rank by
-%\begin{equation}
-%    r_{\varepsilon_m}(g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n) \leq \#\left\{1\leq i\leq n: |t - \lambda_i| < d_{\varepsilon_m, \sigma} \right\},
-%    \label{equ:gaussian-kernel-numerical-rank}
-%\end{equation}
-%with $d_{\varepsilon_m, \sigma} = \sigma \sqrt{-2 \log(\sqrt{2 \pi} \sigma \varepsilon_m)}$ and the eigenvalues $\lambda_1, \dots, \lambda_n$ of $\mtx{A}$ and the constants.
-The expression \refequ{equ:gaussian-kernel-numerical-rank} has a nice visual interpretation: The numerical rank of a matrix is at most equal to the number of eigenvalues which are closer to the parameter $t$ than $d_{\varepsilon_m, \sigma}$.
-This is illustrated in \reffig{fig:numerical-rank}.
-%\begin{figure}[ht]
-%    \centering
-%    \input{figures/numerical-rank.tex}
-%    \caption{The numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$ can be approximately computed by counting the number of eigenvalues $\lambda_{(1)}, \dots, \lambda_{(n)}$ of the matrix $\mtx{A}$ which lie less than a constant $d_{\varepsilon_m, \sigma}$ away from $t$.}
-%    \label{fig:numerical-rank}
-%\end{figure}
-
-%If we additionally assume the eigenvalues of the matrix $\mtx{A}$ to be evenly distributed within $[-1, 1]$, that is, in any subinterval of fixed length in $[a, b]$, $-1 \leq a < b \leq 1$, we can expect to find roughly the same number of eigenvalues, then we can estimate the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$ to be
-%\begin{equation}
-%    r_{\varepsilon_m}(g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n) \lessapprox n d_{\varepsilon_m, \sigma} \approx 6 n \sigma,
-%    \label{equ:gaussian-kernel-numerical-rank-uniform}
-%\end{equation}
-%provided we set $\varepsilon_m = 10^{-16}$ and choose $\sigma$ roughly between $10^{-8}$ and $1$.
-Hence, choosing $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma)$ will yield approximations which can be significantly better than \refthm{thm:chebyshev-nystrom}.

### 2025-02-25 paper/analysis.tex
run spell check on paper

@@
 At this point, we acknowledge the existence of the XNysTrace estimator from \cite{epperly-2024-xtrace-making} which oftentimes outperforms the Nyström++ estimator.
-Moreover, a straight-forward extension of the XNysTrace estimator to the parameter-dependent setting seems to be in reach.
+Moreover, a straightforward extension of the XNysTrace estimator to the parameter-dependent setting seems to be in reach.
 However, what distinguishes \refalg{alg:nystrom-chebyshev-pp} for estimating spectral densities are two key observations --- using the cyclic invariance of the trace and the affine linear form of Chebyshev approximations (see \refsec{subsubsec:chebyshev-nystrom-implementation} for details) --- which it exploits to significantly speed up the computation.
@@
 
-A consequence of~\cite{kressner-2024-randomized-lowrank}, the following result confirms the inituition that the Nystr\"om estimator is very well suited when the singular values of $\mtx{B}(t)$ decay quickly.
+As a consequence of~\cite{kressner-2024-randomized-lowrank}, the following result confirms the intuition that the Nystr\"om estimator is very well suited when the singular values of $\mtx{B}(t)$ decay quickly.
 More specifically, it shows that the $L^1$-error of the estimator stays on the level of the best rank-$r$ approximation error of $B(t)$, measured in the nuclear norm.
@@
 
-Compared to the constant case~\cite[Theorem 8.1]{tropp-2023-randomized-algorithms}, the bound of \refthm{thm:nystrom} features an additional factor $1+r$, which cannot be compensated with oversampling.
+Compared to the constant case~\cite[Theorem 8.1]{tropp-2023-randomized-algorithms}, the bound of \refthm{thm:nystrom} features an additional factor $1+r$, which cannot be compensated for with oversampling.
 

### 2025-02-25 paper/paper.tex
make notation consistent and clean up tex files
run spell check on paper

@@
 \usepackage{stylesheet}
-
-% to speed up compilation speed a bit while drafting
-\pdfcompresslevel=0
-\pdfobjcompresslevel=0
 
@@
 
-%\todo{
-%\begin{itemize}
-%    \item Unify index notation
-%    \item Double-check proofs
-%    \item Add all references
-%    \item Run numerical experiments 
-%    \item Write introduction, abstract, and discussion
-%    \item ArXiv HTML support (check if packages)
-%\end{itemize}
-%}
-
-%\todo{
-%    \tableofcontents
-%}
-
 \begin{abstract}
@@
     When combined with Chebyshev approximation in $t$, the use of such constant random matrices allows one to reuse matrix-vector products across different values of $t$, leading to significant cost reduction.
-    Our analysis shows that the loss of stochastic independence across different $t$ does not lead to deterioriation; in fact, our error bounds for these estimators closely match existing error bounds for the constant case.
+    Our analysis shows that the loss of stochastic independence across different $t$ does not lead to deterioration; in fact, our error bounds for these estimators closely match existing error bounds for the constant case.
 In particular, we show that $\mathcal{O}(\varepsilon^{-1})$ random matrix-vector products suffice to ensure an error of $\varepsilon > 0$ for Nyström++, independent of low-rank properties of $\mtx{B}(t)$.
@@
 
-%\textcolor{red}{TODO for Fabio:
-%\begin{itemize}
- %\item Has~\cite{bhattacharjee-2024-improved-spectral} been published?
- %\item Fix discussion end of Section 3.3.2
- %\item Please reread entire Section 4 and address various red points.
- %\item Please reread abstract.
- %\item Please reread last part of Section 2 (startin from new contributions) and fix red remark.
- %\item Please remove all unreferenced equation labels (using checkref) at the end.
- %\item Perform a spell check at the end.
-%\end{itemize}
-%}
 \input{intro.tex}

### 2025-02-25 paper/spectraldensity.tex
run spell check on paper

@@
 the kernel $g_{\sigma}$ is approximated by a Chebyshev series~\cite[Chapter 4]{trefethen-2020-approximation-theory} truncated after term $m$.
-In constrast to interpolation, the expansion coefficents involve integrals, which need to be approximated by quadrature.
+In contrast to interpolation, the expansion coefficients involve integrals, which need to be approximated by quadrature.
 By choosing $2m+1$ Chebyshev quadrature nodes, this approximation can be carried using the fast Fourier transform (FFT).
@@
 \end{align}
-Each of the annotated terms can be evaluated efficiently for many values of $t$, by precomputing quantities using the Chebyshev reccurrence 
+Each of the annotated terms can be evaluated efficiently for many values of $t$, by precomputing quantities using the Chebyshev recurrence 
 $\mtx{B}(t) = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$, as explained above.
@@
 consider an associated eigenvalue problem and threshold the smallest eigenvalues~\cite{lin-2017-randomized-estimation, epperly-2022-theory-quantum}.
-However, this renders the algorithm significantly more complicated, and we found that it improves accuracy only slightly.
-Instead, we use the standard NumPy least-squares solver when applying $\mtx{K}_1(t_i)^{\dagger}$, which is based on the DGELSD routine from LAPACK, and likewise truncates the smallest eigenvalues by treating them as zero if they are smaller than a certain threshold.
+However, this makes the algorithm significantly more complicated, and we found that it improves accuracy only slightly.
+Instead, we use the standard NumPy least-squares solver when applying $\mtx{K}_1(t_i)^{\dagger}$, which is based on the LAPACK DGELSD routine, and likewise truncates the smallest eigenvalues by treating them as zero if they are smaller than a certain threshold.
 We have found a threshold of $10^{-5}$ times the largest eigenvalue to give the best results in our numerical experiments.

### 2025-02-25 paper/analysis.tex
clean up repository before publication

@@
 The Girard-Hutchinson estimator is a popular choice for spectral density estimation; it has been used in the Haydock method \cite{haydock-1972-electronic-structure}, as well as the Delta-Gauss-Legendre and Delta-Gauss-Chebyshev methods from~\cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation}.
-However, it shares the disadvantage of Monte Carlo methods: Its error is proportional to $n^{-1/2}_{\mtx{\Psi}}$, making it expensive in applications that require decent accuracy.
+However, it shares the disadvantage of Monte Carlo methods: Its error is proportional to $n^{-\sfrac{1}{2}}_{\mtx{\Psi}}$, making it expensive in applications that require decent accuracy.
 

### 2025-02-25 paper/plots/krylov_aware_spin.py

@@ -11,10 +11,6 @@ from matrices.quantum_spin import hamiltonian, partition_function
 
 np.random.seed(0)
 
-# Don't execute this because memory overflow on GitHub due to private repository
-if __name__ == "__main__":
-    quit()
-
 N = 20
 s = 0.5
 h = 0.3

### 2025-02-25 reproduce.py

@@ -6,16 +6,13 @@ parser = argparse.ArgumentParser(
                     description="Usage of the automated setup",
                     epilog="Text at the bottom of help")
 
-parser.add_argument("-a", "--generate-all", action="store_true", help="reproduce the whole project (matrices, results, paper)")
-parser.add_argument("-m", "--generate-matrices", action="store_true", help="download and create all matrices")
-parser.add_argument("-r", "--generate-results", action="store_true", help="reproduce all results (plots/tables)")
-parser.add_argument("-p", "--generate-paper", action="store_true", help="compile the paper (LaTeX)")
+parser.add_argument("-a", "--reproduce-all", action="store_true", help="reproduce the whole project (matrices, results, paper)")
+parser.add_argument("-r", "--reproduce-results", action="store_true", help="reproduce all results (plots/tables)")
+parser.add_argument("-p", "--reproduce-paper", action="store_true", help="compile the paper (LaTeX)")
 
 args = parser.parse_args()
 
-if args.generate_matrices or args.generate_all:
-    os.system("python setup/generate_matrices.py")
-if args.generate_results or args.generate_all:
-    os.system("python setup/generate_results.py")
-if args.generate_paper or args.generate_all:
-    os.system("python setup/generate_paper.py")
+if args.reproduce_results or args.reproduce_all:
+    os.system("python setup/reproduce_results.py")
+if args.reproduce_paper or args.reproduce_all:
+    os.system("python setup/reproduce_paper.py")

### 2025-02-25 reproduce/generate_matrices.py

@@ -1,44 +1,0 @@
-import __context__
-
-import os
-
-import scipy as sp
-import numpy as np
-
-from src.utils import download_matrix
-from src.matrices import hamiltonian, uniform, gaussian_orthogonal_ensemble
-
-matrix_dir = "matrices"
-
-if not os.path.exists(matrix_dir):
-    os.makedirs(matrix_dir)
-
-matrix_urls = [
-    #"https://suitesparse-collection-website.herokuapp.com/MM/ND/nd3k.tar.gz",
-    #"https://suitesparse-collection-website.herokuapp.com/MM/MaxPlanck/shallow_water1.tar.gz",
-    "https://suitesparse-collection-website.herokuapp.com/MM/Pajek/Erdos992.tar.gz",
-    #"https://suitesparse-collection-website.herokuapp.com/MM/Pajek/California.tar.gz",
-]
-
-print("\nDownloading matrices")
-for i, matrix_url in enumerate(matrix_urls):
-    download_matrix(matrix_url, save_path=matrix_dir)
-    print(u"\u2713 Downloaded matrix ({}/{})".format(i + 1, len(matrix_urls)))
-
-ns = [1, 2]#, 3, 4]
-
-print("\nGenerating ModES3D matrices")
-for i, n in enumerate(ns):
-    matrix = hamiltonian(n=n, L=6, h=0.6, dim=3, bc="periodic", beta=2.0, alpha=-4.0)
-    sp.sparse.save_npz(os.path.join(matrix_dir, "ModES3D_{}".format(n**3)), matrix)
-    print("\u2713 Generated matrix ({}/{})".format(i + 1, len(ns)))
-
-print("\nGenerating uniform matrix")
-matrix = uniform(2000, density=0.00015)
-sp.sparse.save_npz(os.path.join(matrix_dir, "uniform"), matrix)
-print("\u2713 Generated matrix")
-
-print("\nGenerating Gaussian orthogonal ensemble")
-matrix = gaussian_orthogonal_ensemble(1000)
-np.savez(os.path.join(matrix_dir, "goe"), matrix)
-print("\u2713 Generated matrix")

### 2025-02-25 paper/re-pro-badge.tex
automatically reproduce paper

@@
-The code for our numerical experiments can be found on \url{https://github.com/FMatti/Rand-TRACE}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/Rand-TRACE/tree/37ec0b7}{37ec0b7} on 2025-02-21 at 17:02:20 UTC.
+The code for our numerical experiments can be found on \url{https://github.com/FMatti/parameter-trace}.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/parameter-trace/tree/0f31e38}{0f31e38} on 2025-02-25 at 17:05:59 UTC.

### 2025-02-25 paper/paper.tex
shorten abstract

@@
 \begin{abstract}
-    Stochastic trace estimation is a well-established tool for approximating the trace of a large symmetric positive semi-definite matrix $\mtx{B}$.
-Several applications involve a matrix that depends continuously on a parameter $t \in [a,b]$, and require trace estimates of $\mtx{B}(t)$ for many different values of $t$.
-This is, for example, the case when approximating the spectral density of a matrix or computing partition functions of quantum systems.
+    Stochastic trace estimation is a well-established tool for approximating the trace of a large symmetric matrix $\mtx{B}$.
+Several applications involve a matrix that depends continuously on a parameter $t \in [a,b]$, and require trace estimates of $\mtx{B}(t)$ for many values of $t$.
+This is, for example, the case when approximating the spectral density of a matrix.
 Approximating the trace separately for each matrix
@@
     When combined with Chebyshev approximation in $t$, the use of such constant random matrices allows one to reuse matrix-vector products across different values of $t$, leading to significant cost reduction.
-    Our analysis shows that the loss of stochastic independence across different $t$ does not lead to deterioration; in fact, our error bounds for these estimators closely match existing error bounds for the constant case.
+    Our analysis shows that the loss of stochastic independence across different $t$ does not lead to deterioration.
 In particular, we show that $\mathcal{O}(\varepsilon^{-1})$ random matrix-vector products suffice to ensure an error of $\varepsilon > 0$ for Nyström++, independent of low-rank properties of $\mtx{B}(t)$.
 We discuss in detail how the combination of Nyström++ with
-    Chebyshev approximation applies to large-scale spectral density estimation and provide an analysis of the resulting method.
+    Chebyshev approximation applies to spectral density estimation and provide an analysis of the resulting method.
 This  improves various aspects of an existing stochastic estimator for spectral density estimation.
-Several numerical experiments for examples from electronic structure interaction, statistical thermodynamics, and neural network optimization illustrate and validate our findings.
+Several numerical experiments from electronic structure interaction, statistical thermodynamics, and neural network optimization validate our findings.
 \end{abstract}

### 2025-02-25 paper/re-pro-badge.tex
automatically reproduce paper

@@
 The code for our numerical experiments can be found on \url{https://github.com/FMatti/parameter-trace}.
-The figures and tables were generated from the commit \href{https://github.com/FMatti/parameter-trace/tree/0f31e38}{0f31e38} on 2025-02-25 at 17:05:59 UTC.
+The figures and tables were generated from the commit \href{https://github.com/FMatti/parameter-trace/tree/334d5c6}{334d5c6} on 2025-02-25 at 19:12:59 UTC.

### 2025-06-27 paper/analysis.tex
switch to SIAM template

@@
 When $\mtx{B}$ is variable, important differences arise from the fact that the random vectors do not depend on $t$.
-On the one hand, as we will see in~\refsec{sec:application}, this allows one to reuse computations.
+On the one hand, as we will see in~\cref{sec:application}, this allows one to reuse computations.
 On the other hand, this significantly complicates the analysis, which will be the main focus of this section.
@@
 
-\paragraph{Girard-Hutchinson estimator.} One of the simplest methods for trace estimation, the Girard-Hutchinson estimator~\cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator} proceeds by sampling $n_{\mtx{\Psi}}$ independent Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$
+\paragraph{Girard-Hutchinson estimator} One of the simplest methods for trace estimation, the Girard-Hutchinson estimator~\cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator} proceeds by sampling $n_{\mtx{\Psi}}$ independent Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$
 and computes the approximation 
@@
 
-\paragraph{Nyström estimator.} Given a good low-rank approximation of a matrix, the trace can be cheaply estimated by simply taking the trace of that low-rank approximation.
+\paragraph{Nyström estimator} Given a good low-rank approximation of a matrix, the trace can be cheaply estimated by simply taking the trace of that low-rank approximation.
 In the context of spectral density estimation, the use of a non-negative smoothing kernel implies that $\mtx{B}(t)$ is symmetric positive semi-definite (SPSD) for every $t$.
@@
 Related randomized low-rank approximations for general parameter-dependent matrices have been analyzed in~\cite{kressner-2024-randomized-lowrank}.
-Exploiting the invariance of the trace under cyclic permutations, taking the trace of~\refequ{equ:nystrom-approximation} gives rise to the Nyström estimator
+Exploiting the invariance of the trace under cyclic permutations, taking the trace of~\cref{equ:nystrom-approximation} gives rise to the Nyström estimator
 \begin{equation}
@@
 
-\paragraph{Nyström++ estimator.}
-The Nyström estimator~\refequ{equ:nystrom-trace-estimator} relies on the feasibility of low-rank approximation, that is, the singular values of $\mtx{B}(t)$ need to decay sufficiently quickly.
+\paragraph{Nyström++ estimator}
+The Nyström estimator~\cref{equ:nystrom-trace-estimator} relies on the feasibility of low-rank approximation, that is, the singular values of $\mtx{B}(t)$ need to decay sufficiently quickly.
 If this property cannot be ensured, it is preferable to use an estimator that combines the Girard-Hutchinson and Nyström estimators.
-The combination suggested in~\cite{lin-2017-randomized-estimation} applies Girard-Hutchinson to correct the error of the Nyström approximation~\refequ{equ:nystrom-approximation}:
+The combination suggested in~\cite{lin-2017-randomized-estimation} applies Girard-Hutchinson to correct the error of the Nyström approximation~\cref{equ:nystrom-approximation}:
 \begin{equation}
@@
 
-For constant $\mtx{B}$, the estimator~\refequ{equ:nystrompp-trace-estimator} coincides with the Nyström++ estimator from~\cite{persson-2022-improved-variants}, which is based on the Hutch++ estimator~\cite{meyer-2021-hutch-optimal}.
+For constant $\mtx{B}$, the estimator~\cref{equ:nystrompp-trace-estimator} coincides with the Nyström++ estimator from~\cite{persson-2022-improved-variants}, which is based on the Hutch++ estimator~\cite{meyer-2021-hutch-optimal}.
 In this situation (constant $\mtx{B}$), these estimators were both shown to achieve a relative $\varepsilon$-error with $\mathcal{O}(\varepsilon^{-1})$ matrix-vector products only, independent of the singular value decay of $\mtx{B}$.
@@
 Moreover, a straightforward extension of the XNysTrace estimator to the parameter-dependent setting seems to be in reach.
-However, what distinguishes \refalg{alg:nystrom-chebyshev-pp} for estimating spectral densities are two key observations --- using the cyclic invariance of the trace and the affine linear form of Chebyshev approximations (see \refsec{subsubsec:chebyshev-nystrom-implementation} for details) --- which it exploits to significantly speed up the computation.
+However, what distinguishes \cref{alg:nystrom-chebyshev-pp} for estimating spectral densities are two key observations --- using the cyclic invariance of the trace and the affine linear form of Chebyshev approximations (see \cref{subsubsec:chebyshev-nystrom-implementation} for details) --- which it exploits to significantly speed up the computation.
 Unfortunately, we do not see a way of marrying these observations with the efficient implementation of XNysTrace described in \cite[Section 2.2]{epperly-2024-xtrace-making}, which limits its suitability for spectral density estimation.
@@
 
-In this section, we derive bounds on the $L^1$-error (with respect to $t$) for each of the trace estimators from \refsec{subsec:methods}.
+In this section, we derive bounds on the $L^1$-error (with respect to $t$) for each of the trace estimators from \cref{subsec:methods}.
 Our bounds parallel existing results in the constant case.
@@
 
-We first derive an $L^1$-error bound for the Girard-Hutchinson estimator~\refequ{equ:hutchinson-trace-estimator} that is relative to the $L^1$-norms of the Frobenius and spectral norms of $\mtx{B}(t)$, defined by
+We first derive an $L^1$-error bound for the Girard-Hutchinson estimator~\cref{equ:hutchinson-trace-estimator} that is relative to the $L^1$-norms of the Frobenius and spectral norms of $\mtx{B}(t)$, defined by
 \begin{equation*}
@@
 
-\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
-For a symmetric matrix $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ that depends continuously on $t \in [a, b]$, consider the  Girard-Hutchinson estimator~\refequ{equ:hutchinson-trace-estimator} with an $n\times n_{\mtx{\Psi}}$ Gaussian random matrix $\mtx{\Psi}$.
+\begin{theorem}[Girard-Hutchinson estimator for parameter-dependent matrices]\label{thm:hutchinson}
+For a symmetric matrix $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ that depends continuously on $t \in [a, b]$, consider the  Girard-Hutchinson estimator~\cref{equ:hutchinson-trace-estimator} with an $n\times n_{\mtx{\Psi}}$ Gaussian random matrix $\mtx{\Psi}$.
 Then for any $p \geq 1$, $p \in \mathbb{R}$ and $\gamma \geq 1$, the following bound holds with probability at least $1 - \gamma^{-p}$:
@@
 From the proof of \cite[Theorem 1]{cortinovis-2022-randomized-trace}, we know that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is a $(2 \lVert \mtx{B}(t) \rVert _F^2 / n_{\mtx{\Psi}}, 2 \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}})$-sub-gamma random variable.
-Therefore, \reflem{lem:sub-gamma-moments} yields
+Therefore, \cref{lem:sub-gamma-moments} yields
     \begin{equation}
@@
     To address the parameter-dependent case, we first note that the continuity assumption implies that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is measurable.
-Therefore, we can apply Minkowski's integral inequality~\cite[Theorem 202]{hardy-1952-inequalities} to conclude from the moment bound~\refequ{equ:moment-bound-hutchinson} that
+Therefore, we can apply Minkowski's integral inequality~\cite[Theorem 202]{hardy-1952-inequalities} to conclude from the moment bound~\cref{equ:moment-bound-hutchinson} that
     \begin{align*}
@@
     \end{align*}
-    This implies~\refequ{equ:girard-hutchinson-bound} by Markov's inequality.
+    This implies~\cref{equ:girard-hutchinson-bound} by Markov's inequality.
 
@@
 
-\refthm{thm:hutchinson} parallels an analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, except for a slightly more restricted choice of the failure probability $\delta$.
+\cref{thm:hutchinson} parallels an analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, except for a slightly more restricted choice of the failure probability $\delta$.
 
 \begin{remark}
-    A variant of \refthm{thm:hutchinson} can be derived when assuming that $\mtx{B}(t)$ is SPSD and nonzero for every $t \in [a, b]$.
-After dividing both sides of \refequ{equ:moment-bound-hutchinson} by $\Trace(\mtx{B}(t))$, one can apply the remaining steps of the proof to deduce that
+    A variant of \cref{thm:hutchinson} can be derived when assuming that $\mtx{B}(t)$ is SPSD and nonzero for every $t \in [a, b]$.
+After dividing both sides of \cref{equ:moment-bound-hutchinson} by $\Trace(\mtx{B}(t))$, one can apply the remaining steps of the proof to deduce that
     \begin{equation*}
@@
 
-\begin{theorem}{Nyström estimator for parameter-dependent matrices}{nystrom}
+\begin{theorem}[Nyström estimator for parameter-dependent matrices]\label{thm:nystrom}
     Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be SPSD and continuous for $t \in [a, b]$.
-For integers $r \geq 2$ and $n_{\mtx{\Omega}} \geq r + 4$, consider the Nystr\"om approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ defined in~\refequ{equ:nystrom-approximation} with an $n\times n_{\mtx{\Omega}}$ Gaussian random matrix $\mtx{\Omega}$.
+For integers $r \geq 2$ and $n_{\mtx{\Omega}} \geq r + 4$, consider the Nystr\"om approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ defined in~\cref{equ:nystrom-approximation} with an $n\times n_{\mtx{\Omega}}$ Gaussian random matrix $\mtx{\Omega}$.
 Then, for any $\gamma \geq 1$, the bound
@@
 
-Compared to the constant case~\cite[Theorem 8.1]{tropp-2023-randomized-algorithms}, the bound of \refthm{thm:nystrom} features an additional factor $1+r$, which cannot be compensated for with oversampling.
+Compared to the constant case~\cite[Theorem 8.1]{tropp-2023-randomized-algorithms}, the bound of \cref{thm:nystrom} features an additional factor $1+r$, which cannot be compensated for with oversampling.
 
@@
 
-\refthm{thm:hutchinson} shows that the Girard-Hutchinson estimator achieves a relative $\varepsilon$-error when using $\mathcal{O}(\varepsilon^{-2})$ queries.
-The aim of this section is to show that the Nyström++ estimator~\refequ{equ:nystrompp-trace-estimator} improves this to $\mathcal{O}(\varepsilon^{-1})$ queries (that is, matrix-vector products with random vectors), \emph{without} requiring any assumption on the singular values of $\mtx{B}(t)$.
+\cref{thm:hutchinson} shows that the Girard-Hutchinson estimator achieves a relative $\varepsilon$-error when using $\mathcal{O}(\varepsilon^{-2})$ queries.
+The aim of this section is to show that the Nyström++ estimator~\cref{equ:nystrompp-trace-estimator} improves this to $\mathcal{O}(\varepsilon^{-1})$ queries (that is, matrix-vector products with random vectors), \emph{without} requiring any assumption on the singular values of $\mtx{B}(t)$.
 This parallels existing results for the constant case~\cite{meyer-2021-hutch-optimal,persson-2022-improved-variants}.
-In particular, \cite[Theorem 3.4]{persson-2022-improved-variants} shows that $|\Trace(\mtx{B}) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})| \leq \varepsilon \Trace(\mtx{B})$ holds with high probability for (constant) SPSD $\mtx{B}$ when $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1})$ in the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})$ from~\refequ{equ:nystrompp-trace-estimator}.
+In particular, \cite[Theorem 3.4]{persson-2022-improved-variants} shows that $|\Trace(\mtx{B}) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})| \leq \varepsilon \Trace(\mtx{B})$ holds with high probability for (constant) SPSD $\mtx{B}$ when $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1})$ in the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})$ from~\cref{equ:nystrompp-trace-estimator}.
 
@@
 
-\begin{lemma}{Nyström approximation for parameter-dependent matrices}{nystrom}
+\begin{lemma}[Nyström approximation for parameter-dependent matrices]\label{lem:nystrom}
     Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be SPSD and continuous for $t \in [a, b]$.
@@
 
-To treat the second term in~\refequ{equ:nystrom-proof-persson-bonud}, we bound 
+To treat the second term in~\cref{equ:nystrom-proof-persson-bonud}, we bound 
 its $q$th moment using the independence of $\mtx{\Omega}_1$ and $\mtx{\Omega}_2$:
@@
 
-    The second factor in~\refequ{equ:nystrom-proof-processed-tail} is bounded using \reflem{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$:
+    The second factor in~\cref{equ:nystrom-proof-processed-tail} is bounded using \cref{lem:spectral-norm-moment} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$:
     \begin{equation*}
@@
     \end{equation*}
-    Inserting this inequality and~\refequ{equ:pinv-spectral-norm-bound} into~\refequ{equ:nystrom-proof-processed-tail} gives
+    Inserting this inequality and~\cref{equ:pinv-spectral-norm-bound} into~\cref{equ:nystrom-proof-processed-tail} gives
     \begin{equation*}
@@
 
-    Inserting~\refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud}, letting $c=154$, and using the triangle inequality for $\mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}[\cdot]$, we obtain
+    Inserting~\cref{equ:nystrom-proof-tail-bound} along with \cref{equ:nystrom-proof-frobenius-trace} in \cref{equ:nystrom-proof-persson-bonud}, letting $c=154$, and using the triangle inequality for $\mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}[\cdot]$, we obtain
     \begin{equation}
@@
     As in \cite{kressner-2024-randomized-lowrank}, one can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable.
-    As in the proof of~\refthm{thm:hutchinson}, the claimed bound~\refequ{equ:nystrompp-theorem-bound}
-    follows from~\refequ{equ:frobenius-moment-bound} using Minkowski's integral inequality and Markov's inequality.
+    As in the proof of~\cref{thm:hutchinson}, the claimed bound~\cref{equ:nystrompp-theorem-bound}
+    follows from~\cref{equ:frobenius-moment-bound} using Minkowski's integral inequality and Markov's inequality.
 
@@
 
-Combining the bound on the Girard-Hutchinson estimator from~\refthm{thm:hutchinson} with the result of~\reflem{lem:nystrom} finally gives the desired error bound for the Nyström++ trace estimator.
-\begin{theorem}{Nyström++ trace estimator for parameter-dependent matrices}{nystrom-pp}
+Combining the bound on the Girard-Hutchinson estimator from~\cref{thm:hutchinson} with the result of~\cref{lem:nystrom} finally gives the desired error bound for the Nyström++ trace estimator.
+\begin{theorem}[Nyström++ trace estimator for parameter-dependent matrices]\label{thm:nystrom-pp}
     Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be SPSD and continuous for $t \in [a, b]$. Choose $\delta \in (0, 2e^{-1}]$ and $\varepsilon \in (0, 1)$.
@@
 \begin{proof}
-    Choosing $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\tilde{\varepsilon}^{-2} \log(\tilde{\delta}^{-1}))$, \refthm{thm:hutchinson} and~\reflem{lem:nystrom} imply that
+    Choosing $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\tilde{\varepsilon}^{-2} \log(\tilde{\delta}^{-1}))$, \cref{thm:hutchinson} and~\cref{lem:nystrom} imply that
     \begin{align*}
@@
 
-Comparing \refthm{thm:nystrom-pp} with the analogous result for constant matrices \cite[Theorem 3.4]{persson-2022-improved-variants}, we notice an additional factor of $\sqrt{\log(\delta^{-1})}$ in the required number of random vectors $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$.
+Comparing \cref{thm:nystrom-pp} with the analogous result for constant matrices \cite[Theorem 3.4]{persson-2022-improved-variants}, we notice an additional factor of $\sqrt{\log(\delta^{-1})}$ in the required number of random vectors $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$.

### 2025-06-27 paper/intro.tex
switch to SIAM template

@@
 
-\paragraph{New contributions.} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices with constant randomness (with respect to $t$): The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}.
+\paragraph{New contributions} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices with constant randomness (with respect to $t$): The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}.
 Combined with Chebyshev approximation, this allows one to reuse the majority of computations across different values of the parameter $t$, making the estimators scale favorably with respect to the number of parameter evaluations.
-We propose several additional modifications to the methods from \cite{lin-2017-randomized-estimation} for the special case of approximating the smoothed spectral density $\phi_{\sigma}$ from~\refequ{equ:smoothed-spectral-density}.
+We propose several additional modifications to the methods from \cite{lin-2017-randomized-estimation} for the special case of approximating the smoothed spectral density $\phi_{\sigma}$ from~\cref{equ:smoothed-spectral-density}.
 In particular, we use a more rigorous approach that preserves non-negativity for approximating matrix functions in terms of Chebyshev polynomials.
@@
 unifies all methods from~\cite{lin-2017-randomized-estimation} in a single algorithm.
-Combining our analysis of the randomized trace estimators with standard error bounds for Chebyshev approximation leads to~\refthm{thm:chebyshev-nystrom}, which shows that $\mathcal O(\varepsilon^{-1})$ matrix-vector products suffice to guarantee an error $\varepsilon$.
+Combining our analysis of the randomized trace estimators with standard error bounds for Chebyshev approximation leads to~\cref{thm:chebyshev-nystrom}, which shows that $\mathcal O(\varepsilon^{-1})$ matrix-vector products suffice to guarantee an error $\varepsilon$.
 
-\paragraph{Structure.} In \refsec{sec:analysis} we will analyze the error of three standard trace estimators --- the Girard-Hutchinson, Nyström, and Nyström++ estimators --- when they are applied to parameter-dependent matrices with constant random matrix.
-In \refsec{sec:application}, we propose and analyze the Chebyshev-Nyström++ estimator, which uses these estimators to approximate the smoothed spectral density of real symmetric matrices.
-Finally, we will illustrate and validate our developments for examples from various applications in \refsec{sec:results}.
+\paragraph{Structure} In \cref{sec:analysis} we will analyze the error of three standard trace estimators --- the Girard-Hutchinson, Nyström, and Nyström++ estimators --- when they are applied to parameter-dependent matrices with constant random matrix.
+In \cref{sec:application}, we propose and analyze the Chebyshev-Nyström++ estimator, which uses these estimators to approximate the smoothed spectral density of real symmetric matrices.
+Finally, we will illustrate and validate our developments for examples from various applications in \cref{sec:results}.
 
-\paragraph{Reproducibility.} \input{re-pro-badge.tex}
+\paragraph{Reproducibility} \input{re-pro-badge.tex}

### 2025-06-27 paper/numresults.tex
run spell check on paper
switch to SIAM template

@@
 
-While theoretically an important tool, we have observed that the preservation of non-negativity discussed in~\refsec{sec:preservnonneg} is not needed in practice.
-The (slight) indefiniteness of the standard Chebyshev approximation $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ from~\refequ{equ:matrix-approximation} does not seem to impede the reliability of the Nyström++ estimator.
+While theoretically an important tool, we have observed that the preservation of non-negativity discussed in~\cref{sec:preservnonneg} is not needed in practice.
+The (slight) indefiniteness of the standard Chebyshev approximation $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ from~\cref{equ:matrix-approximation} does not seem to impede the reliability of the Nyström++ estimator.
 Unless otherwise stated, we transform the input matrices such that their eigenvalues are contained in $[-1, 1]$, by approximating their spectrum with NumPy's Hermitian eigenvalue solver before we apply the methods to them.
@@
 The integral involved in computing the $L^1$-errors is approximated with the composite midpoint quadrature rule using the $n_t$ values of $t$ as nodes.
-Both parameters introduced at the end \refsec{subsubsec:chebyshev-nystrom-implementation},the eigenvalue truncation threshold and the parameter for detecting a vanishing spectral density, are fixed to $10^{-5}$.
+Both parameters introduced at the end \cref{subsubsec:chebyshev-nystrom-implementation},the eigenvalue truncation threshold and the parameter for detecting a vanishing spectral density, are fixed to $10^{-5}$.
 
@@
     $v(r) = v_0 e^{-\lambda r^2}$, 
-with $v_0 = -4$ and $\lambda = 8$, centered in cells of side length $L=6$ which are stacked $n_c \in \mathbb{N}$ times in each spatial dimension; see \reffig{fig:gaussian-well}.
+with $v_0 = -4$ and $\lambda = 8$, centered in cells of side length $L=6$ which are stacked $n_c \in \mathbb{N}$ times in each spatial dimension; see \cref{fig:gaussian-well}.
 The grid width of the finite difference discretization is fixed to $h=0.6$.
@@
     \begin{subfigure}[b]{0.32\columnwidth}
-        \input{plots/gaussian-well-1.pgf}
+        \scalebox{0.7}{\input{plots/gaussian-well-1.pgf}}
         \caption{$n_c=1$}
@@
     \begin{subfigure}[b]{0.32\columnwidth}
-        \input{plots/gaussian-well-2.pgf}
+        \scalebox{0.7}{\input{plots/gaussian-well-2.pgf}}
         \caption{$n_c=2$}
@@
     \begin{subfigure}[b]{0.32\columnwidth}
-        \input{plots/gaussian-well-5.pgf}
+        \scalebox{0.7}{\input{plots/gaussian-well-5.pgf}}
         \caption{$n_c=5$}
@@
 
-\reffig{fig:convergence} displays the convergence behavior of Chebyshev-Nyström++ applied to the matrix $\mtx{A}$ resulting from the discretization of the Hamiltonian for $n_c = 1$ and $n_c = 2$.
-Unlike \cite{lin-2017-randomized-estimation}, where just a small portion of the spectral density is considered, we approximate the whole spectrum of $\mtx{A}$ (transformed to $[-1, 1]$) in accordance with our theoretical results (\refsec{subsubsec:chebyshev-nystrom-analysis}) and to demonstrate the effectiveness of our implementational particularities (\refsec{subsubsec:chebyshev-nystrom-implementation}).
+\cref{fig:convergence} displays the convergence behavior of Chebyshev-Nyström++ applied to the matrix $\mtx{A}$ resulting from the discretization of the Hamiltonian for $n_c = 1$ and $n_c = 2$.
+Unlike \cite{lin-2017-randomized-estimation}, where just a small portion of the spectral density is considered, we approximate the whole spectrum of $\mtx{A}$ (transformed to $[-1, 1]$) in accordance with our theoretical results (\cref{subsubsec:chebyshev-nystrom-analysis}) and to demonstrate the effectiveness of our implementational particularities (\cref{subsubsec:chebyshev-nystrom-implementation}).
 As a reference, we use the eigenvalues computed by NumPy's Hermitian eigenvalue solver.
@@
     \begin{subfigure}[b]{0.495\textwidth}
-        \input{plots/convergence-1.pgf}
+        \scalebox{0.8}{\input{plots/convergence-1.pgf}}
         \caption{$n_c = 1.$}
@@
     \begin{subfigure}[b]{0.495\textwidth}
-        \input{plots/convergence-2.pgf}
+        \scalebox{0.8}{\input{plots/convergence-2.pgf}}
         \caption{$n_c = 2.$}
@@
     \caption{Error vs.
-number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ (with fixed $\sigma=0.005$ and $m=2000$) when applying  Chebyshev-Nyström++ to the Hamiltonian matrix from~\refsec{subsec:hamiltonian}.}
+number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ (with fixed $\sigma=0.005$ and $m=2000$) when applying  Chebyshev-Nyström++ to the Hamiltonian matrix from~\cref{subsec:hamiltonian}.}
     \label{fig:convergence}
 \end{figure}%
-As expected from \refthm{thm:hutchinson}, Girard-Hutchinson alone ($n_{\mtx{\Omega}} = 0$) requires $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ samples to achieve an error of order $\varepsilon$.
-Because the eigenvalues of the matrix are spread out, we observe the better convergence of the Nyström approximation discussed at the end of \refsec{sec:application} once $n_{\mtx{\Omega}}$ is sufficiently large.
-At this point, we can observe the errors stagnate at around $10^{-6}$ as a consequence of the error made in the Chebyshev approximation.
+As expected from \cref{thm:hutchinson}, Girard-Hutchinson alone ($n_{\mtx{\Omega}} = 0$) requires $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ samples to achieve an error of order $\varepsilon$.
+Because the eigenvalues of the matrix are spread out, we observe the better convergence of the Nyström approximation discussed at the end of \cref{sec:application} once $n_{\mtx{\Omega}}$ is sufficiently large.
+At this point, we can observe that the errors stagnate around $10^{-6}$ as a consequence of the error made in the Chebyshev approximation.
 Due to the higher eigenvalue density of $\mtx{A}$ for $n_c = 2$, the stronger convergence can be observed to kick in later in this case, because a larger $n_{\mtx{\Omega}}$ is needed for a good Nyström approximation.
@@
 
-For a fixed budget of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}} = 80$ of random vectors, \reffig{fig:distribution} shows how the value of the smoothing parameter $\sigma$ the choice between Nyström ($n_{\mtx{\Omega}}$) and
+For a fixed budget of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}} = 80$ of random vectors, \cref{fig:distribution} shows how the value of the smoothing parameter $\sigma$ the choice between Nyström ($n_{\mtx{\Omega}}$) and
 Girard-Hutchinson ($n_{\mtx{\Psi}}$). As expected, for small $\sigma$ the Nyström approximation alone suffices, but 
@@
     \centering
-    \input{plots/distribution.pgf}
+    \scalebox{0.8}{ \input{plots/distribution.pgf}}
     \caption{Error vs. $\sigma$ for a fixed budget of $80$ random vectors 
-    when applying  Chebyshev-Nyström++ to the Hamiltonian matrix from~\refsec{subsec:hamiltonian} with $n_c = 1$.
-The Chebyshev approximation error is made negligible by setting $m=16 / \sigma$ (see \reflem{lem:non-negative-chebyshev-error}).}
+    when applying  Chebyshev-Nyström++ to the Hamiltonian matrix from~\cref{subsec:hamiltonian} with $n_c = 1$.
+The Chebyshev approximation error is made negligible by setting $m=16 / \sigma$ (see \cref{lem:non-negative-chebyshev-error}).}
     \label{fig:distribution}
@@
 
-We run our own, faithfully optimized implementation of the Krylov-aware stochastic trace estimator \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} for different parameter settings on the example from \refsec{subsec:hamiltonian} and plot the approximation errors for logarithmically spaced values of the smoothing parameter $\sigma$ in \reffig{fig:krylov-aware-density}.
+We run our own, faithfully optimized implementation of the Krylov-aware stochastic trace estimator \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} for different parameter settings on the example from \cref{subsec:hamiltonian} and plot the approximation errors for logarithmically spaced values of the smoothing parameter $\sigma$ in \cref{fig:krylov-aware-density}.
 For reference, we also report the error of Chebyshev-Nyström++ for parameters that lead to a comparable run-time.
@@
 \begin{figure}[ht]
-    \begin{minipage}[c]{.475\linewidth}
+    \begin{minipage}[c]{.515\linewidth}
         \centering
-        \input{plots/krylov_aware_density.pgf}
+        \scalebox{0.8}{\input{plots/krylov_aware_density.pgf}}
     \end{minipage}\hfill%
@@
         \vspace{-35pt}
-        \input{tables/krylov_aware_density_KA.tex}
+        \scalebox{0.8}{\input{tables/krylov_aware_density_KA.tex}}
         \newline
@@
         \newline
-        \input{tables/krylov_aware_density_CN.tex}
+        \scalebox{0.8}{\input{tables/krylov_aware_density_CN.tex}}
     \end{minipage}
-    \caption{For the example from \refsec{subsec:hamiltonian}, error vs.
+    \caption{For the example from \cref{subsec:hamiltonian}, error vs.
 $\sigma$ the Krylov-Aware (KA) stochastic trace estimator with different parameter settings, compared to the Chebyshev-Nyström++ (CN++) method.}
@@
 
-We plot the spectral density of the Hessian matrix of the untrained neural network, and at different stages of training in \reffig{fig:hessian-density}, as well as the corresponding mean squared error loss.
+We plot the spectral density of the Hessian matrix of the untrained neural network, and at different stages of training in \cref{fig:hessian-density}, as well as the corresponding mean squared error loss.
 It can be observed that the eigenvalues creep towards zero as the training proceeds.
@@
         \centering
-        \input{plots/hessian_density.pgf}
+        \scalebox{0.8}{\input{plots/hessian_density.pgf}}
     \end{minipage}\hfill%
@@
         \centering
-        \input{plots/hessian_density_loss.pgf}
+        \scalebox{0.8}{\input{plots/hessian_density_loss.pgf}}
     \end{minipage}
@@
 To demonstrate that the Chebyshev-Nyström++ estimator is also effective for problems unrelated to spectral density estimation, we consider the problem of approximating the partition function $\Trace(\exp(-\beta \mtx{A}))$ for the $1\,048\,576 \times 1\,048\,576$ Hamiltonian $\mtx{A}$ of a planar Heisenberg spin chain from~\cite[Section 5.1]{chen-2023-krylovaware-stochastic}.
-We compute the pointwise error from the analytic solution as a function of the temperature $t\equiv \beta^{-1}$ and compare it to the error achieved by the Krylov-aware trace estimator in \reffig{fig:krylov-aware-spin}.
+We compute the pointwise error from the analytic solution as a function of the temperature $t\equiv \beta^{-1}$ and compare it to the error achieved by the Krylov-aware trace estimator in \cref{fig:krylov-aware-spin}.
 
 \begin{figure}[ht]
-    \begin{minipage}[c]{.475\linewidth}
+    \begin{minipage}[c]{.515\linewidth}
         \centering
-        \input{plots/krylov_aware_spin.pgf}
-    \end{minipage}\hfill%
+        \scalebox{0.8}{\input{plots/krylov_aware_spin.pgf}}
+    \end{minipage}\hfill
     \begin{minipage}[c]{.475\linewidth}
         \vspace{-35pt}
-        \input{tables/krylov_aware_spin_KA.tex}
+        \scalebox{0.8}{\input{tables/krylov_aware_spin_KA.tex}}
         \newline
@@
         \newline
-        \input{tables/krylov_aware_spin_CN.tex}
+        \scalebox{0.8}{\input{tables/krylov_aware_spin_CN.tex}}
     \end{minipage}
@@
 
-Unlike for the spectral density example from \refsec{subsec:hamiltonian}, where the numerical rank of the matrix is relatively uniform across different $t$, this now changes changes wildly with the parameter $\beta$.
+Unlike for the spectral density example from \cref{subsec:hamiltonian}, where the numerical rank of the matrix is relatively uniform across different $t$, this now changes changes wildly with the parameter $\beta$.
 Recall that the Chebyshev-Nyström++ method requires a fixed allocation of random vectors to the low-rank approximation for all values of the parameter $\beta$.

### 2025-06-27 paper/paper.tex
switch to SIAM template

@@
-\documentclass[11pt]{article}
+% SIAM Article Template
+\documentclass[review,hidelinks,onefignum,onetabnum]{siamart250211}
 
-\usepackage{stylesheet}
+% Information that is shared between the article and the supplement
+% (title and author information, macros, packages, etc.) goes into
+% ex_shared.tex. If there is no supplement, this file can be included
+% directly.
 
-\title{Stochastic trace estimation for parameter-dependent matrices applied to spectral density approximation}
+\input{shared}
 
-\author{Fabio Matti\thanks{Institute of Mathematics, EPFL, 1015 Lausanne, Switzerland.
-{\textbf{(\href{mailto:fabio.matti@epfl.ch}{fabio.matti}, \href{mailto:haoze.he@epfl.ch}{haoze.he}, \href{mailto:daniel.kressner@epfl.ch}{daniel.kressner}, \href{mailto:hysan.lam@epfl.ch}{hysan.lam})@epfl.ch}}.}
-\and Haoze He\footnotemark[1]
-\and Daniel Kressner\footnotemark[1]
-\and Hei Yin Lam\footnotemark[1]}
+%%% FIGURES
+% Figures
+\usepackage{graphicx}
+\usepackage{tikz}
+\usepackage{subcaption}
+\usepackage{booktabs}
+
+%%% COLOR THEME
+
+% Color package
+\usepackage{xcolor}
+
+% Color definitions
+\definecolor{darkblue}{HTML}{2F455C}
+\definecolor{mainblue}{HTML}{0e437c}
+\definecolor{darkorange}{HTML}{F98125}
+\colorlet{lightblue}{darkblue!10!white}
+\colorlet{lightorange}{darkorange!10!white}
+\colorlet{linkcolor}{black}
+
+% Fix undefined contorl sequence issue with pyplot pgf exports
+\def\mathdefault#1{#1}
+
+% Custom commands
+\newcommand{\pp}{\text{\raisebox{0.3ex}{\relscale{0.9}++}}}
+\newcommand*\rfrac[2]{{}^{#1}\!/_{#2}}
+\newcommand{\mtx}[1]{\boldsymbol{#1}}
+\newcommand{\vct}[1]{\boldsymbol{#1}}
+
+% Trace estimator notations
+\DeclareMathOperator{\Trace}{Tr}
+\newcommand\Hutch[1]{\Trace_{#1}}
+\newcommand\Nystr[2]{\widehat{#2}_{#1}}
+\newcommand\Nystrpp[2]{\Trace_{#2, #1}}
+
+% Sideway fractions for simple exponents (e.g. 1/2)
+\newcommand\sfrac[2]{{}^{#1}{\mskip -5mu/\mskip -3mu}_{#2}}
+
+% Algorithms
+%\usepackage{algpseudocodex}
+
+%%% Theorem environments
+
+% Optional PDF information
+\ifpdf
+\hypersetup{
+  pdftitle={Trace estimation for parameter-dependent matrices},
+  pdfauthor={F. Matti, H. He, D. Kressner, H. Lam}
+}
+\fi
+
+% The next statement enables references to information in the
+% supplement. See the xr-hyperref package for details.
+
+\externaldocument[][nocite]{ex_supplement}
+
+% FundRef data to be entered by SIAM
+%<funding-group specific-use="FundRef">
+%<award-group>
+%<funding-source>
+%<named-content content-type="funder-name"> 
+%</named-content> 
+%<named-content content-type="funder-identifier"> 
+%</named-content>
+%</funding-source>
+%<award-id> </award-id>
+%</award-group>
+%</funding-group>
 
 \begin{document}
-
 \maketitle
 
+% REQUIRED
 \begin{abstract}
@@
 
+% REQUIRED
+\begin{keywords}
+trace estimation, low-rank approximation, parameter-dependent matrix, Chebyshev interpolation, spectral density, density of states
+\end{keywords}
+
+% REQUIRED
+\begin{MSCcodes}
+65C05, 65F15, 65Y20, 68W20, 68W25, 68W40
+\end{MSCcodes}
+
 \input{intro.tex}
@@
 
-\paragraph{Acknowledgements.} The authors gratefully acknowledge the help of Lin Lin in reproducing the example in \refsec{subsec:hamiltonian}.
-They thank David Persson for many enlightening discussions relating the proof in \refsec{subsec:nystrom-pp}.
-
-\printbibliography
-
 \input{appendix.tex}
 
+\section*{Acknowledgments}
+The authors gratefully acknowledge the help of Lin Lin in reproducing the example in \cref{subsec:hamiltonian}.
+They thank David Persson for many enlightening discussions relating the proof in \cref{subsec:nystrom-pp}.
+
+%\section*{Acknowledgments}
+%We would like to acknowledge the assistance of volunteers in putting
+%together this example manuscript and supplement.
+
+\bibliographystyle{siamplain}
+\bibliography{bibliography}
+
 \end{document}

### 2025-06-27 paper/spectraldensity.tex
switch to SIAM template

@@
 \end{equation}
-The expression~\refequ{equ:smooth-spectral-density} gives rise to the task of estimating the trace of the parameter-dependent SPSD matrix $g_{\sigma}(t\mtx{I}_n - \mtx{A})$, which has been analyzed in \refsec{sec:analysis}.
+The expression~\cref{equ:smooth-spectral-density} gives rise to the task of estimating the trace of the parameter-dependent SPSD matrix $g_{\sigma}(t\mtx{I}_n - \mtx{A})$, which has been analyzed in \cref{sec:analysis}.
 
@@
 where $\mathcal{S}$ represents an appropriately chosen space of test functions.
-Among others, the reference~\cite{lin-2016-approximating-spectral} uses $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - s), s \in [a, b]\}$, \cite{chen-2021-analysis-stochastic} uses $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [a, b] \}$ with the Heaviside step function $\Theta$, and \cite{braverman-2022-sublinear-time,bhattacharjee-2025-improved-spectral} use $\mathcal{S} = \{f : |f(t) - f(s)| \leq |t - s| \}$ for which the error \refequ{equ:error-metric} is equivalent to the Wasserstein-1 distance between $\phi$ and $\phi_{\sigma}$.
+Among others, the reference~\cite{lin-2016-approximating-spectral} uses $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - s), s \in [a, b]\}$, \cite{chen-2021-analysis-stochastic} uses $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [a, b] \}$ with the Heaviside step function $\Theta$, and \cite{braverman-2022-sublinear-time,bhattacharjee-2025-improved-spectral} use $\mathcal{S} = \{f : |f(t) - f(s)| \leq |t - s| \}$ for which the error \cref{equ:error-metric} is equivalent to the Wasserstein-1 distance between $\phi$ and $\phi_{\sigma}$.
 When choosing 
 the last metric,
-assuming that the eigenvalues are somewhat uniformly distributed and $\sigma$ is rather small, a simple calculation allows one to conclude that the smoothing error \refequ{equ:error-metric} is of the order $\sigma$.
+assuming that the eigenvalues are somewhat uniformly distributed and $\sigma$ is rather small, a simple calculation allows one to conclude that the smoothing error \cref{equ:error-metric} is of the order $\sigma$.
 Consequently, to ensure that the approximation deviates by at most a factor of $\varepsilon > 0$, one needs to choose $\sigma \approx \varepsilon$.
@@
 
-To estimate the trace in \refequ{equ:smooth-spectral-density} efficiently, we approximate the function 
+To estimate the trace in \cref{equ:smooth-spectral-density} efficiently, we approximate the function 
 $s\mapsto g_{\sigma}(t- s)$ by a
@@
 \end{equation}
-Inserting~\refequ{equ:matrix-approximation} into~\refequ{equ:smooth-spectral-density} yields the approximate smoothed spectral density
+Inserting~\cref{equ:matrix-approximation} into~\cref{equ:smooth-spectral-density} yields the approximate smoothed spectral density
 \begin{equation}
@@
 $\tau(t) = \frac{2t - a - b}{b - a}$
-then lets us define a matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ for which the spectrum is contained in $[-1, 1]$ and, hence, the approximation \refequ{equ:matrix-approximation} is sensible.\footnote{Note that one has to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when inverting the transformation.} This allows to assume, from now on and without loss of generality, that the spectrum of $\mtx{A}$ is contained in $[-1, 1]$.
+then lets us define a matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ for which the spectrum is contained in $[-1, 1]$ and, hence, the approximation \cref{equ:matrix-approximation} is sensible.\footnote{Note that one has to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when inverting the transformation.} This allows to assume, from now on and without loss of generality, that the spectrum of $\mtx{A}$ is contained in $[-1, 1]$.
 \end{remark}
@@
 
-\begin{lemma}{Chebyshev approximation error}{chebyshev-error}
-    The Chebyshev approximation $g_{\sigma}^{(m)}$ of the Gaussian smoothing kernel $g_{{\sigma}}$ from  \refequ{equ:smoothing-kernel} satisfies
+\begin{lemma}[Chebyshev approximation error]\label{lem:chebyshev-error}
+    The Chebyshev approximation $g_{\sigma}^{(m)}$ of the Gaussian smoothing kernel $g_{{\sigma}}$ from  \cref{equ:smoothing-kernel} satisfies
     \begin{equation}
@@
     $e^{-\sfrac{(t - x)^2}{2\sigma^2}} \leq 1$ and $|y| \le (\chi - \chi^{-1}) / 2$.
-    Choosing $\chi = 1 + \sigma$ and using $\chi - \chi^{-1} \leq 2\sigma$ completes the proof of~\refequ{equ:chebyshev-interpolation-sup-error-kernel}.
+    Choosing $\chi = 1 + \sigma$ and using $\chi - \chi^{-1} \leq 2\sigma$ completes the proof of~\cref{equ:chebyshev-interpolation-sup-error-kernel}.
 \end{proof}
 
-A simple algebraic manipulation shows that \reflem{lem:chebyshev-error} implies that choosing a degree $m = \mathcal{O}((\log(\varepsilon^{-1}) + \log(\sigma^{-1})) / \log(1 + \sigma))$ ensures an error $\mathcal{O}(\varepsilon)$.
+A simple algebraic manipulation shows that \cref{lem:chebyshev-error} implies that choosing a degree $m = \mathcal{O}((\log(\varepsilon^{-1}) + \log(\sigma^{-1})) / \log(1 + \sigma))$ ensures an error $\mathcal{O}(\varepsilon)$.
 The lemma also allows us to bound the $L^1$-error of the spectral density approximation
-$\phi_{\sigma}^{(m)}$ defined in~\refequ{equ:expanded-spectral-density}:
+$\phi_{\sigma}^{(m)}$ defined in~\cref{equ:expanded-spectral-density}:
 \begin{align}
@@
 (slightly) indefinite.
-This causes issues when attempting to apply the Nyström++ trace estimator (\refalg{equ:nystrompp-trace-estimator}).
+This causes issues when attempting to apply the Nyström++ trace estimator (\cref{equ:nystrompp-trace-estimator}).
 While we have never observed this in our numerical experiments, the Nyström approximation may fail when the SPSD condition is not met; see also the discussion in~\cite{nakatsukasa-2023-randomized-lowrank}.
-On the theoretical side,~\refthm{thm:nystrom-pp}
+On the theoretical side,~\cref{thm:nystrom-pp}
 critically relies on the SPSD condition.
@@
 
-\begin{lemma}{$L^1$-error of non-negative approximate smooth spectral density}{non-negative-chebyshev-error}
-Let $\phi_{\sigma}$ be the smoothed spectral density defined in~\refequ{equ:smooth-spectral-density}
+\begin{lemma}[$L^1$-error of non-negative approximate smooth spectral density]\label{lem:non-negative-chebyshev-error}
+Let $\phi_{\sigma}$ be the smoothed spectral density defined in~\cref{equ:smooth-spectral-density}
 for a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum contained in $[-1, 1]$.
-Then the approximation $\underline{\phi}_{\sigma}^{(m)}$ defined in~\refequ{equ:non-negative-expanded-density} satisfies
+Then the approximation $\underline{\phi}_{\sigma}^{(m)}$ defined in~\cref{equ:non-negative-expanded-density} satisfies
     \begin{equation*}
@@
     \end{equation*}
-    for every \emph{even} degree $m \in \mathbb{N}$, smoothing parameter $\sigma > 0$, and with $E_{\sigma, m}$ defined in \reflem{lem:chebyshev-error}.
+    for every \emph{even} degree $m \in \mathbb{N}$, smoothing parameter $\sigma > 0$, and with $E_{\sigma, m}$ defined in \cref{lem:chebyshev-error}.
 \end{lemma}
@@
     Because of $\sqrt{g_{\sigma}} = \sqrt{2 \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$,
-we can apply \reflem{lem:chebyshev-error} with the substitutions
+we can apply \cref{lem:chebyshev-error} with the substitutions
     $\sigma \gets \sqrt{2}\sigma$ and $m \gets m/2$ to bound 
@@
     for every $s,t \in [-1, 1]$.
-    Bounding the integral as in~\refequ{equ:chebyshev-interpolation-spectral-density} then yields the desired result.
+    Bounding the integral as in~\cref{equ:chebyshev-interpolation-spectral-density} then yields the desired result.
 \end{proof}
 
-Again, \reflem{lem:non-negative-chebyshev-error} implies that choosing a degree $m = \mathcal{O}((\log(\varepsilon^{-1}) + \log(\sigma^{-1})) / \log(1 + \sigma))$ ensures an error $\mathcal{O}(\varepsilon)$.
+Again, \cref{lem:non-negative-chebyshev-error} implies that choosing a degree $m = \mathcal{O}((\log(\varepsilon^{-1}) + \log(\sigma^{-1})) / \log(1 + \sigma))$ ensures an error $\mathcal{O}(\varepsilon)$.
 
-As the non-negative approximation~\refequ{equ:non-negative-chebyshev-approximation} ``wastes'' half of the degrees of freedom, it can be expected to be less accurate than the standard Chebyshev approximation.
-On the other hand, $\sqrt{g_{\sigma}} \propto g_{\sqrt{2}\sigma}$ is somewhat easier to approximate because it is less \enquote{pointy} compared to $g_{\sigma}$.
-Overall, it turns out that one needs to choose the degree of the non-negative approximation roughly a factor $\sqrt{2}$ larger to match the approximation error of the standard Chebyshev approximation (see also \reffig{fig:interpolation-issue}).
+As the non-negative approximation~\cref{equ:non-negative-chebyshev-approximation} ``wastes'' half of the degrees of freedom, it can be expected to be less accurate than the standard Chebyshev approximation.
+On the other hand, $\sqrt{g_{\sigma}} \propto g_{\sqrt{2}\sigma}$ is somewhat easier to approximate because it is less ``pointy" compared to $g_{\sigma}$.
+Overall, it turns out that one needs to choose the degree of the non-negative approximation roughly a factor $\sqrt{2}$ larger to match the approximation error of the standard Chebyshev approximation (see also \cref{fig:interpolation-issue}).
 
@@
 
-It is well known that the coefficients of the Chebyshev approximation \refequ{equ:matrix-approximation} can --- in exact arithmetic --- be computed by inverting the (type I) discrete cosine transform (DCT)
+It is well known that the coefficients of the Chebyshev approximation \cref{equ:matrix-approximation} can --- in exact arithmetic --- be computed by inverting the (type I) discrete cosine transform (DCT)
 \begin{equation*}
@@
 
-A Chebyshev approximation can be squared in $\mathcal{O}(m \log(m))$ operations by using the DCT and its inverse \cite{baszenski-1997-fast-polynomial}; see \refalg{alg:chebyshev-squaring}.
-This allows one to efficiently compute the non-negative Chebyshev approximation \refequ{equ:non-negative-chebyshev-approximation}, and it will find further use in \refsec{subsec:chebyshev-nystrom}.
+A Chebyshev approximation can be squared in $\mathcal{O}(m \log(m))$ operations by using the DCT and its inverse \cite{baszenski-1997-fast-polynomial}; see \cref{alg:chebyshev-squaring}.
+This allows one to efficiently compute the non-negative Chebyshev approximation \cref{equ:non-negative-chebyshev-approximation}, and it will find further use in \cref{subsec:chebyshev-nystrom}.
 
-\begin{algo}{Fast squaring of Chebyshev approximations}{chebyshev-squaring}
+\begin{algorithm}
+    \caption{Fast squaring of Chebyshev approximations}
+    \label{alg:chebyshev-squaring}
+    \hspace*{\algorithmicindent} \textbf{Input:}  Coefficients $\{ \mu_l \}_{l=0}^{m}$ of Chebyshev approximation $\sum_{l=0}^{m} \mu_l T_l$\\
+    \hspace*{\algorithmicindent} \textbf{Output:} Coefficients $\{ \nu_l \}_{l=0}^{2m}$ such that $\sum_{l=0}^{2m} \nu_l T_l = (\sum_{l=0}^{m} \mu_l T_l)^2$
     \begin{algorithmic}[1]
-        \Statex \textbf{Input:} Coefficients $\{ \mu_l \}_{l=0}^{m}$ of Chebyshev approximation $\sum_{l=0}^{m} \mu_l T_l$
-        \Statex \textbf{Output:} Coefficients $\{ \nu_l \}_{l=0}^{2m}$ such that $\sum_{l=0}^{2m} \nu_l T_l = (\sum_{l=0}^{m} \mu_l T_l)^2$
-        \State By appending $m$ zeros, define vector $\vct{\mu} = [\mu_0, \ldots, \mu_m,0,\ldots,0 ] \in \mathbb{R}^{2m + 1}$.
-    \State Compute $\vct{f} = \operatorname{DCT}(\vct{\mu})$
-    \State Compute $[\nu_0, \dots, \nu_{2m}] = \operatorname{DCT}^{-1}(\vct{f} \odot \vct{f})$
+    \STATE{Compute $\vct{f} = \operatorname{DCT}(\vct{\mu})$}
+    \STATE{Compute $[\nu_0, \dots, \nu_{2m}] = \operatorname{DCT}^{-1}(\vct{f} \odot \vct{f})$}
     \label{lin:inverse-DCT}
     \end{algorithmic}
-\end{algo}
-On \reflin{lin:inverse-DCT} of \refalg{alg:chebyshev-squaring}, the symbol $\odot$ denotes the entrywise product.
+\end{algorithm}
+
+On line \ref{lin:inverse-DCT} of \cref{alg:chebyshev-squaring}, the symbol $\odot$ denotes the entrywise product.
 
@@
 Although interpolation and truncated expansion satisfy nearly identical convergence bounds~\cite[Theorem 8.2]{trefethen-2020-approximation-theory}, we observed the procedure from~\cite{lin-2017-randomized-estimation} to be significantly slower due to need of applying the FFT (instead of DCT) to longer vectors.
-See~\reftab{tab:chebyshev-timing-interpolation}, where we compare a benevolent implementation of \cite[Algorithm 1]{lin-2017-randomized-estimation} with our approach, both for the standard and the non-negative Chebyshev approximation.%}
+See~\cref{tab:chebyshev-timing-interpolation}, where we compare a benevolent implementation of \cite[Algorithm 1]{lin-2017-randomized-estimation} with our approach, both for the standard and the non-negative Chebyshev approximation.%}
 
@@
 \emph{DCT}: Applying the discrete cosine transform directly to $g_\sigma$.
-\emph{Non-negative DCT}: Applying the discrete cosine transform to $\sqrt{g_\sigma}$ together with \refalg{alg:chebyshev-squaring}.
+\emph{Non-negative DCT}: Applying the discrete cosine transform to $\sqrt{g_\sigma}$ together with \cref{alg:chebyshev-squaring}.
 We use $\sigma=0.005$, $n_t=1000$ parameter values, and various degrees $m$.
@@
 
-Combining the results of \refsec{sec:analysis} and \refsec{subsec:chebyshev-approximation}, we apply the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$ from~\refequ{equ:nystrompp-trace-estimator}
+Combining the results of \cref{sec:analysis} and \cref{subsec:chebyshev-approximation}, we apply the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$ from~\cref{equ:nystrompp-trace-estimator}
 to the non-negative 
@@
 
-One usually aims at evaluating~\refequ{equ:chebyshev-nystrom-formula} for many values of $t$. In the following, we explain how this can be done efficiently by exploiting 
+One usually aims at evaluating~\cref{equ:chebyshev-nystrom-formula} for many values of $t$. In the following, we explain how this can be done efficiently by exploiting 
 the well-known three-term recurrence relation of Chebyshev polynomials:
@@
 
-We start with the first term in the definition~\refequ{equ:nystrompp-trace-estimator} of $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$, which requires evaluating the trace of the 
-Nyström approximation \refequ{equ:nystrom-approximation}:
+We start with the first term in the definition~\cref{equ:nystrompp-trace-estimator} of $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$, which requires evaluating the trace of the 
+Nyström approximation \cref{equ:nystrom-approximation}:
 \begin{equation}
@@
 \end{equation*}
-Observing that $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ is independent of $t$ allows us to precompute these (small) $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices for $l=0, \dots, m$ using the Chebyshev recurrence \refequ{equ:chebyshev-recurrence}.
+Observing that $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ is independent of $t$ allows us to precompute these (small) $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices for $l=0, \dots, m$ using the Chebyshev recurrence \cref{equ:chebyshev-recurrence}.
 For each value of $t$, we then only need to sum these matrices (multiplied by the corresponding coefficient $\underline{\mu}_l(t)$).
 
-The squaring of Chebyshev approximations discussed in \refsec{sec:preservnonneg} allows us to express $\mtx{B}(t)^2 = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})^2 = \sum_{l=0}^{2m} \underline{\nu}_l(t) T_l(\mtx{A})$, where the coefficients $\underline{\nu}_l(t)$ can be cheaply computed from $\underline{\mu}_l(t)$ using~\refalg{alg:chebyshev-squaring}.
+The squaring of Chebyshev approximations discussed in \cref{sec:preservnonneg} allows us to express $\mtx{B}(t)^2 = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})^2 = \sum_{l=0}^{2m} \underline{\nu}_l(t) T_l(\mtx{A})$, where the coefficients $\underline{\nu}_l(t)$ can be cheaply computed from $\underline{\mu}_l(t)$ using~\cref{alg:chebyshev-squaring}.
 Hence, if we precompute
@@
 We have observed that meticulously ensuring consistency between the approximations of $g_{\sigma}$ and $g_{\sigma}^2$ makes a visible difference in terms of accuracy (cf.
-\reffig{fig:interpolation-issue}).
-This consistency will also allow us to derive bounds for the estimator \refequ{equ:chebyshev-nystrom-formula}.
+\cref{fig:interpolation-issue}).
+This consistency will also allow us to derive bounds for the estimator \cref{equ:chebyshev-nystrom-formula}.
 
@@
     \centering
-    \input{plots/interpolation.pgf}
-    \caption{Accuracy when computing the estimate \refequ{equ:cyclic-property} in three different ways.
+    \scalebox{0.8}{\input{plots/interpolation.pgf}}
+    \caption{Accuracy when computing the estimate \cref{equ:cyclic-property} in three different ways.
 \emph{Inconsistent}: Applying an auxiliary truncated Chebyshev series to $g_{\sigma}^2$, as proposed in~\cite{lin-2017-randomized-estimation}.
-\emph{Consistent}: Squaring the Chebyshev approximation $g_{\sigma}^{(m)}$ using \refalg{alg:chebyshev-squaring}.
-\emph{Non-negative consistent}: Squaring the non-negative Chebyshev approximation $\underline{g}_{\sigma}^{(m)}$ with~\refalg{alg:chebyshev-squaring}.
-As input matrix we use the $100 \times 100$ matrix $\mtx{A}$ coming from a two-dimensional finite difference discretization of the Hamiltonian defined in~\refequ{equ:electronic-hamiltonian} on a $10 \times 10$ grid.
+\emph{Consistent}: Squaring the Chebyshev approximation $g_{\sigma}^{(m)}$ using \cref{alg:chebyshev-squaring}.
+\emph{Non-negative consistent}: Squaring the non-negative Chebyshev approximation $\underline{g}_{\sigma}^{(m)}$ with~\cref{alg:chebyshev-squaring}.
+As input matrix we use the $100 \times 100$ matrix $\mtx{A}$ coming from a two-dimensional finite difference discretization of the Hamiltonian defined in~\cref{equ:electronic-hamiltonian} on a $10 \times 10$ grid.
 We fix $n_{\mtx{\Omega}} = 80$ to make sure the low-rank approximation is accurate, set $\sigma = 0.005$, and compute the $L^1$ error from the approximation of the spectral density at $n_t = 100$ uniformly spaced values of $t$ in the interval $[-1, 1]$.}
@@
 
-To treat the second term in the definition~\refequ{equ:nystrompp-trace-estimator} of $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$, we observe that 
+To treat the second term in the definition~\cref{equ:nystrompp-trace-estimator} of $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$, we observe that 
 \begin{align}
@@
 
-\refalg{alg:nystrom-chebyshev-pp} summarizes the considerations of this section.
+\cref{alg:nystrom-chebyshev-pp} summarizes the considerations of this section.
 
-\begin{algo}{Chebyshev-Nyström++ method}{nystrom-chebyshev-pp}
+\begin{algorithm}
+    \caption{Fast squaring of Chebyshev approximations}
+    \label{alg:nystrom-chebyshev-pp}
+    \hspace*{\algorithmicindent} \textbf{Input:}  Symmetric $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum in $[-1, 1]$, points $\{t_i\}_{i=1}^{n_t} \subset \mathbb{R}$\\
+    \hspace*{\algorithmicindent} \textbf{Parameters:} Even approximation degree $m \in \mathbb{N}$, number of Girard-Hutchinson queries $n_{\mtx{\Psi}} \in \mathbb{N}_0$, Nyström sketch size $n_{\mtx{\Omega}} \in \mathbb{N}_0$,  smoothing parameter $\sigma > 0$\\
+    \hspace*{\algorithmicindent} \textbf{Output:} Spectral density approximation $\{\widetilde{\underline{\phi}}_{\sigma}^{(m)}(t_i)\}_{i=1}^{n_t}$
     \begin{algorithmic}[1]
-    \Statex \textbf{Input:} Symmetric $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum in $[-1, 1]$, points $\{t_i\}_{i=1}^{n_t} \subset \mathbb{R}$
-    \Statex \textbf{Parameters:} Even approximation degree $m \in \mathbb{N}$, number of Girard-Hutchinson queries $n_{\mtx{\Psi}} \in \mathbb{N}_0$, Nyström sketch size $n_{\mtx{\Omega}} \in \mathbb{N}_0$,  smoothing parameter $\sigma > 0$
-    \Statex \textbf{Output:} Spectral density approximation $\{\widetilde{\underline{\phi}}_{\sigma}^{(m)}(t_i)\}_{i=1}^{n_t}$
-    \For {$i = 1, \dots, n_t$}
-        \State Compute coefficients $\{\xi_l(t_i)\}_{l=0}^{\sfrac{m}{2}}$ of Chebyshev approx.~\refequ{equ:square-root-chebyshev-approximation} to $\sqrt{g_{\sigma}}$ with DCT
-        \State Compute coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ of non-negative Chebyshev approx.
-to $g_{\sigma}$ by applying \refalg{alg:chebyshev-squaring} to $\{\xi_l(t_i)\}_{l=0}^{\sfrac{m}{2}}$
-        \State Compute coefficients $\{\underline{\nu}_l(t_i)\}_{l=0}^{2m}$ of squared non-negative Chebyshev approx.
-by applying \refalg{alg:chebyshev-squaring} to $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$
-    \EndFor
-    \State Generate Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$, $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$
-    \State Initialize $[\mtx{V}_1, \mtx{V}_2, \mtx{V}_3] \gets [\mtx{0}_{n \times n_{\mtx{\Omega}}}, \mtx{\mtx{\Omega}}, \mtx{0}_{n \times n_{\mtx{\Omega}}}]$
-    \State Initialize $[\mtx{W}_1, \mtx{W}_2, \mtx{W}_3] \gets [\mtx{0}_{n \times n_{\mtx{\Psi}}}, \mtx{\Psi}, \mtx{0}_{n \times n_{\mtx{\Psi}}}]$
-    \State Initialize $[\mtx{K}_1(t_i), \mtx{K}_2(t_i)] \gets [\mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}}, \mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}}]$ for all $t_i$
-    \State Initialize $[\mtx{L}_1(t_i), \ell(t_i)] \gets [\mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Psi}}}, 0]$ for all $t_i$
-    \For {$l = 0, \dots, 2m$}
-    \State $[\mtx{X}, \mtx{Y}] \gets \mtx{\mtx{\Omega}}^{\top} [\mtx{V}_2, \mtx{W}_2]$  
-      \State $z \gets \Trace(\mtx{\Psi}^{\top} \mtx{W}_2)$
-      \For {$i = 1, \dots, n_t$}
-        \If {$l \leq m$}
-            \State $\mtx{K}_1(t_i) \gets \mtx{K}_1(t_i) + \underline{\mu}_l(t_i) \mtx{X}$ \Comment{assembles $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$}
-            \State $\mtx{L}_1(t_i) \gets \mtx{L}_1(t_i) + \underline{\mu}_l(t_i) \mtx{Y}$ \Comment{assembles $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi}$}
-            \State $\ell(t_i) \gets \ell(t_i) + \underline{\mu}_l(t_i) z$ \Comment{assembles $\Trace(\mtx{\Psi}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi})$}
-        \EndIf
-        \State $\mtx{K}_2(t_i) \gets \mtx{K}_2(t_i) + \underline{\nu}_l(t_i) \mtx{X}$ \Comment{assembles $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})^2 \mtx{\Omega}$}
-      \EndFor
-      \State $[\mtx{V}_3, \mtx{W}_3] \gets (2 - \delta_{l0}) \mtx{A} [\mtx{V}_2, \mtx{W}_2] - [\mtx{V}_1, \mtx{W}_1]$ \Comment{Chebyshev recurrence}
-      \State $[\mtx{V}_1, \mtx{W}_1] \gets [\mtx{V}_2, \mtx{W}_2]$
-      \State $[\mtx{V}_2, \mtx{W}_2] \gets [\mtx{V}_3, \mtx{W}_3]$
-    \EndFor
-    \For {$i = 1, \dots, n_t$}
-      \State $\widetilde{\underline{\phi}}_{\sigma}(t_i) \gets \frac{1}{n} \Trace\left( \mtx{K}_1(t_i)^{\dagger}\mtx{K}_2(t_i) \right) + \frac{1}{n n_{\mtx{\Psi}}} \left( \ell(t_i) + \Trace\left( \mtx{L}_1(t_i)^{\top} \mtx{K}_1(t_i)^{\dagger} \mtx{L}_1(t_i) \right)  \right) $ \label{lin:4-nystromchebyshev-nystrom-pp}
-    \EndFor
+    \FOR{$i = 1, \dots, n_t$}
+        \STATE{Compute coefficients $\{\xi_l(t_i)\}_{l=0}^{\sfrac{m}{2}}$ of Chebyshev approx.~\cref{equ:square-root-chebyshev-approximation} to $\sqrt{g_{\sigma}}$ with DCT}
+        \STATE{Compute coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ of non-negative Chebyshev approx.
+to $g_{\sigma}$ by applying \cref{alg:chebyshev-squaring} to $\{\xi_l(t_i)\}_{l=0}^{\sfrac{m}{2}}$}
+        \STATE{Compute coefficients $\{\underline{\nu}_l(t_i)\}_{l=0}^{2m}$ of squared non-negative Chebyshev approx.
+by applying \cref{alg:chebyshev-squaring} to $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$}
+    \ENDFOR
+    \STATE{Generate Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$, $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$}
+    \STATE{Initialize $[\mtx{V}_1, \mtx{V}_2, \mtx{V}_3] \gets [\mtx{0}_{n \times n_{\mtx{\Omega}}}, \mtx{\mtx{\Omega}}, \mtx{0}_{n \times n_{\mtx{\Omega}}}]$}
+    \STATE{Initialize $[\mtx{W}_1, \mtx{W}_2, \mtx{W}_3] \gets [\mtx{0}_{n \times n_{\mtx{\Psi}}}, \mtx{\Psi}, \mtx{0}_{n \times n_{\mtx{\Psi}}}]$}
+    \STATE{Initialize $[\mtx{K}_1(t_i), \mtx{K}_2(t_i)] \gets [\mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}}, \mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}}]$ for all $t_i$}
+    \STATE{Initialize $[\mtx{L}_1(t_i), \ell(t_i)] \gets [\mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Psi}}}, 0]$ for all $t_i$}
+    \FOR{$l = 0, \dots, 2m$}
+        \STATE{$[\mtx{X}, \mtx{Y}] \gets \mtx{\mtx{\Omega}}^{\top} [\mtx{V}_2, \mtx{W}_2]$  }
+        \STATE{ $z \gets \Trace(\mtx{\Psi}^{\top} \mtx{W}_2)$}
+        \FOR{$i = 1, \dots, n_t$}
+            \IF {$l \leq m$}
+                \STATE{$\mtx{K}_1(t_i) \gets \mtx{K}_1(t_i) + \underline{\mu}_l(t_i) \mtx{X}$ \hfill\COMMENT{assembles $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$}}
+                \STATE{$\mtx{L}_1(t_i) \gets \mtx{L}_1(t_i) + \underline{\mu}_l(t_i) \mtx{Y}$ \hfill\COMMENT{assembles $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi}$}}
+                \STATE{$\ell(t_i) \gets \ell(t_i) + \underline{\mu}_l(t_i) z$ \hfill\COMMENT{assembles $\Trace(\mtx{\Psi}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi})$}}
+            \ENDIF
+            \STATE{$\mtx{K}_2(t_i) \gets \mtx{K}_2(t_i) + \underline{\nu}_l(t_i) \mtx{X}$ \hfill\COMMENT{assembles $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})^2 \mtx{\Omega}$}}
+        \ENDFOR
+        \STATE{$[\mtx{V}_3, \mtx{W}_3] \gets (2 - \delta_{l0}) \mtx{A} [\mtx{V}_2, \mtx{W}_2] - [\mtx{V}_1, \mtx{W}_1]$ \hfill\COMMENT{Chebyshev recurrence}}
+        \STATE{$[\mtx{V}_1, \mtx{W}_1] \gets [\mtx{V}_2, \mtx{W}_2]$}
+        \STATE{$[\mtx{V}_2, \mtx{W}_2] \gets [\mtx{V}_3, \mtx{W}_3]$}
+    \ENDFOR
+    \FOR{$i = 1, \dots, n_t$}
+        \STATE{$\widetilde{\underline{\phi}}_{\sigma}(t_i) \gets \frac{1}{n} \Trace\left( \mtx{K}_1(t_i)^{\dagger}\mtx{K}_2(t_i) \right) + \frac{1}{n n_{\mtx{\Psi}}} \left( \ell(t_i) + \Trace\left( \mtx{L}_1(t_i)^{\top} \mtx{K}_1(t_i)^{\dagger} \mtx{L}_1(t_i) \right)  \right) $} \label{lin:4-nystromchebyshev-nystrom-pp}
+    \ENDFOR
     \end{algorithmic}
-\end{algo}
+\end{algorithm}
 
-\refalg{alg:nystrom-chebyshev-pp} requires $(2m + 1)(n_{\mtx{\Psi}} + n_{\mtx{\Omega}})$ matrix-vector multiplications with $\mtx{A}$, which can be expected to dominate its cost.
+%\begin{algo}{Chebyshev-Nyström++ method}{nystrom-chebyshev-pp}
+%\begin{algorithmic}[1]
+%    \Statex \textbf{Input:} Symmetric $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum in $[-1, 1]$, points $\{t_i\}_{i=1}^{n_t} \subset \mathbb{R}$
+%    \Statex \textbf{Parameters:} Even approximation degree $m \in \mathbb{N}$, number of Girard-Hutchinson queries $n_{\mtx{\Psi}} \in \mathbb{N}_0$, Nyström sketch size $n_{\mtx{\Omega}} \in \mathbb{N}_0$,  smoothing parameter $\sigma > 0$
+%    \Statex \textbf{Output:} Spectral density approximation $\{\widetilde{\underline{\phi}}_{\sigma}^{(m)}(t_i)\}_{i=1}^{n_t}$
+%    f\For {$i = 1, \dots, n_t$}
+%        \State Compute coeficients $\{\xi_l(t_i)\}_{l=0}^{\sfrac{m}{2}}$ of Chebyshev approx.~\cref{equ:square-root-chebyshev-approximation} to $\sqrt{g_{\sigma}}$ with DCT
+%        \State Compute coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ of non-negative Chebyshev approx.
+to $g_{\sigma}$ by applying \cref{alg:chebyshev-squaring} to $\{\xi_l(t_i)\}_{l=0}^{\sfrac{m}{2}}$
+%        \State Compute coefficients $\{\underline{\nu}_l(t_i)\}_{l=0}^{2m}$ of squared non-negative Chebyshev approx.
+by applying \cref{alg:chebyshev-squaring} to $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$
+%    \EndFor
+%    \State Generate Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$, $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$
+%    \State Initialize $[\mtx{V}_1, \mtx{V}_2, \mtx{V}_3] \gets [\mtx{0}_{n \times n_{\mtx{\Omega}}}, \mtx{\mtx{\Omega}}, \mtx{0}_{n \times n_{\mtx{\Omega}}}]$
+%    \State Initialize $[\mtx{W}_1, \mtx{W}_2, \mtx{W}_3] \gets [\mtx{0}_{n \times n_{\mtx{\Psi}}}, \mtx{\Psi}, \mtx{0}_{n \times n_{\mtx{\Psi}}}]$
+%    \State Initialize $[\mtx{K}_1(t_i), \mtx{K}_2(t_i)] \gets [\mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}}, \mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}}]$ for all $t_i$
+%    \State Initialize $[\mtx{L}_1(t_i), \ell(t_i)] \gets [\mtx{0}_{n_{\mtx{\Omega}} \times n_{\mtx{\Psi}}}, 0]$ for all $t_i$
+%    \For {$l = 0, \dots, 2m$}
+%    \State $[\mtx{X}, \mtx{Y}] \gets \mtx{\mtx{\Omega}}^{\top} [\mtx{V}_2, \mtx{W}_2]$  
+%      \State $z \gets \Trace(\mtx{\Psi}^{\top} \mtx{W}_2)$
+%      \For {$i = 1, \dots, n_t$}
+%        \If {$l \leq m$}
+%            \State $\mtx{K}_1(t_i) \gets \mtx{K}_1(t_i) + \underline{\mu}_l(t_i) \mtx{X}$ \hfill\COMMENT{assembles $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$}
+%            \State $\mtx{L}_1(t_i) \gets \mtx{L}_1(t_i) + \underline{\mu}_l(t_i) \mtx{Y}$ \hfill\COMMENT{assembles $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi}$}
+%            \State $\ell(t_i) \gets \ell(t_i) + \underline{\mu}_l(t_i) z$ \hfill\COMMENT{assembles $\Trace(\mtx{\Psi}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi})$}
+%        \EndIf
+%        \State $\mtx{K}_2(t_i) \gets \mtx{K}_2(t_i) + \underline{\nu}_l(t_i) \mtx{X}$ \hfill\COMMENT{assembles $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})^2 \mtx{\Omega}$}
+%      \EndFor
+%      \State $[\mtx{V}_3, \mtx{W}_3] \gets (2 - \delta_{l0}) \mtx{A} [\mtx{V}_2, \mtx{W}_2] - [\mtx{V}_1, \mtx{W}_1]$ \hfill\COMMENT{Chebyshev recurrence}
+%      \State $[\mtx{V}_1, \mtx{W}_1] \gets [\mtx{V}_2, \mtx{W}_2]$
+%      \State $[\mtx{V}_2, \mtx{W}_2] \gets [\mtx{V}_3, \mtx{W}_3]$
+%    \EndFor
+%    \For {$i = 1, \dots, n_t$}
+%      \State $\widetilde{\underline{\phi}}_{\sigma}(t_i) \gets \frac{1}{n} \Trace\left( \mtx{K}_1(t_i)^{\dagger}\mtx{K}_2(t_i) \right) + \frac{1}{n n_{\mtx{\Psi}}} \left( \ell(t_i) + \Trace\left( \mtx{L}_1(t_i)^{\top} \mtx{K}_1(t_i)^{\dagger} \mtx{L}_1(t_i) \right)  \right) $ \label{lin:4-nystromchebyshev-nystrom-pp}
+%    \EndFor
+%\end{algorithmic}
+%\end{algo}
+
+\cref{alg:nystrom-chebyshev-pp} requires $(2m + 1)(n_{\mtx{\Psi}} + n_{\mtx{\Omega}})$ matrix-vector multiplications with $\mtx{A}$, which can be expected to dominate its cost.
 For each of the evaluation points $t_1, \dots, t_{n_t}$ another $\mathcal{O}(m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + (n_{\mtx{\Omega}}^3 + n_{\mtx{\Omega}} n_{\mtx{\Psi}}^2 + 1))$ operations are needed, which is independent of $n$ --- thanks to the precomputations discussed above.
 
-\refalg{alg:nystrom-chebyshev-pp} also accommodates the other two estimators discussed in \refsec{sec:analysis}: For $n_{\mtx{\Omega}} = 0$ we recover the Girard-Hutchinson estimator~\refequ{equ:hutchinson-trace-estimator}.
-For $n_{\mtx{\Psi}} = 0$ we recover the Nyström estimator~\refequ{equ:nystrom-trace-estimator}, provided that $0/0$ is evaluated to $0$ in the second term on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
-In this case, \refalg{alg:nystrom-chebyshev-pp} essentially amounts to the \enquote{spectrum sweeping} method from \cite[Algorithm 5]{lin-2017-randomized-estimation}, featuring a significantly lower computational complexity.
+\cref{alg:nystrom-chebyshev-pp} also accommodates the other two estimators discussed in \cref{sec:analysis}: For $n_{\mtx{\Omega}} = 0$ we recover the Girard-Hutchinson estimator~\cref{equ:hutchinson-trace-estimator}.
+For $n_{\mtx{\Psi}} = 0$ we recover the Nyström estimator~\cref{equ:nystrom-trace-estimator}, provided that $0/0$ is evaluated to $0$ in the second term on line \ref{lin:4-nystromchebyshev-nystrom-pp}.
+In this case, \cref{alg:nystrom-chebyshev-pp} essentially amounts to the ``spectrum sweeping" method from \cite[Algorithm 5]{lin-2017-randomized-estimation}, featuring a significantly lower computational complexity.
 
-The two pseudoinverses on \reflin{lin:4-nystromchebyshev-nystrom-pp} need to be computed with care. One way to ensure numerical stability is to 
+The two pseudoinverses on line \ref{lin:4-nystromchebyshev-nystrom-pp} need to be computed with care. One way to ensure numerical stability is to 
 consider an associated eigenvalue problem and threshold the smallest eigenvalues~\cite{lin-2017-randomized-estimation, epperly-2022-theory-quantum}.
@@
 Especially for small smoothing parameters $\sigma$, the rapid decay of the Gaussian $g_{\sigma}$ will cause $\mtx{K}_1(t) \approx \mtx{\Omega}^{\top} g_{\sigma}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$ to be close to zero for parameter values $t$ located in such regions.
-Hence, on \reflin{lin:4-nystromchebyshev-nystrom-pp} we would compute the pseudoinverse of an almost zero matrix, which can be extremely unstable; see \reffig{fig:zerocheck}.
+Hence, on line \ref{lin:4-nystromchebyshev-nystrom-pp} we would compute the pseudoinverse of an almost zero matrix, which can be extremely unstable; see \cref{fig:zerocheck}.
 To detect this situation, we can cheaply compute the Girard-Hutchinson
 estimate $\Trace(\mtx{K}_1(t)) / n_{\mtx{\Omega}}$ of the nuclear norm.
-If this estimate is small (in our experiments, smaller than $10^{-5}$), we use it as an indicator that $t$ is within a region with a nearly vanishing spectral density, and therefore set $\widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) = 0$ instead of computing the full expression on \reflin{lin:4-nystromchebyshev-nystrom-pp}.
+If this estimate is small (in our experiments, smaller than $10^{-5}$), we use it as an indicator that $t$ is within a region with a nearly vanishing spectral density, and therefore set $\widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) = 0$ instead of computing the full expression on line \ref{lin:4-nystromchebyshev-nystrom-pp}.
 
@@
     \centering
-    \input{plots/zerocheck.pgf}
-    \caption{The difference the non-zero check can make when approximating a spectral density using \refalg{alg:nystrom-chebyshev-pp}.
-As input matrix we use the $1000 \times 1000$ matrix $\mtx{A}$ coming from a three-dimensional finite difference discretization of the Hamiltonian defined in~\refequ{equ:electronic-hamiltonian} on a $10 \times 10$ grid.
+    \scalebox{0.8}{\input{plots/zerocheck.pgf}}
+    \caption{The difference the non-zero check can make when approximating a spectral density using \cref{alg:nystrom-chebyshev-pp}.
+As input matrix we use the $1000 \times 1000$ matrix $\mtx{A}$ coming from a three-dimensional finite difference discretization of the Hamiltonian defined in~\cref{equ:electronic-hamiltonian} on a $10 \times 10$ grid.
 We ran the Chebyshev-Nyström++ method with and without estimating if the matrix function is zero before taking its pseudo-inverse.
@@
 
-In this section, we analyze the accuracy of Chebyshev-Nyström++ estimate returned by~\refalg{alg:nystrom-chebyshev-pp}. 
-\begin{theorem}{Error of Chebyshev-Nyström++ estimate for smoothed spectral density}{chebyshev-nystrom}
-    For a smoothing parameter $\sigma > 0$, let $\phi_{\sigma}$ be the smoothed spectral density~\refequ{equ:smooth-spectral-density} of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum contained in $[-1, 1]$.
+In this section, we analyze the accuracy of Chebyshev-Nyström++ estimate returned by~\cref{alg:nystrom-chebyshev-pp}. 
+\begin{theorem}[Error of Chebyshev-Nyström++ estimate for smoothed spectral density]\label{thm:chebyshev-nystrom}
+    For a smoothing parameter $\sigma > 0$, let $\phi_{\sigma}$ be the smoothed spectral density~\cref{equ:smooth-spectral-density} of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum contained in $[-1, 1]$.
 Consider $\varepsilon \in (0, 3)$ and $\delta \in (0, 2 e^{-1})$.
@@
         &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \int_{-1}^{1} \left| \underline{\phi}_{\sigma}^{(m)}(t) - \widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t &&  \notag \\
-        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \widetilde{\varepsilon} \int_{-1}^{1} \left| \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t && \text{(\refthm{thm:nystrom-pp} w.p.
+        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \widetilde{\varepsilon} \int_{-1}^{1} \left| \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t && \text{(\cref{thm:nystrom-pp} w.p.
 $\geq 1 - \delta$)} \notag \\
         &\leq (1 + \widetilde{\varepsilon}) \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \widetilde{\varepsilon} \int_{-1}^{1} \left| \phi_{\sigma}(t) \right|~\mathrm{d}t &&  \notag \\
-        &\leq (1 + \widetilde{\varepsilon}) \underline{E}_{\sigma, m} + \widetilde{\varepsilon}. && \text{(\reflem{lem:non-negative-chebyshev-error})} \notag 
+        &\leq (1 + \widetilde{\varepsilon}) \underline{E}_{\sigma, m} + \widetilde{\varepsilon}. && \text{(\cref{lem:non-negative-chebyshev-error})} \notag 
     \end{align*}
-    From \reflem{lem:non-negative-chebyshev-error} and the discussion after \reflem{lem:chebyshev-error} we see that the choice $m = \mathcal{O}((\log(\widetilde{\varepsilon}^{-1}) + \log(\sigma^{-1})) / \log(\sigma + 1))$ ensures $\underline{E}_{\sigma, m} \leq \widetilde{\varepsilon}$, and thus
+    From \cref{lem:non-negative-chebyshev-error} and the discussion after \cref{lem:chebyshev-error} we see that the choice $m = \mathcal{O}((\log(\widetilde{\varepsilon}^{-1}) + \log(\sigma^{-1})) / \log(\sigma + 1))$ ensures $\underline{E}_{\sigma, m} \leq \widetilde{\varepsilon}$, and thus
     the result follows by setting $\varepsilon = 3 \widetilde{\varepsilon}$.
@@
 
-In practice, the $\mathcal{O}(\varepsilon^{-1})$ complexity of $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$ in \refthm{thm:chebyshev-nystrom} is often pessimistic, because  the underlying result of \refthm{thm:nystrom-pp} is agnostic to the low-rank structure of the matrix $\mtx{B}(t) = g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
+In practice, the $\mathcal{O}(\varepsilon^{-1})$ complexity of $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$ in \cref{thm:chebyshev-nystrom} is often pessimistic, because  the underlying result of \cref{thm:nystrom-pp} is agnostic to the low-rank structure of the matrix $\mtx{B}(t) = g_{\sigma}(t\mtx{I}_n - \mtx{A})$.
 The singular values of this matrix are given by
@@
 An eigenvalue $\lambda_{(i)}$ distant from $t$ results in an exponentially small singular value $\sigma_i(t)$.
-In particular, for a small value of $\sigma$, all singular values corresponding to eigenvalues outside a certain window $[t - d, t + d]$ are negligibly small; see \reffig{fig:numerical-rank}.
+In particular, for a small value of $\sigma$, all singular values corresponding to eigenvalues outside a certain window $[t - d, t + d]$ are negligibly small; see \cref{fig:numerical-rank}.
 \begin{figure}[ht]
@@
 \end{figure}
-By choosing $n_{\mtx{\Omega}}$ such that the sum $\sum_{i=n_{\mtx{\Omega}}+5}^{n} \sigma_i(t)$ becomes negligibly small,  \refthm{thm:nystrom} implies an excellent approximation with high probability.
+By choosing $n_{\mtx{\Omega}}$ such that the sum $\sum_{i=n_{\mtx{\Omega}}+5}^{n} \sigma_i(t)$ becomes negligibly small,  \cref{thm:nystrom} implies an excellent approximation with high probability.
 In particular, if we assume the eigenvalues of $\mtx{A}$ to be roughly uniformly distributed, a small calculation shows that $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma \sqrt{\log(\varepsilon^{-1}) + \log(\sigma^{-1})})$ will ensure this.
